### 2014-01-27 padicprec.tex
Initial commit from Xavier's e-mail

%\documentclass{sig-alternate}
\documentclass{amsart}
\usepackage[utf8]{inputenc}

\usepackage[plainpages=false,pdfpagelabels,colorlinks=true,citecolor=blue,hypertexnames=false]{hyperref}
\usepackage{color}
\usepackage{stmaryrd}
\usepackage{enumerate}

\begin{document}

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\theoremstyle{definition}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}

\newcommand{\val}{\text{\rm val}}

\def\todo#1{\ \!\!{\color{red} #1}}
\definecolor{purple}{rgb}{0.6,0,0.6}
\def\todofor#1#2{\ \!\!{\color{purple} {\bf #1}: #2}}

%
% --- Author Metadata here ---
%\conferenceinfo{ISSAC}{}

% --- End of Author Metadata ---

\title{p-adic precision}

%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

%\numberofauthors{3} 
%\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
%\alignauthor
%Xavier Caruso\\
%       \affaddr{CNRS/Université de Rennes I}\\
%       \email{xavier.caruso@univ-rennes1.fr}
% 2nd. author
%\alignauthor
%David Roe\\
%       \affaddr{University of Calgary}\\
%       \email{roed.math@gmail.com}
% 3rd. author
%\alignauthor 
%Tristan Vaccon\\
%       \affaddr{Université de Rennes 1}\\
%       \email{tristan.vaccon@univ-rennes1.fr}
%}

\date{november 2013}

\maketitle
\begin{abstract}
A method to handle $p$-adic precision, to handle \textit{precision balance}.
\end{abstract}

% A category with the (minimum) three required fields
%\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulations}[Algebraic Algorithms]

%\terms{Algorithms, Theory}

%\keywords{$p$-adic precision, $p$-adic algorithm}

\section{Introduction}

During the last two decades, $p$-adic methods have become a tremendous tool in computer algebra, since it allows computation not feasible on finite fields, can control the growth of the size of the coefficients when dealing with rationnal numbers, or is needed when performing algorithm of $p$-adic nature.

Yet, $p$-adic algorithms are necessarily numeric or symbolic-numeric: one can not avoid working with finite precision, and thus, keeping track of loss in precision during the computation.

Here are some examples of algorithm that rely on $p$-adic computations:
\begin{itemize}
\item Bostan \cite{Bostan}, on how polynomials in $\mathbb{F}_p$ can be computed from theirs Newton series lifted in $\mathbb{Z}_p$;
\item Gaudry \cite{Gaudry}, on generation of genus $2$ curves via $p$-adic lifting;
\item Kedlaya \cite{Kedlaya}, on point counting on hyperelliptic curves using Monsky--Washnitzer cohomology;
\item Lercier \cite{Lercier}, on computing isogenies between elliptic curves using $p$-adic differential equations.
\end{itemize}

Many usual softwares (\emph{e.g.} {\tt magma} and {\tt sage}) include 
primitives to handle $p$-adic numbers. Since $p$-adic numbers are by 
nature infinite, one needs to truncate them up to some finite level (we 
generally say \emph{finite precision}) in order to manipulate them on a 
computed. The usual technics to achieve this are inspired from case of 
real numbers: the computer works with ``truncated $p$-adic numbers'' 
represented as $x + O(p^n)$ and performs all basic operations (addition, 
product) on this representation. For instance, it computes: 
$$(x + O(p^n)) + (y + O(p^m)) = (x+y) + O(p^{\min(n,m)}).$$
However, in many important cases (\emph{e.g.} those discussed just 
above), this approach is too naive since it creates artificially 
important losses of precision (see \S\ref{ssec:stepbystep}). 
It is the so-called \emph{numerical instability} issue.

The aim of the current paper is to provide a systematical theorical 
study of $p$-adic precision (and more generally nonarchimedian 
precision) and to derive from this study new methods to handle 
precision, that turns out to give nearly optimal results in a quite 
large setting.

\section{Motivations and aims}

\todofor{David}{Add a general introduction to this section.}

\subsection{Separate precision and approximation}

\todofor{David}{
Explain here why it might be useful to work separately with 
approximation on the one hand and precision on the other hand.
}

\subsection{Step-by-step tracking is not sharp}
\label{ssec:stepbystep}

In this subsection, we illustrate by some examples that tracking
$p$-adic precision step by step may lead to important numerical
instability.

\subsubsection*{A first stupid example}

Let $f : \Qp^2 \to \Qp^2$ be the function mapping $(x,y)$ to $(x+y, 
x-y)$. We consider in addition two $p$-adic numbers $x$ and $y$, 
which are known up to precision $O(p^n)$ and $O(p^m)$ respectively.
The question we would like to address in this example is the
computation of $f \circ f(x,y)$.

Two approaches are possible. The first one consists in doing the 
computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
compute then $f(z,t)$ which gives the result. Proceeding this way, we 
see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
are consequently the same holds for the coordinates of the result.

The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. CComputing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
equals $2$.

As a conclusion, the second approach yields a more accurate result.
In other words, computing step-by-step creates ``artificial'' losses 
of precision.

\subsubsection*{The SOMOS 4 sequence}

The SOMOS 4 sequence is a four-term inductive sequence satisfying:
$$u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.$$
In this example, we shall consider the situation where the initial
terms $u_0$, $u_1$, $u_2$ and $u_3$ are \emph{invertible} $p$-adic 
integers, \emph{i.e.} lie in $\Zp^\times$.

Let us first examine what happens (regarding to the precision) when
we are computing the sequence $(u_n)$ naively. The computation of
$u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
decreases the precision by a factor $p^{\val(u_n)}$ where $\val$
stands for the $p$-adic valuation. Hence, if we assume that $u_0$,
$u_1$, $u_2$ and $u_3$ are given with precision $O(p^N)$ then the
naive computation returns the value of $u_n$ with precision
\begin{equation}
\label{eq:SOMOS}
O(p^{N-v_n})
\quad \text{with}\quad
v_n = \val(u_0) + \cdots + \val(u_{n-4}).
\end{equation}

However, on the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon}. This means that, for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
From the latter formula, it follows directly that if $u_0$, $u_1$,
$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
are also known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm ---
which includes the model precision --- we used to compute to SOMOS 
sequence.

\begin{rmk}
From the above discussion, one can easily derive a numerically stable 
algorithm that computes the SOMOS 4 sequence: (1)~we compute the 
Laurent polynomials $P_n$ using the induction formula (and working in 
the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$) and (2)~we 
evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
However this algorithm is not efficient at all since computing the 
$P_n$'s is very time-consuming for two reasons: first, it requires 
division in a polynomial ring with $4$ variables and, second, the
size of the coefficients of $P_n$ explodes as $n$ grows.

In \S \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
the SOMOS 4 sequence which turns out to be, at the same time, efficient 
and numerically stable.
\end{rmk}

\subsubsection*{LU factorization}

Let us first recall that a square matrix $M$ with coefficients in $\Qp$ 
admits a LU factorization if it can be written as a produit $LU$ where 
$L$ and $U$ is lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems.

We first note that a LU decomposition of a given matrix $M$ is unique
if you require further that the $L$-part of this decomposition is
unipotent (\emph{i.e.} all diagonal entries equal $1$). Let us write
$L(M) U(M)$ the LU factorization of $M$ satisfying the above extra
condition (when it exists). There exists Cramer-type formulas that
gives each individual entries of $L(M)$and $U(M)$ in a closed form
(see \cite{Caruso}).

If $M$ is random square matrix over $\Zp$ of side $d$ whose every
coefficient is known up to precision $O(p^N)$, the two following
results are proved in \emph{loc. cit.}:
\begin{itemize}
\item using usual Gauss elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulas yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. This shows once again that ``artifical''
losses of precision appears when one performs Gauss elimination by
tracking precision step-by-step.

%\subsubsection*{Algorithms \emph{\`a la} Kedlaya}

\section{A new philosophy}
\label{sec:philosophy}

The aim of this section is to develop a new point of view on $p$-adic 
precision based on $p$-adic analysis. Actually, we shall generalize a 
bit the setting and work over a general ultrametric field (see defition
below).

\subsection{Ultrametric analysis}
\label{ssec:ultrametric}

In this section, we recall usual definitions and properties related to
ultrametric analysis. We refer to Schneider \cite{Schneider} for a 
complete exposition.

An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
: K \to \R_+$ satisfying the triangle inequality and sending every 
nonzero element of $K$ to a strictly positive real number. It is said 
\emph{ultrametric} if it satisfies the following strong from of the 
triangle inequality:
$$|x+y| \leq \max(|x|, |y|).$$
Given an absolute value $|\cdot|$ on $K$, we note that $|K^\star|$ is a 
multiplicative subgroup of $\R_+^\star$. It is then either discrete or 
dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\star|$ is
discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
|x-y|$) converge.

The formula $|x| = p^{-\val(x)}$ defines an ultrametric absolute value on 
the field $\Qp$ for which it is complete. More generally, the $p$-adic 
absolute value extends uniquely to any finite extension $K$ of $\Qp$ and 
turns $K$ into a complete ultrametric field. Another classical example of 
complete ultrametric field is the field of Laurent series over a finite
field, the absolute value being defined from the valuation of a series.

From now, we fix a field $K$ equipped with an ultrametric absolute value 
denoted by $|\cdot|$ and we assume that $K$ is complete. Given $v$ a 
$K$-vector space (of possibly infinite dimension), an \emph{ultrametric 
norm} on $V$ is a map $\Vert\cdot\Vert : V \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A norm gives rise to a distance and a normed $K$-vector space $V$ is 
said \emph{complete} when any Cauchy sequence with values in $V$ 
converges. A normed $K$-vector space which is complete is called a 
\emph{$K$-Banach}.

Similarly to the real case, any finite dimensional normed $K$-vector 
space is complete and all norms over such a space are equivalent 
(\emph{i.e.} they define the same topology). Many usual Theorems 
concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
Theorem) remain true in the ultrametric setting.

\begin{deftn}
Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a 
continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
$\varepsilon >0$, there exists an open neighbourhoud $U_\varepsilon 
\subset U$ containing $v_0$ with:
\[ 
\forall v, w \in U_\varepsilon, \quad
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

One can derive all usual properties of the differential from this 
definition: unicity, chain rules, \emph{etc}.

\subsection{Lattices}

Let $E$ be a $K$-normed vector space. Given $r \in \R^+$, 
we denote by $B^-_E(r)$ (resp. by $B_E(r)$) the set of elements $x 
\in E$ such that $\Vert x \Vert < r$ (resp. $\Vert x \Vert \leq r$). 
We insist on the following maybe unusual fact: $B^-_E(r)$ and 
$B_E(r)$ are both open and closed for the norm topology on $E$.
Clearly, $B_K(0,1)$ is stable under multiplication. Moreover, the 
ultrametric triangle inequality ensures that $B_K(1)$ is stable under 
addition as well. Thus, $B_K(1)$ is a subring of $K$: it is often 
called the \emph{integer ring} of $K$.

\begin{deftn}
Let $E$ be a $K$-Banach.
A \emph{lattice} in $E$ is an open bounded sub-$B_K(1)$-module of $E$.
\end{deftn}

\begin{rmk}
Any lattice $H$ in $E$ is also closed since its complement is the
union of all cosets $a + H$ (with $a \not\in H$) which are all open.
\end{rmk}

The following Lemma shows that the notion of lattices behaves rather 
well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
\ref{lem:main}) that, under some extra hypothesis, the linearity 
assumption can be drastically relaxed.} transformations.

\begin{lem}
\label{lem:morlat}
Let $E$ and $F$ be two $K$-Banach and $f : 
E \to F$ be a continuous \emph{surjective} linear map.
If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
\end{lem}

\begin{proof}
It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
The fact that it is open is a direct consequence of the open mapping
Theorem.
\end{proof}

To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
\in H$. It follows from the fact that $H$ is open and bounded in $E$
that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
of $I_x$ under the absolute value map is bounded and contains a positive
element. We can then define:
$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
$E$.

\begin{lem} \label{lem:reductionball}
\begin{enumerate}
\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
equivalent.
\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
Then:
$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
and the last inclusion is an equality if the absolute value on $K$ is
discrete.
\end{enumerate}
\end{lem}

\begin{proof}
Left to the reader.
\end{proof}

The first part of Lemma \ref{lem:reductionball} asserts that the 
identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
assuption: it is ``between the open unit ball and the closed unit 
ball''.

\subsubsection*{Lattices and computers}

\todofor{Xavier}{Explain that if $E$ is finite dimensional and $K$ (and 
the absolute value is discrete), a lattice can be encoded by a finite 
amount of informations.}

\subsection{The main Lemma}

\begin{lem} \label{lem:main}
Let $E$ and $F$ be two $K$-Banach and $f : U 
\rightarrow F$ be a function defined on an open subset $U$ of $E$.
We assume that $f$ is differentiable at some point $v_0 \subset 
U$ and that the differential $f'(v_0)$ is surjective.

Then, there exists a positive real number $\delta$ such that, for all $r 
\in (0, \delta)$, the equality
\begin{equation}
\label{eq:main}
f(v_0 + H) = f(v_0) + f'(v_0) \cdot H.
\end{equation}
holds for each lattice $H$ such that $B^-_E(r) \subset H \subset
B_E(r)$.
\end{lem}

\begin{rmk}
Given a lattice $H_0$ in $E$, the equality \eqref{eq:main} holds for $H 
= \lambda H_0$ with $\lambda$ small enough. To prove this, it is enough 
to apply Lemma \ref{lem:main} to the map $f \circ \iota_{H_0}$ where 
$\iota_{H_0} : (E, \Vert \cdot \Vert_{H_0}) \to (E, \Vert \cdot \Vert)$ 
is the ``identity map'' (which is an homeomorphism by Lemma 
\ref{lem:reductionball}).
\end{rmk}

\begin{proof}
We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
C \cdot f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by the fact 
that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
B_E(\delta)$ for some $\delta >0$.

Let $r \in (0, \delta)$.
Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
Let $x \in H$. We must show that $f(x) \in f'(0) \cdot H$. 
By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
\varepsilon \Vert x \Vert $. Therefore, there exists $y \in f'(0) \cdot 
H$ such that 
$\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
surjectify of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
\in B^-_E(r) \subset H$ and $f(x)= f'(0) \cdot (x-x_1) \in f'(0) 
\cdot H$.

We shall now prove surjectivity. Let $y \in f'(0) \cdot H$. 
Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
We define by induction two sequences $(x_n)$ and $(z_n)$ by
\begin{itemize}
\item $z_n \in E$ is such that
$f'(0) \cdot z_n = y - f(x_n)$ and
$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$,
\item $x_{n+1}=x_n+z_n$.
\end{itemize}
By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y_0 
= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y_0 \Vert < \varepsilon 
\Vert x_0 \Vert \leq \varepsilon r$. Thus $\Vert z_0 \Vert \leq 
\varepsilon C \cdot r < r$. Thus, $x_1$ and $z_0$ lie in $B^-_E(r) 
\subset H$. An easy induction involving an analoguous computation
shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
take their values in $H$.

Now, we notice that
$y - f(x_{n+1}) = f(x_n) - f(x_{n+1}) + f'(0) \cdot z_n$.
Therefore, using differentiability, we get:
$$\Vert y  - f(x_{n+1}) \Vert 
\leq \varepsilon \Vert z_n \Vert \leq \varepsilon C \cdot \Vert 
y - f(x_n) \Vert,$$
from what we deduce that $\Vert y - f(x_{n+1}) \Vert = 
O(a^n)$ and $\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < 1$.
This shows that $(x_n)$ is a Cauchy sequence and therefore converges 
(since $E$ is complete). If $x$ denotes the limit of $(x_n)$, we have
$x \in H$ and $y=f(x)$. The result is proven.
\end{proof}

\todofor{Tristan}{Improve the end of this Subsection}

Before discussing applications of Lemma \ref{lem:main} to ultrametric 
precision, we would like to describe a little bit more restrictive 
setting where the constant $\delta$ appearing in this Lemma can be 
computed explicitely.
We refer to Schneider \cite{Schneider}, \S 6, for a definition of a 
locally analytic function.

\begin{rmk}
Let $f$ be locally analytic function $K^n \rightarrow K^m$. Let $x \in K^n$ be such that $f'(x)$ is surjective. Then the radius of the ball needed in can be read on the Newton polygon given by the norms of the Taylor serie expansion of $f$ at $x$. We assume the $C$ given by the Banach-Schauder is given, $\varepsilon = \frac{1}{C}$ and now, we try to recover the link between $\varepsilon$ and $\delta$.
\end{rmk}

\begin{proof}
Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=f'(0) \cdot x +\sum_{n \geq 2} \frac{f^{(n)}(0)}{n!} \cdot (x, \dots,x)$ is the Taylor serie expansion of $f$ at $0$, coinciding on $B_E (0,r)$.

We shall now find $l$ such that if $x \in B_{K^n}(0,l)$, then $\Vert \sum_{n \geq 2} \frac{f^{(n)}(0)}{n!}(x,\dots,x) \Vert \leq \varepsilon$.

By continuity and ultrametricity, it is enough for $l$ to be such that, if $x \in B_{K^n}(0,l)$, then  \[ \sup_{n \geq 2} \interleave f^{(n)}(0) \interleave \frac{\Vert x \Vert^n }{ \vert n! \vert } \leq \varepsilon \Vert x \Vert .\]

Therefore, \[ \sup_{n \geq 2} \interleave f^{(n)}(0) \interleave \frac{l^{n-1} }{\vert n! \vert } \leq \varepsilon . \]

In other words, $\forall n \geq 2$, $\interleave f^{(n)}(0) \interleave \frac{l^{n-1} }{\vert n! \vert } \leq \varepsilon.$

By passing to logarithm and valuation, we get for all $n \geq 2$: \[ \log_p (\interleave f^{(n)}(0) \interleave )-(n-1) \log_p l-val (n!)\geq - \log_p \varepsilon.\]

By Legendre formula, it is enough that for all $n \geq 2$:
\[\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.\]

If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k$, then it is enough that:
\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.\]

If $NP_F^*$ is the Legendre-Fanchel transform of $NP_F$, then it is enough that:
\[ NP_F^* \left( \log_p l - \frac{1}{p-1} \right) \leq \log_p \varepsilon.\]

Therefore, by definition of the Legendre-Fanchel transform, it is enough to:
\begin{itemize}
\item take $M=(0,\log_p \varepsilon)$,
\item draw the line passing by $M$ tangent to the graph of $NP_F$,
\item if we denote by $\mu$ the slope of this line, then take $l$ such that $\mu + \frac{1}{p-1}=\log_p l$.
\end{itemize}
\end{proof}

\subsection{Automatic sharp track of precision}

\todofor{Xavier}{
Explain here how to build a software that tracks precision automatically 
and sharply.
}

\subsubsection*{Working precision}

\section{Generalization to manifolds}

It may happen that natural $p$-adic ``objects'' do not lie in vector 
spaces. Actually, examples are numerous: points on the projective space 
or elliptic curves, subspaces of a fixed vector space (which lie in 
grassmannians), classes of isomorphism of certain curves (which lie in 
various moduli spaces), \emph{etc.} 
That is why we think that it is quite important to extend the formalism 
developped in Section \ref{sec:philosophy} to a more general setting. 
In this Section, we consider the quite general case of a strictly
differentiable ultrametric manifolds locally modeled on (eventually 
infinite dimensional) Banach spaces. This covers all the aforementioned 
examples.

In what follows, the letter $K$ always refers to an ultrametric field 
(\emph{cf} \S \ref{ssec:ultrametric}). The absolute value over $K$ is 
denoted by $|\cdot|$.

\subsection{Strictly differentiable $K$-manifolds}
\label{ssec:manifold}

The theory of finite dimensional $K$-manifolds is presented for example 
in \cite{Schneider}, parts 8 and 9. In this Section, we shall work with
a slightly different notion of manifolds which allows also Banach vector 
spaces of infinite dimension.
More precisely, for us, a \emph{strictly differentiable $K$-manifold} 
(or just \emph{$K$-manifold} for short) is the data of a topological 
space $V$ together with an open covering $V = \bigcup_{i \in I} V_i$ 
(where $I$ is some set) and, for all $i \in I$, an homeomorphism 
$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach 
$E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
nonempty, the composite map
\begin{equation}
\label{eq:psiij}
\psi_{ij} : 
\varphi_i(V_{ij}) \stackrel{\varphi_i^{-1}}{\longrightarrow} 
V_{ij} \stackrel{\varphi_j}{\longrightarrow} \varphi_j(V_{ij})
\quad \text{(with } V_{ij} = V_i \cap V_j \text{)}
\end{equation}
is strictly differentiable at all points in its domain. We recall that
the mappings $\varphi_i$ above are the so-called \emph{charts} and that
their collection defines an \emph{atlas} of $V$. In the sequel, we shall
assume further that the open covering $V = \bigcup_{i \in I} V_i$ is
locally finite, which means that every point $x \in V$ lies only in a
finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
$K$-Banach themselves.

If $V$ is a $K$-manifold and $x$ is a point of $V$, we define the 
\emph{tangent space} $T_x V$ of $V$ at $x$ as the space $E_i$ for some 
$i$ such that $x \in V_i$. We note that if $x$ belongs to $V_i$ and 
$V_j$, the linear map $\psi'_{ij}(\varphi_i(x))$ defines an isomorphism 
between $E_i$ and $E_j$. Furthermore these isomorphisms are compatible 
in an obvious way. This implies that the definition of $T_x V$ given 
above does not depend (up to some canonical isomorphism) on the index 
$i$ such that $x \in V_i$ and then makes sense.

As usual, we can define the notion of strict differentiability (at some 
point) for a continuous mapping between two $K$-manifolds by viewing it 
through the charts. A strictly differentiable map $f : V \to V'$ induces 
a linear map on tangent spaces $f'(x) : T_x V \to T_{f(x)} V'$ for all 
$x$ in the domain $V$. It is called the \emph{differential} of $f$ at
$x$.

\subsection{Precision data}

\todofor{David/Xavier}{Complete this Subsection}

Going back to our problem of precision, given $V$ a $K$-manifold as 
above, we would like to be able to deal with ``approximations up to some 
precision'' of elements in $V$, \emph{i.e.} expressions of the form $x + 
O(H)$ where $x$ belongs to a dense \emph{computable} subset of $V$ and 
$H$ is a ``precision data''.
The aim of this paragraph is to give a precise meaning to the expression 
$x + O(H)$ and to prove several basic properties of it.
For now, we fix a $K$-manifold $V$ and we use freely the notations $I$, 
$V_i$, $\varphi_i$, \emph{etc.} introduced in \S \ref{ssec:manifold}.

\begin{deftn}
Let $x \in V$. 
A \emph{precision data} at $x$ is a lattice in the tangent space 
$T_x V$.
\end{deftn}

\begin{deftn}
Let $x \in V$ and $H$ be a precision data at $x$.
For all $i \in I$ such that $x \in V_i$, we set:
$$x + O_i(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
\subset V.$$
\end{deftn}

We emphasize that if $x$ belongs to $V_i$ and $V_j$ at the same time,
the subsets $x + O_i(H)$ and $x + O_j(H)$ differ in general. It is 
however not the case if $H$ is small enough.

\begin{prop}
\label{prop:independance}
Let $x \in V$ and $H$ be a precision data at $x$.
If $H$ is ``small enough'' \todo{(precise!)}, the subset $x + 
O_i(H)$ does not depend on $i$.
\end{prop}

\begin{proof}
Let $i$ and $j$ be two indices such that $x$ belongs to $V_i$ and $V_j$. 
Set $x_i = \varphi_i(x) \in E_i$ and $H_i = \varphi'_i(x)(H)$. Going 
back to the definitions, we see that the equality $x + O_i(H) = x + 
O_j(H)$ is equivalent to:
$$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$$ 
where we recall that $\psi_{ij}$ is defined by Eq.~\eqref{eq:psiij}.
\todo{Complete the proof and write the correct condition on $H$!}
\end{proof}

Under the assumption of Proposition \ref{prop:independance}, we define
$x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
$x \in V_i$.

\subsubsection*{Change of base point}

In order to restrict ourselves to elements $x$ lying in a dense 
computable subset, we need to compare $x + O(H)$ with some $x' + 
O(H')$ when $x$ and $x'$ are close enough.
Let us first examine the situation in a fixed given chart: we fix some 
index $i \in I$ and pick two elements $x$ and $x'$ in $V_i$. We consider 
in addition a precision data $H$ at $x$ and we want to produce a 
precision data $H'$ at $x'$ such that $x + O_i(H) = x' + O_i(H')$. 
We remark that the tangent spaces $T_x V$ and $T_{x'} V$ are both 
isomorphic to $E_i$ \emph{via} the maps $\varphi'_i(x)$ and 
$\varphi'_i(x')$ respectively. A natural candidate for $H'$ is then:
\begin{equation}
\label{eq:Hprime}
H' = \big(\varphi'_i(x')^{-1} \circ \varphi'_i(x)\big) (H).
\end{equation}
And indeed, one can check that $x + O(H)$ is indeed equal to $x' + 
O(H')$ as soon as $x$ and $x'$ are close enough in the following sense: 
the difference $\varphi_i(x) - \varphi_i(x')$ lies in the lattice $H_i = 
\varphi'_i(x)(H)$. We furthermore have a proprety of independance on 
$i$.

\begin{prop}
We assume that $H$ is ``small enough'' \todo{(precise!)} and that $x$ and 
$x'$ are close enough. Then:
\begin{enumerate}[(i)]
\item the lattice $H'$ defined by Eq.~\eqref{eq:Hprime} does not depend 
on $i$;
\item the couples $(x,H)$ and $(x', H')$ satisfy the assumptions of
Proposition \ref{prop:independance} and we have $x + O(H) = x' + O(H')$.
\end{enumerate}
\end{prop}

\begin{proof}
We first prove (i). For an index $i$ such that $x, x' \in V_i$, let us 
denote by $f_i : T_x V \to T_{x'} V$ the composite $\varphi'_i(x')^{-1} 
\circ \varphi'_i(x)$. Given an extra index $j$ satisfying the same
assumption, we know by \todo{(find a reference)} that the difference
$f_i - f_j$ goes to $0$ when $x'$ converges to $x$. Since $H$ is open
in $T_x V$, this implies that $(f_j - f_i)(H)$ contains $f_i(H)$ and 
$f_j(H)$ if $x'$ and $x$ are close enough. Now, pick $w \in f_j(H)$ and 
write it $w = f_j(v)$ with $v \in H$. Then $w$ is equal to $f_i(v) + 
(f_j - f_i)(v)$ and then belongs to $f_i(H)$ because each summand does. 
Therefore $f_j(H) \subset f_i(H)$. The inverse inclusion can be proved 
in the same way.

We now prove (ii). \todo{Write the proof.}
\end{proof}

\subsection{Generalization of the main Lemma}

\todofor{David/Xavier}{Extend Lemma \ref{lem:main} to manifolds.}

\section{Examples and applications}

%\subsection{Differential tracking of precision}
%
%\todo{
%Explain here that we can track precision by computing differentials step 
%by step. Note also in this subsection that this method can be very time 
%consuming; it's why we often prefer to consider other models of precision.
%}

\todofor{Xavier}{Write this section}

\subsection{Polynomials}

\todo{
Euclidean division, factorization, \emph{etc.}
}

\subsection{Matrices}

\todo{
LU factorization, determinant, characteristic polynomial, kernels, images,
\emph{etc.}
}

\subsection{The SOMOS 4 sequence}
\label{ssec:SOMOS-solution}

\todo{
Study SOMOS sequence
}

\section{Conclusions}

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns

\tiny \bibliographystyle{plain}
\begin{thebibliography}{10}
\addcontentsline{toc}{section}{Références}

\bibitem{Bostan}
{\sc Bostan, Alin, \& al.} 
\newblock From Newton sums to coefficients: complexity issues in characteristic p. Proceedings MEGA'05, 2005.

\bibitem{Caruso}
{\sc Caruso, X.}
\newblock Random matrices over a DVR and LU factorization, preprint (2012)

\bibitem{Gaudry}
{\sc Gaudry,P. \& al.} 
\newblock The 2-adic CM method for genus 2, Asiacrypt 2006, volume 4284 of Lecture Notes in Comput. Sci., pages 114–129. Springer, Berlin, 2006.

\bibitem{Kedlaya}
{\sc Kedlaya,Kiran,} 
\newblock Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology, Journal of the Ramanujan Mathematical Society 16 (2001),

\bibitem{Lercier}
{\sc Lercier, Reynald \&  Sirvent, Thomas,} 
\newblock On elkies subgroups of $\ell$-torsion points in elliptic curves defined over a finite field. Journal de Théorie des Nombres de Bordeaux, 20(3):783-797, December 2008.

\bibitem{Schneider}
{\sc Schneider, Peter}
\newblock $p$-adic Lie Groups, Springer, 2011

\end{thebibliography}

\end{document}

### 2014-01-27 padicprec.tex
Study of SOMOS sequence (not finished)

@@
 
-\usepackage[plainpages=false,pdfpagelabels,colorlinks=true,citecolor=blue,hypertexnames=false]{hyperref}
+\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
 \usepackage{color}
@@
 \usepackage{enumerate}
+\usepackage[algoruled,vlined,french,linesnumbered]{algorithm2e}
 
@@
 \newcommand{\val}{\text{\rm val}}
+\renewcommand{\prec}{\text{\rm prec}}
 
@@
 accurate than the first one. This shows once again that ``artifical''
-losses of precision appears when one performs Gauss elimination by
+losses of precision appear when one performs Gauss elimination by
 tracking precision step-by-step.
@@
 precision based on $p$-adic analysis. Actually, we shall generalize a 
-bit the setting and work over a general ultrametric field (see defition
+bit the setting and work over a general ultrametric field (see definition
 below).
@@
 In this Section, we consider the quite general case of a strictly
-differentiable ultrametric manifolds locally modeled on (eventually 
-infinite dimensional) Banach spaces. This covers all the aforementioned 
-examples.
+differentiable manifolds locally modeled on (eventually infinite 
+dimensional) ultrametric Banach spaces. This covers all the 
+aforementioned examples.
 
@@
 
-\todo{
-Study SOMOS sequence
+Recall (cf \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
+four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
++ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
+Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
+u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
+polynomial with coefficients in $\Z$ in these indeterminates.
+
+From now, we will always consider SOMOS sequences with values in $\Q_p$ 
+(for some prime number $p$) and assume that $u_0, u_1, u_2, u_3$ are 
+chosen such that all terms $u_n$'s are determined (\emph{i.e.} there is 
+no division by zero). We assume also for simplicity that $u_0, u_1, u_2, 
+u_3$ are all units in $\Z_p$. By Laurent phenomenon, this implies first 
+that all $u_n$'s lie in $\Z_p$ and second that if $u_0, u_1, u_2, u_3$ 
+are known with finite precision $O(p^N)$ then all terms $u_n$'s are 
+known with the same precision.
+
+In this last paragraph, we illustrate the theory developed in this 
+article by designing an algorithm for computing the $u_n$'s at precision 
+$O(p^N)$. The algorithm is presented at the top of page 
+\pageref{algo:somos}. We now prove that it is correct.
+%
+\begin{algorithm}[t]
+\SetKwInOut{Assumption}{Assumption}
+\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
+\KwIn{$n, N$ --- two integers}
+\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
+precision $O(p^N)$}
+\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
+\KwOut{$u_n$ at precision $O(p^N)$}
+\BlankLine
+$\prec$ $\leftarrow$ $N$\;
+\For {$i$ from $1$ to $n-3$}{
+  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
+  {\bf lift} (arbitrary) 
+    $b$, $c$ and $d$ to precision $O(p^{\prec})$\;
+  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
+  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
+   \tcp{$e$ is known at precision $O(p^\prec)$}
+  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
 }
+\Return{$d + O(p^N)$}\;
+\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:somos}
+\end{algorithm}
+%
+We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
+$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
+$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
+= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
+differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 
+canonical basis is given by the matrix:
+$$D(a,b,c,d) = \begin{pmatrix}
+0 & 1 & 0 & 0 \\
+0 & 0 & 1 & 0 \\
+0 & 0 & 0 & 1 \\
+-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
+\end{pmatrix}$$
+whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
+of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
+at $(u_0, u_1, u_2, u_3)$ and the matrix its differential at this point 
+is:
+$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
+D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
+%Its determinant is:
+%$$\det D_i = \frac{u_{i+3}}{u_{i-1}} \cdots 
+%\frac{u_5}{u_1} \cdot \frac{u_4}{u_0} =
+%\frac{u_i \cdot u_{i+1} \cdot u_{i+2} \cdot u_{i+3}}
+%{u_0 \cdot u_1 \cdot u_2 \cdot u_3}.$$
+Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
+integral coefficients. Hence, letting $\varphi_i = f'_i(u_0, u_1, u_2, 
+u_3)$, we get 
+$$(\det D_i) \cdot p^N \Z_p^4 \subset \varphi_i(p^N \Z_p^4) 
+\subset p^N \Z_p^4$$
+for all integer $N$.
 

### 2014-01-29 padicprec.tex
Proof of the correctness of Algorithm SOMOS

@@
 
-%\subsection{Differential tracking of precision}
-%
-%\todo{
-%Explain here that we can track precision by computing differentials step 
-%by step. Note also in this subsection that this method can be very time 
-%consuming; it's why we often prefer to consider other models of precision.
-%}
-
 \todofor{Xavier}{Write this section}
 
-\subsection{Polynomials}
+\subsection{Differential of usual operations}
+
+\subsubsection{Numbers}
+
+\todo{sum, product, inverse, $p$-power map}
+
+\subsubsection{Polynomials}
 
 \todo{
-Euclidean division, factorization, \emph{etc.}
+Euclidean division, factorization, Grobner basis (?)
 }
 
-\subsection{Matrices}
+\subsubsection{Matrices}
 
 \todo{
-LU factorization, determinant, characteristic polynomial, kernels, images,
-\emph{etc.}
-}
+determinant, characteristic polynomial, LU factorization, QR 
+factorization, kernels, images}
+
+\subsubsection{Sub vector spaces}
+
+\todo{intersection, sum}
 
@@
 $O(p^N)$. The algorithm is presented at the top of page 
-\pageref{algo:somos}. We now prove that it is correct.
+\pageref{algo:SOMOS}. We now prove that it is correct.
 %
@@
   {\bf lift} (arbitrary) 
-    $b$, $c$ and $d$ to precision $O(p^{\prec})$\;
+    $b$, $c$ and $d$ to precision $O(p^{\prec})$\label{line:lift}\;
   $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
@@
 \Return{$d + O(p^N)$}\;
-\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:somos}
+\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:SOMOS}
 \end{algorithm}
@@
 of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
-at $(u_0, u_1, u_2, u_3)$ and the matrix its differential at this point 
-is:
+at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
+f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix:
 $$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
 D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
-%Its determinant is:
-%$$\det D_i = \frac{u_{i+3}}{u_{i-1}} \cdots 
-%\frac{u_5}{u_1} \cdot \frac{u_4}{u_0} =
-%\frac{u_i \cdot u_{i+1} \cdot u_{i+2} \cdot u_{i+3}}
-%{u_0 \cdot u_1 \cdot u_2 \cdot u_3}.$$
 Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
-integral coefficients. Hence, letting $\varphi_i = f'_i(u_0, u_1, u_2, 
-u_3)$, we get 
-$$(\det D_i) \cdot p^N \Z_p^4 \subset \varphi_i(p^N \Z_p^4) 
-\subset p^N \Z_p^4$$
-for all integer $N$.
+integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
+$\Z_p^4$.
+We are now going to prove by induction on $i$ that, at the end of the 
+$i$-th iteration of the loop, we have $\prec = N + v_p(\det D_i)$ and
+\begin{equation}
+\label{eq:congrSOMOS}
+(a, b, c, d) \equiv (u_i, u_{i+1}, u_{i+2}, u_{i+3}) \pmod
+{p^N \varphi_i(\Z_p^4)}.
+\end{equation}
+The first point is easy. Indeed, from $D_i = D(u_{i-1}, u_i, u_{i+1}, 
+u_{i+2}) D_{i-1}$, we deduce $\det D_i = \det D_{i-1} \cdot 
+\frac{u_i}{u_{i-3}}$ and the assertion follows by taking determinants
+and using the induction hypothesis. 
+Let us now establish Eq.~\eqref{eq:congrSOMOS}. To avoid confusion, 
+let us agree to denote $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
+$a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
+the $i$-th iteration of the loop. By induction hypothesis (or by 
+initialization if $i = 1$), we have:
+\begin{equation}
+\label{eq:congrSOMOS2}
+(a', b', c', d') \equiv (u_{i-1}, u_i, u_{i+1}, u_{i+2}) \pmod
+{p^N \varphi_{i-1}(\Z_p^4)}.
+\end{equation}
+Moreover, we know that the determinant of $\varphi_{i-1}$ has valuation
+$\prec'$. Hence \eqref{eq:congrSOMOS2} remains true if $a'$, $b'$, $c'$
+and $d'$ are replaced by other values which are congruent to them modulo
+$p^{\prec'}$. In particular it holds if $a'$, $b'$, $c'$ and $d'$ denotes
+the values of $a$, $b$, $c$ and $d$ after the execution of line
+\ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
+\ref{prop:} to $\varphi_{i-1}$ and $\varphi_i$ (at the point $(u_0,
+u_1, u_2, u_3)$), we get:
+$$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
+= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).$$
+By the discussion above, this implies in particular that $f(a',b',c',d')$
+belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
+conclude by remarking first that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
+{p^\prec \Z_p^4}$ by construction and second that $p^\prec \Z_p^4 \subset
+p^N \varphi_i(\Z_p^4)$.
+
+Finally Eq.~\eqref{eq:congrSOMOS} applied with $i = n-3$ together with
+the fact that $\varphi_i$ stabilizes $\Z_p^4$ imply that, when we exit
+the loop, the value of $d$ is congruent to $u_n$ modulo $p^N$. Hence,
+our algorithm returns the correct value.
+
+\medskip
+
+Let us conclude this subsection by remarking that Algorithm 
+\ref{algo:SOMOS} performs computations at precision at most $O(p^{N+v})$ 
+where $v$ is the sum of the valuations of five consequence terms among 
+the first $n$ terms of the SOMOS sequence we are considering. 
+Experiments show that the value of $v$ varies like $c \cdot \log n$ 
+where $c$ is some constant. Assuming that we are using a FFT-like 
+algorithm to compute products of integers, the complexity of Algorithm 
+\ref{algo:SOMOS} is then expected to be $\tilde O(N n)$ where the 
+notation $\tilde O$ means that we hide logarithmic factors.
+
+We can compare this with the complexity of the more naive algorithm 
+consisting of lifting the initial terms $u_0, u_1, u_2, u_3$ to enough 
+precision and then do the computation using a naive step-by-step 
+tracking of precision. In this setting, the required original precision 
+is $O(p^{N+2v'})$ where $v'$ is the sum of the valuation of the $u_i$'s 
+for $i$ varying between $0$ and $n$. Experiments show that $v'$ is about 
+$c' \cdot n \log n$ (where $c'$ is a constant), which leads to a 
+complexity in $\tilde O (N n + n^2)$. Our approach is then interesting
+when $n$ is large compared to $N$: under this hypothesis, it saves
+roughly a factor $n$.
 
 \section{Conclusions}
-
-%ACKNOWLEDGMENTS are optional
-\section{Acknowledgments}
 
@@
 \bibitem{Bostan}
-{\sc Bostan, Alin, \& al.} 
-\newblock From Newton sums to coefficients: complexity issues in characteristic p. Proceedings MEGA'05, 2005.
+  A.~Bostan, L.~Gonz\'alez-Vega, H.~Perdry, \'E.~Schost,
+  \emph{From Newton sums to coefficients: complexity issues in characteristic $p$},
+  Proceedings MEGA'05 (2005).
 
 \bibitem{Caruso}
-{\sc Caruso, X.}
-\newblock Random matrices over a DVR and LU factorization, preprint (2012)
+  X.~Caruso,
+  \emph{Random matrices over a DVR and LU factorization},
+  preprint (2012)
 
 \bibitem{Gaudry}
-{\sc Gaudry,P. \& al.} 
-\newblock The 2-adic CM method for genus 2, Asiacrypt 2006, volume 4284 of Lecture Notes in Comput. Sci., pages 114–129. Springer, Berlin, 2006.
+  P.~Gaudry, T.~Houtmann, A.~Weng, C.~Ritzenthaler, D.~Kohel,
+  \emph{The $2$-adic CM method for genus $2$}, 
+  Asiacrypt 2006, vol. 4284 of Lecture Notes in Comput. Sci. (2006), 114--129
 
 \bibitem{Kedlaya}
-{\sc Kedlaya,Kiran,} 
-\newblock Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology, Journal of the Ramanujan Mathematical Society 16 (2001),
+  K.~Kedlaya,
+  \emph{Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology}, 
+  J. Ramanujan Math. Soc. {\bf 16} (2001)
 
 \bibitem{Lercier}
-{\sc Lercier, Reynald \&  Sirvent, Thomas,} 
-\newblock On elkies subgroups of $\ell$-torsion points in elliptic curves defined over a finite field. Journal de Théorie des Nombres de Bordeaux, 20(3):783-797, December 2008.
+  R.~Lercier, T.~Sirvent,
+  \emph{On Elkies subgroups of $\ell$-torsion points in elliptic curves defined over a finite field},
+  J. Th\'eorie des Nombres de Bordeaux {\bf 20} (2008), 783--797
 
 \bibitem{Schneider}
-{\sc Schneider, Peter}
-\newblock $p$-adic Lie Groups, Springer, 2011
-
+  P.~Schneider,
+  \emph{$p$-adic Lie Groups}, Springer (2011)
 \end{thebibliography}

### 2014-01-29 padicprec.tex
Differentials of usual operations on p-adic numbers and polynomials

@@
 \renewcommand{\O}{\mathcal O}
+\newcommand{\XX}{\mathbf X}
 
@@
 
-%
-% --- Author Metadata here ---
-%\conferenceinfo{ISSAC}{}
-
-% --- End of Author Metadata ---
-
-\title{p-adic precision}
-
-%
-% You need the command \numberofauthors to handle the 'placement
-% and alignment' of the authors beneath the title.
-%
-% For aesthetic reasons, we recommend 'three authors at a time'
-% i.e. three 'name/affiliation blocks' be placed beneath the title.
-%
-% NOTE: You are NOT restricted in how many 'rows' of
-% "name/affiliations" may appear. We just ask that you restrict
-% the number of 'columns' to three.
-%
-% Because of the available 'opening page real-estate'
-% we ask you to refrain from putting more than six authors
-% (two rows with three columns) beneath the article title.
-% More than six makes the first-page appear very cluttered indeed.
-%
-% Use the \alignauthor commands to handle the names
-% and affiliations for an 'aesthetic maximum' of six authors.
-% Add names, affiliations, addresses for
-% the seventh etc. author(s) as the argument for the
-% \additionalauthors command.
-% These 'additional authors' will be output/set for you
-% without further effort on your part as the last section in
-% the body of your article BEFORE References or any Appendices.
-
-%\numberofauthors{3} 
-%\author{
-% You can go ahead and credit any number of authors here,
-% e.g. one 'row of three' or two rows (consisting of one row of three
-% and a second row of one, two or three).
-%
-% The command \alignauthor (no curly braces needed) should
-% precede each author name, affiliation/snail-mail address and
-% e-mail address. Additionally, tag each line of
-% affiliation/address with \affaddr, and tag the
-% e-mail address with \email.
-%
-% 1st. author
-%\alignauthor
-%Xavier Caruso\\
-%       \affaddr{CNRS/Université de Rennes I}\\
-%       \email{xavier.caruso@univ-rennes1.fr}
-% 2nd. author
-%\alignauthor
-%David Roe\\
-%       \affaddr{University of Calgary}\\
-%       \email{roed.math@gmail.com}
-% 3rd. author
-%\alignauthor 
-%Tristan Vaccon\\
-%       \affaddr{Université de Rennes 1}\\
-%       \email{tristan.vaccon@univ-rennes1.fr}
-%}
-
-\date{november 2013}
+\title{$p$-adic precision}
+\author{Xavier Caruso}
+\author{David Roe}
+\author{Tristan Vaccon}
+\date\today
 
@@
 \begin{abstract}
-A method to handle $p$-adic precision, to handle \textit{precision balance}.
+We present a new method to handle $p$-adic precision (or more generally
+precision over ultrametric fields) based on $p$-adic calculus.
+We illustrate it by many examples and give a toy application to the
+stable computation of the SOMOS 4 sequence.
 \end{abstract}
 
-% A category with the (minimum) three required fields
-%\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulations}[Algebraic Algorithms]
-
-%\terms{Algorithms, Theory}
-
-%\keywords{$p$-adic precision, $p$-adic algorithm}
+\setcounter{tocdepth}{1}
+\tableofcontents
 
@@
 are consequently the same holds for the coordinates of the result.
-
 The second approach consists in remarking that $f \circ f$ maps
-$(x,y)$ to $(2x, 2y)$. CComputing $f \circ f(x,y)$ using the latter
+$(x,y)$ to $(2x, 2y)$. Computing $f \circ f(x,y)$ using the latter
 formula, it is clear that the result we get is known up to precision
@@
 equals $2$.
-
 As a conclusion, the second approach yields a more accurate result.
@@
 coefficient is known up to precision $O(p^N)$, the two following
-results are proved in \emph{loc. cit.}:
+results are proved in \cite{Caruso}:
 \begin{itemize}
@@
 
-\todofor{Xavier}{Write this section}
+In this Section, we illustrate the theory developed in this paper by 
+giving examples and applications. In \S \ref{ssec:differentials}, we 
+compute the differential of a bunch of usual operations involving 
+$p$-adic objects of different nature. Very often, we will observe that 
+this computation is easy even if the underlying operation is 
+sophisticated (\emph{e.g.} Gr\"obner basis).
+In \S \ref{ssec:SOMOS-solution}, we study the SOMOS 4 sequence 
+introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
+Lemma \ref{lem:main} and Proposition \ref{prop:}, we design a quite 
+\emph{stable} algorithm for computing it.
 
 \subsection{Differential of usual operations}
+\label{ssec:differentials}
 
-\subsubsection{Numbers}
+As in the previous Sections, we always work over a fixed ultrametric 
+field $K$, whose absolute value is denoted by $|\cdot|$. We recall that 
+usual examples are the field of $p$-adic numbers $\Q_p$ and the field of 
+Laurent series $k((X))$ over a base field $k$.
 
-\todo{sum, product, inverse, $p$-power map}
+In what follows, we use freely the ``method of physicists'' to compute 
+differentials: given a function $f$ differentiable at some point $x$, we 
+consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$ 
+by expanding LHS and neglecting terms of order $2$. The differential of 
+$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
 
-\subsubsection{Polynomials}
+\subsubsection*{Numbers}
 
-\todo{
-Euclidean division, factorization, Grobner basis (?)
-}
+The most basic operations are, of course, sums, products and inverses of 
+elements of $K$. Their differential are well-known and quite easy to 
+compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have 
+$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = - 
+\frac{dx}{x^2}$).
 
-\subsubsection{Matrices}
+A little bit more interesting is the $n$-th power map $f$ from $K$ to 
+itself. Its differential $f'(x)$ is obtained by differentiate the 
+equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
+Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
+number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
+means that the behaviour of the precision depends on the absolute value 
+of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
+this inegality might be strict. It happens for instance if the 
+characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
+$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
+not apply. 
+Another situation (which is more interesting) is that of $p$-adic 
+numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
+$|n| = p^{-k}$ and Lemma \ref{lem:main} tells us basically that raising
+a $p$-adic number to the $p^k$-th power increases the precision by $k$
+extra digits (which is actually a well-known fact).
 
-\todo{
-determinant, characteristic polynomial, LU factorization, QR 
+\subsubsection*{Univariate polynomials}
+
+Beyond sums and products (which can be treated as before), a basic 
+operation involving polynomials is Euclidean division. Let us compute 
+the differential of it. We start with two polynomials $A$ and $B$ with 
+$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
+with $\deg R < \deg B$. Differentiating the above equality we find: 
+$dA - dB \cdot Q = B \cdot dQ + dR$,
+which implies that $dQ$ and $dR$ are respectively obtained as the 
+quotient and the remainder of the Euclidean division of $dA - dB \cdot 
+Q$ by $B$. This gives the differential. We note that the discussion 
+above extends readily to convergent series (see also 
+\cite{caruso-lubicz}).
+
+Another usual operations on polynomials is the factorization. Suppose 
+that we are given a monic polynomial $P \in K[X]$ written as a product 
+$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
+lemma implies that there exists a small neighbourhood $\mathcal U$ of 
+$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
+as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
+respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
+defined on the subset of $\mathcal U$ consisting of monic polynomials.
+We want to differentiate $f$ at $P_0$. For this, we differentiate the 
+equality $P = A B$ around $P_0$, obtaining 
+\begin{equation}
+\label{eq:difffactor}
+dP = A_0 \cdot dB + B_0 \cdot dA.
+\end{equation}
+where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
+$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
+between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
+$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
+{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).
+
+An important special case of the previous study occurs when $A_0$ has 
+degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
+application $P \mapsto A$ is then nothing but the mapping that follows 
+the simple root $\alpha$. Of course, its differential around $P_0$ can 
+be computed by the above method. Nevertheless there is a more direct way 
+to obtain it in this particular case: we write $(P + dP)(\alpha + 
+d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
+dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
+is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
+= - \frac{dP(\alpha)}{P'(\alpha)}$.
+
+\subsubsection*{Multivariate polynomials}
+
+We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
+variables over $K$ and fix a monomial order on it. We then have a 
+notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ are polynomials
+in $K[\XX]$, there exists a writing:
+$$f = q_1 f_1 + \cdots + q_s f_s + r$$
+where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
+by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
+that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
+assumption, there still exists a canonical choice of $r$. We do not know
+if the computation of the differential that follows extends to a more
+general setting.} (of the ideal generated by these polynomials), the 
+polynomial $r$ is uniquely determined and called the \emph{remainder} of 
+the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
+$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
+the case of univariate polynomials, we can compute its differential: we 
+find that $dr$ is obtained as the remainder of the division of $f - (q_1 
+\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
+
+We now focus on the computation of Gr\"obner basis. Recall that any 
+ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We 
+can then consider the application mapping a family $(f_1, \ldots, f_s)$ 
+of polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
+\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
+application at $(f_1, \ldots, f_s)$ is continuous provided that:
+\begin{itemize}
+\item the sequence $(f_1, \ldots, f_s)$ is regular
+\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
+$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
+order (\emph{cf} \cite{Vaccon} for more precisions).
+\end{itemize}
+A similar argument proves that it is actually differentiable at such
+points. We are now going to compute the differential. For this we
+remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
+g_t)$ generate the same ideal, we have a relation:
+\begin{equation}
+\label{eq:prodgrobner}
+(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
+\end{equation}
+where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
+Moreover following Vaccon's construction, we see that the entries of $A$ 
+can be chosen in such a way that they define differentiable functions.
+Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
+$$(d f_1, \ldots, d f_s) \cdot A = - (f_1, \ldots, f_s) \cdot dA + 
+(d g_1, \ldots, d g_t).$$
+Finally, using that we are working with \emph{reduced} Gr\"obner
+basis, the above equality implies that $d f_i$ is the remainder in
+the division of the $i$-th coefficient of the product
+$(d f_1, \ldots, d f_s) \cdot A$ by the family $(f_1, \ldots, f_s)$.
+
+\subsubsection*{Matrices}
+
+\todo{Determinant, characteristic polynomial, LU factorization, QR 
 factorization, kernels, images}
 
-\subsubsection{Sub vector spaces}
+\subsubsection*{Vector spaces}
 
-\todo{intersection, sum}
+\todo{Intersection, sum}
 
@@
 
-\section{Conclusions}
+%\section{Conclusions}
 
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
-\bibliographystyle{abbrv}
-\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
-% You must have a proper ".bib" file
-%  and remember to run:
-% latex bibtex latex latex
-% to resolve all references
-%
-% ACM needs 'a single self-contained file'!
-%
-%APPENDICES are optional
-%\balancecolumns
-
-\tiny \bibliographystyle{plain}
 \begin{thebibliography}{10}
-\addcontentsline{toc}{section}{Références}
 
@@
   \emph{$p$-adic Lie Groups}, Springer (2011)
+
+\bibitem{Vaccon}
+  T.~Vaccon,
+  \emph{Matrix-F5 algorithms over finite-precision complete discrete 
+  valuation fields},
+  preprint (2014)
 \end{thebibliography}

### 2014-01-29 padicprec.tex
Small fixes and changes

--- 2014-01-27 padicprec.tex
@@
 The second approach consists in remarking that $f \circ f$ maps
-$(x,y)$ to $(2x, 2y)$. CComputing $f \circ f(x,y)$ using the latter
+$(x,y)$ to $(2x, 2y)$.  Computing $f \circ f(x,y)$ using the latter
 formula, it is clear that the result we get is known up to precision
@@
 Let us first recall that a square matrix $M$ with coefficients in $\Qp$ 
-admits a LU factorization if it can be written as a produit $LU$ where 
+admits a LU factorization if it can be written as a product $LU$ where 
 $L$ and $U$ is lower triangular and upper triangular respectively. 
@@
 $$|x+y| \leq \max(|x|, |y|).$$
-Given an absolute value $|\cdot|$ on $K$, we note that $|K^\star|$ is a 
-multiplicative subgroup of $\R_+^\star$. It is then either discrete or 
-dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\star|$ is
+Given an absolute value $|\cdot|$ on $K$, we note that $|K^\times|$ is a 
+multiplicative subgroup of $\R_+^\times$. It is then either discrete or 
+dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\times|$ is
 discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
@@
 continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
-$\varepsilon >0$, there exists an open neighbourhoud $U_\varepsilon 
+$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
 \subset U$ containing $v_0$ with:
@@
 $H$ is the image under $\iota_H$ of a lattice satisfying an extra 
-assuption: it is ``between the open unit ball and the closed unit 
+assumption: it is ``between the open unit ball and the closed unit 
 ball''.
@@
 $\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
-surjectify of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
+surjectivityy of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
 x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
@@
 \varepsilon C \cdot r < r$. Thus, $x_1$ and $z_0$ lie in $B^-_E(r) 
-\subset H$. An easy induction involving an analoguous computation
+\subset H$. An easy induction involving an analogous computation
 shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
@@
 setting where the constant $\delta$ appearing in this Lemma can be 
-computed explicitely.
+computed explicitly.
 We refer to Schneider \cite{Schneider}, \S 6, for a definition of a 
@@
 Let $x \in K^n$ be such that $f'(x)$ is surjective.
-Then the radius of the ball needed in can be read on the Newton polygon given by the norms of the Taylor serie expansion of $f$ at $x$.
+Then the radius of the ball needed can be read from the Newton polygon given by the norms of the Taylor series expansion of $f$ at $x$.
 We assume the $C$ given by the Banach-Schauder is given, $\varepsilon = \frac{1}{C}$ and now, we try to recover the link between $\varepsilon$ and $\delta$.
@@
 \begin{proof}
-Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=f'(0) \cdot x +\sum_{n \geq 2} \frac{f^{(n)}(0)}{n!} \cdot (x, \dots,x)$ is the Taylor serie expansion of $f$ at $0$, coinciding on $B_E (0,r)$.
+Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=f'(0) \cdot x +\sum_{n \geq 2} \frac{f^{(n)}(0)}{n!} \cdot (x, \dots,x)$ is the Taylor series expansion of $f$ at $0$, coinciding on $B_E (0,r)$.
 
@@
 By Legendre formula, it is enough that for all $n \geq 2$:
-\[\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.\]
+\[
+\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.
+\]
 
-If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k$, then it is enough that:
-\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.\]
+If we denote by $NP_F$ the Newton polygon application of the power series 
+\[
+F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k
+\]
+then it is enough that:
+\[
+\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.
+\]
 
@@
 
-It may happen that natural $p$-adic ``objects'' do not lie in vector 
-spaces. Actually, examples are numerous: points on the projective space 
-or elliptic curves, subspaces of a fixed vector space (which lie in 
-grassmannians), classes of isomorphism of certain curves (which lie in 
-various moduli spaces), \emph{etc.} 
-That is why we think that it is quite important to extend the formalism 
-developped in Section \ref{sec:philosophy} to a more general setting. 
-In this Section, we consider the quite general case of a strictly
-differentiable manifolds locally modeled on (eventually infinite 
-dimensional) ultrametric Banach spaces. This covers all the 
+Many natural $p$-adic objects do not lie in vector spaces:
+points in projective spaces or elliptic curves,
+subspaces of a fixed vector space (which lie in Grassmannians),
+classes of isomorphism of certain curves (which lie in various moduli spaces),
+\emph{etc.}  In this section we extend the formalism 
+developed in Section \ref{sec:philosophy} to a more general setting:
+we consider the quite general case of strictly
+differentiable manifolds locally modeled on ultrametric Banach spaces.
+This covers all the 
 aforementioned examples.
@@
 Let $x \in V$. 
-A \emph{precision data} at $x$ is a lattice in the tangent space 
+A \emph{precision datum} at $x$ is a lattice in the tangent space 
 $T_x V$.
@@
 \begin{deftn}
-Let $x \in V$ and $H$ be a precision data at $x$.
+Let $x \in V$ and $H$ be a precision datum at $x$.
 For all $i \in I$ such that $x \in V_i$, we set:
@@
 \begin{prop}
-\label{prop:independance}
+\label{prop:independence}
 Let $x \in V$ and $H$ be a precision data at $x$.
@@
 
-Under the assumption of Proposition \ref{prop:independance}, we define
+Under the assumption of Proposition \ref{prop:independence}, we define
 $x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
@@
 In order to restrict ourselves to elements $x$ lying in a dense 
-computable subset, we need to compare $x + O(H)$ with some $x' + 
-O(H')$ when $x$ and $x'$ are close enough.
+computable subset, we need to compare $x_0 + O(H_0)$ with varying $x + 
+O(H)$ when $x$ and $x_0$ are close enough.
 Let us first examine the situation in a fixed given chart: we fix some 
-index $i \in I$ and pick two elements $x$ and $x'$ in $V_i$. We consider 
-in addition a precision data $H$ at $x$ and we want to produce a 
-precision data $H'$ at $x'$ such that $x + O_i(H) = x' + O_i(H')$. 
-We remark that the tangent spaces $T_x V$ and $T_{x'} V$ are both 
-isomorphic to $E_i$ \emph{via} the maps $\varphi'_i(x)$ and 
-$\varphi'_i(x')$ respectively. A natural candidate for $H'$ is then:
+index $i \in I$ and pick two elements $x_0$ and $x$ in $V_i$. We consider 
+in addition a precision datum $H_0$ at $x_0$ and we want to produce a 
+precision datum $H$ at $x$ such that $x_0 + O_i(H_0) = x + O_i(H)$. 
+We remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
+isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and 
+$\varphi'_i(x)$ respectively. A natural candidate for $H$ is then:
 \begin{equation}
 \label{eq:Hprime}
-H' = \big(\varphi'_i(x')^{-1} \circ \varphi'_i(x)\big) (H).
+H = \left(\varphi'_i(x)^{-1} \circ \varphi'_i(x_0)\right) (H_0).
 \end{equation}
-And indeed, one can check that $x + O(H)$ is indeed equal to $x' + 
-O(H')$ as soon as $x$ and $x'$ are close enough in the following sense: 
-the difference $\varphi_i(x) - \varphi_i(x')$ lies in the lattice $H_i = 
-\varphi'_i(x)(H)$. We furthermore have a proprety of independance on 
+And indeed, one can check that $x + O(H)$ is indeed equal to $x_0 + 
+O(H_0)$ as soon as $x$ and $x_0$ are close enough in the following sense: 
+the difference $\varphi_i(x) - \varphi_i(x_0)$ lies in the lattice $H_i = 
+\varphi'_i(x_0)(H_0)$. We furthermore have a property of independence on 
 $i$.
@@
 \begin{prop}
-We assume that $H$ is ``small enough'' \todo{(precise!)} and that $x$ and 
-$x'$ are close enough. Then:
+We assume that $H_0$ is ``small enough'' \todo{(precise!)} and that $x$ is close 
+enough to $x_0$. Then
 \begin{enumerate}[(i)]
-\item the lattice $H'$ defined by Eq.~\eqref{eq:Hprime} does not depend 
-on $i$;
-\item the couples $(x,H)$ and $(x', H')$ satisfy the assumptions of
-Proposition \ref{prop:independance} and we have $x + O(H) = x' + O(H')$.
+\item the lattice $H$ defined by \eqref{eq:Hprime} does not depend 
+on $i$,
+\item the pairs $(x,H)$ and $(x_0, H_0)$ satisfy the assumptions of
+Proposition \ref{prop:independence} and we have $x + O(H) = x_0 + O(H_0)$.
 \end{enumerate}
@@
 \begin{proof}
-We first prove (i). For an index $i$ such that $x, x' \in V_i$, let us 
-denote by $f_i : T_x V \to T_{x'} V$ the composite $\varphi'_i(x')^{-1} 
-\circ \varphi'_i(x)$. Given an extra index $j$ satisfying the same
+We first prove (i). For an index $i$ such that $x, x_0 \in V_i$, let us 
+denote by $f_i : T_{x_0} V \to T_x V$ the composite $\varphi'_i(x)^{-1} 
+\circ \varphi'_i(x_0)$. Given an extra index $j$ satisfying the same
 assumption, we know by \todo{(find a reference)} that the difference
-$f_i - f_j$ goes to $0$ when $x'$ converges to $x$. Since $H$ is open
-in $T_x V$, this implies that $(f_j - f_i)(H)$ contains $f_i(H)$ and 
-$f_j(H)$ if $x'$ and $x$ are close enough. Now, pick $w \in f_j(H)$ and 
-write it $w = f_j(v)$ with $v \in H$. Then $w$ is equal to $f_i(v) + 
-(f_j - f_i)(v)$ and then belongs to $f_i(H)$ because each summand does. 
-Therefore $f_j(H) \subset f_i(H)$. The inverse inclusion can be proved 
+$f_i - f_j$ goes to $0$ when $x$ converges to $x_0$. Since $H_0$ is open
+in $T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ and 
+$f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in f_j(H_0)$ and 
+write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal to $f_i(v) + 
+(f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each summand does. 
+Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse inclusion can be proved 
 in the same way.
@@
 $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
-= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
-differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 
+= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is strictly 
+differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
 canonical basis is given by the matrix:

### 2014-01-30 padicprec.tex
Differentials of several operations on matrices

--- 2014-01-29 padicprec.tex
@@
 %\documentclass{sig-alternate}
-\documentclass{amsart}
+\documentclass{lms}
 \usepackage[utf8]{inputenc}
 
+\usepackage{amsmath,amssymb}
 \usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
@@
 \newtheorem{ex}[theo]{Example}
-\theoremstyle{definition}
+%\theoremstyle{definition}
 \newtheorem{deftn}[theo]{Definition}
@@
 \newcommand{\XX}{\mathbf X}
+\newcommand{\trans}{{}^{\text t}}
 
 \newcommand{\val}{\text{\rm val}}
+\newcommand{\tr}{\text{\rm Tr}}
+\newcommand{\com}{\text{\rm Com}}
 \renewcommand{\prec}{\text{\rm prec}}
@@
 formula, it is clear that the result we get is known up to precision
-$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
-equals $2$.
+$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p = 2$.
 As a conclusion, the second approach yields a more accurate result.
@@
 By Legendre formula, it is enough that for all $n \geq 2$:
-\[\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.\]
-
-If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k$, then it is enough that:
-\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.\]
+\[
+\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.
+\]
+
+If we denote by $NP_F$ the Newton polygon application of the power series 
+\[
+F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k
+\]
+then it is enough that:
+\[
+\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.
+\]
 
@@
 \section{Generalization to manifolds}
-
-It may happen that natural $p$-adic ``objects'' do not lie in vector 
-spaces. Actually, examples are numerous: points on the projective space 
-or elliptic curves, subspaces of a fixed vector space (which lie in 
-grassmannians), classes of isomorphism of certain curves (which lie in 
-various moduli spaces), \emph{etc.} 
-That is why we think that it is quite important to extend the formalism 
-developped in Section \ref{sec:philosophy} to a more general setting. 
-In this Section, we consider the quite general case of a strictly
-differentiable manifolds locally modeled on (eventually infinite 
-dimensional) ultrametric Banach spaces. This covers all the 
+\label{sec:manifold}
+
+Many natural $p$-adic objects do not lie in vector spaces:
+points in projective spaces or elliptic curves,
+subspaces of a fixed vector space (which lie in Grassmannians),
+classes of isomorphism of certain curves (which lie in various moduli spaces),
+\emph{etc.}  In this section we extend the formalism 
+developed in Section \ref{sec:philosophy} to a more general setting:
+we consider the quite general case of strictly
+differentiable manifolds locally modeled on ultrametric Banach spaces.
+This covers all the 
 aforementioned examples.
@@
 numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
-$|n| = p^{-k}$ and Lemma \ref{lem:main} tells us basically that raising
-a $p$-adic number to the $p^k$-th power increases the precision by $k$
-extra digits (which is actually a well-known fact).
+$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
+raising a $p$-adic number to the $p^k$-th power increases the precision 
+by $k$ extra digits.
 
@@
 We now focus on the computation of Gr\"obner basis. Recall that any 
-ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We 
-can then consider the application mapping a family $(f_1, \ldots, f_s)$ 
-of polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
+ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
+then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
+\emph{homogeneous}\footnote{This restriction will be convenient for us 
+mainly because we are using the article \cite{Vaccon}, which deals only 
+with homogeneous polynomials. It is however probably not essential.} 
+polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
 \ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
@@
 
-\todo{Determinant, characteristic polynomial, LU factorization, QR 
-factorization, kernels, images}
+Differentiating ring operations (sum, products, inverse) over matrices 
+is again straightforward; we just need to be careful that matrix 
+algebras are (in general) noncommutative.
+
+We now move to determinants, \emph{i.e.} we compute the differential
+of the function $\det : M_n(K) \to K$. It is a standard computation.
+To begin with, we consider an invertible matrix $M$ and write:
+\begin{eqnarray*}
+\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
+&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
+&=& \det(M) + \tr(\com(M) \cdot dM)
+\end{eqnarray*}
+where $\com(M)$ denotes the comatrix of $M$. The differential of 
+$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
+out that this formula is still valid when $M$ is not invertible.
+The same computation extends readily to characteristic polynomials,
+since they are defined as determinants. More precisely, let us 
+consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
+the latter ring denotes the finite dimensional vector space of 
+polynomials over $K$ of degree at most $n$) taking a matrix $M$
+to its characteristic polynomial $\det(X-M)$.
+Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
+differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
+
+Other important tools related to matrices are usual factorizations as, 
+for example, LU factorization. In what follows, let us agree to define a 
+LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
+where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
+Such a writing exists and is unique provided that all principal minors 
+of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
+defined over the Zariski-open set of matrices satisfying the above 
+condition. In order to differentiate it, we differentiate the relation 
+$M = LU$ and rewrite the result as follows:
+$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
+We remark that in the right hand side of the above formula, the first
+summand is lower triangular and all its diagonal entries vanish whereas 
+the second summand is upper triangular. Hence in order to compute $dL$
+and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
+L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
+obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
+recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
+left and $U$ on the right respectively.
+
+A similar argument works for QR factorization.
+For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
+a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
+orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
+a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
+mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
+would like to emphasize at this point that the orthogonality condition
+defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
+vector space (it is defined by equations of degree $2$). The codomain
+of $f$ is then also a manifold; this example then fits to the setting
+of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
+The generalization made in \S \ref{sec:manifold} is then needed.
+Anyway, we can differentiate $f$ by following the method we used for LU 
+factorization: differentiating the relation $M = QR$, we obtain:
+\begin{equation}
+\label{eq:diffQR}
+\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
+\cdot R^{-1}
+\end{equation}
+where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
+Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
+$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
+matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
+triangular with all diagonal entries equal to $0$, we see that 
+Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
+and $dM$.
 
@@
 
-\todo{Intersection, sum}
+\todo{Kernels, images, intersection, sum}
 
@@
 
-Recall (cf \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
+Recall (\emph{cf} \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
 four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 

### 2014-01-30 padicprec.tex
Kernels of matrices

@@
 
+\newcommand{\GL}{\text{\rm GL}}
+
 \newcommand{\val}{\text{\rm val}}
@@
 \newcommand{\com}{\text{\rm Com}}
+\newcommand{\Grass}{\text{\rm Grass}}
 \renewcommand{\prec}{\text{\rm prec}}
@@
 left and $U$ on the right respectively. 
-
-A similar argument works for QR factorization.
+The above discussion extends almost \emph{verbatim} to LUP 
+factorization; the only difference is that LUP factorizations are not 
+unique but they are on a small neighbourhood of $M$ if we fix the matrix 
+$P$.
+
+A similar argument works for QR factorization as well.
 For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
@@
 
-\todo{Kernels, images, intersection, sum}
+Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
+Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
+of dimension $d$. It defines an algebraic variety over $K$ and hence
+\emph{a fortiori} a $K$-manifold in the sense of \S \ref{ssec:manifold}.
+Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
+a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
+and two such matrices $M$ and $M'$ define the same vector space if there 
+exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
+find that we can always choose the above matrix $M$ in the particular
+shape:
+$$M = \begin{pmatrix} I_d & M' \end{pmatrix} \cdot P$$
+where $I_d$ denotes the $(d \times d)$ identity matrix, $M' \in M_{d, 
+n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
+such writings with the same $P$ necessarily coincide. Hence each 
+permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
+is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
+
+Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
+matrices of full rank. The left kernel defines a mapping $\text{LK} : 
+M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
+and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
+course, there exists a neighbourhood $\mathcal U$ of $M$ whose image is 
+entirely contained in a given chart $U_P$. We assume for simplicity that 
+$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
+our chart to the application that takes $M$ to the unique matrix $N \in
+M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
+= 0$. The implicit function theorem then implies that $\text{LK}$ is
+differentiable. Furthermore, its differential satisfies the relation
+$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
+\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
+from which we can compute $dN$ by projecting on the $(n-d)$ last
+columns.
+
+We can develop further this example and study what happens on the 
+closed subset of $M_{n,n-d}(K)$ where matrices have not full rank. On
+this subspace, the left kernel has dimension $< d$ and then no longer
+defines a point in $\Grass(n,d)$. Nevertheless, for all integer $r <
+n-d$, we can consider the subset $\mathcal V_r \subset M_{n,n-d}(K)$
+of matrices whose rank are exactly rank $d$. It is locally closed in
+$M_{n,n-d}(K)$ with respect to the Zariski topology and hence defines
+a $K$-manifold. Furthermore, we have a mapping $\text{LK}_r : \mathcal 
+V_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
+can be computed as before.
+
+The lesson from this study is the following. If we have to compute the 
+kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
+priori} we first need to determine what is the rank of $M$. However, the 
+rank function is not continuous --- it is only lower semi-continuous --- 
+and hence cannot be determined from sure from $M$ (expect if $M$ has 
+full rank). We then need to make an extra hypothesis and the more natural 
+one is arguably to assume that $M$ has actually the minimal rank that it 
+can have. With this convention, for instance, the kernel of the zero 
+matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
+whole space although there exist invertible matrices in any 
+neighbourhood of $M$.
 

### 2014-01-30 padicprec.tex
One paragraph on intersection of subspaces. I think that Section 5 is now almost done.

@@
 a $K$-manifold. Furthermore, we have a mapping $\text{LK}_r : \mathcal 
-V_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
+M_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
 can be computed as before.
@@
 neighbourhood of $M$.
+
+We now focus on the computation of intersections. We pick $n$, $d_1$ and 
+$d_2$ three nonnegative integers such that $d_1 \leq n$, $d_2 \leq n$ 
+and $d_1 + d_2 \geq n$. Two subspaces of $K^n$ of dimension $d_1$ and 
+$d_2$ respectively meet along a subspace of dimension at most $d_1 + d_2 
+- n$. For all $d \leq d_1 + d_2 -n$, we can then define the subspace 
+$\mathcal V_d$ of $\Grass(n,d_1) \times \Grass(n,d_2)$ consisting of 
+pairs $(E_1, E_2)$ such that $\dim (E_1 \cap E_2) = d$ and consider the 
+function $f_d : \mathcal V_d \to \Grass(n, d)$ which sends $(E_1, E_2)$ 
+to $E_1 \cap E_2$. In charts, the function $f_d$ can be interpreted as
+the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
+kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We deduce
+for this that $f_d$ is differentiable and that its differential can be
+computed similarly to what we have done before.
+
+Finally, we note that similar results hold for images of matrices and,
+consequently, sums of subspaces.
 

### 2014-01-30 padicprec.tex
Working on introduction to Section 2

@@
 \section{Motivations and aims}
+
+Two sources of inspiration arise when studying $p$-adic algorithms.
+The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
+$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
+$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
+homomorphism, given unknown elements in two such balls we can
+locate the balls in which their sum and product lie.  Working on a computer
+we must find a way to write elements using only a finite amount of data.
+By lumping elements together into these balls of radius $p^{-n}$, we may
+model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In Sage
+this method for approximating elements is known as ``fixed modulus''.
+
+The second source draws upon parallels between $\Qp$ and $\R$.  Both
+occur as completions of $\Q$, and both have dense subrings:
+the localizations $\Z[\frac1p] \subset \Qp$ and $\Z[\frac12] \subset \R$.
+In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
+subset $S \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
+\[
+\frac{x \circledast y}{x \ast y} - 1 \le 2^{-h}
+\]
+for all $x, y \in S$ and $\ast \in \{+, \cdot\}$.  Interval arithmetic in the reals
+is unwieldy, since arithmetic operations always increase the lengths of the inputs.
+As a consequence, most computations in the real numbers rely on
+statistical cancelation and external estimates of precision loss, rather than
+attempting to track known precision at each step.  This tendency is strengthened
+by the ubiquity of floating point arithmetic in scientific applications, where Gaussian
+distributions are more common than intervals anyway.
+
+In the $p$-adic world, these features are reversed.
 

### 2014-01-30 padicprec.tex

@@
 generally say \emph{finite precision}) in order to manipulate them on a 
-computed. The usual technics to achieve this are inspired from case of 
+computer. The usual technics to achieve this are inspired from case of 
 real numbers: the computer works with ``truncated $p$-adic numbers'' 
@@
 above extends readily to convergent series (see also 
-\cite{caruso-lubicz}).
+\cite{Caruso-Lubicz}).
 
@@
 Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
-$$(d f_1, \ldots, d f_s) \cdot A = - (f_1, \ldots, f_s) \cdot dA + 
+$$(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA = 
 (d g_1, \ldots, d g_t).$$
 Finally, using that we are working with \emph{reduced} Gr\"obner
-basis, the above equality implies that $d f_i$ is the remainder in
+basis, the above equality implies that $d g_i$ is the remainder in
 the division of the $i$-th coefficient of the product
-$(d f_1, \ldots, d f_s) \cdot A$ by the family $(f_1, \ldots, f_s)$.
+$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
 
@@
 
+\bibitem{Caruso-Lubicz}
+  X.~Caruso, D.~Lubicz
+  \emph{Linear Algebra over Zp[[u]] and related rings,} 
+  to appear at LMS J. Comp. and Math.
+
 \bibitem{Gaudry}

### 2014-02-02 padicprec.tex
application -> map

@@
 
-If we denote by $NP_F$ the Newton polygon application of the power series 
+If we denote by $NP_F$ the Newton polygon of the power series 
 \[
@@
 as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
-respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
+respectively. Thus, we can consider the map $f : P \mapsto (A,B)$ 
 defined on the subset of $\mathcal U$ consisting of monic polynomials.
@@
 degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
-application $P \mapsto A$ is then nothing but the mapping that follows 
+map $P \mapsto A$ is then nothing but the mapping that follows 
 the simple root $\alpha$. Of course, its differential around $P_0$ can 
@@
 polynomial $r$ is uniquely determined and called the \emph{remainder} of 
-the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
+the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
 $(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
@@
 ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
-then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
+then consider the map sending a family $(f_1, \ldots, f_s)$ of 
 \emph{homogeneous}\footnote{This restriction will be convenient for us 
@@
 \ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
-application at $(f_1, \ldots, f_s)$ is continuous provided that:
+map at $(f_1, \ldots, f_s)$ is continuous provided that:
 \begin{itemize}
@@
 $P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
-our chart to the application that takes $M$ to the unique matrix $N \in
+our chart to the map that takes $M$ to the unique matrix $N \in
 M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 

### 2014-02-04 padicprec.tex
First tentative for Section 3.4. Comments are welcome.

--- 2014-01-30 padicprec.tex
@@
 
+\begin{prop}
+\label{prop:locanalytic}
+\todo{Write a proposition.}
+\end{prop}
+
 \begin{proof}
@@
 
-\todofor{Xavier}{
-Explain here how to build a software that tracks precision automatically 
-and sharply.
-}
-
-\subsubsection*{Working precision}
+Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have important 
+applications to effective computations with $p$-adic numbers or power 
+series, that we are going to discuss now. Let us recall first that a 
+datum involving elements of a complete ultrametric field $K$ (\emph{e.g.} 
+$p$-adic numbers or power series) cannot be stored entirely on a computed 
+because they basically convey an infinite amount of informations. That is 
+why, instead of this datum itself, we generally work with neighborhoods 
+of it. Such neighborhoods are typically sets of the form $x + H$ where 
+$x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice in 
+$E$. A basic example is the approximate $5$-adic number $1 + O(5^{10})$ 
+which corresponds to the neighborhood of $1$ in $\Z_5$, which is the set 
+of all $5$-adic numbers congruent to $1$ modulo $5^{10}$. In the context
+of computer science, we will often use the notation $x + O(H)$ to refer 
+to $x + H$.
+
+Now, let us consider a function {\tt f} (in the sense of computer 
+science) that takes as input an (approximate) element lying in an open 
+subset $U$ of a $K$-Banach space $E$ and outputs another (approximate) 
+element lying in an open subset $V$ of another $K$-Banach space $F$. Of 
+course, this function can be modelised by a \emph{continuous} 
+mathematical function $f : U \to V$. The relation is the following: when 
+{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
+$f(x+H) \subset x' + H'$. We say that {\tt f} does not generate 
+artificial loss of precision if the above inclusion is an equality; it is 
+often not the case as shown in \S \ref{ssec:stepbystep}.
+
+\smallskip
+
+Let us assume now that $f$ is locally analytic on $U$ and that $x$ and 
+$H$ are chosen so that Lemma \ref{lem:main} applies. Then $f(x+H) = f(x) 
++ f'(x)(H)$ and the smallest possible $H'$ is $f'(x)(H)$. In other 
+words, the function {\tt f} generates artificial loss of precision if 
+the precision $H'$ it outputs is strictly larger that $f'(x)(H)$.
+The aim of this section is to explain how, under the above hypothesis, 
+one can implement the function {\tt f} so that it always outputs the
+optimal precision $O(H')$ with $H' = f'(x)(H)$.
+
+\subsubsection*{Forward computation}
+
+The execution of the function {\tt f} yields a factorization:
+$$f = f_n \circ f_{n-1} \circ \cdots \circ f_1$$
+where the $f_i$'s correspond to each individual basic step (like 
+addition, multiplication or affectation of variables); it is then a
+``nice'' (in particular locally analytic) function. For all 
+$i$, let $U_i$ denote the codomain of $f_i$. Concretely, $U_i$ is the 
+set of all possible values of all variables which are defined in the 
+program after the execution of $i$-th step. Mathematically, it is an 
+open subset in some $K$-Banach space $E_i$. We have $U_n = V$ and the 
+domain of $f_i$ is $U_{i-1}$ where, by convention, we have set $U_0 = 
+U$.
+For all $i$, we set $g_i = f_i \circ \cdots \circ f_1$ and $x_i = 
+g_i(x)$.
+
+When we execute the function {\tt f} on the input $x + O(H)$, we apply 
+first $f_1$ to this input obtaining this way a first result $x_1 + 
+O(H_1)$ and then go on with $f_2, \ldots, f_n$. At each step, we obtain 
+a new intermediate result that we denote by $x_i + O(H_i)$. A way to 
+guarantee that no artificial loss of precision are generated is then to 
+ensure $H_i = f'_i(x)(H_{i-1}) = g_i'(x)(H)$ at each step. This can be 
+achieved by reimplementing all primitives (addition, multiplication, 
+\emph{etc.}) and make them compute at the same time the function $f_i$
+they implement together with its differential and apply the latter to
+the ``current'' lattice $H_i$.
+
+There is nevertheless an important issue with this approach: in order to 
+be sure that Lemma \ref{lem:main} applies, we need \emph{a priori} to 
+compute the exact values of all $x_i$'s, which is of course not possible! 
+Assuming that $g'_i(x)$ is surjective for all $i$, we can fix it as 
+follows. For each $i$, we fix a lattice $\tilde H_i \subset H$ such that 
+Lemma \ref{lem:main} applies to $g_i$, $x$ and $\tilde H_i$. Under our 
+assumption, such lattices always exist and can be computed dynamically 
+using Proposition \ref{prop:locanalytic}. Now, the equality $g_i(x + 
+\tilde H_i) = x_i + g'_i(x)(\tilde H_i)$ means that any perturbation of 
+$x_i$ by an element in $g'_i(x) (\tilde H_i)$ is induced by a 
+perturbation of $x$ by an element in $\tilde H_i \subset H$. Hence, we 
+can freely compute $x_i$ modulo $g'_i(x) (\tilde H_i)$ without changing 
+the final result. Since $g'_i(x)(\tilde H_i)$ is a lattice in $E_i$, this 
+remark makes possible the computation of $x_i$.
+
+\begin{rmk}
+In some cases, it would be possible to determine suitable lattices 
+$\tilde H_i$ together with their images under $g'_i(x)$ (or, at
+least, good approximations of them) before starting the computation 
+by using mathematical arguments. If possible, this generally helps a
+lot. We shall present in \S \ref{ssec:SOMOS-solution} an example of 
+this.
+\end{rmk}
+
+\subsubsection*{Backward computation}
+
+The previous approach works only if the $g'_i(x)$'s are all surjective. 
+Unfortunately, this assumption is in general not fulfilled. Indeed, 
+remember that the dimension of $E_i$ is roughly the number of used 
+variables after the step $i$. It all $g_i'(x)$ were surjective, this 
+would mean that the function {\tt f} never initializes a new variable!
+In what follows, we propose another solution that does not assume the
+surjectivity of $g'_i(x)$.
+
+For $i \in \{1, \ldots, n\}$, define $h_i = f_n \circ \cdots \circ 
+f_{i+1}$, so that we have $f = h_i \circ g_i$. On differentials, we 
+have $f'(x) = h_i'(x_i) \circ g'_i(x)$. Since $f'(x)$ is surjective (by 
+assumption), we deduce that all $h'_i(x_i)$ are surjective. Let $H'_i$ 
+be a lattice in $E_i$ such that:
+\begin{enumerate}[(a)] 
+\item \label{item:Hi1}
+$H'_i$ is contained in $H_i + \ker h'_i(x_i) = h'_i(x_i)^{-1}
+\big(f'(x)(H)\big)$;
+\item \label{item:Hi2}
+Lemma \ref{lem:main} applies with the function $h_i$, the point
+$x_i$ and the lattice $H'_i$.
+\end{enumerate}
+Applying Lemma \ref{lem:main}, we then get:
+$$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H).$$
+Therefore, if after the $i$-th step of the execution of the function
+{\tt f}, we modify the intermediate value $x_i$ by an element of
+$H'_i$, the final result remains unchanged. In other words, it is
+enough to compute $x_i$ modulo $H'_i$.
+
+It is nevertheless not obvious to implement these ideas in practice
+because when we enter in the $i$-th step of the execution of {\tt f},
+we have not computed $h_i$ yet and hence are \emph{a priori} not able
+to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
+and \eqref{item:Hi2} above.
+A possible solution to tackle this problem is to proceed in several
+stages as follows: 
+\begin{enumerate}[(1)]
+\item \label{item:smallprec}
+for $i$ going from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
+at small precision (but enough for the second step) together with an
+approximation of the Newton polygon of the higher order derivatives;
+\item \label{item:determineHi}
+for $i$ going from $n$ to $1$, we compute $h'_i(x_i)$ and
+determines a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
+\eqref{item:Hi2};
+\item \label{item:finalcomp}
+for $i$ going from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
+and finally outputs $x_n + O\big(f'(x)(H)\big)$.
+\end{enumerate}
+Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
+\cite{}), we can reuse in Step~\eqref{item:finalcomp} the computations 
+already performed in Step~\eqref{item:smallprec}. The ``backward'' method 
+we have just presented is then probably not much more expansive than the 
+``forward'' method, although it is certainly much more difficult to 
+implement.
+
+We conclude this Section by remarking that the ``backward'' method seems 
+to be particularly well suited to computations with lazy $p$-adic. Indeed 
+in this setting, a target precision (\emph{i.e.} a precision on the 
+output) is fixed and the software determines automatically the precision 
+it needs on the input to perform the computation. To do this, it first 
+builds the ``skeleton'' of the computation (\emph{i.e.} it determines the 
+functions $f_i$ and eventually computes the $x_i$ at small precision when 
+branching points occur and it needs to decide which branch it follows) 
+and then runs over this skeleton in the reverse direction in order to 
+determine (an upper bound of) the needed precision at each step. The 
+similarity with our approach is noticeably strong.
 
@@
 introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
-Lemma \ref{lem:main} and Proposition \ref{prop:}, we design a quite 
-\emph{stable} algorithm for computing it.
+Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic}, we design
+a quite \emph{stable} algorithm for computing it.
 
@@
 $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
-lemma implies that there exists a small neighbourhood $\mathcal U$ of 
+lemma implies that there exists a small neighborhood $\mathcal U$ of 
 $P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
@@
 factorization; the only difference is that LUP factorizations are not 
-unique but they are on a small neighbourhood of $M$ if we fix the matrix 
+unique but they are on a small neighborhood of $M$ if we fix the matrix 
 $P$.
@@
 and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
-course, there exists a neighbourhood $\mathcal U$ of $M$ whose image is 
+course, there exists a neighborhood $\mathcal U$ of $M$ whose image is 
 entirely contained in a given chart $U_P$. We assume for simplicity that 
@@
 whole space although there exist invertible matrices in any 
-neighbourhood of $M$.
+neighborhood of $M$.
 
@@
 \ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
-\ref{prop:} to $\varphi_{i-1}$ and $\varphi_i$ (at the point $(u_0,
-u_1, u_2, u_3)$), we get:
+\ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
+$(u_0, u_1, u_2, u_3)$), we get:
 $$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)

### 2014-02-05 padicprec.tex
Sections 4 and 5.2 moved to appendices. The paper (without appendices) now fits in 15 pages as wanted by ANTS.

@@
 
-Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have important 
-applications to effective computations with $p$-adic numbers or power 
-series, that we are going to discuss now. Let us recall first that a 
-datum involving elements of a complete ultrametric field $K$ (\emph{e.g.} 
-$p$-adic numbers or power series) cannot be stored entirely on a computed 
-because they basically convey an infinite amount of informations. That is 
-why, instead of this datum itself, we generally work with neighborhoods 
-of it. Such neighborhoods are typically sets of the form $x + H$ where 
-$x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice in 
-$E$. A basic example is the approximate $5$-adic number $1 + O(5^{10})$ 
-which corresponds to the neighborhood of $1$ in $\Z_5$, which is the set 
-of all $5$-adic numbers congruent to $1$ modulo $5^{10}$. In the context
-of computer science, we will often use the notation $x + O(H)$ to refer 
-to $x + H$.
+Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have 
+important applications to effective computations with $p$-adic numbers 
+or power series, that we are going to discuss now. Let us recall first 
+that a datum involving elements of a complete ultrametric field $K$ 
+(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
+on a computed because they basically convey an infinite amount of 
+informations. That is why, instead of this datum itself, we generally 
+work with neighborhoods of it. In what follows, we shall consider 
+neighborhoods of the form $x + H$ where $x$ is some element lying in a 
+$K$-Banach $E$ and $H$ is a lattice in $E$. A very basic example is the 
+approximate $p$-adic number $1 + O(p^N)$ which corresponds to the 
+neighborhood of $1$ in $\Z_p$, which is the set of all $p$-adic numbers 
+congruent to $1$ modulo $p^N$. We nevertheless emphasize that our 
+definition is quite flexible and allows many neighborhoods that we do 
+not consider in general. For instance, an ``admissible'' neighborhood 
+around a monic polynomial $P_0 \in \Q_p[X]$ of degree $d$ may be defined 
+by congruences of the form $P(a_i) \equiv P_0(a_i) \pmod {p^{n_i}}$ ($1 
+\leq i \leq d$) for some fixed elements $a_i \in \Q_p$ and relative 
+integers $n_i$.
+
+In the context of computer science, we will often use the notation $x + 
+O(H)$ to refer to $x + H$.
+
+\medskip
 
@@
 often not the case as shown in \S \ref{ssec:stepbystep}.
-
-\smallskip
 
@@
 \begin{rmk}
-In some cases, it would be possible to determine suitable lattices 
+In some cases, it is actually possible to determine suitable lattices 
 $\tilde H_i$ together with their images under $g'_i(x)$ (or, at
@@
 
-\section{Generalization to manifolds}
-\label{sec:manifold}
-
-Many natural $p$-adic objects do not lie in vector spaces:
-points in projective spaces or elliptic curves,
-subspaces of a fixed vector space (which lie in Grassmannians),
-classes of isomorphism of certain curves (which lie in various moduli spaces),
-\emph{etc.}  In this section we extend the formalism 
-developed in Section \ref{sec:philosophy} to a more general setting:
-we consider the quite general case of strictly
-differentiable manifolds locally modeled on ultrametric Banach spaces.
-This covers all the 
-aforementioned examples.
-
-In what follows, the letter $K$ always refers to an ultrametric field 
-(\emph{cf} \S \ref{ssec:ultrametric}). The absolute value over $K$ is 
-denoted by $|\cdot|$.
-
-\subsection{Strictly differentiable $K$-manifolds}
-\label{ssec:manifold}
-
-The theory of finite dimensional $K$-manifolds is presented for example 
-in \cite{Schneider}, parts 8 and 9. In this Section, we shall work with
-a slightly different notion of manifolds which allows also Banach vector 
-spaces of infinite dimension.
-More precisely, for us, a \emph{strictly differentiable $K$-manifold} 
-(or just \emph{$K$-manifold} for short) is the data of a topological 
-space $V$ together with an open covering $V = \bigcup_{i \in I} V_i$ 
-(where $I$ is some set) and, for all $i \in I$, an homeomorphism 
-$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach 
-$E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
-nonempty, the composite map
-\begin{equation}
-\label{eq:psiij}
-\psi_{ij} : 
-\varphi_i(V_{ij}) \stackrel{\varphi_i^{-1}}{\longrightarrow} 
-V_{ij} \stackrel{\varphi_j}{\longrightarrow} \varphi_j(V_{ij})
-\quad \text{(with } V_{ij} = V_i \cap V_j \text{)}
-\end{equation}
-is strictly differentiable at all points in its domain. We recall that
-the mappings $\varphi_i$ above are the so-called \emph{charts} and that
-their collection defines an \emph{atlas} of $V$. In the sequel, we shall
-assume further that the open covering $V = \bigcup_{i \in I} V_i$ is
-locally finite, which means that every point $x \in V$ lies only in a
-finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
-$K$-Banach themselves.
-
-If $V$ is a $K$-manifold and $x$ is a point of $V$, we define the 
-\emph{tangent space} $T_x V$ of $V$ at $x$ as the space $E_i$ for some 
-$i$ such that $x \in V_i$. We note that if $x$ belongs to $V_i$ and 
-$V_j$, the linear map $\psi'_{ij}(\varphi_i(x))$ defines an isomorphism 
-between $E_i$ and $E_j$. Furthermore these isomorphisms are compatible 
-in an obvious way. This implies that the definition of $T_x V$ given 
-above does not depend (up to some canonical isomorphism) on the index 
-$i$ such that $x \in V_i$ and then makes sense.
-
-As usual, we can define the notion of strict differentiability (at some 
-point) for a continuous mapping between two $K$-manifolds by viewing it 
-through the charts. A strictly differentiable map $f : V \to V'$ induces 
-a linear map on tangent spaces $f'(x) : T_x V \to T_{f(x)} V'$ for all 
-$x$ in the domain $V$. It is called the \emph{differential} of $f$ at
-$x$.
-
-\subsection{Precision data}
-
-\todofor{David/Xavier}{Complete this Subsection}
-
-Going back to our problem of precision, given $V$ a $K$-manifold as 
-above, we would like to be able to deal with ``approximations up to some 
-precision'' of elements in $V$, \emph{i.e.} expressions of the form $x + 
-O(H)$ where $x$ belongs to a dense \emph{computable} subset of $V$ and 
-$H$ is a ``precision data''.
-The aim of this paragraph is to give a precise meaning to the expression 
-$x + O(H)$ and to prove several basic properties of it.
-For now, we fix a $K$-manifold $V$ and we use freely the notations $I$, 
-$V_i$, $\varphi_i$, \emph{etc.} introduced in \S \ref{ssec:manifold}.
-
-\begin{deftn}
-Let $x \in V$. 
-A \emph{precision datum} at $x$ is a lattice in the tangent space 
-$T_x V$.
-\end{deftn}
-
-\begin{deftn}
-Let $x \in V$ and $H$ be a precision datum at $x$.
-For all $i \in I$ such that $x \in V_i$, we set:
-$$x + O_i(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
-\subset V.$$
-\end{deftn}
-
-We emphasize that if $x$ belongs to $V_i$ and $V_j$ at the same time,
-the subsets $x + O_i(H)$ and $x + O_j(H)$ differ in general. It is 
-however not the case if $H$ is small enough.
-
-\begin{prop}
-\label{prop:independence}
-Let $x \in V$ and $H$ be a precision data at $x$.
-If $H$ is ``small enough'' \todo{(precise!)}, the subset $x + 
-O_i(H)$ does not depend on $i$.
-\end{prop}
-
-\begin{proof}
-Let $i$ and $j$ be two indices such that $x$ belongs to $V_i$ and $V_j$. 
-Set $x_i = \varphi_i(x) \in E_i$ and $H_i = \varphi'_i(x)(H)$. Going 
-back to the definitions, we see that the equality $x + O_i(H) = x + 
-O_j(H)$ is equivalent to:
-$$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$$ 
-where we recall that $\psi_{ij}$ is defined by Eq.~\eqref{eq:psiij}.
-\todo{Complete the proof and write the correct condition on $H$!}
-\end{proof}
-
-Under the assumption of Proposition \ref{prop:independence}, we define
-$x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
-$x \in V_i$.
-
-\subsubsection*{Change of base point}
-
-In order to restrict ourselves to elements $x$ lying in a dense 
-computable subset, we need to compare $x_0 + O(H_0)$ with varying $x + 
-O(H)$ when $x$ and $x_0$ are close enough.
-Let us first examine the situation in a fixed given chart: we fix some 
-index $i \in I$ and pick two elements $x_0$ and $x$ in $V_i$. We consider 
-in addition a precision datum $H_0$ at $x_0$ and we want to produce a 
-precision datum $H$ at $x$ such that $x_0 + O_i(H_0) = x + O_i(H)$. 
-We remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
-isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and 
-$\varphi'_i(x)$ respectively. A natural candidate for $H$ is then:
-\begin{equation}
-\label{eq:Hprime}
-H = \left(\varphi'_i(x)^{-1} \circ \varphi'_i(x_0)\right) (H_0).
-\end{equation}
-And indeed, one can check that $x + O(H)$ is indeed equal to $x_0 + 
-O(H_0)$ as soon as $x$ and $x_0$ are close enough in the following sense: 
-the difference $\varphi_i(x) - \varphi_i(x_0)$ lies in the lattice $H_i = 
-\varphi'_i(x_0)(H_0)$. We furthermore have a property of independence on 
-$i$.
-
-\begin{prop}
-We assume that $H_0$ is ``small enough'' \todo{(precise!)} and that $x$ is close 
-enough to $x_0$. Then
-\begin{enumerate}[(i)]
-\item the lattice $H$ defined by \eqref{eq:Hprime} does not depend 
-on $i$,
-\item the pairs $(x,H)$ and $(x_0, H_0)$ satisfy the assumptions of
-Proposition \ref{prop:independence} and we have $x + O(H) = x_0 + O(H_0)$.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}
-We first prove (i). For an index $i$ such that $x, x_0 \in V_i$, let us 
-denote by $f_i : T_{x_0} V \to T_x V$ the composite $\varphi'_i(x)^{-1} 
-\circ \varphi'_i(x_0)$. Given an extra index $j$ satisfying the same
-assumption, we know by \todo{(find a reference)} that the difference
-$f_i - f_j$ goes to $0$ when $x$ converges to $x_0$. Since $H_0$ is open
-in $T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ and 
-$f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in f_j(H_0)$ and 
-write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal to $f_i(v) + 
-(f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each summand does. 
-Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse inclusion can be proved 
-in the same way.
-
-We now prove (ii). \todo{Write the proof.}
-\end{proof}
-
-\subsection{Generalization of the main Lemma}
-
-\todofor{David/Xavier}{Extend Lemma \ref{lem:main} to manifolds.}
-
-\section{Examples and applications}
-
-In this Section, we illustrate the theory developed in this paper by 
-giving examples and applications. In \S \ref{ssec:differentials}, we 
-compute the differential of a bunch of usual operations involving 
-$p$-adic objects of different nature. Very often, we will observe that 
-this computation is easy even if the underlying operation is 
-sophisticated (\emph{e.g.} Gr\"obner basis).
-In \S \ref{ssec:SOMOS-solution}, we study the SOMOS 4 sequence 
-introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
-Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic}, we design
-a quite \emph{stable} algorithm for computing it.
-
-\subsection{Differential of usual operations}
-\label{ssec:differentials}
-
-As in the previous Sections, we always work over a fixed ultrametric 
-field $K$, whose absolute value is denoted by $|\cdot|$. We recall that 
-usual examples are the field of $p$-adic numbers $\Q_p$ and the field of 
-Laurent series $k((X))$ over a base field $k$.
-
-In what follows, we use freely the ``method of physicists'' to compute 
-differentials: given a function $f$ differentiable at some point $x$, we 
-consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$ 
-by expanding LHS and neglecting terms of order $2$. The differential of 
-$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
-
-\subsubsection*{Numbers}
-
-The most basic operations are, of course, sums, products and inverses of 
-elements of $K$. Their differential are well-known and quite easy to 
-compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have 
-$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = - 
-\frac{dx}{x^2}$).
-
-A little bit more interesting is the $n$-th power map $f$ from $K$ to 
-itself. Its differential $f'(x)$ is obtained by differentiate the 
-equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
-Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
-number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
-means that the behaviour of the precision depends on the absolute value 
-of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
-this inegality might be strict. It happens for instance if the 
-characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
-$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
-not apply. 
-Another situation (which is more interesting) is that of $p$-adic 
-numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
-$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
-raising a $p$-adic number to the $p^k$-th power increases the precision 
-by $k$ extra digits.
-
-\subsubsection*{Univariate polynomials}
-
-Beyond sums and products (which can be treated as before), a basic 
-operation involving polynomials is Euclidean division. Let us compute 
-the differential of it. We start with two polynomials $A$ and $B$ with 
-$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
-with $\deg R < \deg B$. Differentiating the above equality we find: 
-$dA - dB \cdot Q = B \cdot dQ + dR$,
-which implies that $dQ$ and $dR$ are respectively obtained as the 
-quotient and the remainder of the Euclidean division of $dA - dB \cdot 
-Q$ by $B$. This gives the differential. We note that the discussion 
-above extends readily to convergent series (see also 
-\cite{Caruso-Lubicz}).
-
-Another usual operations on polynomials is the factorization. Suppose 
-that we are given a monic polynomial $P \in K[X]$ written as a product 
-$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
-lemma implies that there exists a small neighborhood $\mathcal U$ of 
-$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
-as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
-respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
-defined on the subset of $\mathcal U$ consisting of monic polynomials.
-We want to differentiate $f$ at $P_0$. For this, we differentiate the 
-equality $P = A B$ around $P_0$, obtaining 
-\begin{equation}
-\label{eq:difffactor}
-dP = A_0 \cdot dB + B_0 \cdot dA.
-\end{equation}
-where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
-$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
-between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
-$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
-{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).
-
-An important special case of the previous study occurs when $A_0$ has 
-degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
-application $P \mapsto A$ is then nothing but the mapping that follows 
-the simple root $\alpha$. Of course, its differential around $P_0$ can 
-be computed by the above method. Nevertheless there is a more direct way 
-to obtain it in this particular case: we write $(P + dP)(\alpha + 
-d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
-dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
-is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
-= - \frac{dP(\alpha)}{P'(\alpha)}$.
-
-\subsubsection*{Multivariate polynomials}
-
-We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
-variables over $K$ and fix a monomial order on it. We then have a 
-notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ are polynomials
-in $K[\XX]$, there exists a writing:
-$$f = q_1 f_1 + \cdots + q_s f_s + r$$
-where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
-by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
-that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
-assumption, there still exists a canonical choice of $r$. We do not know
-if the computation of the differential that follows extends to a more
-general setting.} (of the ideal generated by these polynomials), the 
-polynomial $r$ is uniquely determined and called the \emph{remainder} of 
-the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
-$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
-the case of univariate polynomials, we can compute its differential: we 
-find that $dr$ is obtained as the remainder of the division of $f - (q_1 
-\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
-
-We now focus on the computation of Gr\"obner basis. Recall that any 
-ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
-then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
-\emph{homogeneous}\footnote{This restriction will be convenient for us 
-mainly because we are using the article \cite{Vaccon}, which deals only 
-with homogeneous polynomials. It is however probably not essential.} 
-polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
-application at $(f_1, \ldots, f_s)$ is continuous provided that:
-\begin{itemize}
-\item the sequence $(f_1, \ldots, f_s)$ is regular
-\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
-$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{Vaccon} for more precisions).
-\end{itemize}
-A similar argument proves that it is actually differentiable at such
-points. We are now going to compute the differential. For this we
-remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
-g_t)$ generate the same ideal, we have a relation:
-\begin{equation}
-\label{eq:prodgrobner}
-(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
-\end{equation}
-where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
-Moreover following Vaccon's construction, we see that the entries of $A$ 
-can be chosen in such a way that they define differentiable functions.
-Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
-$$(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA = 
-(d g_1, \ldots, d g_t).$$
-Finally, using that we are working with \emph{reduced} Gr\"obner
-basis, the above equality implies that $d g_i$ is the remainder in
-the division of the $i$-th coefficient of the product
-$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
-
-\subsubsection*{Matrices}
-
-Differentiating ring operations (sum, products, inverse) over matrices 
-is again straightforward; we just need to be careful that matrix 
-algebras are (in general) noncommutative.
-
-We now move to determinants, \emph{i.e.} we compute the differential
-of the function $\det : M_n(K) \to K$. It is a standard computation.
-To begin with, we consider an invertible matrix $M$ and write:
-\begin{eqnarray*}
-\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
-&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
-&=& \det(M) + \tr(\com(M) \cdot dM)
-\end{eqnarray*}
-where $\com(M)$ denotes the comatrix of $M$. The differential of 
-$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
-out that this formula is still valid when $M$ is not invertible.
-The same computation extends readily to characteristic polynomials,
-since they are defined as determinants. More precisely, let us 
-consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
-the latter ring denotes the finite dimensional vector space of 
-polynomials over $K$ of degree at most $n$) taking a matrix $M$
-to its characteristic polynomial $\det(X-M)$.
-Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
-differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
-
-Other important tools related to matrices are usual factorizations as, 
-for example, LU factorization. In what follows, let us agree to define a 
-LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
-where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
-Such a writing exists and is unique provided that all principal minors 
-of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
-defined over the Zariski-open set of matrices satisfying the above 
-condition. In order to differentiate it, we differentiate the relation 
-$M = LU$ and rewrite the result as follows:
-$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
-We remark that in the right hand side of the above formula, the first
-summand is lower triangular and all its diagonal entries vanish whereas 
-the second summand is upper triangular. Hence in order to compute $dL$
-and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
-L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
-obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
-recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
-left and $U$ on the right respectively. 
-The above discussion extends almost \emph{verbatim} to LUP 
-factorization; the only difference is that LUP factorizations are not 
-unique but they are on a small neighborhood of $M$ if we fix the matrix 
-$P$.
-
-A similar argument works for QR factorization as well.
-For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
-a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
-orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
-a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
-mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
-would like to emphasize at this point that the orthogonality condition
-defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
-vector space (it is defined by equations of degree $2$). The codomain
-of $f$ is then also a manifold; this example then fits to the setting
-of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
-The generalization made in \S \ref{sec:manifold} is then needed.
-Anyway, we can differentiate $f$ by following the method we used for LU 
-factorization: differentiating the relation $M = QR$, we obtain:
-\begin{equation}
-\label{eq:diffQR}
-\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
-\cdot R^{-1}
-\end{equation}
-where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
-Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
-$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
-matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
-triangular with all diagonal entries equal to $0$, we see that 
-Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
-and $dM$.
-
-\subsubsection*{Vector spaces}
-
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
-Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
-of dimension $d$. It defines an algebraic variety over $K$ and hence
-\emph{a fortiori} a $K$-manifold in the sense of \S \ref{ssec:manifold}.
-Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
-a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
-and two such matrices $M$ and $M'$ define the same vector space if there 
-exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
-find that we can always choose the above matrix $M$ in the particular
-shape:
-$$M = \begin{pmatrix} I_d & M' \end{pmatrix} \cdot P$$
-where $I_d$ denotes the $(d \times d)$ identity matrix, $M' \in M_{d, 
-n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
-such writings with the same $P$ necessarily coincide. Hence each 
-permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
-is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
-
-Let us now consider the open subset of $M_{n,n-d}(K)$ consisting of 
-matrices of full rank. The left kernel defines a mapping $\text{LK} : 
-M_{n,n-d}(K) \to \Grass(d,n)$. Let us prove that it is differentiable 
-and compute its differential around some point $M \in M_{n,n-d}(K)$. Of 
-course, there exists a neighborhood $\mathcal U$ of $M$ whose image is 
-entirely contained in a given chart $U_P$. We assume for simplicity that 
-$P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
-our chart to the application that takes $M$ to the unique matrix $N \in
-M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
-= 0$. The implicit function theorem then implies that $\text{LK}$ is
-differentiable. Furthermore, its differential satisfies the relation
-$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
-\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
-from which we can compute $dN$ by projecting on the $(n-d)$ last
-columns.
-
-We can develop further this example and study what happens on the 
-closed subset of $M_{n,n-d}(K)$ where matrices have not full rank. On
-this subspace, the left kernel has dimension $< d$ and then no longer
-defines a point in $\Grass(n,d)$. Nevertheless, for all integer $r <
-n-d$, we can consider the subset $\mathcal V_r \subset M_{n,n-d}(K)$
-of matrices whose rank are exactly rank $d$. It is locally closed in
-$M_{n,n-d}(K)$ with respect to the Zariski topology and hence defines
-a $K$-manifold. Furthermore, we have a mapping $\text{LK}_r : \mathcal 
-M_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
-can be computed as before.
-
-The lesson from this study is the following. If we have to compute the 
-kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
-priori} we first need to determine what is the rank of $M$. However, the 
-rank function is not continuous --- it is only lower semi-continuous --- 
-and hence cannot be determined from sure from $M$ (expect if $M$ has 
-full rank). We then need to make an extra hypothesis and the more natural 
-one is arguably to assume that $M$ has actually the minimal rank that it 
-can have. With this convention, for instance, the kernel of the zero 
-matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
-whole space although there exist invertible matrices in any 
-neighborhood of $M$.
-
-We now focus on the computation of intersections. We pick $n$, $d_1$ and 
-$d_2$ three nonnegative integers such that $d_1 \leq n$, $d_2 \leq n$ 
-and $d_1 + d_2 \geq n$. Two subspaces of $K^n$ of dimension $d_1$ and 
-$d_2$ respectively meet along a subspace of dimension at most $d_1 + d_2 
-- n$. For all $d \leq d_1 + d_2 -n$, we can then define the subspace 
-$\mathcal V_d$ of $\Grass(n,d_1) \times \Grass(n,d_2)$ consisting of 
-pairs $(E_1, E_2)$ such that $\dim (E_1 \cap E_2) = d$ and consider the 
-function $f_d : \mathcal V_d \to \Grass(n, d)$ which sends $(E_1, E_2)$ 
-to $E_1 \cap E_2$. In charts, the function $f_d$ can be interpreted as
-the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
-kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We deduce
-for this that $f_d$ is differentiable and that its differential can be
-computed similarly to what we have done before.
-
-Finally, we note that similar results hold for images of matrices and,
-consequently, sums of subspaces.
-
-\subsection{The SOMOS 4 sequence}
+\subsection{Application to SOMOS sequence}
 \label{ssec:SOMOS-solution}
+
+We illustrate the theory developed in this paper by giving a simple 
+toy application. Other applications will be discussed in subsequent 
+articles. More precisely, we study the SOMOS 4 sequence introduced in \S 
+\ref{ssec:stepbystep}. Making a crucial use of Lemma \ref{lem:main} and 
+Proposition \ref{prop:locanalytic}, we design a \emph{stable} algorithm 
+for computing it.
 
@@
 are known with finite precision $O(p^N)$ then all terms $u_n$'s are 
-known with the same precision.
-
-In this last paragraph, we illustrate the theory developed in this 
-article by designing an algorithm for computing the $u_n$'s at precision 
-$O(p^N)$. The algorithm is presented at the top of page 
-\pageref{algo:SOMOS}. We now prove that it is correct.
-%
+known with the same precision. Algorithm \ref{algo:SOMOS} presented on 
+page \pageref{algo:SOMOS} performs this computation.
+
 \begin{algorithm}[t]
@@
 \end{algorithm}
-%
+
+We now prove that it is correct.
 We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
@@
 $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
-= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is strictly 
+= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is  
 differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
@@
 
+\appendix
+
+\section{Generalization to manifolds}
+\label{sec:manifold}
+
+Many natural $p$-adic objects do not lie in vector spaces:
+points in projective spaces or elliptic curves,
+subspaces of a fixed vector space (which lie in Grassmannians),
+classes of isomorphism of certain curves (which lie in various moduli spaces),
+\emph{etc.}  In this section we extend the formalism 
+developed in Section \ref{sec:philosophy} to a more general setting:
+we consider the quite general case of 
+differentiable manifolds locally modeled on ultrametric Banach spaces.
+This covers all the 
+aforementioned examples.
+
+In what follows, the letter $K$ always refers to an ultrametric field 
+(\emph{cf} \S \ref{ssec:ultrametric}). The absolute value over $K$ is 
+denoted by $|\cdot|$.
+
+\subsection{Differentiable $K$-manifolds}
+\label{ssec:manifold}
+
+The theory of finite dimensional $K$-manifolds is presented for example 
+in \cite{Schneider}, parts 8 and 9. In this Section, we shall work with
+a slightly different notion of manifolds which allows also Banach vector 
+spaces of infinite dimension.
+More precisely, for us, a \emph{ differentiable $K$-manifold} 
+(or just \emph{$K$-manifold} for short) is the data of a topological 
+space $V$ together with an open covering $V = \bigcup_{i \in I} V_i$ 
+(where $I$ is some set) and, for all $i \in I$, an homeomorphism 
+$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach 
+$E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
+nonempty, the composite map
+\begin{equation}
+\label{eq:psiij}
+\psi_{ij} : 
+\varphi_i(V_{ij}) \stackrel{\varphi_i^{-1}}{\longrightarrow} 
+V_{ij} \stackrel{\varphi_j}{\longrightarrow} \varphi_j(V_{ij})
+\quad \text{(with } V_{ij} = V_i \cap V_j \text{)}
+\end{equation}
+is  differentiable at all points in its domain. We recall that
+the mappings $\varphi_i$ above are the so-called \emph{charts} and that
+their collection defines an \emph{atlas} of $V$. In the sequel, we shall
+assume further that the open covering $V = \bigcup_{i \in I} V_i$ is
+locally finite, which means that every point $x \in V$ lies only in a
+finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
+$K$-Banach themselves.
+
+If $V$ is a $K$-manifold and $x$ is a point of $V$, we define the 
+\emph{tangent space} $T_x V$ of $V$ at $x$ as the space $E_i$ for some 
+$i$ such that $x \in V_i$. We note that if $x$ belongs to $V_i$ and 
+$V_j$, the linear map $\psi'_{ij}(\varphi_i(x))$ defines an isomorphism 
+between $E_i$ and $E_j$. Furthermore these isomorphisms are compatible 
+in an obvious way. This implies that the definition of $T_x V$ given 
+above does not depend (up to some canonical isomorphism) on the index 
+$i$ such that $x \in V_i$ and then makes sense.
+
+As usual, we can define the notion of strict differentiability (at some 
+point) for a continuous mapping between two $K$-manifolds by viewing it 
+through the charts. A  differentiable map $f : V \to V'$ induces 
+a linear map on tangent spaces $f'(x) : T_x V \to T_{f(x)} V'$ for all 
+$x$ in the domain $V$. It is called the \emph{differential} of $f$ at
+$x$.
+
+\subsection{Precision data}
+
+\todofor{David/Xavier}{Complete this Subsection}
+
+Going back to our problem of precision, given $V$ a $K$-manifold as 
+above, we would like to be able to deal with ``approximations up to some 
+precision'' of elements in $V$, \emph{i.e.} expressions of the form $x + 
+O(H)$ where $x$ belongs to a dense \emph{computable} subset of $V$ and 
+$H$ is a ``precision data''.
+The aim of this paragraph is to give a precise meaning to the expression 
+$x + O(H)$ and to prove several basic properties of it.
+For now, we fix a $K$-manifold $V$ and we use freely the notations $I$, 
+$V_i$, $\varphi_i$, \emph{etc.} introduced in \S \ref{ssec:manifold}.
+
+\begin{deftn}
+Let $x \in V$. 
+A \emph{precision datum} at $x$ is a lattice in the tangent space 
+$T_x V$.
+\end{deftn}
+
+\begin{deftn}
+Let $x \in V$ and $H$ be a precision datum at $x$.
+For all $i \in I$ such that $x \in V_i$, we set:
+$$x + O_i(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
+\subset V.$$
+\end{deftn}
+
+We emphasize that if $x$ belongs to $V_i$ and $V_j$ at the same time,
+the subsets $x + O_i(H)$ and $x + O_j(H)$ differ in general. It is 
+however not the case if $H$ is small enough.
+
+\begin{prop}
+\label{prop:independence}
+Let $x \in V$ and $H$ be a precision data at $x$.
+If $H$ is ``small enough'' \todo{(precise!)}, the subset $x + 
+O_i(H)$ does not depend on $i$.
+\end{prop}
+
+\begin{proof}
+Let $i$ and $j$ be two indices such that $x$ belongs to $V_i$ and $V_j$. 
+Set $x_i = \varphi_i(x) \in E_i$ and $H_i = \varphi'_i(x)(H)$. Going 
+back to the definitions, we see that the equality $x + O_i(H) = x + 
+O_j(H)$ is equivalent to:
+$$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$$ 
+where we recall that $\psi_{ij}$ is defined by Eq.~\eqref{eq:psiij}.
+\todo{Complete the proof and write the correct condition on $H$!}
+\end{proof}
+
+Under the assumption of Proposition \ref{prop:independence}, we define
+$x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
+$x \in V_i$.
+
+\subsubsection*{Change of base point}
+
+In order to restrict ourselves to elements $x$ lying in a dense 
+computable subset, we need to compare $x_0 + O(H_0)$ with varying $x + 
+O(H)$ when $x$ and $x_0$ are close enough.
+Let us first examine the situation in a fixed given chart: we fix some 
+index $i \in I$ and pick two elements $x_0$ and $x$ in $V_i$. We consider 
+in addition a precision datum $H_0$ at $x_0$ and we want to produce a 
+precision datum $H$ at $x$ such that $x_0 + O_i(H_0) = x + O_i(H)$. 
+We remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
+isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and 
+$\varphi'_i(x)$ respectively. A natural candidate for $H$ is then:
+\begin{equation}
+\label{eq:Hprime}
+H = \left(\varphi'_i(x)^{-1} \circ \varphi'_i(x_0)\right) (H_0).
+\end{equation}
+And indeed, one can check that $x + O(H)$ is indeed equal to $x_0 + 
+O(H_0)$ as soon as $x$ and $x_0$ are close enough in the following sense: 
+the difference $\varphi_i(x) - \varphi_i(x_0)$ lies in the lattice $H_i = 
+\varphi'_i(x_0)(H_0)$. We furthermore have a property of independence on 
+$i$.
+
+\begin{prop}
+We assume that $H_0$ is ``small enough'' \todo{(precise!)} and that $x$ is close 
+enough to $x_0$. Then
+\begin{enumerate}[(i)]
+\item the lattice $H$ defined by \eqref{eq:Hprime} does not depend 
+on $i$,
+\item the pairs $(x,H)$ and $(x_0, H_0)$ satisfy the assumptions of
+Proposition \ref{prop:independence} and we have $x + O(H) = x_0 + O(H_0)$.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}
+We first prove (i). For an index $i$ such that $x, x_0 \in V_i$, let us 
+denote by $f_i : T_{x_0} V \to T_x V$ the composite $\varphi'_i(x)^{-1} 
+\circ \varphi'_i(x_0)$. Given an extra index $j$ satisfying the same
+assumption, we know by \todo{(find a reference)} that the difference
+$f_i - f_j$ goes to $0$ when $x$ converges to $x_0$. Since $H_0$ is open
+in $T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ and 
+$f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in f_j(H_0)$ and 
+write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal to $f_i(v) + 
+(f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each summand does. 
+Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse inclusion can be proved 
+in the same way.
+
+We now prove (ii). \todo{Write the proof.}
+\end{proof}
+
+\subsection{Generalization of the main Lemma}
+
+\todofor{David/Xavier}{Extend Lemma \ref{lem:main} to manifolds.}
+
+\section{Differential of usual operations}
+\label{ssec:differentials}
+
+We continue to work over a fixed ultrametric field $K$, whose absolute 
+value is denoted by $|\cdot|$. We recall that usual examples of such $K$ 
+are the field of $p$-adic numbers $\Q_p$ and the field of Laurent series 
+$k((X))$ over a base field $k$. The aim of this appendix is to compute 
+the differential of many usual operations on numbers, polynomials and 
+matrices. Surprisingly, we observe that all differentials we will 
+consider are rather easy to compute even if the underlying operation is 
+quite involved (\emph{e.g.} Gr\"obner basis).
+
+In what follows, we use freely the ``method of physicists'' to compute 
+differentials: given a function $f$ differentiable at some point $x$, we 
+consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$ 
+by expanding LHS and neglecting terms of order $2$. The differential of 
+$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
+
+\subsection{Numbers}
+
+The most basic operations are, of course, sums, products and inverses of 
+elements of $K$. Their differential are well-known and quite easy to 
+compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have 
+$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = - 
+\frac{dx}{x^2}$).
+
+A little bit more interesting is the $n$-th power map $f$ from $K$ to 
+itself. Its differential $f'(x)$ is obtained by differentiate the 
+equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
+Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
+number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
+means that the behaviour of the precision depends on the absolute value 
+of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
+this inegality might be strict. It happens for instance if the 
+characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
+$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
+not apply. 
+Another situation (which is more interesting) is that of $p$-adic 
+numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
+$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
+raising a $p$-adic number to the $p^k$-th power increases the precision 
+by $k$ extra digits.
+
+\subsection{Univariate polynomials}
+
+For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
+polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
+space of dimension $d+1$. The affine space $X^d + K_{< d}[X]$ is then 
+the set of monic polynomials over $K$ of degree $d$. We denote it by 
+$K_d[X]$.
+
+\subsubsection*{Euclidean division}
+
+Beyond sums and products (which can be treated as before), a basic 
+operation involving polynomials is Euclidean division. Let us compute 
+the differential of it. We start with two polynomials $A$ and $B$ with 
+$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
+with $\deg R < \deg B$. Differentiating the above equality we find: 
+$dA - dB \cdot Q = B \cdot dQ + dR$,
+which implies that $dQ$ and $dR$ are respectively obtained as the 
+quotient and the remainder of the Euclidean division of $dA - dB \cdot 
+Q$ by $B$. This gives the differential. We note that the discussion 
+above extends readily to convergent series (see also 
+\cite{Caruso-Lubicz}).
+
+\subsubsection*{Subresultants and B\'ezout coefficients}
+
+\todo{To be written.}
+
+\subsubsection*{Factorization}
+
+Suppose that we are given a monic polynomial $P \in K[X]$ written as a 
+product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
+Hensel's lemma implies that there exists a small neighborhood $\mathcal 
+U$ of $P_0$ such that any monic polynomial $P \in \mathcal U$ factors 
+uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
+and $B_0$ respectively. Thus, we can consider the application $f : P 
+\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
+polynomials. We want to differentiate $f$ at $P_0$. For this, we 
+differentiate the equality $P = A B$ around $P_0$, obtaining
+\begin{equation}
+\label{eq:difffactor}
+dP = A_0 \cdot dB + B_0 \cdot dA.
+\end{equation}
+where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
+$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
+between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
+$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
+{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).
+
+An important special case of the previous study occurs when $A_0$ has 
+degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
+application $P \mapsto A$ is then nothing but the mapping that follows 
+the simple root $\alpha$. Of course, its differential around $P_0$ can 
+be computed by the above method. Nevertheless there is a more direct way 
+to obtain it in this particular case: we write $(P + dP)(\alpha + 
+d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
+dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
+is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
+= - \frac{dP(\alpha)}{P'(\alpha)}$.
+
+\subsection{Multivariate polynomials}
+
+We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
+variables over $K$ and fix a monomial order on it.
+
+\subsubsection*{Division}
+
+We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
+are polynomials in $K[\XX]$, there exists a writing:
+$$f = q_1 f_1 + \cdots + q_s f_s + r$$
+where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
+by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
+that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
+assumption, there still exists a canonical choice of $r$. We do not know
+if the computation of the differential that follows extends to a more
+general setting.} (of the ideal generated by these polynomials), the 
+polynomial $r$ is uniquely determined and called the \emph{remainder} of 
+the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
+$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
+the case of univariate polynomials, we can compute its differential: we 
+find that $dr$ is obtained as the remainder of the division of $f - (q_1 
+\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
+
+\subsubsection*{Gr\"obner basis}
+
+Recall that any 
+ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
+then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
+\emph{homogeneous}\footnote{This restriction will be convenient for us 
+mainly because we are using the article \cite{Vaccon}, which deals only 
+with homogeneous polynomials. It is however probably not essential.} 
+polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
+\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
+application at $(f_1, \ldots, f_s)$ is continuous provided that:
+\begin{itemize}
+\item the sequence $(f_1, \ldots, f_s)$ is regular
+\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
+$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
+order (\emph{cf} \cite{Vaccon} for more precisions).
+\end{itemize}
+A similar argument proves that it is actually differentiable at such
+points. We are now going to compute the differential. For this we
+remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
+g_t)$ generate the same ideal, we have a relation:
+\begin{equation}
+\label{eq:prodgrobner}
+(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
+\end{equation}
+where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
+Moreover following Vaccon's construction, we see that the entries of $A$ 
+can be chosen in such a way that they define differentiable functions.
+Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
+$$(d g_1, \ldots, d g_t) =
+(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
+Finally, using that we are working with \emph{reduced} Gr\"obner
+basis, the above equality implies that $d g_i$ is the remainder in
+the division of the $i$-th coefficient of the product
+$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
+
+\subsection{Matrices}
+
+Differentiating ring operations (sum, products, inverse) over matrices 
+is again straightforward; we just need to be careful that matrix 
+algebras are (in general) noncommutative.
+
+\subsubsection*{Determinants and characteristic polynomials}
+
+Let us first compute the differential of the function $\det : M_n(K) \to 
+K$. It is a standard computation. To begin with, we consider an 
+invertible matrix $M$ and write:
+\begin{eqnarray*}
+\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
+&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
+&=& \det(M) + \tr(\com(M) \cdot dM)
+\end{eqnarray*}
+where $\com(M)$ denotes the comatrix of $M$. The differential of 
+$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
+out that this formula is still valid when $M$ is not invertible.
+The same computation extends readily to characteristic polynomials,
+since they are defined as determinants. More precisely, let us 
+consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
+the latter ring denotes the finite dimensional vector space of 
+polynomials over $K$ of degree at most $n$) taking a matrix $M$
+to its characteristic polynomial $\det(X-M)$.
+Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
+differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
+
+\subsubsection*{LU factorization}
+
+Let us agree to define a 
+LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
+where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
+Such a writing exists and is unique provided that all principal minors 
+of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
+defined over the Zariski-open set of matrices satisfying the above 
+condition. In order to differentiate it, we differentiate the relation 
+$M = LU$ and rewrite the result as follows:
+$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
+We remark that in the right hand side of the above formula, the first
+summand is lower triangular and all its diagonal entries vanish whereas 
+the second summand is upper triangular. Hence in order to compute $dL$
+and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
+L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
+obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
+recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
+left and $U$ on the right respectively. 
+The above discussion extends almost \emph{verbatim} to LUP 
+factorization; the only difference is that LUP factorizations are not 
+unique but they are on a small neighborhood of $M$ if we fix the matrix 
+$P$.
+
+\subsubsection*{QR factorization}
+
+For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
+a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
+orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
+a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
+mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
+would like to emphasize at this point that the orthogonality condition
+defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
+vector space (it is defined by equations of degree $2$). The codomain
+of $f$ is then also a manifold; this example then fits to the setting
+of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
+The generalization made in \S \ref{sec:manifold} is then needed.
+Anyway, we can differentiate $f$ by following the method we used for LU 
+factorization: differentiating the relation $M = QR$, we obtain:
+\begin{equation}
+\label{eq:diffQR}
+\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
+\cdot R^{-1}
+\end{equation}
+where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
+Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
+$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
+matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
+triangular with all diagonal entries equal to $0$, we see that 
+Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
+and $dM$.
+
+\subsection{Vector spaces}
+
+Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
+Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
+of dimension $d$. It defines an algebraic variety over $K$ and hence
+\emph{a fortiori} a $K$-manifold in the sense of \S \ref{ssec:manifold}.
+Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
+a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
+and two such matrices $M$ and $M'$ define the same vector space if there 
+exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
+find that we can always choose the above matrix $M$ in the particular
+shape:
+$$M = \begin{pmatrix} I_d & M' \end{pmatrix} \cdot P$$
+where $I_d$ denotes the $(d \times d)$ identity matrix, $M' \in M_{d, 
+n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
+such writings with the same $P$ necessarily coincide. Hence each 
+permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
+is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
+
+\subsubsection*{Left kernels}
+
+We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 
+consisting of matrices of full rank, \emph{i.e.} rank $n-d$. The left 
+kernel defines a mapping $\text{LK} : M_{n,n-d}(K) \to \Grass(d,n)$. Let 
+us prove that it is differentiable and compute its differential around 
+some point $M \in M_{n,n-d}(K)$. Of course, there exists a neighborhood 
+$\mathcal U$ of $M$ whose image is entirely contained in a given chart 
+$U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
+U$, the map $\text{LK}$ corresponds in our chart to the application that 
+takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 
+$\begin{pmatrix} I_d & N \end{pmatrix} \cdot M = 0$. The implicit 
+function theorem then implies that $\text{LK}$ is differentiable. 
+Furthermore, its differential satisfies the relation
+$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
+\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
+from which we can compute $dN$ by projecting on the $(n-d)$ last
+columns.
+
+We can develop further this example and study what happens on the 
+closed subset of $M_{n,n-d}(K)$ where matrices have not full rank. On
+this subspace, the left kernel has dimension $< d$ and then no longer
+defines a point in $\Grass(n,d)$. Nevertheless, for all integer $r <
+n-d$, we can consider the subset $\mathcal V_r \subset M_{n,n-d}(K)$
+of matrices whose rank are exactly rank $r$. It is locally closed in
+$M_{n,n-d}(K)$ with respect to the Zariski topology and hence defines
+a $K$-manifold. Furthermore, we have a mapping $\text{LK}_r : \mathcal 
+M_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
+can be computed as before.
+
+The lesson from this study is the following. If we have to compute the 
+kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
+priori} we first need to determine what is the rank of $M$. However, the 
+rank function is not continuous --- it is only lower semi-continuous --- 
+and hence cannot be determined from sure from $M$ (expect if $M$ has 
+full rank). We then need to make an extra hypothesis and the more natural 
+one is arguably to assume that $M$ has actually the minimal rank that it 
+can have. With this convention, for instance, the kernel of the zero 
+matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
+whole space although there exist invertible matrices in any 
+neighborhood of $M$.
+
+\subsubsection*{Intersections}
+
+We pick $n$, $d_1$ and 
+$d_2$ three nonnegative integers such that $d_1 \leq n$, $d_2 \leq n$ 
+and $d_1 + d_2 \geq n$. Two subspaces of $K^n$ of dimension $d_1$ and 
+$d_2$ respectively meet along a subspace of dimension at most $d_1 + d_2 
+- n$. For all $d \leq d_1 + d_2 -n$, we can then define the subspace 
+$\mathcal V_d$ of $\Grass(n,d_1) \times \Grass(n,d_2)$ consisting of 
+pairs $(E_1, E_2)$ such that $\dim (E_1 \cap E_2) = d$ and consider the 
+function $f_d : \mathcal V_d \to \Grass(n, d)$ which sends $(E_1, E_2)$ 
+to $E_1 \cap E_2$. In charts, the function $f_d$ can be interpreted as
+the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
+kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We deduce
+for this that $f_d$ is differentiable and that its differential can be
+computed similarly to what we have done before.
+
+\subsubsection*{Sums and images}
+
+Finally, we note that similar results hold for images of matrices and,
+consequently, sums of subspaces.
+
 \begin{thebibliography}{10}

### 2014-02-05 padicprec.tex
Reference for Somos

--- 2014-02-02 padicprec.tex
@@
 $n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
-1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
+1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{Fomin-Zelevinsky}).
 From the latter formula, it follows directly that if $u_0$, $u_1$,
@@
 
-\begin{deftn}
+\begin{deftn} \label{deftn : diff}
 Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
@@
 C \cdot f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
-$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by the fact 
+$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by definition \ref{deftn : diff} and the fact 
 that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
@@
 $\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
-surjectivityy of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
+surjectivity of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
 x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
@@
 
-If we denote by $NP_F$ the Newton polygon of the power series 
+If we denote by $NP_F$ the Newton polygon application of the power series 
 \[
@@
 as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
-respectively. Thus, we can consider the map $f : P \mapsto (A,B)$ 
+respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
 defined on the subset of $\mathcal U$ consisting of monic polynomials.
@@
 degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
-map $P \mapsto A$ is then nothing but the mapping that follows 
+application $P \mapsto A$ is then nothing but the mapping that follows 
 the simple root $\alpha$. Of course, its differential around $P_0$ can 
@@
 polynomial $r$ is uniquely determined and called the \emph{remainder} of 
-the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
+the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
 $(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
@@
 ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
-then consider the map sending a family $(f_1, \ldots, f_s)$ of 
+then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
 \emph{homogeneous}\footnote{This restriction will be convenient for us 
@@
 \ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
-map at $(f_1, \ldots, f_s)$ is continuous provided that:
+application at $(f_1, \ldots, f_s)$ is continuous provided that:
 \begin{itemize}
@@
 $P$ is the identity. On $\mathcal U$, the map $\text{LK}$ corresponds in 
-our chart to the map that takes $M$ to the unique matrix $N \in
+our chart to the application that takes $M$ to the unique matrix $N \in
 M_{d, n-d}(K)$ such that $\begin{pmatrix} I_d & N \end{pmatrix} \cdot M 
@@
 u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
-polynomial with coefficients in $\Z$ in these indeterminates.
+polynomial with coefficients in $\Z$ in these indeterminates (see \cite{Fomin-Zelevinsky}).
 
@@
 
+\bibitem{Fomin-Zelevinsky}
+  S.~Fomin, A.~Zelevinsky
+  \emph{The Laurent phenomenon,}   Advances in Applied Mathematics 28: 119–144 (2002).
+
 \bibitem{Gaudry}

### 2014-02-05 padicprec.tex
Conflict resolved

--- 2014-02-05 padicprec.tex
@@
 the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
-C \cdot f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
-$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by the fact 
-that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
+C  f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
+$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by by definition \ref{deftn : diff} and the fact that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
 B_E(\delta)$ for some $\delta >0$.
 
-Let $r \in (0, \delta)$.
+Let $r \in (0, \delta)$. Let $B^-_E(r) \subset H \subset
+B_E(r).$
 Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
-Let $x \in H$. We must show that $f(x) \in f'(0) \cdot H$. 
-By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
-\varepsilon \Vert x \Vert $. Therefore, there exists $y \in f'(0) \cdot 
-H$ such that 
-$\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
-surjectivityy of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
-x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
-\in B^-_E(r) \subset H$ and $f(x)= f'(0) \cdot (x-x_1) \in f'(0) 
-\cdot H$.
+
+We first prove that $f(H) \subset f'(0) \cdot H$.
+Let $x \in H$. By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
+\varepsilon \Vert x \Vert $. Therefore, if $y=f(x)-f'(0) \cdot x,$ then $\Vert y \Vert \leq \varepsilon r$. As a consequence of the open mapping Theorem, $B_F(\varepsilon r) \subset 
+\varepsilon C  f'(0) \cdot B_E(r),$ and there exists $x_1 \in \varepsilon C  \cdot B_E(r)$ such that $f'(0) \cdot x_1 =y.$ Since $\varepsilon C <1,$  $x_1 \in B^-_E(r) \subset H.$ and finally, $f(x)= f'(0) \cdot (x-x_1) \in f'(0) \cdot H$.
 
@@
 Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
-We define by induction two sequences $(x_n)$ and $(z_n)$ by
+Since $f(x_0) \in f'(0) \cdot H,$ we can define $z_0 \in E$ to be such that $f'(0) \cdot z_0 = y - f(x_0)$ and $\Vert z_0 \Vert \leq C \cdot \Vert y - f(x_0) \Vert$.
+We then define by induction two sequences $(x_n)$ and $(z_n)$ by
 \begin{itemize}
+\item $x_{n+1}=x_n+z_n$ and 
 \item $z_n \in E$ is such that
 $f'(0) \cdot z_n = y - f(x_n)$ and
-$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$,
-\item $x_{n+1}=x_n+z_n$.
+$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$.
 \end{itemize}
-By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y_0 
-= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y_0 \Vert < \varepsilon 
+By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y 
+= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y \Vert < \varepsilon 
 \Vert x_0 \Vert \leq \varepsilon r$. Thus $\Vert z_0 \Vert \leq 
-\varepsilon C \cdot r < r$. Thus, $x_1$ and $z_0$ lie in $B^-_E(r) 
-\subset H$. An easy induction involving an analogous computation
+\varepsilon C \cdot r < r$. Thus, $z_0$ and $x_1=x_0+z_0$  lie in $B^-_E(r) 
+\subset H$. An easy induction involving an analoguous computation
 shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
@@
 
-Now, we notice that
-$y - f(x_{n+1}) = f(x_n) - f(x_{n+1}) + f'(0) \cdot z_n$.
-Therefore, using differentiability, we get:
+We notice that $y - f(x_{n+1}) = f(x_n) - f(x_{n+1}) + f'(0) \cdot z_n$.
+Since $z_n=x_{n+1}-x_n,$, if we use differentiability, we get:
 $$\Vert y  - f(x_{n+1}) \Vert 
@@
 O(a^n)$ and $\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < 1$.
-This shows that $(x_n)$ is a Cauchy sequence and therefore converges 
-(since $E$ is complete). If $x$ denotes the limit of $(x_n)$, we have
-$x \in H$ and $y=f(x)$. The result is proven.
+This shows that $(x_n)$ is a Cauchy sequence and, since $E$ is complete, it converges.
+Because $H$ is closed, if $x$ denotes the limit of $(x_n)$, we have $x \in H.$ Yet, as a consequence of the differentiability, $f$ is continuous over $U_\varepsilon.$ Therefore, since $x \in H \subset U_\varepsilon,$ then  $y=f(x)$.
+The result is proven.
 \end{proof}
@@
 computed explicitly.
+
+Indeed, when $f$ is a locally analytic function, we prove that we can recover $\delta$ from a Newton polygon defined with the Taylor series expansion of $f.$
+
 We refer to Schneider \cite{Schneider}, \S 6, for a definition of a 
@@
 
-\begin{rmk}
-Let $f$ be locally analytic function $K^n \rightarrow K^m$.
-Let $x \in K^n$ be such that $f'(x)$ is surjective.
-Then the radius of the ball needed can be read from the Newton polygon given by the norms of the Taylor series expansion of $f$ at $x$.
-We assume the $C$ given by the Banach-Schauder is given, $\varepsilon = \frac{1}{C}$ and now, we try to recover the link between $\varepsilon$ and $\delta$.
-\end{rmk}
-
-\begin{prop}
-\label{prop:locanalytic}
-\todo{Write a proposition.}
+\begin{prop}\label{prop:locanalytic}
+Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ be locally analytic map $E \rightarrow F$.
+Let $x \in E$, and let $f(x+h)=f(x)+\sum_{k \geq 0}^{+\infty} L_k \cdot (h,\dots,h)$ be the power series development of $f$ at $x$.
+We assume $L_1$ is surjective.
+Let $C$ be given by the Banach-Schauder theorem for $L_1$, $\varepsilon = \frac{1}{C}$. 
+Let $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L^{k+1}(0) \interleave  t^k \in \mathbb{R}[[t]]$, let $NP_F$ be its Newton polygon.
+Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP_F$.
+Then $r$ such that $\mu=\log_p r$ is enough for lemma \ref{lem:main}.
 \end{prop}
-
 \begin{proof}
-Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=f'(0) \cdot x +\sum_{n \geq 2} \frac{f^{(n)}(0)}{n!} \cdot (x, \dots,x)$ is the Taylor series expansion of $f$ at $0$, coinciding on $B_E (0,r)$.
-
-We shall now find $l$ such that if $x \in B_{K^n}(0,l)$, then $\Vert \sum_{n \geq 2} \frac{f^{(n)}(0)}{n!}(x,\dots,x) \Vert \leq \varepsilon$.
-
-By continuity and ultrametricity, it is enough for $l$ to be such that, if $x \in B_{K^n}(0,l)$, then  \[ \sup_{n \geq 2} \interleave f^{(n)}(0) \interleave \frac{\Vert x \Vert^n }{ \vert n!
-\vert } \leq \varepsilon \Vert x \Vert .\]
-
-Therefore, \[ \sup_{n \geq 2} \interleave f^{(n)}(0) \interleave \frac{l^{n-1} }{\vert n! \vert } \leq \varepsilon . \]
-
-In other words, $\forall n \geq 2$, $\interleave f^{(n)}(0) \interleave \frac{l^{n-1} }{\vert n! \vert } \leq \varepsilon.$
-
-By passing to logarithm and valuation, we get for all $n \geq 2$: \[ \log_p (\interleave f^{(n)}(0) \interleave )-(n-1) \log_p l-val (n!)\geq - \log_p \varepsilon.\]
-
-By Legendre formula, it is enough that for all $n \geq 2$:
-\[
-\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.
-\]
-
-If we denote by $NP_F$ the Newton polygon application of the power series 
-\[
-F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k
-\]
-then it is enough that:
-\[
-\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.
-\]
-
-If $NP_F^*$ is the Legendre-Fanchel transform of $NP_F$, then it is enough that:
-\[ NP_F^* \left( \log_p l - \frac{1}{p-1} \right) \leq \log_p \varepsilon.\]
-
-Therefore, by definition of the Legendre-Fanchel transform, it is enough to:
+Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=L_1 \cdot x +\sum_{n \geq 2} L_n \cdot (x, \dots,x)$ is the Taylor serie expansion of $f$ at $0$, coinciding on $B_E (0,r)$.
+
+We shall now find $l$ such that if $x \in B_{K^n}(0,l)$, then $\Vert \sum_{n \geq 2} L_n \cdot (x,\dots,x) \Vert \leq \varepsilon \vert x \vert $.
+
+By continuity and ultrametricity, it is enough for $l$ to be such that, if $x \in B_{K^n}(0,l)$, then  \[ \sup_{n \geq 2} \interleave L_n \interleave \Vert x \Vert^n \leq \varepsilon \Vert x \Vert .\]
+
+Therefore, \[ \sup_{n \geq 2} \interleave L_n \interleave l^{n-1}  \leq \varepsilon . \]
+
+In other words, $\forall n \geq 2$, $\interleave L_n \interleave \l^{n-1}  \leq \varepsilon.$
+
+By passing to logarithm, we get for all $n \geq 2$ : \[ \log_p (\interleave L_n \interleave )-(n-1) \log_p l \geq - \log_p \varepsilon.\]
+
+If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L_{k+1} \interleave  t^k$, then it is enough that :
+\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \log_p l \right) \geq - \log_p \varepsilon.\]
+
+If $NP_F^*$ is the Legendre-Fanchel transform of $NP_F$, then it is enough that :
+\[ NP_F^* \left( \log_p l  \right) \leq \log_p \varepsilon.\]
+
+Therefore, by definition of the Legendre-Fanchel transform, it is enough to :
 \begin{itemize}
 \item take $M=(0,\log_p \varepsilon)$,
-\item draw the line passing by $M$ tangent to the graph of $NP_F$,
-\item if we denote by $\mu$ the slope of this line, then take $l$ such that $\mu + \frac{1}{p-1}=\log_p l$.
+\item draw the line passing by $M$ tangent to the graph of $NP_F$ (it could be tangent at infinity),
+\item if we denote by $\mu$ the slope of this line, then take $l$ such that $\mu=\log_p l$.
 \end{itemize}
+\end{proof}
+
+We remark that when considering finite-dimensional applications over $p$-adic fields, a more refined proposition is available.
+
+\begin{prop} \label{prop:locanalyticfinitedim}
+If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C$ and $\varepsilon$ are as above.
+It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP_F$.
+Then $r$ such that $\mu+\frac{1}{p-1}=\log_p r$ is enough for lemma \ref{lem:main}. 
+\end{prop}
+\begin{proof}
+In comparison with the previous proposition, $L_n=\frac{f^{(n)}(0)}{n!}$. Therefore, by Legendre formula, it would be enough that for all $n \geq 2$ :
+\[\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon,\]
+and therefore, \[ NP_F^* \left( \log_p l - \frac{1}{p-1} \right) \leq \log_p \varepsilon.\]
+The result follows \textit{mutatis mutandis}.
 \end{proof}

### 2014-02-05 padicprec.tex
Split Section 3 in two Sections

@@
 
-\section{A new philosophy}
-\label{sec:philosophy}
-
-The aim of this section is to develop a new point of view on $p$-adic 
-precision based on $p$-adic analysis. Actually, we shall generalize a 
-bit the setting and work over a general ultrametric field (see definition
-below).
+\section{The main Lemma}
+\label{sec:mainlemma}
+
+The theory of $p$-adic precision we want to develop is mainly based 
+on a single lemma of $p$-adic analysis. The aim of this Section is
+to state and prove this lemma. Consequences will be discussed after
+in Section \ref{sec:tracking}.
+
+From now, we generalize a bit the setting and work over a general 
+ultrametric complete field (see definition below). This includes 
+$p$-adic fields and Laurent series fields.
 
@@
 
-In this section, we recall usual definitions and properties related to
-ultrametric analysis. We refer to Schneider \cite{Schneider} for a 
-complete exposition.
+We recall usual definitions and properties related to ultrametric 
+analysis. We refer to Schneider \cite{Schneider} for a complete 
+exposition.
 
@@
 
-\subsection{The main Lemma}
+\subsection{Images of lattices under differentiable functions}
 
@@
 
-\subsection{Automatic sharp track of precision}
+\section{Tracking precision using differentials}
+\label{sec:tracking}
 
@@
 important applications to effective computations with $p$-adic numbers 
-or power series, that we are going to discuss now. Let us recall first 
+or power series, that we discuss in this Section. Let us recall first 
 that a datum involving elements of a complete ultrametric field $K$ 
@@
 informations. That is why, instead of this datum itself, we generally 
-work with neighborhoods of it. In what follows, we shall consider 
-neighborhoods of the form $x + H$ where $x$ is some element lying in a 
-$K$-Banach $E$ and $H$ is a lattice in $E$. A very basic example is the 
-approximate $p$-adic number $1 + O(p^N)$ which corresponds to the 
-neighborhood of $1$ in $\Z_p$, which is the set of all $p$-adic numbers 
-congruent to $1$ modulo $p^N$. We nevertheless emphasize that our 
-definition is quite flexible and allows many neighborhoods that we do 
-not consider in general. For instance, an ``admissible'' neighborhood 
-around a monic polynomial $P_0 \in \Q_p[X]$ of degree $d$ may be defined 
-by congruences of the form $P(a_i) \equiv P_0(a_i) \pmod {p^{n_i}}$ ($1 
-\leq i \leq d$) for some fixed elements $a_i \in \Q_p$ and relative 
-integers $n_i$.
+work with neighborhoods of it.
+
+In what follows, we shall consider neighborhoods of the form $x + H$ 
+where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
+in $E$. A very basic example is the approximate $p$-adic number $1 + 
+O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
+the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
+nevertheless emphasize that our definition is quite flexible and allows 
+many neighborhoods that we do not consider in general. For instance, an 
+``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
+of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
+P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
+$a_i \in \Q_p$ and relative integers $n_i$.
 
@@
 
-\medskip
-
-Now, let us consider a function {\tt f} (in the sense of computer 
+\subsection{Automatic sharp track of precision}
+
+We consider a function {\tt f} (in the sense of computer 
 science) that takes as input an (approximate) element lying in an open 
@@
 \emph{etc.}  In this section we extend the formalism 
-developed in Section \ref{sec:philosophy} to a more general setting:
+developed in Section \ref{sec:mainlemma} to a more general setting:
 we consider the quite general case of 
@@
 of $f$ is then also a manifold; this example then fits to the setting
-of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
+of \S \ref{sec:manifold} but not to those of \S \ref{sec:mainlemma}. 
 The generalization made in \S \ref{sec:manifold} is then needed.

### 2014-02-05 padicprec.tex
Paragraph "Lattices and computers" written

@@
 
-\todofor{Xavier}{Explain that if $E$ is finite dimensional and $K$ (and 
-the absolute value is discrete), a lattice can be encoded by a finite 
-amount of informations.}
+Although elements of $K$ are not representable in thein entirety on a 
+computer, lattices in finite dimensional $K$-vector spaces are as soon 
+as the elements of the quotient $B_K(1)/B_K(r)$ are for all $r \in 
+(0,1)$. Indeed, given a lattice $L$ in $K^d$, there exists by definition 
+two positive rational numbers $a$ and $b$ such that 
+\begin{equation}
+\label{eq:incllattice}
+B_K(a)^d \subset L \subset B_K(b)^d.
+\end{equation}
+Moreover, a lattice $L$ satisfying \eqref{eq:incllattice} is uniquely 
+determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
+(B_K(1)/B_K(r))^d$ with $r = \frac a b$. Therefore, the data of $L$ can 
+be encoded by a $(d \times d)$ matrix with coefficients in $B_K(1)/
+B_K(r)$. Thus, if elements in the latter ring are representable on a 
+computer, so are the lattices $L$ for which Eq.~\eqref{eq:incllattice} 
+holds. Finally, an arbitrary lattice $L$ in $K^d$ can be represented
+by a triple $(a,b,M)$ where $a$ and $b$ satisfies \eqref{eq:incllattice}
+and $M$ is a $(d \times d)$ matrix over $B_K(1)/B_K(\frac a b)$ 
+representing the image of $L$ in $B_K(b)^d / B_K(a)^d$.
+
+We furthermore remark that if $K = \Q_p$ (for some prime number $p$), 
+the quotient $B_K(1)/B_K(r)$ is isomorphism to some $\Z/p^n\Z$ (with $n 
+= \lfloor - \log_p r \rfloor$) are therefore is representable on a 
+computer. Thus, the above discussion applies when $K = \Q_p$. Similarly 
+it also applies when $K$ is a finite extension of $\Q_p$ or a field of
+Laurent series with coefficients in a finite field.
 
@@
 \end{proof}
+
+\todo{Explain how Newton polygons considered above behave under
+usual operations (sum, multiplication and composition of functions).}
 
@@
 
-\subsubsection*{Subresultants and B\'ezout coefficients}
+\subsubsection*{Greatest common divisors and B\'ezout coefficients}
 

### 2014-02-05 padicprec.tex
Citation to Fomin-Zelevinsky re-added. I removed it by mistake in commit 108658c05f4091c923e605752c39204d2f3cd503

@@
 u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
-polynomial with coefficients in $\Z$ in these indeterminates.
+polynomial with coefficients in $\Z$ in these indeterminates
+(see \cite{Fomin-Zelevinsky}).
 

### 2014-02-06 padicprec.tex
Differentials of gcd's

@@
 \subsection{Univariate polynomials}
+\label{ssec:polynomials}
 
@@
 
-\todo{To be written.}
+We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
+function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
+couple $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
+greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
+coefficients of minimal degrees (they are those computed by extended 
+Euclide algorithm).
+The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
+subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
+$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
+of positive degree. This implies that $f$ is not continuous outside 
+$\mathcal V_0$. On the contrary, on $\mathcal V_0$, the function $f$ is
+differentiable, and actually locally analytic.
+
+Of course, on $\mathcal V_0$, the first component $D$ is constant and 
+therefore $dD = 0$. To compute $dU$ and $dV$, one can simply 
+differentiate the B\'ezout relation $AU + BV = 1$. We get:
+$$A \cdot dU + B \cdot dV = - (dA \cdot U + dB \cdot V)$$
+from which we deduce that $dU$ (resp. $dV$) is obtained as the 
+remainder in the Euclidean division of $U{\cdot}dX$ by $B$ (resp. of 
+$V{\cdot}dX$ by $A$) where $dX = - (dA \cdot U + dB \cdot V)$.
+
+In order to go further and differentiate $f$ outside $\mathcal V_0$, we 
+define, for all integer $i$, the subset $\mathcal V_i$ of $K_n[X] \times 
+K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
+subresultants shows that $\mathcal V_i$ is locally closed with respect to 
+the Zariski topology. In particular, it defines a $K$-manifold in the 
+sense of Appendix \ref{sec:manifold}. This is interesting because the 
+restriction of $f$ to $\mathcal V_i$ is now differentiable. To compute
+its differential, we proceed along the same lines as before: we 
+differentiate the relation $AU + BV = D$ and obtain this way
+$A \cdot dU + B \cdot dV - dD = dX$
+with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
+first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
+whereas the term $dD$ has degree less than $i$. 
+Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
+we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
+-dR$. These relations, together with bounds on the degree of $U$ and $V$,
+imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
+Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
+by $\frac A D$).
+
+\medskip
+
+The lesson we may retain from this study is the following. If we have to 
+compute the greatest common divisor of two polynomials $A$ and $B$ known 
+with finite precision, we first need to determine what is its degree. 
+However, the degree function is not continuous --- it is only upper 
+semi-continuous --- and hence cannot be determined from sure from $A$ and 
+$B$ (expect if they are coprime). We then need to make an extra 
+hypothesis and the more natural one is arguably to assume that 
+$\gcd(A,B)$ has actually the maximal degree that it can have. Once this
+assumption is made, the computation is possible and one can apply Lemma 
+\ref{lem:main} to determine the precision of the result.
+Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
+itself although there exist pair of coprime polynomials in any 
+neighborhood of $(A,A)$.
 
@@
 = - \frac{dP(\alpha)}{P'(\alpha)}$.
+
+\todo{Discuss a bit multiple roots.}
 
@@
 
-We can develop further this example and study what happens on the 
-closed subset of $M_{n,n-d}(K)$ where matrices have not full rank. On
-this subspace, the left kernel has dimension $< d$ and then no longer
-defines a point in $\Grass(n,d)$. Nevertheless, for all integer $r <
-n-d$, we can consider the subset $\mathcal V_r \subset M_{n,n-d}(K)$
-of matrices whose rank are exactly rank $r$. It is locally closed in
-$M_{n,n-d}(K)$ with respect to the Zariski topology and hence defines
-a $K$-manifold. Furthermore, we have a mapping $\text{LK}_r : \mathcal 
-M_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
-can be computed as before.
-
-The lesson from this study is the following. If we have to compute the 
-kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
-priori} we first need to determine what is the rank of $M$. However, the 
-rank function is not continuous --- it is only lower semi-continuous --- 
-and hence cannot be determined from sure from $M$ (expect if $M$ has 
-full rank). We then need to make an extra hypothesis and the more natural 
-one is arguably to assume that $M$ has actually the minimal rank that it 
-can have. With this convention, for instance, the kernel of the zero 
-matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
-whole space although there exist invertible matrices in any 
-neighborhood of $M$.
+Following what we have already done in \S \ref{ssec:polynomials} for 
+greatest common divisors of polynomials, we can develop further this 
+example and study what happens on the closed subset of $M_{n,n-d}(K)$ 
+where matrices have not full rank. On this subspace, the left kernel has 
+dimension $< d$ and then no longer defines a point in $\Grass(n,d)$. 
+Nevertheless, for all integer $r < n-d$, we can consider the subset 
+$\mathcal V_r \subset M_{n,n-d}(K)$ of matrices whose rank are exactly 
+rank $r$. It is locally closed in $M_{n,n-d}(K)$ with respect to the 
+Zariski topology and hence defines a $K$-manifold. Furthermore, we have a 
+mapping $\text{LK}_r : \mathcal M_r \to \Grass(n,n-r)$ which is 
+differentiable and whose differential can be computed as before.
 

### 2014-02-06 padicprec.tex
Precision on a multiple root of a polynomial.

@@
 
-Suppose that we are given a monic polynomial $P \in K[X]$ written as a 
+Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
 product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
 Hensel's lemma implies that there exists a small neighborhood $\mathcal 
-U$ of $P_0$ such that any monic polynomial $P \in \mathcal U$ factors 
+U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
 uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
@@
 
+\subsubsection*{Finding roots}
+
 An important special case of the previous study occurs when $A_0$ has 
-degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
-application $P \mapsto A$ is then nothing but the mapping that follows 
-the simple root $\alpha$. Of course, its differential around $P_0$ can 
-be computed by the above method. Nevertheless there is a more direct way 
-to obtain it in this particular case: we write $(P + dP)(\alpha + 
-d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
-dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
-is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
-= - \frac{dP(\alpha)}{P'(\alpha)}$.
-
-\todo{Discuss a bit multiple roots.}
+degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
+The application $P \mapsto A$ is then nothing but the mapping that 
+follows the simple root $\alpha_0$. Of course, its differential around 
+$P_0$ can be computed by the above method. Nevertheless there is a more 
+direct way to obtain it in this particular case: we write $(P_0 + 
+dP)(\alpha_0 + d\alpha) = 0$ and expand this. We obtain this way 
+$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$ where $P'$ denotes the 
+derivative of $P$. Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
+does not vanish and we find:
+$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$
+
+We now address the case of a multiple roots. Let $P_0$ be a monic 
+polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
+multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
+to follow to root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
+restrict ourselves to polynomials which have a root of multiplicity $m$ 
+close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
+V_m$ consisting of monic polynomials of degree $d$ having a root of 
+multiplicity at least $m$. It is a Zariski close subset of $K_d[X]$ and 
+it contains by assumption the polynomial $P_0$ we started with. 
+Moreover, the irreducible components of $\mathcal V_m$ that meet at 
+$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
+$\geq m$. In particular $\alpha_0$ correspond to one of these 
+irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
+The algebraic variety $\mathcal V_{m,\alpha_0}$ is a \emph{a fortioti} 
+$K$-manifold. Moreover there exists a differentiable function $f$ which 
+is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
+follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
+$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
+multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
+Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
+B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
+alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
+(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
+$P^{(m-1})(\alpha) = 0$ where $P^{(i})$ denotes the $i$-th derivative
+of $P$. We find:
+\begin{equation}
+\label{eq:dalphamult}
+d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
+\end{equation}
+
+The above computation has actually interesting consequences on $p$-adic 
+precision. Let us illustrate this with an exemple. Consider the monic 
+polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
+large integer) and assume that we are asking a softwate to compute one 
+of its roots. What is the right answer? We remark that, if $\alpha$ is 
+any $p$-adic number divisible by $p^n$, then $P(X)$ can be $X^2 - 
+\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
+the two roots of $P$ are necessarily divisible by $p^n$. Hence, the
+right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
+that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
+\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
+O(p^{2N})$ (or $O(p^{2N-1})$ if $p=2$). 
+This is exactly the result we get by applying Lemma \ref{lem:main} and
+using Eq.~\eqref{eq:dalphamult} to simplify the differential.
 

### 2014-02-06 padicprec.tex
Typos corrected and one remark added

@@
 B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
-alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
+\alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
 (along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
-$P^{(m-1})(\alpha) = 0$ where $P^{(i})$ denotes the $i$-th derivative
+$P^{(m-1)}(\alpha) = 0$ where $P^{(i)}$ denotes the $i$-th derivative
 of $P$. We find:
@@
 using Eq.~\eqref{eq:dalphamult} to simplify the differential.
+
+This phenomenon is general: if $P$ is a polynomial over $\Q_p$ known at 
+some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
+multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
+(\alpha)$ vanish at the given precision) can be computed at precision 
+roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
+indeed multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
+where $c$ is some constant.
 

### 2014-02-07 padicprec.tex
Differentials of evaluation and interpolation

@@
 
+\subsubsection*{Evaluation and interpolation}
+
+Beyond sums and products (which can be treated as before), two basic 
+operations involving polynomials are evaluation and interpolation.
+Evaluation is modelised by the function $(P,x) \mapsto P(x)$ where
+$P$ is some polynomial and $x$ some element in $K$. Differentiating
+it can be done by computing
+\begin{equation}
+\label{eq:diffeval}
+(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x)
+\end{equation}
+by neglecting terms of order $2$. Here $P'$ denotes the derivative of 
+$P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
+P'(x) dx + dP(x)$.
+
+As for interpolation, we consider a positive integer $d$ and the partial 
+function $f : K^{2d} \to K_{< d}[X]$ which maps the tuple $(x_1, y_1, 
+\ldots, x_d, y_d)$ to the polynomial $P$ of degree less than $d$ such 
+that $P(x_i) = y_i$ for all $i$. The polynomial $P$ exists and is unique 
+as soon as the $x_i$'s are pairwise distinct; the above function $f$ is 
+then defined on this open set. Furthermore, if $(x_1, y_1, \ldots, x_d, 
+y_d)$ is a point in it and $P$ denotes the corresponding interpolation 
+polynomial, Eq.~\eqref{eq:diffeval} shows that $d y_i = P'(x_i) dx_i + 
+dP(x_i)$ for all $i$. For this, we can compute $dP(x_i)$ from $d x_i$ 
+and $d y_i$ and finally recover $dP$ by performing a new interpolation.
+
 \subsubsection*{Euclidean division}
 
-Beyond sums and products (which can be treated as before), a basic 
-operation involving polynomials is Euclidean division. Let us compute 
-the differential of it. We start with two polynomials $A$ and $B$ with 
-$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
-with $\deg R < \deg B$. Differentiating the above equality we find: 
+Let $A$ and $B$ be two polynomials $B \neq 0$. The Euclidean division of 
+$A$ by $B$ is written $A = BQ + R$ with $\deg R < \deg B$. 
+Differentiating the above equality we find:
 $dA - dB \cdot Q = B \cdot dQ + dR$,

### 2014-02-07 padicprec.tex
Added a short paragraph about comparison with floating point arithmetics.

@@
 similarity with our approach is noticeably strong.
+
+\subsubsection*{Quick comparison with floating point arithmetics}
+
+It worths remarking that the two strategy designed above share many 
+similarities with usual floating point arithmetics over the reals. 
+Indeed, roughly speaking, in each setting, we begin by choosing a large 
+precision, we do all our computations up to this precision understood 
+that when we are not sure about some digit, we choose it ``at random'' 
+or using good heuristics. The main difference is that, in the 
+ultrametric setting, we are able (under some mild hypothesis) to 
+quantify the precision we need at each individual step in order to be 
+sure that the final result is correct up to the required precision.
 
@@
 indeed multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
-where $c$ is some constant.
+where $c$ is some constant (depending on $P$ and $\alpha$ but not on
+$N$).
 

### 2014-02-07 padicprec.tex
All authors are restore!
typo

@@
 \title{$p$-adic precision}
-\author{Xavier Caruso}
-\author{David Roe}
-\author{Tristan Vaccon}
+\author{Xavier Caruso, David Roe \& Tristan Vaccon}
 \date\today
@@
 of its roots. What is the right answer? We remark that, if $\alpha$ is 
-any $p$-adic number divisible by $p^n$, then $P(X)$ can be $X^2 - 
+any $p$-adic number divisible by $p^N$, then $P(X)$ can be $X^2 - 
 \alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
-the two roots of $P$ are necessarily divisible by $p^n$. Hence, the
+the two roots of $P$ are necessarily divisible by $p^N$. Hence, the
 right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
@@
 \alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
-O(p^{2N})$ (or $O(p^{2N-1})$ if $p=2$). 
+O(p^{2N})$ if $p > 2$ and $\alpha = O(p^{2N-1})$ if $p=2$.
 This is exactly the result we get by applying Lemma \ref{lem:main} and

### 2014-02-07 padicprec.tex

@@
 
-If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L_{k+1} \interleave  t^k$, then it is enough that :
-\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \log_p l \right) \geq - \log_p \varepsilon.\]
+If we denote by $NP_F$ the Newton polygon of the power series 
+\[
+F(t) = \sum_{k\geq 0}^{+\infty} \interleave L_{k+1} \interleave  t^k
+\]
+then it is enough that:
+\[
+\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \log_p l \right) \geq - \log_p \varepsilon.
+\]
 
@@
 uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
-and $B_0$ respectively. Thus, we can consider the application $f : P 
+and $B_0$ respectively. Thus, we can consider the map $f : P 
 \mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
@@
 degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
-The application $P \mapsto A$ is then nothing but the mapping that 
+The map $P \mapsto A$ is then nothing but the mapping that 
 follows the simple root $\alpha_0$. Of course, its differential around 
@@
 $U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
-U$, the map $\text{LK}$ corresponds in our chart to the application that 
+U$, the map $\text{LK}$ corresponds in our chart to the map that 
 takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 

### 2014-02-10 padicprec.tex
Improvement of the locally analytic part

--- 2014-02-07 padicprec.tex
@@
 
-\begin{prop}\label{prop:locanalytic}
+\begin{deftn}\label{deftn:analyticNP}
 Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ be locally analytic map $E \rightarrow F$.
 Let $x \in E$, and let $f(x+h)=f(x)+\sum_{k \geq 0}^{+\infty} L_k \cdot (h,\dots,h)$ be the power series development of $f$ at $x$.
-We assume $L_1$ is surjective.
+Let $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L^{k+1}(x) \interleave  t^k \in \mathbb{R}[[t]].$
+
+We define $NP(f,x)$ to be the function defined by its Newton polygon, the lower convex hull of the $(k,\log ( \interleave L^{k+1}(x) \interleave)),$ with $k \geq 0.$
+\end{deftn}
+
+We now prove that the knowledge of $NP(f,x)$ is enough to recover $\delta$.
+
+\begin{prop}\label{prop:locanalytic}
+Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}. We assume $L_1$ is surjective. 
 Let $C$ be given by the Banach-Schauder theorem for $L_1$, $\varepsilon = \frac{1}{C}$. 
-Let $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L^{k+1}(0) \interleave  t^k \in \mathbb{R}[[t]]$, let $NP_F$ be its Newton polygon.
-Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP_F$.
+Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP(f,x)$.
 Then $r$ such that $\mu=\log_p r$ is enough for lemma \ref{lem:main}.

### 2014-02-12 padicprec.tex
typos

--- 2014-02-10 padicprec.tex
@@
 
-\begin{deftn}\label{deftn:analyticNP}
+\begin{prop}\label{prop:locanalytic}
 Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ be locally analytic map $E \rightarrow F$.
 Let $x \in E$, and let $f(x+h)=f(x)+\sum_{k \geq 0}^{+\infty} L_k \cdot (h,\dots,h)$ be the power series development of $f$ at $x$.
-Let $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L^{k+1}(x) \interleave  t^k \in \mathbb{R}[[t]].$
-
-We define $NP(f,x)$ to be the function defined by its Newton polygon, the lower convex hull of the $(k,\log ( \interleave L^{k+1}(x) \interleave)),$ with $k \geq 0.$
-\end{deftn}
-
-We now prove that the knowledge of $NP(f,x)$ is enough to recover $\delta$.
-
-\begin{prop}\label{prop:locanalytic}
-Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}. We assume $L_1$ is surjective. 
+We assume $L_1$ is surjective.
 Let $C$ be given by the Banach-Schauder theorem for $L_1$, $\varepsilon = \frac{1}{C}$. 
-Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP(f,x)$.
+Let $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L^{k+1}(0) \interleave  t^k \in \mathbb{R}[[t]]$, let $NP_F$ be its Newton polygon.
+Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP_F$.
 Then $r$ such that $\mu=\log_p r$ is enough for lemma \ref{lem:main}.
@@
 \ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
-application at $(f_1, \ldots, f_s)$ is continuous provided that:
+application is continuous at $(f_1, \ldots, f_s)$ provided that:
 \begin{itemize}
@@
 since they are defined as determinants. More precisely, let us 
-consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
-the latter ring denotes the finite dimensional vector space of 
-polynomials over $K$ of degree at most $n$) taking a matrix $M$
-to its characteristic polynomial $\det(X-M)$.
+consider the function $\chi : M_n(K) \to K_n[X]$ taking a matrix 
+$M$ to its monic characteristic polynomial $\det(X-M)$.
 Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 

### 2014-02-12 padicprec.tex
Working on Section 2.1

--- 2014-02-12 padicprec.tex
@@
 By lumping elements together into these balls of radius $p^{-n}$, we may
-model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In Sage
-this method for approximating elements is known as ``fixed modulus''.
+model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
+all $p$-adic elements in have constant \emph{absolute precision} $n$.
 
 The second source draws upon parallels between $\Qp$ and $\R$.  Both
-occur as completions of $\Q$, and both have dense subrings:
-the localizations $\Z[\frac1p] \subset \Qp$ and $\Z[\frac12] \subset \R$.
+occur as completions of $\Q$ and we represent elements in both in terms
+of a set of distinguished rational numbers.
 In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
-subset $S \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
+subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
 \[
-\frac{x \circledast y}{x \ast y} - 1 \le 2^{-h}
+\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert \le 2^{-h}
 \]
-for all $x, y \in S$ and $\ast \in \{+, \cdot\}$.  Interval arithmetic in the reals
-is unwieldy, since arithmetic operations always increase the lengths of the inputs.
+for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
+defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
+\[
+\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert_p \le p^{-h}.
+\]
+When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.
+
+Neither of these models allows a user to track the precision of an individual element.
+Over the reals, interval arithmetic is unwieldy, since arithmetic operations always increase the lengths of the inputs.
 As a consequence, most computations in the real numbers rely on
@@
 
-In the $p$-adic world, these features are reversed.
-
-\todofor{David}{Add a general introduction to this section.}
-
-\subsection{Separate precision and approximation}
+In the $p$-adic world, precision tracking using intervals is much more feasible.
+Even a long sequence of operations may not sacrifice any precision.
+The Gaussian distributions of measurement error over $\R$ have no direct analogue in
+$\Qp$.  Instead, number theorists focus on provably determining
+a result within a $p$-adic interval.  As a consequence, interval arithmetic is ubiquitous
+in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.
+
+\subsection{Separating precision from approximation}
 

### 2014-02-12 padicprec.tex

@@
 \ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
-map at $(f_1, \ldots, f_s)$ is continuous provided that:
+map is continuous at $(f_1, \ldots, f_s)$ provided that:
 \begin{itemize}

### 2014-02-13 padicprec.tex
I introduced the notion of first order lattice.

@@
 \label{eq:main}
-f(v_0 + H) = f(v_0) + f'(v_0) \cdot H.
+f(v_0 + H) = f(v_0) + f'(v_0) (H).
 \end{equation}
@@
 \begin{rmk}
+\label{rmk:mainlemma}
 Given a lattice $H_0$ in $E$, the equality \eqref{eq:main} holds for $H 
@@
 B_E(r).$
-Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
-
-We first prove that $f(H) \subset f'(0) \cdot H$.
+Let us show that $f$ maps surjectively $H$ onto $f'(0) (H)$.
+
+We first prove that $f(H) \subset f'(0) (H)$.
 Let $x \in H$. By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
 \varepsilon \Vert x \Vert $. Therefore, if $y=f(x)-f'(0) \cdot x,$ then $\Vert y \Vert \leq \varepsilon r$. As a consequence of the open mapping Theorem, $B_F(\varepsilon r) \subset 
-\varepsilon C  f'(0) \cdot B_E(r),$ and there exists $x_1 \in \varepsilon C  \cdot B_E(r)$ such that $f'(0) \cdot x_1 =y.$ Since $\varepsilon C <1,$  $x_1 \in B^-_E(r) \subset H.$ and finally, $f(x)= f'(0) \cdot (x-x_1) \in f'(0) \cdot H$.
-
-We shall now prove surjectivity. Let $y \in f'(0) \cdot H$. 
+\varepsilon C  f'(0) \cdot B_E(r),$ and there exists $x_1 \in \varepsilon C  \cdot B_E(r)$ such that $f'(0) \cdot x_1 =y.$ Since $\varepsilon C <1,$  $x_1 \in B^-_E(r) \subset H.$ and finally, $f(x)= f'(0) \cdot (x-x_1) \in f'(0) (H)$.
+
+We shall now prove surjectivity. Let $y \in f'(0) (H)$.
 Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
-Since $f(x_0) \in f'(0) \cdot H,$ we can define $z_0 \in E$ to be such that $f'(0) \cdot z_0 = y - f(x_0)$ and $\Vert z_0 \Vert \leq C \cdot \Vert y - f(x_0) \Vert$.
+Since $f(x_0) \in f'(0) (H)$, we can define $z_0 \in E$ to be such that $f'(0) \cdot z_0 = y - f(x_0)$ and $\Vert z_0 \Vert \leq C \cdot \Vert y - f(x_0) \Vert$.
 We then define by induction two sequences $(x_n)$ and $(z_n)$ by
@@
 
-\todofor{Tristan}{Improve the end of this Subsection}
+We introduce the following definition:
+
+\begin{deftn}
+\label{def:firstorder}
+Let $E$ and $F$ be two $K$-Banach and $f : U \rightarrow F$ be a 
+function defined on an open subset $U$ of $E$. Let also $v_0$ be a 
+point in $U$.
+A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
+$v_0$ if
+$$f(v_0 + H) = f(v_0) + f'(v_0) (H).$$
+\end{deftn}
+
+By Lemma \ref{lem:main} and Remark \ref{rmk:mainlemma}, if $f$ is 
+differentiable at $v_0$ and $f'(v_0)$ is surjective, then for all 
+lattice $H$ in $E$, there exists a constant $\delta$ such that $aH$ 
+is a first order lattice for $f$ at $v_0$ for all $a \in B_E(r)$.
+
+\subsection{The case of locally analytic functions}
+
+\todofor{Tristan}{Improve this Subsection}
 
@@
 usual operations (sum, multiplication and composition of functions).}
+
+Using Lemma \ref{lem:reductionball}, one can extend Proposition 
+\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
+a lattice in some $K$-Banach space $E$, we first define the two
+following quantities:
+\begin{itemize}
+\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
+\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
+\end{itemize}
+
+\begin{cor}
+Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
+respectively and $f : U \to V$ be a locally analytic function. Let 
+also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
+Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
+soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
+constant depending only on $f$ and $v_0$ which can be determined 
+explicitely from the Newton polygon of $f$ at $v_0$.
+\end{cor}
+
+\begin{proof}
+We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
+\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
+\Vert)$ is induced by identity map.
+\end{proof}
 
@@
 is differentiable at all points in its domain. We recall that
-the mappings $\varphi_i$ above are the so-called \emph{charts} and that
-their collection defines an \emph{atlas} of $V$. In the sequel, we shall
-assume further that the open covering $V = \bigcup_{i \in I} V_i$ is
-locally finite, which means that every point $x \in V$ lies only in a
+the mappings $\varphi_i$ above are the so-called \emph{charts}. The
+$\psi_{ij}$'s are the transition maps. The collection of $\varphi_i$'s
+and $\psi_{ij}$'s is called an \emph{atlas} of $V$. In the sequel, we 
+shall assume further that the open covering $V = \bigcup_{i \in I} V_i$ 
+is locally finite, which means that every point $x \in V$ lies only in a 
 finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
@@
 \subsection{Precision data}
-
-\todofor{David/Xavier}{Complete this Subsection}
 
@@
 A \emph{precision datum} at $x$ is a lattice in the tangent space 
-$T_x V$.
+$T_x V$ such that for all indices $i$ and $j$ with $x \in U_i \cap
+U_j$, the image of $T_x V$ in $E_i$ is a first order lattice for
+$\psi_{ij}$ at $\varphi_i(x)$ (\emph{cf} Definition \ref{def:firstorder}.
 \end{deftn}
 
-\begin{deftn}
-Let $x \in V$ and $H$ be a precision datum at $x$.
-For all $i \in I$ such that $x \in V_i$, we set:
-$$x + O_i(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
-\subset V.$$
-\end{deftn}
-
-We emphasize that if $x$ belongs to $V_i$ and $V_j$ at the same time,
-the subsets $x + O_i(H)$ and $x + O_j(H)$ differ in general. It is 
-however not the case if $H$ is small enough.
+\begin{rmk}
+The definition of a precision datum at $x$ depends not only on $x$
+and the manifold $V$ where it lies but also on the chosen atlas that
+defines $V$.
+\end{rmk}
 
@@
 \label{prop:independence}
-Let $x \in V$ and $H$ be a precision data at $x$.
-If $H$ is ``small enough'' \todo{(precise!)}, the subset $x + 
-O_i(H)$ does not depend on $i$.
+Let $x \in V$ and $H$ be a precision datum at $x$.
+The subset:
+$$\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
+\subset V$$
+does not depend on the index $i$ such that $x \in V_i$.
 \end{prop}
@@
 Let $i$ and $j$ be two indices such that $x$ belongs to $V_i$ and $V_j$. 
-Set $x_i = \varphi_i(x) \in E_i$ and $H_i = \varphi'_i(x)(H)$. Going 
-back to the definitions, we see that the equality $x + O_i(H) = x + 
-O_j(H)$ is equivalent to:
-$$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$$ 
-where we recall that $\psi_{ij}$ is defined by Eq.~\eqref{eq:psiij}.
-\todo{Complete the proof and write the correct condition on $H$!}
+Set $x_i = \varphi_i(x) \in E_i$ and $H_i = \varphi'_i(x)(H)$. The 
+equality 
+$$\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
+\subset V = \varphi_j^{-1}\big(\varphi_j(x) + \varphi_j'(x)(H)\big) 
+\subset V$$
+is clearly equivalent to
+$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$.
+and the latter holds because $H_i$ is a first order lattice
+of $\psi_{ij}$ at the point $x_i$ by definition.
 \end{proof}
 
-Under the assumption of Proposition \ref{prop:independence}, we define
-$x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
-$x \in V_i$.
+We are now in position to define $x + O(H)$.
+
+\begin{deftn}
+Let $x \in V$ and $H$ be a precision datum at $x$. We set:
+$$x + O(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big)
+\subset V$$
+for some (equivalenty, all) $i$ such that $x \in V_i$.
+\end{deftn}
 
@@
 computable subset, we need to compare $x_0 + O(H_0)$ with varying $x + 
-O(H)$ when $x$ and $x_0$ are close enough.
-Let us first examine the situation in a fixed given chart: we fix some 
-index $i \in I$ and pick two elements $x_0$ and $x$ in $V_i$. We consider 
-in addition a precision datum $H_0$ at $x_0$ and we want to produce a 
-precision datum $H$ at $x$ such that $x_0 + O_i(H_0) = x + O_i(H)$. 
-We remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
-isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and 
-$\varphi'_i(x)$ respectively. A natural candidate for $H$ is then:
+O(H)$ when $x$ and $x_0$ are close enough. Let us first examine the 
+situation in a fixed given chart: we fix some index $i \in I$ and pick 
+two elements $x_0$ and $x$ in $V_i$. We consider in addition a lattice 
+$\tilde H_0$ in $E_i$ --- which should be think as $\varphi'_i(x_0) 
+(H_0)$ --- and we want to produce a lattice $\tilde H$ such that 
+$\varphi_i(x_0) + \tilde H_0 = \varphi_i(x) + \tilde H$. Of course 
+$\tilde H = \tilde H_0$ does the job as soon as 
+$\varphi_i(x) - \varphi_i(x_0) \in \tilde H_0$. 
+Now, we remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
+isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and $\varphi'_i(x)$ 
+respectively. A natural candidate for $H$ is then:
 \begin{equation}
@@
 \end{equation}
-And indeed, one can check that $x + O(H)$ is indeed equal to $x_0 + 
-O(H_0)$ as soon as $x$ and $x_0$ are close enough in the following sense: 
-the difference $\varphi_i(x) - \varphi_i(x_0)$ lies in the lattice $H_i = 
-\varphi'_i(x_0)(H_0)$. We furthermore have a property of independence on 
-$i$.
+With this choice, $x + O(H) = x_0 + O(H_0)$ as soon as $x$ and $x_0$ are 
+close enough in the following sense: the difference $\varphi_i(x) - 
+\varphi_i(x_0)$ lies in the lattice $\varphi'_i(x_0)(H_0)$. We 
+furthermore have a property of independence on $i$.
 
 \begin{prop}
-We assume that $H_0$ is ``small enough'' \todo{(precise!)} and that $x$ is close 
-enough to $x_0$. Then
+Let $x_0 \in V$ and $H_0$ be a precision datum at $x_0$.
+Then, for all $x$ sufficiently close to $x_0$,
 \begin{enumerate}[(i)]
 \item the lattice $H$ defined by \eqref{eq:Hprime} does not depend 
-on $i$,
-\item the pairs $(x,H)$ and $(x_0, H_0)$ satisfy the assumptions of
-Proposition \ref{prop:independence} and we have $x + O(H) = x_0 + O(H_0)$.
+on $i$ and is a precision datum at $x$, and
+\item we have $x + O(H) = x_0 + O(H_0)$.
 \end{enumerate}
@@
 Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse inclusion can be proved 
-in the same way.
-
-We now prove (ii). \todo{Write the proof.}
+in the same way. The fact that $H$ is a precision datum at $x$ is easy
+and left to the reader.
+Finally, if $x$ is close enough to $x_0$, we can check (ii) in the 
+charts but this was already done.
 \end{proof}
@@
 
-\todofor{David/Xavier}{Extend Lemma \ref{lem:main} to manifolds.}
+\todofor{Xavier}{Extend Lemma \ref{lem:main} to manifolds.}
+
+\subsection{Examples}
+
+\todofor{David}{Elliptic curves and grassmannians}
 

### 2014-02-13 padicprec.tex
Working on introduction

--- 2014-02-12 padicprec.tex
@@
 \newcommand{\Qp}{\Q_p}
+\newcommand{\Fp}{\mathbb{F}_p}
 \newcommand{\R}{\mathbb R}
@@
 
-During the last two decades, $p$-adic methods have become a tremendous tool in computer algebra, since it allows computation not feasible on finite fields, can control the growth of the size of the coefficients when dealing with rationnal numbers, or is needed when performing algorithm of $p$-adic nature.
-
-Yet, $p$-adic algorithms are necessarily numeric or symbolic-numeric: one can not avoid working with finite precision, and thus, keeping track of loss in precision during the computation.
-
-Here are some examples of algorithm that rely on $p$-adic computations:
+The last two decades have seen a rise in the popularity of $p$-adic methods in
+computational algebra.  For example,
 \begin{itemize}
-\item Bostan \cite{Bostan}, on how polynomials in $\mathbb{F}_p$ can be computed from theirs Newton series lifted in $\mathbb{Z}_p$;
-\item Gaudry \cite{Gaudry}, on generation of genus $2$ curves via $p$-adic lifting;
-\item Kedlaya \cite{Kedlaya}, on point counting on hyperelliptic curves using Monsky--Washnitzer cohomology;
-\item Lercier \cite{Lercier}, on computing isogenies between elliptic curves using $p$-adic differential equations.
+\item Bostan et al. \cite{Bostan} use Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
+\item Gaudry et al. \cite{Gaudry} use $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
+\item Kedlaya \cite{Kedlaya} uses $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
+\item Lercier and Survent \cite{Lercier} compute isogenies between elliptic curves using $p$-adic differential equations.
 \end{itemize}
+Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
+stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
+that apply across many different algorithms.
+
+Two sources of inspiration arise when studying $p$-adic algorithms.
+The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
+$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
+$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
+homomorphism, given unknown elements in two such balls we can
+locate the balls in which their sum and product lie.  Working on a computer
+we must find a way to write elements using only a finite amount of data.
+By lumping elements together into these balls of radius $p^{-n}$, we may
+model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
+all $p$-adic elements in have constant \emph{absolute precision} $n$.
+
+The second source draws upon parallels between $\Qp$ and $\R$.  Both
+occur as completions of $\Q$ and we represent elements of both in terms
+of a set of distinguished rational numbers.
+In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
+subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
+\[
+\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert \le 2^{-h}
+\]
+for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
+defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
+\[
+\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert_p \le p^{-h}.
+\]
+When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.
+
+Neither of these models allows a user to track the precision of an individual element.
+Over the reals, interval arithmetic is unwieldy, since arithmetic operations always increase the lengths of the inputs.
+As a consequence, most computations in the real numbers rely on
+statistical cancelation and external estimates of precision loss, rather than
+attempting to track known precision at each step.  This tendency is strengthened
+by the ubiquity of floating point arithmetic in scientific applications, where Gaussian
+distributions are more common than intervals anyway.
+
+In the $p$-adic world, precision tracking using intervals is much more feasible.
+Even a long sequence of operations may not sacrifice any precision.  Intervals
+allow number theorists to provably determine a result modulo a given power of $p$,
+and the Gaussian distributions of measurement error over $\R$ have no direct analogue
+over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
+in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.
 
@@
 \section{Motivations and aims}
-
-Two sources of inspiration arise when studying $p$-adic algorithms.
-The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
-$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
-$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
-homomorphism, given unknown elements in two such balls we can
-locate the balls in which their sum and product lie.  Working on a computer
-we must find a way to write elements using only a finite amount of data.
-By lumping elements together into these balls of radius $p^{-n}$, we may
-model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
-all $p$-adic elements in have constant \emph{absolute precision} $n$.
-
-The second source draws upon parallels between $\Qp$ and $\R$.  Both
-occur as completions of $\Q$ and we represent elements in both in terms
-of a set of distinguished rational numbers.
-In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
-subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
-\[
-\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert \le 2^{-h}
-\]
-for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
-defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
-\[
-\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert_p \le p^{-h}.
-\]
-When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.
-
-Neither of these models allows a user to track the precision of an individual element.
-Over the reals, interval arithmetic is unwieldy, since arithmetic operations always increase the lengths of the inputs.
-As a consequence, most computations in the real numbers rely on
-statistical cancelation and external estimates of precision loss, rather than
-attempting to track known precision at each step.  This tendency is strengthened
-by the ubiquity of floating point arithmetic in scientific applications, where Gaussian
-distributions are more common than intervals anyway.
-
-In the $p$-adic world, precision tracking using intervals is much more feasible.
-Even a long sequence of operations may not sacrifice any precision.
-The Gaussian distributions of measurement error over $\R$ have no direct analogue in
-$\Qp$.  Instead, number theorists focus on provably determining
-a result within a $p$-adic interval.  As a consequence, interval arithmetic is ubiquitous
-in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.
 

### 2014-02-13 padicprec.tex
Generalization of main Lemma to manifolds

--- 2014-02-13 padicprec.tex
@@
 
-We introduce the following definition:
-
 \begin{deftn}
@@
 \subsection{The case of locally analytic functions}
+\label{ssec:locanalytic}
 
@@
 
-Let us assume now that $f$ is locally analytic on $U$ and that $x$ and 
-$H$ are chosen so that Lemma \ref{lem:main} applies. Then $f(x+H) = f(x) 
-+ f'(x)(H)$ and the smallest possible $H'$ is $f'(x)(H)$. In other 
-words, the function {\tt f} generates artificial loss of precision if 
-the precision $H'$ it outputs is strictly larger that $f'(x)(H)$.
-The aim of this section is to explain how, under the above hypothesis, 
-one can implement the function {\tt f} so that it always outputs the
-optimal precision $O(H')$ with $H' = f'(x)(H)$.
+Let us assume now that $f$ is locally analytic on $U$ and $H$ is a first 
+order lattice for $f$ at $x$. Then $f(x+H) = f(x) + f'(x)(H)$ and the 
+smallest possible $H'$ is $f'(x)(H)$. In other words, the function {\tt 
+f} generates artificial loss of precision if the precision $H'$ it 
+outputs is strictly larger that $f'(x)(H)$. The aim of this section is 
+to explain how, under the above hypothesis, one can implement the 
+function {\tt f} so that it always outputs the optimal precision $O(H')$ 
+with $H' = f'(x)(H)$.
 
@@
 be sure that Lemma \ref{lem:main} applies, we need \emph{a priori} to 
-compute the exact values of all $x_i$'s, which is of course not possible! 
-Assuming that $g'_i(x)$ is surjective for all $i$, we can fix it as 
-follows. For each $i$, we fix a lattice $\tilde H_i \subset H$ such that 
-Lemma \ref{lem:main} applies to $g_i$, $x$ and $\tilde H_i$. Under our 
-assumption, such lattices always exist and can be computed dynamically 
-using Proposition \ref{prop:locanalytic}. Now, the equality $g_i(x + 
-\tilde H_i) = x_i + g'_i(x)(\tilde H_i)$ means that any perturbation of 
-$x_i$ by an element in $g'_i(x) (\tilde H_i)$ is induced by a 
-perturbation of $x$ by an element in $\tilde H_i \subset H$. Hence, we 
-can freely compute $x_i$ modulo $g'_i(x) (\tilde H_i)$ without changing 
-the final result. Since $g'_i(x)(\tilde H_i)$ is a lattice in $E_i$, this 
-remark makes possible the computation of $x_i$.
+compute the exact values of all $x_i$'s, which is of course not 
+possible! Assuming that $g'_i(x)$ is surjective for all $i$, we can fix 
+it as follows. For each $i$, we fix a first order lattice $\tilde H_i$ 
+for $g_i$ at $x$. Under our assumption, such lattices always exist and 
+can be computed dynamically using Proposition \ref{prop:locanalytic}. 
+Now, the equality $g_i(x + \tilde H_i) = x_i + g'_i(x)(\tilde H_i)$ 
+means that any perturbation of $x_i$ by an element in $g'_i(x) (\tilde 
+H_i)$ is induced by a perturbation of $x$ by an element in $\tilde H_i 
+\subset H$. Hence, we can freely compute $x_i$ modulo $g'_i(x) (\tilde 
+H_i)$ without changing the final result. Since $g'_i(x)(\tilde H_i)$ is 
+a lattice in $E_i$, this remark makes possible the computation of $x_i$.
 
@@
 \item \label{item:Hi2}
-Lemma \ref{lem:main} applies with the function $h_i$, the point
-$x_i$ and the lattice $H'_i$.
+$H'_i$ is a first order lattice for $h_i$ at $x_i$.
 \end{enumerate}
-Applying Lemma \ref{lem:main}, we then get:
-$$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H).$$
+By definition, we have
+$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H)$.
 Therefore, if after the $i$-th step of the execution of the function
@@
 
-\todofor{Xavier}{Extend Lemma \ref{lem:main} to manifolds.}
+With above definitions, Lemma \ref{lem:main} extends readily to 
+manifolds.
+
+\begin{lem}
+Let $V$ and $W$ be two $K$-manifolds. 
+Suppose that we are given a differentiable function $f : V \to W$, 
+together with a point $x \in V$ such that $f'(x) : T_x V \to T_{f(x)} W$ 
+is surjective. Let $H$ be a precision datum at $x$. Then, there exists
+a positive real number $\delta$ such that, for all $\lambda \in 
+B_K(\delta)$, the equality
+$$f(x + \lambda H) = f(x) + \lambda \cdot f'(x)(H)$$
+holds.
+\end{lem}
+
+\begin{proof}
+Apply Lemma \ref{lem:main} in charts.
+\end{proof}
+
+\begin{rmk}
+If $V$ and $W$ are locally analytic $K$-manifolds and the mapping $f$ is 
+locally analytic as well, then one can explicitely determine, using the 
+method of \S \ref{ssec:locanalytic}, a value $\delta$ for which the 
+above Lemma works.
+\end{rmk}
 

### 2014-02-14 padicprec.tex
Working on introduction

--- 2014-02-13 padicprec.tex
@@
 \begin{itemize}
-\item Bostan et al. \cite{Bostan} use Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
-\item Gaudry et al. \cite{Gaudry} use $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
-\item Kedlaya \cite{Kedlaya} uses $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
-\item Lercier and Survent \cite{Lercier} compute isogenies between elliptic curves using $p$-adic differential equations.
+\item Bostan et al. \cite{Bostan} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
+\item Gaudry et al. \cite{Gaudry} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
+\item Kedlaya \cite{Kedlaya} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
+\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves using $p$-adic differential equations.
 \end{itemize}
@@
 In the $p$-adic world, precision tracking using intervals is much more feasible.
-Even a long sequence of operations may not sacrifice any precision.  Intervals
+Even a long sequence of operations with such elements may not sacrifice any precision.  Intervals
 allow number theorists to provably determine a result modulo a given power of $p$,
@@
 over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
-in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.
-
-Many usual softwares (\emph{e.g.} {\tt magma} and {\tt sage}) include 
-primitives to handle $p$-adic numbers. Since $p$-adic numbers are by 
-nature infinite, one needs to truncate them up to some finite level (we 
-generally say \emph{finite precision}) in order to manipulate them on a 
-computer. The usual technics to achieve this are inspired from case of 
-real numbers: the computer works with ``truncated $p$-adic numbers'' 
-represented as $x + O(p^n)$ and performs all basic operations (addition, 
-product) on this representation. For instance, it computes: 
-$$(x + O(p^n)) + (y + O(p^m)) = (x+y) + O(p^{\min(n,m)}).$$
-However, in many important cases (\emph{e.g.} those discussed just 
-above), this approach is too naive since it creates artificially 
-important losses of precision (see \S\ref{ssec:stepbystep}). 
-It is the so-called \emph{numerical instability} issue.
-
-The aim of the current paper is to provide a systematical theorical 
-study of $p$-adic precision (and more generally nonarchimedian 
-precision) and to derive from this study new methods to handle 
-precision, that turns out to give nearly optimal results in a quite 
-large setting.
-
-\section{Motivations and aims}
+in implementations of $p$-adic numbers.  The mathematical software packages
+Sage \cite{sage}, PARI \cite{pari} and Magma \cite{magma} all include $p$-adic elements
+that track precision in this way.
+
+The approach of propagating precision with each arithmetic operation works well, but does
+sometimes underestimate the known precision of a result, as we will discuss in Section
+\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
+implementations of polynomials, vector spaces, matrices and power series.  The practice
+of storing the precision within each entry is not flexible enough for all applications.
+Sometimes only a rough accounting of precision is needed, in which case storing
+and computing the precision of each entry in a large matrix needlessly consumes
+space and time.  Conversely, recording the precision of each entry does not allow
+a constraint such as specifying the precision of $f(0)$, $f(1)$ and $f(2)$ for a
+quadratic polynomial $f$.
+
+For a vector space $V$ over $\Qp$, we propose that the fundamental object used to
+store the precision of an element should be a $\Zp$-lattice $H \subset V$.  By using
+general lattices one can eliminate needless loss of precision.  Moreover, specifying
+the precision of each entry or recording a fixed precision for all entries can both
+be interpreted in terms of lattices.  In Section \ref{sec:prec-proposal} we detail our
+proposal for how to represent the precision of an element of a vector space.
+
+\todofor{David}{Finish summarizing sections}
+
+While we have used $\Qp$ throughout this paper for concreteness, all of our results
+and suggestions apply to arbitrary complete discrete valuation rings.
+
+\section{Tracking Precision} \label{sec:prec-proposal}
 

### 2014-02-14 padicprec.tex
Working on Section 2

@@
 
-While we have used $\Qp$ throughout this paper for concreteness, all of our results
-and suggestions apply to arbitrary complete discrete valuation rings.
-
 \section{Tracking Precision} \label{sec:prec-proposal}
+
+We recall some standard definitions from ultrametric analysis, referring to
+Schneider \cite{Schneider} for a complete exposition.
+
+Let $K$ be a non-Archimedean local field with absolute value
+$|\cdot| : K \to \R_+$ and value group $p^\Z$.   For example, we may take $K = \Qp$ with the
+$p$-adic absolute value or $K = \Fp\lb t \rb$ with the $t$-adic absolute value.
+Write $\OK$ for the ring of integers of $K$ and choose a dense subring $R \subset K$
+consisting of elements that can be represented exactly.  For $K = \Qp$ we may
+choose $R = \Z[\frac1p]$ or $R = \Q$; for $K = \Fp\lb t \rb$ we may choose
+$R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.  Choose a collection of subsets $S_h \subset R$
+indexed by positive integers $h$ so that
+\begin{enumerate}
+\item $S_h \subset S_{h+1}$,
+\item $\{B_K(x, p^{-h}) : x \in S_h\}$ disjointly covers $K$.
+\end{enumerate}
+The choice of the $S_h$ is possible since $K$ is non-Archimedean and locally compact.
+
+An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
+: K \to \R_+$ satisfying the triangle inequality and sending every 
+nonzero element of $K$ to a strictly positive real number. It is said 
+\emph{ultrametric} if it satisfies the following strong from of the 
+triangle inequality:
+$$|x+y| \leq \max(|x|, |y|).$$
+Given an absolute value $|\cdot|$ on $K$, we note that $|K^\times|$ is a 
+multiplicative subgroup of $\R_+^\times$. It is then either discrete or 
+dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\times|$ is
+discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
+Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
+|x-y|$) converge.
+
+The formula $|x| = p^{-\val(x)}$ defines an ultrametric absolute value on 
+the field $\Qp$ for which it is complete. More generally, the $p$-adic 
+absolute value extends uniquely to any finite extension $K$ of $\Qp$ and 
+turns $K$ into a complete ultrametric field. Another classical example of 
+complete ultrametric field is the field of Laurent series over a finite
+field, the absolute value being defined from the valuation of a series.
+
+From now, we fix a field $K$ equipped with an ultrametric absolute value 
+denoted by $|\cdot|$ and we assume that $K$ is complete. Given $v$ a 
+$K$-vector space (of possibly infinite dimension), an \emph{ultrametric 
+norm} on $V$ is a map $\Vert\cdot\Vert : V \to \R^+$ satisfying:
+\begin{enumerate}[(i)]
+\item $\Vert x\Vert = 0$ if and only if $x = 0$;
+\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
+\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
+\end{enumerate}
+A norm gives rise to a distance and a normed $K$-vector space $V$ is 
+said \emph{complete} when any Cauchy sequence with values in $V$ 
+converges. A normed $K$-vector space which is complete is called a 
+\emph{$K$-Banach}.
+
+Similarly to the real case, any finite dimensional normed $K$-vector 
+space is complete and all norms over such a space are equivalent 
+(\emph{i.e.} they define the same topology). Many usual Theorems 
+concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
+Theorem) remain true in the ultrametric setting.
+
+\begin{deftn} \label{deftn : diff}
+Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
+subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
+\emph{differentiable} at $v_0 \in U$ if there exists a 
+continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
+$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
+\subset U$ containing $v_0$ with:
+\[ 
+\forall v, w \in U_\varepsilon, \quad
+\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
+\leq \varepsilon \Vert v-w \Vert. 
+\]
+The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
+\end{deftn}
+
+One can derive all usual properties of the differential from this 
+definition: unicity, chain rules, \emph{etc}.
+
+\subsection{Lattices}
+
+Let $E$ be a $K$-normed vector space. Given $r \in \R^+$, 
+we denote by $B^-_E(r)$ (resp. by $B_E(r)$) the set of elements $x 
+\in E$ such that $\Vert x \Vert < r$ (resp. $\Vert x \Vert \leq r$). 
+We insist on the following maybe unusual fact: $B^-_E(r)$ and 
+$B_E(r)$ are both open and closed for the norm topology on $E$.
+Clearly, $B_K(0,1)$ is stable under multiplication. Moreover, the 
+ultrametric triangle inequality ensures that $B_K(1)$ is stable under 
+addition as well. Thus, $B_K(1)$ is a subring of $K$: it is often 
+called the \emph{integer ring} of $K$.
+
+\begin{deftn}
+Let $E$ be a $K$-Banach.
+A \emph{lattice} in $E$ is an open bounded sub-$B_K(1)$-module of $E$.
+\end{deftn}
+
+\begin{rmk}
+Any lattice $H$ in $E$ is also closed since its complement is the
+union of all cosets $a + H$ (with $a \not\in H$) which are all open.
+\end{rmk}
+
+Let us recall first 
+that a datum involving elements of a complete ultrametric field $K$ 
+(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
+on a computed because they basically convey an infinite amount of 
+informations. That is why, instead of this datum itself, we generally 
+work with neighborhoods of it.
+
+In what follows, we shall consider neighborhoods of the form $x + H$ 
+where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
+in $E$. A very basic example is the approximate $p$-adic number $1 + 
+O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
+the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
+nevertheless emphasize that our definition is quite flexible and allows 
+many neighborhoods that we do not consider in general. For instance, an 
+``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
+of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
+P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
+$a_i \in \Q_p$ and relative integers $n_i$.
+
+In the context of computer science, we will often use the notation $x + 
+O(H)$ to refer to $x + H$.
 
@@
 \label{ssec:ultrametric}
-
-We recall usual definitions and properties related to ultrametric 
-analysis. We refer to Schneider \cite{Schneider} for a complete 
-exposition.
-
-An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
-: K \to \R_+$ satisfying the triangle inequality and sending every 
-nonzero element of $K$ to a strictly positive real number. It is said 
-\emph{ultrametric} if it satisfies the following strong from of the 
-triangle inequality:
-$$|x+y| \leq \max(|x|, |y|).$$
-Given an absolute value $|\cdot|$ on $K$, we note that $|K^\times|$ is a 
-multiplicative subgroup of $\R_+^\times$. It is then either discrete or 
-dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\times|$ is
-discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
-Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
-|x-y|$) converge.
-
-The formula $|x| = p^{-\val(x)}$ defines an ultrametric absolute value on 
-the field $\Qp$ for which it is complete. More generally, the $p$-adic 
-absolute value extends uniquely to any finite extension $K$ of $\Qp$ and 
-turns $K$ into a complete ultrametric field. Another classical example of 
-complete ultrametric field is the field of Laurent series over a finite
-field, the absolute value being defined from the valuation of a series.
-
-From now, we fix a field $K$ equipped with an ultrametric absolute value 
-denoted by $|\cdot|$ and we assume that $K$ is complete. Given $v$ a 
-$K$-vector space (of possibly infinite dimension), an \emph{ultrametric 
-norm} on $V$ is a map $\Vert\cdot\Vert : V \to \R^+$ satisfying:
-\begin{enumerate}[(i)]
-\item $\Vert x\Vert = 0$ if and only if $x = 0$;
-\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
-\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
-\end{enumerate}
-A norm gives rise to a distance and a normed $K$-vector space $V$ is 
-said \emph{complete} when any Cauchy sequence with values in $V$ 
-converges. A normed $K$-vector space which is complete is called a 
-\emph{$K$-Banach}.
-
-Similarly to the real case, any finite dimensional normed $K$-vector 
-space is complete and all norms over such a space are equivalent 
-(\emph{i.e.} they define the same topology). Many usual Theorems 
-concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
-Theorem) remain true in the ultrametric setting.
-
-\begin{deftn} \label{deftn : diff}
-Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
-subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
-\emph{differentiable} at $v_0 \in U$ if there exists a 
-continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
-$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
-\subset U$ containing $v_0$ with:
-\[ 
-\forall v, w \in U_\varepsilon, \quad
-\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
-\leq \varepsilon \Vert v-w \Vert. 
-\]
-The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
-\end{deftn}
-
-One can derive all usual properties of the differential from this 
-definition: unicity, chain rules, \emph{etc}.
-
-\subsection{Lattices}
-
-Let $E$ be a $K$-normed vector space. Given $r \in \R^+$, 
-we denote by $B^-_E(r)$ (resp. by $B_E(r)$) the set of elements $x 
-\in E$ such that $\Vert x \Vert < r$ (resp. $\Vert x \Vert \leq r$). 
-We insist on the following maybe unusual fact: $B^-_E(r)$ and 
-$B_E(r)$ are both open and closed for the norm topology on $E$.
-Clearly, $B_K(0,1)$ is stable under multiplication. Moreover, the 
-ultrametric triangle inequality ensures that $B_K(1)$ is stable under 
-addition as well. Thus, $B_K(1)$ is a subring of $K$: it is often 
-called the \emph{integer ring} of $K$.
-
-\begin{deftn}
-Let $E$ be a $K$-Banach.
-A \emph{lattice} in $E$ is an open bounded sub-$B_K(1)$-module of $E$.
-\end{deftn}
-
-\begin{rmk}
-Any lattice $H$ in $E$ is also closed since its complement is the
-union of all cosets $a + H$ (with $a \not\in H$) which are all open.
-\end{rmk}
 
@@
 important applications to effective computations with $p$-adic numbers 
-or power series, that we discuss in this Section. Let us recall first 
-that a datum involving elements of a complete ultrametric field $K$ 
-(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
-on a computed because they basically convey an infinite amount of 
-informations. That is why, instead of this datum itself, we generally 
-work with neighborhoods of it.
-
-In what follows, we shall consider neighborhoods of the form $x + H$ 
-where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
-in $E$. A very basic example is the approximate $p$-adic number $1 + 
-O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
-the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
-nevertheless emphasize that our definition is quite flexible and allows 
-many neighborhoods that we do not consider in general. For instance, an 
-``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
-of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
-P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
-$a_i \in \Q_p$ and relative integers $n_i$.
-
-In the context of computer science, we will often use the notation $x + 
-O(H)$ to refer to $x + H$.
+or power series, that we discuss in this Section.
 

### 2014-02-14 padicprec.tex
Added missing \newcommands
Some improvement on loc. anal.
Need work...

@@
 \renewcommand{\O}{\mathcal O}
+\newcommand{\OK}{\mathcal{O}_K}
 \newcommand{\XX}{\mathbf X}
@@
 \renewcommand{\prec}{\text{\rm prec}}
+
+\newcommand{\lb}{\ensuremath{\llbracket}}
+\newcommand{\rb}{\ensuremath{\rrbracket}}
 
@@
 
-We refer to Schneider \cite{Schneider}, \S 6, for a definition of a 
-locally analytic function.
-
-\begin{deftn}\label{deftn:analyticNP}
-Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ be locally analytic map $E \rightarrow F$.
-Let $x \in E$, and let $f(x+h)=f(x)+\sum_{k \geq 0}^{+\infty} L_k \cdot (h,\dots,h)$ be the power series development of $f$ at $x$.
-Let $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L^{k+1}(x) \interleave  t^k \in \mathbb{R}[[t]].$
-
-We define $NP(f,x)$ to be the function defined by its Newton polygon, the lower convex hull of the $(k,\log ( \interleave L^{k+1}(x) \interleave)),$ with $k \geq 0.$
+We refer to Schneider \cite{Schneider}, \S 6, for a definition of a locally analytic function.
+
+\begin{deftn} \label{deftn :  serie and norm}
+Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie over $E,$ a complete normed $K$-vector space.
+We assume that for any $\lambda  \in K,$ $x \in E,$ $n \in \mathbb{N},$ $\vert f_n(\lambda x) \vert = \vert \lambda \vert \vert f_n (x) \vert.$
+We define $\vert f_n \vert = \sup_{x \in B_E(0,1)} \vert f_n(x) \vert,$ and we assume that all the $\vert f_n \vert $ are finite. We define $R((f_n)_n)$ to be the radius of convergence of the serie.
+
+We also define : \[\Lambda((f_n)_n) (v) = \log \sup_{x \in B(0,e^v)} \vert f(x) \vert.\]
+We define  $NP((f_n)_n)$ to be the Newton polygon of $f$, that is to say, the lower convex hull in $\mathbb{R}^2$ of the $(k, - \log \vert f_k \vert ),$ $k\geq 0.$ 
 \end{deftn}
 
-We now prove that the knowledge of $NP(f,x)$ is enough to recover $\delta$.
-
-\begin{prop}\label{prop:locanalytic}
-Let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}. We assume $L_1$ is surjective. 
-Let $C$ be given by the Banach-Schauder theorem for $L_1$, $\varepsilon = \frac{1}{C}$. 
-Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP(f,x)$.
-Then $r$ such that $\mu=\log_p r$ is enough for lemma \ref{lem:main}.
+The following proposition is a classical consequence of convex analysis.
+
+\begin{prop} \label{prop : conjugate Newton}
+We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $NP((f_n)_n).$
+We denote by $NP((f_n)_n)^*$ the convex conjugate of $NP((f_n)_n).$ $NP((f_n)_n)^*$ is defined over $\left] -\infty,v_{\infty} \right[$ (at least)
+\end{prop}
+
+$\Lambda((f_n)_n)$ and $NP((f_n)_n)$ are connected :
+
+\begin{prop} \label{prop : connexion sup Newton}
+We have $R((f_n)_n) = e^{v_{\infty}}.$
+Moreover, for any $v < \log \left( R((f_n)_n) \right),$ \[ \Lambda ((f_n)_n) (v) = NP((f_n)_n)^*(v).\]
 \end{prop}
 \begin{proof}
-Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=L_1 \cdot x +\sum_{n \geq 2} L_n \cdot (x, \dots,x)$ is the Taylor serie expansion of $f$ at $0$, coinciding on $B_E (0,r)$.
-
-We shall now find $l$ such that if $x \in B_{K^n}(0,l)$, then $\Vert \sum_{n \geq 2} L_n \cdot (x,\dots,x) \Vert \leq \varepsilon \vert x \vert $.
-
-By continuity and ultrametricity, it is enough for $l$ to be such that, if $x \in B_{K^n}(0,l)$, then  \[ \sup_{n \geq 2} \interleave L_n \interleave \Vert x \Vert^n \leq \varepsilon \Vert x \Vert .\]
-
-Therefore, \[ \sup_{n \geq 2} \interleave L_n \interleave l^{n-1}  \leq \varepsilon . \]
-
-In other words, $\forall n \geq 2$, $\interleave L_n \interleave \l^{n-1}  \leq \varepsilon.$
-
-By passing to logarithm, we get for all $n \geq 2$ : \[ \log_p (\interleave L_n \interleave )-(n-1) \log_p l \geq - \log_p \varepsilon.\]
-
-If we denote by $NP_F$ the Newton polygon of the power series 
-\[
-F(t) = \sum_{k\geq 0}^{+\infty} \interleave L_{k+1} \interleave  t^k
-\]
-then it is enough that:
-\[
-\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \log_p l \right) \geq - \log_p \varepsilon.
-\]
-
-If $NP_F^*$ is the Legendre-Fanchel transform of $NP_F$, then it is enough that :
-\[ NP_F^* \left( \log_p l  \right) \leq \log_p \varepsilon.\]
-
-Therefore, by definition of the Legendre-Fanchel transform, it is enough to :
-\begin{itemize}
-\item take $M=(0,\log_p \varepsilon)$,
-\item draw the line passing by $M$ tangent to the graph of $NP_F$ (it could be tangent at infinity),
-\item if we denote by $\mu$ the slope of this line, then take $l$ such that $\mu=\log_p l$.
-\end{itemize}
+Let $v \in \left] -\infty,v_{\infty} \right[,$ and let us assume that $v$ is not a slope of $NP((f_n)_n).$
+Let $u = NP((f_n)_n)^*(v).$ Then the intersection of the line defined by $y=vx-u$ and $NP((f_n)_n)$ is an extreme point of $NP((f_n)_n),$ say $(n, - \log ( \vert f_n \vert ))$ for some $n \geq 0,$ and $u = vn + \log \vert f_n \vert.$
+
+$v$ is not a slope of $NP((f_n)_n)$ and by convexity of $NP((f_n)_n)$, its difference quotients at $n$ are increasing, therefore, there exists some $c>0$ such that for any $m\geq 0,$ $vm-u \leq - \log \vert f_m \vert - c \vert n-m \vert.$ Since $vm-u=vm-vn-log \vert f_n \vert,$ we get $-vn-\log \vert f_n \vert + c \vert n-m \vert  \leq -vm-\log \vert f_m \vert .$
+
+Now, let $x \in  B(0,e^v),$ and $m \geq 0.$ Definition of $\vert f_m$ yields $ \vert f_m (x) \vert \leq \vert f_m \vert e^{vm}.$ Therefore \[\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn}< \vert f_n \vert e^{vn}.\]
+Therefore, the serie converges and by ultrametricity $ \vert f(x) \vert \leq \vert f_n \vert e^{vn},$ and $\Lambda ((f_l)_l)(v) \leq \log \left( \vert f_n \vert e^{vn} \right).$ Furthermore,  $R((f_n)_n)  \geq e^v.$
+
+Now, definition of $\vert f_n \vert$  provides $(x_i)_{i\geq 0}$ in $B(0,e^v)$ such that $\lim \vert f_n (x_i)= \vert f_n \vert e^{vn}.$
+
+Since for any $i,$ $\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn},$ then for $i$ large enough, $\vert f_m (x) \vert < \vert f_n (x) \vert $ and $\vert f(x) \vert = \vert f_n (x) \vert.$
+
+As a consequence, $\lim \vert f(x_i) \vert = \vert f_n \vert e^{vn},$ and  $\Lambda ((f_l)_l)(v) = \log ( \vert f_n \vert e^{vn}) = u= NP((f_n)_n)^*(v).$
+
+Therefore, for any $v \in \left] -\infty,v_{\infty} \right[,$ with $v$ not a slope of $NP((f_n)_n),$ we get $\Lambda ((f_l)_l)(v) =NP((f_n)_n)^*(v).$
+Since both $\Lambda ((f_l)_l)$ is increasing and $NP((f_n)_n)^*$ is continuous (since it is convex), we deduce that $R((f_n)_n) \geq  e^{v_{\infty}}$ and both functions coincide on $v \in \left] -\infty,v_{\infty} \right[,$
+
+One can show similarily that $R((f_n)_n) \leq  e^{v_{\infty}}.$
 \end{proof}
 
-We remark that when considering finite-dimensional applications over $p$-adic fields, a more refined proposition is available.
-
-\begin{prop} \label{prop:locanalyticfinitedim}
-If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C$ and $\varepsilon$ are as above.
-It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : Let $\mu$ be the slope of the line passing by $(0, \log_p \varepsilon)$ and tangent to $NP_F$.
-Then $r$ such that $\mu+\frac{1}{p-1}=\log_p r$ is enough for lemma \ref{lem:main}. 
+\begin{cor} Let $f(x)= \sum_{n \geq 0} f_n (x)$ and $g(x)= \sum_{n \geq 0} g_n (x)$ be two series as in definition \ref{deftn :  serie and norm}.
+Let $f(x)g(x)=\sum_{n \geq 0} \mu_n (x) $ and $f \circ g(x)=\sum_{n \geq 0} \tau_n (x) .$ Then :
+$\Lambda((f_n+g_n)_n) \leq \max (\Lambda((f_n)_n),\Lambda((g_n)_n))$
+$\Lambda((\mu_n)_n) \leq \Lambda((f_n)_n)+\Lambda((g_n)_n)$
+$\Lambda((\tau_n)_n) \leq \Lambda((f_n)_n) \circ \Lambda((g_n)_n)$
+\end{cor}
+
+\begin{prop} \label{prop : legendre transform}
+If $f$ and $g$ are two real convex functions such that $f \geq g,$ then $f^* \leq g^*.$
+Furthermore, slopes of $f^*$ corresponds to extreme points of $f$.
+\end{prop}
+
+\begin{cor} \label{cor : trunc series}
+Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie as in definition \ref{deftn :  serie and norm}, $n_0 \geq 0$ and $g(x)= \sum_{n \geq n_0} f_n (x).$ Then $\Lambda((g_n)_n) \leq \Lambda((f_n)_n)$ and the slopes of $\Lambda((g_n)_n)$ are bigger than $n_0$.
+As a consequence, a minoration of $\Lambda((g_n)_n)$ can be obtain by extending $\Lambda((f_n)_n)(v)$ with a slope $n_0$ for $v \leq n_0$.
+\end{cor}
+
+Now, let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}.
+We assume $L_1$ is surjective.
+We define, for any $n \geq 1$, $f_n(x)=L_n(x,\dots,x)$ for any $x \in E.$
+Let $C$ be given by the Banach-Schauder theorem for $L_1$, $\varepsilon = \frac{1}{C}$. We provide three methods to find a sufficient $r$ for \ref{lem:main}.
+
+For the first one, we assume that we have know a function that majorates $\Lambda((f_n)_{n \geq 1})$.
+
+\begin{prop} \label{prop : maj Lambda}
+If $m_1$ is a real convex function above $\Lambda((f_n)_{n \geq 1})$.
+We define $m_2$ by extending $m_1$ with slope $2$.
+Let $r$ be the abscissa of the intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x.$ Then $e^r$ is enough for \ref{lem:main}.
 \end{prop}
 \begin{proof}
-In comparison with the previous proposition, $L_n=\frac{f^{(n)}(0)}{n!}$. Therefore, by Legendre formula, it would be enough that for all $n \geq 2$ :
-\[\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon,\]
-and therefore, \[ NP_F^* \left( \log_p l - \frac{1}{p-1} \right) \leq \log_p \varepsilon.\]
-The result follows \textit{mutatis mutandis}.
+$r$ is well-defined thanks to corollary \ref{cor : trunc series}, and $m_2 \geq \Lambda((f_n)_{n \geq 2}).$
+
+One only has to prove that for all $x \in B(0,e^r),$ $\vert \sum_{n\geq 2} f_n(x) \vert \leq \varepsilon \vert x \vert.$
+
+If $x \in B(0,e^r),$ $u = \log \vert x \vert,$ then $\Lambda((f_n)_{n \geq 2})(u) \geq -\log \varepsilon+ u.$
+With proposition \ref{prop : connexion sup Newton}, for any $m \geq 2,$ $-\log \vert f_m \vert +m u \geq -\log \varepsilon+ u.$
+
+Therefore, for any $m \geq 2,$ $\vert f_m \vert  \vert x \vert^m \leq \varepsilon \vert x \vert.$
+By ultrametricity,   $\vert \sum_{n\geq 2} f_n(x) \vert \leq \varepsilon \vert x \vert,$ which concludes the proof.
+\end{proof}
+
+For the second one, we only assume that we have majorations on the $\vert f_n \vert.$
+
+\begin{prop}
+Let $(M_n)_{n \geq 2}$ be real numbers such that for all $n \geq 2$, $\vert f_n \vert \leq M_n.$
+
+If we denote by $NP((M_{n+1})_{n \geq 1})$ the lower convex hull of the $(n, -\log M_{n+1})$ for $n \geq 1$, then $r$ such that $-\log \varepsilon=NP((M_n)_{n \geq 2})^*(r)$ is enough.
+\end{prop}
+\begin{proof}
+Same idea as proposition \ref{prop : maj Lambda}. One can define $g_n$ with $g_n (x) =\frac{f_n(x)}{\vert x \vert}$ and $g_n(0)=0$ and work with $\Lambda ((g_n)_n).$
+\end{proof}
+
+We finally provide a simple method when working with finite dimension and $p$-adic fields.
+
+\begin{prop} \label{prop : locanalyticfinitedim}
+If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C$ and $\varepsilon$ are as above.
+It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}.
+\end{prop}
+\begin{proof}
+Same as before, with $L_n=\frac{f^{(n)}(0)}{n!}$ and Legendre formula. The result follows \textit{mutatis mutandis}.
 \end{proof}

### 2014-02-14 padicprec.tex
Main Lemma revisited: assumption on the lattice weakened

--- 2014-02-14 padicprec.tex
@@
 \newcommand{\val}{\text{\rm val}}
+\newcommand{\pr}{\text{\rm pr}}
 \newcommand{\tr}{\text{\rm Tr}}
@@
 
-The theory of $p$-adic precision we want to develop is mainly based 
-on a single lemma of $p$-adic analysis. The aim of this Section is
-to state and prove this lemma. Consequences will be discussed after
-in Section \ref{sec:tracking}.
+The theory of $p$-adic precision we want to develop is mainly based on a 
+single lemma (Lemma \ref{lem:main}) of $p$-adic analysis. The aim of 
+this Section is to state and prove this lemma. Consequences will be 
+discussed after in Section \ref{sec:tracking}.
 
@@
 
-\subsection{Ultrametric analysis}
-\label{ssec:ultrametric}
-
-The following Lemma shows that the notion of lattices behaves rather 
-well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
-\ref{lem:main}) that, under some extra hypothesis, the linearity 
-assumption can be drastically relaxed.} transformations.
-
-\begin{lem}
-\label{lem:morlat}
-Let $E$ and $F$ be two $K$-Banach and $f : 
-E \to F$ be a continuous \emph{surjective} linear map.
-If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
-\end{lem}
-
-\begin{proof}
-It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
-The fact that it is open is a direct consequence of the open mapping
-Theorem.
-\end{proof}
-
-To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
-on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
-the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
-\in H$. It follows from the fact that $H$ is open and bounded in $E$
-that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
-of $I_x$ under the absolute value map is bounded and contains a positive
-element. We can then define:
-$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
-It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
-$E$.
-
-\begin{lem} \label{lem:reductionball}
-\begin{enumerate}
-\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
-equivalent.
-\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
-\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
-Then:
-$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
-and the last inclusion is an equality if the absolute value on $K$ is
-discrete.
-\end{enumerate}
-\end{lem}
-
-\begin{proof}
-Left to the reader.
-\end{proof}
-
-The first part of Lemma \ref{lem:reductionball} asserts that the 
-identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
-$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
-assumption: it is ``between the open unit ball and the closed unit 
-ball''.
+\subsection{Lattices}
+
+%The following Lemma shows that the notion of lattices behaves rather 
+%well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
+%\ref{lem:main}) that, under some extra hypothesis, the linearity 
+%assumption can be drastically relaxed.} transformations.
+%
+%\begin{lem}
+%\label{lem:morlat}
+%Let $E$ and $F$ be two $K$-Banach and $f : 
+%E \to F$ be a continuous \emph{surjective} linear map.
+%If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
+%\end{lem}
+%
+%\begin{proof}
+%It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
+%The fact that it is open is a direct consequence of the open mapping
+%Theorem.
+%\end{proof}
+%
+%To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
+%on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
+%the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
+%\in H$. It follows from the fact that $H$ is open and bounded in $E$
+%that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
+%of $I_x$ under the absolute value map is bounded and contains a positive
+%element. We can then define:
+%$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
+%It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
+%$E$.
+%
+%\begin{lem} \label{lem:reductionball}
+%\begin{enumerate}
+%\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
+%equivalent.
+%\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
+%\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
+%Then:
+%$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
+%and the last inclusion is an equality if the absolute value on $K$ is
+%discrete.
+%\end{enumerate}
+%\end{lem}
+%
+%\begin{proof}
+%Left to the reader.
+%\end{proof}
+%
+%The first part of Lemma \ref{lem:reductionball} asserts that the 
+%identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
+%\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
+%$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
+%assumption: it is ``between the open unit ball and the closed unit 
+%ball''.
 
@@
 
-\begin{lem} \label{lem:main}
-Let $E$ and $F$ be two $K$-Banach and $f : U 
-\rightarrow F$ be a function defined on an open subset $U$ of $E$.
-We assume that $f$ is differentiable at some point $v_0 \subset 
-U$ and that the differential $f'(v_0)$ is surjective.
-
-Then, there exists a positive real number $\delta$ such that, for all $r 
-\in (0, \delta)$, the equality
-\begin{equation}
-\label{eq:main}
-f(v_0 + H) = f(v_0) + f'(v_0) (H).
-\end{equation}
-holds for each lattice $H$ such that $B^-_E(r) \subset H \subset
-B_E(r)$.
-\end{lem}
-
-\begin{rmk}
-\label{rmk:mainlemma}
-Given a lattice $H_0$ in $E$, the equality \eqref{eq:main} holds for $H 
-= \lambda H_0$ with $\lambda$ small enough. To prove this, it is enough 
-to apply Lemma \ref{lem:main} to the map $f \circ \iota_{H_0}$ where 
-$\iota_{H_0} : (E, \Vert \cdot \Vert_{H_0}) \to (E, \Vert \cdot \Vert)$ 
-is the ``identity map'' (which is an homeomorphism by Lemma 
-\ref{lem:reductionball}).
-\end{rmk}
-
-\begin{proof}
-We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
-the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
-C  f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
-$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by by definition \ref{deftn : diff} and the fact that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
-B_E(\delta)$ for some $\delta >0$.
-
-Let $r \in (0, \delta)$. Let $B^-_E(r) \subset H \subset
-B_E(r).$
-Let us show that $f$ maps surjectively $H$ onto $f'(0) (H)$.
-
-We first prove that $f(H) \subset f'(0) (H)$.
-Let $x \in H$. By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
-\varepsilon \Vert x \Vert $. Therefore, if $y=f(x)-f'(0) \cdot x,$ then $\Vert y \Vert \leq \varepsilon r$. As a consequence of the open mapping Theorem, $B_F(\varepsilon r) \subset 
-\varepsilon C  f'(0) \cdot B_E(r),$ and there exists $x_1 \in \varepsilon C  \cdot B_E(r)$ such that $f'(0) \cdot x_1 =y.$ Since $\varepsilon C <1,$  $x_1 \in B^-_E(r) \subset H.$ and finally, $f(x)= f'(0) \cdot (x-x_1) \in f'(0) (H)$.
-
-We shall now prove surjectivity. Let $y \in f'(0) (H)$.
-Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
-Since $f(x_0) \in f'(0) (H)$, we can define $z_0 \in E$ to be such that $f'(0) \cdot z_0 = y - f(x_0)$ and $\Vert z_0 \Vert \leq C \cdot \Vert y - f(x_0) \Vert$.
-We then define by induction two sequences $(x_n)$ and $(z_n)$ by
-\begin{itemize}
-\item $x_{n+1}=x_n+z_n$ and 
-\item $z_n \in E$ is such that
-$f'(0) \cdot z_n = y - f(x_n)$ and
-$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$.
-\end{itemize}
-By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y 
-= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y \Vert < \varepsilon 
-\Vert x_0 \Vert \leq \varepsilon r$. Thus $\Vert z_0 \Vert \leq 
-\varepsilon C \cdot r < r$. Thus, $z_0$ and $x_1=x_0+z_0$  lie in $B^-_E(r) 
-\subset H$. An easy induction involving an analoguous computation
-shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
-take their values in $H$.
-
-We notice that $y - f(x_{n+1}) = f(x_n) - f(x_{n+1}) + f'(0) \cdot z_n$.
-Since $z_n=x_{n+1}-x_n,$, if we use differentiability, we get:
-$$\Vert y  - f(x_{n+1}) \Vert 
-\leq \varepsilon \Vert z_n \Vert \leq \varepsilon C \cdot \Vert 
-y - f(x_n) \Vert,$$
-from what we deduce that $\Vert y - f(x_{n+1}) \Vert = 
-O(a^n)$ and $\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < 1$.
-This shows that $(x_n)$ is a Cauchy sequence and, since $E$ is complete, it converges.
-Because $H$ is closed, if $x$ denotes the limit of $(x_n)$, we have $x \in H.$ Yet, as a consequence of the differentiability, $f$ is continuous over $U_\varepsilon.$ Therefore, since $x \in H \subset U_\varepsilon,$ then  $y=f(x)$.
-The result is proven.
-\end{proof}
-
 \begin{deftn}
@@
 A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
-$v_0$ if
-$$f(v_0 + H) = f(v_0) + f'(v_0) (H).$$
+$v_0$ if the equality
+$$f(v_0 + H) = f(v_0) + f'(v_0) (H)$$
+holds.
 \end{deftn}
 
-By Lemma \ref{lem:main} and Remark \ref{rmk:mainlemma}, if $f$ is 
-differentiable at $v_0$ and $f'(v_0)$ is surjective, then for all 
-lattice $H$ in $E$, there exists a constant $\delta$ such that $aH$ 
-is a first order lattice for $f$ at $v_0$ for all $a \in B_E(r)$.
+\begin{lem} \label{lem:main}
+Let $E$ and $F$ be two $K$-Banach and $f : U 
+\rightarrow F$ be a function defined on an open subset $U$ of $E$.
+We assume that $f$ is differentiable at some point $v_0 \subset 
+U$ and that the differential $f'(v_0)$ is surjective.
+
+Then, for all $\rho \in (0, 1]$, there exists a positive real 
+number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
+$H$ such that $B^-_E(\rho r) \subset H \subset B_E(r)$ is a first
+order lattice for $f$ at $v_0$.
+\end{lem}
+
+\begin{proof}
+We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
+the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
+C  f'(0)(B_E(1))$. Let $\varepsilon>0$ be such that 
+$\varepsilon C < \rho$. Let $U_\varepsilon \subset E$ given by Definition 
+\ref{deftn : diff} and the fact that $f$ is differentiable at $0$. We 
+may assume $U_\varepsilon = B_E(\delta)$ for some $\delta >0$.
+
+Let $r \in (0, \delta)$. Let $B^-_E(r) \subset H \subset B_E(r).$ Let us 
+show that $f$ maps surjectively $H$ onto $f'(0) (H)$. We first prove 
+that $f(H) \subset f'(0) (H)$. Let $x \in H$. By differentiability at 
+$0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
+Therefore, if $y=f(x)-f'(0)(x)$, then $\Vert y \Vert \leq 
+\varepsilon r$. By definition of $C$, we have $B_F(\varepsilon r) 
+\subset \varepsilon C f'(0) (B_E(r))$. Thus there exists $x_1 \in 
+\varepsilon C \cdot B_E(r)$ such that $f'(0) (x_1) =y$. Since 
+$\varepsilon C < \rho$, we get $x_1 \in B^-_E(\rho r) \subset H$ and 
+then $f(x)= f'(0) (x-x_1) \in f'(0) (H)$.
+
+We shall now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
+be such that $y = f'(0) (x_0)$. We define inductively two sequences 
+$(x_n)$ and $(z_n)$ as follows:
+\begin{itemize}
+\item $z_n$ is an element of $E$ satisfying $f'(0)(z_n) = y - 
+f(x_n)$ and $\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$ (such 
+an element exists by definition of $C$), and
+\item $x_{n+1}=x_n+z_n$.
+\end{itemize}
+For convenience, let us also define $x_{-1} = 0$. We claim that the 
+sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
+$H$. To establish this, it is enough to prove that, assuming that 
+$x_{n-1}$ and $x_n$ belongs to $H$, so do $z_n$ and $x_{n+1}$. Noticing
+that 
+\begin{equation}
+\label{eq:mainlemma}
+y - f(x_n) 
+= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n)
+= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
+\end{equation}
+we deduce using differentiability that
+$\Vert y - f(x_n) \Vert \leq \varepsilon \cdot \Vert x_n - x_{n-1}
+\Vert$.
+Since we are assuming that $x_{n-1}$ and $x_n$ lie in $H \subset
+B_E(r)$, we find $\Vert y - f(x_n) \Vert \leq \varepsilon r$. Thus
+$\Vert z_n \Vert \leq C \cdot \varepsilon r < \rho r$ and then
+$z_n \in H$. From the relation $x_{n+1} = x_n + z_n$, we finally
+deduce $x_{n+1} \in H$.
+
+Now using again Eq.~\eqref{eq:mainlemma} and differentiability at $0$,
+we get, for all positive integer $n$,
+$$\Vert y  - f(x_n) \Vert 
+\leq \varepsilon \cdot \Vert z_{n-1} \Vert \leq \varepsilon C \cdot \Vert 
+y - f(x_{n-1}) \Vert,$$
+from what we deduce that $\Vert y - f(x_n) \Vert = O(a^n)$ and 
+$\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < \rho \leq 1$. 
+This shows that $(x_n)$ is a Cauchy sequence and, since $E$ is complete, 
+it converges. Because $H$ is closed, if $x$ denotes the limit of 
+$(x_n)$, we have $x \in H$. Moreover, as a consequence of the 
+differentiability, $f$ is continuous over $U_\varepsilon$. Therefore, 
+since $x \in H \subset U_\varepsilon$, we get $y=f(x)$ and we are done.
+\end{proof}
+
+Let us conclude this paragraph by a remark on the assumption made in 
+Lemma \ref{lem:main} about the surjectivity of $f'(v_0)$. First let us 
+emphasize that this hypothesis is definitely necessary. Indeed, it would 
+implies otherwise that the image of $f$ is locally contained in a proper 
+sub vector space around each point where the differential of $f$ is not 
+surjective, which is certainly not true! Nevertheless, one can get 
+around this issue in many situations. Indeed, consider the situation of 
+Lemma \ref{lem:main} without assuming that $f'(v_0)$ is surjective. Now, 
+we pick a closed sub vector space $K$ of $F$ such that $K + f'(v_0)(E) = 
+F$. Then, denoting by $\pr_K$ the canonical projection of $F$ onto 
+$F/K$, the composite $\pr_K \circ f$ is certainly differentiable at 
+$v_0$ at its differential is surjective. Hence Lemma \ref{lem:main}
+applies and shows that
+\begin{equation}
+\label{eq:notsurjective}
+f(v_0 + H) \subset f(v_0) + f'(v_0)(H) + K
+\end{equation}
+for all lattices $H$ satisfying a certain condition. Now we can take
+the intersection of the right hand side of \eqref{eq:notsurjective} 
+over all possible $K$ and hope that we get this way a sharp affine upper 
+bound of $f(v_0 + H)$.
 
@@
 
-Using Lemma \ref{lem:reductionball}, one can extend Proposition 
-\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
-a lattice in some $K$-Banach space $E$, we first define the two
-following quantities:
-\begin{itemize}
-\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
-\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
-\end{itemize}
-
-\begin{cor}
-Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
-respectively and $f : U \to V$ be a locally analytic function. Let 
-also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
-Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
-soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
-constant depending only on $f$ and $v_0$ which can be determined 
-explicitely from the Newton polygon of $f$ at $v_0$.
-\end{cor}
-
-\begin{proof}
-We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
-\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-\Vert)$ is induced by identity map.
-\end{proof}
+%Using Lemma \ref{lem:reductionball}, one can extend Proposition 
+%\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
+%a lattice in some $K$-Banach space $E$, we first define the two
+%following quantities:
+%\begin{itemize}
+%\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
+%\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
+%\end{itemize}
+%
+%\begin{cor}
+%Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
+%respectively and $f : U \to V$ be a locally analytic function. Let 
+%also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
+%Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
+%soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
+%constant depending only on $f$ and $v_0$ which can be determined 
+%explicitely from the Newton polygon of $f$ at $v_0$.
+%\end{cor}
+%
+%\begin{proof}
+%We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
+%\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
+%\Vert)$ is induced by identity map.
+%\end{proof}
 
@@
 With above definitions, Lemma \ref{lem:main} extends readily to 
-manifolds.
+manifolds. To state it the context of manifolds, we first need to define 
+a norm on the tangent space $T_x V$ (where $V$ is some $K$-variety and 
+$x$ is a point in $V$). There is actually in general no canonical choice 
+for this. Indeed, let us consider a $K$-manifold $V$ covered by charts 
+$U_i$'s ($i \in I$) which are open subset of $K$-Banach $E_i$'s. If $x$ 
+is a point in $V$, the tangent space $T_x V$ is by definition isomorphic 
+to $E_i$ for each index $i$ such that $x \in V_i$. A natural norm on 
+$T_x V$ is then the one obtained by pulling back the norm on $E_i$. 
+However, since the transition maps are not required to be isometries, 
+this norm do depend on the choice on $i$. They are nevertheless all
+equivalent because the transition maps are required to be continuous.
+
+In the next Lemma, we choose to endow $T_x V$ by one of the above norms.
 
@@
 together with a point $x \in V$ such that $f'(x) : T_x V \to T_{f(x)} W$ 
-is surjective. Let $H$ be a precision datum at $x$. Then, there exists
-a positive real number $\delta$ such that, for all $\lambda \in 
-B_K(\delta)$, the equality
-$$f(x + \lambda H) = f(x) + \lambda \cdot f'(x)(H)$$
-holds.
+is surjective.
+
+Then for all $\rho \in (0,1]$, there exists $\delta \in (0,1)$ such 
+that for all $r \in (0, \delta)$, the equality
+$$f(x + O(H)) = f(x) + O(f'(x)(H))$$
+holds for all lattices $H$ in $T_x V$ such that $B^{\phantom -}_{T_x V}(\rho r)
+\subset H \subset B^-_{T_x V}(r)$.
 \end{lem}
@@
 \begin{rmk}
-If $V$ and $W$ are locally analytic $K$-manifolds and the mapping $f$ is 
-locally analytic as well, then one can explicitely determine, using the 
-method of \S \ref{ssec:locanalytic}, a value $\delta$ for which the 
-above Lemma works.
+The constant $\delta$ that appears in the above Lemma depends (up to
+some multiplicative constant) on the norm that we have chosen on $T_x V$. 
+However, once this norm is fixed, and assuming further that $V$ and $W$ 
+are locally analytic $K$-manifolds and the mapping $f$ is locally 
+analytic as well, the constant $\delta$ can be made explicit using the 
+method of \S \ref{ssec:locanalytic}.
 \end{rmk}

### 2014-02-15 padicprec.tex
Some remarks about non-surjectivity of differentials (to be continued)

@@
 $v_0$ if the equality
-$$f(v_0 + H) = f(v_0) + f'(v_0) (H)$$
+\begin{equation}
+\label{eq:firstorder}
+f(v_0 + H) = f(v_0) + f'(v_0) (H)
+\end{equation}
 holds.
 \end{deftn}
+
+We emphasize that we require an equality in Eq.~\eqref{eq:firstorder}, 
+and not just an inclusion! It will be very important for the 
+applications we have in mind.
+Here comes our main Lemma.
 
@@
 
-Let us assume now that $f$ is locally analytic on $U$ and $H$ is a first 
-order lattice for $f$ at $x$. Then $f(x+H) = f(x) + f'(x)(H)$ and the 
-smallest possible $H'$ is $f'(x)(H)$. In other words, the function {\tt 
-f} generates artificial loss of precision if the precision $H'$ it 
-outputs is strictly larger that $f'(x)(H)$. The aim of this section is 
-to explain how, under the above hypothesis, one can implement the 
-function {\tt f} so that it always outputs the optimal precision $O(H')$ 
-with $H' = f'(x)(H)$.
+Let us assume now that $f$ is locally analytic on $U$ and in addition 
+that $f'(x)$ is surjective. Proposition \ref{prop:locanalytic} then 
+yields a rather simple sufficient condition to decide if a given lattice 
+$H$ is a first order lattice for $f$ at $x$. For such a lattice, by 
+definition, we have $f(x+H) = f(x) + f'(x)(H)$. Thus the smallest 
+possible $H'$ is $f'(x)(H)$. In other words, the function {\tt f} 
+generates artificial loss of precision if the precision $H'$ it outputs 
+is strictly larger that $f'(x)(H)$. The aim of this section is to 
+explain how, under the above hypothesis, one can implement the function 
+{\tt f} so that it always outputs the optimal precision $O(H')$ with $H' 
+= f'(x)(H)$.
 
@@
 
-\subsubsection*{Quick comparison with floating point arithmetics}
-
-It worths remarking that the two strategy designed above share many 
+\subsubsection*{Additional remarks}
+
+From the beggining, we have assumed that $f'(x)$ is surjective. Let us 
+discuss shortly what happens when this assumption is relaxed. As it is 
+explained after the proof of Lemma \ref{lem:main}, the first thing we 
+can do is to project the result onto different quotients, \emph{i.e.} to 
+work with the composites $\pr_K \circ f$ for a sufficiently large family 
+of closed sub vector spaces $K \subset F$ such that $K + f'(x)(E) = F$. 
+(We recall that here $\pr_K$ denotes the canonical projection to $F$ 
+onto the quotient $F/K$.) If $F$ has a natural system of coordinates 
+(which is the case in almost all settings), we can generally take the 
+$\pr_K$'s as the projections on each coordinate. Doing this, we end up 
+with a precision on each individual coordinate. Furthermore, we have the 
+guarantee that all these precisions are sharp (even if the lattice build
+from them is not).
+Let us illustrate the above discussion by an example: suppose that we 
+want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
+family of $n$ vectors to its Graam matrix. The differential of $f$ is
+clearly never surjective because $f$ takes its values in the subspace
+consisting of symmetric matrices. Nevertheless for all couple $(i,j)
+\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
+\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ that takes a matrix 
+$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
+their differentials generally are surjective (\emph{i.e.} they do not 
+vanish). Let $M$ be a matrix known at some finite precision such that 
+$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
+and/or the backward computation and get $f_{ij}(M)$ together with its 
+precision. Putting this together, we get the whole matrix $f(M)$ 
+together with a sharp precision datum on each entry.
+
+The study of the this example actually suggests another solution to 
+tackle the issue of non-surjectivity. Indeed, remark that our $f$ above 
+had not a surjective differential simply because its codomain was too 
+large: if we had replaced $f : (K^n)^n \to M_n(K)$ by $g : (K^n)^n \to 
+S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
+over $K$ of size $n$) defined in the same way, our problem would 
+disappear. Of course the image of a general $f$ is rarely a sub vector 
+space of $F$ but it is often a sub-$K$-manifold of $F$. 
+% Using the results of Appendix
+
+\medskip
+
+We conclude this Section by remarking that the two strategies designed 
+above (\emph{i.e.} forward and backward computations) share some 
 similarities with usual floating point arithmetics over the reals. 

### 2014-02-15 padicprec.tex
Working on Section 2

--- 2014-02-14 padicprec.tex
@@
 \newcommand{\val}{\text{\rm val}}
-\newcommand{\pr}{\text{\rm pr}}
 \newcommand{\tr}{\text{\rm Tr}}
@@
 \newcommand{\rb}{\ensuremath{\rrbracket}}
+\newcommand{\lp}{(\!(}
+\newcommand{\rp}{)\!)}
 
@@
 
+\subsection{Lattices}
+
 We recall some standard definitions from ultrametric analysis, referring to
@@
 
-Let $K$ be a non-Archimedean local field with absolute value
-$|\cdot| : K \to \R_+$ and value group $p^\Z$.   For example, we may take $K = \Qp$ with the
-$p$-adic absolute value or $K = \Fp\lb t \rb$ with the $t$-adic absolute value.
-Write $\OK$ for the ring of integers of $K$ and choose a dense subring $R \subset K$
-consisting of elements that can be represented exactly.  For $K = \Qp$ we may
-choose $R = \Z[\frac1p]$ or $R = \Q$; for $K = \Fp\lb t \rb$ we may choose
-$R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.  Choose a collection of subsets $S_h \subset R$
-indexed by positive integers $h$ so that
-\begin{enumerate}
-\item $S_h \subset S_{h+1}$,
-\item $\{B_K(x, p^{-h}) : x \in S_h\}$ disjointly covers $K$.
-\end{enumerate}
-The choice of the $S_h$ is possible since $K$ is non-Archimedean and locally compact.
-
-An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
-: K \to \R_+$ satisfying the triangle inequality and sending every 
-nonzero element of $K$ to a strictly positive real number. It is said 
-\emph{ultrametric} if it satisfies the following strong from of the 
-triangle inequality:
-$$|x+y| \leq \max(|x|, |y|).$$
-Given an absolute value $|\cdot|$ on $K$, we note that $|K^\times|$ is a 
-multiplicative subgroup of $\R_+^\times$. It is then either discrete or 
-dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\times|$ is
-discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
-Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
-|x-y|$) converge.
-
-The formula $|x| = p^{-\val(x)}$ defines an ultrametric absolute value on 
-the field $\Qp$ for which it is complete. More generally, the $p$-adic 
-absolute value extends uniquely to any finite extension $K$ of $\Qp$ and 
-turns $K$ into a complete ultrametric field. Another classical example of 
-complete ultrametric field is the field of Laurent series over a finite
-field, the absolute value being defined from the valuation of a series.
-
-From now, we fix a field $K$ equipped with an ultrametric absolute value 
-denoted by $|\cdot|$ and we assume that $K$ is complete. Given $v$ a 
-$K$-vector space (of possibly infinite dimension), an \emph{ultrametric 
-norm} on $V$ is a map $\Vert\cdot\Vert : V \to \R^+$ satisfying:
+Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$. 
+We assume that the induced metric is an ultrametric -- 
+\[ \lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert) \]
+-- and that $K$ is complete with respect to it.
+For example, we may take $K = \Qp$ with the
+$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
+Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
+that $K$ contains a dense subring $R \subset K$
+consisting of elements that can be represented with a finite amount of space.
+For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
+for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.
+
+If $E$ is a $K$-vector space, possibly of infinite dimension, then an
+\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
 \begin{enumerate}[(i)]
@@
 \end{enumerate}
-A norm gives rise to a distance and a normed $K$-vector space $V$ is 
-said \emph{complete} when any Cauchy sequence with values in $V$ 
-converges. A normed $K$-vector space which is complete is called a 
-\emph{$K$-Banach}.
-
-Similarly to the real case, any finite dimensional normed $K$-vector 
-space is complete and all norms over such a space are equivalent 
-(\emph{i.e.} they define the same topology). Many usual Theorems 
-concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
-Theorem) remain true in the ultrametric setting.
-
-\begin{deftn} \label{deftn : diff}
-Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
-subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
-\emph{differentiable} at $v_0 \in U$ if there exists a 
-continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
-$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
-\subset U$ containing $v_0$ with:
+A \emph{$K$-Banach space} is a complete normed $K$-vector space.
+
+\begin{deftn} % Separated definitions?
+Let $E$ be a $K$-Banach space.
+A \emph{lattice} in $E$ is an open bounded sub-$\OK$-module of $E$.
+\end{deftn}
+
+For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
+\begin{align*}
+B_E(r) &= \{ x \in E : \Vert x \Vert \le r \}, \\
+B^-_E(r) &= \{ x \in E : \Vert x \Vert < r\}.
+\end{align*}
+Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.
+
+\begin{deftn}
+Suppose $E$ is a $K$-Banach space and $I$ a set.
+A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
+every element $x \in E$ can be written
 \[
-\forall v, w \in U_\varepsilon, \quad
-\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
-\leq \varepsilon \Vert v-w \Vert. 
+x = \sum_{i \in I} \alpha_i x_i
 \]
-The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
+for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
 \end{deftn}
-
-One can derive all usual properties of the differential from this 
-definition: unicity, chain rules, \emph{etc}.
-
-\subsection{Lattices}
-
-Let $E$ be a $K$-normed vector space. Given $r \in \R^+$, 
-we denote by $B^-_E(r)$ (resp. by $B_E(r)$) the set of elements $x 
-\in E$ such that $\Vert x \Vert < r$ (resp. $\Vert x \Vert \leq r$). 
-We insist on the following maybe unusual fact: $B^-_E(r)$ and 
-$B_E(r)$ are both open and closed for the norm topology on $E$.
-Clearly, $B_K(0,1)$ is stable under multiplication. Moreover, the 
-ultrametric triangle inequality ensures that $B_K(1)$ is stable under 
-addition as well. Thus, $B_K(1)$ is a subring of $K$: it is often 
-called the \emph{integer ring} of $K$.
-
-\begin{deftn}
-Let $E$ be a $K$-Banach.
-A \emph{lattice} in $E$ is an open bounded sub-$B_K(1)$-module of $E$.
-\end{deftn}
-
-\begin{rmk}
-Any lattice $H$ in $E$ is also closed since its complement is the
-union of all cosets $a + H$ (with $a \not\in H$) which are all open.
-\end{rmk}
-
-Let us recall first 
-that a datum involving elements of a complete ultrametric field $K$ 
-(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
-on a computed because they basically convey an infinite amount of 
-informations. That is why, instead of this datum itself, we generally 
-work with neighborhoods of it.
-
-In what follows, we shall consider neighborhoods of the form $x + H$ 
-where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
-in $E$. A very basic example is the approximate $p$-adic number $1 + 
-O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
-the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
-nevertheless emphasize that our definition is quite flexible and allows 
-many neighborhoods that we do not consider in general. For instance, an 
-``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
-of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
-P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
-$a_i \in \Q_p$ and relative integers $n_i$.
-
-In the context of computer science, we will often use the notation $x + 
-O(H)$ to refer to $x + H$.
-
-\subsection{Separating precision from approximation}
-
-\todofor{David}{
-Explain here why it might be useful to work separately with 
-approximation on the one hand and precision on the other hand.
-}
-
-\subsection{Step-by-step tracking is not sharp}
-\label{ssec:stepbystep}
-
-In this subsection, we illustrate by some examples that tracking
-$p$-adic precision step by step may lead to important numerical
-instability.
-
-\subsubsection*{A first stupid example}
-
-Let $f : \Qp^2 \to \Qp^2$ be the function mapping $(x,y)$ to $(x+y, 
-x-y)$. We consider in addition two $p$-adic numbers $x$ and $y$, 
-which are known up to precision $O(p^n)$ and $O(p^m)$ respectively.
-The question we would like to address in this example is the
-computation of $f \circ f(x,y)$.
+Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
+% Lattice from Banach basis
+
+\subsection{Approximate elements}
+
+Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.  We say that an element
+$x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
+\begin{equation} \label{eq:exact_elt}
+x = \sum_{j \in J} \alpha_j x_j.
+\end{equation}
+The set of exact elements is dense in $E$.
+
+An approximate element is a pair $(x, H)$ where $x \in E$ is an exact element
+and $H$ is a lattice in $E$.  Such a pair represents an undetermined element of the coset
+$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
+the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
+we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
+
+Note that using an arbitrary lattice to represent the precision of an approximate element
+can reduce precision loss when compared to storing the precision of each coefficient
+$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
+$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
+$f \circ f(a + O(p^n), b + O(p^m))$.
 
@@
 of precision.
+
+Let us recall first 
+that a datum involving elements of a complete ultrametric field $K$ 
+(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
+on a computed because they basically convey an infinite amount of 
+informations. That is why, instead of this datum itself, we generally 
+work with neighborhoods of it.
+
+In what follows, we shall consider neighborhoods of the form $x + H$ 
+where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
+in $E$. A very basic example is the approximate $p$-adic number $1 + 
+O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
+the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
+nevertheless emphasize that our definition is quite flexible and allows 
+many neighborhoods that we do not consider in general. For instance, an 
+``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
+of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
+P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
+$a_i \in \Q_p$ and relative integers $n_i$.
+
+In the context of computer science, we will often use the notation $x + 
+O(H)$ to refer to $x + H$.
+
+\subsection{Separating precision from approximation}
+
+\todofor{David}{
+Explain here why it might be useful to work separately with 
+approximation on the one hand and precision on the other hand.
+}
+
+\subsection{Step-by-step tracking is not sharp}
+\label{ssec:stepbystep}
+
+In this subsection, we illustrate by some examples that tracking
+$p$-adic precision step by step may lead to important numerical
+instability.
+
+\subsubsection*{A first stupid example}
+
+Let
 
@@
 
-The theory of $p$-adic precision we want to develop is mainly based on a 
-single lemma (Lemma \ref{lem:main}) of $p$-adic analysis. The aim of 
-this Section is to state and prove this lemma. Consequences will be 
-discussed after in Section \ref{sec:tracking}.
+The theory of $p$-adic precision we want to develop is mainly based 
+on a single lemma of $p$-adic analysis. The aim of this Section is
+to state and prove this lemma. Consequences will be discussed after
+in Section \ref{sec:tracking}.
 
@@
 
-\subsection{Lattices}
-
-%The following Lemma shows that the notion of lattices behaves rather 
-%well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
-%\ref{lem:main}) that, under some extra hypothesis, the linearity 
-%assumption can be drastically relaxed.} transformations.
-%
-%\begin{lem}
-%\label{lem:morlat}
-%Let $E$ and $F$ be two $K$-Banach and $f : 
-%E \to F$ be a continuous \emph{surjective} linear map.
-%If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
-%\end{lem}
-%
-%\begin{proof}
-%It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
-%The fact that it is open is a direct consequence of the open mapping
-%Theorem.
-%\end{proof}
-%
-%To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
-%on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
-%the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
-%\in H$. It follows from the fact that $H$ is open and bounded in $E$
-%that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
-%of $I_x$ under the absolute value map is bounded and contains a positive
-%element. We can then define:
-%$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
-%It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
-%$E$.
-%
-%\begin{lem} \label{lem:reductionball}
-%\begin{enumerate}
-%\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
-%equivalent.
-%\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
-%\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
-%Then:
-%$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
-%and the last inclusion is an equality if the absolute value on $K$ is
-%discrete.
-%\end{enumerate}
-%\end{lem}
-%
-%\begin{proof}
-%Left to the reader.
-%\end{proof}
-%
-%The first part of Lemma \ref{lem:reductionball} asserts that the 
-%identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-%\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
-%$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
-%assumption: it is ``between the open unit ball and the closed unit 
-%ball''.
+\subsection{Ultrametric analysis}
+\label{ssec:ultrametric}
+
+Similarly to the real case, any finite dimensional normed $K$-vector 
+space is complete and all norms over such a space are equivalent 
+(\emph{i.e.} they define the same topology). Many usual Theorems 
+concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
+Theorem) remain true in the ultrametric setting.
+
+\begin{deftn} \label{deftn : diff}
+Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
+subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
+\emph{differentiable} at $v_0 \in U$ if there exists a 
+continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
+$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
+\subset U$ containing $v_0$ with:
+\[ 
+\forall v, w \in U_\varepsilon, \quad
+\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
+\leq \varepsilon \Vert v-w \Vert. 
+\]
+The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
+\end{deftn}
+
+One can derive all usual properties of the differential from this 
+definition: unicity, chain rules, \emph{etc}.
+
+The following Lemma shows that the notion of lattices behaves rather 
+well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
+\ref{lem:main}) that, under some extra hypothesis, the linearity 
+assumption can be drastically relaxed.} transformations.
+
+\begin{lem}
+\label{lem:morlat}
+Let $E$ and $F$ be two $K$-Banach and $f : 
+E \to F$ be a continuous \emph{surjective} linear map.
+If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
+\end{lem}
+
+\begin{proof}
+It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
+The fact that it is open is a direct consequence of the open mapping
+Theorem.
+\end{proof}
+
+To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
+on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
+the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
+\in H$. It follows from the fact that $H$ is open and bounded in $E$
+that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
+of $I_x$ under the absolute value map is bounded and contains a positive
+element. We can then define:
+$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
+It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
+$E$.
+
+\begin{lem} \label{lem:reductionball}
+\begin{enumerate}
+\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
+equivalent.
+\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
+\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
+Then:
+$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
+and the last inclusion is an equality if the absolute value on $K$ is
+discrete.
+\end{enumerate}
+\end{lem}
+
+\begin{proof}
+Left to the reader.
+\end{proof}
+
+The first part of Lemma \ref{lem:reductionball} asserts that the 
+identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
+\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
+$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
+assumption: it is ``between the open unit ball and the closed unit 
+ball''.
 
@@
 
+\begin{lem} \label{lem:main}
+Let $E$ and $F$ be two $K$-Banach and $f : U 
+\rightarrow F$ be a function defined on an open subset $U$ of $E$.
+We assume that $f$ is differentiable at some point $v_0 \subset 
+U$ and that the differential $f'(v_0)$ is surjective.
+
+Then, there exists a positive real number $\delta$ such that, for all $r 
+\in (0, \delta)$, the equality
+\begin{equation}
+\label{eq:main}
+f(v_0 + H) = f(v_0) + f'(v_0) (H).
+\end{equation}
+holds for each lattice $H$ such that $B^-_E(r) \subset H \subset
+B_E(r)$.
+\end{lem}
+
+\begin{rmk}
+\label{rmk:mainlemma}
+Given a lattice $H_0$ in $E$, the equality \eqref{eq:main} holds for $H 
+= \lambda H_0$ with $\lambda$ small enough. To prove this, it is enough 
+to apply Lemma \ref{lem:main} to the map $f \circ \iota_{H_0}$ where 
+$\iota_{H_0} : (E, \Vert \cdot \Vert_{H_0}) \to (E, \Vert \cdot \Vert)$ 
+is the ``identity map'' (which is an homeomorphism by Lemma 
+\ref{lem:reductionball}).
+\end{rmk}
+
+\begin{proof}
+We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
+the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
+C  f'(0) \cdot B_E(1)$. Let $\varepsilon>0$ be such that 
+$\varepsilon C < 1$. Let $U_\varepsilon \subset E$ be given by by definition \ref{deftn : diff} and the fact that $f$ is differentiable at $0$. We may assume $U_\varepsilon = 
+B_E(\delta)$ for some $\delta >0$.
+
+Let $r \in (0, \delta)$. Let $B^-_E(r) \subset H \subset
+B_E(r).$
+Let us show that $f$ maps surjectively $H$ onto $f'(0) (H)$.
+
+We first prove that $f(H) \subset f'(0) (H)$.
+Let $x \in H$. By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
+\varepsilon \Vert x \Vert $. Therefore, if $y=f(x)-f'(0) \cdot x,$ then $\Vert y \Vert \leq \varepsilon r$. As a consequence of the open mapping Theorem, $B_F(\varepsilon r) \subset 
+\varepsilon C  f'(0) \cdot B_E(r),$ and there exists $x_1 \in \varepsilon C  \cdot B_E(r)$ such that $f'(0) \cdot x_1 =y.$ Since $\varepsilon C <1,$  $x_1 \in B^-_E(r) \subset H.$ and finally, $f(x)= f'(0) \cdot (x-x_1) \in f'(0) (H)$.
+
+We shall now prove surjectivity. Let $y \in f'(0) (H)$.
+Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
+Since $f(x_0) \in f'(0) (H)$, we can define $z_0 \in E$ to be such that $f'(0) \cdot z_0 = y - f(x_0)$ and $\Vert z_0 \Vert \leq C \cdot \Vert y - f(x_0) \Vert$.
+We then define by induction two sequences $(x_n)$ and $(z_n)$ by
+\begin{itemize}
+\item $x_{n+1}=x_n+z_n$ and 
+\item $z_n \in E$ is such that
+$f'(0) \cdot z_n = y - f(x_n)$ and
+$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$.
+\end{itemize}
+By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y 
+= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y \Vert < \varepsilon 
+\Vert x_0 \Vert \leq \varepsilon r$. Thus $\Vert z_0 \Vert \leq 
+\varepsilon C \cdot r < r$. Thus, $z_0$ and $x_1=x_0+z_0$  lie in $B^-_E(r) 
+\subset H$. An easy induction involving an analoguous computation
+shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
+take their values in $H$.
+
+We notice that $y - f(x_{n+1}) = f(x_n) - f(x_{n+1}) + f'(0) \cdot z_n$.
+Since $z_n=x_{n+1}-x_n,$, if we use differentiability, we get:
+$$\Vert y  - f(x_{n+1}) \Vert 
+\leq \varepsilon \Vert z_n \Vert \leq \varepsilon C \cdot \Vert 
+y - f(x_n) \Vert,$$
+from what we deduce that $\Vert y - f(x_{n+1}) \Vert = 
+O(a^n)$ and $\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < 1$.
+This shows that $(x_n)$ is a Cauchy sequence and, since $E$ is complete, it converges.
+Because $H$ is closed, if $x$ denotes the limit of $(x_n)$, we have $x \in H.$ Yet, as a consequence of the differentiability, $f$ is continuous over $U_\varepsilon.$ Therefore, since $x \in H \subset U_\varepsilon,$ then  $y=f(x)$.
+The result is proven.
+\end{proof}
+
 \begin{deftn}
@@
 A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
-$v_0$ if the equality
-$$f(v_0 + H) = f(v_0) + f'(v_0) (H)$$
-holds.
+$v_0$ if
+$$f(v_0 + H) = f(v_0) + f'(v_0) (H).$$
 \end{deftn}
 
-\begin{lem} \label{lem:main}
-Let $E$ and $F$ be two $K$-Banach and $f : U 
-\rightarrow F$ be a function defined on an open subset $U$ of $E$.
-We assume that $f$ is differentiable at some point $v_0 \subset 
-U$ and that the differential $f'(v_0)$ is surjective.
-
-Then, for all $\rho \in (0, 1]$, there exists a positive real 
-number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ such that $B^-_E(\rho r) \subset H \subset B_E(r)$ is a first
-order lattice for $f$ at $v_0$.
-\end{lem}
-
-\begin{proof}
-We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
-the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
-C  f'(0)(B_E(1))$. Let $\varepsilon>0$ be such that 
-$\varepsilon C < \rho$. Let $U_\varepsilon \subset E$ given by Definition 
-\ref{deftn : diff} and the fact that $f$ is differentiable at $0$. We 
-may assume $U_\varepsilon = B_E(\delta)$ for some $\delta >0$.
-
-Let $r \in (0, \delta)$. Let $B^-_E(r) \subset H \subset B_E(r).$ Let us 
-show that $f$ maps surjectively $H$ onto $f'(0) (H)$. We first prove 
-that $f(H) \subset f'(0) (H)$. Let $x \in H$. By differentiability at 
-$0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
-Therefore, if $y=f(x)-f'(0)(x)$, then $\Vert y \Vert \leq 
-\varepsilon r$. By definition of $C$, we have $B_F(\varepsilon r) 
-\subset \varepsilon C f'(0) (B_E(r))$. Thus there exists $x_1 \in 
-\varepsilon C \cdot B_E(r)$ such that $f'(0) (x_1) =y$. Since 
-$\varepsilon C < \rho$, we get $x_1 \in B^-_E(\rho r) \subset H$ and 
-then $f(x)= f'(0) (x-x_1) \in f'(0) (H)$.
-
-We shall now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
-be such that $y = f'(0) (x_0)$. We define inductively two sequences 
-$(x_n)$ and $(z_n)$ as follows:
-\begin{itemize}
-\item $z_n$ is an element of $E$ satisfying $f'(0)(z_n) = y - 
-f(x_n)$ and $\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$ (such 
-an element exists by definition of $C$), and
-\item $x_{n+1}=x_n+z_n$.
-\end{itemize}
-For convenience, let us also define $x_{-1} = 0$. We claim that the 
-sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
-$H$. To establish this, it is enough to prove that, assuming that 
-$x_{n-1}$ and $x_n$ belongs to $H$, so do $z_n$ and $x_{n+1}$. Noticing
-that 
-\begin{equation}
-\label{eq:mainlemma}
-y - f(x_n) 
-= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n)
-= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
-\end{equation}
-we deduce using differentiability that
-$\Vert y - f(x_n) \Vert \leq \varepsilon \cdot \Vert x_n - x_{n-1}
-\Vert$.
-Since we are assuming that $x_{n-1}$ and $x_n$ lie in $H \subset
-B_E(r)$, we find $\Vert y - f(x_n) \Vert \leq \varepsilon r$. Thus
-$\Vert z_n \Vert \leq C \cdot \varepsilon r < \rho r$ and then
-$z_n \in H$. From the relation $x_{n+1} = x_n + z_n$, we finally
-deduce $x_{n+1} \in H$.
-
-Now using again Eq.~\eqref{eq:mainlemma} and differentiability at $0$,
-we get, for all positive integer $n$,
-$$\Vert y  - f(x_n) \Vert 
-\leq \varepsilon \cdot \Vert z_{n-1} \Vert \leq \varepsilon C \cdot \Vert 
-y - f(x_{n-1}) \Vert,$$
-from what we deduce that $\Vert y - f(x_n) \Vert = O(a^n)$ and 
-$\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < \rho \leq 1$. 
-This shows that $(x_n)$ is a Cauchy sequence and, since $E$ is complete, 
-it converges. Because $H$ is closed, if $x$ denotes the limit of 
-$(x_n)$, we have $x \in H$. Moreover, as a consequence of the 
-differentiability, $f$ is continuous over $U_\varepsilon$. Therefore, 
-since $x \in H \subset U_\varepsilon$, we get $y=f(x)$ and we are done.
-\end{proof}
-
-Let us conclude this paragraph by a remark on the assumption made in 
-Lemma \ref{lem:main} about the surjectivity of $f'(v_0)$. First let us 
-emphasize that this hypothesis is definitely necessary. Indeed, it would 
-implies otherwise that the image of $f$ is locally contained in a proper 
-sub vector space around each point where the differential of $f$ is not 
-surjective, which is certainly not true! Nevertheless, one can get 
-around this issue in many situations. Indeed, consider the situation of 
-Lemma \ref{lem:main} without assuming that $f'(v_0)$ is surjective. Now, 
-we pick a closed sub vector space $K$ of $F$ such that $K + f'(v_0)(E) = 
-F$. Then, denoting by $\pr_K$ the canonical projection of $F$ onto 
-$F/K$, the composite $\pr_K \circ f$ is certainly differentiable at 
-$v_0$ at its differential is surjective. Hence Lemma \ref{lem:main}
-applies and shows that
-\begin{equation}
-\label{eq:notsurjective}
-f(v_0 + H) \subset f(v_0) + f'(v_0)(H) + K
-\end{equation}
-for all lattices $H$ satisfying a certain condition. Now we can take
-the intersection of the right hand side of \eqref{eq:notsurjective} 
-over all possible $K$ and hope that we get this way a sharp affine upper 
-bound of $f(v_0 + H)$.
+By Lemma \ref{lem:main} and Remark \ref{rmk:mainlemma}, if $f$ is 
+differentiable at $v_0$ and $f'(v_0)$ is surjective, then for all 
+lattice $H$ in $E$, there exists a constant $\delta$ such that $aH$ 
+is a first order lattice for $f$ at $v_0$ for all $a \in B_E(r)$.
 
@@
 
-%Using Lemma \ref{lem:reductionball}, one can extend Proposition 
-%\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
-%a lattice in some $K$-Banach space $E$, we first define the two
-%following quantities:
-%\begin{itemize}
-%\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
-%\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
-%\end{itemize}
-%
-%\begin{cor}
-%Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
-%respectively and $f : U \to V$ be a locally analytic function. Let 
-%also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
-%Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
-%soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
-%constant depending only on $f$ and $v_0$ which can be determined 
-%explicitely from the Newton polygon of $f$ at $v_0$.
-%\end{cor}
-%
-%\begin{proof}
-%We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
-%\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-%\Vert)$ is induced by identity map.
-%\end{proof}
+Using Lemma \ref{lem:reductionball}, one can extend Proposition 
+\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
+a lattice in some $K$-Banach space $E$, we first define the two
+following quantities:
+\begin{itemize}
+\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
+\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
+\end{itemize}
+
+\begin{cor}
+Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
+respectively and $f : U \to V$ be a locally analytic function. Let 
+also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
+Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
+soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
+constant depending only on $f$ and $v_0$ which can be determined 
+explicitely from the Newton polygon of $f$ at $v_0$.
+\end{cor}
+
+\begin{proof}
+We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
+\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
+\Vert)$ is induced by identity map.
+\end{proof}
 
@@
 With above definitions, Lemma \ref{lem:main} extends readily to 
-manifolds. To state it the context of manifolds, we first need to define 
-a norm on the tangent space $T_x V$ (where $V$ is some $K$-variety and 
-$x$ is a point in $V$). There is actually in general no canonical choice 
-for this. Indeed, let us consider a $K$-manifold $V$ covered by charts 
-$U_i$'s ($i \in I$) which are open subset of $K$-Banach $E_i$'s. If $x$ 
-is a point in $V$, the tangent space $T_x V$ is by definition isomorphic 
-to $E_i$ for each index $i$ such that $x \in V_i$. A natural norm on 
-$T_x V$ is then the one obtained by pulling back the norm on $E_i$. 
-However, since the transition maps are not required to be isometries, 
-this norm do depend on the choice on $i$. They are nevertheless all
-equivalent because the transition maps are required to be continuous.
-
-In the next Lemma, we choose to endow $T_x V$ by one of the above norms.
+manifolds.
 
@@
 together with a point $x \in V$ such that $f'(x) : T_x V \to T_{f(x)} W$ 
-is surjective.
-
-Then for all $\rho \in (0,1]$, there exists $\delta \in (0,1)$ such 
-that for all $r \in (0, \delta)$, the equality
-$$f(x + O(H)) = f(x) + O(f'(x)(H))$$
-holds for all lattices $H$ in $T_x V$ such that $B^{\phantom -}_{T_x V}(\rho r)
-\subset H \subset B^-_{T_x V}(r)$.
+is surjective. Let $H$ be a precision datum at $x$. Then, there exists
+a positive real number $\delta$ such that, for all $\lambda \in 
+B_K(\delta)$, the equality
+$$f(x + \lambda H) = f(x) + \lambda \cdot f'(x)(H)$$
+holds.
 \end{lem}
@@
 \begin{rmk}
-The constant $\delta$ that appears in the above Lemma depends (up to
-some multiplicative constant) on the norm that we have chosen on $T_x V$. 
-However, once this norm is fixed, and assuming further that $V$ and $W$ 
-are locally analytic $K$-manifolds and the mapping $f$ is locally 
-analytic as well, the constant $\delta$ can be made explicit using the 
-method of \S \ref{ssec:locanalytic}.
+If $V$ and $W$ are locally analytic $K$-manifolds and the mapping $f$ is 
+locally analytic as well, then one can explicitely determine, using the 
+method of \S \ref{ssec:locanalytic}, a value $\delta$ for which the 
+above Lemma works.
 \end{rmk}

### 2014-02-15 padicprec.tex

--- 2014-02-14 padicprec.tex
@@
 $p$-adic fields and Laurent series fields.
+
+%\subsection{Ultrametric analysis}
+%\label{ssec:ultrametric}
+
+%Similarly to the real case, any finite dimensional normed $K$-vector 
+%space is complete and all norms over such a space are equivalent 
+%(\emph{i.e.} they define the same topology). Many usual Theorems 
+%concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
+%Theorem) remain true in the ultrametric setting.
+
+%\begin{deftn} \label{deftn : diff}
+%Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
+%subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
+%\emph{differentiable} at $v_0 \in U$ if there exists a 
+%continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
+%$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
+%\subset U$ containing $v_0$ with:
+%\[ 
+%\forall v, w \in U_\varepsilon, \quad
+%\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
+%\leq \varepsilon \Vert v-w \Vert. 
+%\]
+%The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
+%\end{deftn}
+
+%One can derive all usual properties of the differential from this 
+%definition: unicity, chain rules, \emph{etc}.
+
+%The following Lemma shows that the notion of lattices behaves rather 
+%well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
+%\ref{lem:main}) that, under some extra hypothesis, the linearity 
+%assumption can be drastically relaxed.} transformations.
+
+%\begin{lem}
+%\label{lem:morlat}
+%Let $E$ and $F$ be two $K$-Banach and $f : 
+%E \to F$ be a continuous \emph{surjective} linear map.
+%If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
+%\end{lem}
+
+%\begin{proof}
+%It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
+%The fact that it is open is a direct consequence of the open mapping
+%Theorem.
+%\end{proof}
+
+%To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
+%on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
+%the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
+%\in H$. It follows from the fact that $H$ is open and bounded in $E$
+%that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
+%of $I_x$ under the absolute value map is bounded and contains a positive
+%element. We can then define:
+%$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
+%It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
+%$E$.
+
+%\begin{lem} \label{lem:reductionball}
+%\begin{enumerate}
+%\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
+%equivalent.
+%\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
+%\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
+%Then:
+%$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
+%and the last inclusion is an equality if the absolute value on $K$ is
+%discrete.
+%\end{enumerate}
+%\end{lem}
+
+%\begin{proof}
+%Left to the reader.
+%\end{proof}
 

### 2014-02-15 padicprec.tex
Comments on the non-surjectivity of the differential.

@@
 from them is not).
+
 Let us illustrate the above discussion by an example: suppose that we 
@@
 S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
-over $K$ of size $n$) defined in the same way, our problem would 
-disappear. Of course the image of a general $f$ is rarely a sub vector 
-space of $F$ but it is often a sub-$K$-manifold of $F$. 
-% Using the results of Appendix
+over $K$ of size $n$) defined in the same way, our problem would have 
+disappeared. Of course the image of a general $f$ is rarely a sub vector 
+space of $F$ but it is often a sub-$K$-manifold of $F$. We can then use 
+the results of Appendix \ref{sec:manifold} to study $f$ viewed as a 
+function whose codomain is $f(U)$, understood that the differential of 
+it has now good probability to be surjective.
 
@@
 
-We conclude this Section by remarking that the two strategies designed 
+We conclude this Section by briefly remarking that the two strategies designed 
 above (\emph{i.e.} forward and backward computations) share some 

### 2014-02-15 padicprec.tex
todo updated

@@
 
-%\subsection{Ultrametric analysis}
-%\label{ssec:ultrametric}
-
-%Similarly to the real case, any finite dimensional normed $K$-vector 
-%space is complete and all norms over such a space are equivalent 
-%(\emph{i.e.} they define the same topology). Many usual Theorems 
-%concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
-%Theorem) remain true in the ultrametric setting.
-
-%\begin{deftn} \label{deftn : diff}
-%Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
-%subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
-%\emph{differentiable} at $v_0 \in U$ if there exists a 
-%continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
-%$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
-%\subset U$ containing $v_0$ with:
-%\[ 
-%\forall v, w \in U_\varepsilon, \quad
-%\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
-%\leq \varepsilon \Vert v-w \Vert. 
-%\]
-%The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
-%\end{deftn}
-
-%One can derive all usual properties of the differential from this 
-%definition: unicity, chain rules, \emph{etc}.
+\subsection{Ultrametric analysis}
+\label{ssec:ultrametric}
+
+Similarly to the real case, any finite dimensional normed $K$-vector 
+space is complete and all norms over such a space are equivalent 
+(\emph{i.e.} they define the same topology). Many usual Theorems 
+concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
+Theorem) remain true in the ultrametric setting.
+
+\begin{deftn} \label{deftn : diff}
+Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
+subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
+\emph{differentiable} at $v_0 \in U$ if there exists a 
+continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
+$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
+\subset U$ containing $v_0$ with:
+\[ 
+\forall v, w \in U_\varepsilon, \quad
+\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
+\leq \varepsilon \Vert v-w \Vert. 
+\]
+The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
+\end{deftn}
+
+One can derive all usual properties of the differential from this 
+definition: unicity, chain rules, \emph{etc}.
 
@@
 
-\subsection{Lattices}
+\subsubsection*{Lattices}
 
@@
 
-\begin{prop} \label{prop : maj Lambda}
+\begin{prop} \label{prop : maj Lambda} \label{prop:locanalytic}
 If $m_1$ is a real convex function above $\Lambda((f_n)_{n \geq 1})$.

### 2014-02-15 padicprec.tex
Working on Section 2

--- 2014-02-15 padicprec.tex
@@
 \item Kedlaya \cite{Kedlaya} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
-\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves using $p$-adic differential equations.
+\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
 \end{itemize}
@@
 A \emph{$K$-Banach space} is a complete normed $K$-vector space.
-
-\begin{deftn} % Separated definitions?
-Let $E$ be a $K$-Banach space.
-A \emph{lattice} in $E$ is an open bounded sub-$\OK$-module of $E$.
-\end{deftn}
+A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
 
@@
 
-\begin{deftn}
 Suppose $E$ is a $K$-Banach space and $I$ a set.
@@
 for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
-\end{deftn}
 Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
-% Lattice from Banach basis
+
+Given a basis $(x_i)_{i \in I}$ and a sequence $(r_i)_{i \in I}$ with $r_i \in \R_{>0}$, the sets
+\begin{align*}
+B_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert \le r_i \Big\}, \\
+B^-_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert < r_i \Big\}
+\end{align*}
+are lattices precisely when the $r_i$ are bounded.  If we have equipped $E$ with a distinguished basis
+then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.
+% Is every lattice of this shape for some Banach basis?
+
+\subsubsection*{Lattices and computers.}
+
+Suppose that $E \simeq K^d$ is finite dimesional.  Then if $H \subset E$
+is a lattice then there exist $a, b \in \Q_{>0}$ with
+\begin{equation}
+\label{eq:incllattice}
+B_K(a)^d \subset H \subset B_K(b)^d.
+\end{equation}
+Set $r = \frac a b$ and $R_r = \OK/B_K(r)$.  
+Then a lattice $H$ satisfying \eqref{eq:incllattice} is uniquely 
+determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
+R_r^d$.  Since $R \cap \OK$ is dense in $\OK$, elements of $R_r$ may be
+represented exactly.  Thus $H$ may be encoded as a 
+$(d \times d)$ matrix with coefficients in $R_r$.  For example, when
+$K = \Qp$ the ring $R_r$ is just $(\Z / p^n\Z)$ for $n = \lfloor - \log_p r \rfloor$.
 
@@
 
-Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.  We say that an element
-$x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
+Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
+\begin{deftn}
+\begin{itemize}
+\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
 \begin{equation} \label{eq:exact_elt}
@@
 \end{equation}
-The set of exact elements is dense in $E$.
-
-An approximate element is a pair $(x, H)$ where $x \in E$ is an exact element
-and $H$ is a lattice in $E$.  Such a pair represents an undetermined element of the coset
+
+\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
+and $H$ is a lattice in $E$.
+\end{itemize}
+\end{deftn}
+
+The pair $(x, H)$ represents an undetermined element of the coset
 $x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
@@
 we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
-
-Note that using an arbitrary lattice to represent the precision of an approximate element
+Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.
+
+This definition encapsulates the main practical suggestion of the paper with regards to representing
+vector spaces, matrices, polynomials and power series over $K$: one should \textbf{separate} the
+approximation from the precision.  In the rest of this section we discuss some of the benefits made
+possible by such a separation.
+
+Note first that using an arbitrary lattice to represent the precision of an approximate element
 can reduce precision loss when compared to storing the precision of each coefficient
@@
 
-Two approaches are possible. The first one consists in doing the 
-computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
-compute then $f(z,t)$ which gives the result. Proceeding this way, we 
-see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
-are consequently the same holds for the coordinates of the result.
-The second approach consists in remarking that $f \circ f$ maps
-$(x,y)$ to $(2x, 2y)$. Computing $f \circ f(x,y)$ using the latter
-formula, it is clear that the result we get is known up to precision
-$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p = 2$.
-As a conclusion, the second approach yields a more accurate result.
-In other words, computing step-by-step creates ``artificial'' losses 
-of precision.
-
-Let us recall first 
-that a datum involving elements of a complete ultrametric field $K$ 
-(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
-on a computed because they basically convey an infinite amount of 
-informations. That is why, instead of this datum itself, we generally 
-work with neighborhoods of it.
-
-In what follows, we shall consider neighborhoods of the form $x + H$ 
-where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
-in $E$. A very basic example is the approximate $p$-adic number $1 + 
-O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
-the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
-nevertheless emphasize that our definition is quite flexible and allows 
-many neighborhoods that we do not consider in general. For instance, an 
-``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
-of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
-P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
-$a_i \in \Q_p$ and relative integers $n_i$.
-
-In the context of computer science, we will often use the notation $x + 
-O(H)$ to refer to $x + H$.
+Write $(e_1, e_2)$ for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
+the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
+$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
+For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
+Storing precisions on each coordinate of a vector amounts to only allowing precisions of the form
+$B_E((r_i))$ with respect to the standard basis.  One would be forced to truncate the intermediate precision
+to $B_E(p^{-\min(m, n)})$, yielding an artificial loss of precision in the final result.
 
 \subsection{Separating precision from approximation}
+
+In addition to allowing for a more flexible representation of the precision of an element,
+the separation of precision from approximation has other benefits as well.  If the precision
+is encoded with the approximation, certain algorithms become unusable because of their
+numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication
 
@@
 instability.
-
-\subsubsection*{A first stupid example}
-
-Let
 
@@
 %ball''.
-
-\subsubsection*{Lattices and computers}
-
-Although elements of $K$ are not representable in thein entirety on a 
-computer, lattices in finite dimensional $K$-vector spaces are as soon 
-as the elements of the quotient $B_K(1)/B_K(r)$ are for all $r \in 
-(0,1)$. Indeed, given a lattice $L$ in $K^d$, there exists by definition 
-two positive rational numbers $a$ and $b$ such that 
-\begin{equation}
-\label{eq:incllattice}
-B_K(a)^d \subset L \subset B_K(b)^d.
-\end{equation}
-Moreover, a lattice $L$ satisfying \eqref{eq:incllattice} is uniquely 
-determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
-(B_K(1)/B_K(r))^d$ with $r = \frac a b$. Therefore, the data of $L$ can 
-be encoded by a $(d \times d)$ matrix with coefficients in $B_K(1)/
-B_K(r)$. Thus, if elements in the latter ring are representable on a 
-computer, so are the lattices $L$ for which Eq.~\eqref{eq:incllattice} 
-holds. Finally, an arbitrary lattice $L$ in $K^d$ can be represented
-by a triple $(a,b,M)$ where $a$ and $b$ satisfies \eqref{eq:incllattice}
-and $M$ is a $(d \times d)$ matrix over $B_K(1)/B_K(\frac a b)$ 
-representing the image of $L$ in $B_K(b)^d / B_K(a)^d$.
-
-We furthermore remark that if $K = \Q_p$ (for some prime number $p$), 
-the quotient $B_K(1)/B_K(r)$ is isomorphism to some $\Z/p^n\Z$ (with $n 
-= \lfloor - \log_p r \rfloor$) are therefore is representable on a 
-computer. Thus, the above discussion applies when $K = \Q_p$. Similarly 
-it also applies when $K$ is a finite extension of $\Q_p$ or a field of
-Laurent series with coefficients in a finite field.
 

### 2014-02-15 padicprec.tex
Switching to bibtex

@@
 \usepackage{amsmath,amssymb}
+\usepackage{amsrefs}
 \usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
@@
 \usepackage[algoruled,vlined,french,linesnumbered]{algorithm2e}
+
+\newcommand{\noopsort}[1]{}
 
@@
 \begin{itemize}
-\item Bostan et al. \cite{Bostan} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
-\item Gaudry et al. \cite{Gaudry} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
-\item Kedlaya \cite{Kedlaya} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
-\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
+\item Bostan et al. \cite{boston-gonzalez-perdry-schost:05a} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
+\item Gaudry et al. \cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
+\item Kedlaya \cite{kedlaya:01a} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
+\item Lercier and Survent \cite{lercier-sirvent:08a} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
 \end{itemize}
@@
 We recall some standard definitions from ultrametric analysis, referring to
-Schneider \cite{Schneider} for a complete exposition.
+Schneider \cite{schneider:11a} for a complete exposition.
 
@@
 $n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
-1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{Fomin-Zelevinsky}).
+1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{fomin-zelevinsky:02a}).
 From the latter formula, it follows directly that if $u_0$, $u_1$,
@@
 gives each individual entries of $L(M)$and $U(M)$ in a closed form
-(see \cite{Caruso}).
+(see \cite{caruso:12a	}).
 
@@
 coefficient is known up to precision $O(p^N)$, the two following
-results are proved in \cite{Caruso}:
+results are proved in \cite{caruso:12a}:
 \begin{itemize}
@@
 
-We refer to Schneider \cite{Schneider}, \S 6, for a definition of a locally analytic function.
+We refer to Schneider \cite{schneider:11a}, \S 6, for a definition of a locally analytic function.
 
@@
 polynomial with coefficients in $\Z$ in these indeterminates
-(see \cite{Fomin-Zelevinsky}).
+(see \cite{fomin-zelevinsky:02a}).
 
@@
 The theory of finite dimensional $K$-manifolds is presented for example 
-in \cite{Schneider}, parts 8 and 9. In this Section, we shall work with
+in \cite{schneider:11a}*{Ch. 8-9}. In this Section, we shall work with
 a slightly different notion of manifolds which allows also Banach vector 
@@
 above extends readily to convergent series (see also 
-\cite{Caruso-Lubicz}).
+\cite{caruso-lubicz:14a}).
 
@@
 \emph{homogeneous}\footnote{This restriction will be convenient for us 
-mainly because we are using the article \cite{Vaccon}, which deals only 
+mainly because we are using the article \cite{vaccon:14a}, which deals only 
 with homogeneous polynomials. It is however probably not essential.} 
 polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
+\ldots, g_t)$. It follows from Theorem 1.1 of \cite{vaccon:14a} that this 
 map is continuous at $(f_1, \ldots, f_s)$ provided that:
@@
 $f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{Vaccon} for more precisions).
+order (\emph{cf} \cite{vaccon:14a} for more precisions).
 \end{itemize}
@@
 
-\begin{thebibliography}{10}
-
-\bibitem{Bostan}
-  A.~Bostan, L.~Gonz\'alez-Vega, H.~Perdry, \'E.~Schost,
-  \emph{From Newton sums to coefficients: complexity issues in characteristic $p$},
-  Proceedings MEGA'05 (2005).
-
-\bibitem{Caruso}
-  X.~Caruso,
-  \emph{Random matrices over a DVR and LU factorization},
-  preprint (2012)
-
-\bibitem{Caruso-Lubicz}
-  X.~Caruso, D.~Lubicz
-  \emph{Linear Algebra over Zp[[u]] and related rings,} 
-  to appear at LMS J. Comp. and Math.
-
-\bibitem{Fomin-Zelevinsky}
-  S.~Fomin, A.~Zelevinsky
-  \emph{The Laurent phenomenon,}   Advances in Applied Mathematics 28: 119–144 (2002).
-
-\bibitem{Gaudry}
-  P.~Gaudry, T.~Houtmann, A.~Weng, C.~Ritzenthaler, D.~Kohel,
-  \emph{The $2$-adic CM method for genus $2$}, 
-  Asiacrypt 2006, vol. 4284 of Lecture Notes in Comput. Sci. (2006), 114--129
-
-\bibitem{Kedlaya}
-  K.~Kedlaya,
-  \emph{Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology}, 
-  J. Ramanujan Math. Soc. {\bf 16} (2001)
-
-\bibitem{Lercier}
-  R.~Lercier, T.~Sirvent,
-  \emph{On Elkies subgroups of $\ell$-torsion points in elliptic curves defined over a finite field},
-  J. Th\'eorie des Nombres de Bordeaux {\bf 20} (2008), 783--797
-
-\bibitem{Schneider}
-  P.~Schneider,
-  \emph{$p$-adic Lie Groups}, Springer (2011)
-
-\bibitem{Vaccon}
-  T.~Vaccon,
-  \emph{Matrix-F5 algorithms over finite-precision complete discrete 
-  valuation fields},
-  preprint (2014)
-\end{thebibliography}
+\bibliographystyle{plain}
+\bibliography{roebib/Biblio}
 

### 2014-02-16 padicprec.tex
Working on Section 2

@@
 sometimes underestimate the known precision of a result, as we will discuss in Section
-\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
+\ref{ssec:separation}.  Moreover, elements of $\Qp$ provide building blocks for generic
 implementations of polynomials, vector spaces, matrices and power series.  The practice
@@
 \subsection{Separating precision from approximation}
+\label{ssec:separation}
 
@@
 is encoded with the approximation, certain algorithms become unusable because of their
-numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication
+numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication \cite{karatsuba-ofman:62a}
+can needlessly lose precision when operating on polynomials with inexact coefficients.
+However, it works perfectly well on exact approximations, leaving the question of the precision of
+the product to be solved separately.  By separating the precision, more algorithms become available.
+
+There are also
 
@@
 \subsection{Step-by-step tracking is not sharp}
-\label{ssec:stepbystep}
 
@@
 tracking precision step-by-step.
-
-%\subsubsection*{Algorithms \emph{\`a la} Kedlaya}
 
@@
 artificial loss of precision if the above inclusion is an equality; it is 
-often not the case as shown in \S \ref{ssec:stepbystep}.
+often not the case as shown in \S \ref{ssec:separation}.
 
@@
 articles. More precisely, we study the SOMOS 4 sequence introduced in \S 
-\ref{ssec:stepbystep}. Making a crucial use of Lemma \ref{lem:main} and 
+\ref{ssec:separation}. Making a crucial use of Lemma \ref{lem:main} and 
 Proposition \ref{prop:locanalytic}, we design a \emph{stable} algorithm 
@@
 
-Recall (\emph{cf} \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
+Recall (\emph{cf} \S \ref{ssec:separation}) that a SOMOS 4 sequence is a 
 four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 

### 2014-02-16 padicprec.tex
Some examples: elliptic curves and grassmannians

@@
 \subsection{Examples}
-
-\todofor{David}{Elliptic curves and grassmannians}
+\label{ssec:exmanifold}
+
+We illustrate the theory developped above by some classical examples,
+namely elliptic curves and grassmannians.
+
+\subsubsection*{Elliptic curves}
+
+In this example, we assume for simplicity that $K$ has not caracteristic 
+$2$. Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
+\neq 0$ and let $E$ be the subset of $K^2$ consisting of the couples 
+$(x,y)$ satisfying the usual equation $y^2 = x^3 + a x + b$.
+Let also $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
+takes a couple $(x,y)$ to $x$ (resp. to $y$).
+
+\medskip
+
+We first assume that $a$ and $b$ are exactly known; typically the field 
+$K$ is $\Q_p$ and $a$ and $b$ are actual integers. For each point $P_0 = 
+(x_0, y_0)$ on $E$ except possibly a finite number of them, the map 
+$\pr_x$ define a diffeomorphism from an open subset containing $P_0$ to 
+an open subset of $K$. The same is true for $\pr_y$. Moreover around all 
+$P_0 \in E$, one of these projections satisfy the above condition and
+they both do generically. Hence the maps $\pr_x$ and $\pr_y$ defined an 
+various open subsets define together an atlas of $E$. This turns $E$ into 
+a $K$-manifold.
+
+Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
+charts. Proposition \ref{prop:independance} then tells us that a 
+precision datum on $x$ determined a precision datum on $y$ and \emph{vice 
+versa}. Indeed around $P_0$, we can write $y = \sqrt{x^3 + a x + b}$ 
+(with the correct determination of the square root) and find the 
+precision on $y$ from those on $x$ using Lemma \ref{lem:main}. We can go 
+in the other direction as well by writing $x$ locally as a function of 
+$y$. A precision datum at $P_0$ is then nothing but a precision datum
+on the coordinate $x$ or on the coordinate $y$, keeping in mind that 
+each of them determines the other. Viewing a precision datum at $P_0$ as 
+a lattice in the tangent space is a nice way to make it canonical but in 
+practice we can just particularize one coordinate and handle precision
+only on this coordinate.
+
+We conclude this example by showing a simple method to transform a
+precision datum on $x$ to a precision datum on $y$ and \emph{vice versa}.
+Differentiating the equation of the elliptic curve, we get:
+\begin{equation}
+\label{eq:diffec}
+2 y \cdot dy = (3 x^2 + a) \cdot dx.
+\end{equation}
+In the above $dx$ and $dy$ should be thought as a little perturbation of 
+$x$ and $y$ respectively. Eq.~\eqref{eq:diffec} then gives a linear 
+relation between the precision on $x$ (which is represented by $dx$) and 
+those on $y$ (which is represented by $dy$). This relation turns out to
+correspond exactly to the one which is given by Lemma \ref{lem:main}.
+
+\medskip
+
+Let us address now the case where $a$ and $b$ are themselves given with 
+finite precision. The elliptic curve $E$ is then not fully determined. So 
+we cannot consider it as a $K$-manifold and the above discussion does not 
+apply readily. Nevertheless, we can always consider the submanifold of 
+$K^4$ consisting of all tuples $(a,b,x,y)$ satisfying $y^2 = x^3 + a x + 
+b$. The projections on the hyperplanes $a = 0$, $b = 0$, $x = 0$ and $y = 
+0$ respectively define charts of this $K$-manifold. From this, we see 
+that a precision datum on a point of the ``not well determined'' elliptic 
+curve $E$ is a precision datum on a tuple of three variables among $a$, 
+$b$, $x$ and $y$.
+
+\subsubsection*{Grassmannians}
+
+Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
+Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
+of dimension $d$. It defines an algebraic variety over $K$ and hence
+\emph{a fortiori} a $K$-manifold.
+Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
+a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
+and two such matrices $M$ and $M'$ define the same vector space if there 
+exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
+find that we can always choose the above matrix $M$ in the particular
+shape:
+\begin{equation}
+\label{eq:grass}
+M = \begin{pmatrix} I_d & N \end{pmatrix} \cdot P
+\end{equation}
+where $I_d$ denotes the $(d \times d)$ identity matrix, $N \in M_{d, 
+n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
+such writings with the same $P$ necessarily coincide. Hence each 
+permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
+is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
+
+In other words, if $V$ is a subspace of $K^n$ of dimension $d$, we 
+represent it as a matrix $M$ of the shape \eqref{eq:grass} (using row 
+echelon) and a precision datum at $V$ is nothing but a precision datum on 
+the matrix $N$. If we choose another permutation matrix to represent $V$, 
+say $P'$, we end up with another matrix $N'$; the matrices $N$ and $N'$ 
+are then related by a simple relation. Differentiating it, we find a 
+formula for translating the precision datum expressed in the chart $U_P$ 
+to the same precision datum expressed in the chart $U_{P'}$.
+Of course, in practice, when we are doing computations on subspaces of
+$K^n$ (like sum or intersection), we represent the spaces in charts as
+above and perform all the calculations in these charts.
+
+\subsubsection*{Conclusion}
+
+The theory developped in this Appendix is nice because it provides an 
+abstract framework to deal with ultrametric precision in a quite general 
+setting.
+Nevertheless, as we have seen in the above examples, in practice we
+generally carry out all calculations in charts and do not really care
+about the ``real'' nature of the precision data, \emph{i.e.} lattices
+in tangent spaces.
 
@@
 are the field of $p$-adic numbers $\Q_p$ and the field of Laurent series 
-$k((X))$ over a base field $k$. The aim of this appendix is to compute 
-the differential of many usual operations on numbers, polynomials and 
-matrices. Surprisingly, we observe that all differentials we will 
-consider are rather easy to compute even if the underlying operation is 
-quite involved (\emph{e.g.} Gr\"obner basis).
+$k((X))$ over a base field $k$.
+
+We saw in the core of the article that the differential of an operation 
+encodes the intrinsic loss/gain of precision when performing this 
+operation. The aim of this appendix is to compute the differential of 
+many usual operations on numbers, polynomials and matrices. Surprisingly, 
+we observe that all differentials we will consider are rather easy to 
+compute even if the underlying operation is quite involved (\emph{e.g.} 
+Gr\"obner basis).
 
@@
 of $f$ is then also a manifold; this example then fits to the setting
-of \S \ref{sec:manifold} but not to those of \S \ref{sec:mainlemma}. 
-The generalization made in \S \ref{sec:manifold} is then needed.
+of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
+The generalization made in Appendix \ref{sec:manifold} is then needed.
 Anyway, we can differentiate $f$ by following the method we used for LU 
@@
 
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
-Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
-of dimension $d$. It defines an algebraic variety over $K$ and hence
-\emph{a fortiori} a $K$-manifold in the sense of \S \ref{ssec:manifold}.
-Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
-a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
-and two such matrices $M$ and $M'$ define the same vector space if there 
-exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
-find that we can always choose the above matrix $M$ in the particular
-shape:
-$$M = \begin{pmatrix} I_d & M' \end{pmatrix} \cdot P$$
-where $I_d$ denotes the $(d \times d)$ identity matrix, $M' \in M_{d, 
-n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
-such writings with the same $P$ necessarily coincide. Hence each 
-permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
-is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
+In computer science, vector spaces are generally represented as subspaces 
+of $K^n$ for some $n$. Hence they naturally appear as points on some 
+grassmannian $\Grass(n,d)$. These grassmannians are $K$-manifolds in the 
+sense of Appendix \ref{sec:manifold} as it is discussed in \S 
+\ref{ssec:exmanifold}. In what sequel, we use freely the formalism 
+introduced there.
 
@@
 
+Let $d$ and $n$ be two nonnegative integers such that $d \leq n$.
 We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 

### 2014-02-17 padicprec.tex
Paper anonimized
small fix: section -> appendix

@@
 \title{$p$-adic precision}
-\author{Xavier Caruso, David Roe \& Tristan Vaccon}
+%\author{Xavier Caruso, David Roe \& Tristan Vaccon}
 \date\today
@@
 classes of isomorphism of certain curves (which lie in various moduli spaces),
-\emph{etc.}  In this section we extend the formalism 
+\emph{etc.}  In this appendix we extend the formalism 
 developed in Section \ref{sec:mainlemma} to a more general setting:

### 2014-02-17 padicprec.tex
Work on loc. anal.
Improvement of the loc. anal. part.
Consistent with the latest version of the Main lemma.

Still need work on details like \overline{K} versus K, minoration,
majoration, how to obtain them, and make something pretty...

@@
 
-Let $r \in (0, \delta)$. Let $B^-_E(r) \subset H \subset B_E(r).$ Let us 
+Let $r \in (0, \delta)$. Let $B^-_E(\rho r) \subset H \subset B_E(r).$ Let us 
 show that $f$ maps surjectively $H$ onto $f'(0) (H)$. We first prove 
@@
 \end{itemize}
-For convenience, let us also define $x_{-1} = 0$. We claim that the 
+For convenience, let us also define $x_{-1} = 0$ and $z_{-1}=x_0.$ We claim that the 
 sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
@@
 
-We refer to Schneider \cite{schneider:11a}, \S 6, for a definition of a locally analytic function.
+We refer to Schneider \cite{schneider:11a}, \S 6, for a definition of a locally analytic function, and we extend it to any complete $K$ vector space :
+
+\begin{deftn} \label{deftn:analyticNP}
+Let $E$ and $F$ be two complete ultrametric $K$-Banach spaces. Let $U \subset E$ be an open set.
+A function $f \: : \: U \rightarrow F$ is said to be locally analytic if for any $x \in U,$ there exists some open subset $U_x$ of $E$ and some $n$-linear mappings $L_n \: : \: E^n \rightarrow F$ for any $n \geq 1,$ such that : for any $y \in U_x,$ the sum $\sum_{n \geq 1} L_n(y,\dots,y)$ converges in F, and $f(x+y)=f(x)+\sum_{n \geq 1} L_n(y,\dots,y).$
+\end{deftn}
+
+We then mostly study such a serie defined at $x$.
 
 \begin{deftn} \label{deftn :  serie and norm}
-Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie over $E,$ a complete normed $K$-vector space.
-We assume that for any $\lambda  \in K,$ $x \in E,$ $n \in \mathbb{N},$ $\vert f_n(\lambda x) \vert = \vert \lambda \vert \vert f_n (x) \vert.$
-We define $\vert f_n \vert = \sup_{x \in B_E(0,1)} \vert f_n(x) \vert,$ and we assume that all the $\vert f_n \vert $ are finite. We define $R((f_n)_n)$ to be the radius of convergence of the serie.
-
-We also define : \[\Lambda((f_n)_n) (v) = \log \sup_{x \in B(0,e^v)} \vert f(x) \vert.\]
-We define  $NP((f_n)_n)$ to be the Newton polygon of $f$, that is to say, the lower convex hull in $\mathbb{R}^2$ of the $(k, - \log \vert f_k \vert ),$ $k\geq 0.$ 
+Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie over mappings $E \rightarrow F,$ where $E$ and $F$ are complete normed $\overline{K}$-vector space.
+We assume that for any $\lambda  \in \overline{K},$ $x \in E,$ $n \in \mathbb{N},$ $\vert f_n(\lambda x) \vert = \vert \lambda \vert \vert f_n (x) \vert.$
+We define $\vert f_n \vert = \sup_{x \in B_E(0,1)} \vert f_n(x) \vert,$ and we assume that all the $\vert f_n \vert $ are finite. 
+We also define, for any $v \in \mathbb{R}$ such that this $\sup$ is finite : \[\Lambda(f) (v) = \log \sup_{x \in B_E(0,e^v)} \vert f(x) \vert,\]
+and $\Lambda(f) (v) =+\infty$ otherwise.
+We define  $NP(f)$ to be the Newton polygon of $f$, that is to say, the lower convex hull in $\mathbb{R}^2$ of the $(k, - \log \vert f_k \vert ),$ $k\geq 0.$ 
 \end{deftn}
@@
 \begin{prop} \label{prop : conjugate Newton}
-We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $NP((f_n)_n).$
-We denote by $NP((f_n)_n)^*$ the convex conjugate of $NP((f_n)_n).$ $NP((f_n)_n)^*$ is defined over $\left] -\infty,v_{\infty} \right[$ (at least)
+We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $NP(f).$
+We denote by $NP(f)^*$ the convex conjugate of $NP(f).$ $NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$ (at least)
 \end{prop}
 
-$\Lambda((f_n)_n)$ and $NP((f_n)_n)$ are connected :
+Then, $\Lambda(f)$ and $NP(f)$ are connected :
 
 \begin{prop} \label{prop : connexion sup Newton}
-We have $R((f_n)_n) = e^{v_{\infty}}.$
-Moreover, for any $v < \log \left( R((f_n)_n) \right),$ \[ \Lambda ((f_n)_n) (v) = NP((f_n)_n)^*(v).\]
+For any $v \in \left]-\infty, v_\infty \right[,$ \[ \Lambda (f) (v) = NP(f)^*(v).\]
 \end{prop}
 \begin{proof}
-Let $v \in \left] -\infty,v_{\infty} \right[,$ and let us assume that $v$ is not a slope of $NP((f_n)_n).$
-Let $u = NP((f_n)_n)^*(v).$ Then the intersection of the line defined by $y=vx-u$ and $NP((f_n)_n)$ is an extreme point of $NP((f_n)_n),$ say $(n, - \log ( \vert f_n \vert ))$ for some $n \geq 0,$ and $u = vn + \log \vert f_n \vert.$
-
-$v$ is not a slope of $NP((f_n)_n)$ and by convexity of $NP((f_n)_n)$, its difference quotients at $n$ are increasing, therefore, there exists some $c>0$ such that for any $m\geq 0,$ $vm-u \leq - \log \vert f_m \vert - c \vert n-m \vert.$ Since $vm-u=vm-vn-log \vert f_n \vert,$ we get $-vn-\log \vert f_n \vert + c \vert n-m \vert  \leq -vm-\log \vert f_m \vert .$
-
-Now, let $x \in  B(0,e^v),$ and $m \geq 0.$ Definition of $\vert f_m$ yields $ \vert f_m (x) \vert \leq \vert f_m \vert e^{vm}.$ Therefore \[\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn}< \vert f_n \vert e^{vn}.\]
-Therefore, the serie converges and by ultrametricity $ \vert f(x) \vert \leq \vert f_n \vert e^{vn},$ and $\Lambda ((f_l)_l)(v) \leq \log \left( \vert f_n \vert e^{vn} \right).$ Furthermore,  $R((f_n)_n)  \geq e^v.$
-
-Now, definition of $\vert f_n \vert$  provides $(x_i)_{i\geq 0}$ in $B(0,e^v)$ such that $\lim \vert f_n (x_i)= \vert f_n \vert e^{vn}.$
-
-Since for any $i,$ $\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn},$ then for $i$ large enough, $\vert f_m (x) \vert < \vert f_n (x) \vert $ and $\vert f(x) \vert = \vert f_n (x) \vert.$
-
-As a consequence, $\lim \vert f(x_i) \vert = \vert f_n \vert e^{vn},$ and  $\Lambda ((f_l)_l)(v) = \log ( \vert f_n \vert e^{vn}) = u= NP((f_n)_n)^*(v).$
-
-Therefore, for any $v \in \left] -\infty,v_{\infty} \right[,$ with $v$ not a slope of $NP((f_n)_n),$ we get $\Lambda ((f_l)_l)(v) =NP((f_n)_n)^*(v).$
-Since both $\Lambda ((f_l)_l)$ is increasing and $NP((f_n)_n)^*$ is continuous (since it is convex), we deduce that $R((f_n)_n) \geq  e^{v_{\infty}}$ and both functions coincide on $v \in \left] -\infty,v_{\infty} \right[,$
-
-One can show similarily that $R((f_n)_n) \leq  e^{v_{\infty}}.$
+Let $v \in \left] -\infty,v_{\infty} \right[,$ and let us assume that $v$ is not a slope of $NP(f).$
+Let $u = NP(f)^*(v).$ Then the intersection of the line defined by $y=vx-u$ and $NP(f)$ is an extreme point of $NP(f),$ say $(n, - \log ( \vert f_n \vert ))$ for some $n \geq 0,$ and $u = vn + \log \vert f_n \vert.$
+
+$v$ is not a slope of $NP(f)$ and by convexity of $NP(f)$, its difference quotients at $n$ are increasing, therefore, there exists some $c>0$ such that for any $m\geq 0,$ $vm-u \leq - \log \vert f_m \vert - c \vert n-m \vert.$ Since $vm-u=vm-vn-log \vert f_n \vert,$ we get $-vn-\log \vert f_n \vert + c \vert n-m \vert  \leq -vm-\log \vert f_m \vert .$
+
+Now, let $x \in  B_E(0,e^v),$ and $m \geq 0.$ Definition of $\vert f_m \vert$ yields $ \vert f_m (x) \vert \leq \vert f_m \vert e^{vm}.$ Therefore \[\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn}< \vert f_n \vert e^{vn}.\]
+Thus, the serie converges and by ultrametricity $ \vert f(x) \vert \leq \vert f_n \vert e^{vn},$ and $\Lambda (f)(v) \leq \log \left( \vert f_n \vert e^{vn} \right).$
+
+Now, definition of $\vert f_n \vert$  provides $(x_i)_{i\geq 0}$ in $B_E(0,e^v)$ such that $\lim \vert f_n (x_i) \vert= \vert f_n \vert e^{vn}.$
+
+Since for any $i,$ $\vert f_m (x_i) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn},$ then for $i$ large enough, $\vert f_m (x_i) \vert < \vert f_n (x_i) \vert $ and $\vert f(x_i) \vert = \vert f_n (x_i) \vert.$
+
+As a consequence, $\lim_i \vert f(x_i) \vert = \vert f_n \vert e^{vn},$ and  $\Lambda (f)(v) = \log ( \vert f_n \vert e^{vn}) = u= NP(f)^*(v).$
+
+Therefore, for any $v \in \left] -\infty,v_{\infty} \right[,$ with $v$ not a slope of $NP(f),$ we get $\Lambda (f)(v) =NP(f)^*(v).$
+Since $\Lambda (f)$ is increasing, $NP(f)^*$ is continuous (since it is convex), and $NP(f)$ has only finitely many slopes smaller than $v_\infty-\epsilon$, for any $\epsilon >0$, we deduce that both functions coincide on $v \in \left] -\infty,v_{\infty} \right[,$
 \end{proof}
 
-\begin{cor} Let $f(x)= \sum_{n \geq 0} f_n (x)$ and $g(x)= \sum_{n \geq 0} g_n (x)$ be two series as in definition \ref{deftn :  serie and norm}.
-Let $f(x)g(x)=\sum_{n \geq 0} \mu_n (x) $ and $f \circ g(x)=\sum_{n \geq 0} \tau_n (x) .$ Then :
-$\Lambda((f_n+g_n)_n) \leq \max (\Lambda((f_n)_n),\Lambda((g_n)_n))$
-$\Lambda((\mu_n)_n) \leq \Lambda((f_n)_n)+\Lambda((g_n)_n)$
-$\Lambda((\tau_n)_n) \leq \Lambda((f_n)_n) \circ \Lambda((g_n)_n)$
-\end{cor}
+\begin{lem} Let $f(x)= \sum_{n \geq 0} f_n (x)$ and $g(x)= \sum_{n \geq 0} g_n (x)$ be two series as in definition \ref{deftn :  serie and norm}. Then :
+\[\Lambda(f+g) \leq \max (\Lambda(f),\Lambda(g))\]
+\[\Lambda(f \times g) \leq \Lambda(f)+\Lambda(g)\]
+\[\Lambda(f \circ g) \leq \Lambda(f) \circ \Lambda(g)\]
+\end{lem}
+
+The following proposition will reveal crucial in order to minorate or majorate $NP(f),$  $NP(f)^*$ or $\Lambda(f),$ and is classical in convex analysis.
 
@@
 \begin{cor} \label{cor : trunc series}
-Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie as in definition \ref{deftn :  serie and norm}, $n_0 \geq 0$ and $g(x)= \sum_{n \geq n_0} f_n (x).$ Then $\Lambda((g_n)_n) \leq \Lambda((f_n)_n)$ and the slopes of $\Lambda((g_n)_n)$ are bigger than $n_0$.
-As a consequence, a minoration of $\Lambda((g_n)_n)$ can be obtain by extending $\Lambda((f_n)_n)(v)$ with a slope $n_0$ for $v \leq n_0$.
+Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie as in definition \ref{deftn :  serie and norm}, $n_0 \geq 0$ and $g(x)= \sum_{n \geq n_0} f_n (x).$ Then $\Lambda(g) \leq \Lambda(f)$ and the slopes of $\Lambda(g)$ are bigger than $n_0$.
+As a consequence, a minoration of $\Lambda(g)$ can be obtain by extending $\Lambda(f)(v)$ with a slope $n_0$ for $v \leq n_0$.
 \end{cor}
@@
 We assume $L_1$ is surjective.
-We define, for any $n \geq 1$, $f_n(x)=L_n(x,\dots,x)$ for any $x \in E.$
-Let $C$ be given by the Banach-Schauder theorem for $L_1$, $\varepsilon = \frac{1}{C}$. We provide three methods to find a sufficient $r$ for \ref{lem:main}.
-
-For the first one, we assume that we have know a function that majorates $\Lambda((f_n)_{n \geq 1})$.
+We define, the $\overline{L_k},$ $\overline{E}$ and $\overline{F}$ by extending scalars to $\overline{K}.$ We define for any $n \geq 1$, $f_n(x)=\overline{L_n}(x,\dots,x)$ for any $x \in \overline{E}.$
+Let $C$ be given by the open mapping theorem for $L_1$, $\rho \in (0, 1]$ and $\varepsilon < \frac{\rho}{C}$.
+We provide three methods to find a sufficient $\delta$ to define first order lattice as in lemma \ref{lem:main}.
+
+For the first one, we assume that we have know a function that majorates $\Lambda(\sum_{n \geq 1}f_n)$.
 
 \begin{prop} \label{prop : maj Lambda} \label{prop:locanalytic}
-If $m_1$ is a real convex function above $\Lambda((f_n)_{n \geq 1})$.
+If $m_1$ is a real convex function above $\Lambda(\sum_{n \geq 1}f_n)$.
 We define $m_2$ by extending $m_1$ with slope $2$.
-Let $r$ be the abscissa of the intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x.$ Then $e^r$ is enough for \ref{lem:main}.
+Let $\delta$ be the abscissa of the intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x.$ Then $e^\delta$ is enough to define first order lattices as in lemma  \ref{lem:main}.
 \end{prop}
 \begin{proof}
-$r$ is well-defined thanks to corollary \ref{cor : trunc series}, and $m_2 \geq \Lambda((f_n)_{n \geq 2}).$
-
-One only has to prove that for all $x \in B(0,e^r),$ $\vert \sum_{n\geq 2} f_n(x) \vert \leq \varepsilon \vert x \vert.$
-
-If $x \in B(0,e^r),$ $u = \log \vert x \vert,$ then $\Lambda((f_n)_{n \geq 2})(u) \geq -\log \varepsilon+ u.$
+$\delta$ is well-defined thanks to corollary \ref{cor : trunc series}, and $m_2 \geq \Lambda(\sum_{n \geq 2}f_n).$
+
+It is sufficient to prove that for all $x \in B_{\overline{E}}(0,e^\delta),$ $\vert \sum_{n\geq 2} f_n(x) \vert \leq \varepsilon \vert x \vert.$
+
+If $x \in B_{\overline{E}}(0,e^\delta),$ $u = \log \vert x \vert,$ then $\Lambda( \sum_{n \geq 2}f_n)(u) \geq -\log \varepsilon+ u.$
 With proposition \ref{prop : connexion sup Newton}, for any $m \geq 2,$ $-\log \vert f_m \vert +m u \geq -\log \varepsilon+ u.$
@@
 
-If we denote by $NP((M_{n+1})_{n \geq 1})$ the lower convex hull of the $(n, -\log M_{n+1})$ for $n \geq 1$, then $r$ such that $-\log \varepsilon=NP((M_n)_{n \geq 2})^*(r)$ is enough.
+If we denote by $NP((M_{n+1})_{n \geq 1})$ the lower convex hull of the $(n, -\log M_{n+1})$ for $n \geq 1$, then $\delta$ such that $-\log \varepsilon=NP((M_n)_{n \geq 2})^*(r)$ is enough to define first order lattices as in lemma  \ref{lem:main}.
 \end{prop}
 \begin{proof}
-Same idea as proposition \ref{prop : maj Lambda}. One can define $g_n$ with $g_n (x) =\frac{f_n(x)}{\vert x \vert}$ and $g_n(0)=0$ and work with $\Lambda ((g_n)_n).$
+Same idea as proposition \ref{prop : maj Lambda}. One can define $g_n$ with $g_n (x) =\frac{f_n(x)}{\vert x \vert}$ and $g_n(0)=0$ and work with $\Lambda ( \sum g_n).$
 \end{proof}
 
-We finally provide a simple method when working with finite dimension and $p$-adic fields.
+We finally provide a simple method when working with finite-dimensional vector spaces over $p$-adic fields.
 
 \begin{prop} \label{prop : locanalyticfinitedim}
-If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C$ and $\varepsilon$ are as above.
+If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above.
 It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}.

### 2014-02-18 padicprec.tex
Working on Section 2

--- 2014-02-16 padicprec.tex
@@
 
-There are also
-
-\todofor{David}{
-Explain here why it might be useful to work separately with 
-approximation on the one hand and precision on the other hand.
-}
-
-\subsection{Step-by-step tracking is not sharp}
-
-In this subsection, we illustrate by some examples that tracking
-$p$-adic precision step by step may lead to important numerical
-instability.
-
-\subsubsection*{The SOMOS 4 sequence}
-
-The SOMOS 4 sequence is a four-term inductive sequence satisfying:
-$$u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.$$
-In this example, we shall consider the situation where the initial
-terms $u_0$, $u_1$, $u_2$ and $u_3$ are \emph{invertible} $p$-adic 
-integers, \emph{i.e.} lie in $\Zp^\times$.
-
-Let us first examine what happens (regarding to the precision) when
-we are computing the sequence $(u_n)$ naively. The computation of
-$u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
-decreases the precision by a factor $p^{\val(u_n)}$ where $\val$
-stands for the $p$-adic valuation. Hence, if we assume that $u_0$,
-$u_1$, $u_2$ and $u_3$ are given with precision $O(p^N)$ then the
-naive computation returns the value of $u_n$ with precision
+In addition to allowing for more efficient computation of precision, storing the precision separately from the
+approximation makes it easier to determine the precision of the result when step-by-step precision tracking
+does not reflect the actual precision of the result.
+
+The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
+\[
+u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
+\]
+We shall consider the case where the initial
+terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
+
+Let us first examine how the absolute precision of $u_n$ varies with $n$
+if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
+The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
+decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
+step-by-step computation returns the value of $u_n$ with precision
 \begin{equation}
@@
 
-However, on the other hand, one can prove that the SOMOS 4 sequence 
-exhibits the \emph{Laurent phenomenon}. This means that, for all integer 
+On the other hand, one can prove that the SOMOS 4 sequence 
+exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}. This means that, for all integer 
 $n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
-1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{fomin-zelevinsky:02a}).
+1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
 From the latter formula, it follows directly that if $u_0$, $u_1$,
@@
 appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
-precision but some numerical instability related to the algorithm ---
-which includes the model precision --- we used to compute to SOMOS 
-sequence.
+precision but some numerical instability related to the algorithm used to compute $u_n$.
 
@@
 From the above discussion, one can easily derive a numerically stable 
-algorithm that computes the SOMOS 4 sequence: (1)~we compute the 
-Laurent polynomials $P_n$ using the induction formula (and working in 
-the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$) and (2)~we 
-evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
-However this algorithm is not efficient at all since computing the 
-$P_n$'s is very time-consuming for two reasons: first, it requires 
-division in a polynomial ring with $4$ variables and, second, the
+algorithm that computes the SOMOS 4 sequence: 
+\begin{enumerate}
+\item compute the Laurent polynomials $P_n$ using the induction formula (and working in 
+the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$)
+\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
+\end{itemize}
+However, computing the $P_n$'s is very time-consuming since it requires 
+division in a polynomial ring with $4$ variables and the
 size of the coefficients of $P_n$ explodes as $n$ grows.
 
-In \S \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
+In Section \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
 the SOMOS 4 sequence which turns out to be, at the same time, efficient 
@@
 \subsection{Examples}
-\label{ssec:exmanifold}
-
-We illustrate the theory developped above by some classical examples,
-namely elliptic curves and grassmannians.
-
-\subsubsection*{Elliptic curves}
-
-In this example, we assume for simplicity that $K$ has not caracteristic 
-$2$. Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
-\neq 0$ and let $E$ be the subset of $K^2$ consisting of the couples 
-$(x,y)$ satisfying the usual equation $y^2 = x^3 + a x + b$.
-Let also $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
-takes a couple $(x,y)$ to $x$ (resp. to $y$).
-
-\medskip
-
-We first assume that $a$ and $b$ are exactly known; typically the field 
-$K$ is $\Q_p$ and $a$ and $b$ are actual integers. For each point $P_0 = 
-(x_0, y_0)$ on $E$ except possibly a finite number of them, the map 
-$\pr_x$ define a diffeomorphism from an open subset containing $P_0$ to 
-an open subset of $K$. The same is true for $\pr_y$. Moreover around all 
-$P_0 \in E$, one of these projections satisfy the above condition and
-they both do generically. Hence the maps $\pr_x$ and $\pr_y$ defined an 
-various open subsets define together an atlas of $E$. This turns $E$ into 
-a $K$-manifold.
-
-Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
-charts. Proposition \ref{prop:independance} then tells us that a 
-precision datum on $x$ determined a precision datum on $y$ and \emph{vice 
-versa}. Indeed around $P_0$, we can write $y = \sqrt{x^3 + a x + b}$ 
-(with the correct determination of the square root) and find the 
-precision on $y$ from those on $x$ using Lemma \ref{lem:main}. We can go 
-in the other direction as well by writing $x$ locally as a function of 
-$y$. A precision datum at $P_0$ is then nothing but a precision datum
-on the coordinate $x$ or on the coordinate $y$, keeping in mind that 
-each of them determines the other. Viewing a precision datum at $P_0$ as 
-a lattice in the tangent space is a nice way to make it canonical but in 
-practice we can just particularize one coordinate and handle precision
-only on this coordinate.
-
-We conclude this example by showing a simple method to transform a
-precision datum on $x$ to a precision datum on $y$ and \emph{vice versa}.
-Differentiating the equation of the elliptic curve, we get:
-\begin{equation}
-\label{eq:diffec}
-2 y \cdot dy = (3 x^2 + a) \cdot dx.
-\end{equation}
-In the above $dx$ and $dy$ should be thought as a little perturbation of 
-$x$ and $y$ respectively. Eq.~\eqref{eq:diffec} then gives a linear 
-relation between the precision on $x$ (which is represented by $dx$) and 
-those on $y$ (which is represented by $dy$). This relation turns out to
-correspond exactly to the one which is given by Lemma \ref{lem:main}.
-
-\medskip
-
-Let us address now the case where $a$ and $b$ are themselves given with 
-finite precision. The elliptic curve $E$ is then not fully determined. So 
-we cannot consider it as a $K$-manifold and the above discussion does not 
-apply readily. Nevertheless, we can always consider the submanifold of 
-$K^4$ consisting of all tuples $(a,b,x,y)$ satisfying $y^2 = x^3 + a x + 
-b$. The projections on the hyperplanes $a = 0$, $b = 0$, $x = 0$ and $y = 
-0$ respectively define charts of this $K$-manifold. From this, we see 
-that a precision datum on a point of the ``not well determined'' elliptic 
-curve $E$ is a precision datum on a tuple of three variables among $a$, 
-$b$, $x$ and $y$.
-
-\subsubsection*{Grassmannians}
-
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
-Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
-of dimension $d$. It defines an algebraic variety over $K$ and hence
-\emph{a fortiori} a $K$-manifold.
-Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
-a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
-and two such matrices $M$ and $M'$ define the same vector space if there 
-exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
-find that we can always choose the above matrix $M$ in the particular
-shape:
-\begin{equation}
-\label{eq:grass}
-M = \begin{pmatrix} I_d & N \end{pmatrix} \cdot P
-\end{equation}
-where $I_d$ denotes the $(d \times d)$ identity matrix, $N \in M_{d, 
-n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
-such writings with the same $P$ necessarily coincide. Hence each 
-permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
-is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
-
-In other words, if $V$ is a subspace of $K^n$ of dimension $d$, we 
-represent it as a matrix $M$ of the shape \eqref{eq:grass} (using row 
-echelon) and a precision datum at $V$ is nothing but a precision datum on 
-the matrix $N$. If we choose another permutation matrix to represent $V$, 
-say $P'$, we end up with another matrix $N'$; the matrices $N$ and $N'$ 
-are then related by a simple relation. Differentiating it, we find a 
-formula for translating the precision datum expressed in the chart $U_P$ 
-to the same precision datum expressed in the chart $U_{P'}$.
-Of course, in practice, when we are doing computations on subspaces of
-$K^n$ (like sum or intersection), we represent the spaces in charts as
-above and perform all the calculations in these charts.
-
-\subsubsection*{Conclusion}
-
-The theory developped in this Appendix is nice because it provides an 
-abstract framework to deal with ultrametric precision in a quite general 
-setting.
-Nevertheless, as we have seen in the above examples, in practice we
-generally carry out all calculations in charts and do not really care
-about the ``real'' nature of the precision data, \emph{i.e.} lattices
-in tangent spaces.
+
+\todofor{David}{Elliptic curves and grassmannians}
 
@@
 are the field of $p$-adic numbers $\Q_p$ and the field of Laurent series 
-$k((X))$ over a base field $k$.
-
-We saw in the core of the article that the differential of an operation 
-encodes the intrinsic loss/gain of precision when performing this 
-operation. The aim of this appendix is to compute the differential of 
-many usual operations on numbers, polynomials and matrices. Surprisingly, 
-we observe that all differentials we will consider are rather easy to 
-compute even if the underlying operation is quite involved (\emph{e.g.} 
-Gr\"obner basis).
+$k((X))$ over a base field $k$. The aim of this appendix is to compute 
+the differential of many usual operations on numbers, polynomials and 
+matrices. Surprisingly, we observe that all differentials we will 
+consider are rather easy to compute even if the underlying operation is 
+quite involved (\emph{e.g.} Gr\"obner basis).
 
@@
 of $f$ is then also a manifold; this example then fits to the setting
-of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
-The generalization made in Appendix \ref{sec:manifold} is then needed.
+of \S \ref{sec:manifold} but not to those of \S \ref{sec:mainlemma}. 
+The generalization made in \S \ref{sec:manifold} is then needed.
 Anyway, we can differentiate $f$ by following the method we used for LU 
@@
 
-In computer science, vector spaces are generally represented as subspaces 
-of $K^n$ for some $n$. Hence they naturally appear as points on some 
-grassmannian $\Grass(n,d)$. These grassmannians are $K$-manifolds in the 
-sense of Appendix \ref{sec:manifold} as it is discussed in \S 
-\ref{ssec:exmanifold}. In what sequel, we use freely the formalism 
-introduced there.
+Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
+Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
+of dimension $d$. It defines an algebraic variety over $K$ and hence
+\emph{a fortiori} a $K$-manifold in the sense of \S \ref{ssec:manifold}.
+Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
+a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
+and two such matrices $M$ and $M'$ define the same vector space if there 
+exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
+find that we can always choose the above matrix $M$ in the particular
+shape:
+$$M = \begin{pmatrix} I_d & M' \end{pmatrix} \cdot P$$
+where $I_d$ denotes the $(d \times d)$ identity matrix, $M' \in M_{d, 
+n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
+such writings with the same $P$ necessarily coincide. Hence each 
+permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
+is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
 
@@
 
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$.
 We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 

### 2014-02-19 padicprec.tex
Working on Section 3.1

@@
 \newcommand{\noopsort}[1]{}
+\DeclareMathOperator{\NP}{NP}
 
@@
 \newcommand{\rp}{)\!)}
+\newcommand{\col}{\: : \:}
 
@@
 \end{enumerate}
-A \emph{$K$-Banach space} is a complete normed $K$-vector space.
+A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
+normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
 A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
@@
 On the other hand, one can prove that the SOMOS 4 sequence 
-exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}. This means that, for all integer 
+exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}: for all integer 
 $n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
@@
 \begin{enumerate}
-\item compute the Laurent polynomials $P_n$ using the induction formula (and working in 
-the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$)
+\item compute the Laurent polynomials $P_n$ using the recurrence in 
+the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$
 \item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
-\end{itemize}
+\end{enumerate}
 However, computing the $P_n$'s is very time-consuming since it requires 
@@
 \subsubsection*{LU factorization}
+
+% Frame better
 
@@
 
-The theory of $p$-adic precision we want to develop is mainly based on a 
-single lemma (Lemma \ref{lem:main}) of $p$-adic analysis. The aim of 
-this Section is to state and prove this lemma. Consequences will be 
-discussed after in Section \ref{sec:tracking}.
-
-From now, we generalize a bit the setting and work over a general 
-ultrametric complete field (see definition below). This includes 
-$p$-adic fields and Laurent series fields.
-
-\subsection{Ultrametric analysis}
-\label{ssec:ultrametric}
-
-Similarly to the real case, any finite dimensional normed $K$-vector 
-space is complete and all norms over such a space are equivalent 
-(\emph{i.e.} they define the same topology). Many usual Theorems 
-concerning Banach spaces (\emph{e.g.} open mapping Theorem, closed graph 
-Theorem) remain true in the ultrametric setting.
+Our theory of $p$-adic precision rests upon a 
+lemma in $p$-adic analysis: Lemma \ref{lem:main}.  In this section
+we develop the theory surrounding this result, and proceed to practical
+consequences in Section \ref{sec:tracking}.
+
+\subsection{Images of lattices under differentiable functions}
+
+Our goal in this section is to relate the image of a lattice under a differentiable map
+to its image under the derivative.
 
@@
 $\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
-\subset U$ containing $v_0$ with:
+\subset U$ containing $v_0$ with
 \[ 
-\forall v, w \in U_\varepsilon, \quad
 \Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
@@
 \]
-The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
+for all $v, w \in U_\varepsilon$.  The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
 \end{deftn}
@@
 definition: unicity, chain rules, \emph{etc}.
-
-%The following Lemma shows that the notion of lattices behaves rather 
-%well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
-%\ref{lem:main}) that, under some extra hypothesis, the linearity 
-%assumption can be drastically relaxed.} transformations.
-
-%\begin{lem}
-%\label{lem:morlat}
-%Let $E$ and $F$ be two $K$-Banach and $f : 
-%E \to F$ be a continuous \emph{surjective} linear map.
-%If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
-%\end{lem}
-
-%\begin{proof}
-%It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
-%The fact that it is open is a direct consequence of the open mapping
-%Theorem.
-%\end{proof}
-
-%To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
-%on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
-%the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
-%\in H$. It follows from the fact that $H$ is open and bounded in $E$
-%that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
-%of $I_x$ under the absolute value map is bounded and contains a positive
-%element. We can then define:
-%$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
-%It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
-%$E$.
-
-%\begin{lem} \label{lem:reductionball}
-%\begin{enumerate}
-%\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
-%equivalent.
-%\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
-%\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
-%Then:
-%$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
-%and the last inclusion is an equality if the absolute value on $K$ is
-%discrete.
-%\end{enumerate}
-%\end{lem}
-
-%\begin{proof}
-%Left to the reader.
-%\end{proof}
-
-\subsubsection*{Lattices}
 
@@
 
-\subsection{Images of lattices under differentiable functions}
-
 \begin{deftn}
 \label{def:firstorder}
-Let $E$ and $F$ be two $K$-Banach and $f : U \rightarrow F$ be a 
-function defined on an open subset $U$ of $E$. Let also $v_0$ be a 
+Let $E$ and $F$ be two $K$-Banach spaces, $f : U \rightarrow F$ be a 
+function defined on an open subset $U$ of $E$ and $v_0$ be a 
 point in $U$.
@@
 We emphasize that we require an equality in Eq.~\eqref{eq:firstorder}, 
-and not just an inclusion! It will be very important for the 
-applications we have in mind.
-Here comes our main Lemma.
+and not just an inclusion! 
+With this definition in hand, we are able to state our main lemma.
 
 \begin{lem} \label{lem:main}
-Let $E$ and $F$ be two $K$-Banach and $f : U 
+Let $E$ and $F$ be two $K$-Banach spaces and $f : U 
 \rightarrow F$ be a function defined on an open subset $U$ of $E$.
-We assume that $f$ is differentiable at some point $v_0 \subset 
+We assume that $f$ is differentiable at some point $v_0 \in 
 U$ and that the differential $f'(v_0)$ is surjective.
@@
 
-\begin{proof}
-We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, by 
-the open mapping Theorem, there exists $C>0$ such that $B_F(1) \subset 
-C  f'(0)(B_E(1))$. Let $\varepsilon>0$ be such that 
-$\varepsilon C < \rho$. Let $U_\varepsilon \subset E$ given by Definition 
-\ref{deftn : diff} and the fact that $f$ is differentiable at $0$. We 
-may assume $U_\varepsilon = B_E(\delta)$ for some $\delta >0$.
-
-Let $r \in (0, \delta)$. Let $B^-_E(\rho r) \subset H \subset B_E(r).$ Let us 
-show that $f$ maps surjectively $H$ onto $f'(0) (H)$. We first prove 
-that $f(H) \subset f'(0) (H)$. Let $x \in H$. By differentiability at 
+\begin{proof} % Call it the Banach-Schauder theorem?
+We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 
+\[
+B_F(1) \subset C  f'(0)(B_E(1)).
+\]
+Let $\varepsilon>0$ be such that 
+$\varepsilon C < \rho$, and choose $U_\varepsilon \subset E$ as in Definition 
+\ref{deftn : diff}.  We may assume $U_\varepsilon = B_E(\delta)$ for some $\delta >0$.
+
+Let $r \in (0, \delta)$ and suppose $H$ is a lattice with $B^-_E(\rho r) \subset H \subset B_E(r).$ 
+We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
+that $f(H) \subset f'(0) (H)$. Suppose $x \in H$; we first show that $f(x) \in f'(0)(H)$.  By differentiability at 
 $0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
-Therefore, if $y=f(x)-f'(0)(x)$, then $\Vert y \Vert \leq 
-\varepsilon r$. By definition of $C$, we have $B_F(\varepsilon r) 
-\subset \varepsilon C f'(0) (B_E(r))$. Thus there exists $x_1 \in 
-\varepsilon C \cdot B_E(r)$ such that $f'(0) (x_1) =y$. Since 
-$\varepsilon C < \rho$, we get $x_1 \in B^-_E(\rho r) \subset H$ and 
-then $f(x)= f'(0) (x-x_1) \in f'(0) (H)$.
+Setting $y=f(x)-f'(0)(x)$, we have $\Vert y \Vert \leq 
+\varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r) 
+\subset \varepsilon C f'(0) (B_E(r))$. Thus there exists $x' \in 
+\varepsilon C \cdot B_E(r)$ such that $f'(0) (x') =y$. Since 
+$\varepsilon C < \rho$, we get $x' \in B^-_E(\rho r) \subset H$ and 
+then $f(x)= f'(0) (x-x') \in f'(0) (H)$.
 
 We shall now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
-be such that $y = f'(0) (x_0)$. We define inductively two sequences 
+be such that $y = f'(0) (x_0)$. We inductively define two sequences 
 $(x_n)$ and $(z_n)$ as follows:
@@
 sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
-$H$. To establish this, it is enough to prove that, assuming that 
-$x_{n-1}$ and $x_n$ belongs to $H$, so do $z_n$ and $x_{n+1}$. Noticing
+$H$. We do so by induction, assuming that $x_{n-1}$ and $x_n$ belong to $H$
+and showing that $z_n$ and $x_{n+1}$ do as well. Noticing
 that 
@@
 \label{eq:mainlemma}
-y - f(x_n) 
-= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n)
-= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
+\begin{aligned}
+y - f(x_n) &= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n) \\
+&= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
+\end{aligned}
 \end{equation}
@@
 
-Now using again Eq.~\eqref{eq:mainlemma} and differentiability at $0$,
-we get, for all positive integer $n$,
+Using Eq.~\eqref{eq:mainlemma} and differentiability at $0$ once more,
+we get
 $$\Vert y  - f(x_n) \Vert 
@@
 y - f(x_{n-1}) \Vert,$$
-from what we deduce that $\Vert y - f(x_n) \Vert = O(a^n)$ and 
+for all positive integer $n$.  We therefore deduce that $\Vert y - f(x_n) \Vert = O(a^n)$ and 
 $\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < \rho \leq 1$. 
-This shows that $(x_n)$ is a Cauchy sequence and, since $E$ is complete, 
-it converges. Because $H$ is closed, if $x$ denotes the limit of 
-$(x_n)$, we have $x \in H$. Moreover, as a consequence of the 
-differentiability, $f$ is continuous over $U_\varepsilon$. Therefore, 
-since $x \in H \subset U_\varepsilon$, we get $y=f(x)$ and we are done.
+These conditions show that $(x_n)$ is a Cauchy sequence, which converges since $E$ is complete.
+Write $x$ for the limit of the $x_n$; we have $x \in H$ because $H$ is closed.
+Moreover, $f$ is continuous on $H \subseteq U_\varepsilon$ since it is differentiable and thus $y=f(x)$.
 \end{proof}
 
-Let us conclude this paragraph by a remark on the assumption made in 
-Lemma \ref{lem:main} about the surjectivity of $f'(v_0)$. First let us 
-emphasize that this hypothesis is definitely necessary. Indeed, it would 
-implies otherwise that the image of $f$ is locally contained in a proper 
-sub vector space around each point where the differential of $f$ is not 
-surjective, which is certainly not true! Nevertheless, one can get 
-around this issue in many situations. Indeed, consider the situation of 
-Lemma \ref{lem:main} without assuming that $f'(v_0)$ is surjective. Now, 
-we pick a closed sub vector space $K$ of $F$ such that $K + f'(v_0)(E) = 
-F$. Then, denoting by $\pr_K$ the canonical projection of $F$ onto 
-$F/K$, the composite $\pr_K \circ f$ is certainly differentiable at 
-$v_0$ at its differential is surjective. Hence Lemma \ref{lem:main}
-applies and shows that
+We end this section with a remark on the surjectivity of $f'(v_0)$ assumed in
+Lemma \ref{lem:main}. First, let us 
+emphasize that this hypothesis is definitely necessary. Indeed, the lemma would 
+otherwise imply that the image of $f$ is locally contained in a proper 
+sub-vector-space around each point where the differential of $f$ is not 
+surjective, which is certainly not true! Nevertheless, one can use lemma \ref{lem:main}
+to prove a weaker result in the context that $f'(v_0)$ is not surjective.
+To do so, choose a closed sub-vector-space $W$ of $F$ such that $W + f'(v_0)(E) = 
+F$. Denoting by $\pr_W$ the canonical projection of $F$ onto 
+$F/W$, the composite $\pr_W \circ f$ is differentiable at 
+$v_0$ with surjective differential. For a given lattice $H$, there will be various choices of $W$ to which Lemma \ref{lem:main}
+applies.  For each such $W$,
 \begin{equation}
 \label{eq:notsurjective}
-f(v_0 + H) \subset f(v_0) + f'(v_0)(H) + K
+f(v_0 + H) \subset f(v_0) + f'(v_0)(H) + W;
 \end{equation}
-for all lattices $H$ satisfying a certain condition. Now we can take
-the intersection of the right hand side of \eqref{eq:notsurjective} 
-over all possible $K$ and hope that we get this way a sharp affine upper 
-bound of $f(v_0 + H)$.
+taking the intersection of the right hand side
+over all possible $W$ yields an upper bound on $f(v_0 + H)$.
 
@@
 Before discussing applications of Lemma \ref{lem:main} to ultrametric 
-precision, we would like to describe a little bit more restrictive 
-setting where the constant $\delta$ appearing in this Lemma can be 
-computed explicitly.
-
-Indeed, when $f$ is a locally analytic function, we prove that we can recover $\delta$ from a Newton polygon defined with the Taylor series expansion of $f.$
-
-We refer to Schneider \cite{schneider:11a}, \S 6, for a definition of a locally analytic function, and we extend it to any complete $K$ vector space :
+precision, we discuss the case that $f$ is locally analytic.  In this context
+we prove that we can recover the constant $\delta$ appearing in the lemma
+from the Newton polygon determined by the Taylor series expansion of $f$.
+
+Schneider \cite{schneider:11a}*{\S 6} defines locally analytic functions on finite dimensional vector spaces.
+His definition extends easily to $K$-Banach spaces:
 
 \begin{deftn} \label{deftn:analyticNP}
-Let $E$ and $F$ be two complete ultrametric $K$-Banach spaces. Let $U \subset E$ be an open set.
-A function $f \: : \: U \rightarrow F$ is said to be locally analytic if for any $x \in U,$ there exists some open subset $U_x$ of $E$ and some $n$-linear mappings $L_n \: : \: E^n \rightarrow F$ for any $n \geq 1,$ such that : for any $y \in U_x,$ the sum $\sum_{n \geq 1} L_n(y,\dots,y)$ converges in F, and $f(x+y)=f(x)+\sum_{n \geq 1} L_n(y,\dots,y).$
+Let $E$ and $F$ be $K$-Banach spaces, and let $U \subset E$ be an open set.
+A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} if for any $x \in U,$ there exists some open subset $U_x$ of $E$ and some $n$-linear mappings $L_n \: : \: E^n \rightarrow F$ for any $n \geq 1,$ such that : for any $y \in U_x,$ the sum $\sum_{n \geq 1} L_n(y,\dots,y)$ converges in F, and $f(x+y)=f(x)+\sum_{n \geq 1} L_n(y,\dots,y).$
 \end{deftn}
@@
 and $\Lambda(f) (v) =+\infty$ otherwise.
-We define  $NP(f)$ to be the Newton polygon of $f$, that is to say, the lower convex hull in $\mathbb{R}^2$ of the $(k, - \log \vert f_k \vert ),$ $k\geq 0.$ 
+We define  $\NP(f)$ to be the Newton polygon of $f$, that is to say, the lower convex hull in $\mathbb{R}^2$ of the $(k, - \log \vert f_k \vert ),$ $k\geq 0.$ 
 \end{deftn}
@@
 \begin{prop} \label{prop : conjugate Newton}
-We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $NP(f).$
-We denote by $NP(f)^*$ the convex conjugate of $NP(f).$ $NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$ (at least)
+We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $\NP(f).$
+We denote by $\NP(f)^*$ the convex conjugate of $\NP(f).$ $\NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$ (at least)
 \end{prop}
 
-Then, $\Lambda(f)$ and $NP(f)$ are connected :
+Then, $\Lambda(f)$ and $\NP(f)$ are connected :
 
 \begin{prop} \label{prop : connexion sup Newton}
-For any $v \in \left]-\infty, v_\infty \right[,$ \[ \Lambda (f) (v) = NP(f)^*(v).\]
+For any $v \in \left]-\infty, v_\infty \right[,$ \[ \Lambda (f) (v) = \NP(f)^*(v).\]
 \end{prop}
 \begin{proof}
-Let $v \in \left] -\infty,v_{\infty} \right[,$ and let us assume that $v$ is not a slope of $NP(f).$
-Let $u = NP(f)^*(v).$ Then the intersection of the line defined by $y=vx-u$ and $NP(f)$ is an extreme point of $NP(f),$ say $(n, - \log ( \vert f_n \vert ))$ for some $n \geq 0,$ and $u = vn + \log \vert f_n \vert.$
-
-$v$ is not a slope of $NP(f)$ and by convexity of $NP(f)$, its difference quotients at $n$ are increasing, therefore, there exists some $c>0$ such that for any $m\geq 0,$ $vm-u \leq - \log \vert f_m \vert - c \vert n-m \vert.$ Since $vm-u=vm-vn-log \vert f_n \vert,$ we get $-vn-\log \vert f_n \vert + c \vert n-m \vert  \leq -vm-\log \vert f_m \vert .$
+Let $v \in \left] -\infty,v_{\infty} \right[,$ and let us assume that $v$ is not a slope of $\NP(f).$
+Let $u = \NP(f)^*(v).$ Then the intersection of the line defined by $y=vx-u$ and $\NP(f)$ is an extreme point of $\NP(f),$ say $(n, - \log ( \vert f_n \vert ))$ for some $n \geq 0,$ and $u = vn + \log \vert f_n \vert.$
+
+$v$ is not a slope of $\NP(f)$ and by convexity of $\NP(f)$, its difference quotients at $n$ are increasing, therefore, there exists some $c>0$ such that for any $m\geq 0,$ $vm-u \leq - \log \vert f_m \vert - c \vert n-m \vert.$ Since $vm-u=vm-vn-log \vert f_n \vert,$ we get $-vn-\log \vert f_n \vert + c \vert n-m \vert  \leq -vm-\log \vert f_m \vert .$
 
@@
 
-As a consequence, $\lim_i \vert f(x_i) \vert = \vert f_n \vert e^{vn},$ and  $\Lambda (f)(v) = \log ( \vert f_n \vert e^{vn}) = u= NP(f)^*(v).$
-
-Therefore, for any $v \in \left] -\infty,v_{\infty} \right[,$ with $v$ not a slope of $NP(f),$ we get $\Lambda (f)(v) =NP(f)^*(v).$
-Since $\Lambda (f)$ is increasing, $NP(f)^*$ is continuous (since it is convex), and $NP(f)$ has only finitely many slopes smaller than $v_\infty-\epsilon$, for any $\epsilon >0$, we deduce that both functions coincide on $v \in \left] -\infty,v_{\infty} \right[,$
+As a consequence, $\lim_i \vert f(x_i) \vert = \vert f_n \vert e^{vn},$ and  $\Lambda (f)(v) = \log ( \vert f_n \vert e^{vn}) = u= \NP(f)^*(v).$
+
+Therefore, for any $v \in \left] -\infty,v_{\infty} \right[,$ with $v$ not a slope of $\NP(f),$ we get $\Lambda (f)(v) =\NP(f)^*(v).$
+Since $\Lambda (f)$ is increasing, $\NP(f)^*$ is continuous (since it is convex), and $\NP(f)$ has only finitely many slopes smaller than $v_\infty-\epsilon$, for any $\epsilon >0$, we deduce that both functions coincide on $v \in \left] -\infty,v_{\infty} \right[,$
 \end{proof}
@@
 
-The following proposition will reveal crucial in order to minorate or majorate $NP(f),$  $NP(f)^*$ or $\Lambda(f),$ and is classical in convex analysis.
+The following proposition will reveal crucial in order to minorate or majorate $\NP(f),$  $\NP(f)^*$ or $\Lambda(f),$ and is classical in convex analysis.
 
@@
 
-If we denote by $NP((M_{n+1})_{n \geq 1})$ the lower convex hull of the $(n, -\log M_{n+1})$ for $n \geq 1$, then $\delta$ such that $-\log \varepsilon=NP((M_n)_{n \geq 2})^*(r)$ is enough to define first order lattices as in lemma  \ref{lem:main}.
+If we denote by $\NP((M_{n+1})_{n \geq 1})$ the lower convex hull of the $(n, -\log M_{n+1})$ for $n \geq 1$, then $\delta$ such that $-\log \varepsilon=\NP((M_n)_{n \geq 2})^*(r)$ is enough to define first order lattices as in lemma  \ref{lem:main}.
 \end{prop}
@@
 If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above.
-It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}.
+It is enough to work with the Newton polygon $\NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $\NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}.
 \end{prop}
@@
 
-In what follows, the letter $K$ always refers to an ultrametric field 
-(\emph{cf} \S \ref{ssec:ultrametric}). The absolute value over $K$ is 
-denoted by $|\cdot|$.
-
 \subsection{Differentiable $K$-manifolds}
@@
 (where $I$ is some set) and, for all $i \in I$, an homeomorphism 
-$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach 
+$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach space
 $E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
@@
 finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
-$K$-Banach themselves.
+$K$-Banach spaces themselves.
 
@@
 for this. Indeed, let us consider a $K$-manifold $V$ covered by charts 
-$U_i$'s ($i \in I$) which are open subset of $K$-Banach $E_i$'s. If $x$ 
+$U_i$'s ($i \in I$) which are open subset of $K$-Banach spaces $E_i$'s. If $x$ 
 is a point in $V$, the tangent space $T_x V$ is by definition isomorphic 

### 2014-02-19 padicprec.tex
Working on Section 3.2

@@
 Let $E$ and $F$ be $K$-Banach spaces, and let $U \subset E$ be an open set.
-A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} if for any $x \in U,$ there exists some open subset $U_x$ of $E$ and some $n$-linear mappings $L_n \: : \: E^n \rightarrow F$ for any $n \geq 1,$ such that : for any $y \in U_x,$ the sum $\sum_{n \geq 1} L_n(y,\dots,y)$ converges in F, and $f(x+y)=f(x)+\sum_{n \geq 1} L_n(y,\dots,y).$
+A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} if one may find
+an open set $U_x \subset E$ for each $x \in U$ satisfying the following condition.
+For each $n \ge 1$ there exists an $n$-linear mapping
+$L_n \col E^n \rightarrow F$ so that 
+\begin{enumerate}
+\item the sum $\sum_{n \geq 1} L_n(y,\ldots,y)$ converges in F for every $y \in U_x,$ 
+\item $f(x+y)=f(x)+\sum_{n \geq 1} L_n(y,\dots,y).$
+\end{enumerate}
 \end{deftn}
@@
 \begin{deftn} \label{deftn :  serie and norm}
-Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie over mappings $E \rightarrow F,$ where $E$ and $F$ are complete normed $\overline{K}$-vector space.
-We assume that for any $\lambda  \in \overline{K},$ $x \in E,$ $n \in \mathbb{N},$ $\vert f_n(\lambda x) \vert = \vert \lambda \vert \vert f_n (x) \vert.$
-We define $\vert f_n \vert = \sup_{x \in B_E(0,1)} \vert f_n(x) \vert,$ and we assume that all the $\vert f_n \vert $ are finite. 
-We also define, for any $v \in \mathbb{R}$ such that this $\sup$ is finite : \[\Lambda(f) (v) = \log \sup_{x \in B_E(0,e^v)} \vert f(x) \vert,\]
-and $\Lambda(f) (v) =+\infty$ otherwise.
-We define  $\NP(f)$ to be the Newton polygon of $f$, that is to say, the lower convex hull in $\mathbb{R}^2$ of the $(k, - \log \vert f_k \vert ),$ $k\geq 0.$ 
+Suppose that $E$ and $F$ are $\overline{K}$-Banach spaces and $f_n \col E \to F$ are maps satisfying
+$\lvert f_n(\lambda x) \rvert = \lvert \lambda \rvert \cdot \lvert f_n (x) \rvert$ for all $\lambda \in \overline{K}$ and $x \in E$.
+We define $\lVert f_n \rVert = \sup_{x \in B_E(1)} \vert f_n(x) \vert$ and further assume that all the $\lVert f_n \rVert$ are finite.
+Write $f = \sum_n f_n$ and define $\Lambda(f)(v) \in \R \cup \{\infty\}$ for each $v \in \R$ by
+\[
+\Lambda(f)(v) = \log \sup_{x \in B_E(e^v)} \vert f(x) \vert.
+\]
+Finally, we define the Newton polygon $\NP(f)$ of $f$ to be the lower convex hull in $\mathbb{R}^2$
+of the points $(k, - \log \lVert f_k \rVert )$ for $k \in \Z_{\ge 0}.$ 
 \end{deftn}
 
-The following proposition is a classical consequence of convex analysis.
+We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $\NP(f)$, and
+denote by $\NP(f)^*$ the convex conjugate of $\NP(f)$.  The following proposition is a classical consequence of convex analysis.
 
 \begin{prop} \label{prop : conjugate Newton}
-We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $\NP(f).$
-We denote by $\NP(f)^*$ the convex conjugate of $\NP(f).$ $\NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$ (at least)
+Considered as a function on $\R$, $\NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$, though it may be defined for larger values.
 \end{prop}
 
-Then, $\Lambda(f)$ and $\NP(f)$ are connected :
+We can interpret the function $\Lambda(f)$ in terms of the conjugate Newton polygon:
 
@@
 \begin{proof}
-Let $v \in \left] -\infty,v_{\infty} \right[,$ and let us assume that $v$ is not a slope of $\NP(f).$
-Let $u = \NP(f)^*(v).$ Then the intersection of the line defined by $y=vx-u$ and $\NP(f)$ is an extreme point of $\NP(f),$ say $(n, - \log ( \vert f_n \vert ))$ for some $n \geq 0,$ and $u = vn + \log \vert f_n \vert.$
-
-$v$ is not a slope of $\NP(f)$ and by convexity of $\NP(f)$, its difference quotients at $n$ are increasing, therefore, there exists some $c>0$ such that for any $m\geq 0,$ $vm-u \leq - \log \vert f_m \vert - c \vert n-m \vert.$ Since $vm-u=vm-vn-log \vert f_n \vert,$ we get $-vn-\log \vert f_n \vert + c \vert n-m \vert  \leq -vm-\log \vert f_m \vert .$
-
-Now, let $x \in  B_E(0,e^v),$ and $m \geq 0.$ Definition of $\vert f_m \vert$ yields $ \vert f_m (x) \vert \leq \vert f_m \vert e^{vm}.$ Therefore \[\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn}< \vert f_n \vert e^{vn}.\]
-Thus, the serie converges and by ultrametricity $ \vert f(x) \vert \leq \vert f_n \vert e^{vn},$ and $\Lambda (f)(v) \leq \log \left( \vert f_n \vert e^{vn} \right).$
-
-Now, definition of $\vert f_n \vert$  provides $(x_i)_{i\geq 0}$ in $B_E(0,e^v)$ such that $\lim \vert f_n (x_i) \vert= \vert f_n \vert e^{vn}.$
+We prove the equality first for $v$ that are not slopes of $\NP(f)$.  Assume that $v < v_\infty$ is such a slope and set $u = \NP(f)^*(v)$.
+Then the intersection of $\NP(f)$ with the line defined by $y=vx-u$ is a vertex of $\NP(f),$ say
+$(n, - \log \lVert f_n \rVert )$ for some $n \geq 0$ with $u = vn + \log \lVert f_n \rVert.$
+
+Since $v$ is not a slope of $\NP(f)$, the difference quotients of $\NP(f)$ at $n$ are increasing.
+Therefore, we may find a $c>0$ so that $vm-u \leq - \log \lVert f_m \rVert - c \vert n-m \vert$ for any $m\geq 0$.
+Since $vm-u=vm-vn-\log \lVert f_n \rVert,$ we get $-vn-\log \lVert f_n \rVert + c \vert n-m \vert  \leq -vm-\log \lVert f_m \rVert .$
+
+For any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
+\[
+\vert f_m (x) \vert \leq \lVert f_m \rVert e^{vm}
+\]
+by the definition of $\lVert f_m \rVert$. Therefore
+\[
+\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \lVert f_n \rVert  e^{vn}< \lVert f_n \rVert e^{vn}.
+\]
+Thus, the series converges and by ultrametricity $ \vert f(x) \vert \leq \vert f_n \vert e^{vn},$ and $\Lambda (f)(v) \leq \log \left( \vert f_n \vert e^{vn} \right).$
+
+Now, the definition of $\lVert f_n \rVert$ provides a sequence $(x_i)_{i\geq 0}$ in $B_E(e^v)$ such that $\lim \vert f_n (x_i) \vert= \lVert f_n \rVert e^{vn}.$
 
@@
 
-As a consequence, $\lim_i \vert f(x_i) \vert = \vert f_n \vert e^{vn},$ and  $\Lambda (f)(v) = \log ( \vert f_n \vert e^{vn}) = u= \NP(f)^*(v).$
+As a consequence, $\lim_i \vert f(x_i) \vert = \lVert f_n \rVert e^{vn},$ and  $\Lambda (f)(v) = \log ( \lVert f_n \rVert e^{vn}) = u= \NP(f)^*(v).$
 
@@
 
-\begin{lem} Let $f(x)= \sum_{n \geq 0} f_n (x)$ and $g(x)= \sum_{n \geq 0} g_n (x)$ be two series as in definition \ref{deftn :  serie and norm}. Then :
-\[\Lambda(f+g) \leq \max (\Lambda(f),\Lambda(g))\]
-\[\Lambda(f \times g) \leq \Lambda(f)+\Lambda(g)\]
-\[\Lambda(f \circ g) \leq \Lambda(f) \circ \Lambda(g)\]
+\begin{lem} Let $f(x)= \sum_{n \geq 0} f_n (x)$ and $g(x)= \sum_{n \geq 0} g_n (x)$ be two series as in definition \ref{deftn :  serie and norm}. Then
+\begin{align*}
+\Lambda(f+g) &\leq \max (\Lambda(f),\Lambda(g)) \\
+\Lambda(f \times g) &\leq \Lambda(f)+\Lambda(g) \\
+\Lambda(f \circ g) &\leq \Lambda(f) \circ \Lambda(g)
+\end{align*}
 \end{lem}
 
-The following proposition will reveal crucial in order to minorate or majorate $\NP(f),$  $\NP(f)^*$ or $\Lambda(f),$ and is classical in convex analysis.
+The following classical result from convex analysis will play a crucial role in bounding $\NP(f),$  $\NP(f)^*$ and $\Lambda(f)$.
 
@@
 If $f$ and $g$ are two real convex functions such that $f \geq g,$ then $f^* \leq g^*.$
-Furthermore, slopes of $f^*$ corresponds to extreme points of $f$.
+Furthermore, the slopes of $f^*$ corresponds to vertices of $f$.
 \end{prop}

### 2014-02-19 padicprec.tex
Work on loc. anal.

--- 2014-02-18 padicprec.tex
@@
 
-In addition to allowing for more efficient computation of precision, storing the precision separately from the
-approximation makes it easier to determine the precision of the result when step-by-step precision tracking
-does not reflect the actual precision of the result.
-
-The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
-\[
-u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
-\]
-We shall consider the case where the initial
-terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
-
-Let us first examine how the absolute precision of $u_n$ varies with $n$
-if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
-The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
-decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
-step-by-step computation returns the value of $u_n$ with precision
+There are also
+
+\todofor{David}{
+Explain here why it might be useful to work separately with 
+approximation on the one hand and precision on the other hand.
+}
+
+\subsection{Step-by-step tracking is not sharp}
+
+In this subsection, we illustrate by some examples that tracking
+$p$-adic precision step by step may lead to important numerical
+instability.
+
+\subsubsection*{The SOMOS 4 sequence}
+
+The SOMOS 4 sequence is a four-term inductive sequence satisfying:
+$$u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.$$
+In this example, we shall consider the situation where the initial
+terms $u_0$, $u_1$, $u_2$ and $u_3$ are \emph{invertible} $p$-adic 
+integers, \emph{i.e.} lie in $\Zp^\times$.
+
+Let us first examine what happens (regarding to the precision) when
+we are computing the sequence $(u_n)$ naively. The computation of
+$u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
+decreases the precision by a factor $p^{\val(u_n)}$ where $\val$
+stands for the $p$-adic valuation. Hence, if we assume that $u_0$,
+$u_1$, $u_2$ and $u_3$ are given with precision $O(p^N)$ then the
+naive computation returns the value of $u_n$ with precision
 \begin{equation}
@@
 
-On the other hand, one can prove that the SOMOS 4 sequence 
-exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}. This means that, for all integer 
+However, on the other hand, one can prove that the SOMOS 4 sequence 
+exhibits the \emph{Laurent phenomenon}. This means that, for all integer 
 $n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
-1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
+1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$ (see \cite{fomin-zelevinsky:02a}).
 From the latter formula, it follows directly that if $u_0$, $u_1$,
@@
 appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
-precision but some numerical instability related to the algorithm used to compute $u_n$.
+precision but some numerical instability related to the algorithm ---
+which includes the model precision --- we used to compute to SOMOS 
+sequence.
 
@@
 From the above discussion, one can easily derive a numerically stable 
-algorithm that computes the SOMOS 4 sequence: 
-\begin{enumerate}
-\item compute the Laurent polynomials $P_n$ using the induction formula (and working in 
-the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$)
-\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
-\end{itemize}
-However, computing the $P_n$'s is very time-consuming since it requires 
-division in a polynomial ring with $4$ variables and the
+algorithm that computes the SOMOS 4 sequence: (1)~we compute the 
+Laurent polynomials $P_n$ using the induction formula (and working in 
+the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$) and (2)~we 
+evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
+However this algorithm is not efficient at all since computing the 
+$P_n$'s is very time-consuming for two reasons: first, it requires 
+division in a polynomial ring with $4$ variables and, second, the
 size of the coefficients of $P_n$ explodes as $n$ grows.
 
-In Section \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
+In \S \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
 the SOMOS 4 sequence which turns out to be, at the same time, efficient 
@@
 \begin{prop} \label{prop : legendre transform}
-If $f$ and $g$ are two real convex functions such that $f \geq g,$ then $f^* \leq g^*.$
-Furthermore, slopes of $f^*$ corresponds to extreme points of $f$.
+Slopes of $f^*$ corresponds to extreme points of $f$.
 \end{prop}
@@
 We define $m_2$ by extending $m_1$ with slope $2$.
-Let $\delta$ be the abscissa of the intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x.$ Then $e^\delta$ is enough to define first order lattices as in lemma  \ref{lem:main}.
+Let $\delta$ be the abscissa of the smallest intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x,$ or $v_\infty$ if there is no such intersection.
+Then $e^\delta$ is enough to define first order lattices as in lemma  \ref{lem:main}.
 \end{prop}
@@
 
-If $x \in B_{\overline{E}}(0,e^\delta),$ $u = \log \vert x \vert,$ then $\Lambda( \sum_{n \geq 2}f_n)(u) \geq -\log \varepsilon+ u.$
+By comparing slopes, if $x \in B_{\overline{E}}(0,e^\delta),$ $u = \log \vert x \vert,$  $\Lambda( \sum_{n \geq 2}f_n)(u) \geq -\log \varepsilon+ u.$
 With proposition \ref{prop : connexion sup Newton}, for any $m \geq 2,$ $-\log \vert f_m \vert +m u \geq -\log \varepsilon+ u.$
@@
 If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above.
-It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}.
+It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $\delta$ be such that $NP_F^* \left( \log_p \delta - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $\delta$ is enough for lemma \ref{lem:main}.
 \end{prop}

### 2014-02-19 padicprec.tex

--- 2014-02-19 padicprec.tex
@@
 \begin{prop} \label{prop : legendre transform}
-If $f$ and $g$ are two real convex functions such that $f \geq g,$ then $f^* \leq g^*.$
-Furthermore, the slopes of $f^*$ corresponds to vertices of $f$.
+Slopes of $f^*$ corresponds to extreme points of $f$.
 \end{prop}
@@
 If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above.
-It is enough to work with the Newton polygon $\NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $r$ be such that $\NP_F^* \left( \log_p r - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $r$ is enough for lemma \ref{lem:main}.
+It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $\delta$ be such that $NP_F^* \left( \log_p \delta - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $\delta$ is enough for lemma \ref{lem:main}.
 \end{prop}

### 2014-02-19 padicprec.tex
Working on Section 4

--- 2014-02-19 padicprec.tex
@@
 
-\todofor{David}{Finish summarizing sections}
+Section \ref{sec:mainlemma} centers on Lemma \ref{lem:main}, our main result.
+By describing how lattices transform under non-linear maps, it allows us to propogate
+precision using differentials.  In Section
 
@@
 
-%Using Lemma \ref{lem:reductionball}, one can extend Proposition 
-%\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
-%a lattice in some $K$-Banach space $E$, we first define the two
-%following quantities:
-%\begin{itemize}
-%\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
-%\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
-%\end{itemize}
-%
-%\begin{cor}
-%Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
-%respectively and $f : U \to V$ be a locally analytic function. Let 
-%also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
-%Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
-%soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
-%constant depending only on $f$ and $v_0$ which can be determined 
-%explicitely from the Newton polygon of $f$ at $v_0$.
-%\end{cor}
-%
-%\begin{proof}
-%We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
-%\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-%\Vert)$ is induced by identity map.
-%\end{proof}
-
 \section{Tracking precision using differentials}
@@
 
-Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have 
-important applications to effective computations with $p$-adic numbers 
-or power series, that we discuss in this Section.
-
-\subsection{Automatic sharp track of precision}
+In this section we discuss applications of Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic}
+to effective computations with $p$-adic numbers and power series.
+
+\subsection{Optimal precision tracking}
 
 We consider a function {\tt f} (in the sense of computer 
-science) that takes as input an (approximate) element lying in an open 
-subset $U$ of a $K$-Banach space $E$ and outputs another (approximate) 
-element lying in an open subset $V$ of another $K$-Banach space $F$. Of 
-course, this function can be modelised by a \emph{continuous} 
-mathematical function $f : U \to V$. The relation is the following: when 
+science) that takes as input an approximate element lying in an open 
+subset $U$ of a $K$-Banach space $E$ and outputs another approximate
+element lying in an open subset $V$ of another $K$-Banach space $F$.
+In applications, this function models a continuous
+mathematical function $f : U \to V$: when 
 {\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
-$f(x+H) \subset x' + H'$. We say that {\tt f} does not generate 
-artificial loss of precision if the above inclusion is an equality; it is 
-often not the case as shown in \S \ref{ssec:separation}.
-
-Let us assume now that $f$ is locally analytic on $U$ and in addition 
+$f(x+H) \subseteq x' + H'$. We say that {\tt f} \emph{preserves precision}
+if the above inclusion is an equality; it is 
+often not the case as shown in Section \ref{ssec:separation}.
+
+Let us assume now that $f$ is locally analytic on $U$ and
 that $f'(x)$ is surjective. Proposition \ref{prop:locanalytic} then 
@@
 $H$ is a first order lattice for $f$ at $x$. For such a lattice, by 
-definition, we have $f(x+H) = f(x) + f'(x)(H)$. Thus the smallest 
-possible $H'$ is $f'(x)(H)$. In other words, the function {\tt f} 
-generates artificial loss of precision if the precision $H'$ it outputs 
-is strictly larger that $f'(x)(H)$. The aim of this section is to 
+definition, we have $f(x+H) = f(x) + f'(x)(H)$ and thus {\tt f}
+must output $f'(x)(H)$ if it preserves precision. In this section we
 explain how, under the above hypothesis, one can implement the function 
@@
 
-\subsubsection*{Forward computation}
+\subsubsection*{One-pass computation}
 
@@
 where the $f_i$'s correspond to each individual basic step (like 
-addition, multiplication or affectation of variables); it is then a
+addition, multiplication or creation of variables); it is then a
 ``nice'' (in particular locally analytic) function. For all 
@@
 
-\subsubsection*{Backward computation}
+\subsubsection*{Two-pass computation}
 
@@
 
-It is nevertheless not obvious to implement these ideas in practice
+It is nevertheless not obvious how to implement these ideas in practice
 because when we enter in the $i$-th step of the execution of {\tt f},
@@
 
-We conclude this Section by remarking that the ``backward'' method seems 
+We conclude this section by remarking that the ``backward'' method seems 
 to be particularly well suited to computations with lazy $p$-adic. Indeed 
@@
 
-\subsubsection*{Additional remarks}
-
-From the beggining, we have assumed that $f'(x)$ is surjective. Let us 
+\subsubsection*{Non-surjectivity}
+
+From the beginning, we have assumed that $f'(x)$ is surjective. Let us 
 discuss shortly what happens when this assumption is relaxed. As it is 
@@
 can do is to project the result onto different quotients, \emph{i.e.} to 
-work with the composites $\pr_K \circ f$ for a sufficiently large family 
-of closed sub vector spaces $K \subset F$ such that $K + f'(x)(E) = F$. 
-(We recall that here $\pr_K$ denotes the canonical projection to $F$ 
-onto the quotient $F/K$.) If $F$ has a natural system of coordinates 
-(which is the case in almost all settings), we can generally take the 
-$\pr_K$'s as the projections on each coordinate. Doing this, we end up 
+work with the composites $\pr_W \circ f$ for a sufficiently large family 
+of closed sub-vector-spaces $W \subset F$ such that $W + f'(x)(E) = F$. 
+If $F$ has a natural system of coordinates, we may generally take the 
+$\pr_W$'s as the projections on each coordinate. Doing this, we end up 
 with a precision on each individual coordinate. Furthermore, we have the 
-guarantee that all these precisions are sharp (even if the lattice build
-from them is not).
+guarantee that each coordinate-wise precision is sharp, even if the lattice built
+from them is not.
 
@@
 want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
-family of $n$ vectors to its Graam matrix. The differential of $f$ is
+family of $n$ vectors to its Gram matrix. The differential of $f$ is
 clearly never surjective because $f$ takes its values in the subspace
-consisting of symmetric matrices. Nevertheless for all couple $(i,j)
+consisting of symmetric matrices. Nevertheless, for all pairs $(i,j)
 \in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
-\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ that takes a matrix 
+\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ takes a matrix 
 $M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
-their differentials generally are surjective (\emph{i.e.} they do not 
-vanish). Let $M$ be a matrix known at some finite precision such that 
+their differentials are generically surjective. Let $M$ be a matrix known
+at some finite precision such that 
 $f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
@@
 
-We conclude this Section by briefly remarking that the two strategies designed 
+We conclude this section by briefly remarking that the two strategies designed 
 above (\emph{i.e.} forward and backward computations) share some 
@@
 The theory of finite dimensional $K$-manifolds is presented for example 
-in \cite{schneider:11a}*{Ch. 8-9}. In this Section, we shall work with
+in \cite{schneider:11a}*{Ch. 8-9}. In this section, we shall work with
 a slightly different notion of manifolds which allows also Banach vector 

### 2014-02-20 padicprec.tex
Work on Section 3

--- 2014-02-19 padicprec.tex
@@
 We first note that a LU decomposition of a given matrix $M$ is unique
-if you require further that the $L$-part of this decomposition is
+if we require further that the $L$-part of this decomposition is
 unipotent (\emph{i.e.} all diagonal entries equal $1$). Let us write
 $L(M) U(M)$ the LU factorization of $M$ satisfying the above extra
-condition (when it exists). There exists Cramer-type formulas that
-gives each individual entries of $L(M)$and $U(M)$ in a closed form
-(see \cite{caruso:12a	}).
-
+condition (when it exists). There exists Cramer-type formulae that
+gives each individual entry of $L(M)$ and $U(M)$ in a closed form
+(see \cite{caruso:12a}).
 If $M$ is random square matrix over $\Zp$ of side $d$ whose every
@@
 $O(p^{N - \frac{2d}{p-1}})$ on average;
-\item computing $L(M)$ by evaluating Cramer-type formulas yields a
+\item computing $L(M)$ by evaluating Cramer-type formulae yields a
 result whose every entry is known up to precision $O(p^{N - 2 \log_p 
@@
 
-Our theory of $p$-adic precision rests upon a 
-lemma in $p$-adic analysis: Lemma \ref{lem:main}.  In this section
-we develop the theory surrounding this result, and proceed to practical
-consequences in Section \ref{sec:tracking}.
+Our theory of $p$-adic precision rests upon a lemma in $p$-adic 
+analysis: Lemma \ref{lem:main}.  In this section we develop the theory 
+surrounding this result, and proceed to practical consequences in 
+Section \ref{sec:tracking}.
 
@@
 
-Our goal in this section is to relate the image of a lattice under a differentiable map
-to its image under the derivative.
+Our goal in this section is to relate the image of a lattice under a 
+differentiable map to its image under the derivative.
 
 \begin{deftn} \label{deftn : diff}
-Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
-subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
-\emph{differentiable} at $v_0 \in U$ if there exists a 
-continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
-$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
-\subset U$ containing $v_0$ with
+Let $E$ and $F$ be two normed $K$-vector spaces, let $U$ be an open 
+subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
+\emph{differentiable} at $v_0 \in U$ if there exists a continuous linear 
+map $f'(v_0) : U \rightarrow W$ such that for any $\varepsilon >0$, 
+there exists an open neighborhood $U_\varepsilon \subset U$ containing 
+$v_0$ with
 \[ 
@@
 \end{deftn}
+
+\begin{rem}
+This notion of differentiability is stronger than the usual notion with 
+which we are familiar. It implies for instance the following: if $f$ is 
+differentiable on $U$, then the function $x \mapsto f'(x)$ defined on 
+$U$ is continuous. 
+It is sometimes called \emph{strict differentiability} but, for this
+article, we prefer a simplier terminology since we shall only use this
+notion of differentiability.
+\end{rem}
 
@@
 We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 
-\[
-B_F(1) \subset C  f'(0)(B_E(1)).
-\]
+$$B_F(1) \subset f'(0)(B_E(C)).$$
 Let $\varepsilon>0$ be such that 
@@
 
-Let $r \in (0, \delta)$ and suppose $H$ is a lattice with $B^-_E(\rho r) \subset H \subset B_E(r).$ 
+Let $r \in (0, \delta)$. We suppose that $H$ is a lattice with 
+$B^-_E(\rho r) \subset H \subset B_E(r).$
 We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
@@
 \varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r) 
-\subset \varepsilon C f'(0) (B_E(r))$. Thus there exists $x' \in 
-\varepsilon C \cdot B_E(r)$ such that $f'(0) (x') =y$. Since 
+\subset \varepsilon f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
+\in \cdot B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
 $\varepsilon C < \rho$, we get $x' \in B^-_E(\rho r) \subset H$ and 
@@
 
-\todofor{Tristan}{Improve this Subsection}
-
-Before discussing applications of Lemma \ref{lem:main} to ultrametric 
-precision, we discuss the case that $f$ is locally analytic.  In this context
-we prove that we can recover the constant $\delta$ appearing in the lemma
-from the Newton polygon determined by the Taylor series expansion of $f$.
-
-Schneider \cite{schneider:11a}*{\S 6} defines locally analytic functions on finite dimensional vector spaces.
-His definition extends easily to $K$-Banach spaces:
-
-\begin{deftn} \label{deftn:analyticNP}
-Let $E$ and $F$ be $K$-Banach spaces, and let $U \subset E$ be an open set.
-A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} if one may find
-an open set $U_x \subset E$ for each $x \in U$ satisfying the following condition.
-For each $n \ge 1$ there exists an $n$-linear mapping
-$L_n \col E^n \rightarrow F$ so that 
-\begin{enumerate}
-\item the sum $\sum_{n \geq 1} L_n(y,\ldots,y)$ converges in F for every $y \in U_x,$ 
-\item $f(x+y)=f(x)+\sum_{n \geq 1} L_n(y,\dots,y).$
-\end{enumerate}
+We consider in this subsection the case where $f$ is locally analytic: 
+under this extra assumption, we make explicit the constant $\delta$ 
+which appears in Lemma \ref{lem:main}.
+Locally analytic functions over finite dimensional $K$-vector spaces are 
+defined in Schneider's book \cite {schneider:11a}, \S 6. His definition 
+extends readily to $K$-Banach spaces as follows:
+
+\begin{deftn}
+\label{def:locanal}
+Let $E$ and $F$ be $K$-Banach spaces. Let $U$ be an open subset of $E$
+and let $x \in U$.
+A function $f \col U \rightarrow F$ is said \emph{locally analytic} at 
+$x$ if there exist an open subset $U_x \subset E$ and, for all $n \ge 
+1$, a continuous $n$-linear map $L_n \col E^n \rightarrow F$ such that
+$$f(x+h) = f(x) + \sum_{n \geq 1} L_n(h,\ldots,h)$$
+for all $h$ with $x + h \in U_x$.
 \end{deftn}
 
-We then mostly study such a serie defined at $x$.
-
-\begin{deftn} \label{deftn :  serie and norm}
-Suppose that $E$ and $F$ are $\overline{K}$-Banach spaces and $f_n \col E \to F$ are maps satisfying
-$\lvert f_n(\lambda x) \rvert = \lvert \lambda \rvert \cdot \lvert f_n (x) \rvert$ for all $\lambda \in \overline{K}$ and $x \in E$.
-We define $\lVert f_n \rVert = \sup_{x \in B_E(1)} \vert f_n(x) \vert$ and further assume that all the $\lVert f_n \rVert$ are finite.
-Write $f = \sum_n f_n$ and define $\Lambda(f)(v) \in \R \cup \{\infty\}$ for each $v \in \R$ by
-\[
-\Lambda(f)(v) = \log \sup_{x \in B_E(e^v)} \vert f(x) \vert.
-\]
-Finally, we define the Newton polygon $\NP(f)$ of $f$ to be the lower convex hull in $\mathbb{R}^2$
-of the points $(k, - \log \lVert f_k \rVert )$ for $k \in \Z_{\ge 0}.$ 
-\end{deftn}
-
-We define $v_\infty \in \mathbb{R} \cup \left\lbrace \infty \right\rbrace$ to be the limit of the slopes of $\NP(f)$, and
-denote by $\NP(f)^*$ the convex conjugate of $\NP(f)$.  The following proposition is a classical consequence of convex analysis.
-
-\begin{prop} \label{prop : conjugate Newton}
-Considered as a function on $\R$, $\NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$, though it may be defined for larger values.
-\end{prop}
-
-We can interpret the function $\Lambda(f)$ in terms of the conjugate Newton polygon:
-
-\begin{prop} \label{prop : connexion sup Newton}
-For any $v \in \left]-\infty, v_\infty \right[,$ \[ \Lambda (f) (v) = \NP(f)^*(v).\]
-\end{prop}
-\begin{proof}
-We prove the equality first for $v$ that are not slopes of $\NP(f)$.  Assume that $v < v_\infty$ is such a slope and set $u = \NP(f)^*(v)$.
-Then the intersection of $\NP(f)$ with the line defined by $y=vx-u$ is a vertex of $\NP(f),$ say
-$(n, - \log \lVert f_n \rVert )$ for some $n \geq 0$ with $u = vn + \log \lVert f_n \rVert.$
-
-Since $v$ is not a slope of $\NP(f)$, the difference quotients of $\NP(f)$ at $n$ are increasing.
-Therefore, we may find a $c>0$ so that $vm-u \leq - \log \lVert f_m \rVert - c \vert n-m \vert$ for any $m\geq 0$.
-Since $vm-u=vm-vn-\log \lVert f_n \rVert,$ we get $-vn-\log \lVert f_n \rVert + c \vert n-m \vert  \leq -vm-\log \lVert f_m \rVert .$
-
-For any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
-\[
-\vert f_m (x) \vert \leq \lVert f_m \rVert e^{vm}
-\]
-by the definition of $\lVert f_m \rVert$. Therefore
-\[
-\vert f_m (x) \vert \leq e^{-c \vert n-m \vert } \lVert f_n \rVert  e^{vn}< \lVert f_n \rVert e^{vn}.
-\]
-Thus, the series converges and by ultrametricity $ \vert f(x) \vert \leq \vert f_n \vert e^{vn},$ and $\Lambda (f)(v) \leq \log \left( \vert f_n \vert e^{vn} \right).$
-
-Now, the definition of $\lVert f_n \rVert$ provides a sequence $(x_i)_{i\geq 0}$ in $B_E(e^v)$ such that $\lim \vert f_n (x_i) \vert= \lVert f_n \rVert e^{vn}.$
-
-Since for any $i,$ $\vert f_m (x_i) \vert \leq e^{-c \vert n-m \vert } \vert f_n \vert  e^{vn},$ then for $i$ large enough, $\vert f_m (x_i) \vert < \vert f_n (x_i) \vert $ and $\vert f(x_i) \vert = \vert f_n (x_i) \vert.$
-
-As a consequence, $\lim_i \vert f(x_i) \vert = \lVert f_n \rVert e^{vn},$ and  $\Lambda (f)(v) = \log ( \lVert f_n \rVert e^{vn}) = u= \NP(f)^*(v).$
-
-Therefore, for any $v \in \left] -\infty,v_{\infty} \right[,$ with $v$ not a slope of $\NP(f),$ we get $\Lambda (f)(v) =\NP(f)^*(v).$
-Since $\Lambda (f)$ is increasing, $\NP(f)^*$ is continuous (since it is convex), and $\NP(f)$ has only finitely many slopes smaller than $v_\infty-\epsilon$, for any $\epsilon >0$, we deduce that both functions coincide on $v \in \left] -\infty,v_{\infty} \right[,$
-\end{proof}
-
-\begin{lem} Let $f(x)= \sum_{n \geq 0} f_n (x)$ and $g(x)= \sum_{n \geq 0} g_n (x)$ be two series as in definition \ref{deftn :  serie and norm}. Then
+\begin{rmk}
+A function $f$ which is locally analytic at $x$ is \emph{a fortiori} 
+differentiable at $x$. Furthermore, in that case, the linear map $L_1$ 
+appearing in the above definition is nothing but the derivative $f'(x)$.
+\end{rmk}
+
+From now, we assume that $K$ is algebraically closed. As in Definition 
+\ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
+family of continuous $n$-linear maps $L_n : E^n \to F$. For $n \ge 1$ 
+and $h \in E$, we set $f_n(h) = L_n(h, \ldots, h)$ and
+$$\Vert f_n \Vert = \sup_{h \in B_E(1)} \Vert f_n(h) \Vert.$$
+When the series $\sum_n f_n(h)$ converges, we denote by $f(h)$ its sum;
+we shall write $f = \sum_{n \geq 0} f_n$. We assume that $f$ is defined
+in a neighborhood of $0$. Under this assumption, the datum of $f$ 
+uniquely determines the $f_n$'s (it is a consequence of Propostion 
+\ref{prop:lambdaNP} below).
+To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
+\cup \{+\infty\}$ defined by:
+$$\begin{array}{rcll}
+\Lambda(f)(v) & = & 
+\log \big( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \big)
+& \text{if } f \text{ is defined on } B^-_E(e^v) \\
+& = & +\infty & \text{otherwise}
+\end{array}$$
+The following lemma is easy and left to the reader.
+
+\begin{lem}
+\label{lem:opLambda}
+Let $f = \sum_{n \geq 0} f_n$ and $g = \sum_{n \geq 0} g_n$ be two 
+series as above. Then:
 \begin{align*}
@@
 
-The following classical result from convex analysis will play a crucial role in bounding $\NP(f),$  $\NP(f)^*$ and $\Lambda(f)$.
-
-\begin{prop} \label{prop : legendre transform}
-Slopes of $f^*$ corresponds to extreme points of $f$.
+\begin{rem}
+\label{rem:opLambda}
+Using Lemma \ref{lem:opLambda}, one can easily derive an upper 
+bound of $\Lambda(f)$ from a formula describing $f$.
+\end{rem}
+
+The function $\Lambda(f)$ we have just defined is actually closely 
+related to the Newton polygon of $f$. Recall that, by definition, the 
+Newton polygon of $f$ is the convex hull in $\R^2$ of the points of 
+coordinates $(n, - \log \lVert f_n \rVert )$ for $n \geq 0$ together 
+with the extra point $(0, +\infty)$. We denote by $\NP(f) : \R \to \R 
+\cup \{+\infty\}$ the convex function whose epigraph is the Newton 
+polygon of $f$.
+
+We recall that the Legendre transform of a \emph{convex} function 
+$\varphi : \R \to \R \cup \{+\infty\}$ is the function $\varphi^\star : 
+\R \to \R \cup \{+\infty\}$ defined by:
+$$\textstyle \varphi^\star(v) = \sup_{u \in \R} \:
+\big(uv - \varphi(u)\big), \qquad
+\forall v \in \R.$$
+One can check that the map $\varphi \mapsto \varphi^\star$ is an
+involution (\emph{i.e.} $(\varphi^\star)^\star = \varphi$) and
+reverses the order (\emph{i.e.} if $\varphi \leq \psi$, then
+$\varphi^\star \geq \psi^\star$). 
+%Furthermore, if $\varphi$ is
+%piecewise affine (and convex), the slopes of $\varphi^\star$ are
+%exactly the points where $\varphi$ is not differentiable and \emph{vice
+%versa}.
+
+%\begin{prop} \label{prop : conjugate Newton}
+%Considered as a function on $\R$, $\NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$, though it may be defined for larger values.
+%\end{prop}
+%
+%We can interpret the function $\Lambda(f)$ in terms of the conjugate Newton polygon:
+
+\begin{prop} \label{prop:lambdaNP}
+Keeping the above notation, we have $\Lambda (f) = \NP(f)^\star$.
 \end{prop}
 
-\begin{cor} \label{cor : trunc series}
-Let $f(x)= \sum_{n \geq 0} f_n (x)$ be a serie as in definition \ref{deftn :  serie and norm}, $n_0 \geq 0$ and $g(x)= \sum_{n \geq n_0} f_n (x).$ Then $\Lambda(g) \leq \Lambda(f)$ and the slopes of $\Lambda(g)$ are bigger than $n_0$.
-As a consequence, a minoration of $\Lambda(g)$ can be obtain by extending $\Lambda(f)(v)$ with a slope $n_0$ for $v \leq n_0$.
-\end{cor}
-
-Now, let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}.
+\begin{proof}
+Let $v \in \R$. We assume first that $\NP(f)^\star(v) = +\infty$. Then, 
+using a multilinear version of Banach-Steinhaus Theorem, we can prove 
+that $\sum_{n \geq 0} f_n$ does not converge on $B_E^-(e^v)$ and 
+therefore that $\Lambda(f)(v) = +\infty$ as well.
+
+We now suppose that $\NP(f)^\star(v)$ is finite. To begin with, we 
+assume in addition that $v$ is not a slope of $\NP(f)$. We set $u = 
+\NP(f)^*(v)$. The function $m \mapsto \NP(f)(m) - vm + u$ has the 
+following properties: (1)~it is piecewise affine and everywhere 
+nonnegative, (2)~it does not admit $0$ as a slope and (3)~it vanishes at 
+$x = n$ for some integer $n$ and $u = vn + \log \Vert f_n \Vert$.
+We deduce from this that there exists $c>0$ such that $vm-u \leq - \log 
+\lVert f_m \rVert - c \cdot \vert n-m \vert$ for any $m\geq 0$.
+Since $vm-u=vm-vn-\log \lVert f_n \rVert$, we get $-vn-\log \lVert 
+f_n \rVert + c \cdot \vert n-m \vert \leq -vm-\log \lVert f_m \rVert$.
+Therefore, for any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
+$$\Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
+\lVert f_n \rVert \cdot e^{vn}< \lVert f_n \rVert \cdot e^{vn}.$$
+Thus, the series $\sum_{m \geq 0} f_m(x)$ converges and $\Vert f(x) 
+\Vert \leq \Vert f_n \Vert \cdot e^{vn}$. We then get:
+\begin{equation}
+\label{eq:lowerLambda}
+\Lambda (f)(v) \leq \log \big( \Vert f_n \Vert e^{vn} \big)
+= vn + \log \Vert f_n \Vert = u.
+\end{equation}
+On the other hand, it follows from the definition of $\lVert f_n \rVert$ 
+that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
+$\lim \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
+furthermore $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} 
+\Vert f_n \Vert \cdot e^{vn}$ for all $i$, we get $\Vert f_m (x_i) \Vert 
+< \Vert f_n (x_i) \Vert$ for $i$ large enough. For these $i$, we then 
+have $\Vert f(x_i) \Vert = \Vert f_n (x_i) \Vert$. Passing to the limit 
+on $i$, we find $\Lambda (f)(v) \geq u$. Comparing with 
+Eq.~\eqref{eq:lowerLambda}, we get $\Lambda (f)(v) = u = 
+\NP(f)^\star(v)$.
+
+We have then proved that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
+coincide expect maybe on the set consisting of slopes of $\NP(f)$. But 
+we remark that this set is countable and that both functions $\Lambda 
+(f)$ and $\NP(f)^\star$ are left continuous. This is enough to ensure 
+that $\Lambda (f)$ and $\NP(f)^\star$ coincide everywhere.
+\end{proof}
+
+\begin{rem}
+It follows from Proposition \ref{prop:lambdaNP} that $\Lambda(f)$ is a
+convex function (which was not clear at all \emph{a priori}).
+\end{rem}
+
+We now study the effect of truncation on series: given $f$ as above
+and a nonnegative integer $n_0$, we set:
+$$f_{\geq n_0} = \sum_{n \geq n_0} f_n = f - (f_0 + f_1 + \cdots +
+f_{n_0 - 1}).$$
+On the other hand, given a convex function $\varphi : \R \to \R \cup
+\{+\infty\}$ and a real number $v$, we define $\varphi_{\geq v} : \R \to 
+\R \cup \{\pm \infty\}$ as the highest convex function such that 
+$\varphi_{\geq v} \leq \varphi$ and the function $x \mapsto 
+\varphi_{\geq v}(x) - v x$ is not decreasing. Concretely, we have:
+$$\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y 
+\big).$$
+When $v$ is fixed, the construction $\varphi \mapsto \varphi_{\geq v}$
+is nondecreasing: if $\varphi$ and $\psi$ are two convex functions such
+that $\varphi \leq \psi$, we deduce $\varphi_{\geq v} \leq \psi_{\geq v}$.
+
+\begin{prop} \label{prop:trunc}
+With the above notations, we have $\Lambda(f_{\geq n_0}) \leq
+\Lambda(f)_{\geq n_0}$ for all $n_0 \in \mathbb N$.
+\end{prop}
+
+\begin{proof}
+It follows easily from Proposition \ref{prop:lambdaNP} and the fact that 
+the slopes of the Legendre transform of a convex piecewise affine 
+function $f$ are exactly the points where $f$ is not differentiable.
+\end{proof}
+
+%Now, let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}.
 We assume $L_1$ is surjective.
 We define, the $\overline{L_k},$ $\overline{E}$ and $\overline{F}$ by extending scalars to $\overline{K}.$ We define for any $n \geq 1$, $f_n(x)=\overline{L_n}(x,\dots,x)$ for any $x \in \overline{E}.$
-Let $C$ be given by the open mapping theorem for $L_1$, $\rho \in (0, 1]$ and $\varepsilon < \frac{\rho}{C}$.
+%Let $C$ be given by the open mapping theorem for $L_1$, $\rho \in (0, 1]$ and $\varepsilon < \frac{\rho}{C}$.
 We provide three methods to find a sufficient $\delta$ to define first order lattice as in lemma \ref{lem:main}.
-
-For the first one, we assume that we have know a function that majorates $\Lambda(\sum_{n \geq 1}f_n)$.
-
-\begin{prop} \label{prop : maj Lambda} \label{prop:locanalytic}
-If $m_1$ is a real convex function above $\Lambda(\sum_{n \geq 1}f_n)$.
-We define $m_2$ by extending $m_1$ with slope $2$.
-Let $\delta$ be the abscissa of the smallest intersection of $m_2$ and the line defined by $y=-\log \varepsilon+ x,$ or $v_\infty$ if there is no such intersection.
-Then $e^\delta$ is enough to define first order lattices as in lemma  \ref{lem:main}.
+%
+%For the first one, we assume that we have know a function that majorates $\Lambda(\sum_{n \geq 1}f_n)$.
+
+\begin{prop}
+\label{prop:locanalytic}
+Let $f = \sum_{n \geq 0} f_n$ be a function as above.
+Let $C$ be a positive real number satisfying 
+$B_F(1) \subset f_1(B_E(C))$.
+Let $\rho \in (0, 1]$ and $\nu$ be a real number such that 
+\begin{equation}
+\label{eq:locanalytic}
+\Lambda(f)_{\geq 2} (\nu) < \nu + \log \Big( \frac \rho C \Big).
+\end{equation}
+Then the conclusion of Lemma \ref{lem:main} holds with $\delta = e^\nu$.
 \end{prop}
+
+\begin{rem}
+On a neighborhood of $-\infty$, the function $x \mapsto \Lambda(f)_{\geq 
+2} (x) - x$ is affine of slope $1$. This implies that, for all $\rho \in
+(0,1]$, there exists $\nu$ satisfying \eqref{eq:locanalytic}. Moreover,
+if $\rho$ is close enough to $0$, then one can take $\delta = e^\nu$ as a 
+linear function of $\rho$.
+\end{rem}
+
 \begin{proof}
-$\delta$ is well-defined thanks to corollary \ref{cor : trunc series}, and $m_2 \geq \Lambda(\sum_{n \geq 2}f_n).$
-
-It is sufficient to prove that for all $x \in B_{\overline{E}}(0,e^\delta),$ $\vert \sum_{n\geq 2} f_n(x) \vert \leq \varepsilon \vert x \vert.$
-
-By comparing slopes, if $x \in B_{\overline{E}}(0,e^\delta),$ $u = \log \vert x \vert,$  $\Lambda( \sum_{n \geq 2}f_n)(u) \geq -\log \varepsilon+ u.$
-With proposition \ref{prop : connexion sup Newton}, for any $m \geq 2,$ $-\log \vert f_m \vert +m u \geq -\log \varepsilon+ u.$
-
-Therefore, for any $m \geq 2,$ $\vert f_m \vert  \vert x \vert^m \leq \varepsilon \vert x \vert.$
-By ultrametricity,   $\vert \sum_{n\geq 2} f_n(x) \vert \leq \varepsilon \vert x \vert,$ which concludes the proof.
+Pick $\varepsilon$ in the interval $(e^{\Lambda(f)_{\geq 2} (\nu) - \nu}, 
+\frac \rho C)$. Going back to the proof of Lemma \ref{lem:main}, we 
+observe that it is enough to prove that
+\begin{equation}
+\label{eq:boundqueue} 
+\Vert f_{\geq 2}(x) \Vert \leq \varepsilon \cdot \Vert x \Vert.
+\end{equation}
+for all $x \in B_E(\delta)$.
+This follows directly from Proposition \ref{prop:lambdaNP} and Proposition
+\ref{prop:trunc}.
 \end{proof}
 
-For the second one, we only assume that we have majorations on the $\vert f_n \vert.$
-
-\begin{prop}
-Let $(M_n)_{n \geq 2}$ be real numbers such that for all $n \geq 2$, $\vert f_n \vert \leq M_n.$
-
-If we denote by $\NP((M_{n+1})_{n \geq 1})$ the lower convex hull of the $(n, -\log M_{n+1})$ for $n \geq 1$, then $\delta$ such that $-\log \varepsilon=\NP((M_n)_{n \geq 2})^*(r)$ is enough to define first order lattices as in lemma  \ref{lem:main}.
-\end{prop}
-\begin{proof}
-Same idea as proposition \ref{prop : maj Lambda}. One can define $g_n$ with $g_n (x) =\frac{f_n(x)}{\vert x \vert}$ and $g_n(0)=0$ and work with $\Lambda ( \sum g_n).$
-\end{proof}
-
-We finally provide a simple method when working with finite-dimensional vector spaces over $p$-adic fields.
-
-\begin{prop} \label{prop : locanalyticfinitedim}
-If $f$ is a locally analytic map $K^n \rightarrow K^m$, with $K$ a $p$-adic field, if $C,$ $\rho$ and $\varepsilon$ are as above.
-It is enough to work with the Newton polygon $NP_F$ of $F(t) = \sum_{k\geq 0}^{+\infty} \interleave f^{(k+1)}(0) \interleave  t^k \in \mathbb{R}[[t]]$ : let $\delta$ be such that $NP_F^* \left( \log_p \delta - \frac{1}{p-1} \right) \leq \log_p \varepsilon.$ Then $\delta$ is enough for lemma \ref{lem:main}.
-\end{prop}
-\begin{proof}
-Same as before, with $L_n=\frac{f^{(n)}(0)}{n!}$ and Legendre formula. The result follows \textit{mutatis mutandis}.
-\end{proof}
-
-\todo{Explain how Newton polygons considered above behave under
-usual operations (sum, multiplication and composition of functions).}
-
-%Using Lemma \ref{lem:reductionball}, one can extend Proposition 
-%\ref{prop:locanalytic} to arbitrary lattices $H$ as follows. Given $H$
-%a lattice in some $K$-Banach space $E$, we first define the two
-%following quantities:
-%\begin{itemize}
-%\item $\max(H)$ is the smallest $r$ such that $H \subset B_E(r)$, and
-%\item $\min(H)$ is the greatest $r$ such that $B_E(r) \subset H$.
-%\end{itemize}
-%
-%\begin{cor}
-%Let $U$ are $V$ be open subsets in $K$-Banach spaces $E$ and $F$ 
-%respectively and $f : U \to V$ be a locally analytic function. Let 
-%also $v_0$ be a point in $U$ such that $f'(v_0)$ is surjective. 
-%Then a lattice $H$ in $E$ is a first order lattice for $f$ at $v_0$ as 
-%soon as $\max(H)^2 < C_{f, v_0} \cdot \min(H)$ where $C_{f, v_0}$ is a 
-%constant depending only on $f$ and $v_0$ which can be determined 
-%explicitely from the Newton polygon of $f$ at $v_0$.
-%\end{cor}
-%
-%\begin{proof}
-%We apply Proposition \ref{prop:locanalytic} to the composite $f \circ 
-%\iota_H$ where $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-%\Vert)$ is induced by identity map.
-%\end{proof}
+\begin{rem}
+It follows from the proof that Proposition \ref{prop:locanalytic} is 
+still valid if $K$ is not assumed to be algebraically closed. Indeed, 
+the functions $f_n$ --- and then $f$ also --- extend to an algebraic
+closure $\bar K$ of $K$ and Eq.~\eqref{eq:boundqueue} holds over $\bar 
+K$, which is enough to conclude.
+\end{rem}
+
+\begin{cor}
+We keep the notations of Proposition \ref{prop:locanalytic} and consider 
+in addition a sequence $(M_n)_{n \geq 2}$ such that $\Vert f_n \Vert 
+\leq M_n$ for all $n \geq 2$. Let $\NP(M_n)$ denote the convex function
+whose epigraph is the convex hull in $\R^2$ of the points of coordinates
+$(n, -\log M_n)$ for $n \geq 2$ together with the extra point $(0,
++\infty)$.
+
+Let $\rho \in (0, 1]$ and $\nu$ be a real number such that 
+$$NP(M_n)^\star (\nu) < \nu + \log \Big( \frac \rho C \Big).$$
+Then the conclusion of Lemma \ref{lem:main} holds with $\delta = e^\nu$.
+\end{cor}
+
+\begin{rem}
+If $K$ has characteristic $0$ and the vector spaces $E$ and $F$ are
+finite dimensional, then the $M_n$'s defined by
+$$M_n = \frac 1 {|n!|} \cdot 
+\sup_{\substack{1 \leq i \leq \dim E \\ \vert \underline n
+\vert = n}} \,
+\Big\Vert \frac{\partial^n f_i}{\partial x^{\underline n}}(0) 
+\Big\Vert$$
+do the job. Here $f_i$ denotes the $i$-th coordinate of $f$, the
+notation $\underline n$ refers to a tuple of $(\dim F)$ nonnegative
+integers and $\vert \underline n \vert$ is the sum of the coordinates
+of $\underline n$.
+\end{rem}
 
@@
 for $g_i$ at $x$. Under our assumption, such lattices always exist and 
-can be computed dynamically using Proposition \ref{prop:locanalytic}. 
+can be computed dynamically using Proposition \ref{prop:locanalytic}
+and Lemma \ref{lem:opLambda} (see also Remark \ref{rem:opLambda}).
 Now, the equality $g_i(x + \tilde H_i) = x_i + g'_i(x)(\tilde H_i)$ 
@@
 Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
-\cite{}), we can reuse in Step~\eqref{item:finalcomp} the computations 
-already performed in Step~\eqref{item:smallprec}. The ``backward'' method 
-we have just presented is then probably not much more expansive than the 
-``forward'' method, although it is certainly much more difficult to 
-implement.
+\cite{hoeven1, hoeven2, berthomieu}), we can reuse in 
+Step~\eqref{item:finalcomp} the computations already performed in 
+Step~\eqref{item:smallprec}. The ``backward'' method we have just 
+presented is then probably not much more expansive than the ``forward'' 
+method, although it is certainly much more difficult to implement.
 
@@
 The theory of finite dimensional $K$-manifolds is presented for example 
-in \cite{schneider:11a}*{Ch. 8-9}. In this Section, we shall work with
+in \cite[\S 8, 9]{schneider:11a}. In this Section, we shall work with
 a slightly different notion of manifolds which allows also Banach vector 
@@
 
-\subsubsection*{Conclusion}
-
-The theory developped in this Appendix is nice because it provides an 
-abstract framework to deal with ultrametric precision in a quite general 
-setting.
-Nevertheless, as we have seen in the above examples, in practice we
-generally carry out all calculations in charts and do not really care
-about the ``real'' nature of the precision data, \emph{i.e.} lattices
-in tangent spaces.
+%\subsubsection*{Conclusion}
+%
+%The theory developped in this Appendix is nice because it provides an 
+%abstract framework to deal with ultrametric precision in a quite general 
+%setting.
+%Nevertheless, as we have seen in the above examples, in practice we
+%generally carry out all calculations in charts and do not really care
+%about the ``real'' nature of the precision data, \emph{i.e.} lattices
+%in tangent spaces.
 
@@
 \bibliographystyle{plain}
-\bibliography{roebib/Biblio}
+\bibliography{Biblio}
 

### 2014-02-20 padicprec.tex
Bibliography updated
Working on Appendix A

--- 2014-02-19 padicprec.tex
@@
 \usepackage{enumerate}
-\usepackage[algoruled,vlined,french,linesnumbered]{algorithm2e}
+\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
 
@@
 no division by zero). We assume also for simplicity that $u_0, u_1, u_2, 
-u_3$ are all units in $\Z_p$. By Laurent phenomenon, this implies first 
-that all $u_n$'s lie in $\Z_p$ and second that if $u_0, u_1, u_2, u_3$ 
-are known with finite precision $O(p^N)$ then all terms $u_n$'s are 
-known with the same precision. Algorithm \ref{algo:SOMOS} presented on 
+u_3$ are all units in $\Z_p$. By the Laurent phenomenon, this implies 
+that all $u_n$'s lie in $\Z_p$, and that if $u_0, u_1, u_2, u_3$ 
+are known with finite precision $O(p^N)$ then all $u_n$ are 
+known with the same absolute precision. Algorithm \ref{algo:SOMOS} presented on 
 page \pageref{algo:SOMOS} performs this computation.
@@
 precision $O(p^N)$}
-\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
+\Assumption{None of the $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
 \KwOut{$u_n$ at precision $O(p^N)$}
@@
 \For {$i$ from $1$ to $n-3$}{
-  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
-  {\bf lift} (arbitrary) 
-    $b$, $c$ and $d$ to precision $O(p^{\prec})$\label{line:lift}\;
+  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2)$\;
+  {\bf lift} 
+    $b$, $c$ and $d$ arbitrarily to precision $O(p^{\prec})$\label{line:lift}\;
   $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
@@
 
-We now prove that it is correct.
-We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
+We now prove that Algorithm \ref{algo:SOMOS} produces the correct result.
+We introduce the function $f : \Qp^\times \times \Q_p^3 \to \Q_p^4$ defined by 
 $f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
@@
 D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
-Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
+Thanks to the Laurent phenomenon, we know that $D_i$ has
 integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
@@
 
-Going back to our problem of precision, given $V$ a $K$-manifold as 
+Returning to our problem of precision, given $V$ a $K$-manifold as 
 above, we would like to be able to deal with ``approximations up to some 
@@
 O(H)$ where $x$ belongs to a dense \emph{computable} subset of $V$ and 
-$H$ is a ``precision data''.
-The aim of this paragraph is to give a precise meaning to the expression 
-$x + O(H)$ and to prove several basic properties of it.
+$H$ is a ``precision datum''.
 For now, we fix a $K$-manifold $V$ and we use freely the notations $I$, 
@@
 Let $x \in V$ and $H$ be a precision datum at $x$.
-The subset:
+The subset
 $$\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
@@
 equality 
-$$\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
-\subset V = \varphi_j^{-1}\big(\varphi_j(x) + \varphi_j'(x)(H)\big) 
-\subset V$$
+\[
+\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
+ = \varphi_j^{-1}\big(\varphi_j(x) + \varphi_j'(x)(H)\big)
+ \]
 is clearly equivalent to
@@
 \begin{deftn}
-Let $x \in V$ and $H$ be a precision datum at $x$. We set:
+Let $x \in V$ and $H$ be a precision datum at $x$. We set
 $$x + O(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big)
@@
 However, since the transition maps are not required to be isometries, 
-this norm do depend on the choice on $i$. They are nevertheless all
+this norm depends on the choice of $i$. They are nevertheless all
 equivalent because the transition maps are required to be continuous.
 
-In the next Lemma, we choose to endow $T_x V$ by one of the above norms.
+In the next lemma, we choose any of the above norms for $T_x V$.
 
@@
 
-Then for all $\rho \in (0,1]$, there exists $\delta \in (0,1)$ such 
-that for all $r \in (0, \delta)$, the equality
-$$f(x + O(H)) = f(x) + O(f'(x)(H))$$
-holds for all lattices $H$ in $T_x V$ such that $B^{\phantom -}_{T_x V}(\rho r)
-\subset H \subset B^-_{T_x V}(r)$.
+Then, for all $\rho \in (0, 1]$, there exists a positive real 
+number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
+$H$ in $T_x V$ such that $B^-_{T_x V}(\rho r) \subset H \subset B_{T_x V}(r)$ is a first
+order lattice for $f$ at $x$.
 \end{lem}

### 2014-02-20 padicprec.tex

--- 2014-02-20 padicprec.tex
@@
 of $\underline n$.
-\end{rem}
+\end{rem}x
 
@@
 The theory of finite dimensional $K$-manifolds is presented for example 
-in \cite[\S 8, 9]{schneider:11a}. In this Section, we shall work with
+in \cite{schneider:11a}*{Ch. 8-9}. In this section, we shall work with
 a slightly different notion of manifolds which allows also Banach vector 
@@
 Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
-charts. Proposition \ref{prop:independance} then tells us that a 
+charts. Proposition \ref{prop:independence} then tells us that a 
 precision datum on $x$ determined a precision datum on $y$ and \emph{vice 

### 2014-02-20 padicprec.tex
Small fixes

--- 2014-02-20 padicprec.tex
@@
 \begin{rem}
+\label{rem:differentiability}
 This notion of differentiability is stronger than the usual notion with 
@@
 U_j$, the image of $T_x V$ in $E_i$ is a first order lattice for
-$\psi_{ij}$ at $\varphi_i(x)$ (\emph{cf} Definition \ref{def:firstorder}.
+$\psi_{ij}$ at $\varphi_i(x)$ (\emph{cf} Definition \ref{def:firstorder}).
 \end{deftn}
@@
 \circ \varphi'_i(x_0)$. Given an extra index $j$ satisfying the same
-assumption, we know by \todo{(find a reference)} that the difference
-$f_i - f_j$ goes to $0$ when $x$ converges to $x_0$. Since $H_0$ is open
-in $T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ and 
-$f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in f_j(H_0)$ and 
-write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal to $f_i(v) + 
-(f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each summand does. 
-Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse inclusion can be proved 
-in the same way. The fact that $H$ is a precision datum at $x$ is easy
-and left to the reader.
+assumption, the difference $f_i - f_j$ goes to $0$ when $x$ converges to 
+$x_0$ (see Remark \ref{rem:differentiability}). Since $H_0$ is open in 
+$T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ 
+and $f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in 
+f_j(H_0)$ and write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal 
+to $f_i(v) + (f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each 
+summand does. Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse 
+inclusion can be proved in the same way. The fact that $H$ is a 
+precision datum at $x$ is easy and left to the reader.
 Finally, if $x$ is close enough to $x_0$, we can check (ii) in the 

### 2014-02-20 padicprec.tex
Working on Appendix B

--- 2014-02-20 padicprec.tex
@@
 of $\underline n$.
-\end{rem}x
+\end{rem}
 
@@
 \begin{rmk}
-The constant $\delta$ that appears in the above Lemma depends (up to
+The constant $\delta$ that appears in the lemma depends (up to
 some multiplicative constant) on the norm that we have chosen on $T_x V$. 
@@
 analytic as well, the constant $\delta$ can be made explicit using the 
-method of \S \ref{ssec:locanalytic}.
+method of Section \ref{ssec:locanalytic}.
 \end{rmk}
@@
 
-In this example, we assume for simplicity that $K$ has not caracteristic 
-$2$. Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
-\neq 0$ and let $E$ be the subset of $K^2$ consisting of the couples 
+In this example, we assume for simplicity that $K$ does not have characteristic~$2$.
+Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
+\neq 0$ and let $E$ be the subset of $K^2$ consisting of the pairs 
 $(x,y)$ satisfying the usual equation $y^2 = x^3 + a x + b$.
-Let also $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
-takes a couple $(x,y)$ to $x$ (resp. to $y$).
-
-\medskip
-
-We first assume that $a$ and $b$ are exactly known; typically the field 
-$K$ is $\Q_p$ and $a$ and $b$ are actual integers. For each point $P_0 = 
+Let $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
+takes a pair $(x,y)$ to $x$ (resp. to $y$).
+
+We first assume that $a$ and $b$ lie in the subring $R$ of exact elements. For each point $P_0 = 
 (x_0, y_0)$ on $E$ except possibly a finite number of them, the map 
 $\pr_x$ define a diffeomorphism from an open subset containing $P_0$ to 
-an open subset of $K$. The same is true for $\pr_y$. Moreover around all 
-$P_0 \in E$, one of these projections satisfy the above condition and
-they both do generically. Hence the maps $\pr_x$ and $\pr_y$ defined an 
-various open subsets define together an atlas of $E$. This turns $E$ into 
+an open subset of $K$; the same is true for $\pr_y$. Moreover, around each 
+$P_0 \in E$, at least one of these projections satisfies the above condition.
+Hence the maps $\pr_x$ and $\pr_y$ define together an atlas of $E$, giving $E$ the structure of 
 a $K$-manifold.
@@
 charts. Proposition \ref{prop:independence} then tells us that a 
-precision datum on $x$ determined a precision datum on $y$ and \emph{vice 
-versa}. Indeed around $P_0$, we can write $y = \sqrt{x^3 + a x + b}$ 
-(with the correct determination of the square root) and find the 
-precision on $y$ from those on $x$ using Lemma \ref{lem:main}. We can go 
+precision datum on $x$ determines a precision datum on $y$ and vice 
+versa. Indeed, in a neighborhood of $P_0$ we can write $y = \sqrt{x^3 + a x + b}$ 
+(for some choice of square root) and find the 
+precision on $y$ from the precision on $x$ using Lemma \ref{lem:main}. We can go 
 in the other direction as well by writing $x$ locally as a function of 
@@
 a lattice in the tangent space is a nice way to make it canonical but in 
-practice we can just particularize one coordinate and handle precision
+practice we can just choose one coordinate and track precision
 only on this coordinate.
@@
 
-\medskip
-
-Let us address now the case where $a$ and $b$ are themselves given with 
-finite precision. The elliptic curve $E$ is then not fully determined. So 
+Finally, consider the case where $a$ and $b$ are themselves given with 
+finite precision and $E$ is not fully determined. So 
 we cannot consider it as a $K$-manifold and the above discussion does not 
@@
 n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
-such writings with the same $P$ necessarily coincide. Hence each 
+such expressions with the same $P$ necessarily coincide. Hence each 
 permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
-is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
+is canonically diffeomorphic to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
 
@@
 
-We continue to work over a fixed ultrametric field $K$, whose absolute 
-value is denoted by $|\cdot|$. We recall that usual examples of such $K$ 
-are the field of $p$-adic numbers $\Q_p$ and the field of Laurent series 
-$k((X))$ over a base field $k$.
-
 We saw in the core of the article that the differential of an operation 
@@
 operation. The aim of this appendix is to compute the differential of 
-many usual operations on numbers, polynomials and matrices. Surprisingly, 
+many common operations on numbers, polynomials and matrices. Surprisingly, 
 we observe that all differentials we will consider are rather easy to 
@@
 
-A little bit more interesting is the $n$-th power map $f$ from $K$ to 
-itself. Its differential $f'(x)$ is obtained by differentiate the 
-equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
-Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
-number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
-means that the behaviour of the precision depends on the absolute value 
-of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
-this inegality might be strict. It happens for instance if the 
+\todofor{David}{Change to account for valuation of $x$}
+
+Slightly more interesting is the $n$-th power map $f$ from $K$ to 
+itself. Its differential $f'(x)$ is found by differentiating the 
+equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
+Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
+means that the behavior of the precision depends on the absolute value 
+of $n$. By the ultrametric inequality, we always have $|n| \leq 1$ but 
+this inequality might be strict. It happens for instance if the 
 characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
@@
 polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
-space of dimension $d+1$. The affine space $X^d + K_{< d}[X]$ is then 
+space of dimension $d$. The affine space $X^d + K_{< d}[X]$ is then 
 the set of monic polynomials over $K$ of degree $d$. We denote it by 
@@
 operations involving polynomials are evaluation and interpolation.
-Evaluation is modelised by the function $(P,x) \mapsto P(x)$ where
-$P$ is some polynomial and $x$ some element in $K$. Differentiating
+Evaluation models the function $(P,x) \mapsto P(x)$, where
+$P$ is some polynomial and $x \in K$. Differentiating
 it can be done by computing
@@
 
-Let $A$ and $B$ be two polynomials $B \neq 0$. The Euclidean division of 
-$A$ by $B$ is written $A = BQ + R$ with $\deg R < \deg B$. 
-Differentiating the above equality we find:
-$dA - dB \cdot Q = B \cdot dQ + dR$,
+Let $A$ and $B$ be two polynomials with $B \neq 0$. The Euclidean division of 
+$A$ by $B$ is the problem of finding $Q$ and $R$ with $A = BQ + R$ and $\deg R < \deg B$. 
+Differentiating the above equality we find
+\[
+dA - dB \cdot Q = B \cdot dQ + dR,
+\]
 which implies that $dQ$ and $dR$ are respectively obtained as the 
@@
 function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
-couple $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
+pair $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
 greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
-coefficients of minimal degrees (they are those computed by extended 
-Euclide algorithm).
+coefficients of minimal degrees, computed by the extended 
+Euclidean algorithm.
 The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
@@
 $1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
-of positive degree. This implies that $f$ is not continuous outside 
-$\mathcal V_0$. On the contrary, on $\mathcal V_0$, the function $f$ is
-differentiable, and actually locally analytic.
+of positive degree.  Since $\mathcal V_0$ is dense, $f$ is not continuous outside 
+$\mathcal V_0$. Conversely, the function $f$ is differentiable, and even locally analytic,
+on $\mathcal V_0$.
 
@@
 
-In order to go further and differentiate $f$ outside $\mathcal V_0$, we 
-define, for all integer $i$, the subset $\mathcal V_i$ of $K_n[X] \times 
+In order to differentiate $f$ outside $\mathcal V_0$, we 
+define the subset $\mathcal V_i$ of $K_n[X] \times 
 K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
@@
 the Zariski topology. In particular, it defines a $K$-manifold in the 
-sense of Appendix \ref{sec:manifold}. This is interesting because the 
-restriction of $f$ to $\mathcal V_i$ is now differentiable. To compute
+sense of Appendix \ref{sec:manifold}, and the 
+restriction of $f$ to $\mathcal V_i$ is differentiable. To compute
 its differential, we proceed along the same lines as before: we 
@@
 However, the degree function is not continuous --- it is only upper 
-semi-continuous --- and hence cannot be determined from sure from $A$ and 
-$B$ (expect if they are coprime). We then need to make an extra 
-hypothesis and the more natural one is arguably to assume that 
-$\gcd(A,B)$ has actually the maximal degree that it can have. Once this
-assumption is made, the computation is possible and one can apply Lemma 
+semi-continuous --- and hence cannot be determined with certainty from $A$ and 
+$B$, unless the approximation to $(A,B)$ lies entirely within $\mathcal V_0$.
+
+We therefore need to make an extra hypothesis.  The most natural hypothesis to make
+is that $\gcd(A, B)$ has the maximal possible degree.  The main reason for choosing
+this convention is that if the actual polynomials $A, B \in K[X]$ have a greatest common
+divisor of degree $i$ then there is some precision for which the maximal degree will be
+equal to $i$, whereas if they have a positive
+degree common divisor then no amount of increased precision will eliminate an intersection
+with $\mathcal V_0$.  A second justification is that any other choice would yield a result
+with no precision, since $A$ and $B$ appear to lie in $\mathcal V_i$ at the known precision.
+
+Once this assumption is made, the computation is possible and one can apply Lemma 
 \ref{lem:main} to determine the precision of the result.
 Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
-itself although there exist pair of coprime polynomials in any 
+itself although there exist pairs of coprime polynomials in any 
 neighborhood of $(A,A)$.
@@
 Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
-product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
+product $P_0 = A_0 B_0$, where $A_0$ and $B_0$ are monic and coprime. 
 Hensel's lemma implies that there exists a small neighborhood $\mathcal 
@@
 follows the simple root $\alpha_0$. Of course, its differential around 
-$P_0$ can be computed by the above method. Nevertheless there is a more 
-direct way to obtain it in this particular case: we write $(P_0 + 
-dP)(\alpha_0 + d\alpha) = 0$ and expand this. We obtain this way 
-$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$ where $P'$ denotes the 
-derivative of $P$. Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
+$P_0$ can be computed by the above method. Nevertheless, we may
+obtain it more directly by expanding the equation $(P_0 + 
+dP)(\alpha_0 + d\alpha) = 0$, obtaining
+\[
+P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0.
+\]
+Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
 does not vanish and we find:
@@
 multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
-to follow to root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
+to follow the root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
 restrict ourselves to polynomials which have a root of multiplicity $m$ 
@@
 V_m$ consisting of monic polynomials of degree $d$ having a root of 
-multiplicity at least $m$. It is a Zariski close subset of $K_d[X]$ and 
-it contains by assumption the polynomial $P_0$ we started with. 
+multiplicity at least $m$. It is a Zariski closed subset of $K_d[X]$ and
+contains by assumption the polynomial $P_0$. 
 Moreover, the irreducible components of $\mathcal V_m$ that meet at 
 $P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
-$\geq m$. In particular $\alpha_0$ correspond to one of these 
+$\geq m$. In particular, $\alpha_0$ corresponds to one of these 
 irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
@@
 
-The above computation has actually interesting consequences on $p$-adic 
-precision. Let us illustrate this with an exemple. Consider the monic 
+The above computation has interesting consequences for $p$-adic 
+precision. For example, consider the monic 
 polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
-large integer) and assume that we are asking a softwate to compute one 
+large integer) and suppose that we are asking a computer to compute one 
 of its roots. What is the right answer? We remark that, if $\alpha$ is 
@@
 roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
-indeed multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
+multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
 where $c$ is some constant (depending on $P$ and $\alpha$ but not on
@@
 We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
-are polynomials in $K[\XX]$, there exists a writing:
+are polynomials in $K[\XX]$, then one may decompose $f$ as
 $$f = q_1 f_1 + \cdots + q_s f_s + r$$
@@
 by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
-that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
-assumption, there still exists a canonical choice of $r$. We do not know
-if the computation of the differential that follows extends to a more
-general setting.} (of the ideal generated by these polynomials), the 
+that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{A canonical choice of $r$ exists even without this
+assumption. However, we do not know
+if the computation of the differential that follows extends to this more
+general setting.} of the ideal generated by these polynomials, the 
 polynomial $r$ is uniquely determined and called the \emph{remainder} of 
 the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
-$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
-the case of univariate polynomials, we can compute its differential: we 
-find that $dr$ is obtained as the remainder of the division of $f - (q_1 
+$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and we can compute its differential,
+finding that $dr$ is obtained as the remainder of the division of $f - (q_1 
 \cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
@@
 Recall that any 
-ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
-then consider the map sending a family $(f_1, \ldots, f_s)$ of 
+ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We may
+consider the map sending a family $(f_1, \ldots, f_s)$ of 
 \emph{homogeneous}\footnote{This restriction will be convenient for us 
 mainly because we are using the article \cite{vaccon:14a}, which deals only 
-with homogeneous polynomials. It is however probably not essential.} 
+with homogeneous polynomials. It is probably not essential.} 
 polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from Theorem 1.1 of \cite{vaccon:14a} that this 
+\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
 map is continuous at $(f_1, \ldots, f_s)$ provided that:
@@
 \item for all $i \in \{1, \ldots, s\}$, the ideal generated by
-$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{vaccon:14a} for more precisions).
+$f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
+order (\emph{cf} \cite{vaccon:14a} for more detail).
 \end{itemize}
@@
 \end{equation}
-where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
-Moreover following Vaccon's construction, we see that the entries of $A$ 
+where $A$ is an $(s \times t)$ matrix with coefficients in $K[\XX]$.
+Following Vaccon's construction, we see that the entries of $A$ 
 can be chosen in such a way that they define differentiable functions.
-Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
+Differentiating \eqref{eq:prodgrobner}, we get
 $$(d g_1, \ldots, d g_t) =
 (d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
-Finally, using that we are working with \emph{reduced} Gr\"obner
+Finally, since $(g_1, \ldots, g_t)$ is a \emph{reduced} Gr\"obner
 basis, the above equality implies that $d g_i$ is the remainder in
-the division of the $i$-th coefficient of the product
+the division of the $i$-th entry of the product
 $(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

### 2014-02-20 padicprec.tex
Added periods to subsubsections

@@
 
-\subsubsection*{LU factorization}
+\subsubsection*{LU factorization.}
 
@@
 
-\subsubsection*{One-pass computation}
+\subsubsection*{One-pass computation.}
 
@@
 
-\subsubsection*{Two-pass computation}
+\subsubsection*{Two-pass computation.}
 
@@
 
-\subsubsection*{Non-surjectivity}
+\subsubsection*{Non-surjectivity.}
 
@@
 
-\subsubsection*{Change of base point}
+\subsubsection*{Change of base point.}
 
@@
 
-\subsubsection*{Elliptic curves}
+\subsubsection*{Elliptic curves.}
 
@@
 
-\subsubsection*{Grassmannians}
+\subsubsection*{Grassmannians.}
 
@@
 
-%\subsubsection*{Conclusion}
+%\subsubsection*{Conclusion.}
 %
@@
 
-\subsubsection*{Evaluation and interpolation}
+\subsubsection*{Evaluation and interpolation.}
 
@@
 
-\subsubsection*{Euclidean division}
+\subsubsection*{Euclidean division.}
 
@@
 
-\subsubsection*{Greatest common divisors and B\'ezout coefficients}
+\subsubsection*{Greatest common divisors and B\'ezout coefficients.}
 
@@
 
-\subsubsection*{Factorization}
+\subsubsection*{Factorization.}
 
@@
 
-\subsubsection*{Finding roots}
+\subsubsection*{Finding roots.}
 
@@
 
-\subsubsection*{Division}
+\subsubsection*{Division.}
 
@@
 
-\subsubsection*{Gr\"obner basis}
+\subsubsection*{Gr\"obner basis.}
 
@@
 
-\subsubsection*{Determinants and characteristic polynomials}
+\subsubsection*{Determinants and characteristic polynomials.}
 
@@
 
-\subsubsection*{QR factorization}
+\subsubsection*{QR factorization.}
 
@@
 
-\subsubsection*{Left kernels}
+\subsubsection*{Left kernels.}
 
@@
 
-\subsubsection*{Intersections}
+\subsubsection*{Intersections.}
 
@@
 
-\subsubsection*{Sums and images}
+\subsubsection*{Sums and images.}
 

### 2014-02-20 padicprec.tex
Finish introduction

@@
 Section \ref{sec:mainlemma} centers on Lemma \ref{lem:main}, our main result.
-By describing how lattices transform under non-linear maps, it allows us to propogate
-precision using differentials.  In Section
+By describing how lattices transform under non-linear maps, it allows us to propagate
+precision using differentials.  More specifically, it describes a class of first order lattices,
+whose image under a map of Banach spaces is precisely described by applying the
+differential of that map.  In Section \ref{ssec:locanalytic} we make the conditions
+of Lemma \ref{lem:main} more explicit in the case of locally analytic functions.
+
+In Section \ref{sec:tracking} we describe methods for tracking precision in practice.
+Section \ref{ssec:opt-tracking} includes a discussion of two models of precision tracking:
+one-pass tracking, where the precision lattice is propagated at each step of the algorithm,
+and two-pass tracking, where an initial pass computing rough approximations is used
+in computing the precision lattices.  In Section \ref{ssec:SOMOS-solution}, we give
+an application of these ideas to an algorithm for computing terms of the SOMOS sequence.
+
+In Appendix \ref{sec:manifold}, we extend the results of Section \ref{sec:mainlemma}
+to $p$-adic manifolds, describing how to specify precisions for points on elliptic curves
+and Grassmannians.  Finally, Appendix \ref{sec:differentials} describes how to compute
+the derivative of many common operations on polynomials and matrices, with an eye
+toward applying Lemma \ref{lem:main}.
 
@@
 
-We emphasize that we require an equality in Eq.~\eqref{eq:firstorder}, 
+We emphasize that we require an equality in \eqref{eq:firstorder}, 
 and not just an inclusion! 
@@
 
-Using Eq.~\eqref{eq:mainlemma} and differentiability at $0$ once more,
+Using \eqref{eq:mainlemma} and differentiability at $0$ once more,
 we get
@@
 on $i$, we find $\Lambda (f)(v) \geq u$. Comparing with 
-Eq.~\eqref{eq:lowerLambda}, we get $\Lambda (f)(v) = u = 
+\eqref{eq:lowerLambda}, we get $\Lambda (f)(v) = u = 
 \NP(f)^\star(v)$.
@@
 the functions $f_n$ --- and then $f$ also --- extend to an algebraic
-closure $\bar K$ of $K$ and Eq.~\eqref{eq:boundqueue} holds over $\bar 
+closure $\bar K$ of $K$ and \eqref{eq:boundqueue} holds over $\bar 
 K$, which is enough to conclude.
@@
 \subsection{Optimal precision tracking}
+\label{ssec:opt-tracking}
 
@@
 \item \label{item:smallprec}
-for $i$ going from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
+for $i$ from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
 at small precision (but enough for the second step) together with an
@@
 \item \label{item:determineHi}
-for $i$ going from $n$ to $1$, we compute $h'_i(x_i)$ and
-determines a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
+for $i$ from $n$ to $1$, we compute $h'_i(x_i)$ and
+determine a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
 \eqref{item:Hi2};
 \item \label{item:finalcomp}
-for $i$ going from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
+for $i$ from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
 and finally outputs $x_n + O\big(f'(x)(H)\big)$.
@@
 and using the induction hypothesis. 
-Let us now establish Eq.~\eqref{eq:congrSOMOS}. To avoid confusion, 
+Let us now establish \eqref{eq:congrSOMOS}. To avoid confusion, 
 let us agree to denote $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
@@
 
-Finally Eq.~\eqref{eq:congrSOMOS} applied with $i = n-3$ together with
+Finally \eqref{eq:congrSOMOS} applied with $i = n-3$ together with
 the fact that $\varphi_i$ stabilizes $\Z_p^4$ imply that, when we exit
@@
 In the above $dx$ and $dy$ should be thought as a little perturbation of 
-$x$ and $y$ respectively. Eq.~\eqref{eq:diffec} then gives a linear 
+$x$ and $y$ respectively. Equation \eqref{eq:diffec} then gives a linear 
 relation between the precision on $x$ (which is represented by $dx$) and 
@@
 \section{Differential of usual operations}
-\label{ssec:differentials}
+\label{sec:differentials}
 
@@
 y_d)$ is a point in it and $P$ denotes the corresponding interpolation 
-polynomial, Eq.~\eqref{eq:diffeval} shows that $d y_i = P'(x_i) dx_i + 
+polynomial, \eqref{eq:diffeval} shows that $d y_i = P'(x_i) dx_i + 
 dP(x_i)$ for all $i$. For this, we can compute $dP(x_i)$ from $d x_i$ 
@@
 $\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
-between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
+between $A_0$ and $B_0$, it follows from \eqref{eq:difffactor} that
 $dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
@@
 This is exactly the result we get by applying Lemma \ref{lem:main} and
-using Eq.~\eqref{eq:dalphamult} to simplify the differential.
+using \eqref{eq:dalphamult} to simplify the differential.
 
@@
 Differentiating ring operations (sum, products, inverse) over matrices 
-is again straightforward; we just need to be careful that matrix 
-algebras are (in general) noncommutative.
+is again straightforward, keeping in mind that matrix algebras are not
+commutative.
 
@@
 
-Let us first compute the differential of the function $\det : M_n(K) \to 
-K$. It is a standard computation. To begin with, we consider an 
-invertible matrix $M$ and write:
-\begin{eqnarray*}
-\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
-&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
-&=& \det(M) + \tr(\com(M) \cdot dM)
-\end{eqnarray*}
-where $\com(M)$ denotes the comatrix of $M$. The differential of 
-$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
+We first outline the standard computation of the differential of the function
+$\det : M_n(K) \to K$. Suppose that $M \in \GL_n(K)$ and that $\com(M) = \det(M) M^{-1}$.
+Then
+\begin{align*}
+\det(M + dM) &= \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
+&= \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
+&= \det(M) + \tr(\com(M) \cdot dM).
+\end{align*}
+The differential of $\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
 out that this formula is still valid when $M$ is not invertible.
@@
 
-Let us agree to define a 
-LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
+Define the LU factorization of a square matrix $M \in M_n(K)$ as a decomposition $M = LU$ 
 where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
-Such a writing exists and is unique provided that all principal minors 
-of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
+Such a decomposition exists and is unique provided that no principal minor
+of $M$ vanishes. We can then consider the mapping $M \mapsto (L,U)$ 
 defined over the Zariski-open set of matrices satisfying the above 
 condition. In order to differentiate it, we differentiate the relation 
-$M = LU$ and rewrite the result as follows:
+$M = LU$ and rewrite the result as
 $$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
 We remark that in the right hand side of the above formula, the first
-summand is lower triangular and all its diagonal entries vanish whereas 
+summand is lower triangular with zero diagonal whereas 
 the second summand is upper triangular. Hence in order to compute $dL$
-and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
-L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
-obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
-recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
-left and $U$ on the right respectively. 
+and $dU$, one can proceed as follows: 
+\begin{enumerate}
+\item compute the product $dX = L^{-1} dM \: U^{-1}$,
+\item separate the lower and upper part of $dX$ obtaining $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$
+\item recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the left and $U$ on the right respectively.
+\end{enumerate}
 The above discussion extends almost \emph{verbatim} to LUP 
@@
 
-For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
-a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
+A QR factorization of a square matrix $M \in M_n(K)$ will be
+a decomposition $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
 orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
-a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
+a decomposition exists and is unique on a Zariski-open subset of $M_n(K)$. The
 mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
 would like to emphasize at this point that the orthogonality condition
-defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
-vector space (it is defined by equations of degree $2$). The codomain
-of $f$ is then also a manifold; this example then fits to the setting
+defines a sub-manifold of $M_n(K)$ which is \emph{not} a
+vector space: it is defined by equations of degree $2$. The codomain
+of $f$ is then also a manifold; this example then fits into the setting
 of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
-The generalization made in Appendix \ref{sec:manifold} is then needed.
-Anyway, we can differentiate $f$ by following the method we used for LU 
-factorization: differentiating the relation $M = QR$, we obtain:
+We can differentiate $f$ by following the method used for LU 
+factorization.  Differentiating the relation $M = QR$, we obtain
 \begin{equation}
@@
 triangular with all diagonal entries equal to $0$, we see that 
-Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
+\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
 and $dM$.
@@
 sense of Appendix \ref{sec:manifold} as it is discussed in \S 
-\ref{ssec:exmanifold}. In what sequel, we use freely the formalism 
+\ref{ssec:exmanifold}. In the sequel, we use freely the formalism 
 introduced there.
@@
 the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
-kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We deduce
-for this that $f_d$ is differentiable and that its differential can be
-computed similarly to what we have done before.
+kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We thus deduce
+that $f_d$ is differentiable and that its differential can be
+computed as above.
 

### 2014-02-20 padicprec.tex
Switch to one-pass and two-pass notation

@@
 Step~\eqref{item:finalcomp} the computations already performed in 
-Step~\eqref{item:smallprec}. The ``backward'' method we have just 
-presented is then probably not much more expansive than the ``forward'' 
-method, although it is certainly much more difficult to implement.
-
-We conclude this section by remarking that the ``backward'' method seems 
-to be particularly well suited to computations with lazy $p$-adic. Indeed 
-in this setting, a target precision (\emph{i.e.} a precision on the 
-output) is fixed and the software determines automatically the precision 
-it needs on the input to perform the computation. To do this, it first 
+Step~\eqref{item:smallprec}. The two-pass method we have just 
+presented is then probably not much more expansive than the one-pass 
+method, although it is more difficult to implement.
+
+We conclude this section by remarking that the two-pass method seems 
+to be particularly well suited to computations with lazy $p$-adics.
+In this setting, a target precision is fixed and the software determines automatically the precision 
+it needs on the input to achieve this output precision. To do this, it first 
 builds the ``skeleton'' of the computation (\emph{i.e.} it determines the 
@@
 and then runs over this skeleton in the reverse direction in order to 
-determine (an upper bound of) the needed precision at each step. The 
-similarity with our approach is noticeably strong.
+determine (an upper bound of) the needed precision at each step.
 
@@
 at some finite precision such that 
-$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
-and/or the backward computation and get $f_{ij}(M)$ together with its 
+$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply a one- or two-pass 
+computation and get $f_{ij}(M)$ together with its 
 precision. Putting this together, we get the whole matrix $f(M)$ 
@@
 
-We conclude this section by briefly remarking that the two strategies designed 
-above (\emph{i.e.} forward and backward computations) share some 
+We conclude this section by briefly remarking that the two strategies described 
+above share some 
 similarities with usual floating point arithmetics over the reals. 

### 2014-02-20 padicprec.tex
Update Section 2.3

@@
 then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.
-% Is every lattice of this shape for some Banach basis?
 
@@
 
+\subsubsection*{SOMOS 4.}
+
 The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
@@
 
-% Frame better
-
-Let us first recall that a square matrix $M$ with coefficients in $\Qp$ 
+Let us first recall that a square matrix $M$ with coefficients in $K$ 
 admits a LU factorization if it can be written as a product $LU$ where 
@@
 The computation of a LU factorization appears as an important tool to 
-tackle many classical questions about matrices or linear systems.
-
-We first note that a LU decomposition of a given matrix $M$ is unique
-if we require further that the $L$-part of this decomposition is
-unipotent (\emph{i.e.} all diagonal entries equal $1$). Let us write
-$L(M) U(M)$ the LU factorization of $M$ satisfying the above extra
-condition (when it exists). There exists Cramer-type formulae that
-gives each individual entry of $L(M)$ and $U(M)$ in a closed form
-(see \cite{caruso:12a}).
-If $M$ is random square matrix over $\Zp$ of side $d$ whose every
-coefficient is known up to precision $O(p^N)$, the two following
-results are proved in \cite{caruso:12a}:
+tackle many classical questions about matrices or linear systems, and is
+discussed further in Appendix \ref{ssec:matrices}.
+
+When computing the entries of $L$ and $U$ from a $d \times d$ matrix
+over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
 \begin{itemize}
-\item using usual Gauss elimination and tracking $p$-adic precision 
+\item using usual Gaussian elimination and tracking $p$-adic precision 
 step-by-step, the smallest precision on an entry of $L(M)$ is about
@@
 result whose every entry is known up to precision $O(p^{N - 2 \log_p 
-d})$.
+d})$ \cite{caruso:12a}.
 \end{itemize}
 If $d$ is large compared to $p$, the second precision is much more
-accurate than the first one. This shows once again that ``artifical''
-losses of precision appear when one performs Gauss elimination by
-tracking precision step-by-step.
+accurate than the first one.  The precision loss inherent in an algorithm
+is an important factor in analyzing its performance.  However,
+since this precision loss depends heavily on the details of the problem
+and the algorithm, we will not focus on such questions in this paper.
 
@@
 \end{proof}
-
-%Now, let $E$ and $F$ be two complete normed $K$-vector spaces, and let $f$ and the $L_k$'s be as in definition \ref{deftn:analyticNP}.
-We assume $L_1$ is surjective.
-We define, the $\overline{L_k},$ $\overline{E}$ and $\overline{F}$ by extending scalars to $\overline{K}.$ We define for any $n \geq 1$, $f_n(x)=\overline{L_n}(x,\dots,x)$ for any $x \in \overline{E}.$
-%Let $C$ be given by the open mapping theorem for $L_1$, $\rho \in (0, 1]$ and $\varepsilon < \frac{\rho}{C}$.
-We provide three methods to find a sufficient $\delta$ to define first order lattice as in lemma \ref{lem:main}.
-%
-%For the first one, we assume that we have know a function that majorates $\Lambda(\sum_{n \geq 1}f_n)$.
 
@@
 
-%\section{Conclusions}
-
 \appendix
@@
 
-%\subsubsection*{Conclusion.}
-%
-%The theory developped in this Appendix is nice because it provides an 
-%abstract framework to deal with ultrametric precision in a quite general 
-%setting.
-%Nevertheless, as we have seen in the above examples, in practice we
-%generally carry out all calculations in charts and do not really care
-%about the ``real'' nature of the precision data, \emph{i.e.} lattices
-%in tangent spaces.
-
 \section{Differential of usual operations}
@@
 
-\todofor{David}{Change to account for valuation of $x$}
-
 Slightly more interesting is the $n$-th power map $f$ from $K$ to 
@@
 equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
-Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
+Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(\lvert nx^{n-1} \rvert \cdot r)$. According to Lemma \ref{lem:main}, this 
 means that the behavior of the precision depends on the absolute value 
-of $n$. By the ultrametric inequality, we always have $|n| \leq 1$ but 
-this inequality might be strict. It happens for instance if the 
-characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
-$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
-not apply. 
-Another situation (which is more interesting) is that of $p$-adic 
-numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
-$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
+of $n$. By the ultrametric inequality, we always have $\lvert n \rvert \leq 1$ but 
+may have $\lvert n \rvert = 0$ if the characteristic of $K$ divides $n$ or 
+$\lvert n \rvert = p^{-k}$ if $K$ is $p$-adic with $\val(n) = k$.  In the first case,
+$f'(x)$ also vanishes and Lemma \ref{lem:main} does 
+not apply.  In the second, Lemma \ref{lem:main} reflects the well-known fact that
 raising a $p$-adic number to the $p^k$-th power increases the precision 
@@
 \subsection{Matrices}
+\label{ssec:matrices}
 

### 2014-02-20 padicprec.tex
Update title and abstract

@@
 
-\title{$p$-adic precision}
+\title{Tracking $p$-adic precision}
 %\author{Xavier Caruso, David Roe \& Tristan Vaccon}
@@
 \begin{abstract}
-We present a new method to handle $p$-adic precision (or more generally
-precision over ultrametric fields) based on $p$-adic calculus.
-We illustrate it by many examples and give a toy application to the
+We present a new method to propagate $p$-adic precision in computations, which also
+applies to other ultrametric fields.
+We illustrate it with many examples and give a toy application to the
 stable computation of the SOMOS 4 sequence.

### 2014-02-20 padicprec.tex
A few changes to Section 3

@@
 \label{rem:differentiability}
-This notion of differentiability is stronger than the usual notion with 
-which we are familiar. It implies for instance the following: if $f$ is 
+This notion of differentiability is stronger than the standard definition.
+It implies for instance the following: if $f$ is 
 differentiable on $U$, then the function $x \mapsto f'(x)$ defined on 
@@
 It is sometimes called \emph{strict differentiability} but, for this
-article, we prefer a simplier terminology since we shall only use this
+article, we prefer a simpler terminology since we shall only use this
 notion of differentiability.
@@
 
-One can derive all usual properties of the differential from this 
+One can derive all usual properties of the derivative from this 
 definition: unicity, chain rules, \emph{etc}.
@@
 \varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r) 
-\subset \varepsilon f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
+\subset f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
 \in \cdot B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
@@
 
-We consider in this subsection the case where $f$ is locally analytic: 
-under this extra assumption, we make explicit the constant $\delta$ 
+We consider in this subsection the case where $f$ is locally analytic.
+Under this extra assumption, we make explicit the constant $\delta$ 
 which appears in Lemma \ref{lem:main}.
@@
 A function $f \col U \rightarrow F$ is said \emph{locally analytic} at 
-$x$ if there exist an open subset $U_x \subset E$ and, for all $n \ge 
-1$, a continuous $n$-linear map $L_n \col E^n \rightarrow F$ such that
-$$f(x+h) = f(x) + \sum_{n \geq 1} L_n(h,\ldots,h)$$
+$x$ if there exist an open subset $U_x \subset E$ and continuous $n$-linear
+maps $L_n \col E^n \rightarrow F$ for $n \ge 1$ such that
+\[
+f(x+h) = f(x) + \sum_{n \geq 1} L_n(h,\ldots,h)
+\]
 for all $h$ with $x + h \in U_x$.
@@
 
-From now, we assume that $K$ is algebraically closed. As in Definition 
+For the rest of this section, we assume that $K$ is algebraically closed. As in Definition 
 \ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
@@
 in a neighborhood of $0$. Under this assumption, the datum of $f$ 
-uniquely determines the $f_n$'s (it is a consequence of Propostion 
-\ref{prop:lambdaNP} below).
+uniquely determines the $f_n$'s (a consequence of Proposition 
+\ref{prop:lambdaNP}).
 To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
@@
 The function $\Lambda(f)$ we have just defined is actually closely 
-related to the Newton polygon of $f$. Recall that, by definition, the 
+related to the Newton polygon of $f$. Recall that the 
 Newton polygon of $f$ is the convex hull in $\R^2$ of the points of 
@@
 $\varphi : \R \to \R \cup \{+\infty\}$ is the function $\varphi^\star : 
-\R \to \R \cup \{+\infty\}$ defined by:
-$$\textstyle \varphi^\star(v) = \sup_{u \in \R} \:
-\big(uv - \varphi(u)\big), \qquad
-\forall v \in \R.$$
+\R \to \R \cup \{+\infty\}$ defined by
+\[
+\textstyle \varphi^\star(v) = \sup_{u \in \R} \:
+\big(uv - \varphi(u)\big),
+\]
+for $v \in \R$.
 One can check that the map $\varphi \mapsto \varphi^\star$ is an
@@
 \NP(f)^*(v)$. The function $m \mapsto \NP(f)(m) - vm + u$ has the 
-following properties: (1)~it is piecewise affine and everywhere 
-nonnegative, (2)~it does not admit $0$ as a slope and (3)~it vanishes at 
-$x = n$ for some integer $n$ and $u = vn + \log \Vert f_n \Vert$.
-We deduce from this that there exists $c>0$ such that $vm-u \leq - \log 
-\lVert f_m \rVert - c \cdot \vert n-m \vert$ for any $m\geq 0$.
-Since $vm-u=vm-vn-\log \lVert f_n \rVert$, we get $-vn-\log \lVert 
-f_n \rVert + c \cdot \vert n-m \vert \leq -vm-\log \lVert f_m \rVert$.
+following properties: 
+\begin{enumerate}
+\item it is piecewise affine and everywhere nonnegative,
+\item it does not admit $0$ as a slope and
+\item it vanishes at $x = n$ for some integer $n$ and $u = vn + \log \Vert f_n \Vert$.
+\end{enumerate}
+We deduce from these facts that there exists $c>0$ such that 
+\[
+vm-u \leq - \log \lVert f_m \rVert - c \cdot \vert n-m \vert
+\]
+for any $m\geq 0$.
+Since $vm-u=vm-vn-\log \lVert f_n \rVert$, we get 
+\[
+-vn-\log \lVert f_n \rVert + c \cdot \vert n-m \vert \leq -vm-\log \lVert f_m \rVert.
+\]
 Therefore, for any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
-$$\Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
-\lVert f_n \rVert \cdot e^{vn}< \lVert f_n \rVert \cdot e^{vn}.$$
+\[
+\Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
+\lVert f_n \rVert \cdot e^{vn}< \lVert f_n \rVert \cdot e^{vn}.
+\]
 Thus, the series $\sum_{m \geq 0} f_m(x)$ converges and $\Vert f(x) 
@@
 It follows from Proposition \ref{prop:lambdaNP} that $\Lambda(f)$ is a
-convex function (which was not clear at all \emph{a priori}).
+convex function.
 \end{rem}
@@
 We now study the effect of truncation on series: given $f$ as above
-and a nonnegative integer $n_0$, we set:
-$$f_{\geq n_0} = \sum_{n \geq n_0} f_n = f - (f_0 + f_1 + \cdots +
-f_{n_0 - 1}).$$
+and a nonnegative integer $n_0$, we set
+\[
+f_{\geq n_0} = \sum_{n \geq n_0} f_n = f - (f_0 + f_1 + \cdots +
+f_{n_0 - 1}).
+\]
 On the other hand, given a convex function $\varphi : \R \to \R \cup
@@
 closure $\bar K$ of $K$ and \eqref{eq:boundqueue} holds over $\bar 
-K$, which is enough to conclude.
+K$, which is enough to conclude the result.
 \end{rem}

### 2014-02-20 padicprec.tex
Add section on precision types

@@
 \DeclareMathOperator{\NP}{NP}
+\DeclareMathOperator{\GL}{GL}
+\DeclareMathOperator{\val}{val}
+\DeclareMathOperator{\pr}{pr}
+\DeclareMathOperator{\tr}{Tr}
+\DeclareMathOperator{\com}{Com}
+\DeclareMathOperator{\Grass}{Grass}
+\DeclareMathOperator{\Lat}{Lat}
+\DeclareMathOperator{\round}{round}
 
@@
 \newcommand{\trans}{{}^{\text t}}
-
-\newcommand{\GL}{\text{\rm GL}}
-
-\newcommand{\val}{\text{\rm val}}
-\newcommand{\pr}{\text{\rm pr}}
-\newcommand{\tr}{\text{\rm Tr}}
-\newcommand{\com}{\text{\rm Com}}
-\newcommand{\Grass}{\text{\rm Grass}}
+\newcommand{\T}{\mathcal{T}}
+
 \renewcommand{\prec}{\text{\rm prec}}
@@
 
+\subsection{Precision Types}
+
+Using an arbitrary lattice to record the precision of an approximate element
+has the benefit of allowing computations to proceed without unnecessary
+precision loss using Lemma \ref{lem:main}.  However, while recording a
+lattice exactly is possible it does require a lot of space.  For example,
+the space required to store a lattice precision for a single $n \times n$
+matrix with entries of size $O(p^N)$ is $O(Nn^4)$.  Conversely, the space
+needed to record that every entry has precision $O(p^N)$ is just $O(\log N)$.
+
+\begin{deftn}
+Suppose that $E$ is a $K$-Banach space, and write $\Lat(E)$ for the set of lattices in $E$.
+A \emph{precision type} for a $K$-Banach space $E$ is a set $\T \subseteq \Lat(E)$ together
+with a function $\round : \Lat(E) \to \T$ such that
+\begin{itemize}
+\item[$(\ast)$] For every lattice $H \in \Lat(E)$, the lattice $\round(H)$ is a least upper bound for $H$ under the inclusion order:
+$H \subseteq \round(H)$ and if $T \in \T$ satisfies $T \subset \round(H)$ then $H \not\subseteq T$.
+\end{itemize}
+\end{deftn}
+
+Different precision types are appropriate for different problems.  For example,
+the final step of Kedlaya's algorithm for computing zeta functions of hyperelliptic
+curves involves taking the characteristic polynomial of the matrix of Frobenius acting
+on a $p$-adic cohomology space.  Obtaining extra precision on the entries of the
+matrix requires a long computation, so it is advisable to work with a precision type
+that does not round too much.
+
+The following list gives examples of useful precision types.  A description of the $\round$ function has been omitted for brevity.
+\begin{itemize}
+\item The \emph{lattice} precision type has $\T = \Lat(E)$.
+\item In the \emph{jagged} precision type, $\T$ consists of lattices of the shape $B_E((e_i),(r_i))$ for a fixed Banach basis $(e_i)$ of $E$.
+\item In the \emph{flat} precision type, $\T$ consists of lattices $B_E(r)$.  The flat precision type is useful since it takes so little space to store and it easy to compute with.
+\item If $E = K_{<d}[X]$ is the space of polynomials of degree less than $d$, the \emph{Newton} precision type consists of lattices $B_E((X^i),(r_i))$ where $-\log r_i$ is a convex function of $i$.
+The Newton precision type is sensible if one thinks of polynomials as functions $K \to K$, since extra precision above the Newton polygon never increases the precision of an evaluation.
+\item If $E = M_{m \times n}(K)$, the \emph{column} precision type consists of lattices with identical image under all projections $\pr_i : E \to K^m$ sending a matrix to its $i$th column.
+It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
+\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.
+It is important when working with overconvergent modular symbols since these lattices are stable under certain Hecke operators.
+\end{itemize}
+
+A final benefit of separating precision from approximation is that it makes
+it much easier to implement algorithms capable of processing different precision types.
+
 \section{The main Lemma}
@@
 
-%The following Lemma shows that the notion of lattices behaves rather 
-%well under linear\footnote{We shall see in the sequel (\emph{cf} Lemma 
-%\ref{lem:main}) that, under some extra hypothesis, the linearity 
-%assumption can be drastically relaxed.} transformations.
-%
-%\begin{lem}
-%\label{lem:morlat}
-%Let $E$ and $F$ be two $K$-Banach and $f : 
-%E \to F$ be a continuous \emph{surjective} linear map.
-%If $H$ is a lattice in $E$, then $f(H)$ is a lattice in $F$.
-%\end{lem}
-%
-%\begin{proof}
-%It is clear that $f(H)$ is a module over $B_K(1)$ which is bounded.
-%The fact that it is open is a direct consequence of the open mapping
-%Theorem.
-%\end{proof}
-%
-%To each lattice $H$ in $E$, one can attach a norm $\Vert \cdot \Vert_H$ 
-%on $E$ defined as follows. We pick $x \in E$, $x \neq 0$. We introduce
-%the set $I_x$ consisting of scalars $\lambda \in K$ such that $\lambda x 
-%\in H$. It follows from the fact that $H$ is open and bounded in $E$
-%that $I_x$ is nonempty and bounded in $K$. Therefore, the image $|I_x|$ 
-%of $I_x$ under the absolute value map is bounded and contains a positive
-%element. We can then define:
-%$$\Vert x \Vert_H = (\sup |I_x|)^{-1}.$$
-%It is an exercise to check that $\Vert \cdot \Vert_H$ is a norm over 
-%$E$.
-%
-%\begin{lem} \label{lem:reductionball}
-%\begin{enumerate}
-%\item The norms $\Vert \cdot \Vert_H$ and $\Vert \cdot \Vert$ are 
-%equivalent.
-%\item Let $B^-_{E,H}(1)$ (resp. by $B_{E,H}(1)$) the set of elements $x 
-%\in E$ such that $\Vert x \Vert_H < 1$ (resp. $\Vert x \Vert_H \leq 1$).
-%Then:
-%$$B^-_{E,H}(1) \subset H \subset B_{E,H}(1)$$
-%and the last inclusion is an equality if the absolute value on $K$ is
-%discrete.
-%\end{enumerate}
-%\end{lem}
-%
-%\begin{proof}
-%Left to the reader.
-%\end{proof}
-%
-%The first part of Lemma \ref{lem:reductionball} asserts that the 
-%identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
-%\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
-%$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
-%assumption: it is ``between the open unit ball and the closed unit 
-%ball''.
-
 \begin{deftn}
@@
 
-\begin{proof} % Call it the Banach-Schauder theorem?
+\begin{proof}
 We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 

### 2014-02-20 padicprec.tex
Switched back to roebib/Biblio

@@
 Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
-\cite{hoeven1, hoeven2, berthomieu}), we can reuse in 
+\cites{hoeven:02a, hoeven:07a, berthomieu-hoeven-lecerf:11a}), we can reuse in 
 Step~\eqref{item:finalcomp} the computations already performed in 
@@
 \bibliographystyle{plain}
-\bibliography{Biblio}
+\bibliography{roebib/Biblio}
 

### 2014-02-20 padicprec.tex
Small fixes

@@
 \item Kedlaya \cite{kedlaya:01a} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
-\item Lercier and Survent \cite{lercier-sirvent:08a} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
+\item Lercier and Sirvent \cite{lercier-sirvent:08a} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
 \end{itemize}
@@
 \[
-\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert \le 2^{-h}
+\Big\vert \frac{x \circledast y}{x \ast y} - 1 \Big\vert \le 2^{-h}
 \]
@@
 \[
-\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert_p \le p^{-h}.
+\Big\vert \frac{x \circledast y}{x \ast y} - 1 \Big\vert_p \le p^{-h}.
 \]
@@
 
-Section \ref{sec:mainlemma} centers on Lemma \ref{lem:main}, our main result.
-By describing how lattices transform under non-linear maps, it allows us to propagate
+In Section \ref{sec:mainlemma}, we develop the mathematical background on which our proposal is based.
+The most notable result of this section is Lemma \ref{lem:main} which describes how lattices transform under non-linear maps and allows us to propagate
 precision using differentials.  More specifically, it describes a class of first order lattices,
-whose image under a map of Banach spaces is precisely described by applying the
+whose image under a map of Banach spaces is obtained by applying the
 differential of that map.  In Section \ref{ssec:locanalytic} we make the conditions
@@
 
-In Section \ref{sec:tracking} we describe methods for tracking precision in practice.
+In Section \ref{sec:tracking} we propose methods for tracking precision in practice.
 Section \ref{ssec:opt-tracking} includes a discussion of two models of precision tracking:
@@
 A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
+We underline that any lattice $H$ in $E$ is also closed since its 
+complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
+are all open.
 
@@
 
-Suppose that $E \simeq K^d$ is finite dimesional.  Then if $H \subset E$
+Suppose that $E \simeq K^d$ is finite dimensional.  Then if $H \subset E$
 is a lattice then there exist $a, b \in \Q_{>0}$ with
@@
 discussed further in Appendix \ref{ssec:matrices}.
-
 When computing the entries of $L$ and $U$ from a $d \times d$ matrix
@@
 If $d$ is large compared to $p$, the second precision is much more
-accurate than the first one.  The precision loss inherent in an algorithm
-is an important factor in analyzing its performance.  However,
-since this precision loss depends heavily on the details of the problem
-and the algorithm, we will not focus on such questions in this paper.
+accurate than the first one. On the other hand, the second algorithm
+is less efficient than the first one because evaluating Cramer-type
+formulae requires many computations. 
+%In \cite{caruso:12a}, the author designs a new algorithm which combines 
+%both advantages: efficiency and stability.
+%The precision loss inherent in an algorithm is an important factor in 
+%analyzing its performance.
+%However, since this precision loss depends heavily on the details of the 
+%problem and the algorithm, we will not focus on such questions in this 
+%paper.
 
@@
 
-Using an arbitrary lattice to record the precision of an approximate element
-has the benefit of allowing computations to proceed without unnecessary
-precision loss using Lemma \ref{lem:main}.  However, while recording a
-lattice exactly is possible it does require a lot of space.  For example,
-the space required to store a lattice precision for a single $n \times n$
-matrix with entries of size $O(p^N)$ is $O(Nn^4)$.  Conversely, the space
-needed to record that every entry has precision $O(p^N)$ is just $O(\log N)$.
+Using an arbitrary lattice to record the precision of an approximate 
+element has the benefit of allowing computations to proceed without 
+unnecessary precision loss using Lemma \ref{lem:main}.  However, while 
+recording a lattice exactly is possible it does require a lot of space.  
+For example, the space required to store a lattice precision for a 
+single $n \times n$ matrix with entries of size $O(p^N)$ is $O(Nn^4 
+\cdot \log p)$.  Conversely, the space needed to record that every entry 
+has precision $O(p^N)$ is just $O(\log N)$.
 
@@
 We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 
-$$B_F(1) \subset f'(0)(B_E(C)).$$
+$B_F(1) \subset f'(0)(B_E(C))$.
 Let $\varepsilon>0$ be such that 
@@
 We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
-that $f(H) \subset f'(0) (H)$. Suppose $x \in H$; we first show that $f(x) \in f'(0)(H)$.  By differentiability at 
+that $f(H) \subset f'(0) (H)$. Suppose $x \in H$. By differentiability at 
 $0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
@@
 \subset f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
-\in \cdot B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
+\in B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
 $\varepsilon C < \rho$, we get $x' \in B^-_E(\rho r) \subset H$ and 
@@
 sub-vector-space around each point where the differential of $f$ is not 
-surjective, which is certainly not true! Nevertheless, one can use lemma \ref{lem:main}
+surjective, which is certainly not true! Nevertheless, one can use Lemma \ref{lem:main}
 to prove a weaker result in the context that $f'(v_0)$ is not surjective.
@@
 defined in Schneider's book \cite {schneider:11a}, \S 6. His definition 
-extends readily to $K$-Banach spaces as follows:
+extends to $K$-Banach spaces as follows:
 
@@
 reverses the order (\emph{i.e.} if $\varphi \leq \psi$, then
-$\varphi^\star \geq \psi^\star$). 
+$\varphi^\star \geq \psi^\star$). We refer to \cite{Rockafellar:97}
+for a complete exposition on Legendre transforms.
 %Furthermore, if $\varphi$ is
@@
 that $\sum_{n \geq 0} f_n$ does not converge on $B_E^-(e^v)$ and 
-therefore that $\Lambda(f)(v) = +\infty$ as well.
+therefore that $\Lambda(f)(v) = +\infty$.
 
@@
 that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
-$\lim \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
+$\lim_{i \to \infty} \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
 furthermore $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} 
@@
 $\varphi_{\geq v} \leq \varphi$ and the function $x \mapsto 
-\varphi_{\geq v}(x) - v x$ is not decreasing. Concretely, we have:
+\varphi_{\geq v}(x) - v x$ is nondecreasing. Concretely, we have:
 $$\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y 
@@
 \end{proof}
+
+We are now ready to provide two sufficient conditions to recognize 
+effectively first order lattices.
 
@@
 
+\begin{rem}
+In the statement of Proposition \ref{prop:locanalytic}, one can of 
+course replace the function $\Lambda(f)$ by any convex function 
+$\varphi$ with $\varphi \geq \Lambda(f)$. If $f$ is given by some
+formula or some algorithm, such a function $\varphi$ can be obtained
+using Remark \ref{rem:opLambda}.
+\end{rem}
+
 \begin{proof}
@@
 for all $x \in B_E(\delta)$.
-This follows directly from Proposition \ref{prop:lambdaNP} and Proposition
-\ref{prop:trunc}.
+This follows from Proposition \ref{prop:lambdaNP} and Proposition
+\ref{prop:trunc} applied to the function $x \mapsto \frac {\Lambda_{\geq 
+2}(x)} x$.
 \end{proof}
@@
 definition, we have $f(x+H) = f(x) + f'(x)(H)$ and thus {\tt f}
-must output $f'(x)(H)$ if it preserves precision. In this section we
+must output $O(f'(x)(H))$ if it preserves precision. In this section we
 explain how, under the above hypothesis, one can implement the function 
-{\tt f} so that it always outputs the optimal precision $O(H')$ with $H' 
-= f'(x)(H)$.
+{\tt f} so that it always outputs the optimal precision.
 
@@
 ``nice'' (in particular locally analytic) function. For all 
-$i$, let $U_i$ denote the codomain of $f_i$. Concretely, $U_i$ is the 
-set of all possible values of all variables which are defined in the 
-program after the execution of $i$-th step. Mathematically, it is an 
+$i$, let $U_i$ denote the codomain of $f_i$. Of course $U_i$ must
+contains all possible values of all variables which are defined in the 
+program after the execution of $i$-th step. Mathematically, we assume
+that it is an 
 open subset in some $K$-Banach space $E_i$. We have $U_n = V$ and the 
@@
 a new intermediate result that we denote by $x_i + O(H_i)$. A way to 
-guarantee that no artificial loss of precision are generated is then to 
+guarantee that precision is preserved is then to 
 ensure $H_i = f'_i(x)(H_{i-1}) = g_i'(x)(H)$ at each step. This can be 
@@
 have $f'(x) = h_i'(x_i) \circ g'_i(x)$. Since $f'(x)$ is surjective (by 
-assumption), we deduce that all $h'_i(x_i)$ are surjective. Let $H'_i$ 
+assumption), we deduce that $h'_i(x_i)$ is surjective for all $i$. Let $H'_i$ 
 be a lattice in $E_i$ such that:
@@
 
-It is nevertheless not obvious how to implement these ideas in practice
+It is nevertheless not obvious to implement these ideas in practice
 because when we enter in the $i$-th step of the execution of {\tt f},
@@
 disappeared. Of course the image of a general $f$ is rarely a sub vector 
-space of $F$ but it is often a sub-$K$-manifold of $F$. We can then use 
+space of $F$ but it is often a sub-$K$-manifold of $F$ (see Appendix
+\ref{sec:manifold}). We can then use 
 the results of Appendix \ref{sec:manifold} to study $f$ viewed as a 
@@
 
-We illustrate the theory developed in this paper by giving a simple 
+We illustrate the theory developed above by giving a simple 
 toy application. Other applications will be discussed in subsequent 
@@
 + u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
-Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
-u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
+Laurent phenomenon: it means that, if the four initial terms $u_0, u_1, 
+u_2, u_3$ are some indeterminates, then each $u_n$ is a Laurent 
 polynomial with coefficients in $\Z$ in these indeterminates
@@
 From now, we will always consider SOMOS sequences with values in $\Q_p$ 
-(for some prime number $p$) and assume that $u_0, u_1, u_2, u_3$ are 
-chosen such that all terms $u_n$'s are determined (\emph{i.e.} there is 
-no division by zero). We assume also for simplicity that $u_0, u_1, u_2, 
+(for some prime number $p$). We assume for simplicity that $u_0, u_1, u_2, 
 u_3$ are all units in $\Z_p$. By the Laurent phenomenon, this implies 
@@
 We introduce the function $f : \Qp^\times \times \Q_p^3 \to \Q_p^4$ defined by 
-$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
+$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$. For all $i$, we have 
 $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
@@
 
-Let us conclude this subsection by remarking that Algorithm 
+We conclude this section by remarking that Algorithm 
 \ref{algo:SOMOS} performs computations at precision at most $O(p^{N+v})$ 
@@
 consisting of lifting the initial terms $u_0, u_1, u_2, u_3$ to enough 
-precision and then do the computation using a naive step-by-step 
+precision and then doing the computation using a naive step-by-step 
 tracking of precision. In this setting, the required original precision 
-is $O(p^{N+2v'})$ where $v'$ is the sum of the valuation of the $u_i$'s 
+is $O(p^{N+v'})$ where $v'$ is the sum of the valuation of the $u_i$'s 
 for $i$ varying between $0$ and $n$. Experiments show that $v'$ is about 
@@
 \end{equation}
-is differentiable at all points in its domain. We recall that
+is differentiable. We recall that
 the mappings $\varphi_i$ above are the so-called \emph{charts}. The
@@
 
-As usual, we can define the notion of strict differentiability (at some 
+As usual, we can define the notion of differentiability (at some 
 point) for a continuous mapping between two $K$-manifolds by viewing it 
@@
 
-\begin{prop}
-\label{prop:independence}
+\begin{lem}
+\label{lem:independence}
 Let $x \in V$ and $H$ be a precision datum at $x$.
@@
 does not depend on the index $i$ such that $x \in V_i$.
-\end{prop}
+\end{lem}
 
@@
 \end{equation}
-With this choice, $x + O(H) = x_0 + O(H_0)$ as soon as $x$ and $x_0$ are 
-close enough in the following sense: the difference $\varphi_i(x) - 
+With this choice, $x + O(H) = x_0 + O(H_0)$ provided that $x$ and $x_0$ 
+are close enough in the following sense: the difference $\varphi_i(x) - 
 \varphi_i(x_0)$ lies in the lattice $\varphi'_i(x_0)(H_0)$. We 
@@
 f_j(H_0)$ and write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal 
-to $f_i(v) + (f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each 
+to $f_i(v) + (f_j - f_i)(v)$ and thus belongs to $f_i(H_0)$ because each 
 summand does. Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse 
-inclusion can be proved in the same way. The fact that $H$ is a 
+inclusion is proved in the same way. The fact that $H$ is a 
 precision datum at $x$ is easy and left to the reader.
@@
 
-With above definitions, Lemma \ref{lem:main} extends readily to 
-manifolds. To state it the context of manifolds, we first need to define 
+With above definitions, Lemma \ref{lem:main} extends to 
+manifolds. To do so, we first need to define 
 a norm on the tangent space $T_x V$ (where $V$ is some $K$-variety and 
@@
 Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
-charts. Proposition \ref{prop:independence} then tells us that a 
+charts. Lemma \ref{lem:independence} then tells us that a 
 precision datum on $x$ determines a precision datum on $y$ and vice 
@@
 \end{equation}
-by neglecting terms of order $2$. Here $P'$ denotes the derivative of 
+and neglecting terms of order $2$. Here $P'$ denotes the derivative of 
 $P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
@@
 
-We now address the case of a multiple roots. Let $P_0$ be a monic 
+We now address the case of a multiple root. Let $P_0$ be a monic 
 polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
@@
 irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
-The algebraic variety $\mathcal V_{m,\alpha_0}$ is a \emph{a fortioti} 
+The algebraic variety $\mathcal V_{m,\alpha_0}$ is \emph{a fortioti} a
 $K$-manifold. Moreover there exists a differentiable function $f$ which 
@@
 $\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
-matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
+matrix and a diagonal one. Since moreover $dR \cdot R^{-1}$ is upper
 triangular with all diagonal entries equal to $0$, we see that 
@@
 \bibliographystyle{plain}
-\bibliography{roebib/Biblio}
+%\bibliography{roebib/Biblio}
+\bibliography{Biblio}
 

### 2014-02-22 padicprec.tex
Proof of Proposition 3.9 improved

@@
 
+\newcommand{\N}{\mathbb N}
 \newcommand{\Z}{\mathbb Z}
@@
 \begin{proof}
-Let $v \in \R$. We assume first that $\NP(f)^\star(v) = +\infty$. Then, 
-using a multilinear version of Banach-Steinhaus Theorem, we can prove 
-that $\sum_{n \geq 0} f_n$ does not converge on $B_E^-(e^v)$ and 
-therefore that $\Lambda(f)(v) = +\infty$.
-
-We now suppose that $\NP(f)^\star(v)$ is finite. To begin with, we 
-assume in addition that $v$ is not a slope of $\NP(f)$. We set $u = 
-\NP(f)^*(v)$. The function $m \mapsto \NP(f)(m) - vm + u$ has the 
-following properties: 
+Let $v \in \R$. We assume first that $\NP(f)^\star(v)$ is finite and 
+that $v$ is not a slope of $\NP(f)$. We set $u = \NP(f)^*(v)$. The 
+function $m \mapsto \NP(f)(m) - vm + u$ has the following properties:
 \begin{enumerate}
@@
 On the other hand, it follows from the definition of $\lVert f_n \rVert$ 
+and the fact that $\vert K^\star \vert$ is dense in $\R$ (which holds
+because $K$ is algebraically closed)
 that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
@@
 \NP(f)^\star(v)$.
+
+\smallskip
+
+We now assume that $\NP(f)^\star(v) = +\infty$. The function $x \mapsto 
+\NP(f)(x) - vx$ is then not bounded from below. Since it is convex, it 
+goes to $-\infty$ when $x$ goes to $+\infty$. Going back to the 
+definition of $\NP(f)$, this implies that $v n + \log \Vert f_n \Vert$
+goes to infinity when $n$ grows. It is then enough to establish the
+following claim:
+\begin{equation}
+\label{eq:claim}
+\forall n \in \N, \quad 
+\Lambda(f)(v) \geq v n + \log \Vert f_n \Vert - \log 2.
+\end{equation}
+Let $n$ be a fixed integer. If $\Vert f_n 
+\Vert = 0$, there is nothing to prove. Otherwise, let us consider an
+element $x_n \in B^-_E(e^v)$ such that $\Vert f_n(x_n) \Vert \geq \frac 
+1 2 \Vert f_n \Vert \cdot e^{vn}$. If the series $\sum_{m \geq 0} 
+f_m(x_n)$ diverges, then $\Lambda(f)(v) = +\infty$ by definition and 
+Eq.~\eqref{eq:claim} holds. On the other hand, if it converges, the
+sequence $\Vert f_m(x_n) \Vert$ goes to $0$ when $m$ goes to infinity.
+Hence it takes its maximum value $R$ a finite number of times; let us 
+denote by $I \subset \N$ the set of the corresponding indices. For any
+$\lambda \in B_K(1)$, the series defining $f(\lambda x_n)$ converges 
+and
+$$f(\lambda x_n) \in B_F(R)
+\quad \text{and} \quad
+f(\lambda x_n) \equiv \sum_{m \in I} \lambda^m f_m(x_n)
+\pmod {B^-_F(R)}.$$
+The quotient $B_F(R) / B^-_F(R)$ is a vector space over the residue
+field of $K$, that is the field $k$ defined by $k = B_K(1)/B^-_K(1)$.
+Hensel's lemma implies that $k$ is algebraically closed (because $K$ 
+is) and then infinite. Therefore, there must exist $\lambda \in 
+B_K(1)$ such that $\sum_{m \in I} \lambda^m f_m(x_n)$ does not vanish
+in $B_F(R) / B^-_F(R)$. For such an element $\lambda$, we have $\Vert 
+f(\lambda x_n) \Vert = R \geq \frac 1 2 \Vert f_n \Vert \cdot e^{vn}$. 
+The claim \eqref{eq:claim} follows.
+
+\smallskip
 

### 2014-02-24 padicprec.tex
Fix bibiography
Xavier's suggested changes

@@
 
-\section{Tracking Precision} \label{sec:prec-proposal}
-
-\subsection{Lattices}
-
-We recall some standard definitions from ultrametric analysis, referring to
-Schneider \cite{schneider:11a} for a complete exposition.
-
-Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
-We assume that the induced metric is an ultrametric -- 
-\[ \lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert) \]
--- and that $K$ is complete with respect to it.
-For example, we may take $K = \Qp$ with the
-$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
-Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
-that $K$ contains a dense subring $R \subset K$
-consisting of elements that can be represented with a finite amount of space.
-For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
-for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.
-
-If $E$ is a $K$-vector space, possibly of infinite dimension, then an
-\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
-\begin{enumerate}[(i)]
-\item $\Vert x\Vert = 0$ if and only if $x = 0$;
-\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
-\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
-\end{enumerate}
-A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
-normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
-A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
-We underline that any lattice $H$ in $E$ is also closed since its 
-complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
-are all open.
-
-For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
-\begin{align*}
-B_E(r) &= \{ x \in E : \Vert x \Vert \le r \}, \\
-B^-_E(r) &= \{ x \in E : \Vert x \Vert < r\}.
-\end{align*}
-Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.
-
-Suppose $E$ is a $K$-Banach space and $I$ a set.
-A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
-every element $x \in E$ can be written
-\[
-x = \sum_{i \in I} \alpha_i x_i
-\]
-for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
-Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
-
-Given a basis $(x_i)_{i \in I}$ and a sequence $(r_i)_{i \in I}$ with $r_i \in \R_{>0}$, the sets
-\begin{align*}
-B_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert \le r_i \Big\}, \\
-B^-_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert < r_i \Big\}
-\end{align*}
-are lattices precisely when the $r_i$ are bounded.  If we have equipped $E$ with a distinguished basis
-then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.
-
-\subsubsection*{Lattices and computers.}
-
-Suppose that $E \simeq K^d$ is finite dimensional.  Then if $H \subset E$
-is a lattice then there exist $a, b \in \Q_{>0}$ with
-\begin{equation}
-\label{eq:incllattice}
-B_K(a)^d \subset H \subset B_K(b)^d.
-\end{equation}
-Set $r = \frac a b$ and $R_r = \OK/B_K(r)$.  
-Then a lattice $H$ satisfying \eqref{eq:incllattice} is uniquely 
-determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
-R_r^d$.  Since $R \cap \OK$ is dense in $\OK$, elements of $R_r$ may be
-represented exactly.  Thus $H$ may be encoded as a 
-$(d \times d)$ matrix with coefficients in $R_r$.  For example, when
-$K = \Qp$ the ring $R_r$ is just $(\Z / p^n\Z)$ for $n = \lfloor - \log_p r \rfloor$.
-
-\subsection{Approximate elements}
-
-Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
-\begin{deftn}
-\begin{itemize}
-\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
-\begin{equation} \label{eq:exact_elt}
-x = \sum_{j \in J} \alpha_j x_j.
-\end{equation}
-
-\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
-and $H$ is a lattice in $E$.
-\end{itemize}
-\end{deftn}
-
-The pair $(x, H)$ represents an undetermined element of the coset
-$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
-the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
-we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
-Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.
-
-This definition encapsulates the main practical suggestion of the paper with regards to representing
-vector spaces, matrices, polynomials and power series over $K$: one should \textbf{separate} the
-approximation from the precision.  In the rest of this section we discuss some of the benefits made
-possible by such a separation.
-
-Note first that using an arbitrary lattice to represent the precision of an approximate element
-can reduce precision loss when compared to storing the precision of each coefficient
-$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
-$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
-$f \circ f(a + O(p^n), b + O(p^m))$.
-
-Write $(e_1, e_2)$ for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
-the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
-$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
-For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
-Storing precisions on each coordinate of a vector amounts to only allowing precisions of the form
-$B_E((r_i))$ with respect to the standard basis.  One would be forced to truncate the intermediate precision
-to $B_E(p^{-\min(m, n)})$, yielding an artificial loss of precision in the final result.
-
-\subsection{Separating precision from approximation}
-\label{ssec:separation}
-
-In addition to allowing for a more flexible representation of the precision of an element,
-the separation of precision from approximation has other benefits as well.  If the precision
-is encoded with the approximation, certain algorithms become unusable because of their
-numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication \cite{karatsuba-ofman:62a}
-can needlessly lose precision when operating on polynomials with inexact coefficients.
-However, it works perfectly well on exact approximations, leaving the question of the precision of
-the product to be solved separately.  By separating the precision, more algorithms become available.
-
-In addition to allowing for more efficient computation of precision, storing the precision separately from the
-approximation makes it easier to determine the precision of the result when step-by-step precision tracking
-does not reflect the actual precision of the result.
+\section{Problems in Precision} \label{sec:prec-proposal}
+
+We begin by offering two examples that illustrate key features of $p$-adic precision.
+Recall that an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if
+it is well defined up to adding an unknown element of $p^N\Zp$.
 
@@
 
-\subsection{Precision Types}
-
-Using an arbitrary lattice to record the precision of an approximate 
-element has the benefit of allowing computations to proceed without 
-unnecessary precision loss using Lemma \ref{lem:main}.  However, while 
-recording a lattice exactly is possible it does require a lot of space.  
-For example, the space required to store a lattice precision for a 
-single $n \times n$ matrix with entries of size $O(p^N)$ is $O(Nn^4 
-\cdot \log p)$.  Conversely, the space needed to record that every entry 
-has precision $O(p^N)$ is just $O(\log N)$.
-
+\subsection{Lattices}
+
+In order to make our proposals for tracking precision clear, we need
+some definitions from ultrametric analysis.  See 
+Schneider \cite{schneider:11a} for a more complete exposition.
+
+Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
+We assume that the induced metric is an ultrametric -- 
+\[ \lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert) \]
+-- and that $K$ is complete with respect to it.
+For example, we may take $K = \Qp$ with the
+$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
+Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
+that $K$ contains a dense subring $R \subset K$
+consisting of elements that can be represented with a finite amount of space.
+For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
+for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.
+
+If $E$ is a $K$-vector space, possibly of infinite dimension, then an
+\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
+\begin{enumerate}[(i)]
+\item $\Vert x\Vert = 0$ if and only if $x = 0$;
+\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
+\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
+\end{enumerate}
+A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
+normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
+A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
+We underline that any lattice $H$ in $E$ is also closed since its 
+complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
+are all open.
+
+For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
+\begin{align*}
+B_E(r) &= \{ x \in E : \Vert x \Vert \le r \}, \\
+B^-_E(r) &= \{ x \in E : \Vert x \Vert < r\}.
+\end{align*}
+Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.
+
+Suppose $E$ is a $K$-Banach space and $I$ a set.
+A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
+every element $x \in E$ can be written
+\[
+x = \sum_{i \in I} \alpha_i x_i
+\]
+for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
+Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
+
+Given a basis $(x_i)_{i \in I}$ and a sequence $(r_i)_{i \in I}$ with $r_i \in \R_{>0}$, the sets
+\begin{align*}
+B_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert \le r_i \Big\}, \\
+B^-_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert < r_i \Big\}
+\end{align*}
+are lattices precisely when the $r_i$ are bounded.  If we have equipped $E$ with a distinguished basis
+then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.
+
+\subsubsection*{Lattices and computers.}
+
+Suppose that $E \simeq K^d$ is finite dimensional.  Then if $H \subset E$
+is a lattice then there exist $a, b \in \Q_{>0}$ with
+\begin{equation}
+\label{eq:incllattice}
+B_K(a)^d \subset H \subset B_K(b)^d.
+\end{equation}
+Set $r = \frac a b$ and $R_r = \OK/B_K(r)$.  
+Then a lattice $H$ satisfying \eqref{eq:incllattice} is uniquely 
+determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
+R_r^d$.  Since $R \cap \OK$ is dense in $\OK$, elements of $R_r$ may be
+represented exactly.  Thus $H$ may be encoded as a 
+$(d \times d)$ matrix with coefficients in $R_r$.  For example, when
+$K = \Qp$ the ring $R_r$ is just $(\Z / p^n\Z)$ for $n = \lfloor - \log_p r \rfloor$.
+
+\subsection{Approximate elements}
+
+Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
 \begin{deftn}
-Suppose that $E$ is a $K$-Banach space, and write $\Lat(E)$ for the set of lattices in $E$.
-A \emph{precision type} for a $K$-Banach space $E$ is a set $\T \subseteq \Lat(E)$ together
-with a function $\round : \Lat(E) \to \T$ such that
 \begin{itemize}
-\item[$(\ast)$] For every lattice $H \in \Lat(E)$, the lattice $\round(H)$ is a least upper bound for $H$ under the inclusion order:
-$H \subseteq \round(H)$ and if $T \in \T$ satisfies $T \subset \round(H)$ then $H \not\subseteq T$.
+\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
+\begin{equation} \label{eq:exact_elt}
+x = \sum_{j \in J} \alpha_j x_j.
+\end{equation}
+
+\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
+and $H$ is a lattice in $E$.
 \end{itemize}
@@
 
-Different precision types are appropriate for different problems.  For example,
-the final step of Kedlaya's algorithm for computing zeta functions of hyperelliptic
-curves involves taking the characteristic polynomial of the matrix of Frobenius acting
-on a $p$-adic cohomology space.  Obtaining extra precision on the entries of the
-matrix requires a long computation, so it is advisable to work with a precision type
-that does not round too much.
-
-The following list gives examples of useful precision types.  A description of the $\round$ function has been omitted for brevity.
-\begin{itemize}
-\item The \emph{lattice} precision type has $\T = \Lat(E)$.
-\item In the \emph{jagged} precision type, $\T$ consists of lattices of the shape $B_E((e_i),(r_i))$ for a fixed Banach basis $(e_i)$ of $E$.
-\item In the \emph{flat} precision type, $\T$ consists of lattices $B_E(r)$.  The flat precision type is useful since it takes so little space to store and it easy to compute with.
-\item If $E = K_{<d}[X]$ is the space of polynomials of degree less than $d$, the \emph{Newton} precision type consists of lattices $B_E((X^i),(r_i))$ where $-\log r_i$ is a convex function of $i$.
-The Newton precision type is sensible if one thinks of polynomials as functions $K \to K$, since extra precision above the Newton polygon never increases the precision of an evaluation.
-\item If $E = M_{m \times n}(K)$, the \emph{column} precision type consists of lattices with identical image under all projections $\pr_i : E \to K^m$ sending a matrix to its $i$th column.
-It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
-\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.
-It is important when working with overconvergent modular symbols since these lattices are stable under certain Hecke operators.
-\end{itemize}
-
-A final benefit of separating precision from approximation is that it makes
-it much easier to implement algorithms capable of processing different precision types.
+The pair $(x, H)$ represents an undetermined element of the coset
+$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
+the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
+we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
+Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.
+
+\subsection{Separating precision from approximation}
+\label{ssec:separation}
+
+The definition of an approximate element encapsulates the two main practical suggestions of this paper
+with regards to representing vector spaces, matrices, polynomials and power series over $K$:
+\begin{enumerate}
+\item one should \textbf{separate} the approximation from the precision,
+\item the appropriate object to represent precision is a \textbf{lattice}.
+\end{enumerate}
+In the rest of this section we discuss some of the benefits made possible these choices.
+
+Note first that using an arbitrary lattice to represent the precision of an approximate element
+can reduce precision loss when compared to storing the precision of each coefficient
+$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
+$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
+$f \circ f(a + O(p^n), b + O(p^m))$.
+
+Write $(e_1, e_2)$ for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
+the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
+$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
+For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
+Storing precisions on each coordinate of a vector amounts to only allowing precisions of the form
+$B_E((r_i))$ with respect to the standard basis.  One would be forced to truncate the intermediate precision
+to $B_E(p^{-\min(m, n)})$, yielding an artificial loss of precision in the final result.
+
+In addition to allowing for a more flexible representation of the precision of an element,
+the separation of precision from approximation has other benefits as well.  If the precision
+is encoded with the approximation, certain algorithms become unusable because of their
+numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication \cite{karatsuba-ofman:62a}
+can needlessly lose precision when operating on polynomials with inexact coefficients.
+However, it works perfectly well on exact approximations, leaving the question of the precision of
+the product to be solved separately.  By separating the precision, more algorithms become available.
+
+In addition to allowing for more efficient computation of precision, storing the precision separately from the
+approximation makes it easier to determine the precision of the result when step-by-step precision tracking
+does not reflect the actual precision of the result.
 
@@
 
+\subsection{Precision Types}
+
+Using an arbitrary lattice to record the precision of an approximate 
+element has the benefit of allowing computations to proceed without 
+unnecessary precision loss using Lemma \ref{lem:main}.  However, while 
+recording a lattice exactly is possible it does require a lot of space.  
+For example, the space required to store a lattice precision for a 
+single $n \times n$ matrix with entries of size $O(p^N)$ is $O(Nn^4 
+\cdot \log p)$.  Conversely, the space needed to record that every entry 
+has precision $O(p^N)$ is just $O(\log N)$.
+
+\begin{deftn}
+Suppose that $E$ is a $K$-Banach space, and write $\Lat(E)$ for the set of lattices in $E$.
+A \emph{precision type} for a $K$-Banach space $E$ is a set $\T \subseteq \Lat(E)$ together
+with a function $\round : \Lat(E) \to \T$ such that
+\begin{itemize}
+\item[$(\ast)$] For every lattice $H \in \Lat(E)$, the lattice $\round(H)$ is a least upper bound for $H$ under the inclusion order:
+$H \subseteq \round(H)$ and if $T \in \T$ satisfies $T \subset \round(H)$ then $H \not\subseteq T$.
+\end{itemize}
+\end{deftn}
+
+Different precision types are appropriate for different problems.  For example,
+the final step of Kedlaya's algorithm for computing zeta functions of hyperelliptic
+curves involves taking the characteristic polynomial of the matrix of Frobenius acting
+on a $p$-adic cohomology space.  Obtaining extra precision on the entries of the
+matrix requires a long computation, so it is advisable to work with a precision type
+that does not round too much.
+
+The following list gives examples of useful precision types.  A description of the $\round$ function has been omitted for brevity.
+\begin{itemize}
+\item The \emph{lattice} precision type has $\T = \Lat(E)$.
+\item In the \emph{jagged} precision type, $\T$ consists of lattices of the shape $B_E((e_i),(r_i))$ for a fixed Banach basis $(e_i)$ of $E$.
+\item In the \emph{flat} precision type, $\T$ consists of lattices $B_E(r)$.  The flat precision type is useful since it takes so little space to store and it easy to compute with.
+\item If $E = K_{<d}[X]$ is the space of polynomials of degree less than $d$, the \emph{Newton} precision type consists of lattices $B_E((X^i),(r_i))$ where $-\log r_i$ is a convex function of $i$.
+The Newton precision type is sensible if one thinks of polynomials as functions $K \to K$, since extra precision above the Newton polygon never increases the precision of an evaluation.
+\item If $E = M_{m \times n}(K)$, the \emph{column} precision type consists of lattices with identical image under all projections $\pr_i : E \to K^m$ sending a matrix to its $i$th column.
+It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
+\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.
+It is important when working with overconvergent modular symbols since these lattices are stable under certain Hecke operators.
+\end{itemize}
+
+A final benefit of separating precision from approximation is that it makes
+it much easier to implement algorithms capable of processing different precision types.
+
 \subsection{Application to SOMOS sequence}
@@
 \bibliographystyle{plain}
-%\bibliography{roebib/Biblio}
-\bibliography{Biblio}
+\bibliography{roebib/Biblio,extras}
 

### 2014-02-24 padicprec.tex
Changes from Monday meeting

@@
 
-\section{Problems in Precision} \label{sec:prec-proposal}
+\section{Precision proposals} \label{sec:prec-proposal}
+
+\subsection{Problems in precision}
 
@@
 it is well defined up to adding an unknown element of $p^N\Zp$.
+The following problems illustrate cases where step-by-step precision tracking
+does not yield optimal results.
+
+\subsubsection*{A linear map.}
+
+Consider the function $f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem
+of computing $f \circ f(a + O(p^n), b + O(p^m))$.  Applying $f$ twice, computing precision
+with each step, yields $\left(2a + O(p^{\min(m,n)}), 2b + O(p^{\min(m, n)})\right)$.
+On the other hand, $f \circ f(x,y) = (2x, 2y)$, so one may compute the result more
+accurately as $(2a + O(p^n), 2b + O(p^m))$, with even more precision when $p=2$.
 
@@
 
+\subsubsection*{Approximate elements.}
+
+Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
+\begin{deftn}
+\begin{itemize}
+\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
+\begin{equation} \label{eq:exact_elt}
+x = \sum_{j \in J} \alpha_j x_j.
+\end{equation}
+
+\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
+and $H$ is a lattice in $E$.
+\end{itemize}
+\end{deftn}
+
+The pair $(x, H)$ represents an undetermined element of the coset
+$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
+the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
+we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
+Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.
+
 \subsubsection*{Lattices and computers.}
@@
 
-\subsection{Approximate elements}
-
-Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
-\begin{deftn}
-\begin{itemize}
-\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
-\begin{equation} \label{eq:exact_elt}
-x = \sum_{j \in J} \alpha_j x_j.
-\end{equation}
-
-\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
-and $H$ is a lattice in $E$.
-\end{itemize}
-\end{deftn}
-
-The pair $(x, H)$ represents an undetermined element of the coset
-$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
-the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
-we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
-Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.
-
 \subsection{Separating precision from approximation}
@@
 can reduce precision loss when compared to storing the precision of each coefficient
-$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
-$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
-$f \circ f(a + O(p^n), b + O(p^m))$.
-
-Write $(e_1, e_2)$ for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
+$\alpha_i$ in \eqref{eq:exact_elt} separately.  Recall the map 
+$f : (x,y) \mapsto (x+y, x-y)$ from the beginning of the section, and write $(e_1, e_2)$
+for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
 the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
@@
 For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
-Storing precisions on each coordinate of a vector amounts to only allowing precisions of the form
-$B_E((r_i))$ with respect to the standard basis.  One would be forced to truncate the intermediate precision
-to $B_E(p^{-\min(m, n)})$, yielding an artificial loss of precision in the final result.
+By using lattices one eliminates the loss of precision seen previously.
 
@@
 
-In addition to allowing for more efficient computation of precision, storing the precision separately from the
-approximation makes it easier to determine the precision of the result when step-by-step precision tracking
-does not reflect the actual precision of the result.
-
-\section{The main Lemma}
+\section{Lattices and differentials}
 \label{sec:mainlemma}
@@
 
-\section{Tracking precision using differentials}
+\section{Precision in practice}
 \label{sec:tracking}
@@
 
-A final benefit of separating precision from approximation is that it makes
-it much easier to implement algorithms capable of processing different precision types.
+Note that sometimes the precision of a final result can be computed
+a priori (using the methods of Section \ref{sec:differentials} for example).
+Taking advantage of such knowledge can minimize artificial precision loss
+even when using rougher precision types such as flat or jagged.
+Separating precision from approximation also makes
+it much easier to implement algorithms capable of processing different precision types,
+since one can implement the arithmetic of the approximation separately from the logic
+handling the precision.
 

### 2014-02-24 padicprec.tex
Small corrections.

@@
 sometimes underestimate the known precision of a result, as we will discuss in Section
-\ref{ssec:separation}.  Moreover, elements of $\Qp$ provide building blocks for generic
+\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
 implementations of polynomials, vector spaces, matrices and power series.  The practice
@@
 \subsection{Problems in precision}
-
-We begin by offering two examples that illustrate key features of $p$-adic precision.
-Recall that an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if
-it is well defined up to adding an unknown element of $p^N\Zp$.
-The following problems illustrate cases where step-by-step precision tracking
-does not yield optimal results.
+\label{ssec:stepbystep}
+
+The usual way to track $p$-adic precision consists in replacing $p$-adic 
+numbers by approximate elements of the form $a + O(p^N)$ and performing 
+all usual arithmetical operations on these approximations.
+%Recall that 
+%an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if it 
+%is well defined up to adding an unknown element of $p^N\Zp$.
+We offer below three examples that illustrate cases where this way to 
+track precision does not yield optimal results.
 
@@
 terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
-
 Let us first examine how the absolute precision of $u_n$ varies with $n$
@@
 Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
-We assume that the induced metric is an ultrametric -- 
-\[ \lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert) \]
--- and that $K$ is complete with respect to it.
+We assume that the induced metric is an ultrametric (\emph{i.e.}
+$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
+and that $K$ is complete with respect to it.
 For example, we may take $K = \Qp$ with the
@@
 are all open.
-
 For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
-\begin{align*}
-B_E(r) &= \{ x \in E : \Vert x \Vert \le r \}, \\
-B^-_E(r) &= \{ x \in E : \Vert x \Vert < r\}.
-\end{align*}
+$$B_E(r) = \{ x \in E : \Vert x \Vert \le r \}, 
+\quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
 Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.
@@
 A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
-every element $x \in E$ can be written
-\[
-x = \sum_{i \in I} \alpha_i x_i
-\]
+every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
 for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
@@
 \begin{deftn}
+\label{def:approximate}
 \begin{itemize}
@@
 
-The definition of an approximate element encapsulates the two main practical suggestions of this paper
+Definition \ref{def:approximate} encapsulates the two main practical suggestions of this paper
 with regards to representing vector spaces, matrices, polynomials and power series over $K$:
@@
 By using lattices one eliminates the loss of precision seen previously.
+We shall see in the next section that a similar phenomenon occurs for non-linear mappings as well.
 
@@
 \begin{deftn} \label{deftn : diff}
-Let $E$ and $F$ be two normed $K$-vector spaces, let $U$ be an open 
+Let $E$ and $F$ be two $K$-Banach spaces, let $U$ be an open 
 subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
@@
 
-One can derive all usual properties of the derivative from this 
-definition: unicity, chain rules, \emph{etc}.
+%One can derive all usual properties of the derivative from this 
+%definition: unicity, chain rules, \emph{etc}.
 
@@
 number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ such that $B^-_E(\rho r) \subset H \subset B_E(r)$ is a first
+$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ is a first
 order lattice for $f$ at $v_0$.
@@
 Let $r \in (0, \delta)$. We suppose that $H$ is a lattice with 
-$B^-_E(\rho r) \subset H \subset B_E(r).$
+$B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r).$
 We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
@@
 
-We shall now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
+We now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
 be such that $y = f'(0) (x_0)$. We inductively define two sequences 
@@
 uniquely determines the $f_n$'s (a consequence of Proposition 
-\ref{prop:lambdaNP}).
+\ref{prop:lambdaNP} below).
 To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
@@
 \begin{proof}
-Let $v \in \R$. We assume first that $\NP(f)^\star(v)$ is finite and 
-that $v$ is not a slope of $\NP(f)$. We set $u = \NP(f)^*(v)$. The 
+We first remark that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
+are both left continuous. It is then enough to prove that they are
+equal on a dense subset of $\R$. We are going to prove that they
+coincide expect maybe on the set consisting of slopes of $\NP(f)$.
+
+\smallskip
+
+Let $v \in \R$ which is not a slope of $\NP(f)$. We assume first that 
+$\NP(f)^\star(v)$ is finite. We set $u = \NP(f)^*(v)$. The 
 function $m \mapsto \NP(f)(m) - vm + u$ has the following properties:
@@
 Let $n$ be a fixed integer. If $\Vert f_n 
-\Vert = 0$, there is nothing to prove. Otherwise, let us consider an
+\Vert = 0$, there is nothing to prove. Otherwise, we consider an
 element $x_n \in B^-_E(e^v)$ such that $\Vert f_n(x_n) \Vert \geq \frac 
@@
 field of $K$, that is the field $k$ defined by $k = B_K(1)/B^-_K(1)$.
-Hensel's lemma implies that $k$ is algebraically closed (because $K$ 
+It is clearly algebraically closed (because $K$ 
 is) and then infinite. Therefore, there must exist $\lambda \in 
@@
 The claim \eqref{eq:claim} follows.
-
-\smallskip
-
-We have then proved that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
-coincide expect maybe on the set consisting of slopes of $\NP(f)$. But 
-we remark that this set is countable and that both functions $\Lambda 
-(f)$ and $\NP(f)^\star$ are left continuous. This is enough to ensure 
-that $\Lambda (f)$ and $\NP(f)^\star$ coincide everywhere.
 \end{proof}
@@
 On a neighborhood of $-\infty$, the function $x \mapsto \Lambda(f)_{\geq 
-2} (x) - x$ is affine of slope $1$. This implies that, for all $\rho \in
-(0,1]$, there exists $\nu$ satisfying \eqref{eq:locanalytic}. Moreover,
+2} (x) - x$ is affine with slope $1$. This implies that, for all $\rho 
+\in (0,1]$, there exists $\nu$ satisfying \eqref{eq:locanalytic}. Moreover,
 if $\rho$ is close enough to $0$, then one can take $\delta = e^\nu$ as a 
@@
 if the above inclusion is an equality; it is 
-often not the case as shown in Section \ref{ssec:separation}.
+often not the case as shown in Section \ref{ssec:stepbystep}.
 
@@
 where the $f_i$'s correspond to each individual basic step (like 
-addition, multiplication or creation of variables); it is then a
-``nice'' (in particular locally analytic) function. For all 
+addition, multiplication or creation of variables); they are then
+``nice'' (in particular locally analytic) functions. For all 
 $i$, let $U_i$ denote the codomain of $f_i$. Of course $U_i$ must
@@
 at small precision (but enough for the second step) together with an
-approximation of the Newton polygon of the higher order derivatives;
+upper bound of the function $\Lambda(h \mapsto f_i(x_{i-1}+h) - 
+f_i(x_{i-1}))$;
 \item \label{item:determineHi}
@@
 
-We conclude this section by briefly remarking that the two strategies described 
-above share some 
-similarities with usual floating point arithmetics over the reals. 
-Indeed, roughly speaking, in each setting, we begin by choosing a large 
-precision, we do all our computations up to this precision understood 
-that when we are not sure about some digit, we choose it ``at random'' 
-or using good heuristics. The main difference is that, in the 
-ultrametric setting, we are able (under some mild hypothesis) to 
-quantify the precision we need at each individual step in order to be 
-sure that the final result is correct up to the required precision.
+\subsubsection*{Quick comparison with floating point arithmetics.}
+
+The two strategies described above share some similarities with usual 
+floating point arithmetics over the reals. Indeed, roughly speaking, in 
+each setting, we begin by choosing a large precision, we do all our 
+computations up to this precision understood that when we are not sure 
+about some digit, we choose it ``at random'' or using good heuristics. 
+The main difference is that, in the ultrametric setting, we are able 
+(under some mild hypothesis) to quantify the precision we need at each 
+individual step in order to be sure that the final result is correct up 
+to the required precision.
 
@@
 articles. More precisely, we study the SOMOS 4 sequence introduced in \S 
-\ref{ssec:separation}. Making a crucial use of Lemma \ref{lem:main} and 
+\ref{ssec:stepbystep}. Making a crucial use of Lemma \ref{lem:main} and 
 Proposition \ref{prop:locanalytic}, we design a \emph{stable} algorithm 
@@
 
-Recall (\emph{cf} \S \ref{ssec:separation}) that a SOMOS 4 sequence is a 
+Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
 four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
@@
 (see \cite{fomin-zelevinsky:02a}).
-
 From now, we will always consider SOMOS sequences with values in $\Q_p$ 
@@
 
-We now prove that Algorithm \ref{algo:SOMOS} produces the correct result.
+We now prove that it is correct.
 We introduce the function $f : \Qp^\times \times \Q_p^3 \to \Q_p^4$ defined by 
@@
 at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
-f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix:
-$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
-D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
+f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix
+$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
+D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3)$.
 Thanks to the Laurent phenomenon, we know that $D_i$ has
@@
 Let us now establish \eqref{eq:congrSOMOS}. To avoid confusion, 
-let us agree to denote $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
+let us agree to denote by $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
 $a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
@@
 \ref{algo:SOMOS} performs computations at precision at most $O(p^{N+v})$ 
-where $v$ is the sum of the valuations of five consequence terms among 
+where $v$ is the maximum of the sum of the valuations of five consecutive terms among 
 the first $n$ terms of the SOMOS sequence we are considering. 
@@
 is clearly equivalent to
-$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$.
+$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$
 and the latter holds because $H_i$ is a first order lattice
-of $\psi_{ij}$ at the point $x_i$ by definition.
+for $\psi_{ij}$ at $x_i$.
 \end{proof}
@@
 precision datum at $x$ is easy and left to the reader.
-Finally, if $x$ is close enough to $x_0$, we can check (ii) in the 
-charts but this was already done.
+Finally, if $x$ is close enough to $x_0$, it is enough to check (ii) in 
+the charts but this was already done.
 \end{proof}
@@
 number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ in $T_x V$ such that $B^-_{T_x V}(\rho r) \subset H \subset B_{T_x V}(r)$ is a first
+$H$ in $T_x V$ such that $B^-_{T_x V}(\rho r) \subset H \subset 
+B^{\phantom -}_{T_x V}(r)$ is a first
 order lattice for $f$ at $x$.
@@
 \label{eq:diffeval}
-(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x)
-\end{equation}
-and neglecting terms of order $2$. Here $P'$ denotes the derivative of 
+(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x).
+\end{equation}
+Here $P'$ denotes the derivative of 
 $P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
@@
 of positive degree.  Since $\mathcal V_0$ is dense, $f$ is not continuous outside 
-$\mathcal V_0$. Conversely, the function $f$ is differentiable, and even locally analytic,
+$\mathcal V_0$. On the contrary, the function $f$ is differentiable, and even locally analytic,
 on $\mathcal V_0$.
@@
 with no precision, since $A$ and $B$ appear to lie in $\mathcal V_i$ at the known precision.
-
 Once this assumption is made, the computation is possible and one can apply Lemma 
@@
 dP)(\alpha_0 + d\alpha) = 0$, obtaining
-\[
-P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0.
-\]
+$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$.
 Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
@@
 mainly because we are using the article \cite{vaccon:14a}, which deals only 
-with homogeneous polynomials. It is probably not essential.} 
+with homogeneous polynomials. It is however probably not essential.} 
 polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 

### 2014-02-26 padicprec.tex
Responses to Xavier's comments, updates on Section 3

@@
 \usepackage{amsrefs}
-\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
 \usepackage{color}
@@
 \usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
+\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
+\renewcommand*{\theHsection}{\thesection}
 
@@
 
-Neither of these models allows a user to track the precision of an individual element.
-Over the reals, interval arithmetic is unwieldy, since arithmetic operations always increase the lengths of the inputs.
-As a consequence, most computations in the real numbers rely on
+In both of these models, precision (absolute or relative) is constant across all elements.
+Since some operations lose precision, it can be useful to attach a precision to each element.
+Over the reals, such interval arithmetic is unwieldy, since arithmetic operations always increase
+the lengths of the inputs.  As a consequence, most computations in the real numbers rely on
 statistical cancelation and external estimates of precision loss, rather than
@@
 and two-pass tracking, where an initial pass computing rough approximations is used
-in computing the precision lattices.  In Section \ref{ssec:SOMOS-solution}, we give
+in computing the precision lattices.  We introduce precision types in Section \ref{ssec:types},
+which allow a tradeoff between flexibility, space and time in computing with precision.
+In Section \ref{ssec:SOMOS-solution}, we give
 an application of these ideas to an algorithm for computing terms of the SOMOS sequence.
@@
 Our theory of $p$-adic precision rests upon a lemma in $p$-adic 
-analysis: Lemma \ref{lem:main}.  In this section we develop the theory 
-surrounding this result, and proceed to practical consequences in 
+analysis: Lemma \ref{lem:main}.  This section develops the theory 
+surrounding this result; we proceed to practical consequences in 
 Section \ref{sec:tracking}.
@@
 \label{rem:differentiability}
-This notion of differentiability is stronger than the standard definition.
-It implies for instance the following: if $f$ is 
-differentiable on $U$, then the function $x \mapsto f'(x)$ defined on 
-$U$ is continuous. 
-It is sometimes called \emph{strict differentiability} but, for this
-article, we prefer a simpler terminology since we shall only use this
-notion of differentiability.
+This notion of differentiability is sometimes called \emph{strict differentiability};
+it implies that the function $x \mapsto f'(x)$ is continuous on $U$. 
 \end{rem}
@@
 A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
-$v_0$ if the equality
+$v_0$ if the following equality holds:
 \begin{equation}
 \label{eq:firstorder}
-f(v_0 + H) = f(v_0) + f'(v_0) (H)
-\end{equation}
-holds.
+f(v_0 + H) = f(v_0) + f'(v_0) (H).
+\end{equation}
 \end{deftn}
@@
 \begin{proof}
-We may assume that $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 
+Without loss of generality, $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 
 $B_F(1) \subset f'(0)(B_E(C))$.
@@
 we get
-$$\Vert y  - f(x_n) \Vert 
+\[
+\Vert y  - f(x_n) \Vert 
 \leq \varepsilon \cdot \Vert z_{n-1} \Vert \leq \varepsilon C \cdot \Vert 
-y - f(x_{n-1}) \Vert,$$
-for all positive integer $n$.  We therefore deduce that $\Vert y - f(x_n) \Vert = O(a^n)$ and 
-$\Vert z_n \Vert = O(a^n)$ with $a = \varepsilon C < \rho \leq 1$. 
+y - f(x_{n-1}) \Vert,
+\]
+for all $n > 0$.  Therefore, $\Vert y - f(x_n) \Vert = O(a^n)$ and 
+$\Vert z_n \Vert = O(a^n)$ for $a = \varepsilon C < \rho \leq 1$. 
 These conditions show that $(x_n)$ is a Cauchy sequence, which converges since $E$ is complete.
 Write $x$ for the limit of the $x_n$; we have $x \in H$ because $H$ is closed.
-Moreover, $f$ is continuous on $H \subseteq U_\varepsilon$ since it is differentiable and thus $y=f(x)$.
+Moreover, $f$ is continuous on $H \subseteq U_\varepsilon$ since it is differentiable, and thus $y=f(x)$.
 \end{proof}
@@
 taking the intersection of the right hand side
-over all possible $W$ yields an upper bound on $f(v_0 + H)$.
+over many $W$ yields an upper bound on $f(v_0 + H)$.
 
@@
 
-We consider in this subsection the case where $f$ is locally analytic.
-Under this extra assumption, we make explicit the constant $\delta$ 
-which appears in Lemma \ref{lem:main}.
-Locally analytic functions over finite dimensional $K$-vector spaces are 
-defined in Schneider's book \cite {schneider:11a}, \S 6. His definition 
-extends to $K$-Banach spaces as follows:
+In this section we make the constant $\delta$ in Lemma \ref{lem:main} explicit,
+under the additional assumption that $f$ is locally analytic.
+We extend the definition of such functions from finite-dimensional
+$K$-vector spaces \cite {schneider:11a}*{\S 6} to $K$-Banach spaces.
 
@@
 A function $f \col U \rightarrow F$ is said \emph{locally analytic} at 
-$x$ if there exist an open subset $U_x \subset E$ and continuous $n$-linear
+$x$ if there exists an open subset $U_x \subset E$ and continuous $n$-linear
 maps $L_n \col E^n \rightarrow F$ for $n \ge 1$ such that
@@
 A function $f$ which is locally analytic at $x$ is \emph{a fortiori} 
-differentiable at $x$. Furthermore, in that case, the linear map $L_1$ 
-appearing in the above definition is nothing but the derivative $f'(x)$.
+differentiable at $x$, with derivative given by $L_1$.
 \end{rmk}
@@
 
-The function $\Lambda(f)$ we have just defined is actually closely 
+The function $\Lambda(f)$ we have just defined is closely 
 related to the Newton polygon of $f$. Recall that the 
-Newton polygon of $f$ is the convex hull in $\R^2$ of the points of 
-coordinates $(n, - \log \lVert f_n \rVert )$ for $n \geq 0$ together 
+Newton polygon of $f$ is the convex hull in $\R^2$ of the points
+$(n, - \log \lVert f_n \rVert )$ for $n \geq 0$, together 
 with the extra point $(0, +\infty)$. We denote by $\NP(f) : \R \to \R 
@@
 One can check that the map $\varphi \mapsto \varphi^\star$ is an
-involution (\emph{i.e.} $(\varphi^\star)^\star = \varphi$) and
-reverses the order (\emph{i.e.} if $\varphi \leq \psi$, then
-$\varphi^\star \geq \psi^\star$). We refer to \cite{Rockafellar:97}
-for a complete exposition on Legendre transforms.
-%Furthermore, if $\varphi$ is
-%piecewise affine (and convex), the slopes of $\varphi^\star$ are
-%exactly the points where $\varphi$ is not differentiable and \emph{vice
-%versa}.
-
-%\begin{prop} \label{prop : conjugate Newton}
-%Considered as a function on $\R$, $\NP(f)^*$ is defined over $\left] -\infty,v_{\infty} \right[$, though it may be defined for larger values.
-%\end{prop}
-%
-%We can interpret the function $\Lambda(f)$ in terms of the conjugate Newton polygon:
+order-reversing involution: $(\varphi^\star)^\star = \varphi$ and
+$\varphi^\star \geq \psi^\star$ whenever $\varphi \leq \psi$.
+We refer to \cite{Rockafellar:97} for a complete exposition on Legendre transforms.
 
@@
 \begin{proof}
-We first remark that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
-are both left continuous. It is then enough to prove that they are
-equal on a dense subset of $\R$. We are going to prove that they
-coincide expect maybe on the set consisting of slopes of $\NP(f)$.
+Note that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
+are both left continuous. It is then enough to prove that they
+coincide expect possibly on the set of slopes of $\NP(f)$,
+a dense subset of $\R$.
 
@@
 
-Let $v \in \R$ which is not a slope of $\NP(f)$. We assume first that 
+Let $v \in \R$, not a slope of $\NP(f)$. We assume first that 
 $\NP(f)^\star(v)$ is finite. We set $u = \NP(f)^*(v)$. The 
@@
 \Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
-\lVert f_n \rVert \cdot e^{vn}< \lVert f_n \rVert \cdot e^{vn}.
+\lVert f_n \rVert \cdot e^{vn} \le \lVert f_n \rVert \cdot e^{vn}.
 \]
 Thus, the series $\sum_{m \geq 0} f_m(x)$ converges and $\Vert f(x) 
-\Vert \leq \Vert f_n \Vert \cdot e^{vn}$. We then get:
+\Vert \leq \Vert f_n \Vert \cdot e^{vn}$. We then get
 \begin{equation}
@@
 On the other hand, it follows from the definition of $\lVert f_n \rVert$ 
-and the fact that $\vert K^\star \vert$ is dense in $\R$ (which holds
-because $K$ is algebraically closed)
+and the fact that $\vert K^\times \vert$ is dense in $\R$ ($K$ is algebraically closed)
 that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
 $\lim_{i \to \infty} \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
-furthermore $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} 
-\Vert f_n \Vert \cdot e^{vn}$ for all $i$, we get $\Vert f_m (x_i) \Vert 
+ $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} \cdot
+\Vert f_n \Vert \cdot e^{vn}$ for all $m$ and $i$, we get $\Vert f_m (x_i) \Vert 
 < \Vert f_n (x_i) \Vert$ for $i$ large enough. For these $i$, we then 
@@
 \NP(f)(x) - vx$ is then not bounded from below. Since it is convex, it 
-goes to $-\infty$ when $x$ goes to $+\infty$. Going back to the 
-definition of $\NP(f)$, this implies that $v n + \log \Vert f_n \Vert$
-goes to infinity when $n$ grows. It is then enough to establish the
+goes to $-\infty$ when $x$ goes to $+\infty$. By the
+definition of $\NP(f)$, the expression $v n + \log \Vert f_n \Vert$
+goes to infinity as $n$ grows. It is then enough to establish the
 following claim:
@@
 Eq.~\eqref{eq:claim} holds. On the other hand, if it converges, the
-sequence $\Vert f_m(x_n) \Vert$ goes to $0$ when $m$ goes to infinity.
+sequence $\Vert f_m(x_n) \Vert$ goes to $0$ as $m$ goes to infinity.
 Hence it takes its maximum value $R$ a finite number of times; let us 
 denote by $I \subset \N$ the set of the corresponding indices. For any
-$\lambda \in B_K(1)$, the series defining $f(\lambda x_n)$ converges 
+$\lambda \in \OK$, the series defining $f(\lambda x_n)$ converges 
 and
@@
 The quotient $B_F(R) / B^-_F(R)$ is a vector space over the residue
-field of $K$, that is the field $k$ defined by $k = B_K(1)/B^-_K(1)$.
-It is clearly algebraically closed (because $K$ 
-is) and then infinite. Therefore, there must exist $\lambda \in 
-B_K(1)$ such that $\sum_{m \in I} \lambda^m f_m(x_n)$ does not vanish
+field $k$ of $K$.  Since $k$ is infinite, there must exist $\lambda \in 
+\OK$ such that $\sum_{m \in I} \lambda^m f_m(x_n)$ does not vanish
 in $B_F(R) / B^-_F(R)$. For such an element $\lambda$, we have $\Vert 
@@
 
-We are now ready to provide two sufficient conditions to recognize 
-effectively first order lattices.
+We may now provide two sufficient conditions to effectively recognize 
+first order lattices.
 
@@
 for all $x \in B_E(\delta)$.
-This follows from Proposition \ref{prop:lambdaNP} and Proposition
+This inequality follows from Propositions \ref{prop:lambdaNP} and
 \ref{prop:trunc} applied to the function $x \mapsto \frac {\Lambda_{\geq 
@@
 $h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H)$.
-Therefore, if after the $i$-th step of the execution of the function
-{\tt f}, we modify the intermediate value $x_i$ by an element of
-$H'_i$, the final result remains unchanged. In other words, it is
+Therefore, modifying the intermediate value $x_i$ by an element
+of $H'_i$ after the $i$-th step of the execution of {\tt f} leaves
+the final result remains unchanged. In other words, it is
 enough to compute $x_i$ modulo $H'_i$.
@@
 \subsection{Precision Types}
+\label{ssec:types}
 
@@
 the final step of Kedlaya's algorithm for computing zeta functions of hyperelliptic
-curves involves taking the characteristic polynomial of the matrix of Frobenius acting
+curves \cite{kedlaya:01a}*{\S 4: Step 3} involves taking the characteristic polynomial of the matrix of Frobenius acting
 on a $p$-adic cohomology space.  Obtaining extra precision on the entries of the

### 2014-02-26 padicprec.tex
Typos

--- 2014-02-24 padicprec.tex
@@
 equal on a dense subset of $\R$. We are going to prove that they
-coincide expect maybe on the set consisting of slopes of $\NP(f)$.
+coincide except maybe on the set consisting of slopes of $\NP(f)$.
 
@@
 the slopes of the Legendre transform of a convex piecewise affine 
-function $f$ are exactly the points where $f$ is not differentiable.
+function $f$ are exactly the abscissae of the points where $f$ is not 
+differentiable.
 \end{proof}
@@
 Note that sometimes the precision of a final result can be computed
-a priori (using the methods of Section \ref{sec:differentials} for example).
+a priori (using the methods of Appendix \ref{sec:differentials} for example).
 Taking advantage of such knowledge can minimize artificial precision loss

### 2014-02-26 padicprec.tex

--- 2014-02-26 padicprec.tex
@@
 \item Gaudry et al. \cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
-\item Kedlaya \cite{kedlaya:01a} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
+\item Kedlaya \cite{kedlaya:01a}, Lauder \cite{lauder:04a} and many followers used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
 \item Lercier and Sirvent \cite{lercier-sirvent:08a} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
@@
 \emph{homogeneous}\footnote{This restriction will be convenient for us 
-mainly because we are using the article \cite{vaccon:14a}, which deals only 
+mainly because we are using the article \cite{vaccon:14b}, which deals only 
 with homogeneous polynomials. It is however probably not essential.} 
 polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
+\ldots, g_t)$. It follows from \cite{vaccon:14b}*{Thm. 1.1} that this 
 map is continuous at $(f_1, \ldots, f_s)$ provided that:
@@
 $f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{vaccon:14a} for more detail).
+order (\emph{cf} \cite{vaccon:14b} for more detail).
 \end{itemize}

### 2014-05-14 padicprec.tex
First pass fixing typos pointed out by referee

@@
 model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
-all $p$-adic elements in have constant \emph{absolute precision} $n$.
+all $p$-adic elements have constant \emph{absolute precision} $n$.
 
@@
 \[
-u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}.
+u_n = \frac{u_{n-3} u_{n-1} + u_{n-2}^2}{u_{n-4}}.
 \]
@@
 if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
-The computation of $u_{n+4}$ involves a division by $u_n$ and hence, roughly speaking,
-decreases the precision by a factor $p^{\val(u_n)}$.  Hence the
+The computation of $u_n$ involves a division by $u_{n-4}$ and hence, roughly speaking,
+decreases the precision by a factor $p^{\val(u_{n-4})}$.  Hence the
 step-by-step computation returns the value of $u_n$ with precision
@@
 
-Let us first recall that a square matrix $M$ with coefficients in $K$ 
+Let us first recall that a square matrix $M$
 admits a LU factorization if it can be written as a product $LU$ where 
-$L$ and $U$ is lower triangular and upper triangular respectively. 
+$L$ and $U$ are lower triangular and upper triangular respectively. 
 The computation of a LU factorization appears as an important tool to 
@@
 For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
-for $K = \Fp\lb t \rb$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.
+for $K = \Fp\lp t \rp$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.
 
@@
 \[ 
-\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
+\Vert f(v)-f(w)-f'(v_0) \left( v-w \right) \Vert 
 \leq \varepsilon \Vert v-w \Vert. 
@@
 and let $x \in U$.
-A function $f \col U \rightarrow F$ is said \emph{locally analytic} at 
+A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} at 
 $x$ if there exists an open subset $U_x \subset E$ and continuous $n$-linear
@@
 \{+\infty\}$ and a real number $v$, we define $\varphi_{\geq v} : \R \to 
-\R \cup \{\pm \infty\}$ as the highest convex function such that 
-$\varphi_{\geq v} \leq \varphi$ and the function $x \mapsto 
-\varphi_{\geq v}(x) - v x$ is nondecreasing. Concretely, we have:
-$$\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y 
-\big).$$
+\R \cup \{\pm \infty\}$ by
+\[
+\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y \big).
+\]
+The function $\varphi_{\geq v}$ is the maximum among all functions $\varphi'$
+with $\varphi' \leq \varphi$ and $x \mapsto \varphi'(x) - v x$ nondecreasing.
 When $v$ is fixed, the construction $\varphi \mapsto \varphi_{\geq v}$
 is nondecreasing: if $\varphi$ and $\psi$ are two convex functions such
-that $\varphi \leq \psi$, we deduce $\varphi_{\geq v} \leq \psi_{\geq v}$.
+that $\varphi \leq \psi$ then $\varphi_{\geq v} \leq \psi_{\geq v}$.
 
@@
 contains all possible values of all variables which are defined in the 
-program after the execution of $i$-th step. Mathematically, we assume
+program after the execution of the $i$-th step. Mathematically, we assume
 that it is an 
@@
 remember that the dimension of $E_i$ is roughly the number of used 
-variables after the step $i$. It all $g_i'(x)$ were surjective, this 
+variables after the step $i$. If all $g_i'(x)$ were surjective, this 
 would mean that the function {\tt f} never initializes a new variable!
@@
 of $H'_i$ after the $i$-th step of the execution of {\tt f} leaves
-the final result remains unchanged. In other words, it is
+the final result unchanged. In other words, it is
 enough to compute $x_i$ modulo $H'_i$.
@@
 Step~\eqref{item:smallprec}. The two-pass method we have just 
-presented is then probably not much more expansive than the one-pass 
+presented is then probably not much more expensive than the one-pass 
 method, although it is more difficult to implement.
@@
 
-The study of the this example actually suggests another solution to 
-tackle the issue of non-surjectivity. Indeed, remark that our $f$ above 
-had not a surjective differential simply because its codomain was too 
-large: if we had replaced $f : (K^n)^n \to M_n(K)$ by $g : (K^n)^n \to 
-S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
-over $K$ of size $n$) defined in the same way, our problem would have 
-disappeared. Of course the image of a general $f$ is rarely a sub vector 
-space of $F$ but it is often a sub-$K$-manifold of $F$ (see Appendix
-\ref{sec:manifold}). We can then use 
-the results of Appendix \ref{sec:manifold} to study $f$ viewed as a 
-function whose codomain is $f(U)$, understood that the differential of 
-it has now good probability to be surjective.
+The study of this example actually suggests another solution to 
+tackle the issue of non-surjectivity. Indeed, we remark that our $f$ above 
+did not have a surjective differential simply because its codomain was too 
+large: replacing the codomain with the $K$-vector space $S_n(K)$
+of $n \times n$ symmetric matrices over $K$ makes the differential of $f$ surjective.
+While the image of a general $f$ is rarely a sub vector 
+space of $F$, it is often a sub-$K$-manifold of $F$. We can then use 
+the results of Appendix \ref{sec:manifold} to study $f : U \to f(U)$, which
+likely has surjective differential.
 
@@
 
-\subsubsection*{Quick comparison with floating point arithmetics.}
-
-The two strategies described above share some similarities with usual 
-floating point arithmetics over the reals. Indeed, roughly speaking, in 
-each setting, we begin by choosing a large precision, we do all our 
-computations up to this precision understood that when we are not sure 
-about some digit, we choose it ``at random'' or using good heuristics. 
-The main difference is that, in the ultrametric setting, we are able 
-(under some mild hypothesis) to quantify the precision we need at each 
-individual step in order to be sure that the final result is correct up 
-to the required precision.
+\subsubsection*{Quick comparison with floating point arithmetic.}
+
+The two strategies described above share some similarities with standard 
+floating point arithmetic over the reals. In each setting, we begin
+by choosing a large precision for all computations, and when we
+encounter an unconstrained digit we choose it ``at random'' or using good heuristics. 
+However, in the ultrametric setting, mild hypotheses allow us
+to quantify the precision needed at each 
+individual step in order to ensure a specified final precision.
 
@@
 Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
-four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
-+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
-Laurent phenomenon: it means that, if the four initial terms $u_0, u_1, 
-u_2, u_3$ are some indeterminates, then each $u_n$ is a Laurent 
-polynomial with coefficients in $\Z$ in these indeterminates
-(see \cite{fomin-zelevinsky:02a}).
-From now, we will always consider SOMOS sequences with values in $\Q_p$ 
-(for some prime number $p$). We assume for simplicity that $u_0, u_1, u_2, 
-u_3$ are all units in $\Z_p$. By the Laurent phenomenon, this implies 
-that all $u_n$'s lie in $\Z_p$, and that if $u_0, u_1, u_2, u_3$ 
+four-term inductive sequence defined by $u_n = \frac{u_{n-3} u_{n-1} 
++ u_{n-2}^2}{u_n}$ exhibiting the Laurent phenomenon.
+We will focus on SOMOS sequences with values in $\Q_p$, and 
+assume for simplicity that $u_0, u_1, u_2, u_3 \in \Z_p^\times$.
+By the Laurent phenomenon, $u_n \in \Z_p$ for all $n$, and if $u_0, u_1, u_2, u_3$ 
 are known with finite precision $O(p^N)$ then all $u_n$ are 

### 2014-05-16 padicprec.tex
More changes suggested by referee

@@
 1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
-From the latter formula, it follows directly that if $u_0$, $u_1$,
-$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
-are also known with the same precision. Thus, the term $v_n$ that 
+From the latter formula and our assumption that $u_0$, $u_1$, 
+$u_2$ and $u_3$ are units known up to precision $O(p^N)$, it follows that all $u_n$'s
+are known with the same precision. Thus, the term $v_n$ that 
 appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
@@
 If $d$ is large compared to $p$, the second precision is much more
-accurate than the first one. On the other hand, the second algorithm
-is less efficient than the first one because evaluating Cramer-type
-formulae requires many computations. 
+accurate than the first one. On the other hand, the Cramer-type formulae
+in the second algorithm yield a substantially longer running time than the first.
 %In \cite{caruso:12a}, the author designs a new algorithm which combines 
@@
 Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
-that $K$ contains a dense subring $R \subset K$
-consisting of elements that can be represented with a finite amount of space.
-For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
-for $K = \Fp\lp t \rp$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.
+that $K$ contains a computable dense subring $R \subset K$ \cite{rabin:60a}.
+This assumption holds for $K = \Qp$ and $K=\Fp\lp t \rp$ by setting $R = \Z[\frac1p]$ or $R = \Q$
+in the case of $\Qp$ and $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$ in the case of $\Fp\lp t \rp$.
 
@@
 \quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
-Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.
+Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.  We will also set $B_E(\infty) = B^-_E(\infty) = E$.
 
@@
 every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
-for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
+for scalars $\alpha_i \in K$ with $\alpha_i \to 0$ (according to the filter of cofinite subsets), and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
 Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
@@
 \ref{prop:lambdaNP} below).
-To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
+To such a series $f$, we attach the function $\Lambda(f) : \R \cup \{+\infty\} \to \R 
 \cup \{+\infty\}$ defined by:
-$$\begin{array}{rcll}
-\Lambda(f)(v) & = & 
-\log \big( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \big)
-& \text{if } f \text{ is defined on } B^-_E(e^v) \\
-& = & +\infty & \text{otherwise}
-\end{array}$$
+\[
+\Lambda(f)(v) = \begin{cases} \log \left( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \right) & \mbox{if $f$ is defined on} B^-_E(e^v), \\
++\infty & \mbox{otherwise.} \end{cases}
+\]
 The following lemma is easy and left to the reader.
@@
 Let $f = \sum_{n \geq 0} f_n$ and $g = \sum_{n \geq 0} g_n$ be two 
-series as above. Then:
+series as above. Then
 \begin{align*}
-\Lambda(f+g) &\leq \max (\Lambda(f),\Lambda(g)) \\
-\Lambda(f \times g) &\leq \Lambda(f)+\Lambda(g) \\
-\Lambda(f \circ g) &\leq \Lambda(f) \circ \Lambda(g)
+\Lambda(f+g) &\leq \max (\Lambda(f),\Lambda(g)), \\
+\Lambda(f \times g) &\leq \Lambda(f)+\Lambda(g), \\
+\Lambda(f \circ g) &\leq \Lambda(f) \circ \Lambda(g).
 \end{align*}
@@
 four-term inductive sequence defined by $u_n = \frac{u_{n-3} u_{n-1} 
-+ u_{n-2}^2}{u_n}$ exhibiting the Laurent phenomenon.
++ u_{n-2}^2}{u_{n-4}}$ exhibiting the Laurent phenomenon.
 We will focus on SOMOS sequences with values in $\Q_p$, and 
@@
 \emph{homogeneous}\footnote{This restriction will be convenient for us 
-mainly because we are using the article \cite{vaccon:14b}, which deals only 
+mainly because we are using the article \cite{vaccon:14a}, which deals only 
 with homogeneous polynomials. It is however probably not essential.} 
 polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from \cite{vaccon:14b}*{Thm. 1.1} that this 
+\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
 map is continuous at $(f_1, \ldots, f_s)$ provided that:
@@
 $f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{vaccon:14b} for more detail).
+order (\emph{cf} \cite{vaccon:14a} for more detail).
 \end{itemize}

### 2014-05-21 padicprec.tex
Factor 2 missing in SOMOS algorithm

@@
     $b$, $c$ and $d$ arbitrarily to precision $O(p^{\prec})$\label{line:lift}\;
-  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
+  $\prec$ $\leftarrow$ $\prec - 2 \: v_p(a)$\;
   $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}

### 2014-05-22 padicprec.tex
Small fixes

@@
 A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
-$v_0$ if the following equality holds:
+$v_0$ if $v_0 + H \subset U$ and the following equality holds:
 \begin{equation}
@@
 order-reversing involution: $(\varphi^\star)^\star = \varphi$ and
-$\varphi^\star \geq \psi^\star$ whenever $\varphi \leq \psi$.
+$\varphi^\star \geq \psi^\star$ whenever $\varphi \leq \psi$. When
+necessary, we extend $\varphi^\star$ to $\R \cup \{+\infty\}$ by
+left continuity.
 We refer to \cite{Rockafellar:97} for a complete exposition on Legendre transforms.

### 2014-05-24 sage/precision/coerce.py
A new Sage implementation of padics (very incomplete)

@@ -1,0 +1,26 @@
+def coerce_bigoh(left,right,op):
+    parent = left.parent()
+    path = [ ]
+    try:
+        # Todo: find something better
+        path = parent._models.shortest_path(left.__class__,right.__class__)
+        if len(path) > 0:
+            for i in range(1,len(path)):
+                map = parent._models.edge_label(path[i-1],path[i])
+                if map == None:
+                    left = path[i](parent,left)
+                else:
+                    left = map(parent,left)
+            return op(left,right)
+        path = parent._models.shortest_path(right.__class__,left.__class__)
+        if len(path) > 0:
+            for i in range(1,len(path)):
+                map = parent._models.edge_label(path[i-1],path[i])
+                if map == None:
+                    right = path[i](parent,right)
+                else:
+                    right = map(parent,right)
+            return op(left,right)
+    except RuntimeError:
+        pass
+    raise TypeError("bigoh coercion error: models are %s and %s" % (left._repr_model(), right._repr_model()))

### 2014-05-24 sage/precision/constructor.py

@@ -1,0 +1,17 @@
+from precision.parent_inexact import Parent_inexact
+
+
+def PolynomialRing(parent,*args,**kwargs):
+    if isinstance(parent,Parent_inexact):
+        from precision.polynomial.polynomial_ring import PolynomialRing_inexact as constructor
+    else:
+        from sage.rings.polynomial.polynomial_ring_constructor import PolynomialRing as constructor
+    return constructor(parent,*args,**kwargs)
+
+
+def MatrixSpace(parent,*args,**kwargs):
+    if isinstance(parent,Parent_inexact):
+        from precision.matrix.matrix_space import MatrixSpace_inexact as constructor
+    else:
+        from sage.matrix.matrix_space import MatrixSpace as constructor
+    return constructor(parent,*args,**kwargs)

### 2014-05-24 sage/precision/element_inexact.py

@@ -1,0 +1,453 @@
+#include "/Users/xavier/sage/devel/sage/sage/ext/stdsage.pxi"
+
+from operator import add, mul
+
+from sage.misc.cachefunc import cached_function
+from sage.structure.element import Element 
+from sage.rings.infinity import Infinity
+
+from bigoh import BigOh
+from lazy import LazyApproximation, LazyApproximationWrapper
+
+from grouping.mode import PrecisionMode_Disable, PrecisionMode_Individual, PrecisionMode_Collective
+from grouping.mode import LazyMode_Disable, LazyMode_ASAP, LazyMode_Printing, LazyMode_Lazy
+from grouping.exception import GroupError
+
+from limits import limit_lazyprecision 
+from exception import PrecisionError, ApproximationError
+
+
+class Element_inexact(Element):
+    # for cdef classes, don't forget:
+    # cdef object __weakref__
+
+    def __init__(self, parent, x, prec=None, workprec=None, group=None):
+        from parent_inexact import Parent_inexact
+        if not isinstance(parent, Parent_inexact):
+            raise TypeError("parent must be an instance of Parent_inexact: %s" % parent)
+        Element.__init__(self,parent)
+        base = parent.base_ring()
+        baseapprox = base.approximation()
+        parentapprox = parent.approximation()
+
+        # determine group
+        if group is None:
+            from grouping.context import stack
+            group = stack[-1]
+        else:
+            from grouping.group import GroupElements_inexact
+            if not isinstance(group, GroupElements_inexact):
+                raise TypeError("group not valid")
+        self._group = group
+        if group.precision_mode() is PrecisionMode_Disable:
+            self._precision = None
+        else:
+            self._precision = parent._exact_precision
+
+        # build approximation
+        coeffs = [ ]
+        if isinstance(x,list):
+            dimension = self.parent().dimension()
+            if len(x) > dimension:
+                raise ValueError("list too long")
+            for i in range(len(x)):
+                try:
+                    if x[i].parent() is base:
+                        coeffs.append(x[i])
+                    else:
+                        coeffs.append(base(x[i], group=self._group))
+                except (TypeError,ValueError,PrecisionError):
+                    raise TypeError("unable to create an element from %s" % x)
+            approximation = [ c._approximation for c in coeffs ]
+            if group.precision_mode() is not PrecisionMode_Disable:
+                self._precision = parent.precision()([ c._precision for c in coeffs ])
[…]

### 2014-05-24 sage/precision/grouping/all.py

@@ -1,0 +1,2 @@
+from context import set_default_context, unset_default_context
+from context import Context, context

### 2014-05-24 sage/precision/grouping/bigoh.py

@@ -1,0 +1,39 @@
+from precision.bigoh import BigOh
+from precision.exception import PrecisionError
+
+class GroupBigOh(BigOh):
+    def __init__(self,parent,group):
+        self._group = group
+        from parents import ParentGroupBigOh
+        BigOh.__init__(self, ParentGroupBigOh)
+
+    def __repr__(self):
+        return "BigOh for group %s" % self._group
+
+    def __add__(self,other):
+        raise PrecisionError("Can't add two GroupBigOh")
+
+    def add_element(self,key,prec):
+        #print "Add element %s" % key()
+        pass
+
+    def delete_element(self,key):
+        #print "Delete element %s" % key()
+        pass
+
+    def precision(self,key):
+        # key can be a single element or a list of keys
+        raise NotImplementedError
+
+    def to_workprec(self):
+        return 0
+
+
+class FlatGroupBigOh(GroupBigOh):
+    pass
+
+class JaggedGroupBigOh(GroupBigOh):
+    pass
+
+class LatticeGroupBigOh(GroupBigOh):
+    pass

### 2014-05-24 sage/precision/grouping/exception.py

@@ -1,0 +1,2 @@
+class GroupError(Exception):
+    pass

### 2014-05-24 sage/precision/grouping/group.py

@@ -1,0 +1,109 @@
+import _weakref as weakref
+from sage.structure.sage_object import SageObject
+from sage.rings.infinity import Infinity
+from sage.rings.integer_ring import ZZ
+
+from precision.element_inexact import Element_inexact
+
+from mode import PrecisionMode, LazyMode
+from mode import PrecisionMode_Collective
+
+from exception import GroupError
+
+class GroupElements_inexact(SageObject):
+    def __init__(self, label=None, precision_mode=None, capped_relative=None, lazy_mode=None, remember_members=None):
+        if label is not None and not isinstance(label,str):
+            raise TypeError("label must be None or a string")
+        self._label = label
+        try:
+            from context import stack
+            current_group = stack[-1]
+        except ImportError:
+            pass
+        if precision_mode is None:
+            self._precision_mode = current_group.precision_mode()
+        else:
+            self._precision_mode = PrecisionMode(precision_mode)
+        if capped_relative is None:
+            self._capped_relative = current_group.capped_relative()
+        elif capped_relative is Infinity:
+            self._capped_relative = Infinity
+        else:
+            self._capped_relative = ZZ(capped_relative)
+        if lazy_mode is None:
+            self._lazy_mode = current_group.lazy_mode()
+        else:
+            self._lazy_mode = LazyMode(lazy_mode)
+        if remember_members is None:
+            self._remember = self._precision_mode is PrecisionMode_Collective
+        else:
+            self._remember = remember_members
+        if self._remember:
+            self._members = set()
+
+    def label(self):
+        return self.label
+
+    def precision_mode(self):
+        return self._precision_mode
+
+    def capped_relative(self):
+        return self._capped_relative
+
+    def lazy_mode(self):
+        return self._lazy_mode
+
+    def remember_members(self):
+        return self._remember
+
+    def __repr__(self):
+        import re
+        if self._label is not None:
+            return "Group '%s'" % self._label
+        if self._remember:
+            members = self.members()
+            if len(members) > 0:
+                s = "Group consisting of the following element(s):"
+                for element in self.members():
+                    s += "\n * " + re.sub("\n", "\n   ", element.__repr__())
+            else:
+                s = "Empty group of inexact elements"
+            return s
+        else:
+            return "Group of inexact elements"
+
+    def members(self,safemode=False):
+        if not self._remember:
+            if safemode:
+                return
+            else:
+                raise GroupError("This group does not remember its members")
+        return [ ref() for ref in self._members ]
+
[…]

### 2014-05-24 sage/precision/grouping/mode.py

@@ -1,0 +1,74 @@
+from sage.structure.sage_object import SageObject
+from sage.misc.classcall_metaclass import ClasscallMetaclass
+
+class PrecisionMode(SageObject):
+    __metaclass__ = ClasscallMetaclass
+    _mode = [ None for _ in range(3) ]
+
+    @staticmethod
+    def __classcall__(cls,mode):
+        if isinstance(mode,PrecisionMode):
+            return mode
+        if not isinstance(mode,str):
+            raise TypeError("mode must be an instance of PrecisionMode or a string")
+        mode = mode.lower()
+        if mode == 'disable': mode = 0
+        elif mode == 'individual': mode = 1
+        elif mode == 'collective': mode = 2
+        else:
+            raise ValueError
+        if cls._mode[mode] is None:
+            cls._mode[mode] = cls.__new__(PrecisionMode)
+            cls._mode[mode]._mode = mode
+        return cls._mode[mode]
+
+    def _repr_(self):
+        if self._mode == 0:
+            return "Disable"
+        elif self._mode == 1:
+            return "Individual"
+        elif self._mode == 2:
+            return "Collective"
+
+PrecisionMode_Disable = PrecisionMode("disable")
+PrecisionMode_Individual = PrecisionMode("individual")
+PrecisionMode_Collective = PrecisionMode("collective")
+
+
+class LazyMode(SageObject):
+    __metaclass__ = ClasscallMetaclass
+    _mode = [ None for _ in range(4) ]
+
+    @staticmethod
+    def __classcall__(cls,mode):
+        if isinstance(mode,LazyMode):
+            return mode
+        if not isinstance(mode,str):
+            raise TypeError("mode must be an instance of LazyMode or a string")
+        mode = mode.lower()
+        if mode == 'disable': mode = 0
+        elif mode == 'asap': mode = 1
+        elif mode == 'print' or mode == 'printing': mode = 2
+        elif mode == 'lazy' or mode == 'never': mode = 3
+        else:
+            raise ValueError
+        if cls._mode[mode] is None:
+            cls._mode[mode] = cls.__new__(LazyMode)
+            cls._mode[mode]._mode = mode
+        return cls._mode[mode]
+
+    def _repr_(self):
+        if self._mode == 0:
+            return "Disable"
+        elif self._mode == 1:
+            return "Evaluate as soon as possible"
+        elif self._mode == 2:
[…]

### 2014-05-24 sage/precision/lazy.py

@@ -1,0 +1,486 @@
+import _weakref as weakref
+import inspect
+from types import FunctionType
+from sage.misc.cachefunc import cached_function
+
+from sage.structure.sage_object import SageObject
+from sage.structure.element import Element
+from sage.rings.integer import Integer
+from sage.rings.infinity import Infinity
+from approximation import Approximation
+from limits import maxworkprec, maxworkprec_cache
+
+from exception import PrecisionError
+
+
+class LazyApproximation(Approximation):
+    def __init__(self, parent, x, max_workprec=None, **kwargs):
+        Approximation.__init__(self, parent)
+        self._length = -1
+        self._current_workprec = None
+        self._current_approximation = None
+        self._dict_max_workprec = { }
+        if isinstance(x,Approximation) and x.parent() is parent:
+            if isinstance(x,LazyApproximation):
+                raise TypeError("can't lazy something already lazy")
+            self._current_workprec = Infinity
+            self._current_approximation = x
+            self.repr = lambda: "%s" % x
+            self._length = length = x._length
+            base = parent.base_ring()
+            if length >= 0:
+                zero = base._lazy_zero
+                lazy_class = base._lazy_class
+                @cached_function
+                def getitem_by_num(i):
+                    if i < length:
+                        return lazy_class(base, x._getitem_by_num(i)).new_wrapper()
+                    else:
+                        return zero
+                self._getitem_by_num = getitem_by_num
+        elif isinstance(x,list):
+            self.name = 'element_from_list'
+            if parent.dimension() < Infinity:
+                zeros = [ parent.base_ring()._zero ] * (parent.dimension() - len(x))
+                self.function = lambda workprec: parent([ item.at_workprec(workprec) for item in x ] + zeros)
+            else:
+                self.function = lambda workprec: parent([ item.at_workprec(workprec) for item in x ])
+            x_wrapper = [ c.new_wrapper() for c in x ]
+            self._length = length = len(x)
+            zero = parent.base_ring()._lazy_zero
+            def getitem_by_num(i):
+                if i < length:
+                    return x_wrapper[i]
+                else:
+                    return zero
+            self._getitem_by_num = getitem_by_num
+            self.args = self._dependances = x_wrapper
+            if len(x) == 0:
+                if max_workprec is None:
+                    max_workprec = 0
+            else:
+                if max_workprec is None:
[…]

### 2014-05-24 sage/precision/limits.py

@@ -1,0 +1,4 @@
+maxworkprec = 2**20
+maxworkprec_cache = 2**10
+
+limit_lazyprecision = 100

### 2014-05-24 sage/precision/matrix/approximation.py

@@ -1,0 +1,63 @@
+from sage.structure.unique_representation import UniqueRepresentation
+from sage.structure.mutability import Mutability
+from sage.matrix.matrix_space import MatrixSpace
+from sage.matrix.matrix_generic_dense import Matrix_generic_dense
+
+from precision.approximation import Approximation
+
+class MatrixSpaceApprox(MatrixSpace):
+    def __init__(self, *args, **kwargs):
+        MatrixSpace.__init__(self, *args, **kwargs)
+        from lazy import LazyApproximation_matrix
+        self._lazy_class = LazyApproximation_matrix
+        self._zero = self(0)
+        self._lazy_zero = self._lazy_class(self,[])
+
+    def _get_matrix_class(self):
+        return MatrixApprox
+
+    def indices_basis(self):
+        return [ (i,j) for i in range(self.nrows()) for j in range(self.ncols()) ]
+
+
+class MatrixApprox(Matrix_generic_dense, Approximation):
+    def __init__(self, parent, entries, copy, coerce):
+        Matrix_generic_dense.__init__(self, parent, entries, copy, coerce)
+        Approximation.__init__(self,parent)
+        self._length = parent.dimension()
+        self.set_immutable()
+
+    def matrix_space(self, nrows=None, ncols=None, sparse=None):
+        if nrows is None:
+            nrows = self.nrows()
+        if ncols is None:
+            ncols = self.ncols()
+        if sparse is None:
+            sparse = self.is_sparse()
+        return MatrixSpaceApprox(self.base_ring(), nrows, ncols, sparse=sparse)
+
+    def truncate(self,workprec):
+        return self.__class__(self.parent(), [ c.truncate(workprec) for c in self.list() ], False, False)
+
+    def parenthesis_level(self):
+        return 3
+
+    def __repr__(self):
+        from printing import repr_matrix
+        return repr_matrix([ [ self[i,j] for j in range(self.ncols()) ] for i in range(self.nrows()) ])
+
+    #def valuation_entries(self,lazylimit=None):
+    #    return [ self[i,j].valuation() for i in range(self.nrows()) for j in range(self.ncols()) ]
+
+    def _getitem_by_num(self,n):
+        ncols = self.ncols()
+        j = n % ncols
+        i = (n-j) / ncols
+        return self[i,j]
+
+    # Grrr
+    def _mul_(self,other):
+        res = Matrix_generic_dense._mul_(self,other)
+        Approximation.__init__(res,res.parent())
+        res.set_immutable()
+        return res

### 2014-05-24 sage/precision/matrix/matrix.py

@@ -1,0 +1,126 @@
+from operator import mul
+
+from sage.rings.infinity import Infinity
+from sage.structure.element import Element
+
+from precision.element_inexact import Element_inexact, ModuleElement_inexact
+from precision.bigoh import BigOh, ExactBigOh
+from precision.exception import PrecisionError
+from precision.grouping.mode import LazyMode_Disable
+
+from precision.limits import limit_lazyprecision
+
+
+class Matrix_inexact(ModuleElement_inexact):
+    def __init__(self, parent, x, prec=None, workprec=None, group=None):
+        if parent.is_square():
+            base = parent.base_ring()
+            if isinstance(x,Element) and x.parent() is base:
+                c = x
+            else:
+                try:
+                    c = base(x)
+                except (TypeError,ValueError,PrecisionError):
+                    c = None
+            if c is not None:
+                n = parent.nrows()
+                x = (n**2) * [ base(0) ]
+                for i in range(0,n**2,n+1):
+                    x[i] = c
+        ModuleElement_inexact.__init__(self, parent, x, prec, workprec, group)
+
+    def is_square(self):
+        return self.parent().is_square()
+
+    def __mul__(self,other):
+        parent = self.parent()
+        if isinstance(other,Matrix_inexact) and parent.base_ring() is other.parent().base_ring():
+            group = self.group()
+            if other.group() != group:
+                raise TypeError("Can multiply only two matrices in the same group")
+            return self._matrix_times_matrix_(other)
+        else:
+            from sage.structure.element import get_coercion_model
+            return get_coercion_model().bin_op(self,other,mul)
+
+    def _matrix_times_matrix_(self,other):
+        if self.ncols() != other.nrows():
+            raise TypeError("incompatible dimensions")
+        from matrix_space import MatrixSpace_inexact
+        parent = MatrixSpace_inexact(self.parent().base_ring(), self.nrows(), other.ncols())
+        selfval = self.valuation_bigoh(lazylimit=limit_lazyprecision)
+        otherval = other.valuation_bigoh(lazylimit=limit_lazyprecision)
+        precision = self._precision * otherval + selfval * other._precision
+        if self._group.lazy_mode() is not LazyMode_Disable:
+            selfv = selfval.flat_below()
+            otherv = otherval.flat_below()
+            approximation = self._approximation._mul_(other._approximation, workprec_shifts=[-otherv,-selfv], expected_valuation = selfv+otherv)
+        else:
[…]

### 2014-05-24 sage/precision/padics/ZZp_element.py

@@ -1,0 +1,42 @@
+from sage.rings.rational_field import QQ
+from sage.rings.infinity import Infinity
+
+from sage.structure.element import RingElement
+
+from precision.element_inexact import RingElement_inexact
+from approximation import QQpApprox_element
+from precision.bigoh import ExactBigOh, ValuationBigOh
+from QQp_element import QQp_element
+from lazy import LazyApproximation_padics
+
+from precision.exception import ApproximationError
+
+
+class ZZp_element(QQp_element):
+    def __init__(self, parent, x, prec=None, *args, **kwargs):
+        approxring = parent.approximation()
+        if isinstance(x, QQpApprox_element) or isinstance(x, LazyApproximation_padics):
+            approximation = x
+        elif x in QQ:
+            if x.valuation(parent._p) < 0:
+                raise ValueError("%s is not in Z_%s" % (x, parent._p))
+            approximation = QQpApprox_element(approxring, x, 0)
+        else:
+            raise TypeError
+        RingElement_inexact.__init__(self, parent, approximation, prec, *args, **kwargs)
+
+    def is_unit(self):
+        return self.valuation() == 0
+
+    def __invert__(self):
+        if self._group is not None:
+            raise NotImplementedError
+        parent = self.parent()
+        if self.valuation() != 0:
+            raise ValueError("%s is not invertible in Z_%s" % (self, parent._p))
+        precision = self._precision
+        def lazy_invert(prec):
+            return ~self.approximation(prec)
+        default_precision = ValuationBigOh(parent.precision(), 1)
+        approximation = parent._lazy_class(parent.approximation(), lazy_invert, default_precision, args=[self])
+        return self._new_fast(self.parent(), approximation, precision)

### 2014-05-24 sage/precision/parent_inexact.py

@@ -1,0 +1,102 @@
+from sage.structure.unique_representation import UniqueRepresentation
+
+from sage.structure.parent import Parent
+from sage.rings.integer_ring import ZZ
+from sage.rings.infinity import Infinity
+
+from parent_precision import ParentBigOh
+#from precision_mode import *
+from bigoh import ExactBigOh, ValuationBigOh, FlatBigOh, FlatIntervalBigOh, JaggedBigOh
+from element_inexact import Element_inexact
+
+from exception import PrecisionError
+
+
+class Parent_inexact(Parent,UniqueRepresentation):
+    def __init__(self, base, parent_approximation, models=None, exact=Infinity,
+                 category=None, element_constructor=None, gens=None, names=None, normalize=True, facade=None):
+        Parent.__init__(self,base,category=category,element_constructor=element_constructor,
+                        gens=gens, names=names, normalize=normalize, facade=facade)
+        self._approximation = parent_approximation
+        self._precision = ParentBigOh(self)
+        if models is None:
+            dimension = self.dimension()
+            if dimension == 1:
+                self._precision.add_model(ValuationBigOh)
+            else:
+                if self.dimension() < Infinity:
+                    self._precision.add_model(FlatBigOh)
+                self._precision.add_model(JaggedBigOh)
+                try:
+                    if self.indices_basis() is None:
+                        self._precision.add_model(FlatIntervalBigOh)
+                        self._precision.add_modelconversion(FlatIntervalBigOh,JaggedBigOh)
+                except NotImplementedError:
+                    pass
+                self._precision.add_modelconversion(FlatBigOh,JaggedBigOh,safemode=True)
+                self._precision.add_modelconversion(FlatBigOh,FlatIntervalBigOh,safemode=True)
+        else:
+            for model in models.vertices():
+                self._precision.add_model(model)
+            for (mfrom,mto,map) in models.edges():
+                self._precision.add_modelconversion(mfrom,mto,map)
+        try:
+            self._exact_precision = self._precision(exact)
+        except PrecisionError:
+            self._exact_precision = ExactBigOh(self._precision)
+
+    def __hash__(self):
+        return id(self)
+
+    def uniformizer_name(self):
+        base = self._base
+        if hasattr(base, "uniformizer_name"):
+            return base.uniformizer_name()
+        elif hasattr(base, "uniformizer"):
+            import re
+            return re.sub("\s*\+\s*O(.*)", "", base.uniformizer().__repr__())
+        else:
[…]

### 2014-05-24 sage/precision/parent_precision.py

@@ -1,0 +1,144 @@
+from sage.structure.unique_representation import UniqueRepresentation
+from sage.structure.parent import Parent
+
+from sage.graphs.digraph import DiGraph
+
+from exception import PrecisionError
+
+
+class ParentBigOh(Parent,UniqueRepresentation):
+    def __init__(self,ambiant):
+        try:
+            self._uniformizer = ambiant.uniformizer_name()
+        except NotImplementedError:
+            raise TypeError("Impossible to determine the name of the uniformizer")
+        self._ambiant_space = ambiant
+        self._models = DiGraph(loops=False)
+        self._default_model = None
+        Parent.__init__(self,ambiant.base_ring())
+        if self.base_ring() is None:
+            self._base_precision = None
+        else:
+            if self.base_ring() == ambiant:
+                self._base_precision = self
+            else:
+                self._base_precision = ParentBigOh(self.base_ring())
+
+    def __hash__(self):
+        return id(self)
+
+    def base_precision(self):
+        return self._base_precision
+
+    def precision(self):
+        return self._precision
+
+    def default_model(self):
+        if self._default_mode is None:
+            self.set_default_model()
+        return self._default_model
+    
+    def set_default_model(self,model=None):
+        if model is None:
+            self._default_model = self._models.topological_sort()[-1]
+        else:
+            if self._models.has_vertex(model):
+                self._default_model = model
+            else:
+                raise ValueError
+
+    def add_model(self,model):
+        from bigoh import BigOh
+        if not isinstance(model,list):
+            model = [model]
+        for m in model:
+            if not issubclass(m,BigOh):
+                raise TypeError("A precision model must derive from BigOh but '%s' is not"%m)
+            self._models.add_vertex(m)
+
+    def delete_model(self,model):
+        if isinstance(model,list):
+            model = [model]
+        for m in model:
+            if self._models.has_vertex(m):
+                self._models.delete_vertex(m)
+
+    def update_model(self,old,new):
+        from bigoh import BigOh
+        if self._models.has_vertex(old):
+            if not issubclass(new,BigOh):
+                raise TypeError("A precision model must derive from BigOh but '%s' is not"%new)
+            self._models.relabel({old:new})
+        else:
+            raise ValueError("Model '%m' does not exist"%old)
+
+    def add_modelconversion(self,domain,codomain,constructor=None,safemode=False):
+        if not self._models.has_vertex(domain):
[…]

### 2014-05-24 sage/precision/polynomial/polynomial_element.py

@@ -1,0 +1,91 @@
+from sage.rings.infinity import Infinity
+
+from precision.element_inexact import RingElement_inexact
+from precision.bigoh import ExactBigOh
+
+from precision.exception import PrecisionError
+from precision.limits import limit_lazyprecision
+
+
+class Polynomial_inexact(RingElement_inexact):
+    def _normalize(self):
+        pass
+
+    def variable_name(self):
+        return self.parent().variable_name()
+
+    def valuation_bigoh(self,model=None,lazylimit=Infinity):
+        from bigoh import FlatIntervalBigOh_polynomial, JaggedBigOh_polynomial, NewtonBigOh_polynomial, LatticeBigOh_polynomial
+        deg = self.degree(minimal=False)
+        valuations = [ self[i].valuation(minimal=True,lazylimit=lazylimit) for i in range(deg+1) ]
+        valuations.append(Infinity)
+        parentprec = self._precision.parent()
+        if model is None:
+            model = self.precision_model()
+        if model is FlatIntervalBigOh_polynomial:
+            val = min(valuations)
+            first = len(valuations)
+            for i in range(len(valuations)):
+                if valuations[i] < Infinity:
+                    first = i
+                    break
+            last = 0
+            for i in range(len(valuations), 0, -1):
+                if valuations[i-1] < Infinity:
+                    last = i
+                    break
+            return FlatIntervalBigOh_polynomial(parentprec, val, last, first)
+        elif model is JaggedBigOh_polynomial:
+            return JaggedBigOh_polynomial(parentprec, valuations)
+        elif model is NewtonBigOh_polynomial or model is ExactBigOh:
+            return NewtonBigOh_polynomial(parentprec, valuations)
+        else:
+            raise TypeError("invalid model")
+
+    def degree(self,minimal=True,lazylimit=Infinity):
+        deg = -1
+        for d in range(self._approximation._expected_degree,-1,-1):
+            if not self[d].is_zero(lazylimit=lazylimit): 
+                deg = d
+                break
+        if minimal:
+            return deg
+        else:
+            return max(deg, self.precision().last()-1)
+
+    def add_bigoh(self,prec,degree=None):
+        from bigoh import FlatIntervalBigOh_polynomial
+        parentprec = self.parent().precision()
+        precision = self._precision
+        try:
+            precscal = parentprec._base_precision(prec)
+            if degree == None:
+                degree = self.degree(minimal=False)
[…]

### 2014-05-24 sage/precision/polynomial/printing.py

@@ -1,0 +1,33 @@
+from precision.printing import repr_operator, repr_parenthesis, repr_exponant, simplify
+
+def repr_polynomial(coeffs, variable_name, variable_level_parenthesis=3):
+    length = len(coeffs)
+    if length == 0: return '0'
+    terms = [ ]
+    for i in range(length):
+        coeff = coeffs[i]
+        s = str(coeff)
+        if s == '0': continue
+        if coeff.parenthesis_level() < 1:
+            s = repr_parenthesis(s)
+        if i == 1:
+            if variable_level_parenthesis < 1:
+                monome = repr_parenthesis(variable_name)
+            else:
+                monome = variable_name
+        elif i > 1:
+            if variable_level_parenthesis < 2:
+                monome = repr_parenthesis(variable_name)
+            else:
+                monome = variable_name
+            monome = repr_exponant(monome, str(i))
+        if i == 0:
+            terms.append(s)
+        elif s == '1':
+            terms.append(monome)
+        else:
+            terms.append(repr_operator("*", [s, monome]))
+    terms.reverse()
+    s = repr_operator(" + ", terms)
+    if s == '': s = '0'
+    return simplify(s)

### 2014-05-24 sage/precision/tutorial.py

@@ -1,0 +1,635 @@
+r"""
+All modules in this directory (and subdirectories) are devoted to a 
+new implementation of inexact objects (e.g. p-adics, formal series)
+in Sage.
+
+WARNING:
+
+    Everything is written in Python and very very slow. Do not use 
+    this package for performance!
+
+
+INSTALLATION::
+
+Just copy the package in some directory /my/directory/precision
+and add the three following lines to $HOME/.sage/init.sage::
+
+    import sys
+    sys.path.append("/my/directory/")
+    from precision.all import *
+
+
+WORKING WITH p-ADIC NUMBERS::
+
+We first create the p-adic field Q_5::
+
+    sage: K = QQp(5)
+    sage: K
+    5-adic Field
+
+Elements in K are created by the usual syntax::
+
+    sage: x = K(0)
+    sage: x
+    0
+    sage: x.parent()
+    5-adic Field
+
+Note that the element `x` is exactly known (i.e. it does not carry 
+a precision - or, more exactly, it carries an exact precision as we 
+shall see later)::
+
+    sage: x.is_exact()
+    True
+
+Of course, it is also possible to create inexact objects.
+The simplest way to do this is to specify the precision when the
+object is created::
+
+    sage: y = K(1,20)
+    sage: y
+    1 + O(5^20)
+    sage: y.is_exact()
+    False
+
+Another possibility to create inexact p-adics is to use the
+method ``add_bigoh`` as follows ::
+
+    sage: z = K(27); z
+    27
+    sage: z.add_bigoh(50)
+    27 + O(5^50)
+
+Of course, we can do basic operations with p-adics::
+
+    sage: x+y
+    1 + O(5^20)
+    sage: x*y
+    0
+    sage: x+y-y
+    O(5^20)
+    sage: y^25
+    1 + O(5^22)
+
+Other usual functions are availiable::
+
+    sage: x = K(6,20); x
+    6 + O(5^20)
+    sage: y = log(x); y
+    9146849050361*5^1 + O(5^20)
+    sage: exp(y)
+    6 + O(5^20)
+    sage: x.teichmuller()
+    1
+
+
+LAZY p-ADICS::
+
+It may happen that computing a value requires an infinite amount
+of computation. For example, if `x` is `5`-adic number `6` known
+with infinite precision, computing `\log(x)` requires to expand 
[…]

### 2014-05-24 sage/precision/window.py

@@ -1,0 +1,10 @@
+import struct, fcntl, termios
+
+from sage.rings.infinity import Infinity
+
+def width():
+    s = struct.pack('HHHH', 0, 0, 0, 0)
+    x = fcntl.ioctl(1, termios.TIOCGWINSZ, s)
+    width = struct.unpack('HHHH', x)[1]
+    if width == 0: width = Infinity
+    return width

### 2014-05-26 padicprec.tex
Changes for final copy, including removing most of Appendix B

@@
 \title{Tracking $p$-adic precision}
-%\author{Xavier Caruso, David Roe \& Tristan Vaccon}
+\author{Xavier Caruso, David Roe and Tristan Vaccon}
 \date\today
@@
 to $p$-adic manifolds, describing how to specify precisions for points on elliptic curves
-and Grassmannians.  Finally, Appendix \ref{sec:differentials} describes how to compute
-the derivative of many common operations on polynomials and matrices, with an eye
+and Grassmannians.  Finally, Appendix \ref{sec:matrices} describes how to compute
+the derivative of some common operations on matrices as an example, with an eye
 toward applying Lemma \ref{lem:main}.
@@
 \begin{enumerate}
-\item compute the Laurent polynomials $P_n$ using the recurrence in 
+\item compute the polynomials $P_n$ using the recurrence in 
 the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$
@@
 tackle many classical questions about matrices or linear systems, and is
-discussed further in Appendix \ref{ssec:matrices}.
+discussed further in Appendix \ref{sec:matrices}.
 When computing the entries of $L$ and $U$ from a $d \times d$ matrix
@@
 It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
-\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.
-It is important when working with overconvergent modular symbols since these lattices are stable under certain Hecke operators.
+\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices of the form $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.
+These lattices are stable under certain Hecke operators, which is necessary for computing with overconvergent modular symbols.
 \end{itemize}
@@
 Note that sometimes the precision of a final result can be computed
-a priori (using the methods of Appendix \ref{sec:differentials} for example).
+a priori (using the methods of Appendix \ref{sec:matrices} for example).
 Taking advantage of such knowledge can minimize artificial precision loss
@@
 the values of $a$, $b$, $c$ and $d$ after the execution of line
-\ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
+\ref{line:lift}. Applying Lemma \ref{lem:main} and Proposition
 \ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
 $(u_0, u_1, u_2, u_3)$), we get:
-$$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
-= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).$$
-By the discussion above, this implies in particular that $f(a',b',c',d')$
+\[
+f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
+= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).
+\]
+By the discussion above, this equation implies in particular that $f(a',b',c',d')$
 belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
-conclude by remarking first that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
-{p^\prec \Z_p^4}$ by construction and second that $p^\prec \Z_p^4 \subset
+conclude by remarking that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
+{p^\prec \Z_p^4}$ by construction and that $p^\prec \Z_p^4 \subset
 p^N \varphi_i(\Z_p^4)$.
@@
 above and perform all the calculations in these charts.
+
+\section{Example: Matrices}
+\label{sec:matrices}
+
+We saw in the core of the article that the differential of an operation 
+encodes the intrinsic loss/gain of precision when performing this 
+operation. In this appendix we compute the differential of various
+common operations on matrices. Surprisingly, 
+we observe that all differentials we will consider are rather easy to 
+compute even if the underlying operation is quite involved.
+
+In what follows, we use freely the ``method of physicists'' to compute 
+differentials: given a function $f$ differentiable at some point $x$, we 
+consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$ 
+by expanding LHS and neglecting terms of order $2$. The differential of 
+$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
+
+\subsubsection*{Determinants and characteristic polynomials.}
+
+We first outline the standard computation of the differential of the function
+$\det : M_n(K) \to K$. Suppose that $M \in \GL_n(K)$ and that $\com(M) = \det(M) M^{-1}$.
+Then
+\begin{align*}
+\det(M + dM) &= \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
+&= \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
+&= \det(M) + \tr(\com(M) \cdot dM).
+\end{align*}
+The differential of $\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
+out that this formula is still valid when $M$ is not invertible.
+The same computation extends readily to characteristic polynomials,
+since they are defined as determinants. More precisely, let us 
+consider the function $\chi : M_n(K) \to K_n[X]$ taking a matrix 
+$M$ to its monic characteristic polynomial $\det(X-M)$.
+Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
+differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
+
+\subsubsection*{LU factorization.}
+
+Define the LU factorization of a square matrix $M \in M_n(K)$ as a decomposition $M = LU$ 
+where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
+Such a decomposition exists and is unique provided that no principal minor
+of $M$ vanishes. We can then consider the mapping $M \mapsto (L,U)$ 
+defined over the Zariski-open set of matrices satisfying the above 
+condition. In order to differentiate it, we differentiate the relation 
+$M = LU$ and rewrite the result as
+$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
+We remark that in the right hand side of the above formula, the first
+summand is lower triangular with zero diagonal whereas 
+the second summand is upper triangular. Hence in order to compute $dL$
+and $dU$, one can proceed as follows: 
+\begin{enumerate}
+\item compute the product $dX = L^{-1} dM \: U^{-1}$,
+\item separate the lower and upper part of $dX$ obtaining $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$
+\item recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the left and $U$ on the right respectively.
+\end{enumerate}
+The above discussion extends almost \emph{verbatim} to LUP 
+factorization; the only difference is that LUP factorizations are not 
+unique but they are on a small neighborhood of $M$ if we fix the matrix 
+$P$.
+
+\subsubsection*{QR factorization.}
+
+A QR factorization of a square matrix $M \in M_n(K)$ will be
+a decomposition $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
+orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
+a decomposition exists and is unique on a Zariski-open subset of $M_n(K)$. The
+mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
+would like to emphasize at this point that the orthogonality condition
+defines a sub-manifold of $M_n(K)$ which is \emph{not} a
+vector space: it is defined by equations of degree $2$. The codomain
+of $f$ is then also a manifold; this example then fits into the setting
+of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
+We can differentiate $f$ by following the method used for LU 
+factorization.  Differentiating the relation $M = QR$, we obtain
+\begin{equation}
+\label{eq:diffQR}
+\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
+\cdot R^{-1}
+\end{equation}
+where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
+Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
+$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
+matrix and a diagonal one. Since moreover $dR \cdot R^{-1}$ is upper
+triangular with all diagonal entries equal to $0$, we see that 
+\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
+and $dM$.
+
+\bibliographystyle{plain}
+\bibliography{roebib/Biblio,extras}
+
+\affiliationone{
+Xavier Caruso and Tristan Vaccon \\
+Institut de recheche en math\'ematiques de Rennes (IRMAR) \\
+Universit\'e Rennes 1 \\
+Campus de Beaulieu \\
+35042 Rennes Cedex \\
+France}
+\affiliationtwo{
+David Roe \\
+Department of Mathematics \\
+University of Calgary \\
+2500 University Dr NW \\
+Calgary AB T2N 1N4 \\
+Canada}
+
+\end{document}
 
@@
 consequently, sums of subspaces.
-
-\bibliographystyle{plain}
-\bibliography{roebib/Biblio,extras}
-
-\end{document}

### 2014-08-04 examples.tex
Starting a new paper: examples.tex. Until now:  - I wrote a technical result which may be useful (hopefully)    to bound many growing functions Lambda(f), and  - I copy-pasted the removed appendix of our latest paper.

--- 2014-05-21 padicprec.tex
@@
-%\documentclass{sig-alternate}
-\documentclass{lms}
+\documentclass{sig-alternate}
+%\documentclass{lms}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{amsrefs}
-\usepackage{color}
+\usepackage[usenames,dvipsnames]{color}
 \usepackage{stmaryrd}
@@
 \usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
+\usepackage{comment}
+\usepackage{tikz}
 \renewcommand*{\theHsection}{\thesection}
@@
 \newtheorem{quest}[theo]{Question}
+%\theoremstyle{definition}
 \newtheorem{rem}[theo]{Remark}
 \newtheorem{ex}[theo]{Example}
-%\theoremstyle{definition}
 \newtheorem{deftn}[theo]{Definition}
@@
 
+\newcommand{\id}{\textrm{id}}
+\newcommand{\Epi}{\textrm{Epi}}
+
 \newcommand{\lb}{\ensuremath{\llbracket}}
@@
 
-\title{Tracking $p$-adic precision}
-%\author{Xavier Caruso, David Roe \& Tristan Vaccon}
-\date\today
+\def\binom#1#2{\Big(\begin{array}{cc} #1 \\ #2 \end{array}\Big)}
+
+\title{Tracking p-adic precision: examples}
+
+\numberofauthors{3}
+\author{
+\alignauthor Xavier Caruso\\
+  \affaddr{Universit\'e Rennes 1}\\
+  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
+\alignauthor Tristan Vaccon\\
+  \affaddr{Universit\'e Rennes 1}\\
+  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
+\alignauthor David Roe \\
+  \affaddr{University of Calgary}\\
+  \email{\normalsize \textsf{roed.math@gmail.com}}
+}
 
 \maketitle
+
 \begin{abstract}
-We present a new method to propagate $p$-adic precision in computations, which also
-applies to other ultrametric fields.
-We illustrate it with many examples and give a toy application to the
-stable computation of the SOMOS 4 sequence.
 \end{abstract}
 
-\setcounter{tocdepth}{1}
-\tableofcontents
+\vspace{1mm}
+ \noindent
+ {\bf Categories and Subject Descriptors:} \\
+\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
+  Manipulation -- \emph{Algebraic Algorithms}
+
+ \vspace{1mm}
+ \noindent
+ {\bf General Terms:} Algorithms, Theory
+
+ \vspace{1mm}
+ \noindent
+ {\bf Keywords:} 
+\medskip
 
@@
 
-The last two decades have seen a rise in the popularity of $p$-adic methods in
-computational algebra.  For example,
-\begin{itemize}
-\item Bostan et al. \cite{boston-gonzalez-perdry-schost:05a} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
-\item Gaudry et al. \cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
-\item Kedlaya \cite{kedlaya:01a}, Lauder \cite{lauder:04a} and many followers used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
-\item Lercier and Sirvent \cite{lercier-sirvent:08a} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
-\end{itemize}
-Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
-stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
-that apply across many different algorithms.
-
-Two sources of inspiration arise when studying $p$-adic algorithms.
-The first relates $\Zp$ to its quotients $\Z/p^n\Z$.  The preimage in
-$\Zp$ of an element $a \in \Z / p^n \Z$ is a ball, and these balls cover
-$\Zp$ for any fixed $n$.  Since the projection $\Zp \to \Z/p^n\Z$ is a
-homomorphism, given unknown elements in two such balls we can
-locate the balls in which their sum and product lie.  Working on a computer
-we must find a way to write elements using only a finite amount of data.
-By lumping elements together into these balls of radius $p^{-n}$, we may
-model arithmetic in $\Zp$ using the finite ring $\Z / p^n \Z$.  In this representation,
-all $p$-adic elements have constant \emph{absolute precision} $n$.
-
-The second source draws upon parallels between $\Qp$ and $\R$.  Both
-occur as completions of $\Q$ and we represent elements of both in terms
-of a set of distinguished rational numbers.
-In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
-subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
-\[
-\Big\vert \frac{x \circledast y}{x \ast y} - 1 \Big\vert \le 2^{-h}
-\]
-for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
-defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
-\[
-\Big\vert \frac{x \circledast y}{x \ast y} - 1 \Big\vert_p \le p^{-h}.
-\]
-When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.
-
-In both of these models, precision (absolute or relative) is constant across all elements.
-Since some operations lose precision, it can be useful to attach a precision to each element.
-Over the reals, such interval arithmetic is unwieldy, since arithmetic operations always increase
-the lengths of the inputs.  As a consequence, most computations in the real numbers rely on
-statistical cancelation and external estimates of precision loss, rather than
-attempting to track known precision at each step.  This tendency is strengthened
-by the ubiquity of floating point arithmetic in scientific applications, where Gaussian
-distributions are more common than intervals anyway.
-
-In the $p$-adic world, precision tracking using intervals is much more feasible.
-Even a long sequence of operations with such elements may not sacrifice any precision.  Intervals
-allow number theorists to provably determine a result modulo a given power of $p$,
-and the Gaussian distributions of measurement error over $\R$ have no direct analogue
-over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
-in implementations of $p$-adic numbers.  The mathematical software packages
-Sage \cite{sage}, PARI \cite{pari} and Magma \cite{magma} all include $p$-adic elements
-that track precision in this way.
-
-The approach of propagating precision with each arithmetic operation works well, but does
-sometimes underestimate the known precision of a result, as we will discuss in Section
-\ref{ssec:stepbystep}.  Moreover, elements of $\Qp$ provide building blocks for generic
-implementations of polynomials, vector spaces, matrices and power series.  The practice
-of storing the precision within each entry is not flexible enough for all applications.
-Sometimes only a rough accounting of precision is needed, in which case storing
-and computing the precision of each entry in a large matrix needlessly consumes
-space and time.  Conversely, recording the precision of each entry does not allow
-a constraint such as specifying the precision of $f(0)$, $f(1)$ and $f(2)$ for a
-quadratic polynomial $f$.
-
-For a vector space $V$ over $\Qp$, we propose that the fundamental object used to
-store the precision of an element should be a $\Zp$-lattice $H \subset V$.  By using
-general lattices one can eliminate needless loss of precision.  Moreover, specifying
-the precision of each entry or recording a fixed precision for all entries can both
-be interpreted in terms of lattices.  In Section \ref{sec:prec-proposal} we detail our
-proposal for how to represent the precision of an element of a vector space.
-
-In Section \ref{sec:mainlemma}, we develop the mathematical background on which our proposal is based.
-The most notable result of this section is Lemma \ref{lem:main} which describes how lattices transform under non-linear maps and allows us to propagate
-precision using differentials.  More specifically, it describes a class of first order lattices,
-whose image under a map of Banach spaces is obtained by applying the
-differential of that map.  In Section \ref{ssec:locanalytic} we make the conditions
-of Lemma \ref{lem:main} more explicit in the case of locally analytic functions.
-
-In Section \ref{sec:tracking} we propose methods for tracking precision in practice.
-Section \ref{ssec:opt-tracking} includes a discussion of two models of precision tracking:
-one-pass tracking, where the precision lattice is propagated at each step of the algorithm,
-and two-pass tracking, where an initial pass computing rough approximations is used
-in computing the precision lattices.  We introduce precision types in Section \ref{ssec:types},
-which allow a tradeoff between flexibility, space and time in computing with precision.
-In Section \ref{ssec:SOMOS-solution}, we give
-an application of these ideas to an algorithm for computing terms of the SOMOS sequence.
-
-In Appendix \ref{sec:manifold}, we extend the results of Section \ref{sec:mainlemma}
-to $p$-adic manifolds, describing how to specify precisions for points on elliptic curves
-and Grassmannians.  Finally, Appendix \ref{sec:differentials} describes how to compute
-the derivative of many common operations on polynomials and matrices, with an eye
-toward applying Lemma \ref{lem:main}.
-
-\section{Precision proposals} \label{sec:prec-proposal}
-
-\subsection{Problems in precision}
-\label{ssec:stepbystep}
-
-The usual way to track $p$-adic precision consists in replacing $p$-adic 
-numbers by approximate elements of the form $a + O(p^N)$ and performing 
-all usual arithmetical operations on these approximations.
-%Recall that 
-%an element $a + O(p^N)$ of $\Qp$ has \emph{absolute precision} $N$ if it 
-%is well defined up to adding an unknown element of $p^N\Zp$.
-We offer below three examples that illustrate cases where this way to 
-track precision does not yield optimal results.
-
-\subsubsection*{A linear map.}
-
-Consider the function $f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem
-of computing $f \circ f(a + O(p^n), b + O(p^m))$.  Applying $f$ twice, computing precision
-with each step, yields $\left(2a + O(p^{\min(m,n)}), 2b + O(p^{\min(m, n)})\right)$.
-On the other hand, $f \circ f(x,y) = (2x, 2y)$, so one may compute the result more
-accurately as $(2a + O(p^n), 2b + O(p^m))$, with even more precision when $p=2$.
-
-\subsubsection*{SOMOS 4.}
-
-The SOMOS 4 sequence \cite{somos:89a} is defined by the recurrence
-\[
-u_n = \frac{u_{n-3} u_{n-1} + u_{n-2}^2}{u_{n-4}}.
-\]
-We shall consider the case where the initial
-terms $u_0$, $u_1$, $u_2$ and $u_3$ lie in $\Zp^\times$ and have precision $O(p^N)$.
-Let us first examine how the absolute precision of $u_n$ varies with $n$
-if it is computed from the precision of $u_{n-4}, \ldots, u_{n-1}$ using the recurrence.
-The computation of $u_n$ involves a division by $u_{n-4}$ and hence, roughly speaking,
-decreases the precision by a factor $p^{\val(u_{n-4})}$.  Hence the
-step-by-step computation returns the value of $u_n$ with precision
-\begin{equation}
-\label{eq:SOMOS}
-O(p^{N-v_n})
-\quad \text{with}\quad
-v_n = \val(u_0) + \cdots + \val(u_{n-4}).
-\end{equation}
-
-On the other hand, one can prove that the SOMOS 4 sequence 
-exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}: for all integer 
-$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
-1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
-From the latter formula and our assumption that $u_0$, $u_1$, 
-$u_2$ and $u_3$ are units known up to precision $O(p^N)$, it follows that all $u_n$'s
-are known with the same precision. Thus, the term $v_n$ that 
-appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
-precision but some numerical instability related to the algorithm used to compute $u_n$.
-
-\begin{rmk}
-From the above discussion, one can easily derive a numerically stable 
-algorithm that computes the SOMOS 4 sequence: 
-\begin{enumerate}
-\item compute the Laurent polynomials $P_n$ using the recurrence in 
-the ring $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 1}, T^{\pm 1}]$
-\item evaluate $P_n$ at the point $(u_0, u_1, u_2, u_3)$.
-\end{enumerate}
-However, computing the $P_n$'s is very time-consuming since it requires 
-division in a polynomial ring with $4$ variables and the
-size of the coefficients of $P_n$ explodes as $n$ grows.
-
-In Section \ref{ssec:SOMOS-solution}, we shall design an algorithm computing 
-the SOMOS 4 sequence which turns out to be, at the same time, efficient 
-and numerically stable.
-\end{rmk}
-
-\subsubsection*{LU factorization.}
-
-Let us first recall that a square matrix $M$
-admits a LU factorization if it can be written as a product $LU$ where 
-$L$ and $U$ are lower triangular and upper triangular respectively. 
-The computation of a LU factorization appears as an important tool to 
-tackle many classical questions about matrices or linear systems, and is
-discussed further in Appendix \ref{ssec:matrices}.
-When computing the entries of $L$ and $U$ from a $d \times d$ matrix
-over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
-\begin{itemize}
-\item using usual Gaussian elimination and tracking $p$-adic precision 
-step-by-step, the smallest precision on an entry of $L(M)$ is about
-$O(p^{N - \frac{2d}{p-1}})$ on average;
-\item computing $L(M)$ by evaluating Cramer-type formulae yields a
-result whose every entry is known up to precision $O(p^{N - 2 \log_p 
-d})$ \cite{caruso:12a}.
-\end{itemize}
-If $d$ is large compared to $p$, the second precision is much more
-accurate than the first one. On the other hand, the Cramer-type formulae
-in the second algorithm yield a substantially longer running time than the first.
-%In \cite{caruso:12a}, the author designs a new algorithm which combines 
-%both advantages: efficiency and stability.
-%The precision loss inherent in an algorithm is an important factor in 
-%analyzing its performance.
-%However, since this precision loss depends heavily on the details of the 
-%problem and the algorithm, we will not focus on such questions in this 
-%paper.
-
-\subsection{Lattices}
-
-In order to make our proposals for tracking precision clear, we need
-some definitions from ultrametric analysis.  See 
-Schneider \cite{schneider:11a} for a more complete exposition.
-
-Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
-We assume that the induced metric is an ultrametric (\emph{i.e.}
-$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
-and that $K$ is complete with respect to it.
-For example, we may take $K = \Qp$ with the
-$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
-Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
-that $K$ contains a computable dense subring $R \subset K$ \cite{rabin:60a}.
-This assumption holds for $K = \Qp$ and $K=\Fp\lp t \rp$ by setting $R = \Z[\frac1p]$ or $R = \Q$
-in the case of $\Qp$ and $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$ in the case of $\Fp\lp t \rp$.
-
-If $E$ is a $K$-vector space, possibly of infinite dimension, then an
-\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
-\begin{enumerate}[(i)]
-\item $\Vert x\Vert = 0$ if and only if $x = 0$;
-\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
-\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
-\end{enumerate}
-A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
-normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
-A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
-We underline that any lattice $H$ in $E$ is also closed since its 
-complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
-are all open.
-For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
-$$B_E(r) = \{ x \in E : \Vert x \Vert \le r \}, 
-\quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
-Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.  We will also set $B_E(\infty) = B^-_E(\infty) = E$.
-
-Suppose $E$ is a $K$-Banach space and $I$ a set.
-A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
-every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
-for scalars $\alpha_i \in K$ with $\alpha_i \to 0$ (according to the filter of cofinite subsets), and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
-Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
-
-Given a basis $(x_i)_{i \in I}$ and a sequence $(r_i)_{i \in I}$ with $r_i \in \R_{>0}$, the sets
+\section{Preliminaries}
+
+In the next section, we will see that the derivative at some point $x$ 
+of many standard operations, modelised by a function $f$, has a simple 
+expression in term of $x$ and $f(x)$. In other terms, such functions $f$
+satisfy:
+\begin{equation}
+\label{eq:diffequa}
+f' = g \circ (f, \id)
+\end{equation}
+where $g$ is a given --- and hopefully rather simple --- function. 
+The aim of these preliminaries is to study the differential equation
+\eqref{eq:diffequa} (or, actually, a slight generalization of it) and
+to derive from it some bounds on the growing function $\Lambda(f)$ 
+defined in \cite{padicprec}.
+
+\begin{comment}
+
+\subsection{Legendre transform of a compositum}
+
+For $i \in \{1,2\}$, let $\varphi : \R \to \R \cap \{+\infty\}$ denote a 
+convex function and $\varphi_i^\star$ be its Legendre transform defined 
+by:
+\begin{equation}
+\label{eq:defLegendre}
+\varphi_i^\star(u) = \sup_{v \in \R} \varphi(v) - uv.
+\end{equation}
+Set $\varphi = \varphi_1 \circ \varphi_2$. The aim of this subsection is 
+to relate $\varphi^\star$ with $\varphi_1^\star$ and $\varphi_2^\star$.
+
+\begin{lem}
+\label{lem:Legendrecomp}
+We assume that $\varphi_1$ and $\varphi_2$ are nondecreasing and that 
+their slopes are all in $\N$. Then, for all nonnegative interger $s$:
+$$\varphi^\star(s) =
+\inf_{\ell, (n_i)} \varphi_1^\star(\ell) + \varphi_2^\star(n_1) + 
+\cdots + \varphi_2^\star(n_\ell)$$
+where the infimum is taken over all finite sequences of nonnegative
+integers (of variable length) $n_1, \ldots, n_\ell$ with $n_1 + \cdots +
+n_\ell = s$.
+\end{lem}
+
+\begin{proof}
+Let $n_1, \ldots, n_\ell$ be a sequence as above. Taking the supremum 
+in $v$ in the relation:
+$$\varphi(v) - s v =
+\varphi_1(\varphi_2(v)) - \ell \varphi_2(v) + \sum_{i=1}^\ell
+(\varphi_2(v) - n_i v)$$
+we get:
+\begin{equation}
+\label{eq:inegvarphi}
+\varphi^\star(s) \leq \varphi_1^\star(\ell) + 
+\varphi_2^\star(n_1) + \cdots + \varphi_2^\star(n_\ell).
+\end{equation}
+It remains then to prove that one can choose $\ell$ and $n_1, \ldots,
+n_\ell$ such that the inequality above is actually an equality.
+
+Let $x$ be the unique value such that the left (resp. right) derivative 
+of $\varphi$ at $x$ is less than (resp. greater than) or equal to $s$. 
+Let $\ell^-$ (resp. $\ell^+$) denote the left (resp. right) derivative 
+of $\varphi_1$ at $\varphi_2(x)$ and $n^-$ (resp. $n^+$) denote the left 
+(resp. right) derivative of $\varphi_2$ at $x$. On the one hand, $\ell^- 
+n^- \leq s \leq \ell^+ n^+$ since $\ell^- n^-$ (resp. $\ell^+ n^+$) 
+equals the left (resp. right) derivative of $\varphi$ at $x$). On the 
+other hand:
+$$\begin{array}{ll}
+\varphi_1^\star(u) = \varphi(x) - u \varphi_2(x) & 
+\text{for all } u \in [\ell^-,\, \ell^+] \\
+\varphi_2^\star(u) = \varphi_2(x) - u x & 
+\text{for all } u \in [n^-,\, n^+] \\
+\varphi^\star(u) = \varphi(x) - u x & 
+\text{for all } u \in [\ell^- n^-,\, \ell^+ n^+] \\
+\end{array}$$
+Therefore, the inequality \eqref{eq:inegvarphi} is an equality for all 
+integer $\ell \in [\ell^-, \ell^+]$ and all sequences of integers 
+$(n_i)_{1 \leq i \leq \ell}$ with $n_i \in [n^-, n^+]$ for all $i$ and 
+$n_1 + \cdots + n_\ell = s$. It is then enough to prove that such a 
+couple $(\ell, (n_i))$ exists but it is clear.
+\end{proof}
+
+\end{comment}
+
+\subsection{Composite of locally analytic functions}
+
+Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces
+$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$ and
+consider two locally analytic functions $f : U \to V$ and $g : V \to W$ 
+two locally analytic functions around $0$. The composition $h = g \circ
+f$ is then locally analytic around $0$ as well. Let us write the
+analytic expansion of $f$, $g$ and $h$ as follows:
+$$f = \sum_{n \geq 0} f_n, \quad 
+g = \sum_{n \geq 0} g_n, \quad
+h = \sum_{n \geq 0} h_n, \quad$$
+where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
+some $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The aim of
+this subsection is to prove the following proposition.
+
+\begin{prop}
+\label{prop:boundhr}
+With the above notations, we have for all nonnegative integer $r$:
+$$\Vert h_r \Vert \leq \sup_{m, (n_i)}
+  \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert$$
+where the supremum is taken over all couples $(m, (n_i))$ where $m$
+is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
+length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
+\end{prop}
+
+During the proof, multinomial coefficients will play an important role. 
+Given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
+k_1 + \cdots + k_\ell$, it is defined by:
+$$\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
+  \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.$$
+We recall that they are always positive integers.
+We now expand $g \circ f$ as follows:
+\begin{align}
+g \circ f & = \sum_{m \geq 0} g_m \Big(\sum_{n \geq 1} f_n\Big) \nonumber \\
+& = \sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
+G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell}, \ldots, f_{n_\ell})
+\label{eq:expansiongf}
+\end{align}
+where the latest sum runs over:
+
+\noindent
+(a) all finite sequences $(k_i)$ of positive integers whose length 
+(resp. sum) is denoted by $\ell$ (resp. $m$), and
+
+\noindent
+(b) all finite sequences $(n_i)$ of positive integers of length
+$\ell$.
+
+\noindent
+Moreover, in the argument of $G_m$, the variable $f_{n_i}$ 
+is repeated $k_i$ times.
+
+The degree of $G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell},
+\ldots, f_{n_\ell})$ is $r = k_1 n_1 + \ldots + k_\ell n_\ell$ and 
+then contributes to $h_r$. As a consequence $h_r$ is equal to 
+\eqref{eq:expansiongf} where the sum is restricted to sequences
+$(k_i)$, $(n_i)$ such that $k_1 n_1 + \ldots + k_\ell n_\ell = r$.
+Proposition \ref{prop:boundhr} now follows from the next lemma.
+
+\begin{lem}
+Let $\mathcal E$ be a $K$-vector space. Let $\varphi : \mathcal E^n \to 
+K$ be a symmetric $s$-linear form and $\psi: E \to K$ defined by 
+$\psi(x) = \varphi(x, x, \ldots, x)$.
+Given $k_1, \ldots, k_\ell$ some positive integers whose sum is $m$ and 
+$x_1, \ldots, x_\ell$ some elements of $\mathcal E$, we have:
+$$\begin{array}{l}
+\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
+\varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
+x_\ell) \Big\Vert  \\
+\hspace{4.3cm} \leq \Vert \psi \Vert \cdot \Vert x_1 \Vert^{k_1} \cdots
+ \Vert x_\ell \Vert^{k_\ell}
+\end{array}$$
+where, in LHS, the variable $x_i$ is repeated $k_i$ times.
+\end{lem}
+
+\begin{proof}
+It is enough to prove that:
+\begin{equation}
+\label{eq:normpolar}
+\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
+\varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
+x_\ell) \Big\Vert \leq \Vert \psi \Vert
+\end{equation}
+provided that all the $x_i$'s have norm at most $1$. We proceed by 
+induction on 
+$\ell$. If $\ell = 1$, Eq.~\eqref{eq:normpolar} follows directly from
+the definition of $\Vert \psi \Vert$. 
+We now pick $(\ell+1)$ integers $k_1, \ldots, k_{\ell+1}$ whose sum 
+equals $m$ together with $(\ell+1)$ elements $x_1, \ldots, x_{\ell+1}$
+lying in the unit ball of $\mathcal E$.
+We also consider a new variable $\lambda$ varying in
+$\O_K$. We set $x'_i = x_i$, $k'_i = k_i$ when $i < \ell$ and $x'_\ell 
+= x_\ell + \lambda x_{\ell+1}$, $k_\ell = k_\ell + k_{\ell+1}$. By the 
+induction hypothesis, we know that the inequality:
+$$\Big\Vert \binom s {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
+\varphi(x'_1, \ldots, x'_1, \ldots, x'_\ell, \ldots, x'_\ell) \Big\Vert
+\leq \Vert \psi \Vert$$
+holds for all $\lambda \in K$. Furthermore, LHS of the above inequality
+is a polynomial $P(\lambda)$ of degree $k'_\ell$ whose coefficient in 
+$\lambda^j$ is:
+$$\binom s {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
+\binom {k'_\ell} {j} \cdot
+\varphi(\underline x_j) = 
+\binom s {k_1 \,\, \cdots \,\, k_{\ell-1} \,\, j} \cdot
+\varphi(\underline x_j)$$
+with
+$$\underline x_j = (x_1, \ldots, x_1, \ldots, x_{\ell+1}, \ldots, 
+x_{\ell+1})$$
+where $x_i$ is repeated $k_i$ times if $i < \ell$ and $x_\ell$ 
+(resp. $x_{\ell+1}$) is repeated $j$ times (resp. $k'_\ell - j$ times).
+Since $\Vert P(\lambda) \Vert \leq \Vert \psi \Vert$ for all $\lambda$ in 
+the unit ball, the norm of all its coefficients must be at most $\Vert \psi
+\Vert$ as well. Looking at the coefficient in $\lambda^{k_i}$, we find
+Eq.~\eqref{eq:normpolar} for the families $(k_1, \ldots, k_{\ell+1})$ 
+and $(x_1, \ldots, x_{\ell+1})$. The induction follows.
+\end{proof}
+
+\subsection{Bounding a growing function}
+\label{ssec:boundLambdaf}
+
+We now study a slight generalization of the differential equation 
+\eqref{eq:diffequa}, which is:
+\begin{equation}
+\label{eq:diffequah}
+f' = g \circ (f, h).
+\end{equation}
+Here $g$ and $h$ are known locally analytic functions and $f$ is the 
+unknown, which is assumed to be locally analytic as well. The function
+$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
+and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
+is again an open subset in a $K$-Banach space $G$. Consequently, $g$
+goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
+all continuous linear applications $E \to F$.
+In what follows, we always assume that $V$ and $W$ contains the origin, 
+$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are armless 
+because (1)~we can always shift $f$ and $h$ (and $g$ accordingly) so 
+that they both vanish at $0$ and (2)~in order to apply the theory of 
+\cite{padicprec}, the derivative $f'(0)$ needs to be surjective and 
+therefore \emph{a fortiori} nonzero.
+
+Let us write the analytic expansion of $f$, $g$ and $h$ as follows: 
+$$f = \sum_{n \geq 0} f_n, \quad
+g = \sum_{n \geq 0} g_n, \quad
+h = \sum_{n \geq 0} h_n$$
+where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
+some $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
+Following \cite{padicprec}, we define $\NP(f) : \R \to \R \cup 
+\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
+\log \Vert f_n \Vert$ for all integer $n$ and denote by $\Lambda(f)$ its 
+Legendre transform. We define $\NP(g)$, $\Lambda(g)$, $\NP(h)$ and
+$\Lambda(h)$ similarly.
+
+We now assume that \textbf{$K$ has caracteristic $0$} and we consider a 
+nonnegative real number $\alpha$ such that $\Vert n! \Vert \geq 
+e^{-\alpha n}$ for all positive integer $n$. A suitable value for 
+$\alpha$ is
+$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$
+where $p$ denotes the characteristic of the residue field and
+where, by convention, the above expression equals $0$ if $p = 0$.
+
+\begin{prop}
+\label{prop:boundLambdaf}
+We keep the above notations. If the couple $(a,b)$ satisfies:
+\begin{equation}
+\label{eq:condAB}
+b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)
+\end{equation}
+then $b \geq \Lambda(f)(a - \alpha)$.
+\end{prop}
+
+\begin{proof}
+Assuming that $F_n$ is symmetric (which is armless in characteristic 
+$0$), we get $f' = \sum_{n \geq 0} f'_n$ where
+$$f'_n : U \to \mathcal L(E,F), \quad
+x \mapsto \big(h \mapsto n \cdot F_n(h, x, x, \ldots, x)\big).$$
+In particular, taking $h = x$, we find:
+\begin{equation}
+\label{eq:normderivative}
+\Vert f'_n \Vert \geq \Vert n f_n \Vert = 
+\Vert n \Vert \cdot \Vert f_n \Vert.
+\end{equation}
+Combining this with Proposition \ref{prop:boundhr}, we get, for all
+nonnegative integer $r$:
+$$\Vert (r+1) f_{r+1} \Vert \leq
+  \sup_{m, (n_i)} \Vert g_m \Vert \cdot 
+  \prod_{i=1}^m \max(\Vert f_{n_i} \Vert, \Vert h_{n_i} \Vert)$$
+where the supremum runs over all couples $(m, (n_i))$ where $m$
+is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
+length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
+We set $u_r = \Vert r! f_r \Vert$. Multiplying the above inequality by
+$\Vert r! \Vert$, we obtain:
+\begin{equation}
+\label{eq:boundurrec}
+u_{r+1} \leq
+  \sup_{m, (n_i)} \Vert g_m \Vert \cdot 
+  \prod_{i=1}^m \max(u_{n_i}, \Vert n_i! h_{n_i} \Vert)
+\end{equation}
+since the multinomial coefficient $\binom r {\!n_1 \, \cdots \, n_m\!}$
+is an integer and hence as norm at most $1$.
+
+We now pick two real numbers $A$ and $B$ satisfying \eqref{eq:condAB}.
+Set $d = \Lambda(h)(a)$. Going back to the definitions, we remark that
+the above equality implies:
+\begin{equation}
+\label{eq:boundhn}
+\Vert h_n \Vert \leq e^{- a n + d},
+  \quad \forall n \in \N.
+\end{equation}
+In the same way, from Eq.~\eqref{eq:condAB}, we get:
+\begin{equation}
+\label{eq:boundgm}
+\Vert g_m \Vert \leq e^{- \max(b,d)\cdot m + b - a},
+  \quad \forall m \in \N.
+\end{equation}
+We are now ready to prove $u_r \leq e^{-ar + b}$ by induction on $r$.
+When $r = 0$, it is obvious because $u_0$ vanishes. Otherwise, it
+follows from \eqref{eq:boundurrec}, \eqref{eq:boundhn}, \eqref{eq:boundgm}, 
+and the induction hypothesis that:
 \begin{align*}
-B_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert \le r_i \Big\}, \\
-B^-_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert < r_i \Big\}
+u_{r+1} 
+& \leq \sup_{m, (n_i)}
+    e^{ -\max(b,d)\cdot m + b - a + \sum_{i=1}^m (-a n_i + \max(b,d))} \\
+& = e^{ b - a - a r } = e^{ -a (r+1) + b}
 \end{align*}
-are lattices precisely when the $r_i$ are bounded.  If we have equipped $E$ with a distinguished basis
-then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.
-
-\subsubsection*{Approximate elements.}
-
-Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
-\begin{deftn}
-\label{def:approximate}
-\begin{itemize}
-\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
-\begin{equation} \label{eq:exact_elt}
-x = \sum_{j \in J} \alpha_j x_j.
-\end{equation}
-
-\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
-and $H$ is a lattice in $E$.
-\end{itemize}
-\end{deftn}
-
-The pair $(x, H)$ represents an undetermined element of the coset
-$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
-the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
-we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
-Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.
-
-\subsubsection*{Lattices and computers.}
-
-Suppose that $E \simeq K^d$ is finite dimensional.  Then if $H \subset E$
-is a lattice then there exist $a, b \in \Q_{>0}$ with
-\begin{equation}
-\label{eq:incllattice}
-B_K(a)^d \subset H \subset B_K(b)^d.
-\end{equation}
-Set $r = \frac a b$ and $R_r = \OK/B_K(r)$.  
-Then a lattice $H$ satisfying \eqref{eq:incllattice} is uniquely 
-determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
-R_r^d$.  Since $R \cap \OK$ is dense in $\OK$, elements of $R_r$ may be
-represented exactly.  Thus $H$ may be encoded as a 
-$(d \times d)$ matrix with coefficients in $R_r$.  For example, when
-$K = \Qp$ the ring $R_r$ is just $(\Z / p^n\Z)$ for $n = \lfloor - \log_p r \rfloor$.
-
-\subsection{Separating precision from approximation}
-\label{ssec:separation}
-
-Definition \ref{def:approximate} encapsulates the two main practical suggestions of this paper
-with regards to representing vector spaces, matrices, polynomials and power series over $K$:
-\begin{enumerate}
-\item one should \textbf{separate} the approximation from the precision,
-\item the appropriate object to represent precision is a \textbf{lattice}.
-\end{enumerate}
-In the rest of this section we discuss some of the benefits made possible these choices.
-
-Note first that using an arbitrary lattice to represent the precision of an approximate element
-can reduce precision loss when compared to storing the precision of each coefficient
-$\alpha_i$ in \eqref{eq:exact_elt} separately.  Recall the map 
-$f : (x,y) \mapsto (x+y, x-y)$ from the beginning of the section, and write $(e_1, e_2)$
-for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
-the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
-$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
-For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
-By using lattices one eliminates the loss of precision seen previously.
-We shall see in the next section that a similar phenomenon occurs for non-linear mappings as well.
-
-In addition to allowing for a more flexible representation of the precision of an element,
-the separation of precision from approximation has other benefits as well.  If the precision
-is encoded with the approximation, certain algorithms become unusable because of their
-numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication \cite{karatsuba-ofman:62a}
-can needlessly lose precision when operating on polynomials with inexact coefficients.
-However, it works perfectly well on exact approximations, leaving the question of the precision of
-the product to be solved separately.  By separating the precision, more algorithms become available.
-
-\section{Lattices and differentials}
-\label{sec:mainlemma}
-
-Our theory of $p$-adic precision rests upon a lemma in $p$-adic 
-analysis: Lemma \ref{lem:main}.  This section develops the theory 
-surrounding this result; we proceed to practical consequences in 
-Section \ref{sec:tracking}.
-
-\subsection{Images of lattices under differentiable functions}
-
-Our goal in this section is to relate the image of a lattice under a 
-differentiable map to its image under the derivative.
-
-\begin{deftn} \label{deftn : diff}
-Let $E$ and $F$ be two $K$-Banach spaces, let $U$ be an open 
-subset of $E$ and let $f : U \rightarrow F$ be a map. Then $f$ is called 
-\emph{differentiable} at $v_0 \in U$ if there exists a continuous linear 
-map $f'(v_0) : U \rightarrow W$ such that for any $\varepsilon >0$, 
-there exists an open neighborhood $U_\varepsilon \subset U$ containing 
-$v_0$ with
-\[ 
-\Vert f(v)-f(w)-f'(v_0) \left( v-w \right) \Vert 
-\leq \varepsilon \Vert v-w \Vert. 
-\]
-for all $v, w \in U_\varepsilon$.  The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
-\end{deftn}
+and the induction goes.
+Finally, keeping in mind the definition of $u_r$, from $u_r \leq 
+e^{-a r + b}$, we obtain:
+$$\Vert f_r \Vert \leq u_r \cdot \Vert r! \Vert^{-1} \leq
+e^{-(a - \alpha) r + b}$$
+which means that $b \geq \Lambda(f)(a - \alpha)$.
+\end{proof}
 
 \begin{rem}
-\label{rem:differentiability}
-This notion of differentiability is sometimes called \emph{strict differentiability};
-it implies that the function $x \mapsto f'(x)$ is continuous on $U$. 
+A simpler proof is possible if $\Z \subset \O_K$, \emph{i.e.} if
+one can choose $\alpha = 0$. Indeed, under this extra assumption,
+one can prove using Eq.~\eqref{eq:normderivative} that:
+$$\Lambda(f') \geq \Lambda(f) - \id.$$
+Combining this with Lemma 3.7 of \cite{padicprec}, we find:
+$$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
+from which we can easily derive Proposition \ref{prop:boundLambdaf}.
 \end{rem}
 
-%One can derive all usual properties of the derivative from this 
-%definition: unicity, chain rules, \emph{etc}.
-
-\begin{deftn}
-\label{def:firstorder}
-Let $E$ and $F$ be two $K$-Banach spaces, $f : U \rightarrow F$ be a 
-function defined on an open subset $U$ of $E$ and $v_0$ be a 
-point in $U$.
-A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
-$v_0$ if the following equality holds:
-\begin{equation}
-\label{eq:firstorder}
-f(v_0 + H) = f(v_0) + f'(v_0) (H).
-\end{equation}
-\end{deftn}
-
-We emphasize that we require an equality in \eqref{eq:firstorder}, 
-and not just an inclusion! 
-With this definition in hand, we are able to state our main lemma.
-
-\begin{lem} \label{lem:main}
-Let $E$ and $F$ be two $K$-Banach spaces and $f : U 
-\rightarrow F$ be a function defined on an open subset $U$ of $E$.
-We assume that $f$ is differentiable at some point $v_0 \in 
-U$ and that the differential $f'(v_0)$ is surjective.
-
-Then, for all $\rho \in (0, 1]$, there exists a positive real 
-number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ is a first
-order lattice for $f$ at $v_0$.
-\end{lem}
+We want now to rephrase Proposition \ref{prop:boundLambdaf} in a more 
+workable way. We assume that we are given two nondecreasing convex 
+functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
+\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We assume in addition that 
+there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
+$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
+we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
+does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
+defined by:
+$$\begin{array}{ll}
+\tau_\nu(x) = x & \text{if } x \leq \nu \\
+\hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise}
+\end{array}$$
+and:
+$$\Lambda_f(x) = 
+  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)$$
+where we recall that $\alpha$ satisfies $\Vert n! \Vert \geq e^{-\alpha 
+n}$ for all $n$.
+
+\begin{cor}
+\label{cor:boundLambdaf}
+We have $\Lambda(f) \leq \Lambda_f$.
+\end{cor}
 
 \begin{proof}
-Without loss of generality, $v_0=0$ and $f(0)=0$. Since $f'(0)$ is surjective, the open mapping theorem provides a $C>0$ such that 
-$B_F(1) \subset f'(0)(B_E(C))$.
-Let $\varepsilon>0$ be such that 
-$\varepsilon C < \rho$, and choose $U_\varepsilon \subset E$ as in Definition 
-\ref{deftn : diff}.  We may assume $U_\varepsilon = B_E(\delta)$ for some $\delta >0$.
-
-Let $r \in (0, \delta)$. We suppose that $H$ is a lattice with 
-$B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r).$
-We seek to show that $f$ maps $H$ surjectively onto $f'(0) (H)$. We first prove 
-that $f(H) \subset f'(0) (H)$. Suppose $x \in H$. By differentiability at 
-$0$, $\Vert f(x)-f'(0)(x) \Vert \leq \varepsilon \Vert x \Vert $. 
-Setting $y=f(x)-f'(0)(x)$, we have $\Vert y \Vert \leq 
-\varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r) 
-\subset f'(0) (B_E(\varepsilon rC))$. Thus there exists $x' 
-\in B_E(\varepsilon r C)$ such that $f'(0) (x') =y$. Since 
-$\varepsilon C < \rho$, we get $x' \in B^-_E(\rho r) \subset H$ and 
-then $f(x)= f'(0) (x-x') \in f'(0) (H)$.
-
-We now prove surjectivity. Let $y \in f'(0) (H)$. Let $x_0 \in H$ 
-be such that $y = f'(0) (x_0)$. We inductively define two sequences 
-$(x_n)$ and $(z_n)$ as follows:
-\begin{itemize}
-\item $z_n$ is an element of $E$ satisfying $f'(0)(z_n) = y - 
-f(x_n)$ and $\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$ (such 
-an element exists by definition of $C$), and
-\item $x_{n+1}=x_n+z_n$.
-\end{itemize}
-For convenience, let us also define $x_{-1} = 0$ and $z_{-1}=x_0.$ We claim that the 
-sequences $(x_n)$ and $(z_n)$ are well defined and take their values in 
-$H$. We do so by induction, assuming that $x_{n-1}$ and $x_n$ belong to $H$
-and showing that $z_n$ and $x_{n+1}$ do as well. Noticing
-that 
-\begin{equation}
-\label{eq:mainlemma}
-\begin{aligned}
-y - f(x_n) &= f(x_{n-1}) + f'(0)(z_{n-1}) - f(x_n) \\
-&= f(x_{n-1}) - f(x_n) - f'(0)(x_{n-1} - x_n)
-\end{aligned}
-\end{equation}
-we deduce using differentiability that
-$\Vert y - f(x_n) \Vert \leq \varepsilon \cdot \Vert x_n - x_{n-1}
-\Vert$.
-Since we are assuming that $x_{n-1}$ and $x_n$ lie in $H \subset
-B_E(r)$, we find $\Vert y - f(x_n) \Vert \leq \varepsilon r$. Thus
-$\Vert z_n \Vert \leq C \cdot \varepsilon r < \rho r$ and then
-$z_n \in H$. From the relation $x_{n+1} = x_n + z_n$, we finally
-deduce $x_{n+1} \in H$.
-
-Using \eqref{eq:mainlemma} and differentiability at $0$ once more,
-we get
-\[
-\Vert y  - f(x_n) \Vert 
-\leq \varepsilon \cdot \Vert z_{n-1} \Vert \leq \varepsilon C \cdot \Vert 
-y - f(x_{n-1}) \Vert,
-\]
-for all $n > 0$.  Therefore, $\Vert y - f(x_n) \Vert = O(a^n)$ and 
-$\Vert z_n \Vert = O(a^n)$ for $a = \varepsilon C < \rho \leq 1$. 
-These conditions show that $(x_n)$ is a Cauchy sequence, which converges since $E$ is complete.
-Write $x$ for the limit of the $x_n$; we have $x \in H$ because $H$ is closed.
-Moreover, $f$ is continuous on $H \subseteq U_\varepsilon$ since it is differentiable, and thus $y=f(x)$.
+Given $a \in \R$ and $b = a + \Lambda_g \circ \Lambda_h(a)$, we have to 
+prove that $\Lambda(f)(a-\alpha) \leq b$ provided that $b \leq \nu$. 
+Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to check 
+that such couples $(a,b)$ satisfy \eqref{eq:condAB}. Clearly:
+$$b \geq a + \Lambda(g) \circ \Lambda(h)(a)$$
+since $\Lambda_g \geq \Lambda(g)$, $\Lambda_h \geq \Lambda(h)$ and
+$\Lambda_g$ is nondecreasing. Furthermore, from $b \leq \nu$, we get:
+$$\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
+\Lambda_h(a).$$
+Thus $a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) =
+b$ and we are done.
 \end{proof}
 
-We end this section with a remark on the surjectivity of $f'(v_0)$ assumed in
-Lemma \ref{lem:main}. First, let us 
-emphasize that this hypothesis is definitely necessary. Indeed, the lemma would 
-otherwise imply that the image of $f$ is locally contained in a proper 
-sub-vector-space around each point where the differential of $f$ is not 
-surjective, which is certainly not true! Nevertheless, one can use Lemma \ref{lem:main}
-to prove a weaker result in the context that $f'(v_0)$ is not surjective.
-To do so, choose a closed sub-vector-space $W$ of $F$ such that $W + f'(v_0)(E) = 
-F$. Denoting by $\pr_W$ the canonical projection of $F$ onto 
-$F/W$, the composite $\pr_W \circ f$ is differentiable at 
-$v_0$ with surjective differential. For a given lattice $H$, there will be various choices of $W$ to which Lemma \ref{lem:main}
-applies.  For each such $W$,
-\begin{equation}
-\label{eq:notsurjective}
-f(v_0 + H) \subset f(v_0) + f'(v_0)(H) + W;
-\end{equation}
-taking the intersection of the right hand side
-over many $W$ yields an upper bound on $f(v_0 + H)$.
-
-\subsection{The case of locally analytic functions}
-\label{ssec:locanalytic}
-
-In this section we make the constant $\delta$ in Lemma \ref{lem:main} explicit,
-under the additional assumption that $f$ is locally analytic.
-We extend the definition of such functions from finite-dimensional
-$K$-vector spaces \cite {schneider:11a}*{\S 6} to $K$-Banach spaces.
-
-\begin{deftn}
-\label{def:locanal}
-Let $E$ and $F$ be $K$-Banach spaces. Let $U$ be an open subset of $E$
-and let $x \in U$.
-A function $f \col U \rightarrow F$ is said to be \emph{locally analytic} at 
-$x$ if there exists an open subset $U_x \subset E$ and continuous $n$-linear
-maps $L_n \col E^n \rightarrow F$ for $n \ge 1$ such that
-\[
-f(x+h) = f(x) + \sum_{n \geq 1} L_n(h,\ldots,h)
-\]
-for all $h$ with $x + h \in U_x$.
-\end{deftn}
-
-\begin{rmk}
-A function $f$ which is locally analytic at $x$ is \emph{a fortiori} 
-differentiable at $x$, with derivative given by $L_1$.
-\end{rmk}
-
-For the rest of this section, we assume that $K$ is algebraically closed. As in Definition 
-\ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
-family of continuous $n$-linear maps $L_n : E^n \to F$. For $n \ge 1$ 
-and $h \in E$, we set $f_n(h) = L_n(h, \ldots, h)$ and
-$$\Vert f_n \Vert = \sup_{h \in B_E(1)} \Vert f_n(h) \Vert.$$
-When the series $\sum_n f_n(h)$ converges, we denote by $f(h)$ its sum;
-we shall write $f = \sum_{n \geq 0} f_n$. We assume that $f$ is defined
-in a neighborhood of $0$. Under this assumption, the datum of $f$ 
-uniquely determines the $f_n$'s (a consequence of Proposition 
-\ref{prop:lambdaNP} below).
-To such a series $f$, we attach the function $\Lambda(f) : \R \cup \{+\infty\} \to \R 
-\cup \{+\infty\}$ defined by:
-\[
-\Lambda(f)(v) = \begin{cases} \log \left( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \right) & \mbox{if $f$ is defined on} B^-_E(e^v), \\
-+\infty & \mbox{otherwise.} \end{cases}
-\]
-The following lemma is easy and left to the reader.
-
-\begin{lem}
-\label{lem:opLambda}
-Let $f = \sum_{n \geq 0} f_n$ and $g = \sum_{n \geq 0} g_n$ be two 
-series as above. Then
-\begin{align*}
-\Lambda(f+g) &\leq \max (\Lambda(f),\Lambda(g)), \\
-\Lambda(f \times g) &\leq \Lambda(f)+\Lambda(g), \\
-\Lambda(f \circ g) &\leq \Lambda(f) \circ \Lambda(g).
-\end{align*}
-\end{lem}
-
-\begin{rem}
-\label{rem:opLambda}
-Using Lemma \ref{lem:opLambda}, one can easily derive an upper 
-bound of $\Lambda(f)$ from a formula describing $f$.
-\end{rem}
-
-The function $\Lambda(f)$ we have just defined is closely 
-related to the Newton polygon of $f$. Recall that the 
-Newton polygon of $f$ is the convex hull in $\R^2$ of the points
-$(n, - \log \lVert f_n \rVert )$ for $n \geq 0$, together 
-with the extra point $(0, +\infty)$. We denote by $\NP(f) : \R \to \R 
-\cup \{+\infty\}$ the convex function whose epigraph is the Newton 
-polygon of $f$.
-
-We recall that the Legendre transform of a \emph{convex} function 
-$\varphi : \R \to \R \cup \{+\infty\}$ is the function $\varphi^\star : 
-\R \to \R \cup \{+\infty\}$ defined by
-\[
-\textstyle \varphi^\star(v) = \sup_{u \in \R} \:
-\big(uv - \varphi(u)\big),
-\]
-for $v \in \R$.
-One can check that the map $\varphi \mapsto \varphi^\star$ is an
-order-reversing involution: $(\varphi^\star)^\star = \varphi$ and
-$\varphi^\star \geq \psi^\star$ whenever $\varphi \leq \psi$.
-We refer to \cite{Rockafellar:97} for a complete exposition on Legendre transforms.
-
-\begin{prop} \label{prop:lambdaNP}
-Keeping the above notation, we have $\Lambda (f) = \NP(f)^\star$.
+\begin{figure}
+\null \hfill
+\begin{tikzpicture}[xscale=0.8,yscale=0.8]
+\draw[thick,green!30,fill=green!20] 
+   (-2,-1)--(-1,2)--(-1,3)--(3.5,3)
+ --(2.5,2)--(-3.5,-4)--cycle;
+\draw[->] (-5,0)--(3.5,0);
+\draw[->] (0,-5)--(0,3);
+\node[below, scale=0.75] at (3.5,0) { $x$ };
+\node[right, scale=0.75] at (0,3) { $y$ };
+\draw[black!80,dotted] (-1,2)--(0,2);
+\node[right,scale=0.75] at (0,2) { $\nu$ };
+\draw[black!80,dotted] (-2,0)--(-2,-1)--(0,-1);
+\node[above,scale=0.75] at (-2.4,0) { $\Lambda_h^{-1}(\nu){-}\alpha$ };
+\node[right,scale=0.75] at (0,-1) { $\Lambda_h^{-1}(\nu){+}\mu$ };
+\draw[Blue,very thick] (-5,-4)--(-2,-1)--(-1,2);
+\draw[Sepia, very thick] (-4.5,-5)--(3.5,3);
+\draw[Blue, very thick,dashed] (-2,-1)--(-3.5,-4);
+\node[below right,rotate=45,scale=0.9] at (-4.3,-4.8) 
+  { $y = x + \log \Vert g(0) \Vert$ };
+\node[above right,rotate=45,scale=0.9] at (-4.8,-3.8) 
+  { $y = x + \alpha + \mu$ };
+\end{tikzpicture}
+\hfill \null
+
+\caption{Admissible region for the graph of $\Lambda(f)$}
+\label{fig:area}
+\end{figure}
+
+Figure \ref{fig:area} illustrates Corollary \ref{cor:boundLambdaf}. The 
+blue plain line represents the graph of the function $\Lambda_f$. A 
+quick computation shows that, on a neighborhood of ${-}\infty$, this 
+function is given by:
+$$\Lambda_f(x) = x + \alpha + \mu$$
+where $\mu$ is the value that $\Lambda_g$ takes on the interval 
+$]{-}\infty, \nu]$. Corollary \ref{cor:boundLambdaf} says that the
+graph of $\Lambda(f)$ lies below the plain blue line. We remark
+moreover that the Taylor expansion of $f(x)$ starts with the term
+$g(0)(x)$. Hence, on a neighborhood on ${-}\infty$, we have 
+$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
+get:
+$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
+  \quad \forall x \in \R.$$
+In other words, the graph of $\Lambda(f)$ lies above the brown line.
+Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
+because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
+the dashed blue line defined as the line of slope $2$ passing through
+the first break point of the blue plain line --- which has coordinate 
+$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
+\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
+must coincide with the brown line until it meets the dashed blue line 
+and then has to stay in the green area.
+
+As a consequence of the above discussion, we derive the following
+proposition which can be used directly in combination with Proposition
+3.12 of \cite{padicprec}.
+
+\begin{prop}
+\label{prop:boundLambdaf2}
+Keeping the above notations, we have:
+$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) + 
+\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
+for all $x < \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
 \end{prop}
@@
 \begin{proof}
-Note that the functions $\Lambda(f)$ and $\NP(f)^\star$ 
-are both left continuous. It is then enough to prove that they
-coincide expect possibly on the set of slopes of $\NP(f)$,
-a dense subset of $\R$.
+Just remark that $y = 2(x + \alpha + \mu) + y_0$ is a equation of the 
+dashed blue line.
+\end{proof}
+
+\noindent
+\textbf{About the choice of the function $h$}
 
@@
 
-Let $v \in \R$, not a slope of $\NP(f)$. We assume first that 
-$\NP(f)^\star(v)$ is finite. We set $u = \NP(f)^*(v)$. The 
-function $m \mapsto \NP(f)(m) - vm + u$ has the following properties:
-\begin{enumerate}
-\item it is piecewise affine and everywhere nonnegative,
-\item it does not admit $0$ as a slope and
-\item it vanishes at $x = n$ for some integer $n$ and $u = vn + \log \Vert f_n \Vert$.
-\end{enumerate}
-We deduce from these facts that there exists $c>0$ such that 
-\[
-vm-u \leq - \log \lVert f_m \rVert - c \cdot \vert n-m \vert
-\]
-for any $m\geq 0$.
-Since $vm-u=vm-vn-\log \lVert f_n \rVert$, we get 
-\[
--vn-\log \lVert f_n \rVert + c \cdot \vert n-m \vert \leq -vm-\log \lVert f_m \rVert.
-\]
-Therefore, for any $x \in  B_E(e^v)$ and $m \geq 0,$ we have
-\[
-\Vert f_m (x) \Vert \leq e^{-c \cdot \vert n-m \vert } \cdot
-\lVert f_n \rVert \cdot e^{vn} \le \lVert f_n \rVert \cdot e^{vn}.
-\]
-Thus, the series $\sum_{m \geq 0} f_m(x)$ converges and $\Vert f(x) 
-\Vert \leq \Vert f_n \Vert \cdot e^{vn}$. We then get
-\begin{equation}
-\label{eq:lowerLambda}
-\Lambda (f)(v) \leq \log \big( \Vert f_n \Vert e^{vn} \big)
-= vn + \log \Vert f_n \Vert = u.
-\end{equation}
-On the other hand, it follows from the definition of $\lVert f_n \rVert$ 
-and the fact that $\vert K^\times \vert$ is dense in $\R$ ($K$ is algebraically closed)
-that there exists a sequence $(x_i)_{i\geq 0}$ in $B^-_E(e^v)$ such that 
-$\lim_{i \to \infty} \Vert f_n (x_i) \Vert= \lVert f_n \rVert \cdot e^{vn}$. Since 
- $\Vert f_m (x_i) \Vert \leq e^{-c \cdot \vert n-m \vert} \cdot
-\Vert f_n \Vert \cdot e^{vn}$ for all $m$ and $i$, we get $\Vert f_m (x_i) \Vert 
-< \Vert f_n (x_i) \Vert$ for $i$ large enough. For these $i$, we then 
-have $\Vert f(x_i) \Vert = \Vert f_n (x_i) \Vert$. Passing to the limit 
-on $i$, we find $\Lambda (f)(v) \geq u$. Comparing with 
-\eqref{eq:lowerLambda}, we get $\Lambda (f)(v) = u = 
-\NP(f)^\star(v)$.
-
-\smallskip
-
-We now assume that $\NP(f)^\star(v) = +\infty$. The function $x \mapsto 
-\NP(f)(x) - vx$ is then not bounded from below. Since it is convex, it 
-goes to $-\infty$ when $x$ goes to $+\infty$. By the
-definition of $\NP(f)$, the expression $v n + \log \Vert f_n \Vert$
-goes to infinity as $n$ grows. It is then enough to establish the
-following claim:
-\begin{equation}
-\label{eq:claim}
-\forall n \in \N, \quad 
-\Lambda(f)(v) \geq v n + \log \Vert f_n \Vert - \log 2.
-\end{equation}
-Let $n$ be a fixed integer. If $\Vert f_n 
-\Vert = 0$, there is nothing to prove. Otherwise, we consider an
-element $x_n \in B^-_E(e^v)$ such that $\Vert f_n(x_n) \Vert \geq \frac 
-1 2 \Vert f_n \Vert \cdot e^{vn}$. If the series $\sum_{m \geq 0} 
-f_m(x_n)$ diverges, then $\Lambda(f)(v) = +\infty$ by definition and 
-Eq.~\eqref{eq:claim} holds. On the other hand, if it converges, the
-sequence $\Vert f_m(x_n) \Vert$ goes to $0$ as $m$ goes to infinity.
-Hence it takes its maximum value $R$ a finite number of times; let us 
-denote by $I \subset \N$ the set of the corresponding indices. For any
-$\lambda \in \OK$, the series defining $f(\lambda x_n)$ converges 
-and
-$$f(\lambda x_n) \in B_F(R)
-\quad \text{and} \quad
-f(\lambda x_n) \equiv \sum_{m \in I} \lambda^m f_m(x_n)
-\pmod {B^-_F(R)}.$$
-The quotient $B_F(R) / B^-_F(R)$ is a vector space over the residue
-field $k$ of $K$.  Since $k$ is infinite, there must exist $\lambda \in 
-\OK$ such that $\sum_{m \in I} \lambda^m f_m(x_n)$ does not vanish
-in $B_F(R) / B^-_F(R)$. For such an element $\lambda$, we have $\Vert 
-f(\lambda x_n) \Vert = R \geq \frac 1 2 \Vert f_n \Vert \cdot e^{vn}$. 
-The claim \eqref{eq:claim} follows.
-\end{proof}
-
-\begin{rem}
-It follows from Proposition \ref{prop:lambdaNP} that $\Lambda(f)$ is a
-convex function.
-\end{rem}
-
-We now study the effect of truncation on series: given $f$ as above
-and a nonnegative integer $n_0$, we set
-\[
-f_{\geq n_0} = \sum_{n \geq n_0} f_n = f - (f_0 + f_1 + \cdots +
-f_{n_0 - 1}).
-\]
-On the other hand, given a convex function $\varphi : \R \to \R \cup
-\{+\infty\}$ and a real number $v$, we define $\varphi_{\geq v} : \R \to 
-\R \cup \{\pm \infty\}$ by
-\[
-\varphi_{\geq v}(x) = \inf_{y \geq 0} \, \big(\varphi(x + y) - v y \big).
-\]
-The function $\varphi_{\geq v}$ is the maximum among all functions $\varphi'$
-with $\varphi' \leq \varphi$ and $x \mapsto \varphi'(x) - v x$ nondecreasing.
-When $v$ is fixed, the construction $\varphi \mapsto \varphi_{\geq v}$
-is nondecreasing: if $\varphi$ and $\psi$ are two convex functions such
-that $\varphi \leq \psi$ then $\varphi_{\geq v} \leq \psi_{\geq v}$.
-
-\begin{prop} \label{prop:trunc}
-With the above notations, we have $\Lambda(f_{\geq n_0}) \leq
-\Lambda(f)_{\geq n_0}$ for all $n_0 \in \mathbb N$.
-\end{prop}
-
-\begin{proof}
-It follows easily from Proposition \ref{prop:lambdaNP} and the fact that 
-the slopes of the Legendre transform of a convex piecewise affine 
-function $f$ are exactly the abscissae of the points where $f$ is not 
-differentiable.
-\end{proof}
-
-We may now provide two sufficient conditions to effectively recognize 
-first order lattices.
-
-\begin{prop}
-\label{prop:locanalytic}
-Let $f = \sum_{n \geq 0} f_n$ be a function as above.
-Let $C$ be a positive real number satisfying 
-$B_F(1) \subset f_1(B_E(C))$.
-Let $\rho \in (0, 1]$ and $\nu$ be a real number such that 
-\begin{equation}
-\label{eq:locanalytic}
-\Lambda(f)_{\geq 2} (\nu) < \nu + \log \Big( \frac \rho C \Big).
-\end{equation}
-Then the conclusion of Lemma \ref{lem:main} holds with $\delta = e^\nu$.
-\end{prop}
-
-\begin{rem}
-On a neighborhood of $-\infty$, the function $x \mapsto \Lambda(f)_{\geq 
-2} (x) - x$ is affine with slope $1$. This implies that, for all $\rho 
-\in (0,1]$, there exists $\nu$ satisfying \eqref{eq:locanalytic}. Moreover,
-if $\rho$ is close enough to $0$, then one can take $\delta = e^\nu$ as a 
-linear function of $\rho$.
-\end{rem}
-
-\begin{rem}
-In the statement of Proposition \ref{prop:locanalytic}, one can of 
-course replace the function $\Lambda(f)$ by any convex function 
-$\varphi$ with $\varphi \geq \Lambda(f)$. If $f$ is given by some
-formula or some algorithm, such a function $\varphi$ can be obtained
-using Remark \ref{rem:opLambda}.
-\end{rem}
-
-\begin{proof}
-Pick $\varepsilon$ in the interval $(e^{\Lambda(f)_{\geq 2} (\nu) - \nu}, 
-\frac \rho C)$. Going back to the proof of Lemma \ref{lem:main}, we 
-observe that it is enough to prove that
-\begin{equation}
-\label{eq:boundqueue} 
-\Vert f_{\geq 2}(x) \Vert \leq \varepsilon \cdot \Vert x \Vert.
-\end{equation}
-for all $x \in B_E(\delta)$.
-This inequality follows from Propositions \ref{prop:lambdaNP} and
-\ref{prop:trunc} applied to the function $x \mapsto \frac {\Lambda_{\geq 
-2}(x)} x$.
-\end{proof}
-
-\begin{rem}
-It follows from the proof that Proposition \ref{prop:locanalytic} is 
-still valid if $K$ is not assumed to be algebraically closed. Indeed, 
-the functions $f_n$ --- and then $f$ also --- extend to an algebraic
-closure $\bar K$ of $K$ and \eqref{eq:boundqueue} holds over $\bar 
-K$, which is enough to conclude the result.
-\end{rem}
-
-\begin{cor}
-We keep the notations of Proposition \ref{prop:locanalytic} and consider 
-in addition a sequence $(M_n)_{n \geq 2}$ such that $\Vert f_n \Vert 
-\leq M_n$ for all $n \geq 2$. Let $\NP(M_n)$ denote the convex function
-whose epigraph is the convex hull in $\R^2$ of the points of coordinates
-$(n, -\log M_n)$ for $n \geq 2$ together with the extra point $(0,
-+\infty)$.
-
-Let $\rho \in (0, 1]$ and $\nu$ be a real number such that 
-$$NP(M_n)^\star (\nu) < \nu + \log \Big( \frac \rho C \Big).$$
-Then the conclusion of Lemma \ref{lem:main} holds with $\delta = e^\nu$.
-\end{cor}
-
-\begin{rem}
-If $K$ has characteristic $0$ and the vector spaces $E$ and $F$ are
-finite dimensional, then the $M_n$'s defined by
-$$M_n = \frac 1 {|n!|} \cdot 
-\sup_{\substack{1 \leq i \leq \dim E \\ \vert \underline n
-\vert = n}} \,
-\Big\Vert \frac{\partial^n f_i}{\partial x^{\underline n}}(0) 
-\Big\Vert$$
-do the job. Here $f_i$ denotes the $i$-th coordinate of $f$, the
-notation $\underline n$ refers to a tuple of $(\dim F)$ nonnegative
-integers and $\vert \underline n \vert$ is the sum of the coordinates
-of $\underline n$.
-\end{rem}
-
-\section{Precision in practice}
-\label{sec:tracking}
-
-In this section we discuss applications of Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic}
-to effective computations with $p$-adic numbers and power series.
-
-\subsection{Optimal precision tracking}
-\label{ssec:opt-tracking}
-
-We consider a function {\tt f} (in the sense of computer 
-science) that takes as input an approximate element lying in an open 
-subset $U$ of a $K$-Banach space $E$ and outputs another approximate
-element lying in an open subset $V$ of another $K$-Banach space $F$.
-In applications, this function models a continuous
-mathematical function $f : U \to V$: when 
-{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
-$f(x+H) \subseteq x' + H'$. We say that {\tt f} \emph{preserves precision}
-if the above inclusion is an equality; it is 
-often not the case as shown in Section \ref{ssec:stepbystep}.
-
-Let us assume now that $f$ is locally analytic on $U$ and
-that $f'(x)$ is surjective. Proposition \ref{prop:locanalytic} then 
-yields a rather simple sufficient condition to decide if a given lattice 
-$H$ is a first order lattice for $f$ at $x$. For such a lattice, by 
-definition, we have $f(x+H) = f(x) + f'(x)(H)$ and thus {\tt f}
-must output $O(f'(x)(H))$ if it preserves precision. In this section we
-explain how, under the above hypothesis, one can implement the function 
-{\tt f} so that it always outputs the optimal precision.
-
-\subsubsection*{One-pass computation.}
-
-The execution of the function {\tt f} yields a factorization:
-$$f = f_n \circ f_{n-1} \circ \cdots \circ f_1$$
-where the $f_i$'s correspond to each individual basic step (like 
-addition, multiplication or creation of variables); they are then
-``nice'' (in particular locally analytic) functions. For all 
-$i$, let $U_i$ denote the codomain of $f_i$. Of course $U_i$ must
-contains all possible values of all variables which are defined in the 
-program after the execution of the $i$-th step. Mathematically, we assume
-that it is an 
-open subset in some $K$-Banach space $E_i$. We have $U_n = V$ and the 
-domain of $f_i$ is $U_{i-1}$ where, by convention, we have set $U_0 = 
-U$.
-For all $i$, we set $g_i = f_i \circ \cdots \circ f_1$ and $x_i = 
-g_i(x)$.
-
-When we execute the function {\tt f} on the input $x + O(H)$, we apply 
-first $f_1$ to this input obtaining this way a first result $x_1 + 
-O(H_1)$ and then go on with $f_2, \ldots, f_n$. At each step, we obtain 
-a new intermediate result that we denote by $x_i + O(H_i)$. A way to 
-guarantee that precision is preserved is then to 
-ensure $H_i = f'_i(x)(H_{i-1}) = g_i'(x)(H)$ at each step. This can be 
-achieved by reimplementing all primitives (addition, multiplication, 
-\emph{etc.}) and make them compute at the same time the function $f_i$
-they implement together with its differential and apply the latter to
-the ``current'' lattice $H_i$.
-
-There is nevertheless an important issue with this approach: in order to 
-be sure that Lemma \ref{lem:main} applies, we need \emph{a priori} to 
-compute the exact values of all $x_i$'s, which is of course not 
-possible! Assuming that $g'_i(x)$ is surjective for all $i$, we can fix 
-it as follows. For each $i$, we fix a first order lattice $\tilde H_i$ 
-for $g_i$ at $x$. Under our assumption, such lattices always exist and 
-can be computed dynamically using Proposition \ref{prop:locanalytic}
-and Lemma \ref{lem:opLambda} (see also Remark \ref{rem:opLambda}).
-Now, the equality $g_i(x + \tilde H_i) = x_i + g'_i(x)(\tilde H_i)$ 
-means that any perturbation of $x_i$ by an element in $g'_i(x) (\tilde 
-H_i)$ is induced by a perturbation of $x$ by an element in $\tilde H_i 
-\subset H$. Hence, we can freely compute $x_i$ modulo $g'_i(x) (\tilde 
-H_i)$ without changing the final result. Since $g'_i(x)(\tilde H_i)$ is 
-a lattice in $E_i$, this remark makes possible the computation of $x_i$.
-
-\begin{rmk}
-In some cases, it is actually possible to determine suitable lattices 
-$\tilde H_i$ together with their images under $g'_i(x)$ (or, at
-least, good approximations of them) before starting the computation 
-by using mathematical arguments. If possible, this generally helps a
-lot. We shall present in \S \ref{ssec:SOMOS-solution} an example of 
-this.
-\end{rmk}
-
-\subsubsection*{Two-pass computation.}
-
-The previous approach works only if the $g'_i(x)$'s are all surjective. 
-Unfortunately, this assumption is in general not fulfilled. Indeed, 
-remember that the dimension of $E_i$ is roughly the number of used 
-variables after the step $i$. If all $g_i'(x)$ were surjective, this 
-would mean that the function {\tt f} never initializes a new variable!
-In what follows, we propose another solution that does not assume the
-surjectivity of $g'_i(x)$.
-
-For $i \in \{1, \ldots, n\}$, define $h_i = f_n \circ \cdots \circ 
-f_{i+1}$, so that we have $f = h_i \circ g_i$. On differentials, we 
-have $f'(x) = h_i'(x_i) \circ g'_i(x)$. Since $f'(x)$ is surjective (by 
-assumption), we deduce that $h'_i(x_i)$ is surjective for all $i$. Let $H'_i$ 
-be a lattice in $E_i$ such that:
-\begin{enumerate}[(a)] 
-\item \label{item:Hi1}
-$H'_i$ is contained in $H_i + \ker h'_i(x_i) = h'_i(x_i)^{-1}
-\big(f'(x)(H)\big)$;
-\item \label{item:Hi2}
-$H'_i$ is a first order lattice for $h_i$ at $x_i$.
-\end{enumerate}
-By definition, we have
-$h_i(x_i + H'_i) = x_n + h'_i(x_i)(H'_i) \subset x_n + f'(x)(H)$.
-Therefore, modifying the intermediate value $x_i$ by an element
-of $H'_i$ after the $i$-th step of the execution of {\tt f} leaves
-the final result unchanged. In other words, it is
-enough to compute $x_i$ modulo $H'_i$.
-
-It is nevertheless not obvious to implement these ideas in practice
-because when we enter in the $i$-th step of the execution of {\tt f},
-we have not computed $h_i$ yet and hence are \emph{a priori} not able
-to determine a lattice $H'_i$ satisfying the axioms \eqref{item:Hi1}
-and \eqref{item:Hi2} above.
-A possible solution to tackle this problem is to proceed in several
-stages as follows: 
-\begin{enumerate}[(1)]
-\item \label{item:smallprec}
-for $i$ from $1$ to $n$, we compute $x_i$, $f'_i(x_{i-1})$ 
-at small precision (but enough for the second step) together with an
-upper bound of the function $\Lambda(h \mapsto f_i(x_{i-1}+h) - 
-f_i(x_{i-1}))$;
-\item \label{item:determineHi}
-for $i$ from $n$ to $1$, we compute $h'_i(x_i)$ and
-determine a lattice $H'_i$ satisfying \eqref{item:Hi1} and 
-\eqref{item:Hi2};
-\item \label{item:finalcomp}
-for $i$ from $1$ to $n$, we recompute $x_i$ modulo $H'_i$
-and finally outputs $x_n + O\big(f'(x)(H)\big)$.
-\end{enumerate}
-Using relaxed algorithms for computing with elements in $K$ (\emph{cf} 
-\cites{hoeven:02a, hoeven:07a, berthomieu-hoeven-lecerf:11a}), we can reuse in 
-Step~\eqref{item:finalcomp} the computations already performed in 
-Step~\eqref{item:smallprec}. The two-pass method we have just 
-presented is then probably not much more expensive than the one-pass 
-method, although it is more difficult to implement.
-
-We conclude this section by remarking that the two-pass method seems 
-to be particularly well suited to computations with lazy $p$-adics.
-In this setting, a target precision is fixed and the software determines automatically the precision 
-it needs on the input to achieve this output precision. To do this, it first 
-builds the ``skeleton'' of the computation (\emph{i.e.} it determines the 
-functions $f_i$ and eventually computes the $x_i$ at small precision when 
-branching points occur and it needs to decide which branch it follows) 
-and then runs over this skeleton in the reverse direction in order to 
-determine (an upper bound of) the needed precision at each step.
-
-\subsubsection*{Non-surjectivity.}
-
-From the beginning, we have assumed that $f'(x)$ is surjective. Let us 
-discuss shortly what happens when this assumption is relaxed. As it is 
-explained after the proof of Lemma \ref{lem:main}, the first thing we 
-can do is to project the result onto different quotients, \emph{i.e.} to 
-work with the composites $\pr_W \circ f$ for a sufficiently large family 
-of closed sub-vector-spaces $W \subset F$ such that $W + f'(x)(E) = F$. 
-If $F$ has a natural system of coordinates, we may generally take the 
-$\pr_W$'s as the projections on each coordinate. Doing this, we end up 
-with a precision on each individual coordinate. Furthermore, we have the 
-guarantee that each coordinate-wise precision is sharp, even if the lattice built
-from them is not.
-
-Let us illustrate the above discussion by an example: suppose that we 
-want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
-family of $n$ vectors to its Gram matrix. The differential of $f$ is
-clearly never surjective because $f$ takes its values in the subspace
-consisting of symmetric matrices. Nevertheless, for all pairs $(i,j)
-\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
-\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ takes a matrix 
-$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
-their differentials are generically surjective. Let $M$ be a matrix known
-at some finite precision such that 
-$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply a one- or two-pass 
-computation and get $f_{ij}(M)$ together with its 
-precision. Putting this together, we get the whole matrix $f(M)$ 
-together with a sharp precision datum on each entry.
-
-The study of this example actually suggests another solution to 
-tackle the issue of non-surjectivity. Indeed, we remark that our $f$ above 
-did not have a surjective differential simply because its codomain was too 
-large: replacing the codomain with the $K$-vector space $S_n(K)$
-of $n \times n$ symmetric matrices over $K$ makes the differential of $f$ surjective.
-While the image of a general $f$ is rarely a sub vector 
-space of $F$, it is often a sub-$K$-manifold of $F$. We can then use 
-the results of Appendix \ref{sec:manifold} to study $f : U \to f(U)$, which
-likely has surjective differential.
-
-\medskip
-
-\subsubsection*{Quick comparison with floating point arithmetic.}
-
-The two strategies described above share some similarities with standard 
-floating point arithmetic over the reals. In each setting, we begin
-by choosing a large precision for all computations, and when we
-encounter an unconstrained digit we choose it ``at random'' or using good heuristics. 
-However, in the ultrametric setting, mild hypotheses allow us
-to quantify the precision needed at each 
-individual step in order to ensure a specified final precision.
-
-\subsection{Precision Types}
-\label{ssec:types}
-
-Using an arbitrary lattice to record the precision of an approximate 
-element has the benefit of allowing computations to proceed without 
-unnecessary precision loss using Lemma \ref{lem:main}.  However, while 
-recording a lattice exactly is possible it does require a lot of space.  
-For example, the space required to store a lattice precision for a 
-single $n \times n$ matrix with entries of size $O(p^N)$ is $O(Nn^4 
-\cdot \log p)$.  Conversely, the space needed to record that every entry 
-has precision $O(p^N)$ is just $O(\log N)$.
-
-\begin{deftn}
-Suppose that $E$ is a $K$-Banach space, and write $\Lat(E)$ for the set of lattices in $E$.
-A \emph{precision type} for a $K$-Banach space $E$ is a set $\T \subseteq \Lat(E)$ together
-with a function $\round : \Lat(E) \to \T$ such that
-\begin{itemize}
-\item[$(\ast)$] For every lattice $H \in \Lat(E)$, the lattice $\round(H)$ is a least upper bound for $H$ under the inclusion order:
-$H \subseteq \round(H)$ and if $T \in \T$ satisfies $T \subset \round(H)$ then $H \not\subseteq T$.
-\end{itemize}
-\end{deftn}
-
-Different precision types are appropriate for different problems.  For example,
-the final step of Kedlaya's algorithm for computing zeta functions of hyperelliptic
-curves \cite{kedlaya:01a}*{\S 4: Step 3} involves taking the characteristic polynomial of the matrix of Frobenius acting
-on a $p$-adic cohomology space.  Obtaining extra precision on the entries of the
-matrix requires a long computation, so it is advisable to work with a precision type
-that does not round too much.
-
-The following list gives examples of useful precision types.  A description of the $\round$ function has been omitted for brevity.
-\begin{itemize}
-\item The \emph{lattice} precision type has $\T = \Lat(E)$.
-\item In the \emph{jagged} precision type, $\T$ consists of lattices of the shape $B_E((e_i),(r_i))$ for a fixed Banach basis $(e_i)$ of $E$.
-\item In the \emph{flat} precision type, $\T$ consists of lattices $B_E(r)$.  The flat precision type is useful since it takes so little space to store and it easy to compute with.
-\item If $E = K_{<d}[X]$ is the space of polynomials of degree less than $d$, the \emph{Newton} precision type consists of lattices $B_E((X^i),(r_i))$ where $-\log r_i$ is a convex function of $i$.
-The Newton precision type is sensible if one thinks of polynomials as functions $K \to K$, since extra precision above the Newton polygon never increases the precision of an evaluation.
-\item If $E = M_{m \times n}(K)$, the \emph{column} precision type consists of lattices with identical image under all projections $\pr_i : E \to K^m$ sending a matrix to its $i$th column.
-It is appropriate when considering linear maps where the image of each basis vector has the same lattice precision.
-\item If $E = \Qp \lb X \rb$, the \emph{Pollack-Stevens} precision type consists of lattices $H_N := B_E((X^i),(p^{\min(i-N,0)}))$ \cite{pollack-stevens:11a}*{\S 1.5}.
-It is important when working with overconvergent modular symbols since these lattices are stable under certain Hecke operators.
-\end{itemize}
-
-Note that sometimes the precision of a final result can be computed
-a priori (using the methods of Appendix \ref{sec:differentials} for example).
-Taking advantage of such knowledge can minimize artificial precision loss
-even when using rougher precision types such as flat or jagged.
-Separating precision from approximation also makes
-it much easier to implement algorithms capable of processing different precision types,
-since one can implement the arithmetic of the approximation separately from the logic
-handling the precision.
-
-\subsection{Application to SOMOS sequence}
-\label{ssec:SOMOS-solution}
-
-We illustrate the theory developed above by giving a simple 
-toy application. Other applications will be discussed in subsequent 
-articles. More precisely, we study the SOMOS 4 sequence introduced in \S 
-\ref{ssec:stepbystep}. Making a crucial use of Lemma \ref{lem:main} and 
-Proposition \ref{prop:locanalytic}, we design a \emph{stable} algorithm 
-for computing it.
-
-Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
-four-term inductive sequence defined by $u_n = \frac{u_{n-3} u_{n-1} 
-+ u_{n-2}^2}{u_{n-4}}$ exhibiting the Laurent phenomenon.
-We will focus on SOMOS sequences with values in $\Q_p$, and 
-assume for simplicity that $u_0, u_1, u_2, u_3 \in \Z_p^\times$.
-By the Laurent phenomenon, $u_n \in \Z_p$ for all $n$, and if $u_0, u_1, u_2, u_3$ 
-are known with finite precision $O(p^N)$ then all $u_n$ are 
-known with the same absolute precision. Algorithm \ref{algo:SOMOS} presented on 
-page \pageref{algo:SOMOS} performs this computation.
-
-\begin{algorithm}[t]
-\SetKwInOut{Assumption}{Assumption}
-\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
-\KwIn{$n, N$ --- two integers}
-\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
-precision $O(p^N)$}
-\Assumption{None of the $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
-\KwOut{$u_n$ at precision $O(p^N)$}
-\BlankLine
-$\prec$ $\leftarrow$ $N$\;
-\For {$i$ from $1$ to $n-3$}{
-  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2)$\;
-  {\bf lift} 
-    $b$, $c$ and $d$ arbitrarily to precision $O(p^{\prec})$\label{line:lift}\;
-  $\prec$ $\leftarrow$ $\prec - 2 \: v_p(a)$\;
-  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
-   \tcp{$e$ is known at precision $O(p^\prec)$}
-  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
-}
-\Return{$d + O(p^N)$}\;
-\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:SOMOS}
-\end{algorithm}
-
-We now prove that it is correct.
-We introduce the function $f : \Qp^\times \times \Q_p^3 \to \Q_p^4$ defined by 
-$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$. For all $i$, we have 
-$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
-= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is  
-differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
-canonical basis is given by the matrix:
-$$D(a,b,c,d) = \begin{pmatrix}
-0 & 1 & 0 & 0 \\
-0 & 0 & 1 & 0 \\
-0 & 0 & 0 & 1 \\
--\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
-\end{pmatrix}$$
-whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
-of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
-at $(u_0, u_1, u_2, u_3)$ and its differential $\varphi_i = 
-f'_i(u_0, u_1, u_2, u_3)$ is given by the matrix
-$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
-D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3)$.
-Thanks to the Laurent phenomenon, we know that $D_i$ has
-integral coefficients, \emph{i.e.} $\varphi_i$ stabilizes the lattice
-$\Z_p^4$.
-We are now going to prove by induction on $i$ that, at the end of the 
-$i$-th iteration of the loop, we have $\prec = N + v_p(\det D_i)$ and
-\begin{equation}
-\label{eq:congrSOMOS}
-(a, b, c, d) \equiv (u_i, u_{i+1}, u_{i+2}, u_{i+3}) \pmod
-{p^N \varphi_i(\Z_p^4)}.
-\end{equation}
-The first point is easy. Indeed, from $D_i = D(u_{i-1}, u_i, u_{i+1}, 
-u_{i+2}) D_{i-1}$, we deduce $\det D_i = \det D_{i-1} \cdot 
-\frac{u_i}{u_{i-3}}$ and the assertion follows by taking determinants
-and using the induction hypothesis. 
-Let us now establish \eqref{eq:congrSOMOS}. To avoid confusion, 
-let us agree to denote by $a'$, $b'$, $c'$, $d'$ and $\prec'$ the values of 
-$a$, $b$, $c$, $d$ and $\prec$ respectively at the \emph{beginning} of 
-the $i$-th iteration of the loop. By induction hypothesis (or by 
-initialization if $i = 1$), we have:
-\begin{equation}
-\label{eq:congrSOMOS2}
-(a', b', c', d') \equiv (u_{i-1}, u_i, u_{i+1}, u_{i+2}) \pmod
-{p^N \varphi_{i-1}(\Z_p^4)}.
-\end{equation}
-Moreover, we know that the determinant of $\varphi_{i-1}$ has valuation
-$\prec'$. Hence \eqref{eq:congrSOMOS2} remains true if $a'$, $b'$, $c'$
-and $d'$ are replaced by other values which are congruent to them modulo
-$p^{\prec'}$. In particular it holds if $a'$, $b'$, $c'$ and $d'$ denotes
-the values of $a$, $b$, $c$ and $d$ after the execution of line
-\ref{line:lift}. Now applying Lemma \ref{lem:main} and Proposition
-\ref{prop:locanalytic} to $\varphi_{i-1}$ and $\varphi_i$ (at the point 
-$(u_0, u_1, u_2, u_3)$), we get:
-$$f\big((u_{i-1}, u_i, u_{i+1}, u_{i+2}) + p^N \varphi_{i-1}(\Z_p^4)\big)
-= (u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4).$$
-By the discussion above, this implies in particular that $f(a',b',c',d')$
-belongs to $(u_i, u_{i+1}, u_{i+2}, u_{i+3}) + p^N \varphi_i(\Z_p^4)$. We
-conclude by remarking first that $(a,b,c,d) \equiv f(a',b',c',d') \pmod 
-{p^\prec \Z_p^4}$ by construction and second that $p^\prec \Z_p^4 \subset
-p^N \varphi_i(\Z_p^4)$.
-
-Finally \eqref{eq:congrSOMOS} applied with $i = n-3$ together with
-the fact that $\varphi_i$ stabilizes $\Z_p^4$ imply that, when we exit
-the loop, the value of $d$ is congruent to $u_n$ modulo $p^N$. Hence,
-our algorithm returns the correct value.
-
-\medskip
-
-We conclude this section by remarking that Algorithm 
-\ref{algo:SOMOS} performs computations at precision at most $O(p^{N+v})$ 
-where $v$ is the maximum of the sum of the valuations of five consecutive terms among 
-the first $n$ terms of the SOMOS sequence we are considering. 
-Experiments show that the value of $v$ varies like $c \cdot \log n$ 
-where $c$ is some constant. Assuming that we are using a FFT-like 
-algorithm to compute products of integers, the complexity of Algorithm 
-\ref{algo:SOMOS} is then expected to be $\tilde O(N n)$ where the 
-notation $\tilde O$ means that we hide logarithmic factors.
-
-We can compare this with the complexity of the more naive algorithm 
-consisting of lifting the initial terms $u_0, u_1, u_2, u_3$ to enough 
-precision and then doing the computation using a naive step-by-step 
-tracking of precision. In this setting, the required original precision 
-is $O(p^{N+v'})$ where $v'$ is the sum of the valuation of the $u_i$'s 
-for $i$ varying between $0$ and $n$. Experiments show that $v'$ is about 
-$c' \cdot n \log n$ (where $c'$ is a constant), which leads to a 
-complexity in $\tilde O (N n + n^2)$. Our approach is then interesting
-when $n$ is large compared to $N$: under this hypothesis, it saves
-roughly a factor $n$.
-
-\appendix
-
-\section{Generalization to manifolds}
-\label{sec:manifold}
-
-Many natural $p$-adic objects do not lie in vector spaces:
-points in projective spaces or elliptic curves,
-subspaces of a fixed vector space (which lie in Grassmannians),
-classes of isomorphism of certain curves (which lie in various moduli spaces),
-\emph{etc.}  In this appendix we extend the formalism 
-developed in Section \ref{sec:mainlemma} to a more general setting:
-we consider the quite general case of 
-differentiable manifolds locally modeled on ultrametric Banach spaces.
-This covers all the 
-aforementioned examples.
-
-\subsection{Differentiable $K$-manifolds}
-\label{ssec:manifold}
-
-The theory of finite dimensional $K$-manifolds is presented for example 
-in \cite{schneider:11a}*{Ch. 8-9}. In this section, we shall work with
-a slightly different notion of manifolds which allows also Banach vector 
-spaces of infinite dimension.
-More precisely, for us, a \emph{ differentiable $K$-manifold} 
-(or just \emph{$K$-manifold} for short) is the data of a topological 
-space $V$ together with an open covering $V = \bigcup_{i \in I} V_i$ 
-(where $I$ is some set) and, for all $i \in I$, an homeomorphism 
-$\varphi_i : V_i \to U_i$ where $U_i$ is an open subset of a $K$-Banach space
-$E_i$ such that for all $i,j \in I$ for which $V_i \cap V_j$ is 
-nonempty, the composite map
-\begin{equation}
-\label{eq:psiij}
-\psi_{ij} : 
-\varphi_i(V_{ij}) \stackrel{\varphi_i^{-1}}{\longrightarrow} 
-V_{ij} \stackrel{\varphi_j}{\longrightarrow} \varphi_j(V_{ij})
-\quad \text{(with } V_{ij} = V_i \cap V_j \text{)}
-\end{equation}
-is differentiable. We recall that
-the mappings $\varphi_i$ above are the so-called \emph{charts}. The
-$\psi_{ij}$'s are the transition maps. The collection of $\varphi_i$'s
-and $\psi_{ij}$'s is called an \emph{atlas} of $V$. In the sequel, we 
-shall assume further that the open covering $V = \bigcup_{i \in I} V_i$ 
-is locally finite, which means that every point $x \in V$ lies only in a 
-finite number of $V_i$'s. Trivial examples of $K$-manifolds are 
-$K$-Banach spaces themselves.
-
-If $V$ is a $K$-manifold and $x$ is a point of $V$, we define the 
-\emph{tangent space} $T_x V$ of $V$ at $x$ as the space $E_i$ for some 
-$i$ such that $x \in V_i$. We note that if $x$ belongs to $V_i$ and 
-$V_j$, the linear map $\psi'_{ij}(\varphi_i(x))$ defines an isomorphism 
-between $E_i$ and $E_j$. Furthermore these isomorphisms are compatible 
-in an obvious way. This implies that the definition of $T_x V$ given 
-above does not depend (up to some canonical isomorphism) on the index 
-$i$ such that $x \in V_i$ and then makes sense.
-
-As usual, we can define the notion of differentiability (at some 
-point) for a continuous mapping between two $K$-manifolds by viewing it 
-through the charts. A  differentiable map $f : V \to V'$ induces 
-a linear map on tangent spaces $f'(x) : T_x V \to T_{f(x)} V'$ for all 
-$x$ in the domain $V$. It is called the \emph{differential} of $f$ at
-$x$.
-
-\subsection{Precision data}
-
-Returning to our problem of precision, given $V$ a $K$-manifold as 
-above, we would like to be able to deal with ``approximations up to some 
-precision'' of elements in $V$, \emph{i.e.} expressions of the form $x + 
-O(H)$ where $x$ belongs to a dense \emph{computable} subset of $V$ and 
-$H$ is a ``precision datum''.
-For now, we fix a $K$-manifold $V$ and we use freely the notations $I$, 
-$V_i$, $\varphi_i$, \emph{etc.} introduced in \S \ref{ssec:manifold}.
-
-\begin{deftn}
-Let $x \in V$.
-A \emph{precision datum} at $x$ is a lattice in the tangent space 
-$T_x V$ such that for all indices $i$ and $j$ with $x \in U_i \cap
-U_j$, the image of $T_x V$ in $E_i$ is a first order lattice for
-$\psi_{ij}$ at $\varphi_i(x)$ (\emph{cf} Definition \ref{def:firstorder}).
-\end{deftn}
-
-\begin{rmk}
-The definition of a precision datum at $x$ depends not only on $x$
-and the manifold $V$ where it lies but also on the chosen atlas that
-defines $V$.
-\end{rmk}
-
-\begin{lem}
-\label{lem:independence}
-Let $x \in V$ and $H$ be a precision datum at $x$.
-The subset
-$$\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
-\subset V$$
-does not depend on the index $i$ such that $x \in V_i$.
-\end{lem}
-
-\begin{proof}
-Let $i$ and $j$ be two indices such that $x$ belongs to $V_i$ and $V_j$. 
-Set $x_i = \varphi_i(x) \in E_i$ and $H_i = \varphi'_i(x)(H)$. The 
-equality 
-\[
-\varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
- = \varphi_j^{-1}\big(\varphi_j(x) + \varphi_j'(x)(H)\big)
- \]
-is clearly equivalent to
-$\psi_{ij}(x_i + H_i) = \psi_{ij}(x_i) + \psi'_{ij}(x_i)(H_i)$
-and the latter holds because $H_i$ is a first order lattice
-for $\psi_{ij}$ at $x_i$.
-\end{proof}
-
-We are now in position to define $x + O(H)$.
-
-\begin{deftn}
-Let $x \in V$ and $H$ be a precision datum at $x$. We set
-$$x + O(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big)
-\subset V$$
-for some (equivalenty, all) $i$ such that $x \in V_i$.
-\end{deftn}
-
-\subsubsection*{Change of base point.}
-
-In order to restrict ourselves to elements $x$ lying in a dense 
-computable subset, we need to compare $x_0 + O(H_0)$ with varying $x + 
-O(H)$ when $x$ and $x_0$ are close enough. Let us first examine the 
-situation in a fixed given chart: we fix some index $i \in I$ and pick 
-two elements $x_0$ and $x$ in $V_i$. We consider in addition a lattice 
-$\tilde H_0$ in $E_i$ --- which should be think as $\varphi'_i(x_0) 
-(H_0)$ --- and we want to produce a lattice $\tilde H$ such that 
-$\varphi_i(x_0) + \tilde H_0 = \varphi_i(x) + \tilde H$. Of course 
-$\tilde H = \tilde H_0$ does the job as soon as 
-$\varphi_i(x) - \varphi_i(x_0) \in \tilde H_0$. 
-Now, we remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
-isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and $\varphi'_i(x)$ 
-respectively. A natural candidate for $H$ is then:
-\begin{equation}
-\label{eq:Hprime}
-H = \left(\varphi'_i(x)^{-1} \circ \varphi'_i(x_0)\right) (H_0).
-\end{equation}
-With this choice, $x + O(H) = x_0 + O(H_0)$ provided that $x$ and $x_0$ 
-are close enough in the following sense: the difference $\varphi_i(x) - 
-\varphi_i(x_0)$ lies in the lattice $\varphi'_i(x_0)(H_0)$. We 
-furthermore have a property of independence on $i$.
-
-\begin{prop}
-Let $x_0 \in V$ and $H_0$ be a precision datum at $x_0$.
-Then, for all $x$ sufficiently close to $x_0$,
-\begin{enumerate}[(i)]
-\item the lattice $H$ defined by \eqref{eq:Hprime} does not depend 
-on $i$ and is a precision datum at $x$, and
-\item we have $x + O(H) = x_0 + O(H_0)$.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}
-We first prove (i). For an index $i$ such that $x, x_0 \in V_i$, let us 
-denote by $f_i : T_{x_0} V \to T_x V$ the composite $\varphi'_i(x)^{-1} 
-\circ \varphi'_i(x_0)$. Given an extra index $j$ satisfying the same
-assumption, the difference $f_i - f_j$ goes to $0$ when $x$ converges to 
-$x_0$ (see Remark \ref{rem:differentiability}). Since $H_0$ is open in 
-$T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ 
-and $f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in 
-f_j(H_0)$ and write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal 
-to $f_i(v) + (f_j - f_i)(v)$ and thus belongs to $f_i(H_0)$ because each 
-summand does. Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse 
-inclusion is proved in the same way. The fact that $H$ is a 
-precision datum at $x$ is easy and left to the reader.
-Finally, if $x$ is close enough to $x_0$, it is enough to check (ii) in 
-the charts but this was already done.
-\end{proof}
-
-\subsection{Generalization of the main Lemma}
-
-With above definitions, Lemma \ref{lem:main} extends to 
-manifolds. To do so, we first need to define 
-a norm on the tangent space $T_x V$ (where $V$ is some $K$-variety and 
-$x$ is a point in $V$). There is actually in general no canonical choice 
-for this. Indeed, let us consider a $K$-manifold $V$ covered by charts 
-$U_i$'s ($i \in I$) which are open subset of $K$-Banach spaces $E_i$'s. If $x$ 
-is a point in $V$, the tangent space $T_x V$ is by definition isomorphic 
-to $E_i$ for each index $i$ such that $x \in V_i$. A natural norm on 
-$T_x V$ is then the one obtained by pulling back the norm on $E_i$. 
-However, since the transition maps are not required to be isometries, 
-this norm depends on the choice of $i$. They are nevertheless all
-equivalent because the transition maps are required to be continuous.
-
-In the next lemma, we choose any of the above norms for $T_x V$.
-
-\begin{lem}
-Let $V$ and $W$ be two $K$-manifolds. 
-Suppose that we are given a differentiable function $f : V \to W$, 
-together with a point $x \in V$ such that $f'(x) : T_x V \to T_{f(x)} W$ 
-is surjective.
-
-Then, for all $\rho \in (0, 1]$, there exists a positive real 
-number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ in $T_x V$ such that $B^-_{T_x V}(\rho r) \subset H \subset 
-B^{\phantom -}_{T_x V}(r)$ is a first
-order lattice for $f$ at $x$.
-\end{lem}
-
-\begin{proof}
-Apply Lemma \ref{lem:main} in charts.
-\end{proof}
-
-\begin{rmk}
-The constant $\delta$ that appears in the lemma depends (up to
-some multiplicative constant) on the norm that we have chosen on $T_x V$. 
-However, once this norm is fixed, and assuming further that $V$ and $W$ 
-are locally analytic $K$-manifolds and the mapping $f$ is locally 
-analytic as well, the constant $\delta$ can be made explicit using the 
-method of Section \ref{ssec:locanalytic}.
-\end{rmk}
-
-\subsection{Examples}
-\label{ssec:exmanifold}
-
-We illustrate the theory developped above by some classical examples,
-namely elliptic curves and grassmannians.
-
-\subsubsection*{Elliptic curves.}
-
-In this example, we assume for simplicity that $K$ does not have characteristic~$2$.
-Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
-\neq 0$ and let $E$ be the subset of $K^2$ consisting of the pairs 
-$(x,y)$ satisfying the usual equation $y^2 = x^3 + a x + b$.
-Let $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
-takes a pair $(x,y)$ to $x$ (resp. to $y$).
-
-We first assume that $a$ and $b$ lie in the subring $R$ of exact elements. For each point $P_0 = 
-(x_0, y_0)$ on $E$ except possibly a finite number of them, the map 
-$\pr_x$ define a diffeomorphism from an open subset containing $P_0$ to 
-an open subset of $K$; the same is true for $\pr_y$. Moreover, around each 
-$P_0 \in E$, at least one of these projections satisfies the above condition.
-Hence the maps $\pr_x$ and $\pr_y$ define together an atlas of $E$, giving $E$ the structure of 
-a $K$-manifold.
-
-Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
-charts. Lemma \ref{lem:independence} then tells us that a 
-precision datum on $x$ determines a precision datum on $y$ and vice 
-versa. Indeed, in a neighborhood of $P_0$ we can write $y = \sqrt{x^3 + a x + b}$ 
-(for some choice of square root) and find the 
-precision on $y$ from the precision on $x$ using Lemma \ref{lem:main}. We can go 
-in the other direction as well by writing $x$ locally as a function of 
-$y$. A precision datum at $P_0$ is then nothing but a precision datum
-on the coordinate $x$ or on the coordinate $y$, keeping in mind that 
-each of them determines the other. Viewing a precision datum at $P_0$ as 
-a lattice in the tangent space is a nice way to make it canonical but in 
-practice we can just choose one coordinate and track precision
-only on this coordinate.
-
-We conclude this example by showing a simple method to transform a
-precision datum on $x$ to a precision datum on $y$ and \emph{vice versa}.
-Differentiating the equation of the elliptic curve, we get:
-\begin{equation}
-\label{eq:diffec}
-2 y \cdot dy = (3 x^2 + a) \cdot dx.
-\end{equation}
-In the above $dx$ and $dy$ should be thought as a little perturbation of 
-$x$ and $y$ respectively. Equation \eqref{eq:diffec} then gives a linear 
-relation between the precision on $x$ (which is represented by $dx$) and 
-those on $y$ (which is represented by $dy$). This relation turns out to
-correspond exactly to the one which is given by Lemma \ref{lem:main}.
-
-Finally, consider the case where $a$ and $b$ are themselves given with 
-finite precision and $E$ is not fully determined. So 
-we cannot consider it as a $K$-manifold and the above discussion does not 
-apply readily. Nevertheless, we can always consider the submanifold of 
-$K^4$ consisting of all tuples $(a,b,x,y)$ satisfying $y^2 = x^3 + a x + 
-b$. The projections on the hyperplanes $a = 0$, $b = 0$, $x = 0$ and $y = 
-0$ respectively define charts of this $K$-manifold. From this, we see 
-that a precision datum on a point of the ``not well determined'' elliptic 
-curve $E$ is a precision datum on a tuple of three variables among $a$, 
-$b$, $x$ and $y$.
-
-\subsubsection*{Grassmannians.}
-
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
-Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
-of dimension $d$. It defines an algebraic variety over $K$ and hence
-\emph{a fortiori} a $K$-manifold.
-Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
-a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
-and two such matrices $M$ and $M'$ define the same vector space if there 
-exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
-find that we can always choose the above matrix $M$ in the particular
-shape:
-\begin{equation}
-\label{eq:grass}
-M = \begin{pmatrix} I_d & N \end{pmatrix} \cdot P
-\end{equation}
-where $I_d$ denotes the $(d \times d)$ identity matrix, $N \in M_{d, 
-n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
-such expressions with the same $P$ necessarily coincide. Hence each 
-permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
-is canonically diffeomorphic to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.
-
-In other words, if $V$ is a subspace of $K^n$ of dimension $d$, we 
-represent it as a matrix $M$ of the shape \eqref{eq:grass} (using row 
-echelon) and a precision datum at $V$ is nothing but a precision datum on 
-the matrix $N$. If we choose another permutation matrix to represent $V$, 
-say $P'$, we end up with another matrix $N'$; the matrices $N$ and $N'$ 
-are then related by a simple relation. Differentiating it, we find a 
-formula for translating the precision datum expressed in the chart $U_P$ 
-to the same precision datum expressed in the chart $U_{P'}$.
-Of course, in practice, when we are doing computations on subspaces of
-$K^n$ (like sum or intersection), we represent the spaces in charts as
-above and perform all the calculations in these charts.
-
-\section{Differential of usual operations}
-\label{sec:differentials}
-
-We saw in the core of the article that the differential of an operation 
-encodes the intrinsic loss/gain of precision when performing this 
-operation. The aim of this appendix is to compute the differential of 
-many common operations on numbers, polynomials and matrices. Surprisingly, 
-we observe that all differentials we will consider are rather easy to 
-compute even if the underlying operation is quite involved (\emph{e.g.} 
-Gr\"obner basis).
-
-In what follows, we use freely the ``method of physicists'' to compute 
-differentials: given a function $f$ differentiable at some point $x$, we 
-consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$ 
-by expanding LHS and neglecting terms of order $2$. The differential of 
-$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
-
-\subsection{Numbers}
-
-The most basic operations are, of course, sums, products and inverses of 
-elements of $K$. Their differential are well-known and quite easy to 
-compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have 
-$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = - 
-\frac{dx}{x^2}$).
-
-Slightly more interesting is the $n$-th power map $f$ from $K$ to 
-itself. Its differential $f'(x)$ is found by differentiating the 
-equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
-Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(\lvert nx^{n-1} \rvert \cdot r)$. According to Lemma \ref{lem:main}, this 
-means that the behavior of the precision depends on the absolute value 
-of $n$. By the ultrametric inequality, we always have $\lvert n \rvert \leq 1$ but 
-may have $\lvert n \rvert = 0$ if the characteristic of $K$ divides $n$ or 
-$\lvert n \rvert = p^{-k}$ if $K$ is $p$-adic with $\val(n) = k$.  In the first case,
-$f'(x)$ also vanishes and Lemma \ref{lem:main} does 
-not apply.  In the second, Lemma \ref{lem:main} reflects the well-known fact that
-raising a $p$-adic number to the $p^k$-th power increases the precision 
-by $k$ extra digits.
-
-\subsection{Univariate polynomials}
-\label{ssec:polynomials}
+At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
+function $h$ in our differential equation. The aim of this was to be 
+more flexible and possibly get this way better bounds on $\Lambda(f)$. 
+In this paragraph, we discuss a recommended choice for this function 
+$h$.
+
+Actually, we do recommend taking $h$ as the identity function 
+\emph{but} changing the norm on the codomain of $h$ at the same time.
+More precisely, if $F$ --- which is the $K$-Banach in which $f$ takes
+its values --- is equipped with the norm $\Vert \cdot \Vert_F$, we 
+recommand endowing it with the second norm defined by
+$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
+where $\lambda$ is a positive real number that we shall choose later 
+and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
+acting as the identity on the underlying vector spaces. Doing this, the 
+function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
+can certainly choose $\Lambda_h = \Lambda(h)$.
+
+We believe that a good way to opmitize $\lambda$ is to choose it so that 
+the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
+agree. With the expression we got just above for $\Lambda_h$, our 
+heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
+With this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
+at least, on the domain we are interested in --- and we do not loose so 
+much when we write that the norm of the couple $(f(x), h(x))$ is the 
+maximum of those of $f(x)$ and $h(x)$ respectively.
+
+{\color{Bittersweet}
+
+\section{Univariate polynomials}
+\label{sec:polynomials}
 
@@
 
-\subsubsection*{Evaluation and interpolation.}
+\subsection*{Evaluation and interpolation.}
 
@@
 
-\subsubsection*{Euclidean division.}
+\subsection*{Euclidean division.}
 
@@
 
-\subsubsection*{Greatest common divisors and B\'ezout coefficients.}
+\subsection*{Greatest common divisors and B\'ezout coefficients.}
 
@@
 
-\subsubsection*{Factorization.}
+\subsection*{Factorization.}
 
@@
 
-\subsubsection*{Finding roots.}
+\subsection*{Finding roots.}
 
@@
 
-\subsection{Multivariate polynomials}
+\section{Multivariate polynomials}
 
@@
 
-\subsubsection*{Division.}
+\subsection*{Division.}
 
@@
 
-\subsubsection*{Gr\"obner basis.}
+\subsection*{Gr\"obner basis.}
 
@@
 
-\subsection{Matrices}
-\label{ssec:matrices}
+\section{Matrices}
+\label{sec:matrices}
 
@@
 
-\subsubsection*{Determinants and characteristic polynomials.}
+\subsection*{Determinants and characteristic polynomials.}
 
@@
 
-\subsubsection*{LU factorization.}
+\subsection*{LU factorization.}
 
@@
 
-\subsubsection*{QR factorization.}
+\subsection*{QR factorization.}
 
@@
 
-\subsection{Vector spaces}
+\section{Vector spaces}
 
@@
 sense of Appendix \ref{sec:manifold} as it is discussed in \S 
-\ref{ssec:exmanifold}. In the sequel, we use freely the formalism 
+\ref{sec:exmanifold}. In the sequel, we use freely the formalism 
 introduced there.
 
-\subsubsection*{Left kernels.}
+\subsection*{Left kernels.}
 
@@
 
-Following what we have already done in \S \ref{ssec:polynomials} for 
+Following what we have already done in \S \ref{sec:polynomials} for 
 greatest common divisors of polynomials, we can develop further this 
@@
 
-\subsubsection*{Intersections.}
+\subsection*{Intersections.}
 
@@
 
-\subsubsection*{Sums and images.}
+\subsection*{Sums and images.}
 
@@
 
-\bibliographystyle{plain}
-\bibliography{roebib/Biblio,extras}
+}
+
+%\bibliographystyle{plain}
+%\bibliography{roebib/Biblio,extras}
 

### 2014-08-04 examples.tex
typo

@@
 During the proof, multinomial coefficients will play an important role. 
-Given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
-k_1 + \cdots + k_\ell$, it is defined by:
+We recall that they are defined as follows: 
+given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
+k_1 + \cdots + k_\ell$, we set:
 $$\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
   \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.$$
-We recall that they are always positive integers.
+We also recall that these numbers are all positive integers.
+
 We now expand $g \circ f$ as follows:
@@
 \null \hfill
-\begin{tikzpicture}[xscale=0.8,yscale=0.8]
+\begin{tikzpicture}[scale=0.8]
 \draw[thick,green!30,fill=green!20] 
@@
 \draw[Blue,very thick] (-5,-4)--(-2,-1)--(-1,2);
+\fill[Blue] (-1,2) circle (0.06);
+\draw[Blue] (-1,2)--(-1,3);
 \draw[Sepia, very thick] (-4.5,-5)--(3.5,3);
@@
 As a consequence of the above discussion, we derive the following 
-proposition which can be used directly in combination with Proposition
-3.12 of \cite{padicprec}.
+proposition which can be directly combined with Proposition 3.12 of 
+\cite{padicprec}.
 
@@
 \min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
-for all $x < \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
+for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
 \end{prop}
@@
 function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
-can certainly choose $\Lambda_h = \Lambda(h)$.
+choose $\Lambda_h = \Lambda(h)$.
 
@@
 heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
-With this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
+Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
 at least, on the domain we are interested in --- and we do not loose so 
 much when we write that the norm of the couple $(f(x), h(x))$ is the 
-maximum of those of $f(x)$ and $h(x)$ respectively.
+maximum of those of $f(x)$ and $h(x)$.
 

### 2014-08-17 examples.tex
A stable algorithm for computing subresultants

@@
 Keeping the above notations, we have:
-$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) + 
+$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
 \min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
@@
 \begin{proof}
-Just remark that $y = 2(x + \alpha + \mu) + y_0$ is a equation of the 
-dashed blue line.
+Just remark that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
+the dashed blue line.
 \end{proof}

### 2014-08-21 examples.tex
Typos
Correction of some typos, plus some small details added.

@@
 consider two locally analytic functions $f : U \to V$ and $g : V \to W$ 
-two locally analytic functions around $0$. The composition $h = g \circ
+two locally analytic functions around $0$, with $f(0)=0.$ The composition $h = g \circ
 f$ is then locally analytic around $0$ as well. Let us write the
@@
 $\O_K$. We set $x'_i = x_i$, $k'_i = k_i$ when $i < \ell$ and $x'_\ell 
-= x_\ell + \lambda x_{\ell+1}$, $k_\ell = k_\ell + k_{\ell+1}$. By the 
+= x_\ell + \lambda x_{\ell+1}$, $k'_\ell = k_\ell + k_{\ell+1}$. By the 
 induction hypothesis, we know that the inequality:
@@
 the unit ball, the norm of all its coefficients must be at most $\Vert \psi
-\Vert$ as well. Looking at the coefficient in $\lambda^{k_i}$, we find
+\Vert$ as well. Looking at the coefficient in $\lambda^{k_{l+1}}$, we find
 Eq.~\eqref{eq:normpolar} for the families $(k_1, \ldots, k_{\ell+1})$ 
@@
 In what follows, we always assume that $V$ and $W$ contains the origin, 
-$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are armless 
+$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
 because (1)~we can always shift $f$ and $h$ (and $g$ accordingly) so 
@@
 
-We now pick two real numbers $A$ and $B$ satisfying \eqref{eq:condAB}.
-Set $d = \Lambda(h)(a)$. Going back to the definitions, we remark that
+We now pick two real numbers $a$ and $b$ satisfying \eqref{eq:condAB}.
+Set $d = \Lambda(h)(a)$. Going back to the definitions of $\Lambda (h)$ and Legendre transform, we remark that
 the above equality implies:
@@
 moreover that the Taylor expansion of $f(x)$ starts with the term
-$g(0)(x)$. Hence, on a neighborhood on ${-}\infty$, we have 
+$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
 $\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
@@
 
-We believe that a good way to opmitize $\lambda$ is to choose it so that 
+We believe that a good way to optimize $\lambda$ is to choose it so that 
 the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 

### 2014-11-17 examples/examples.tex
Typos in examples.tex

--- 2014-08-17 examples.tex
@@
 
+\emph{In all this section, we assume that \textbf{the base field $K$ 
+has characteristic $0$}.}
+
 \begin{comment}
@@
 Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces 
-$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$ and
-consider two locally analytic functions $f : U \to V$ and $g : V \to W$ 
-two locally analytic functions around $0$. The composition $h = g \circ
-f$ is then locally analytic around $0$ as well. Let us write the
-analytic expansion of $f$, $g$ and $h$ as follows:
+$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
+$f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
+$0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
+analytic around $0$ as well. Let us write the analytic expansion of $f$, 
+$g$ and $h$ as follows:
 $$f = \sum_{n \geq 0} f_n, \quad 
@@
 where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
-some $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The aim of
-this subsection is to prove the following proposition.
+some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
+aim of this subsection is to prove the following proposition.
 
@@
 where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
-some $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
+some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
 Following \cite{padicprec}, we define $\NP(f) : \R \to \R \cup 
@@
 
-We now assume that \textbf{$K$ has caracteristic $0$} and we consider a 
-nonnegative real number $\alpha$ such that $\Vert n! \Vert \geq 
-e^{-\alpha n}$ for all positive integer $n$. A suitable value for 
+We consider a nonnegative real number $\alpha$ such that $\Vert n! \Vert 
+\geq e^{-\alpha n}$ for all positive integer $n$. A suitable value for 
 $\alpha$ is
@@
 where $p$ denotes the characteristic of the residue field and
-where, by convention, the above expression equals $0$ if $p = 0$.
+where, by convention, the above expression equals $0$ if $p = 0$
+(we recall that we have assumed that $K$ has characteristic $0$).
 
@@
 \begin{proof}
-Assuming that $F_n$ is symmetric (which is armless in characteristic 
-$0$), we get $f' = \sum_{n \geq 0} f'_n$ where
+We have $f' = \sum_{n \geq 0} f'_n$ where
 $$f'_n : U \to \mathcal L(E,F), \quad
@@
 
-We now pick two real numbers $A$ and $B$ satisfying \eqref{eq:condAB}.
+We now pick two real numbers $a$ and $b$ satisfying \eqref{eq:condAB}.
 Set $d = \Lambda(h)(a)$. Going back to the definitions, we remark that
@@
 \label{eq:boundgm}
-\Vert g_m \Vert \leq e^{- \max(b,d)\cdot m + b - a},
+\Vert g_m \Vert \leq e^{-\!\max(b,d)\cdot m + b - a},
   \quad \forall m \in \N.

### 2014-11-17 examples.tex
Changes for submission

--- 2014-08-04 examples.tex
@@
 consider two locally analytic functions $f : U \to V$ and $g : V \to W$ 
-two locally analytic functions around $0$. The composition $h = g \circ
+around $0$. The composition $h = g \circ
 f$ is then locally analytic around $0$ as well. Let us write the
@@
 \label{prop:boundhr}
-With the above notations, we have for all nonnegative integer $r$:
-$$\Vert h_r \Vert \leq \sup_{m, (n_i)}
-  \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert$$
-where the supremum is taken over all couples $(m, (n_i))$ where $m$
-is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
-length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
+For any nonnegative integer $r$, 
+\[
+\Vert h_r \Vert \leq \sup_{m, (n_i)}
+  \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert,
+\]
+where the supremum is taken over nonnegative integers $m$
+and sequences $(n_i)_{1 \leq i \leq m}$ of nonnegative integers
+with $n_1 + \cdots + n_m = r$.
 \end{prop}
 
-During the proof, multinomial coefficients will play an important role. 
-We recall that they are defined as follows: 
-given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
-k_1 + \cdots + k_\ell$, we set:
-$$\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
-  \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.$$
-We also recall that these numbers are all positive integers.
+Multinomial coefficients will play an important role in the proof, so we recall their definition. 
+Given positive integers $s, k_1, \ldots, k_\ell$ with $s \geq k_1 + \cdots + k_\ell$, set
+\[
+\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
+  \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.
+\]
+These numbers are all positive integers.
 
@@
 \end{align}
-where the latest sum runs over:
+where the second sum runs over
 

### 2014-11-26 resultants/Caruso_resume.tex
Paper on resultants and subresultants (preliminary version)

\documentclass[12pt,a4paper]{article}

%***************************************************************
% Choisissez l'encodage:
%\usepackage[latin1]{inputenc}            % pour Windows
\usepackage[utf8]{inputenc}                % pour Linux
%\usepackage[applemac]{inputenc}        % pour Mac
%****************************************************************

\usepackage[T1]{fontenc}
\usepackage[english,francais]{babel}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsmath,amssymb,amsthm}

\newcommand{\Z}{\mathbb Z}
\newcommand{\E}{\mathbb E}
\newcommand{\prob}{\mathbb P}
\newcommand{\pgcd}{\text{\sc pgcd}}
\newcommand{\Res}{\text{\rm Rés}}

\newtheorem{theo}{Théorème}

\begin{document}

\thispagestyle{empty}

\title{\textbf{Résultants et sous-résultants de polynômes $p$-adiques}}

\author{\textbf{X.~Caruso} \\
IRMAR \\
Université Rennes 1 \\
Campus de Beaulieu \\
35042 Rennes Cedex \\
\texttt{xavier.caruso@normalesup.org}}
\date{}

\maketitle

Soient $p$ un nombre premier et $\Z_p$ l'anneau des entiers 
$p$-adiques. À l'origine de ce travail est la volonté d'obtenir
des algorithmes à la fois efficaces et stables numériquement pour
le calcul du $\pgcd$ --- ainsi que des coefficients de Bézout ---
de polynômes à coefficients dans $\Z_p$.

\medskip

Un premier candidat est, bien entendu, l'algorithme d'Euclide usuel. 
Malheureusement, s'il est plutôt efficace, on s'aperçoit rapidement 
qu'il ne fait pas le poids au niveau de la stabilité numérique. En 
effet, on observe expérimentalement que sur des entrées aléatoires $A$ 
et $B$ piochées parmi les polynômes unitaires de degré $d$ à 
coefficients dans $\Z_p$, l'algorithme d'Euclide étendu calcule les 
coefficients de Bézout correspondants $U$ et $V$ avec une perte moyenne 
d'un nombre de chiffres significatifs sur chaque coefficient qui croît 
proportionnellement à $d$.
De surcroît, on obtient fréquemment des exemples pour lesquels on 
observe une chute importante de la précision alors que le résultant 
$\Res(A,B)$ est inversible dans $\Z_p$. Or, avec un algorithme stable,
ceci ne devrait pas se produire car la théorie des résultants affirme 
que les coefficients de $U$ et $V$ s'écrivent comme le quotient d'une 
expression polynômiale en les coefficient de $A$ et $B$ par $\Res(A,B)$ ; 
ainsi, si $\Res(A,B)$ est inversible dans $\Z_p$, on s'attend à 
connaître les coefficients de Bézout avec la même précision que les 
entrées.

\bigskip

La première partie de mon exposé sera consacrée à l'explication des 
phénomènes qui viennent d'être décrits. Plus précisément, je 
montrerai que les pertes de précision qui s'accumulent au cours de 
l'exécution de l'algorithme d'Euclide valent approximativement :
\begin{equation}
\label{eq:perteEuclide}
2 \cdot \sum_{j=0}^{d-1} V_j(A,B)
\end{equation}
où $V_j(A,B)$ désigne la valuation de coefficient dominant du $j$-ième
sous-résultant de $(A,B)$.
Cette quantité est à mettre en comparaison avec la valeur $2 \cdot 
V_0(A,B)$ qui est la perte \og théorique \fg\ donnée par l'argument
des résultants. J'étudierai ensuite les fonctions $V_j$ considérées 
comme des variables aléatoires : j'énoncerai un théorème qui décrit 
leur loi et, comme corollaire, en déduirai les résultats descriptifs 
que voici.

\begin{theo}
Pour tout $j \in \{0, \ldots, d-1\}$, on a :
\begin{enumerate}[i)]
\item $\frac 1 {p-1} \leq \E[V_j] \leq \frac p {(p-1)^2}$ ;
\item $\sigma(V_j) = O\big(\frac 1 {\sqrt p}\big)$ ;
\item $\prob[V_j \leq m] = O\big( p^{-m + O(\sqrt m)} \big)$
\end{enumerate}
où les constantes dans tous les $O(\cdot)$ sont absolues (et, en
particulier, ne dépendent ni de $j$, ni de $d$).
\end{theo}

Il résulte du théorème que l'expression \eqref{eq:perteEuclide} vaut en 
moyenne $\simeq \frac{2d}{p-1}$, en accord avec ce qui avait été 
observé initialement. En comparaison, la perte \og théorique \fg\ $2 
\cdot V_0(A,B)$ ne vaut en moyenne que $\simeq \frac 2 {p-1}$. En
d'autres termes, l'algorithme d'Euclide surestime les pertes de
précision d'un facteur $d$.

\bigskip

Enfin, dans une deuxième partie de mon exposé, je présenterai une 
variante \og stabilisée \fg\ de l'algorithme d'Euclide qui conserve sa 
complexité mais atteint également la perte de précision donnée par la 
théorie des résultants. Cette variante repose de façon essentielle sur
la théorie de la précision $p$-adique développée dans \cite{padicprec}.

\vspace{30pt}

\renewcommand{\section}[2]{}%
\textbf{Bibliographie}
\begin{thebibliography}{0}
\bibitem{padicprec}
{\sc X.~Caruso, D.~Roe, T.~Vaccon},
{\em Tracking $p$-adic precision},
LMS J. Comput. Math. {\bf 17} (Special issue A), 2014,  274--294
\end{thebibliography}

\end{document}

### 2014-11-26 resultants/resultants.tex
Paper on resultants and subresultants (preliminary version)

\documentclass{sig-alternate}
%\documentclass{lms}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb}
\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\renewcommand*{\theHsection}{\thesection}

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}

\begin{document}

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
%\theoremstyle{definition}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\F}{\mathbb F}
\newcommand{\Fp}{\F_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}
\newcommand{\m}{\mathfrak m}

\newcommand{\M}{\text{\tt M}}

\newcommand{\ring}{\mathfrak A}
\newcommand{\fracring}{\mathfrak K}
\renewcommand{\prec}{\text{\rm prec}}
\renewcommand{\val}{\text{\rm val}}
\newcommand{\id}{\text{\rm id}}
\newcommand{\Res}{\text{\rm Res}}
\newcommand{\lc}{\text{\rm lc}}

\renewcommand{\P}{\mathbb P}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\text{\rm Var}}
\newcommand{\Cov}{\text{\rm Cov}}

\newcommand{\lb}{\ensuremath{\llbracket}}
\newcommand{\rb}{\ensuremath{\rrbracket}}
\newcommand{\lp}{(\!(}
\newcommand{\rp}{)\!)}
\newcommand{\col}{\: : \:}

\newcommand{\A}{W}

\def\todo#1{\ \!\!{\color{red} #1}}
\definecolor{purple}{rgb}{0.6,0,0.6}
\def\todofor#1#2{\ \!\!{\color{purple} {\bf #1}: #2}}

\def\binom#1#2{\Big(\begin{array}{cc} #1 \\ #2 \end{array}\Big)}

\title{Resultants and subresultants\\of p-adic polynomials}

\numberofauthors{1}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
}

\maketitle

\begin{abstract}
\end{abstract}

\vspace{1mm}
 \noindent
 {\bf Categories and Subject Descriptors:} \\
\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
  Manipulation -- \emph{Algebraic Algorithms}

 \vspace{1mm}
 \noindent
 {\bf General Terms:} Algorithms, Theory

 \vspace{1mm}
 \noindent
 {\bf Keywords:} 
\medskip

\section{Introduction}

The strating point of this paper is the willing to desing a reasonable 
algorithm to compute \textsc{gcd}s and B\'ezout coefficients of 
polynomials with coefficients in $\Zp$ (the ring of $p$-adic integers). 
Indeed, easy experimentations show that all standard algorithms for this 
(\emph{e.g.} extended Euclide's algorithm) are not satisfying because of 
stability.

Stability issues occur in this context because, just like real numbers, 
$p$-adic numbers are by nature infinite and thus need to be truncated in 
order to fit in the memory of a computer. The level of truncation is 
called the \emph{precision}. Usual software implementing $p$-adics (as 
\textsc{magma}, \textsc{sage}) generally tracks the precision as follows:
an individual precision is attached to any $p$-adic variable and this
precision is updated after each basic arithmetic operation (addition
and multiplication). This way of tracking precision can be seen as the
analogue of the arithmetic intervals in the real setting.

Figure \ref{fig:} illustrates the unstability of Euclide's algorithm 
which it is called on random inputs which are monic $2$-adic polynomials 
of fixed degree. Let us explain how this table should be read by 
commenting on its last line. It indicates that (extended) Euclide's 
algorithm outputs the B\'ezout coefficients of two monic $2$-adic 
polynomials of degree $500$ with an average loss of $...$ significant 
digits by coefficient whereas a stable algorithm should only loose $...$ 
digits in average. This ``theoretical'' loss is computed as the double 
of the valuation of the resultant. Indeed Cramer-like formulae imply 
that B\'ezout coefficients can be computed by performing a unique 
division by the resultant, inducing then only the aforementioned loss 
of precision. Examining the table a bit more, we furthermore observe 
that the ``practical'' loss of precision due to Euclide's algorithm 
seems to grow linearly with respect to the degree whereas the 
``theoretical'' loss seems to be independant on the degree. In other
words, the unstability of Euclide's algorithm is becoming more and
more critical when the degree of the input increases.

\medskip

[...]

\section{The setting}

\subsection{Complete discrete valuation rings}

bla

\medskip

\noindent
\textbf{Definitions.}
Let $\A$ be a \emph{discrete valuation ring}. We recall that it is a 
domain equipped with a map $\val : \A \to \Z \cup \{+\infty\}$ --- the 
so-called \emph{valuation} --- satisfying the four axioms: (1)~$\val(x) = 
+\infty$ iff $x = 0$, (2)~$\val(xy) = \val(x) + \val(y)$ and (3)~$\val(x+y) \geq 
\min(\val(x), \val(y))$, (4)~any element of valuation $0$ is invertible.
We also recall that a discrete valuation ring admits a unique maximal 
ideal $\m$, consisting of elements of positive valuation. 
Furthermore, this ideal is principal and generated by any element of 
minimal valuation. Such an element is called a \emph{uniformizer}.
Let us fix one of them and denote it by $\pi$. The \emph{residue
field} of $W$ is the quotient $W/\m = W/\pi W$ and we shall denote
it by $k$.

The valuation defines a distance $d$ on $\A$ by letting $d(x,y) = 
e^{-\val(x-y)}$ for all $x, y \in \A$. We say that $\A$ is \emph{complete}
if it is complete with respect to the previous distance. Assuming that
$\A$ is complete, any element $x \in \A$ can be written uniquely as a 
convergent series:
\begin{equation}
\label{eq:expandCDVR}
x = \sum_{i=0}^\infty x_i \pi^i
\end{equation}
where the $x_i$'s lie in a fixed set of representatives of classes
modulo $\pi$. Therefore, as an additive group, $W$ is isomorphic to 
the set of sequences $\N \to k$. On the contrary, the multiplicative
structure may vary.

Let $K$ denote the fraction field of $\A$. The valuation $v$ extends 
uniquely to $K$ by letting $\val(\frac x y) = \val(x) - \val(y)$. 
Moreover, it follows from the axiom (4) that $K$ is obtain from $\A$ by 
inverting $\pi$. Thus, any element of $K$ can be uniquely written as an 
infinite sum:
\begin{equation}
\label{eq:expandCDVF}
x = \sum_{i=v}^\infty x_i \pi^i
\end{equation}
where $v$ is some relative integer and the $x_i$'s are as above. The
valuation of $x$ can be easily read on this writing: it is the smallest 
integer $v$ such that $x_i \not\equiv 0 \pmod \pi$.

\medskip

\noindent
\textbf{Examples.}
The two main examples of complete discrete valuation rings are:
\begin{enumerate}[(1)]
\vspace{-0.25cm}
\item the ring of $p$-adic integers $\Zp$: the valuation on it is
the usual $p$-adic valuation, a uniformizer is $p$, its residue
field is $\Fp = \Z/p\Z$ and Eq.~\eqref{eq:expandCDVR} corresponds
to standard expansion in $p$-basis of a $p$-adic number; the fraction 
field of $\Zp$ is the so-called field of $p$-adic numbers and is usually 
denoted by $\Qp$;
\vspace{-0.25cm}
\item the ring of formal series $k[[x]]$ where $k$ is a field: the
valuation on it is the usual valuation of series, a uniformizer is
$x$, the residue field is $k$ and Eq.~\eqref{eq:expandCDVR} is the
usual writing of formal series; the fraction field of $k[[x]]$ is
the field of Laurent series $k((x))$.
\end{enumerate}
\vspace{-0.2cm}

\medskip

\noindent
\textbf{Precision.}
The memory of a computer being necessarily finite, it cannot be possible 
to find a way to represent exhaustively all elements of $\A$. 
Very often, mimicing what we do for real numbers, we choose to truncate 
the expansion \eqref{eq:expandCDVR} at some finite level. Concretely,
this means that we work with approximations of elements of $\A$ of the
form:
$$\sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N)$$
where $N$ is some integer called the \emph{precision} of $x$ and 
the term $O(\pi^N)$ means that the $x_i$'s with $i \geq N$ are not 
specified. Hence, the notation above does not represent a unique
element of $\A$ but all elements lying in the ball having center
$\sum_{i=0}^{N-1} x_i \pi^i$ and radius $e^{-N v(\pi)}$. In other
words, on a computer, we do not work with elements of $\A$ but with
balls in $\A$ which are much more tractable.

\subsection{Subresultants}

A first issue when dealing with numerical computations of \textsc{gcd}s 
of polynomials over $\A$ is that the $\gcd$ function is not continuous: 
it takes the value $1$ on an open dense subset without being constant. 
This of course annihilate any hope of computing \textsc{gcd}s of 
polynomials when only approximations of them are known. Fortunately, 
there exists a standard way to recover continuity in this context: it 
consists in replacing \textsc{gcd}s by the so-called subresultants 
(whose definitions are main properties are quickly recalled below) which 
are playing an analoguous role. For this reason, in what follows, we 
will exclusively consider the problem of computing subresultants.

\medskip

\noindent
\textbf{Notations.} We refer to \cite{} for a nice and rather complete 
introduction to resultants and subresultants and seek to fix notations
in this paragraph. For now, we work over a 
general ring $\ring$. We denote by $\ring_{<n}[X]$ (resp. $\ring_{\leq 
n}[X]$) the set of polynomials of degree less than $n$ (resp. less than 
or equal to $n$) with coefficients in $\ring$. Both $\ring_{<n}[X]$ and 
$\ring_{\leq n}[X]$ are free $\ring$-modules of finite rank.

Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
$d_A$ and $d_B$ be two integer greater than or equal to the degree of 
$A$ and $B$ respectively. Given an integer $j \in \{0, \ldots, \min 
(d_A, d_B)-1\}$, we denote by $\Res_j^{d_A, d_B}(A,B)$ the $j$-th 
\emph{subresultant} of $(A,B)$ ``computing in degree $(d_A, d_B)$'' and 
by $U_j^{d_A, d_B}(A,B)$ and $V_j^{d_A, d_B}(A,B)$ the corresponding 
\emph{cofactors}. In what follows, we will freely drop the exponent 
$(d_A,d_B)$ when $d_A$ and $d_B$ are the degrees of $A$ and $B$ 
respectively.

We recall briefly that $\Res_j^{d_A, d_B}(A,B)$ (resp. $U_j^{d_A, 
d_B}(A,B)$, resp. $V_j^{d_A, d_B}(A,B)$) is a polynomial with 
coefficients in $\ring$ of degree at most $j$ (resp. $d-j-1$, resp. 
$d-j-1$) whose coefficients can be expressed, up to a sign, as a minor 
of the Sylvester matrix. These polynomials satisfy in addition the
relation:
$$A \cdot U_j^{d_A,d_B}(A,B) + B \cdot V_j^{d_A,d_B}(A,B) = 
\Res^{d_A,d_B}_j(A,B).$$
The $j$-th coefficient of $R_j^{d_A,d_B}(A,B)$ is sometimes called the
$j$-th \emph{scalar subresultant} of $(A,B)$.
When $j=0$, the polynomial $\Res^{d_A,d_B}_0(A,B)$ is a constant which
is nothing but the resultant of $(A,B)$. In the sequel, we shall denote
it simply $\Res^{d_A,d_B}(A,B)$ (thus removing the index $0$).
The relation between subresultants and \textsc{gcd}s is given by the
following well-known theorem.

\begin{theo}
We assume that $\ring$ is a field. Let $A$ and $B$ be two polynomials
with coefficients in $\ring$. Let $j$ be the smallest integer such
that $R_j(A,B)$ does not vanish. Then $R_j(A,B)$ is a \textsc{gcd}
of $A$ and $B$.
\end{theo}

Since they are defined as determinants, subresultants behave well with 
respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
and $A$ and $B$ are polynomials over $\ring$ then:
$$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$$
where $f(A)$ and $f(B)$ denotes the polynomials deduced from $A$
and $B$ respectively by applying $f$ coefficient-wise. This property
is sometimes referred to as the \emph{functoriality} of subresultants.

\medskip

\noindent
\textbf{The subresultant pseudo-remainder sequence.}
When $\ring$ is a domain, there exists a standard nice Euclide-like
reinterpreation of subresultants, which provides in particular an 
efficient algorithm for computing them. Since it will play an important 
role in this paper, we take a few lines to recall it.

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above and assume that $d_A = \deg A$. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
$$\left\{\begin{array}{ll}
S_0 = A, \, S_1 = B, \, c_0 = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\delta_i + 1}
s_{i-1}^{-1} \: c_i^{-\delta_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 1 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\delta_{i+1}} \cdot c_i^{1-\delta_{i+1}}
& \text{for } i \geq 0. 
\end{array}\right.$$
Here $n_i = \deg S_i$ and $s_i$ is the leading coefficient of 
$S_i$ if $i > 0$ and $s_0 = 1$ by convention. These sequences are
finite and the above recurrence applies until $S_i$ has reached the
value $0$.

% \delta_i = n_{i-1} - n_i

\begin{prop}
\label{prop:prem}
We keep the above notations. For all positive integer $i$ such that
$S_i$ is defined, we have:
$$S_i = \Res_{\deg S_{i-1} - 1} (A,B).$$
\end{prop}

\begin{rem}
Mimicing ideas behind extended Euclide's algorithm, one can define the 
``extended subresultant pseudo-remainder sequence'' as well and compute
this way at the same time the cofactors.
\end{rem}

Important simplifications occur in the common case where all scalar 
subresultants do not vanish. Under this additional assumption, one can 
prove that the degrees of the $S_i$'s decrease by one at each step, 
meaning that we $\deg S_i = d_B + 1 - i$ for all $i$. The sequence 
$(S_i)$ then stops at $i = d_B + 1$. Moreover, the $\delta_i$'s and the 
$c_i$'s are no longer necessary since we have $\delta_i = 1$ and $c_i = 
s_i$ for all $i$. The recurrence formula then becomes:
$$\left\{\begin{array}{ll}
S_0 = A, \, S_1 = B \smallskip \\
\displaystyle S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
\cdot (S_{i-1} \,\%\, S_i) & \text{for } i \geq 1.
\end{array}\right.$$
To conclude with, we remark that Proposition \ref{prop:prem} simply 
states that $R_j = S_{d_B+1-j}$; we then get this way all subresultants.

\section{Euclide-like algorithms\\are unstable}

%The aim of this section is to study the loss of precision

\subsection{A lower bound on losses of precision}

We consider two fixed polynomials $A$ and $B$ with coefficients in $\A$ 
whose coefficients are known with precision $O(p^N)$ for some positive
integer $N$. For simplicity, we assume further that $A$ and $B$ are both
monic and share the same degree $d$. 
For any integer $j$ between $0$ and $d-1$, we denote by $R_j$ the $j$-th 
subresultant of $A$ and $B$.

In this subsection, we estimate the loss of precision if we compute the 
$R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
In what follows, we are going use a \emph{flat precision model}: this 
means that a polynomial $P(X)$ is represented as:
$$P(X) = \sum_{i=1}^n a_i X^i + O(p^N)
\quad \text{with } a_i \in \Qp \text{ and } N \in \Z.$$
In other words, we shall assume that the software we are using does not
carry a precision data on each coefficient but only a unique precision
data for the whole polynomial. Concretely this means that, after having
computing a polynomial, the software truncates the precision on each
coefficients to the smallest precision. 
One can argue that this assumption is too strong (compared to usual
implementations of $p$-adic numbers). Nevertheless, it defines a 
simplified framework where computations can be performed and experiments 
show that it rather well reflects the behaviour of the loss of precision 
in Euclide-like algorithms.

Let $V_j$ be the valuation of the leading coefficient of $R_j$ and $W_j$ 
be the smallest valuation of a coefficient of $R_j$. We of course have 
$V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.

\begin{prop}
\label{prop:precEuclide}
If $A$ and $B$ are known at precision $O(p^N)$ then, in the flat 
precision model, the subresultant pseudo-remainder sequence computes 
$R_j$ at precision $O(p^{N_j})$ with
$$N_j \leq N + V_{j+1} - 2 \sum_{k=j+1}^d \delta_j.$$
\end{prop}

\begin{proof}
\end{proof}

\begin{rem}
\end{rem}

\subsection{Behaviour on random inputs}

Proposition \ref{prop:precEuclide} gives an estimation of the loss
of precision in Euclide-like algorithms in terms of the quantities
$V_j$ and $\delta_j$. It is nevertheless \emph{a priori} not clear
how large these numbers are. The aim of this paragraph is to compute
their order of magnitude when $A$ and $B$ are picked randomly among
the set of monic polynomials of degree $d$ with coefficients in $\A$.

We endow $\A$ with its Haar measure. The set $\Omega$ of couples of 
monic polynomial of degree $d$ with coefficients in $A$ is canonically 
in bijection with $\A^{2d}$ and hence inherits the product measure. 
We consider $V_j$, $W_j$ and $\delta_j$ as random variables defined on 
$\Omega$.

\begin{theo}
\label{th:lawVj}
We fix $j \in \{0, \ldots, d-1\}$.
Let $X_0, \ldots, X_{d-1}$ be $d$ pairwise independant discrete random 
variables with geometric law of parameter $(1 - p^{-1})$, \emph{i.e.}
$$\P[X_i = k] = (1 - p^{-1}) \cdot p^{-k} \quad 
\text{(with } 0 \leq i < d \text{ and } k \in \N \text{)}.$$
Then $V_j$ is distributed as the random variable
$$Y_j = \sum_{i=0}^d \min(X_{j-i}, X_{j-i+1}, \ldots, X_{j+i})$$
where, by convention, $X_i = +\infty$ if $i < 0$ and $X_i = 0$ if 
$i \geq d$.
\end{theo}

\begin{rem}
The above Theorem does not say anything about the correlations between
the $X_j$'s. In particular, we emphasize that it is \emph{false} that
the tuple $(V_{d-1}, \ldots, V_0)$ is distributed as $(Y_{d-1}, \ldots,
Y_0)$. For instance, an explicit computation shows that $(V_{d-1}, 
V_{d-2})$ is distributed as 
$$(X, \, Y + \min(Y, [X/2]))$$
where $X$ and $Y$ are two independant discrete random variables with 
geometric law of parameter $(1 - p^{-1})$ and the notation $[\cdot]$ 
stands for the integer part function. In particular, we observe that 
the couple $(2,1)$ does not occur almost surely\footnote{Actually, one
can prove that it never occurs.} although the events $\{V_{d-1} = 2\}$ 
and $\{V_{d-2} = 1\}$ both occur with positive probability.
\end{rem}

\begin{theo}
\label{th:estimWj}
For all $j \in \{0, \ldots, d-1\}$ and all $m \in \N$, we have:
$$\P[W_j \geq m] \geq (1 - p^{-1}) \cdot p^{-jm}.$$
\end{theo}

The proof of these two Theorem is given in \S \ref{subsec:proof}. 
Let us now derive some consequences. We will need the permutation
$\sigma$ defined by:
$$\begin{array}{rcl}
\sigma & = & \left(
\begin{array}{cccccccc}
1 & 2 & \cdots & \frac d 2 & \frac d 2 + 1 & \frac d 2 + 2 & \cdots & d \smallskip \\
1 & 3 & \cdots & d-1 & d & d-2 & \cdots & 2 
\end{array} \right) \\
& & \hspace{5cm} \text{if $d$ is even} \medskip \\
\sigma & = & \left(
\begin{array}{cccccccc}
1 & 2 & \cdots & \frac{d+1} 2 & \frac{d+3} 2 & \frac{d+5} 2 & \cdots & d \smallskip \\
1 & 3 & \cdots & d & d-1 & d-3 & \cdots & 2 
\end{array} \right) \\
& & \hspace{5cm} \text{if $d$ is odd.} \\
\end{array}$$
In other words, $\sigma$ takes first the odd values in $[1,d]$ in 
increasing order and then the even values in the same range in 
decreasing order.

\begin{cor}
\label{cor:Vj}
For all $j \in \{0, \ldots, d-1\}$, we have:

\vspace{-0.15cm}
\begin{enumerate}[\hspace{0.3cm}(1)]
\item $\E[V_j] = 
\displaystyle \sum_{i=1}^{d-j} \frac 1 {p^{\sigma(i)} - 1}$ \\
in particular $\frac 1 {p-1} \leq \E[V_j] < \frac p {(p-1)^2}$
\item $\sigma[V_j]^2 = 
\displaystyle \sum_{i=1}^{d-j} \frac {(2i-1) \cdot p^{\sigma(i)}}
{(p^{\sigma(i)} - 1)^2}$ \\
in particular $\frac{\sqrt p}{p-1} \leq \sigma[V_j] < 
\frac {p \:\sqrt{p+1}} {(p-1)^2}$
\item $\P[V_j \geq m] \leq p^{-m + O(\sqrt m)}$
\item $\E[\max(V_0, \ldots, V_{d-1})] \leq \log_p d + O(\sqrt{\log_p d})$
\end{enumerate}
\end{cor}

\begin{proof}
By Theorem \ref{th:lawVj}, the expected value of $V_j$ is equal to 
$\sum_{i=0}^d \E[Z_i]$ with $Z_i = \min(X_{j-i}, \ldots, X_{j+i})$ ($j$ 
is fixed during all the proof). Our conventions imply that $Z_i$ 
vanishes if $i > d-j$. On the contrary, if $i \leq d-j$, let us define 
$\tau(1), \ldots, \tau(d-j)$ as the numbers $\sigma(1), \ldots, 
\sigma(d-j)$ sorted in increasing order. The random variable $Z_i$ 
is then the minimum of $\tau_j(i)$ independant random variables with 
geometric distribution of parameter $(1 - p^{-1})$ and thus its 
distribution is geometric of parameter $(1 - p^{-\tau(i)})$.
Its expected value is then $\frac 1 {p^{\tau(i) - 1}}$ and the first
formula follows. The inequality $\frac 1 {p-1} \leq \E[V_j]$ is clear
because $\frac 1 {p-1}$ is the first summand in the expression of 
$\E[V_j]$. The upper bound is obtained as follows:
$$\E[V_j] < \sum_{i=0}^\infty \frac 1{p^i - 1}
\leq \sum_{i=0}^\infty \frac 1{p^i - p^{i-1}} = \frac p {(p-1)^2}.$$

The variance of $V_j$ is related to the covariance 
of $Z_i$'s thanks to the formula
$\Var(V_j) = \sum_{1 \leq i,i' \leq d-j} \Cov(Z_i, Z_{i'})$.
Moreover, given $X$ and $X'$ two independant variables having geometric
distribution of parameter $(1 - a^{-1})$ and $(1 - b^{-1})$ respectively,
a direct computation gives:
$$\Cov(X, \min(X,X')) = \frac{ab}{(ab-1)^2}.$$
Applying this to our setting, we get:
$$\Cov(Z_i, Z_{i'}) = \frac{p^{e(i,i')}}{(p^{e(i,i')}-1)^2}$$
where $e(i,i') = \min(\tau(i), \tau(i')) = \tau(\min(i,i'))$. Summing
up these contributions, we get the equality in~(2). The inequalities
are derived from this similarly to what we have done in~(1).

Let $(Z_i)_{i \geq 0}$ be a countable family 
of independant random variable having all geometric distribution of 
parameter $(1 - p^{-1})$. We set $Z = \sum_{i=1}^\infty \min(Z_1, \ldots, 
Z_i)$. Cleary $V_j \leq Z$ and it is then enough to prove:
$$\P[Z \geq m] \leq p^{-m + O(\sqrt m)}.$$
We introduce the event $E_m$ formulated as follows:
\emph{there exists a partition $(m_1, \ldots, m_\ell)$ of $m$ such that $X_i 
\geq m_i$ for all $i \leq \ell$}.
Up to a neglectable subset, $E_m$ contains the event $\{ Z \geq m\}$.
Thus $\P[Z \geq m] \leq \P[E_m]$. Furthermore, the probability of $E_m$
is bounded from above by $\sum \prod_{i=1}^\ell \P[X_1 \geq m_i]$ where
the sum runs over all partitions $(m_1, \ldots, m_\ell)$ of $m$. 
Replacing $\P[X_1 \geq m_i]$ by its value $p^{-m_i}$, we get:
$$\P[E_m] \leq p(m) \cdot p^{-m}$$
where $p(m)$ denotes the number of partitions of $m$. By a famous 
formula \cite{andrews}, we know that $\log p(m)$ is equivalent to $\pi 
\sqrt{2m/3}$. In particular it is in $p^{O(\sqrt m)}$ and~(3) is proved.

We now derive (4) by a standard argument. It follows from~(3) that:
$$\P[\max(V_0, \ldots, V_{d-1})] \leq d \cdot p^{-m + c\sqrt m}$$
for some constant $c$. Therefore:
$$\E[\max(V_0, \ldots, V_{d-1}] \leq \sum_{m=1}^\infty \min(1,
d \cdot p^{-m + c\sqrt m}).$$
Let $m_0$ denote the smallest index such that $d \cdot p^{-m_0 + c\sqrt 
m_0}$, \emph{i.e.} $m_0 - c \sqrt{m_0} \geq \log_p d$. Solving the latest
equation, we get $m_0 = \log_p + O(\sqrt{\log_p d})$. Moreover
$\sum_{m=m_0}^\infty d\: p^{-m + c\sqrt m}$ is bounded independantly of
$d$. The result follows.
\end{proof}

\begin{cor}
\label{cor:Wj}
For all $j \in \{0, \ldots, d-1\}$, we have:
$$\E[W_j] \leq (1 - p^{-1}) \cdot \frac 1 {p^j - 1}.$$
\end{cor}

\begin{proof}
Writing $\E[W_j] = \sum_{m=1}^\infty \P[W_j \geq m]$, the corollary
follows directly from Theorem \ref{th:estimWj}.
\end{proof}

\subsection{Proof of Theorems \ref{th:lawVj} and \ref{th:estimWj}}
\label{subsec:proof}

In what follows, we use the notation $\lc(P)$ fot the leading 
coefficient of a polynomial $P$. During the proof, $A$ and $B$ will 
always refer to monic polynomials of degree $d$ and $R_j$ (resp. $U_j$ 
and $V_j$) denote their $j$-th subresultant (resp. their $j$-th 
cofactors). If $P$ is a polynomial and $n$ is a positive integer, we 
use the notation $P[n]$ to refer to the coefficient of $X^n$ in $P$.
We set $r_j = R_j[j]$.

\medskip

\noindent
\textbf{Preliminaries on subresultants.}
We collect here various useful relations between subresultants and 
cofactors. During all these preliminaries, we work over an arbitrary
base $\ring$.

\begin{prop}
\label{prop:relations}
The following relations holds:
\begin{itemize}
\item $U_{j-1} V_j - U_j V_{j-1} = r_j^2$;
\item $U_j[d{-}j{-}1] = -V_j[d{-}j{-}1] = (-1)^j r_{j+1}$;
\item $\Res^{j,j-1}_k(R_j, R_{j-1}) = r_j^{2(j-k-1)} \: R_k$ for $k < j$;
\item $\Res^{d-j,d-j-1}_k(U_{j-1}, U_j) = r_j^{2(d-j-k-1)} \: U_{d-1-k}$\\
for $k < d-j$.
\end{itemize}
Moreover $r_j$ depends only on the $2j$ coefficients of highest 
degree of $A$ and $B$.
\end{prop}

\begin{proof}
By functoriality of subresultants, we may assume that $\ring = 
\Z[a_0, \ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$
are the two generic monic polynomials of degree $d$:
$$A = X^d + \sum_{i=0} a_i X^i \quad \text{and} \quad
B = X^d + \sum_{i=0} b_i X^i.$$
All relations then follows from the extended subresultants 
pseudo-remainder sequence algorithm (noting in addition that all
scalar subresultants are nonzero).
\end{proof}

For any fixed index $j \in \{1, \ldots, d-1\}$, we consider the 
function $\psi_j$ that takes a couple $(A,B) \in \ring_d[X]^2$ to the
quadruple $(U_j, U_{j-1}, R_j, R_{j-1})$. It follows from Proposition
\ref{prop:relations} that $\psi_j$ takes its values in the subset 
$\mathcal E_j$ of
$$\big(\ring_{\leq d-j-1}[X]\big) \times \big(\ring_{\leq d-j}[X]\big) \times 
\big(\ring_{\leq j}[X]\big) \times \big(\ring_{\leq j-1}[X]\big)$$
gathering the quadruples $(\mathcal U_j, \mathcal U_{j-1}, \mathcal
R_j, \mathcal R_{j-1})$ satisfying the condition:
$$\begin{array}{ll}
&\mathcal U_{j-1}[d{-}j] = (-1)^{j-1} \: \mathcal R_j[j] \smallskip \\
\text{and} &
\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) = -\mathcal R_j[j]^{2(d-j-1)}.
\end{array}$$
Let $\mathcal E_j^\times$ be the subset of $\mathcal E_j$ defined by 
requiring that $\mathcal R_j[j]$ is invertible in $\ring$. In the same 
way, we define $\Omega_j^\times$ as the subset of $\ring_d[X]^2$ 
consisting of couples $(A,B)$ whose $j$-th scalar subresultants (in 
degree $(d,d)$) is invertible in $\ring$.

\begin{prop}
\label{prop:bijection}
With the above notations, the function $\psi_j$ induces a bijection between
$\Omega_j^\times$ and $\mathcal E_j^\times$.
\end{prop}

\begin{proof} 
We are going to define the inverse of $\psi_j$. We fix a quadruple 
$(\mathcal U_j, \mathcal U_{j-1}, \mathcal R_j, \mathcal R_{j-1})$ in 
$\mathcal E_j^\times$ and set $a = \mathcal R_j[j]$.
Let $\mathcal W_j$ and $\mathcal W_{j-1}$ denote 
the $j$-th cofactors of $(\mathcal U_{j-1}, \mathcal U_j)$ in degree 
$(d{-}j, d{-}j{-}1)$. Define $\mathcal V_j = \alpha \mathcal W_j$ and 
$\mathcal V_{j-1} = -\alpha \mathcal W_{j-1}$ where $\alpha =
a^{4j-4d+6}$. The relation:
\begin{equation}
\label{eq:missingcof}
\mathcal U_{j-1} \mathcal V_j - 
\mathcal U_j \mathcal V_{j-1} = a^2.
\end{equation}
then holds. We now define $A$ and $B$ using the formulae:
\begin{equation}
\label{eq:systemAB}
\left\{\begin{array}{l}
A = \todo{\text{something}} \\
B = \todo{\text{something}} \\
\end{array}\right.
\end{equation}
and let $\varphi_j$ be the function mapping $(\mathcal U_j, \mathcal 
U_{j-1}, \mathcal R_j, \mathcal R_{j-1})$ to $(A,B)$. The composite 
$\varphi_j \circ \psi_j$ is easily checked to be the identity: indeed, 
if $\psi_j(A,B) = (\mathcal U_j, \mathcal U_{j-1}, \mathcal R_j, 
\mathcal R_{j-1})$, the relation~\eqref{eq:missingcof} implies that 
$\mathcal V_{j-1}$ and $\mathcal V_j$ are the missing cofactors and, 
consequently, $A$ and $B$ have to be given by the 
system~\eqref{eq:systemAB}.

To conclude the proof, it remains to prove that the composite in 
the other direction $\psi_j \circ \varphi_j$ is the identity as well. 
Since both $\varphi_j$ and $\psi_j$ are componant-wise given by 
polynomials, we can use functoriality and assume that $\ring$ is 
the field $\Q(c_0, c_1, \ldots, c_n)$ (with $n = 2d$) and that 
each variable $c_i$ corresponds to one coefficient of $\mathcal U_j$, 
$\mathcal U_{j-1}$, $\mathcal R_j$ and $\mathcal R_{j-1}$ with the
convention that $c_0$ (resp. $(-1)^{j-1} c_0$) is used for the leading
coefficients of $\mathcal R_j$ (resp. $\mathcal U_{j-1}$). Set:
$$\begin{array}{ll}
& (A,B) = \varphi_j(\mathcal U_j, \mathcal U_{j-1}, 
\mathcal R_j, \mathcal R_{j-1}) \smallskip \\
\text{and} & (U_j, U_{j-1}, R_j, R_{j-1}) = \psi_j(A,B).
\end{array}$$
Since $\ring$ is a field and $\mathcal R_j[j]$ does not vanish, the 
Sylvester mapping
$$\begin{array}{rcl}
\ring_{<d-j}[X] \times \ring_{<d-j}[X] & \to & 
\ring_{<2d-j}[X]/\ring_{<j}[X] \smallskip \\
(U,V) & \mapsto & AU + BV
\end{array}$$
has to be bijective. Therefore there must exist $\lambda \in \ring$ 
such that $\mathcal R_j = \lambda \cdot R_j$ and $\mathcal U_j = \lambda 
\cdot U_j$. Similarly $(\mathcal R_{j-1}, \mathcal U_{j-1}) = \mu \cdot
(R_{j-1}, U_{j-1})$ for some $\mu \in \ring$. Identifying the leadings 
coefficients, we get $\lambda = \mu$. Writing
$\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) = 
\Res^{d-j,d-j-1}(U_{j-1}, U_j)$,
we get $\lambda^{2(d-j)-1} = 1$. Since the exponent is odd, this implies 
$\lambda = 1$ and we are done.
\end{proof}

\noindent 
\textbf{The distribution in the residue field.}
We assume in this paragraph that $\ring$ is a finite field and we equip 
$\Omega_\ring = \ring_d[X]^2$ with the uniform distribution. Just as 
before, we consider each scalar subresultant $r_j$ as a random variable 
on $\Omega_\ring$ taking its values in $\ring$.

\begin{prop}
\label{prop:distribk}
Assuming that $\ring$ is a finite field, the tuple $(r_0, \ldots, 
r_{d-1})$ is uniformly distributed in $\ring^d$.
\end{prop}

\begin{proof}
\end{proof}

\begin{rem}
In the next paragraph, we will use Proposition \ref{prop:distribk} 
with $\ring = k$ (the residue field of $\A$). As for now, we can 
notice that it implies (using functoriality of subresultants) that
the variables $\bar V_j = \mathbbm{1}_{\{V_j = 0\}}$ follows the 
Bernoulli distribution of parameter $(1 - p^{-1})$, in accordance
with Theorem \ref{th:lawVj}. Proposition \ref{prop:distribk} 
furthermore imply that the $\bar V_j$'s are mutually independant.
\end{rem}

\noindent
\textbf{End of the proof.}

\section{The method of adaptative lifts}

We have seen in the previous sections that Euclide's algorithm (and 
their variants) are very unstable in practice. On the other hand, 
one can compute subresultants in a very stable way by evaluating
the corresponding minors of the Sylvester matrix. Doing so, we do
not loose any significant digit. Of course, the downside is the
quite bad efficiency.

In this section, we design an algorithm which combines the two
advantages: it has the same complexity than Euclide's algorithm
and it is very stable in the sense that it does not loose any
significant digit. 
This algorithm is deduced from the subresultant pseudo-remainder
sequence algorithm by applying a ``stabilization process'', whose 
inspiration comes from \cite{}.

\subsection{Crash course on ultrametric precision}

In this subsection, we briefly report on the results of \cite{}. In this 
paper, the authors draw the lines of a general framework to handle a 
sharp (often optimal) track of ultrametric precision. In order to do so, 
they proposed to attach a single precision datum --- which is a lattice 
in a huge vector space --- to the set of all variables, instead of 
attaching a precision datum to each individual variable as we do 
usually. This proposal is supported by the following lemma in ultrametric
analysis.

\begin{lem}
\label{lem:padicprec}
Let $f : K^n \to K^m$ be a strictly differentiable function. Let $x$
be a point in $K^n$ at which the differential of $f$ is surjective.
Then, for all sufficiently large $N$, the following equality holds:
\begin{equation}
\label{eq:padicprec}
f(x + p^N \A^n) = f(x) + f'(x)(p^N \A^n)
\end{equation}
where $f'(x)$ stands for the differential of $f$ at $x$.
\end{lem}

An important feature of Lemma \ref{lem:padicprec} is that 
Eq.~\eqref{eq:padicprec} is an equality, and not only an inclusion. 
This reflects the optimality of the method and, as we shall see later, 
is also the origin of the ``stabilization process'' mentionned above.
Moreover, if the function $f$ appearing in Lemma \ref{lem:padicprec} 
has more regularity, one can exhibit a lower bound on $N$ ensuring
that Eq.~\eqref{eq:padicprec} holds. In \cite{}, the case of locally 
analytic functions is considered. For our purpose, it will be enough 
to restrict ourselves to \emph{integral polynomial} functions, that
are functions of the shape:
$$f(x) = \big(F_1(x), \ldots, F_m(x)\big)
\quad \text{with } x = (x_1, \ldots, x_n)$$
where $F_1, \ldots, F_m$ are multivariate polynomials in $n$ variables
with coefficients in $\A$. Under this assumption, we have:

\begin{prop}
\label{prop:padicprec}
Let $f : K^n \to K^m$ be an integral polynomial function and $x \in
\A^n$. Let $v$ be an integer such that $f'(x)(\A^n) \supset p^v \A^m$.
Then, for all $N > v$, Eq.~\eqref{eq:padicprec} holds.
\end{prop}

\begin{proof}
It is a direct corollary of \cite[Proposition 3.12]{}.
\end{proof}

\subsection{The stabilization process}

In order to get a stabilized version of the subresultant 
pseudo-remainder sequence algorithm, we are going to estimate 
the loss --- and the gain --- of precision at each step of this 
algorithm using Proposition \ref{prop:padicprec}. In order to do
so, we introduce the functions
$$\begin{array}{rcl}
f_j \, : \, K_d[X] \times K_d[X] & \to & K_{\leq j}[X] \times 
K_{\leq j-1}[X] \smallskip \\
(A,B) & \mapsto & (\Res_j(A,B), \Res_{j-1}(A,B)).
\end{array}$$

\noindent
{\bf Differential computation.}
The first step consists in estimating the differential of $f_j$. 
Set $\mathcal L = \A_d[X] \times \A_d[X]$ and $\mathcal L_j = 
\A_{\leq j}[X] \times \A_{\leq j-1}[X]$; they are standard lattices
in the source and the target of $f_j$ respectively.

\begin{lem}
\label{lem:diff}
For all $(A,B) \in K_d[X]^2$, we have:
$$r_j^2 \cdot \mathcal L_j 
\subset f'_j(A,B)(\mathcal L) \subset \mathcal L_j$$
where $r_j$ is the $j$-th scalar subresultant of $(A,B)$.
\end{lem}

\begin{proof}
The second inclusion is clear because $f_j$ is a polynomial function. 
Let us prove the first inclusion. One may of course assume that $r_j$ 
does not vanish, otherwise there is nothing to prove. Now, we remark 
that $f_j$ factors through the function $\psi_j$ introduced in \S 
\ref{subsec:proof}. By continuity, the $j$-th scalar subresultant 
function does not vanish on a neighborhood of $(A,B)$. By Proposition 
\ref{prop:bijection}, $\psi_j$ is injective on this neighborhood. 
Therefore so is $f_j$. Furthermore, a close look at the proof of 
Proposition \ref{prop:bijection} indicates that a left inverse of $f_i$ 
is:
$$(S_j, S_{j-1}) \mapsto \todo{\text{something}}$$
where $U_j, V_j$ (resp. $U_{j-1}$, $V_{j-1}$) are the $j$-th (resp 
$(j-1)$-th) cofactors of $(A,B)$. Differenting this, we get the
announced result.
\end{proof}

\noindent
{\bf The algorithm.}
The input of our algorithm is a couple of polynomials $(A,B)$ with 
coefficients in $\A$ whose coefficients are known at precision 
$O(\pi^N)$ for some given positive integer $N$. We make the following 
hypothesis (and refer to \S \ref{subsec:questions} for a brief 
discussion about it):

\medskip

\begin{tabular}{rl}
& (a)~$A$ and $B$ are monic of the same degree $d$ \\
{\bf (H)}: & (b)~all scalar subresultants of $(A,B)$ \\
& \phantom{(b)~}do not vanish modulo $\pi^{N/2}$.
\end{tabular}

\medskip

\noindent
The subresultants $R_j = \Res_j(A,B)$ can then be computed recursively 
using the decreasing recurrence:
$$\begin{array}{l}
R_d = B, \, R_{d-1} = A-B \\
R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})
\end{array}$$
Moreover Lemma \ref{lem:diff} and Proposition \ref{prop:padicprec} imply
that any perturbation of $(R_j, R_{j-1})$ by an element of $p^N r_j^2 
\mathcal L_j$ will not affect the values of the next $R_k$'s modulo 
$\pi^N$. Indeed, given $k > j$, let $f_{j,k} : (R_j, R_{j-1}) \mapsto 
(R_k, R_{k+1})$ be the function modeling the above recurrence. We have
$f_k = f_{j,k} \circ f_j$ and thus we can write:
\begin{align*}
f_{j,k}\big((R_j, R_{j-1}) + \pi^N r_j^2 \mathcal L_j\big) 
& \subset f_k\big((A,B) + \pi^N \mathcal L\big) \\
& \subset (R_k, R_{k+1}) + \pi^N \mathcal L_k.
\end{align*}
This suggests the following algorithm.

\noindent\hrulefill

\noindent {\bf Algorithm 1:} {\tt subresultants\_stable}

\noindent{\bf Input:} an integer $N$, a couple $(A,B)$ satisfying {\bf (H)}

\noindent{\bf Output:} the subresultants of $(A,B)$ at precision $O(\pi^N)$

\smallskip\noindent 1.\ Set $R_d = B$, $v_d = 0$ and $R_{d-1} = A-B$

\smallskip\noindent 2.\ {\bf for} $j=d,\dots,2$

\smallskip\noindent 3.\ \hspace{0.3cm}{\bf compute} $v_{j-1} = \val(\lc(R_{j-1}))$

\smallskip\noindent 4.\ \hspace{0.3cm}{\bf lift} $(R_j, R_{j-1})$ at precision $O(\pi^{N+2(v_j + v_{j-1})})$

\smallskip\noindent 5.\ \hspace{0.3cm}{\bf compute} $R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})$\\
\phantom{6.\ }\hspace{0.3cm}at precision $O(\pi^{N+2 v_{j-1}})$

\smallskip\noindent 6.\ {\bf return} $R_{d-1} + O(\pi^N), \ldots,
R_0 + O(\pi^N)$

\vspace{-1ex}\noindent\hrulefill

\begin{prop}
Algorithm 1 is correct and runs in
$$O(d^2 \cdot \M(N + \max(V_0, \ldots, V_{d-1}))$$
bit operations where $V_j$ denotes the valuation of $r_j$ and 
$\M(n)$ is the number of bit operations needed to perform an 
arithmetic operation (addition, product) in $\A$ at precision 
$O(\pi^n)$.
\end{prop}

\begin{rem}
\label{rem:M}
In all usual examples ($p$-adic numbers, Laurent series), one can 
choose $\M(n)$ is quasi-linear in $n$ and the size of the residue
field $k$.
\end{rem}

\begin{proof}
Regarding to the correctness, it remains only to justify that $R_{j-2}$ 
can be computed at the given precision in line 5 but this follows 
from a naive track of precision (\emph{cf} proof of Proposition
\ref{prop:precEuclide} for a similar computation).
As usual Euclide's algorithm, Algorithm 1 requires $O(d^2)$ operations 
in the base ring $\A$. Moreover, we observe that the maximal precision 
at which we are computing is upper bounded by $N + 2 \max(V_0, \ldots, 
V_{d-1})$. The proposition follows.
\end{proof}

According to Corollary \ref{cor:Vj}, the expected value of the variable 
$\max(V_0, \ldots, V_{d-1})$ is in $O(\log_p d)$. Thus, the average 
complexity of Algorithm 1 is $O(d^2 \cdot \M(N + \log d))$ bit 
operations. In all usual cases (\emph{cf} Remark \ref{rem:M}), this
complexity is also $\tilde O(d^2 N \cdot \log |k|)$ bit operations.

\subsection{Open questions}
\label{subsec:questions}

Of course, the first question we would like to raise is to possibility 
to drop the hypothesis {\bf (H)}. It is actually rather easy to relax
the part~(a) of the hypothesis, replacing it by:

\medskip

(a')~$\lc(B)$ is invertible in $\A$ and $\deg A \geq \deg B$.

\medskip

\noindent
Assuming only this weaker assumption, Algorithm 1 still works except 
that, in line 1, we have to set
$$R_{d-1} = (-1)^{\deg A - \deg B} (A \,\%\, B)$$
(instead of $R_{d-1} = A-B$).
We remark that this value can be computed at precision $O(\pi^N)$
because the leading coefficient of $B$ is assume to be invertible.
Dropping entirely~(a') and~(b) seems nevertheless to be more difficult. 
We may however remark that~(b) is satisfied with high probability if
$N$ is large compared to $2 \cdot \log_d p$. Thus, replacing eventually
$N$ by $3 \cdot \log_d p$ (which does not affect the complexity), the 
assumption~(b) is armless in average. Nonetheless, it may cause
important difficulties on particularly bad instances.

Beyond this, one may wonder if one can use similar technics to compute 
not only subresultants but cofactors as well. For those indexes $j$ such 
that $r_j$ is invertible in $\A$, the same analysis applies almost 
\emph{verbatim}. However for other indexes $j$, the differential 
computation seems to be much more subtle. One can get around this issue 
by using lifting technics only when $r_j$ is invertible and tracking 
precision naively otherwise: it is possible to get this way a stable 
algorithm whose average running time is acceptable but which seems to be 
bad in the worst case. Can we do better?

Another quite interesting question is those of designing an algorithm 
which combines the precision technology developed in this paper with the 
``half-\textsc{gcd}'' methods. It is actually closely related to the 
previous question because ``half-\textsc{gcd}'' methods make an 
intensive use of cofactors in order to speed up the computation.

\section{Towards p-adic floats}

When computing with real numbers, computers very often use floating 
point arithmetic. 
The rough idea of this model consists in representating all real numbers 
using the same number of digits (the so-called \emph{precision}) and to
apply rounding heuristics when final digits are unsettled.
In comparison with arithmetic interval, floating point arithmetic has
two main advantages. First, it allows simple and fast implementations.
Second, experiments show the results obtained this way have generally
more much correct digits that those predicted by arithmetic interval.
The counterpart is that, expect on small examples, obtaining proved
results is generally intractable.

In the $p$-adic setting, the analogue of floating point arithmetic has
not been developed yet. One reason for this is probably the well-known
saying: ``in the $p$-adic world, rounding errors do not accumulate''.
Consequently one might expect that interval arithmetic would provide
sharp results.
Nonetheless it already appears on basic examples that this hope is
failing. For instance:
\begin{enumerate}[a)]
\vspace{-0.25cm}
\item in this paper, we showed that Euclide-like algorithms for 
computing resultants loose a number of digits which is fairly linear in 
the degree whereas no loss of precision is expected (because resultants
can be computed by evaluating determinants);
\vspace{-0.25cm}
\item in \cite{}, it is shown that a similar phenomenom occurs for
the computation of the SOMOS 4 sequence;
\vspace{-0.25cm}
\item in \cite{}, a stable algorithm is presented for computing the LU 
factorization of a matrix: it looses a number of digits which is linear 
in $\log d$ (where $d$ is the dimension of the input matrix) whereas 
standard Gauss elimination looses a number of digits which is linear in 
$d$;
\vspace{-0.25cm}
\item in \cite{}, the authors studied certain $p$-adic differential
equations and designed specific stable algorithms to solve them after
having noticed that usual (naive) technics may loose much more digits 
that needed.
\end{enumerate}
\vspace{-0.2cm}
All these examples suggest that interval arithmetic is not as good as 
one might have expected at first. Therefore, it probably makes sense 
to seriously study the analogue of floating point arithmetic in a
ultrametric context.

\medskip

Let us describe quickly what might be this analogue and what are its
advantages and disadvantages. We keep the notations of previous
sections: the letter $\A$ denotes a complete discrete valuation ring
with uniformizer $\pi$ and $K$ is its fraction field.
In the model of ultrametric floating point arithmetic, we fix a
positive integer $N$ (the \emph{precision}) and represent elements 
of $K$ by approximations of the form:
\begin{equation}
\label{eq:padicfloat}
\pi^e \cdot \sum_{i=0}^{N-1} x_i \pi^i
\end{equation}
where $e$ is a relative integer and the $x_i$'s are elements of a fixed 
set of representatives of $\A$ modulo $\pi$ with the convention that the 
representative of $0 \in k$ is $0 \in \A$.
We further assume that $x_0 \neq 0$, \emph{i.e.} $e$ is the valuation
of the sum \eqref{eq:padicfloat}. We see that this framework is quite
similar to usual floating point arithmetics: the integer $e$ plays the
role of exponent, the uniformizer $\pi$ plays the role of the basis
and the value $\sum_{i=0}^{N-1} x_i \pi^i$ plays the role of the
significand (the mantissa). It remains to define operations $\oplus$
and $\odot$ on approximations modeling addition and multiplication
on $K$ respectively. We do this as follows: given $x$ and $y$ two
elements of $K$ of the form Eq.~\eqref{eq:padicfloat}, we compute
$x+y$ (resp. $xy$) in $K$, expand it as a convergent series 
$\sum_{i=v}^{\infty} s_i \pi^i$ (with $s_v \neq 0$) and define $x 
\oplus y$ (resp. $x \odot y$) by truncating the series at $i = v+N$.

Similarly to real floating point arithmetic, the main advantages of 
ultrametric floating point arithmetic are the simplicity and the 
efficiency and the counterpart is the difficulty to get proved results. 
Moreover, the aforementioned examples are evidences that ultrametric 
floating point arithmetic may often compute much more correct digits 
than those predicted by an analysis based on interval arithmetic. In 
order to illustrate this last assertion, let us go back to the case of 
resultants discussed earlier in this paper. Let $A$ and $B$ be two monic 
polynomials of degree $d$ (picked at random) whose coefficients are all 
known at precision $O(\pi^N)$. We have proved that if we are using the
model of interval arithmetic, then the subresultant pseudo-remainder
sequence algorithm will output $\Res(A,B)$ at precision 
$O(\pi^{N-N_{\text{int}}})$ where $N_{\text{int}}$ grows linearly 
with respect to $d$ in average. On the other hand, if we are using 
ultrametric floating point arithmetic, then the same algorithm will
output $\Res(A,B)$ at precision $O(\pi^{N-N_{\text{float}}})$ where 
$N_{\text{float}}$ grows linearly with respect to $\log d$ in average.
We emphasize furthermore that this result is \emph{proved}! From this
point of view, floating point arithmetics seems to behave better in
the ultrametric setting: we may hope to get proved results relatively
cheaply.

\begin{thebibliography}{99}
\bibitem[a]{b}
  Blah
\end{thebibliography}

\end{document}

### 2014-11-26 examples/examples.tex

@@
 We now pick two real numbers $a$ and $b$ satisfying \eqref{eq:condAB}.
-Set $d = \Lambda(h)(a)$. Going back to the definitions, we remark that
+Set $d = \Lambda(h)(a)$. Going back to the definitions of $\Lambda (h)$ and Legendre transform, we remark that
 the above equality implies:

### 2014-12-03 examples/stability.tex
Paper reorganized and introduction (first version) added

--- 2014-11-26 examples/examples.tex
@@
 \documentclass{sig-alternate}
-%\documentclass{lms}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{tikz}
-\renewcommand*{\theHsection}{\thesection}
 
@@
 
-\title{Tracking p-adic precision: examples}
+\title{About p-adic stability}
 
@@
 \begin{abstract}
+Based on the framework developed in \cite{caruso-roe-vaccon:14a}, we study the 
+$p$-adic stability of many classical algorithms involving many usual
+algebraic objects (matrices, univariate and multivariate polynomials,
+vector spaces). We observe in particular that generic algorithms ---
+which are designed to work over an arbitrary field --- are generally
+very unstable.
 \end{abstract}
@@
  \noindent
- {\bf Keywords:} 
+ {\bf Keywords:} $p$-adic precision
 \medskip
@@
 
-\section{Preliminaries}
-
-In the next section, we will see that the derivative at some point $x$ 
-of many standard operations, modelised by a function $f$, has a simple 
-expression in term of $x$ and $f(x)$. In other terms, such functions $f$
-satisfy:
-\begin{equation}
-\label{eq:diffequa}
-f' = g \circ (f, \id)
-\end{equation}
-where $g$ is a given --- and hopefully rather simple --- function. 
-The aim of these preliminaries is to study the differential equation
-\eqref{eq:diffequa} (or, actually, a slight generalization of it) and
-to derive from it some bounds on the growing function $\Lambda(f)$ 
-defined in \cite{padicprec}.
-
-\emph{In all this section, we assume that \textbf{the base field $K$ 
-has characteristic $0$}.}
-
-\begin{comment}
-
-\subsection{Legendre transform of a compositum}
-
-For $i \in \{1,2\}$, let $\varphi : \R \to \R \cap \{+\infty\}$ denote a 
-convex function and $\varphi_i^\star$ be its Legendre transform defined 
-by:
-\begin{equation}
-\label{eq:defLegendre}
-\varphi_i^\star(u) = \sup_{v \in \R} \varphi(v) - uv.
-\end{equation}
-Set $\varphi = \varphi_1 \circ \varphi_2$. The aim of this subsection is 
-to relate $\varphi^\star$ with $\varphi_1^\star$ and $\varphi_2^\star$.
-
-\begin{lem}
-\label{lem:Legendrecomp}
-We assume that $\varphi_1$ and $\varphi_2$ are nondecreasing and that 
-their slopes are all in $\N$. Then, for all nonnegative interger $s$:
-$$\varphi^\star(s) =
-\inf_{\ell, (n_i)} \varphi_1^\star(\ell) + \varphi_2^\star(n_1) + 
-\cdots + \varphi_2^\star(n_\ell)$$
-where the infimum is taken over all finite sequences of nonnegative
-integers (of variable length) $n_1, \ldots, n_\ell$ with $n_1 + \cdots +
-n_\ell = s$.
-\end{lem}
+For about twenty years, the use of $p$-adic methods in symbolic 
+computation was gaining more and more popularity: they were succesfully 
+used for instance 
+to compute compose products of polynomials 
+\cite{boston-gonzalez-perdry-schost:05a}, 
+to produce hyperelliptic curves of genus $2$ with complex multiplication 
+\cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a},
+to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
+and, last but not least,
+to count points on certain varieties (mainly curves) using $p$-adic 
+cohomologies (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
+followers).
+However, a general framework allowing a precise study of $p$-adic 
+precision --- which is a main issue we encounter when trying to work 
+with $p$-adic numbers --- was designed only recently in 
+\cite{caruso-roe-vaccon:14a}.
+
+The present paper, which may be considered as a continuation of 
+\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. cit.} in 
+order to analyse the $p$-adic stability of many standard algorithms
+for computing with all basic algebraic structures that are numbers,
+matrices, univariate and multivariate polynomials, series and vector
+spaces. For each considered algorithm, we will follow the same protocol: 
+first, we will compute the theoretical loss of precision of the 
+underlying question using the framework of \cite{caruso-roe-vaccon:14a} and then 
+compare it to those observed in available implementations in 
+\textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
+\textsc{sage} \cite{sage}.
+
+The conclusion of our analysis is that generic algorithms which are 
+designed to work over an arbitrary field are often quite unstable. Very 
+roughly, denoting by $D$ the number of divisions performed by the 
+algorithm under analysis and by $q$ the cardinality of the residue 
+field, it appears that the numbers of lost digits is about $O(\frac D 
+q)$ whereas the theoretical loss of precision is closer than $O(\log_q 
+D)$ --- and sometimes even $O(1)$. We would like to underline in 
+particular that the instability phenomema are much more apparent when 
+the residue field is small. It is the reason why we will always take
+$K = \Q_2$ is our examples.
+
+\medskip
+
+\noindent
+{\bf Organization of the paper.}
+The section \ref{sec:theory} introduces the theory of ultrametric 
+precision developped in \cite{caruso-roe-vaccon:14a} and completes it on a technical 
+point. The analysis of standard algorithms are treated in the next
+sections: \S\S \ref{sec:numbers}--\ref{sec:vectorspaces} are 
+successively devoted to the case of numbers, matrices, univariate 
+polynomials, multivariate polynomials and finally vector spaces.
+
+\medskip
+
+\noindent
+{\bf Notations.}
+Throughout the paper, the letter $K$ will refer to a \emph{complete 
+discrete valuation field of characteristic $0$}, that is either a finite 
+extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
+series over a field of characteristic $0$. We denote by $\val : K \to \Z 
+\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
+and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
+$\Vert \cdot \Vert$ the corresponding norm.
+
+\section{The theory of p-adic precision}
+\label{sec:theory}
+
+\subsection{The main lemma and its consequences}
+
+\todo{Recall results of \cite{caruso-roe-vaccon:14a}.}
+
+{\color{Bittersweet}
+
+We saw in the core of the article that the differential of an operation
+encodes the intrinsic loss/gain of precision when performing this
+operation. The aim of this appendix is to compute the differential of
+many common operations on numbers, polynomials and matrices. Surprisingly,
+we observe that all differentials we will consider are rather easy to
+compute even if the underlying operation is quite involved (\emph{e.g.}
+Gr\"obner basis).
+
+In what follows, we use freely the ``method of physicists'' to compute
+differentials: given a function $f$ differentiable at some point $x$, we
+consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$
+by expanding LHS and neglecting terms of order $2$. The differential of
+$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
+
+}
+
+\subsection{A bound on a growing function}
+\label{ssec:boundLambdaf}
+
+In the next section, we will compute the derivative of many standard 
+operations and remark that it often has a simple expression in term of 
+the input and the output. In other terms, the function $f$ modeling
+such an operation satisfies a differential equation of the form
+$f' = g \circ (f, \id)$
+where $g$ is a given --- and hopefully rather simple --- function. The 
+aim of this subsection is to study this differential equation and to 
+derive from it some bounds on the growing function $\Lambda(f)$.
+We shall actually work with the slight more general differential 
+equation:
+\begin{equation}
+\label{eq:diffequah}
+f' = g \circ (f, h).
+\end{equation}
+Here $g$ and $h$ are known locally analytic functions and $f$ is the 
+unknown, which is assumed to be locally analytic as well. The function
+$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
+and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
+is again an open subset in a $K$-Banach space $G$. Consequently, $g$
+goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
+all continuous linear applications $E \to F$.
+In what follows, we always assume that $V$ and $W$ contains the origin, 
+$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
+for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
+accordingly) so that they both vanish at $0$ and second, in order to 
+apply the theory of \cite{caruso-roe-vaccon:14a}, the derivative $f'(0)$ needs to be 
+surjective and therefore \emph{a fortiori} nonzero.
+
+We assume that we are given in addition two nondecreasing convex 
+functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
+\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We assume in addition that 
+there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
+$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
+we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
+does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
+defined by:
+$$\begin{array}{rll}
+&\tau_\nu(x) = x & \text{if } x \leq \nu \\
+& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
+\text{and} &
+\multicolumn{2}{l}{\Lambda_f(x) = 
+  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
+\end{array}$$
+where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
+e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is 
+$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$ 
+where $p$ denotes the characteristic of the residue field and where, by 
+convention, the above expression equals $0$ if $p = 0$ (we recall that 
+we have assumed that $K$ has characteristic $0$).
+
+\begin{prop}
+\label{prop:boundLambdaf}\label{PROP:BOUNDLAMBDAF}
+We have $\Lambda(f) \leq \Lambda_f$.
+\end{prop}
 
 \begin{proof}
-Let $n_1, \ldots, n_\ell$ be a sequence as above. Taking the supremum 
-in $v$ in the relation:
-$$\varphi(v) - s v =
-\varphi_1(\varphi_2(v)) - \ell \varphi_2(v) + \sum_{i=1}^\ell
-(\varphi_2(v) - n_i v)$$
-we get:
-\begin{equation}
-\label{eq:inegvarphi}
-\varphi^\star(s) \leq \varphi_1^\star(\ell) + 
-\varphi_2^\star(n_1) + \cdots + \varphi_2^\star(n_\ell).
-\end{equation}
-It remains then to prove that one can choose $\ell$ and $n_1, \ldots,
-n_\ell$ such that the inequality above is actually an equality.
-
-Let $x$ be the unique value such that the left (resp. right) derivative 
-of $\varphi$ at $x$ is less than (resp. greater than) or equal to $s$. 
-Let $\ell^-$ (resp. $\ell^+$) denote the left (resp. right) derivative 
-of $\varphi_1$ at $\varphi_2(x)$ and $n^-$ (resp. $n^+$) denote the left 
-(resp. right) derivative of $\varphi_2$ at $x$. On the one hand, $\ell^- 
-n^- \leq s \leq \ell^+ n^+$ since $\ell^- n^-$ (resp. $\ell^+ n^+$) 
-equals the left (resp. right) derivative of $\varphi$ at $x$). On the 
-other hand:
-$$\begin{array}{ll}
-\varphi_1^\star(u) = \varphi(x) - u \varphi_2(x) & 
-\text{for all } u \in [\ell^-,\, \ell^+] \\
-\varphi_2^\star(u) = \varphi_2(x) - u x & 
-\text{for all } u \in [n^-,\, n^+] \\
-\varphi^\star(u) = \varphi(x) - u x & 
-\text{for all } u \in [\ell^- n^-,\, \ell^+ n^+] \\
-\end{array}$$
-Therefore, the inequality \eqref{eq:inegvarphi} is an equality for all 
-integer $\ell \in [\ell^-, \ell^+]$ and all sequences of integers 
-$(n_i)_{1 \leq i \leq \ell}$ with $n_i \in [n^-, n^+]$ for all $i$ and 
-$n_1 + \cdots + n_\ell = s$. It is then enough to prove that such a 
-couple $(\ell, (n_i))$ exists but it is clear.
+See Appendix \ref{app:proof}.
 \end{proof}
 
-\end{comment}
+\begin{figure}
+\null \hfill
+\begin{tikzpicture}[scale=0.8]
+\draw[thick,green!30,fill=green!20] 
+   (-2,-1)--(-1,2)--(-1,3)--(3.5,3)
+ --(2.5,2)--(-3.5,-4)--cycle;
+\draw[->] (-5,0)--(3.5,0);
+\draw[->] (0,-5)--(0,3);
+\node[below, scale=0.75] at (3.5,0) { $x$ };
+\node[right, scale=0.75] at (0,3) { $y$ };
+\draw[black!80,dotted] (-1,2)--(0,2);
+\node[right,scale=0.75] at (0,2) { $\nu$ };
+\draw[black!80,dotted] (-2,0)--(-2,-1)--(0,-1);
+\node[above,scale=0.75] at (-2.4,0) { $\Lambda_h^{-1}(\nu){-}\alpha$ };
+\node[right,scale=0.75] at (0,-1) { $\Lambda_h^{-1}(\nu){+}\mu$ };
+\draw[Blue,very thick] (-5,-4)--(-2,-1)--(-1,2);
+\fill[Blue] (-1,2) circle (0.06);
+\draw[Blue] (-1,2)--(-1,3);
+\draw[Sepia, very thick] (-4.5,-5)--(3.5,3);
+\draw[Blue, very thick,dashed] (-2,-1)--(-3.5,-4);
+\node[below right,rotate=45,scale=0.9] at (-4.3,-4.8) 
+  { $y = x + \log \Vert g(0) \Vert$ };
+\node[above right,rotate=45,scale=0.9] at (-4.8,-3.8) 
+  { $y = x + \alpha + \mu$ };
+\end{tikzpicture}
+\hfill \null
+
+\caption{Admissible region for the graph of $\Lambda(f)$}
+\label{fig:area}
+\end{figure}
+
+Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
+The blue plain line represents the graph of the function $\Lambda_f$. A 
+quick computation shows that, on a neighborhood of ${-}\infty$, this 
+function is given by $\Lambda_f(x) = x + \alpha + \mu$
+where $\mu$ is the value that $\Lambda_g$ takes on the interval 
+$]{-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
+graph of $\Lambda(f)$ lies below the plain blue line. We remark
+moreover that the Taylor expansion of $f(x)$ starts with the term
+$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
+$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
+get:
+$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
+  \quad \forall x \in \R.$$
+In other words, the graph of $\Lambda(f)$ lies above the brown line.
+Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
+because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
+the dashed blue line defined as the line of slope $2$ passing through
+the first break point of the blue plain line --- which has coordinate 
+$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
+\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
+must coincide with the brown line until it meets the dashed blue line 
+and then has to stay in the green area.
+
+As a consequence of the above discussion, we derive the following 
+proposition which can be directly combined with Proposition 3.12 of 
+\cite{caruso-roe-vaccon:14a}.
+
+\begin{prop}
+\label{prop:boundLambdaf2}
+Keeping the above notations, we have:
+$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
+\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
+for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
+\end{prop}
+
+\begin{proof}
+Just remark that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
+the dashed blue line.
+\end{proof}
+
+\noindent
+{\bf About the choice of the function $h$}.
+At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
+function $h$ in our differential equation. The aim of this was to be 
+more flexible and possibly get this way better bounds on $\Lambda(f)$. 
+In this paragraph, we discuss a recommended choice for it.
+
+Actually, we do recommend taking $h$ as the identity function 
+\emph{but} changing the norm on the codomain of $h$ at the same time.
+More precisely, if $F$ --- which is the $K$-Banach in which $f$ takes
+its values --- is equipped with the norm $\Vert \cdot \Vert_F$, we 
+recommand endowing it with the second norm defined by
+$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
+where $\lambda$ is a positive real number that we shall choose later 
+and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
+acting as the identity on the underlying vector spaces. Doing this, the 
+function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
+choose $\Lambda_h = \Lambda(h)$.
+
+We believe that a good way to optimize $\lambda$ is to choose it so that 
+the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
+agree. With the expression we got just above for $\Lambda_h$, our 
+heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
+Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
+at least, on the domain we are interested in --- and we do not loose so 
+much when we write that the norm of the couple $(f(x), h(x))$ is the 
+maximum of those of $f(x)$ and $h(x)$.
+
+{\color{Bittersweet}
+
+\section{Numbers}
+\label{sec:numbers}
+
+The most basic operations are, of course, sums, products and inverses of
+elements of $K$. Their differential are well-known and quite easy to
+compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have
+$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = -
+\frac{dx}{x^2}$).
+
+Slightly more interesting is the $n$-th power map $f$ from $K$ to
+itself. Its differential $f'(x)$ is found by differentiating the
+equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
+Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(\lvert nx^{n-1} \rvert \cdot r)$. According to Lemma \ref{lem:main}, this
+means that the behavior of the precision depends on the absolute value
+of $n$. By the ultrametric inequality, we always have $\lvert n \rvert \leq 1$ but
+may have $\lvert n \rvert = 0$ if the characteristic of $K$ divides $n$ or
+$\lvert n \rvert = p^{-k}$ if $K$ is $p$-adic with $\val(n) = k$.  In the first case,
+$f'(x)$ also vanishes and Lemma \ref{lem:main} does
+not apply.  In the second, Lemma \ref{lem:main} reflects the well-known fact that
+raising a $p$-adic number to the $p^k$-th power increases the precision
+by $k$ extra digits.
+
+\section{Matrices}
+\label{sec:matrices}
+
+Differentiating ring operations (sum, products, inverse) over matrices 
+is again straightforward, keeping in mind that matrix algebras are not
+commutative.
+
+\subsection*{Determinants and characteristic polynomials.}
+
+We first outline the standard computation of the differential of the function
+$\det : M_n(K) \to K$. Suppose that $M \in \GL_n(K)$ and that $\com(M) = \det(M) M^{-1}$.
+Then
+\begin{align*}
+\det(M + dM) &= \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
+&= \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
+&= \det(M) + \tr(\com(M) \cdot dM).
+\end{align*}
+The differential of $\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
+out that this formula is still valid when $M$ is not invertible.
+The same computation extends readily to characteristic polynomials,
+since they are defined as determinants. More precisely, let us 
+consider the function $\chi : M_n(K) \to K_n[X]$ taking a matrix 
+$M$ to its monic characteristic polynomial $\det(X-M)$.
+Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
+differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
+
+\subsection*{LU factorization.}
+
+Define the LU factorization of a square matrix $M \in M_n(K)$ as a decomposition $M = LU$ 
+where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
+Such a decomposition exists and is unique provided that no principal minor
+of $M$ vanishes. We can then consider the mapping $M \mapsto (L,U)$ 
+defined over the Zariski-open set of matrices satisfying the above 
+condition. In order to differentiate it, we differentiate the relation 
+$M = LU$ and rewrite the result as
+$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
+We remark that in the right hand side of the above formula, the first
+summand is lower triangular with zero diagonal whereas 
+the second summand is upper triangular. Hence in order to compute $dL$
+and $dU$, one can proceed as follows: 
+\begin{enumerate}
+\item compute the product $dX = L^{-1} dM \: U^{-1}$,
+\item separate the lower and upper part of $dX$ obtaining $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$
+\item recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the left and $U$ on the right respectively.
+\end{enumerate}
+The above discussion extends almost \emph{verbatim} to LUP 
+factorization; the only difference is that LUP factorizations are not 
+unique but they are on a small neighborhood of $M$ if we fix the matrix 
+$P$.
+
+\subsection*{QR factorization.}
+
+A QR factorization of a square matrix $M \in M_n(K)$ will be
+a decomposition $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
+orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
+a decomposition exists and is unique on a Zariski-open subset of $M_n(K)$. The
+mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
+would like to emphasize at this point that the orthogonality condition
+defines a sub-manifold of $M_n(K)$ which is \emph{not} a
+vector space: it is defined by equations of degree $2$. The codomain
+of $f$ is then also a manifold; this example then fits into the setting
+of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
+We can differentiate $f$ by following the method used for LU 
+factorization.  Differentiating the relation $M = QR$, we obtain
+\begin{equation}
+\label{eq:diffQR}
+\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
+\cdot R^{-1}
+\end{equation}
+where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
+Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
+$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
+matrix and a diagonal one. Since moreover $dR \cdot R^{-1}$ is upper
+triangular with all diagonal entries equal to $0$, we see that 
+\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
+and $dM$.
+
+\section{Univariate polynomials}
+\label{sec:polynomials}
+
+For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
+polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
+space of dimension $d$. The affine space $X^d + K_{< d}[X]$ is then 
+the set of monic polynomials over $K$ of degree $d$. We denote it by 
+$K_d[X]$.
+
+\subsection*{Evaluation and interpolation.}
+
+Beyond sums and products (which can be treated as before), two basic 
+operations involving polynomials are evaluation and interpolation.
+Evaluation models the function $(P,x) \mapsto P(x)$, where
+$P$ is some polynomial and $x \in K$. Differentiating
+it can be done by computing
+\begin{equation}
+\label{eq:diffeval}
+(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x).
+\end{equation}
+Here $P'$ denotes the derivative of 
+$P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
+P'(x) dx + dP(x)$.
+
+As for interpolation, we consider a positive integer $d$ and the partial 
+function $f : K^{2d} \to K_{< d}[X]$ which maps the tuple $(x_1, y_1, 
+\ldots, x_d, y_d)$ to the polynomial $P$ of degree less than $d$ such 
+that $P(x_i) = y_i$ for all $i$. The polynomial $P$ exists and is unique 
+as soon as the $x_i$'s are pairwise distinct; the above function $f$ is 
+then defined on this open set. Furthermore, if $(x_1, y_1, \ldots, x_d, 
+y_d)$ is a point in it and $P$ denotes the corresponding interpolation 
+polynomial, \eqref{eq:diffeval} shows that $d y_i = P'(x_i) dx_i + 
+dP(x_i)$ for all $i$. For this, we can compute $dP(x_i)$ from $d x_i$ 
+and $d y_i$ and finally recover $dP$ by performing a new interpolation.
+
+\subsection*{Euclidean division.}
+
+Let $A$ and $B$ be two polynomials with $B \neq 0$. The Euclidean division of 
+$A$ by $B$ is the problem of finding $Q$ and $R$ with $A = BQ + R$ and $\deg R < \deg B$. 
+Differentiating the above equality we find
+\[
+dA - dB \cdot Q = B \cdot dQ + dR,
+\]
+which implies that $dQ$ and $dR$ are respectively obtained as the 
+quotient and the remainder of the Euclidean division of $dA - dB \cdot 
+Q$ by $B$. This gives the differential. We note that the discussion 
+above extends readily to convergent series (see also 
+\cite{caruso-lubicz:14a}).
+
+\subsection*{Greatest common divisors and B\'ezout coefficients.}
+
+We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
+function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
+pair $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
+greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
+coefficients of minimal degrees, computed by the extended 
+Euclidean algorithm.
+The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
+subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
+$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
+of positive degree.  Since $\mathcal V_0$ is dense, $f$ is not continuous outside 
+$\mathcal V_0$. On the contrary, the function $f$ is differentiable, and even locally analytic,
+on $\mathcal V_0$.
+
+Of course, on $\mathcal V_0$, the first component $D$ is constant and 
+therefore $dD = 0$. To compute $dU$ and $dV$, one can simply 
+differentiate the B\'ezout relation $AU + BV = 1$. We get:
+$$A \cdot dU + B \cdot dV = - (dA \cdot U + dB \cdot V)$$
+from which we deduce that $dU$ (resp. $dV$) is obtained as the 
+remainder in the Euclidean division of $U{\cdot}dX$ by $B$ (resp. of 
+$V{\cdot}dX$ by $A$) where $dX = - (dA \cdot U + dB \cdot V)$.
+
+In order to differentiate $f$ outside $\mathcal V_0$, we 
+define the subset $\mathcal V_i$ of $K_n[X] \times 
+K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
+subresultants shows that $\mathcal V_i$ is locally closed with respect to 
+the Zariski topology. In particular, it defines a $K$-manifold in the 
+sense of Appendix \ref{sec:manifold}, and the 
+restriction of $f$ to $\mathcal V_i$ is differentiable. To compute
+its differential, we proceed along the same lines as before: we 
+differentiate the relation $AU + BV = D$ and obtain this way
+$A \cdot dU + B \cdot dV - dD = dX$
+with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
+first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
+whereas the term $dD$ has degree less than $i$. 
+Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
+we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
+-dR$. These relations, together with bounds on the degree of $U$ and $V$,
+imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
+Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
+by $\frac A D$).
+
+\medskip
+
+The lesson we may retain from this study is the following. If we have to 
+compute the greatest common divisor of two polynomials $A$ and $B$ known 
+with finite precision, we first need to determine what is its degree. 
+However, the degree function is not continuous --- it is only upper 
+semi-continuous --- and hence cannot be determined with certainty from $A$ and 
+$B$, unless the approximation to $(A,B)$ lies entirely within $\mathcal V_0$.
+
+We therefore need to make an extra hypothesis.  The most natural hypothesis to make
+is that $\gcd(A, B)$ has the maximal possible degree.  The main reason for choosing
+this convention is that if the actual polynomials $A, B \in K[X]$ have a greatest common
+divisor of degree $i$ then there is some precision for which the maximal degree will be
+equal to $i$, whereas if they have a positive
+degree common divisor then no amount of increased precision will eliminate an intersection
+with $\mathcal V_0$.  A second justification is that any other choice would yield a result
+with no precision, since $A$ and $B$ appear to lie in $\mathcal V_i$ at the known precision.
+Once this assumption is made, the computation is possible and one can apply Lemma 
+\ref{lem:main} to determine the precision of the result.
+Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
+itself although there exist pairs of coprime polynomials in any 
+neighborhood of $(A,A)$.
+
+\subsection*{Factorization.}
+
+Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
+product $P_0 = A_0 B_0$, where $A_0$ and $B_0$ are monic and coprime. 
+Hensel's lemma implies that there exists a small neighborhood $\mathcal 
+U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
+uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
+and $B_0$ respectively. Thus, we can consider the map $f : P 
+\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
+polynomials. We want to differentiate $f$ at $P_0$. For this, we 
+differentiate the equality $P = A B$ around $P_0$, obtaining
+\begin{equation}
+\label{eq:difffactor}
+dP = A_0 \cdot dB + B_0 \cdot dA.
+\end{equation}
+where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
+$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
+between $A_0$ and $B_0$, it follows from \eqref{eq:difffactor} that
+$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
+{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).
+
+\subsection*{Finding roots.}
+
+An important special case of the previous study occurs when $A_0$ has 
+degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
+The map $P \mapsto A$ is then nothing but the mapping that 
+follows the simple root $\alpha_0$. Of course, its differential around 
+$P_0$ can be computed by the above method. Nevertheless, we may
+obtain it more directly by expanding the equation $(P_0 + 
+dP)(\alpha_0 + d\alpha) = 0$, obtaining
+$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$.
+Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
+does not vanish and we find:
+$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$
+
+We now address the case of a multiple root. Let $P_0$ be a monic 
+polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
+multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
+to follow the root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
+restrict ourselves to polynomials which have a root of multiplicity $m$ 
+close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
+V_m$ consisting of monic polynomials of degree $d$ having a root of 
+multiplicity at least $m$. It is a Zariski closed subset of $K_d[X]$ and
+contains by assumption the polynomial $P_0$. 
+Moreover, the irreducible components of $\mathcal V_m$ that meet at 
+$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
+$\geq m$. In particular, $\alpha_0$ corresponds to one of these 
+irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
+The algebraic variety $\mathcal V_{m,\alpha_0}$ is \emph{a fortioti} a
+$K$-manifold. Moreover there exists a differentiable function $f$ which 
+is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
+follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
+$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
+multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
+Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
+B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
+\alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
+(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
+$P^{(m-1)}(\alpha) = 0$ where $P^{(i)}$ denotes the $i$-th derivative
+of $P$. We find:
+\begin{equation}
+\label{eq:dalphamult}
+d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
+\end{equation}
+
+The above computation has interesting consequences for $p$-adic 
+precision. For example, consider the monic 
+polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
+large integer) and suppose that we are asking a computer to compute one 
+of its roots. What is the right answer? We remark that, if $\alpha$ is 
+any $p$-adic number divisible by $p^N$, then $P(X)$ can be $X^2 - 
+\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
+the two roots of $P$ are necessarily divisible by $p^N$. Hence, the
+right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
+that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
+\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
+O(p^{2N})$ if $p > 2$ and $\alpha = O(p^{2N-1})$ if $p=2$.
+This is exactly the result we get by applying Lemma \ref{lem:main} and
+using \eqref{eq:dalphamult} to simplify the differential.
+
+This phenomenon is general: if $P$ is a polynomial over $\Qp$ known at 
+some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
+multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
+(\alpha)$ vanish at the given precision) can be computed at precision 
+roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
+multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
+where $c$ is some constant (depending on $P$ and $\alpha$ but not on
+$N$).
+
+\section{Multivariate polynomials}
+\label{sec:multivar}
+
+We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
+variables over $K$ and fix a monomial order on it.
+
+\subsection*{Division.}
+
+We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
+are polynomials in $K[\XX]$, then one may decompose $f$ as
+$$f = q_1 f_1 + \cdots + q_s f_s + r$$
+where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
+by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
+that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{A canonical choice of $r$ exists even without this
+assumption. However, we do not know
+if the computation of the differential that follows extends to this more
+general setting.} of the ideal generated by these polynomials, the 
+polynomial $r$ is uniquely determined and called the \emph{remainder} of 
+the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
+$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and we can compute its differential,
+finding that $dr$ is obtained as the remainder of the division of $f - (q_1 
+\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
+
+\subsection*{Gr\"obner basis.}
+
+Recall that any 
+ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We may
+consider the map sending a family $(f_1, \ldots, f_s)$ of 
+\emph{homogeneous}\footnote{This restriction will be convenient for us 
+mainly because we are using the article \cite{vaccon:14a}, which deals only 
+with homogeneous polynomials. It is however probably not essential.} 
+polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
+\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
+map is continuous at $(f_1, \ldots, f_s)$ provided that:
+\begin{itemize}
+\item the sequence $(f_1, \ldots, f_s)$ is regular
+\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
+$f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
+order (\emph{cf} \cite{vaccon:14a} for more detail).
+\end{itemize}
+A similar argument proves that it is actually differentiable at such
+points. We are now going to compute the differential. For this we
+remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
+g_t)$ generate the same ideal, we have a relation:
+\begin{equation}
+\label{eq:prodgrobner}
+(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
+\end{equation}
+where $A$ is an $(s \times t)$ matrix with coefficients in $K[\XX]$.
+Following Vaccon's construction, we see that the entries of $A$ 
+can be chosen in such a way that they define differentiable functions.
+Differentiating \eqref{eq:prodgrobner}, we get
+$$(d g_1, \ldots, d g_t) =
+(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
+Finally, since $(g_1, \ldots, g_t)$ is a \emph{reduced} Gr\"obner
+basis, the above equality implies that $d g_i$ is the remainder in
+the division of the $i$-th entry of the product
+$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
+
+\section{Vector spaces}
+\label{sec:vectorspaces}
+
+In computer science, vector spaces are generally represented as subspaces 
+of $K^n$ for some $n$. Hence they naturally appear as points on some 
+grassmannian $\Grass(n,d)$. These grassmannians are $K$-manifolds in the 
+sense of Appendix \ref{sec:manifold} as it is discussed in \S 
+\ref{sec:exmanifold}. In the sequel, we use freely the formalism 
+introduced there.
+
+\subsection*{Left kernels.}
+
+Let $d$ and $n$ be two nonnegative integers such that $d \leq n$.
+We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 
+consisting of matrices of full rank, \emph{i.e.} rank $n-d$. The left 
+kernel defines a mapping $\text{LK} : M_{n,n-d}(K) \to \Grass(d,n)$. Let 
+us prove that it is differentiable and compute its differential around 
+some point $M \in M_{n,n-d}(K)$. Of course, there exists a neighborhood 
+$\mathcal U$ of $M$ whose image is entirely contained in a given chart 
+$U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
+U$, the map $\text{LK}$ corresponds in our chart to the map that 
+takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 
+$\begin{pmatrix} I_d & N \end{pmatrix} \cdot M = 0$. The implicit 
+function theorem then implies that $\text{LK}$ is differentiable. 
+Furthermore, its differential satisfies the relation
+$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
+\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
+from which we can compute $dN$ by projecting on the $(n-d)$ last
+columns.
+
+Following what we have already done in \S \ref{sec:polynomials} for 
+greatest common divisors of polynomials, we can develop further this 
+example and study what happens on the closed subset of $M_{n,n-d}(K)$ 
+where matrices have not full rank. On this subspace, the left kernel has 
+dimension $< d$ and then no longer defines a point in $\Grass(n,d)$. 
+Nevertheless, for all integer $r < n-d$, we can consider the subset 
+$\mathcal V_r \subset M_{n,n-d}(K)$ of matrices whose rank are exactly 
+rank $r$. It is locally closed in $M_{n,n-d}(K)$ with respect to the 
+Zariski topology and hence defines a $K$-manifold. Furthermore, we have a 
+mapping $\text{LK}_r : \mathcal M_r \to \Grass(n,n-r)$ which is 
+differentiable and whose differential can be computed as before.
+
+\subsection*{Intersections.}
+
+We pick $n$, $d_1$ and 
+$d_2$ three nonnegative integers such that $d_1 \leq n$, $d_2 \leq n$ 
+and $d_1 + d_2 \geq n$. Two subspaces of $K^n$ of dimension $d_1$ and 
+$d_2$ respectively meet along a subspace of dimension at most $d_1 + d_2 
+- n$. For all $d \leq d_1 + d_2 -n$, we can then define the subspace 
+$\mathcal V_d$ of $\Grass(n,d_1) \times \Grass(n,d_2)$ consisting of 
+pairs $(E_1, E_2)$ such that $\dim (E_1 \cap E_2) = d$ and consider the 
+function $f_d : \mathcal V_d \to \Grass(n, d)$ which sends $(E_1, E_2)$ 
+to $E_1 \cap E_2$. In charts, the function $f_d$ can be interpreted as
+the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
+kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We thus deduce
+that $f_d$ is differentiable and that its differential can be
+computed as above.
+
+\subsection*{Sums and images.}
+
+Finally, we note that similar results hold for images of matrices and,
+consequently, sums of subspaces.
+
+}
+
+\appendix
+
+\section{Proof of Proposition \ref{prop:boundLambdaf}}
+\label{app:proof}
 
@@
 some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
-aim of this subsection is to prove the following proposition.
+aim of this subsection is to prove the following intermediate result.
 
@@
 \subsection{Bounding a growing function}
-\label{ssec:boundLambdaf}
-
-We now study a slight generalization of the differential equation 
-\eqref{eq:diffequa}, which is:
-\begin{equation}
-\label{eq:diffequah}
-f' = g \circ (f, h).
-\end{equation}
-Here $g$ and $h$ are known locally analytic functions and $f$ is the 
-unknown, which is assumed to be locally analytic as well. The function
-$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
-and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
-is again an open subset in a $K$-Banach space $G$. Consequently, $g$
-goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
-all continuous linear applications $E \to F$.
-In what follows, we always assume that $V$ and $W$ contains the origin, 
-$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
-because (1)~we can always shift $f$ and $h$ (and $g$ accordingly) so 
-that they both vanish at $0$ and (2)~in order to apply the theory of 
-\cite{padicprec}, the derivative $f'(0)$ needs to be surjective and 
-therefore \emph{a fortiori} nonzero.
-
-Let us write the analytic expansion of $f$, $g$ and $h$ as follows: 
+
+We now go back to the setting of Proposition \ref{prop:boundLambdaf}.
+We write the analytic expansion of $f$, $g$ and $h$ as follows: 
 $$f = \sum_{n \geq 0} f_n, \quad
@@
 some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
-Following \cite{padicprec}, we define $\NP(f) : \R \to \R \cup 
+Following \cite{caruso-roe-vaccon:14a}, we define $\NP(f) : \R \to \R \cup 
 \{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
@@
 $\Lambda(h)$ similarly.
-
-We consider a nonnegative real number $\alpha$ such that $\Vert n! \Vert 
-\geq e^{-\alpha n}$ for all positive integer $n$. A suitable value for 
-$\alpha$ is
-$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$
-where $p$ denotes the characteristic of the residue field and
-where, by convention, the above expression equals $0$ if $p = 0$
-(we recall that we have assumed that $K$ has characteristic $0$).
-
-\begin{prop}
-\label{prop:boundLambdaf}
+We recall that $\alpha$ is a nonnegative real number such that $\Vert n! 
+\Vert \geq e^{-\alpha n}$ for all positive integer $n$.
+
+\begin{lem}
+\label{lem:boundLambdaf}
 We keep the above notations. If the couple $(a,b)$ satisfies:
@@
 then $b \geq \Lambda(f)(a - \alpha)$.
-\end{prop}
+\end{lem}
 
@@
 $$\Lambda(f') \geq \Lambda(f) - \id.$$
-Combining this with Lemma 3.7 of \cite{padicprec}, we find:
+Combining this with Lemma 3.7 of \cite{caruso-roe-vaccon:14a}, we find:
 $$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
-from which we can easily derive Proposition \ref{prop:boundLambdaf}.
+from which we can easily derive Lemma \ref{lem:boundLambdaf}.
 \end{rem}
 
-We want now to rephrase Proposition \ref{prop:boundLambdaf} in a more 
-workable way. We assume that we are given two nondecreasing convex 
-functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
-\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We assume in addition that 
-there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
-$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
-we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
-does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
-defined by:
-$$\begin{array}{ll}
-\tau_\nu(x) = x & \text{if } x \leq \nu \\
-\hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise}
-\end{array}$$
-and:
-$$\Lambda_f(x) = 
-  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)$$
-where we recall that $\alpha$ satisfies $\Vert n! \Vert \geq e^{-\alpha 
-n}$ for all $n$.
-
-\begin{cor}
-\label{cor:boundLambdaf}
-We have $\Lambda(f) \leq \Lambda_f$.
-\end{cor}
-
-\begin{proof}
-Given $a \in \R$ and $b = a + \Lambda_g \circ \Lambda_h(a)$, we have to 
-prove that $\Lambda(f)(a-\alpha) \leq b$ provided that $b \leq \nu$. 
-Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to check 
-that such couples $(a,b)$ satisfy \eqref{eq:condAB}. Clearly:
-$$b \geq a + \Lambda(g) \circ \Lambda(h)(a)$$
+We can now conclude the proof of Proposition \ref{prop:boundLambdaf} as 
+follows. Given $a \in \R$ and $b = a + \Lambda_g \circ \Lambda_h(a)$, we 
+have to prove that $\Lambda(f)(a-\alpha) \leq b$ provided that $b \leq 
+\nu$. Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to 
+check that such couples $(a,b)$ satisfy \eqref{eq:condAB}. Clearly:
+$b \geq a + \Lambda(g) \circ \Lambda(h)(a)$
 since $\Lambda_g \geq \Lambda(g)$, $\Lambda_h \geq \Lambda(h)$ and
@@
 b$ and we are done.
-\end{proof}
-
-\begin{figure}
-\null \hfill
-\begin{tikzpicture}[scale=0.8]
-\draw[thick,green!30,fill=green!20] 
-   (-2,-1)--(-1,2)--(-1,3)--(3.5,3)
- --(2.5,2)--(-3.5,-4)--cycle;
-\draw[->] (-5,0)--(3.5,0);
-\draw[->] (0,-5)--(0,3);
-\node[below, scale=0.75] at (3.5,0) { $x$ };
-\node[right, scale=0.75] at (0,3) { $y$ };
-\draw[black!80,dotted] (-1,2)--(0,2);
-\node[right,scale=0.75] at (0,2) { $\nu$ };
-\draw[black!80,dotted] (-2,0)--(-2,-1)--(0,-1);
-\node[above,scale=0.75] at (-2.4,0) { $\Lambda_h^{-1}(\nu){-}\alpha$ };
-\node[right,scale=0.75] at (0,-1) { $\Lambda_h^{-1}(\nu){+}\mu$ };
-\draw[Blue,very thick] (-5,-4)--(-2,-1)--(-1,2);
-\fill[Blue] (-1,2) circle (0.06);
-\draw[Blue] (-1,2)--(-1,3);
-\draw[Sepia, very thick] (-4.5,-5)--(3.5,3);
-\draw[Blue, very thick,dashed] (-2,-1)--(-3.5,-4);
-\node[below right,rotate=45,scale=0.9] at (-4.3,-4.8) 
-  { $y = x + \log \Vert g(0) \Vert$ };
-\node[above right,rotate=45,scale=0.9] at (-4.8,-3.8) 
-  { $y = x + \alpha + \mu$ };
-\end{tikzpicture}
-\hfill \null
-
-\caption{Admissible region for the graph of $\Lambda(f)$}
-\label{fig:area}
-\end{figure}
-
-Figure \ref{fig:area} illustrates Corollary \ref{cor:boundLambdaf}. The 
-blue plain line represents the graph of the function $\Lambda_f$. A 
-quick computation shows that, on a neighborhood of ${-}\infty$, this 
-function is given by:
-$$\Lambda_f(x) = x + \alpha + \mu$$
-where $\mu$ is the value that $\Lambda_g$ takes on the interval 
-$]{-}\infty, \nu]$. Corollary \ref{cor:boundLambdaf} says that the
-graph of $\Lambda(f)$ lies below the plain blue line. We remark
-moreover that the Taylor expansion of $f(x)$ starts with the term
-$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
-$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
-get:
-$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
-  \quad \forall x \in \R.$$
-In other words, the graph of $\Lambda(f)$ lies above the brown line.
-Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
-because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
-the dashed blue line defined as the line of slope $2$ passing through
-the first break point of the blue plain line --- which has coordinate 
-$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
-\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
-must coincide with the brown line until it meets the dashed blue line 
-and then has to stay in the green area.
-
-As a consequence of the above discussion, we derive the following 
-proposition which can be directly combined with Proposition 3.12 of 
-\cite{padicprec}.
-
-\begin{prop}
-\label{prop:boundLambdaf2}
-Keeping the above notations, we have:
-$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
-\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
-for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
-\end{prop}
-
-\begin{proof}
-Just remark that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
-the dashed blue line.
-\end{proof}
-
-\noindent
-\textbf{About the choice of the function $h$}
-
-\smallskip
-
-At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
-function $h$ in our differential equation. The aim of this was to be 
-more flexible and possibly get this way better bounds on $\Lambda(f)$. 
-In this paragraph, we discuss a recommended choice for this function 
-$h$.
-
-Actually, we do recommend taking $h$ as the identity function 
-\emph{but} changing the norm on the codomain of $h$ at the same time.
-More precisely, if $F$ --- which is the $K$-Banach in which $f$ takes
-its values --- is equipped with the norm $\Vert \cdot \Vert_F$, we 
-recommand endowing it with the second norm defined by
-$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
-where $\lambda$ is a positive real number that we shall choose later 
-and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
-acting as the identity on the underlying vector spaces. Doing this, the 
-function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
-choose $\Lambda_h = \Lambda(h)$.
-
-We believe that a good way to optimize $\lambda$ is to choose it so that 
-the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
-agree. With the expression we got just above for $\Lambda_h$, our 
-heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
-Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
-at least, on the domain we are interested in --- and we do not loose so 
-much when we write that the norm of the couple $(f(x), h(x))$ is the 
-maximum of those of $f(x)$ and $h(x)$.
-
-{\color{Bittersweet}
-
-\section{Univariate polynomials}
-\label{sec:polynomials}
-
-For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
-polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
-space of dimension $d$. The affine space $X^d + K_{< d}[X]$ is then 
-the set of monic polynomials over $K$ of degree $d$. We denote it by 
-$K_d[X]$.
-
-\subsection*{Evaluation and interpolation.}
-
-Beyond sums and products (which can be treated as before), two basic 
-operations involving polynomials are evaluation and interpolation.
-Evaluation models the function $(P,x) \mapsto P(x)$, where
-$P$ is some polynomial and $x \in K$. Differentiating
-it can be done by computing
-\begin{equation}
-\label{eq:diffeval}
-(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x).
-\end{equation}
-Here $P'$ denotes the derivative of 
-$P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
-P'(x) dx + dP(x)$.
-
-As for interpolation, we consider a positive integer $d$ and the partial 
-function $f : K^{2d} \to K_{< d}[X]$ which maps the tuple $(x_1, y_1, 
-\ldots, x_d, y_d)$ to the polynomial $P$ of degree less than $d$ such 
-that $P(x_i) = y_i$ for all $i$. The polynomial $P$ exists and is unique 
-as soon as the $x_i$'s are pairwise distinct; the above function $f$ is 
-then defined on this open set. Furthermore, if $(x_1, y_1, \ldots, x_d, 
-y_d)$ is a point in it and $P$ denotes the corresponding interpolation 
-polynomial, \eqref{eq:diffeval} shows that $d y_i = P'(x_i) dx_i + 
-dP(x_i)$ for all $i$. For this, we can compute $dP(x_i)$ from $d x_i$ 
-and $d y_i$ and finally recover $dP$ by performing a new interpolation.
-
-\subsection*{Euclidean division.}
-
-Let $A$ and $B$ be two polynomials with $B \neq 0$. The Euclidean division of 
-$A$ by $B$ is the problem of finding $Q$ and $R$ with $A = BQ + R$ and $\deg R < \deg B$. 
-Differentiating the above equality we find
-\[
-dA - dB \cdot Q = B \cdot dQ + dR,
-\]
-which implies that $dQ$ and $dR$ are respectively obtained as the 
-quotient and the remainder of the Euclidean division of $dA - dB \cdot 
-Q$ by $B$. This gives the differential. We note that the discussion 
-above extends readily to convergent series (see also 
-\cite{caruso-lubicz:14a}).
-
-\subsection*{Greatest common divisors and B\'ezout coefficients.}
-
-We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
-function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
-pair $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
-greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
-coefficients of minimal degrees, computed by the extended 
-Euclidean algorithm.
-The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
-subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
-$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
-of positive degree.  Since $\mathcal V_0$ is dense, $f$ is not continuous outside 
-$\mathcal V_0$. On the contrary, the function $f$ is differentiable, and even locally analytic,
-on $\mathcal V_0$.
-
-Of course, on $\mathcal V_0$, the first component $D$ is constant and 
-therefore $dD = 0$. To compute $dU$ and $dV$, one can simply 
-differentiate the B\'ezout relation $AU + BV = 1$. We get:
-$$A \cdot dU + B \cdot dV = - (dA \cdot U + dB \cdot V)$$
-from which we deduce that $dU$ (resp. $dV$) is obtained as the 
-remainder in the Euclidean division of $U{\cdot}dX$ by $B$ (resp. of 
-$V{\cdot}dX$ by $A$) where $dX = - (dA \cdot U + dB \cdot V)$.
-
-In order to differentiate $f$ outside $\mathcal V_0$, we 
-define the subset $\mathcal V_i$ of $K_n[X] \times 
-K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
-subresultants shows that $\mathcal V_i$ is locally closed with respect to 
-the Zariski topology. In particular, it defines a $K$-manifold in the 
-sense of Appendix \ref{sec:manifold}, and the 
-restriction of $f$ to $\mathcal V_i$ is differentiable. To compute
-its differential, we proceed along the same lines as before: we 
-differentiate the relation $AU + BV = D$ and obtain this way
-$A \cdot dU + B \cdot dV - dD = dX$
-with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
-first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
-whereas the term $dD$ has degree less than $i$. 
-Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
-we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
--dR$. These relations, together with bounds on the degree of $U$ and $V$,
-imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
-Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
-by $\frac A D$).
-
-\medskip
-
-The lesson we may retain from this study is the following. If we have to 
-compute the greatest common divisor of two polynomials $A$ and $B$ known 
-with finite precision, we first need to determine what is its degree. 
-However, the degree function is not continuous --- it is only upper 
-semi-continuous --- and hence cannot be determined with certainty from $A$ and 
-$B$, unless the approximation to $(A,B)$ lies entirely within $\mathcal V_0$.
-
-We therefore need to make an extra hypothesis.  The most natural hypothesis to make
-is that $\gcd(A, B)$ has the maximal possible degree.  The main reason for choosing
-this convention is that if the actual polynomials $A, B \in K[X]$ have a greatest common
-divisor of degree $i$ then there is some precision for which the maximal degree will be
-equal to $i$, whereas if they have a positive
-degree common divisor then no amount of increased precision will eliminate an intersection
-with $\mathcal V_0$.  A second justification is that any other choice would yield a result
-with no precision, since $A$ and $B$ appear to lie in $\mathcal V_i$ at the known precision.
-Once this assumption is made, the computation is possible and one can apply Lemma 
-\ref{lem:main} to determine the precision of the result.
-Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
-itself although there exist pairs of coprime polynomials in any 
-neighborhood of $(A,A)$.
-
-\subsection*{Factorization.}
-
-Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
-product $P_0 = A_0 B_0$, where $A_0$ and $B_0$ are monic and coprime. 
-Hensel's lemma implies that there exists a small neighborhood $\mathcal 
-U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
-uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
-and $B_0$ respectively. Thus, we can consider the map $f : P 
-\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
-polynomials. We want to differentiate $f$ at $P_0$. For this, we 
-differentiate the equality $P = A B$ around $P_0$, obtaining
-\begin{equation}
-\label{eq:difffactor}
-dP = A_0 \cdot dB + B_0 \cdot dA.
-\end{equation}
-where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
-$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
-between $A_0$ and $B_0$, it follows from \eqref{eq:difffactor} that
-$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
-{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).
-
-\subsection*{Finding roots.}
-
-An important special case of the previous study occurs when $A_0$ has 
-degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
-The map $P \mapsto A$ is then nothing but the mapping that 
-follows the simple root $\alpha_0$. Of course, its differential around 
-$P_0$ can be computed by the above method. Nevertheless, we may
-obtain it more directly by expanding the equation $(P_0 + 
-dP)(\alpha_0 + d\alpha) = 0$, obtaining
-$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$.
-Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
-does not vanish and we find:
-$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$
-
-We now address the case of a multiple root. Let $P_0$ be a monic 
-polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
-multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
-to follow the root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
-restrict ourselves to polynomials which have a root of multiplicity $m$ 
-close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
-V_m$ consisting of monic polynomials of degree $d$ having a root of 
-multiplicity at least $m$. It is a Zariski closed subset of $K_d[X]$ and
-contains by assumption the polynomial $P_0$. 
-Moreover, the irreducible components of $\mathcal V_m$ that meet at 
-$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
-$\geq m$. In particular, $\alpha_0$ corresponds to one of these 
-irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
-The algebraic variety $\mathcal V_{m,\alpha_0}$ is \emph{a fortioti} a
-$K$-manifold. Moreover there exists a differentiable function $f$ which 
-is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
-follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
-$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
-multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
-Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
-B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
-\alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
-(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
-$P^{(m-1)}(\alpha) = 0$ where $P^{(i)}$ denotes the $i$-th derivative
-of $P$. We find:
-\begin{equation}
-\label{eq:dalphamult}
-d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
-\end{equation}
-
-The above computation has interesting consequences for $p$-adic 
-precision. For example, consider the monic 
-polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
-large integer) and suppose that we are asking a computer to compute one 
-of its roots. What is the right answer? We remark that, if $\alpha$ is 
-any $p$-adic number divisible by $p^N$, then $P(X)$ can be $X^2 - 
-\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
-the two roots of $P$ are necessarily divisible by $p^N$. Hence, the
-right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
-that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
-\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
-O(p^{2N})$ if $p > 2$ and $\alpha = O(p^{2N-1})$ if $p=2$.
-This is exactly the result we get by applying Lemma \ref{lem:main} and
-using \eqref{eq:dalphamult} to simplify the differential.
-
-This phenomenon is general: if $P$ is a polynomial over $\Q_p$ known at 
-some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
-multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
-(\alpha)$ vanish at the given precision) can be computed at precision 
-roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
-multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
-where $c$ is some constant (depending on $P$ and $\alpha$ but not on
-$N$).
-
-\section{Multivariate polynomials}
-
-We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
-variables over $K$ and fix a monomial order on it.
-
-\subsection*{Division.}
-
-We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
-are polynomials in $K[\XX]$, then one may decompose $f$ as
-$$f = q_1 f_1 + \cdots + q_s f_s + r$$
-where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
-by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
-that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{A canonical choice of $r$ exists even without this
-assumption. However, we do not know
-if the computation of the differential that follows extends to this more
-general setting.} of the ideal generated by these polynomials, the 
-polynomial $r$ is uniquely determined and called the \emph{remainder} of 
-the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
-$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and we can compute its differential,
-finding that $dr$ is obtained as the remainder of the division of $f - (q_1 
-\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
-
-\subsection*{Gr\"obner basis.}
-
-Recall that any 
-ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We may
-consider the map sending a family $(f_1, \ldots, f_s)$ of 
-\emph{homogeneous}\footnote{This restriction will be convenient for us 
-mainly because we are using the article \cite{vaccon:14a}, which deals only 
-with homogeneous polynomials. It is however probably not essential.} 
-polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
-map is continuous at $(f_1, \ldots, f_s)$ provided that:
-\begin{itemize}
-\item the sequence $(f_1, \ldots, f_s)$ is regular
-\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
-$f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{vaccon:14a} for more detail).
-\end{itemize}
-A similar argument proves that it is actually differentiable at such
-points. We are now going to compute the differential. For this we
-remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
-g_t)$ generate the same ideal, we have a relation:
-\begin{equation}
-\label{eq:prodgrobner}
-(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
-\end{equation}
-where $A$ is an $(s \times t)$ matrix with coefficients in $K[\XX]$.
-Following Vaccon's construction, we see that the entries of $A$ 
-can be chosen in such a way that they define differentiable functions.
-Differentiating \eqref{eq:prodgrobner}, we get
-$$(d g_1, \ldots, d g_t) =
-(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
-Finally, since $(g_1, \ldots, g_t)$ is a \emph{reduced} Gr\"obner
-basis, the above equality implies that $d g_i$ is the remainder in
-the division of the $i$-th entry of the product
-$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
-
-\section{Matrices}
-\label{sec:matrices}
-
-Differentiating ring operations (sum, products, inverse) over matrices 
-is again straightforward, keeping in mind that matrix algebras are not
-commutative.
-
-\subsection*{Determinants and characteristic polynomials.}
-
-We first outline the standard computation of the differential of the function
-$\det : M_n(K) \to K$. Suppose that $M \in \GL_n(K)$ and that $\com(M) = \det(M) M^{-1}$.
-Then
-\begin{align*}
-\det(M + dM) &= \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
-&= \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
-&= \det(M) + \tr(\com(M) \cdot dM).
-\end{align*}
-The differential of $\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
-out that this formula is still valid when $M$ is not invertible.
-The same computation extends readily to characteristic polynomials,
-since they are defined as determinants. More precisely, let us 
-consider the function $\chi : M_n(K) \to K_n[X]$ taking a matrix 
-$M$ to its monic characteristic polynomial $\det(X-M)$.
-Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
-differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
-
-\subsection*{LU factorization.}
-
-Define the LU factorization of a square matrix $M \in M_n(K)$ as a decomposition $M = LU$ 
-where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
-Such a decomposition exists and is unique provided that no principal minor
-of $M$ vanishes. We can then consider the mapping $M \mapsto (L,U)$ 
-defined over the Zariski-open set of matrices satisfying the above 
-condition. In order to differentiate it, we differentiate the relation 
-$M = LU$ and rewrite the result as
-$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
-We remark that in the right hand side of the above formula, the first
-summand is lower triangular with zero diagonal whereas 
-the second summand is upper triangular. Hence in order to compute $dL$
-and $dU$, one can proceed as follows: 
-\begin{enumerate}
-\item compute the product $dX = L^{-1} dM \: U^{-1}$,
-\item separate the lower and upper part of $dX$ obtaining $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$
-\item recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the left and $U$ on the right respectively.
-\end{enumerate}
-The above discussion extends almost \emph{verbatim} to LUP 
-factorization; the only difference is that LUP factorizations are not 
-unique but they are on a small neighborhood of $M$ if we fix the matrix 
-$P$.
-
-\subsection*{QR factorization.}
-
-A QR factorization of a square matrix $M \in M_n(K)$ will be
-a decomposition $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
-orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
-a decomposition exists and is unique on a Zariski-open subset of $M_n(K)$. The
-mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
-would like to emphasize at this point that the orthogonality condition
-defines a sub-manifold of $M_n(K)$ which is \emph{not} a
-vector space: it is defined by equations of degree $2$. The codomain
-of $f$ is then also a manifold; this example then fits into the setting
-of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
-We can differentiate $f$ by following the method used for LU 
-factorization.  Differentiating the relation $M = QR$, we obtain
-\begin{equation}
-\label{eq:diffQR}
-\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
-\cdot R^{-1}
-\end{equation}
-where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
-Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
-$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
-matrix and a diagonal one. Since moreover $dR \cdot R^{-1}$ is upper
-triangular with all diagonal entries equal to $0$, we see that 
-\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
-and $dM$.
-
-\section{Vector spaces}
-
-In computer science, vector spaces are generally represented as subspaces 
-of $K^n$ for some $n$. Hence they naturally appear as points on some 
-grassmannian $\Grass(n,d)$. These grassmannians are $K$-manifolds in the 
-sense of Appendix \ref{sec:manifold} as it is discussed in \S 
-\ref{sec:exmanifold}. In the sequel, we use freely the formalism 
-introduced there.
-
-\subsection*{Left kernels.}
-
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$.
-We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 
-consisting of matrices of full rank, \emph{i.e.} rank $n-d$. The left 
-kernel defines a mapping $\text{LK} : M_{n,n-d}(K) \to \Grass(d,n)$. Let 
-us prove that it is differentiable and compute its differential around 
-some point $M \in M_{n,n-d}(K)$. Of course, there exists a neighborhood 
-$\mathcal U$ of $M$ whose image is entirely contained in a given chart 
-$U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
-U$, the map $\text{LK}$ corresponds in our chart to the map that 
-takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 
-$\begin{pmatrix} I_d & N \end{pmatrix} \cdot M = 0$. The implicit 
-function theorem then implies that $\text{LK}$ is differentiable. 
-Furthermore, its differential satisfies the relation
-$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
-\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
-from which we can compute $dN$ by projecting on the $(n-d)$ last
-columns.
-
-Following what we have already done in \S \ref{sec:polynomials} for 
-greatest common divisors of polynomials, we can develop further this 
-example and study what happens on the closed subset of $M_{n,n-d}(K)$ 
-where matrices have not full rank. On this subspace, the left kernel has 
-dimension $< d$ and then no longer defines a point in $\Grass(n,d)$. 
-Nevertheless, for all integer $r < n-d$, we can consider the subset 
-$\mathcal V_r \subset M_{n,n-d}(K)$ of matrices whose rank are exactly 
-rank $r$. It is locally closed in $M_{n,n-d}(K)$ with respect to the 
-Zariski topology and hence defines a $K$-manifold. Furthermore, we have a 
-mapping $\text{LK}_r : \mathcal M_r \to \Grass(n,n-r)$ which is 
-differentiable and whose differential can be computed as before.
-
-\subsection*{Intersections.}
-
-We pick $n$, $d_1$ and 
-$d_2$ three nonnegative integers such that $d_1 \leq n$, $d_2 \leq n$ 
-and $d_1 + d_2 \geq n$. Two subspaces of $K^n$ of dimension $d_1$ and 
-$d_2$ respectively meet along a subspace of dimension at most $d_1 + d_2 
-- n$. For all $d \leq d_1 + d_2 -n$, we can then define the subspace 
-$\mathcal V_d$ of $\Grass(n,d_1) \times \Grass(n,d_2)$ consisting of 
-pairs $(E_1, E_2)$ such that $\dim (E_1 \cap E_2) = d$ and consider the 
-function $f_d : \mathcal V_d \to \Grass(n, d)$ which sends $(E_1, E_2)$ 
-to $E_1 \cap E_2$. In charts, the function $f_d$ can be interpreted as
-the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
-kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We thus deduce
-that $f_d$ is differentiable and that its differential can be
-computed as above.
-
-\subsection*{Sums and images.}
-
-Finally, we note that similar results hold for images of matrices and,
-consequently, sums of subspaces.
-
-}
-
-%\bibliographystyle{plain}
-%\bibliography{roebib/Biblio,extras}
+
+\bibliographystyle{plain}
+\bibliography{../roebib/Biblio,extras}
 

### 2014-12-09 examples/stability.tex
Differentials for vector spaces

@@
 \DeclareMathOperator{\NP}{NP}
+\DeclareMathOperator{\Hom}{Hom}
+\DeclareMathOperator{\End}{End}
 \DeclareMathOperator{\GL}{GL}
@@
 \newcommand{\Epi}{\textrm{Epi}}
+\renewcommand{\c}{\text{\rm c}}
+
+\newcommand{\DI}{\text{\rm DI}}
+\newcommand{\II}{\text{\rm II}}
 
@@
 $p$-adic stability of many classical algorithms involving many usual
-algebraic objects (matrices, univariate and multivariate polynomials,
+algebraic objects (matrices, univariate polynomials,
 vector spaces). We observe in particular that generic algorithms ---
@@
 The present paper, which may be considered as a continuation of 
-\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. cit.} in 
-order to analyse the $p$-adic stability of many standard algorithms
-for computing with all basic algebraic structures that are numbers,
-matrices, univariate and multivariate polynomials, series and vector
-spaces. For each considered algorithm, we will follow the same protocol: 
+\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. 
+cit.} in order to analyse the $p$-adic stability of many standard 
+algorithms for computing with basic algebraic structures: numbers, 
+matrices, univariate polynomials and vector spaces. For each considered 
+algorithm, we will follow the same protocol: 
 first, we will compute the theoretical loss of precision of the 
-underlying question using the framework of \cite{caruso-roe-vaccon:14a} and then 
-compare it to those observed in available implementations in 
+underlying question using the framework of \cite{caruso-roe-vaccon:14a} 
+and then compare it to those observed in available implementations in 
 \textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
@@
 
-\section{Multivariate polynomials}
-\label{sec:multivar}
-
-We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
-variables over $K$ and fix a monomial order on it.
-
-\subsection*{Division.}
-
-We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
-are polynomials in $K[\XX]$, then one may decompose $f$ as
-$$f = q_1 f_1 + \cdots + q_s f_s + r$$
-where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
-by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
-that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{A canonical choice of $r$ exists even without this
-assumption. However, we do not know
-if the computation of the differential that follows extends to this more
-general setting.} of the ideal generated by these polynomials, the 
-polynomial $r$ is uniquely determined and called the \emph{remainder} of 
-the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
-$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and we can compute its differential,
-finding that $dr$ is obtained as the remainder of the division of $f - (q_1 
-\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.
-
-\subsection*{Gr\"obner basis.}
-
-Recall that any 
-ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We may
-consider the map sending a family $(f_1, \ldots, f_s)$ of 
-\emph{homogeneous}\footnote{This restriction will be convenient for us 
-mainly because we are using the article \cite{vaccon:14a}, which deals only 
-with homogeneous polynomials. It is however probably not essential.} 
-polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
-\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
-map is continuous at $(f_1, \ldots, f_s)$ provided that:
-\begin{itemize}
-\item the sequence $(f_1, \ldots, f_s)$ is regular
-\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
-$f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
-order (\emph{cf} \cite{vaccon:14a} for more detail).
-\end{itemize}
-A similar argument proves that it is actually differentiable at such
-points. We are now going to compute the differential. For this we
-remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
-g_t)$ generate the same ideal, we have a relation:
-\begin{equation}
-\label{eq:prodgrobner}
-(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
-\end{equation}
-where $A$ is an $(s \times t)$ matrix with coefficients in $K[\XX]$.
-Following Vaccon's construction, we see that the entries of $A$ 
-can be chosen in such a way that they define differentiable functions.
-Differentiating \eqref{eq:prodgrobner}, we get
-$$(d g_1, \ldots, d g_t) =
-(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
-Finally, since $(g_1, \ldots, g_t)$ is a \emph{reduced} Gr\"obner
-basis, the above equality implies that $d g_i$ is the remainder in
-the division of the $i$-th entry of the product
-$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
+}
 
@@
 
-In computer science, vector spaces are generally represented as subspaces 
-of $K^n$ for some $n$. Hence they naturally appear as points on some 
-grassmannian $\Grass(n,d)$. These grassmannians are $K$-manifolds in the 
-sense of Appendix \ref{sec:manifold} as it is discussed in \S 
-\ref{sec:exmanifold}. In the sequel, we use freely the formalism 
-introduced there.
-
-\subsection*{Left kernels.}
-
-Let $d$ and $n$ be two nonnegative integers such that $d \leq n$.
-We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 
-consisting of matrices of full rank, \emph{i.e.} rank $n-d$. The left 
-kernel defines a mapping $\text{LK} : M_{n,n-d}(K) \to \Grass(d,n)$. Let 
-us prove that it is differentiable and compute its differential around 
-some point $M \in M_{n,n-d}(K)$. Of course, there exists a neighborhood 
-$\mathcal U$ of $M$ whose image is entirely contained in a given chart 
-$U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
-U$, the map $\text{LK}$ corresponds in our chart to the map that 
-takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 
-$\begin{pmatrix} I_d & N \end{pmatrix} \cdot M = 0$. The implicit 
-function theorem then implies that $\text{LK}$ is differentiable. 
-Furthermore, its differential satisfies the relation
-$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
-\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
-from which we can compute $dN$ by projecting on the $(n-d)$ last
-columns.
-
-Following what we have already done in \S \ref{sec:polynomials} for 
-greatest common divisors of polynomials, we can develop further this 
-example and study what happens on the closed subset of $M_{n,n-d}(K)$ 
-where matrices have not full rank. On this subspace, the left kernel has 
-dimension $< d$ and then no longer defines a point in $\Grass(n,d)$. 
-Nevertheless, for all integer $r < n-d$, we can consider the subset 
-$\mathcal V_r \subset M_{n,n-d}(K)$ of matrices whose rank are exactly 
-rank $r$. It is locally closed in $M_{n,n-d}(K)$ with respect to the 
-Zariski topology and hence defines a $K$-manifold. Furthermore, we have a 
-mapping $\text{LK}_r : \mathcal M_r \to \Grass(n,n-r)$ which is 
-differentiable and whose differential can be computed as before.
-
-\subsection*{Intersections.}
-
-We pick $n$, $d_1$ and 
-$d_2$ three nonnegative integers such that $d_1 \leq n$, $d_2 \leq n$ 
-and $d_1 + d_2 \geq n$. Two subspaces of $K^n$ of dimension $d_1$ and 
-$d_2$ respectively meet along a subspace of dimension at most $d_1 + d_2 
-- n$. For all $d \leq d_1 + d_2 -n$, we can then define the subspace 
-$\mathcal V_d$ of $\Grass(n,d_1) \times \Grass(n,d_2)$ consisting of 
-pairs $(E_1, E_2)$ such that $\dim (E_1 \cap E_2) = d$ and consider the 
-function $f_d : \mathcal V_d \to \Grass(n, d)$ which sends $(E_1, E_2)$ 
-to $E_1 \cap E_2$. In charts, the function $f_d$ can be interpreted as
-the kernel of a matrix simply because $E_1 \cap E_2$ appears as the
-kernel of the canonical linear map $E_1 \oplus E_2 \to K^n$. We thus deduce
-that $f_d$ is differentiable and that its differential can be
-computed as above.
-
-\subsection*{Sums and images.}
-
-Finally, we note that similar results hold for images of matrices and,
-consequently, sums of subspaces.
-
-}
+Vector spaces are generally represented as subspaces of $K^n$ for some 
+$n$. Hence they naturally appear as points on grassmannians. Therefore, 
+one can use the framework of \cite[Appendix A]{caruso-roe-vaccon:14a} to 
+study the $p$-adic in this context.
+
+\subsection{Geometry of grassmannians}
+
+Given $E$ be a finite dimensional vector space over $K$ and $d$ an 
+integer in the range $[0, \dim E]$ we denote by $\Grass(E,d)$ the 
+set of $d$-dimensional subspaces of $E$. It is the so-called 
+\emph{grassmannian}. It is well-known that $\Grass(E,d)$ has a 
+natural structure of $K$-manifold. The aim of this subsection is to 
+recall standard facts about its geometry. In what follows, we set
+$n = \dim E$ and equip $E$ with a distinguished basis $(e_1, \ldots, 
+e_n)$.
+
+\medskip
+
+\noindent
+{\bf Description and tangent space.}
+Let $V$ denote a fixed sub-vector space of $E$ of dimension $d$. The 
+grassmannian $\Grass(E,d)$ can be viewed as the quotient of the set 
+of linear embeddings $f: V \hookrightarrow E$ modulo the action (by 
+precomposition) of $\GL(V)$: the mapping $f$ represents its image 
+$f(V)$. It follows from this description that the tangent space of 
+$\Grass(E,d)$ is canonically isomorphic to $$\Hom(V, E) / \End(V) 
+\simeq \Hom(V, E/V).$$
+
+\medskip
+
+\noindent
+{\bf Charts.}
+Let $V$ and $V^\c$ be two complementary subspaces of $E$ 
+(\emph{i.e.} $V \oplus V^\c = E$). We assume that $V$ has 
+dimension $d$ and denote by $\pi$ the projection $E \to V$ 
+corresponding to the above decomposition. We introduce the set 
+$\mathcal U_{V,V^\c}$ of all embeddings $f : V \hookrightarrow E$ 
+such that $\pi \circ f = \id_V$. Clearly it is an affine space over
+$\Hom(V,V^\c)$. 
+Furthermore, we can embed it into $\Grass(E,d)$ by taking $f$ as
+above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
+on open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
+$W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
+at each such $W$ becomes isomorphic to $\Hom(V,V^\c)$. The
+identification $\Hom(V,V^\c) \to \Hom(W, E/W)$ is given by
+$du \mapsto (du \circ f^{-1}) \text{ mod } W$ where $f : V 
+\stackrel{\sim}{\to} W$ is the linear mapping defining $W$.
+
+When the couple $(V, V^\c)$ varies, the open subsets $\mathcal 
+U_{V,V^\c}$ cover the whole grassmaniann and then define an atlas 
+of it. When implementing vector spaces on computes, we usually restrict 
+ourselves to the subatlas consisting of all charts of the form $(V_I, 
+V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
+\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspaces spanned by 
+the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
+least one $\mathcal U_{V_I, V_{I^{\text{c}}}}$ and, given a family of
+generators of $W$, we can determine such an $I$ together with the
+corresponding embedding $f : V_I \hookrightarrow E$ by echelonizing 
+the matrix of generators of $W$.
+
+\medskip
+
+\noindent
+{\bf A variant.}
+Alternatively, one can describe $\Grass(E,d)$ as the set of linear 
+surjective morphisms $f : E \to E/V$ modulo the action (by 
+postcomposition) of $\GL(E/V)$ as well. This identification presents the 
+tangent space at a given point $V$ as the quotient $\Hom(V, E/V) / 
+\End(E/V)$, which turns out to be again isomorphic to $\Hom(V, E/V)$. 
+Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
+U^\star_{V, V^\c}$ denote the set of surjective linear maps $f : E 
+\to V^\c$ whose restriction to $V^\c$ is the identity. It is an 
+affine space over $\Hom(V, V^\c)$ which can be identified to an open
+subset of $\Grass(E,d)$ \emph{via} the map $f \mapsto \ker f$.
+
+It is actually easily seen that $\mathcal U_{V, V^\c}$ and $\mathcal 
+U^\star_{V, V^\c}$ define the same open subset in $\Grass(E,d)$. 
+Indeed, given $f \in \mathcal U_{V, V^\c}$, one can write $f = \id_V 
++ h$ with $h \in \Hom(V, V^\c)$ and define the morphism $g = \id_E -
+h \circ \pi \in \mathcal U^\star_{V, V^\c}$. The association $f \mapsto
+g$ then defines a bijection $\mathcal U_{V, V^\c} \to \mathcal
+U^\star_{V, V^\c}$ which commutes with the embeddings into the
+grassmannian. We emphasize that the above bijection is quite easy to
+compute: it consists in replacing the matrix 
+$\Big(\begin{matrix} I \\ H \end{matrix}\Big)$ by $(\begin{matrix} -H & 
+I \end{matrix})$ (in a good choice of coordinates).
+
+\medskip
+
+\noindent
+{\bf Duality.}
+If $E$ is a finite dimension vector space over $K$, we use the notation 
+$E^\star$ for its dual (\emph{i.e.} $E^\star = \Hom(E,K)$) and, given in 
+addition a subspace $V \subset E$, we denote by $V^\perp$ the subspace 
+of $E^\star$ consisting of linear maps that vanish on $V$. We recall
+that the dual of $V^\perp$ (resp. $E^\star/V^\perp$) is canonically
+isomorphic to $E/V$ (resp. $V$).
+For all $d$, the association $V \mapsto V^\perp$ defines a continuous 
+morphism $\psi_E : \Grass(E,d) \to \Grass(E^\star, n-d)$. The action of 
+$\psi_E$ on tangent spaces is easily described. Indeed, the differential 
+of $\psi_E$ at $V$ is nothing but the canonical identification between
+$\Hom(V, E/V)$ and $\Hom(V^\perp, E^\star/V^\perp)$ induced by 
+``transposition''. Furthermore, we observe that $\psi_E$ respects
+the charts we have defined above, in the sense that it maps bijectively 
+$\mathcal U_{V, V^\c}$ to $\mathcal U^\star_{V^\perp, (V^\c)^\perp}
+\simeq \mathcal U_{V^\perp, (V^\c)^\perp}$.
+
+\subsection{Differential computations}
+
+We compute in this subsection the differential of various operations on 
+vector spaces. For brevity, we skip the estimation of the corresponding 
+growing functions (but this can be done using Proposition 
+\ref{prop:boundLambdaf2} as before).
+
+\medskip
+
+\noindent
+{\bf Direct images.}
+Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
+$n$ and $m$ respectively. Let $d$ be an integer in $[0,n]$. We are
+interested here in the function $\DI$ (for ``direct image'') defined 
+over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
+couple $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
+priori} vary, the map $\DI$ does not take its values in well-defined 
+grassmannian. To tackle this problem, we stratify $\mathcal M$ as
+follows: for each integer $r \in [0,d]$, we introduce the subset
+$\mathcal M_r \subset \mathcal M$ consisting of those couples $(f,V)$
+for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
+locally closed in $\mathcal M$ are therefore are submanifolds. 
+Moreover, $\DI$ induces by restriction respectable differentiable 
+functions between $K$-manifolds:
+$$\DI_r : \mathcal M_r \to \Grass(F,r).$$
+We would like to differentiate $\DI_r$ around some point $(f,V) \in 
+\mathcal M_r$. To do so, we use the first description of the 
+grassmannians we gave above: we shall see points in $\Grass(E,d)$ 
+(resp. $\Grass(F,d)$) as embeddings $V \hookrightarrow E$ (resp. $W 
+\hookrightarrow F$) modulo the action of $\GL(V)$ (resp. $\GL(W)$).
+The point $V \in \Grass(E,d)$ is then represented by the canonical 
+inclusion $v : V \to E$ whereas a representative $w$ of $W$ satisfies
+$w \circ \varphi = f \circ v$
+where $\varphi : V \to W$ is the linear mapping induced by $f$. The 
+previous relation still holds if $(f,v)$ is replaced by a couple $(f', 
+v') \in \mathcal M_r$ sufficiently close to $(f,v)$.
+Differentiating it and passing to the quotient we find, first, that the 
+tangent space of $\mathcal M_r$ at $(f,v)$ consists of couples $(df, dv) 
+\in \Hom(E,F) \times \Hom(V,E/V)$ such that
+$$d\tilde w = \big((df \circ v + f \circ dv) \text{ mod } W\big)
+\in \Hom(V, F/W)$$
+factors through $\varphi$ (\emph{i.e.} vanishes on $\ker \varphi = V 
+\cap \ker f$) and, second, that the differential of $\DI_r$ at $(f,V)$ 
+is the linear mapping sending $(df,dv)$ as above to the unique element 
+$dw \in \Hom(W,F/W)$ such that $dw \circ \varphi = d \tilde w$.
+
+\medskip
+
+\noindent
+{\bf Inverse images.}
+We now consider the mapping $\II$ (for ``inverse image'') sending a 
+couple $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
+$f^{-1}(W)$. As before, this map does not takes its value in a single 
+grassmannian and we need to stratify $\mathcal W$ in order to get 
+differentiable functions. For each integer $r \in [0,n]$, we introduce 
+the sub-manifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
+couples $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
+induces a continuous function
+$\II_r : \mathcal W_r \to \Grass(E,s)$.
+Pick $(f,W) \in \mathcal W_r$. Set $V = f^{-1}(W)$ and denote by $w : F 
+\to F/W$ the canonical projection.
+Similarly to what we have done for direct images, one can prove that
+the tangent space of $\mathcal W_r$ at some point $(f,W) \in \mathcal
+W_r$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
+couples $(df,dw)$ such that
+$d \tilde v = (w \circ df + dw \circ f)_{|W}$
+factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
+$f$. Furthermore $\II_r$ is differentiable at $(f,W)$ and its 
+differential is the linear mapping that takes $(df,dw)$ as above to the 
+unique element $dv \in \Hom(V,E/V)$ satisfying $\varphi \circ dv =
+d\tilde v$.
+
+The cases of direct images and inverse images are related by duality
+as follows: if $f : E \to F$ is any linear map and $W$ is a subspace
+of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. Using this, one 
+can deduce the differential of $\DI_r$ from those of $\II_s$ and 
+\emph{vice versa}.
+
+%\begin{rem}
+%A particularly interesting case of inverse images is of course those
+%of kernels.
+%\end{rem}
+
+\medskip
+
+\noindent
+{\bf Sums and intersections.}
+Let $d_1$ and $d_2$ be two nonnegative integers. We consider the 
+function $\Sigma$ defined on the manifold $\mathcal C = \Grass(E,d_1) 
+\times \Grass(E,d_2)$ by $\Sigma(V_1, V_2) = V_1 + V_2$. As before, in 
+order to study $\Sigma$, we stratify $\mathcal C$ according to the 
+dimension of the sum: for each integer $d \in [0, d_1+d_2]$, we define 
+$\mathcal C_d$ as the sub-manifold of $\mathcal C$ consisting of those 
+couples $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get the way a 
+well-defined mapping $\mathcal C_d \to \Grass(E,d)$ whose differential
+can be computed similarly as before. We seek to state the result: the
+tangent space of $\mathcal C_d$ at a given point $(V_1, V_2)$ consists 
+of couples $(dv_1, dv_2) \in \Hom(V_1, E/V_1)
+\times \Hom(V_2, E/V_2)$ such that $dv_1 \equiv dv_2 \pmod{V_1 + V_2}$ 
+on the intersection $V_1 \cap V_2$ and the differential of $\Sigma_r$
+at $(V_1, V_2)$ maps $(dv_1, dv_2)$ to $dv \in \Hom(V, E/V)$ (with $V
+= V_1 + V_2$) defined by $dv(v_1 + v_2) = dv_1(v_1) + dv_2(v_2)$ ($v_1
+\in V_1$, $v_2 \in V_2$).
+
+Using duality, we derive from this a similar result for the mapping 
+$(V_1, V_2) \mapsto V_1 \cap V_2$ (left to the reader).
+
+\subsection{Stability}
 

### 2014-12-10 examples/stability.tex
Smith form for matrices

@@
 
+}
+
 \section{Matrices}
@@
 
-Differentiating ring operations (sum, products, inverse) over matrices 
-is again straightforward, keeping in mind that matrix algebras are not
-commutative.
-
-\subsection*{Determinants and characteristic polynomials.}
+\todo{Write a short introduction.}
+
+The notation $M_{n,m}(K)$ refers to the $K$-vector space of matrices 
+with $n$ rows and $m$ columns with coefficients in $K$.
+
+\medskip
+
+\noindent
+{\bf Smith form.}
+In what follows, we will often use the \emph{Smith decomposition} of a 
+matrix $M \in M_{n,m}(K)$: we recall that it is a factorization of $M$ 
+as a product $U_M \cdot \Delta_M \cdot V_M$ when $U_M$ and $V_M$ are 
+unimodular (square) matrices of size $n$ and $m$ respectively, with 
+coefficients in $\O_K$ and $\Delta_M$ satisfies the two following
+conditions:
+
+\noindent
+(i) its $(i,j)$-th entry vanishes as soon as $i \neq j$
+
+\noindent
+(ii) its $(i,i)$-th entry divides its $(j,j)$-th entry when $i
+\leq j$.
+
+\noindent
+Such a factorization is unique and can be easily computed by performing 
+elementary operations on rows and columns.
+Regarding to precision, one can prove that if the entries of $M$ are all 
+known up to some $O(\pi^N)$, it is possible to compute a factorization 
+$U_M \cdot \Delta_M \cdot V_M$ where $U_M$ and $V_M$ are \emph{exact}
+matrices satisfying the above conditions and where $\Delta_M$ is known
+at precision $O(\pi^N)$ and satisfies~(i) and~(ii) modulo $\pi^N$.
+
+\subsection{Product of matrices}
+
+{\color{Bittersweet}
+
+\subsection{Determinant}
 
@@
 out that this formula is still valid when $M$ is not invertible.
+
+\subsection{Characteristic polynomials.}
+
 The same computation extends readily to characteristic polynomials,
@@
 
-\subsection{Stability}
+\subsection{Application to precision}
 

### 2014-12-10 examples/stability.tex
Multiplication and determinant of matrices

@@
 
+\newcommand{\detp}{\text{\rm det}'}
 \newcommand{\DI}{\text{\rm DI}}
@@
 \noindent
-Such a factorization is unique and can be easily computed by performing 
-elementary operations on rows and columns.
+Such a factorization is \emph{not} unique but the valuations of
+the ``diagonal'' entries of $\Delta_M$ are. Smith decompositions
+can be easily computed by performing elementary operations on rows and columns.
 Regarding to precision, one can prove that if the entries of $M$ are all 
@@
 
-\subsection{Product of matrices}
+\subsection{Multiplication}
+
+To begin with, we want to study the behaviour of the precision when 
+performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
+positive integers with $s \geq \min(r,t)$ and assume that we want to 
+multiply a matrix $A \in M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. 
+This operation is of course modeled by the polynomial function:
+$$\begin{array}{rcl}
+\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
+M_{r,t}(K) \smallskip \\
+(A,B) & \mapsto & AB.
+\end{array}$$
+According to Proposition \ref{}, the behaviour of the precision when 
+computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$. It is obvious 
+that this differential is the linear mapping that takes a couple 
+$(dA,dB)$ to $A \cdot dB + dA \cdot B$.
+
+To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
+lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
+to apply Proposition \ref{}, we then need to compute the image 
+of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
+M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
+contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
+entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
+Nonetheless, it may happen that the above inclusion is strict, meaning 
+that we are \emph{gaining} precision in those cases. In order to compute 
+exactly $\mathcal P'_{a,b,c}(A,B)(\mathcal L_0)$, we introduce the Smith 
+factorization of $A$ and $B$:
+$$A = U_A \cdot \Delta_A \cdot V_A
+\text{ and } B = U_B \cdot \Delta_B \cdot V_B$$
+and denote by $a_1, \ldots, a_{\min(r,s)}$ (resp. $b_1, \ldots, 
+b_{\min(s,t)}$) the valuation of the diagonal entries of $\Delta_A$
+(resp. $\Delta_B$). We also agree to set $a_n = b_m = +\infty$ if
+$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}(\O_K)$ as 
+the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
+such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.
+
+\begin{prop}
+\label{prop:mulmatrix}
+With the above notations, we have:
+$$\begin{array}{rl}
+& \mathcal P'_{a,b,c}(A,B)(\mathcal L_0)
+= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
+\text{and} &
+\displaystyle 
+[\mathcal P'_{a,b,c}(A,B)(\mathcal L_0) : M_{r,t}(\O_K)] =
+\sum_{m=1}^\infty a^{(m)} b^{(m)}
+\end{array}$$
+where $a^{(m)}$ (resp. $b^{(m)}$) is the number of $a_i$'s (resp.
+of $b_j$'s) which are $\geq m$.
+\end{prop}
+
+\begin{proof}
+We write $A \cdot dB + dA \cdot B = U_A \cdot M \cdot V_B$ with
+$$M = \Delta_A \cdot V_A \cdot dA \cdot V_B^{-1} 
++ U_A^{-1} \cdot dB \cdot U_B \cdot \Delta_B.$$
+When $dA$ varies in $M_{a,b}(\O_K)$ so does $V_A \cdot dA \cdot V_B^{-1}$
+and therefore the first summand in $M$ varies in the subspace of 
+$M_{a,c}(\O_K)$ consisting of matrices whose all entries on the $i$-th
+row have valuation at least $a_i$. Arguing similarly for the second
+summand, we deduce the first statement of the Proposition. The second
+statement is left to the reader.
+\end{proof}
+
+Let us briefly comment on Proposition \ref{prop:mulmatrix} from the 
+perspective of precision. The second statement roughly means that when 
+we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
+a^{(m)} b^{(m)}$ significant digits (at least if $N$ is large enough).
+However these digits are not --- 
+even individually --- attached to some particular entry of $AB$ but are 
+``spread out'' among all its entries. Thus, a usual coordinate-wise 
+precision data does not see this gain of precision in general.
+There is a more pedestrian way to visualize this phenomenon. Remark that
+the product $AB$ can be rewritten:
+$$AB = U_A \cdot P \cdot V_B
+\quad \text{with} \quad 
+P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$$
+Tracking precision in the usual way, we see that the $(i,j)$-th entry
+of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
+gain of precision is visible here. However it is then ``spread out'' 
+when we are computing $U_A \cdot P \cdot V_B$ and mostly disappears in 
+the usual coordinate-wise precision model. The lattice precision model 
+then appears as a way to keep track of this ``diffused'' gain of
+precision.
+
+\medskip
+
+One may wonder to what extend it is important to take care of these 
+``diffused'' gains of precision. Of course, this strongly depends on the 
+problem we are interested in.
+Here is nevertheless a simple --- and hopefully persuasive --- example 
+illustrating that we may gain a lot by considering lattice precision
+data. \todo{Write the example.}
+
+\subsection{Determinant}
+
+We now come to the computation of the determinant of the square $n 
+\times n$ with coefficients in $K$. The differential computation is 
+quite classical: the differential of the function $\det : M_{n,n}(K) \to 
+K$ at some point $M$ is the linear mapping:
+$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM)$$
+where $\com(M)$ stands for the comatrix of $M$. If $M$ is not 
+invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
+\ref{}. Therefore, from now on, we suppose that $M$ is invertible.
+
+As we did for matrix multiplication, we first determine the image of
+the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.
+In order to do so, we write $M = U_M \cdot \Delta_M \cdot V_M$ the 
+Smith factorization of $M$ and denote by $m_1 \leq \ldots \leq m_n$ 
+the valuation of the diagonal entries of $\Delta_M$.
+
+\begin{prop}
+\label{prop:detmatrix}
+Setting $v = m_1 + \cdots + m_{n-1}$, we have
+$\detp(M)(\mathcal L_0) = \pi^v \O_K$.
+\end{prop}
+
+\begin{proof}
+From the description of $\detp(M)$, we see that it is enough to prove 
+that the smallest valuation on an entry of $\com(M)$ is $v$
+or, equivalently, that the ideal of $\O_K$ generated by all 
+minors of $M$ of size $(n-1)$ is $\pi^v
+\O_K$. But this ideal remains unchanged when we multiply $M$ 
+(on the left or on the right) by any unimodular matrix. Thus we may 
+assume that $M$ is $\Delta_M$ and the result becomes clear.
+\end{proof}
+
+In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
+$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
+is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
+digits in absolute precision\footnote{We note that the valuation of 
+$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
+precisions then we should say that we are loosing $m_n$ significant 
+digits.}.
+Furthermore, there exists a easy method for computing the determinant
+of $M$ with this optimal precision. It consists in first computing an
+approximate Smith decomposition of $M$:
+$$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M.$$
+Here $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
+$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
+$\pi^N$ and its diagonal entries $a_1, \ldots, a_m$ have valuations 
+$m_1, \ldots, m_n$ respectively.
+We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
+the $i$-th row of $\tilde \Delta_M$, we obtain
+$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
+where $U$ is a matrix which is congruent to the identity modulo 
+$\pi^{N-m_n}$. Therefore we get:
+$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
+from what we see that we can compute $\det(M)$ at the required precision
+by simply evaluating the product $a_1 \cdots a_m$.
 
 {\color{Bittersweet}
-
-\subsection{Determinant}
-
-We first outline the standard computation of the differential of the function
-$\det : M_n(K) \to K$. Suppose that $M \in \GL_n(K)$ and that $\com(M) = \det(M) M^{-1}$.
-Then
-\begin{align*}
-\det(M + dM) &= \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
-&= \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
-&= \det(M) + \tr(\com(M) \cdot dM).
-\end{align*}
-The differential of $\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
-out that this formula is still valid when $M$ is not invertible.
 
@@
 
-\subsection*{LU factorization.}
+\subsection{LU factorization.}
 
@@
 
-\subsection*{QR factorization.}
+\subsection{QR factorization.}
 
@@
 and $dM$.
+
+\begin{comment}
 
@@
 $N$).
+
+\end{comment}
 

### 2014-12-11 examples/stability.tex
Quick description of algorithms for vector spaces

@@
 \usepackage{comment}
+\usepackage{multirow}
 \usepackage{tikz}
@@
 
-\title{About p-adic stability}
+\title{About p-adic stability in linear algebra}
 
@@
 \begin{abstract}
-Based on the framework developed in \cite{caruso-roe-vaccon:14a}, we study the 
-$p$-adic stability of many classical algorithms involving many usual
-algebraic objects (matrices, univariate polynomials,
-vector spaces). We observe in particular that generic algorithms ---
-which are designed to work over an arbitrary field --- are generally
-very unstable.
+Based on the framework developed in \cite{caruso-roe-vaccon:14a}, we study 
+the $p$-adic stability of standard operations on matrices and vector 
+spaces. 
+%We observe in particular that generic algorithms --- which are 
+%designed to work over an arbitrary field --- are generally very 
+%unstable.
 \end{abstract}
@@
 
+\todo{Rewrite the introduction}
+
+\begin{comment}
 The present paper, which may be considered as a continuation of 
@@
 $K = \Q_2$ is our examples.
+\end{comment}
 
@@
 The section \ref{sec:theory} introduces the theory of ultrametric 
-precision developped in \cite{caruso-roe-vaccon:14a} and completes it on a technical 
-point. The analysis of standard algorithms are treated in the next
-sections: \S\S \ref{sec:numbers}--\ref{sec:vectorspaces} are 
+precision developped in \cite{caruso-roe-vaccon:14a} and completes it 
+on a technical point. In \S \ref{sec:matrices}
+
+The analysis of standard algorithms are treated in 
+the next sections: \S\S \ref{sec:numbers}--\ref{sec:vectorspaces} are 
 successively devoted to the case of numbers, matrices, univariate 
@@
 
-\subsection{The main lemma and its consequences}
-
-\todo{Recall results of \cite{caruso-roe-vaccon:14a}.}
-
-{\color{Bittersweet}
-
-We saw in the core of the article that the differential of an operation
-encodes the intrinsic loss/gain of precision when performing this
-operation. The aim of this appendix is to compute the differential of
-many common operations on numbers, polynomials and matrices. Surprisingly,
-we observe that all differentials we will consider are rather easy to
-compute even if the underlying operation is quite involved (\emph{e.g.}
-Gr\"obner basis).
-
-In what follows, we use freely the ``method of physicists'' to compute
-differentials: given a function $f$ differentiable at some point $x$, we
-consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$
-by expanding LHS and neglecting terms of order $2$. The differential of
-$f$ at $x$ is then the linear mapping $dx \mapsto dy$.
-
-}
+The aim of this section is to summerize briefly the content of 
+\cite{caruso-roe-vaccon:14a} and complete it on certain points.
+
+\subsection{Lattices as precision data}
+
+Roughly speaking, the main proposal of \cite{caruso-roe-vaccon:14a} is 
+to use lattices to track precision when computing with elements lying in 
+vector spaces over $K$.
+
+To make this sentence more explcit, let us consider be a finite 
+dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
+actually those of Banach spaces. However we will not need infinite 
+dimensional spaces in the present paper.} normed vector space $E$ 
+defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
+on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
+closed) ball of radius $r$ centered at the origin. A lattice $L \subset 
+E$ is, by definition, a sub-$\O_K$-module which generates $E$ over $K$. 
+Since we are working in a ultrametric world, the balls $B^{\phantom 
+-}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
+should be thought as special neighborhoods of $0$ and therefore are good 
+candidates to model precision data. Moreover, as revealed in 
+\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
+differentiable maps:
+
+\begin{prop}
+\label{prop:precision}
+Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
+and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
+$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
+that the differential $f'(v_0)$ is surjective.
+Then, for all $\rho \in (0, 1]$, there exists a positive real
+number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
+$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
+satisfies:
+\begin{equation}
+\label{eq:firstorder}
+f(v_0 + H) = f(v_0) + f'(v_0) (H).
+\end{equation}
+\end{prop}
+
+Proposition \ref{prop:precision} can be rephrased in terms of precision 
+as follows: if the input $v_0$ is known at precision $H$, then $f(v_0)$ 
+is known up to $f'(v_0)(H)$ and this precision is optimal. Optimality is 
+reflected by the equality sign in Eq.~\eqref{eq:firstorder}.
+
+In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
+$f$ is locally analytic, then the constant $\delta$ appearing in
+Proposition \ref{prop:precision} can be made explicit in terms of the
+growing function $\Lambda(f)$ of $f$ defined by:
+$$\textstyle \Lambda(f)(v) = 
+\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
+f(h) \Vert \big)$$
+with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
+converge on $B^-_E(e^v)$.
+We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
+precise statement and seek here to state the particular case of
+polynomial functions. A function $f : E \to F$ is said \emph{polynomial} 
+if it is coordinate-wise given by multivariate polynomials in any 
+(equivalently all) system of coordinates associated to orthonormal 
+basis.
+
+\begin{prop}
+\label{prop:precision2}
+We keep the notations of Proposition \ref{prop:precision} and assume 
+in addition that $f$ is polynomial. Let $C$ be a positive real
+number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
+Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
+\rho^{-1}$.
+\end{prop}
 
@@
 
-In the next section, we will compute the derivative of many standard 
-operations and remark that it often has a simple expression in term of 
-the input and the output. In other terms, the function $f$ modeling
+In the next sections, we will compute the derivative of several standard 
+operations and remark that it sometimes has a simple expression in term 
+of the input and the output. In other terms, the function $f$ modeling
 such an operation satisfies a differential equation of the form
@@
 derive from it some bounds on the growing function $\Lambda(f)$. 
-We shall actually work with the slight more general differential 
-equation:
+\emph{In the following, we shall always assume that $K$ has 
+characteristic $0$.}
+
+We actually generalize a bit the setting above and consider the 
+differential equation:
 \begin{equation}
@@
 unknown, which is assumed to be locally analytic as well. The function
-$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
+$f$ goes from $U$ to $V$, two open subsets in two finite dimensional
+normed vector spaces $E$ 
 and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
 is again an open subset in a $K$-Banach space $G$. Consequently, $g$
-goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
-all continuous linear applications $E \to F$.
+goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
+all (continuous) linear applications $E \to F$.
 In what follows, we always assume that $V$ and $W$ contains the origin, 
@@
 accordingly) so that they both vanish at $0$ and second, in order to 
-apply the theory of \cite{caruso-roe-vaccon:14a}, the derivative $f'(0)$ needs to be 
-surjective and therefore \emph{a fortiori} nonzero.
+apply Proposition \ref{prop:precision2}, the derivative $f'(0)$ 
+needs to be surjective and therefore \emph{a fortiori} nonzero.
 
@@
 functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
-\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We assume in addition that 
+\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
 there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
@@
 where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
-e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is 
-$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$ 
-where $p$ denotes the characteristic of the residue field and where, by 
-convention, the above expression equals $0$ if $p = 0$ (we recall that 
-we have assumed that $K$ has characteristic $0$).
+e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is $\alpha = - 
+\frac p {p-1} \cdot \log \Vert p \Vert > 0$ where $p$ denotes the 
+characteristic of the residue field and where, by convention, the above 
+expression equals $0$ if $p = 0$.
+The next Proposition is proved in Appendix \ref{app:proof}.
 
@@
 \end{prop}
-
-\begin{proof}
-See Appendix \ref{app:proof}.
-\end{proof}
 
@@
 
-{\color{Bittersweet}
-
-\section{Numbers}
-\label{sec:numbers}
-
-The most basic operations are, of course, sums, products and inverses of
-elements of $K$. Their differential are well-known and quite easy to
-compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have
-$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = -
-\frac{dx}{x^2}$).
-
-Slightly more interesting is the $n$-th power map $f$ from $K$ to
-itself. Its differential $f'(x)$ is found by differentiating the
-equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
-Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(\lvert nx^{n-1} \rvert \cdot r)$. According to Lemma \ref{lem:main}, this
-means that the behavior of the precision depends on the absolute value
-of $n$. By the ultrametric inequality, we always have $\lvert n \rvert \leq 1$ but
-may have $\lvert n \rvert = 0$ if the characteristic of $K$ divides $n$ or
-$\lvert n \rvert = p^{-k}$ if $K$ is $p$-adic with $\val(n) = k$.  In the first case,
-$f'(x)$ also vanishes and Lemma \ref{lem:main} does
-not apply.  In the second, Lemma \ref{lem:main} reflects the well-known fact that
-raising a $p$-adic number to the $p^k$-th power increases the precision
-by $k$ extra digits.
-
-}
-
 \section{Matrices}
@@
 \end{array}$$
-According to Proposition \ref{}, the behaviour of the precision when 
+According to Proposition \ref{prop:precision}, the behaviour of the precision when 
 computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$. It is obvious 
@@
 lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
-to apply Proposition \ref{}, we then need to compute the image 
+to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
 of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
@@
 we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
-a^{(m)} b^{(m)}$ significant digits (at least if $N$ is large enough).
-However these digits are not --- 
+a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
+note that, on the other side, the valuation of the entries of $AB$ may 
+increase, meaning that we are also loosing some significant digits if 
+we are reasoning in relative precision.} (under the assumption of 
+Proposition \ref{prop:precision2}). However these digits are not --- 
 even individually --- attached to some particular entry of $AB$ but are 
@@
 One may wonder to what extend it is important to take care of these 
-``diffused'' gains of precision. Of course, this strongly depends on the 
-problem we are interested in.
-Here is nevertheless a simple --- and hopefully persuasive --- example 
-illustrating that we may gain a lot by considering lattice precision
-data. \todo{Write the example.}
+``diffused'' gains of precision. For one single product of matrices, 
+it is not so much. However, the benefits becomes really substantial
+when we are executing an algorithm that computes many products of
+matrices. Let us illustrate this with the following quite simple
+example.
+
+\noindent\hrulefill
+
+\noindent {\bf Algorithm 1:} {\tt example\_product}
+
+\noindent{\bf Input:} a list $(M_1, \ldots, M_n)$ of square matrices
+of size $d$.
+
+\smallskip
+
+\noindent 1.\ Set $P$ to the identity matrix of size $d$
+
+\noindent 2.\ {\bf for} $j=1,\dots,n$ {\bf do} {\bf compute} $P = P M_i$
+
+\noindent 3.\ {\bf return} the top left entry of $P$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
+
+\noindent
+Figure \ref{fig:mulmatrix} compares the number of significant digits in 
+\emph{relative} precision we are loosing on the output of Algorithm 1 
+when we are using, on the one hand, a standard component-wise track of 
+precision and, on the other hand, a lattice-based method to handle 
+precision.
+%
+\begin{figure}
+\begin{center}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{|c|c|c|c|}
+\hline
+\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
+\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
+\multicolumn{2}{|c|}{Average loss of precision} \\
+\cline{3-4}
+& & Comp.-wise method & Lattice method \\
+\hline 
+$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
+$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
+$2$ & $1000$ & $157.8$ & $7.9$ \\
+\hline
+$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
+$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
+$3$ & $1000$ & $122.5$ & $7.0$ \\
+\hline
+\end{tabular}
+
+\smallskip
+
+{\small
+Results for a sample of $1000$ random inputs in $M_{d,d}(\Z_2)^n$}
+\end{center}
+\renewcommand{\arraystretch}{1}
+
+\vspace{-0.3cm}
+
+\caption{Average loss of precision in Algorithm 1}
+\label{fig:mulmatrix}
+\end{figure}
+%
+We observe that, in the first case, the number of lost digits seems to 
+grow linearly with respect to the number of multiplications we are 
+performing (that is $n$) whereas, in the second case, the growing seems 
+to be only logarithmic. It would be nice to have a precise formulation 
+(and proof) of this heuristic.
 
@@
 invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
-\ref{}. Therefore, from now on, we suppose that $M$ is invertible.
+\ref{prop:precision}. Therefore, from now on, we suppose that $M$ is invertible.
 
@@
 
-{\color{Bittersweet}
-
-\subsection{Characteristic polynomials.}
-
-The same computation extends readily to characteristic polynomials,
-since they are defined as determinants. More precisely, let us 
-consider the function $\chi : M_n(K) \to K_n[X]$ taking a matrix 
-$M$ to its monic characteristic polynomial $\det(X-M)$.
-Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
-differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.
-
-\subsection{LU factorization.}
-
-Define the LU factorization of a square matrix $M \in M_n(K)$ as a decomposition $M = LU$ 
-where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
-Such a decomposition exists and is unique provided that no principal minor
-of $M$ vanishes. We can then consider the mapping $M \mapsto (L,U)$ 
-defined over the Zariski-open set of matrices satisfying the above 
-condition. In order to differentiate it, we differentiate the relation 
-$M = LU$ and rewrite the result as
-$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
-We remark that in the right hand side of the above formula, the first
-summand is lower triangular with zero diagonal whereas 
-the second summand is upper triangular. Hence in order to compute $dL$
-and $dU$, one can proceed as follows: 
-\begin{enumerate}
-\item compute the product $dX = L^{-1} dM \: U^{-1}$,
-\item separate the lower and upper part of $dX$ obtaining $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$
-\item recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the left and $U$ on the right respectively.
-\end{enumerate}
-The above discussion extends almost \emph{verbatim} to LUP 
-factorization; the only difference is that LUP factorizations are not 
-unique but they are on a small neighborhood of $M$ if we fix the matrix 
-$P$.
-
-\subsection{QR factorization.}
-
-A QR factorization of a square matrix $M \in M_n(K)$ will be
-a decomposition $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
-orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
-a decomposition exists and is unique on a Zariski-open subset of $M_n(K)$. The
-mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
-would like to emphasize at this point that the orthogonality condition
-defines a sub-manifold of $M_n(K)$ which is \emph{not} a
-vector space: it is defined by equations of degree $2$. The codomain
-of $f$ is then also a manifold; this example then fits into the setting
-of Appendix \ref{sec:manifold} but not to those of Section \ref{sec:mainlemma}. 
-We can differentiate $f$ by following the method used for LU 
-factorization.  Differentiating the relation $M = QR$, we obtain
-\begin{equation}
-\label{eq:diffQR}
-\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
-\cdot R^{-1}
-\end{equation}
-where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
-Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
-$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
-matrix and a diagonal one. Since moreover $dR \cdot R^{-1}$ is upper
-triangular with all diagonal entries equal to $0$, we see that 
-\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
-and $dM$.
-
-\begin{comment}
-
-\section{Univariate polynomials}
-\label{sec:polynomials}
-
-For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
-polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
-space of dimension $d$. The affine space $X^d + K_{< d}[X]$ is then 
-the set of monic polynomials over $K$ of degree $d$. We denote it by 
-$K_d[X]$.
-
-\subsection*{Evaluation and interpolation.}
-
-Beyond sums and products (which can be treated as before), two basic 
-operations involving polynomials are evaluation and interpolation.
-Evaluation models the function $(P,x) \mapsto P(x)$, where
-$P$ is some polynomial and $x \in K$. Differentiating
-it can be done by computing
-\begin{equation}
-\label{eq:diffeval}
-(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x).
-\end{equation}
-Here $P'$ denotes the derivative of 
-$P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
-P'(x) dx + dP(x)$.
-
-As for interpolation, we consider a positive integer $d$ and the partial 
-function $f : K^{2d} \to K_{< d}[X]$ which maps the tuple $(x_1, y_1, 
-\ldots, x_d, y_d)$ to the polynomial $P$ of degree less than $d$ such 
-that $P(x_i) = y_i$ for all $i$. The polynomial $P$ exists and is unique 
-as soon as the $x_i$'s are pairwise distinct; the above function $f$ is 
-then defined on this open set. Furthermore, if $(x_1, y_1, \ldots, x_d, 
-y_d)$ is a point in it and $P$ denotes the corresponding interpolation 
-polynomial, \eqref{eq:diffeval} shows that $d y_i = P'(x_i) dx_i + 
-dP(x_i)$ for all $i$. For this, we can compute $dP(x_i)$ from $d x_i$ 
-and $d y_i$ and finally recover $dP$ by performing a new interpolation.
-
-\subsection*{Euclidean division.}
-
-Let $A$ and $B$ be two polynomials with $B \neq 0$. The Euclidean division of 
-$A$ by $B$ is the problem of finding $Q$ and $R$ with $A = BQ + R$ and $\deg R < \deg B$. 
-Differentiating the above equality we find
-\[
-dA - dB \cdot Q = B \cdot dQ + dR,
-\]
-which implies that $dQ$ and $dR$ are respectively obtained as the 
-quotient and the remainder of the Euclidean division of $dA - dB \cdot 
-Q$ by $B$. This gives the differential. We note that the discussion 
-above extends readily to convergent series (see also 
-\cite{caruso-lubicz:14a}).
-
-\subsection*{Greatest common divisors and B\'ezout coefficients.}
-
-We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
-function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
-pair $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
-greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
-coefficients of minimal degrees, computed by the extended 
-Euclidean algorithm.
-The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
-subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
-$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
-of positive degree.  Since $\mathcal V_0$ is dense, $f$ is not continuous outside 
-$\mathcal V_0$. On the contrary, the function $f$ is differentiable, and even locally analytic,
-on $\mathcal V_0$.
-
-Of course, on $\mathcal V_0$, the first component $D$ is constant and 
-therefore $dD = 0$. To compute $dU$ and $dV$, one can simply 
-differentiate the B\'ezout relation $AU + BV = 1$. We get:
-$$A \cdot dU + B \cdot dV = - (dA \cdot U + dB \cdot V)$$
-from which we deduce that $dU$ (resp. $dV$) is obtained as the 
-remainder in the Euclidean division of $U{\cdot}dX$ by $B$ (resp. of 
-$V{\cdot}dX$ by $A$) where $dX = - (dA \cdot U + dB \cdot V)$.
-
-In order to differentiate $f$ outside $\mathcal V_0$, we 
-define the subset $\mathcal V_i$ of $K_n[X] \times 
-K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
-subresultants shows that $\mathcal V_i$ is locally closed with respect to 
-the Zariski topology. In particular, it defines a $K$-manifold in the 
-sense of Appendix \ref{sec:manifold}, and the 
-restriction of $f$ to $\mathcal V_i$ is differentiable. To compute
-its differential, we proceed along the same lines as before: we 
-differentiate the relation $AU + BV = D$ and obtain this way
-$A \cdot dU + B \cdot dV - dD = dX$
-with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
-first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
-whereas the term $dD$ has degree less than $i$. 
-Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
-we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
--dR$. These relations, together with bounds on the degree of $U$ and $V$,
-imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
-Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
-by $\frac A D$).
-
-\medskip
-
-The lesson we may retain from this study is the following. If we have to 
-compute the greatest common divisor of two polynomials $A$ and $B$ known 
-with finite precision, we first need to determine what is its degree. 
-However, the degree function is not continuous --- it is only upper 
-semi-continuous --- and hence cannot be determined with certainty from $A$ and 
-$B$, unless the approximation to $(A,B)$ lies entirely within $\mathcal V_0$.
-
-We therefore need to make an extra hypothesis.  The most natural hypothesis to make
-is that $\gcd(A, B)$ has the maximal possible degree.  The main reason for choosing
-this convention is that if the actual polynomials $A, B \in K[X]$ have a greatest common
-divisor of degree $i$ then there is some precision for which the maximal degree will be
-equal to $i$, whereas if they have a positive
-degree common divisor then no amount of increased precision will eliminate an intersection
-with $\mathcal V_0$.  A second justification is that any other choice would yield a result
-with no precision, since $A$ and $B$ appear to lie in $\mathcal V_i$ at the known precision.
-Once this assumption is made, the computation is possible and one can apply Lemma 
-\ref{lem:main} to determine the precision of the result.
-Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
-itself although there exist pairs of coprime polynomials in any 
-neighborhood of $(A,A)$.
-
-\subsection*{Factorization.}
-
-Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
-product $P_0 = A_0 B_0$, where $A_0$ and $B_0$ are monic and coprime. 
-Hensel's lemma implies that there exists a small neighborhood $\mathcal 
-U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
-uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
-and $B_0$ respectively. Thus, we can consider the map $f : P 
-\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
-polynomials. We want to differentiate $f$ at $P_0$. For this, we 
-differentiate the equality $P = A B$ around $P_0$, obtaining
-\begin{equation}
-\label{eq:difffactor}
-dP = A_0 \cdot dB + B_0 \cdot dA.
-\end{equation}
-where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
-$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
-between $A_0$ and $B_0$, it follows from \eqref{eq:difffactor} that
-$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
-{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).
-
-\subsection*{Finding roots.}
-
-An important special case of the previous study occurs when $A_0$ has 
-degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
-The map $P \mapsto A$ is then nothing but the mapping that 
-follows the simple root $\alpha_0$. Of course, its differential around 
-$P_0$ can be computed by the above method. Nevertheless, we may
-obtain it more directly by expanding the equation $(P_0 + 
-dP)(\alpha_0 + d\alpha) = 0$, obtaining
-$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$.
-Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
-does not vanish and we find:
-$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$
-
-We now address the case of a multiple root. Let $P_0$ be a monic 
-polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
-multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
-to follow the root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
-restrict ourselves to polynomials which have a root of multiplicity $m$ 
-close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
-V_m$ consisting of monic polynomials of degree $d$ having a root of 
-multiplicity at least $m$. It is a Zariski closed subset of $K_d[X]$ and
-contains by assumption the polynomial $P_0$. 
-Moreover, the irreducible components of $\mathcal V_m$ that meet at 
-$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
-$\geq m$. In particular, $\alpha_0$ corresponds to one of these 
-irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
-The algebraic variety $\mathcal V_{m,\alpha_0}$ is \emph{a fortioti} a
-$K$-manifold. Moreover there exists a differentiable function $f$ which 
-is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
-follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
-$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
-multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
-Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
-B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
-\alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
-(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
-$P^{(m-1)}(\alpha) = 0$ where $P^{(i)}$ denotes the $i$-th derivative
-of $P$. We find:
-\begin{equation}
-\label{eq:dalphamult}
-d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
-\end{equation}
-
-The above computation has interesting consequences for $p$-adic 
-precision. For example, consider the monic 
-polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
-large integer) and suppose that we are asking a computer to compute one 
-of its roots. What is the right answer? We remark that, if $\alpha$ is 
-any $p$-adic number divisible by $p^N$, then $P(X)$ can be $X^2 - 
-\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
-the two roots of $P$ are necessarily divisible by $p^N$. Hence, the
-right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
-that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
-\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
-O(p^{2N})$ if $p > 2$ and $\alpha = O(p^{2N-1})$ if $p=2$.
-This is exactly the result we get by applying Lemma \ref{lem:main} and
-using \eqref{eq:dalphamult} to simplify the differential.
-
-This phenomenon is general: if $P$ is a polynomial over $\Qp$ known at 
-some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
-multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
-(\alpha)$ vanish at the given precision) can be computed at precision 
-roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
-multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
-where $c$ is some constant (depending on $P$ and $\alpha$ but not on
-$N$).
-
-\end{comment}
-
-}
+\subsection{Characteristic polynomials}
+
+\todo{Write this subsection}
+
+\subsection{LU factorization}
+
+We refer to \cite[Appendix B]{caruso-roe-vaccon:14a} for the computation
+of the derivative of the LU factorization.
+\todo{Complete this subsection.}
+
+%\subsection{QR factorization}
+%
+%We refer to \cite[Appendix B]{caruso-roe-vaccon:14a} for the computation
+%of the derivative of the QR factorization.
+%\todo{Complete this subsection.}
 
@@
 U^\star_{V, V^\c}$ which commutes with the embeddings into the
-grassmannian. We emphasize that the above bijection is quite easy to
-compute: it consists in replacing the matrix 
-$\Big(\begin{matrix} I \\ H \end{matrix}\Big)$ by $(\begin{matrix} -H & 
-I \end{matrix})$ (in a good choice of coordinates).
+grassmannian.
 
@@
 
-\subsection{Application to precision}
+\subsection{Implementations and experiments}
+
+\noindent
+{\bf Standard representation of vector spaces.}
+A standard way to represent subvector spaces of $K^n$ is to view them 
+through the charts $\mathcal U_{V_I, V_{I^\c}}$ (where $I$ is a subset 
+of $\{1, \ldots, n\}$) introduced above. This means more concretely 
+that a subspace $V \subset K^n$ is represented as the span of the rows
+of a matrix $G_V$ having the following extra property: there 
+exists some $I \subset \{1, \ldots, n\}$ such that the submatrix of 
+$G_V$ obtained by keeping only columns with index in $I$ is the identity 
+matrix. We recall that such a representation always exists and, when 
+the set of indices $I$ is fixed (and has good cardinality), there 
+exists at most one $G_V$ satisfying the above condition.
+Given a family of generators of $V$, one can compute $G_V$ and $I$ as 
+above by performing standard row-echelon. If we are choosing always the 
+first non-vanishing pivot, we get this way the set $I$ with the smallest 
+possible elements. It is the usual choice to get a normalized 
+representation. However, in the context of inexact base fields, it is 
+much more clever to always choose the pivot with maximal norm.
+
+\medskip
+
+\noindent
+{\bf The dual representation.}
+Of course, one may alternatively use the charts $\mathcal U^\star_{V_I, 
+V_{I^\c}}$. Concretely, this means that we represent $V$ as the left 
+kernel of a matrix $H_V$ having the following extra property: there 
+exists some $I \subset \{1, \ldots, n\}$ such that the submatrix of 
+$H_V$ obtained by \emph{deleting} rows with index in $I$ is the identity 
+matrix. Similarly as above, if $V$ is presented as the vanishing of a 
+family of linear equations, we can compute $I$ and $H_V$ by performing 
+column-echelon.
+
+We emphasize that passing from a representation to the other is almost 
+free (and also quite stable). To simplify notations, let us suppose 
+that $I = \{1, \ldots, d\}$ with $d = \dim V$. Then, denoting by $I_d$ 
+the identity matrix of size $d$, the matrix $G_V$ has the shape
+$(\begin{matrix} I_d & G'_V \end{matrix})$
+and it is quite easy to check that one can represent $V$ with the same
+$I$ and the matrix
+$H_V = \Big(\begin{matrix} -G'_V \\ I_{n-d} \end{matrix}\Big)$. Of
+course, for a general $I$, a similar formula exists.
+
+\medskip
+
+\noindent
+{\bf Operations on vector spaces.}
+The first representation we gave is well suited for the computation
+of direct images and sums. For instance, in order to compute $f(V)$, 
+we just have to apply $f$ to each row of $G_V$ (getting this way a
+family of generators of $f(V)$) and then row-echelonize. Dually, the
+second representation is nice for computing inverse images --- and 
+in particular kernels --- and intersections. Since, one know moreover
+how to go back and forth between the two dual representations, we have 
+a complete picture of the underlying algorithmic.
+
+Regarding to precision, one can either track it in the standard way or 
+by using Proposition \ref{prop:precision} and the differential 
+computations we have presented before. Of course, the second method 
+is more costly but also more accurate (as illustrated above).
+
+\medskip
+
+\noindent
+{\bf Some experiments.}
 

### 2014-12-12 examples.tex

--- 2014-08-21 examples.tex
@@
 consider two locally analytic functions $f : U \to V$ and $g : V \to W$ 
-two locally analytic functions around $0$, with $f(0)=0.$ The composition $h = g \circ
+around $0$. The composition $h = g \circ
 f$ is then locally analytic around $0$ as well. Let us write the

### 2014-12-12 resultants/resultants.tex

@@
 \usepackage{amsmath,amssymb}
-\usepackage{amsrefs}
+%\usepackage{amsrefs}
 \usepackage[usenames,dvipsnames]{color}
@@
 \usepackage{bbm}
+\usepackage{multirow}
 \renewcommand*{\theHsection}{\thesection}
@@
 \newcommand{\lc}{\text{\rm lc}}
+\newcommand{\Card}{\text{\rm Card}\:}
 
@@
 
-\title{Resultants and subresultants\\of p-adic polynomials}
+\title{Resultants and subresultants of p-adic polynomials}
 
@@
 \begin{abstract}
+We address the problem of the stability of the computations of resultants 
+and subresultants of polynomials defined over complete discrete valuation 
+rings (\emph{e.g.} $\Zp$ or $k[[t]]$ where $k$ is a field). We prove that 
+Euclide-like algorithms are highly unstable in average and we explain, in
+many cases, how one can stabilize them without sacrifying the complexity.
+On the way, we completely determine the distribution of the valuation of
+the subresultants of two random monic $p$-adic polynomials having the
+same degree.
 \end{abstract}
@@
  \noindent
- {\bf Keywords:} 
+ {\bf Keywords:} $p$-adic precision, subresultants
 \medskip
@@
 
-The strating point of this paper is the willing to desing a reasonable 
-algorithm to compute \textsc{gcd}s and B\'ezout coefficients of 
-polynomials with coefficients in $\Zp$ (the ring of $p$-adic integers). 
-Indeed, easy experimentations show that all standard algorithms for this 
-(\emph{e.g.} extended Euclide's algorithm) are not satisfying because of 
-stability.
-
-Stability issues occur in this context because, just like real numbers, 
-$p$-adic numbers are by nature infinite and thus need to be truncated in 
-order to fit in the memory of a computer. The level of truncation is 
-called the \emph{precision}. Usual software implementing $p$-adics (as 
-\textsc{magma}, \textsc{sage}) generally tracks the precision as follows:
-an individual precision is attached to any $p$-adic variable and this
-precision is updated after each basic arithmetic operation (addition
-and multiplication). This way of tracking precision can be seen as the
-analogue of the arithmetic intervals in the real setting.
-
-Figure \ref{fig:} illustrates the unstability of Euclide's algorithm 
-which it is called on random inputs which are monic $2$-adic polynomials 
-of fixed degree. Let us explain how this table should be read by 
-commenting on its last line. It indicates that (extended) Euclide's 
-algorithm outputs the B\'ezout coefficients of two monic $2$-adic 
-polynomials of degree $500$ with an average loss of $...$ significant 
-digits by coefficient whereas a stable algorithm should only loose $...$ 
-digits in average. This ``theoretical'' loss is computed as the double 
-of the valuation of the resultant. Indeed Cramer-like formulae imply 
-that B\'ezout coefficients can be computed by performing a unique 
-division by the resultant, inducing then only the aforementioned loss 
-of precision. Examining the table a bit more, we furthermore observe 
-that the ``practical'' loss of precision due to Euclide's algorithm 
-seems to grow linearly with respect to the degree whereas the 
-``theoretical'' loss seems to be independant on the degree. In other
-words, the unstability of Euclide's algorithm is becoming more and
-more critical when the degree of the input increases.
+As wonderfully illustrated by the success of Kedlaya-type counting points 
+algorithms \cite{kedlaya}, $p$-adic technics are gaining nowadays more and more 
+popularity in computer science, and more specifically in symbolic 
+computation. A crucial issue when dealing with $p$-adics is those of 
+stability. Indeed, just like real numbers, $p$-adic numbers are by nature 
+infinite and thus need to be truncated in order to fit in the memory of a 
+computer. The level of truncation is called the \emph{precision}. Usual 
+softwares implementing $p$-adics (\emph{e.g.} \textsc{magma} \cite{magma}, 
+\textsc{pari} \cite{pari}, \textsc{sage} \cite{sage}) 
+generally tracks the precision as follows: an individual precision is 
+attached to any $p$-adic variable and this precision is updated after 
+each basic arithmetic operation (addition and multiplication). This way 
+of tracking precision can be seen as the analogue of the arithmetic 
+intervals in the real setting.
+
+In the paper \cite{padicprec}, the authors propose a new framework to control 
+$p$-adic precision. The aim of this paper is to illustrate the technics 
+of \emph{loc. cit.} on the concrete example of computation of 
+\textsc{gcd}s and subresultants of $p$-adic polynomials. There is 
+actually a real need to do this due to the combination of two reasons: on 
+the one hand, computating \textsc{gcd}s is a very basic operation for 
+which it cannot be acceptable to have important instability whereas, on 
+the other hand, easy experimentations show that all standard algorithms 
+for this task (\emph{e.g.} extended Euclide's algorithm) are \emph{very} 
+unstable.
+%
+\begin{figure}
+\renewcommand{\arraystretch}{1.2}
+\begin{center}
+\begin{tabular}{|c||c|c|}
+\hline
+\multirow{2}{*}{\hspace{0.2cm}Degree\hspace{0.2cm}\null} & 
+\multicolumn{2}{c|}{\hspace{0.3cm}Loss of precision (in number of digits)\hspace{0.3cm}\null}\\
+\cline{2-3}
+& \hspace{0.3cm}Euclide algorithm\hspace{0.3cm}\null & expected \\
+\hline
+$5$ & $\phantom{00}6.3$ & $3.1$ \\
+$10$ & $\phantom{0}14.3$ & $3.2$ \\
+$25$ & $\phantom{0}38.9$ & $3.2$ \\
+$50$ & $\phantom{0}79.9$ & $3.2$ \\
+$100$ & $160.0$ & $3.2$ \\
+\hline
+\end{tabular}
+\end{center}
+
+\vspace{-0.3cm}
+
+\caption{Average loss of precision when computing the {\sc gcd}
+of two random monic polynomial of fixed degree over $\Z_2$.}
+\label{fig:precision}
+
+\vspace{-0.3cm}
+\end{figure}
+%
+Figure \ref{fig:precision} illustrates the instability of extended Euclide's 
+algorithm when it is called on random inputs which are monic $2$-adic 
+polynomials of fixed degree. Looking at the last line, we see that 
+extended Euclide's algorithm outputs the B\'ezout coefficients of two 
+monic $2$-adic polynomials of degree $100$ with an average loss of 
+$161.1$ significant digits by coefficient whereas a stable 
+algorithm should only loose \todo{$4$} digits in average. This 
+``theoretical'' loss is computed as the double of the valuation of the 
+resultant. Indeed Cramer-like formulae imply that B\'ezout coefficients 
+can be computed by performing a unique division by the resultant, 
+inducing then only the aforementioned loss of precision. Examining the 
+table a bit more, we furthermore observe that the ``practical'' loss of 
+precision due to Euclide's algorithm seems to grow linearly with respect 
+to the degree of the input polynomials whereas the ``theoretical'' loss 
+seems to be independant of it. In other words, the unstability of 
+Euclide's algorithm is becoming more and more critical when the degree 
+of the input increases.
 
@@
 
-[...]
+\noindent
+{\bf Content of the paper.}
+The aim of this article is twofold. We first provide in \S 
+\ref{sec:unstable} a theoretical study of the instability phenomenon 
+described above and give strong evidences that the loss of precision 
+grows linearly with respect to the degree of the input polynomials, as we 
+observed empirically. In doing so, we determine the distribution of the 
+valuation of the subresultants of random monic polynomials over $\Zp$ 
+(\emph{cf} Theorem \ref{th:lawVj}). This is an independant result which 
+has its own interest.
+
+Our second goal, which is carried out in \S \ref{sec:stable}, is to rub 
+out these unexpected losses of precision. Making slight changes to the 
+standard subresultants pseudo-remainder sequence algorithm and using in 
+an essential way the results of \cite{padicprec}, we manage to design a stable 
+algorithm for computing all subresultants of two monic polynomials over 
+$\Zp$ (satisfying an additional mild assumption). This basically allows 
+to stably compute \textsc{gcd}s assuming that the degree of the 
+\textsc{gcd} is known in advance.
 
@@
 
+The aim of this section is to introduce the setting we shall work in 
+throughout this paper (which is a bit more general than those considered 
+in the introduction).
+
 \subsection{Complete discrete valuation rings}
 
-bla
-
-\medskip
-
 \noindent
 \textbf{Definitions.}
-Let $\A$ be a \emph{discrete valuation ring}. We recall that it is a 
+Let $\A$ be a \emph{discrete valuation ring} (DVR for short), that is a 
 domain equipped with a map $\val : \A \to \Z \cup \{+\infty\}$ --- the 
-so-called \emph{valuation} --- satisfying the four axioms: (1)~$\val(x) = 
-+\infty$ iff $x = 0$, (2)~$\val(xy) = \val(x) + \val(y)$ and (3)~$\val(x+y) \geq 
-\min(\val(x), \val(y))$, (4)~any element of valuation $0$ is invertible.
-We also recall that a discrete valuation ring admits a unique maximal 
-ideal $\m$, consisting of elements of positive valuation. 
-Furthermore, this ideal is principal and generated by any element of 
-minimal valuation. Such an element is called a \emph{uniformizer}.
-Let us fix one of them and denote it by $\pi$. The \emph{residue
-field} of $W$ is the quotient $W/\m = W/\pi W$ and we shall denote
-it by $k$.
+so-called \emph{valuation} --- satisfying the four axioms: (1)~$\val(x) 
+= +\infty$ iff $x = 0$, (2)~$\val(xy) = \val(x) + \val(y)$ and 
+(3)~$\val(x+y) \geq \min(\val(x), \val(y))$, (4)~any element of 
+valuation $0$ is invertible. In the sequel, we shall assume that the
+valuation is normalized so that it takes the value $1$. We recall that a 
+DVR admits a unique maximal ideal $\m$, consisting of elements of 
+positive valuation. Furthermore, this ideal is principal and generated 
+by any element of valuation $1$. Such an element is called a 
+\emph{uniformizer}. Let us fix one of them and denote it by $\pi$. The 
+\emph{residue field} of $W$ is the quotient $W/\m = W/\pi W$ and we 
+shall denote it by $k$.
 
@@
 \label{eq:expandCDVR}
-x = \sum_{i=0}^\infty x_i \pi^i
+x = x_0 + x_1 \pi + x_2 \pi^2 + \cdots + x_n \pi^n + \cdots
 \end{equation}
@@
 structure may vary.
-
 Let $K$ denote the fraction field of $\A$. The valuation $v$ extends 
 uniquely to $K$ by letting $\val(\frac x y) = \val(x) - \val(y)$. 
-Moreover, it follows from the axiom (4) that $K$ is obtain from $\A$ by 
+Moreover, it follows from the axiom~(4) that $K$ is obtain from $\A$ by 
 inverting $\pi$. Thus, any element of $K$ can be uniquely written as an 
-infinite sum:
-\begin{equation}
-\label{eq:expandCDVF}
-x = \sum_{i=v}^\infty x_i \pi^i
-\end{equation}
-where $v$ is some relative integer and the $x_i$'s are as above. The
-valuation of $x$ can be easily read on this writing: it is the smallest 
-integer $v$ such that $x_i \not\equiv 0 \pmod \pi$.
-
-\medskip
-
-\noindent
-\textbf{Examples.}
-The two main examples of complete discrete valuation rings are:
-\begin{enumerate}[(1)]
-\vspace{-0.25cm}
-\item the ring of $p$-adic integers $\Zp$: the valuation on it is
-the usual $p$-adic valuation, a uniformizer is $p$, its residue
-field is $\Fp = \Z/p\Z$ and Eq.~\eqref{eq:expandCDVR} corresponds
-to standard expansion in $p$-basis of a $p$-adic number; the fraction 
-field of $\Zp$ is the so-called field of $p$-adic numbers and is usually 
-denoted by $\Qp$;
-\vspace{-0.25cm}
-\item the ring of formal series $k[[x]]$ where $k$ is a field: the
-valuation on it is the usual valuation of series, a uniformizer is
-$x$, the residue field is $k$ and Eq.~\eqref{eq:expandCDVR} is the
-usual writing of formal series; the fraction field of $k[[x]]$ is
-the field of Laurent series $k((x))$.
-\end{enumerate}
-\vspace{-0.2cm}
+infinite sum $x = \sum_{i=v}^\infty x_i \pi^i$ where $v$ is some 
+relative integer and the $x_i$'s are as above. The valuation of $x$ can 
+be easily read on this writing: it is the smallest integer $v$ such that 
+$x_i \not\equiv 0 \pmod \pi$.
+
+The two main examples of complete discrete valuation rings are the 
+ring of $p$-adic integers $\Zp$ and the ring of formal series 
+$k[[x]]$ over a field $k$. Their fraction fields are $\Q_p$ and
+$k((x))$ respectively.
 
@@
 this means that we work with approximations of elements of $\A$ of the
-form:
-$$\sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N)$$
+form
+$\sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N)$
 where $N$ is some integer called the \emph{precision} of $x$ and 
@@
 \subsection{Subresultants}
+\label{subsec:subres}
 
 A first issue when dealing with numerical computations of \textsc{gcd}s 
-of polynomials over $\A$ is that the $\gcd$ function is not continuous: 
-it takes the value $1$ on an open dense subset without being constant. 
-This of course annihilate any hope of computing \textsc{gcd}s of 
-polynomials when only approximations of them are known. Fortunately, 
+of polynomials over $\A$ is that the \textsc{gcd} function is not 
+continuous: it takes the value $1$ on an open dense subset without being 
+constant. This of course annihilate any hope of computing \textsc{gcd}s 
+of polynomials when only approximations of them are known. Fortunately, 
 there exists a standard way to recover continuity in this context: it 
-consists in replacing \textsc{gcd}s by the so-called subresultants 
-(whose definitions are main properties are quickly recalled below) which 
-are playing an analoguous role. For this reason, in what follows, we 
-will exclusively consider the problem of computing subresultants.
+consists in replacing \textsc{gcd}s by the so-called subresultants which 
+are playing an analoguous role. For this reason, in what follows, we will 
+exclusively consider the problem of computing subresultants.
 
@@
 \noindent
-\textbf{Notations.} We refer to \cite{} for a nice and rather complete 
+\textbf{Notations.} We refer to \cite{resultant} for a nice and rather complete 
 introduction to resultants and subresultants and seek to fix notations
@@
 with coefficients in $\ring$. Let $j$ be the smallest integer such
-that $R_j(A,B)$ does not vanish. Then $R_j(A,B)$ is a \textsc{gcd}
+that $\Res_j(A,B)$ does not vanish. Then $\Res_j(A,B)$ is a \textsc{gcd}
 of $A$ and $B$.
@@
 respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
-and $A$ and $B$ are polynomials over $\ring$ then:
-$$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$$
+and $A$ and $B$ are polynomials over $\ring$ then
+$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$
 where $f(A)$ and $f(B)$ denotes the polynomials deduced from $A$
@@
 $$\left\{\begin{array}{ll}
-S_0 = A, \, S_1 = B, \, c_0 = 1 \smallskip \\
+S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
 \displaystyle S_{i+1} = (-s_i)^{\delta_i + 1}
@@
 Here $n_i = \deg S_i$ and $s_i$ is the leading coefficient of 
-$S_i$ if $i > 0$ and $s_0 = 1$ by convention. These sequences are
-finite and the above recurrence applies until $S_i$ has reached the
-value $0$.
-
-% \delta_i = n_{i-1} - n_i
+$S_i$ if $i \geq 0$ and $s_{-1} = 1$ by convention. These sequences 
+are finite and the above recurrence applies until $S_i$ has reached 
+the value $0$.
 
@@
 \label{prop:prem}
-We keep the above notations. For all positive integer $i$ such that
-$S_i$ is defined, we have:
-$$S_i = \Res_{\deg S_{i-1} - 1} (A,B).$$
+With the above notations, we have:
+$$\begin{array}{rcll}
+\Res_j(A,B) & = & S_i & \text{if } j = n_{i-1} - 1 \smallskip \\
+& = & 0 & \text{if } n_i < j < n_{i-1} - 1 < j \\
+& = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\delta_i-1} \cdot S_i & 
+\text{if } j = n_i
+\end{array}$$
+for all $i$ such that $S_i$ is defined.
 \end{prop}
@@
 \begin{rem}
+The Proposition \ref{prop:prem} computes \emph{all} subresultants. We
+note moreover that, in the common case where $n_{i-1} = n_i - 1$, the
+two formulas giving $\Res_{n_i}(A,B)$ agree.
+
 Mimicing ideas behind extended Euclide's algorithm, one can define the 
@@
 prove that the degrees of the $S_i$'s decrease by one at each step, 
-meaning that we $\deg S_i = d_B + 1 - i$ for all $i$. The sequence 
-$(S_i)$ then stops at $i = d_B + 1$. Moreover, the $\delta_i$'s and the 
-$c_i$'s are no longer necessary since we have $\delta_i = 1$ and $c_i = 
-s_i$ for all $i$. The recurrence formula then becomes:
-$$\left\{\begin{array}{ll}
-S_0 = A, \, S_1 = B \smallskip \\
-\displaystyle S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
-\cdot (S_{i-1} \,\%\, S_i) & \text{for } i \geq 1.
-\end{array}\right.$$
+meaning that we $\deg S_i = d_B - i$ for all $i$. The sequence 
+$(S_i)$ then stops at $i = d_B$. Moreover, the $\delta_i$'s and the 
+$c_i$'s are no longer necessary since we have $\delta_i = 1$ and $c_i
+= s_i$ for all $i$. The recurrence formula then becomes:
+$$S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
+\cdot (S_{i-1} \,\%\, S_i) \quad \text{for } i \geq 1.$$
 To conclude with, we remark that Proposition \ref{prop:prem} simply 
-states that $R_j = S_{d_B+1-j}$; we then get this way all subresultants.
+states that $R_j = S_{d_B-j}$.
 
 \section{Euclide-like algorithms\\are unstable}
+\label{sec:unstable}
 
@@
 \quad \text{with } a_i \in \Qp \text{ and } N \in \Z.$$
-In other words, we shall assume that the software we are using does not
+In other words, we assume that the software we are using does not
 carry a precision data on each coefficient but only a unique precision
@@
 
-Let $V_j$ be the valuation of the leading coefficient of $R_j$ and $W_j$ 
-be the smallest valuation of a coefficient of $R_j$. We of course have 
-$V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
+Let $V_j$ be the valuation of the scalar $j$-th subresultant of $(A,B$) 
+and $W_j$ be the smallest valuation of a coefficient of $R_j$. We of 
+course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
 
@@
 \label{prop:precEuclide}
-If $A$ and $B$ are known at precision $O(p^N)$ then, in the flat 
-precision model, the subresultant pseudo-remainder sequence computes 
-$R_j$ at precision $O(p^{N_j})$ with
-$$N_j \leq N + V_{j+1} - 2 \sum_{k=j+1}^d \delta_j.$$
+Let $A$ and $B$ as above. We assume that all scalar subresultants of 
+$(A,B)$ do not vanish. Then, in the flat precision model, the 
+subresultant pseudo-remainder sequence computes $R_j$ at precision 
+$O(p^{N_j})$ with:
+$$N_j \leq N + V_{j+1} - 2 \cdot (\delta_{j+1} + \delta_{j+2} + \cdots
++ \delta_{d-1}).$$
 \end{prop}
@@
 \begin{proof}
+By our assumptions, the subresultant pseudo-remainder sequence 
+is defined by the recurrence:
+$$R_{j-1} = r_j^2 \cdot r_{j+1}^{-2} \cdot 
+(R_{j+1} \,\%\, R_j).$$
+Using that $R_j$ and $R_{j-1}$ have the expected degrees, the
+remainder $(R_{j+1} \,\%\, R_j)$ is computed as follows:
+$$\begin{array}{rrcl}
+\text{we set:} & S & = & R_{j+1} - r_{j+1} \cdot r_j^{-1} \cdot R_j \smallskip \\
+\text{and we have:} & R_{j+1} \,\%\, R_j & = & S - s \cdot r_j^{-1} \cdot R_j
+\end{array}$$
+where $s$ is the coefficient of degree $j$ of $S$. Let us first estimate 
+the precision of $S$. Recall for that the relative precision is defined 
+as the difference between the (absolute) precision and the valuation: 
+intuitively, it is the number of significant digits. It is easy to check 
+that the relative precision of a product is the minimum of those of the 
+factors. 
+Thus the relation precision of $r_{j+1} \cdot r_j^{-1} \cdot R_j$ is
+$\min(N_{j+1} - V_{j+1}, N_j - V_j)$. Its absolute precision is then
+$M = \min(N_{j+1} - \delta_j, N_j - \delta_j + V_{j+1} - V_j)$. This
+quantity is also the precision of $S$ since the other summand
+$R_{j+1}$ is known with higher precision. Repeating the argument, we
+find that the precision of $(R_{j+1} \,\%\, R_j)$ is equal to 
+$\min(M - \delta_j, N_j - \delta_j + \val(s) - V_j)$ and therefore is 
+lower bounded by $M - \delta_j \leq N_j - 2 \delta_j + V_{j+1} - V_j$.
+We derive for this $N_{j-1} \leq N_j - 2 \delta_j - V_{j+1} + V_j$ and
+the proposition finally follows by summing up these inequalities.
 \end{proof}
 
-\begin{rem}
-\end{rem}
+The difference 
+$N - N_0 = - V_1 + 2 \sum_{k=1}^d \delta_j$
+is a lower bound on the number of digits lost after having computed the 
+resultant using the subresultant pseudo-remainder sequence algorithm. In 
+the next subsection (\emph{cf} Corollary \ref{cor:Vj}), we shall see that 
+$V_1$ and all $\delta_j$'s are 
+approximatively equal to $\frac 1 {p-1}$ in average. The loss of 
+precision then grows linearly with respect to $d$. This confirms the 
+precision benchmarks shown in Figure \ref{fig:precision}.
+We emphasize one more time that this loss of precision is \emph{not}
+intrinsic but an artefact of the algorithm we have used; indeed, one
+should not loose any precision when computing resultants because they
+are given by polynomial expressions.
 
@@
 the set of monic polynomials of degree $d$ with coefficients in $\A$.
+In what follows, we assume that the residue field of $\A$ is finite
+and we use the letter $q$ to denote its cardinality.
 
@@
 Let $X_0, \ldots, X_{d-1}$ be $d$ pairwise independant discrete random 
-variables with geometric law of parameter $(1 - p^{-1})$, \emph{i.e.}
-$$\P[X_i = k] = (1 - p^{-1}) \cdot p^{-k} \quad 
+variables with geometric law of parameter $(1 - q^{-1})$, \emph{i.e.}
+$$\P[X_i = k] = (1 - q^{-1}) \cdot q^{-k} \quad 
 \text{(with } 0 \leq i < d \text{ and } k \in \N \text{)}.$$
 Then $V_j$ is distributed as the random variable
-$$Y_j = \sum_{i=0}^d \min(X_{j-i}, X_{j-i+1}, \ldots, X_{j+i})$$
-where, by convention, $X_i = +\infty$ if $i < 0$ and $X_i = 0$ if 
-$i \geq d$.
+\begin{center}
+$\displaystyle Y_j = \sum_{i=0}^d \min(X_{j-i}, X_{j-i+1}, \ldots, X_{j+i})$
+\end{center}
+\noindent
+with $X_i = +\infty$ if $i < 0$ and $X_i = 0$ if $i \geq d$.
 \end{theo}
@@
 the tuple $(V_{d-1}, \ldots, V_0)$ is distributed as $(Y_{d-1}, \ldots,
-Y_0)$. For instance, an explicit computation shows that $(V_{d-1}, 
+Y_0)$. For instance, one can prove that $(V_{d-1}, 
 V_{d-2})$ is distributed as 
-$$(X, \, Y + \min(Y, [X/2]))$$
-where $X$ and $Y$ are two independant discrete random variables with 
-geometric law of parameter $(1 - p^{-1})$ and the notation $[\cdot]$ 
+$(X, \, X' + \min(X', [X/2]))$
+where $X$ and $X'$ are two independant discrete random variables with 
+geometric law of parameter $(1 - q^{-1})$ and the notation $[\cdot]$ 
 stands for the integer part function. In particular, we observe that 
-the couple $(2,1)$ does not occur almost surely\footnote{Actually, one
-can prove that it never occurs.} although the events $\{V_{d-1} = 2\}$ 
-and $\{V_{d-2} = 1\}$ both occur with positive probability.
+the couple $(2,1)$ does not occur almost surely although the events 
+$\{V_{d-1} = 2\}$ and $\{V_{d-2} = 1\}$ both occur with positive 
+probability.
+
+Nonetheless, a consequence of Proposition \ref{prop:distribk} below is 
+that the variables $\bar V_j = \mathbbm{1}_{\{V_j = 0\}}$ are mutually 
+independant.
 \end{rem}
@@
 \begin{theo}
-\label{th:estimWj}
+\label{th:deltaj}
 For all $j \in \{0, \ldots, d-1\}$ and all $m \in \N$, we have:
-$$\P[W_j \geq m] \geq (1 - p^{-1}) \cdot p^{-jm}.$$
+$$\P[\delta_j \geq m] \geq \frac{(q-1)(q^j-1)}{q^{j+1}-1} q^{-m}.$$
 \end{theo}
@@
 The proof of these two Theorem is given in \S \ref{subsec:proof}. 
-Let us now derive some consequences. We will need the permutation
-$\sigma$ defined by:
-$$\begin{array}{rcl}
-\sigma & = & \left(
+We now derive some consequences. Let $\sigma$ denote the following
+permutation:
+$$\begin{array}{cl}
+\left(
 \begin{array}{cccccccc}
@@
 1 & 3 & \cdots & d-1 & d & d-2 & \cdots & 2 
-\end{array} \right) \\
-& & \hspace{5cm} \text{if $d$ is even} \medskip \\
-\sigma & = & \left(
+\end{array} \right) & \text{if } 2 \mid d \medskip \\
+\left(
 \begin{array}{cccccccc}
@@
 1 & 3 & \cdots & d & d-1 & d-3 & \cdots & 2 
-\end{array} \right) \\
-& & \hspace{5cm} \text{if $d$ is odd.} \\
+\end{array} \right) & \text{if } 2 \nmid d.
 \end{array}$$
@@
 
-\vspace{-0.15cm}
 \begin{enumerate}[\hspace{0.3cm}(1)]
+\setlength\itemsep{0.1em}
 \item $\E[V_j] = 
-\displaystyle \sum_{i=1}^{d-j} \frac 1 {p^{\sigma(i)} - 1}$ \\
-in particular $\frac 1 {p-1} \leq \E[V_j] < \frac p {(p-1)^2}$
+\displaystyle \sum_{i=1}^{d-j} \frac 1 {q^{\sigma(i)} - 1}$ \\
+in particular $\frac 1 {q-1} \leq \E[V_j] < \frac q {(q-1)^2}$
+\item $\frac {q^j - 1}{q^{j+1} - 1} \leq \E[\delta_j] \leq \E[V_j]$
 \item $\sigma[V_j]^2 = 
-\displaystyle \sum_{i=1}^{d-j} \frac {(2i-1) \cdot p^{\sigma(i)}}
-{(p^{\sigma(i)} - 1)^2}$ \\
-in particular $\frac{\sqrt p}{p-1} \leq \sigma[V_j] < 
-\frac {p \:\sqrt{p+1}} {(p-1)^2}$
-\item $\P[V_j \geq m] \leq p^{-m + O(\sqrt m)}$
-\item $\E[\max(V_0, \ldots, V_{d-1})] \leq \log_p d + O(\sqrt{\log_p d})$
+\displaystyle \sum_{i=1}^{d-j} \frac {(2i-1) \cdot q^{\sigma(i)}}
+{(q^{\sigma(i)} - 1)^2}$ \\
+in particular $\frac{\sqrt q}{q-1} \leq \sigma[V_j] < 
+\frac {q \:\sqrt{q+1}} {(q-1)^2}$
+\item $\P[V_j \geq m] \leq q^{-m + O(\sqrt m)}$
+\item $\E[\max(V_0, \ldots, V_{d-1})] \leq \log_q d + O(\sqrt{\log_q d})$
 \end{enumerate}
@@
 is then the minimum of $\tau_j(i)$ independant random variables with 
-geometric distribution of parameter $(1 - p^{-1})$ and thus its 
-distribution is geometric of parameter $(1 - p^{-\tau(i)})$.
-Its expected value is then $\frac 1 {p^{\tau(i) - 1}}$ and the first
-formula follows. The inequality $\frac 1 {p-1} \leq \E[V_j]$ is clear
-because $\frac 1 {p-1}$ is the first summand in the expression of 
+geometric distribution of parameter $(1 - q^{-1})$ and thus its 
+distribution is geometric of parameter $(1 - q^{-\tau(i)})$.
+Its expected value is then $\frac 1 {q^{\tau(i) - 1}}$ and the first
+formula follows. The inequality $\frac 1 {q-1} \leq \E[V_j]$ is clear
+because $\frac 1 {q-1}$ is the first summand in the expansion of
 $\E[V_j]$. The upper bound is obtained as follows:
-$$\E[V_j] < \sum_{i=0}^\infty \frac 1{p^i - 1}
-\leq \sum_{i=0}^\infty \frac 1{p^i - p^{i-1}} = \frac p {(p-1)^2}.$$
-
-The variance of $V_j$ is related to the covariance 
-of $Z_i$'s thanks to the formula
+$$\E[V_j] < \sum_{i=0}^\infty \frac 1{q^i - 1}
+\leq \sum_{i=0}^\infty \frac 1{q^i - q^{i-1}} = \frac q {(q-1)^2}.$$
+
+The first inequality of claim~(2) is obtained from the relation 
+$\E[\delta_j] = \sum_{m=1}^\infty \P[\delta_j \geq m]$ using the 
+estimation of Theorem \ref{th:deltaj}. The second inequality is 
+clear because $\delta_j \leq V_j$.
+
+The variance of $V_j$ is related to the covariance of $Z_i$'s thanks to 
+the formula
 $\Var(V_j) = \sum_{1 \leq i,i' \leq d-j} \Cov(Z_i, Z_{i'})$.
@@
 Applying this to our setting, we get:
-$$\Cov(Z_i, Z_{i'}) = \frac{p^{e(i,i')}}{(p^{e(i,i')}-1)^2}$$
+$$\Cov(Z_i, Z_{i'}) = \frac{q^{e(i,i')}}{(q^{e(i,i')}-1)^2}$$
 where $e(i,i') = \min(\tau(i), \tau(i')) = \tau(\min(i,i'))$. Summing
-up these contributions, we get the equality in~(2). The inequalities
+up these contributions, we get the equality in~(3). The inequalities
 are derived from this similarly to what we have done in~(1).
@@
 of independant random variable having all geometric distribution of 
-parameter $(1 - p^{-1})$. We set $Z = \sum_{i=1}^\infty \min(Z_1, \ldots, 
+parameter $(1 - q^{-1})$. We set $Z = \sum_{i=1}^\infty \min(Z_1, \ldots, 
 Z_i)$. Cleary $V_j \leq Z$ and it is then enough to prove:
-$$\P[Z \geq m] \leq p^{-m + O(\sqrt m)}.$$
+$$\P[Z \geq m] \leq q^{-m + O(\sqrt m)}.$$
 We introduce the event $E_m$ formulated as follows:
@@
 the sum runs over all partitions $(m_1, \ldots, m_\ell)$ of $m$. 
-Replacing $\P[X_1 \geq m_i]$ by its value $p^{-m_i}$, we get:
-$$\P[E_m] \leq p(m) \cdot p^{-m}$$
+Replacing $\P[X_1 \geq m_i]$ by $q^{-m_i}$, we get
+$\P[E_m] \leq p(m) \cdot q^{-m}$
 where $p(m)$ denotes the number of partitions of $m$. By a famous 
 formula \cite{andrews}, we know that $\log p(m)$ is equivalent to $\pi 
-\sqrt{2m/3}$. In particular it is in $p^{O(\sqrt m)}$ and~(3) is proved.
-
-We now derive (4) by a standard argument. It follows from~(3) that:
-$$\P[\max(V_0, \ldots, V_{d-1})] \leq d \cdot p^{-m + c\sqrt m}$$
+\sqrt{2m/3}$. In particular it is in $q^{O(\sqrt m)}$ and~(4) is proved.
+
+We now derive~(5) by a standard argument. It follows from~(4) that
+$\P[\max(V_0, \ldots, V_{d-1})] \leq d \cdot q^{-m + c\sqrt m}$
 for some constant $c$. Therefore:
-$$\E[\max(V_0, \ldots, V_{d-1}] \leq \sum_{m=1}^\infty \min(1,
-d \cdot p^{-m + c\sqrt m}).$$
-Let $m_0$ denote the smallest index such that $d \cdot p^{-m_0 + c\sqrt 
-m_0}$, \emph{i.e.} $m_0 - c \sqrt{m_0} \geq \log_p d$. Solving the latest
-equation, we get $m_0 = \log_p + O(\sqrt{\log_p d})$. Moreover
-$\sum_{m=m_0}^\infty d\: p^{-m + c\sqrt m}$ is bounded independantly of
+$$\E[\max(V_0, \ldots, V_{d-1})] \leq \sum_{m=1}^\infty \min(1,
+d \cdot q^{-m + c\sqrt m}).$$
+Let $m_0$ denote the smallest index such that $d \cdot q^{-m_0 + c\sqrt 
+m_0}$, \emph{i.e.} $m_0 - c \sqrt{m_0} \geq \log_q d$. Solving the latest
+equation, we get $m_0 = \log_q + O(\sqrt{\log_q d})$. Moreover
+$\sum_{m=m_0}^\infty d\: q^{-m + c\sqrt m}$ is bounded independantly of
 $d$. The result follows.
@@
 
-\begin{cor}
-\label{cor:Wj}
-For all $j \in \{0, \ldots, d-1\}$, we have:
-$$\E[W_j] \leq (1 - p^{-1}) \cdot \frac 1 {p^j - 1}.$$
-\end{cor}
-
-\begin{proof}
-Writing $\E[W_j] = \sum_{m=1}^\infty \P[W_j \geq m]$, the corollary
-follows directly from Theorem \ref{th:estimWj}.
-\end{proof}
-
-\subsection{Proof of Theorems \ref{th:lawVj} and \ref{th:estimWj}}
+\subsection{Proof of Theorems \ref{th:lawVj} and \ref{th:deltaj}}
 \label{subsec:proof}
 
-In what follows, we use the notation $\lc(P)$ fot the leading 
-coefficient of a polynomial $P$. During the proof, $A$ and $B$ will 
-always refer to monic polynomials of degree $d$ and $R_j$ (resp. $U_j$ 
-and $V_j$) denote their $j$-th subresultant (resp. their $j$-th 
-cofactors). If $P$ is a polynomial and $n$ is a positive integer, we 
-use the notation $P[n]$ to refer to the coefficient of $X^n$ in $P$.
-We set $r_j = R_j[j]$.
+During the proof, $A$ and $B$ will always refer to monic polynomials of 
+degree $d$ and $R_j$ (resp. $U_j$ and $V_j$) denote their $j$-th 
+subresultant (resp. their $j$-th cofactors). If $P$ is a polynomial and 
+$n$ is a positive integer, we use the notation $P[n]$ to refer to the 
+coefficient of $X^n$ in $P$. We set $r_j = R_j[j]$.
 
@@
 \label{prop:relations}
-The following relations holds:
+The following relations hold:
 \begin{itemize}
-\item $U_{j-1} V_j - U_j V_{j-1} = r_j^2$;
+\setlength\itemsep{0.1em}
+\item $U_{j-1} V_j - U_j V_{j-1} = (-1)^j r_j^2$;
 \item $U_j[d{-}j{-}1] = -V_j[d{-}j{-}1] = (-1)^j r_{j+1}$;
@@
 \Z[a_0, \ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$
-are the two generic monic polynomials of degree $d$:
-$$A = X^d + \sum_{i=0} a_i X^i \quad \text{and} \quad
-B = X^d + \sum_{i=0} b_i X^i.$$
+are the two generic monic polynomials 
+$A = X^d + \sum_{i=0} a_i X^i$ and
+$B = X^d + \sum_{i=0} b_i X^i$.
 All relations then follows from the extended subresultants 
@@
 gathering the quadruples $(\mathcal U_j, \mathcal U_{j-1}, \mathcal
-R_j, \mathcal R_{j-1})$ satisfying the condition:
+R_j, \mathcal R_{j-1})$ such that:
 $$\begin{array}{ll}
@@
 \left\{\begin{array}{l}
-A = \todo{\text{something}} \\
-B = \todo{\text{something}} \\
+A = (-1)^j \cdot a^{-2} \cdot (V_j R_{j-1} - V_{j-1} R_j) \smallskip \\
+B = (-1)^{j-1} \cdot a^{-2} \cdot (U_j R_{j-1} - U_{j-1} R_j) \\
 \end{array}\right.
@@
 \mathcal R_j, \mathcal R_{j-1}) \smallskip \\
-\text{and} & (U_j, U_{j-1}, R_j, R_{j-1}) = \psi_j(A,B).
+\text{and} & (U_j, U_{j-1}, R_j, R_{j-1}) = \psi_j(A,B)
 \end{array}$$
@@
 
+\begin{cor}
+\label{cor:bijection}
+We assume that $\ring = \A$. Then the map $\psi_j : \Omega_j^\times
+\to \mathcal E_j^\times$ preserves the Haar measure.
+\end{cor}
+
+\begin{proof}
+Proposition \ref{prop:bijection} applied with the quotient rings $\ring 
+= \A/\pi^n\A$ shows that $(\psi_j \text{ mod } \pi^n)$ is a bijection 
+for all $n$. This proves the corollary.
+\end{proof}
+
 \noindent 
 \textbf{The distribution in the residue field.}
-We assume in this paragraph that $\ring$ is a finite field and we equip 
-$\Omega_\ring = \ring_d[X]^2$ with the uniform distribution. Just as 
-before, we consider each scalar subresultant $r_j$ as a random variable 
-on $\Omega_\ring$ taking its values in $\ring$.
+We assume in this paragraph that $\ring$ is a finite field of 
+cardinality $q$. We equip $\Omega_\ring = \ring_d[X]^2$ with the uniform 
+distribution. For $j \in \{0, \ldots, d-1\}$ and $(A,B) \in \Omega_\ring$,
+we set $\bar V_j(A,B) = 1$ if $r_j(A,B)$ vanishes and $\bar V_j(A,B) = 0$
+otherwise. The functions $\bar V_j$'s define random variables over 
+$\Omega_\ring$.
 
@@
 \label{prop:distribk}
-Assuming that $\ring$ is a finite field, the tuple $(r_0, \ldots, 
-r_{d-1})$ is uniformly distributed in $\ring^d$.
+With the above notations, the $\bar V_j$'s are mutually independant
+and they all follow a Bernoulli distribution of parameter $\frac 1 q$.
 \end{prop}
@@
 \begin{proof} 
+Given a subset $J$ of $\{0, \ldots, d-1\}$, we denote by 
+$\Omega_\ring(J)$ the subset of $\Omega_\ring$ consisting of those 
+couples $(A,B)$ for which $r_j(A,B)$ does not vanish if and only if $j 
+\in J$. We want to prove that $\Omega_\ring(J)$ has cardinality 
+$q^{2d-\Card J} (q-1)^{\Card J}$. To do this, we introduce several 
+additional notations. First, we fix $J$ and write $J = \{n_1, \ldots, 
+n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$. we also set $n_{\ell+1} = 
+0$ by convention. Then, given $n$ and $m$ two integers with $m < n$, let 
+$V_{(m,n)}$ denote the set of polynomials of the form
+$a_m X^m + a_{m+1} X^{m+1} \cdots + a_n X^n$
+with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{n,m}$ has
+cardinality $(q-1) q^{n-m}$. If $P$ is any polynomial of 
+degree $n$ and $m < n$ is an integer, we further define $P[m{:}] \in 
+V_{(m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
+with coefficients $< m$. Third, given $(A,B)$ in $\Omega_\ring$, we 
+denote by $(S_i(A,B))$ its subresultant pseudo-remainder sequence as 
+defined in \S \ref{subsec:subres}. We note that, if $(A,B) \in 
+\Omega_\ring(J)$, the sequence $(S_i(A,B))$ stops at $i = \ell$ and we 
+have $\deg S_i = n_i$ for all $i$. We now claim that the mapping
+$$\begin{array}{rcl} 
+\Lambda_J : \,\, 
+\Omega_\ring(J) & \to & 
+V_{(n_1,n_2)} \times \cdots \times V_{(n_\ell,n_{\ell+1})} \smallskip \\
+(A,B) & \mapsto & 
+\big(S_i(A,B)[n_{i+1}{:}]\big)_{1 \leq i \leq \ell}
+\end{array}$$ 
+is injective. In order to establish the claim, we remark that the 
+knowledge of $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$ (for some $i$) 
+are enough to reconstruct the quotient of the Euclidean division of 
+$S_i(A,B)$ by $S_{i-1}(A,B)$. Thus, one can recover $S_i(A,B)$ from 
+$S_{i-2}(A,B)$, $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$. We deduce 
+from this that $\Lambda_J(A,B)$ determines uniquely all $S_i(A,B)$'s 
+and finally $A$ and $B$ themselves. This proves the claim.
+
+Finally, we note that the claim implies that the cardinality of 
+$\Omega_\ring(J)$ is at most $q^{2d-\ell} (q-1)^\ell$. Summing up these 
+inequalities over all possible $J$, we get $\Card \Omega_\ring \leq 
+q^{2d}$. This latest inequality being an equality, we need to have
+$\Card \Omega_\ring(J) = q^{2d-\Card J} (q-1)^{\Card J}$ for all $J$.
 \end{proof}
 
-\begin{rem}
-In the next paragraph, we will use Proposition \ref{prop:distribk} 
-with $\ring = k$ (the residue field of $\A$). As for now, we can 
-notice that it implies (using functoriality of subresultants) that
-the variables $\bar V_j = \mathbbm{1}_{\{V_j = 0\}}$ follows the 
-Bernoulli distribution of parameter $(1 - p^{-1})$, in accordance
-with Theorem \ref{th:lawVj}. Proposition \ref{prop:distribk} 
-furthermore imply that the $\bar V_j$'s are mutually independant.
-\end{rem}
-
-\noindent
-\textbf{End of the proof.}
+\noindent
+\textbf{Proof of Theorem \ref{th:deltaj}.}
+We assume first that $j < d-1$. Proposition \ref{prop:distribk} above 
+then says that $r_{j+1}$ is invertible in $\A$ with probability 
+$(1 - q^{-1})$. Moreover, assuming that this event holds, Corollary 
+\ref{cor:bijection} implies that $R_j$ is distributed in $\A_{\leq 
+j}[X]$ according to the Haar measure. An easy computation gives
+$\P[\delta_j \geq m\,|\, r_{j+1} \in \A^\times ] =
+\frac{q(q^j - 1)}{q^{j+1}-1}$
+and therefore:
+$$\P[\delta_j \geq m] \geq (1 - q^{-1}) \cdot
+\frac{q(q^j - 1)}{q^{j+1}-1} =
+\frac{(q-1)(q^j - 1)}{q^{j+1}-1}.$$
+The case $j = d-1$ is actually simpler. Indeed, the same argument works 
+expect that we know for sure that $r_{j+1} = r_d$ is invertible since it 
+is equal to $1$ by convention. In that case, the probability is then 
+equal to $\frac{q(q^j - 1)}{q^{j+1}-1}$.
+
+\medskip
+
+\noindent
+\textbf{Proof of Theorem \ref{th:lawVj}.}
+We fix $j \in \{0, \ldots, d-1\}$. 
+We define a random variable $V_j^{(0)}$ as the greatest (nonnegative) 
+integer $v$ such that all scalar subresultants $r_{j'}$ have positive 
+valuation for $j'$ varying in the open range $(j-v, j+v)$ (with the
+convention that $r_{j'} = 0$ whenever $j < 0$). It is clear from the 
+definition that $r_{j-v}$ or $r_{j+v}$ (with $v = V_j^{(0)}$) has 
+valuation $0$. Moreover, assuming first that $\val(r_{j+v}) = 0$, we 
+get by Proposition \ref{prop:relations}:
+$$\begin{array}{rl}
+& \val(r_j) = v + \val\big(r_v^{j-v,j-v+1}(A^{(0)}, B^{(0)})\big) 
+\smallskip \\
+\text{with} &
+A^{(1)} = \frac 1 {r_{j+v} X^{j-v-1}} \cdot R_{j+v}[j{-}v{-}1\:{:}], \smallskip \\
+\text{and} &
+B^{(1)} = A^{(1)} + \frac 1 {\pi X^{j-v-1}} \cdot R_{j+v-1}[j{-}v{-}1\:{:}].
+\end{array}$$
+We notice that $B^{(1)}$ has actually all its coefficients in $\A$ 
+because $r_{j'}$ has positive valuation for $j' \in (j-v, j+v)$.
+Furthermore, Corollary \ref{cor:bijection} shows that the 
+couple $(A^{(1)}, B^{(1)})$ is distributed according to the Haar measure 
+on $(\A_{2v-1}[X])^2$. If $\val(r_{j+v}) = 0$, one can argue similarly 
+by replacing $R_{j+v}$ and $R_{j+v-1}$ by the cofactors $U_{j-v}$ and 
+$U_{j-v+1}$ respectively. Replacing $(A,B)$ by $(A^{(1)}, B^{(1)})$, we 
+can now define a new random variable $V_j^{(1)}$ and, continuing this 
+way, we construct an infinite sequence $V_j^{(m)}$ such that:
+$V_j = \sum_{m \geq 0} V_j^{(m)}$.
+
+We now introduce a double sequence $(X_i^{(m)})_{0 \leq i < d, m 
+\geq 0}$ of mutually independant random variables with Bernoulli
+distribution of parameter $\frac 1 q$ and we agree to set 
+$X_{j'}^{(m)} = 0$ for $j' < 0$ and $X_{j'}^{(m)} = 1$ for $j \geq d$.
+It follows from Proposition \ref{prop:distribk} (applied with $\ring = 
+k$) that $V_j^{(0)}$ has the same distribution than 
+$Y_j^{(0)} = \sum_{i=1}^d \min(X_{j-i}^{(0)}, \ldots, X_{j+i}^{(0)})$.
+In the same way, keeping in mind that $A^{(1)}$ and $B^{(1)}$ have both 
+degree $2V_j^{(0)} - 1$, we find that $V_j^{(1)}$ has the same 
+distribution than $\sum_{i=1}^{V_j^{(0)} - 1} \min(X_{j-i}^{(1)}, 
+\ldots, X_{j+i}^{(1)})$, which can be rewritten as:
+$Y_j^{(1)} = \sum_{i=1}^d \min(X_{j-i}^{(0)}, X_{j-i}^{(1)}, \ldots, 
+X_{j+i}^{(0)}, X_{j+i}^{(1)})$
+and the equidistribution of $(A^{(1)}, B^{(1)})$ shows that the 
+joint distribution $(V_j^{(0)}, V_j^{(1)})$ is the same as those
+of $(Y_j^{(0)}, Y_j^{(1)})$. Repeating the argument, we see that
+$(V_j^{(m)})_{m \geq 0}$ is distributed as $(Y_j^{(m)})_{m \geq 0}$
+where:
+$$Y_j^{(m)} = \sum_{i=1}^d \min(X_{j-i}^{(0)}, \ldots X_{j-i}^{(m)}, 
+\ldots, X_{j+i}^{(0)}, \ldots, X_{j+i}^{(m)}).$$
+Setting finally $X_i = \sum_{m \geq 0} \min(X_1^{(0)}, \ldots, 
+X_i^{(m)})$, we find the $X_i$'s ($0 \leq i < d$) are mutually 
+independant and that they all follow a geometric distribution of 
+parameter $(1 - q^{-1})$. We now conclude the proof by noting that 
+$Y_j$ equals $\sum_{i=1}^d \min(X_{j-i}, \ldots, X_{j+i})$ (recall
+that the $X_i^{(m)}$'s only take the values $0$ and $1$).
 
 \section{The method of adaptative lifts}
+\label{sec:stable}
 
@@
 sequence algorithm by applying a ``stabilization process'', whose 
-inspiration comes from \cite{}.
+inspiration comes from \cite{padicprec}.
 
@@
 
-In this subsection, we briefly report on the results of \cite{}. In this 
+In this subsection, we briefly report on the results of \cite{padicprec}. In this 
 paper, the authors draw the lines of a general framework to handle a 
@@
 has more regularity, one can exhibit a lower bound on $N$ ensuring
-that Eq.~\eqref{eq:padicprec} holds. In \cite{}, the case of locally 
+that Eq.~\eqref{eq:padicprec} holds. In \cite{padicprec}, the case of locally 
 analytic functions is considered. For our purpose, it will be enough 
@@
 Proposition \ref{prop:bijection} indicates that a left inverse of $f_i$ 
-is:
-$$(S_j, S_{j-1}) \mapsto \todo{\text{something}}$$
+is the function mapping $(S_j, S_{j-1})$ to
+$$(-1)^j \cdot r_j^{-2} \cdot 
+(V_j R_{j-1}{-}V_{j-1} R_j,\: -U_j R_{j-1}{+}U_{j-1} R_j)$$
 where $U_j, V_j$ (resp. $U_{j-1}$, $V_{j-1}$) are the $j$-th (resp 
@@
 \noindent
-The subresultants $R_j = \Res_j(A,B)$ can then be computed recursively 
-using the decreasing recurrence:
-$$\begin{array}{l}
-R_d = B, \, R_{d-1} = A-B \\
-R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})
-\end{array}$$
+The subresultants $R_j = \Res_j(A,B)$ can then be computed recursively:
+setting $R_d = B$, we have $R_{d-1} = A-B$ and the recurrence
+$R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})$.
 Moreover Lemma \ref{lem:diff} and Proposition \ref{prop:padicprec} imply
@@
 because the leading coefficient of $B$ is assume to be invertible.
-Dropping entirely~(a') and~(b) seems nevertheless to be more difficult. 
-We may however remark that~(b) is satisfied with high probability if
-$N$ is large compared to $2 \cdot \log_d p$. Thus, replacing eventually
-$N$ by $3 \cdot \log_d p$ (which does not affect the complexity), the 
-assumption~(b) is armless in average. Nonetheless, it may cause
-important difficulties on particularly bad instances.
+Dropping entirely~(a') seems nevertheless to be more difficult. Let us 
+move to assumption~(b) and first remark that it is satisfied with high 
+probability if $N$ is large compared to $2 \cdot \log_d p$. Thus, 
+replacing eventually $N$ by $3 \cdot \log_d p$ (which does not affect 
+the complexity), the assumption~(b) is harmless in average --- but maybe
+not on particularly bad instances. We finally underline that, if we are
+just interested in computing the $j$-th subresultant for a particular
+$j$, then we just need to assume the non-vanishing of the scalar
+subresultants in the range $[j{+}1, d{-}1]$.
 
@@
 
-\section{Towards p-adic floats}
+\section{Conclusion: p-adic floats?}
 
@@
 Consequently one might expect that interval arithmetic would provide 
-sharp results.
-Nonetheless it already appears on basic examples that this hope is
-failing. For instance:
-\begin{enumerate}[a)]
-\vspace{-0.25cm}
-\item in this paper, we showed that Euclide-like algorithms for 
-computing resultants loose a number of digits which is fairly linear in 
-the degree whereas no loss of precision is expected (because resultants
-can be computed by evaluating determinants);
-\vspace{-0.25cm}
-\item in \cite{}, it is shown that a similar phenomenom occurs for
-the computation of the SOMOS 4 sequence;
-\vspace{-0.25cm}
-\item in \cite{}, a stable algorithm is presented for computing the LU 
-factorization of a matrix: it looses a number of digits which is linear 
-in $\log d$ (where $d$ is the dimension of the input matrix) whereas 
-standard Gauss elimination looses a number of digits which is linear in 
-$d$;
-\vspace{-0.25cm}
-\item in \cite{}, the authors studied certain $p$-adic differential
-equations and designed specific stable algorithms to solve them after
-having noticed that usual (naive) technics may loose much more digits 
-that needed.
-\end{enumerate}
-\vspace{-0.2cm}
-All these examples suggest that interval arithmetic is not as good as 
-one might have expected at first. Therefore, it probably makes sense 
-to seriously study the analogue of floating point arithmetic in a
-ultrametric context.
+sharp results. Nonetheless this hope is failing and examples are basic 
+and numerous: $p$-adic differential equations \cite{bostan, padicdiff}, 
+LU factorization \cite{LU}, SOMOS 4 sequence \cite{padicprec}, 
+resultants (this paper), \emph{etc.} Consequently, interval arithmetic 
+is not as good as one might have expected at first. Therefore, it 
+probably makes sense to seriously study the analogue of floating point 
+arithmetic in a ultrametric context.
 
@@
 \begin{thebibliography}{99}
-\bibitem[a]{b}
-  Blah
+\bibitem{andrews}
+  G.~Andrews,
+  \emph{The Theory of Partitions},
+  Cambridge University Press (1976)
+\bibitem{pari}
+  C.~Batut, K.~Belabas, D.~Benardi, H.~Cohen, M.~Olivier, 
+  \emph{User’s guide to PARI-GP} (1985--2013)
+\bibitem{bostan}
+  A.~Bostan, L.~Gonz\'alez-Vega, H.~Perdry, É.~Schost, 
+  \emph{From Newton sums to coefficients: complexity issues in characteristic $p$}, 
+  MEGA’05 (2005)
+\bibitem{magma}
+  W.~Bosma, J.~Cannon, C.~Payoust, 
+  \emph{The Magma algebra system. I. The user language.}
+  J. Symbolic Comput. {\bf 24} (1997), 235--265
+\bibitem{padicprec}
+  X.~Caruso, D.~Roe, T.~Vaccon,
+  \emph{Tracking $p$-adic precision},
+  LMS J. Comp. and Math. {\bf 17}, 274--294
+\bibitem{LU}
+  X.~Caruso,
+  \emph{Random matrices over a DVR and LU factorization},
+  to appear at J. Symb. Comp.
+\bibitem{kedlaya}
+  K.~Kedlaya,
+  \emph{Counting points on hyperelliptic curves using Monsky--Washnitzer cohomology}, 
+  J. Ramanujan Math. Soc. {\bf 16} (2001), 323--338
+\bibitem{padicdiff}
+  P.~Lairez, T.~Vaccon,
+  \emph{...}
+  preprint (2014)
+\bibitem{sage}
+  W.~Stein et al.
+  \emph{Sage Mathematics Software}, 
+  The Sage Development Team (2005--2013)
+
+\bibitem{resultant}
+
+\bibitem{}
 \end{thebibliography}

### 2014-12-12 examples/stability.tex
More work on vector spaces

@@
 The section \ref{sec:theory} introduces the theory of ultrametric 
-precision developped in \cite{caruso-roe-vaccon:14a} and completes it 
-on a technical point. In \S \ref{sec:matrices}
-
-The analysis of standard algorithms are treated in 
-the next sections: \S\S \ref{sec:numbers}--\ref{sec:vectorspaces} are 
-successively devoted to the case of numbers, matrices, univariate 
-polynomials, multivariate polynomials and finally vector spaces.
+precision developped in \cite{caruso-roe-vaccon:14a} and completes 
+it on a technical point.
+In \S \ref{sec:matrices}, we focus on matrices... \todo{Complete
+this}.
 
@@
 precise statement and seek here to state the particular case of
-polynomial functions. A function $f : E \to F$ is said \emph{polynomial} 
-if it is coordinate-wise given by multivariate polynomials in any 
-(equivalently all) system of coordinates associated to orthonormal 
-basis.
+integral polynomial functions. A function $f : E \to F$ is said 
+\emph{integral polynomial} if it is given by multivariate polynomial 
+functions with coefficients in $\O_K$ in any (equivalently all) system 
+of coordinates associated to orthonormal basis.
 
@@
 We keep the notations of Proposition \ref{prop:precision} and assume 
-in addition that $f$ is polynomial. Let $C$ be a positive real
+in addition that $f$ is integral polynomial. Let $C$ be a positive real
 number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
@@
 \end{prop}
+
+In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
+to manifolds over $K$ and covers this way a very large families of
+situations (\emph{e.g.} computation with vector spaces as discussed
+in \S \ref{sec:vectorspaces}). In this extended framework, the precision 
+datum at some point $x$ is a lattice in the tangent space at $x$. 
+Propositions \ref{prop:precision} and \ref{prop:precision2} have
+analogues as well, simply obtained by working in charts.
 
@@
 and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
-is again an open subset in a $K$-Banach space $G$. Consequently, $g$
-goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
+is again an open subset in a normed $K$-vector space $G$. Consequently,
+$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
 all (continuous) linear applications $E \to F$.
@@
 \emph{but} changing the norm on the codomain of $h$ at the same time.
-More precisely, if $F$ --- which is the $K$-Banach in which $f$ takes
-its values --- is equipped with the norm $\Vert \cdot \Vert_F$, we 
-recommand endowing it with the second norm defined by
+More precisely, if $F$ is equipped with the norm $\Vert \cdot \Vert_F$, 
+we recommand endowing it with the second norm defined by
 $\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
@@
 \subsection{Multiplication}
+\label{subsec:mulmatrix}
 
@@
 performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
-positive integers with $s \geq \min(r,t)$ and assume that we want to 
-multiply a matrix $A \in M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. 
-This operation is of course modeled by the polynomial function:
+positive integers and assume that we want to multiply a matrix $A \in 
+M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. This operation is of course 
+modeled by the integral polynomial function:
 $$\begin{array}{rcl}
@@
 (resp. $\Delta_B$). We also agree to set $a_n = b_m = +\infty$ if
-$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}(\O_K)$ as 
-the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
+$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}((a_i),(b_j))$ 
+as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
 such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.
@@
 note that, on the other side, the valuation of the entries of $AB$ may 
-increase, meaning that we are also loosing some significant digits if 
-we are reasoning in relative precision.} (under the assumption of 
-Proposition \ref{prop:precision2}). However these digits are not --- 
-even individually --- attached to some particular entry of $AB$ but are 
-``spread out'' among all its entries. Thus, a usual coordinate-wise 
-precision data does not see this gain of precision in general.
-There is a more pedestrian way to visualize this phenomenon. Remark that
-the product $AB$ can be rewritten:
-$$AB = U_A \cdot P \cdot V_B
-\quad \text{with} \quad 
-P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$$
+increase, meaning that we are also loosing some significant digits if we 
+are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
+(\emph{cf} Proposition \ref{prop:precision2}). However these digits are 
+not --- even individually --- attached to some particular entry of $AB$ 
+but are ``spread out'' among all its entries. Thus, a usual 
+coordinate-wise precision data does not see this gain of precision in 
+general. There is a more pedestrian way to visualize this phenomenon. 
+Remark that the product $AB$ can be rewritten:
+$AB = U_A \cdot P \cdot V_B$ with
+$P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
 Tracking precision in the usual way, we see that the $(i,j)$-th entry
@@
 \emph{relative} precision we are loosing on the output of Algorithm 1 
-when we are using, on the one hand, a standard component-wise track of 
+when we are using, on the one hand, a standard coordinate-wise track of 
 precision and, on the other hand, a lattice-based method to handle 
@@
 \cline{3-4}
-& & Comp.-wise method & Lattice method \\
+& & Coord-wise method & Lattice method \\
 \hline 
@@
 Moreover, $\DI$ induces by restriction respectable differentiable 
-functions between $K$-manifolds:
-$$\DI_r : \mathcal M_r \to \Grass(F,r).$$
+functions between $K$-manifolds
+$\DI_r : \mathcal M_r \to \Grass(F,r)$.
+
 We would like to differentiate $\DI_r$ around some point $(f,V) \in 
 \mathcal M_r$. To do so, we use the first description of the 
-grassmannians we gave above: we shall see points in $\Grass(E,d)$ 
+grassmannians we gave above: we see points in $\Grass(E,d)$ 
 (resp. $\Grass(F,d)$) as embeddings $V \hookrightarrow E$ (resp. $W 
@@
 couple $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
-$f^{-1}(W)$. As before, this map does not takes its value in a single 
+$f^{-1}(W)$. As before, this map does not take its value in a single 
 grassmannian and we need to stratify $\mathcal W$ in order to get 
@@
 
-\subsection{Implementations and experiments}
+\subsection{Implementation and experiments}
 
@@
 {\bf Some experiments.}
+Let us consider the example of computation given by the following
+algorithm.
+
+\noindent\hrulefill
+
+\noindent {\bf Algorithm 2:} {\tt example\_vector\_space}
+
+\noindent{\bf Input:} two integers $n$ and $N$
+
+\smallskip
+
+\noindent 1.\ Set $L_0 = \langle (1 + O(2^N),  O(2^N), O(2^N)) \rangle
+\subset \Q_2^3$
+
+\noindent 2.\ {\bf for} $i=0,\dots,n-1$
+
+\noindent 3.\ \hspace{0.3cm}pick randomly $\alpha, \beta, \gamma, \delta
+\in M_{3,3}(\Z_2)$ with high precision
+
+\noindent 4.\ \hspace{0.3cm}{\bf compute} $L_{i+1} = 
+\big(\alpha(L_i) + \beta(L_i)\big) \cap \big(\gamma(L_i) + \delta(L_i)\big)$
+
+\noindent 5.\ {\bf return} $L_n$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
+
+\noindent
+The expression ``high precision'' on line 3 means that the precision on 
+$\alpha$, $\beta$, $\gamma$ and $\delta$ is set in such a way that it 
+does not affect the resulting precision on $L_{i+1}$.
+%
+\begin{figure}
+\begin{center}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{|c|c|c|c|}
+\hline
+\multirow{3}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
+\multicolumn{3}{|c|}{Average loss of precision} \\
+\cline{2-4}
+& \raisebox{-0.05cm}{Coord-wise} & 
+\multicolumn{2}{|c|}{Lattice method} \\
+\cline{3-4}
+& \raisebox{0.04cm}{method} & 
+\hspace{0.2cm}projected\hspace{0.2cm} & 
+\hspace{0.2cm}diffused\hspace{0.2cm} \\
+\hline 
+$10$ & $\hphantom{0}7.3$ & $\hphantom{0}2.7$ & $\hphantom{0}{-}2.4 \times 2$ \\
+$20$ & $14.8$ & $\hphantom{0}5.5$ & $\hphantom{0}{-}4.7 \times 2$ \\
+$50$ & $38.6$ & $13.1$ & $-12.0 \times 2$ \\
+$100$ & $78.1$ & $26.5$ & $-23.5 \times 2$ \\
+\hline
+\end{tabular}
+
+\smallskip
+
+{\small
+Results for a sample of $1000$ executions (with $N \gg n$)}
+\end{center}
+\renewcommand{\arraystretch}{1}
+
+\vspace{-0.3cm}
+
+\caption{Average loss of precision in Algorithm 2}
+\label{fig:vectorspace}
+\end{figure}
+%
+Figure \ref{fig:vectorspace} shows the losses of precision when
+executing Algorithm 2 with various inputs $n$ (the input $N$ is
+always chosen sufficiently large so that it does not affect the
+behaviour of precision). The ``coord-wise column'' corresponds to
+the standard way to track precision. On the other hand, in the two 
+last columns, the precision is tracked using lattices. The meaning of 
+``diffused'' is the same than those explained in \S 
+\ref{subsec:mulmatrix} when we were dealing with matrix multiplication: 
+it then corresponds to significant digits which are ``spread out'' among 
+the two coordinates of the precision lattice. In the present situation,
+it turns out that these ``diffused'' digits may be ``separated'' by
+working in another basis of the underlying space $\Q_2^3$
+The fact that only negative numbers appear in this column means that we 
+are actually always gaining precision with this model! Finally, in the 
+``projected'' column, we have written the loss of precision we get after 
+having projected the lattice precision data onto each coordinate (in 
+order to get an answer which is coordinate-wise).
 
@@
 \label{app:proof}
+
+We prove Proposition \ref{prop:boundLambdaf} in the slightly more
+general context of $K$-Banach.
 
@@
 $0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
-analytic around $0$ as well. Let us write the analytic expansion of $f$, 
-$g$ and $h$ as follows:
-$$f = \sum_{n \geq 0} f_n, \quad 
-g = \sum_{n \geq 0} g_n, \quad
-h = \sum_{n \geq 0} h_n, \quad$$
-where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
+analytic around $0$ as well. Let $f = \sum_{n \geq 0} f_n$
+$g = \sum_{n \geq 0} g_n$ and $h = \sum_{n \geq 0} h_n$ be the
+analytic expansions of $f$, $g$ and $h$.
+Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
 some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
@@
 
-During the proof, multinomial coefficients will play an important role. 
-We recall that they are defined as follows: 
-given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
-k_1 + \cdots + k_\ell$, we set:
-$$\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
-  \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.$$
-We also recall that these numbers are all positive integers.
-
-We now expand $g \circ f$ as follows:
-\begin{align}
-g \circ f & = \sum_{m \geq 0} g_m \Big(\sum_{n \geq 1} f_n\Big) \nonumber \\
-& = \sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
+A computation gives the following expansion for $g \circ f$:
+\begin{equation}
+\label{eq:expansiongf}
+\sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
 G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell}, \ldots, f_{n_\ell})
-\label{eq:expansiongf}
-\end{align}
-where the latest sum runs over:
+\end{equation}
+where $\binom m {\!k_1 \,\, \cdots \,\, k_\ell\!}$ denotes the
+multinomial coefficient and the sum runs over:
 
@@
 It is enough to prove that:
-\begin{equation}
-\label{eq:normpolar}
-\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
+$$\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
 \varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
-x_\ell) \Big\Vert \leq \Vert \psi \Vert
-\end{equation}
+x_\ell) \Big\Vert \leq \Vert \psi \Vert$$
 provided that all the $x_i$'s have norm at most $1$. We proceed by 
 induction on 
-$\ell$. If $\ell = 1$, Eq.~\eqref{eq:normpolar} follows directly from
+$\ell$. If $\ell = 1$, this follows directly from
 the definition of $\Vert \psi \Vert$. 
@@
 with
-$$\underline x_j = (x_1, \ldots, x_1, \ldots, x_{\ell+1}, \ldots, 
-x_{\ell+1})$$
+$\underline x_j = (x_1, \ldots, x_1, \ldots, x_{\ell+1}, \ldots, 
+x_{\ell+1})$
 where $x_i$ is repeated $k_i$ times if $i < \ell$ and $x_\ell$ 
@@
 \Vert$ as well. Looking at the coefficient in $\lambda^{k_{l+1}}$, we find
-Eq.~\eqref{eq:normpolar} for the families $(k_1, \ldots, k_{\ell+1})$ 
-and $(x_1, \ldots, x_{\ell+1})$. The induction follows.
+the desired result.
 \end{proof}
@@
 We now go back to the setting of Proposition \ref{prop:boundLambdaf}.
-We write the analytic expansion of $f$, $g$ and $h$ as follows: 
-$$f = \sum_{n \geq 0} f_n, \quad
-g = \sum_{n \geq 0} g_n, \quad
-h = \sum_{n \geq 0} h_n$$
-where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
+Let $f = \sum_{n \geq 0} f_n$, $g = \sum_{n \geq 0} g_n$ and
+$h = \sum_{n \geq 0} h_n$ be the analytice expansions of
+$f$, $g$ and $h$.
+Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
 some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
@@
 \{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
-\log \Vert f_n \Vert$ for all integer $n$ and denote by $\Lambda(f)$ its 
-Legendre transform. We define $\NP(g)$, $\Lambda(g)$, $\NP(h)$ and
-$\Lambda(h)$ similarly.
+\log \Vert f_n \Vert$ for all integer $n$. We recall that $\Lambda(f)$ 
+is the Legendre transform of $\NP(f)$ 
+\cite[Proposition 3.9]{caruso-roe-vaccon:14a}.
+We define $\NP(g)$ and $\NP(h)$ similarly.
 We recall that $\alpha$ is a nonnegative real number such that $\Vert n! 
@@
 \label{lem:boundLambdaf}
-We keep the above notations. If the couple $(a,b)$ satisfies:
-\begin{equation}
-\label{eq:condAB}
-b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)
-\end{equation}
+We keep the above notations. 
+If $(a,b)$ satisfies
+$b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)$
 then $b \geq \Lambda(f)(a - \alpha)$.
@@
 x \mapsto \big(h \mapsto n \cdot F_n(h, x, x, \ldots, x)\big).$$
-In particular, taking $h = x$, we find:
-\begin{equation}
-\label{eq:normderivative}
-\Vert f'_n \Vert \geq \Vert n f_n \Vert = 
-\Vert n \Vert \cdot \Vert f_n \Vert.
-\end{equation}
+Taking $h = x$, we find
+$\Vert f'_n \Vert \geq \Vert n f_n \Vert = 
+\Vert n \Vert \cdot \Vert f_n \Vert$.
 Combining this with Proposition \ref{prop:boundhr}, we get, for all
@@
 since the multinomial coefficient $\binom r {\!n_1 \, \cdots \, n_m\!}$
-is an integer and hence as norm at most $1$.
-
-We now pick two real numbers $a$ and $b$ satisfying \eqref{eq:condAB}.
-Set $d = \Lambda(h)(a)$. Going back to the definitions of $\Lambda (h)$ and Legendre transform, we remark that
-the above equality implies:
-\begin{equation}
-\label{eq:boundhn}
-\Vert h_n \Vert \leq e^{- a n + d},
-  \quad \forall n \in \N.
-\end{equation}
-In the same way, from Eq.~\eqref{eq:condAB}, we get:
-\begin{equation}
-\label{eq:boundgm}
-\Vert g_m \Vert \leq e^{-\!\max(b,d)\cdot m + b - a},
-  \quad \forall m \in \N.
-\end{equation}
+is an integer and hence has norm at most $1$.
+We now pick two real numbers $a$ and $b$ satisfying the hypothesis of 
+the Lemma. Set $d = \Lambda(h)(a)$. Going back to the definitions of 
+$\Lambda (h)$ and Legendre transform, we get $\Vert h_n \Vert \leq e^{- 
+a n + d}$ for all $n$.
+Similarly, from our hypothesis on $(a,b)$, we find
+$\Vert g_m \Vert \leq e^{-\!\max(b,d)\cdot m + b - a}$ for all $m$.
 We are now ready to prove $u_r \leq e^{-ar + b}$ by induction on $r$.
-When $r = 0$, it is obvious because $u_0$ vanishes. Otherwise, it
-follows from \eqref{eq:boundurrec}, \eqref{eq:boundhn}, \eqref{eq:boundgm}, 
-and the induction hypothesis that:
+When $r = 0$, it is obvious because $u_0$ vanishes. Otherwise, we 
+can write:
 \begin{align*}
@@
 \end{align*}
-and the induction goes.
-Finally, keeping in mind the definition of $u_r$, from $u_r \leq 
-e^{-a r + b}$, we obtain:
-$$\Vert f_r \Vert \leq u_r \cdot \Vert r! \Vert^{-1} \leq
-e^{-(a - \alpha) r + b}$$
-which means that $b \geq \Lambda(f)(a - \alpha)$.
+and the induction goes. Going back to the definition of $u_r$, we obtain 
+$\Vert f_r \Vert \leq u_r \cdot \Vert r! \Vert^{-1} \leq e^{-(a - \alpha) 
+r + b}$, which means $b \geq \Lambda(f)(a - \alpha)$.
 \end{proof}
 
-\begin{rem}
-A simpler proof is possible if $\Z \subset \O_K$, \emph{i.e.} if
-one can choose $\alpha = 0$. Indeed, under this extra assumption,
-one can prove using Eq.~\eqref{eq:normderivative} that:
-$$\Lambda(f') \geq \Lambda(f) - \id.$$
-Combining this with Lemma 3.7 of \cite{caruso-roe-vaccon:14a}, we find:
-$$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
-from which we can easily derive Lemma \ref{lem:boundLambdaf}.
-\end{rem}
+%\begin{rem}
+%A simpler proof is possible if $\Z \subset \O_K$, \emph{i.e.} if
+%one can choose $\alpha = 0$. Indeed, under this extra assumption,
+%one can prove using Eq.~\eqref{eq:normderivative} that:
+%$$\Lambda(f') \geq \Lambda(f) - \id.$$
+%Combining this with Lemma 3.7 of \cite{caruso-roe-vaccon:14a}, we find:
+%$$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
+%from which we can easily derive Lemma \ref{lem:boundLambdaf}.
+%\end{rem}
 
@@
 \nu$. Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to 
-check that such couples $(a,b)$ satisfy \eqref{eq:condAB}. Clearly:
+check that such couples $(a,b)$ satisfy the hypothesis of Lemma
+\ref{lem:boundLambdaf}. Clearly:
 $b \geq a + \Lambda(g) \circ \Lambda(h)(a)$
 since $\Lambda_g \geq \Lambda(g)$, $\Lambda_h \geq \Lambda(h)$ and
-$\Lambda_g$ is nondecreasing. Furthermore, from $b \leq \nu$, we get:
-$$\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
-\Lambda_h(a).$$
-Thus $a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) =
-b$ and we are done.
+$\Lambda_g$ is nondecreasing. Furthermore, from $b \leq \nu$, we get
+$\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
+\Lambda_h(a)$, from what we derive 
+$a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) = b$.
 

### 2014-12-19 examples/stability.tex
Added bibliography on (sub)resultants
Extended version of abstract :-)
Small fixes

@@
 \begin{abstract}
-Based on the framework developed in \cite{caruso-roe-vaccon:14a}, we study 
+Based on the framework of ``differential precision'' developed in a
+previous work by the same authors, we study 
 the $p$-adic stability of standard operations on matrices and vector 
-spaces. 
-%We observe in particular that generic algorithms --- which are 
-%designed to work over an arbitrary field --- are generally very 
-%unstable.
+spaces. In particular, we highlight that lattice-based methods can be
+quite profitable in many cases (\emph{e.g.} matrix multiplication). We 
+underpin this observation by numerous numerical evidences.
 \end{abstract}
@@
 \text{and} &
+\text{\emph{length}}\Big(\frac{\mathcal P'_{a,b,c}(A,B)(\mathcal L_0) }
+{M_{r,t}(\O_K)}\Big) =
 \displaystyle 
-[\mathcal P'_{a,b,c}(A,B)(\mathcal L_0) : M_{r,t}(\O_K)] =
 \sum_{m=1}^\infty a^{(m)} b^{(m)}

### 2014-12-19 examples/stability.tex
First draft over LU factorization
Need to be improved…
Lacks numerical experiments.

@@
 
-We refer to \cite[Appendix B]{caruso-roe-vaccon:14a} for the computation
-of the derivative of the LU factorization.
-\todo{Complete this subsection.}
+\todo{Improve this Subsection.}
+
+In this Subsection, we apply our framework to the case of $LU$ factorization.
+The computation of the differential of this factorization has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}.
+It remains to determine under which condition Proposition \ref{prop:precision} applies.
+
+\begin{deftn}
+Let $n \in \mathbb{Z}_{\geq 0}$ and  $O_n \subset K^{n \times n}$ be the set of the matrices such that all principal minors are non-vanishing.
+Let $U_n$ (resp.
+$L_n$)  be the $K$-vector spaces of the upper-triangular (resp.
+the strictly lower-triangular) matrices in $K^{n \times n}.$ Let $L_n^u \subset K^{n \times n} $ be the set of lower-triangular matrices with $1$'s one the diagonal.
+We define $D : O_n \rightarrow L_n^u \times U_n$ to be the mapping sending a matrix $M$ to $(L,M)$ such that $LU$ is the LU factorization of $M.$
+\end{deftn}
+
+This example has been previously studied by X.Caruso in \cite{caruso:12a}, but without using differential precision.
+We show here that the framework of Subsection \ref{ssec:boundLambdaf} applies directly.
+
+\subsubsection*{Computation of the differential of $D$} 
+We begin with some notations:
+\begin{deftn}
+We denote by $\textrm{low}$ (resp. $\textrm{up}$) the canonical projection $K^{n \times n} \rightarrow L_n$ (resp. $K^{n \times n} \rightarrow U_n$).
+\end{deftn}
+
+Then, an explicit formula for $D'$ has been computed in \cite[Appendix B]{caruso-roe-vaccon:14a} :
+\begin{prop}
+Let $(M,L,U) \in O_n \times L_n^u \times U_n$ be such that $M=LU.$ Then,
+
+\begin{tabular}{cccc}
+$D'(M) :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
+ & $\mathrm{d}M$ & $ \mapsto$ & $(L \textrm{low}(L^{*-1} \mathrm{d}M U^{-1}), \textrm{up}(L^{*-1} \mathrm{dM} U^{-1}) U).$ \\ 
+\end{tabular} 
+\end{prop}
+
+A right inverse of $D'(M)$ can be made explicit:
+\begin{rem} \label{rem:rightinverseLU}
+Let \begin{tabular}{cccc}
+$E(M) :$ & $L_n \times U_n$  & $\rightarrow$ &  $K^{n \times n}$\\ 
+ & $(A,B)$ & $ \mapsto$ & $AU+LB$. \\ 
+\end{tabular}
+
+Then $D'(M)  \circ E(M) = Id.$
+\end{rem}
+
+\subsubsection*{Applying Subsection \ref{ssec:boundLambdaf}}
+
+In order to apply Subsection \ref{ssec:boundLambdaf}, $f$ will be a translation of $D,$ so as to ensure $g(0) \neq 0.$ Namely,  let $(M_0,L_{0},U_{0}) \in O_n \times L_n^u \times U_n$ be such that $M_0 = L_{0}U_{0},$ we define
+
+\begin{tabular}{cccc}
+$f :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
+ & $M$ & $ \mapsto$ & $(L_{M+M_0}-L_{0}, U_{M+M_0}-U_{0}).$ \\ 
+\end{tabular}
+
+Then, $f(0)=0$ and \[f'(M) \cdot \mathrm{d}M = (L_{M+M_0} \textrm{low}(L_{M+M_0}^{-1} \mathrm{d}M U_{M+M_0}^{-1}), \textrm{up}(L_{M+M_0}^{-1} \mathrm{dM} U_{M+M_0}^{-1}) U).\]
+
+We now proceed with the three steps of the framework of Subsection \ref{ssec:boundLambdaf}.
+
+\textbf{The $C$ constant:}
+
+\begin{lem}
+For $C = \max (\Vert U_{0} \Vert,\Vert L_{0} \Vert ),$ $B_{K^{n \times n}}(1) \subset f'(0) \cdot B_{K^{n \times n}}(C).$ 
+\end{lem}
+\begin{proof}
+Apply Remark \ref{rem:rightinverseLU}.
+\end{proof}
+
+\textbf{The differential equation:}
+
+\begin{lem} Let $g=g_{M_0}$ be defined by:
+\begin{align*}
+g(A,B) \cdot X =& ( (L_0^*+A) \textrm{low} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right), \\
+& \: \textrm{up} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right) (U_0+B) ).\\
+\end{align*}
+Then, \[f' = g \circ f.\]
+\end{lem}
+
+\textbf{Determining $\mu$ and $\nu$}
+
+Let $\vert \: \vert $ be an ultrametric norm on $K$ and $\Vert \: \Vert$ be the corresponding subordinate matrix norm. To determine $\mu$ and $\nu,$ we first use the following lemma
+
+\begin{lem}
+Let $(U,V) \in K^{n \times n} \times K^{n \times n}.$ If $\Vert V \Vert < \Vert U \Vert$ then $\Vert U+V \Vert = \Vert U \Vert.$
+
+If $U \in GL_n(K)$ and $\Vert U \Vert < \Vert U^{-1} \Vert^{-1},$ then $\Vert (U+V)^{-1} \Vert = \Vert U^{-1} \Vert.$
+
+Finally $\Vert U^{-1} \Vert^{-1} \leq \Vert U \Vert.$
+\end{lem}
+
+This leads to the following upper-bound on the norm of $g(A,B)$.
+
+\begin{prop}
+Let $(A,B) \in L_n \times U_n.$ be such that $\Vert A \Vert < \Vert L_0^{*-1} \Vert $ and $\Vert B \Vert <\Vert U_0^{*-1} \Vert. $ Then 
+\[ \Vert g(A,B) \cdot X \Vert \leq \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right) \times \Vert X \Vert .
+\]
+\end{prop}
+
+\begin{prop}
+For $f$ and $g$ as before, we can take $\nu = \min ( \Vert L_0^{-1} \Vert ^{-1} ,\Vert U_0^{-1} \Vert ^{-1} ),$ and \[ \mu = \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right).\]
+\end{prop}
+
+\textbf{Conclusion}
+
+Consequently, we can apply Proposition \ref{prop:boundLambdaf2} and Proposition 3.12 of \cite{caruso-roe-vaccon:14a}:
+
+\begin{prop}
+For any $\rho \in (0,1],$ if $r<e^{ \nu -2(\alpha +\mu)+\log \left( \frac{\rho}{C} \right)},$ the equation \ref{eq:firstorder} for $f$ is satisfied.
+\end{prop}
+
+\begin{ex}
+If $\Vert L_0\Vert = \Vert L_0^{-1}\Vert =\Vert U_0\Vert =  \Vert U_0^{-1}\Vert = \rho=1$ and $p=2,$ then Proposition \ref{prop:precision} can be applied with $r=2^{-4}.$
+\end{ex}
+
+\subsubsection*{Numerical experiments}
+
+\todo{Add numerical experiments...}
 

### 2014-12-30 resultants/resultants.tex

@@
 $161.1$ significant digits by coefficient whereas a stable 
-algorithm should only loose \todo{$4$} digits in average. This 
+algorithm should only loose $3.2$ digits in average. This 
 ``theoretical'' loss is computed as the double of the valuation of the 
@@
 to the degree of the input polynomials whereas the ``theoretical'' loss 
-seems to be independant of it. In other words, the unstability of 
+seems to be independant of it. In other words, the instability of 
 Euclide's algorithm is becoming more and more critical when the degree 
@@
 \noindent
-\textbf{Notations.} We refer to \cite{resultant} for a nice and rather complete 
+\textbf{Notations.} We refer to \cite[\S 3.3]{cohen} and \cite[\S 4.1]{winkler}
+for a nice 
 introduction to resultants and subresultants and seek to fix notations
@@
 
-\medskip
-
 Let us describe quickly what might be this analogue and what are its
@@
 \begin{thebibliography}{99}
+\footnotesize
 \bibitem{andrews}
@@
   to appear at J. Symb. Comp.
+\bibitem{cohen}
+  H.~Cohen,
+  \emph{A course in Computational Algebraic Number Theory},
+  Springer (1996)
 \bibitem{kedlaya}
@@
   P.~Lairez, T.~Vaccon,
-  \emph{...}
+  \emph{Computation of power series solutions with $p$-adic coefficients of certain differential equations},
   preprint (2014)
@@
   The Sage Development Team (2005--2013)
-
-\bibitem{resultant}
-
+\bibitem{winkler}
+  F.~Winkler,
+  \emph{Polynomial Algorithms in Computer Algebra},
+  Springer Wien New Work (1996)
 \bibitem{}

### 2015-01-05 examples/stability.tex
Changed abstract, sage file for experimenting with characteristic polynomials

@@
 
-\title{About p-adic stability in linear algebra}
+\title{p-adic stability in linear algebra}
 
@@
 \begin{abstract}
-Based on the framework of ``differential precision'' developed in a
-previous work by the same authors, we study 
-the $p$-adic stability of standard operations on matrices and vector 
-spaces. In particular, we highlight that lattice-based methods can be
-quite profitable in many cases (\emph{e.g.} matrix multiplication). We 
-underpin this observation by numerous numerical evidences.
+Using the differential precision methods developed previously by the same authors,
+we study the $p$-adic stability of standard operations on matrices and vector 
+spaces. We demonstrate that lattice-based methods surpass naive methods in many
+applications, such as matrix multiplication and sums and intersections of subspaces.
+We also analyze determinants and characteristic polynomials using these differential methods.
+We supplement these observations with numerical experiments.
 \end{abstract}

### 2015-01-05 examples/stability.tex
Minor typoes corrected

@@
 
-To make this sentence more explcit, let us consider be a finite 
+To make this sentence more explicit, let us consider a finite 
 dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
@@
 above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
-on open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
+an open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
 $W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
@@
 U_{V,V^\c}$ cover the whole grassmaniann and then define an atlas 
-of it. When implementing vector spaces on computes, we usually restrict 
+of it. When implementing vector spaces on computer, we usually restrict 
 ourselves to the subatlas consisting of all charts of the form $(V_I, 
@@
 for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
-locally closed in $\mathcal M$ are therefore are submanifolds. 
+locally closed in $\mathcal M$ and therefore are submanifolds. 
 Moreover, $\DI$ induces by restriction respectable differentiable 

### 2015-01-05 examples/stability.tex
Section on LU factorization compactified

--- 2015-01-05 examples/stability.tex
@@
 
+\newcommand{\low}{\text{\rm low}}
+\newcommand{\up}{\text{\rm up}}
+
 \newcommand{\detp}{\text{\rm det}'}
@@
 applications, such as matrix multiplication and sums and intersections of subspaces.
-We also analyze determinants and characteristic polynomials using these differential methods.
-We supplement these observations with numerical experiments.
+We also analyze determinants, characteristic polynomials and LU factorization using these differential methods.
+We supplement our observations with numerical experiments.
 \end{abstract}
@@
 
-\noindent
-{\bf About the choice of the function $h$}.
-At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
-function $h$ in our differential equation. The aim of this was to be 
-more flexible and possibly get this way better bounds on $\Lambda(f)$. 
-In this paragraph, we discuss a recommended choice for it.
-
-Actually, we do recommend taking $h$ as the identity function 
-\emph{but} changing the norm on the codomain of $h$ at the same time.
-More precisely, if $F$ is equipped with the norm $\Vert \cdot \Vert_F$, 
-we recommand endowing it with the second norm defined by
-$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
-where $\lambda$ is a positive real number that we shall choose later 
-and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
-acting as the identity on the underlying vector spaces. Doing this, the 
-function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
-choose $\Lambda_h = \Lambda(h)$.
-
-We believe that a good way to optimize $\lambda$ is to choose it so that 
-the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
-agree. With the expression we got just above for $\Lambda_h$, our 
-heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
-Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
-at least, on the domain we are interested in --- and we do not loose so 
-much when we write that the norm of the couple $(f(x), h(x))$ is the 
-maximum of those of $f(x)$ and $h(x)$.
+\begin{rem}
+In certain situations,
+it may happen that the function $f$ is solution of a simplier 
+differential equation of the form $f' = g \circ f$. If this holds, 
+Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
+2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.
+
+Beyond this particular case, our general recommandation for choosing the 
+function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
+e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
+\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
+underlying vector spaces. The function $\Lambda(h)$ the maps $x$ to $x + 
+\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
+\end{rem}
 
@@
 with $n$ rows and $m$ columns with coefficients in $K$.
-
-\medskip
-
-\noindent
-{\bf Smith form.}
-In what follows, we will often use the \emph{Smith decomposition} of a 
-matrix $M \in M_{n,m}(K)$: we recall that it is a factorization of $M$ 
-as a product $U_M \cdot \Delta_M \cdot V_M$ when $U_M$ and $V_M$ are 
-unimodular (square) matrices of size $n$ and $m$ respectively, with 
-coefficients in $\O_K$ and $\Delta_M$ satisfies the two following
-conditions:
-
-\noindent
-(i) its $(i,j)$-th entry vanishes as soon as $i \neq j$
-
-\noindent
-(ii) its $(i,i)$-th entry divides its $(j,j)$-th entry when $i
-\leq j$.
-
-\noindent
-Such a factorization is \emph{not} unique but the valuations of
-the ``diagonal'' entries of $\Delta_M$ are. Smith decompositions
-can be easily computed by performing elementary operations on rows and columns.
-Regarding to precision, one can prove that if the entries of $M$ are all 
-known up to some $O(\pi^N)$, it is possible to compute a factorization 
-$U_M \cdot \Delta_M \cdot V_M$ where $U_M$ and $V_M$ are \emph{exact}
-matrices satisfying the above conditions and where $\Delta_M$ is known
-at precision $O(\pi^N)$ and satisfies~(i) and~(ii) modulo $\pi^N$.
 
@@
 of $M$ with this optimal precision. It consists in first computing an
-approximate Smith decomposition of $M$:
-$$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M.$$
-Here $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
+approximate Smith decomposition of $M$, that is a writing
+$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
+where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
 $\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
-$\pi^N$ and its diagonal entries $a_1, \ldots, a_m$ have valuations 
+$\pi^N$ and has diagonal entries $a_1, \ldots, a_m$ of valuations 
 $m_1, \ldots, m_n$ respectively.
@@
 
-\todo{Improve this Subsection.}
-
-In this Subsection, we apply our framework to the case of $LU$ factorization.
-The computation of the differential of this factorization has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}.
-It remains to determine under which condition Proposition \ref{prop:precision} applies.
-
-\begin{deftn}
-Let $n \in \mathbb{Z}_{\geq 0}$ and  $O_n \subset K^{n \times n}$ be the set of the matrices such that all principal minors are non-vanishing.
-Let $U_n$ (resp.
-$L_n$)  be the $K$-vector spaces of the upper-triangular (resp.
-the strictly lower-triangular) matrices in $K^{n \times n}.$ Let $L_n^u \subset K^{n \times n} $ be the set of lower-triangular matrices with $1$'s one the diagonal.
-We define $D : O_n \rightarrow L_n^u \times U_n$ to be the mapping sending a matrix $M$ to $(L,M)$ such that $LU$ is the LU factorization of $M.$
-\end{deftn}
-
-This example has been previously studied by X.Caruso in \cite{caruso:12a}, but without using differential precision.
-We show here that the framework of Subsection \ref{ssec:boundLambdaf} applies directly.
-
-\subsubsection*{Computation of the differential of $D$} 
-We begin with some notations:
-\begin{deftn}
-We denote by $\textrm{low}$ (resp. $\textrm{up}$) the canonical projection $K^{n \times n} \rightarrow L_n$ (resp. $K^{n \times n} \rightarrow U_n$).
-\end{deftn}
-
-Then, an explicit formula for $D'$ has been computed in \cite[Appendix B]{caruso-roe-vaccon:14a} :
-\begin{prop}
-Let $(M,L,U) \in O_n \times L_n^u \times U_n$ be such that $M=LU.$ Then,
-
-\begin{tabular}{cccc}
-$D'(M) :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
- & $\mathrm{d}M$ & $ \mapsto$ & $(L \textrm{low}(L^{*-1} \mathrm{d}M U^{-1}), \textrm{up}(L^{*-1} \mathrm{dM} U^{-1}) U).$ \\ 
-\end{tabular} 
-\end{prop}
-
-A right inverse of $D'(M)$ can be made explicit:
-\begin{rem} \label{rem:rightinverseLU}
-Let \begin{tabular}{cccc}
-$E(M) :$ & $L_n \times U_n$  & $\rightarrow$ &  $K^{n \times n}$\\ 
- & $(A,B)$ & $ \mapsto$ & $AU+LB$. \\ 
-\end{tabular}
-
-Then $D'(M)  \circ E(M) = Id.$
-\end{rem}
-
-\subsubsection*{Applying Subsection \ref{ssec:boundLambdaf}}
-
-In order to apply Subsection \ref{ssec:boundLambdaf}, $f$ will be a translation of $D,$ so as to ensure $g(0) \neq 0.$ Namely,  let $(M_0,L_{0},U_{0}) \in O_n \times L_n^u \times U_n$ be such that $M_0 = L_{0}U_{0},$ we define
-
-\begin{tabular}{cccc}
-$f :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
- & $M$ & $ \mapsto$ & $(L_{M+M_0}-L_{0}, U_{M+M_0}-U_{0}).$ \\ 
-\end{tabular}
-
-Then, $f(0)=0$ and \[f'(M) \cdot \mathrm{d}M = (L_{M+M_0} \textrm{low}(L_{M+M_0}^{-1} \mathrm{d}M U_{M+M_0}^{-1}), \textrm{up}(L_{M+M_0}^{-1} \mathrm{dM} U_{M+M_0}^{-1}) U).\]
-
-We now proceed with the three steps of the framework of Subsection \ref{ssec:boundLambdaf}.
-
-\textbf{The $C$ constant:}
-
-\begin{lem}
-For $C = \max (\Vert U_{0} \Vert,\Vert L_{0} \Vert ),$ $B_{K^{n \times n}}(1) \subset f'(0) \cdot B_{K^{n \times n}}(C).$ 
-\end{lem}
-\begin{proof}
-Apply Remark \ref{rem:rightinverseLU}.
-\end{proof}
-
-\textbf{The differential equation:}
-
-\begin{lem} Let $g=g_{M_0}$ be defined by:
-\begin{align*}
-g(A,B) \cdot X =& ( (L_0^*+A) \textrm{low} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right), \\
-& \: \textrm{up} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right) (U_0+B) ).\\
-\end{align*}
-Then, \[f' = g \circ f.\]
-\end{lem}
-
-\textbf{Determining $\mu$ and $\nu$}
-
-Let $\vert \: \vert $ be an ultrametric norm on $K$ and $\Vert \: \Vert$ be the corresponding subordinate matrix norm. To determine $\mu$ and $\nu,$ we first use the following lemma
-
-\begin{lem}
-Let $(U,V) \in K^{n \times n} \times K^{n \times n}.$ If $\Vert V \Vert < \Vert U \Vert$ then $\Vert U+V \Vert = \Vert U \Vert.$
-
-If $U \in GL_n(K)$ and $\Vert U \Vert < \Vert U^{-1} \Vert^{-1},$ then $\Vert (U+V)^{-1} \Vert = \Vert U^{-1} \Vert.$
-
-Finally $\Vert U^{-1} \Vert^{-1} \leq \Vert U \Vert.$
-\end{lem}
-
-This leads to the following upper-bound on the norm of $g(A,B)$.
-
-\begin{prop}
-Let $(A,B) \in L_n \times U_n.$ be such that $\Vert A \Vert < \Vert L_0^{*-1} \Vert $ and $\Vert B \Vert <\Vert U_0^{*-1} \Vert. $ Then 
-\[ \Vert g(A,B) \cdot X \Vert \leq \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right) \times \Vert X \Vert .
-\]
-\end{prop}
-
-\begin{prop}
-For $f$ and $g$ as before, we can take $\nu = \min ( \Vert L_0^{-1} \Vert ^{-1} ,\Vert U_0^{-1} \Vert ^{-1} ),$ and \[ \mu = \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right).\]
-\end{prop}
-
-\textbf{Conclusion}
-
-Consequently, we can apply Proposition \ref{prop:boundLambdaf2} and Proposition 3.12 of \cite{caruso-roe-vaccon:14a}:
-
-\begin{prop}
-For any $\rho \in (0,1],$ if $r<e^{ \nu -2(\alpha +\mu)+\log \left( \frac{\rho}{C} \right)},$ the equation \ref{eq:firstorder} for $f$ is satisfied.
-\end{prop}
-
-\begin{ex}
-If $\Vert L_0\Vert = \Vert L_0^{-1}\Vert =\Vert U_0\Vert =  \Vert U_0^{-1}\Vert = \rho=1$ and $p=2,$ then Proposition \ref{prop:precision} can be applied with $r=2^{-4}.$
-\end{ex}
-
-\subsubsection*{Numerical experiments}
-
-\todo{Add numerical experiments...}
+In this subsection, we study the LU factorization.
+%This example has been previously studied by Caruso in \cite{caruso:12a}, 
+%but without using differential precision. We are going to show here that 
+%the framework of \S \ref{ssec:boundLambdaf} applies directly.
+We first introduce several notations.
+We denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
+and, given a positive real number $C$, we let $B(C)$ be the closed ball 
+in $M_n(K)$ centered at the origin of radius $C$. We consider in addition
+the following subsets of $M_n(K)$:
+
+\noindent $\bullet$
+$O_n$ is the open subset of matrices whose all principal 
+minors do not vanish (we recall that the latest condition implies the
+existence and the unicity of a LU factorization);
+
+\noindent $\bullet$
+$U_n$ is the sub-vector space of upper-triangular 
+matrices;
+
+\noindent $\bullet$
+$L_n$ (resp. $L_n^u$) is the sub-affine space space of 
+nilpotent (resp. unipotent) lower-triangular matrices.
+
+\medskip
+
+\noindent
+{\bf Calculus and precision.}
+We choose to normalize the LU factorization by requiring that $L$ is 
+unipotent and denote by $D : O_n \to L_n^u \times U_n$ the function 
+modeling this decomposition. The computation of the differential of $D$ 
+has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. For
+$M \in \O_n$ with $D(M) = (L,U)$, the linear mapping $D'(M)$ is given 
+by:
+$$\begin{array}{l}
+dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U) \smallskip \\
+\hspace{1cm}\text{with } dX = L^{-1} \cdot dM \cdot U^{-1}
+\end{array}$$
+where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
+onto $L_n$ (resp. $U_n$). 
+It is easily checked that $D'(M)$ is bijective with inverse given by
+$(A,B) \mapsto AU+LB$.
+
+We now want to apply Proposition \ref{prop:boundLambdaf2}
+in order to derive a concrete result on precision. We pick $M_0 \in
+O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the ``translated'' 
+function $f$ taking $M$ to $D(M_0+M)-D(M_0)$. We then have $f(0) = 0$ 
+and $f'(M) = D'(M_0+M)$. Using the explicit description of the inverse
+of $D'(M_0+M)$, we find $B(1) \subset f'(0) \cdot B(C)$
+for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover $f$ satisfies 
+the differential equation $f' = g \circ f$ where $g$ is defined by:
+\begin{equation}
+\label{eq:gAB}
+\begin{array}{l}
+g(A,B)(X) = \big( (L_0+A) \cdot \low(Y), \up(Y) \cdot (U_0+B) \big)
+\smallskip \\
+\hspace{1.8cm}\text{with } Y = (L_0+A)^{-1} \cdot X \cdot (U_0+B)^{-1}.
+\end{array}
+\end{equation}
+Let $\kappa(S) = \Vert S \Vert \cdot \Vert S^{-1} \Vert$ denote the 
+condition number of a matrix $S$. Remarking that $\Vert S + T 
+\Vert = \Vert S \Vert$ if $\Vert T \Vert < \Vert S \Vert$ and $\Vert (S 
++ T)^{-1} \Vert = \Vert S^{-1} \Vert$ if $\Vert T \Vert < \Vert S^{-1} 
+\Vert^{-1}$, we deduce from \eqref{eq:gAB} that
+$\Vert g(A,B) \Vert \leq \max 
+\big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
+\kappa(U_0) \Vert L_0^{-1} \Vert \big)$
+provided that $\Vert A \Vert < \Vert L_0^{-1} \Vert^{-1}$ and 
+$\Vert B \Vert < \Vert U_0^{-1} \Vert^{-1}$.
+Combining this with Proposition \ref{prop:boundLambdaf2} and 
+\cite[Proposition.~3.12]{caruso-roe-vaccon:14a}, we finally find that 
+Eq.~\eqref{eq:firstorder} holds as soon as 
+$$\begin{array}{l}
+\displaystyle \frac \rho r > \Vert p \Vert^{-\frac{2p}{p-1}} \cdot
+\max ( \Vert L_0 \Vert, \Vert U_0 \Vert) \cdot
+\max ( \Vert L_0 ^{-1} \Vert, \Vert U_0 ^{-1} \Vert) \cdot{} \\
+\hspace{3.5cm} \max \big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
+  \kappa(U_0) \Vert L_0^{-1} \Vert \big)^2.
+\end{array}$$
+
+\noindent
+{\bf Numerical experiments.}
+\todo{Do it.}
 

### 2015-01-08 examples/stability.tex
Numerical experiments on LU
Mean and standard deviation for loss in precision over LU factorization.

@@
 {\bf Numerical experiments.}
-\todo{Do it.}
+Let $B_n=(E_{1,1},\dots,E_{1,n},\dots,E_{n,1},\dots,E_{n,n})$ be the canonical basis of $M_n(K)$.
+It can be naturally seen as a basis of $L_n \times U_n.$ For a given $M \in O_n,$ we write $D'(M)$ in $B_n,$ and we can remark that $D'(M)$ is lower-triangular in this basis.
+We can compute the coordinate-wise loss of precision of $D'(M),$ given by $\sum_u \left( \max_v \left( val(D'(M)_{u,v}) \right) \right),$ and compare it to the lattice-method loss of precision, given by $val(\det(D'(M))).$ The following array sums up the the mean and standard deviation of those losses for 2000 random matrices in $M_d(\mathbb{Z}_2)$ known at precision $100.$
+
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+ \multirow{2}{*}{ \hspace{0.2cm}$d$\hspace{0.2cm}} & \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
+  & mean & std & mean & std \\ \hline
+$2$ & $-3$& $5$ & $-1.5$ & $1.4$ \\ \hline
+$3$& $-9.4$& $11$ & $-2.3$ & $2.3$\\ \hline
+$4$ & $-20$& $20$ & $-3.8$ & $3.1$\\ \hline
+\end{tabular}
 

### 2015-01-09 examples/stability.tex
Figure for the table of benchmarks for LU

--- 2015-01-05 examples/stability.tex
@@
 {\bf Numerical experiments.}
-\todo{Do it.}
+Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
+$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
+well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
+for the matrix of this linear mapping in the above basis. We remark that
+$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
+coordinate, we find the best coordinate-wise loss of precision we can 
+hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
+val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
+$val(\det(D'(M)))$, which is the lattice-method loss of precision.
+Figure \ref{fig:LU} sums up the mean and standard deviation of those 
+losses for a sample of 2000 random matrices in $M_d(\mathbb{Z}_2)$.
+%
+\begin{figure}
+\begin{center}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{|c|c|c|c|c|}
+\hline 
+& \multicolumn{4}{|c|}{Loss of precision} \\
+\cline{2-5}
+\raisebox{0.2em}{matrix}
+& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
+\smash{\raisebox{0.6em}{size}} 
+& \hspace{0.5em}mean\hspace{0.5em} & deviation 
+& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
+$2$ & $\phantom{2}{-}3.0$& $5$ & $-1.5$ & $1.4$ \\% \hline
+$3$& $\phantom{2}{-}9.4$& $11$ & $-2.3$ & $2.3$\\% \hline
+$4$ & $-20\phantom{.4}$& $20$ & $-3.8$ & $3.1$\\ \hline
+\end{tabular} 
+\smallskip
+
+{\small
+Results for a sample of $2000$ instances}
+\end{center}
+\renewcommand{\arraystretch}{1}
+
+\vspace{-0.3cm}
+
+\caption{Loss of precision for LU factorization}
+\label{fig:LU}
+\end{figure}
+%
 

### 2015-01-13 examples/stability.tex
Working on characteristic polynomial section

--- 2015-01-05 examples/stability.tex
@@
 
-\begin{document}
-
 \newtheorem{theo}{Theorem}[section]
@@
 \newtheorem{quest}[theo]{Question}
+\newtheorem{conj}[theo]{Conjecture}
 %\theoremstyle{definition}
@@
 
-\newcommand{\low}{\text{\rm low}}
-\newcommand{\up}{\text{\rm up}}
-
-\newcommand{\detp}{\text{\rm det}'}
+%\newcommand{\detp}{\text{\rm det}'}
+
+%\DeclareMathOperator{\det}{det}
+\newcommand{\detp}{\det'}
 \newcommand{\DI}{\text{\rm DI}}
 \newcommand{\II}{\text{\rm II}}
+\DeclareMathOperator{\charpoly}{char}
+\newcommand{\charp}{\charpoly'}
 
@@
 \def\binom#1#2{\Big(\begin{array}{cc} #1 \\ #2 \end{array}\Big)}
+
+\begin{document}
 
@@
 \alignauthor David Roe \\
-  \affaddr{University of Calgary}\\
+  \affaddr{University of British Columbia}\\
   \email{\normalsize \textsf{roed.math@gmail.com}}
@@
 applications, such as matrix multiplication and sums and intersections of subspaces.
-We also analyze determinants, characteristic polynomials and LU factorization using these differential methods.
-We supplement our observations with numerical experiments.
+We also analyze determinants and characteristic polynomials using these differential methods.
+We supplement these observations with numerical experiments.
 \end{abstract}
@@
 
-To make this sentence more explicit, let us consider a finite 
+To make this sentence more explcit, let us consider be a finite 
 dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
@@
 
-\begin{rem}
-In certain situations,
-it may happen that the function $f$ is solution of a simplier 
-differential equation of the form $f' = g \circ f$. If this holds, 
-Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
-2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.
-
-Beyond this particular case, our general recommandation for choosing the 
-function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
-e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
-\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
-underlying vector spaces. The function $\Lambda(h)$ the maps $x$ to $x + 
-\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
-\end{rem}
+\noindent
+{\bf About the choice of the function $h$}.
+At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
+function $h$ in our differential equation. The aim of this was to be 
+more flexible and possibly get this way better bounds on $\Lambda(f)$. 
+In this paragraph, we discuss a recommended choice for it.
+
+Actually, we do recommend taking $h$ as the identity function 
+\emph{but} changing the norm on the codomain of $h$ at the same time.
+More precisely, if $F$ is equipped with the norm $\Vert \cdot \Vert_F$, 
+we recommand endowing it with the second norm defined by
+$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
+where $\lambda$ is a positive real number that we shall choose later 
+and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
+acting as the identity on the underlying vector spaces. Doing this, the 
+function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
+choose $\Lambda_h = \Lambda(h)$.
+
+We believe that a good way to optimize $\lambda$ is to choose it so that 
+the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
+agree. With the expression we got just above for $\Lambda_h$, our 
+heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
+Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
+at least, on the domain we are interested in --- and we do not loose so 
+much when we write that the norm of the couple $(f(x), h(x))$ is the 
+maximum of those of $f(x)$ and $h(x)$.
 
@@
 with $n$ rows and $m$ columns with coefficients in $K$.
+
+\medskip
+
+\noindent
+{\bf Smith form.}
+In what follows, we will often use the \emph{Smith decomposition} of a 
+matrix $M \in M_{n,m}(K)$: we recall that it is a factorization of $M$ 
+as a product $U_M \cdot \Delta_M \cdot V_M$ when $U_M$ and $V_M$ are 
+unimodular (square) matrices of size $n$ and $m$ respectively, with 
+coefficients in $\O_K$ and $\Delta_M$ satisfies the two following
+conditions:
+
+\noindent
+(i) its $(i,j)$-th entry vanishes as soon as $i \neq j$
+
+\noindent
+(ii) its $(i,i)$-th entry divides its $(j,j)$-th entry when $i
+\leq j$.
+
+\noindent
+Such a factorization is \emph{not} unique but the valuations of
+the ``diagonal'' entries of $\Delta_M$ are. Smith decompositions
+can be easily computed by performing elementary operations on rows and columns.
+Regarding to precision, one can prove that if the entries of $M$ are all 
+known up to some $O(\pi^N)$, it is possible to compute a factorization 
+$U_M \cdot \Delta_M \cdot V_M$ where $U_M$ and $V_M$ are \emph{exact}
+matrices satisfying the above conditions and where $\Delta_M$ is known
+at precision $O(\pi^N)$ and satisfies~(i) and~(ii) modulo $\pi^N$.
 
@@
 \times n$ with coefficients in $K$. The differential computation is 
-quite classical: the differential of the function $\det : M_{n,n}(K) \to 
-K$ at some point $M$ is the linear mapping:
-$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM)$$
-where $\com(M)$ stands for the comatrix of $M$. If $M$ is not 
+quite classical.  The differential of the function $\det : M_{n,n}(K) \to 
+K$ at some point $M$ is the linear mapping
+$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),$$
+where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
 invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
-\ref{prop:precision}. Therefore, from now on, we suppose that $M$ is invertible.
+\ref{prop:precision}. Therefore, we suppose that $M$ is invertible for the rest of this section.
 
@@
 of $M$ with this optimal precision. It consists in first computing an
-approximate Smith decomposition of $M$, that is a writing
-$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
-where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
+approximate Smith decomposition of $M$:
+$$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M.$$
+Here $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
 $\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
-$\pi^N$ and has diagonal entries $a_1, \ldots, a_m$ of valuations 
+$\pi^N$ and its diagonal entries $a_1, \ldots, a_m$ have valuations 
 $m_1, \ldots, m_n$ respectively.
@@
 
-\todo{Write this subsection}
+Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of polynomials of degree less than $n$.
+Then the determinant of $\charpoly$ at a point $M$ is
+\[
+\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
+\]
+The image is the $K$-span of the entries of $\com(X - M)$, which is clearly contained within $K_{<n}[X]$.
+In fact, the image will equal $K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same generalized eigenvalue.
+
+For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.
+In the standard case that the precision lattice for $M$ is just $H = \pi^k M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.
+
+Recall that the \emph{Newton polygon} of a polynomial $\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} of a matrix is the Newton polygon of its characteristic polynomial.
+If $m_1 \le \cdots \le m_n$ are the valuations of the diagonal entries of $\Delta_M$ in a Smith factorization $U_M \cdot \Delta_M \cdot V_M$ of $M$ as in the previous section, the \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = i+1}^n m_i)$.
+
+\begin{deftn}
+The \emph{precision polygon} of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
+\end{deftn}
+
+Any such polygon $P$ naturally yields a lattice in $K_{<n}[X]$, consisting of polynomials whose Newton polygons lie above $P$.
+The lattice associated to $P$ is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.
+
+If a polygon $P$ has vertices $(x_i, y_i)$, we $T_n(P)$ to be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.
+
+\begin{prop}
+Suppose a matrix $M$ is known with precision $H = \pi^k M_{n,n}(\O_K)$.
+Then the precision of the characteristic polynomial of $M$ is contained within $S_k$ applied to the lattice of the precision polygon of $M$.
+\end{prop}
+\begin{proof}
+The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(H)$, the result follows.
+\end{proof}
+
+\begin{prop} \label{prop:prec_hodge}
+The precision polygon of $M$ lies above $T_1$ applied to the Hodge polygon of $M$.
+\end{prop}
+
+\begin{proof}
+\todo{Add this proof from Xavier's e-mail}
+\end{proof}
+
+For many matrices, the precision polygon equals the Hodge polygon.
+For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.
+\todo{p=3?}  Moreover, experiments support the following conjectures.
+
+\begin{conj}
+\begin{enumerate}
+\item The precision polygon is equal to the Hodge polygon if the determinant of $M$ has valuation $0$ or $1$, or if $M$ is $3 \times 3$.
+\item The height at $0$ of the precision polygon is equal to the height at $1$ of the Hodge polygon.
+\item The precision polygon lies below $T_1$ applied to the Newton polygon.
+\end{enumerate}
+\end{conj}
+Empirically, the precision polynomial seems most likely to differ from the Hodge polygon at $1$, corresponding to the precision of the linear term of the characteristic polynomial.
+
+Of course, the precision polygon does necessarily agree with the full precision of the characteristic polynomial of $M$.
+In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the lattice associated to the precision polygon contains the actual precision lattice with index greater than $1$.
+This percentage increases as the valuation of the determinant increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.
+
+Moreover, one can construct examples with arbitrarily large extra lattice precision.
+If $M$ has elementary divisors with valuations $m_1 \le \cdots \le m_n$, Proposition \ref{prop:prec_hodge} implies that the precision lattice of the characteristic polynomial of $M$ is contained within $\O_{K,n}[X]$ with index at least $\sum_{i=1}^{n-1} m_i$.
+The precision lattice of $1 + M$ is obtained from the precision lattice of $M$ via the transformation $X \mapsto X + 1$, but the precision polygon is now flat with height $0$.
 
@@
 
-In this subsection, we study the LU factorization.
-%This example has been previously studied by Caruso in \cite{caruso:12a}, 
-%but without using differential precision. We are going to show here that 
-%the framework of \S \ref{ssec:boundLambdaf} applies directly.
-We first introduce several notations.
-We denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
-and, given a positive real number $C$, we let $B(C)$ be the closed ball 
-in $M_n(K)$ centered at the origin of radius $C$. We consider in addition
-the following subsets of $M_n(K)$:
-
-\noindent $\bullet$
-$O_n$ is the open subset of matrices whose all principal 
-minors do not vanish (we recall that the latest condition implies the
-existence and the unicity of a LU factorization);
-
-\noindent $\bullet$
-$U_n$ is the sub-vector space of upper-triangular 
-matrices;
-
-\noindent $\bullet$
-$L_n$ (resp. $L_n^u$) is the sub-affine space space of 
-nilpotent (resp. unipotent) lower-triangular matrices.
-
-\medskip
-
-\noindent
-{\bf Calculus and precision.}
-We choose to normalize the LU factorization by requiring that $L$ is 
-unipotent and denote by $D : O_n \to L_n^u \times U_n$ the function 
-modeling this decomposition. The computation of the differential of $D$ 
-has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. For
-$M \in \O_n$ with $D(M) = (L,U)$, the linear mapping $D'(M)$ is given 
-by:
-$$\begin{array}{l}
-dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U) \smallskip \\
-\hspace{1cm}\text{with } dX = L^{-1} \cdot dM \cdot U^{-1}
-\end{array}$$
-where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
-onto $L_n$ (resp. $U_n$). 
-It is easily checked that $D'(M)$ is bijective with inverse given by
-$(A,B) \mapsto AU+LB$.
-
-We now want to apply Proposition \ref{prop:boundLambdaf2}
-in order to derive a concrete result on precision. We pick $M_0 \in
-O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the ``translated'' 
-function $f$ taking $M$ to $D(M_0+M)-D(M_0)$. We then have $f(0) = 0$ 
-and $f'(M) = D'(M_0+M)$. Using the explicit description of the inverse
-of $D'(M_0+M)$, we find $B(1) \subset f'(0) \cdot B(C)$
-for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover $f$ satisfies 
-the differential equation $f' = g \circ f$ where $g$ is defined by:
-\begin{equation}
-\label{eq:gAB}
-\begin{array}{l}
-g(A,B)(X) = \big( (L_0+A) \cdot \low(Y), \up(Y) \cdot (U_0+B) \big)
-\smallskip \\
-\hspace{1.8cm}\text{with } Y = (L_0+A)^{-1} \cdot X \cdot (U_0+B)^{-1}.
-\end{array}
-\end{equation}
-Let $\kappa(S) = \Vert S \Vert \cdot \Vert S^{-1} \Vert$ denote the 
-condition number of a matrix $S$. Remarking that $\Vert S + T 
-\Vert = \Vert S \Vert$ if $\Vert T \Vert < \Vert S \Vert$ and $\Vert (S 
-+ T)^{-1} \Vert = \Vert S^{-1} \Vert$ if $\Vert T \Vert < \Vert S^{-1} 
-\Vert^{-1}$, we deduce from \eqref{eq:gAB} that
-$\Vert g(A,B) \Vert \leq \max 
-\big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
-\kappa(U_0) \Vert L_0^{-1} \Vert \big)$
-provided that $\Vert A \Vert < \Vert L_0^{-1} \Vert^{-1}$ and 
-$\Vert B \Vert < \Vert U_0^{-1} \Vert^{-1}$.
-Combining this with Proposition \ref{prop:boundLambdaf2} and 
-\cite[Proposition.~3.12]{caruso-roe-vaccon:14a}, we finally find that 
-Eq.~\eqref{eq:firstorder} holds as soon as 
-$$\begin{array}{l}
-\displaystyle \frac \rho r > \Vert p \Vert^{-\frac{2p}{p-1}} \cdot
-\max ( \Vert L_0 \Vert, \Vert U_0 \Vert) \cdot
-\max ( \Vert L_0 ^{-1} \Vert, \Vert U_0 ^{-1} \Vert) \cdot{} \\
-\hspace{3.5cm} \max \big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
-  \kappa(U_0) \Vert L_0^{-1} \Vert \big)^2.
-\end{array}$$
-
-\noindent
-{\bf Numerical experiments.}
-\todo{Do it.}
+\todo{Improve this Subsection.}
+
+In this Subsection, we apply our framework to the case of $LU$ factorization.
+The computation of the differential of this factorization has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}.
+It remains to determine under which condition Proposition \ref{prop:precision} applies.
+
+\begin{deftn}
+Let $n \in \mathbb{Z}_{\geq 0}$ and  $O_n \subset K^{n \times n}$ be the set of the matrices such that all principal minors are non-vanishing.
+Let $U_n$ (resp.
+$L_n$)  be the $K$-vector spaces of the upper-triangular (resp.
+the strictly lower-triangular) matrices in $K^{n \times n}.$ Let $L_n^u \subset K^{n \times n} $ be the set of lower-triangular matrices with $1$'s one the diagonal.
+We define $D : O_n \rightarrow L_n^u \times U_n$ to be the mapping sending a matrix $M$ to $(L,M)$ such that $LU$ is the LU factorization of $M.$
+\end{deftn}
+
+This example has been previously studied by X.Caruso in \cite{caruso:12a}, but without using differential precision.
+We show here that the framework of Subsection \ref{ssec:boundLambdaf} applies directly.
+
+\subsubsection*{Computation of the differential of $D$} 
+We begin with some notations:
+\begin{deftn}
+We denote by $\textrm{low}$ (resp. $\textrm{up}$) the canonical projection $K^{n \times n} \rightarrow L_n$ (resp. $K^{n \times n} \rightarrow U_n$).
+\end{deftn}
+
+Then, an explicit formula for $D'$ has been computed in \cite[Appendix B]{caruso-roe-vaccon:14a} :
+\begin{prop}
+Let $(M,L,U) \in O_n \times L_n^u \times U_n$ be such that $M=LU.$ Then,
+
+\begin{tabular}{cccc}
+$D'(M) :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
+ & $\mathrm{d}M$ & $ \mapsto$ & $(L \textrm{low}(L^{*-1} \mathrm{d}M U^{-1}), \textrm{up}(L^{*-1} \mathrm{dM} U^{-1}) U).$ \\ 
+\end{tabular} 
+\end{prop}
+
+A right inverse of $D'(M)$ can be made explicit:
+\begin{rem} \label{rem:rightinverseLU}
+Let \begin{tabular}{cccc}
+$E(M) :$ & $L_n \times U_n$  & $\rightarrow$ &  $K^{n \times n}$\\ 
+ & $(A,B)$ & $ \mapsto$ & $AU+LB$. \\ 
+\end{tabular}
+
+Then $D'(M)  \circ E(M) = Id.$
+\end{rem}
+
+\subsubsection*{Applying Subsection \ref{ssec:boundLambdaf}}
+
+In order to apply Subsection \ref{ssec:boundLambdaf}, $f$ will be a translation of $D,$ so as to ensure $g(0) \neq 0.$ Namely,  let $(M_0,L_{0},U_{0}) \in O_n \times L_n^u \times U_n$ be such that $M_0 = L_{0}U_{0},$ we define
+
+\begin{tabular}{cccc}
+$f :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
+ & $M$ & $ \mapsto$ & $(L_{M+M_0}-L_{0}, U_{M+M_0}-U_{0}).$ \\ 
+\end{tabular}
+
+Then, $f(0)=0$ and \[f'(M) \cdot \mathrm{d}M = (L_{M+M_0} \textrm{low}(L_{M+M_0}^{-1} \mathrm{d}M U_{M+M_0}^{-1}), \textrm{up}(L_{M+M_0}^{-1} \mathrm{dM} U_{M+M_0}^{-1}) U).\]
+
+We now proceed with the three steps of the framework of Subsection \ref{ssec:boundLambdaf}.
+
+\textbf{The $C$ constant:}
+
+\begin{lem}
+For $C = \max (\Vert U_{0} \Vert,\Vert L_{0} \Vert ),$ $B_{K^{n \times n}}(1) \subset f'(0) \cdot B_{K^{n \times n}}(C).$ 
+\end{lem}
+\begin{proof}
+Apply Remark \ref{rem:rightinverseLU}.
+\end{proof}
+
+\textbf{The differential equation:}
+
+\begin{lem} Let $g=g_{M_0}$ be defined by:
+\begin{align*}
+g(A,B) \cdot X =& ( (L_0^*+A) \textrm{low} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right), \\
+& \: \textrm{up} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right) (U_0+B) ).\\
+\end{align*}
+Then, \[f' = g \circ f.\]
+\end{lem}
+
+\textbf{Determining $\mu$ and $\nu$}
+
+Let $\vert \: \vert $ be an ultrametric norm on $K$ and $\Vert \: \Vert$ be the corresponding subordinate matrix norm. To determine $\mu$ and $\nu,$ we first use the following lemma
+
+\begin{lem}
+Let $(U,V) \in K^{n \times n} \times K^{n \times n}.$ If $\Vert V \Vert < \Vert U \Vert$ then $\Vert U+V \Vert = \Vert U \Vert.$
+
+If $U \in GL_n(K)$ and $\Vert U \Vert < \Vert U^{-1} \Vert^{-1},$ then $\Vert (U+V)^{-1} \Vert = \Vert U^{-1} \Vert.$
+
+Finally $\Vert U^{-1} \Vert^{-1} \leq \Vert U \Vert.$
+\end{lem}
+
+This leads to the following upper-bound on the norm of $g(A,B)$.
+
+\begin{prop}
+Let $(A,B) \in L_n \times U_n.$ be such that $\Vert A \Vert < \Vert L_0^{*-1} \Vert $ and $\Vert B \Vert <\Vert U_0^{*-1} \Vert. $ Then 
+\[ \Vert g(A,B) \cdot X \Vert \leq \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right) \times \Vert X \Vert .
+\]
+\end{prop}
+
+\begin{prop}
+For $f$ and $g$ as before, we can take $\nu = \min ( \Vert L_0^{-1} \Vert ^{-1} ,\Vert U_0^{-1} \Vert ^{-1} ),$ and \[ \mu = \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right).\]
+\end{prop}
+
+\textbf{Conclusion}
+
+Consequently, we can apply Proposition \ref{prop:boundLambdaf2} and Proposition 3.12 of \cite{caruso-roe-vaccon:14a}:
+
+\begin{prop}
+For any $\rho \in (0,1],$ if $r<e^{ \nu -2(\alpha +\mu)+\log \left( \frac{\rho}{C} \right)},$ the equation \ref{eq:firstorder} for $f$ is satisfied.
+\end{prop}
+
+\begin{ex}
+If $\Vert L_0\Vert = \Vert L_0^{-1}\Vert =\Vert U_0\Vert =  \Vert U_0^{-1}\Vert = \rho=1$ and $p=2,$ then Proposition \ref{prop:precision} can be applied with $r=2^{-4}.$
+\end{ex}
+
+\subsubsection*{Numerical experiments}
+
+\todo{Add numerical experiments...}
 
@@
 above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
-an open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
+on open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
 $W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
@@
 U_{V,V^\c}$ cover the whole grassmaniann and then define an atlas 
-of it. When implementing vector spaces on computer, we usually restrict 
+of it. When implementing vector spaces on computes, we usually restrict 
 ourselves to the subatlas consisting of all charts of the form $(V_I, 
@@
 for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
-locally closed in $\mathcal M$ and therefore are submanifolds. 
+locally closed in $\mathcal M$ are therefore are submanifolds. 
 Moreover, $\DI$ induces by restriction respectable differentiable 

### 2015-01-13 examples/stability.tex

--- 2015-01-09 examples/stability.tex
@@
 
+\newcommand{\detp}{\det'}
 \newcommand{\low}{\text{\rm low}}
 \newcommand{\up}{\text{\rm up}}
-
-\newcommand{\detp}{\text{\rm det}'}
 \newcommand{\DI}{\text{\rm DI}}
 \newcommand{\II}{\text{\rm II}}
+\DeclareMathOperator{\charpoly}{char}
+\newcommand{\charp}{\charpoly'}
 

### 2015-01-15 examples/stability.tex
Mostly spelling and grammar changes

@@
 
-\newcommand{\detp}{\det'}
+\newcommand{\detp}{\det{'}}
 \newcommand{\low}{\text{\rm low}}
@@
 For about twenty years, the use of $p$-adic methods in symbolic 
-computation was gaining more and more popularity: they were succesfully 
+computation has been gaining popularity: they were successfully 
 used for instance 
@@
 to count points on certain varieties (mainly curves) using $p$-adic 
-cohomologies (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
+cohomology theories (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
 followers).
@@
 \cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. 
-cit.} in order to analyse the $p$-adic stability of many standard 
+cit.} in order to analyze the $p$-adic stability of many standard 
 algorithms for computing with basic algebraic structures: numbers, 
@@
 The section \ref{sec:theory} introduces the theory of ultrametric 
-precision developped in \cite{caruso-roe-vaccon:14a} and completes 
+precision developed in \cite{caruso-roe-vaccon:14a} and completes 
 it on a technical point.
@@
 Throughout the paper, the letter $K$ will refer to a \emph{complete 
-discrete valuation field of characteristic $0$}, that is either a finite 
+discrete valuation field of characteristic $0$}, \todo{why characteristic 0?} that is either a finite 
 extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
@@
 
-The aim of this section is to summerize briefly the content of 
+The aim of this section is to summarize briefly the content of 
 \cite{caruso-roe-vaccon:14a} and complete it on certain points.
@@
 functions with coefficients in $\O_K$ in any (equivalently all) system 
-of coordinates associated to orthonormal basis.
+of coordinates associated to orthonormal bases.
 
@@
 is again an open subset in a normed $K$-vector space $G$. Consequently,
-$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
-all (continuous) linear applications $E \to F$.
-In what follows, we always assume that $V$ and $W$ contains the origin, 
+$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ of
+(continuous) linear maps $E \to F$.
+In what follows, we always assume that $V$ and $W$ contain the origin, 
 $f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
@@
 there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
-$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
+$({-}\infty, \nu]$\footnote{We note that this assumption is fullfilled if 
 we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
@@
 characteristic of the residue field and where, by convention, the above 
-expression equals $0$ if $p = 0$.
+expression equals $0$ if $p = 0$. \todo{which expression; the one defining $\alpha$?}
 The next Proposition is proved in Appendix \ref{app:proof}.
@@
 \begin{prop}
-\label{prop:boundLambdaf}\label{PROP:BOUNDLAMBDAF}
+\label{prop:boundLambdaf}
 We have $\Lambda(f) \leq \Lambda_f$.
@@
 Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
-The blue plain line represents the graph of the function $\Lambda_f$. A 
+The blue \todo{do we have color?} plain line represents the graph of the function $\Lambda_f$. A 
 quick computation shows that, on a neighborhood of ${-}\infty$, this 
@@
 where $\mu$ is the value that $\Lambda_g$ takes on the interval 
-$]{-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
+$({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
 graph of $\Lambda(f)$ lies below the plain blue line. We remark
-moreover that the Taylor expansion of $f(x)$ starts with the term
-$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
+moreover that the Taylor expansion of $f(z)$ starts with the term
+$g(0) z$. Hence, on a neighborhood on ${-}\infty$, we have 
 $\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
@@
 \min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
-for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
+for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$. \todo{What is $\Lambda(f)_{\leq 2}$?}
 \end{prop}
@@
 In certain situations,
-it may happen that the function $f$ is solution of a simplier 
+it may happen that the function $f$ is solution of a simpler 
 differential equation of the form $f' = g \circ f$. If this holds, 
@@
 
-Beyond this particular case, our general recommandation for choosing the 
+Beyond this particular case, our general recommendation for choosing the 
 function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
@@
 \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
-underlying vector spaces. The function $\Lambda(h)$ the maps $x$ to $x + 
+underlying vector spaces. The function $\Lambda(h)$ then maps $x$ to $x + 
 \mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
@@
 
-To begin with, we want to study the behaviour of the precision when 
+To begin with, we want to study the behavior of the precision when 
 performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
@@
 \end{array}$$
-According to Proposition \ref{prop:precision}, the behaviour of the precision when 
-computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$. It is obvious 
-that this differential is the linear mapping that takes a couple 
+According to Proposition \ref{prop:precision}, the behavior of the precision when 
+computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$, the linear mapping that takes a pair 
 $(dA,dB)$ to $A \cdot dB + dA \cdot B$.
@@
 $$\begin{array}{rl}
-& \mathcal P'_{a,b,c}(A,B)(\mathcal L_0)
+& \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
 = U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
 \text{and} &
-\text{\emph{length}}\Big(\frac{\mathcal P'_{a,b,c}(A,B)(\mathcal L_0) }
-{M_{r,t}(\O_K)}\Big) =
+\text{\emph{length}}\Big(\frac{M_{r,t}(\O_K)}{\mathcal P'_{r,s,t}(A,B)(\mathcal L_0) }
+\Big) =
 \displaystyle 
@@
 
-One may wonder to what extend it is important to take care of these 
+One may wonder to what extent it is important to take care of these 
 ``diffused'' gains of precision. For one single product of matrices, 
-it is not so much. However, the benefits becomes really substantial
+it is not so much. However, the benefits become substantial
 when we are executing an algorithm that computes many products of
@@
 Figure \ref{fig:mulmatrix} compares the number of significant digits in 
-\emph{relative} precision we are loosing on the output of Algorithm 1 
+\emph{relative} precision we are losing on the output of Algorithm 1 
 when we are using, on the one hand, a standard coordinate-wise track of 
@@
 
-We now come to the computation of the determinant of the square $n 
-\times n$ with coefficients in $K$. The differential computation is 
+We now come to the computation of the determinant of a square $n 
+\times n$ matrix with coefficients in $K$. The differential computation is 
 quite classical.  The differential of the function $\det : M_{n,n}(K) \to 
@@
 From the description of $\detp(M)$, we see that it is enough to prove 
-that the smallest valuation on an entry of $\com(M)$ is $v$
+that the smallest valuation of an entry of $\com(M)$ is $v$
 or, equivalently, that the ideal of $\O_K$ generated by all 
@@
 $\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
-$\pi^N$ and has diagonal entries $a_1, \ldots, a_m$ of valuations 
+$\pi^N$ and has diagonal entries $a_1, \ldots, a_n$ of valuations 
 $m_1, \ldots, m_n$ respectively.
@@
 For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.
-In the standard case that the precision lattice for $M$ is just $H = \pi^k M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.
+In the standard case that the precision lattice for $M$ is just $H = \pi^N M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.
 
@@
 
-If a polygon $P$ has vertices $(x_i, y_i)$, we $T_n(P)$ to be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.
+If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.
 
 \begin{prop}
-Suppose a matrix $M$ is known with precision $H = \pi^k M_{n,n}(\O_K)$.
-Then the precision of the characteristic polynomial of $M$ is contained within $S_k$ applied to the lattice of the precision polygon of $M$.
+Suppose a matrix $M$ is known with precision $H = \pi^N M_{n,n}(\O_K)$.
+Then the precision of the characteristic polynomial of $M$ is contained within $S_N$ applied to the lattice of the precision polygon of $M$.
 \end{prop}
@@
 \noindent $\bullet$
-$L_n$ (resp. $L_n^u$) is the sub-affine space space of 
+$L_n$ (resp. $L_n^u$) is the sub-affine space of 
 nilpotent (resp. unipotent) lower-triangular matrices.
@@
 hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
-val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
-$val(\det(D'(M)))$, which is the lattice-method loss of precision.
+\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
+$\val(\det(D'(M)))$, which is the lattice-method loss of precision.
 Figure \ref{fig:LU} sums up the mean and standard deviation of those 
@@
 \hline 
-& \multicolumn{4}{|c|}{Loss of precision} \\
+& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
 \cline{2-5}
@@
 Vector spaces are generally represented as subspaces of $K^n$ for some 
-$n$. Hence they naturally appear as points on grassmannians. Therefore, 
+$n$. Hence they naturally appear as points on Grassmannians. Therefore, 
 one can use the framework of \cite[Appendix A]{caruso-roe-vaccon:14a} to 
@@
 
-\subsection{Geometry of grassmannians}
+\subsection{Geometry of Grassmannians}
 
@@
 set of $d$-dimensional subspaces of $E$. It is the so-called 
-\emph{grassmannian}. It is well-known that $\Grass(E,d)$ has a 
+\emph{Grassmannian}. It is well-known that $\Grass(E,d)$ has a 
 natural structure of $K$-manifold. The aim of this subsection is to 
@@
 Let $V$ denote a fixed sub-vector space of $E$ of dimension $d$. The 
-grassmannian $\Grass(E,d)$ can be viewed as the quotient of the set 
+Grassmannian $\Grass(E,d)$ can be viewed as the quotient of the set 
 of linear embeddings $f: V \hookrightarrow E$ modulo the action (by 
@@
 above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
-an open subset of $\Grass(E,d)$ consistant exactly of those subspaces 
+an open subset of $\Grass(E,d)$ consisting exactly of those subspaces 
 $W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
@@
 When the couple $(V, V^\c)$ varies, the open subsets $\mathcal 
-U_{V,V^\c}$ cover the whole grassmaniann and then define an atlas 
+U_{V,V^\c}$ cover the whole Grassmaniann and then define an atlas 
 of it. When implementing vector spaces on computer, we usually restrict 
@@
 U^\star_{V, V^\c}$ which commutes with the embeddings into the
-grassmannian.
+Grassmannian.
 
@@
 priori} vary, the map $\DI$ does not take its values in well-defined 
-grassmannian. To tackle this problem, we stratify $\mathcal M$ as
+Grassmannian. To tackle this problem, we stratify $\mathcal M$ as
 follows: for each integer $r \in [0,d]$, we introduce the subset
@@
 \mathcal M_r$. To do so, we use the first description of the 
-grassmannians we gave above: we see points in $\Grass(E,d)$ 
+Grassmannians we gave above: we see points in $\Grass(E,d)$ 
 (resp. $\Grass(F,d)$) as embeddings $V \hookrightarrow E$ (resp. $W 
@@
 $f^{-1}(W)$. As before, this map does not take its value in a single 
-grassmannian and we need to stratify $\mathcal W$ in order to get 
+Grassmannian and we need to stratify $\mathcal W$ in order to get 
 differentiable functions. For each integer $r \in [0,n]$, we introduce 
@@
 always chosen sufficiently large so that it does not affect the
-behaviour of precision). The ``coord-wise column'' corresponds to
+behavior of precision). The ``coord-wise column'' corresponds to
 the standard way to track precision. On the other hand, in the two 
@@
 Let $f = \sum_{n \geq 0} f_n$, $g = \sum_{n \geq 0} g_n$ and
-$h = \sum_{n \geq 0} h_n$ be the analytice expansions of
+$h = \sum_{n \geq 0} h_n$ be the analytic expansions of
 $f$, $g$ and $h$.

### 2015-01-17 examples/stability.tex
Updating charpoly section, fixing typos in A.1

@@
 \DeclareMathOperator{\NP}{NP}
+\DeclareMathOperator{\HP}{HP}
+\DeclareMathOperator{\PP}{PP}
 \DeclareMathOperator{\Hom}{Hom}
@@
 {\bf Notations.}
-Throughout the paper, the letter $K$ will refer to a \emph{complete 
-discrete valuation field of characteristic $0$}, \todo{why characteristic 0?} that is either a finite 
+Throughout the paper, the letter $K$ will refer to a complete 
+discrete valuation field of characteristic $0$, that is either a finite 
 extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
-series over a field of characteristic $0$. We denote by $\val : K \to \Z 
+series over a field of characteristic $0$ \todo{what about mixed characteristic with infinite residue field?}.
+We denote by $\val : K \to \Z 
 \cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
@@
 where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
-e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is $\alpha = - 
-\frac p {p-1} \cdot \log \Vert p \Vert > 0$ where $p$ denotes the 
-characteristic of the residue field and where, by convention, the above 
-expression equals $0$ if $p = 0$. \todo{which expression; the one defining $\alpha$?}
+e^{-\alpha n}$ for all $n$. If $p$ is the characteristic of the residue field,
+a suitable value for $\alpha$ is $\alpha = - 
+\frac p {p-1} \cdot \log \Vert p \Vert$ if $p > 0$ and $\alpha = 0$ if $p = 0$.
 The next Proposition is proved in Appendix \ref{app:proof}.
@@
 \begin{prop}
-\label{prop:boundLambdaf}
+\label{prop:boundLambdaf} \label{PROP:BOUNDLAMBDAF}
 We have $\Lambda(f) \leq \Lambda_f$.
@@
 Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
-The blue \todo{do we have color?} plain line represents the graph of the function $\Lambda_f$. A 
+The blue plain line represents the graph of the function $\Lambda_f$. A 
 quick computation shows that, on a neighborhood of ${-}\infty$, this 
@@
 
-\todo{Write a short introduction.}
-
-The notation $M_{n,m}(K)$ refers to the $K$-vector space of matrices 
-with $n$ rows and $m$ columns with coefficients in $K$.
+The notation $M_{m,n}(K)$ refers to the $K$-vector space of matrices 
+with $m$ rows and $n$ columns with coefficients in $K$.  We will repeatedly
+use the Smith decomposition for $M \in M_{m,n}(K)$:
+\[
+M = U_M \cdot \Delta_M \cdot V_M,
+\]
+with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
+for the valuation of the $(i,i)$-th entry of $\Delta_M$, and by convention set
+$\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.
 
@@
 Nonetheless, it may happen that the above inclusion is strict, meaning 
-that we are \emph{gaining} precision in those cases. In order to compute 
-exactly $\mathcal P'_{a,b,c}(A,B)(\mathcal L_0)$, we introduce the Smith 
-factorization of $A$ and $B$:
-$$A = U_A \cdot \Delta_A \cdot V_A
-\text{ and } B = U_B \cdot \Delta_B \cdot V_B$$
-and denote by $a_1, \ldots, a_{\min(r,s)}$ (resp. $b_1, \ldots, 
-b_{\min(s,t)}$) the valuation of the diagonal entries of $\Delta_A$
-(resp. $\Delta_B$). We also agree to set $a_n = b_m = +\infty$ if
-$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}((a_i),(b_j))$ 
+that we are \emph{gaining} precision in those cases.
+
+Set $a_i = \sigma_i(A)$ and $b_i = \sigma_i(B)$, and define $M_{r,t}((a_i),(b_j))$ 
 as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
-such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.
+such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.  Write $a^{(m)}$ (resp. $b^{(m)}$)
+for the number of $a_i$'s (resp. $b_i$'s) which are at least $m$.
 
@@
 \label{prop:mulmatrix}
-With the above notations, we have:
-$$\begin{array}{rl}
+With the above notations, we have
+\[
+\begin{array}{rl}
 & \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
@@
 \displaystyle 
-\sum_{m=1}^\infty a^{(m)} b^{(m)}
-\end{array}$$
-where $a^{(m)}$ (resp. $b^{(m)}$) is the number of $a_i$'s (resp.
-of $b_j$'s) which are $\geq m$.
+\sum_{m=1}^\infty a^{(m)} b^{(m)}.
+\end{array}
+\]
 \end{prop}
@@
 the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.
-In order to do so, we write $M = U_M \cdot \Delta_M \cdot V_M$ the 
-Smith factorization of $M$ and denote by $m_1 \leq \ldots \leq m_n$ 
-the valuation of the diagonal entries of $\Delta_M$.
 
@@
 \label{prop:detmatrix}
-Setting $v = m_1 + \cdots + m_{n-1}$, we have
+Setting $v = \sigma_1(M) + \cdots + \sigma_{n-1}(M)$, we have
 $\detp(M)(\mathcal L_0) = \pi^v \O_K$.
@@
 \O_K$. But this ideal remains unchanged when we multiply $M$ 
-(on the left or on the right) by any unimodular matrix. Thus we may 
-assume that $M$ is $\Delta_M$ and the result becomes clear.
+on the left or on the right by a unimodular matrix. Thus we may 
+assume that $M = \Delta_M$, and the result becomes clear.
 \end{proof}
@@
 $M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
-is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
-digits in absolute precision\footnote{We note that the valuation of 
-$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
-precisions then we should say that we are loosing $m_n$ significant 
-digits.}.
-Furthermore, there exists a easy method for computing the determinant
-of $M$ with this optimal precision. It consists in first computing an
-approximate Smith decomposition of $M$, that is a writing
-$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
-where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
-$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
-$\pi^N$ and has diagonal entries $a_1, \ldots, a_n$ of valuations 
-$m_1, \ldots, m_n$ respectively.
-We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
-the $i$-th row of $\tilde \Delta_M$, we obtain
-$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
-where $U$ is a matrix which is congruent to the identity modulo 
-$\pi^{N-m_n}$. Therefore we get:
-$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
-from what we see that we can compute $\det(M)$ at the required precision
-by simply evaluating the product $a_1 \cdots a_m$.
-
-\subsection{Characteristic polynomials}
+is known at precision $O(\pi^{N+v})$.  Thus we are gaining $v$ significant 
+digits in absolute precision or, equivalently, losing $\sigma_n(M)$ digits
+of relative precision.
+Furthermore, one may easily compute the determinant
+of $M$ with this optimal precision by finding an approximate
+Smith decomposition with $\Delta_M$ known at precision $O(\pi^N)$ and
+multiplying its diagonal entries.
+
+\subsection{Characteristic polynomials} \label{ssec:charpoly}
 
@@
 
-For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.
-In the standard case that the precision lattice for $M$ is just $H = \pi^N M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.
-
-Recall that the \emph{Newton polygon} of a polynomial $\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} of a matrix is the Newton polygon of its characteristic polynomial.
-If $m_1 \le \cdots \le m_n$ are the valuations of the diagonal entries of $\Delta_M$ in a Smith factorization $U_M \cdot \Delta_M \cdot V_M$ of $M$ as in the previous section, the \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = i+1}^n m_i)$.
+Write $\mathcal{E}$ for the precision lattice in $K_{<n}[X]$ of the characteristic polynomial of $M$, where $M$ is known with precision lattice $\mathcal{L}$.
+As long as $\mathcal{L}$ is sufficiently small, Proposition \ref{prop:precision} implies that $\mathcal{E}$ will be generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $\mathcal{L}$.
+When $\mathcal{L} = \mathcal{L}_N := \pi^N M_{n,n}(\O_K)$, we can further analyze $\mathcal{E}$.
+
+Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = \sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is $\NP(\charpoly(M))$.
+The \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$.
+Such polygons arise naturally in tracking the precision of polynomials \cite{caruso-roe-vaccon:14a}*{\S 4.2}.
+Any such polygon $P$ yields a lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose Newton polygons lie above $P$.
+This lattice is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.
+One uses these polygons to do a certain kind of coordinate-wise precision tracking for polynomials, bounding the actual precision lattice.
+
+For any matrix $M$, the Newton polygon lies above the Hodge polygon \cite{kedlaya:padicDiffEq}*{Thm.
+4.3.11}.
+We now introduce another polygon that will be bounded by between these two, and will provide an estimate for $\mathcal{E}$.
 
 \begin{deftn}
-The \emph{precision polygon} of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
+The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
 \end{deftn}
 
-Any such polygon $P$ naturally yields a lattice in $K_{<n}[X]$, consisting of polynomials whose Newton polygons lie above $P$.
-The lattice associated to $P$ is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.
-
 If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.
@@
 \begin{prop}
-Suppose a matrix $M$ is known with precision $H = \pi^N M_{n,n}(\O_K)$.
-Then the precision of the characteristic polynomial of $M$ is contained within $S_N$ applied to the lattice of the precision polygon of $M$.
+Suppose a matrix $M$ is known with precision $\mathcal{L}_N$ with $N$ sufficiently large.  Then $\mathcal{E} \subseteq \mathcal{L}_{S_N(\PP(M))}$.
 \end{prop}
 \begin{proof}
-The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(H)$, the result follows.
+The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.
+Since these generate $\charp(M)(\mathcal{L}_0)$, the result follows from Proposition \ref{prop:precision}.
 \end{proof}
@@
 \begin{prop} \label{prop:prec_hodge}
-The precision polygon of $M$ lies above $T_1$ applied to the Hodge polygon of $M$.
+The precision polygon $\PP(M)$ is bounded below by the translated Hodge polygon $T_1(\HP(M))$.
 \end{prop}
@@
 \begin{proof}
-\todo{Add this proof from Xavier's e-mail}
+The coefficients of $\charpoly(M)$ can be expressed as traces of exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M)) = \tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.
+The entries of $\Lambda^i(\Delta_M)$ can be computed easily, and now the result follows from Proposition \ref{prop:precision2}.
+% Clarify and expand.
 \end{proof}
 
-For many matrices, the precision polygon equals the Hodge polygon.
+For many matrices, $\PP(M) = T_1(\HP(M))$.
 For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.
-\todo{p=3?}  Moreover, experiments support the following conjectures.
+For $p=3$, the fraction rises to $99.98\%$, with no clear dependence on dimension.
+
+We may also bound $\PP(M)$ above:
+
+\begin{prop} \label{prop:newton_prec}
+If $M \in M_{n,n}(\O_K)$, the precision polygon $\PP(M)$ is bounded above by the Newton polygon $\NP(M)$.
+\end{prop}
+\begin{proof}
+Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$ and $f_{i,j}$ as the $(i,j)$th entry of $\com(X - M)$.
+We write $f[k]$ for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = +\infty$.
+The equation $(X - M) \cdot \com(X - M) = f \cdot I$  yields, for all $i$ and $k$,
+\begin{align*}
+f[k] &= \inf(f_{i,i}[k-1], \val(m_{i,0}) + f_{0,i}[k], \ldots, \val(m_{i,n}) + f_{n,i}[k]) \\
+&\ge \inf(f_{j,i}[k]),
+\end{align*}
+with the infimum taken over $j$.  Taking lower convex hulls yields the result.
+\end{proof}
+
+Moreover, experiments support the following stronger results.
 
@@
 \begin{enumerate}
-\item The precision polygon is equal to the Hodge polygon if the determinant of $M$ has valuation $0$ or $1$, or if $M$ is $3 \times 3$.
-\item The height at $0$ of the precision polygon is equal to the height at $1$ of the Hodge polygon.
-\item The precision polygon lies below $T_1$ applied to the Newton polygon.
+\item $\PP(M) = T_1(\HP(M))$ if $\val(\det(M))$ is $0$ or $1$, or if $M \in M_{3,3}(\O_K)$.
+\item The precision polygon $\PP(M)$ is bounded above by $T_1(\NP(M))$.
 \end{enumerate}
 \end{conj}
-Empirically, the precision polynomial seems most likely to differ from the Hodge polygon at $1$, corresponding to the precision of the linear term of the characteristic polynomial.
-
-Of course, the precision polygon does necessarily agree with the full precision of the characteristic polynomial of $M$.
-In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the lattice associated to the precision polygon contains the actual precision lattice with index greater than $1$.
-This percentage increases as the valuation of the determinant increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.
-
-Moreover, one can construct examples with arbitrarily large extra lattice precision.
-If $M$ has elementary divisors with valuations $m_1 \le \cdots \le m_n$, Proposition \ref{prop:prec_hodge} implies that the precision lattice of the characteristic polynomial of $M$ is contained within $\O_{K,n}[X]$ with index at least $\sum_{i=1}^{n-1} m_i$.
-The precision lattice of $1 + M$ is obtained from the precision lattice of $M$ via the transformation $X \mapsto X + 1$, but the precision polygon is now flat with height $0$.
+Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at $1$, corresponding to the precision of the linear term of the characteristic polynomial.
+
+Of course, $\PP(M)$ does necessarily agree with $\mathcal{E}$.
+In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the index of $\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.
+This percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.
+
+Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : \mathcal{E}]$ arbitrarily large.
+Suppose the $\sigma_i(M)$ are large.
+Proposition \ref{prop:prec_hodge} implies that $\mathcal{E}$ is contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} \sigma_i(M)$.
+The precision lattice of $1 + M$ is obtained from $\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ is now flat with height $0$.
+
+In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ using the Hodge polygon rarely loses much precision.
+However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.
 
@@
 $\val(\det(D'(M)))$, which is the lattice-method loss of precision.
-Figure \ref{fig:LU} sums up the mean and standard deviation of those 
-losses for a sample of 2000 random matrices in $M_d(\mathbb{Z}_2)$.
+Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
+losses for a sample of 2000 random matrices in $M_{d,d}(\mathbb{Z}_2)$.
 %
@@
 
-When the couple $(V, V^\c)$ varies, the open subsets $\mathcal 
+When the pair $(V, V^\c)$ varies, the open subsets $\mathcal 
 U_{V,V^\c}$ cover the whole Grassmaniann and then define an atlas 
-of it. When implementing vector spaces on computer, we usually restrict 
+of it. When implementing vector spaces on a computer, we usually restrict 
 ourselves to the subatlas consisting of all charts of the form $(V_I, 
 V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
-\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspaces spanned by 
+\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspace spanned by 
 the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
@@
 \to V^\c$ whose restriction to $V^\c$ is the identity. It is an 
-affine space over $\Hom(V, V^\c)$ which can be identified to an open
+affine space over $\Hom(V, V^\c)$ which can be identified with an open
 subset of $\Grass(E,d)$ \emph{via} the map $f \mapsto \ker f$.
 
-It is actually easily seen that $\mathcal U_{V, V^\c}$ and $\mathcal 
+It is easily seen that $\mathcal U_{V, V^\c}$ and $\mathcal 
 U^\star_{V, V^\c}$ define the same open subset in $\Grass(E,d)$. 
@@
 If $E$ is a finite dimension vector space over $K$, we use the notation 
-$E^\star$ for its dual (\emph{i.e.} $E^\star = \Hom(E,K)$) and, given in 
-addition a subspace $V \subset E$, we denote by $V^\perp$ the subspace 
+$E^\star$ for its dual (\emph{i.e.} $E^\star = \Hom(E,K)$). If we are also given
+a subspace $V \subset E$, we denote by $V^\perp$ the subspace 
 of $E^\star$ consisting of linear maps that vanish on $V$. We recall
@@
 over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
-couple $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
-priori} vary, the map $\DI$ does not take its values in well-defined 
+pair $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
+priori} vary, the map $\DI$ does not take its values in a well-defined 
 Grassmannian. To tackle this problem, we stratify $\mathcal M$ as
 follows: for each integer $r \in [0,d]$, we introduce the subset
-$\mathcal M_r \subset \mathcal M$ consisting of those couples $(f,V)$
+$\mathcal M_r \subset \mathcal M$ consisting of those pairs $(f,V)$
 for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
 locally closed in $\mathcal M$ and therefore are submanifolds. 
-Moreover, $\DI$ induces by restriction respectable differentiable 
+Moreover, $\DI$ induces by restriction differentiable 
 functions between $K$-manifolds
@@
 where $\varphi : V \to W$ is the linear mapping induced by $f$. The 
-previous relation still holds if $(f,v)$ is replaced by a couple $(f', 
+previous relation still holds if $(f,v)$ is replaced by a pair $(f', 
 v') \in \mathcal M_r$ sufficiently close to $(f,v)$.
 Differentiating it and passing to the quotient we find, first, that the 
-tangent space of $\mathcal M_r$ at $(f,v)$ consists of couples $(df, dv) 
+tangent space of $\mathcal M_r$ at $(f,v)$ consists of pairs $(df, dv) 
 \in \Hom(E,F) \times \Hom(V,E/V)$ such that
@@
 We now consider the mapping $\II$ (for ``inverse image'') sending a 
-couple $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
+pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
 $f^{-1}(W)$. As before, this map does not take its value in a single 
@@
 the sub-manifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
-couples $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
+pairs $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
 induces a continuous function
@@
 W_r$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
-couples $(df,dw)$ such that
+pairs $(df,dw)$ such that
 $d \tilde v = (w \circ df + dw \circ f)_{|W}$
@@
 $\mathcal C_d$ as the sub-manifold of $\mathcal C$ consisting of those 
-couples $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get the way a 
+pairs $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get a 
 well-defined mapping $\mathcal C_d \to \Grass(E,d)$ whose differential
-can be computed similarly as before. We seek to state the result: the
+can be computed as before. The
 tangent space of $\mathcal C_d$ at a given point $(V_1, V_2)$ consists 
-of couples $(dv_1, dv_2) \in \Hom(V_1, E/V_1)
+of pairs $(dv_1, dv_2) \in \Hom(V_1, E/V_1)
 \times \Hom(V_2, E/V_2)$ such that $dv_1 \equiv dv_2 \pmod{V_1 + V_2}$ 
@@
 
-Using duality, we derive from this a similar result for the mapping 
+Using duality, we derive a similar result for the mapping 
 $(V_1, V_2) \mapsto V_1 \cap V_2$ (left to the reader).
@@
 through the charts $\mathcal U_{V_I, V_{I^\c}}$ (where $I$ is a subset 
-of $\{1, \ldots, n\}$) introduced above. This means more concretely 
-that a subspace $V \subset K^n$ is represented as the span of the rows
+of $\{1, \ldots, n\}$) introduced above. More concretely,
+a subspace $V \subset K^n$ is represented as the span of the rows
 of a matrix $G_V$ having the following extra property: there 
@@
 matrix. We recall that such a representation always exists and, when 
-the set of indices $I$ is fixed (and has good cardinality), there 
+the set of indices $I$ is fixed, there 
 exists at most one $G_V$ satisfying the above condition.
 Given a family of generators of $V$, one can compute $G_V$ and $I$ as 
-above by performing standard row-echelon. If we are choosing always the 
+above by performing standard row reduction. If we are choosing always the 
 first non-vanishing pivot, we get this way the set $I$ with the smallest 
@@
 representation. However, in the context of inexact base fields, it is 
-much more clever to always choose the pivot with maximal norm.
+much better to always choose the pivot with maximal norm.
 
@@
 family of linear equations, we can compute $I$ and $H_V$ by performing 
-column-echelon.
-
-We emphasize that passing from a representation to the other is almost 
-free (and also quite stable). To simplify notations, let us suppose 
+column reduction.
+
+We emphasize that passing from one representation to the other is almost 
+free (and also quite stable). To simplify notation, let us suppose 
 that $I = \{1, \ldots, d\}$ with $d = \dim V$. Then, denoting by $I_d$ 
@@
 we just have to apply $f$ to each row of $G_V$ (getting this way a
-family of generators of $f(V)$) and then row-echelonize. Dually, the
-second representation is nice for computing inverse images --- and 
-in particular kernels --- and intersections. Since, one know moreover
-how to go back and forth between the two dual representations, we have 
-a complete picture of the underlying algorithmic.
-
-Regarding to precision, one can either track it in the standard way or 
+family of generators of $f(V)$) and then row reduce. Dually, the
+second representation works well for computing inverse images, including 
+kernels, and intersections. Since translating between the two dual
+representations is straightforward, we get algorithms for solving both
+problems using either representation.
+
+With regard to precision, one can either track it in the standard way or 
 by using Proposition \ref{prop:precision} and the differential 
@@
 {\bf Some experiments.}
-Let us consider the example of computation given by the following
+Let us consider the example computation given in the following
 algorithm.
@@
 always chosen sufficiently large so that it does not affect the
-behavior of precision). The ``coord-wise column'' corresponds to
-the standard way to track precision. On the other hand, in the two 
-last columns, the precision is tracked using lattices. The meaning of 
-``diffused'' is the same than those explained in \S 
+behavior of precision). The ``coord-wise'' column corresponds to
+the standard way of tracking precision. On the other hand, in the two 
+last columns, the precision is tracked using lattices. The term 
+``diffused'' has the same meaning as in \S 
 \ref{subsec:mulmatrix} when we were dealing with matrix multiplication: 
@@
 it turns out that these ``diffused'' digits may be ``separated'' by
-working in another basis of the underlying space $\Q_2^3$
+working in another basis of the underlying space $\Q_2^3$.
 The fact that only negative numbers appear in this column means that we 
@@
 We prove Proposition \ref{prop:boundLambdaf} in the slightly more
-general context of $K$-Banach.
+general context of $K$-Banach spaces.
 
@@
 
-Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces 
+Let $U$, $V$ and $W$ be three open subsets in $K$-Banach spaces 
 $E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
@@
 \label{prop:boundhr}
-With the above notations, we have for all nonnegative integer $r$:
+With the above notation, we have
 $$\Vert h_r \Vert \leq \sup_{m, (n_i)}
   \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert$$
-where the supremum is taken over all couples $(m, (n_i))$ where $m$
+for all nonnegative integer $r$, where the supremum is taken over all pairs $(m, (n_i))$ where $m$
 is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
@@
 \begin{lem}
-Let $\mathcal E$ be a $K$-vector space. Let $\varphi : \mathcal E^n \to 
-K$ be a symmetric $s$-linear form and $\psi: E \to K$ defined by 
+Let $E$ be a $K$-vector space. Let $\varphi : E^m \to 
+K$ be a symmetric $m$-linear form and $\psi: E \to K$ defined by 
 $\psi(x) = \varphi(x, x, \ldots, x)$.
-Given $k_1, \ldots, k_\ell$ some positive integers whose sum is $m$ and 
-$x_1, \ldots, x_\ell$ some elements of $\mathcal E$, we have:
+Given positive integers $k_1, \ldots, k_\ell$ whose sum is $m$ and 
+$x_1, \ldots, x_\ell \in E$, we have:
 $$\begin{array}{l}
@@
 equals $m$ together with $(\ell+1)$ elements $x_1, \ldots, x_{\ell+1}$
-lying in the unit ball of $\mathcal E$.
+lying in the unit ball of $E$.
 We also consider a new variable $\lambda$ varying in
 $\O_K$. We set $x'_i = x_i$, $k'_i = k_i$ when $i < \ell$ and $x'_\ell 
-= x_\ell + \lambda x_{\ell+1}$, $k'_\ell = k_\ell + k_{\ell+1}$. By the 
+= x_\ell + \lambda x_{\ell+1}$ and $k'_\ell = k_\ell + k_{\ell+1}$. By the 
 induction hypothesis, we know that the inequality:
-$$\Big\Vert \binom s {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
+$$\Big\Vert \binom m {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
 \varphi(x'_1, \ldots, x'_1, \ldots, x'_\ell, \ldots, x'_\ell) \Big\Vert
@@
 $\lambda^j$ is:
-$$\binom s {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
+$$\binom m {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
 \binom {k'_\ell} {j} \cdot
 \varphi(\underline x_j) = 
-\binom s {k_1 \,\, \cdots \,\, k_{\ell-1} \,\, j} \cdot
+\binom m {k_1 \,\, \cdots \,\, k_{\ell-1} \,\, j} \cdot
 \varphi(\underline x_j)$$
@@
 the unit ball, the norm of all its coefficients must be at most $\Vert \psi
-\Vert$ as well. Looking at the coefficient in $\lambda^{k_{l+1}}$, we find
+\Vert$ as well. Looking at the coefficient of $\lambda^{k_l}$, we find
 the desired result.
@@
 some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
-Following \cite{caruso-roe-vaccon:14a}, we define $\NP(f) : \R \to \R \cup 
-\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
-\log \Vert f_n \Vert$ for all integer $n$. We recall that $\Lambda(f)$ 
-is the Legendre transform of $\NP(f)$ 
-\cite[Proposition 3.9]{caruso-roe-vaccon:14a}.
-We define $\NP(g)$ and $\NP(h)$ similarly.
-We recall that $\alpha$ is a nonnegative real number such that $\Vert n! 
-\Vert \geq e^{-\alpha n}$ for all positive integer $n$.
+We recall that $\Lambda(f)$ 
+is the Legendre transform of the Newton polygon $\NP(f)$ defined in Section \ref{sec:matrices}
+\cite[Proposition 3.9]{caruso-roe-vaccon:14a} and that $\alpha$ is a nonnegative
+real number such that $\Vert n! \Vert \geq e^{-\alpha n}$ for all positive integer $n$.
 
@@
 \Vert n \Vert \cdot \Vert f_n \Vert$.
-Combining this with Proposition \ref{prop:boundhr}, we get, for all
-nonnegative integer $r$:
+Combining this with Proposition \ref{prop:boundhr}, we get
 $$\Vert (r+1) f_{r+1} \Vert \leq
@@
   \prod_{i=1}^m \max(\Vert f_{n_i} \Vert, \Vert h_{n_i} \Vert)$$
-where the supremum runs over all couples $(m, (n_i))$ where $m$
+for all nonnegative integer $r$,
+where the supremum runs over all pairs $(m, (n_i))$ where $m$
 is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
@@
 \nu$. Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to 
-check that such couples $(a,b)$ satisfy the hypothesis of Lemma
+check that such pairs $(a,b)$ satisfy the hypothesis of Lemma
 \ref{lem:boundLambdaf}. Clearly:

### 2015-01-18 examples/stability.tex
Small rewriting of Section 3.3 (on characteristic polynomials)

@@
 
+Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic 
+polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of 
+polynomials of degree less than $n$.  Then the differential of 
+$\charpoly$ at a point $M$ is
+\[
+\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
+\]
+The image is the $K$-span of the entries of $\com(X{-}M)$, which is 
+clearly contained within $K_{<n}[X]$.  In fact, the image will equal 
+$K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same 
+generalized eigenvalue. For now on, we make this assumption.
+
+Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = 
+\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, 
+\val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
+$\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower 
+convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$. For 
+any matrix $M$, the Newton polygon lies above the Hodge polygon 
+\cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.
+
+Such polygons arise naturally in tracking the precision of polynomials 
+\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
+lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose 
+Newton polygons lie above $P$.  This lattice is generated by monomials 
+$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
+One uses these polygons to do a certain kind of coordinate-wise 
+precision tracking for polynomials, bounding the actual precision 
+lattice. We now introduce another polygon that will be bounded by 
+between these two, and will provide an estimate for $\mathcal{E}$.
+
+\begin{deftn}
+The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of 
+the Newton polygons of the entries of $\com(X{-}M)$.
+\end{deftn}
+
+It is clear from the definition $\mathcal L_{\PP(M)} \subset 
+\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
+$M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
+which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
+Proposition \ref{prop:precision}, the precision polygon determines the 
+best precision loses we can hope if we are encoding precision on 
+polynomials by polygons.
+
+It turns out that the precision polygon is related to Hodge and Newton 
+polygon. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
+the translated polygon with vertices $(x_i - n, y_i)$.
+
+\begin{prop} \label{prop:polygons}
+The precision polygon $\PP(M)$ lies between $T_1(\HP(M))$
+and $\NP(M)$.
+
+Moreover $\PP(M)$ and $T_1(\HP(M))$ meet at $0$ and $n{-}1$.
+\end{prop}
+
+\begin{proof}
+The coefficients of $\charpoly(M)$ can be expressed as traces of 
+exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M))$,
+which is $\tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.  
+Computing $\Lambda^i(\Delta_M)$, we get the first statement of
+the Proposition. For $i = 1$, we further find that $\PP(M)$ vanishes
+at the abscissa $n{-}1$. By definition so does $T_1(\HP(M))$. The fact 
+that $\PP(M)$ and $T_1(\HP(M))$ meet at abscissa $0$ follows from
+Proposition \ref{prop:detmatrix}.
+
+It remains to prove the comparison with Newton polygon.
+Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$, 
+$f_{i,j}$ as the $(i,j)$th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
+\val(m_{i,j})$. We write $f[k]$ 
+for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
++\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
+for all $i$ and $k$,
+\begin{align*}
+f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
+\mu_{i,n} + f_{n,i}[k]) \\
+&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
+\end{align*}
+with the infimum is taken over $j$. 
+Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing
+(it follows from the comparison with the Hodge polygon), we get the 
+result.
+\end{proof}
+
+\begin{rem}
+Experiments actually support the following stronger result: $\PP(M)$ is 
+bounded above by $T_1(\HP(M))$.
+\end{rem}
+
+For many matrices, $\PP(M) = T_1(\HP(M))$.  For random matrices over 
+$\Z_2$, the $2$-adic precision polygon is equal to the Hodge polygon in 
+$99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  
+Over $\Z_3$, the fraction rises to $99.98\%$, with no clear dependence on 
+dimension.
+Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at 
+$1$, corresponding to the precision of the linear term of the 
+characteristic polynomial.
+
+Of course, $\PP(M)$ does necessarily agree with the precision lattice 
+$\mathcal{E} = \charp(M)(\mathcal L_0)$.  In $11\%$ of cases in 
+dimension $3$, up to $15\%$ of cases in dimension $8$, the index of 
+$\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.  This 
+percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in 
+dimension $9$ for matrices constrained to have determinant with $2$-adic 
+valuation $12$.
+Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : 
+\mathcal{E}]$ arbitrarily large.  Suppose the $\sigma_i(M)$ are large.  
+Proposition \ref{prop:polygons} implies that $\mathcal{E}$ is 
+contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} 
+\sigma_i(M)$.  The precision lattice of $1 + M$ is obtained from 
+$\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ 
+is now flat with height $0$.
+
+In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
+using the Hodge polygon loses precision but not so much. Nonetheless, as 
+already illustrated for matrix multiplication, this loses of precision 
+may accumulate and become significant if we need to compute many 
+characteristic polynomials. In the case of a single characteristic 
+polynomial, we note moreover that if the $\sigma_i(M)$'s are known to be 
+large or if $M$ is a translate of such a matrix, using lattice precision 
+can be very useful.
+
+\begin{comment}
+\subsection{Characteristic polynomials} \label{ssec:charpoly}
+
 Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of polynomials of degree less than $n$.
@@
 However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.
+\end{comment}
 

### 2015-01-18 examples/stability.tex
\todo{} fixed

@@
 
-\medskip
+\smallskip
 
@@
 {\bf Notations.}
-Throughout the paper, the letter $K$ will refer to a complete 
-discrete valuation field of characteristic $0$, that is either a finite 
-extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
-series over a field of characteristic $0$ \todo{what about mixed characteristic with infinite residue field?}.
-We denote by $\val : K \to \Z 
-\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
-and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
-$\Vert \cdot \Vert$ the corresponding norm.
+Throughout the paper, the letter $K$ will refer to a complete discrete 
+valuation field of characteristic $0$. Usual examples are finite 
+extensions of $\Qp$ and Laurent series fields over a field of 
+characteristic $0$. We denote by $\val : K \to \Z \cup \{+\infty\}$ of 
+$K$ --- it is the $p$-adic valuation when $K = \Qp$ and the usual 
+valuation of a Laurent series when $K = k((t))$ --- and by $\Vert \cdot 
+\Vert$ the corresponding norm.
 
@@
 proposition which can be directly combined with Proposition 3.12 of 
-\cite{caruso-roe-vaccon:14a}.
+\cite{caruso-roe-vaccon:14a}. Following \cite{caruso-roe-vaccon:14a},
+if $\varphi$ is a convex function and $v$ is a real number, we define
+$$\varphi_{\geq v} : x \mapsto \inf_{y \geq 0} \big( \varphi(x+y)
+- vy \big).$$
+It is highest convex function with $\varphi_{\geq v} \leq \varphi$ and 
+$\varphi'_{\geq v} \geq v$.
 
@@
 \min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
-for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$. \todo{What is $\Lambda(f)_{\leq 2}$?}
+for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
 \end{prop}

### 2015-01-18 examples/stability.tex
Update charpoly section

--- 2015-01-18 examples/stability.tex
@@
 precision tracking for polynomials, bounding the actual precision 
-lattice. We now introduce another polygon that will be bounded by 
-between these two, and will provide an estimate for $\mathcal{E}$.
+lattice. We now introduce another polygon, bounded 
+between these two, that will provide an estimate on the precision of $\charpoly(M)$.
 
@@
 which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
-Proposition \ref{prop:precision}, the precision polygon determines the 
-best precision loses we can hope if we are encoding precision on 
-polynomials by polygons.
-
-It turns out that the precision polygon is related to Hodge and Newton 
-polygon. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
+Proposition \ref{prop:precision}, the precision polygon determines the minimal 
+precision losses possible when encoding polynomial precision using polygons.
+
+It turns out that the precision polygon is related to the Hodge and Newton 
+polygons. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
 the translated polygon with vertices $(x_i - n, y_i)$.
@@
 \end{align*}
-with the infimum is taken over $j$. 
+with the infimum taken over $j$. 
 Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing
-(it follows from the comparison with the Hodge polygon), we get the 
+(which follows from the comparison with the Hodge polygon), we get the 
 result.
@@
 Experiments actually support the following stronger result: $\PP(M)$ is 
-bounded above by $T_1(\HP(M))$.
+bounded above by $T_1(\NP(M))$.
 \end{rem}
@@
 In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
-using the Hodge polygon loses precision but not so much. Nonetheless, as 
-already illustrated for matrix multiplication, this loses of precision 
-may accumulate and become significant if we need to compute many 
-characteristic polynomials. In the case of a single characteristic 
-polynomial, we note moreover that if the $\sigma_i(M)$'s are known to be 
+using the Hodge polygon loses only a small amount of precision. However,
+if the $\sigma_i(M)$'s are known to be 
 large or if $M$ is a translate of such a matrix, using lattice precision 
 can be very useful.
-
-\begin{comment}
-\subsection{Characteristic polynomials} \label{ssec:charpoly}
-
-Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of polynomials of degree less than $n$.
-Then the determinant of $\charpoly$ at a point $M$ is
-\[
-\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
-\]
-The image is the $K$-span of the entries of $\com(X - M)$, which is clearly contained within $K_{<n}[X]$.
-In fact, the image will equal $K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same generalized eigenvalue.
-
-Write $\mathcal{E}$ for the precision lattice in $K_{<n}[X]$ of the characteristic polynomial of $M$, where $M$ is known with precision lattice $\mathcal{L}$.
-As long as $\mathcal{L}$ is sufficiently small, Proposition \ref{prop:precision} implies that $\mathcal{E}$ will be generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $\mathcal{L}$.
-When $\mathcal{L} = \mathcal{L}_N := \pi^N M_{n,n}(\O_K)$, we can further analyze $\mathcal{E}$.
-
-Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = \sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is $\NP(\charpoly(M))$.
-The \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$.
-Such polygons arise naturally in tracking the precision of polynomials \cite{caruso-roe-vaccon:14a}*{\S 4.2}.
-Any such polygon $P$ yields a lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose Newton polygons lie above $P$.
-This lattice is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.
-One uses these polygons to do a certain kind of coordinate-wise precision tracking for polynomials, bounding the actual precision lattice.
-
-For any matrix $M$, the Newton polygon lies above the Hodge polygon \cite{kedlaya:padicDiffEq}*{Thm.
-4.3.11}.
-We now introduce another polygon that will be bounded by between these two, and will provide an estimate for $\mathcal{E}$.
-
-\begin{deftn}
-The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
-\end{deftn}
-
-If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.
-
-\begin{prop}
-Suppose a matrix $M$ is known with precision $\mathcal{L}_N$ with $N$ sufficiently large.  Then $\mathcal{E} \subseteq \mathcal{L}_{S_N(\PP(M))}$.
-\end{prop}
-\begin{proof}
-The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.
-Since these generate $\charp(M)(\mathcal{L}_0)$, the result follows from Proposition \ref{prop:precision}.
-\end{proof}
-
-\begin{prop} \label{prop:prec_hodge}
-The precision polygon $\PP(M)$ is bounded below by the translated Hodge polygon $T_1(\HP(M))$.
-\end{prop}
-
-\begin{proof}
-The coefficients of $\charpoly(M)$ can be expressed as traces of exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M)) = \tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.
-The entries of $\Lambda^i(\Delta_M)$ can be computed easily, and now the result follows from Proposition \ref{prop:precision2}.
-% Clarify and expand.
-\end{proof}
-
-For many matrices, $\PP(M) = T_1(\HP(M))$.
-For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.
-For $p=3$, the fraction rises to $99.98\%$, with no clear dependence on dimension.
-
-We may also bound $\PP(M)$ above:
-
-\begin{prop} \label{prop:newton_prec}
-If $M \in M_{n,n}(\O_K)$, the precision polygon $\PP(M)$ is bounded above by the Newton polygon $\NP(M)$.
-\end{prop}
-\begin{proof}
-Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$ and $f_{i,j}$ as the $(i,j)$th entry of $\com(X - M)$.
-We write $f[k]$ for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = +\infty$.
-The equation $(X - M) \cdot \com(X - M) = f \cdot I$  yields, for all $i$ and $k$,
-\begin{align*}
-f[k] &= \inf(f_{i,i}[k-1], \val(m_{i,0}) + f_{0,i}[k], \ldots, \val(m_{i,n}) + f_{n,i}[k]) \\
-&\ge \inf(f_{j,i}[k]),
-\end{align*}
-with the infimum taken over $j$.  Taking lower convex hulls yields the result.
-\end{proof}
-
-Moreover, experiments support the following stronger results.
-
-\begin{conj}
-\begin{enumerate}
-\item $\PP(M) = T_1(\HP(M))$ if $\val(\det(M))$ is $0$ or $1$, or if $M \in M_{3,3}(\O_K)$.
-\item The precision polygon $\PP(M)$ is bounded above by $T_1(\NP(M))$.
-\end{enumerate}
-\end{conj}
-Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at $1$, corresponding to the precision of the linear term of the characteristic polynomial.
-
-Of course, $\PP(M)$ does necessarily agree with $\mathcal{E}$.
-In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the index of $\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.
-This percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.
-
-Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : \mathcal{E}]$ arbitrarily large.
-Suppose the $\sigma_i(M)$ are large.
-Proposition \ref{prop:prec_hodge} implies that $\mathcal{E}$ is contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} \sigma_i(M)$.
-The precision lattice of $1 + M$ is obtained from $\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ is now flat with height $0$.
-
-In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ using the Hodge polygon rarely loses much precision.
-However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.
-\end{comment}
 

### 2015-01-18 examples/stability.tex
Small fixes

--- 2015-01-18 examples/stability.tex
@@
 with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
-for the valuation of the $(i,i)$-th entry of $\Delta_M$, and by convention set
+for the valuation of the $(i,i)$th entry of $\Delta_M$, and by convention set
 $\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.
@@
 and therefore the first summand in $M$ varies in the subspace of 
-$M_{a,c}(\O_K)$ consisting of matrices whose all entries on the $i$-th
+$M_{a,c}(\O_K)$ consisting of matrices whose all entries on the $i$th
 row have valuation at least $a_i$. Arguing similarly for the second
@@
 general. There is a more pedestrian way to visualize this phenomenon. 
-Remark that the product $AB$ can be rewritten:
+Remark that the product $AB$ can be rewritten
 $AB = U_A \cdot P \cdot V_B$ with
 $P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
-Tracking precision in the usual way, we see that the $(i,j)$-th entry
+Tracking precision in the usual way, we see that the $(i,j)$th entry
 of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
@@
 precision.
-
-\smallskip
 

### 2015-01-19 examples/stability.tex
Added intro, Def 2.3 and many small changes

--- 2015-01-18 examples/stability.tex
@@
+%% from this description could be cut for space
+
 \documentclass{sig-alternate}
@@
 \usepackage{amsmath,amssymb}
+\usepackage{amsthmnoproof}
 \usepackage{amsrefs}
@@
 
-\newtheorem{theo}{Theorem}[section]
-\newtheorem{lem}[theo]{Lemma}
-\newtheorem{prop}[theo]{Proposition}
-\newtheorem{cor}[theo]{Corollary}
-\newtheorem{quest}[theo]{Question}
-\newtheorem{conj}[theo]{Conjecture}
-%\theoremstyle{definition}
-\newtheorem{rem}[theo]{Remark}
-\newtheorem{ex}[theo]{Example}
-\newtheorem{deftn}[theo]{Definition}
-\newtheorem{rmk}[theo]{Remark}
-
 \newcommand{\N}{\mathbb N}
@@
 
+\newtheorem{theo}{Theorem}[section]
+\newtheorem{lem}[theo]{Lemma}
+\newtheorem{prop}[theo]{Proposition}
+\newtheorem{cor}[theo]{Corollary}
+\newtheorem{quest}[theo]{Question}
+\newtheorem{conj}[theo]{Conjecture}
+\theoremstyle{definition}
+\newtheorem{rem}[theo]{Remark}
+\newtheorem{ex}[theo]{Example}
+\newtheorem{deftn}[theo]{Definition}
+
 \title{p-adic stability in linear algebra}
@@
 For about twenty years, the use of $p$-adic methods in symbolic 
-computation has been gaining popularity: they were successfully 
-used for instance 
-to compute compose products of polynomials 
+computation has been gaining popularity. Such methods were 
+used to compute composed products of polynomials 
 \cite{boston-gonzalez-perdry-schost:05a}, 
@@
 to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
-and, last but not least,
-to count points on certain varieties (mainly curves) using $p$-adic 
-cohomology theories (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
+and to count points on varieties using $p$-adic 
+cohomology theories (\emph{cf.} \cite{kedlaya:01a,lauder:04a} and many
 followers).
 However, a general framework allowing a precise study of $p$-adic 
-precision --- which is a main issue we encounter when trying to work 
+precision --- the main issue encountered when computing 
 with $p$-adic numbers --- was designed only recently in 
@@
 
-\todo{Rewrite the introduction}
+In \cite{caruso-roe-vaccon:14a}, we advocate the use of lattices
+to track the precision of vectors and points on $p$-adic manifolds.
+The main result of \emph{loc. cit.}, recalled here as Proposition \ref{prop:precision},
+allows for the propagation of precision using differentials.  While
+allowing arbitrary lattices in representing precision carries a high
+cost in space and time requirements, the reduced precision loss
+can sometimes make up for these costs as part of a larger algorithm.
+In this paper we apply the ideas of \cite{caruso-roe-vaccon:14a} to
+a variety of linear algebraic tasks.
+
+\smallskip
+
+\noindent
+{\bf Main Results.}
+We give a number of contexts where lattice-based precision methods
+outperform the standard coordinate-wise methods.  The two most striking
+examples are an analysis of matrix multiplication and of 
+sums and intersections of subspaces.  In Proposition \ref{prop:mulmatrix},
+we give a formula for the lattice-precision of the product of two matrices.
+In Figure \ref{fig:mulmatrix}, we describe the precision loss in multiplying
+$n$ matrices, which appears linear in $n$ when using standard methods
+and logarithmic in $n$ when using lattice methods.  We also give an
+example in Figure \ref{fig:vectorspace} where lattice methods
+actually yield an \emph{increase} in precision as the computation
+proceeds.
 
@@
 D)$ --- and sometimes even $O(1)$. We would like to underline in 
-particular that the instability phenomema are much more apparent when 
+particular that the instability phenomena are much more apparent when 
 the residue field is small. It is the reason why we will always take
@@
 {\bf Organization of the paper.}
-The section \ref{sec:theory} introduces the theory of ultrametric 
-precision developed in \cite{caruso-roe-vaccon:14a} and completes 
-it on a technical point.
-In \S \ref{sec:matrices}, we focus on matrices... \todo{Complete
-this}.
-
-\smallskip
-
-\noindent
-{\bf Notations.}
+Section \ref{ssec:latticeprec} recalls the theory of 
+precision developed in \cite{caruso-roe-vaccon:14a} and defines
+a notion of diffuse precision for comparing to coordinate-wise methods.
+In particular, we recall Proposition \ref{prop:precision}, which
+allows the computation of precision using differentials.
+Proposition \ref{prop:boundLambdaf2} in Section \ref{ssec:boundLambdaf}
+is a technical result that will be used to determine the applicability of
+Proposition \ref{prop:precision}.
+
+In Section \ref{ssec:mulmatrix}, we analyze matrix multiplication and report
+on experiments that demonstrate the utility of lattice-based precision tracking.
+Section \ref{ssec:det} finds the precision of the determinant of a
+matrix, and Section \ref{ssec:charpoly} applies the resulting formula to
+characteristic polynomials.  We define the precision polygon of a matrix, which
+gives a lower bound on the precision of the characteristic polynomial.
+The precision polygon lies above the Hodge polygon of the matrix, and
+we give statistics on the difference and the amount of diffuse precision.
+Finally, we find the precision of LU factorization, determine the conditions under which
+Proposition \ref{prop:precision} may be applied, and again describe experiments
+with lattice-based precision analysis.
+
+Section \ref{ssec:grassgeo} reviews the geometry of Grassmannians,
+which we apply in Section \ref{ssec:grassdiff} to differentiating
+the direct image, inverse image, sum and intersection maps
+between Grassmannians.  We then report in Section \ref{grassimpl}
+on tracking the precision of subspace arithmetic in practice.
+
+In the appendix, we give a proof of Proposition \ref{prop:boundLambdaf}.
+
+\smallskip
+
+\noindent
+{\bf Notation.}
 Throughout the paper, the letter $K$ will refer to a complete discrete 
@@
 
-The aim of this section is to summarize briefly the content of 
-\cite{caruso-roe-vaccon:14a} and complete it on certain points.
+The aim of this section is to briefly summarize the content of 
+\cite{caruso-roe-vaccon:14a} and fill in certain details.
 
 \subsection{Lattices as precision data}
+\label{ssec:latticeprec}
 
@@
 
-To make this sentence more explicit, let us consider a finite 
+More explicitly, let us consider a finite 
 dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
-actually those of Banach spaces. However we will not need infinite 
+actually those of Banach spaces. However, we will not need infinite 
 dimensional spaces in the present paper.} normed vector space $E$ 
@@
 We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
-precise statement and seek here to state the particular case of
-integral polynomial functions. A function $f : E \to F$ is said 
+precise statement.  We state here the case of
+integral polynomial functions. A function $f : E \to F$ is said to be
 \emph{integral polynomial} if it is given by multivariate polynomial 
@@
 \label{prop:precision2}
-We keep the notations of Proposition \ref{prop:precision} and assume 
+We keep the notation of Proposition \ref{prop:precision} and assume 
 in addition that $f$ is integral polynomial. Let $C$ be a positive real
@@
 In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
-to manifolds over $K$ and covers this way a very large families of
+to manifolds over $K$ and covers in this way a very large familiy of
 situations (\emph{e.g.} computation with vector spaces as discussed
@@
 
+In this paper, we will compare the use of lattices to track precision
+with the standard coordinate-wise methods.  The following definition
+will be useful in discussing the differences.  Suppose $E$ is equipped
+with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.
+
+\begin{deftn} \label{def:diffused}
+If $H \subset E$ is a lattice, let
+\[
+H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H).
+\]
+Set $k = \log_p([H_0 : H])$.  Then we say that $H$ has $k$
+\emph{diffused digits of precision}.
+\end{deftn}
+If $H$ represents the actual precision of some object, then
+$H_0$ is the best coordinate-wise approximation to $H$.
+The diffused precision is not visible coordinate-wise in the
+specified basis of $E$.
+
 \subsection{A bound on a growing function}
@@
 In the next sections, we will compute the derivative of several standard 
-operations and remark that it sometimes has a simple expression in term 
-of the input and the output. In other terms, the function $f$ modeling
+operations and sometimes give a simple expression in term 
+of the input and the output. In other words, the function $f$ modeling
 such an operation satisfies a differential equation of the form
-$f' = g \circ (f, \id)$
+$f' = g \circ (f, \id)$  %%%%%%%% HERE
 where $g$ is a given --- and hopefully rather simple --- function. The 
 aim of this subsection is to study this differential equation and to 
-derive from it some bounds on the growing function $\Lambda(f)$. 
-\emph{In the following, we shall always assume that $K$ has 
+derive from it certain bounds on the growing function $\Lambda(f)$. 
+\emph{We will assume that $K$ has 
 characteristic $0$.}
 
-We actually generalize a bit the setting above and consider the 
+Let $E, F$ and $G$ be finite-dimensional normed vector spaces with
+$U \subseteq E$ and $V \subseteq F$ and $W \subset G$ open subsets.
+Generalizing the setting above, we consider the 
 differential equation:
@@
 \end{equation}
-Here $g$ and $h$ are known locally analytic functions and $f$ is the 
-unknown, which is assumed to be locally analytic as well. The function
-$f$ goes from $U$ to $V$, two open subsets in two finite dimensional
-normed vector spaces $E$ 
-and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
-is again an open subset in a normed $K$-vector space $G$. Consequently,
-$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ of
-(continuous) linear maps $E \to F$.
+Here $g : V \times W \to \Hom(E, F)$ and $h : U \to W$ are known
+locally analytic functions and $f : U \to V$ is the 
+unknown locally analytic function. 
 In what follows, we always assume that $V$ and $W$ contain the origin, 
@@
 for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
-accordingly) so that they both vanish at $0$ and second, in order to 
-apply Proposition \ref{prop:precision2}, the derivative $f'(0)$ 
+accordingly) so that they both vanish at $0$, and second, in order to 
+apply Proposition \ref{prop:precision2} the derivative $f'(0)$ 
 needs to be surjective and therefore \emph{a fortiori} nonzero.
@@
 $({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
-graph of $\Lambda(f)$ lies below the plain blue line. We remark
-moreover that the Taylor expansion of $f(z)$ starts with the term
+graph of $\Lambda(f)$ lies below the plain blue line. Moreover, we remark
+that the Taylor expansion of $f(z)$ starts with the term
 $g(0) z$. Hence, on a neighborhood on ${-}\infty$, we have 
@@
 get:
-$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
-  \quad \forall x \in \R.$$
-In other words, the graph of $\Lambda(f)$ lies above the brown line.
+\[
+\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert,
+\]
+%  \quad \forall x \in \R.$$
+for all $x \in \R$.  In other words, the graph of $\Lambda(f)$ lies above the brown line.
 Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
@@
 the dashed blue line defined as the line of slope $2$ passing through
-the first break point of the blue plain line --- which has coordinate 
+the first break point of the blue plain line, which has coordinate 
 $(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
@@
 
-As a consequence of the above discussion, we derive the following 
-proposition which can be directly combined with Proposition 3.12 of 
+As a consequence, we derive the following 
+proposition, which can be combined with Proposition 3.12 of 
 \cite{caruso-roe-vaccon:14a}. Following \cite{caruso-roe-vaccon:14a},
-if $\varphi$ is a convex function and $v$ is a real number, we define
+if $\varphi$ is a convex function and $v \in \R$, we define
 $$\varphi_{\geq v} : x \mapsto \inf_{y \geq 0} \big( \varphi(x+y)
@@
 \label{prop:boundLambdaf2}
-Keeping the above notations, we have:
+Keeping the above notation, we have:
 $$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
@@
 \begin{proof}
-Just remark that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
+The inequality follows from the fact that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
 the dashed blue line.
@@
 
-Beyond this particular case, our general recommendation for choosing the 
-function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
-e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
-\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
-underlying vector spaces. The function $\Lambda(h)$ then maps $x$ to $x + 
-\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
+Beyond this particular case, we recommend choosing the 
+function $h$ by endowing $F$ with the second norm $\Vert x \Vert'_F = 
+e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to taking $h : (F, \Vert 
+\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ to be the identity on the
+underlying vector spaces. The function $\Lambda(h) : \R \to \R$ then maps $x$ to $x + 
+\mu$ and we can choose $\Lambda_h = \Lambda(h)$.
 \end{rem}
@@
 
-The notation $M_{m,n}(K)$ refers to the $K$-vector space of matrices 
-with $m$ rows and $n$ columns with coefficients in $K$.  We will repeatedly
+Let $M_{m,n}(K)$ denote the space of $m \times n$ matrices over $K$.  We will repeatedly
 use the Smith decomposition for $M \in M_{m,n}(K)$:
@@
 \subsection{Multiplication}
-\label{subsec:mulmatrix}
+\label{ssec:mulmatrix}
 
@@
 \label{prop:mulmatrix}
-With the above notations, we have
+With the above notation, we have
 \[
@@
 and therefore the first summand in $M$ varies in the subspace of 
-$M_{a,c}(\O_K)$ consisting of matrices whose all entries on the $i$-th
+$M_{a,c}(\O_K)$ consisting of matrices whose entries on the $i$-th
 row have valuation at least $a_i$. Arguing similarly for the second
@@
 
-Let us briefly comment on Proposition \ref{prop:mulmatrix} from the 
-perspective of precision. The second statement roughly means that when 
-we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
+From the perspective of precision, the second statement of Proposition \ref{prop:mulmatrix}
+means that the computation of $AB$ gains $\sum_{m=1}^\infty 
 a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
@@
 are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
-(\emph{cf} Proposition \ref{prop:precision2}). However these digits are 
-not --- even individually --- attached to some particular entry of $AB$ 
-but are ``spread out'' among all its entries. Thus, a usual 
-coordinate-wise precision data does not see this gain of precision in 
-general. There is a more pedestrian way to visualize this phenomenon. 
-Remark that the product $AB$ can be rewritten:
+(\emph{cf.} Proposition \ref{prop:precision2}). However, many of these digits are 
+diffused in the sense of Definition \ref{def:diffused}.
+To change bases in order to make this increased precision visible
+with coordinates, write 
 $AB = U_A \cdot P \cdot V_B$ with
 $P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
-Tracking precision in the usual way, we see that the $(i,j)$-th entry
-of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
-gain of precision is visible here. However it is then ``spread out'' 
-when we are computing $U_A \cdot P \cdot V_B$ and mostly disappears in 
-the usual coordinate-wise precision model. The lattice precision model 
-then appears as a way to keep track of this ``diffused'' gain of
-precision.
-
-\smallskip
-
-One may wonder to what extent it is important to take care of these 
-``diffused'' gains of precision. For one single product of matrices, 
-it is not so much. However, the benefits become substantial
-when we are executing an algorithm that computes many products of
-matrices. Let us illustrate this with the following quite simple
+Tracking precision in the usual way, the $(i,j)$-th entry
+of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$.  Multiplication by
+$U_A$ and $V_B$ then diffuses the precision across the entries of $AB$.
+
+\smallskip
+
+We now consider the impact of tracking this diffuse precision.
+Although the benefit is not substantial for
+a single product of matrices, it accumulates as we multiply a large
+number of matrices.  We illustrate this phenomenon with the following simple
 example.
@@
 \subsection{Determinant}
-
-We now come to the computation of the determinant of a square $n 
-\times n$ matrix with coefficients in $K$. The differential computation is 
-quite classical.  The differential of the function $\det : M_{n,n}(K) \to 
-K$ at some point $M$ is the linear mapping
-$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),$$
+\label{ssec:det}
+
+The computation of the differential of $\det : M_{n,n}(K) \to K$
+is classical: at a point $M$ it is the linear map
+\[
+\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),
+\]
 where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
@@
 
-As we did for matrix multiplication, we first determine the image of
+As with matrix multiplication, we first determine the image of
 the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.
@@
 $M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
-is known at precision $O(\pi^{N+v})$.  Thus we are gaining $v$ significant 
+is known at precision $O(\pi^{N+v})$.  Thus we are gaining $v$  
 digits in absolute precision or, equivalently, losing $\sigma_n(M)$ digits
 of relative precision.
-Furthermore, one may easily compute the determinant
-of $M$ with this optimal precision by finding an approximate
+Furthermore, one may compute $\det(M)$ with this optimal precision by finding an approximate
 Smith decomposition with $\Delta_M$ known at precision $O(\pi^N)$ and
@@
 
-Moreover $\PP(M)$ and $T_1(\HP(M))$ meet at $0$ and $n{-}1$.
+Moreover, $\PP(M)$ and $T_1(\HP(M))$ meet at $0$ and $n{-}1$.
 \end{prop}
@@
 It remains to prove the comparison with Newton polygon.
-Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$, 
-$f_{i,j}$ as the $(i,j)$th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
+Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$-th entry of $M$, 
+$f_{i,j}$ as the $(i,j)$-th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
 \val(m_{i,j})$. We write $f[k]$ 
@@
 \end{align*}
-with the infimum taken over $j$. 
-Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing
-(which follows from the comparison with the Hodge polygon), we get the 
+with the infimum over $j$. 
+Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing,
+which follows from the comparison with the Hodge polygon, we get the 
 result.
@@
 
-Of course, $\PP(M)$ does necessarily agree with the precision lattice 
-$\mathcal{E} = \charp(M)(\mathcal L_0)$.  In $11\%$ of cases in 
-dimension $3$, up to $15\%$ of cases in dimension $8$, the index of 
-$\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.  This 
+Of course, the precision lattice $\mathcal{E} = \charp(M)(\mathcal L_0)$ 
+may contain diffuse precision which is not encapsulated in $\PP(M)$.
+Diffuse precision arises in $11\%$ of cases in 
+dimension $3$, up to $15\%$ of cases in dimension $8$.  This 
 percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in 
@@
 valuation $12$.
-Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : 
-\mathcal{E}]$ arbitrarily large.  Suppose the $\sigma_i(M)$ are large.  
+Moreover, one can construct examples with arbitrarily large amounts of diffuse precision.
+Suppose the $\sigma_i(M)$ are large.  
 Proposition \ref{prop:polygons} implies that $\mathcal{E}$ is 
@@
 using the Hodge polygon loses only a small amount of precision. However,
-if the $\sigma_i(M)$'s are known to be 
-large or if $M$ is a translate of such a matrix, using lattice precision 
-can be very useful.
+if the $\sigma_i(M)$'s are known to be large or if $M$ is a translate of
+such a matrix, using lattice precision can be very useful.
 
@@
 
-In this subsection, we study the LU factorization.
 %This example has been previously studied by Caruso in \cite{caruso:12a}, 
@@
 %the framework of \S \ref{ssec:boundLambdaf} applies directly.
-We first introduce several notations.
-We denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
+In this section, we denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
 and, given a positive real number $C$, we let $B(C)$ be the closed ball 
@@
 \noindent $\bullet$
-$O_n$ is the open subset of matrices whose all principal 
+$O_n$ is the open subset of matrices whose principal 
 minors do not vanish (we recall that the latest condition implies the
@@
 in order to derive a concrete result on precision. We pick $M_0 \in
-O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the ``translated'' 
+O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the translated
 function $f$ taking $M$ to $D(M_0+M)-D(M_0)$. We then have $f(0) = 0$ 
@@
 of $D'(M_0+M)$, we find $B(1) \subset f'(0) \cdot B(C)$
-for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover $f$ satisfies 
+for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover, $f$ satisfies 
 the differential equation $f' = g \circ f$ where $g$ is defined by:
@@
 Vector spaces are generally represented as subspaces of $K^n$ for some 
-$n$. Hence they naturally appear as points on Grassmannians. Therefore, 
+$n$ and hence naturally appear as points on Grassmannians. Therefore, 
 one can use the framework of \cite[Appendix A]{caruso-roe-vaccon:14a} to 
@@
 \subsection{Geometry of Grassmannians}
-
-Given $E$ be a finite dimensional vector space over $K$ and $d$ an 
-integer in the range $[0, \dim E]$ we denote by $\Grass(E,d)$ the 
-set of $d$-dimensional subspaces of $E$. It is the so-called 
-\emph{Grassmannian}. It is well-known that $\Grass(E,d)$ has a 
-natural structure of $K$-manifold. The aim of this subsection is to 
-recall standard facts about its geometry. In what follows, we set
-$n = \dim E$ and equip $E$ with a distinguished basis $(e_1, \ldots, 
-e_n)$.
+\label{ssec:grassgeo}
+
+Given $E$, a finite dimensional vector space over $K$, and $d$, an 
+integer in the range $[0, \dim E]$, we write $\Grass(E,d)$ for the
+Grassmannian of $d$-dimensional subspaces of $E$. It is well-known
+that $\Grass(E,d)$ has a natural structure of $K$-manifold. The aim of
+this subsection is to recall standard facts about its geometry. In what follows,
+we set $n = \dim E$ and equip $E$ with a distinguished basis
+$(e_1, \ldots, e_n)$.
 
@@
 {\bf Description and tangent space.}
-Let $V$ denote a fixed sub-vector space of $E$ of dimension $d$. The 
+Let $V$ denote a fixed subspace of $E$ of dimension $d$. The 
 Grassmannian $\Grass(E,d)$ can be viewed as the quotient of the set 
@@
 $\mathcal U_{V,V^\c}$ of all embeddings $f : V \hookrightarrow E$ 
-such that $\pi \circ f = \id_V$. Clearly it is an affine space over
+such that $\pi \circ f = \id_V$. Clearly, it is an affine space over
 $\Hom(V,V^\c)$. 
@@
 When the pair $(V, V^\c)$ varies, the open subsets $\mathcal 
-U_{V,V^\c}$ cover the whole Grassmaniann and then define an atlas 
-of it. When implementing vector spaces on a computer, we usually restrict 
+U_{V,V^\c}$ cover the whole Grassmaniann and define an atlas.
+When implementing vector spaces on a computer, we usually restrict 
 ourselves to the subatlas consisting of all charts of the form $(V_I, 
@@
 generators of $W$, we can determine such an $I$ together with the
-corresponding embedding $f : V_I \hookrightarrow E$ by echelonizing 
+corresponding embedding $f : V_I \hookrightarrow E$ by row reducing 
 the matrix of generators of $W$.
@@
 surjective morphisms $f : E \to E/V$ modulo the action (by 
-postcomposition) of $\GL(E/V)$ as well. This identification presents the 
-tangent space at a given point $V$ as the quotient $\Hom(V, E/V) / 
-\End(E/V)$, which turns out to be again isomorphic to $\Hom(V, E/V)$. 
+postcomposition) of $\GL(E/V)$. This identification presents the 
+tangent space at a given point $V$ as the quotient $\Hom(E, E/V) / 
+\End(E/V) \cong \Hom(V, E/V)$.
 Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
@@
 $\Hom(V, E/V)$ and $\Hom(V^\perp, E^\star/V^\perp)$ induced by 
-``transposition''. Furthermore, we observe that $\psi_E$ respects
+transposition. Furthermore, we observe that $\psi_E$ respects
 the charts we have defined above, in the sense that it maps bijectively 
@@
 \subsection{Differential computations}
-
-We compute in this subsection the differential of various operations on 
+\label{ssec:grassdiff}
+
+In this subsection, we compute the differential of various operations on 
 vector spaces. For brevity, we skip the estimation of the corresponding 
@@
 Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
-$n$ and $m$ respectively. Let $d$ be an integer in $[0,n]$. We are
-interested here in the function $\DI$ (for ``direct image'') defined 
-over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
-pair $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
-priori} vary, the map $\DI$ does not take its values in a well-defined 
-Grassmannian. To tackle this problem, we stratify $\mathcal M$ as
-follows: for each integer $r \in [0,d]$, we introduce the subset
-$\mathcal M_r \subset \mathcal M$ consisting of those pairs $(f,V)$
-for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
-locally closed in $\mathcal M$ and therefore are submanifolds. 
-Moreover, $\DI$ induces by restriction differentiable 
-functions between $K$-manifolds
-$\DI_r : \mathcal M_r \to \Grass(F,r)$.
+$n$ and $m$, respectively. Let $d$ be an integer in $[0,n]$. We are
+interested in the direct image function $\DI$ defined 
+on $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
+pair $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may vary,
+the map $\DI$ does not take its values in a well-defined 
+Grassmannian. We therefore stratify $\mathcal M$ as
+follows: for each integer $r \in [0,d]$, let $\mathcal M_r \subset \mathcal M$
+be the subset of pairs $(f,V)$ for which $f(V)$ has dimension $r$.
+The $\mathcal M_r$'s are locally closed in $\mathcal M$ and are therefore
+submanifolds. Moreover, $\DI$ induces differentiable 
+functions $\DI_r : \mathcal M_r \to \Grass(F,r)$.
 
@@
 {\bf Inverse images.}
-We now consider the mapping $\II$ (for ``inverse image'') sending a 
+We now consider the inverse image mapping $\II$ sending a 
 pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
@@
 differentiable functions. For each integer $r \in [0,n]$, we introduce 
-the sub-manifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
+the submanifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
 pairs $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
@@
 
-The cases of direct images and inverse images are related by duality
+Direct images and inverse images are related by duality
 as follows: if $f : E \to F$ is any linear map and $W$ is a subspace
-of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. Using this, one 
-can deduce the differential of $\DI_r$ from those of $\II_s$ and 
+of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. We 
+can thus deduce the differentials of $\DI_r$ from those of $\II_s$ and 
 \emph{vice versa}.
@@
 dimension of the sum: for each integer $d \in [0, d_1+d_2]$, we define 
-$\mathcal C_d$ as the sub-manifold of $\mathcal C$ consisting of those 
+$\mathcal C_d$ as the submanifold of $\mathcal C$ consisting of those 
 pairs $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get a 
@@
 \subsection{Implementation and experiments}
-
-\noindent
-{\bf Standard representation of vector spaces.}
-A standard way to represent subvector spaces of $K^n$ is to view them 
-through the charts $\mathcal U_{V_I, V_{I^\c}}$ (where $I$ is a subset 
+\label{grassimpl}
+
+\noindent
+{\bf Standard representation of vector spaces.}  %%%%%%%%% Repetitive
+One commonly represents subspaces of $K^n$  
+using the charts $\mathcal U_{V_I, V_{I^\c}}$ (where $I$ is a subset 
 of $\{1, \ldots, n\}$) introduced above. More concretely,
@@
 matrix. We recall that such a representation always exists and, when 
-the set of indices $I$ is fixed, there 
-exists at most one $G_V$ satisfying the above condition.
+the set of indices $I$ is fixed, at most one $G_V$ satisfies the above condition.
 Given a family of generators of $V$, one can compute $G_V$ and $I$ as 
-above by performing standard row reduction. If we are choosing always the 
-first non-vanishing pivot, we get this way the set $I$ with the smallest 
-possible elements. It is the usual choice to get a normalized 
-representation. However, in the context of inexact base fields, it is 
-much better to always choose the pivot with maximal norm.
+above by performing standard row reduction.  Choosing the first non-vanishing
+pivot at every stage provides a canonical choice for $I$, but in the context of
+inexact base fields, choosing the pivot with the maximal norm yields
+a more stable algorithm.
 
@@
 $H_V$ obtained by \emph{deleting} rows with index in $I$ is the identity 
-matrix. Similarly as above, if $V$ is presented as the vanishing of a 
-family of linear equations, we can compute $I$ and $H_V$ by performing 
+matrix. As above, we can then compute $I$ and $H_V$ by performing 
 column reduction.
 
-We emphasize that passing from one representation to the other is almost 
-free (and also quite stable). To simplify notation, let us suppose 
-that $I = \{1, \ldots, d\}$ with $d = \dim V$. Then, denoting by $I_d$ 
-the identity matrix of size $d$, the matrix $G_V$ has the shape
-$(\begin{matrix} I_d & G'_V \end{matrix})$
-and it is quite easy to check that one can represent $V$ with the same
-$I$ and the matrix
-$H_V = \Big(\begin{matrix} -G'_V \\ I_{n-d} \end{matrix}\Big)$. Of
-course, for a general $I$, a similar formula exists.
+Note that switching representations is cheap and stable.
+If $I = \{1, \ldots, d\}$ with $d = \dim V$ and $I_d$ 
+is the identity matrix of size $d$, the matrix $G_V$ has the form
+$(\begin{matrix} I_d & G'_V \end{matrix})$.
+One can represent $V$ with the same $I$ and the matrix
+$H_V = \Big(\begin{matrix} -G'_V \\ I_{n-d} \end{matrix}\Big)$.
+A similar formula exists for general $I$.
 
@@
 The first representation we gave is well suited for the computation
-of direct images and sums. For instance, in order to compute $f(V)$, 
-we just have to apply $f$ to each row of $G_V$ (getting this way a
-family of generators of $f(V)$) and then row reduce. Dually, the
+of direct images and sums. For instance, to compute $f(V)$
+we apply $f$ to each row of $G_V$, obtaining a
+family of generators of $f(V)$, and then row reduce. Dually, the
 second representation works well for computing inverse images, including 
@@
 
-With regard to precision, one can either track it in the standard way or 
-by using Proposition \ref{prop:precision} and the differential 
-computations we have presented before. Of course, the second method 
-is more costly but also more accurate (as illustrated above).
+%One can either track precision in the standard way or 
+%by using Proposition \ref{prop:precision}.  While the second method 
+%is more costly, it is also more accurate.
 
@@
 \noindent
-The expression ``high precision'' on line 3 means that the precision on 
+The expression \emph{high precision} on line 3 means that the precision on 
 $\alpha$, $\beta$, $\gamma$ and $\delta$ is set in such a way that it 
@@
 & \raisebox{0.04cm}{method} & 
-\hspace{0.2cm}projected\hspace{0.2cm} & 
-\hspace{0.2cm}diffused\hspace{0.2cm} \\
+\hspace{0.2cm}Projected\hspace{0.2cm} & 
+\hspace{0.2cm}Diffused\hspace{0.2cm} \\
 \hline 
@@
 always chosen sufficiently large so that it does not affect the
-behavior of precision). The ``coord-wise'' column corresponds to
+behavior of precision). The \emph{Coord-wise} column corresponds to
 the standard way of tracking precision. On the other hand, in the two 
-last columns, the precision is tracked using lattices. The term 
-``diffused'' has the same meaning as in \S 
-\ref{subsec:mulmatrix} when we were dealing with matrix multiplication: 
-it then corresponds to significant digits which are ``spread out'' among 
-the two coordinates of the precision lattice. In the present situation,
-it turns out that these ``diffused'' digits may be ``separated'' by
-working in another basis of the underlying space $\Q_2^3$.
+last columns, the precision is tracked using lattices. The \emph{Diffused}
+column gives the amount of diffused precision, factored to be comparible
+to the Coord-wise column. 
 The fact that only negative numbers appear in this column means that we 
-are actually always gaining precision with this model! Finally, in the 
-``projected'' column, we have written the loss of precision we get after 
-having projected the lattice precision data onto each coordinate (in 
-order to get an answer which is coordinate-wise).
+are actually always gaining precision with this model! Finally, the 
+\emph{Projected} column gives the precision loss after projecting the lattice
+precision onto coordinates.
 
@@
 Let $U$, $V$ and $W$ be three open subsets in $K$-Banach spaces 
-$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
+$E$, $F$ and $G$, respectively. We assume that $0 \in U$, $0 \in V$. Let 
 $f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
@@
 Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
-some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
+some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$, respectively. The 
 aim of this subsection is to prove the following intermediate result.
@@
 Given positive integers $k_1, \ldots, k_\ell$ whose sum is $m$ and 
-$x_1, \ldots, x_\ell \in E$, we have:
+$x_1, \ldots, x_\ell \in E$, we have
 $$\begin{array}{l}
@@
 \end{array}$$
-where, in LHS, the variable $x_i$ is repeated $k_i$ times.
+where, in the LHS, the variable $x_i$ is repeated $k_i$ times.
 \end{lem}
@@
 \begin{proof}
-It is enough to prove that:
+It is enough to prove that
 $$\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
@@
 induction on 
-$\ell$. If $\ell = 1$, this follows directly from
+$\ell$. The $\ell = 1$ case follows directly from
 the definition of $\Vert \psi \Vert$. 
 We now pick $(\ell+1)$ integers $k_1, \ldots, k_{\ell+1}$ whose sum 
-equals $m$ together with $(\ell+1)$ elements $x_1, \ldots, x_{\ell+1}$
+equals $m$, together with $(\ell+1)$ elements $x_1, \ldots, x_{\ell+1}$
 lying in the unit ball of $E$.
@@
 = x_\ell + \lambda x_{\ell+1}$ and $k'_\ell = k_\ell + k_{\ell+1}$. By the 
-induction hypothesis, we know that the inequality:
+induction hypothesis, we know that the inequality
 $$\Big\Vert \binom m {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
@@
 \leq \Vert \psi \Vert$$
-holds for all $\lambda \in K$. Furthermore, LHS of the above inequality
+holds for all $\lambda \in K$. Furthermore, the LHS of the inequality
 is a polynomial $P(\lambda)$ of degree $k'_\ell$ whose coefficient in 
-$\lambda^j$ is:
+$\lambda^j$ is
 $$\binom m {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
@@
 Since $\Vert P(\lambda) \Vert \leq \Vert \psi \Vert$ for all $\lambda$ in 
-the unit ball, the norm of all its coefficients must be at most $\Vert \psi
-\Vert$ as well. Looking at the coefficient of $\lambda^{k_l}$, we find
-the desired result.
+the unit ball, the norm of all its coefficients must also be at most $\Vert \psi
+\Vert$. From the coefficient of $\lambda^{k_l}$, the result follows.
 \end{proof}
@@
 
-We now go back to the setting of Proposition \ref{prop:boundLambdaf}.
+We return to the setting of Proposition \ref{prop:boundLambdaf}.
 Let $f = \sum_{n \geq 0} f_n$, $g = \sum_{n \geq 0} g_n$ and
@@
 Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
-some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
+some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$, respectively.
 We recall that $\Lambda(f)$ 
 is the Legendre transform of the Newton polygon $\NP(f)$ defined in Section \ref{sec:matrices}
-\cite[Proposition 3.9]{caruso-roe-vaccon:14a} and that $\alpha$ is a nonnegative
-real number such that $\Vert n! \Vert \geq e^{-\alpha n}$ for all positive integer $n$.
+\cite[Proposition 3.9]{caruso-roe-vaccon:14a}, and that $\alpha$ is a
+real number such that $\Vert n! \Vert \geq e^{-\alpha n}$ for all positive integers $n$.
 
@@
 \label{lem:boundLambdaf}
-We keep the above notations. 
+We keep the above notation. 
 If $(a,b)$ satisfies
@@
 We are now ready to prove $u_r \leq e^{-ar + b}$ by induction on $r$.
-When $r = 0$, it is obvious because $u_0$ vanishes. Otherwise, we 
-can write:
+When $r = 0$, it is obvious because $u_0$ vanishes. Otherwise, the
+induction follows from
 \begin{align*}
@@
     e^{ -\max(b,d)\cdot m + b - a + \sum_{i=1}^m (-a n_i + \max(b,d))} \\
-& = e^{ b - a - a r } = e^{ -a (r+1) + b}
+& = e^{ b - a - a r } = e^{ -a (r+1) + b}.
 \end{align*}
-and the induction goes. Going back to the definition of $u_r$, we obtain 
+From the definition of $u_r$, we obtain 
 $\Vert f_r \Vert \leq u_r \cdot \Vert r! \Vert^{-1} \leq e^{-(a - \alpha) 
-r + b}$, which means $b \geq \Lambda(f)(a - \alpha)$.
+r + b}$. Thus $b \geq \Lambda(f)(a - \alpha)$.
 \end{proof}
@@
 $\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
-\Lambda_h(a)$, from what we derive 
+\Lambda_h(a)$, from which we derive 
 $a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) = b$.

### 2015-01-19 examples/stability.tex
Typo

@@
 \cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
-lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose 
+lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting of polynomials whose 
 Newton polygons lie above $P$.  This lattice is generated by monomials 

### 2015-01-19 examples/stability.tex
Small corrections

--- 2015-01-19 examples/stability.tex
@@
   \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
+\alignauthor David Roe \\
+  \affaddr{University of British Columbia}\\
+  \email{\normalsize \textsf{roed.math@gmail.com}}
 \alignauthor Tristan Vaccon\\
@@
   \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
-\alignauthor David Roe \\
-  \affaddr{University of British Columbia}\\
-  \email{\normalsize \textsf{roed.math@gmail.com}}
 }
@@
 Throughout the paper, the letter $K$ will refer to a complete discrete 
-valuation field of characteristic $0$. Usual examples are finite 
-extensions of $\Qp$ and Laurent series fields over a field of 
-characteristic $0$. We denote by $\val : K \to \Z \cup \{+\infty\}$ of 
-$K$ --- it is the $p$-adic valuation when $K = \Qp$ and the usual 
-valuation of a Laurent series when $K = k((t))$ --- and by $\Vert \cdot 
-\Vert$ the corresponding norm.
+valuation field. Usual examples are finite extensions of $\Qp$ and 
+Laurent series fields over a field. We denote by $\val : K \to \Z \cup 
+\{+\infty\}$ the valuation on $K$ normalized by $\val(K^\star) = \Z$, by 
+$\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
+$1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.
 
@@
 functions with coefficients in $\O_K$ in any (equivalently all) system 
-of coordinates associated to orthonormal bases.
+of coordinates associated to a $\OK$-basis of $B_E(1)$.
 
@@
 Keeping the above notation, we have:
-$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
+$$\Lambda(f)_{\geq 2} (x) \leq 2(x + \alpha + \mu) -
 \min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
@@
 differential equation of the form $f' = g \circ f$. If this holds, 
-Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
+Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\geq 
 2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.
@@
 $M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
-which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
+which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
 Proposition \ref{prop:precision}, the precision polygon determines the minimal 
@@
 \noindent $\bullet$
-$L_n$ (resp. $L_n^u$) is the sub-affine space of 
+$L_n^0$ (resp. $L_n^u$) is the sub-affine space of 
 nilpotent (resp. unipotent) lower-triangular matrices.
@@
 by:
-$$\begin{array}{l}
-dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U) \smallskip \\
-\hspace{1cm}\text{with } dX = L^{-1} \cdot dM \cdot U^{-1}
-\end{array}$$
+$$
+%\begin{array}{l}
+dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U)
+%\smallskip \\ \hspace{1cm}
+%\quad 
+\text{ with } dX = L^{-1} \cdot dM \cdot U^{-1}
+%\end{array}
+$$
 where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
-onto $L_n$ (resp. $U_n$). 
+onto $L_n^0$ (resp. $U_n$). 
 It is easily checked that $D'(M)$ is bijective with inverse given by
@@
 
-We now want to apply Proposition \ref{prop:boundLambdaf2}
-in order to derive a concrete result on precision. We pick $M_0 \in
+We now want to apply Proposition \ref{prop:boundLambdaf2} in order to 
+derive a concrete result on precision. \emph{We then assume that $K$ has 
+characteristic $0$.} We pick $M_0 \in
 O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the translated
@@
 Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
-$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
+$M_n(K)$. It can be naturally seen as a basis of $L_n^0 \times U_n$ as
 well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
@@
 $\val(\det(D'(M)))$, which is the lattice-method loss of precision. 
+%With the language of Definition \ref{def:diffused},
+The number of diffused digits of precision of $D'(M)(M_{n,n}(\OK))$ is 
+then the difference between these two numbers.
 Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
@@
 tangent space at a given point $V$ as the quotient $\Hom(E, E/V) / 
-\End(E/V) \cong \Hom(V, E/V)$.
+\End(E/V) \simeq \Hom(V, E/V)$.
 Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
@@
 growing functions (but this can be done using Proposition 
-\ref{prop:boundLambdaf2} as before).
+\ref{prop:boundLambdaf2} as before if $\text{char}(K) = 0$).
 

### 2015-03-12 examples/stability.tex
add keywords
Slides of my talk at ALEA 2015

@@
  \noindent
- {\bf Keywords:} $p$-adic precision
+ {\bf Keywords:} $p$-adic precision, linear algebra, ultrametric analysis
 \medskip
@@
 \]
-Set $k = \log_p([H_0 : H])$.  Then we say that $H$ has $k$
+Set $k = \text{length}(H_0/H)$.  Then we say that $H$ has $k$
 \emph{diffused digits of precision}.
@@
 and therefore the first summand in $M$ varies in the subspace of 
-$M_{a,c}(\O_K)$ consisting of matrices whose entries on the $i$-th
+$M_{r,t}(\O_K)$ consisting of matrices whose entries on the $i$-th
 row have valuation at least $a_i$. Arguing similarly for the second
@@
 & \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
-$2$ & $\phantom{2}{-}3.0$& $5$ & $-1.5$ & $1.4$ \\% \hline
-$3$& $\phantom{2}{-}9.4$& $11$ & $-2.3$ & $2.3$\\% \hline
-$4$ & $-20\phantom{.4}$& $20$ & $-3.8$ & $3.1$\\ \hline
+$2$ & $\phantom{2}3.0$& $5$ & $1.5$ & $1.4$ \\% \hline
+$3$& $\phantom{2}9.4$& $11$ & $2.3$ & $2.3$\\% \hline
+$4$ & $20\phantom{.4}$& $20$ & $3.8$ & $3.1$\\ \hline
 \end{tabular} 
@@
 one can use the framework of \cite[Appendix A]{caruso-roe-vaccon:14a} to 
-study the $p$-adic in this context.
+study $p$-adic precision in this context.
 
@@
 the unit ball, the norm of all its coefficients must also be at most $\Vert \psi
-\Vert$. From the coefficient of $\lambda^{k_l}$, the result follows.
+\Vert$. From the coefficient of $\lambda^{k_\ell}$, the result follows.
 \end{proof}

### 2015-03-21 examples/stability.tex
Typos spotted by reviewers

@@
 to track the precision of vectors and points on $p$-adic manifolds.
-The main result of \emph{loc. cit.}, recalled here as Proposition \ref{prop:precision},
+The main result of \emph{loc.~cit.}, recalled here as Proposition \ref{prop:precision},
 allows for the propagation of precision using differentials.  While
@@
 The present paper, which may be considered as a continuation of 
-\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. 
-cit.} in order to analyze the $p$-adic stability of many standard 
+\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc.~cit.} in order to analyze the $p$-adic stability of many standard 
 algorithms for computing with basic algebraic structures: numbers, 
@@
 -}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
-should be thought as special neighborhoods of $0$ and therefore are good 
+should be thought of as special neighborhoods of $0$ and therefore are good 
 candidates to model precision data. Moreover, as revealed in 
@@
 In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
-to manifolds over $K$ and covers in this way a very large familiy of
+to manifolds over $K$ and covers in this way a very large family of
 situations (\emph{e.g.} computation with vector spaces as discussed
@@
 In what follows, we always assume that $V$ and $W$ contain the origin, 
-$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
+$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. These assumptions are harmless 
 for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
@@
 there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
-$({-}\infty, \nu]$\footnote{We note that this assumption is fullfilled if 
+$({-}\infty, \nu]$\footnote{We note that this assumption is fulfilled if 
 we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
@@
 that the Taylor expansion of $f(z)$ starts with the term
-$g(0) z$. Hence, on a neighborhood on ${-}\infty$, we have 
+$g(0) z$. Hence, on a neighborhood of ${-}\infty$, we have 
 $\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
@@
 the dashed blue line defined as the line of slope $2$ passing through
-the first break point of the blue plain line, which has coordinate 
+the first break point of the blue plain line, which has coordinates
 $(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
@@
 - vy \big).$$
-It is highest convex function with $\varphi_{\geq v} \leq \varphi$ and 
+It is the highest convex function with $\varphi_{\geq v} \leq \varphi$ and 
 $\varphi'_{\geq v} \geq v$.
@@
 function $h$ by endowing $F$ with the second norm $\Vert x \Vert'_F = 
-e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to taking $h : (F, \Vert 
+e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and taking $h : (F, \Vert 
 \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ to be the identity on the
@@
 According to Proposition \ref{prop:precision}, the behavior of the precision when 
-computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$, the linear mapping that takes a pair 
+computing $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$, the linear mapping that takes a pair 
 $(dA,dB)$ to $A \cdot dB + dA \cdot B$.
@@
 note that, on the other side, the valuation of the entries of $AB$ may 
-increase, meaning that we are also loosing some significant digits if we 
+increase, meaning that we are also losing some significant digits if we 
 are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
@@
 grow linearly with respect to the number of multiplications we are 
-performing (that is $n$) whereas, in the second case, the growing seems 
+performing (that is $n$) whereas, in the second case, the growth seems 
 to be only logarithmic. It would be nice to have a precise formulation 
@@
 $\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower 
-convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$. For 
+convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_j(M))$. For 
 any matrix $M$, the Newton polygon lies above the Hodge polygon 
@@
 The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of 
-the Newton polygons of the entries of $\com(X{-}M)$.
+the Newton polygon of the entries of $\com(X{-}M)$.
 \end{deftn}
@@
 \charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
-$M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
+$M_{n,n}(\OK)$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
 which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
@@
 minors do not vanish (we recall that the latest condition implies the
-existence and the unicity of a LU factorization);
+existence and the uniqueness of a LU factorization);
 
@@
 Grassmannian of $d$-dimensional subspaces of $E$. It is well-known
-that $\Grass(E,d)$ has a natural structure of $K$-manifold. The aim of
+that $\Grass(E,d)$ has the natural structure of a $K$-manifold. The aim of
 this subsection is to recall standard facts about its geometry. In what follows,
@@
 When the pair $(V, V^\c)$ varies, the open subsets $\mathcal 
-U_{V,V^\c}$ cover the whole Grassmaniann and define an atlas.
+U_{V,V^\c}$ cover the whole Grassmannian and define an atlas.
 When implementing vector spaces on a computer, we usually restrict 
@@
 {\bf Duality.}
-If $E$ is a finite dimension vector space over $K$, we use the notation 
+If $E$ is a finite dimensional vector space over $K$, we use the notation 
 $E^\star$ for its dual (\emph{i.e.} $E^\star = \Hom(E,K)$). If we are also given
@@
 pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
-$f^{-1}(W)$. As before, this map does not take its value in a single 
+$f^{-1}(W)$. As before, this map does not take its values in a single 
 Grassmannian and we need to stratify $\mathcal W$ in order to get 
 differentiable functions. For each integer $r \in [0,n]$, we introduce 
-the submanifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
+the submanifold $\mathcal W_s$ of $\mathcal W$ consisting of those 
 pairs $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
@@
   \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert$$
-for all nonnegative integer $r$, where the supremum is taken over all pairs $(m, (n_i))$ where $m$
+for all nonnegative integers $r$, where the supremum is taken over all pairs $(m, (n_i))$ where $m$
 is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
@@
   \prod_{i=1}^m \max(\Vert f_{n_i} \Vert, \Vert h_{n_i} \Vert)$$
-for all nonnegative integer $r$,
+for all nonnegative integers $r$,
 where the supremum runs over all pairs $(m, (n_i))$ where $m$

### 2015-04-07 examples/stability.tex
Removing a comment

--- 2015-01-19 examples/stability.tex
@@
-%% from this description could be cut for space
-
 \documentclass{sig-alternate}
@@
   \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
+\alignauthor Tristan Vaccon\\
+  \affaddr{Universit\'e Rennes 1}\\
+  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
 \alignauthor David Roe \\
@@
   \email{\normalsize \textsf{roed.math@gmail.com}}
-\alignauthor Tristan Vaccon\\
-  \affaddr{Universit\'e Rennes 1}\\
-  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
 }
@@
 Throughout the paper, the letter $K$ will refer to a complete discrete 
-valuation field. Usual examples are finite extensions of $\Qp$ and 
-Laurent series fields over a field. We denote by $\val : K \to \Z \cup 
-\{+\infty\}$ the valuation on $K$ normalized by $\val(K^\star) = \Z$, by 
-$\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
-$1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.
+valuation field of characteristic $0$. Usual examples are finite 
+extensions of $\Qp$ and Laurent series fields over a field of 
+characteristic $0$. We denote by $\val : K \to \Z \cup \{+\infty\}$ of 
+$K$ --- it is the $p$-adic valuation when $K = \Qp$ and the usual 
+valuation of a Laurent series when $K = k((t))$ --- and by $\Vert \cdot 
+\Vert$ the corresponding norm.
 
@@
 functions with coefficients in $\O_K$ in any (equivalently all) system 
-of coordinates associated to a $\OK$-basis of $B_E(1)$.
+of coordinates associated to orthonormal bases.
 
@@
 Keeping the above notation, we have:
-$$\Lambda(f)_{\geq 2} (x) \leq 2(x + \alpha + \mu) -
+$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
 \min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
@@
 differential equation of the form $f' = g \circ f$. If this holds, 
-Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\geq 
+Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
 2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.
@@
 $M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
-which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
+which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
 Proposition \ref{prop:precision}, the precision polygon determines the minimal 
@@
 \noindent $\bullet$
-$L_n^0$ (resp. $L_n^u$) is the sub-affine space of 
+$L_n$ (resp. $L_n^u$) is the sub-affine space of 
 nilpotent (resp. unipotent) lower-triangular matrices.
@@
 by:
-$$
-%\begin{array}{l}
-dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U)
-%\smallskip \\ \hspace{1cm}
-%\quad 
-\text{ with } dX = L^{-1} \cdot dM \cdot U^{-1}
-%\end{array}
-$$
+$$\begin{array}{l}
+dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U) \smallskip \\
+\hspace{1cm}\text{with } dX = L^{-1} \cdot dM \cdot U^{-1}
+\end{array}$$
 where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
-onto $L_n^0$ (resp. $U_n$). 
+onto $L_n$ (resp. $U_n$). 
 It is easily checked that $D'(M)$ is bijective with inverse given by
@@
 
-We now want to apply Proposition \ref{prop:boundLambdaf2} in order to 
-derive a concrete result on precision. \emph{We then assume that $K$ has 
-characteristic $0$.} We pick $M_0 \in
+We now want to apply Proposition \ref{prop:boundLambdaf2}
+in order to derive a concrete result on precision. We pick $M_0 \in
 O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the translated
@@
 Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
-$M_n(K)$. It can be naturally seen as a basis of $L_n^0 \times U_n$ as
+$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
 well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
@@
 $\val(\det(D'(M)))$, which is the lattice-method loss of precision.
-%With the language of Definition \ref{def:diffused},
-The number of diffused digits of precision of $D'(M)(M_{n,n}(\OK))$ is 
-then the difference between these two numbers.
 Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
@@
 tangent space at a given point $V$ as the quotient $\Hom(E, E/V) / 
-\End(E/V) \simeq \Hom(V, E/V)$.
+\End(E/V) \cong \Hom(V, E/V)$.
 Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
@@
 growing functions (but this can be done using Proposition 
-\ref{prop:boundLambdaf2} as before if $\text{char}(K) = 0$).
+\ref{prop:boundLambdaf2} as before).
 

### 2015-04-07 resultants/resultants.tex

@@
 rings (\emph{e.g.} $\Zp$ or $k[[t]]$ where $k$ is a field). We prove that 
-Euclide-like algorithms are highly unstable in average and we explain, in
+Euclide-like algorithms are highly unstable on average and we explain, in
 many cases, how one can stabilize them without sacrifying the complexity.
@@
  \noindent
- {\bf Keywords:} $p$-adic precision, subresultants
+ {\bf Keywords:} $p$-adic precision, subresultants, random $p$-adic 
+polynomials
 \medskip
@@
 $161.1$ significant digits by coefficient whereas a stable 
-algorithm should only loose $3.2$ digits in average. This 
+algorithm should only loose $3.2$ digits on average. This 
 ``theoretical'' loss is computed as the double of the valuation of the 
@@
 words, on a computer, we do not work with elements of $\A$ but with
-balls in $\A$ which are much more tractable.
+balls in $\A$ which are more tractable.
 
@@
 coefficients in $\ring$ of degree at most $j$ (resp. $d-j-1$, resp. 
-$d-j-1$) whose coefficients can be expressed, up to a sign, as a minor 
+$d-j-1$) whose coefficients can be expressed, up to a sign, as minors
 of the Sylvester matrix. These polynomials satisfy in addition the
@@
 $(S_i)$ then stops at $i = d_B$. Moreover, the $\delta_i$'s and the 
-$c_i$'s are no longer necessary since we have $\delta_i = 1$ and $c_i
+$c_i$'s are now all ``trivial'' since we have $\delta_i = 1$ and $c_i
 = s_i$ for all $i$. The recurrence formula then becomes:
@@
 To conclude with, we remark that Proposition \ref{prop:prem} simply 
-states that $R_j = S_{d_B-j}$.
+states that $R_j = S_{d_B-j}$ in this case.
 
@@
 means that a polynomial $P(X)$ is represented as:
-$$P(X) = \sum_{i=1}^n a_i X^i + O(p^N)
-\quad \text{with } a_i \in \Qp \text{ and } N \in \Z.$$
+$$P(X) = \sum_{i=1}^n a_i X^i + O(\pi^N)
+\quad \text{with } a_i \in K \text{ and } N \in \Z.$$
 In other words, we assume that the software we are using does not
@@
 subresultant pseudo-remainder sequence computes $R_j$ at precision 
-$O(p^{N_j})$ with:
+$O(\pi^{N_j})$ with:
 $$N_j \leq N + V_{j+1} - 2 \cdot (\delta_{j+1} + \delta_{j+2} + \cdots
@@
 lower bounded by $M - \delta_j \leq N_j - 2 \delta_j + V_{j+1} - V_j$.
-We derive for this $N_{j-1} \leq N_j - 2 \delta_j - V_{j+1} + V_j$ and
+From this, we derive $N_{j-1} \leq N_j - 2 \delta_j - V_{j+1} + V_j$ and
 the proposition finally follows by summing up these inequalities.
@@
 $V_1$ and all $\delta_j$'s are 
-approximatively equal to $\frac 1 {p-1}$ in average. The loss of 
+approximatively equal to $\frac 1 {p-1}$ on average. The loss of 
 precision then grows linearly with respect to $d$. This confirms the 
@@
 We endow $\A$ with its Haar measure. The set $\Omega$ of couples of 
-monic polynomial of degree $d$ with coefficients in $A$ is canonically 
+monic polynomial of degree $d$ with coefficients in $\A$ is canonically 
 in bijection with $\A^{2d}$ and hence inherits the product measure. 
@@
 stands for the integer part function. In particular, we observe that 
-the couple $(2,1)$ does not occur almost surely although the events 
+$(V_{d-2}, V_{d-1}) \neq (2,1)$ almost surely although the events 
 $\{V_{d-1} = 2\}$ and $\{V_{d-2} = 1\}$ both occur with positive 
@@
 \sigma(d-j)$ sorted in increasing order. The random variable $Z_i$ 
-is then the minimum of $\tau_j(i)$ independant random variables with 
+is then the minimum of $\tau(i)$ independant random variables with 
 geometric distribution of parameter $(1 - q^{-1})$ and thus its 
@@
 During the proof, $A$ and $B$ will always refer to monic polynomials of 
-degree $d$ and $R_j$ (resp. $U_j$ and $V_j$) denote their $j$-th 
+degree $d$ and $R_j$ (resp. $U_j$ and $V_j$) to their $j$-th 
 subresultant (resp. their $j$-th cofactors). If $P$ is a polynomial and 
@@
 = \A/\pi^n\A$ shows that $(\psi_j \text{ mod } \pi^n)$ is a bijection 
-for all $n$. This proves the corollary.
+for all $n$. This proves the Corollary.
 \end{proof}
@@
 additional notations. First, we fix $J$ and write $J = \{n_1, \ldots, 
-n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$. we also set $n_{\ell+1} = 
+n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$. We also set $n_{\ell+1} = 
 0$ by convention. Then, given $n$ and $m$ two integers with $m < n$, let 
@@
 knowledge of $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$ (for some $i$) 
-are enough to reconstruct the quotient of the Euclidean division of 
+is enough to reconstruct the quotient of the Euclidean division of 
 $S_i(A,B)$ by $S_{i-1}(A,B)$. Thus, one can recover $S_i(A,B)$ from 
@@
 valuation for $j'$ varying in the open range $(j-v, j+v)$ (with the
-convention that $r_{j'} = 0$ whenever $j < 0$). It is clear from the 
+convention that $r_{j'} = 0$ whenever $j' < 0$). It is clear from the 
 definition that $r_{j-v}$ or $r_{j+v}$ (with $v = V_j^{(0)}$) has 
@@
 \end{array}$$
-We notice that $B^{(1)}$ has actually all its coefficients in $\A$ 
+We notice that all the coefficients of $B^{(1)}$ lie in $\A$ 
 because $r_{j'}$ has positive valuation for $j' \in (j-v, j+v)$.
@@
 can now define a new random variable $V_j^{(1)}$ and, continuing this 
-way, we construct an infinite sequence $V_j^{(m)}$ such that:
+way, we construct an infinite sequence $V_j^{(m)}$ such that
 $V_j = \sum_{m \geq 0} V_j^{(m)}$.
@@
 distribution than $\sum_{i=1}^{V_j^{(0)} - 1} \min(X_{j-i}^{(1)}, 
-\ldots, X_{j+i}^{(1)})$, which can be rewritten as:
+\ldots, X_{j+i}^{(1)})$, which can be rewritten as
 $Y_j^{(1)} = \sum_{i=1}^d \min(X_{j-i}^{(0)}, X_{j-i}^{(1)}, \ldots, 
-X_{j+i}^{(0)}, X_{j+i}^{(1)})$
-and the equidistribution of $(A^{(1)}, B^{(1)})$ shows that the 
+X_{j+i}^{(0)}, X_{j+i}^{(1)})$. More precisely,
+the equidistribution of $(A^{(1)}, B^{(1)})$ shows that the 
 joint distribution $(V_j^{(0)}, V_j^{(1)})$ is the same as those
@@
 replacing eventually $N$ by $3 \cdot \log_d p$ (which does not affect 
-the complexity), the assumption~(b) is harmless in average --- but maybe
+the complexity), the assumption~(b) is harmless on average --- but maybe
 not on particularly bad instances. We finally underline that, if we are
@@
 $O(\pi^{N-N_{\text{int}}})$ where $N_{\text{int}}$ grows linearly 
-with respect to $d$ in average. On the other hand, if we are using 
+with respect to $d$ on average. On the other hand, if we are using 
 ultrametric floating point arithmetic, then the same algorithm will
 output $\Res(A,B)$ at precision $O(\pi^{N-N_{\text{float}}})$ where 
-$N_{\text{float}}$ grows linearly with respect to $\log d$ in average.
+$N_{\text{float}}$ grows linearly with respect to $\log d$ on average.
 We emphasize furthermore that this result is \emph{proved}! From this

### 2015-04-12 examples/stability.tex
Notes to David on where to insert additional info requested by referee

--- 2015-03-21 examples/stability.tex
@@
 The aim of this section is to briefly summarize the content of 
-\cite{caruso-roe-vaccon:14a} and fill in certain details.
+\cite{caruso-roe-vaccon:14a} and fill in certain details. % Add usual way to track precision.
 
@@
 is known up to $f'(v_0)(H)$ and this precision is optimal. Optimality is 
-reflected by the equality sign in Eq.~\eqref{eq:firstorder}.
+reflected by the equality sign in Eq.~\eqref{eq:firstorder}. % define coordinate-wise and lattice-wise here.
 
@@
 \emph{diffused digits of precision}.
-\end{deftn}
+\end{deftn} % Add explanation here - smallest lattice containing H here.
 If $H$ represents the actual precision of some object, then

### 2015-04-20 examples/stability.tex
Updates for final version

--- 2015-03-21 examples/stability.tex
@@
 \DeclareMathOperator{\round}{round}
+\DeclareMathOperator{\rank}{rank}
 
@@
 
+\conferenceinfo{ISSAC'15,}{July 6--9, 2015, Bath, United Kingdom.}
+\copyrightetc{Copyright \copyright~2015 ACM \the\acmcopyr}
+\crdata{978-1-4503-3435-8/15/07}
+
 \begin{document}
@@
 
-\vspace{1mm}
- \noindent
- {\bf Categories and Subject Descriptors:} \\
-\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
-  Manipulation -- \emph{Algebraic Algorithms}
-
- \vspace{1mm}
- \noindent
- {\bf General Terms:} Algorithms, Theory
-
- \vspace{1mm}
- \noindent
- {\bf Keywords:} $p$-adic precision, linear algebra, ultrametric analysis
-\medskip
+\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulation -- \emph{Algebraic Algorithms}}
+\terms{Algorithms, Theory}
+\keywords{$p$-adic precision, linear algebra, ultrametric analysis}
+
+%\vspace{1mm}
+% \noindent
+% {\bf Categories and Subject Descriptors:} \\
+%\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
+%  Manipulation -- \emph{Algebraic Algorithms}
+%
+% \vspace{1mm}
+% \noindent
+% {\bf General Terms:} Algorithms, Theory
+%
+% \vspace{1mm}
+% \noindent
+% {\bf Keywords:} $p$-adic precision, linear algebra, ultrametric analysis
+%\medskip
 
@@
 to track the precision of vectors and points on $p$-adic manifolds.
-The main result of \emph{loc.~cit.}, recalled here as Proposition \ref{prop:precision},
+The main result of \emph{loc.~cit.}, recalled in Proposition \ref{prop:precision},
 allows for the propagation of precision using differentials.  While
-allowing arbitrary lattices in representing precision carries a high
+representing precision by lattices carries a high
 cost in space and time requirements, the reduced precision loss
-can sometimes make up for these costs as part of a larger algorithm.
-In this paper we apply the ideas of \cite{caruso-roe-vaccon:14a} to
-a variety of linear algebraic tasks.
+can sometimes overwhelm these costs within a larger algorithm.
+In this paper, we apply the ideas of \cite{caruso-roe-vaccon:14a} to
+certain linear algebraic tasks.
 
@@
 gives a lower bound on the precision of the characteristic polynomial.
-The precision polygon lies above the Hodge polygon of the matrix, and
+This polygon lies above the Hodge polygon of the matrix;
 we give statistics on the difference and the amount of diffuse precision.
-Finally, we find the precision of LU factorization, determine the conditions under which
-Proposition \ref{prop:precision} may be applied, and again describe experiments
-with lattice-based precision analysis.
+Finally, we apply Proposition \ref{prop:precision} to LU factorization and
+describe experiments with lattice-based precision analysis.
 
@@
 {\bf Notation.}
-Throughout the paper, the letter $K$ will refer to a complete discrete 
+Throughout the paper, $K$ will refer to a complete discrete 
 valuation field. Usual examples are finite extensions of $\Qp$ and 
-Laurent series fields over a field. We denote by $\val : K \to \Z \cup 
-\{+\infty\}$ the valuation on $K$ normalized by $\val(K^\star) = \Z$, by 
+Laurent series fields over a field. We denote by $\val : K \twoheadrightarrow \Z \cup 
+\{+\infty\}$ the valuation on $K$, by 
 $\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
@@
 
-Roughly speaking, the main proposal of \cite{caruso-roe-vaccon:14a} is 
-to use lattices to track precision when computing with elements lying in 
-vector spaces over $K$.
-
-More explicitly, let us consider a finite 
+In \cite{caruso-roe-vaccon:14a}, we suggest the 
+use of lattices to represent the precision of elements in 
+$K$-vector spaces.  We shall contrast this approach with the
+\emph{coordinate-wise method} used in Sage and Magma,
+where the precision of an element is specified by giving the precision
+of each coordinate separately.
+
+Consider a finite 
 dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
 actually those of Banach spaces. However, we will not need infinite 
-dimensional spaces in the present paper.} normed vector space $E$ 
+dimensional spaces here.} normed vector space $E$ 
 defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
 on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
-closed) ball of radius $r$ centered at the origin. A lattice $L \subset 
-E$ is, by definition, a sub-$\O_K$-module which generates $E$ over $K$. 
+closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
+E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
 Since we are working in a ultrametric world, the balls $B^{\phantom 
@@
 
-Proposition \ref{prop:precision} can be rephrased in terms of precision 
-as follows: if the input $v_0$ is known at precision $H$, then $f(v_0)$ 
-is known up to $f'(v_0)(H)$ and this precision is optimal. Optimality is 
-reflected by the equality sign in Eq.~\eqref{eq:firstorder}.
-
-In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
+This proposition enables the \emph{lattice method} of tracking precision,
+where the precision of the input is specified as a lattice $H$ and precision
+is tracked via differentials of the steps within a given algorithm.
+The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
+yields the optimum possible precision.
+
+In \cite{caruso-roe-vaccon:14a}, we also explained that if 
 $f$ is locally analytic, then the constant $\delta$ appearing in
-Proposition \ref{prop:precision} can be made explicit in terms of the
-growing function $\Lambda(f)$ of $f$ defined by:
-$$\textstyle \Lambda(f)(v) = 
+Proposition \ref{prop:precision} can be expressed in terms of the
+growing function $\Lambda(f)$ defined by
+\[
+\textstyle \Lambda(f)(v) = 
 \log \big( \sup_{h \in B^-_E(e^v)} \Vert 
-f(h) \Vert \big)$$
+f(h) \Vert \big)
+\]
 with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
@@
 In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
-to manifolds over $K$ and covers in this way a very large family of
-situations (\emph{e.g.} computation with vector spaces as discussed
-in \S \ref{sec:vectorspaces}). In this extended framework, the precision 
+to manifolds over $K$, where the precision 
 datum at some point $x$ is a lattice in the tangent space at $x$. 
 Propositions \ref{prop:precision} and \ref{prop:precision2} have
-analogues as well, simply obtained by working in charts.
-
-In this paper, we will compare the use of lattices to track precision
-with the standard coordinate-wise methods.  The following definition
-will be useful in discussing the differences.  Suppose $E$ is equipped
-with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.
+analogues obtained by working in charts.  We use this extension
+to compute with vector spaces in \S \ref{sec:vectorspaces}.
+
+We will use the following definition in contrasting lattice and coordinate-wise methods. 
+Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.
 
@@
 If $H$ represents the actual precision of some object, then
-$H_0$ is the best coordinate-wise approximation to $H$.
-The diffused precision is not visible coordinate-wise in the
-specified basis of $E$.
+$H_0$ is the smallest diagonal lattice containing $H$.  Since
+coordinate-wise methods cannot yield a precision better than $H_0$,
+$k$ provides a lower bound on the number of $p$-adic digits
+gained by lattice methods over standard methods.
 
@@
 defined by:
-$$\begin{array}{rll}
-&\tau_\nu(x) = x & \text{if } x \leq \nu \\
-& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
-\text{and} &
-\multicolumn{2}{l}{\Lambda_f(x) = 
-  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
-\end{array}$$
+\begin{align*}
+\tau_\nu(x) &= \smash{\left\{\begin{array}{ll} x & \mbox{ if } x \le \nu, \\
++\infty & \mbox{ otherwise;} \end{array}\right.} \\[8pt]
+\mbox{and } \Lambda_f(x) &= \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha),
+\end{align*}
+%$$\begin{array}{rll}
+%&\tau_\nu(x) = x & \text{if } x \leq \nu \\
+%& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
+%\text{and} &
+%\multicolumn{2}{l}{\Lambda_f(x) = 
+%  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
+%\end{array}$$
 where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
@@
 
+Note that multiplication of many random matrices plays a central
+role in the theory of random walks on homogeneous spaces \cite{Benoist:2012}.
+Better stability in computing such products helps estimate
+Lyapunov exponents in that context.
+
 \subsection{Determinant}
@@
 \]
-where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
-invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
-\ref{prop:precision}. Therefore, we suppose that $M$ is invertible for the rest of this section.
+where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $\rank(M) < n - 1$,
+then $\detp(M)$ is not surjective and we cannot apply Proposition 
+\ref{prop:precision}. Therefore, we suppose that $\rank(M) \ge n-1$ for the rest of this section.
 
@@
 \val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
-$\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower 
+$\NP(\charpoly(M))$.  The \emph{Hodge polygon} $\HP(M)$ of $M$ is the lower 
 convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_j(M))$. For 
-any matrix $M$, the Newton polygon lies above the Hodge polygon 
+any matrix $M$, the polygon $\NP(M)$ lies above $\HP(M)$ 
 \cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.
@@
 $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
-One uses these polygons to do a certain kind of coordinate-wise 
-precision tracking for polynomials, bounding the actual precision 
-lattice. We now introduce another polygon, bounded 
-between these two, that will provide an estimate on the precision of $\charpoly(M)$.
+These polygons are used in a coordinate-wise precision tracking for polynomial
+arithmetic. We now introduce another polygon, bounded 
+between $\NP(M)$ and $\HP(M)$, that will provide an estimate on the precision of $\charpoly(M)$.
 
@@
 \charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
-$M_{n,n}(\OK)$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
+$M_{n,n}(\OK)$. More precisely, $\PP(M)$ is the smallest polygon $P$ for 
 which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
@@
 
-It remains to prove the comparison with Newton polygon.
+It remains to prove the comparison with the Newton polygon.
 Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$-th entry of $M$, 
@@
 
-In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
+For randomly chosen matrices, approximating $\mathcal{E}$ 
 using the Hodge polygon loses only a small amount of precision. However,
-if the $\sigma_i(M)$'s are known to be large or if $M$ is a translate of
+if the $\sigma_i(M)$'s are large or if $M$ is a translate of
 such a matrix, using lattice precision can be very useful.
@@
 and, given a positive real number $C$, we let $B(C)$ be the closed ball 
-in $M_n(K)$ centered at the origin of radius $C$. We consider in addition
+in $M_n(K)$ centered at the origin of radius $C$. We consider
 the following subsets of $M_n(K)$:
@@
 \val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
-$\val(\det(D'(M)))$, which is the lattice-method loss of precision. 
+$\val(\det(D'(M)))$, which is precision lost in the lattice method. 
 %With the language of Definition \ref{def:diffused},
@@
 pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
-$f^{-1}(W)$. As before, this map does not take its values in a single 
-Grassmannian and we need to stratify $\mathcal W$ in order to get 
-differentiable functions. For each integer $r \in [0,n]$, we introduce 
+$f^{-1}(W)$. As before, this map does not take values in a single 
+Grassmannian, so we need to stratify $\mathcal W$ in order to get 
+differentiable functions. For each integer $s \in [0,n]$, we introduce 
 the submanifold $\mathcal W_s$ of $\mathcal W$ consisting of those 
@@
 induces a continuous function
-$\II_r : \mathcal W_r \to \Grass(E,s)$.
-Pick $(f,W) \in \mathcal W_r$. Set $V = f^{-1}(W)$ and denote by $w : F 
+$\II_s : \mathcal W_s \to \Grass(E,s)$.
+Pick $(f,W) \in \mathcal W_s$. Set $V = f^{-1}(W)$ and denote by $w : F 
 \to F/W$ the canonical projection.
 Similarly to what we have done for direct images, one can prove that
-the tangent space of $\mathcal W_r$ at some point $(f,W) \in \mathcal
-W_r$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
+the tangent space of $\mathcal W_s$ at some point $(f,W) \in \mathcal
+W_s$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
 pairs $(df,dw)$ such that
@@
 factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
-$f$. Furthermore $\II_r$ is differentiable at $(f,W)$ and its 
+$f$. Furthermore $\II_s$ is differentiable at $(f,W)$ and its 
 differential is the linear mapping that takes $(df,dw)$ as above to the 
@@
 of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. We 
-can thus deduce the differentials of $\DI_r$ from those of $\II_s$ and 
+can thus deduce the differentials of $\DI_s$ from those of $\II_s$ and 
 \emph{vice versa}.
@@
 \times \Hom(V_2, E/V_2)$ such that $dv_1 \equiv dv_2 \pmod{V_1 + V_2}$ 
-on the intersection $V_1 \cap V_2$ and the differential of $\Sigma_r$
+on the intersection $V_1 \cap V_2$, and the differential of $\Sigma$
 at $(V_1, V_2)$ maps $(dv_1, dv_2)$ to $dv \in \Hom(V, E/V)$ (with $V

### 2015-04-26 examples/stability.tex
Small fix

@@
 as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
-such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.  Write $a^{(m)}$ (resp. $b^{(m)}$)
-for the number of $a_i$'s (resp. $b_i$'s) which are at least $m$.
+such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.  
+%Write $a^{(m)}$ (resp. $b^{(m)}$)
+%for the number of $a_i$'s (resp. $b_i$'s) which are at least $m$.
 
@@
 \displaystyle 
-\sum_{m=1}^\infty a^{(m)} b^{(m)}.
+\sum_{i,j} \min(a_i,b_j)
 \end{array}
@@
 summand, we deduce the first statement of the Proposition. The second
-statement is left to the reader.
+statement is now clear.
 \end{proof}
 
-From the perspective of precision, the second statement of Proposition \ref{prop:mulmatrix}
-means that the computation of $AB$ gains $\sum_{m=1}^\infty 
-a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
+From the perspective of precision, the second statement of Proposition 
+\ref{prop:mulmatrix} means that the computation of $AB$ gains 
+$\sum_{i,j} \min(a_i,b_j)$ significant digits in absolute 
+precision\footnote{We
 note that, on the other side, the valuation of the entries of $AB$ may 

### 2015-04-26 resultants/resultants.tex
Small fix

@@
-\documentclass{sig-alternate}
+%\documentclass{sig-alternate}
 %\documentclass{lms}
+\documentclass{article}
 \usepackage[utf8]{inputenc}
 
-\usepackage{amsmath,amssymb}
+\usepackage{vmargin}
+
+\usepackage{amsmath,amssymb,amsthm}
 %\usepackage{amsrefs}
@@
 
-\title{Resultants and subresultants of p-adic polynomials}
-
-\numberofauthors{1}
-\author{
-\alignauthor Xavier Caruso\\
-  \affaddr{Universit\'e Rennes 1}\\
-  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
-}
+\title{Resultants and subresultants of $p$-adic polynomials}
+
+%\numberofauthors{1}
+%\author{
+%\alignauthor Xavier Caruso\\
+%  \affaddr{Universit\'e Rennes 1}\\
+%  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
+%}
+\author{Xavier Caruso}
 
@@
 \end{abstract}
-
-\vspace{1mm}
- \noindent
- {\bf Categories and Subject Descriptors:} \\
-\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
-  Manipulation -- \emph{Algebraic Algorithms}
-
- \vspace{1mm}
- \noindent
- {\bf General Terms:} Algorithms, Theory
-
- \vspace{1mm}
- \noindent
- {\bf Keywords:} $p$-adic precision, subresultants, random $p$-adic 
-polynomials
-\medskip
 
@@
 algorithm when it is called on random inputs which are monic $2$-adic 
-polynomials of fixed degree. Looking at the last line, we see that 
+polynomials of fixed degree. \todo{(Donner l'algorithme.)} Looking at the last line, we see that 
 extended Euclide's algorithm outputs the B\'ezout coefficients of two 
@@
 
-\medskip
-
-\noindent
-{\bf Content of the paper.}
+\paragraph{Content of the paper}
+
 The aim of this article is twofold. We first provide in \S 
@@
 
+\paragraph{Notations}
+
+Figure~\ref{fig:notations} summerizes the main notations used in
+this paper. The definitions of many of them will be recalled in 
+\S \ref{subsec:setting}.
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{rcl}
+$\ring$ &--- & a commutative ring (without any further assumption) \\
+$\A$ &---& a complete discrete valuation ring \\
+$\pi$ &---& a uniformizer of $\A$ \\
+$K$ &---& the fraction field of $\A$ \\
+$k$ &---& the residue field of $\A$ \smallskip \\
+
+$\ring_{<n}[X]$ &---& the free $\ring$-module consisting of
+polynomials over $\ring$ of degree $<n$ \\
+$\ring_{\leq n}[X]$ &---& the free $\ring$-module consisting of
+polynomials over $\ring$ of degree $\leq n$ \\
+$\ring_n[X]$ &---& the affine space consisting of \emph{monic}
+polynomials over $\ring$ of degree $n$. \smallskip \\
+
+$\Res^{d_A,d_B}(A,B)$ &---& The resultant of $A$ and
+$B$ ``computed in degree $(d_A,d_B)$'' \\
+$\Res_j^{d_A,d_B}(A,B)$ &---& The $j$-th subresultant of $A$ and
+$B$ ``computed in degree $(d_A,d_B)$'' \\
+\end{tabular}
+\end{center}
+
+\vspace{-0.2cm}
+
+\caption{Notations used in the paper}
+\label{fig:notations}
+\end{figure}
+
 \section{The setting}
+\label{subsec:setting}
 
@@
 
-\noindent
-\textbf{Definitions.}
-Let $\A$ be a \emph{discrete valuation ring} (DVR for short), that is a 
-domain equipped with a map $\val : \A \to \Z \cup \{+\infty\}$ --- the 
-so-called \emph{valuation} --- satisfying the four axioms: (1)~$\val(x) 
-= +\infty$ iff $x = 0$, (2)~$\val(xy) = \val(x) + \val(y)$ and 
-(3)~$\val(x+y) \geq \min(\val(x), \val(y))$, (4)~any element of 
-valuation $0$ is invertible. In the sequel, we shall assume that the
-valuation is normalized so that it takes the value $1$. We recall that a 
-DVR admits a unique maximal ideal $\m$, consisting of elements of 
-positive valuation. Furthermore, this ideal is principal and generated 
-by any element of valuation $1$. Such an element is called a 
+\begin{deftn}
+A \emph{discrete valuation ring} (DVR for short) is a domain $\A$
+equipped with a map $\val : \A \to \Z \cup \{+\infty\}$ --- the 
+so-called \emph{valuation} --- satisfying the four axioms: 
+\begin{enumerate}
+\item $\val(x) = +\infty$ iff $x = 0$
+\item $\val(xy) = \val(x) + \val(y)$
+\item $\val(x+y) \geq \min(\val(x), \val(y))$
+\item any element of valuation $0$ is invertible.
+\end{enumerate}
+\end{deftn}
+
+Throughout this paper, we fix a discrete valuation ring $\A$ and assume 
+that the valuation on it is normalized so that it takes the value $1$. 
+We recall that $\A$ admits a unique maximal ideal $\m$, consisting of 
+elements of positive valuation. Furthermore, this ideal is principal and 
+generated by any element of valuation $1$. Such an element is called a 
 \emph{uniformizer}. Let us fix one of them and denote it by $\pi$. The 
-\emph{residue field} of $W$ is the quotient $W/\m = W/\pi W$ and we 
+\emph{residue field} of $\A$ is the quotient $\A/\m = \A/\pi\A$ and we 
 shall denote it by $k$.
@@
 The valuation defines a distance $d$ on $\A$ by letting $d(x,y) = 
-e^{-\val(x-y)}$ for all $x, y \in \A$. We say that $\A$ is \emph{complete}
-if it is complete with respect to the previous distance. Assuming that
-$\A$ is complete, any element $x \in \A$ can be written uniquely as a 
-convergent series:
+e^{-\val(x-y)}$ for all $x, y \in \A$. We say that $\A$ is 
+\emph{complete} if it is complete with respect to $d$, in the sense that 
+every Cauchy sequence converges. Assuming that $\A$ is complete, any 
+element $x \in \A$ can be written uniquely as a convergent series:
 \begin{equation}
@@
 \end{equation}
-where the $x_i$'s lie in a fixed set of representatives of classes
-modulo $\pi$. Therefore, as an additive group, $W$ is isomorphic to 
+where the $x_i$'s lie in a fixed set $S$ of representatives of classes
+modulo $\pi$. Therefore, as an additive group, $\A$ is isomorphic to 
 the set of sequences $\N \to k$. On the contrary, the multiplicative
@@
 uniquely to $K$ by letting $\val(\frac x y) = \val(x) - \val(y)$. 
-Moreover, it follows from the axiom~(4) that $K$ is obtain from $\A$ by 
+Moreover, it follows from axiom 4 that $K$ is obtain from $\A$ by 
 inverting $\pi$. Thus, any element of $K$ can be uniquely written as an 
-infinite sum $x = \sum_{i=v}^\infty x_i \pi^i$ where $v$ is some 
-relative integer and the $x_i$'s are as above. The valuation of $x$ can 
-be easily read on this writing: it is the smallest integer $v$ such that 
-$x_i \not\equiv 0 \pmod \pi$.
-
-The two main examples of complete discrete valuation rings are the 
-ring of $p$-adic integers $\Zp$ and the ring of formal series 
-$k[[x]]$ over a field $k$. Their fraction fields are $\Q_p$ and
-$k((x))$ respectively.
+infinite sum:
+\begin{equation}
+\label{eq:expandCDVF}
+x = \sum_{i=i_0}^\infty x_i \pi^i
+\end{equation}
+where $i_0$ is some relative integer and the $x_i$'s are as above. The 
+valuation of $x$ can be easily read off this writing: it is the smallest 
+integer $i$ such that $x_i \not\equiv 0 \pmod \pi$.
+
+\subsubsection*{Examples}
+
+A first quite important class of examples of discrete valuation rings 
+are rings of formal power series over a field. They are equipped with 
+the standard valuation defined as follows: $\val(\sum_{i \geq 0} a_i 
+t^i)$ is the smallest integer $i$ with $a_i \geq 0$. The corresponding 
+distance on $k[[t]]$ is complete. Indeed, denoting by $f[i]$ the term in 
+$t^i$ in a series $f \in k[[t]]$, we observe that a sequence $(f_n)_{n 
+\geq 0}$ is Cauchy if and only if the sequences $(f_n[i])_{n \geq 0}$ 
+are all ultimately constant. A Cauchy sequence $(f_n)_{n \geq 0}$ 
+therefore converges to $\sum_{i \geq 0} a_i t^i$ where $a_i$ is the 
+limit of $f_n[i]$ when $n$ goes to $+\infty$.
+The DVR $k[[t]]$ has a distinguished uniformizer, namely $t$. Its 
+maximal ideal is then the principal ideal $(t)$ and its residue field
+is canonically isomorphic to $k$. 
+If one chooses $\pi = t$ and constant polynomials as representatives of 
+classes modulo $t$, the expansion \eqref{eq:expandCDVR} is nothing but 
+the standard writing of formal series.
+The fraction field of $k[[t]]$ is the ring of Laurent series over $k$
+and, once again, the expansion \eqref{eq:expandCDVF} corresponds to the
+usual writing of Laurent series.
 
@@
 
-\noindent
-\textbf{Precision.}
+The above example is quite important because it models all complete 
+discrete valuation rings of equal characteristic, \emph{i.e.} whose 
+fraction field and residue field have the same characteristic. On the 
+contrary, in the mixed characteristic case (\emph{i.e.} when the 
+fraction field has characteristic $0$ and the residue field has positive 
+characteristic), the picture is not that simple.
+Nevertheless, one can construct several examples and, among them, the 
+most important is certainly the ring of $p$-adic integers $\Zp$ (where 
+$p$ is a fixed prime number). It is defined as the projective limit of 
+the finite rings $\Z/p^n\Z$ for $n$ varying in $\N$. In concrete terms, 
+an element of $\Zp$ is a sequence $(x_n)_{n \geq 0}$ with $x_n \in 
+\Z/p^n\Z$ and $x_{n+1} \equiv x_n \pmod{p^n}$. The addition (resp. 
+multiplication) on $\Zp$ is the usual coordinate-wise addition (resp. 
+multiplication) on the sequences. The $p$-adic valuation of $(x_n)_{n 
+\geq 0}$ as above is defined as the smallest integer $i$ such that $x_i 
+\neq 0$. We can easily check that $\Zp$ equipped with the $p$-adic 
+valuation satisfies the four above axioms and hence is a DVR. A
+uniformizer of $\Zp$ is $p$ and its residue field is $\Z/p\Z$. A
+canonical set of representatives of classes modulo $p$ is $\{0, 1,
+\ldots, p-1\}$.
+
+Given a $p$-adic integer $x = (x_n)_{n \geq 0}$, the $i$-th digit of 
+$x_n$ in $p$-basis is well defined as soon as $i<n$ and the 
+compatibility condition $x_{n+1} \equiv x_n \pmod{p^n}$ implies that 
+it does not depend on $n$. As a consequence, a $p$-adic integer can 
+alternatively be represented as a ``number'' written in $p$-basis having 
+an infinite number of digits, that is a formal sum of the shape:
+\begin{equation}
+\label{eq:expandZp}
+a_0 + a_1 p + a_2 p^2 + \cdots + a_n p^n + \cdots
+\quad \text{with } a_i \in \{0, 1, \ldots, p-1\}.
+\end{equation}
+Additions and multiplications can be performed on the above writing
+according to the rules we all studied at school (and therefore taking
+care of caries). Similarly to the equal characteristic case, we prove
+that $\Zp$ is complete with respect to the distance associated to the 
+$p$-adic valuation. The writing \eqref{eq:expandZp} corresponds to the 
+expansion \eqref{eq:expandCDVR} provided that we have chosen $\pi = p$ 
+and $S = \{0, 1, \ldots, p-1\}$.
+The fraction field of $\Zp$ is denoted by $\Qp$.
+
+\subsubsection*{Symbolic computations on DVR}
+
+\todo{Détailler cette partie}
+% parler notamment de précision absolue et de précision relative
+
 The memory of a computer being necessarily finite, it cannot be possible 
@@
 continuous: it takes the value $1$ on an open dense subset without being 
-constant. This of course annihilate any hope of computing \textsc{gcd}s 
+constant. This of course annihilates any hope of computing \textsc{gcd}s 
 of polynomials when only approximations of them are known. Fortunately, 
@@
 
-\medskip
-
-\noindent
-\textbf{Notations.} We refer to \cite[\S 3.3]{cohen} and \cite[\S 4.1]{winkler}
-for a nice 
-introduction to resultants and subresultants and seek to fix notations
-in this paragraph. For now, we work over a 
-general ring $\ring$. We denote by $\ring_{<n}[X]$ (resp. $\ring_{\leq 
-n}[X]$) the set of polynomials of degree less than $n$ (resp. less than 
-or equal to $n$) with coefficients in $\ring$. Both $\ring_{<n}[X]$ and 
-$\ring_{\leq n}[X]$ are free $\ring$-modules of finite rank.
-
+\subsubsection*{Definitions and notations}
+
+We recall briefly basic definitions and results about resultants and 
+subresultants. For a more complete exposition, we refer to \cite[\S 
+3.3]{cohen} and \cite[\S 4.1]{winkler}.
 Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
 $d_A$ and $d_B$ be two integer greater than or equal to the degree of 
-$A$ and $B$ respectively. Given an integer $j \in \{0, \ldots, \min 
-(d_A, d_B)-1\}$, we denote by $\Res_j^{d_A, d_B}(A,B)$ the $j$-th 
-\emph{subresultant} of $(A,B)$ ``computing in degree $(d_A, d_B)$'' and 
-by $U_j^{d_A, d_B}(A,B)$ and $V_j^{d_A, d_B}(A,B)$ the corresponding 
-\emph{cofactors}. In what follows, we will freely drop the exponent 
-$(d_A,d_B)$ when $d_A$ and $d_B$ are the degrees of $A$ and $B$ 
-respectively.
-
-We recall briefly that $\Res_j^{d_A, d_B}(A,B)$ (resp. $U_j^{d_A, 
-d_B}(A,B)$, resp. $V_j^{d_A, d_B}(A,B)$) is a polynomial with 
-coefficients in $\ring$ of degree at most $j$ (resp. $d-j-1$, resp. 
-$d-j-1$) whose coefficients can be expressed, up to a sign, as minors
-of the Sylvester matrix. These polynomials satisfy in addition the
-relation:
-$$A \cdot U_j^{d_A,d_B}(A,B) + B \cdot V_j^{d_A,d_B}(A,B) = 
-\Res^{d_A,d_B}_j(A,B).$$
-The $j$-th coefficient of $R_j^{d_A,d_B}(A,B)$ is sometimes called the
-$j$-th \emph{scalar subresultant} of $(A,B)$.
-When $j=0$, the polynomial $\Res^{d_A,d_B}_0(A,B)$ is a constant which
-is nothing but the resultant of $(A,B)$. In the sequel, we shall denote
-it simply $\Res^{d_A,d_B}(A,B)$ (thus removing the index $0$).
-The relation between subresultants and \textsc{gcd}s is given by the
-following well-known theorem.
+$A$ and $B$ respectively. We consider the Sylvester application:
+$$\begin{array}{rcl}
+\psi : \quad
+\ring_{< d_B}[X] \times \ring_{< d_A}[X] & \to &
+\ring_{< d_A+d_B}[X] \smallskip \\
+(U,V) & \mapsto & AU + BV
+\end{array}$$
+The Sylvester matrix is the matrix of $\psi$ in the canonical ordered
+basis, which are
+$$\begin{array}{rrl}
+& ((X^{d_B-1},0), \ldots, (X,0), (1,0), (0,X^{d_A-1}), \ldots, (0,1))
+& \text{for the source} \smallskip \\
+\text{and} & (X^{d_A+d_B-1}, \ldots, X, 1) &
+\text{for the target.}
+\end{array}$$
+The \emph{resultant} of $A$ and $B$ (computed in degree $d_A, d_B$) is the 
+determinant of the $\psi$; we denote it by $\Res^{d_A,d_B}(A,B)$. We 
+observe that it vanishes if $d_A > \deg A$ or $d_B > \deg B$. In what 
+follows, we will freely drop the exponent $d_A, d_B$ if $d_A$ and $d_B$ 
+are the degrees of $A$ and $B$ respectively.
+Using Cramer formulae, we can build polynomials $U^{d_A, d_B}(A,B) \in
+\ring_{< d_B}[X]$ and $V^{d_A, d_B}(A,B) \in \ring_{< d_A}[X]$ whose 
+coefficients are, up to a sing, maximal minors of the Sylvester matrix 
+such that:
+$$A \cdot U^{d_A, d_B}(A,B) + B \cdot V^{d_A, d_B}(A,B) = 
+\Res^{d_A, d_B}(A,B).$$
+There polynomials are called the \emph{cofactors} of $A$ and $B$
+(computed in degree $d_A, d_B$).
+
+\medskip
+
+The subresultants are defined in the similar fashion. Given an
+integer $j$ in the range $[0, d)$ where $d = \min(d_A, d_B)$, we
+consider the ``truncated'' Sylvester application:
+$$\begin{array}{rcl}
+\psi_j : \quad
+\ring_{< d_B-j}[X] \times \ring_{< d_A-j}[X] & \to &
+\ring_{< d_A+d_B-j}[X]/\ring_{< j}[X] \smallskip \\
+(U,V) & \mapsto & AU + BV.
+\end{array}$$
+Its determinant (in the canonical basis) is the $j$-th \emph{principal 
+subresultant} of $A$ and $B$ (computed in degree $d_A, d_B$). Just 
+as before, we can construct polynomials $U_j^{d_A, d_B}(A,B) \in
+\ring_{< d_B-j}[X]$ and $V_j^{d_A, d_B}(A,B) \in \ring_{< d_A-j}[X]$
+whose coefficients are, up to a sign, minors of the Sylvester 
+matrix\footnote{Indeed, observe that the matrix of $\psi_j$ is a
+submatrix of the Sylvester matrix.} such that:
+$$A \cdot U_j^{d_A, d_B}(A,B) + B \cdot V_j^{d_A, d_B}(A,B) 
+\equiv \det \psi_j \pmod {\ring_{< j}[X]}.$$
+We set $R_j^{d_A, d_B}(A,B) = A \cdot U_j^{d_A, d_B}(A,B) + B \cdot 
+V_j^{d_A, d_B}(A,B)$: it is the $j$-th \emph{subresultant} of $A$
+and $B$ (computed in degree $d_A, d_B$). The above congruence implies
+that $R_j^{d_A, d_B}(A,B)$ has degree at most $j$ and that its
+coefficient of degree $j$ is the $j$-th principal subresultant of $A$
+and $B$. As before, we freely drop the exponent $d_A, d_B$ when
+$d_A$ and $d_B$ are equal to the degrees of $A$ and $B$ respectively.
+When $j=0$, the application $\psi_j$ is nothing but $\psi$. Therefore,
+$\Res_0^{d_A, d_B}(A,B) = \Res^{d_A, d_B}(A,B)$ and, similarly, the
+cofactors agree: we have $U_0^{d_A, d_B}(A,B) = U^{d_A, d_B}(A,B)$
+and $V_0^{d_A, d_B}(A,B) = V^{d_A, d_B}(A,B)$.
 
@@
 
-\medskip
-
-\noindent
-\textbf{The subresultant pseudo-remainder sequence.}
+\subsubsection*{The subresultant pseudo-remainder sequence}
+
+% Expliquer qu'on ne donne qu'un résultat, pas un algorithme
+
 When $\ring$ is a domain, there exists a standard nice Euclide-like
@@
 s_{i-1}^{-1} \: c_i^{-\delta_i} \cdot (S_{i-1} \,\%\, S_i)
-& \text{for } i \geq 1 \smallskip \\
+& \text{for } i \geq 0 \smallskip \\
 \displaystyle c_{i+1} = s_{i+1}^{\delta_{i+1}} \cdot c_i^{1-\delta_{i+1}}
-& \text{for } i \geq 0. 
+& \text{for } i \geq -1. 
 \end{array}\right.$$
-Here $n_i = \deg S_i$ and $s_i$ is the leading coefficient of 
+Here $n_i = \deg S_i$ \todo{define $\delta_i$ and change the notation} 
+and $s_i$ is the leading coefficient of 
 $S_i$ if $i \geq 0$ and $s_{-1} = 1$ by convention. These sequences 
@@
 
-Important simplifications occur in the common case where all scalar 
+Important simplifications occur in the common case where all principal 
 subresultants do not vanish. Under this additional assumption, one can 
@@
 
-\section{Euclide-like algorithms\\are unstable}
+\section{Euclide-like algorithms are unstable}
 \label{sec:unstable}
@@
 
+% Expliquer que les pertes avec l'algorithme des pseudo-restes
+% sont les mêmes que ceux avec l'algorithme d'Euclide
+
 \subsection{A lower bound on losses of precision}
@@
 We consider two fixed polynomials $A$ and $B$ with coefficients in $\A$ 
-whose coefficients are known with precision $O(p^N)$ for some positive
+whose coefficients are known with precision $O(\pi^N)$ for some positive
 integer $N$. For simplicity, we assume further that $A$ and $B$ are both
@@
 $R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
-In what follows, we are going use a \emph{flat precision model}: this 
+In what follows, we are going to use a \emph{flat precision model}: this 
 means that a polynomial $P(X)$ is represented as:
@@
 
-Let $V_j$ be the valuation of the scalar $j$-th subresultant of $(A,B$) 
-and $W_j$ be the smallest valuation of a coefficient of $R_j$. We of 
-course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
+Let $V_j$ be the valuation of the principal $j$-th subresultant of $A,B$ 
+and $W_j$ be the minimum of the valuations of the coefficients of $R_j$. 
+We of course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
+\todo{Change these notations.}
 
@@
 \label{prop:precEuclide}
-Let $A$ and $B$ as above. We assume that all scalar subresultants of 
+Let $A$ and $B$ as above. We assume that all principal subresultants of 
 $(A,B)$ do not vanish. Then, in the flat precision model, the 
@@
 (R_{j+1} \,\%\, R_j).$$
-Using that $R_j$ and $R_{j-1}$ have the expected degrees, the
+Using that $R_{j+1}$ and $R_j$ have the expected degrees, the
 remainder $(R_{j+1} \,\%\, R_j)$ is computed as follows:
@@
 where $s$ is the coefficient of degree $j$ of $S$. Let us first estimate 
-the precision of $S$. Recall for that the relative precision is defined 
+the precision of $S$. Recall that the relative precision is defined 
 as the difference between the (absolute) precision and the valuation: 
@@
 \item $\E[V_j] = 
-\displaystyle \sum_{i=1}^{d-j} \frac 1 {q^{\sigma(i)} - 1}$ \\
+\displaystyle \sum_{i=1}^{d-j} \frac 1 {q^{\sigma(i)} - 1}$;
 in particular $\frac 1 {q-1} \leq \E[V_j] < \frac q {(q-1)^2}$
@@
 \displaystyle \sum_{i=1}^{d-j} \frac {(2i-1) \cdot q^{\sigma(i)}}
-{(q^{\sigma(i)} - 1)^2}$ \\
+{(q^{\sigma(i)} - 1)^2}$;
 in particular $\frac{\sqrt q}{q-1} \leq \sigma[V_j] < 
@@
 is fixed during all the proof). Our conventions imply that $Z_i$ 
-vanishes if $i > d-j$. On the contrary, if $i \leq d-j$, let us define 
+vanishes if $i \geq d-j$. On the contrary, if $i < d-j$, let us define 
 $\tau(1), \ldots, \tau(d-j)$ as the numbers $\sigma(1), \ldots, 
@@
 The first inequality of claim~(2) is obtained from the relation 
-$\E[\delta_j] = \sum_{m=1}^\infty \P[\delta_j \geq m]$ using the 
-estimation of Theorem \ref{th:deltaj}. The second inequality is 
-clear because $\delta_j \leq V_j$.
+$$\E[\delta_j] 
+= \sum_{m=1}^\infty m \cdot \P[\delta_j = m]
+= \sum_{m=1}^\infty \P[\delta_j \geq m]$$ 
+using the estimation of Theorem \ref{th:deltaj}. The second inequality 
+is clear because $\delta_j \leq V_j$.
 
@@
 \item $\Res^{j,j-1}_k(R_j, R_{j-1}) = r_j^{2(j-k-1)} \: R_k$ for $k < j$;
-\item $\Res^{d-j,d-j-1}_k(U_{j-1}, U_j) = r_j^{2(d-j-k-1)} \: U_{d-1-k}$\\
+\item $\Res^{d-j,d-j-1}_k(U_{j-1}, U_j) = r_j^{2(d-j-k-1)} \: U_{d-1-k}$
 for $k < d-j$.
@@
 pseudo-remainder sequence algorithm (noting in addition that all
-scalar subresultants are nonzero).
+principal subresultants are nonzero).
 \end{proof}
@@
 way, we define $\Omega_j^\times$ as the subset of $\ring_d[X]^2$ 
-consisting of couples $(A,B)$ whose $j$-th scalar subresultants (in 
+consisting of couples $(A,B)$ whose $j$-th principal subresultants (in 
 degree $(d,d)$) is invertible in $\ring$.
@@
 $a_m X^m + a_{m+1} X^{m+1} \cdots + a_n X^n$
-with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{n,m}$ has
+with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{(m,n)}$ has
 cardinality $(q-1) q^{n-m}$. If $P$ is any polynomial of 
@@
 We define a random variable $V_j^{(0)}$ as the greatest (nonnegative) 
-integer $v$ such that all scalar subresultants $r_{j'}$ have positive 
+integer $v$ such that all principal subresultants $r_{j'}$ have positive 
 valuation for $j'$ varying in the open range $(j-v, j+v)$ (with the
@@
 \label{eq:padicprec}
-f(x + p^N \A^n) = f(x) + f'(x)(p^N \A^n)
+f(x + \pi^N \A^n) = f(x) + f'(x)(\pi^N \A^n)
 \end{equation}
@@
 has more regularity, one can exhibit a lower bound on $N$ ensuring
-that Eq.~\eqref{eq:padicprec} holds. In \cite{padicprec}, the case of locally 
-analytic functions is considered. For our purpose, it will be enough 
-to restrict ourselves to \emph{integral polynomial} functions, that
-are functions of the shape:
+that Eq.~\eqref{eq:padicprec} holds. In \cite{padicprec}, the case of 
+locally analytic functions is considered. For our purpose, it will be 
+enough to restrict ourselves to \emph{integral polynomial} functions, 
+that are functions of the shape:
 $$f(x) = \big(F_1(x), \ldots, F_m(x)\big)
@@
 \begin{proof}
-It is a direct corollary of \cite[Proposition 3.12]{}.
+It is a direct corollary of \cite[Proposition 3.12]{padicprec}.
 \end{proof}
@@
 \subset f'_j(A,B)(\mathcal L) \subset \mathcal L_j$$
-where $r_j$ is the $j$-th scalar subresultant of $(A,B)$.
+where $r_j$ is the $j$-th principal subresultant of $(A,B)$.
 \end{lem}
@@
 that $f_j$ factors through the function $\psi_j$ introduced in \S 
-\ref{subsec:proof}. By continuity, the $j$-th scalar subresultant 
+\ref{subsec:proof}. By continuity, the $j$-th principal subresultant 
 function does not vanish on a neighborhood of $(A,B)$. By Proposition 
@@
 & (a)~$A$ and $B$ are monic of the same degree $d$ \\
-{\bf (H)}: & (b)~all scalar subresultants of $(A,B)$ \\
-& \phantom{(b)~}do not vanish modulo $\pi^{N/2}$.
+{\bf (H)}: & (b)~all principal subresultants of $(A,B)$ 
+do not vanish modulo $\pi^{N/2}$.
 \end{tabular}
@@
 Moreover Lemma \ref{lem:diff} and Proposition \ref{prop:padicprec} imply
-that any perturbation of $(R_j, R_{j-1})$ by an element of $p^N r_j^2 
+that any perturbation of $(R_j, R_{j-1})$ by an element of $\pi^N r_j^2 
 \mathcal L_j$ will not affect the values of the next $R_k$'s modulo 
@@
 
-\smallskip\noindent 5.\ \hspace{0.3cm}{\bf compute} $R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})$\\
-\phantom{6.\ }\hspace{0.3cm}at precision $O(\pi^{N+2 v_{j-1}})$
+\smallskip\noindent 5.\ \hspace{0.3cm}{\bf compute} $R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})$ at precision $O(\pi^{N+2 v_{j-1}})$
 
@@
 We remark that this value can be computed at precision $O(\pi^N)$
-because the leading coefficient of $B$ is assume to be invertible.
-Dropping entirely~(a') seems nevertheless to be more difficult. Let us 
+because the leading coefficient of $B$ is assumed to be invertible.
+Dropping entirely~(a') seems nevertheless more difficult. Let us 
 move to assumption~(b) and first remark that it is satisfied with high 
@@
 just interested in computing the $j$-th subresultant for a particular
-$j$, then we just need to assume the non-vanishing of the scalar
+$j$, then we just need to assume the non-vanishing of the principal
 subresultants in the range $[j{+}1, d{-}1]$.
@@
 
-\section{Conclusion: p-adic floats?}
+\section{Conclusion: $p$-adic floats?}
 
@@
 \begin{thebibliography}{99}
-\footnotesize
+%\footnotesize
 \bibitem{andrews}

### 2015-04-26 examples/stability.tex
Added 2 remarks + 1 item in the biblio

@@
 $K$-vector spaces.  We shall contrast this approach with the
-\emph{coordinate-wise method} used in Sage and Magma,
+\emph{coordinate-wise method} used in Sage, % and Magma,
 where the precision of an element is specified by giving the precision
-of each coordinate separately.
+of each coordinate separately and is updated after each basic
+operation.
 
@@
 yields the optimum possible precision. 
+We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
+exposition.
 
@@
 
-\vspace{-0.3cm}
+\vspace{-0.4cm}
 
@@
 \label{fig:LU}
+
 \end{figure}
@@
 
-\vspace{-1ex}\noindent\hrulefill
-
-\medskip
+%\vspace{-1ex}\noindent\hrulefill
+
+%\medskip
 
@@
 \label{fig:vectorspace}
+
+\vspace{-0.2cm}
+
 \end{figure}

### 2015-04-27 examples/stability.tex
Better vertical spacing (?)
Added a link to our code (on github)

@@
 describe experiments with lattice-based precision analysis.
-
 Section \ref{ssec:grassgeo} reviews the geometry of Grassmannians,
@@
 on tracking the precision of subspace arithmetic in practice.
-
 In the appendix, we give a proof of Proposition \ref{prop:boundLambdaf}.
+
+The code used to make experiments presented in this paper is 
+available at \url{https://github.com/CETHop/padicprec}.
 
@@
 $1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.
+
+\medskip
 
@@
 \begin{deftn} \label{def:diffused}
-If $H \subset E$ is a lattice, let
-\[
-H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H).
-\]
-Set $k = \text{length}(H_0/H)$.  Then we say that $H$ has $k$
-\emph{diffused digits of precision}.
+Let $H \subset E$ be a lattice. The number of 
+\emph{diffused digits of precision} of $H$ is the length of
+$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
 \end{deftn}
@@
 defined by:
+
+\vspace{-0.2cm}
+
 \begin{align*}
@@
 
+\vspace{-0.1cm}
+
 \caption{Admissible region for the graph of $\Lambda(f)$}
 \label{fig:area}
+
 \end{figure}
@@
 get:
-\[
-\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert,
-\]
+$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert$
 %  \quad \forall x \in \R.$$
@@
 
-Let $M_{m,n}(K)$ denote the space of $m \times n$ matrices over $K$.  We will repeatedly
-use the Smith decomposition for $M \in M_{m,n}(K)$:
-\[
-M = U_M \cdot \Delta_M \cdot V_M,
-\]
+Let $M_{m,n}(K)$ denote the space of $m \times n$ matrices over $K$.  
+We will repeatedly use the Smith decomposition for $M \in M_{m,n}(K)$,
+which is $M = U_M \cdot \Delta_M \cdot V_M$
 with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
@@
 
-\vspace{-0.4cm}
+\vspace{-0.3cm}
 

### 2015-04-27 examples/stability.tex
ACM Publishing Licence

@@
 
+\permission{%
+Permission to make digital or hard copies of all or part of this work for
+personal or classroom use is granted without fee provided that copies
+are not made or distributed for profit or commercial advantage and
+that copies bear this notice and the full citation on the first page.
+Copyrights for components of this work owned by others than the
+author(s) must be honored. Abstracting with credit is permitted. To
+copy otherwise, or republish, to post on servers or to redistribute to
+lists, requires prior specific permission and/or a fee.
+Request permissions from Permissions@acm.org.}
 \conferenceinfo{ISSAC'15,}{July 6--9, 2015, Bath, United Kingdom.}
 \copyrightetc{Copyright \copyright~2015 ACM \the\acmcopyr}
-\crdata{978-1-4503-3435-8/15/07}
+\crdata{978-1-4503-3435-8/15/07\ ...\$15.00\\
+DOI: http://dx.doi.org/10.1145/2755996.2756655}
 

### 2015-05-22 examples/stability.tex
US letter format
ACM remarks

@@
-\documentclass{sig-alternate}
+\documentclass{sig-alternate-2013}
+
+\setlength{\paperheight}{11in}
+\setlength{\paperwidth}{8.5in}
+
 \usepackage[utf8]{inputenc}
@@
 
-\title{p-adic stability in linear algebra}
+\title{p-Adic Stability In Linear Algebra}
 
@@
   \affaddr{Universit\'e Rennes 1}\\
-  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
+  \affaddr{\textsf{xavier.caruso@normalesup.org}}
 \alignauthor David Roe \\
   \affaddr{University of British Columbia}\\
-  \email{\normalsize \textsf{roed.math@gmail.com}}
+  \affaddr{\textsf{roed.math@gmail.com}}
 \alignauthor Tristan Vaccon\\
   \affaddr{Universit\'e Rennes 1}\\
-  \email{\normalsize \textsf{tristan.vaccon@univ-rennes1.fr}}
+  \affaddr{\textsf{tristan.vaccon@univ-rennes1.fr}}
 }
@@
 \terms{Algorithms, Theory}
-\keywords{$p$-adic precision, linear algebra, ultrametric analysis}
+\keywords{$p$-adic precision; linear algebra; ultrametric analysis}
 
@@
 $f(V)$. It follows from this description that the tangent space of 
-$\Grass(E,d)$ is canonically isomorphic to $$\Hom(V, E) / \End(V) 
-\simeq \Hom(V, E/V).$$
+$\Grass(E,d)$ is canonically isomorphic to $\Hom(V, E) / \End(V)$,
+which is $\Hom(V, E/V)$.
 
@@
 
+\smallskip
+
 \noindent
@@
 
-\smallskip
+\medskip
 
@@
 
-%\vspace{-1ex}\noindent\hrulefill
-
-%\medskip
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
 
@@
 
+{\raggedright
 \bibliographystyle{plain}
 \bibliography{../roebib/Biblio,extras}
+}
 

### 2015-06-25 resultants/resultants.tex
Talk at ISSAC: preliminary version of slides

@@
 \usepackage{enumerate}
-%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
+\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
 \usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
@@
 \newcommand{\id}{\text{\rm id}}
-\newcommand{\Res}{\text{\rm Res}}
+\renewcommand{\Res}{\text{\rm Res}}
 \newcommand{\lc}{\text{\rm lc}}
@@
 %
-Figure \ref{fig:precision} illustrates the instability of extended Euclide's 
-algorithm when it is called on random inputs which are monic $2$-adic 
-polynomials of fixed degree. \todo{(Donner l'algorithme.)} Looking at the last line, we see that 
+Figure \ref{fig:precision} illustrates the instability of the classical
+extended Euclide's algorithm (\emph{cf} Algorithm \ref{algo:Euclide})
+when it is called on random inputs which are monic $2$-adic 
+polynomials of fixed degree. Looking at the last line, we see that 
 extended Euclide's algorithm outputs the B\'ezout coefficients of two 
@@
 of the input increases.
+%
+\begin{algorithm}
+\todo{Write it}
+
+\caption{Extended Euclide's algorithm}
+\label{algo:Euclide}
+\end{algorithm}
 
@@
 
-A first quite important class of examples of discrete valuation rings 
+A first class of examples of discrete valuation rings 
 are rings of formal power series over a field. They are equipped with 
 the standard valuation defined as follows: $\val(\sum_{i \geq 0} a_i 
-t^i)$ is the smallest integer $i$ with $a_i \geq 0$. The corresponding 
+t^i)$ is the smallest integer $i$ with $a_i \neq 0$. The corresponding 
 distance on $k[[t]]$ is complete. Indeed, denoting by $f[i]$ the term in 
@@
 classes modulo $t$, the expansion \eqref{eq:expandCDVR} is nothing but 
-the standard writing of formal series.
+the standard writing of a formal series.
 The fraction field of $k[[t]]$ is the ring of Laurent series over $k$
@@
 % parler notamment de précision absolue et de précision relative
+% expliquer comment la précision se comporte vis-à-vis des opérations
+% traiter un exemple de calcul de pgcd avec l'algorithme d'Euclide
 
@@
 Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
-$d_A$ and $d_B$ be two integer greater than or equal to the degree of 
+$d_A$ and $d_B$ be two integers greater than or equal to the degree of 
 $A$ and $B$ respectively. We consider the Sylvester application:
@@
 respect to base change: if $f : \ring \to \ring'$ is a morphism of rings 
-and $A$ and $B$ are polynomials over $\ring$ then
-$\Res_j^{d_A,d_B}(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$
-where $f(A)$ and $f(B)$ denotes the polynomials deduced from $A$
-and $B$ respectively by applying $f$ coefficient-wise. This property
-is sometimes referred to as the \emph{functoriality} of subresultants.
+and $A$ and $B$ are polynomials over $\ring$ then $\Res_j^{d_A,d_B} 
+(f(A),f(B)) = f\big(\Res_j^{d_A,d_B}(A,B)\big)$ where $f(A)$ and $f(B)$ 
+denotes the polynomials deduced from $A$ and $B$ respectively by 
+applying $f$ coefficient-wise. This property is sometimes referred to as 
+the \emph{functoriality} of subresultants.
+We emphasize that, when $f$ is not injective, the relation 
+$\Res_j(f(A),f(B)) = f\big(\Res_j(A,B)\big)$ does \emph{not} hold in 
+general since applying $f$ may decrease the degree. Nevertheless, if
+$d_A$ and $d_B$ remained fixed, this issue cannot happen.
 
@@
 S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
-\displaystyle S_{i+1} = (-s_i)^{\delta_i + 1}
-s_{i-1}^{-1} \: c_i^{-\delta_i} \cdot (S_{i-1} \,\%\, S_i)
+\displaystyle S_{i+1} = (-s_i)^{\varepsilon_i + 1}
+s_{i-1}^{-1} \: c_i^{-\varepsilon_i} \cdot (S_{i-1} \,\%\, S_i)
 & \text{for } i \geq 0 \smallskip \\
-\displaystyle c_{i+1} = s_{i+1}^{\delta_{i+1}} \cdot c_i^{1-\delta_{i+1}}
+\displaystyle c_{i+1} = s_{i+1}^{\varepsilon_{i+1}} \cdot c_i^{1-\varepsilon_{i+1}}
 & \text{for } i \geq -1. 
 \end{array}\right.$$
-Here $n_i = \deg S_i$ \todo{define $\delta_i$ and change the notation} 
-and $s_i$ is the leading coefficient of 
-$S_i$ if $i \geq 0$ and $s_{-1} = 1$ by convention. These sequences 
-are finite and the above recurrence applies until $S_i$ has reached 
-the value $0$.
+Here $n_i = \deg S_i$, $\varepsilon_i = n_{i+1} - n_i$ and $s_i$ is the 
+leading coefficient of $S_i$ if $i \geq 0$ and $s_{-1} = 1$ by 
+convention. These sequences are finite and the above recurrence applies 
+until $S_i$ has reached the value $0$.
 
@@
 & = & 0 & \text{if } n_i < j < n_{i-1} - 1 < j \\
-& = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\delta_i-1} \cdot S_i & 
+& = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\varepsilon_i-1} \cdot S_i & 
 \text{if } j = n_i
@@
 \begin{rem}
-The Proposition \ref{prop:prem} computes \emph{all} subresultants. We
+The Proposition \ref{prop:prem} gives a formula for \emph{all} subresultants. We
 note moreover that, in the common case where $n_{i-1} = n_i - 1$, the
@@
 Mimicing ideas behind extended Euclide's algorithm, one can define the 
-``extended subresultant pseudo-remainder sequence'' as well and compute
-this way at the same time the cofactors.
+``extended subresultant pseudo-remainder sequence'' as well and obtains
+recursive formulae for cofactors at the same time.
 \end{rem}
@@
 meaning that we $\deg S_i = d_B - i$ for all $i$. The sequence 
-$(S_i)$ then stops at $i = d_B$. Moreover, the $\delta_i$'s and the 
-$c_i$'s are now all ``trivial'' since we have $\delta_i = 1$ and $c_i
+$(S_i)$ then stops at $i = d_B$. Moreover, the $\varepsilon_i$'s and the 
+$c_i$'s are now all ``trivial'' since we have $\varepsilon_i = 1$ and $c_i
 = s_i$ for all $i$. The recurrence formula then becomes:
@@
 \cdot (S_{i-1} \,\%\, S_i) \quad \text{for } i \geq 1.$$
-To conclude with, we remark that Proposition \ref{prop:prem} simply 
-states that $R_j = S_{d_B-j}$ in this case.
+and Proposition \ref{prop:prem} now simply states that $R_j = S_{d_B-j}$ 
+in this case.
 
@@
 
-%The aim of this section is to study the loss of precision
-
-% Expliquer que les pertes avec l'algorithme des pseudo-restes
-% sont les mêmes que ceux avec l'algorithme d'Euclide
+\todo{Write a short introduction}
+
+\todo{
+Expliquer que les pertes avec l'algorithme des pseudo-restes
+sont les mêmes que ceux avec l'algorithme d'Euclide.
+}
 
@@
 computing a polynomial, the software truncates the precision on each
-coefficients to the smallest precision. 
+coefficient to the smallest precision. 
 One can argue that this assumption is too strong (compared to usual
@@
 We of course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
-\todo{Change these notations.}
 
@@
 pseudo-remainder sequence algorithm (noting in addition that all
-principal subresultants are nonzero).
+principal subresultants are nonzero). \todo{Give details.}
 \end{proof}
@@
 Given a subset $J$ of $\{0, \ldots, d-1\}$, we denote by 
-$\Omega_\ring(J)$ the subset of $\Omega_\ring$ consisting of those 
+$\Omega_\ring(J)$ the subset of $\Omega_\ring$ consisting of
 couples $(A,B)$ for which $r_j(A,B)$ does not vanish if and only if $j 
@@
 additional notations. First, we fix $J$ and write $J = \{n_1, \ldots, 
-n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$. We also set $n_{\ell+1} = 
-0$ by convention. Then, given $n$ and $m$ two integers with $m < n$, let 
+n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$ and set $n_{\ell+1} = 
+0$ by convention. Given $n$ and $m$ two integers with $m < n$, let 
 $V_{(m,n)}$ denote the set of polynomials of the form
@@
 V_{(m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
-with coefficients $< m$. Third, given $(A,B)$ in $\Omega_\ring$, we 
+with coefficients $< m$. Finally, given $(A,B)$ in $\Omega_\ring$, we 
 denote by $(S_i(A,B))$ its subresultant pseudo-remainder sequence as 
@@
 is enough to reconstruct the quotient of the Euclidean division of 
-$S_i(A,B)$ by $S_{i-1}(A,B)$. Thus, one can recover $S_i(A,B)$ from 
+$S_i(A,B)$ by $S_{i-1}(A,B)$. Thus, one can reconstruct $S_i(A,B)$ from 
+the knowledge of 
 $S_{i-2}(A,B)$, $S_{i-1}(A,B)$ and $S_i(A,B)[n_{i+1}{:}]$. We deduce 
-from this that $\Lambda_J(A,B)$ determines uniquely all $S_i(A,B)$'s 
+that $\Lambda_J(A,B)$ determines uniquely all $S_i(A,B)$'s 
 and finally $A$ and $B$ themselves. This proves the claim.
 
-Finally, we note that the claim implies that the cardinality of 
+To conclude the proof, we note that the claim implies that the cardinality of 
 $\Omega_\ring(J)$ is at most $q^{2d-\ell} (q-1)^\ell$. Summing up these 
@@
 We have seen in the previous sections that Euclide's algorithm (and 
-their variants) are very unstable in practice. On the other hand, 
+their variants) are unstable in practice. On the other hand, 
 one can compute subresultants in a very stable way by evaluating
@@
 
-In this subsection, we briefly report on the results of \cite{padicprec}. In this 
-paper, the authors draw the lines of a general framework to handle a 
-sharp (often optimal) track of ultrametric precision. In order to do so, 
+In this subsection, we briefly report on the results of \cite{padicprec}. In 
+\emph{loc. cit.}, the authors draw the lines of a general framework to handle a 
+sharp (often optimal) track of ultrametric precision.
+\todo{Give more details here.}
+
+ In order to do so, 
 they proposed to attach a single precision datum --- which is a lattice 
@@
 \end{proof}
+
+\todo{Add a subsection to explain the stabilization process in
+general.}
 

### 2015-07-20 resultants/resultants.tex
Work on the paper on (sub)resultants

@@
 \newtheorem{quest}[theo]{Question}
-%\theoremstyle{definition}
+\theoremstyle{definition}
 \newtheorem{rem}[theo]{Remark}
@@
 
-As wonderfully illustrated by the success of Kedlaya-type counting points 
-algorithms \cite{kedlaya}, $p$-adic technics are gaining nowadays more and more 
-popularity in computer science, and more specifically in symbolic 
-computation. A crucial issue when dealing with $p$-adics is those of 
-stability. Indeed, just like real numbers, $p$-adic numbers are by nature 
-infinite and thus need to be truncated in order to fit in the memory of a 
-computer. The level of truncation is called the \emph{precision}. Usual 
-softwares implementing $p$-adics (\emph{e.g.} \textsc{magma} \cite{magma}, 
-\textsc{pari} \cite{pari}, \textsc{sage} \cite{sage}) 
-generally tracks the precision as follows: an individual precision is 
-attached to any $p$-adic variable and this precision is updated after 
-each basic arithmetic operation (addition and multiplication). This way 
-of tracking precision can be seen as the analogue of the arithmetic 
-intervals in the real setting.
-
-In the paper \cite{padicprec}, the authors propose a new framework to control 
-$p$-adic precision. The aim of this paper is to illustrate the technics 
-of \emph{loc. cit.} on the concrete example of computation of 
+As wonderfully illustrated by the success of Kedlaya-type counting 
+points algorithms \cite{kedlaya}, $p$-adic technics are gaining nowadays 
+more and more popularity in computer science, and more specifically in 
+symbolic computation. A crucial issue when dealing with $p$-adics is 
+those of stability. Indeed, just like real numbers, $p$-adic numbers are 
+by nature infinite and thus need to be truncated in order to fit in the 
+memory of a computer. The level of truncation is called the 
+\emph{precision}. Usual softwares implementing $p$-adics (\emph{e.g.} 
+\textsc{magma} \cite{magma}, \textsc{pari} \cite{pari}, \textsc{sage} 
+\cite{sage}) generally tracks the precision as follows: an individual 
+precision is attached to any $p$-adic variable and this precision is 
+updated after each basic arithmetic operation (addition and 
+multiplication). This way of tracking precision can be seen as the 
+analogue of the arithmetic intervals in the real setting. We refer
+to \S \ref{subsubsec:computDVR} for more details.
+
+In the paper \cite{padicprec}, the authors propose a new framework to 
+control $p$-adic precision. The aim of this paper is to illustrate the 
+technics of \emph{loc. cit.} on the concrete example of computation of 
 \textsc{gcd}s and subresultants of $p$-adic polynomials. There is 
-actually a real need to do this due to the combination of two reasons: on 
-the one hand, computating \textsc{gcd}s is a very basic operation for 
+actually a real need to do this due to the combination of two reasons: 
+on the one hand, computating \textsc{gcd}s is a very basic operation for 
 which it cannot be acceptable to have important instability whereas, on 
@@
 \multirow{2}{*}{\hspace{0.2cm}Degree\hspace{0.2cm}\null} & 
-\multicolumn{2}{c|}{\hspace{0.3cm}Loss of precision (in number of digits)\hspace{0.3cm}\null}\\
+\multicolumn{2}{c|}{\hspace{0.1cm}Loss of precision (in number of significant digits)\hspace{0.1cm}\null}\\
 \cline{2-3}
-& \hspace{0.3cm}Euclide algorithm\hspace{0.3cm}\null & expected \\
+& \hspace{0.5cm}Euclide algorithm\hspace{0.5cm}\null & expected \\
 \hline
@@
 when it is called on random inputs which are monic $2$-adic 
-polynomials of fixed degree. Looking at the last line, we see that 
+polynomials of fixed degree (see also Example \ref{ex:Euclide}). Looking at the last line, we see that 
 extended Euclide's algorithm outputs the B\'ezout coefficients of two 
@@
 \begin{algorithm}
-\todo{Write it}
-
+  \SetKwInOut{Input}{Input}
+  \SetKwInOut{Output}{Output}
+  \Input{Two polynomials $A, B \in \Qp[X]$ (whose coefficients are known at given precision)}
+  \Output{A triple $D, U, V$ such that $D = AU + BV = \gcd(A,B)$}
+
+  \BlankLine
+
+  $R_1 \leftarrow A$; $U_1 \leftarrow 1$; $V_1 \leftarrow 0$;
+
+  $R_2 \leftarrow B$; $U_2 \leftarrow 0$; $V_2 \leftarrow 1$;
+
+  $k \leftarrow 2$;
+
+  \While{$R_k \neq 0$}
+    {$Q, R_{k+1} \leftarrow$ quotient and remainder in the Euclidean
+     division of $R_{k-1}$ by $R_k$;
+
+     $U_{k+1} \leftarrow U_{k-1} - Q U_k$;
+
+     $V_{k+1} \leftarrow V_{k-1} - Q V_k$;
+
+     $k \leftarrow k+1$;}
+
+  \Return{$R_{k-1}, U_{k-1}, V_{k-1}$};
 \caption{Extended Euclide's algorithm}
@@
 
-\subsubsection*{Examples}
+\subsubsection{Examples}
 
@@
 
-\subsubsection*{Symbolic computations on DVR}
-
-\todo{Détailler cette partie}
-% parler notamment de précision absolue et de précision relative
-% expliquer comment la précision se comporte vis-à-vis des opérations
-% traiter un exemple de calcul de pgcd avec l'algorithme d'Euclide
-
-The memory of a computer being necessarily finite, it cannot be possible 
-to find a way to represent exhaustively all elements of $\A$. 
-Very often, mimicing what we do for real numbers, we choose to truncate 
-the expansion \eqref{eq:expandCDVR} at some finite level. Concretely,
-this means that we work with approximations of elements of $\A$ of the
-form
-$\sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N)$
-where $N$ is some integer called the \emph{precision} of $x$ and 
-the term $O(\pi^N)$ means that the $x_i$'s with $i \geq N$ are not 
-specified. Hence, the notation above does not represent a unique
-element of $\A$ but all elements lying in the ball having center
-$\sum_{i=0}^{N-1} x_i \pi^i$ and radius $e^{-N v(\pi)}$. In other
-words, on a computer, we do not work with elements of $\A$ but with
-balls in $\A$ which are more tractable.
+\subsubsection{Symbolic computations over DVR}
+\label{subsubsec:computDVR}
+
+The memory of a computer being necessarily finite, it is not possible to 
+represent exhaustively all elements of $\A$. Very often, mimicing 
+what we do for real numbers, we choose to truncate the expansion 
+\eqref{eq:expandCDVR} at some finite level. Concretely, this means 
+that we work with approximations of elements of $\A$ of the form
+\begin{equation}
+\label{eq:approx}
+x = \sum_{i=0}^{N-1} x_i \pi^i + O(\pi^N) \quad \text{with } N \in \N
+\end{equation}
+where the notation $O(\pi^N)$ means that the $x_i$'s with $i \geq N$ are 
+not specified.
+
+\begin{rem}
+\label{rem:balls}
+From a theoretical point of view, the expression~\eqref{eq:approx}
+does not represent a single element $x$ of $\A$ but an open ball in
+$\A$, namely the ball of radius $e^{-N}$ centered at $\sum_{i=0}^{N-1} 
+x_i \pi^i$ (or actually any element congruent to it modulo $\pi^N$).
+In other words, on a computer, we cannot work with actual $p$-adic
+numbers and we replace them by balls which are more tractable (at
+least, they can be encoded by a finite amount of information).
+\end{rem}
+
+The integer $N$ appearing in Eq.~\eqref{eq:approx} is the so-called 
+\emph{absolute precision} of $x$. The \emph{relative precision} of $x$ 
+is defined as the difference $N-v$ where $v$ denotes the valuation of 
+$x$. Continuing the comparison with real numbers, the relative precision 
+corresponds to the number of significant digits since $x$ can be 
+alternatively written:
+$$x = p^v \sum_{j=0}^{N-v-1} y_j \pi^j + O(\pi^N)
+\quad \text{with } y_j = x_{j+v} \text{ and } y_0 \neq 0.$$
+Of course, it may happen that all the $x_i$'s ($0 \leq i < N$) vanish,
+in which case the valuation of $x$ is undetermined. In this particular
+case, the relative precision of $x$ is undefined.
+
+There exist simple formulas to following precision after each single
+elementary computation. For instance, basic arithmetic operations can
+be handled using:
+\begin{align}
+\big(a + O(\pi^{N_a})\big) + \big(b + O(\pi^{N_b})\big) 
+& = a + b + O(\pi^{\min(N_a,N_b)}), 
+\label{eq:precadd} \\
+\big(a + O(\pi^{N_a})\big) - \big(b + O(\pi^{N_b})\big) 
+& = a - b + O(\pi^{\min(N_a,N_b)}), 
+\label{eq:precsub} \\
+\big(a + O(\pi^{N_a})\big) \times \big(b + O(\pi^{N_b})\big) 
+& = ab + O(\pi^{\min(N_a + \val(b), N_b + \val(a))}).
+\label{eq:precmul} \\
+\big(a + O(\pi^{N_a})\big) \div \big(b + O(\pi^{N_b})\big) 
+& = \frac a b + O(\pi^{\min(N_a - \val(b), N_b + \val(a) - 2 \val(b))}).\footnotemark
+\label{eq:precdiv} 
+\end{align}
+
+\footnotetext{We observe that these formulas can be rephrased as follows: 
+the absolute (resp. relative) precision on the result of a sum or a
+substraction (resp. a product or a division) is the minimum of the 
+absolute (resp. relative) precisions on .}
+
+\noindent
+Combining these formulas, one can track the precision while executing 
+any given algorithm. This is the analogue of the standard \emph{interval 
+arithmetic} over the reals. Many usual softwares (as \textsc{sage}, 
+\textsc{magma}) implement $p$-adic numbers this way. We shall see later
+that this often results in overestimating the losses of precision.
+
+\begin{ex}
+\label{ex:Euclide}
+As an illustration, let us examine the behaviour of the precision
+on the sequence $(R_i)$ while Algorithm~\ref{algo:Euclide} is
+executed with the input:
+\begin{align*}
+A & = X^5 + 
+      \big(27 + O(2^5)\big) X^4 +
+      \big(11 + O(2^5)\big) X^3 +
+      \big(5 + O(2^5)\big) X^2 +
+      \big(18 + O(2^5)\big) X +
+      \big(25 + O(2^5)\big) \\
+B & = X^5 + 
+      \big(24 + O(2^5)\big) X^4 +
+      \big(25 + O(2^5)\big) X^3 +
+      \big(12 + O(2^5)\big) X^2 +
+      \big(3 + O(2^5)\big) X +
+      \big(10 + O(2^5)\big).
+\end{align*}
+The remainder in the Euclidean division of $A$ by $B$ is $R_3 = A-B$.
+We do not loose precision while performing this substraction and the
+result we get is:
+$$R_3 = 
+      \big(3 + O(2^5)\big) X^4 +
+      \big(18 + O(2^5)\big) X^3 +
+      \big(25 + O(2^5)\big) X^2 +
+      \big(15 + O(2^5)\big) X +
+      \big(15 + O(2^5)\big).$$ 
+In order to compute $R_4$, we have now to perform the Euclidean division 
+of $R_2 = B$ by $R_3$. Noting that the leading coefficient of $R_2$ has 
+valuation $0$ and using Eq~\eqref{eq:precadd}--\eqref{eq:precdiv}, we 
+deduce that this operation does not loose precision again. We get:
+$$R_4 = 
+      \big(26 + O(2^5)\big) X^3 +
+      \big(17 + O(2^5)\big) X^2 +
+      \big(4 + O(2^5)\big) X +
+      \big(16 + O(2^5)\big).$$
+We observe now that the leading coefficient of $R_4$ has valuation $1$. 
+According to Eq.~\eqref{eq:precdiv}, divising by this coefficient ---
+and therefore \emph{a fortioti} computing the euclidean division of 
+$R_3$ by $R_4$ --- will result in loosing at least one digit in
+relative precision. The result we find is:
+$$R_5 = 
+      \underbrace{\big(\textstyle \frac 3 4 + O(2^2)\big)}_{\text{rel. prec.} = 4} X^2 +
+      \underbrace{\big(6 + O(2^3)\big)}_{\text{rel. prec.} = 2} X +
+      \underbrace{\big(3 + O(2^3)\big)}_{\text{rel. prec.} = 3}.$$
+Continuing this process, we obtain:
+$$R_6 = 
+      \big(20 + O(2^5)\big) X +
+      \big(12 + O(2^5)\big) 
+\quad \text{and} \quad
+R_7 =
+      \textstyle \frac 7 4 + O(2).$$
+The relative precision on the final result $R_7$ is then $3$, which is
+less than the initial precision which was $5$.
+\end{ex}
 
@@
 
-% Expliquer qu'on ne donne qu'un résultat, pas un algorithme
-
 When $\ring$ is a domain, there exists a standard nice Euclide-like
@@
 \Res_j(A,B) & = & S_i & \text{if } j = n_{i-1} - 1 \smallskip \\
-& = & 0 & \text{if } n_i < j < n_{i-1} - 1 < j \\
+& = & 0 & \text{if } n_i < j < n_{i-1} - 1 \\
 & = & \!\big(\frac{s_i}{s_{i-1}}\big)^{\varepsilon_i-1} \cdot S_i & 
@@
 \begin{rem}
-The Proposition \ref{prop:prem} gives a formula for \emph{all} subresultants. We
-note moreover that, in the common case where $n_{i-1} = n_i - 1$, the
-two formulas giving $\Res_{n_i}(A,B)$ agree.
+The Proposition \ref{prop:prem} provides a formula for \emph{all} 
+subresultants. We note moreover that, in the common case where $n_{i-1} 
+= n_i - 1$, the two formulas giving $\Res_{n_i}(A,B)$ agree.
 
@@
 
-Important simplifications occur in the common case where all principal 
-subresultants do not vanish. Under this additional assumption, one can 
-prove that the degrees of the $S_i$'s decrease by one at each step, 
-meaning that we $\deg S_i = d_B - i$ for all $i$. The sequence 
-$(S_i)$ then stops at $i = d_B$. Moreover, the $\varepsilon_i$'s and the 
-$c_i$'s are now all ``trivial'' since we have $\varepsilon_i = 1$ and $c_i
-= s_i$ for all $i$. The recurrence formula then becomes:
+Important simplifications occur in the ``normal'' case, which is the 
+case where all principal subresultants do not vanish. Under this 
+additional assumption, one can prove that the degrees of the $S_i$'s 
+decrease by one at each step, meaning that we $\deg S_i = d_B - i$ for 
+all $i$. The sequence $(S_i)$ then stops at $i = d_B$. Moreover, the 
+$\varepsilon_i$'s and the $c_i$'s are now all ``trivial'' since we have 
+$\varepsilon_i = 1$ and $c_i = s_i$ for all $i$. The recurrence formula 
+then becomes:
 $$S_{i+1} = s_i^2 \cdot s_{i-1}^{-2}
@@
 
+\begin{algorithm}
+  \SetKwInOut{Input}{Input}
+  \SetKwInOut{Output}{Output}
+  \Input{Two polynomials $A, B \in K_d[X]$ (given at finie precision)}
+  \Output{The complete sequence of subresultants of $A$ and $B$.}
+
+  \BlankLine
+
+  $R_d \leftarrow B$; $r_d \leftarrow 1$;
+
+  $R_{d-1} \leftarrow B-A$;
+
+  \For{$j = (d-1), (d-2), \ldots, 1$}
+    {$r_j \leftarrow$ coefficient in $X^j$ of $R_j$;
+
+     \lIf{$r_j = 0$}{\textbf{raise} NotImplementedError;}
+
+     $R_{j-1} \leftarrow \texttt{prem}(R_{j+1}, R_j) / r_{j+1}^2$
+    }
+
+  \Return{$R_{d-1}, \ldots, R_0$};
+\caption{Subresultant pseudo remainder sequence algorithm}
+\label{algo:subres}
+\end{algorithm}
+
+Proposition \ref{prop:prem} of course yields an algorithm for computing 
+subresultants. In the normal case and assuming further for simplicity 
+that the input polynomials are monic of same degree, it is 
+Algorithm~\ref{algo:subres}. Note that this algorithm uses the primitive 
+\texttt{prem} which computes the pseudo-remainder: if $A$ and $B$ are 
+two polynomials, by definition $\texttt{prem}(A,B) = \lc(B)^{\deg B-\deg 
+A+1} (A \% B)$ where $\lc(B)$ denotes the leading coefficient of $B$. 
+Unfortunately, when working over a complete discrete valuation field 
+$K$, the stability of Algorithm~\ref{algo:subres} is comparable --- and 
+actually even a bit worse --- than standard Euclide algorithm. The main 
+interest of using Algorithm~\ref{algo:subres} is to avoid denominators 
+(\emph{i.e.} always work over $\A$ instead $K$) but its use almost does 
+not affect the behaviour of the \emph{relative} precision. We shall 
+explain later (\emph{cf} \S~\ref{subsec:stabilization}) a method for 
+stabilizing Algorithm~\ref{algo:subres}.
+
 \section{Euclide-like algorithms are unstable}
@@
 \todo{Write a short introduction}
-
-\todo{
-Expliquer que les pertes avec l'algorithme des pseudo-restes
-sont les mêmes que ceux avec l'algorithme d'Euclide.
-}
 
@@
 $B = X^d + \sum_{i=0} b_i X^i$.
-All relations then follows from the extended subresultants 
+All relations then follow from the extended subresultants 
 pseudo-remainder sequence algorithm (noting in addition that all
@@
 
-In this subsection, we briefly report on the results of \cite{padicprec}. In 
-\emph{loc. cit.}, the authors draw the lines of a general framework to handle a 
+In this subsection, we briefly report on the results of \cite{padicprec} 
+where the authors draw the lines of a general framework to handle a 
 sharp (often optimal) track of ultrametric precision.
-\todo{Give more details here.}
-
- In order to do so, 
-they proposed to attach a single precision datum --- which is a lattice 
-in a huge vector space --- to the set of all variables, instead of 
-attaching a precision datum to each individual variable as we do 
-usually. This proposal is supported by the following lemma in ultrametric
-analysis.
-
-\begin{lem}
+
+\subsubsection*{The notion of lattice}
+
+As underlined in Remark~\ref{rem:balls}, the usual way of tracking 
+precision consists in replacing elements of $\A$ --- which cannot fit 
+entirely in the memory of a computer --- by balls around them. Using 
+this framework, a software manipulating $d$ variables in $\A$ will work 
+with $d$ ``independant'' balls. The main proposal of \cite{padicprec} is 
+to get rid of this ``independance'' and model precision using a unique
+object contained in a $d$-dimensional vector space. In order to be more
+precise, we need the following definition.
+
+\begin{deftn}
+A $\A$-lattice in a finite dimensional vector space $E$ over $K$ is
+a $\A$-submodule of $E$ generated by a $K$-basis of $E$.
+\end{deftn}
+
+Although the defintion of a lattice is similar to that of $\Z$-lattice 
+in $\R^d$, the geometrical representation of it is quite different since 
+the elements of $\A$ themselves are not equidistributed in $K$ but form 
+a ball inside $K$ (they are exactly elements of norm $\leq 1$). More 
+generally, assume that $E$ is equipped with a ultrametric norm $\Vert 
+\cdot \Vert_E$ compatible with that on $K$ (\emph{i.e.} $\Vert \lambda x 
+\Vert_E = |\lambda| \cdot \Vert x \Vert_E$ for $\lambda \in K$, $x \in 
+E$). (A typical example is $E = K^n$ equipped with the sup norm.) One
+checks that the subsets
+$$B_E(r) = \big\{ \,\, x \in E \quad|\quad \Vert x \Vert_E \leq r \,\,\big\}$$
+are all lattices in $E$. Moreover, any lattice is deduced from $B_E(1)$ 
+by applying a bijective linear endomorphism of $E$. Therefore, lattices 
+should be thought as special neighborhoods of $0$. As a consequence, 
+cosets of the form $x + H$, where $H$ is a lattice, appear as 
+interesting candidates to model precision. This feeling is consolidated 
+by the following result which roughly speaking claims that such cosets 
+behave quite nicely under differentiable maps.
+
+\begin{lem}[\cite{padicprec}, Lemma 3.4]
 \label{lem:padicprec}
-Let $f : K^n \to K^m$ be a strictly differentiable function. Let $x$
-be a point in $K^n$ at which the differential of $f$ is surjective.
-Then, for all sufficiently large $N$, the following equality holds:
+Let $E$ and $F$ be two normed finite dimensional $K$-vector spaces.
+Let $f : E \to F$ be a function of class $C^1$ and let $x$
+be a point in $K^n$ at which the differential of $f$, denoted by
+$f'(x)$, is surjective.
+Then, for all $\rho \in (0,1]$, there exists $\delta > 0$ such that
+the following equality holds:
 \begin{equation}
 \label{eq:padicprec}
-f(x + \pi^N \A^n) = f(x) + f'(x)(\pi^N \A^n)
+f(x + H) = f(x) + f'(x)(H)
 \end{equation}
-where $f'(x)$ stands for the differential of $f$ at $x$.
+for any lattice $H$ satisfying $B_E(\rho r) \subset H \subset B_E(r)$
+for some $r < \delta$.
 \end{lem}
 
-An important feature of Lemma \ref{lem:padicprec} is that 
-Eq.~\eqref{eq:padicprec} is an equality, and not only an inclusion. 
-This reflects the optimality of the method and, as we shall see later, 
-is also the origin of the ``stabilization process'' mentionned above.
-Moreover, if the function $f$ appearing in Lemma \ref{lem:padicprec} 
-has more regularity, one can exhibit a lower bound on $N$ ensuring
-that Eq.~\eqref{eq:padicprec} holds. In \cite{padicprec}, the case of 
-locally analytic functions is considered. For our purpose, it will be 
-enough to restrict ourselves to \emph{integral polynomial} functions, 
-that are functions of the shape:
-$$f(x) = \big(F_1(x), \ldots, F_m(x)\big)
-\quad \text{with } x = (x_1, \ldots, x_n)$$
-where $F_1, \ldots, F_m$ are multivariate polynomials in $n$ variables
-with coefficients in $\A$. Under this assumption, we have:
+In what follows, we will mainly use Lemma \ref{lem:padicprec} with
+$\rho = 1$. It states in this particular case that 
+\begin{equation}
+\label{eq:padicprec2}
+f(x + B_E(r)) = f(x) + f'(x)(B_E(r))
+\end{equation}
+as soon as $r$ is small enough. It is moreover possible to provide
+an explicit upper bound on $r$ assuming that $f$ has more regularity.
+The case of locally analytic functions is treated in \cite{padicprec}
+in full generality. Nevertheless, for the application we have in mind, 
+it will be enough to restrict ourselves to the simpler case of 
+\emph{integral polynomial} functions. In order to proceed, we assume 
+that $E$ is endowed with distingushed ``orthonormal'' 
+basis\footnote{One can prove that such a basis always exists.}, that is 
+a basis $(e_1, \ldots, e_n)$ with the property that $\Vert \sum_{i=1}^n 
+x_i e_i \Vert_E = \max_{1\leq i \leq n} |x_i|$ for all 
+families of $\lambda_i$'s lying in $K$. In other words, the choice of
+this distingushed ``orthonormal'' basis defines a norm-preserving 
+isomorphism between $E$ and $K^n$ endowed with the sup norm. We assume
+similarly that we are given a distingushed ``orthonormal'' basis
+$(f_1, \ldots, f_m)$ of $F$. Then any function $f : E \to F$ can
+be written in our distinguished system of coordinates as follows:
+$$f(x) = \sum_{j=1}^m F_j(x_1, \ldots, x_n) f_j
+\quad \text{with} \quad x = \sum_{i=1}^n x_i e_i.$$
+
+\begin{deftn}
+The function $f$ is \emph{integral polynomial} if all $F_j$'s are
+polynomials functions with coefficients in $\A$.
+\end{deftn}
+
+Obviously, all integral polynomial functions are of class $C^1$ (and
+even locally analytic), so that Lemma~\ref{lem:padicprec} applies to 
+them.
+
+\begin{ex}
+Let us examine more closely the case of polynomial spaces since it will 
+be considered repeadtly in the sequel. We take $E = K_{<n}[X]$ and $F = 
+K_{<m}[X]$ and endow both with the Gauss norm, which is defined by:
+\begin{align*}
+\Vert a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} \Vert_E & 
+  = \max \big(|a_0|, |a_1|, \ldots, |a_{n-1}|\big) \\
+\Vert b_0 + b_1 X + \cdots + b_{m-1} X^{m-1} \Vert_F & 
+  = \max \big(|b_0|, |b_1|, \ldots, |b_{m-1}|\big)
+\end{align*}
+It is clear from these definitions that the canonical basis $(1, X, 
+\ldots, X^{n-1})$ and $(1, X, \ldots, X^{m-1})$ of $E$ and $F$ 
+respectively are ``orthonormal''. Moreover the coordinates in these
+basis are the $a_i$'s and the $b_i$'s respectively. Hence, an integral
+polynomial function $f : E \to F$ is nothing but a function mapping a
+a polynomial $P$ to a polynomial $Q$ whose coefficients are given by
+polynomial expressions which involve only the coefficients of $P$ and 
+some constants in $\A$.
+\end{ex}
 
@@
 \label{prop:padicprec}
-Let $f : K^n \to K^m$ be an integral polynomial function and $x \in
-\A^n$. Let $v$ be an integer such that $f'(x)(\A^n) \supset p^v \A^m$.
-Then, for all $N > v$, Eq.~\eqref{eq:padicprec} holds.
+Let $f : E \to F$ be an integral polynomial function and $x \in B_E(1)$. 
+Then, Eq.~\eqref{eq:padicprec2} holds as soon as 
+$B_F(r) \subset f'(x)(B_E(1))$.
 \end{prop}
@@
 
+\subsubsection*{Application to precision}
+
 \todo{Add a subsection to explain the stabilization process in
@@
 \subsection{The stabilization process}
+\label{subsec:stabilization}
 

### 2015-07-21 resultants/resultants.tex
The stabilization process in general

@@
 
+\definecolor{beige}{rgb}{0.9,0.9,0.7}
+
 \newcommand{\noopsort}[1]{}
@@
 \newcommand{\M}{\text{\tt M}}
+
+\renewcommand{\min}{\text{\rm min}}
+\renewcommand{\max}{\text{\rm max}}
 
@@
 
-\subsubsection*{The notion of lattice}
+\subsubsection{The notion of lattice}
 
@@
 by applying a bijective linear endomorphism of $E$. Therefore, lattices 
-should be thought as special neighborhoods of $0$. As a consequence, 
-cosets of the form $x + H$, where $H$ is a lattice, appear as 
-interesting candidates to model precision. This feeling is consolidated 
-by the following result which roughly speaking claims that such cosets 
-behave quite nicely under differentiable maps.
+should be thought as special neighborhoods of $0$ (see Figure~\ref{fig:lattice}).
+%
+\begin{figure}
+\hfill
+\begin{tikzpicture}
+\begin{scope}[beige, rounded corners=10pt]
+\fill (0.1,0.2) rectangle (3.9,5.8);
+\end{scope}
+
+\begin{scope}[xshift=2cm,yshift=3cm,rotate=20]
+\draw[black,fill=green!50] (-0.7,-1.5) rectangle (0.7,1.5);
+\draw[black,->] (-1,0)--(1,0);
+\draw[black,->] (0,-2)--(0,2);
+\end{scope}
+
+\begin{scope}[very thick, rounded corners=10pt]
+\draw (0.1,0.2) rectangle (3.9,5.8);
+\end{scope} 
+\end{tikzpicture}
+\hfill \null
+
+\caption{Picture of a lattice in the ultrametric world}
+\label{fig:lattice}
+\end{figure}
+%
+As a consequence, cosets of the form $x + H$, where $H$ is a lattice, 
+appear as interesting candidates to model precision. This feeling is 
+consolidated by the following result which roughly speaking claims that 
+such cosets behave quite nicely under differentiable maps.
 
@@
 
-\subsubsection*{Application to precision}
-
-\todo{Add a subsection to explain the stabilization process in
-general.}
+\subsubsection{Application to precision}
+
+Let us now briefly explain how one can use Lemma~\ref{lem:padicprec} for 
+tracking precision.
+
+\paragraph{Tracking precision \emph{locally}}
+
+Assume first that we want to perform a given rather simple operation --- 
+corresponding, say, to an elementary step (\emph{e.g.} an iteration of 
+the main loop) of the algorithm we are executing --- modeled by a 
+stricty differentiable function $g$ defined on an open subset $U$ of a 
+normed $K$-vector space $E$ and taking values in another normed 
+$K$-vector space $F$.
+Our input is an approximated element of $U$ which is represented 
+as a coset $C$ with respect to some lattice $H$, that is a subset of $U$ 
+of the form $C = x+H$ for some $x \in U$. We would like to insist on the 
+following: the value of $x$ is a priori \emph{not} given; only is given 
+the subset $C$. However, since $H$ is stable under addtion, we have $C = 
+x+H$ \emph{any} element $x \in C$.\footnote{This assertion means that 
+any element of the ``rectangle'' $C$ is a center of it... which might be 
+surprising if we are accustomed to real numbers.} As explained in \S 
+\ref{subsubsec:computDVR}, assuming that $g$ is given as an algebraic 
+expression, the naive solution for evaluating $g(C)$ consists in using 
+formulas \eqref{eq:precadd}--\eqref{eq:precdiv}. However, this often 
+results in an overestimation on the precision, in the following sense: 
+this method leads to some inclusion
+$$g(C) = g(x+H) \subset y + H_{\text{naive}}$$
+where $y \in F$ and $H_{\text{naive}}$ is a lattice which is generally 
+much more larger that the best one, namely $g'(x)(H)$ according to
+Lemma~\ref{lem:padicprec} (assuming that the assumptions of this Lemma
+are fullfiled). In order to avoid this and be sharp on
+precision, another solution consists in splitting the computation of 
+$g(C)$ into two parts as follows:
+\begin{enumerate}[(A)]
+\item \label{item:gpx} compute $g'(x)(H)$, and
+\item \label{item:gx} compute $g(x)$ for some $x \in C$.
+\end{enumerate}
+Part~\eqref{item:gpx} is not easy to handle in full generality: in order 
+to be efficient, a special close analysis taking advantage of the 
+particular problem under consideration is often necessary. For the 
+example considered in the paper (which is that of subresultants), we will 
+use a \emph{theoretical} argument in order to exhibit a good approximation 
+of $g'(x)(H)$, that are lattices $H_\min$ and $H_\max$ such that:
+\begin{equation}
+\label{eq:Hminmax}
+H_\min \subset g'(x)(H) \subset H_\max
+\end{equation}
+for a negligible cost. Up to allowing in some cases a \emph{very slight} 
+leak of precision, this would be enough for the application we have in 
+mind. We refer to \S \ref{subsec:stabilization} for more details about
+this part.
+
+For now on, we assume that $H_\min$ and $H_\max$ satisfying 
+Eq.~\eqref{eq:Hminmax} have been computed and we focus on 
+part~\eqref{item:gx}, which also requires some discussion. Indeed, 
+computing $g(x)$ is not straightforward because $x$ itself lies in a 
+$K$-vector space and therefore cannot be stored and manipulated on a 
+computer. Nevertheless, one can take advantage of the fact that $x$ may 
+be chosen arbitratily in $C$. More precisely, we pick a sublattice $H'$ 
+of $H$ and consider the new approximated element $x+H' \subset x+H$. 
+Concretely, this means that we arbitrarily increase the precision on the 
+given input $x$. Now, applying the naive method with $x+H'$, we compute 
+some $y \in F$ and some lattice $H'_{\text{naive}} \subset F$ with the 
+property that:
+$$g(x+H') \subset y + H'_{\text{naive}}.$$
+If furthemore $H'$ is chosen in such a way that $H'_{\text{naive}} 
+\subset H_\min$, the two cosets $y + g'(x)(H)$ and $g(C)$ have a 
+non-empty intersection because $g(x)$ lies in both. Therefore they 
+must coincide. We deduce that $y \in g(C)$. This exactly means that
+$y$ is an acceptable value for $g(x)$ and we are done.
+Moreover, estimating the dependance of $H'_{\text{naive}}$ in terms 
+of $H'$ is usually rather easy (remember that $f$ is supposed to model
+a simple operation). Hence since $H_\min$ is known --- as
+we had assumed --- finding $H'$ satisfying the required assumption is
+generally not difficult. 
+The whole method is illustrating on Figure~\ref{fig:adaptative}.
+
+\begin{figure}
+\hfill
+\begin{tikzpicture}[scale=1.2]
+
+\begin{scope}[beige, rounded corners=10pt]
+\fill (0.1,0.2) rectangle (3.9,5.8);
+\fill (6.1,0.2) rectangle (9.9,5.8);
+\end{scope}
+
+\draw[->,very thick] (4.1,3)--(5.9,3)
+  node[midway,above] { $g$ };
+
+\begin{scope}[xshift=2cm,yshift=3.5cm,rotate=20]
+\draw[black,fill=green!50] (-0.8,-1.8) rectangle (0.8,1.8);
+\node[scale=0.75] at (0,0) { $x{+}H$ };
+\end{scope}
+\draw[black,fill=purple!50] (2,2.8) rectangle (2.8,2);
+\node[scale=0.75] at (2.4,2.4) { $x{+}H'$ };
+
+\begin{scope}[xshift=8cm,yshift=3cm]
+\draw[black,fill=orange!50] (-1.7,-2.3) rectangle (1.7,1.8);
+\node[above left,scale=0.75] at (1.7,-2.3) { $g(x){+}H_\max$ };
+\draw[black,fill=green!50,rotate=-40] (-1,-1.3) rectangle (1,1.3);
+\node[scale=0.75] at (0.15,1.2) { $g(x){+}{}$ };
+\node[scale=0.75] at (0.1,0.9) { $g'(x)(H)$ };
+\draw[black,fill=red!50] (-0.8,-0.8) rectangle (0.4,0.4);
+\node[scale=0.75] at (-0.2,0.1) { $g(x){+}H_\min$ };
+\draw[black,fill=purple!50] (-0.7,-0.7) rectangle (0.3,-0.3);
+\node[scale=0.75] at (-0.2,-0.5) { $y{+}H'_{\text{naive}}$ };
+\end{scope}
+
+\begin{scope}[very thick, rounded corners=10pt]
+\draw (0.1,0.2) rectangle (3.9,5.8);
+\draw (6.1,0.2) rectangle (9.9,5.8);
+\end{scope} 
+\end{tikzpicture}
+\hfill \null
+
+\caption{Method for tracking precision based on Lemma~\ref{lem:padicprec}}
+\label{fig:adaptative}
+\end{figure}
+
+\paragraph{Tracking precision \emph{globally}}
+
+As already said, we shall use the above method for tracking precision 
+while executing a single step in a complete algorithm. Let us now 
+address the problem of ``glueing''. We consider an algorithm \texttt{F} 
+consisting in a succession of $n$ steps $\texttt{G}_0, \ldots,
+\texttt{G}_{n-1}$. It is modeled by a strictly differentiable function $f : 
+U \to F$ where $U$ is an open subset in a normed $K$-vector space $E$ 
+and $F$ is a normed $K$-vector space. The input of \texttt{F} is an 
+approximated element in $U$ represented as a coset $C = x+H$ where $x 
+\in U$ and $H$ is a lattice.
+We also introduce notations for each individual step. For all $i$, we
+assume that $\texttt{G}_i$ is modeled by a strictly differentiable
+function $g_i : U_i \to U_{i+1}$ where $U_i$ is an open subset is 
+some normed $K$-vector space $E_i$ and, by convention, $U_0 = U$, $E_0 
+= E$ and $U_n = E_n = F$. We thus have:
+
+$$f = g_{n-1} \circ g_{n-2} \circ \cdots \circ g_1 \circ g_0.$$
+For all $i$, we set $f_i = g_{i-1} \circ \cdots \circ g_0$. It is the 
+function modeling the execution of the $i$ first steps of our algorithm. 
+We further define $x_i = f_i(x)$ and $H_i = f_i'(x)(H)$. The chain rule 
+for composing differentials readily implies the recurrence
+\begin{equation}
+\label{eq:giprime}
+H_{i+1} = g_i'(x_i)(H_i)
+\end{equation}
+For simplicity, we make the following assumptions:
+\begin{itemize}
+\item the $\Zp$-submodule $H_i$ is a lattice in $E_i$ such that $x_i + 
+H_i \subset U_i$;
+\item the triple $(g_i, x_i, H_i)$ satisfies the assumptions of 
+Lemma~\ref{lem:padicprec};
+\item for all $i$, we have succeeded in finding (good enough) explicit 
+lattices $H_{\min,i}$ and $H_{\max,i}$ such that $H_{\min,i} \subset H_i 
+\subset H_{\max,i}$;
+\item for all $i$, we have succeeded in finding an explicit lattice
+$H'_i$ such that, while tracking naively precision, we end up with an
+inclusion
+$$g_i(x_i + H'_i) = x_{i+1} + H_{\text{naive},i+1}$$
+with $H_{\text{naive},i+1} \subset H_{\min,i+1}$.
+\end{itemize}
+We note that the first and the second assumptions are quite strong 
+because it implies in particular that the sequence of $\dim E_i$ is 
+non-increasing. However, it really simplifies the forthcoming discussion 
+and will be harmless for the application developed in this paper. As
+already mentionned, the construction of $H_{\min,i}$ and $H_{\max,i}$
+will generally follow from a \emph{theoretical} argument depending on
+the setting, while exhibiting $H'_i$ will often be straightforward.
+Anyway, we are now in position to apply the method for 
+tracking precision locally we have discussed earlier to all $g_i$'s. 
+This leads to a \emph{stabilized} version of the algorithm \texttt{G} 
+whose skeleton is depicted in Algorithm~\ref{algo:stabilizedG}.
+%
+\begin{algorithm}
+  \SetKwInOut{Input}{Input}
+  \SetKwInOut{Output}{Output}
+  \Input{$x$ given at precision $O(H)$}
+  \Output{$g(x)$ given at precision $O(H_{\max,n})$}
+
+  \BlankLine
+
+  $x_0 \leftarrow x$;
+
+  \For{$i = 0, \ldots, n-1$}
+    {\textbf{lift} $x_i$ to precision $O(H'_i)$;
+
+     $x_{i+1} \leftarrow \texttt{F}_i(x_i)$;}
+
+  \Return{$x_n + O(H_{\max,n})$};
+\caption{Stabilized version of \texttt{G}}
+\label{algo:stabilizedG}
+\end{algorithm}
+%
 

### 2015-07-22 resultants/resultants.tex
Stabilized version of the subresultant pseudo-remainder sequence algorithm

@@
 \newcommand{\m}{\mathfrak m}
+\newcommand{\calL}{\mathcal L}
 
@@
 \definecolor{purple}{rgb}{0.6,0,0.6}
-\def\todofor#1#2{\ \!\!{\color{purple} {\bf #1}: #2}}
 
@@
 
-  $R_1 \leftarrow A$; $U_1 \leftarrow 1$; $V_1 \leftarrow 0$;
-
-  $R_2 \leftarrow B$; $U_2 \leftarrow 0$; $V_2 \leftarrow 1$;
-
-  $k \leftarrow 2$;
-
-  \While{$R_k \neq 0$}
-    {$Q, R_{k+1} \leftarrow$ quotient and remainder in the Euclidean
-     division of $R_{k-1}$ by $R_k$;
-
-     $U_{k+1} \leftarrow U_{k-1} - Q U_k$;
-
-     $V_{k+1} \leftarrow V_{k-1} - Q V_k$;
-
-     $k \leftarrow k+1$;}
-
-  \Return{$R_{k-1}, U_{k-1}, V_{k-1}$};
+  $S_1 \leftarrow A$; $U_1 \leftarrow 1$; $V_1 \leftarrow 0$
+
+  $S_2 \leftarrow B$; $U_2 \leftarrow 0$; $V_2 \leftarrow 1$
+
+  $k \leftarrow 2$
+
+  \While{$S_k \neq 0$}
+    {$Q, S_{k+1} \leftarrow$ quotient and remainder in the Euclidean
+     division of $S_{k-1}$ by $S_k$
+
+     $U_{k+1} \leftarrow U_{k-1} - Q U_k$
+
+     $V_{k+1} \leftarrow V_{k-1} - Q V_k$
+
+     $k \leftarrow k+1$}
+
+  \Return{$S_{k-1}, U_{k-1}, V_{k-1}$}
 \caption{Extended Euclide's algorithm}
@@
 \end{align*}
-The remainder in the Euclidean division of $A$ by $B$ is $R_3 = A-B$.
+The remainder in the Euclidean division of $A$ by $B$ is $S_3 = A-B$.
 We do not loose precision while performing this substraction and the
 result we get is:
-$$R_3 = 
+$$S_3 = 
       \big(3 + O(2^5)\big) X^4 +
@@
       \big(15 + O(2^5)\big).$$ 
-In order to compute $R_4$, we have now to perform the Euclidean division 
-of $R_2 = B$ by $R_3$. Noting that the leading coefficient of $R_2$ has 
+In order to compute $S_4$, we have now to perform the Euclidean division 
+of $S_2 = B$ by $S_3$. Noting that the leading coefficient of $S_2$ has 
 valuation $0$ and using Eq~\eqref{eq:precadd}--\eqref{eq:precdiv}, we 
 deduce that this operation does not loose precision again. We get:
-$$R_4 = 
+$$S_4 = 
       \big(26 + O(2^5)\big) X^3 +
@@
       \big(16 + O(2^5)\big).$$
-We observe now that the leading coefficient of $R_4$ has valuation $1$. 
+We observe now that the leading coefficient of $S_4$ has valuation $1$. 
 According to Eq.~\eqref{eq:precdiv}, divising by this coefficient ---
 and therefore \emph{a fortioti} computing the euclidean division of 
-$R_3$ by $R_4$ --- will result in loosing at least one digit in
+$S_3$ by $S_4$ --- will result in loosing at least one digit in
 relative precision. The result we find is:
-$$R_5 = 
+$$S_5 = 
       \underbrace{\big(\textstyle \frac 3 4 + O(2^2)\big)}_{\text{rel. prec.} = 4} X^2 +
@@
 Continuing this process, we obtain:
-$$R_6 = 
+$$S_6 = 
       \big(20 + O(2^5)\big) X +
@@
 \quad \text{and} \quad
-R_7 =
+S_7 =
       \textstyle \frac 7 4 + O(2).$$
-The relative precision on the final result $R_7$ is then $3$, which is
+The relative precision on the final result $S_7$ is then $3$, which is
 less than the initial precision which was $5$.
@@
 recursive sequences $(S_i)$ and $(c_i)$ as follows:
-$$\left\{\begin{array}{ll}
+\begin{equation}
+\label{eq:subressequence}
+\left\{\begin{array}{ll}
 S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
@@
 & \text{for } i \geq -1. 
-\end{array}\right.$$
+\end{array}\right.
+\end{equation}
 Here $n_i = \deg S_i$, $\varepsilon_i = n_{i+1} - n_i$ and $s_i$ is the 
@@
 
-  $R_d \leftarrow B$; $r_d \leftarrow 1$;
-
-  $R_{d-1} \leftarrow B-A$;
+  $R_d \leftarrow B$; $r_d \leftarrow 1$
+
+  $R_{d-1} \leftarrow B-A$
 
   \For{$j = (d-1), (d-2), \ldots, 1$}
-    {$r_j \leftarrow$ coefficient in $X^j$ of $R_j$;
-
-     \lIf{$r_j = 0$}{\textbf{raise} NotImplementedError;}
+    {$r_j \leftarrow$ coefficient in $X^j$ of $R_j$
+
+     \lIf{$r_j = 0$}{\textbf{raise} NotImplementedError}
 
@@
 
-  \Return{$R_{d-1}, \ldots, R_0$};
+  \Return{$R_{d-1}, \ldots, R_0$}
 \caption{Subresultant pseudo remainder sequence algorithm}
@@
 interest of using Algorithm~\ref{algo:subres} is to avoid denominators 
-(\emph{i.e.} always work over $\A$ instead $K$) but its use almost does 
-not affect the behaviour of the \emph{relative} precision. We shall 
-explain later (\emph{cf} \S~\ref{subsec:stabilization}) a method for 
-stabilizing Algorithm~\ref{algo:subres}.
-
-\section{Euclide-like algorithms are unstable}
+(\emph{i.e.} always work over $\A$ instead $K$) but this stability
+are comparable --- actually even worse --- than that of 
+Algorithm~\ref{algo:Euclide}.
+
+\begin{ex}
+\label{ex:subres}
+Applying Algorithm~\ref{algo:subres} with the input $(A,B)$ of 
+Example~\ref{ex:Euclide}, we obtain:
+\begin{align*}
+R_4 & = 
+  \big(29 + O(2^5)\big) X^4 + 
+  \big(14 + O(2^5)\big) X^3 + 
+  \big(5 + O(2^5)\big) X^2 + 
+  \big(17 + O(2^5)\big) X + 
+  \big(17 + O(2^5)\big) \\
+R_3 & = 
+  \big(4 + O(2^5)\big) X^3 + 
+  \big(13 + O(2^5)\big) X^2 + 
+  \big(4 + O(2^5)\big) X + 
+  \big(16 + O(2^5)\big) \\
+R_2 & = 
+  \big(5 + O(2^5)\big) X^2 + 
+  \big(20 + O(2^5)\big) X + 
+  O(2^5) \\
+R_1 & = 
+  \big(1 + O(2)\big) X + 
+  \big(1 + O(2)\big) \\
+R_0 & = 1 + O(2)
+\end{align*}
+We observe in particular that the absolute precision on $R_0$ is $1$, 
+although it should be at least $5$ since $R_0$ is given by an integral
+polynomial expression in terms of the coefficients of $A$ and $B$. We
+note moreover that the relative precision on $R_0$ (which is $1$ as
+well) is worse that the relative precision we got on $S_7$ (which was
+$3$) while executing Algorithm~\ref{algo:Euclide} (\emph{cf} 
+Example~\ref{ex:Euclide}).
+\end{ex}
+
+\section{Unstability of Euclide-like algorithms}
 \label{sec:unstable}
 
-\todo{Write a short introduction}
+In this section, we provide strong evidences for explaining the average 
+loss of precision observed while executing Algorithm~\ref{algo:subres} 
+(see Figure~\ref{fig:precision}).
+Concretely, in \S \ref{subsec:lowerbound} we establish\footnote{in a 
+model of precision which is slightly weaker that the usual one; we refer 
+to \S \ref{subsec:lowerbound} for a complete discussion about this.} a 
+lower bound of the losses of precision which depends on extra 
+parameters, that are the valuations of the principal subresultants. 
+The next subsections (\S\S \ref{subsec:proba} and \ref{subsec:proof})
+aim at studying the behaviour of these valuations on random inputs;
+they thus have a strong probabilistic flavour.
+
+\begin{rem}
+The locution \emph{Euclide-like algorithms} (which appears in the 
+title of the Section) refers to the family of
+algorithms computed $\gcd$ or subresultants by means of successive
+Euclidean divisions. We believe that the stability of algorithms in
+this family is comparable since we are precisely loosing precision
+while performing Euclidean divisions.
+Among all algorithms in this family, we chose to concentrale ourselves
+on Algorithm~\ref{algo:subres} because it is simpler due to the fact
+that it only manipulates polynomials with coefficients in $\A$.
+Nevertheless, our method extends to many other Euclide-like algorithms
+including Algorithm~\ref{algo:Euclide}; this is left as an exercice to
+the reader.
+\end{rem}
 
 \subsection{A lower bound on losses of precision}
+\label{subsec:lowerbound}
 
@@
 \subsection{Behaviour on random inputs}
+\label{subsec:proba}
 
@@
 
-\section{The method of adaptative lifts}
+\section{Stabilization}
 \label{sec:stable}
@@
 not loose any significant digit. Of course, the downside is the
-quite bad efficiency.
+rather bad efficiency.
 
@@
 \subsection{Crash course on ultrametric precision}
-
-In this subsection, we briefly report on the results of \cite{padicprec} 
-where the authors draw the lines of a general framework to handle a 
-sharp (often optimal) track of ultrametric precision.
+\label{subsec:crashcourse}
+
+In this subsection, we briefly report on and complete the results of 
+\cite{padicprec} where the authors draw the lines of a general framework 
+to handle a sharp (often optimal) track of ultrametric precision.
 
@@
 tracking precision locally we have discussed earlier to all $g_i$'s. 
-This leads to a \emph{stabilized} version of the algorithm \texttt{G} 
+This leads to a \emph{stabilized} version of the algorithm \texttt{F} 
 whose skeleton is depicted in Algorithm~\ref{algo:stabilizedG}.
@@
 
-  $x_0 \leftarrow x$;
+  $x_0 \leftarrow x$
 
   \For{$i = 0, \ldots, n-1$}
-    {\textbf{lift} $x_i$ to precision $O(H'_i)$;
-
-     $x_{i+1} \leftarrow \texttt{F}_i(x_i)$;}
-
-  \Return{$x_n + O(H_{\max,n})$};
-\caption{Stabilized version of \texttt{G}}
+    {\textbf{lift} $x_i$ to precision $O(H'_i)$
+
+     $x_{i+1} \leftarrow \texttt{G}_i(x_i)$}
+
+  \Return{$x_n + O(H_{\max,n})$}
+\caption{Stabilized version of \texttt{F}}
 \label{algo:stabilizedG}
@@
 
-\subsection{The stabilization process}
+\subsection{Application to subresultants}
 \label{subsec:stabilization}
 
-In order to get a stabilized version of the subresultant 
-pseudo-remainder sequence algorithm, we are going to estimate 
-the loss --- and the gain --- of precision at each step of this 
-algorithm using Proposition \ref{prop:padicprec}. In order to do
-so, we introduce the functions
+We now apply the theory presented in \S \ref{subsec:crashcourse} above 
+to the problem of computing subresultants, \emph{i.e.} the abstract 
+Algorithm \texttt{F} is now instantiated to Algorithm~\ref{algo:subres}. 
+We split this algorithm into steps in the obvious manner, each step 
+corresponding to an iteration of the main loop. We thus consider the
+functions:
+
+$$\begin{array}{rrclc}
+&g_d \, : \, K_d[X] \times K_d[X] &\to &K_d[X] \times K_{\leq d-1}[X] 
+\smallskip \\
+&(A,B) & \mapsto & (B, A-B) \bigskip \\
+\text{and} &g_j \, : \, 
+K_{\leq j+1}[X] \times K_{\leq j}[X]
+& \to & K_{\leq j}[X] \times K_{\leq j-1}[X] \smallskip \\
+&(R_{j+1}, R_j) & \mapsto & (R_j, R_{j-1})
+& 
+\end{array}$$
+where $R_{j-1}$ is defined as usual by
+$R_{j-1} = r_j^2 \cdot r_{j+1}^{-2} \cdot 
+(R_{j+1} \,\%\, R_j)$
+where $r_j$ (resp. $r_{j+1}$) stands for the coefficient of degree $j$ 
+in $R_j$ (resp. of degree $j+1$ in $R_{j+1}$). We remark that $g_j$ is 
+only defined on the subset consisting of pairs $(R_{j+1}, R_j)$ for 
+which $R_{j+1}$ has degree $j+1$; this reflects the fact that 
+Algorithm~\ref{algo:subres} fails on inputs for which at least one 
+principal
+subresultant vanishes. The composite function $f = g_1 \circ \cdots 
+\circ g_d$ (be careful with the order of the indices) models (a slight 
+variant of) Algorithm~\ref{algo:subres}. For all $j$, we set $f_j = 
+g_{j+1} \circ \cdots \circ g_d$ for all $j$. It is the function:
 $$\begin{array}{rcl}
@@
 
-\noindent
-{\bf Differential computation.}
-The first step consists in estimating the differential of $f_j$. 
-Set $\mathcal L = \A_d[X] \times \A_d[X]$ and $\mathcal L_j = 
-\A_{\leq j}[X] \times \A_{\leq j-1}[X]$; they are standard lattices
-in the source and the target of $f_j$ respectively.
+For simplicity, we assume in addition that the precision on the input 
+$(A,B)$ is \emph{flat}, meaning that all coefficients of $A$ and $B$ are 
+known with the same absolute precision $N$. In the language of \S 
+\ref{subsec:crashcourse}, this flat precision corresponds to the lattice 
+$H = \pi^N \calL$ where $\calL = \A_{<d}[X] \times \A_{<d}[X]$ is the 
+standard lattice in $K_d[X] \times K_d[X]$. Following \S 
+\ref{subsec:crashcourse}, our first task consists in finding two 
+lattices $H_{\min,j}$ and $H_{\max,j}$ having the property that 
+$H_{\min,j} \subset f'_j(A,B)(H) \subset H_{\max,j}$.
 
@@
 \subset f'_j(A,B)(\mathcal L) \subset \mathcal L_j$$
-where $r_j$ is the $j$-th principal subresultant of $(A,B)$.
+where $r_j$ is the $j$-th principal subresultant of $(A,B)$ and
+$\calL_j = \A_{\leq j}[X] \times \A_{\leq j-1}[X]$ is the standard
+lattice in $K_{\leq j}[X] \times K_{\leq j-1}[X]$.
 \end{lem}
@@
 Therefore so is $f_j$. Furthermore, a close look at the proof of 
-Proposition \ref{prop:bijection} indicates that a left inverse of $f_i$ 
+Proposition \ref{prop:bijection} indicates that a left inverse of $f_j$ 
 is the function mapping $(S_j, S_{j-1})$ to
@@
 
-\noindent
-{\bf The algorithm.}
-The input of our algorithm is a couple of polynomials $(A,B)$ with 
-coefficients in $\A$ whose coefficients are known at precision 
-$O(\pi^N)$ for some given positive integer $N$. We make the following 
-hypothesis (and refer to \S \ref{subsec:questions} for a brief 
-discussion about it):
+Lemma \ref{lem:diff} ensures that one can safely take $H_{\min,j} = 
+r_j^2 \cdot \pi^N \calL_j$ and $H_{\max,j} = \pi^N \calL_j$. It finally 
+remains to construct the lattice $H'_j \subset K_{\leq j}[X] \times 
+K_{\leq j-1}[X]$. For this, we remark that a naive track of precision 
+leads to a loss of at most $2 \cdot \val(r_{j+1})$ digits while 
+executing the step $\texttt{G}_j$ (see also proof of Proposition 
+\ref{prop:precEuclide} for similar considerations). Therefore, one can 
+take $H'_j = r_j^2 r_{j+1}^2 \pi^N \calL_j$.
+
+\begin{algorithm}
+  \SetKwInOut{Input}{Input}
+  \SetKwInOut{Output}{Output}
+  \Input{Two polynomials $A, B \in K_d[X]$ given at flat precision $O(\pi^n)$}
+  \Output{The sequence of subresultants of $A$ and $B$ given at flat precision $O(\pi^n)$}
+
+  \BlankLine
+
+  $R_d \leftarrow B$; $r_d \leftarrow 1$
+
+  $R_{d-1} \leftarrow B-A$
+
+  \For{$j = (d-1), (d-2), \ldots, 1$}
+    {$r_j \leftarrow$ coefficient in $X^j$ of $R_j$
+
+     \lIf{$v_j \geq \frac N 2$}{\textbf{raise} NotImplementedError}
+
+     \textbf{lift} $(R_{j+1}, R_j)$ at precision $O(\pi^{N+2\val(r_j) + 2\val(r_{j+1}))})$
+
+     $R_{j-1} \leftarrow \texttt{prem}(R_{j+1}, R_j) / r_{j+1}^2$
+    }
+
+  \Return{$R_{d-1} + O(\pi^N), \ldots, R_0 + O(\pi^N)$}
+\caption{Stabilized version of Algorithm~\ref{algo:subres}}
+\label{algo:stabsubres}
+\end{algorithm}
+
+\begin{prop}
+Algorithm \ref{algo:stabsubres} computes all subresultants of
+$(A,B)$ at precision $O(\pi^N)$ under the following 
+assumption\footnote{If this assumption is not fullfiled, the
+algorithms fails and returns an error.}
 
@@
 \begin{tabular}{rl}
-& (a)~$A$ and $B$ are monic of the same degree $d$ \\
-{\bf (H)}: & (b)~all principal subresultants of $(A,B)$ 
+{\bf (H)}: & all principal subresultants of $(A,B)$ 
 do not vanish modulo $\pi^{N/2}$.
@@
 \noindent
-The subresultants $R_j = \Res_j(A,B)$ can then be computed recursively:
-setting $R_d = B$, we have $R_{d-1} = A-B$ and the recurrence
-$R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})$.
-Moreover Lemma \ref{lem:diff} and Proposition \ref{prop:padicprec} imply
-that any perturbation of $(R_j, R_{j-1})$ by an element of $\pi^N r_j^2 
-\mathcal L_j$ will not affect the values of the next $R_k$'s modulo 
-$\pi^N$. Indeed, given $k > j$, let $f_{j,k} : (R_j, R_{j-1}) \mapsto 
-(R_k, R_{k+1})$ be the function modeling the above recurrence. We have
-$f_k = f_{j,k} \circ f_j$ and thus we can write:
-\begin{align*}
-f_{j,k}\big((R_j, R_{j-1}) + \pi^N r_j^2 \mathcal L_j\big) 
-& \subset f_k\big((A,B) + \pi^N \mathcal L\big) \\
-& \subset (R_k, R_{k+1}) + \pi^N \mathcal L_k.
-\end{align*}
-This suggests the following algorithm.
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm 1:} {\tt subresultants\_stable}
-
-\noindent{\bf Input:} an integer $N$, a couple $(A,B)$ satisfying {\bf (H)}
-
-\noindent{\bf Output:} the subresultants of $(A,B)$ at precision $O(\pi^N)$
-
-\smallskip\noindent 1.\ Set $R_d = B$, $v_d = 0$ and $R_{d-1} = A-B$
-
-\smallskip\noindent 2.\ {\bf for} $j=d,\dots,2$
-
-\smallskip\noindent 3.\ \hspace{0.3cm}{\bf compute} $v_{j-1} = \val(\lc(R_{j-1}))$
-
-\smallskip\noindent 4.\ \hspace{0.3cm}{\bf lift} $(R_j, R_{j-1})$ at precision $O(\pi^{N+2(v_j + v_{j-1})})$
-
-\smallskip\noindent 5.\ \hspace{0.3cm}{\bf compute} $R_{j-2} = r_{j-1}^2 \cdot r_j^{-2} \cdot (R_j \,\%\, R_{j-1})$ at precision $O(\pi^{N+2 v_{j-1}})$
-
-\smallskip\noindent 6.\ {\bf return} $R_{d-1} + O(\pi^N), \ldots,
-R_0 + O(\pi^N)$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\begin{prop}
-Algorithm 1 is correct and runs in
-$$O(d^2 \cdot \M(N + \max(V_0, \ldots, V_{d-1}))$$
-bit operations where $V_j$ denotes the valuation of $r_j$ and 
-$\M(n)$ is the number of bit operations needed to perform an 
-arithmetic operation (addition, product) in $\A$ at precision 
-$O(\pi^n)$.
+It runs in $O(d^2 \cdot \M(N + \max(V_0, \ldots, V_{d-1}))$ bit 
+operations where $V_j$ denotes the valuation of $r_j$ and $\M(n)$ is the 
+number of bit operations needed to perform an arithmetic operation 
+(addition, product) in $\A$ at precision $O(\pi^n)$.
 \end{prop}
@@
 \begin{proof}
-Regarding to the correctness, it remains only to justify that $R_{j-2}$ 
-can be computed at the given precision in line 5 but this follows 
-from a naive track of precision (\emph{cf} proof of Proposition
-\ref{prop:precEuclide} for a similar computation).
+Correctness has been already proved (the assumption {\bf (H)} ensures
+that Proposition~\ref{prop:padicprec} applies to each $g_j$).
 As usual Euclide's algorithm, Algorithm 1 requires $O(d^2)$ operations 
@@
 at which we are computing is upper bounded by $N + 2 \max(V_0, \ldots, 
-V_{d-1})$. The proposition follows.
+V_{d-1})$. This justifies the announced complexity.
 \end{proof}
@@
 
-\subsection{Open questions}
+To conclude with, let us comment on briefly the hypothesis {\bf (H)}. We 
+first remark that it is satisfied with high probability if $N$ is large 
+compared to $2 \cdot \log_d p$. Thus, replacing eventually $N$ by $3 
+\cdot \log_d p$ (which does not affect the complexity), the 
+assumption~(b) is harmless on average --- but maybe not on particularly 
+bad instances. Let us also underline that, if we are just interested in 
+computing the $j$-th subresultant for a particular $j$, then we just 
+need to assume the non-vanishing of the principal subresultants in the 
+range $[j{+}1, d{-}1]$.
+
+\subsubsection*{Open questions}
 \label{subsec:questions}
 
-Of course, the first question we would like to raise is to possibility 
-to drop the hypothesis {\bf (H)}. It is actually rather easy to relax
-the part~(a) of the hypothesis, replacing it by:
-
-\medskip
-
-(a')~$\lc(B)$ is invertible in $\A$ and $\deg A \geq \deg B$.
-
-\medskip
-
-\noindent
-Assuming only this weaker assumption, Algorithm 1 still works except 
-that, in line 1, we have to set
-$$R_{d-1} = (-1)^{\deg A - \deg B} (A \,\%\, B)$$
-(instead of $R_{d-1} = A-B$).
-We remark that this value can be computed at precision $O(\pi^N)$
-because the leading coefficient of $B$ is assumed to be invertible.
-Dropping entirely~(a') seems nevertheless more difficult. Let us 
-move to assumption~(b) and first remark that it is satisfied with high 
-probability if $N$ is large compared to $2 \cdot \log_d p$. Thus, 
-replacing eventually $N$ by $3 \cdot \log_d p$ (which does not affect 
-the complexity), the assumption~(b) is harmless on average --- but maybe
-not on particularly bad instances. We finally underline that, if we are
-just interested in computing the $j$-th subresultant for a particular
-$j$, then we just need to assume the non-vanishing of the principal
-subresultants in the range $[j{+}1, d{-}1]$.
+The first hypothesis we would like to relax is of course {\bf (H)}. 
+Actually, it seems quite plausible that one can produce a stabilized 
+version of the ``complete''\footnote{\emph{I.e.} dealing with abnormal 
+sequences as well.} subresultant pseudo-remainder sequence algorithm 
+following the same strategy. Nevertheless, this extension will not 
+completely straightforward because one will need to understand precisely 
+how the coefficients $c_i$'s (appearing in Eq.~\ref{eq:subressequence}) 
+alter the behaviour of the precision.
+
+As it was presented, Algorithm~\ref{algo:stabsubres} only accepts inputs 
+consisting of a pair of \emph{monic} polynomials having the same degree. 
+It is actually not difficult to make it work with all couples of 
+polynomials $(A,B)$ such that $\lc(B)$ is invertible in $\A$ and $\deg A 
+\geq \deg B$. Indeed, it is enough for that to replace line 2 by:
+$$R_{d-1} \leftarrow (-1)^{\deg A - \deg B} (A \,\%\, B).$$
+We remark that this can be computed at precision $O(\pi^N)$ thanks to 
+the assumption on the leading coefficient of $B$.
+However, writing an extension of Algorithm~\ref{algo:stabsubres} that 
+accepts all inputs seems much more tricky and this is the first question 
+we would like to address.
 

### 2015-07-22 resultants/resultants.tex
More details in the proof of Proposition 3.7

@@
 subresultants. For a more complete exposition, we refer to \cite[\S
-3.3]{cohen} and \cite[\S 4.1]{winkler}.
+4.2]{real}, \cite[\S 3.3]{cohen} and \cite[\S 4.1]{winkler}.
 Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
@@
 and Proposition \ref{prop:prem} now simply states that $R_j = S_{d_B-j}$ 
-in this case.
+in this case. In other words, still assuming that all principal 
+subresultants do not vanish, the sequence of subresultants obeys to 
+the recurrence:
+\begin{align}
+\label{eq:recRj}
+R_{d+1} = A, \quad R_d = B, & \qquad
+R_{j-1} = r_j^2 \cdot r_{j+1}^{-2}
+\cdot (R_{j+1} \,\%\, R_j) 
+\end{align}
+where $r_j$ is the leading coefficient of $R_j$ for $j \leq d$ and $r_{d+1}
+= 1$ by convention. Moreover, a similar 
+recurrence exists for cofactors as well:
+\begin{align}
+U_{d+1} = 1, \quad U_d = 0, & \qquad
+U_{j-1} = r_j^2 \cdot r_{j+1}^{-2}
+\cdot (U_{j+1} - Q_j U_j) \label{eq:recUj} \\
+V_{d+1} = 0, \quad U_d = 1, &\qquad
+V_{j-1} = r_j^2 \cdot r_{j+1}^{-2}
+\cdot (V_{j+1} - Q_j V_j) \label{eq:recVj}
+\end{align}
+where $Q_j$ is quotient in the Euclidean division of $R_{j+1}$ by $R_j$.
+Proposition \ref{prop:prem} of course yields an algorithm for computing 
+subresultants. In the normal case and assuming further for simplicity 
+that the input polynomials are monic of same degree, it is 
+Algorithm~\ref{algo:subres}, which uses the primitive 
+\texttt{prem} which computes the pseudo-remainder: if $A$ and $B$ are 
+two polynomials, by definition $\texttt{prem}(A,B) = \lc(B)^{\deg B-\deg 
+A+1} (A \% B)$ where $\lc(B)$ denotes the leading coefficient of $B$.
 
@@
 
-Proposition \ref{prop:prem} of course yields an algorithm for computing 
-subresultants. In the normal case and assuming further for simplicity 
-that the input polynomials are monic of same degree, it is 
-Algorithm~\ref{algo:subres}. Note that this algorithm uses the primitive 
-\texttt{prem} which computes the pseudo-remainder: if $A$ and $B$ are 
-two polynomials, by definition $\texttt{prem}(A,B) = \lc(B)^{\deg B-\deg 
-A+1} (A \% B)$ where $\lc(B)$ denotes the leading coefficient of $B$. 
 Unfortunately, when working over a complete discrete valuation field 
-$K$, the stability of Algorithm~\ref{algo:subres} is comparable --- and 
-actually even a bit worse --- than standard Euclide algorithm. The main 
-interest of using Algorithm~\ref{algo:subres} is to avoid denominators 
-(\emph{i.e.} always work over $\A$ instead $K$) but this stability
-are comparable --- actually even worse --- than that of 
-Algorithm~\ref{algo:Euclide}.
+$K$, the stability of Algorithm~\ref{algo:subres} is as bad as that of 
+standard Euclide algorithm. The main interest of using 
+Algorithm~\ref{algo:subres} is to avoid denominators, \emph{i.e.} we
+always work over $\A$ instead $K$.
 
@@
 \begin{proof}
-By functoriality of subresultants, we may assume that $\ring = 
-\Z[a_0, \ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$
-are the two generic monic polynomials 
-$A = X^d + \sum_{i=0} a_i X^i$ and
-$B = X^d + \sum_{i=0} b_i X^i$.
-All relations then follow from the extended subresultants 
-pseudo-remainder sequence algorithm (noting in addition that all
-principal subresultants are nonzero). \todo{Give details.}
+
+By functoriality of subresultants, we may assume that $\ring = \Z[a_0, 
+\ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$ are the two 
+generic monic polynomials $A = X^d + \sum_{i=0} a_i X^i$ and $B = X^d + 
+\sum_{i=0} b_i X^i$. Under this additional assumption, all principal
+subresultant are nonzero. Therefore, the sequences $(R_j)_j$, $(U_j)_j$
+and $(V_j)_j$ satisfy the recurrences \eqref{eq:recRj}--\eqref{eq:recVj}.
+The two first announced relations follow easily. Let now focus on the
+third one. We set $\tilde R_j = R_j$ and $\tilde R_k = r_j^{2(j-k-1)} 
+\: R_k$ for $k < j$. An easy decreasing induction on $k$ shows that this 
+sequence obeys to the recurrence:
+$$\tilde R_{k-1} = \tilde r_k^2 \cdot \tilde r_{k+1}^{-2} \cdot 
+(\tilde R_{k+1} \,\%\, \tilde R_k)$$
+where $\tilde r_j = 1$ and $\tilde r_k$ is the coefficient of $\tilde 
+R_k$ of degree $k$ for all $k < j$. Comparing with \eqref{eq:recRj},
+this implies that $\tilde R_k$ is the $k$-th subresultant of the pair 
+$(R_j, R_{j-1})$ and we are done. The last equality is proved in a 
+similar fashion using Eq.~\eqref{eq:recUj}.
 \end{proof}
@@
   Cambridge University Press (1976)
+\bibitem{real}
+  S.~Basu, R.~Pollack, M.-F.~Roy
+  \emph{Algorithms in Real Algebraic Geometry},
+  Springer-Verlag (2008), second edition
 \bibitem{pari}
@@
   Springer Wien New Work (1996)
-\bibitem{}
 \end{thebibliography}

### 2015-07-23 resultants/resultants.tex
Minor changes

@@
 precision is attached to any $p$-adic variable and this precision is 
-updated after each basic arithmetic operation (addition and 
-multiplication). This way of tracking precision can be seen as the 
-analogue of the arithmetic intervals in the real setting. We refer
-to \S \ref{subsubsec:computDVR} for more details.
+updated after each basic arithmetic operation. This way of tracking 
+precision can be seen as the analogue of the arithmetic intervals in the 
+real setting. We refer to \S \ref{subsubsec:computDVR} for more details.
 
@@
 when it is called on random inputs which are monic $2$-adic 
-polynomials of fixed degree (see also Example \ref{ex:Euclide}). Looking at the last line, we see that 
-extended Euclide's algorithm outputs the B\'ezout coefficients of two 
-monic $2$-adic polynomials of degree $100$ with an average loss of 
-$161.1$ significant digits by coefficient whereas a stable 
-algorithm should only loose $3.2$ digits on average. This 
-``theoretical'' loss is computed as the double of the valuation of the 
-resultant. Indeed Cramer-like formulae imply that B\'ezout coefficients 
-can be computed by performing a unique division by the resultant, 
-inducing then only the aforementioned loss of precision. Examining the 
-table a bit more, we furthermore observe that the ``practical'' loss of 
-precision due to Euclide's algorithm seems to grow linearly with respect 
-to the degree of the input polynomials whereas the ``theoretical'' loss 
-seems to be independant of it. In other words, the instability of 
-Euclide's algorithm is becoming more and more critical when the degree 
-of the input increases.
+polynomials of fixed degree (see also Example \ref{ex:Euclide}). 
 %
@@
 \end{algorithm}
+%
+Looking at the last line, we see that 
+extended Euclide's algorithm outputs the B\'ezout coefficients of two 
+monic $2$-adic polynomials of degree $100$ with an average loss of 
+$160$ significant digits by coefficient whereas a stable 
+algorithm should only loose $3.2$ digits on average. This 
+``theoretical'' loss is computed as the double of the valuation of the 
+resultant. Indeed Cramer-like formulae imply that B\'ezout coefficients 
+can be computed by performing a unique division by the resultant, 
+inducing then only the aforementioned loss of precision (see 
+\S \ref{subsubsec:computDVR}, Eq.~\eqref{eq:precdiv} for a full
+justification). Examining the 
+table a bit more, we observe that the ``practical'' loss of 
+precision due to Euclide's algorithm seems to grow linearly with respect 
+to the degree of the input polynomials whereas the ``theoretical'' loss 
+seems to be independant of it. In other words, the instability of 
+Euclide's algorithm is becoming more and more critical when the degree 
+of the input increases.
 
@@
 out these unexpected losses of precision. Making slight changes to the 
-standard subresultants pseudo-remainder sequence algorithm and using in 
+standard subresultant pseudo-remainder sequence algorithm and using in 
 an essential way the results of \cite{padicprec}, we manage to design a stable 
 algorithm for computing all subresultants of two monic polynomials over 
-$\Zp$ (satisfying an additional mild assumption). This basically allows 
+$\Zp$ (satisfying an additional assumption). This basically allows 
 to stably compute \textsc{gcd}s assuming that the degree of the 
@@
 We recall that $\A$ admits a unique maximal ideal $\m$, consisting of 
-elements of positive valuation. Furthermore, this ideal is principal and 
+elements of positive valuation. This ideal is principal and 
 generated by any element of valuation $1$. Such an element is called a 
@@
 structure may vary.
+
 Let $K$ denote the fraction field of $\A$. The valuation $v$ extends 
 uniquely to $K$ by letting $\val(\frac x y) = \val(x) - \val(y)$. 
-Moreover, it follows from axiom 4 that $K$ is obtain from $\A$ by 
+Moreover, it follows from axiom 4 that $K$ is obtained from $\A$ by 
 inverting $\pi$. Thus, any element of $K$ can be uniquely written as an 
@@
 according to the rules we all studied at school (and therefore taking
-care of caries). Similarly to the equal characteristic case, we prove
+care of carries). Similarly to the equal characteristic case, we prove
 that $\Zp$ is complete with respect to the distance associated to the 
@@
 
+We now go back to a general complete discrete valuation ring $\A$, 
+whose fraction field is still denoted by $K$.
 The memory of a computer being necessarily finite, it is not possible to 
@@
 \noindent
+with the convention that $\val(a) = N_a$ (resp. $\val(b) = N_b$) if all
+known digits of $a$ (resp. $b$) are zero.
 Combining these formulas, one can track the precision while executing 
@@
 arithmetic} over the reals. Many usual softwares (as \textsc{sage}, 
-\textsc{magma}) implement $p$-adic numbers this way. We shall see later
-that this often results in overestimating the losses of precision.
+\textsc{magma}) implement $p$-adic numbers and formal series this way. 
+We shall see later that this often results in overestimating the losses 
+of precision.
 
@@
 As an illustration, let us examine the behaviour of the precision
-on the sequence $(R_i)$ while Algorithm~\ref{algo:Euclide} is
-executed with the input:
+on the sequence $(R_i)$ while executing Algorithm~\ref{algo:Euclide} 
+with the input:
 \begin{align*}
@@
 The remainder in the Euclidean division of $A$ by $B$ is $S_3 = A-B$.
-We do not loose precision while performing this substraction and the
+According to Eq.~\eqref{eq:precsub},
+we do not loose precision while performing this substraction and the
 result we get is:
@@
 there exists a standard way to recover continuity in this context: it 
-consists in replacing \textsc{gcd}s by the so-called subresultants which 
+consists in replacing \textsc{gcd}s by subresultants which 
 are playing an analoguous role. For this reason, in what follows, we will 
@@
 subresultants. For a more complete exposition, we refer to \cite[\S 
-4.2]{real}, \cite[\S 3.3]{cohen} and \cite[\S 4.1]{winkler}.
-Let $A$ and $B$ be two polynomials with coefficients in $\ring$ and 
-$d_A$ and $d_B$ be two integers greater than or equal to the degree of 
-$A$ and $B$ respectively. We consider the Sylvester application:
+4.2]{real}, \cite[\S 3.3]{cohen} and \cite[\S 4.1]{winkler}. Let $\ring$ 
+be an arbitrary ring and let $A$ and $B$ be two polynomials with 
+coefficients in $\ring$. We pick in addition two integers $d_A$ and 
+$d_B$ greater than or equal to the degree of $A$ and $B$ respectively. 
+We consider the Sylvester application:
 $$\begin{array}{rcl}
@@
 \end{array}$$
+where $\ring_{<d}[X]$ refers to the finite free $\ring$-module of
+rank $d$ consisting of polynomials over $\ring$ of degree strictly 
+less than $d$.
 The Sylvester matrix is the matrix of $\psi$ in the canonical ordered
 basis, which are
-$$\begin{array}{rrl}
-& ((X^{d_B-1},0), \ldots, (X,0), (1,0), (0,X^{d_A-1}), \ldots, (0,1))
-& \text{for the source} \smallskip \\
-\text{and} & (X^{d_A+d_B-1}, \ldots, X, 1) &
+$$\begin{array}{rl}
+((X^{d_B-1},0), \ldots, (X,0), (1,0), (0,X^{d_A-1}), \ldots, (0,1)) &
+\text{for the source} \smallskip \\
+\text{and} \quad (X^{d_A+d_B-1}, \ldots, X, 1) &
 \text{for the target.}
@@
 Using Cramer formulae, we can build polynomials $U^{d_A, d_B}(A,B) \in
-\ring_{< d_B}[X]$ and $V^{d_A, d_B}(A,B) \in \ring_{< d_A}[X]$ whose 
-coefficients are, up to a sing, maximal minors of the Sylvester matrix 
-such that:
-$$A \cdot U^{d_A, d_B}(A,B) + B \cdot V^{d_A, d_B}(A,B) = 
-\Res^{d_A, d_B}(A,B).$$
-There polynomials are called the \emph{cofactors} of $A$ and $B$
+\ring_{< d_B}[X]$ and $V^{d_A, d_B}(A,B) \in \ring_{< d_A}[X]$ 
+satisfying the two following conditions:
+\begin{enumerate}[i)]
+\item their coefficients are, up to a sing, maximal minors of the 
+Sylvester matrix, and
+\item $A \cdot U^{d_A, d_B}(A,B) + B \cdot V^{d_A, d_B}(A,B) = 
+\Res^{d_A, d_B}(A,B)$.
+\end{enumerate}
+These polynomials are called the \emph{cofactors} of $A$ and $B$
 (computed in degree $d_A, d_B$).
@@
 \ring_{< d_B-j}[X]$ and $V_j^{d_A, d_B}(A,B) \in \ring_{< d_A-j}[X]$
-whose coefficients are, up to a sign, minors of the Sylvester 
-matrix\footnote{Indeed, observe that the matrix of $\psi_j$ is a
-submatrix of the Sylvester matrix.} such that:
-$$A \cdot U_j^{d_A, d_B}(A,B) + B \cdot V_j^{d_A, d_B}(A,B) 
-\equiv \det \psi_j \pmod {\ring_{< j}[X]}.$$
+such that:
+\begin{enumerate}[i)]
+\item their coefficients are, up to a sing, maximal minors of the 
+Sylvester matrix\footnote{Indeed, observe that the matrix of $\psi_j$ is a
+submatrix of the Sylvester matrix.}, and
+\item $A \cdot U_j^{d_A, d_B}(A,B) + B \cdot V_j^{d_A, d_B}(A,B) 
+\equiv \det \psi_j \pmod {\ring_{< j}[X]}$.
+\end{enumerate}
 We set $R_j^{d_A, d_B}(A,B) = A \cdot U_j^{d_A, d_B}(A,B) + B \cdot 
@@
 
+We recall the following very classical result.
+
 \begin{theo}
@@
 pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
-$B$ as above and assume that $d_A = \deg A$. Denoting by $(P \,\%\, Q)$ 
+$B$ as above. Denoting by $(P \,\%\, Q)$ 
 the remainder in the Euclidean division of $P$ by $Q$, we define two 
@@
 additional assumption, one can prove that the degrees of the $S_i$'s 
-decrease by one at each step, meaning that we $\deg S_i = d_B - i$ for 
+decrease by one at each step; in other words, $\deg S_i = d_B - i$ for 
 all $i$. The sequence $(S_i)$ then stops at $i = d_B$. Moreover, the 
-$\varepsilon_i$'s and the $c_i$'s are now all ``trivial'' since we have 
+$\varepsilon_i$'s and the $c_i$'s are now all ``trivial'': we have 
 $\varepsilon_i = 1$ and $c_i = s_i$ for all $i$. The recurrence formula 
@@
 \cdot (S_{i-1} \,\%\, S_i) \quad \text{for } i \geq 1.$$
-and Proposition \ref{prop:prem} now simply states that $R_j = S_{d_B-j}$ 
-in this case. In other words, still assuming that all principal 
-subresultants do not vanish, the sequence of subresultants obeys to 
-the recurrence:
+and Proposition \ref{prop:prem} now simply states that $R_j = 
+S_{d_B-j}$. In other words, still assuming that all principal 
+subresultants do not vanish, the sequence of subresultants obeys to the 
+recurrence:
 \begin{align}
@@
 where $Q_j$ is quotient in the Euclidean division of $R_{j+1}$ by $R_j$.
+
 Proposition \ref{prop:prem} of course yields an algorithm for computing 
@@
 that the input polynomials are monic of same degree, it is 
-Algorithm~\ref{algo:subres}, which uses the primitive 
-\texttt{prem} which computes the pseudo-remainder: if $A$ and $B$ are 
-two polynomials, by definition $\texttt{prem}(A,B) = \lc(B)^{\deg B-\deg 
-A+1} (A \% B)$ where $\lc(B)$ denotes the leading coefficient of $B$.
+Algorithm~\ref{algo:subres}, which uses the primitive \texttt{prem} for 
+computing pseudo-remainders. We recall that the pseudo-remainder of the 
+division of $A$ by $B$ is the polynomial $\texttt{prem}(A,B)$ defined by 
+$\texttt{prem}(A,B) = \lc(B)^{\deg B-\deg A+1} (A \% B)$ where $\lc(B)$ 
+denotes the leading coefficient of $B$.
 
@@
 
-Unfortunately, when working over a complete discrete valuation field 
+Unfortunately, while working over a complete discrete valuation field 
 $K$, the stability of Algorithm~\ref{algo:subres} is as bad as that of 
-standard Euclide algorithm. The main interest of using 
-Algorithm~\ref{algo:subres} is to avoid denominators, \emph{i.e.} we
-always work over $\A$ instead $K$.
+standard Euclide algorithm. The use of Algorithm~\ref{algo:subres} is 
+interesting because it avoids denominators (\emph{i.e.} we always work 
+over $\A$ instead $K$) but it does not improve the stability.
 
@@
 In this section, we provide strong evidences for explaining the average 
-loss of precision observed while executing Algorithm~\ref{algo:subres} 
-(see Figure~\ref{fig:precision}).
+loss of precision observed while executing Algorithm~\ref{algo:subres}.
 Concretely, in \S \ref{subsec:lowerbound} we establish\footnote{in a 
@@
 to \S \ref{subsec:lowerbound} for a complete discussion about this.} a 
-lower bound of the losses of precision which depends on extra 
+lower bound on the losses of precision which depends on extra 
 parameters, that are the valuations of the principal subresultants. 
@@
 title of the Section) refers to the family of
-algorithms computed $\gcd$ or subresultants by means of successive
-Euclidean divisions. We believe that the stability of algorithms in
+algorithms computed \textsc{gcd}s or subresultants by means of successive
+Euclidean divisions. We believe that the stability of all algorithms in
 this family is comparable since we are precisely loosing precision
@@
 Nevertheless, our method extends to many other Euclide-like algorithms
-including Algorithm~\ref{algo:Euclide}; this is left as an exercice to
+including Algorithm~\ref{algo:Euclide}; this extension is left as an exercice to
 the reader.
@@
 In this subsection, we estimate the loss of precision if we compute the 
-$R_j$'s using the recurrence formula given by Proposition \ref{prop:prem}. 
+$R_j$'s using the recurrence \eqref{eq:recRj}.
 In what follows, we are going to use a \emph{flat precision model}: this 
-means that a polynomial $P(X)$ is represented as:
+means that a polynomial $P(X)$ is internally represented as:
 $$P(X) = \sum_{i=1}^n a_i X^i + O(\pi^N)
@@
 computing a polynomial, the software truncates the precision on each
-coefficient to the smallest precision. 
+coefficient to the smallest one.
 One can argue that this assumption is too strong (compared to usual
@@
 \label{prop:precEuclide}
-Let $A$ and $B$ as above. We assume that all principal subresultants of 
-$(A,B)$ do not vanish. Then, in the flat precision model, the 
-subresultant pseudo-remainder sequence computes $R_j$ at precision 
-$O(\pi^{N_j})$ with:
+Let $A$ and $B$ as above. Either Algorithm~\ref{algo:subres} fails or
+it outputs the subresultants $R_j$'s at precision $O(\pi^{N_j})$ with:
 $$N_j \leq N + V_{j+1} - 2 \cdot (\delta_{j+1} + \delta_{j+2} + \cdots
@@
 \begin{proof}
-By our assumptions, the subresultant pseudo-remainder sequence 
-is defined by the recurrence:
-$$R_{j-1} = r_j^2 \cdot r_{j+1}^{-2} \cdot 
-(R_{j+1} \,\%\, R_j).$$
 Using that $R_{j+1}$ and $R_j$ have the expected degrees, the
@@
 where $s$ is the coefficient of degree $j$ of $S$. Let us first estimate 
-the precision of $S$. Recall that the relative precision is defined 
-as the difference between the (absolute) precision and the valuation: 
-intuitively, it is the number of significant digits. It is easy to check 
-that the relative precision of a product is the minimum of those of the 
-factors. 
-Thus the relation precision of $r_{j+1} \cdot r_j^{-1} \cdot R_j$ is
-$\min(N_{j+1} - V_{j+1}, N_j - V_j)$. Its absolute precision is then
+the precision of $S$. Using \eqref{eq:precmul}--\eqref{eq:precdiv}, we 
+find that the computed relation precision on $r_{j+1} \cdot r_j^{-1} 
+\cdot R_j$ is $\min(N_{j+1} - V_{j+1}, N_j - V_j)$. The absolute
+precision of this value is then
 $M = \min(N_{j+1} - \delta_j, N_j - \delta_j + V_{j+1} - V_j)$. This
@@
 
-Proposition \ref{prop:precEuclide} gives an estimation of the loss
-of precision in Euclide-like algorithms in terms of the quantities
-$V_j$ and $\delta_j$. It is nevertheless \emph{a priori} not clear
-how large these numbers are. The aim of this paragraph is to compute
-their order of magnitude when $A$ and $B$ are picked randomly among
-the set of monic polynomials of degree $d$ with coefficients in $\A$.
-In what follows, we assume that the residue field of $\A$ is finite
-and we use the letter $q$ to denote its cardinality.
+Proposition \ref{prop:precEuclide} gives an estimation of the loss of 
+precision in Euclide-like algorithms in terms of the quantities $V_j$ 
+and $\delta_j$. It is nevertheless \emph{a priori} not clear how large 
+these numbers are. The aim of this paragraph is to compute their order 
+of magnitude when $A$ and $B$ are picked randomly among the set of monic 
+polynomials of degree $d$ with coefficients in $\A$. In what follows, we 
+assume that the residue field $k = \A/\pi\A$ is finite and we use the 
+letter $q$ to denote its cardinality.
 
@@
 
-The proof of these two Theorem is given in \S \ref{subsec:proof}. 
+The proof of these two theorems will be given in \S \ref{subsec:proof}. 
 We now derive some consequences. Let $\sigma$ denote the following
@@
 \begin{proof}
-By Theorem \ref{th:lawVj}, the expected value of $V_j$ is equal to 
-$\sum_{i=0}^d \E[Z_i]$ with $Z_i = \min(X_{j-i}, \ldots, X_{j+i})$ ($j$ 
+By Theorem \ref{th:lawVj}, we have $\E[V_j] =
+\sum_{i=0}^d \E[Z_i]$ with $Z_i = \min(X_{j-i}, \ldots, X_{j+i})$ ($j$ 
 is fixed during all the proof). Our conventions imply that $Z_i$ 
@@
 because $\frac 1 {q-1}$ is the first summand in the expansion of
-$\E[V_j]$. The upper bound is obtained as follows:
+$\E[V_j]$. The upper bound is derived as follows:
 $$\E[V_j] < \sum_{i=0}^\infty \frac 1{q^i - 1}
@@
 the formula
-$\Var(V_j) = \sum_{1 \leq i,i' \leq d-j} \Cov(Z_i, Z_{i'})$.
+$$\Var(V_j) = \sum_{1 \leq i,i' \leq d-j} \Cov(Z_i, Z_{i'}).$$
 Moreover, given $X$ and $X'$ two independant variables having geometric
@@
 
-Let $(Z_i)_{i \geq 0}$ be a countable family 
+We now prove~(4). Let $(Z_i)_{i \geq 0}$ be a countable family 
 of independant random variable having all geometric distribution of 
@@
 \geq m_i$ for all $i \leq \ell$}.
-Up to a neglectable subset, $E_m$ contains the event $\{ Z \geq m\}$.
-Thus $\P[Z \geq m] \leq \P[E_m]$. Furthermore, the probability of $E_m$
-is bounded from above by $\sum \prod_{i=1}^\ell \P[X_1 \geq m_i]$ where
-the sum runs over all partitions $(m_1, \ldots, m_\ell)$ of $m$. 
-Replacing $\P[X_1 \geq m_i]$ by $q^{-m_i}$, we get
+Up to a measure-zero subset, $E_m$ contains the event $\{ Z \geq m\}$.
+We obtain this way:
+$$\P[Z \geq m] \leq \P[E_m] \leq
+\sum \, \prod_{i=1}^\ell \P[X_1 \geq m_i]$$
+where the latter sum runs over all partitions $(m_1, \ldots, m_\ell)$ 
+of $m$. Replacing $\P[X_1 \geq m_i]$ by $q^{-m_i}$, we get
 $\P[E_m] \leq p(m) \cdot q^{-m}$
@@
 We now derive~(5) by a standard argument. It follows from~(4) that
-$\P[\max(V_0, \ldots, V_{d-1})] \leq d \cdot q^{-m + c\sqrt m}$
+$$\P[\max(V_0, \ldots, V_{d-1})] \leq d \cdot q^{-m + c\sqrt m}$$
 for some constant $c$. Therefore:
@@
 cofactors. During all these preliminaries, we work over an arbitrary
-base $\ring$.
+base ring $\ring$.
 
@@
 \end{itemize}
-Moreover $r_j$ depends only on the $2j$ coefficients of highest 
+Moreover $r_j$ depends only on the $2(d-j)-1$ coefficients of highest 
 degree of $A$ and $B$.
@@
 subresultant are nonzero. Therefore, the sequences $(R_j)_j$, $(U_j)_j$
-and $(V_j)_j$ satisfy the recurrences \eqref{eq:recRj}--\eqref{eq:recVj}.
+and $(V_j)_j$ are given by the recurrences \eqref{eq:recRj}--\eqref{eq:recVj}.
 The two first announced relations follow easily. Let now focus on the
@@
 this implies that $\tilde R_k$ is the $k$-th subresultant of the pair 
-$(R_j, R_{j-1})$ and we are done. The last equality is proved in a 
-similar fashion using Eq.~\eqref{eq:recUj}.
+$(R_j, R_{j-1})$ and we are done. The fourth equality and the last
+statement are proved in a similar fashion.
 \end{proof}
@@
 \big(\ring_{\leq j}[X]\big) \times \big(\ring_{\leq j-1}[X]\big)$$
-gathering the quadruples $(\mathcal U_j, \mathcal U_{j-1}, \mathcal
+consisting of the quadruples $(\mathcal U_j, \mathcal U_{j-1}, \mathcal
 R_j, \mathcal R_{j-1})$ such that:
@@
 \label{prop:bijection}
-With the above notations, the function $\psi_j$ induces a bijection between
+The function $\psi_j$ induces a bijection between
 $\Omega_j^\times$ and $\mathcal E_j^\times$.
@@
 \left\{\begin{array}{l}
-A = (-1)^j \cdot a^{-2} \cdot (V_j R_{j-1} - V_{j-1} R_j) \smallskip \\
-B = (-1)^{j-1} \cdot a^{-2} \cdot (U_j R_{j-1} - U_{j-1} R_j) \\
+A = (-1)^j \cdot a^{-2} \cdot (\mathcal V_j \mathcal R_{j-1} - \mathcal V_{j-1} \mathcal R_j) \smallskip \\
+B = (-1)^{j-1} \cdot a^{-2} \cdot (\mathcal U_j \mathcal R_{j-1} - \mathcal U_{j-1} \mathcal R_j) \\
 \end{array}\right.
@@
 \begin{proof} 
-Given a subset $J$ of $\{0, \ldots, d-1\}$, we denote by 
+Given $J \subset \{0, \ldots, d-1\}$, we denote by 
 $\Omega_\ring(J)$ the subset of $\Omega_\ring$ consisting of
@@
 $q^{2d-\Card J} (q-1)^{\Card J}$. To do this, we introduce several 
-additional notations. First, we fix $J$ and write $J = \{n_1, \ldots, 
+additional notations. First, we write $J = \{n_1, \ldots, 
 n_\ell\}$ with $n_1 > n_2 > \cdots > n_\ell$ and set $n_{\ell+1} = 
-0$ by convention. Given $n$ and $m$ two integers with $m < n$, let 
+0$ by convention. Given $n$ and $m$ two integers with $m < n$, we let 
 $V_{(m,n)}$ denote the set of polynomials of the form
@@
 V_{(m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
-with coefficients $< m$. Finally, given $(A,B)$ in $\Omega_\ring$, we 
+of degree $< m$. Finally, given $(A,B)$ in $\Omega_\ring$, we 
 denote by $(S_i(A,B))$ its subresultant pseudo-remainder sequence as 
@@
 inequalities over all possible $J$, we get $\Card \Omega_\ring \leq 
-q^{2d}$. This latest inequality being an equality, we need to have
+q^{2d}$. This latest inequality being an equality, we must have
 $\Card \Omega_\ring(J) = q^{2d-\Card J} (q-1)^{\Card J}$ for all $J$.
@@
 We assume first that $j < d-1$. Proposition \ref{prop:distribk} above 
-then says that $r_{j+1}$ is invertible in $\A$ with probability 
+ensures that $r_{j+1}$ is invertible in $\A$ with probability 
 $(1 - q^{-1})$. Moreover, assuming that this event holds, Corollary 
@@
 We fix $j \in \{0, \ldots, d-1\}$. 
-We define a random variable $V_j^{(0)}$ as the greatest (nonnegative) 
+We define the random variable $V_j^{(0)}$ as the greatest (nonnegative) 
 integer $v$ such that all principal subresultants $r_{j'}$ have positive 
@@
 \text{and} &
-B^{(1)} = A^{(1)} + \frac 1 {\pi X^{j-v-1}} \cdot R_{j+v-1}[j{-}v{-}1\:{:}].
+B^{(1)} = A^{(1)} + \frac 1 {\pi X^{j-v-1}} \cdot R_{j+v-1}[j{-}v{-}1\:{:}]
 \end{array}$$
+where we recall that, given a polynomial $P$ and an integer $m$, the 
+notation $P[m{:}]$ refers to the polynomial obtained from $P$ by
+removing its monomials of degree strictly less than $m$.
 We notice that all the coefficients of $B^{(1)}$ lie in $\A$ 
@@
 
-\section{Stabilization}
+\section{A stabilized algorithm for computing subresultants}
 \label{sec:stable}
 
-We have seen in the previous sections that Euclide's algorithm (and 
-their variants) are unstable in practice. On the other hand, 
-one can compute subresultants in a very stable way by evaluating
-the corresponding minors of the Sylvester matrix. Doing so, we do
-not loose any significant digit. Of course, the downside is the
-rather bad efficiency.
+We have seen in the previous sections that Euclide-like algorithm are 
+unstable in practice. On the other hand, one can compute subresultants 
+in a very stable way by evaluating the corresponding minors of the 
+Sylvester matrix. Doing so, we do not loose any significant digit. Of 
+course, the downside is the rather bad efficiency.
 
@@
 \cite{padicprec} where the authors draw the lines of a general framework 
-to handle a sharp (often optimal) track of ultrametric precision.
+to handle a sharp (often optimal) track of ultrametric precision. In 
+what follows, the letter $W$ still refers to a complete DVR while the 
+letter $K$ is used for its fraction field.
 
@@
 Although the defintion of a lattice is similar to that of $\Z$-lattice 
-in $\R^d$, the geometrical representation of it is quite different since 
-the elements of $\A$ themselves are not equidistributed in $K$ but form 
-a ball inside $K$ (they are exactly elements of norm $\leq 1$). More 
+in $\R^d$, the geometrical representation of it is quite different. 
+Indeed, the elements of $\A$ themselves are not distributed as $\Z$ is 
+in $\R$ but rather from a ball inside $K$ (they are exactly elements of 
+norm $\leq 1$). More 
 generally, assume that $E$ is equipped with a ultrametric norm $\Vert 
@@
 E$). (A typical example is $E = K^n$ equipped with the sup norm.) One
-checks that the subsets
-$$B_E(r) = \big\{ \,\, x \in E \quad|\quad \Vert x \Vert_E \leq r \,\,\big\}$$
+checks that the balls
+$$B_E(r) = \big\{ \,\, x \in E \quad\big|\quad \Vert x \Vert_E \leq r \,\,\big\}$$
 are all lattices in $E$. Moreover, any lattice is deduced from $B_E(1)$ 
@@
 consolidated by the following result which roughly speaking claims that 
-such cosets behave quite nicely under differentiable maps.
+such cosets behave quite well under differentiable maps.
 
@@
 
-In what follows, we will mainly use Lemma \ref{lem:padicprec} with
+In what follows, we will often use Lemma \ref{lem:padicprec} with
 $\rho = 1$. It states in this particular case that 
@@
 
-Obviously, all integral polynomial functions are of class $C^1$ (and
-even locally analytic), so that Lemma~\ref{lem:padicprec} applies to 
-them.
-
 \begin{ex}
+\label{ex:Gaussnorm}
 Let us examine more closely the case of polynomial spaces since it will 
@@
 
+Obviously, all integral polynomial functions are function of class $C^1$ 
+(and even locally analytic), so that Lemma~\ref{lem:padicprec} applies 
+to them. Proposition \ref{prop:padicprec} below exhibits an explcit 
+value for the bound $\delta$ appearing in Lemma~\ref{lem:padicprec} when 
+$f$ is integral polynomial and $r=1$.
+
 \begin{prop}
@@
 
-Let us now briefly explain how one can use Lemma~\ref{lem:padicprec} for 
-tracking precision.
+Let us now briefly explain how Lemma~\ref{lem:padicprec} can be utilized 
+for tracking precision.
 
@@
 the main loop) of the algorithm we are executing --- modeled by a 
-stricty differentiable function $g$ defined on an open subset $U$ of a 
-normed $K$-vector space $E$ and taking values in another normed 
-$K$-vector space $F$.
+function $g$ of class $C^1$ defined on an open subset $U$ of a finite 
+dimensional normed $K$-vector space $E$ and taking values in another 
+finite dimensional normed $K$-vector space $F$.
 Our input is an approximated element of $U$ which is represented 
-as a coset $C$ with respect to some lattice $H$, that is a subset of $U$ 
+by a coset $C$ with respect to some lattice $H$, that is a subset of $U$ 
 of the form $C = x+H$ for some $x \in U$. We would like to insist on the 
@@
 the subset $C$. However, since $H$ is stable under addtion, we have $C = 
-x+H$ \emph{any} element $x \in C$.\footnote{This assertion means that 
+x+H$ for \emph{any} element $x \in C$.\footnote{This assertion means that 
 any element of the ``rectangle'' $C$ is a center of it... which might be 
@@
 where $y \in F$ and $H_{\text{naive}}$ is a lattice which is generally 
-much more larger that the best one, namely $g'(x)(H)$ according to
+much more larger that $g'(x)(H)$, the latter being the best possible
+one according to
 Lemma~\ref{lem:padicprec} (assuming that the assumptions of this Lemma
@@
 to be efficient, a special close analysis taking advantage of the 
-particular problem under consideration is often necessary. For the 
-example considered in the paper (which is that of subresultants), we will 
-use a \emph{theoretical} argument in order to exhibit a good approximation 
-of $g'(x)(H)$, that are lattices $H_\min$ and $H_\max$ such that:
+particular problem under consideration is often necessary. For now, let 
+us simply assume that we have given two lattices $H_\min$ and $H_\max$ 
+with the property that:
 \begin{equation}
 \label{eq:Hminmax}
-H_\min \subset g'(x)(H) \subset H_\max
+H_\min \subset g'(x)(H) \subset H_\max.
 \end{equation}
-for a negligible cost. Up to allowing in some cases a \emph{very slight} 
-leak of precision, this would be enough for the application we have in 
-mind. We refer to \S \ref{subsec:stabilization} for more details about
-this part.
-
-For now on, we assume that $H_\min$ and $H_\max$ satisfying 
-Eq.~\eqref{eq:Hminmax} have been computed and we focus on 
-part~\eqref{item:gx}, which also requires some discussion. Indeed, 
-computing $g(x)$ is not straightforward because $x$ itself lies in a 
-$K$-vector space and therefore cannot be stored and manipulated on a 
-computer. Nevertheless, one can take advantage of the fact that $x$ may 
-be chosen arbitratily in $C$. More precisely, we pick a sublattice $H'$ 
-of $H$ and consider the new approximated element $x+H' \subset x+H$. 
-Concretely, this means that we arbitrarily increase the precision on the 
-given input $x$. Now, applying the naive method with $x+H'$, we compute 
-some $y \in F$ and some lattice $H'_{\text{naive}} \subset F$ with the 
-property that:
+We shall see later (\emph{cf} \S \ref{subsec:stabilization}) how these
+lattices can be constructed --- for a negligible cost --- in the special 
+case of subresultants.
+
+We now focus on part~\eqref{item:gx}, which also requires some 
+discussion. Indeed, computing $g(x)$ is not straightforward because $x$ 
+itself lies in a $K$-vector space and therefore cannot be stored and 
+manipulated on a computer. Nevertheless, one can take advantage of the 
+fact that $x$ may be chosen arbitrarily in $C$. More precisely, we pick 
+a sublattice $H'$ of $H$ and consider the new approximated element $x+H' 
+\subset x+H$. Concretely, this means that we arbitrarily increase the 
+precision on the given input $x$. Now, applying the naive method with 
+$x+H'$, we compute some $y \in F$ and some lattice $H'_{\text{naive}} 
+\subset F$ with the property that:
 $$g(x+H') \subset y + H'_{\text{naive}}.$$
@@
 Moreover, estimating the dependance of $H'_{\text{naive}}$ in terms 
-of $H'$ is usually rather easy (remember that $f$ is supposed to model
+of $H'$ is usually rather easy (remember that $g$ is supposed to model
 a simple operation). Hence since $H_\min$ is known --- as
@@
 generally not difficult.
-The whole method is illustrating on Figure~\ref{fig:adaptative}.
 
@@
 consisting in a succession of $n$ steps $\texttt{G}_0, \ldots, 
-\texttt{G}_{n-1}$. It is modeled by a strictly differentiable function $f : 
-U \to F$ where $U$ is an open subset in a normed $K$-vector space $E$ 
-and $F$ is a normed $K$-vector space. The input of \texttt{F} is an 
-approximated element in $U$ represented as a coset $C = x+H$ where $x 
-\in U$ and $H$ is a lattice.
+\texttt{G}_{n-1}$. It is modeled by a function $f : U \to F$ of class 
+$C^1$ where $U$ is an open subset in a finite dimensional normed 
+$K$-vector space $E$ and $F$ is a finite dimensional normed $K$-vector 
+space. The input of \texttt{F} is an approximated element in $U$ 
+represented as a coset $C = x+H$ where $x \in U$ and $H$ is a lattice.
 We also introduce notations for each individual step. For all $i$, we
-assume that $\texttt{G}_i$ is modeled by a strictly differentiable
-function $g_i : U_i \to U_{i+1}$ where $U_i$ is an open subset is 
+assume that $\texttt{G}_i$ is modeled by a function $g_i : U_i \to
+U_{i+1}$ of class $C^1$ where $U_i$ is an open subset is 
 some normed $K$-vector space $E_i$ and, by convention, $U_0 = U$, $E_0 
@@
 We note that the first and the second assumptions are quite strong 
-because it implies in particular that the sequence of $\dim E_i$ is 
+because they imply in particular that the sequence of $\dim E_i$ is 
 non-increasing. However, it really simplifies the forthcoming discussion 
@@
 This leads to a \emph{stabilized} version of the algorithm \texttt{F} 
-whose skeleton is depicted in Algorithm~\ref{algo:stabilizedG}.
-%
+whose skeleton is depicted in Algorithm~\ref{algo:stabilizedF}.
+
 \begin{algorithm}
@@
 \caption{Stabilized version of \texttt{F}}
-\label{algo:stabilizedG}
+\label{algo:stabilizedF}
 \end{algorithm}
-%
+
+The correctness of Algorithm~\ref{algo:stabilizedF} (under the 
+assumptions listed above) is clear after Lemma~\ref{eq:padicprec}.
 
@@
 \circ g_d$ (be careful with the order of the indices) models (a slight 
-variant of) Algorithm~\ref{algo:subres}. For all $j$, we set $f_j = 
-g_{j+1} \circ \cdots \circ g_d$ for all $j$. It is the function:
+variant of) Algorithm~\ref{algo:subres}. For all $j$, we put $f_j = 
+g_{j+1} \circ \cdots \circ g_d$; it is the function:
 $$\begin{array}{rcl}
@@
 $H = \pi^N \calL$ where $\calL = \A_{<d}[X] \times \A_{<d}[X]$ is the 
-standard lattice in $K_d[X] \times K_d[X]$. Following \S 
+unit ball in $K_d[X] \times K_d[X]$ with respect to the Gauss norm
+(\emph{cf} Example~\ref{ex:Gaussnorm}). Following \S 
 \ref{subsec:crashcourse}, our first task consists in finding two 
@@
 where $r_j$ is the $j$-th principal subresultant of $(A,B)$ and
-$\calL_j = \A_{\leq j}[X] \times \A_{\leq j-1}[X]$ is the standard
-lattice in $K_{\leq j}[X] \times K_{\leq j-1}[X]$.
+$\calL_j = \A_{\leq j}[X] \times \A_{\leq j-1}[X]$ is the unit ball
+in $K_{\leq j}[X] \times K_{\leq j-1}[X]$.
 \end{lem}
@@
 $$(-1)^j \cdot r_j^{-2} \cdot 
-(V_j R_{j-1}{-}V_{j-1} R_j,\: -U_j R_{j-1}{+}U_{j-1} R_j)$$
+(V_j S_{j-1}{-}V_{j-1} S_j,\: -U_j S_{j-1}{+}U_{j-1} S_j)$$
 where $U_j, V_j$ (resp. $U_{j-1}$, $V_{j-1}$) are the $j$-th (resp 
@@
 \ref{prop:precEuclide} for similar considerations). Therefore, one can 
-take $H'_j = r_j^2 r_{j+1}^2 \pi^N \calL_j$.
+take $H'_j = r_j^2 r_{j+1}^2 \cdot \pi^N \calL_j$. Instantiating
+Algorithm~\ref{algo:stabilizedF} in this particular case, we end up
+with Algorithm~\ref{algo:stabsubres} below which then appears as a 
+stable version of Algorithm~\ref{algo:subres}.
 
@@
 number of bit operations needed to perform an arithmetic operation 
-(addition, product) in $\A$ at precision $O(\pi^n)$.
+(addition, product, division) in $\A$ at precision $O(\pi^n)$.
 \end{prop}
@@
 In all usual examples ($p$-adic numbers, Laurent series), one can 
-choose $\M(n)$ is quasi-linear in $n$ and the size of the residue
+choose $\M(n)$ to be quasi-linear in $n$ and the size of the residue
 field $k$.
@@
 assumption~(b) is harmless on average --- but maybe not on particularly 
-bad instances. Let us also underline that, if we are just interested in 
+bad instances. We moreover underline that, if we are just interested in 
 computing the $j$-th subresultant for a particular $j$, then we just 
 need to assume the non-vanishing of the principal subresultants in the 
-range $[j{+}1, d{-}1]$.
+range $[j+1, d-1]$.
 
@@
 sequences as well.} subresultant pseudo-remainder sequence algorithm 
-following the same strategy. Nevertheless, this extension will not 
-completely straightforward because one will need to understand precisely 
-how the coefficients $c_i$'s (appearing in Eq.~\ref{eq:subressequence}) 
-alter the behaviour of the precision.
+following the same strategy. Nevertheless, this extension is not 
+completely straightforward because designing it requires to understand 
+precisely how the coefficients $c_i$'s (appearing in 
+Eq.~\ref{eq:subressequence}) alter the behaviour of the precision. We
+therefore let it as an open question.
 
@@
 polynomials $(A,B)$ such that $\lc(B)$ is invertible in $\A$ and $\deg A 
-\geq \deg B$. Indeed, it is enough for that to replace line 2 by:
+\geq \deg B$. Indeed, it is enough for this to replace line 2 by:
 $$R_{d-1} \leftarrow (-1)^{\deg A - \deg B} (A \,\%\, B).$$
-We remark that this can be computed at precision $O(\pi^N)$ thanks to 
-the assumption on the leading coefficient of $B$.
 However, writing an extension of Algorithm~\ref{algo:stabsubres} that 
-accepts all inputs seems much more tricky and this is the first question 
-we would like to address.
+accepts all inputs seems much more tricky and this is the second open
+question we would like to point out.
 
@@
 computation seems to be much more subtle. One can get around this issue 
-by using lifting technics only when $r_j$ is invertible and tracking 
+by using lifting technics only when $r_j$ is a unit in $\A$ and tracking 
 precision naively otherwise: it is possible to get this way a stable 
@@
 
-\section{Conclusion: $p$-adic floats?}
+\section{Conclusion: towards $p$-adic floats}
 
@@
 two main advantages. First, it allows simple and fast implementations.
-Second, experiments show the results obtained this way have generally
+Second, experiments show that the obtained results have generally
 more much correct digits that those predicted by arithmetic interval.
@@
 Let us describe quickly what might be this analogue and what are its
-advantages and disadvantages. We keep the notations of previous
+advantages and disadvantages. We keep the notations of the previous
 sections: the letter $\A$ denotes a complete discrete valuation ring
@@
 ultrametric floating point arithmetic are the simplicity and the 
-efficiency and the counterpart is the difficulty to get proved results. 
+efficiency while the counterpart is the difficulty to get proved results. 
 Moreover, the aforementioned examples are evidences that ultrametric 

### 2015-07-23 resultants/resultants.tex
formulas -> formulae

@@
 
-There exist simple formulas to following precision after each single
+There exist simple formulae to following precision after each single
 elementary computation. For instance, basic arithmetic operations can
@@
 
-\footnotetext{We observe that these formulas can be rephrased as follows: 
+\footnotetext{We observe that these formulae can be rephrased as follows: 
 the absolute (resp. relative) precision on the result of a sum or a
@@
 known digits of $a$ (resp. $b$) are zero.
-Combining these formulas, one can track the precision while executing 
+Combining these formulae, one can track the precision while executing 
 any given algorithm. This is the analogue of the standard \emph{interval 
@@
 subresultants. We note moreover that, in the common case where $n_{i-1} 
-= n_i - 1$, the two formulas giving $\Res_{n_i}(A,B)$ agree.
+= n_i - 1$, the two formulae giving $\Res_{n_i}(A,B)$ agree.
 
@@
 expression, the naive solution for evaluating $g(C)$ consists in using 
-formulas \eqref{eq:precadd}--\eqref{eq:precdiv}. However, this often 
+formulae \eqref{eq:precadd}--\eqref{eq:precdiv}. However, this often 
 results in an overestimation on the precision, in the following sense: 

### 2015-09-25 polynomials/muldivfact.tex
New paper of p-adic polynomials (empty for now)

--- 2015-04-27 examples/stability.tex
@@
 
-\conferenceinfo{ISSAC'15,}{July 6--9, 2015, Bath, United Kingdom.}
-\copyrightetc{Copyright \copyright~2015 ACM \the\acmcopyr}
-\crdata{978-1-4503-3435-8/15/07}
+\permission{%
+}
 
@@
 
-\title{p-adic stability in linear algebra}
+\title{Multiplication, division and factorization\\of p-adic polynomials}
 
@@
 \begin{abstract}
-Using the differential precision methods developed previously by the same authors,
-we study the $p$-adic stability of standard operations on matrices and vector 
-spaces. We demonstrate that lattice-based methods surpass naive methods in many
-applications, such as matrix multiplication and sums and intersections of subspaces.
-We also analyze determinants, characteristic polynomials and LU factorization using these differential methods.
-We supplement our observations with numerical experiments.
 \end{abstract}
@@
 \terms{Algorithms, Theory}
-\keywords{$p$-adic precision, linear algebra, ultrametric analysis}
+\keywords{}
 
@@
 
-For about twenty years, the use of $p$-adic methods in symbolic 
-computation has been gaining popularity. Such methods were 
-used to compute composed products of polynomials 
-\cite{boston-gonzalez-perdry-schost:05a}, 
-to produce hyperelliptic curves of genus $2$ with complex multiplication 
-\cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a},
-to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
-and to count points on varieties using $p$-adic 
-cohomology theories (\emph{cf.} \cite{kedlaya:01a,lauder:04a} and many
-followers).
-However, a general framework allowing a precise study of $p$-adic 
-precision --- the main issue encountered when computing 
-with $p$-adic numbers --- was designed only recently in 
-\cite{caruso-roe-vaccon:14a}.
+\section{Modular multiplication}
 
-In \cite{caruso-roe-vaccon:14a}, we advocate the use of lattices
-to track the precision of vectors and points on $p$-adic manifolds.
-The main result of \emph{loc.~cit.}, recalled in Proposition \ref{prop:precision},
-allows for the propagation of precision using differentials.  While
-representing precision by lattices carries a high
-cost in space and time requirements, the reduced precision loss
-can sometimes overwhelm these costs within a larger algorithm.
-In this paper, we apply the ideas of \cite{caruso-roe-vaccon:14a} to
-certain linear algebraic tasks.
+\section{Euclidean division}
 
-\smallskip
-
-\noindent
-{\bf Main Results.}
-We give a number of contexts where lattice-based precision methods
-outperform the standard coordinate-wise methods.  The two most striking
-examples are an analysis of matrix multiplication and of 
-sums and intersections of subspaces.  In Proposition \ref{prop:mulmatrix},
-we give a formula for the lattice-precision of the product of two matrices.
-In Figure \ref{fig:mulmatrix}, we describe the precision loss in multiplying
-$n$ matrices, which appears linear in $n$ when using standard methods
-and logarithmic in $n$ when using lattice methods.  We also give an
-example in Figure \ref{fig:vectorspace} where lattice methods
-actually yield an \emph{increase} in precision as the computation
-proceeds.
-
-\begin{comment}
-The present paper, which may be considered as a continuation of 
-\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc.~cit.} in order to analyze the $p$-adic stability of many standard 
-algorithms for computing with basic algebraic structures: numbers, 
-matrices, univariate polynomials and vector spaces. For each considered 
-algorithm, we will follow the same protocol: 
-first, we will compute the theoretical loss of precision of the 
-underlying question using the framework of \cite{caruso-roe-vaccon:14a} 
-and then compare it to those observed in available implementations in 
-\textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
-\textsc{sage} \cite{sage}.
-
-The conclusion of our analysis is that generic algorithms which are 
-designed to work over an arbitrary field are often quite unstable. Very 
-roughly, denoting by $D$ the number of divisions performed by the 
-algorithm under analysis and by $q$ the cardinality of the residue 
-field, it appears that the numbers of lost digits is about $O(\frac D 
-q)$ whereas the theoretical loss of precision is closer than $O(\log_q 
-D)$ --- and sometimes even $O(1)$. We would like to underline in 
-particular that the instability phenomena are much more apparent when 
-the residue field is small. It is the reason why we will always take
-$K = \Q_2$ is our examples.
-\end{comment}
-
-\smallskip
-
-\noindent
-{\bf Organization of the paper.}
-Section \ref{ssec:latticeprec} recalls the theory of 
-precision developed in \cite{caruso-roe-vaccon:14a} and defines
-a notion of diffuse precision for comparing to coordinate-wise methods.
-In particular, we recall Proposition \ref{prop:precision}, which
-allows the computation of precision using differentials.
-Proposition \ref{prop:boundLambdaf2} in Section \ref{ssec:boundLambdaf}
-is a technical result that will be used to determine the applicability of
-Proposition \ref{prop:precision}.
-
-In Section \ref{ssec:mulmatrix}, we analyze matrix multiplication and report
-on experiments that demonstrate the utility of lattice-based precision tracking.
-Section \ref{ssec:det} finds the precision of the determinant of a
-matrix, and Section \ref{ssec:charpoly} applies the resulting formula to
-characteristic polynomials.  We define the precision polygon of a matrix, which
-gives a lower bound on the precision of the characteristic polynomial.
-This polygon lies above the Hodge polygon of the matrix;
-we give statistics on the difference and the amount of diffuse precision.
-Finally, we apply Proposition \ref{prop:precision} to LU factorization and
-describe experiments with lattice-based precision analysis.
-Section \ref{ssec:grassgeo} reviews the geometry of Grassmannians,
-which we apply in Section \ref{ssec:grassdiff} to differentiating
-the direct image, inverse image, sum and intersection maps
-between Grassmannians.  We then report in Section \ref{grassimpl}
-on tracking the precision of subspace arithmetic in practice.
-In the appendix, we give a proof of Proposition \ref{prop:boundLambdaf}.
-
-The code used to make experiments presented in this paper is 
-available at \url{https://github.com/CETHop/padicprec}.
-
-\smallskip
-
-\noindent
-{\bf Notation.}
-Throughout the paper, $K$ will refer to a complete discrete 
-valuation field. Usual examples are finite extensions of $\Qp$ and 
-Laurent series fields over a field. We denote by $\val : K \twoheadrightarrow \Z \cup 
-\{+\infty\}$ the valuation on $K$, by 
-$\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
-$1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.
-
-\medskip
-
-\section{The theory of p-adic precision}
-\label{sec:theory}
-
-The aim of this section is to briefly summarize the content of 
-\cite{caruso-roe-vaccon:14a} and fill in certain details.
-
-\subsection{Lattices as precision data}
-\label{ssec:latticeprec}
-
-In \cite{caruso-roe-vaccon:14a}, we suggest the 
-use of lattices to represent the precision of elements in 
-$K$-vector spaces.  We shall contrast this approach with the
-\emph{coordinate-wise method} used in Sage, % and Magma,
-where the precision of an element is specified by giving the precision
-of each coordinate separately and is updated after each basic
-operation.
-
-Consider a finite 
-dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
-actually those of Banach spaces. However, we will not need infinite 
-dimensional spaces here.} normed vector space $E$ 
-defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
-on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
-closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
-E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
-Since we are working in a ultrametric world, the balls $B^{\phantom 
--}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
-should be thought of as special neighborhoods of $0$ and therefore are good 
-candidates to model precision data. Moreover, as revealed in 
-\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
-differentiable maps:
-
-\begin{prop}
-\label{prop:precision}
-Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
-and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
-$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
-that the differential $f'(v_0)$ is surjective.
-Then, for all $\rho \in (0, 1]$, there exists a positive real
-number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
-satisfies:
-\begin{equation}
-\label{eq:firstorder}
-f(v_0 + H) = f(v_0) + f'(v_0) (H).
-\end{equation}
-\end{prop}
-
-This proposition enables the \emph{lattice method} of tracking precision,
-where the precision of the input is specified as a lattice $H$ and precision
-is tracked via differentials of the steps within a given algorithm.
-The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
-yields the optimum possible precision. 
-We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
-exposition.
-
-In \cite{caruso-roe-vaccon:14a}, we also explained that if 
-$f$ is locally analytic, then the constant $\delta$ appearing in
-Proposition \ref{prop:precision} can be expressed in terms of the
-growing function $\Lambda(f)$ defined by
-\[
-\textstyle \Lambda(f)(v) = 
-\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
-f(h) \Vert \big)
-\]
-with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
-converge on $B^-_E(e^v)$.
-We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
-precise statement.  We state here the case of
-integral polynomial functions. A function $f : E \to F$ is said to be
-\emph{integral polynomial} if it is given by multivariate polynomial 
-functions with coefficients in $\O_K$ in any (equivalently all) system 
-of coordinates associated to a $\OK$-basis of $B_E(1)$.
-
-\begin{prop}
-\label{prop:precision2}
-We keep the notation of Proposition \ref{prop:precision} and assume 
-in addition that $f$ is integral polynomial. Let $C$ be a positive real
-number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
-Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
-\rho^{-1}$.
-\end{prop}
-
-In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
-to manifolds over $K$, where the precision 
-datum at some point $x$ is a lattice in the tangent space at $x$. 
-Propositions \ref{prop:precision} and \ref{prop:precision2} have
-analogues obtained by working in charts.  We use this extension
-to compute with vector spaces in \S \ref{sec:vectorspaces}.
-
-We will use the following definition in contrasting lattice and coordinate-wise methods. 
-Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.
-
-\begin{deftn} \label{def:diffused}
-Let $H \subset E$ be a lattice. The number of 
-\emph{diffused digits of precision} of $H$ is the length of
-$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
-\end{deftn}
-If $H$ represents the actual precision of some object, then
-$H_0$ is the smallest diagonal lattice containing $H$.  Since
-coordinate-wise methods cannot yield a precision better than $H_0$,
-$k$ provides a lower bound on the number of $p$-adic digits
-gained by lattice methods over standard methods.
-
-\subsection{A bound on a growing function}
-\label{ssec:boundLambdaf}
-
-In the next sections, we will compute the derivative of several standard 
-operations and sometimes give a simple expression in term 
-of the input and the output. In other words, the function $f$ modeling
-such an operation satisfies a differential equation of the form
-$f' = g \circ (f, \id)$  %%%%%%%% HERE
-where $g$ is a given --- and hopefully rather simple --- function. The 
-aim of this subsection is to study this differential equation and to 
-derive from it certain bounds on the growing function $\Lambda(f)$. 
-\emph{We will assume that $K$ has 
-characteristic $0$.}
-
-Let $E, F$ and $G$ be finite-dimensional normed vector spaces with
-$U \subseteq E$ and $V \subseteq F$ and $W \subset G$ open subsets.
-Generalizing the setting above, we consider the 
-differential equation:
-\begin{equation}
-\label{eq:diffequah}
-f' = g \circ (f, h).
-\end{equation}
-Here $g : V \times W \to \Hom(E, F)$ and $h : U \to W$ are known
-locally analytic functions and $f : U \to V$ is the 
-unknown locally analytic function. 
-In what follows, we always assume that $V$ and $W$ contain the origin, 
-$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. These assumptions are harmless 
-for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
-accordingly) so that they both vanish at $0$, and second, in order to 
-apply Proposition \ref{prop:precision2} the derivative $f'(0)$ 
-needs to be surjective and therefore \emph{a fortiori} nonzero.
-
-We assume that we are given in addition two nondecreasing convex 
-functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
-\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
-there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
-$({-}\infty, \nu]$\footnote{We note that this assumption is fulfilled if 
-we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
-does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
-defined by:
-
-\vspace{-0.2cm}
-
-\begin{align*}
-\tau_\nu(x) &= \smash{\left\{\begin{array}{ll} x & \mbox{ if } x \le \nu, \\
-+\infty & \mbox{ otherwise;} \end{array}\right.} \\[8pt]
-\mbox{and } \Lambda_f(x) &= \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha),
-\end{align*}
-%$$\begin{array}{rll}
-%&\tau_\nu(x) = x & \text{if } x \leq \nu \\
-%& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
-%\text{and} &
-%\multicolumn{2}{l}{\Lambda_f(x) = 
-%  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
-%\end{array}$$
-where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
-e^{-\alpha n}$ for all $n$. If $p$ is the characteristic of the residue field,
-a suitable value for $\alpha$ is $\alpha = - 
-\frac p {p-1} \cdot \log \Vert p \Vert$ if $p > 0$ and $\alpha = 0$ if $p = 0$.
-The next Proposition is proved in Appendix \ref{app:proof}.
-
-\begin{prop}
-\label{prop:boundLambdaf} \label{PROP:BOUNDLAMBDAF}
-We have $\Lambda(f) \leq \Lambda_f$.
-\end{prop}
-
-\begin{figure}
-\null \hfill
-\begin{tikzpicture}[scale=0.8]
-\draw[thick,green!30,fill=green!20] 
-   (-2,-1)--(-1,2)--(-1,3)--(3.5,3)
- --(2.5,2)--(-3.5,-4)--cycle;
-\draw[->] (-5,0)--(3.5,0);
-\draw[->] (0,-5)--(0,3);
-\node[below, scale=0.75] at (3.5,0) { $x$ };
-\node[right, scale=0.75] at (0,3) { $y$ };
-\draw[black!80,dotted] (-1,2)--(0,2);
-\node[right,scale=0.75] at (0,2) { $\nu$ };
-\draw[black!80,dotted] (-2,0)--(-2,-1)--(0,-1);
-\node[above,scale=0.75] at (-2.4,0) { $\Lambda_h^{-1}(\nu){-}\alpha$ };
-\node[right,scale=0.75] at (0,-1) { $\Lambda_h^{-1}(\nu){+}\mu$ };
-\draw[Blue,very thick] (-5,-4)--(-2,-1)--(-1,2);
-\fill[Blue] (-1,2) circle (0.06);
-\draw[Blue] (-1,2)--(-1,3);
-\draw[Sepia, very thick] (-4.5,-5)--(3.5,3);
-\draw[Blue, very thick,dashed] (-2,-1)--(-3.5,-4);
-\node[below right,rotate=45,scale=0.9] at (-4.3,-4.8) 
-  { $y = x + \log \Vert g(0) \Vert$ };
-\node[above right,rotate=45,scale=0.9] at (-4.8,-3.8) 
-  { $y = x + \alpha + \mu$ };
-\end{tikzpicture}
-\hfill \null
-
-\vspace{-0.1cm}
-
-\caption{Admissible region for the graph of $\Lambda(f)$}
-\label{fig:area}
-
-\end{figure}
-
-Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
-The blue plain line represents the graph of the function $\Lambda_f$. A 
-quick computation shows that, on a neighborhood of ${-}\infty$, this 
-function is given by $\Lambda_f(x) = x + \alpha + \mu$
-where $\mu$ is the value that $\Lambda_g$ takes on the interval 
-$({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
-graph of $\Lambda(f)$ lies below the plain blue line. Moreover, we remark
-that the Taylor expansion of $f(z)$ starts with the term
-$g(0) z$. Hence, on a neighborhood of ${-}\infty$, we have 
-$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
-get:
-$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert$
-%  \quad \forall x \in \R.$$
-for all $x \in \R$.  In other words, the graph of $\Lambda(f)$ lies above the brown line.
-Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
-because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
-the dashed blue line defined as the line of slope $2$ passing through
-the first break point of the blue plain line, which has coordinates
-$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
-\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
-must coincide with the brown line until it meets the dashed blue line 
-and then has to stay in the green area.
-
-As a consequence, we derive the following 
-proposition, which can be combined with Proposition 3.12 of 
-\cite{caruso-roe-vaccon:14a}. Following \cite{caruso-roe-vaccon:14a},
-if $\varphi$ is a convex function and $v \in \R$, we define
-$$\varphi_{\geq v} : x \mapsto \inf_{y \geq 0} \big( \varphi(x+y)
-- vy \big).$$
-It is the highest convex function with $\varphi_{\geq v} \leq \varphi$ and 
-$\varphi'_{\geq v} \geq v$.
-
-\begin{prop}
-\label{prop:boundLambdaf2}
-Keeping the above notation, we have:
-$$\Lambda(f)_{\geq 2} (x) \leq 2(x + \alpha + \mu) -
-\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
-for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
-\end{prop}
-
-\begin{proof}
-The inequality follows from the fact that $y = 2(x + \alpha + \mu) - y_0$ is the equation of 
-the dashed blue line.
-\end{proof}
-
-\begin{rem}
-In certain situations,
-it may happen that the function $f$ is solution of a simpler 
-differential equation of the form $f' = g \circ f$. If this holds, 
-Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\geq 
-2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.
-
-Beyond this particular case, we recommend choosing the 
-function $h$ by endowing $F$ with the second norm $\Vert x \Vert'_F = 
-e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and taking $h : (F, \Vert 
-\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ to be the identity on the
-underlying vector spaces. The function $\Lambda(h) : \R \to \R$ then maps $x$ to $x + 
-\mu$ and we can choose $\Lambda_h = \Lambda(h)$.
-\end{rem}
-
-\section{Matrices}
-\label{sec:matrices}
-
-Let $M_{m,n}(K)$ denote the space of $m \times n$ matrices over $K$.  
-We will repeatedly use the Smith decomposition for $M \in M_{m,n}(K)$,
-which is $M = U_M \cdot \Delta_M \cdot V_M$
-with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
-for the valuation of the $(i,i)$-th entry of $\Delta_M$, and by convention set
-$\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.
-
-\subsection{Multiplication}
-\label{ssec:mulmatrix}
-
-To begin with, we want to study the behavior of the precision when 
-performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
-positive integers and assume that we want to multiply a matrix $A \in 
-M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. This operation is of course 
-modeled by the integral polynomial function:
-$$\begin{array}{rcl}
-\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
-M_{r,t}(K) \smallskip \\
-(A,B) & \mapsto & AB.
-\end{array}$$
-According to Proposition \ref{prop:precision}, the behavior of the precision when 
-computing $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$, the linear mapping that takes a pair 
-$(dA,dB)$ to $A \cdot dB + dA \cdot B$.
-
-To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
-lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
-to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
-of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
-M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
-contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
-entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
-Nonetheless, it may happen that the above inclusion is strict, meaning 
-that we are \emph{gaining} precision in those cases.
-
-Set $a_i = \sigma_i(A)$ and $b_i = \sigma_i(B)$, and define $M_{r,t}((a_i),(b_j))$ 
-as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
-such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.  
-%Write $a^{(m)}$ (resp. $b^{(m)}$)
-%for the number of $a_i$'s (resp. $b_i$'s) which are at least $m$.
-
-\begin{prop}
-\label{prop:mulmatrix}
-With the above notation, we have
-\[
-\begin{array}{rl}
-& \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
-= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
-\text{and} &
-\text{\emph{length}}\Big(\frac{M_{r,t}(\O_K)}{\mathcal P'_{r,s,t}(A,B)(\mathcal L_0) }
-\Big) =
-\displaystyle 
-\sum_{i,j} \min(a_i,b_j)
-\end{array}
-\]
-\end{prop}
-
-\begin{proof}
-We write $A \cdot dB + dA \cdot B = U_A \cdot M \cdot V_B$ with
-$$M = \Delta_A \cdot V_A \cdot dA \cdot V_B^{-1} 
-+ U_A^{-1} \cdot dB \cdot U_B \cdot \Delta_B.$$
-When $dA$ varies in $M_{a,b}(\O_K)$ so does $V_A \cdot dA \cdot V_B^{-1}$
-and therefore the first summand in $M$ varies in the subspace of 
-$M_{r,t}(\O_K)$ consisting of matrices whose entries on the $i$-th
-row have valuation at least $a_i$. Arguing similarly for the second
-summand, we deduce the first statement of the Proposition. The second
-statement is now clear.
-\end{proof}
-
-From the perspective of precision, the second statement of Proposition 
-\ref{prop:mulmatrix} means that the computation of $AB$ gains 
-$\sum_{i,j} \min(a_i,b_j)$ significant digits in absolute 
-precision\footnote{We
-note that, on the other side, the valuation of the entries of $AB$ may 
-increase, meaning that we are also losing some significant digits if we 
-are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
-(\emph{cf.} Proposition \ref{prop:precision2}). However, many of these digits are 
-diffused in the sense of Definition \ref{def:diffused}.
-To change bases in order to make this increased precision visible
-with coordinates, write 
-$AB = U_A \cdot P \cdot V_B$ with
-$P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
-Tracking precision in the usual way, the $(i,j)$-th entry
-of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$.  Multiplication by
-$U_A$ and $V_B$ then diffuses the precision across the entries of $AB$.
-
-\smallskip
-
-We now consider the impact of tracking this diffuse precision.
-Although the benefit is not substantial for
-a single product of matrices, it accumulates as we multiply a large
-number of matrices.  We illustrate this phenomenon with the following simple
-example.
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm 1:} {\tt example\_product}
-
-\noindent{\bf Input:} a list $(M_1, \ldots, M_n)$ of square matrices
-of size $d$.
-
-\smallskip
-
-\noindent 1.\ Set $P$ to the identity matrix of size $d$
-
-\noindent 2.\ {\bf for} $j=1,\dots,n$ {\bf do} {\bf compute} $P = P M_i$
-
-\noindent 3.\ {\bf return} the top left entry of $P$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\medskip
-
-\noindent
-Figure \ref{fig:mulmatrix} compares the number of significant digits in 
-\emph{relative} precision we are losing on the output of Algorithm 1 
-when we are using, on the one hand, a standard coordinate-wise track of 
-precision and, on the other hand, a lattice-based method to handle 
-precision.
-%
-\begin{figure}
-\begin{center}
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|c|c|c|c|}
-\hline
-\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
-\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
-\multicolumn{2}{|c|}{Average loss of precision} \\
-\cline{3-4}
-& & Coord-wise method & Lattice method \\
-\hline 
-$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
-$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
-$2$ & $1000$ & $157.8$ & $7.9$ \\
-\hline
-$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
-$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
-$3$ & $1000$ & $122.5$ & $7.0$ \\
-\hline
-\end{tabular}
-
-\smallskip
-
-{\small
-Results for a sample of $1000$ random inputs in $M_{d,d}(\Z_2)^n$}
-\end{center}
-\renewcommand{\arraystretch}{1}
-
-\vspace{-0.3cm}
-
-\caption{Average loss of precision in Algorithm 1}
-\label{fig:mulmatrix}
-\end{figure}
-%
-We observe that, in the first case, the number of lost digits seems to 
-grow linearly with respect to the number of multiplications we are 
-performing (that is $n$) whereas, in the second case, the growth seems 
-to be only logarithmic. It would be nice to have a precise formulation 
-(and proof) of this heuristic.
-
-Note that multiplication of many random matrices plays a central
-role in the theory of random walks on homogeneous spaces \cite{Benoist:2012}.
-Better stability in computing such products helps estimate
-Lyapunov exponents in that context.
-
-\subsection{Determinant}
-\label{ssec:det}
-
-The computation of the differential of $\det : M_{n,n}(K) \to K$
-is classical: at a point $M$ it is the linear map
-\[
-\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),
-\]
-where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $\rank(M) < n - 1$,
-then $\detp(M)$ is not surjective and we cannot apply Proposition 
-\ref{prop:precision}. Therefore, we suppose that $\rank(M) \ge n-1$ for the rest of this section.
-
-As with matrix multiplication, we first determine the image of
-the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.
-
-\begin{prop}
-\label{prop:detmatrix}
-Setting $v = \sigma_1(M) + \cdots + \sigma_{n-1}(M)$, we have
-$\detp(M)(\mathcal L_0) = \pi^v \O_K$.
-\end{prop}
-
-\begin{proof}
-From the description of $\detp(M)$, we see that it is enough to prove 
-that the smallest valuation of an entry of $\com(M)$ is $v$
-or, equivalently, that the ideal of $\O_K$ generated by all 
-minors of $M$ of size $(n-1)$ is $\pi^v
-\O_K$. But this ideal remains unchanged when we multiply $M$ 
-on the left or on the right by a unimodular matrix. Thus we may 
-assume that $M = \Delta_M$, and the result becomes clear.
-\end{proof}
-
-In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
-$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
-is known at precision $O(\pi^{N+v})$.  Thus we are gaining $v$  
-digits in absolute precision or, equivalently, losing $\sigma_n(M)$ digits
-of relative precision.
-Furthermore, one may compute $\det(M)$ with this optimal precision by finding an approximate
-Smith decomposition with $\Delta_M$ known at precision $O(\pi^N)$ and
-multiplying its diagonal entries.
-
-\subsection{Characteristic polynomials} \label{ssec:charpoly}
-
-Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic 
-polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of 
-polynomials of degree less than $n$.  Then the differential of 
-$\charpoly$ at a point $M$ is
-\[
-\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
-\]
-The image is the $K$-span of the entries of $\com(X{-}M)$, which is 
-clearly contained within $K_{<n}[X]$.  In fact, the image will equal 
-$K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same 
-generalized eigenvalue. For now on, we make this assumption.
-
-Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = 
-\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, 
-\val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
-$\NP(\charpoly(M))$.  The \emph{Hodge polygon} $\HP(M)$ of $M$ is the lower 
-convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_j(M))$. For 
-any matrix $M$, the polygon $\NP(M)$ lies above $\HP(M)$ 
-\cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.
-
-Such polygons arise naturally in tracking the precision of polynomials 
-\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
-lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting of polynomials whose 
-Newton polygons lie above $P$.  This lattice is generated by monomials 
-$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
-These polygons are used in a coordinate-wise precision tracking for polynomial
-arithmetic. We now introduce another polygon, bounded 
-between $\NP(M)$ and $\HP(M)$, that will provide an estimate on the precision of $\charpoly(M)$.
-
-\begin{deftn}
-The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of 
-the Newton polygon of the entries of $\com(X{-}M)$.
-\end{deftn}
-
-It is clear from the definition $\mathcal L_{\PP(M)} \subset 
-\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
-$M_{n,n}(\OK)$. More precisely, $\PP(M)$ is the smallest polygon $P$ for 
-which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
-Proposition \ref{prop:precision}, the precision polygon determines the minimal 
-precision losses possible when encoding polynomial precision using polygons.
-
-It turns out that the precision polygon is related to the Hodge and Newton 
-polygons. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
-the translated polygon with vertices $(x_i - n, y_i)$.
-
-\begin{prop} \label{prop:polygons}
-The precision polygon $\PP(M)$ lies between $T_1(\HP(M))$
-and $\NP(M)$.
-
-Moreover, $\PP(M)$ and $T_1(\HP(M))$ meet at $0$ and $n{-}1$.
-\end{prop}
-
-\begin{proof}
-The coefficients of $\charpoly(M)$ can be expressed as traces of 
-exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M))$,
-which is $\tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.  
-Computing $\Lambda^i(\Delta_M)$, we get the first statement of
-the Proposition. For $i = 1$, we further find that $\PP(M)$ vanishes
-at the abscissa $n{-}1$. By definition so does $T_1(\HP(M))$. The fact 
-that $\PP(M)$ and $T_1(\HP(M))$ meet at abscissa $0$ follows from
-Proposition \ref{prop:detmatrix}.
-
-It remains to prove the comparison with the Newton polygon.
-Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$-th entry of $M$, 
-$f_{i,j}$ as the $(i,j)$-th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
-\val(m_{i,j})$. We write $f[k]$ 
-for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
-+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
-for all $i$ and $k$,
-\begin{align*}
-f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
-\mu_{i,n} + f_{n,i}[k]) \\
-&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
-\end{align*}
-with the infimum over $j$. 
-Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing,
-which follows from the comparison with the Hodge polygon, we get the 
-result.
-\end{proof}
-
-\begin{rem}
-Experiments actually support the following stronger result: $\PP(M)$ is 
-bounded above by $T_1(\NP(M))$.
-\end{rem}
-
-For many matrices, $\PP(M) = T_1(\HP(M))$.  For random matrices over 
-$\Z_2$, the $2$-adic precision polygon is equal to the Hodge polygon in 
-$99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  
-Over $\Z_3$, the fraction rises to $99.98\%$, with no clear dependence on 
-dimension.
-Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at 
-$1$, corresponding to the precision of the linear term of the 
-characteristic polynomial.
-
-Of course, the precision lattice $\mathcal{E} = \charp(M)(\mathcal L_0)$ 
-may contain diffuse precision which is not encapsulated in $\PP(M)$.
-Diffuse precision arises in $11\%$ of cases in 
-dimension $3$, up to $15\%$ of cases in dimension $8$.  This 
-percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in 
-dimension $9$ for matrices constrained to have determinant with $2$-adic 
-valuation $12$.
-Moreover, one can construct examples with arbitrarily large amounts of diffuse precision.
-Suppose the $\sigma_i(M)$ are large.  
-Proposition \ref{prop:polygons} implies that $\mathcal{E}$ is 
-contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} 
-\sigma_i(M)$.  The precision lattice of $1 + M$ is obtained from 
-$\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ 
-is now flat with height $0$.
-
-For randomly chosen matrices, approximating $\mathcal{E}$ 
-using the Hodge polygon loses only a small amount of precision. However,
-if the $\sigma_i(M)$'s are large or if $M$ is a translate of
-such a matrix, using lattice precision can be very useful.
-
-\subsection{LU factorization}
-
-%This example has been previously studied by Caruso in \cite{caruso:12a}, 
-%but without using differential precision. We are going to show here that 
-%the framework of \S \ref{ssec:boundLambdaf} applies directly.
-In this section, we denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
-and, given a positive real number $C$, we let $B(C)$ be the closed ball 
-in $M_n(K)$ centered at the origin of radius $C$. We consider
-the following subsets of $M_n(K)$:
-
-\noindent $\bullet$
-$O_n$ is the open subset of matrices whose principal 
-minors do not vanish (we recall that the latest condition implies the
-existence and the uniqueness of a LU factorization);
-
-\noindent $\bullet$
-$U_n$ is the sub-vector space of upper-triangular 
-matrices;
-
-\noindent $\bullet$
-$L_n^0$ (resp. $L_n^u$) is the sub-affine space of 
-nilpotent (resp. unipotent) lower-triangular matrices.
-
-\smallskip
-
-\noindent
-{\bf Calculus and precision.}
-We choose to normalize the LU factorization by requiring that $L$ is 
-unipotent and denote by $D : O_n \to L_n^u \times U_n$ the function 
-modeling this decomposition. The computation of the differential of $D$ 
-has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. For
-$M \in \O_n$ with $D(M) = (L,U)$, the linear mapping $D'(M)$ is given 
-by:
-$$
-%\begin{array}{l}
-dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U)
-%\smallskip \\ \hspace{1cm}
-%\quad 
-\text{ with } dX = L^{-1} \cdot dM \cdot U^{-1}
-%\end{array}
-$$
-where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
-onto $L_n^0$ (resp. $U_n$). 
-It is easily checked that $D'(M)$ is bijective with inverse given by
-$(A,B) \mapsto AU+LB$.
-
-We now want to apply Proposition \ref{prop:boundLambdaf2} in order to 
-derive a concrete result on precision. \emph{We then assume that $K$ has 
-characteristic $0$.} We pick $M_0 \in
-O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the translated
-function $f$ taking $M$ to $D(M_0+M)-D(M_0)$. We then have $f(0) = 0$ 
-and $f'(M) = D'(M_0+M)$. Using the explicit description of the inverse
-of $D'(M_0+M)$, we find $B(1) \subset f'(0) \cdot B(C)$
-for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover, $f$ satisfies 
-the differential equation $f' = g \circ f$ where $g$ is defined by:
-\begin{equation}
-\label{eq:gAB}
-\begin{array}{l}
-g(A,B)(X) = \big( (L_0+A) \cdot \low(Y), \up(Y) \cdot (U_0+B) \big)
-\smallskip \\
-\hspace{1.8cm}\text{with } Y = (L_0+A)^{-1} \cdot X \cdot (U_0+B)^{-1}.
-\end{array}
-\end{equation}
-Let $\kappa(S) = \Vert S \Vert \cdot \Vert S^{-1} \Vert$ denote the 
-condition number of a matrix $S$. Remarking that $\Vert S + T 
-\Vert = \Vert S \Vert$ if $\Vert T \Vert < \Vert S \Vert$ and $\Vert (S 
-+ T)^{-1} \Vert = \Vert S^{-1} \Vert$ if $\Vert T \Vert < \Vert S^{-1} 
-\Vert^{-1}$, we deduce from \eqref{eq:gAB} that
-$\Vert g(A,B) \Vert \leq \max 
-\big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
-\kappa(U_0) \Vert L_0^{-1} \Vert \big)$
-provided that $\Vert A \Vert < \Vert L_0^{-1} \Vert^{-1}$ and 
-$\Vert B \Vert < \Vert U_0^{-1} \Vert^{-1}$.
-Combining this with Proposition \ref{prop:boundLambdaf2} and 
-\cite[Proposition.~3.12]{caruso-roe-vaccon:14a}, we finally find that 
-Eq.~\eqref{eq:firstorder} holds as soon as 
-$$\begin{array}{l}
-\displaystyle \frac \rho r > \Vert p \Vert^{-\frac{2p}{p-1}} \cdot
-\max ( \Vert L_0 \Vert, \Vert U_0 \Vert) \cdot
-\max ( \Vert L_0 ^{-1} \Vert, \Vert U_0 ^{-1} \Vert) \cdot{} \\
-\hspace{3.5cm} \max \big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
-  \kappa(U_0) \Vert L_0^{-1} \Vert \big)^2.
-\end{array}$$
-
-\noindent
-{\bf Numerical experiments.}
-Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
-$M_n(K)$. It can be naturally seen as a basis of $L_n^0 \times U_n$ as
-well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
-for the matrix of this linear mapping in the above basis. We remark that
-$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
-coordinate, we find the best coordinate-wise loss of precision we can 
-hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
-\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
-$\val(\det(D'(M)))$, which is precision lost in the lattice method. 
-%With the language of Definition \ref{def:diffused},
-The number of diffused digits of precision of $D'(M)(M_{n,n}(\OK))$ is 
-then the difference between these two numbers.
-Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
-losses for a sample of 2000 random matrices in $M_{d,d}(\mathbb{Z}_2)$.
-%
-\begin{figure}
-\begin{center}
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|c|c|c|c|c|}
-\hline 
-& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
-\cline{2-5}
-\raisebox{0.2em}{matrix}
-& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
-\smash{\raisebox{0.6em}{size}} 
-& \hspace{0.5em}mean\hspace{0.5em} & deviation 
-& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
-$2$ & $\phantom{2}3.0$& $5$ & $1.5$ & $1.4$ \\% \hline
-$3$& $\phantom{2}9.4$& $11$ & $2.3$ & $2.3$\\% \hline
-$4$ & $20\phantom{.4}$& $20$ & $3.8$ & $3.1$\\ \hline
-\end{tabular} 
-\smallskip
-
-{\small
-Results for a sample of $2000$ instances}
-\end{center}
-\renewcommand{\arraystretch}{1}
-
-\vspace{-0.3cm}
-
-\caption{Loss of precision for LU factorization}
-\label{fig:LU}
-
-\end{figure}
-%
-
-%\subsection{QR factorization}
-%
-%We refer to \cite[Appendix B]{caruso-roe-vaccon:14a} for the computation
-%of the derivative of the QR factorization.
-%\todo{Complete this subsection.}
-
-\section{Vector spaces}
-\label{sec:vectorspaces}
-
-Vector spaces are generally represented as subspaces of $K^n$ for some 
-$n$ and hence naturally appear as points on Grassmannians. Therefore, 
-one can use the framework of \cite[Appendix A]{caruso-roe-vaccon:14a} to 
-study $p$-adic precision in this context.
-
-\subsection{Geometry of Grassmannians}
-\label{ssec:grassgeo}
-
-Given $E$, a finite dimensional vector space over $K$, and $d$, an 
-integer in the range $[0, \dim E]$, we write $\Grass(E,d)$ for the
-Grassmannian of $d$-dimensional subspaces of $E$. It is well-known
-that $\Grass(E,d)$ has the natural structure of a $K$-manifold. The aim of
-this subsection is to recall standard facts about its geometry. In what follows,
-we set $n = \dim E$ and equip $E$ with a distinguished basis
-$(e_1, \ldots, e_n)$.
-
-\smallskip
-
-\noindent
-{\bf Description and tangent space.}
-Let $V$ denote a fixed subspace of $E$ of dimension $d$. The 
-Grassmannian $\Grass(E,d)$ can be viewed as the quotient of the set 
-of linear embeddings $f: V \hookrightarrow E$ modulo the action (by 
-precomposition) of $\GL(V)$: the mapping $f$ represents its image 
-$f(V)$. It follows from this description that the tangent space of 
-$\Grass(E,d)$ is canonically isomorphic to $$\Hom(V, E) / \End(V) 
-\simeq \Hom(V, E/V).$$
-
-\smallskip
-
-\noindent
-{\bf Charts.}
-Let $V$ and $V^\c$ be two complementary subspaces of $E$ 
-(\emph{i.e.} $V \oplus V^\c = E$). We assume that $V$ has 
-dimension $d$ and denote by $\pi$ the projection $E \to V$ 
-corresponding to the above decomposition. We introduce the set 
-$\mathcal U_{V,V^\c}$ of all embeddings $f : V \hookrightarrow E$ 
-such that $\pi \circ f = \id_V$. Clearly, it is an affine space over
-$\Hom(V,V^\c)$. 
-Furthermore, we can embed it into $\Grass(E,d)$ by taking $f$ as
-above to its image. This way, $\mathcal U_{V,V^\c}$ appears as
-an open subset of $\Grass(E,d)$ consisting exactly of those subspaces 
-$W$ such that $W \cap V^\c = 0$. As a consequence, the tangent space 
-at each such $W$ becomes isomorphic to $\Hom(V,V^\c)$. The
-identification $\Hom(V,V^\c) \to \Hom(W, E/W)$ is given by
-$du \mapsto (du \circ f^{-1}) \text{ mod } W$ where $f : V 
-\stackrel{\sim}{\to} W$ is the linear mapping defining $W$.
-
-When the pair $(V, V^\c)$ varies, the open subsets $\mathcal 
-U_{V,V^\c}$ cover the whole Grassmannian and define an atlas.
-When implementing vector spaces on a computer, we usually restrict 
-ourselves to the subatlas consisting of all charts of the form $(V_I, 
-V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
-\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspace spanned by 
-the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
-least one $\mathcal U_{V_I, V_{I^{\text{c}}}}$ and, given a family of
-generators of $W$, we can determine such an $I$ together with the
-corresponding embedding $f : V_I \hookrightarrow E$ by row reducing 
-the matrix of generators of $W$.
-
-\smallskip
-
-\noindent
-{\bf A variant.}
-Alternatively, one can describe $\Grass(E,d)$ as the set of linear 
-surjective morphisms $f : E \to E/V$ modulo the action (by 
-postcomposition) of $\GL(E/V)$. This identification presents the 
-tangent space at a given point $V$ as the quotient $\Hom(E, E/V) / 
-\End(E/V) \simeq \Hom(V, E/V)$.
-Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
-U^\star_{V, V^\c}$ denote the set of surjective linear maps $f : E 
-\to V^\c$ whose restriction to $V^\c$ is the identity. It is an 
-affine space over $\Hom(V, V^\c)$ which can be identified with an open
-subset of $\Grass(E,d)$ \emph{via} the map $f \mapsto \ker f$.
-
-It is easily seen that $\mathcal U_{V, V^\c}$ and $\mathcal 
-U^\star_{V, V^\c}$ define the same open subset in $\Grass(E,d)$. 
-Indeed, given $f \in \mathcal U_{V, V^\c}$, one can write $f = \id_V 
-+ h$ with $h \in \Hom(V, V^\c)$ and define the morphism $g = \id_E -
-h \circ \pi \in \mathcal U^\star_{V, V^\c}$. The association $f \mapsto
-g$ then defines a bijection $\mathcal U_{V, V^\c} \to \mathcal
-U^\star_{V, V^\c}$ which commutes with the embeddings into the
-Grassmannian.
-
-\smallskip
-
-\noindent
-{\bf Duality.}
-If $E$ is a finite dimensional vector space over $K$, we use the notation 
-$E^\star$ for its dual (\emph{i.e.} $E^\star = \Hom(E,K)$). If we are also given
-a subspace $V \subset E$, we denote by $V^\perp$ the subspace 
-of $E^\star$ consisting of linear maps that vanish on $V$. We recall
-that the dual of $V^\perp$ (resp. $E^\star/V^\perp$) is canonically
-isomorphic to $E/V$ (resp. $V$).
-For all $d$, the association $V \mapsto V^\perp$ defines a continuous 
-morphism $\psi_E : \Grass(E,d) \to \Grass(E^\star, n-d)$. The action of 
-$\psi_E$ on tangent spaces is easily described. Indeed, the differential 
-of $\psi_E$ at $V$ is nothing but the canonical identification between
-$\Hom(V, E/V)$ and $\Hom(V^\perp, E^\star/V^\perp)$ induced by 
-transposition. Furthermore, we observe that $\psi_E$ respects
-the charts we have defined above, in the sense that it maps bijectively 
-$\mathcal U_{V, V^\c}$ to $\mathcal U^\star_{V^\perp, (V^\c)^\perp}
-\simeq \mathcal U_{V^\perp, (V^\c)^\perp}$.
-
-\subsection{Differential computations}
-\label{ssec:grassdiff}
-
-In this subsection, we compute the differential of various operations on 
-vector spaces. For brevity, we skip the estimation of the corresponding 
-growing functions (but this can be done using Proposition 
-\ref{prop:boundLambdaf2} as before if $\text{char}(K) = 0$).
-
-\smallskip
-
-\noindent
-{\bf Direct images.}
-Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
-$n$ and $m$, respectively. Let $d$ be an integer in $[0,n]$. We are
-interested in the direct image function $\DI$ defined 
-on $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
-pair $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may vary,
-the map $\DI$ does not take its values in a well-defined 
-Grassmannian. We therefore stratify $\mathcal M$ as
-follows: for each integer $r \in [0,d]$, let $\mathcal M_r \subset \mathcal M$
-be the subset of pairs $(f,V)$ for which $f(V)$ has dimension $r$.
-The $\mathcal M_r$'s are locally closed in $\mathcal M$ and are therefore
-submanifolds. Moreover, $\DI$ induces differentiable 
-functions $\DI_r : \mathcal M_r \to \Grass(F,r)$.
-
-We would like to differentiate $\DI_r$ around some point $(f,V) \in 
-\mathcal M_r$. To do so, we use the first description of the 
-Grassmannians we gave above: we see points in $\Grass(E,d)$ 
-(resp. $\Grass(F,d)$) as embeddings $V \hookrightarrow E$ (resp. $W 
-\hookrightarrow F$) modulo the action of $\GL(V)$ (resp. $\GL(W)$).
-The point $V \in \Grass(E,d)$ is then represented by the canonical 
-inclusion $v : V \to E$ whereas a representative $w$ of $W$ satisfies
-$w \circ \varphi = f \circ v$
-where $\varphi : V \to W$ is the linear mapping induced by $f$. The 
-previous relation still holds if $(f,v)$ is replaced by a pair $(f', 
-v') \in \mathcal M_r$ sufficiently close to $(f,v)$.
-Differentiating it and passing to the quotient we find, first, that the 
-tangent space of $\mathcal M_r$ at $(f,v)$ consists of pairs $(df, dv) 
-\in \Hom(E,F) \times \Hom(V,E/V)$ such that
-$$d\tilde w = \big((df \circ v + f \circ dv) \text{ mod } W\big)
-\in \Hom(V, F/W)$$
-factors through $\varphi$ (\emph{i.e.} vanishes on $\ker \varphi = V 
-\cap \ker f$) and, second, that the differential of $\DI_r$ at $(f,V)$ 
-is the linear mapping sending $(df,dv)$ as above to the unique element 
-$dw \in \Hom(W,F/W)$ such that $dw \circ \varphi = d \tilde w$.
-
-\smallskip
-
-\noindent
-{\bf Inverse images.}
-We now consider the inverse image mapping $\II$ sending a 
-pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
-$f^{-1}(W)$. As before, this map does not take values in a single 
-Grassmannian, so we need to stratify $\mathcal W$ in order to get 
-differentiable functions. For each integer $s \in [0,n]$, we introduce 
-the submanifold $\mathcal W_s$ of $\mathcal W$ consisting of those 
-pairs $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
-induces a continuous function
-$\II_s : \mathcal W_s \to \Grass(E,s)$.
-Pick $(f,W) \in \mathcal W_s$. Set $V = f^{-1}(W)$ and denote by $w : F 
-\to F/W$ the canonical projection.
-Similarly to what we have done for direct images, one can prove that
-the tangent space of $\mathcal W_s$ at some point $(f,W) \in \mathcal
-W_s$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
-pairs $(df,dw)$ such that
-$d \tilde v = (w \circ df + dw \circ f)_{|W}$
-factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
-$f$. Furthermore $\II_s$ is differentiable at $(f,W)$ and its 
-differential is the linear mapping that takes $(df,dw)$ as above to the 
-unique element $dv \in \Hom(V,E/V)$ satisfying $\varphi \circ dv =
-d\tilde v$.
-
-Direct images and inverse images are related by duality
-as follows: if $f : E \to F$ is any linear map and $W$ is a subspace
-of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. We 
-can thus deduce the differentials of $\DI_s$ from those of $\II_s$ and 
-\emph{vice versa}.
-
-%\begin{rem}
-%A particularly interesting case of inverse images is of course those
-%of kernels.
-%\end{rem}
-
-\smallskip
-
-\noindent
-{\bf Sums and intersections.}
-Let $d_1$ and $d_2$ be two nonnegative integers. We consider the 
-function $\Sigma$ defined on the manifold $\mathcal C = \Grass(E,d_1) 
-\times \Grass(E,d_2)$ by $\Sigma(V_1, V_2) = V_1 + V_2$. As before, in 
-order to study $\Sigma$, we stratify $\mathcal C$ according to the 
-dimension of the sum: for each integer $d \in [0, d_1+d_2]$, we define 
-$\mathcal C_d$ as the submanifold of $\mathcal C$ consisting of those 
-pairs $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get a 
-well-defined mapping $\mathcal C_d \to \Grass(E,d)$ whose differential
-can be computed as before. The
-tangent space of $\mathcal C_d$ at a given point $(V_1, V_2)$ consists 
-of pairs $(dv_1, dv_2) \in \Hom(V_1, E/V_1)
-\times \Hom(V_2, E/V_2)$ such that $dv_1 \equiv dv_2 \pmod{V_1 + V_2}$ 
-on the intersection $V_1 \cap V_2$, and the differential of $\Sigma$
-at $(V_1, V_2)$ maps $(dv_1, dv_2)$ to $dv \in \Hom(V, E/V)$ (with $V
-= V_1 + V_2$) defined by $dv(v_1 + v_2) = dv_1(v_1) + dv_2(v_2)$ ($v_1
-\in V_1$, $v_2 \in V_2$).
-
-Using duality, we derive a similar result for the mapping 
-$(V_1, V_2) \mapsto V_1 \cap V_2$ (left to the reader).
-
-\subsection{Implementation and experiments}
-\label{grassimpl}
-
-\noindent
-{\bf Standard representation of vector spaces.}  %%%%%%%%% Repetitive
-One commonly represents subspaces of $K^n$  
-using the charts $\mathcal U_{V_I, V_{I^\c}}$ (where $I$ is a subset 
-of $\{1, \ldots, n\}$) introduced above. More concretely,
-a subspace $V \subset K^n$ is represented as the span of the rows
-of a matrix $G_V$ having the following extra property: there 
-exists some $I \subset \{1, \ldots, n\}$ such that the submatrix of 
-$G_V$ obtained by keeping only columns with index in $I$ is the identity 
-matrix. We recall that such a representation always exists and, when 
-the set of indices $I$ is fixed, at most one $G_V$ satisfies the above condition.
-Given a family of generators of $V$, one can compute $G_V$ and $I$ as 
-above by performing standard row reduction.  Choosing the first non-vanishing
-pivot at every stage provides a canonical choice for $I$, but in the context of
-inexact base fields, choosing the pivot with the maximal norm yields
-a more stable algorithm.
-
-\smallskip
-
-\noindent
-{\bf The dual representation.}
-Of course, one may alternatively use the charts $\mathcal U^\star_{V_I, 
-V_{I^\c}}$. Concretely, this means that we represent $V$ as the left 
-kernel of a matrix $H_V$ having the following extra property: there 
-exists some $I \subset \{1, \ldots, n\}$ such that the submatrix of 
-$H_V$ obtained by \emph{deleting} rows with index in $I$ is the identity 
-matrix. As above, we can then compute $I$ and $H_V$ by performing 
-column reduction.
-
-Note that switching representations is cheap and stable.
-If $I = \{1, \ldots, d\}$ with $d = \dim V$ and $I_d$ 
-is the identity matrix of size $d$, the matrix $G_V$ has the form
-$(\begin{matrix} I_d & G'_V \end{matrix})$.
-One can represent $V$ with the same $I$ and the matrix
-$H_V = \Big(\begin{matrix} -G'_V \\ I_{n-d} \end{matrix}\Big)$.
-A similar formula exists for general $I$.
-
-\smallskip
-
-\noindent
-{\bf Operations on vector spaces.}
-The first representation we gave is well suited for the computation
-of direct images and sums. For instance, to compute $f(V)$
-we apply $f$ to each row of $G_V$, obtaining a
-family of generators of $f(V)$, and then row reduce. Dually, the
-second representation works well for computing inverse images, including 
-kernels, and intersections. Since translating between the two dual
-representations is straightforward, we get algorithms for solving both
-problems using either representation.
-
-%One can either track precision in the standard way or 
-%by using Proposition \ref{prop:precision}.  While the second method 
-%is more costly, it is also more accurate.
-
-\smallskip
-
-\noindent
-{\bf Some experiments.}
-Let us consider the example computation given in the following
-algorithm.
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm 2:} {\tt example\_vector\_space}
-
-\noindent{\bf Input:} two integers $n$ and $N$
-
-\smallskip
-
-\noindent 1.\ Set $L_0 = \langle (1 + O(2^N),  O(2^N), O(2^N)) \rangle
-\subset \Q_2^3$
-
-\noindent 2.\ {\bf for} $i=0,\dots,n-1$
-
-\noindent 3.\ \hspace{0.3cm}pick randomly $\alpha, \beta, \gamma, \delta
-\in M_{3,3}(\Z_2)$ with high precision
-
-\noindent 4.\ \hspace{0.3cm}{\bf compute} $L_{i+1} = 
-\big(\alpha(L_i) + \beta(L_i)\big) \cap \big(\gamma(L_i) + \delta(L_i)\big)$
-
-\noindent 5.\ {\bf return} $L_n$
-
-%\vspace{-1ex}\noindent\hrulefill
-
-%\medskip
-
-\noindent
-The expression \emph{high precision} on line 3 means that the precision on 
-$\alpha$, $\beta$, $\gamma$ and $\delta$ is set in such a way that it 
-does not affect the resulting precision on $L_{i+1}$.
-%
-\begin{figure}
-\begin{center}
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular}{|c|c|c|c|}
-\hline
-\multirow{3}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
-\multicolumn{3}{|c|}{Average loss of precision} \\
-\cline{2-4}
-& \raisebox{-0.05cm}{Coord-wise} & 
-\multicolumn{2}{|c|}{Lattice method} \\
-\cline{3-4}
-& \raisebox{0.04cm}{method} & 
-\hspace{0.2cm}Projected\hspace{0.2cm} & 
-\hspace{0.2cm}Diffused\hspace{0.2cm} \\
-\hline 
-$10$ & $\hphantom{0}7.3$ & $\hphantom{0}2.7$ & $\hphantom{0}{-}2.4 \times 2$ \\
-$20$ & $14.8$ & $\hphantom{0}5.5$ & $\hphantom{0}{-}4.7 \times 2$ \\
-$50$ & $38.6$ & $13.1$ & $-12.0 \times 2$ \\
-$100$ & $78.1$ & $26.5$ & $-23.5 \times 2$ \\
-\hline
-\end{tabular}
-
-\smallskip
-
-{\small
-Results for a sample of $1000$ executions (with $N \gg n$)}
-\end{center}
-\renewcommand{\arraystretch}{1}
-
-\vspace{-0.4cm}
-
-\caption{Average loss of precision in Algorithm 2}
-\label{fig:vectorspace}
-
-\vspace{-0.2cm}
-
-\end{figure}
-%
-Figure \ref{fig:vectorspace} shows the losses of precision when
-executing Algorithm 2 with various inputs $n$ (the input $N$ is
-always chosen sufficiently large so that it does not affect the
-behavior of precision). The \emph{Coord-wise} column corresponds to
-the standard way of tracking precision. On the other hand, in the two 
-last columns, the precision is tracked using lattices. The \emph{Diffused}
-column gives the amount of diffused precision, factored to be comparible
-to the Coord-wise column. 
-The fact that only negative numbers appear in this column means that we 
-are actually always gaining precision with this model! Finally, the 
-\emph{Projected} column gives the precision loss after projecting the lattice
-precision onto coordinates.
-
-\appendix
-
-\section{Proof of Proposition \ref{prop:boundLambdaf}}
-\label{app:proof}
-
-We prove Proposition \ref{prop:boundLambdaf} in the slightly more
-general context of $K$-Banach spaces.
-
-\subsection{Composite of locally analytic functions}
-
-Let $U$, $V$ and $W$ be three open subsets in $K$-Banach spaces 
-$E$, $F$ and $G$, respectively. We assume that $0 \in U$, $0 \in V$. Let 
-$f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
-$0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
-analytic around $0$ as well. Let $f = \sum_{n \geq 0} f_n$
-$g = \sum_{n \geq 0} g_n$ and $h = \sum_{n \geq 0} h_n$ be the
-analytic expansions of $f$, $g$ and $h$.
-Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
-some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$, respectively. The 
-aim of this subsection is to prove the following intermediate result.
-
-\begin{prop}
-\label{prop:boundhr}
-With the above notation, we have
-$$\Vert h_r \Vert \leq \sup_{m, (n_i)}
-  \Vert g_m \Vert \cdot \Vert f_{n_1} \Vert \cdots \Vert f_{n_m} \Vert$$
-for all nonnegative integers $r$, where the supremum is taken over all pairs $(m, (n_i))$ where $m$
-is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
-length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
-\end{prop}
-
-A computation gives the following expansion for $g \circ f$:
-\begin{equation}
-\label{eq:expansiongf}
-\sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
-G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell}, \ldots, f_{n_\ell})
-\end{equation}
-where $\binom m {\!k_1 \,\, \cdots \,\, k_\ell\!}$ denotes the
-multinomial coefficient and the sum runs over:
-
-\noindent
-(a) all finite sequences $(k_i)$ of positive integers whose length 
-(resp. sum) is denoted by $\ell$ (resp. $m$), and
-
-\noindent
-(b) all finite sequences $(n_i)$ of positive integers of length
-$\ell$.
-
-\noindent
-Moreover, in the argument of $G_m$, the variable $f_{n_i}$ 
-is repeated $k_i$ times.
-
-The degree of $G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell},
-\ldots, f_{n_\ell})$ is $r = k_1 n_1 + \ldots + k_\ell n_\ell$ and 
-then contributes to $h_r$. As a consequence $h_r$ is equal to 
-\eqref{eq:expansiongf} where the sum is restricted to sequences
-$(k_i)$, $(n_i)$ such that $k_1 n_1 + \ldots + k_\ell n_\ell = r$.
-Proposition \ref{prop:boundhr} now follows from the next lemma.
-
-\begin{lem}
-Let $E$ be a $K$-vector space. Let $\varphi : E^m \to 
-K$ be a symmetric $m$-linear form and $\psi: E \to K$ defined by 
-$\psi(x) = \varphi(x, x, \ldots, x)$.
-Given positive integers $k_1, \ldots, k_\ell$ whose sum is $m$ and 
-$x_1, \ldots, x_\ell \in E$, we have
-$$\begin{array}{l}
-\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
-\varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
-x_\ell) \Big\Vert  \\
-\hspace{4.3cm} \leq \Vert \psi \Vert \cdot \Vert x_1 \Vert^{k_1} \cdots
- \Vert x_\ell \Vert^{k_\ell}
-\end{array}$$
-where, in the LHS, the variable $x_i$ is repeated $k_i$ times.
-\end{lem}
-
-\begin{proof}
-It is enough to prove that
-$$\Big\Vert \binom m {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} \cdot
-\varphi(x_1, \ldots, x_1, \ldots, x_\ell, \ldots,
-x_\ell) \Big\Vert \leq \Vert \psi \Vert$$
-provided that all the $x_i$'s have norm at most $1$. We proceed by 
-induction on 
-$\ell$. The $\ell = 1$ case follows directly from
-the definition of $\Vert \psi \Vert$. 
-We now pick $(\ell+1)$ integers $k_1, \ldots, k_{\ell+1}$ whose sum 
-equals $m$, together with $(\ell+1)$ elements $x_1, \ldots, x_{\ell+1}$
-lying in the unit ball of $E$.
-We also consider a new variable $\lambda$ varying in
-$\O_K$. We set $x'_i = x_i$, $k'_i = k_i$ when $i < \ell$ and $x'_\ell 
-= x_\ell + \lambda x_{\ell+1}$ and $k'_\ell = k_\ell + k_{\ell+1}$. By the 
-induction hypothesis, we know that the inequality
-$$\Big\Vert \binom m {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
-\varphi(x'_1, \ldots, x'_1, \ldots, x'_\ell, \ldots, x'_\ell) \Big\Vert
-\leq \Vert \psi \Vert$$
-holds for all $\lambda \in K$. Furthermore, the LHS of the inequality
-is a polynomial $P(\lambda)$ of degree $k'_\ell$ whose coefficient in 
-$\lambda^j$ is
-$$\binom m {k'_1 \,\, \cdots \,\, k'_\ell} \cdot
-\binom {k'_\ell} {j} \cdot
-\varphi(\underline x_j) = 
-\binom m {k_1 \,\, \cdots \,\, k_{\ell-1} \,\, j} \cdot
-\varphi(\underline x_j)$$
-with
-$\underline x_j = (x_1, \ldots, x_1, \ldots, x_{\ell+1}, \ldots, 
-x_{\ell+1})$
-where $x_i$ is repeated $k_i$ times if $i < \ell$ and $x_\ell$ 
-(resp. $x_{\ell+1}$) is repeated $j$ times (resp. $k'_\ell - j$ times).
-Since $\Vert P(\lambda) \Vert \leq \Vert \psi \Vert$ for all $\lambda$ in 
-the unit ball, the norm of all its coefficients must also be at most $\Vert \psi
-\Vert$. From the coefficient of $\lambda^{k_\ell}$, the result follows.
-\end{proof}
-
-\subsection{Bounding a growing function}
-
-We return to the setting of Proposition \ref{prop:boundLambdaf}.
-Let $f = \sum_{n \geq 0} f_n$, $g = \sum_{n \geq 0} g_n$ and
-$h = \sum_{n \geq 0} h_n$ be the analytic expansions of
-$f$, $g$ and $h$.
-Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
-some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$, respectively.
-We recall that $\Lambda(f)$ 
-is the Legendre transform of the Newton polygon $\NP(f)$ defined in Section \ref{sec:matrices}
-\cite[Proposition 3.9]{caruso-roe-vaccon:14a}, and that $\alpha$ is a
-real number such that $\Vert n! \Vert \geq e^{-\alpha n}$ for all positive integers $n$.
-
-\begin{lem}
-\label{lem:boundLambdaf}
-We keep the above notation. 
-If $(a,b)$ satisfies
-$b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)$
-then $b \geq \Lambda(f)(a - \alpha)$.
-\end{lem}
-
-\begin{proof}
-We have $f' = \sum_{n \geq 0} f'_n$ where
-$$f'_n : U \to \mathcal L(E,F), \quad
-x \mapsto \big(h \mapsto n \cdot F_n(h, x, x, \ldots, x)\big).$$
-Taking $h = x$, we find
-$\Vert f'_n \Vert \geq \Vert n f_n \Vert = 
-\Vert n \Vert \cdot \Vert f_n \Vert$.
-Combining this with Proposition \ref{prop:boundhr}, we get
-$$\Vert (r+1) f_{r+1} \Vert \leq
-  \sup_{m, (n_i)} \Vert g_m \Vert \cdot 
-  \prod_{i=1}^m \max(\Vert f_{n_i} \Vert, \Vert h_{n_i} \Vert)$$
-for all nonnegative integers $r$,
-where the supremum runs over all pairs $(m, (n_i))$ where $m$
-is a nonnegative integer and $(n_i)_{1 \leq i \leq m}$ is a sequence of
-length $m$ of nonnegative integers such that $n_1 + \ldots + n_m = r$.
-We set $u_r = \Vert r! f_r \Vert$. Multiplying the above inequality by
-$\Vert r! \Vert$, we obtain:
-\begin{equation}
-\label{eq:boundurrec}
-u_{r+1} \leq
-  \sup_{m, (n_i)} \Vert g_m \Vert \cdot 
-  \prod_{i=1}^m \max(u_{n_i}, \Vert n_i! h_{n_i} \Vert)
-\end{equation}
-since the multinomial coefficient $\binom r {\!n_1 \, \cdots \, n_m\!}$
-is an integer and hence has norm at most $1$.
-We now pick two real numbers $a$ and $b$ satisfying the hypothesis of 
-the Lemma. Set $d = \Lambda(h)(a)$. Going back to the definitions of 
-$\Lambda (h)$ and Legendre transform, we get $\Vert h_n \Vert \leq e^{- 
-a n + d}$ for all $n$.
-Similarly, from our hypothesis on $(a,b)$, we find
-$\Vert g_m \Vert \leq e^{-\!\max(b,d)\cdot m + b - a}$ for all $m$.
-We are now ready to prove $u_r \leq e^{-ar + b}$ by induction on $r$.
-When $r = 0$, it is obvious because $u_0$ vanishes. Otherwise, the
-induction follows from
-\begin{align*}
-u_{r+1} 
-& \leq \sup_{m, (n_i)}
-    e^{ -\max(b,d)\cdot m + b - a + \sum_{i=1}^m (-a n_i + \max(b,d))} \\
-& = e^{ b - a - a r } = e^{ -a (r+1) + b}.
-\end{align*}
-From the definition of $u_r$, we obtain 
-$\Vert f_r \Vert \leq u_r \cdot \Vert r! \Vert^{-1} \leq e^{-(a - \alpha) 
-r + b}$. Thus $b \geq \Lambda(f)(a - \alpha)$.
-\end{proof}
-
-%\begin{rem}
-%A simpler proof is possible if $\Z \subset \O_K$, \emph{i.e.} if
-%one can choose $\alpha = 0$. Indeed, under this extra assumption,
-%one can prove using Eq.~\eqref{eq:normderivative} that:
-%$$\Lambda(f') \geq \Lambda(f) - \id.$$
-%Combining this with Lemma 3.7 of \cite{caruso-roe-vaccon:14a}, we find:
-%$$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
-%from which we can easily derive Lemma \ref{lem:boundLambdaf}.
-%\end{rem}
-
-We can now conclude the proof of Proposition \ref{prop:boundLambdaf} as 
-follows. Given $a \in \R$ and $b = a + \Lambda_g \circ \Lambda_h(a)$, we 
-have to prove that $\Lambda(f)(a-\alpha) \leq b$ provided that $b \leq 
-\nu$. Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to 
-check that such pairs $(a,b)$ satisfy the hypothesis of Lemma
-\ref{lem:boundLambdaf}. Clearly:
-$b \geq a + \Lambda(g) \circ \Lambda(h)(a)$
-since $\Lambda_g \geq \Lambda(g)$, $\Lambda_h \geq \Lambda(h)$ and
-$\Lambda_g$ is nondecreasing. Furthermore, from $b \leq \nu$, we get
-$\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
-\Lambda_h(a)$, from which we derive 
-$a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) = b$.
-
-\bibliographystyle{plain}
-\bibliography{../roebib/Biblio,extras}
+\section{Slope factorization}
 

### 2015-11-29 polynomials/muldivfact.tex
First steps for Slope Factorization
First formulae and NP drawing toward the proof of Slope Factorization

@@
 \DeclareMathOperator{\rank}{rank}
+
+\newcommand{\slp}{\text{\rm slp}}
+\newcommand{\RS}{\text{\rm RS}}
 
@@
 \alignauthor David Roe \\
-  \affaddr{University of British Columbia}\\
+  \affaddr{University of Pittsburgh}\\
   \affaddr{\textsf{roed.math@gmail.com}}
 \alignauthor Tristan Vaccon\\
-  \affaddr{Universit\'e Rennes 1}\\
-  \affaddr{\textsf{tristan.vaccon@univ-rennes1.fr}}
+  \affaddr{Rikkyo University}\\
+  \affaddr{\textsf{vaccon@rikkyo.ac.jp}}
 }
@@
 % \noindent
-% {\bf Categories and Subject Descriptors:} \\
+% {\bf Categories and Subject Descripto\RS:} \\
 %\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
@@
 
+\begin{theo} \label{theo:EDivisionNP}
+Here should be the theorem for the Newton polygons during a Euclidean division.
+\end{theo}
+
 \section{Slope factorization}
+\subsection{Main theorem}
+
+\begin{theo} \label{theo:slope-factor}
+Under some assumptions... $\textrm{coeff}(P,d)=1,$ $\NP (A_0)' <0$ (???)
+
+\begin{align*}
+A_0 &= \textrm{trunc}(P,d), \\
+B_{0} &= P \div A_{0}, \\
+V_{0} &= 1.
+\end{align*}
+
+\begin{align*}
+A_{i+1} &= A_i + (V_i P \% A_i), \\
+B_{i+1} &= P \div A_{i+1}, \\
+V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \% A_{i+1}.
+\end{align*}
+$A_i$ convergences to $A_\infty$ such that $\NP (A_\infty)= \NP (A_0)$ and $A_\infty$ divides $P.$
+\end{theo}
+
+\begin{rem}
+Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in \ref{theo:slope-factor} do not induce any loss in generality.
+\end{rem}
+
+\subsection{Useful formulae}
+
+\begin{lem}
+Some definitions :
+\begin{align}
+R_i &= A_{i+1} -A_i, \label{eqdef:Ri}\\
+S_i &= P \% A_i, \label{eqdef:Si} \\
+Q_i &= V_i P \div A_i. \label{eqdef:Qi}. 
+\end{align}
+Some formulae :
+\begin{align}
+B_i-B_{i+1} &= R_i B_{i+1} \div A_i, \label{eqdef:Biminus} \\
+S_i-S_{i+1} &= R_i B_{i+1} \div A_i,  \label{eqdef:Siminus} \\
+S_{i} &= (B_i R_i + (1-V_i B_i) S_{i-1}+(1-V_i B_i)(S_i-S_{i-1})) \% A_{i}, \label{eqdef:Si2}  \\
+V_{i+1}-V_i &= (V_i (1-V_i B_i)+V_i^2(B_i-B_{i+1})) \% A_{i}, \label{eqdef:Viminus}\\
+1-Q_i &= (1-V_i B_i)+((R_i-V_i S_i) \div A_i), \label{eqdef:Qiminus}\\
+R_{i+1} &= ((V_{i+1}-V_i)S_{i+1}+(1-Q_i)R_i) \% A_{i+1}, \label{eq:Riplus}\\
+1-V_{i+1}B_{i+1} &= ((1-V_i B_i)+V_i (B_i-B_{i+1}))^2. \label{eq:ViBiplus}
+\end{align}
+\end{lem}
+\begin{proof}
+We first remark that by definition,
+\begin{align*}
+P &= A_i B_i + S_i, \\
+P V_i&= A_i Q_i + R_i.
+\end{align*}
+Since $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1},$ we have \[R_i B_{i+1}= (B_i -B_{i+1})A_i + (S_i+S_{i+1}).\]
+Hence, by consideration of degree, what we have written is the Euclidean division of $R_i B_{i+1}$ by $A_i$ and we obtain \eqref{eqdef:Biminus} and \eqref{eqdef:Siminus}.
+
+From $V_i P =  A_i Q_i + R_i=V_i (A_i B_i + S_i),$ we get \begin{equation}(V_i B_i -Q_i) A_i=R_i-V_i S_i. \label{eq:ViBiQi} \end{equation} Thus, \[ R_i=V_i S_i \% A_i.\]
+Hence, $B_i R_i =S_i (V_iB_i-1)+S_i \% A_i,$ and $S_i=B_i R_i-S_i (V_iB_i-1) \% A_i.$ \eqref{eqdef:Si2} follows directly.
+
+By definition of $V_i,$ we get $V_{i+1}-V_i=V_i (1-V_i B_{i+1}) \% A_{i+1}.$ \eqref{eqdef:Viminus} follows immediately.
+
+We can write $1-Q_i=(1-V_i B_i)+(V_i B_i - Q_i).$ Using \eqref{eq:ViBiQi}, we get \eqref{eqdef:Qiminus}.
+
+We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i,$ and $V_{i+1} P=A_{i+1} Q_{i+1}+R_{i+1}.$ As a consequence, 
+\begin{align*}
+R_{i+1} &= (V_{i+1}-V_i)P+(1-Q_i)R_i \\
+ &= (V_{i+1}-V_i)S_{i+1}+(1-Q_i)R_i) \% A_{i+1},
+\end{align*} and \eqref{eq:Riplus} is proved.
+
+Finally, \begin{align*}
+1-V_{i+1}B_{i+1} &= 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \% A_{i+1}, \\
+&= (1-V_i B_{i+1})^2 \% A_{i+1}, \\
+&= ((1-V_i B_i)+V_i (B_i-B_{i+1}))^2 \% A_{i+1},
+\end{align*} which concludes the proof.
+\end{proof}
+
+\begin{deftn}
+Notation:
+\[P=-d \lambda_0+d \slp(\lambda_0) + l_1 \slp (\lambda_1)+\RS \]
+($\RS$ for remaining slopes)
+
+\begin{tikzpicture}
+\draw[->] (0,-1) -- (0,2);
+
+\draw[->] (-0.05,0) -- (4,0);
+
+\node[left] at (0,0) {$0$};
+
+\draw (1.5,-.1) -- (1.5,.1);
+\node[above] at (1.5,0) {$d$};
+
+\draw (3,-.1) -- (3,.1);
+\node[above] at (2.8,0) {$d+l_1$};
+
+\node[left] at (0,1.5) {$-d \lambda_0$};
+
+\draw (0,1.5)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
+\node[below] at (.75,0.5) {$\lambda_0$};
+\node[below] at (2.25,-0.25) {$\lambda_1$};
+\node[right] at (3.5,0.25) {$\RS$};
+\end{tikzpicture}
+\end{deftn}
+
+\subsection{Proof of the Main Theorem}
+
+\subsubsection{Initial Newton polygons}
+
+\begin{lem}
+$\NP(R_0)=\NP(S_0)$:
+\[ \NP (R_0)=\NP(S_0)=(\lambda_1-\lambda_0)- d \lambda_0+(d-1)\slp( \lambda_0). \]
+
+\begin{tikzpicture}
+\draw[->] (0,-.5) -- (0,2.5);
+
+\draw[->] (-0.05,0) -- (2,0);
+
+\draw[->] (-0.05,1.75) -- (.05,1.75);
+\node[left] at (0,1.75) {$(\lambda_1-\lambda_0)$};
+\node[left] at (-.4,1.4) {$-d \lambda_0$};
+
+\node[left] at (0,0) {$0$};
+
+\draw (1,-.1) -- (1,.1);
+\node[below, right] at (1,-.2) {$d-1$};
+
+\draw (0,1.75) -- (1,.75) -- (1,2.5);
+\node[below] at (.5,1.3) {$\lambda_0$};
+\end{tikzpicture}
+
+$\NP (1-V_0B_0 \% A_0)$:
+\begin{align*}
+\NP ((1-V_0B_0)\% A_0) &\geq (\lambda_1-\lambda_0)+1\slp( \lambda_0)+(d-1) \slp(\lambda_1), \\
+&\geq (\lambda_1-\lambda_0)+d\slp( \lambda_0). \\
+\end{align*} 
+\begin{tikzpicture}
+\draw[->] (0,-1) -- (0,1);
+
+\draw[->] (-0.05,0) -- (2.5,0);
+
+\node[left] at (0,0) {$0$};
+
+\node[above] at (1.5,0) {$d-1$};
+
+\draw (.5,-.1) -- (.5,.1);
+\node[above] at (.5,0) {$1$};
+
+\draw (-.1,.25) -- (.1,.25);
+\node[left] at (0,.25) {$(\lambda_1-\lambda_0)$};
+
+\draw (0,.25) -- (.5,-.25) -- (1,-.375)-- (1,1.5);
+\draw (0,.25) -- (.5,-.25) -- (1,-.75)-- (1,1.5);
+
+\node[below] at (.25,-0.1) {$\lambda_0$};
+\node[below] at (.75,0.1) {$\lambda_1$};
+\end{tikzpicture}
+
+\end{lem}
+
+\subsubsection{Unchanged Newton polygons}
+
+\begin{lem}
+$\NP (V_i)$ for $i>0$ ($V_0=1$):
+\[\NP(V_i)=0+(d-1)\slp( \lambda_1). \]
+
+\begin{tikzpicture}
+\draw[->] (0,-1) -- (0,1);
+
+\draw[->] (-0.05,0) -- (2,0);
+
+\node[left] at (0,0) {$0$};
+
+\node[below, right] at (1,-.2) {$d-1$};
+
+\draw (0,0) -- (1,-.25) -- (1,1);
+\node[below] at (.5,-0.25) {$\lambda_1$};
+\end{tikzpicture}
+
+$\NP (B_i)$:
+\[\NP(B_i)=0+(l_1)\slp( \lambda_1)+\RS. \]
+
+\begin{tikzpicture}
+\draw[->] (0,-1) -- (0,1);
+
+\draw[->] (-0.05,0) -- (2.5,0);
+
+\node[left] at (0,0) {$0$};
+
+\draw (1.5,-.1) -- (1.5,.1);
+\node[above] at (1.5,0) {$l_1$};
+
+\draw (0,0) -- (1.5,-.375) -- (2,.625);
+\node[below] at (.5,-0.25) {$\lambda_1$};
+\node[below] at (2,-0.25) {$\lambda_2$};
+\end{tikzpicture}
+\end{lem}
+
+\subsubsection{Newton polygons inside the loop}
+
+\begin{lem}
+\begin{align*}
+\NP (R_i), \NP(S_i) & \geq -\lambda_0+2^i(\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0), \\
+\NP (1-V_i B_i) & \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
+\NP (B_i-B_{i+1}) & \geq 2^i (\lambda_1-\lambda_0)+(\lambda_1-\lambda_0)+(l_1-1) \slp (\lambda_1)+\RS, \\
+\NP (S_i-S_{i+1}) & \geq -\lambda_0 d+2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
+\NP (V_i-V_{i+1}) & \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
+\NP (1-Q_i) & \geq  2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\\
+\end{align*}
+\end{lem}
 

### 2015-12-04 polynomials/muldivfact.tex
Proof of main algorithm
Getting closer to a first version of the proof.

@@
 \begin{rem}
-Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in \ref{theo:slope-factor} do not induce any loss in generality.
+Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in Theorem \ref{theo:slope-factor} do not induce any loss in generality.
+\end{rem}
+
+\begin{rem}
+Non necessity to have $A_0$ as the first slope. It is possible to use minorations.
+
+\begin{tikzpicture}
+\draw[->] (0,-1) -- (0,2);
+
+\draw[->] (-0.05,0) -- (4,0);
+
+\node[left] at (0,0) {$0$};
+
+\draw (1.5,-.1) -- (1.5,.1);
+\node[above] at (1.5,0) {$d$};
+
+\draw (0,1)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
+\draw[green] (0,1.5)-- (1,0.33)-- (1.5,0) --(2,-.125)-- (3,0)-- (3.5,1);
+
+\node[below] at (.75,-0.2) {$\lambda_0$};
+\end{tikzpicture}
 \end{rem}
@@
 
-\begin{lem}
+\begin{lem} \label{lem:formulae}
 Some definitions :
@@
 
+\label{lem:initial_NP}
 \end{lem}
+\begin{proof}
+Clear by Theorem \ref{theo:EDivisionNP}, and the fact that $coeff(P,d)=1.$
+\end{proof}
 
@@
 
-\begin{lem}
+\begin{lem} \label{lem_unchanged_NP}
 $\NP (V_i)$ for $i>0$ ($V_0=1$):
@@
 \end{tikzpicture}
+
+$\NP (A_i)$:
+\[\NP(A_i)=-d \lambda_0+d \slp( \lambda_0). \]
+
+\begin{tikzpicture}
+\draw[->] (0,-1) -- (0,2);
+
+\draw[->] (-0.05,0) -- (2,0);
+
+\node[left] at (0,0) {$0$};
+
+\draw (1.5,-.1) -- (1.5,.1);
+\node[above] at (1.5,0) {$d$};
+
+\node[left] at (0,1.5) {$-d \lambda_0$};
+
+\draw (0,1.5)-- (1.5,0) ;
+\node[below] at (.75,0.5) {$\lambda_0$};
+\end{tikzpicture}
 \end{lem}
@@
 
-\begin{lem}
+\begin{lem} \label{lem:iteration_inside_the_loop}
 \begin{align*}
@@
 \NP (1-V_i B_i) & \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
-\NP (B_i-B_{i+1}) & \geq 2^i (\lambda_1-\lambda_0)+(\lambda_1-\lambda_0)+(l_1-1) \slp (\lambda_1)+\RS, \\
-\NP (S_i-S_{i+1}) & \geq -\lambda_0 d+2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
-\NP (V_i-V_{i+1}) & \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
-\NP (1-Q_i) & \geq  2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\\
 \end{align*}
@@
 
+We prove the last two lemmas by induction.
+
+\begin{proof}
+The case $i=0$ is clear thanks to Theorem \ref{theo:EDivisionNP} and Lemma \ref{lem:initial_NP}.
+
+Now, we assume that for some $i \geq 0$ the formulae defined in the Lemmae \ref{lem_unchanged_NP} and \ref{lem:iteration_inside_the_loop} are satisfied. 
+We prove them for $i+1$.
+
+Firstly, thanks to the minoration on $\NP (R_i)$ and the fact that $R_i=A_{i+1}-A_i$, we have $\NP (A_{i+1}= \NP (A_i).$
+Since $B_{i+1} = P \div A_{i+1},$ we have $\NP (B_{i+1}) = \NP (B_i).$
+We apply \eqref{eqdef:Biminus} to obtain 
+\[\NP (B_i-B_{i+1})  \geq 2^i (\lambda_1-\lambda_0)+(\lambda_1-\lambda_0)+(l_1-1) \slp (\lambda_1)+\RS. \]
+Then, we can apply \eqref{eqdef:Siminus} to obtain 
+\[\NP (S_i-S_{i+1})  \geq -\lambda_0 d+2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
+
+As a consequence, we get that $\NP (S_{i+1}) \geq \NP (S_i)$ (pointwise).
+
+Now, we can apply \eqref{eqdef:Viminus} to obtain that 
+\[\NP (V_i-V_{i+1}) \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
+As a consequence, we get that $\NP(V_i)=\NP(V_{i+1}).$
+Similarly, we use \eqref{eqdef:Qiminus} to obtain that
+\[\NP (1-Q_i)  \geq  2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
+
+This is enough to conclude on the Newton polygon of $R_{i+1}.$
+Indeed, thanks to \eqref{eq:Riplus}, we obtain that 
+\[\NP (R_i) \geq -\lambda_0+2^i(\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0). \]
+We can proceed similarly for $1-V_{i+1} B_{i+1}$ using \eqref{eq:ViBiplus}.
+
+Finally, we can apply \eqref{eqdef:Si2} for $i+1,$
+\[S_{i+1} = (B_{i+1} R_{i+1} + (1-V_{i+1} B_{i+1}) S_{i}+(1-V_{i+1} B_{i+1})(S_{i+1}-S_{i})) \% A_{i+1},\] to obtain the desired minoration on $\NP (S_{i+1}).$ This concludes the proof.
+\end{proof}
+
 \end{document}

### 2015-12-23 polynomials/muldivfact.tex
Definition of jagged and Newton precision

@@
 \renewcommand{\prec}{\text{\rm prec}}
+\newcommand{\NF}{\text{\rm NF}}
 
@@
 
-\section{Modular multiplication}
+\noindent
+{\bf Notations.}
+Throughout this paper, we fix a complete discrete valuation field $K$; 
+we denote by $\val : K \to \Z \cup \{+\infty\}$ the valuation on it and 
+by $W$ its ring of integers (\emph{i.e.} the set of elements with 
+nonnegative valuation). We assume that $\val$ is normalized so that it
+is surjective and denote by $\pi$ a uniformizer of $K$, that is an 
+element of valuation $1$. Denoting by $S \subset W$ a fixed set of 
+representatives of the classes modulo $\pi$ and assuming $0 \in S$, 
+one can prove that each element in $x \in K$ can be represented 
+uniquely as a convergent series:
+\begin{equation}
+\label{eq:CDVFseries}
+x = \sum_{i=\val(x)}^{+\infty} a_i \pi^i
+\quad \text{with} \quad a_i \in S.
+\end{equation}
+The two most important examples are the field of $p$-adic numbers $K = 
+\Qp$ and the field of Laurent series $K = k((t))$ over a field $k$. The 
+valuation on them are the $p$-adic valuation and the usual valuation of 
+a Laurent series respectively. Their ring of integers are therefore 
+$\Zp$ and $k[[t]]$ respectively. A distinguished uniformizer is $p$ 
+and $t$ whereas a possible set $S$ is $\{0, \ldots, p-1\}$ and $k$
+respectively.
+The reader who is not familiar with complete discrete valuation fields
+may assume (without sacrifying too much to the generality) that $K$ is
+one of the two aforementioned examples.
+
+In what follows, the notation $K[X]$ refers to the ring of univariate 
+polynomials with coefficients in $K$. The subspace of polynomials of 
+degree at most $n$ (resp. exactly $n$) is denoted by $K_{\leq n}[X]$ 
+(resp. $K_{=n}[X]$).
+
+\section{Precision data}
+
+Elements in $K$ (and \emph{a fortiori} in $K[X]$) carry an infinite 
+amount of information. They thus cannot be stored entirely in the
+memory of a computer and have to be truncated. Elements of $K$ are
+usually represented by truncating Eq.\eqref{eq:CDVFseries} as
+follows:
+\begin{equation}
+\label{eq:CDVFseriesO}
+x = \sum_{i=v}^{N-1} a_i \pi^i + O(\pi^N)
+\end{equation}
+where $N$ is an integer called the \emph{absolute precision} and 
+the notation $O(\pi^N)$ means that the coefficients $a_i$ for $i
+\geq N$ are discarded. If $N > v$ and $a_v \neq 0$, the integer $v$ 
+is the valuation of $x$ and the difference $N-v$ is called the
+\emph{relative precision}.
+Alternatively, one may thing that the writing~\eqref{eq:CDVFseriesO}
+represents a subset of $K$ which consists of all elements in $K$ for
+which the $a_i$'s in the range $[v,N-1]$ are those specified. From the
+metric point of view, this is a ball (centered at any point inside it).
+
+It worths be noting that tracking precision using this representation is 
+rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
+relative) precision $N_x$ and $N_y$ respectively, one can compute the 
+sum $x+y$ (resp. the product $xy$) at absolute (resp. relative) 
+precision $\min(N_x,N_y)$. Computations with $p$-adic and Laurent
+series are often handle this way on symbolic computation softwares.
+
+\subsection{Precision for polynomials}
+
+The situation is much more subtle when we are working with a collection 
+of elements of $K$ (\emph{e.g.} a polynomial) and not just a single one.
+Indeed, several precision data may be considered and, as we shall see
+later, each of them has is own interest. Below we detail three of them
+for the special case of polynomials.
+
+\medskip
+
+\noindent
+{\bf Jagged precision.}
+The most naive precision we may consider is the so-called \emph{jagged 
+precision}. It consists in attaching an individual precision (of the
+shape $O(\pi^{N_i})$) to each coefficient separatedly.
+
+\medskip
+
+\noindent
+{\bf Newton precision.} \todo{Write a one-line introduction.}
+
+\begin{deftn}
+A \emph{Newton function of degree $n$} is a convex function 
+$f : [0,n] \to \R$ which is piecewise affine and whose epigraph
+$\Epi(f)$ have integral extremal points.
+\end{deftn}
+
+\begin{rem}
+The datum of $f$ is equivalent to that of $\Epi(f)$ and they can
+easily represented and manipulated on a computer.
+\end{rem}
+
+We recall that one can attach a Newton function to each polynomial.
+If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
+polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
+($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
+and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
+function whose epigraph is $\NP(P)$. It is well known \todo{(Give a
+reference?)} that:
+$$\begin{array}{r@{\,\,}c@{\,\,}l}
+\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
+\NP(PQ) & = & \NP(P) + \NP(Q)
+\end{array}$$
+where $\text{Conv}$ denotes the convex hull and the plus sign stands 
+for the Minkowski sum. This translates to:
+$$\begin{array}{r@{\,\,}c@{\,\,}l}
+\NF(P+Q) \geq \NF(P) \oplus \NF(Q) \smallskip \\
+\NF(PQ) = \NF(P) \odot \NF(Q)
+\end{array}$$
+where the operations $\oplus$ and $\odot$ are defined accordingly.
+It turns out that there exist efficient algorithms for computing these 
+operations.
+
+In a similar fashion, Newton functions can be used to model precision: 
+given a Newton function $f$ of degree $n$, we agree that a polynomial 
+of degree at most $n$ is given at precision $O(f)$ when, for all $i$,
+its $i$-th coefficient is given at precision $O\big(\pi^{\lceil f(i)
+\rceil}\big)$ (where $\lceil \cdot \rceil$ is the ceiling function).
+In the sequel, we shall write
+$O(f) = \sum_{i=0}^n O\big(\pi^{\lceil f(i) \rceil}\big) \cdot X^i$
+and use the notation $\sum_{i=0}^n a_i X^i + O(f)$
+to refer to a polynomial given at precision $O(f)$.
+
+We observe in particular that Newton precision data are particular
+instances of jagged precision. The main interest of using them is the
+efficiency.
+
+\medskip
+
+\noindent
+{\bf Lattice precision.}
 
@@
 \end{theo}
+
+\section{Modular multiplication}
 

### 2015-12-23 polynomials/muldivfact.tex
Definition of lattice precision

@@
 \noindent
-{\bf Newton precision.} \todo{Write a one-line introduction.}
+{\bf Newton precision.} 
+We now move to \emph{Newton precision} data. They can be actually seen 
+as particular instances of jagged precision but there exist for them 
+better representations and better algorithms.
 
@@
 A \emph{Newton function of degree $n$} is a convex function 
-$f : [0,n] \to \R$ which is piecewise affine and whose epigraph
-$\Epi(f)$ have integral extremal points.
+$\varphi : [0,n] \to \R$ which is piecewise affine and whose epigraph
+$\Epi(\varphi)$ have integral extremal points.
 \end{deftn}
@@
 \begin{rem}
-The datum of $f$ is equivalent to that of $\Epi(f)$ and they can
-easily represented and manipulated on a computer.
+The datum of $\varphi$ is equivalent to that of $\Epi(\varphi)$ and they 
+can easily represented and manipulated on a computer.
 \end{rem}
@@
 In a similar fashion, Newton functions can be used to model precision: 
-given a Newton function $f$ of degree $n$, we agree that a polynomial 
-of degree at most $n$ is given at precision $O(f)$ when, for all $i$,
-its $i$-th coefficient is given at precision $O\big(\pi^{\lceil f(i)
+given a Newton function $\varphi$ of degree $n$, we agree that a polynomial 
+of degree at most $n$ is given at precision $O(\varphi)$ when, for all $i$,
+its $i$-th coefficient is given at precision $O\big(\pi^{\lceil \varphi(i)
 \rceil}\big)$ (where $\lceil \cdot \rceil$ is the ceiling function).
 In the sequel, we shall write
-$O(f) = \sum_{i=0}^n O\big(\pi^{\lceil f(i) \rceil}\big) \cdot X^i$
-and use the notation $\sum_{i=0}^n a_i X^i + O(f)$
-to refer to a polynomial given at precision $O(f)$.
+$O(\varphi) = \sum_{i=0}^n O\big(\pi^{\lceil \varphi(i) \rceil}\big) \cdot X^i$
+and use the notation $\sum_{i=0}^n a_i X^i + O(\varphi)$ (where the
+coefficients $a_i$ are given by truncated series) to refer to a 
+polynomial given at precision $O(\varphi)$.
 
-We observe in particular that Newton precision data are particular
-instances of jagged precision. The main interest of using them is the
-efficiency.
+It is easily checked that if $P$ and $Q$ are two polynomials known at 
+precision $O(\varphi_P)$ and $O(\varphi_Q)$ respectively, then $P+Q$ is 
+known at precision $O(\varphi_P \oplus \varphi_Q)$ and $PQ$ is known at 
+precision $O\big(\varphi_P \odot \NF(Q) \oplus \NF(P) \odot 
+\varphi_Q\big)$.
 
@@
 {\bf Lattice precision.}
+The notion of \emph{lattice precision} was developed in 
+\cite{padicprec}. It encompasses the two previous models and has
+the decisive advantage of optimality. As a counterpart, it might be
+very space-consuming and time-consuming for polynomials of large
+degree.
+
+\begin{deftn}
+Let $V$ be a finite dimensional vector space over $K$. A lattice
+in $V$ is a sub-$W$-module of $V$ generated by a $K$-basis of
+$V$.
+\end{deftn}
+
+\noindent
+We fix an integer $n$. A lattice precision datum for a polynomial of 
+degree $n$ is a lattice $H$ lying in the vector space $K_{\leq n}[X]$. 
+We shall sometimes denote it $O(H)$ in order to emphasize that it should 
+be considered as a precision datum. The notation $P(X) + O(H)$ then 
+refers to any polynomial in the $W$-affine space $P(X) + H$. Tracking
+lattice precision can be done using differentials as shown in
+\cite[Lemma~3.4 and Proposition~3.12]{padicprec}: if $f : K_{\leq n}[X] 
+\to K_{\leq m}[X]$ denotes any strictly differentiable function with
+surjective differential, under mild assumption on $H$, we have:
+$$f(P(X)+H) = f(P(X)) + f'(P(X))(H)$$
+where $f'(P(X))$ denotes the differential of $f$ at $P(X)$. The
+equality sign reflets the optimality of the method.
+
+As already mentioned, the jagged precision model is a particular case
+of the lattice precision. Indeed, a precision of the shape
+$\sum_{i=0}^n O(\pi^{N_i}) X^i$ corresponds to the lattice
+generated by the elements $\pi^{N_i} X^i$ ($0 \leq i \leq n$).
 

### 2016-01-03 polynomials/muldivfact.tex

@@
 
+\renewcommand{\mod}{\,\%\,}
+\renewcommand{\div}{\,/\hspace{-0.3em}/\,}
+\newcommand{\nfop}[1]{%
+\raisebox{-0.1mm}{\begin{tikzpicture}
+\draw[transparent] (-0.18,0)--(0.18,0);
+\draw[very thin] (0,0) circle (0.11cm);
+\begin{scope}
+\clip (0,0) circle (0.11cm);
+\node[scale=0.75] at (0,0) { $#1$ };
+\end{scope}
+\end{tikzpicture}}}
+\newcommand{\nfplus}{\nfop+}
+\newcommand{\nftimes}{\nfop\times}
+\newcommand{\nfmod}{\nfop\mod}
+\newcommand{\nfdiv}{\nfop\div}
+
 \newcommand{\id}{\textrm{id}}
@@
 \newcommand{\col}{\: : \:}
+
+\newcommand{\tinyplus}{\raisebox{0.4mm}{\hspace{0.4mm}\tiny +\hspace{0.4mm}}}
 
@@
 
-\title{Multiplication, division and factorization\\of p-adic polynomials}
+\title{Division and Factorization of p-Adic Polynomials}
 
@@
 A \emph{Newton function of degree $n$} is a convex function 
-$\varphi : [0,n] \to \R$ which is piecewise affine and whose epigraph
-$\Epi(\varphi)$ have integral extremal points.
+$\varphi : [0,n] \to \R \cup \{+\infty\}$ which is piecewise affine, 
+which takes a finite value at $n$ and whose epigraph $\Epi(\varphi)$ 
+have integral extremal points.
 \end{deftn}
@@
 $$\begin{array}{r@{\,\,}c@{\,\,}l}
-\NF(P+Q) \geq \NF(P) \oplus \NF(Q) \smallskip \\
-\NF(PQ) = \NF(P) \odot \NF(Q)
+\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
+\NF(PQ) = \NF(P) \nftimes \NF(Q)
 \end{array}$$
-where the operations $\oplus$ and $\odot$ are defined accordingly.
-It turns out that there exist efficient algorithms for computing these 
-operations.
+where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
+There exist classical algorithms for computing these two operations
+whose complexity is quasi-linear with respect to the degree.
 
@@
 precision $O(\varphi_P)$ and $O(\varphi_Q)$ respectively, then $P+Q$ is 
-known at precision $O(\varphi_P \oplus \varphi_Q)$ and $PQ$ is known at 
-precision $O\big(\varphi_P \odot \NF(Q) \oplus \NF(P) \odot 
-\varphi_Q\big)$.
+known at precision $O(\varphi_P \nfplus \varphi_Q)$ and $PQ$ is known at 
+precision $O\big((\varphi_P \nftimes \NF(Q)) \nfplus (\NF(P) \nftimes 
+\varphi_Q)\big)$.
+
+\begin{deftn}
+\label{def:nondeg}
+Let $P = P_0 + O(\varphi_P)$. We say that the Newton precision 
+$O(\varphi_P)$ on $P$ is \emph{nondegenerate} if $\varphi_P \geq 
+\NF(P_0)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
+$\NP(P_0)$.
+\end{deftn}
+
+We notice that, under the conditions of the above definition, the
+Newton polygon of $P$ is well defined. Indeed, if $\delta P$ is any
+polynomial whose Newton function is not less than $\varphi_P$, we
+have $\NP(P_0 + \delta P) = \NP(P_0)$.
 
@@
 
-As already mentioned, the jagged precision model is a particular case
-of the lattice precision. Indeed, a precision of the shape
-$\sum_{i=0}^n O(\pi^{N_i}) X^i$ corresponds to the lattice
-generated by the elements $\pi^{N_i} X^i$ ($0 \leq i \leq n$).
+As already mentioned, the jagged precision model is a particular case of 
+the lattice precision. Indeed, a precision of the shape $\sum_{i=0}^n 
+O(\pi^{N_i}) X^i$ corresponds to the lattice generated by the elements 
+$\pi^{N_i} X^i$ ($0 \leq i \leq n$). This remark is the origin of the 
+notion of \emph{diffused digits of precision} introduced in 
+\cite[Definition 2.3]{preclinalg}. We shall use it repeatedly in the 
+sequel in order to compare the behaviour of the three aforementioned 
+precision data in concrete situations.
 
@@
 
-\begin{theo} \label{theo:EDivisionNP}
-Here should be the theorem for the Newton polygons during a Euclidean division.
-\end{theo}
+\todo{Add introduction.}
+In the sequel, we shall will the notation $A \div B$ (resp $A \mod B$)
+for the quotient (resp. the remainder) in the Euclidean division of $A$ 
+by $B$.
 
-\section{Modular multiplication}
+\subsection{Euclidean division of Newton functions}
 
-\section{Slope factorization}
-\subsection{Main theorem}
+\begin{deftn}
+Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
+$d$ respectively. Letting $\mu = \psi(d) - \psi(d-1)$ and $\Delta$
+be the greatest affine function of slope $\mu$ with $\Delta \leq
+\varphi$, we define:
+$$\begin{array}{rcl}
+\varphi \nfmod \psi \, : \, [0, d-1] & \to & \R \cup \{+\infty\} \smallskip \\
+x & \mapsto & \psi(x) + \Delta(d) - \psi(d) \medskip \\
+\varphi \nfdiv \psi \, : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
+x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
+\end{array}$$
+\end{deftn}
 
-\begin{theo} \label{theo:slope-factor}
-Under some assumptions... $\textrm{coeff}(P,d)=1,$ $\NP (A_0)' <0$ (???)
+\begin{figure}
+\hfill
+\begin{tikzpicture}[xscale=0.7, yscale=0.5]
+\fill[blue!30] (4,9.5)--(4,2.667)--(2,2)--(0.5,4)--(0,8)--(0,9.5)--cycle;
+\fill[yellow!50] (5,9.5)--(5,3)--(4,2.667)--(4,9.5)--cycle;
+\fill[green!30] (5,9.5)--(5,3)--(8,4)--(11,8)--(11,9.5)--cycle;
+\draw[->] (-0.5,0)--(11.5,0);
+\draw[->] (0,-0.5)--(0,9.5);
+\draw[very thick] (0,6)--(0.5,2)--(2,0)--(5,1);
+\draw[very thick] (0,9)--(3,5)--(8,4)--(11,8);
+\draw[dashed] (11,5)--(8,4);
+\draw (8,4)--(2,2)--(0.5,4)--(0,8);
+\draw[dotted,thick] (5,0)--(5,9.5);
+\draw[dotted,thick] (4,0)--(4,9.5);
+\node[scale=0.8, below left] at (0,0) { $0$ };
+\node[scale=0.8, below] at (5,0) { $d$ };
+\node[scale=0.8, below] at (4,0) { $d{-}1$ };
+\node[scale=0.9, below right] at (2.5,1.2) { $\psi$ };
+\node[scale=0.9, right] at (9.8,7.3) { $\varphi$ };
+\node[scale=0.9, right] at (10.5,4.4) { $\Delta$ };
+\end{tikzpicture}
+\hfill \null
 
-\begin{align*}
-A_0 &= \textrm{trunc}(P,d), \\
-B_{0} &= P \div A_{0}, \\
-V_{0} &= 1.
-\end{align*}
+\vspace{-6mm}
 
-\begin{align*}
-A_{i+1} &= A_i + (V_i P \% A_i), \\
-B_{i+1} &= P \div A_{i+1}, \\
-V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \% A_{i+1}.
-\end{align*}
-$A_i$ convergences to $A_\infty$ such that $\NP (A_\infty)= \NP (A_0)$ and $A_\infty$ divides $P.$
+\caption{Euclidean division of Newton functions}
+\label{fig:NewtonEuclide}
+\end{figure}
+
+Figure~\ref{fig:NewtonEuclide} illustrates the definition: if $\varphi$ 
+and $\psi$ are the functions represented on the diagram, the epigraph of 
+$\varphi \nfmod \psi$ is the blue area whereas that of $\varphi \nfdiv \psi$ 
+is the green area translated by $(-d,0)$.
+It is an easy exercise (left to the reader) to design algorithms for
+computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$ for a cost
+which is at most $n + O(1)$.
+
+\begin{theo}
+\label{theo:EDivisionNP}
+Let $A, B \in K[X]$ with $B \neq 0$. Let $d$ be the degree of $B$.
+Let $\mu = \NF(B)(d) - \NF(B)(d{-}1)$ be the last slope of $B$ and 
+$y = \Delta(x)$ be the
+equation of the highest line of slope $\mu$ lying below $\NP(A)$.
+Then:
+\begin{align}
+\NF(A \mod B) & \geq \NF(B) \nfmod \NF(A) \label{eq:AmodB} \\
+\text{and} \hspace{4mm}
+\NF(A \div B) & \geq \NF(B) \nfdiv \NF(A) \label{eq:AdivB}
+\end{align}
 \end{theo}
@@
 \begin{rem}
-Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in Theorem \ref{theo:slope-factor} do not induce any loss in generality.
+Keeping the notations of the Theorem and decomposing $A$ as a sum 
+$A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$) consists of
+monomials of $A$ of degree $< d$ (resp of degree $\geq d$), we have:
+$$A \div B = A_{\geq d} \div B 
+\quad \text{and} \quad
+A \mod B = A_{<d} + (A_{\geq d} \mod B).$$
+Using the above relations and applying Theorem~\ref{theo:EDivisionNP}
+with $A_{\geq d}$ and $B$ (instead of $A$ and $B$ respectively) may lead
+to a better lower bound on the Newton function of $A \div B$ and $A \mod
+B$.
 \end{rem}
 
+\begin{proof}[of Theorem~\ref{theo:EDivisionNP}]
+Let us first prove Eq.~\eqref{eq:AmodB}. 
+Replacing $B$ by $c^{-1} B$ where $c$ denotes the leading coefficient
+of $B$, we may assume that $B$ is monic. Using linearity, we may further
+assume that $A$ is a monomial. Set $R_n = X^n \mod B$. The relation
+we have to prove is:
+$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d).$$
+When $n < d$, we have $R_n = 0$ and the assertion of the Theorem holds. 
+For larger $n$, we proceed by induction. We have the relation
+$R_{n+1} = X R_n - c_n B$
+where $c_n$ is the coefficient in $X^{d-1}$ of $R_n$. Thanks to the
+induction hypothesis, we have:
+\begin{align*}
+\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \mu(n{-}d) \\
+& = - \mu(n{+}1{-}d)
+\end{align*}
+since $\mu = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
+$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n+1-d)$ for all $x$. On the 
+other hand, for all $x$, we have:
+$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \mu$$
+from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n+1-d)$. As a
+consequence
+$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n+1-d)$
+and the induction follows.
+Eq.~\eqref{eq:AdivB} is now derived from:
+$$\NF(A \div B) \nftimes \NF(B) \geq \NF(A) \nfplus \NF(A \mod B)$$
+using the estimation on $\NF(A\mod B)$ we have just proved
+(see Figure~\ref{fig:NewtonEuclide}).
+\end{proof}
+
+\subsection{Tracking precision}
+
+\medskip
+
+\noindent
+{\bf Jagged precision.}
+\todo{Write something... Explain in particular that the fast methods
+(based on fast multipliction) may lead to important looses of precision.}
+
+\medskip
+
+\noindent
+{\bf Newton precision.}
+We consider here the case where the precision is given by Newton
+functions. Concretely, we pick $A, B \in K[X]$ two polynomials which 
+are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
+$$A = A_0 + O(\varphi_A)
+\quad \text{and} \quad
+B = B_0 + O(\varphi_B)$$
+Here $A_0$ and $B_0$ are some approximations of $A$ and $B$ respectively 
+and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
+$\deg A$ and $\deg B$ respectively.
+We are interesting in determining the precision on $A \mod B$ and $A
+\div B$. The following proposition gives a theoretical answer under
+mild assumptions.
+
+\begin{prop}
+\label{prop:NewtonprecEuclide}
+We keep the above notations and assume that the Newton precisions
+$O(\varphi_A)$ and $O(\varphi_B)$ on $A$ and $B$ respectively are 
+both nondegenerate (\emph{cf} Definition~\ref{def:nondeg}). Then,
+setting:
+$$\varphi = \varphi_A \nfplus 
+\big[ \varphi_B \nftimes \big(\NF(A) \nfdiv \NF(B)\big) \big]$$
+the polynomials $A \div B$ and $A \mod B$ are known at precision 
+$O(\varphi \nfdiv \NF(B))$ and $O(\varphi \nfmod \NF(B))$ respectively.
+\end{prop}
+
+\begin{proof}
+Let $\delta A$ (resp. $\delta B$) be a polynomial whose Newton 
+function is not less than $\varphi_A$ (resp. $\varphi_B$) and define
+$\delta Q$ and $\delta R$ by:
+\begin{align*}
+Q_0 + \delta Q & = (A_0 + \delta A) \div (B_0 + \delta B) \\
+R_0 + \delta R & = (A_0 + \delta A) \mod (B_0 + \delta B)
+\end{align*}
+where $Q_0 = A_0 \div B_0$ and $R_0 = A_0 \mod B_0$.
+We have to show that $\NF(\delta Q) \geq \varphi \nfdiv \NF(B)$ and
+$\NF(\delta R) \geq \varphi \nfmod \NF(B)$.
+Set $\delta X = \delta A - Q_0 \delta B$. 
+Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_0) \geq
+\NF(A) \div \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
+On the other hand, an easy computation yields
+$$\delta X = (B + \delta B) \cdot \delta Q + \delta R$$
+so that $\delta Q = \delta X \div (B + \delta B)$ and 
+$\delta R = \delta X \mod (B + \delta B)$. Using again 
+Theorem~\ref{theo:EDivisionNP}, we get the desired result.
+\end{proof}
+
+\todo{Explain how one can use this proposition: we split the computation
+into two parts (approximation and precision)...\\
+Don't forget to give the complexity.}
+
+\medskip
+
+\noindent
+{\bf Lattice precision.}
+We now move to lattice precision. We pick $A$ and $B$ two polynomials 
+of respective degree $n$ and $d$ and assume that they are known at 
+precision $O(H_A)$ and $O(H_B)$ respectively:
+$$A = A_0 + O(H_A)
+\quad \text{and} \quad
+B = B_0 + O(H_B).$$
+where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
+According to the results of \cite{padicprec}, in order to determine the 
+precision on $A \div B$ and $A \mod B$, we need to compute the 
+differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
+\mapsto X \mod Y$ at the point $(A_0, B_0)$. Writing $Q_0 = A_0 \div
+B_0$ and $R_0 = A_0 \mod B_0$, this can be done by
+expanding the relation:
+$$A_0 + dA = (B_0 + dB) (Q_0 + dQ) + (R_0 + dR)$$
+and neglecting the terms of order $\geq 2$. We get this way 
+$dA = B_0 dQ + Q_0 dB + dR$
+meaning that $dQ$ and $dR$ appears respectively as the quotient and
+the remainder of the Euclidean division of $dX = dA - B_0 dQ$ by $B_0$.
+
+Once this has been done, the strategy is quite similar to that explained 
+for Newton precision: \todo{summerize this in one sentence.}
+
+\subsection{An example: modular multiplication}
+
+For this example, we work over $W = \Z_2$ and fix a monic polynomial $M 
+\in \Z[X]$ (known exactly) of degree $5$. Our aim is to compare the 
+numerical stability of the multiplication in the quotient $\Z_2[X]/M(X)$ 
+depending on the precision model we are using. In order to do so, we 
+pick $n$ random polynomials $P_1, \ldots, P_n$ in $\Z_2[X]/M(X)$ 
+(according to the Haar measure) whose coefficients are all known at 
+precision $O(2^N)$ for some large integer $N$. We then compute the 
+product of the $P_i$'s using the following quite naive algorithm.
+
+\noindent\hrulefill
+
+%\noindent {\bf Algorithm 1:} {\tt example\_product}
+%
+%\smallskip
+
+\noindent 1.\ {\bf set} $P = 1$
+
+\noindent 2.\ {\bf for} $i=1,\dots,n$ {\bf do} {\bf compute} $P = (P 
+\cdot P_i) \mod M$
+
+\noindent 3.\ {\bf return} $P$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
+
+The table below reports the average gain of \emph{absolute} 
+precision $G$ which is observed while executing the algorithm above 
+for various modulus and $n$. The average is taken on a sample of $1000$ 
+random inputs. We recall that $G$ is defined as follows:
+
+\noindent $\bullet$
+in the case of jagged and Newton precision, the precision on the output 
+may be written into the form $\sum_{i=1}^4 O(2^{N_i}) \: X^i$ and 
+$G = \sum_{i=1}^4 (N_i - N)$;
+
+\noindent $\bullet$
+in the case of lattice precision, the precision on the output is a 
+lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
+$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
+$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
+index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
+generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
+(The term $G_d$ corresponds to diffused digits according to 
+\cite[Definition 2.3]{preclinalg}.)
+
+\medskip
+
+{\small%
+\noindent \hfill%
+\renewcommand{\arraystretch}{1.2}%
+\begin{tabular}{|c|c|c|c|c|}%
+\hline
+\multirow{3}{*}{Modulus $M$} & 
+\multirow{3}{*}{\hspace{2mm}$n$\hspace{2mm}} & 
+\multicolumn{3}{c|}{Gain of precision} \\
+\cline{3-5}
+& & \multirow{2}{*}{Jagged} & \multirow{2}{*}{Newton}
+      & \raisebox{-0.5mm}{Lattice} \\
+& & & & \raisebox{0.5mm}{\scriptsize (not dif.\tinyplus dif.)} \\
+\hline
+\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + X^2 + 1$ \\ {\scriptsize (Irred. mod $2$)} \end{tabular}}
+& $10$ & $\phantom{00}0.2$ & $\phantom{00}0.2$ & $\phantom{00}0.2\tinyplus\phantom{00}0.0$ \\
+& $50$ & $\phantom{00}4.2$ & $\phantom{00}4.2$ & $\phantom{00}4.2\tinyplus\phantom{00}0.0$ \\
+& $100$ & $\phantom{0}11.2$ & $\phantom{0}11.2$ & $\phantom{0}11.2\tinyplus\phantom{00}0.0$ \\
+\hline
+\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + 1$ \\ {\scriptsize (Sep. mod $2$)} \end{tabular}}
+& $10$ & $\phantom{00}0.4$ & $\phantom{00}0.4$ & $\phantom{00}0.9\tinyplus\phantom{00}6.0$ \\
+& $50$ & $\phantom{00}5.6$ & $\phantom{00}5.6$ & $\phantom{0}11.1\tinyplus\phantom{0}42.0$ \\
+& $100$ & $\phantom{0}13.6$ & $\phantom{0}13.6$ & $\phantom{0}27.0\tinyplus\phantom{0}87.0$ \\
+\hline
+\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + 2$ \\ {\scriptsize (Eisenstein)} \end{tabular}}
+& $10$ & $\phantom{00}6.2$ & $\phantom{00}6.2$ & $\phantom{00}6.2\tinyplus\phantom{00}0.0$ \\
+& $50$ & $\phantom{0}44.0$ & $\phantom{0}44.0$ & $\phantom{0}44.0\tinyplus\phantom{00}0.0$ \\
+& $100$ & $\phantom{0}92.5$ & $\phantom{0}92.5$ & $\phantom{0}92.5\tinyplus\phantom{00}0.0$ \\
+\hline
+\multirow{3}{*}{\begin{tabular}{@{}c@{}} $(X+1)^5 + 2$ \\ {\scriptsize (Shift Eisenstein)} \end{tabular}}
+& $10$ & $\phantom{00}0.6$ & $\phantom{00}0.6$ & $\phantom{00}4.7\tinyplus\phantom{00}1.4$ \\
+& $50$ & $\phantom{00}7.1$ & $\phantom{00}7.1$ & $\phantom{0}42.6\tinyplus\phantom{00}1.4$ \\
+& $100$ & $\phantom{0}15.1$ & $\phantom{0}15.1$ & $\phantom{0}91.8\tinyplus\phantom{00}1.4$ \\
+\hline
+\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + X + 2$ \\ {\scriptsize (Two slopes)} \end{tabular}}
+& $10$ & $\phantom{00}1.7$ & $\phantom{00}1.7$ & $\phantom{00}7.9\tinyplus\phantom{00}9.8$ \\
+& $50$ & $\phantom{00}8.1$ & $\phantom{00}8.1$ & $\phantom{0}70.7\tinyplus\phantom{0}59.8$ \\
+& $100$ & $\phantom{0}16.1$ & $\phantom{0}16.1$ & $152.6\tinyplus125.9$ \\
+\hline
+\end{tabular}
+\hfill \null}
+
+\medskip
+
+We observe several interesting properties. First of all, the gains for 
+Newton precision and jagged precision always agree though one may has 
+thought at first that Newton precision is weaker. Since performing
+precision computations in the Newton framework is cheaper, it seems
+(at least on this example) that using the jagged precision model is not
+relevant.
+
+On the other hand, the lattice precision may end up with better results. 
+Nevertheless this strongly depends on the modulus $M$. For instance, 
+when $M$ is irreducible modulo $p=2$ or Eiseistein, there is apparently 
+no benefit of using the lattice precision model. We emphasize that 
+these two particular cases correspond to modulus that are usually used 
+to define (unramified and totally ramified respectively) extensions of 
+$\Q_2$.
+
+For other modulus, the situation is quite different and the benefit of 
+using the lattice precision model becomes more apparent. 
+The comparison between the gain of precision in the jagged model and 
+the number of not diffused digits in the lattice model makes sense:
+indeed the latter appears as a theoretical upper bound of the former
+and the difference between them quantifies the quality of the way we
+track precision in the jagged (or the Newton) precision model. We 
+observe that this difference is usually not neglictible (\emph{cf} 
+notably the case of $M(X) = (X+1)^5 + 2$) meaning that this quality 
+is not very good in general.
+As for diffused digits, they correspond to digits that cannot be 
+``seen'' in the jagged precision model. Their number then measures the 
+intrinsic limitations of this model. We observe that it can be very 
+important as well in several cases.
+
+\todo{Add some words about the comparison between $X^5 + 2$ and
+$(X+1)^5 + 2$?}
+
+\section{Slope factorization}
+
+A well-known theorem \todo{(Give a reference?)} asserts that each 
+extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
+K[X]$ corresponds to a factorization $P = AB$ where the Newton polygon 
+of $A$ (resp. $B$) is given by the part of $\NP(P)$ located at the left 
+(resp. the right) of $M$. Such a factorization is often called a 
+\emph{slope factorization}.
+
+The aim of this section is to design efficient and stable algorithms 
+for computing these factorizations. \todo{Compare with previous
+results...}
+
+\subsection{A Newton iteration}
+
+In the previous references, the factor $A$ defined above is always 
+obtained \emph{via} a Newton iteration which takes place over a finite 
+extension of $K$. We introduce here a variant of this Newton iteration 
+whose main advantage to be defined over $K$ itself.
+
+\begin{theo} \label{theo:slope-factor}
+Let $P(X) = \sum_{i=0}^n a_i X^i$ be a polynomial of degree $n$ with
+coefficients in $K$. We assume that $\NP(P)$ has an extremal point 
+whose abscissa is $d$.
+We define the sequences $(A_i)_{i \geq 0}$ and $(V_i)_{i \geq 0}$
+recursively by:
+\begin{align*}
+A_0 & = \sum_{i=0}^d a_i X^i, \quad V_{0} = 1 \medskip \\
+A_{i+1} &= A_i + (V_i P \:\mod\: A_i), \smallskip \\
+V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\
+& \hspace{2cm}\text{where } B_{i+1} = P \div A_{i+1}.
+\end{align*}
+Then the sequence $(A_i)$ convergences to a divisor $A$ of $P$ 
+of degree $d$ whose leading coefficient is $a_d$ and whose Newton 
+function agrees with $\NF(P)$ on $[0,d]$.
+\end{theo}
+
+%\begin{rem}
+%Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in Theorem \ref{theo:slope-factor} do not induce any loss in generality.
+%\end{rem}
+
+%\begin{rem}
+%Non necessity to have $A_0$ as the first slope. It is possible to use minorations.
+%
+%\begin{tikzpicture}
+%\draw[->] (0,-1) -- (0,2);
+%\draw[->] (-0.05,0) -- (4,0);
+%\node[left] at (0,0) {$0$};
+%\draw (1.5,-.1) -- (1.5,.1);
+%\node[above] at (1.5,0) {$d$};
+%\draw (0,1)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
+%\draw[green] (0,1.5)-- (1,0.33)-- (1.5,0) --(2,-.125)-- (3,0)-- (3.5,1);
+%\node[below] at (.75,-0.2) {$\lambda_0$};
+%\end{tikzpicture}
+%\end{rem}
+
 \begin{rem}
-Non necessity to have $A_0$ as the first slope. It is possible to use minorations.
-
-\begin{tikzpicture}
-\draw[->] (0,-1) -- (0,2);
-
-\draw[->] (-0.05,0) -- (4,0);
-
-\node[left] at (0,0) {$0$};
-
-\draw (1.5,-.1) -- (1.5,.1);
-\node[above] at (1.5,0) {$d$};
-
-\draw (0,1)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
-\draw[green] (0,1.5)-- (1,0.33)-- (1.5,0) --(2,-.125)-- (3,0)-- (3.5,1);
-
-\node[below] at (.75,-0.2) {$\lambda_0$};
-\end{tikzpicture}
+The divisor $A$ is uniquely determined by the conditions of 
+Theorem~\ref{theo:slope-factor}.
 \end{rem}
 
-\subsection{Useful formulae}
+The rest of this subsection is devoted to the proof of the theorem.
+If $d = n$ (resp. $d = 1$), the sequence $A_i$ is constant equal to
+$P$ (resp. to the constant coefficient of $P$) and theorem is clear.
+We then assume $1 < d < n$. Without loss of generality, we may further 
+assume that the coefficient $a_d$ is equal to $1$. We set:
+\begin{align*}
+\lambda_0 & = \NF(P)(d) - \NF(P)(d{-}1) \\
+\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d).
+\end{align*}
+The existence of an extremal point of $\NP(P)$ located at abscissa
+$d$ ensures that $\lambda_1 > \lambda_0$.
+For all indices $i$, we define:
+$$\begin{array}{r@{\,\,}lr@{\,\,}l}
+Q_i &= V_i P \div A_i, &
+R_i & = V_i P \mod A_i = A_{i+1} - A_i, 
+\smallskip \\
+S_i &= P \mod A_i, &
+T_i &= 1 - V_i B_i. 
+\end{array}$$
+and when $\square$ is some letter, we put $\Delta \square_i = \square_{i+1} 
+- \square_i$.
+%We use the next Lemma to record some useful formulae.
 
 \begin{lem} \label{lem:formulae}
-Some definitions :
+The following relations hold:
 \begin{align}
-R_i &= A_{i+1} -A_i, \label{eqdef:Ri}\\
-S_i &= P \% A_i, \label{eqdef:Si} \\
-Q_i &= V_i P \div A_i. \label{eqdef:Qi}. 
-\end{align}
-Some formulae :
-\begin{align}
-B_i-B_{i+1} &= R_i B_{i+1} \div A_i, \label{eqdef:Biminus} \\
-S_i-S_{i+1} &= R_i B_{i+1} \div A_i,  \label{eqdef:Siminus} \\
-S_{i} &= (B_i R_i + (1-V_i B_i) S_{i-1}+(1-V_i B_i)(S_i-S_{i-1})) \% A_{i}, \label{eqdef:Si2}  \\
-V_{i+1}-V_i &= (V_i (1-V_i B_i)+V_i^2(B_i-B_{i+1})) \% A_{i}, \label{eqdef:Viminus}\\
-1-Q_i &= (1-V_i B_i)+((R_i-V_i S_i) \div A_i), \label{eqdef:Qiminus}\\
-R_{i+1} &= ((V_{i+1}-V_i)S_{i+1}+(1-Q_i)R_i) \% A_{i+1}, \label{eq:Riplus}\\
-1-V_{i+1}B_{i+1} &= ((1-V_i B_i)+V_i (B_i-B_{i+1}))^2. \label{eq:ViBiplus}
+\Delta B_i &= -(R_i B_{i+1}) \div A_i, \label{eqdef:Biminus} \\
+\Delta S_i &= -(R_i B_{i+1}) \mod A_i,  \label{eqdef:Siminus} \\
+S_{i} &= (B_i R_i + T_i S_{i-1} - T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
+\Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
+1-Q_i &= T_i + (R_i-V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
+R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
+T_{i+1} &\equiv T_i + V_i \: (\Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
 \end{align}
 \end{lem}
+
 \begin{proof}
-We first remark that by definition,
+From $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1}$, we get
+$- R_i B_{i+1}= \Delta B_i \cdot A_i + \Delta S_i$.
+Hence, by consideration of degree, we obtain \eqref{eqdef:Biminus} 
+and \eqref{eqdef:Siminus}.
+On the other hand, from $V_i P = A_i Q_i + R_i=V_i (A_i B_i + S_i)$, we 
+derive
+\begin{equation}
+\label{eq:ViBiQi} 
+(V_i B_i -Q_i) \cdot A_i=R_i-V_i S_i. 
+\end{equation} 
+Thus $R_i=V_i S_i \mod A_i$.
+Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
+\mod A_i$, from what \eqref{eqdef:Si2} follows directly.
+By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
+We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi}, we 
+get \eqref{eqdef:Qiminus}.
+
+We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i$ and $V_{i+1} P=A_{i+1} 
+Q_{i+1}+R_{i+1}$. Thus:
 \begin{align*}
-P &= A_i B_i + S_i, \\
-P V_i&= A_i Q_i + R_i.
+R_{i+1} &= \Delta V_i \: P + (1-Q_i)R_i \\
+ &= (\Delta V_i \: S_{i+1} + (1-Q_i)R_i) \mod A_{i+1},
 \end{align*} 
-Since $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1},$ we have \[R_i B_{i+1}= (B_i -B_{i+1})A_i + (S_i+S_{i+1}).\]
-Hence, by consideration of degree, what we have written is the Euclidean division of $R_i B_{i+1}$ by $A_i$ and we obtain \eqref{eqdef:Biminus} and \eqref{eqdef:Siminus}.
-
-From $V_i P =  A_i Q_i + R_i=V_i (A_i B_i + S_i),$ we get \begin{equation}(V_i B_i -Q_i) A_i=R_i-V_i S_i. \label{eq:ViBiQi} \end{equation} Thus, \[ R_i=V_i S_i \% A_i.\]
-Hence, $B_i R_i =S_i (V_iB_i-1)+S_i \% A_i,$ and $S_i=B_i R_i-S_i (V_iB_i-1) \% A_i.$ \eqref{eqdef:Si2} follows directly.
-
-By definition of $V_i,$ we get $V_{i+1}-V_i=V_i (1-V_i B_{i+1}) \% A_{i+1}.$ \eqref{eqdef:Viminus} follows immediately.
-
-We can write $1-Q_i=(1-V_i B_i)+(V_i B_i - Q_i).$ Using \eqref{eq:ViBiQi}, we get \eqref{eqdef:Qiminus}.
-
-We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i,$ and $V_{i+1} P=A_{i+1} Q_{i+1}+R_{i+1}.$ As a consequence, 
+and \eqref{eq:Riplus} is proved. Finally
 \begin{align*}
-R_{i+1} &= (V_{i+1}-V_i)P+(1-Q_i)R_i \\
- &= (V_{i+1}-V_i)S_{i+1}+(1-Q_i)R_i) \% A_{i+1},
-\end{align*} and \eqref{eq:Riplus} is proved.
-
-Finally, \begin{align*}
-1-V_{i+1}B_{i+1} &= 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \% A_{i+1}, \\
-&= (1-V_i B_{i+1})^2 \% A_{i+1}, \\
-&= ((1-V_i B_i)+V_i (B_i-B_{i+1}))^2 \% A_{i+1},
-\end{align*} which concludes the proof.
+T_{i+1} &\equiv 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \pmod {A_{i+1}} \\
+&\equiv (1-V_i B_{i+1})^2 \pmod {A_{i+1}} \\
+&\equiv T_i + V_i (\Delta B_i)^2 \pmod {A_{i+1}}
+\end{align*}
+which concludes the proof.
 \end{proof}
@@
 
-\subsection{Proof of the Main Theorem}
+\subsubsection{Proof of the Main Theorem}
 
-\subsubsection{Initial Newton polygons}
+{\bf Initial Newton polygons}
 
@@
 
-\subsubsection{Unchanged Newton polygons}
+\medskip
+
+{\bf Unchanged Newton polygons}
 
@@
 
-\subsubsection{Newton polygons inside the loop}
+\medskip
+
+{\bf Newton polygons inside the loop}
 
@@
 
+\subsection{A slope factorization algorithm}
+
+Let $P \in K_n[X]$ and $d$ be the abscissa of an extremal point of 
+$\NP(P)$. Previously (\emph{cf} Theorem~\ref{theo:slope-factor}), we 
+have defined a sequence $(A_i, V_i)$ converging to $(A,V)$ where $A$ is 
+a factor of $P$ whose Newton function is $\NF(P)_{|[0,d]}$ and $V$ is the inverse of $B = P/A$ modulo $A$.
+We now assume that $P$ is known up to some finite precision: $P = P_0 + 
+O(\hspace{0.5mm}\cdots)$ where the object inside the $O$ depends on the 
+chosen precision model. We address the two following questions:
+(1)~what is the precision on the factor $A$, and
+(2)~how one can compute in practive $A$ at this precision?
+
+In the sequel, it will be convenient to use a different normalization on 
+$A$ and $B$: if $a_d$ is the coefficient of $P$ of degree $d$, we set
+$A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
+so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We shall also 
+always assume that $P$ is monic in the sense that its leading
+coefficient is \emph{exactly} $1$; the precision datum on $P$ then 
+concerns only the coefficients up to degree $n{-}1$. Similarly, noting
+that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
+precision datum up to degree $d{-}1$ and $n{-}d{-}1$ respectively.
+
+\medskip
+
+\noindent
+{\bf About the first question.}
+We first assume that we are working with lattice precision.
+The answer to the first question is then given by the theory of 
+\cite{padicprec} and is controlled by the differential of the
+application sending $P$ to $A^{(1)}$.
+
+\begin{prop}
+\label{prop:precA1}
+The application $P \mapsto A^{(1)}$ (where $A^{(1)}$ is defined as before) is 
+of class $C^1$ in a neighborhood of $P_0$ and its differential 
+at $P_0$ is the linear mapping:
+$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
+where $V^{(1)}$ is the inverse of $B^{(1)}$ modulo $A^{(1)}$.
+\end{prop}
+
+\begin{proof}
+There certain exists a neighborhood $\mathcal U$ of $P_0$ on which the 
+Newton polygon does not vary. On $\mathcal U$, the function $f : P 
+\mapsto (A^{(1)},B^{(1)})$ is well defined. It is furthermore bijective 
+and its inverse $g$ is given by $(A^{(1)},B^{(1)}) \mapsto A^{(1)}B^{(1)}$.
+Clearly $g$ is of class $C^1$ and its differential is given by
+\begin{equation}
+\label{eq:dPdAdB}
+(dA^{(1)}, dB^{(1)}) \mapsto dP = A^{(1)} \: dB^{(1)} + B^{(1)} \: dA^{(1)}.
+\end{equation}
+Thanks to Bézout Theorem, it is invertible as soon as $A^{(1)}$ and $B^{(1)}$ are
+coprime. This happens on $f(\mathcal U) = g^{-1}(\mathcal U)$ because 
+$\NP(A^{(1)})$ and $\NP(B^{(1)})$ do not shape a common slope. As a consequence
+$f$ is of class $C^1$ on $\mathcal U$. Moreover, its differential is
+obtained by inverting Eq.~\eqref{eq:dPdAdB}. Reducing modulo $A^{(1)}$, we
+get $dP \equiv B^{(1)} \: dA^{(1)} \pmod {A^{(1)}}$.
+The claimed result follows after having noticed that $dA^{(1)}$
+has degree at most $d{-}1$.
+\end{proof}
+
+\todo{Derive from Proposition~\ref{prop:precA1} a theoretical
+result on Newton precision.}
+
+\medskip
+
+\noindent
+{\bf About the second question.}
+We now study the problem of the effective computation of $A^{(1)}$.
+
 \end{document}

### 2016-01-05 polynomials/muldivfact.tex
Draft for the proof of the slope factorization theorem

@@
 Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
-$d$ respectively. Letting $\mu = \psi(d) - \psi(d-1)$ and $\Delta$
+$d$ respectively. Set $\mu = \psi(d) - \psi(d-1)$. Letting $\Delta$
 be the greatest affine function of slope $\mu$ with $\Delta \leq
-\varphi$, we define:
-$$\begin{array}{rcl}
-\varphi \nfmod \psi \, : \, [0, d-1] & \to & \R \cup \{+\infty\} \smallskip \\
-x & \mapsto & \psi(x) + \Delta(d) - \psi(d) \medskip \\
-\varphi \nfdiv \psi \, : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
-x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
+\varphi_{|[d,n]}$ and $\delta = \Delta(d) - \psi(d)$, we define:
+$$\begin{array}{r@{\,\,}rcl}
+\varphi \nfmod \psi & 
+\multicolumn{3}{@{}l}{= \varphi_{|[0,d{-}1]} \nfplus \big( 
+\psi_{|[0,d{-}1]} + \delta\big)} \medskip \\
+\varphi \nfdiv \psi & : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
+& x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
 \end{array}$$
@@
 is the green area translated by $(-d,0)$.
-It is an easy exercise (left to the reader) to design algorithms for
-computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$ for a cost
-which is at most $n + O(1)$.
+It is an easy exercise (left to the reader) to design quasi-linear
+algorithms for computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$.
 
@@
 
-\begin{rem}
-Keeping the notations of the Theorem and decomposing $A$ as a sum 
-$A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$) consists of
-monomials of $A$ of degree $< d$ (resp of degree $\geq d$), we have:
-$$A \div B = A_{\geq d} \div B 
+\begin{proof}
+Write $A = A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$)
+consists of monomials of $A$ of degree at least $d$ (resp. less than
+$d$). Noting that:
+$$A \mod B = A_{<d} + (A_{\geq d} \mod B)
 \quad \text{and} \quad
-A \mod B = A_{<d} + (A_{\geq d} \mod B).$$
-Using the above relations and applying Theorem~\ref{theo:EDivisionNP}
-with $A_{\geq d}$ and $B$ (instead of $A$ and $B$ respectively) may lead
-to a better lower bound on the Newton function of $A \div B$ and $A \mod
-B$.
-\end{rem}
-
-\begin{proof}[of Theorem~\ref{theo:EDivisionNP}]
-Let us first prove Eq.~\eqref{eq:AmodB}. 
+A \div B = A_{\geq d} \div B$$
+we may assume that $A = A_{\geq d}$.
+
+Let us now prove Eq.~\eqref{eq:AmodB}. 
 Replacing $B$ by $c^{-1} B$ where $c$ denotes the leading coefficient
@@
 we have to prove is:
-$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d).$$
-When $n < d$, we have $R_n = 0$ and the assertion of the Theorem holds. 
-For larger $n$, we proceed by induction. We have the relation
+$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d)
+\quad \text{for } n\geq d.$$
+We proceed by induction. The initialisation is clear because $R_d$
+agrees with $(-B)$ up to degree $d{-}1$. We have the relation
 $R_{n+1} = X R_n - c_n B$
@@
 since $\mu = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
-$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n+1-d)$ for all $x$. On the 
+$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$ for all $x$. On the 
 other hand, for all $x$, we have:
 $$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \mu$$
-from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n+1-d)$. As a
+from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$. As a
 consequence
-$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n+1-d)$
+$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$
 and the induction follows.
@@
 Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_0) \geq
-\NF(A) \div \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
+\NF(A) \nfdiv \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
 On the other hand, an easy computation yields
@@
 \end{align*}
-Then the sequence $(A_i)$ convergences to a divisor $A$ of $P$ 
+Then the sequence $(A_i)$ converges to a divisor $A_\infty$ of $P$ 
 of degree $d$ whose leading coefficient is $a_d$ and whose Newton 
 function agrees with $\NF(P)$ on $[0,d]$.
+Moreover, setting:
+$$\kappa = \frac{\NF(P)(d{+}1) + \NF(P)(d{-}1)} 2 - \NF(P)(d)$$
+(with $\NF(P)(-1) = \NF(P)(n{+}1) = +\infty$ if necessary),
+we have $\kappa > 0$ and the following rate of convergence:
+$$\forall i \geq 1, \quad
+\NF(A_\infty - A_i) \geq \NF(P)_{|[0,d]} \,+\, 2^i \kappa.$$
 \end{theo}
@@
 R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
-T_{i+1} &\equiv T_i + V_i \: (\Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
+T_{i+1} &\equiv (T_i + V_i \: \Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
 \end{align}
@@
 &\equiv (1-V_i B_{i+1})^2 \pmod {A_{i+1}} \\
-&\equiv T_i + V_i (\Delta B_i)^2 \pmod {A_{i+1}}
+&\equiv (T_i + V_i \: \Delta B_i)^2 \pmod {A_{i+1}}
 \end{align*}

### 2016-01-05 polynomials/muldivfact.tex
Draft corrected and included in the paper

@@
 Moreover, setting:
-$$\kappa = \frac{\NF(P)(d{+}1) + \NF(P)(d{-}1)} 2 - \NF(P)(d)$$
+$$\kappa = \NF(P)(d{+}1) + \NF(P)(d{-}1) - 2 \cdot \NF(P)(d)$$
 (with $\NF(P)(-1) = \NF(P)(n{+}1) = +\infty$ if necessary),
 we have $\kappa > 0$ and the following rate of convergence:
-$$\forall i \geq 1, \quad
+$$\forall i \geq 0, \quad
 \NF(A_\infty - A_i) \geq \NF(P)_{|[0,d]} \,+\, 2^i \kappa.$$
@@
 \lambda_0 & = \NF(P)(d) - \NF(P)(d{-}1) \\
-\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d).
+\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d),
 \end{align*}
+so that $\kappa = \lambda_1 - \lambda_0$.
 The existence of an extremal point of $\NP(P)$ located at abscissa
-$d$ ensures that $\lambda_1 > \lambda_0$.
+$d$ ensures that $\lambda_1 > \lambda_0$, \emph{i.e.} $\kappa > 0$.
 For all indices $i$, we define:
@@
 S_i &= P \mod A_i, &
-T_i &= 1 - V_i B_i. 
+T_i &= (1 - V_i B_i) \mod A_i
 \end{array}$$
@@
 \Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
-1-Q_i &= T_i + (R_i-V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
+1-Q_i &= T_i - (V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
 R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
-T_{i+1} &\equiv (T_i + V_i \: \Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
+T_{i+1} &= (T_i + V_i \: \Delta B_i)^2 \mod{A_{i+1}}. \label{eq:ViBiplus}
 \end{align}
@@
 By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
-We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi}, we 
+We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi} and
+noting that $\deg R_i < \deg A_i = d$, we 
 get \eqref{eqdef:Qiminus}.
@@
 
-\subsubsection{Proof of the Main Theorem}
+\subsubsection{Tristan's proof}
 
@@
 Finally, we can apply \eqref{eqdef:Si2} for $i+1,$
-\[S_{i+1} = (B_{i+1} R_{i+1} + (1-V_{i+1} B_{i+1}) S_{i}+(1-V_{i+1} B_{i+1})(S_{i+1}-S_{i})) \% A_{i+1},\] to obtain the desired minoration on $\NP (S_{i+1}).$ This concludes the proof.
+\begin{align*}
+S_{i+1} & = (B_{i+1} R_{i+1} + (1-V_{i+1} B_{i+1}) S_{i} \\
+& + (1-V_{i+1} B_{i+1})(S_{i+1}-S_{i})) \% A_{i+1},
+\end{align*}
+to obtain the desired minoration on $\NP (S_{i+1}).$ This concludes the proof.
 \end{proof}
+
+\subsubsection{Xavier's proof}
+
+If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
+is constant and the theorem is obvious. We then assume that $\lambda_0$
+and $\lambda_1$ are both finite.
+We notice that, when the polynomial is changed into $cP$ where $c$ is a 
+nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
+multiplied by $c$ as well whereas the $B_i$'s and the $V_i$'s remained 
+unchanged. Therefore, the theorem holds for $P$ if and only if it holds 
+for $cP$. As a consequence we may assume that $P$ is normalized so that 
+$\NF(P)(d) = d \lambda_0$. We define the function $\varphi : \R^+ \to \R 
+\cup \{+\infty\}$ by:
+$$\begin{array}{r@{\,\,}ll}
+\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
+& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
+\end{array}$$
+For a polynomial $Q \in K[X]$ of degree $n$, we further define:
+\begin{align*}
+b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) \\
+\text{and} \quad
+b_0(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_0 x.
+\end{align*}
+Set also $b_\varphi(0) = b_0(0) = +\infty$ by convention.
+With the normalization of $P$ we have chosen above, we have $b_0(P)
+= b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
+$Q$.
+
+\begin{lem}
+\label{lem:bphib0}
+Let $b \in \{b_\varphi, b_0\}$. For $Q_1, Q_2 \in K[X]$:
+
+\smallskip
+
+a) $b(Q_1+Q_2) \geq \min \big(b(Q_1), b(Q_2)\big)$
+
+\smallskip
+
+b) $b(Q_1Q_2) \geq \min b(Q_1) + b(Q_2)$
+
+\smallskip
+
+c) $b_\varphi(Q_1 Q_2) \geq b_\varphi(Q_1) + b_0(Q_2) \geq
+b_\varphi(Q_1) + b_\varphi(Q_2)$
+
+\smallskip
+
+\noindent
+For $Q, A \in K[X]$ with $\deg A = d$ and $\NF(A) =
+\NF(P)_{|[0,d]}$:
+
+\smallskip
+
+d) $b(Q \mod A) \geq b(Q)$
+
+\smallskip
+
+e) $b_0(Q \div A) \geq b_\varphi(Q)$.
+\end{lem}
+
+\begin{proof}
+\emph{a)} is clear.
+
+We skip the proof of \emph{b)} which is similar to that of \emph{c)}.
+
+Let $t_1$ (resp. $t_2$) be the translation of vector $(0, b_\varphi
+(Q_1))$ (resp. $(0, b_0(Q_2))$. It follows from the definition of
+$b_\varphi$ that $\NP(Q_1)$ is a subset of $t_1(\Epi(\varphi))$ where
+$\Epi(\varphi)$ denotes the epigraph of $\varphi$. Similarly $\NP(Q_2)
+\subset t_2(C)$ where $C$ is the convex cone generated by the vectors
+$(0,1)$ and $(1, \lambda_0)$. Thus
+$$\NP(Q) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
+t_2 \circ t_1\big(\Epi(\varphi)\big)$$
+and \emph{c)} follows.
+
+Finally \emph{d)} and \emph{e)} follows from
+Theorem~\ref{theo:EDivisionNP}.
+\end{proof}
+
+We are now going to prove by induction on $i$ the conjonction of all
+equalities and inequalities below:
+\begin{equation}
+\label{eq:induction}
+\begin{array}{l}
+\NF(A_i) = \varphi_{|[0,d]}, \smallskip \\
+b_0(V_i) \geq 0, \quad
+b_\varphi(R_i) \geq 2^i \kappa \smallskip \\
+b_\varphi(S_i) \geq 0, \quad
+b_0(T_i) \geq 2^i \kappa.
+\end{array}
+\end{equation}
+
+Noting that $A_0$ and $P$ agree up to degree $d$ and that $\NP(P)$ has 
+an extremal point at abscissa $d$, we get $\NF(A_0) = \varphi_{|[0,d]}$.
+Clearly $b_0(V_0) \geq 0$ since $V_0 = 1$. It follows from the 
+definitions that $R_0 = S_0 = P \mod A_0 = (P{-}A_0) \mod A_0$. 
+Noting that $P{-}A_0 = \sum_{i=d+1}^n a_i X^i$, we obtain
+$b_\varphi(P{-}A_0) \geq \kappa$ and then $b_\varphi(R_0) = b_\varphi(S_0)
+\geq \kappa$. Finally observe that:
+$$T_0 = 1 - B_0 = (A_0 - P) \div A_0.$$
+Therefore $b_0(T_0) \geq \kappa$ results from $b_\varphi(A{-}P_0) \geq 
+\kappa$ thanks to Lemma~\ref{lem:bphib0}. 
+We have then established~\eqref{eq:induction} when $i = 0$.
+
+We now assume~\eqref{eq:induction} for the index $i$.
+From $A_{i+1} = A_i + R_i$ and the estimation $b_\varphi(R_i) \geq 2^i 
+\kappa > 0$, we derive $\NF(A_{i+1}) = \varphi_{|[0,d]}$. Therefore,
+Lemma~\ref{lem:bphib0} applies with $A = A_i$ and $A = A_{i+1}$.
+Now using the definition of $B_{i+1}$, we get $b_0(B_{i+1}) \geq 0$.
+As a consequence:
+$$b_\varphi(R_i B_{i+1}) \geq b_\varphi(R_i) + b_0(B_{i+1})
+\geq 2^i \kappa$$
+by Lemma~\ref{lem:bphib0} and the induction hypothesis.
+Using again Lemma~\ref{lem:bphib0}, we then derive from 
+\eqref{eqdef:Biminus} and \eqref{eqdef:Siminus}, $b_0(\Delta B_i) \geq 
+2^i \kappa$ and $b_\varphi(\Delta S_i) \geq 2^i \kappa$.
+Similarly, using \eqref{eqdef:Si2} and the estimations we already know,
+we obtain $b_\varphi(S_i) \geq 2^i \kappa$. Combining this with
+$b_\varphi(\Delta S_i) \geq 2^i \kappa$, we find
+$b_\varphi(S_{i+1}) \geq 2^i \kappa$ as well. Applying again and again
+the same strategy, we deduce successively
+$b_0(\Delta V_i) \geq 2^i \kappa$ using \eqref{eqdef:Viminus},
+$b_0(1{-}Q_i) \geq 2^i \kappa$ using \eqref{eqdef:Qiminus},
+$b_\varphi(R_{i+1}) \geq 2^{i+1} \kappa$ using \eqref{eq:Riplus},
+and finally $b_0(T_{i+1}) \geq 2^{i+1} \kappa$
+using \eqref{eq:ViBiplus}. The equalities and inequalities
+of Eq.~\eqref{eq:induction} have then all been established for
+the index $i+1$ and the induction goes.
+
+From the inequalities $b_\varphi(R_i) \geq 2^i \kappa$, we deduce that 
+the sequence $(A_i)$ is Cauchy and therefore converges. Its limit 
+$A_\infty$ certain satisfies $\NF(A_\infty) = \varphi_{|[0,d]}$ because 
+all the $A_i$'s do. Moreover we know that $b_\varphi(S_i) \geq 2^i 
+\kappa$ from what we derive that the sequence $(S_i)$ goes to $0$.
+Coming back to the definition of $S_i$, we find $P \mod A_\infty = 0$,
+\emph{i.e.} $A_\infty$ divides $P$.
+
+\begin{comment}
+
+Induction hypothesis:
+
+Assume lambda_i > 0 
+(we need a slightly different argument for i=0)
+
+* NF(A_(i+1)) = NF(A_i)
+* B_(i+1) = P // A_(i+1)
+     => b_mu(B_(i+1)) >= 0  (because b_phi(P) >= 0)
+* b_phi(R_i B_(i+1)) >= lambda_i
+     => b_phi(Delta S_i) >= lambda_i
+    and b_mu(Delta B_i) >= lambda_i
+
+* b_phi(B_i R_i) >= lambda_i
+  b_phi(S_i T_i) >= lambda_i
+     => b_phi(S_i) >= lambda_i
+     => b_phi(S_(i+1)) >= lambda_i
+
+* b_phi(V_i T_i) >= lambda_i
+  b_phi(V_i^2 Delta B_i) >= lambda_i
+     => b_phi(Delta V_i) >= lambda_i
+
+* b_phi(T_i) >= lambda_i
+  b_phi(V_i S_i) >= lambda_i
+     => b_mu(V_i S_i // A_i) >= lambda_i
+     => b_mu(1 - Q_i) >= lambda_i
+
+* b_phi(Delta V_i) >= lambda_i
+  b_phi(S_(i+1)) >= lambda_i
+  b_mu(1 - Q_i) >= lambda_i
+  b_phi(R_i) >= lambda_i
+     => b_phi(R_(i+1)) >= 2*lambda_i
+
+* b_phi(T_i) >= lambda_i
+  b_phi(V_i Delta B_i) => lambda_i
+     => b_phi(T_(i+1)) >= 2*lambda_i
+
+And induction goes.
+\end{comment}
 

### 2016-01-05 polynomials/muldivfact.tex
Newton precision and slope factorisation

@@
 \newcommand{\id}{\textrm{id}}
+\newcommand{\app}{\textrm{app}}
 \newcommand{\Epi}{\textrm{Epi}}
@@
 \subsection{A Newton iteration}
+\label{ssec:Newtoniter}
 
@@
 we have $\kappa > 0$ and the following rate of convergence:
-$$\forall i \geq 0, \quad
-\NF(A_\infty - A_i) \geq \NF(P)_{|[0,d]} \,+\, 2^i \kappa.$$
+\begin{equation}
+\label{eq:rateconv}
+\forall i \geq 0, \quad
+\NF(A_\infty - A_i) \geq \NF(P)_{|[0,d{-}1]} \,+\, 2^i \kappa.
+\end{equation}
 \end{theo}
@@
 
+\subsubsection{Tristan's proof}
+
 \begin{deftn}
@@
 \end{deftn}
-
-\subsubsection{Tristan's proof}
 
@@
 \cup \{+\infty\}$ by:
-$$\begin{array}{r@{\,\,}ll}
+\begin{equation}
+\label{eqdef:functionphi}
+\begin{array}{r@{\,\,}ll}
 \varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
 & = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
-\end{array}$$
+\end{array}
+\end{equation}
 For a polynomial $Q \in K[X]$ of degree $n$, we further define:
-\begin{align*}
-b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) \\
+\begin{align}
+b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) 
+\label{eqdef:bphi} \\
 \text{and} \quad
 b_0(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_0 x.
-\end{align*}
+\label{eqdef:b0}
+\end{align}
 Set also $b_\varphi(0) = b_0(0) = +\infty$ by convention.
@@
 \geq \kappa$. Finally observe that:
-$$T_0 = 1 - B_0 = (A_0 - P) \div A_0.$$
+$$T_0 = (1 - B_0) \mod A_0 = \big((A_0 - P) \div A_0\big) \mod A_0.$$
 Therefore $b_0(T_0) \geq \kappa$ results from $b_\varphi(A{-}P_0) \geq 
@@
 the sequence $(A_i)$ is Cauchy and therefore converges. Its limit 
-$A_\infty$ certain satisfies $\NF(A_\infty) = \varphi_{|[0,d]}$ because 
+$A_\infty$ certainly satisfies $\NF(A_\infty) = \varphi_{|[0,d]}$ because 
 all the $A_i$'s do. Moreover we know that $b_\varphi(S_i) \geq 2^i 
@@
 Coming back to the definition of $S_i$, we find $P \mod A_\infty = 0$,
-\emph{i.e.} $A_\infty$ divides $P$.
-
-\begin{comment}
-
-Induction hypothesis:
-
-Assume lambda_i > 0 
-(we need a slightly different argument for i=0)
-
-* NF(A_(i+1)) = NF(A_i)
-* B_(i+1) = P // A_(i+1)
-     => b_mu(B_(i+1)) >= 0  (because b_phi(P) >= 0)
-* b_phi(R_i B_(i+1)) >= lambda_i
-     => b_phi(Delta S_i) >= lambda_i
-    and b_mu(Delta B_i) >= lambda_i
-
-* b_phi(B_i R_i) >= lambda_i
-  b_phi(S_i T_i) >= lambda_i
-     => b_phi(S_i) >= lambda_i
-     => b_phi(S_(i+1)) >= lambda_i
-
-* b_phi(V_i T_i) >= lambda_i
-  b_phi(V_i^2 Delta B_i) >= lambda_i
-     => b_phi(Delta V_i) >= lambda_i
-
-* b_phi(T_i) >= lambda_i
-  b_phi(V_i S_i) >= lambda_i
-     => b_mu(V_i S_i // A_i) >= lambda_i
-     => b_mu(1 - Q_i) >= lambda_i
-
-* b_phi(Delta V_i) >= lambda_i
-  b_phi(S_(i+1)) >= lambda_i
-  b_mu(1 - Q_i) >= lambda_i
-  b_phi(R_i) >= lambda_i
-     => b_phi(R_(i+1)) >= 2*lambda_i
-
-* b_phi(T_i) >= lambda_i
-  b_phi(V_i Delta B_i) => lambda_i
-     => b_phi(T_(i+1)) >= 2*lambda_i
-
-And induction goes.
-\end{comment}
+\emph{i.e.} $A_\infty$ divides $P$. Finally, Eq.~\eqref{eq:rateconv}
+giving the rate of convergence follows from the writing:
+$$A_\infty - A_i = R_i + R_{i+1} + \cdots + R_j + \cdots$$
+together with the facts that $b_\varphi(R_j) \geq 2^i \kappa$ and $\deg 
+R_j \leq d{-}1$ for all $j \geq i$.
+
+\begin{rem}
+\label{rem:slope-factor}
+It follows from the proof above that the conclusion of 
+Theorem~\ref{theo:slope-factor} is still correct as soon as $A_0$ 
+and $V_0$ satisfies:
+$$b_\varphi\big(V_0 P \mod A_0\big) > 0
+\quad \text{and} \quad
+b_0\big((1 - V_0 B_0) \mod A_0\big) > 0$$
+except that the constant $\kappa$ giving the rate of convergence
+should be now defined as:
+$$\kappa = \min \big( 
+b_\varphi\big(V_0 P \mod A_0\big), \,
+b_0\big((1 - V_0 B_0) \mod A_0\big)\big).$$
+\end{rem}
 
@@
 a factor of $P$ whose Newton function is $\NF(P)_{|[0,d]}$ and $V$ is the inverse of $B = P/A$ modulo $A$.
-We now assume that $P$ is known up to some finite precision: $P = P_0 + 
+We now assume that $P$ is known up to some finite precision: $P = P_\app + 
 O(\hspace{0.5mm}\cdots)$ where the object inside the $O$ depends on the 
@@
 always assume that $P$ is monic in the sense that its leading
-coefficient is \emph{exactly} $1$; the precision datum on $P$ then 
-concerns only the coefficients up to degree $n{-}1$. Similarly, noting
+coefficient is \emph{exactly}~$1$; the precision datum on $P$ then 
+only concerns the coefficients up to degree $n{-}1$. Similarly, noting
 that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
@@
 \noindent
-{\bf About the first question.}
-We first assume that we are working with lattice precision.
-The answer to the first question is then given by the theory of 
-\cite{padicprec} and is controlled by the differential of the
-application sending $P$ to $A^{(1)}$.
+{\bf Newton precision.}
+We assume that the precision on the input $P$ has the shape 
+$O(\varphi_P)$ where $\varphi_P$ is a Newton function of degree 
+$n{-}1$. From now on, we assume that the precision $O(\varphi_P)$ 
+is nondegenerate in the sense of Definition~\ref{def:nondeg}. This
+ensures in particular that the Newton polygon of $P$ is well defined. 
+We import the notations $\varphi$, $b_\varphi$ and $b_0$ from \S 
+\ref{ssec:Newtoniter} and refer to 
+Eqs.~\eqref{eqdef:functionphi}--\eqref{eqdef:b0} for the definitions.
+
+\begin{prop}
+We keep all the above notations and assumptions. We set:
+$$\delta = \min_{x \in [0,n{-}1]} \varphi_P(x) - \varphi(x)$$
+and assume that $\delta > 0$. 
+Then the factor $A^{(1)}$ is known with precision at least 
+$O(\varphi_{A^{(1)}})$ with
+$\varphi_{A^{(1)}} = \varphi_{|[0,d{-}1]} - \varphi(d) + \delta$.
+\end{prop}
+
+\begin{proof}
+Let $\delta_P \in K[X]$ be such that $\NP(\delta_P) \geq \varphi_P$.
+Let $A_\app^{(1)}$ and $A^{(1)}$ be the monic factors of $P_\app$
+and $P_\app + \delta P$ respectively whose Newton functions are
+$\varphi_{|[0,d]} - \varphi(d)$.
+
+We define the sequences $(A_i)$ and $(V_i)$ by the recurrence: 
+\begin{align*} 
+A_0 & = A_\app, \quad V_{0} = V_\app \medskip \\ A_{i+1} 
+&= A_i + \big(V_i (P_\app + \delta_P) \:\mod\: A_i\big), \smallskip \\ 
+V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\ 
+&  \hspace{1cm}\text{where } B_{i+1} = (P_\app + \delta_P) \div A_{i+1}. 
+\end{align*} 
+where $A_\app$ and $V_\app$ are those related to $P_\app$. Note that 
+$A_\app = a_d \cdot A^{(1)}_\app$ if $a_d$ denotes the coefficient of 
+$X^d$ in $P$. By Remark~\ref{rem:slope-factor}, we know that the 
+sequence $(A_i)$ converges to $A = a_d \cdot A^{(1)}$ and furthermore:
+$$\NP(A - A_\app) \geq \varphi_{[0,d{-}1]} + 
+b_\varphi\big((V_\app \cdot \delta P) \mod A_\app\big)$$ 
+since $(V_\app P_\app) \mod A_\app = (1 - V_\app B_\app) \mod A_\app = 
+0$. Using repeatedly Lemma~\ref{lem:bphib0}, we obtain:
+$$b_\varphi\big((V_\app \cdot \delta P) \mod A_\app\big)
+  \geq b_0(V_\app) + b_\varphi(\delta P) \\
+  \geq b_\varphi(\delta P) \geq \delta.$$
+Thus $\NP(A - A_\app) \geq \varphi_{[0,d{-}1]} + \delta$. Dividing
+by $a_d$, we find
+$\NP(A^{(1)} - A^{(1)}_\app) \geq \varphi_{A^{(1)}}$ and we are done.
+\end{proof}
+
+\medskip
+
+\noindent
+{\bf Lattice precision.}
+In the case of lattice precision, the answer to the first question is 
+then given by the theory of \cite{padicprec} and is controlled by the 
+differential of the application sending $P$ to $A^{(1)}$.
 
@@
 
-\todo{Derive from Proposition~\ref{prop:precA1} a theoretical
-result on Newton precision.}
-
-\medskip
-
-\noindent
-{\bf About the second question.}
-We now study the problem of the effective computation of $A^{(1)}$.
-
 \end{document}

### 2016-01-08 polynomials/muldivfact.tex

@@
 \label{def:nondeg}
-Let $P = P_0 + O(\varphi_P)$. We say that the Newton precision 
+Let $P = P_\app + O(\varphi_P)$. We say that the Newton precision 
 $O(\varphi_P)$ on $P$ is \emph{nondegenerate} if $\varphi_P \geq 
-\NF(P_0)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
-$\NP(P_0)$.
+\NF(P_\app)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
+$\NP(P_\app)$.
 \end{deftn}
@@
 polynomial whose Newton function is not less than $\varphi_P$, we
-have $\NP(P_0 + \delta P) = \NP(P_0)$.
+have $\NP(P_\app + \delta P) = \NP(P_\app)$.
 
@@
 We shall sometimes denote it $O(H)$ in order to emphasize that it should 
-be considered as a precision datum. The notation $P(X) + O(H)$ then 
-refers to any polynomial in the $W$-affine space $P(X) + H$. Tracking
-lattice precision can be done using differentials as shown in
+be considered as a precision datum. The notation $P_\app(X) + O(H)$ then 
+refers to any polynomial in the $W$-affine space $P_\app(X) + H$. 
+Tracking lattice precision can be done using differentials as shown in
 \cite[Lemma~3.4 and Proposition~3.12]{padicprec}: if $f : K_{\leq n}[X] 
@@
 surjective differential, under mild assumption on $H$, we have:
-$$f(P(X)+H) = f(P(X)) + f'(P(X))(H)$$
-where $f'(P(X))$ denotes the differential of $f$ at $P(X)$. The
-equality sign reflets the optimality of the method.
+$$f(P_\app(X)+H) = f(P_\app(X)) + f'(P_\app(X))(H)$$
+where $f'(P_\app(X))$ denotes the differential of $f$ at $P_\app(X)$. 
+The equality sign reflets the optimality of the method.
 
@@
 Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
-$d$ respectively. Set $\mu = \psi(d) - \psi(d-1)$. Letting $\Delta$
-be the greatest affine function of slope $\mu$ with $\Delta \leq
+$d$ respectively. Set $\lambda = \psi(d) - \psi(d-1)$. Letting $\Delta$
+be the greatest affine function of slope $\lambda$ with $\Delta \leq
 \varphi_{|[d,n]}$ and $\delta = \Delta(d) - \psi(d)$, we define:
@@
 \varphi \nfdiv \psi & : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
-& x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
+& x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \lambda h.
 \end{array}$$
@@
 \label{theo:EDivisionNP}
-Let $A, B \in K[X]$ with $B \neq 0$. Let $d$ be the degree of $B$.
-Let $\mu = \NF(B)(d) - \NF(B)(d{-}1)$ be the last slope of $B$ and 
-$y = \Delta(x)$ be the
-equation of the highest line of slope $\mu$ lying below $\NP(A)$.
-Then:
+Given $A, B \in K[X]$ with $B \neq 0$, we have:
 \begin{align}
-\NF(A \mod B) & \geq \NF(B) \nfmod \NF(A) \label{eq:AmodB} \\
+\NF(A \mod B) & \geq \NF(A) \nfmod \NF(B) \label{eq:AmodB} \\
 \text{and} \hspace{4mm}
-\NF(A \div B) & \geq \NF(B) \nfdiv \NF(A) \label{eq:AdivB}
+\NF(A \div B) & \geq \NF(A) \nfdiv \NF(B) \label{eq:AdivB}
 \end{align}
@@
 we have to prove is:
-$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d)
+$$\NF(R_n)(x) \geq \NF(B)(x) - \lambda(n-d)
 \quad \text{for } n\geq d.$$
@@
 \begin{align*}
-\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \mu(n{-}d) \\
-& = - \mu(n{+}1{-}d)
+\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \lambda(n{-}d) \\
+& = - \lambda(n{+}1{-}d)
 \end{align*}
-since $\mu = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
-$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$ for all $x$. On the 
+since $\lambda = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
+$\NF(c_n B)(x) \geq \NF(B)(x) - \lambda(n{+}1{-}d)$ for all $x$. On the 
 other hand, for all $x$, we have:
-$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \mu$$
-from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$. As a
+$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \lambda$$
+from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \lambda(n{+}1{-}d)$. As a
 consequence
-$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$
+$\NF(R_{n+1})(x) \geq \NF(B)(x) - \lambda(n{+}1{-}d)$
 and the induction follows.
@@
 are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
-$$A = A_0 + O(\varphi_A)
+$$A = A_\app + O(\varphi_A)
 \quad \text{and} \quad
-B = B_0 + O(\varphi_B)$$
-Here $A_0$ and $B_0$ are some approximations of $A$ and $B$ respectively 
+B = B_\app + O(\varphi_B)$$
+Here $A_\app$ and $B_\app$ are some approximations of $A$ and $B$ respectively 
 and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
@@
 \begin{align*}
-Q_0 + \delta Q & = (A_0 + \delta A) \div (B_0 + \delta B) \\
-R_0 + \delta R & = (A_0 + \delta A) \mod (B_0 + \delta B)
+Q_\app + \delta Q & = (A_\app + \delta A) \div (B_\app + \delta B) \\
+R_\app + \delta R & = (A_\app + \delta A) \mod (B_\app + \delta B)
 \end{align*}
-where $Q_0 = A_0 \div B_0$ and $R_0 = A_0 \mod B_0$.
+where $Q_\app = A_\app \div B_\app$ and $R_\app = A_\app \mod B_\app$.
 We have to show that $\NF(\delta Q) \geq \varphi \nfdiv \NF(B)$ and
 $\NF(\delta R) \geq \varphi \nfmod \NF(B)$.
-Set $\delta X = \delta A - Q_0 \delta B$. 
-Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_0) \geq
+Set $\delta X = \delta A - Q_\app \delta B$. 
+Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_\app) \geq
 \NF(A) \nfdiv \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
@@
 precision $O(H_A)$ and $O(H_B)$ respectively:
-$$A = A_0 + O(H_A)
+$$A = A_\app + O(H_A)
 \quad \text{and} \quad
-B = B_0 + O(H_B).$$
+B = B_\app + O(H_B).$$
 where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
@@
 differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
-\mapsto X \mod Y$ at the point $(A_0, B_0)$. Writing $Q_0 = A_0 \div
-B_0$ and $R_0 = A_0 \mod B_0$, this can be done by
+\mapsto X \mod Y$ at the point $(A_\app, B_\app)$. Writing $Q_\app = A_\app \div
+B_\app$ and $R_\app = A_\app \mod B_\app$, this can be done by
 expanding the relation:
-$$A_0 + dA = (B_0 + dB) (Q_0 + dQ) + (R_0 + dR)$$
+$$A_\app + dA = (B_\app + dB) (Q_\app + dQ) + (R_\app + dR)$$
 and neglecting the terms of order $\geq 2$. We get this way 
-$dA = B_0 dQ + Q_0 dB + dR$
+$dA = B_\app dQ + Q_\app dB + dR$
 meaning that $dQ$ and $dR$ appears respectively as the quotient and
-the remainder of the Euclidean division of $dX = dA - B_0 dQ$ by $B_0$.
+the remainder of the Euclidean division of $dX = dA - B_\app dQ$ by $B_\app$.
 
@@
 \in \Z[X]$ (known exactly) of degree $5$. Our aim is to compare the 
-numerical stability of the multiplication in the quotient $\Z_2[X]/M(X)$ 
+numerical stability of the multiplication in the quotient $\Z_2[X]/M$ 
 depending on the precision model we are using. In order to do so, we 
@@
 in the case of jagged and Newton precision, the precision on the output 
-may be written into the form $\sum_{i=1}^4 O(2^{N_i}) \: X^i$ and 
-$G = \sum_{i=1}^4 (N_i - N)$;
+may be written into the form $\sum_{i=0}^4 O(2^{N_i}) \: X^i$ and 
+$G = \sum_{i=0}^4 (N_i - N)$;
 
@@
 $A_\app = a_d \cdot A^{(1)}_\app$ if $a_d$ denotes the coefficient of 
-$X^d$ in $P$. By Remark~\ref{rem:slope-factor}, we know that the 
+$X^d$ in $P_\app$. By Remark~\ref{rem:slope-factor}, we know that the 
 sequence $(A_i)$ converges to $A = a_d \cdot A^{(1)}$ and furthermore:
-$$\NP(A - A_\app) \geq \varphi_{[0,d{-}1]} + 
+$$\NP(A - A_\app) \geq \varphi_{|[0,d{-}1]} + 
 b_\varphi\big((V_\app \cdot \delta P) \mod A_\app\big)$$ 
@@
 
+\todo{Answer here the second question.}
+
 \medskip
@@
 {\bf Lattice precision.}
-In the case of lattice precision, the answer to the first question is 
-then given by the theory of \cite{padicprec} and is controlled by the 
-differential of the application sending $P$ to $A^{(1)}$.
+The precision datum is given here by a lattice $H_P$ in $K_{\leq n{-}1}
+[X]$; we shall then write:
+$$P = P_\app + O(H_P)$$
+where $P_\app$ is a \emph{monic} approximation of the inexact polynomial 
+$P$ we want to factor. We assume from now that $H_P$ is sufficiently 
+small so that the Newton polygon of $P$ is well defined. We then can
+define the function:
+$$F : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
+mapping a polynomial $P$ to the couple $(A^{(1)}, B^{(1)})$ obtained
+from it. We set $(A_\app^{(1)}, B_\app^{(1)}) = F(P_\app)$.
+
+\smallskip
+
+\noindent
+We make the following hypothesis \textbf{(H)}:
+
+\smallskip
+
+\noindent
+\hfill
+\begin{minipage}{8cm}
+The lattice $H_P$ is a first order lattice at every point
+of $P_\app + H_P$ in the sense of \cite[Definition~3.3]{padicprec},
+\emph{i.e.} for all $P \in P_\app + H_P$:
+
+\vspace{-5mm}
+
+$$F(P + H_P) = F(P) + F'(P)(H_P).$$
+\end{minipage}
+\hfill \null
+
+\bigskip
+
+\noindent
+Obviously \textbf{(H)} gives an answer to the first question we have 
+raised above: the precision on the couple $A^{(1)}$ is the lattice 
+$H_{A^{(1)}}$ defined as projection on the first component of 
+$F'(P_\app)(H_P)$. It turns out that it can be computed explicitely as 
+shown by the next proposition.
 
@@
 \label{prop:precA1}
-The application $P \mapsto A^{(1)}$ (where $A^{(1)}$ is defined as before) is 
-of class $C^1$ in a neighborhood of $P_0$ and its differential 
-at $P_0$ is the linear mapping:
+
+The application $P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
+and its differential at at some point $P$ is the linear mapping:
 $$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
-where $V^{(1)}$ is the inverse of $B^{(1)}$ modulo $A^{(1)}$.
+where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
+$B^{(1)}$ modulo $A^{(1)}$.
 \end{prop}
@@
 \begin{proof}
-There certain exists a neighborhood $\mathcal U$ of $P_0$ on which the 
-Newton polygon does not vary. On $\mathcal U$, the function $f : P 
-\mapsto (A^{(1)},B^{(1)})$ is well defined. It is furthermore bijective 
-and its inverse $g$ is given by $(A^{(1)},B^{(1)}) \mapsto A^{(1)}B^{(1)}$.
-Clearly $g$ is of class $C^1$ and its differential is given by
+The function $F$ is injective and a left inverse of it is
+$G : (A^{(1)},B^{(1)}) \mapsto A^{(1)}B^{(1)}$.
+Clearly $G$ is of class $C^1$ and its differential is given by
 \begin{equation}
@@
 \end{equation}
-Thanks to Bézout Theorem, it is invertible as soon as $A^{(1)}$ and $B^{(1)}$ are
-coprime. This happens on $f(\mathcal U) = g^{-1}(\mathcal U)$ because 
-$\NP(A^{(1)})$ and $\NP(B^{(1)})$ do not shape a common slope. As a consequence
-$f$ is of class $C^1$ on $\mathcal U$. Moreover, its differential is
-obtained by inverting Eq.~\eqref{eq:dPdAdB}. Reducing modulo $A^{(1)}$, we
-get $dP \equiv B^{(1)} \: dA^{(1)} \pmod {A^{(1)}}$.
-The claimed result follows after having noticed that $dA^{(1)}$
-has degree at most $d{-}1$.
+Thanks to Bézout Theorem, it is invertible as soon as $A^{(1)}$ and 
+$B^{(1)}$ are coprime, which is true because $\NP(A^{(1)})$ and 
+$\NP(B^{(1)})$ do not shape a common slope. As a consequence $F$ is of 
+class $C^1$ and its differential is obtained by inverting 
+Eq.~\eqref{eq:dPdAdB}. Reducing modulo $A^{(1)}$, we get $dP \equiv 
+B^{(1)} \: dA^{(1)} \pmod {A^{(1)}}$. The claimed result follows after 
+having noticed that $dA^{(1)}$ has degree at most $d{-}1$.
 \end{proof}
 
+\todo{Answer here the second question. (The Proposition below should
+be useful at some point.)}
+
+\begin{prop}
+We assume \textbf{(H)}.
+
+\vspace{-2mm}
+
+\begin{enumerate}[(i)]
+\item Let $\tilde A^{(1)} \in K_{=d}[X]$ such that
+$\tilde A^{(1)} \tilde B^{(1)} \in P_\app + O(H_P)$ with
+$\tilde B^{(1)} = P_\app \div \tilde A^{(1)}$.
+Then $\tilde A^{(1)} \in A_\app^{(1)} + H_{A^{(1)}}$.
+\item Let in addition $\tilde V^{(1)} \in K_{\leq d{-}1}[X]$ such that:
+$$\big(\tilde B^{(1)} \tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}
+= H_P \mod \tilde A^{(1)}.$$
+Then $H_{A^{(1)}} = \big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}$.
+\end{enumerate}
+\end{prop}
+
+\begin{proof}
+\todo{Write it.}
+\end{proof}
+
 \end{document}

### 2016-01-08 polynomials/muldivfact.tex
Proof of Proposition 4.11 added

@@
 \begin{rem}
+\label{rem:unicity}
 The divisor $A$ is uniquely determined by the conditions of 
-Theorem~\ref{theo:slope-factor}.
+Theorem~\ref{theo:slope-factor}. Indeed, consider two divisors
+$A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
+Then $L = \text{llcm}(A,A')$ is a divisor of $P$ as well and the 
+slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
+\NF(P)(d{-}1)$. Therefore $\deg L = d$ and $L$ differs from $A$ and
+$A'$ by a multiplicative nonzero constant. Then, if $A$ and $A'$ 
+share in addition the same leading coefficient, they must coincide.
 \end{rem}
@@
 \label{prop:precA1}
-
 The application $P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
@@
 \begin{proof}
-\todo{Write it.}
+\emph{(i)} 
+Set $\tilde P = \tilde A^{(1)} \tilde B^{(1)}$. We know by assumption
+that $\tilde P = P_\app + O(H_P)$. Thus $F(\tilde P)$ is well defined.
+The unicity of the slope factorization (\emph{cf} 
+Remark~\ref{rem:unicity}) further implies that $F(\tilde P) = 
+(\tilde A^{(1)}, \tilde B^{(1)})$. The claim result now follows from
+the hypothesis~\textbf{(H)}.
+
+\smallskip
+
+\emph{(ii)} By applying~\textbf{(H)} with $P = \tilde P$ and replacing 
+$F'(\tilde P)$ by its expression given by Proposition~\ref{prop:precA1}, 
+we find:
+$$H_{A^{(1)}} = \big((\tilde B^{(1)})^{-1} \cdot H_P\big) \mod \tilde A^{(1)}.$$
+Dividing $\big(\tilde B^{(1)} \tilde V^{(1)} H_P\big) \mod \tilde 
+A^{(1)} = H_P \mod \tilde A^{(1)}$ by $\tilde B^{(1)}$ modulo $\tilde
+A^{(1)}$, we get
+$H_{A^{(1)}} = \big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}$
+as expected.
 \end{proof}

### 2016-01-08 polynomials/muldivfact.tex
Minor corrections
Typos, etc...

@@
 \section{Introduction}
+
+\todo{Write introduction.}
+
+\medskip
+
+\noindent
+{\bf Main Results.}
+
+\medskip
+
+\noindent
+{\bf Organization of the paper.}
+
+\medskip
 
@@
 
-It worths be noting that tracking precision using this representation is 
+It is worth noting that tracking precision using this representation is 
 rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
@@
 which takes a finite value at $n$ and whose epigraph $\Epi(\varphi)$ 
-have integral extremal points.
+have extremal points with integral abscissa.
 \end{deftn}
@@
 \todo{Add introduction.}
-In the sequel, we shall will the notation $A \div B$ (resp $A \mod B$)
+In the sequel, we use the notation $A \div B$ (resp $A \mod B$)
 for the quotient (resp. the remainder) in the Euclidean division of $A$ 
@@
 $\deg A$ and $\deg B$ respectively.
-We are interesting in determining the precision on $A \mod B$ and $A
+We are interested in determining the precision on $A \mod B$ and $A
 \div B$. The following proposition gives a theoretical answer under
@@
 We observe several interesting properties. First of all, the gains for 
-Newton precision and jagged precision always agree though one may has 
+Newton precision and jagged precision always agree though one may have 
 thought at first that Newton precision is weaker. Since performing
@@
 
+\begin{comment}
 \subsubsection{Tristan's proof}
@@
 \subsubsection{Xavier's proof}
+\end{comment}
 
@@
 \text{and} \quad
-b_0(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_0 x.
+b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
+\quad \text{for } i \in \{0,1\}.
 \label{eqdef:b0}
 \end{align}
-Set also $b_\varphi(0) = b_0(0) = +\infty$ by convention.
-With the normalization of $P$ we have chosen above, we have $b_0(P)
+Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
+With the normalization of $P$ we chose above, we have $b_0(P)
 = b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
-$Q$.
+$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.
 
@@
 \label{lem:bphib0}
-Let $b \in \{b_\varphi, b_0\}$. For $Q_1, Q_2 \in K[X]$:
+Let $b \in \{b_\varphi, b_0, b_1\}$. For $Q_1, Q_2 \in K[X]$:
 
@@
 
-c) $b_\varphi(Q_1 Q_2) \geq b_\varphi(Q_1) + b_0(Q_2) \geq
-b_\varphi(Q_1) + b_\varphi(Q_2)$
+c) $b_\varphi(Q_1 Q_2) \geq b_\varphi(Q_1) + b_0(Q_2)$
 
@@
 \subset t_2(C)$ where $C$ is the convex cone generated by the vectors 
-$(0,1)$ and $(1, \lambda_0)$. Thus
-$$\NP(Q) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
+starting from $(0,0)$ to $(0,1)$ and $(1, \lambda_0)$. Thus
+$$\NP(Q_1 Q_2) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
 t_2 \circ t_1\big(\Epi(\varphi)\big)$$
@@
 \NF(A_i) = \varphi_{|[0,d]}, \smallskip \\
-b_0(V_i) \geq 0, \quad
+b_1(V_i) \geq 0, \quad
 b_\varphi(R_i) \geq 2^i \kappa \smallskip \\
@@
 an extremal point at abscissa $d$, we get $\NF(A_0) = \varphi_{|[0,d]}$.
-Clearly $b_0(V_0) \geq 0$ since $V_0 = 1$. It follows from the 
+Clearly $b_1(V_0) \geq 0$ since $V_0 = 1$. It follows from the 
 definitions that $R_0 = S_0 = P \mod A_0 = (P{-}A_0) \mod A_0$. 
@@
 Lemma~\ref{lem:bphib0} applies with $A = A_i$ and $A = A_{i+1}$.
-Now using the definition of $B_{i+1}$, we get $b_0(B_{i+1}) \geq 0$.
-As a consequence:
+Now coming back to the the definition of $B_{i+1}$ and using
+Theorem~\ref{theo:EDivisionNP}, we get $b_0(B_{i+1}) \geq b_1(B_{i+1}) 
+\geq 0$. As a consequence:
 $$b_\varphi(R_i B_{i+1}) \geq b_\varphi(R_i) + b_0(B_{i+1})
@@
 the same strategy, we deduce successively
-$b_0(\Delta V_i) \geq 2^i \kappa$ using \eqref{eqdef:Viminus},
-$b_0(1{-}Q_i) \geq 2^i \kappa$ using \eqref{eqdef:Qiminus},
-$b_\varphi(R_{i+1}) \geq 2^{i+1} \kappa$ using \eqref{eq:Riplus},
-and finally $b_0(T_{i+1}) \geq 2^{i+1} \kappa$
-using \eqref{eq:ViBiplus}. The equalities and inequalities
+$b_0(\Delta V_i) \geq 2^i \kappa$ using~\eqref{eqdef:Viminus},
+$b_0(1{-}Q_i) \geq 2^i \kappa$ using~\eqref{eqdef:Qiminus},
+$b_\varphi(R_{i+1}) \geq 2^{i+1} \kappa$ using~\eqref{eq:Riplus},
+and then $b_0(T_{i+1}) \geq 2^{i+1} \kappa$
+using~\eqref{eq:ViBiplus}. Finally, coming back to the recurrence
+defining $V_{i+1}$ and remembering that $b_1(V_i)$ and $b_1(B_{i+1})$
+are both nonnegative, we find $b_1(V_{i+1}) \geq 0$.
+The equalities and inequalities
 of Eq.~\eqref{eq:induction} have then all been established for
@@
 \label{rem:slope-factor}
-It follows from the proof above that the conclusion of 
-Theorem~\ref{theo:slope-factor} is still correct as soon as $A_0$ 
-and $V_0$ satisfies:
+It follows from the proof above that the sequence $(V_i)_{i \geq 0}$
+converges as well. Its limit $V_\infty$ is an inverse of $B_\infty
+= P \div A_\infty$ modulo $A_\infty$ and it satisfies in addition
+$b_1(V_\infty) \geq 0$.
+
+Moreover, the conclusion of 
+Theorem~\ref{theo:slope-factor} is still correct if $A_0$ is any 
+polynomial of degree $d$ and leading coefficient $a_d$ and $V_0$
+is any polynomial as soon as they satisfy:
 $$b_\varphi\big(V_0 P \mod A_0\big) > 0
@@
 \begin{prop}
+\label{prop:Newtonprecslope}
 We keep all the above notations and assumptions. We set:
@@
 
+\begin{rem}
+Under the hypothesis~\textbf{(H)} introduced below, a correct precision 
+on $A^{(1)}$ is also $O(\psi_{A^{(1)}})$ where:
+$$\psi_{A^{(1)}} = \big(\varphi_P \nftimes \big(\NF(V_\app)
+- \NF(P)(d)\big)\big) \nfmod \NF(P)_{|[0,d]}.$$
+This follows from Proposition~\ref{prop:precA1} using
+$V^{(1)}_\app = a_d^{-1} V_\app$. It follows in addition from 
+Remark~\ref{rem:slope-factor} that $\NF(V_\app)$ is bounded from below 
+by $x \mapsto \lambda_1 x$. This yields the bound
+$\psi_{A^{(1)}} \geq \big(\varphi_P \nftimes \psi\big) \nfmod 
+\NF(P)_{|[0,d]}$
+where $\psi : [0,d{-}1] \to \R$ is the affine function mapping $x$
+to $\lambda_1 x - \NF(P)(d)$. 
+\end{rem}
+
 \todo{Answer here the second question.}
@@
 define the function:
-$$F : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
+$$F = (F_A, F_B) : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
 mapping a polynomial $P$ to the couple $(A^{(1)}, B^{(1)})$ obtained
@@
 \label{prop:precA1}
-The application $P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
+The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
 and its differential at at some point $P$ is the linear mapping:
@@
 
+A remarkable Corollary of Proposition~\ref{prop:precA1} asserts the
+optimality of Proposition~\ref{prop:Newtonprecslope} in a particular 
+case.
+
+\begin{cor}
+We assume~\textbf{(H)}.
+When the precision of $P$ is given by $O(\varphi_P)$ with:
+$$\varphi_P = (\NF(P) + \delta)_{|[0,d-1]}
+\quad \text{where $\delta \in \R$}$$
+the precision of $A^{(1)}$ given by 
+Proposition~\ref{prop:Newtonprecslope} is optimal.
+\end{cor}
+
+\begin{proof}
+First remark that the $\delta$ defined in the statement of 
+Proposition~\ref{prop:Newtonprecslope} coincide with the $\delta$
+introduced in the Corollary. Define
+$\varphi_{A^{(1)}} = \varphi_{|[0,d{-}1]} - \varphi(d) + \delta$.
+Let $H_P$ (resp. $H_{A^{(1)}})$ be the lattice consisting of
+polynomials of degree at most $n{-}1$ (resp. at most $d{-}1$)
+whose Newton function is not less that $\varphi_P$ (resp. 
+$\varphi_{A^{(1)}}$). We have to show that $F'_A(P_\app)(H_P) = 
+H_{A^{(1)}}$. But, according to Proposition~\ref{prop:precA1}
+the mapping $G : K_{\leq d{-}1}[X] \to K_{\leq n{-}1}[X]$,
+$$dA^{(1)} \mapsto (B_\app^{(1)} \: dA^{(1)}) \mod A_\app^{(1)}$$
+is a right inverse of $F'_A(P_\app)$. It is then enough to prove
+that $G$ takes $H_{A^{(1)}}$ to $H_P$, which can be done easily
+using Theorem~\ref{theo:EDivisionNP}.
+\end{proof}
+
 \todo{Answer here the second question. (The Proposition below should

### 2016-01-09 polynomials/muldivfact.tex
Some remarks added

@@
 \section{Introduction}
-
-\todo{Write introduction.}
-
-\medskip
-
-\noindent
-{\bf Main Results.}
-
-\medskip
-
-\noindent
-{\bf Organization of the paper.}
-
-\medskip
 
@@
 
-It is worth noting that tracking precision using this representation is 
+It worths be noting that tracking precision using this representation is 
 rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
@@
 \todo{Add introduction.}
-In the sequel, we use the notation $A \div B$ (resp $A \mod B$)
+In the sequel, we shall will the notation $A \div B$ (resp $A \mod B$)
 for the quotient (resp. the remainder) in the Euclidean division of $A$ 
@@
 $\deg A$ and $\deg B$ respectively.
-We are interested in determining the precision on $A \mod B$ and $A
+We are interesting in determining the precision on $A \mod B$ and $A
 \div B$. The following proposition gives a theoretical answer under
@@
 We observe several interesting properties. First of all, the gains for 
-Newton precision and jagged precision always agree though one may have 
+Newton precision and jagged precision always agree though one may has 
 thought at first that Newton precision is weaker. Since performing
@@
 \subset t_2(C)$ where $C$ is the convex cone generated by the vectors
-starting from $(0,0)$ to $(0,1)$ and $(1, \lambda_0)$. Thus
-$$\NP(Q_1 Q_2) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
+$(0,1)$ and $(1, \lambda_0)$. Thus
+$$\NP(Q) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
 t_2 \circ t_1\big(\Epi(\varphi)\big)$$
@@
 define the function:
-$$F = (F_A, F_B) : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
+$$F : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
 mapping a polynomial $P$ to the couple $(A^{(1)}, B^{(1)})$ obtained
@@
 \label{prop:precA1}
-The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
+The application $P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
 and its differential at at some point $P$ is the linear mapping:
@@
 
-A remarkable Corollary of Proposition~\ref{prop:precA1} asserts the
-optimality of Proposition~\ref{prop:Newtonprecslope} in a particular 
-case.
-
-\begin{cor}
-We assume~\textbf{(H)}.
-When the precision of $P$ is given by $O(\varphi_P)$ with:
-$$\varphi_P = (\NF(P) + \delta)_{|[0,d-1]}
-\quad \text{where $\delta \in \R$}$$
-the precision of $A^{(1)}$ given by 
-Proposition~\ref{prop:Newtonprecslope} is optimal.
-\end{cor}
-
-\begin{proof}
-First remark that the $\delta$ defined in the statement of 
-Proposition~\ref{prop:Newtonprecslope} coincide with the $\delta$
-introduced in the Corollary. Define
-$\varphi_{A^{(1)}} = \varphi_{|[0,d{-}1]} - \varphi(d) + \delta$.
-Let $H_P$ (resp. $H_{A^{(1)}})$ be the lattice consisting of
-polynomials of degree at most $n{-}1$ (resp. at most $d{-}1$)
-whose Newton function is not less that $\varphi_P$ (resp. 
-$\varphi_{A^{(1)}}$). We have to show that $F'_A(P_\app)(H_P) = 
-H_{A^{(1)}}$. But, according to Proposition~\ref{prop:precA1}
-the mapping $G : K_{\leq d{-}1}[X] \to K_{\leq n{-}1}[X]$,
-$$dA^{(1)} \mapsto (B_\app^{(1)} \: dA^{(1)}) \mod A_\app^{(1)}$$
-is a right inverse of $F'_A(P_\app)$. It is then enough to prove
-that $G$ takes $H_{A^{(1)}}$ to $H_P$, which can be done easily
-using Theorem~\ref{theo:EDivisionNP}.
-\end{proof}
-
 \todo{Answer here the second question. (The Proposition below should

### 2016-01-11 polynomials/muldivfact.tex
Drawing in main proof
Toward a drawing in initialization of main proof.

--- 2016-01-08 polynomials/muldivfact.tex
@@
 definitions that $R_0 = S_0 = P \mod A_0 = (P{-}A_0) \mod A_0$. 
-Noting that $P{-}A_0 = \sum_{i=d+1}^n a_i X^i$, we obtain
+We remark that $P{-}A_0 = \sum_{i=d+1}^n a_i X^i.$
+Using Theorem \ref{theo:EDivisionNP}, we obtain that
 $b_\varphi(P{-}A_0) \geq \kappa$ and then $b_\varphi(R_0) = b_\varphi(S_0)
-\geq \kappa$. Finally observe that:
+\geq \kappa$. 
+%Mettre le dessin ici
+%
+%
+%
+
+\begin{tikzpicture}
+[xscale=0.7, yscale=0.5]
+\fill[blue!30] (4,9.5)--(4,1.6)--(2,1.2)--(0.5,2.8)--(0,6.8)--(0,9.5)--cycle;
+\fill[yellow!50] (6,9.5)--(6,2)--(4,1.6)--(4,9.5)--cycle;
+\fill[green!30] (6,9.5)--(6,2)--(8,4)--(11,9)--(11,9.5)--cycle;
+\draw[->] (-0.5,0)--(11.5,0);
+\draw[->] (0,-0.5)--(0,9.5);
+\draw[very thick] (0,6)--(0.5,2)--(2,.4)--(5,1);
+\draw(5,1)--(8,4)--(11,9);
+\draw[thick] (4,9.5)-- (4,1.6)--(2,1.2)--(0.5,2.8)--(0,6.8);
+\draw[dotted] (6,2)--(4,1.6);
+%useless part?
+\draw[dotted] (7,3)--(2,2)--(0.5,3.6)--(0,7.6);
+\draw[dotted] (8,4)--(2,2.8)--(0.5,4.4)--(0,8.4);
+\draw[dotted] (9,5.667)--(2,4.267)--(0.5,5.867)--(0.046,9.5); %%(0,9.867);
+\draw[dotted] (10,7.333)--(2,5.733)--(0.5,7.333)--(0.229,9.5); %%--(0,11.333);
+\draw[dotted] (11,9)--(2,7.2)--(0.5,8.8)-- (.413,9.5);%%(0,12.8);
+
+\draw[<->] (4,.8)--(4,1.6);
+\draw[dotted,thick] (5,0)--(5,9.5);
+\draw[dotted,thick] (4,0)--(4,9.5);
+\draw[dotted,thick] (6,0)--(6,9.5);
+\node[scale=0.8, below left] at (0,0) { $0$ };
+\node[scale=0.8, below] at (5,0) { $d$ };
+\node[scale=0.8, below] at (4,0) { $d{-}1$ };
+\node[scale=0.8, below] at (6,0) { $d{+}1$ };
+\node[scale=0.9, below right] at (2.5,.6) { $\mathbf{A_0}$ };
+\node[scale=0.9, right] at (10.2,7.3) { $P$ };
+%\node[scale=0.9, right] at (10.5,4.4) { $\Delta$ };
+\node[scale=0.9] at (4.2,1.2) { $\kappa$ };
+\node[thick] at (3,9) { $R_0$ };
+\node[thick] at (3,8.5) { $S_0$ };
+%\draw[dashed] (11,5)--(8,4);
+%\draw (8,4)--(2,2)--(0.5,4)--(0,8);
+\end{tikzpicture}
+
+Finally observe that:
 $$T_0 = (1 - B_0) \mod A_0 = \big((A_0 - P) \div A_0\big) \mod A_0.$$

### 2016-01-12 polynomials/muldivfact.tex
Drawing in main proof
Continued.
Tristan's figure is now Figure 2
Abstract added

@@
 \begin{abstract}
+We study two important operations on polynomials defined over complete 
+discrete valuation fields (\emph{e.g.} $\Q_p$ or $k((t))$ for a 
+field $k$), namely Euclidean division and factorization. We consider 
+notably the question of computing slope factorizations and design a new 
+simple and efficient algorithm based on Newton iteration to solve it.
+One of its main feature consists in avoiding working over an extension 
+of the base field.
+We constantly pay particular attention to stability: we design several
+precision models, compare them and always discuss optimality.
 \end{abstract}
@@
 $b_\varphi(P{-}A_0) \geq \kappa$ and then $b_\varphi(R_0) = b_\varphi(S_0)
-\geq \kappa$. 
-%Mettre le dessin ici
+\geq \kappa$ (see Figure~\ref{fig:boundR0S0}).
 %
-%
-%
-
-\begin{tikzpicture}
-[xscale=0.7, yscale=0.5]
+\begin{figure}
+\hfill
+\begin{tikzpicture}[xscale=0.7, yscale=0.5]
 \fill[blue!30] (4,9.5)--(4,1.6)--(2,1.2)--(0.5,2.8)--(0,6.8)--(0,9.5)--cycle;
@@
 \fill[green!30] (6,9.5)--(6,2)--(8,4)--(11,9)--(11,9.5)--cycle;
-\draw[->] (-0.5,0)--(11.5,0);
+\draw[->] (-0.5,0)--(11,0);
 \draw[->] (0,-0.5)--(0,9.5);
@@
 \draw[dotted] (11,9)--(2,7.2)--(0.5,8.8)-- (.413,9.5);%%(0,12.8);
+\draw[very thick,dotted] (5,1)--(11,2.2);
+\draw[dotted] (0,0)-- (5,1);
 
@@
 \node[scale=0.8, below] at (6,0) { $d{+}1$ };
-\node[scale=0.9, below right] at (2.5,.6) { $\mathbf{A_0}$ };
-\node[scale=0.9, right] at (10.2,7.3) { $P$ };
+\node[scale=0.85, below left] at (1.5,1.3) { $\mathbf{A_0}$ };
+\node[scale=0.85, right] at (10.1,7.3) { $NF(P)$ };
 %\node[scale=0.9, right] at (10.5,4.4) { $\Delta$ };
 \node[scale=0.9] at (4.2,1.2) { $\kappa$ };
-\node[thick] at (3,9) { $R_0$ };
-\node[thick] at (3,8.5) { $S_0$ };
+\node[scale=0.8] at (3,9) { $NP(R_0)$ };
+\node[scale=0.8] at (3,8.3) { $NP(S_0)$ };
+\node[right,scale=0.9,rotate=35] at (6.2,1.7) { $\text{slope} = \lambda_1$ };
+\node[right,scale=0.9,rotate=8] at (6.8,.9) { $\text{slope} = \lambda_0$ };
+\node[thick] at (10,1.6) { $\varphi$ };
 %\draw[dashed] (11,5)--(8,4);
@@
 \end{tikzpicture}
-
+\hfill\null
+
+\vspace{-5mm}
+
+\caption{Bound on $\NF(R_0)$ and $\NF(S_0)$}
+\label{fig:boundR0S0}
+\end{figure}
+%
 Finally observe that:

### 2016-01-14 polynomials/muldivfact.tex
Fixing some typos

--- 2016-01-10 polynomials/muldivfact.tex
@@
 which takes a finite value at $n$ and whose epigraph $\Epi(\varphi)$ 
-have extremal points with integral abscissa.
+have integral extremal points.
 \end{deftn}
@@
 \varphi \nfdiv \psi & : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
-& x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \lambda h.
+& x & \mapsto & \inf_{h \geq 0} \varphi(x+d+h) - \lambda h.
 \end{array}$$
@@
 Write $A = A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$)
-consists of monomials of $A$ of degree at least $d$ (resp. less than
+consists of monomials of $A$ of degree less than $d$ (resp. at least
 $d$). Noting that:
@@
 meaning that $dQ$ and $dR$ appears respectively as the quotient and
-the remainder of the Euclidean division of $dX = dA - B_\app dQ$ by $B_\app$.
+the remainder of the Euclidean division of $dX = dA - Q_\app dB$ by $B_\app$.
 
@@
 track precision in the jagged (or the Newton) precision model. We 
-observe that this difference is usually not neglictible (\emph{cf} 
+observe that this difference is usually not negligible (\emph{cf} 
 notably the case of $M(X) = (X+1)^5 + 2$) meaning that this quality 
@@
 \begin{rem}
-\label{rem:unicity}
 The divisor $A$ is uniquely determined by the conditions of 
-Theorem~\ref{theo:slope-factor}. Indeed, consider two divisors
-$A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
-Then $L = \text{llcm}(A,A')$ is a divisor of $P$ as well and the 
-slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
-\NF(P)(d{-}1)$. Therefore $\deg L = d$ and $L$ differs from $A$ and
-$A'$ by a multiplicative nonzero constant. Then, if $A$ and $A'$ 
-share in addition the same leading coefficient, they must coincide.
+Theorem~\ref{theo:slope-factor}.
 \end{rem}
@@
 
-\begin{comment}
 \subsubsection{Tristan's proof}
@@
 \subsubsection{Xavier's proof}
-\end{comment}
 
@@
 \text{and} \quad
-b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
-\quad \text{for } i \in \{0,1\}.
+b_0(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_0 x.
 \label{eqdef:b0}
 \end{align}
-Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
-With the normalization of $P$ we chose above, we have $b_0(P)
+Set also $b_\varphi(0) = b_0(0) = +\infty$ by convention.
+With the normalization of $P$ we have chosen above, we have $b_0(P)
 = b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
-$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.
+$Q$.
 
@@
 \label{lem:bphib0}
-Let $b \in \{b_\varphi, b_0, b_1\}$. For $Q_1, Q_2 \in K[X]$:
+Let $b \in \{b_\varphi, b_0\}$. For $Q_1, Q_2 \in K[X]$:
 
@@
 
-c) $b_\varphi(Q_1 Q_2) \geq b_\varphi(Q_1) + b_0(Q_2)$
+c) $b_\varphi(Q_1 Q_2) \geq b_\varphi(Q_1) + b_0(Q_2) \geq
+b_\varphi(Q_1) + b_\varphi(Q_2)$
 
@@
 \NF(A_i) = \varphi_{|[0,d]}, \smallskip \\
-b_1(V_i) \geq 0, \quad
+b_0(V_i) \geq 0, \quad
 b_\varphi(R_i) \geq 2^i \kappa \smallskip \\
@@
 an extremal point at abscissa $d$, we get $\NF(A_0) = \varphi_{|[0,d]}$.
-Clearly $b_1(V_0) \geq 0$ since $V_0 = 1$. It follows from the 
+Clearly $b_0(V_0) \geq 0$ since $V_0 = 1$. It follows from the 
 definitions that $R_0 = S_0 = P \mod A_0 = (P{-}A_0) \mod A_0$. 
@@
 Lemma~\ref{lem:bphib0} applies with $A = A_i$ and $A = A_{i+1}$.
-Now coming back to the the definition of $B_{i+1}$ and using
-Theorem~\ref{theo:EDivisionNP}, we get $b_0(B_{i+1}) \geq b_1(B_{i+1}) 
-\geq 0$. As a consequence:
+Now using the definition of $B_{i+1}$, we get $b_0(B_{i+1}) \geq 0$.
+As a consequence:
 $$b_\varphi(R_i B_{i+1}) \geq b_\varphi(R_i) + b_0(B_{i+1})
@@
 the same strategy, we deduce successively
-$b_0(\Delta V_i) \geq 2^i \kappa$ using~\eqref{eqdef:Viminus},
-$b_0(1{-}Q_i) \geq 2^i \kappa$ using~\eqref{eqdef:Qiminus},
-$b_\varphi(R_{i+1}) \geq 2^{i+1} \kappa$ using~\eqref{eq:Riplus},
-and then $b_0(T_{i+1}) \geq 2^{i+1} \kappa$
-using~\eqref{eq:ViBiplus}. Finally, coming back to the recurrence
-defining $V_{i+1}$ and remembering that $b_1(V_i)$ and $b_1(B_{i+1})$
-are both nonnegative, we find $b_1(V_{i+1}) \geq 0$.
-The equalities and inequalities
+$b_0(\Delta V_i) \geq 2^i \kappa$ using \eqref{eqdef:Viminus},
+$b_0(1{-}Q_i) \geq 2^i \kappa$ using \eqref{eqdef:Qiminus},
+$b_\varphi(R_{i+1}) \geq 2^{i+1} \kappa$ using \eqref{eq:Riplus},
+and finally $b_0(T_{i+1}) \geq 2^{i+1} \kappa$
+using \eqref{eq:ViBiplus}. The equalities and inequalities
 of Eq.~\eqref{eq:induction} have then all been established for
@@
 \label{rem:slope-factor}
-It follows from the proof above that the sequence $(V_i)_{i \geq 0}$
-converges as well. Its limit $V_\infty$ is an inverse of $B_\infty
-= P \div A_\infty$ modulo $A_\infty$ and it satisfies in addition
-$b_1(V_\infty) \geq 0$.
-
-Moreover, the conclusion of 
-Theorem~\ref{theo:slope-factor} is still correct if $A_0$ is any 
-polynomial of degree $d$ and leading coefficient $a_d$ and $V_0$
-is any polynomial as soon as they satisfy:
+It follows from the proof above that the conclusion of 
+Theorem~\ref{theo:slope-factor} is still correct as soon as $A_0$ 
+and $V_0$ satisfies:
 $$b_\varphi\big(V_0 P \mod A_0\big) > 0
@@
 \begin{prop}
-\label{prop:Newtonprecslope}
 We keep all the above notations and assumptions. We set:
@@
 
-\begin{rem}
-Under the hypothesis~\textbf{(H)} introduced below, a correct precision 
-on $A^{(1)}$ is also $O(\psi_{A^{(1)}})$ where:
-$$\psi_{A^{(1)}} = \big(\varphi_P \nftimes \big(\NF(V_\app)
-- \NF(P)(d)\big)\big) \nfmod \NF(P)_{|[0,d]}.$$
-This follows from Proposition~\ref{prop:precA1} using
-$V^{(1)}_\app = a_d^{-1} V_\app$. It follows in addition from 
-Remark~\ref{rem:slope-factor} that $\NF(V_\app)$ is bounded from below 
-by $x \mapsto \lambda_1 x$. This yields the bound
-$\psi_{A^{(1)}} \geq \big(\varphi_P \nftimes \psi\big) \nfmod 
-\NF(P)_{|[0,d]}$
-where $\psi : [0,d{-}1] \to \R$ is the affine function mapping $x$
-to $\lambda_1 x - \NF(P)(d)$. 
-\end{rem}
-
 \todo{Answer here the second question.}
@@
 \label{prop:precA1}
-The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
+
+The application $P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
 and its differential at at some point $P$ is the linear mapping:
@@
 \begin{proof}
-\emph{(i)} 
-Set $\tilde P = \tilde A^{(1)} \tilde B^{(1)}$. We know by assumption
-that $\tilde P = P_\app + O(H_P)$. Thus $F(\tilde P)$ is well defined.
-The unicity of the slope factorization (\emph{cf} 
-Remark~\ref{rem:unicity}) further implies that $F(\tilde P) = 
-(\tilde A^{(1)}, \tilde B^{(1)})$. The claim result now follows from
-the hypothesis~\textbf{(H)}.
-
-\smallskip
-
-\emph{(ii)} By applying~\textbf{(H)} with $P = \tilde P$ and replacing 
-$F'(\tilde P)$ by its expression given by Proposition~\ref{prop:precA1}, 
-we find:
-$$H_{A^{(1)}} = \big((\tilde B^{(1)})^{-1} \cdot H_P\big) \mod \tilde A^{(1)}.$$
-Dividing $\big(\tilde B^{(1)} \tilde V^{(1)} H_P\big) \mod \tilde 
-A^{(1)} = H_P \mod \tilde A^{(1)}$ by $\tilde B^{(1)}$ modulo $\tilde
-A^{(1)}$, we get
-$H_{A^{(1)}} = \big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}$
-as expected.
+\todo{Write it.}
 \end{proof}

### 2016-01-14 polynomials/muldivfact.tex
Working on todos

@@
 and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
-function whose epigraph is $\NP(P)$. It is well known \todo{(Give a
-reference?)} that:
+function whose epigraph is $\NP(P)$. It is well known \cite{dwork-geratto-sullivan:Gfunctions}*{\S 1.6} that:
 $$\begin{array}{r@{\,\,}c@{\,\,}l}
@@
 The notion of \emph{lattice precision} was developed in 
-\cite{padicprec}. It encompasses the two previous models and has
+\cite{caruso-roe-vaccon:14a}. It encompasses the two previous models and has
 the decisive advantage of optimality. As a counterpart, it might be
@@
 Tracking lattice precision can be done using differentials as shown in
-\cite[Lemma~3.4 and Proposition~3.12]{padicprec}: if $f : K_{\leq n}[X] 
+\cite{caruso-roe-vaccon:14a}*{Lemma~3.4 and Proposition~3.12}: if $f : K_{\leq n}[X] 
 \to K_{\leq m}[X]$ denotes any strictly differentiable function with
@@
 notion of \emph{diffused digits of precision} introduced in 
-\cite[Definition 2.3]{preclinalg}. We shall use it repeatedly in the 
+\cite{caruso-roe-vaccon:15a}*{Definition 2.3}. We shall use it repeatedly in the 
 sequel in order to compare the behaviour of the three aforementioned 
@@
 
-\todo{Add introduction.}
-In the sequel, we use the notation $A \div B$ (resp $A \mod B$)
-for the quotient (resp. the remainder) in the Euclidean division of $A$ 
-by $B$.
+Euclidean division provides a building block for many algorithms associated
+to polynomials in one variable.  In order to analyze the precision behavior
+of such algorithms, we need to first understand the precision attached to
+the quotient and remainder when dividing two polynomials.
+In the sequel, we use the notation $A \div B$ and $A \mod B$
+for the polynomials satisfying
+\begin{align*}
+A &= (A \div B) \cdot B + (A \mod B) \\
+\deg(A \mod B) &< \deg(B).
+\end{align*}
 
@@
 \noindent
-{\bf Jagged precision.}
-\todo{Write something... Explain in particular that the fast methods
-(based on fast multipliction) may lead to important looses of precision.}
+{\bf Flat and Jagged precision.}
+
+The simplest method for tracking the precision of a polynomial is
+to record each coefficient modulo a fixed power of $p$, yet this
+method suffers when applied to polynomials with small roots.
+The next obvious approach is to record the precision of each
+coefficient individually, a method that we will refer to as \emph{jagged}
+precision.  Jagged precision is the easiest to implement in
+a computer algebra system, since standard polynomial algorithms,
+such as Karatsuba multiplication, can be written for generic coefficient rings.
+
+However, this ease of implementation has a cost:
+methods designed for exact rings will often have suboptimal
+precision behavior, since combining intermediate expressions
+into a final answer may lose precision.  Moreover, extra precision
+in the middle coefficients of a polynomial, above the Newton polygon
+of the remaining terms, will have no effect on any of the values
+of that polynomial.  As a consequence of these deficiencies, we
+do not present a detailed analysis of this precision model here.
 
@@
 
-\todo{Explain how one can use this proposition: we split the computation
-into two parts (approximation and precision)...\\
-Don't forget to give the complexity.}
+With this result in hand, we may split the computation of
+Euclidean division into two pieces, first computing approximations
+$Q_\app$ and $R_\app$ and separately computing $\delta Q$ and $\delta R$.
+Both the approximations and the precision can be computing in time that is
+quasi-linear in the degree. \todo{Is this sufficiently precision complexity analysis?}
 
@@
 where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
-According to the results of \cite{padicprec}, in order to determine the 
+According to the results of \cite{caruso-roe-vaccon:14a}, in order to determine the 
 precision on $A \div B$ and $A \mod B$, we need to compute the 
@@
 Once this has been done, the strategy is quite similar to that explained 
-for Newton precision: \todo{summerize this in one sentence.}
+for Newton precision: compute approximations and precision lattices
+separately for quotient and remainder.
 
@@
 (The term $G_d$ corresponds to diffused digits according to 
-\cite[Definition 2.3]{preclinalg}.)
+\cite{caruso-roe-vaccon:15a}*{Definition 2.3}.)
 
@@
 
-\todo{Add some words about the comparison between $X^5 + 2$ and
-$(X+1)^5 + 2$?}
+The modulus $(X+1)^5 + 2$ shows the advantage of working
+with lattice precision in intermediate computations.  While
+a small amount of the precision gain is diffuse, the number of
+diffuse $p$-adic digits remains constant as $n$ increases,
+and the increase in precision on the actual coefficients
+increases almost on pace with $X^5 + 2$.  In contrast,
+non-lattice methods for $(X+1)^5 + 2$ are on par
+with the other non-Eisenstein examples.
 
@@
 
-A well-known theorem \todo{(Give a reference?)} asserts that each 
+A well-known theorem \cite{dwork-geratto-sullivan:Gfunctions}*{Thm. 6.1} asserts that each 
 extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
@@
 The lattice $H_P$ is a first order lattice at every point
-of $P_\app + H_P$ in the sense of \cite[Definition~3.3]{padicprec},
+of $P_\app + H_P$ in the sense of \cite{caruso-roe-vaccon:14a}*{Definition~3.3},
 \emph{i.e.} for all $P \in P_\app + H_P$:
@@
 
+\bibliographystyle{plain}
+\bibliography{../roebib/Biblio}
+
 \end{document}

### 2016-01-22 polynomials/muldivfact.tex
Adding some todos

@@
 \DeclareMathOperator{\rank}{rank}
+\DeclareMathOperator{\lcm}{lcm}
 
@@
 $A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
-Then $L = \text{llcm}(A,A')$ is a divisor of $P$ as well and the 
+Then $L = \lcm(A,A')$ is a divisor of $P$ as well and the 
 slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
@@
 If $d = n$ (resp. $d = 1$), the sequence $A_i$ is constant equal to
-$P$ (resp. to the constant coefficient of $P$) and theorem is clear.
+$P$ (resp. to the constant coefficient of $P$) and theorem is clear. \todo{Should this be $d=0$?}
 We then assume $1 < d < n$. Without loss of generality, we may further 
@@
 Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
-\mod A_i$, from what \eqref{eqdef:Si2} follows directly.
+\mod A_i$, from which \eqref{eqdef:Si2} follows directly \todo{(there is a sign error here)}.
 By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
@@
 \cup \{+\infty\}$ by:
-\begin{equation}
-\label{eqdef:functionphi}
-\begin{array}{r@{\,\,}ll}
-\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
-& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
-\end{array}
+\begin{equation} \label{eqdef:functionphi}
+\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
+\lambda_0 (x-d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
+%\begin{array}{r@{\,\,}ll}
+%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
+%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
+%\end{array}
 \end{equation}
-For a polynomial $Q \in K[X]$ of degree $n$, we further define:
+\todo{For $x>d$, why not say $\varphi(x) = \lambda_0 x$?} For a polynomial $Q \in K[X]$ of degree $n$, we further define:
 \begin{align}

### 2016-01-22 polynomials/muldivfact.tex
Updating abstract

@@
 We study two important operations on polynomials defined over complete 
-discrete valuation fields (\emph{e.g.} $\Q_p$ or $k((t))$ for a 
-field $k$), namely Euclidean division and factorization. We consider 
-notably the question of computing slope factorizations and design a new 
-simple and efficient algorithm based on Newton iteration to solve it.
-One of its main feature consists in avoiding working over an extension 
-of the base field.
-We constantly pay particular attention to stability: we design several
-precision models, compare them and always discuss optimality.
+discrete valuation fields: Euclidean division and factorization. In particular,
+we design a simple and efficient algorithm for computing slope factorizations,
+based on Newton iteration.  One of its main features is that we avoid working
+over an extension of the base field.
+We pay particular attention to stability, and analyze the behavior
+of the algorithm using several precision models.
 \end{abstract}

### 2016-01-26 polynomials/muldivfact.tex
Toward numerical results for slope factorization
Tentatively mplementation of slope factorization ;
Minor typoes and grammar errors corrected in main article.

--- 2016-01-14 polynomials/muldivfact.tex
@@
 \alignauthor Tristan Vaccon\\
-  \affaddr{Rikkyo University}\\
+  \affaddr{JSPS--Rikkyo University}\\
   \affaddr{\textsf{vaccon@rikkyo.ac.jp}}
@@
 \emph{relative precision}.
-Alternatively, one may thing that the writing~\eqref{eq:CDVFseriesO}
+Alternatively, one may think that the writing~\eqref{eq:CDVFseriesO}
 represents a subset of $K$ which consists of all elements in $K$ for
@@
 precision $\min(N_x,N_y)$. Computations with $p$-adic and Laurent
-series are often handle this way on symbolic computation softwares.
+series are often handled this way on symbolic computation softwares.
 
@@
 Indeed, several precision data may be considered and, as we shall see
-later, each of them has is own interest. Below we detail three of them
-for the special case of polynomials.
+later, each of them has its own interest. Below we detail three models 
+of precision for the special case of polynomials.
 
@@
 The datum of $\varphi$ is equivalent to that of $\Epi(\varphi)$ and they 
-can easily represented and manipulated on a computer.
+can easily be represented and manipulated on a computer.
 \end{rem}
@@
 \cite{caruso-roe-vaccon:14a}. It encompasses the two previous models and has
-the decisive advantage of optimality. As a counterpart, it might be
-very space-consuming and time-consuming for polynomials of large
-degree.
+the decisive advantage of precision optimality.
+ As a counterpart, it might be very space-consuming and time-consuming 
+ for polynomials of large degree.
 
@@
 extension of $K$. We introduce here a variant of this Newton iteration 
-whose main advantage to be defined over $K$ itself.
+whose main advantage is to be defined over $K$ itself.
 
@@
 $A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
-Then $L = \text{llcm}(A,A')$ is a divisor of $P$ as well and the 
+Then $L = \text{lcm}(A,A')$ is a divisor of $P$ as well and the 
 slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
@@
 (1)~what is the precision on the factor $A$, and
-(2)~how one can compute in practive $A$ at this precision?
+(2)~how can one compute in practice $A$ at this precision?
 
@@
 Remark~\ref{rem:unicity}) further implies that $F(\tilde P) = 
-(\tilde A^{(1)}, \tilde B^{(1)})$. The claim result now follows from
+(\tilde A^{(1)}, \tilde B^{(1)})$. The claimed result now follows from
 the hypothesis~\textbf{(H)}.

### 2016-01-28 polynomials/muldivfact.tex
Small progress toward num. results
More interesting exemples.
Need a better understanding of the computation, and then would be ready
to give an array of results.

@@
 
-\todo{Write introduction.}
+Polynomial factorization is a fundamental problem in computational algebra.
+The algorithms used to solve it depend on the ring of coefficients, with finite fields,
+local fields, number fields and rings of integers of particular interest to number theorists.
+In this article, we focus on a task that forms a building block for factorization
+algorithms over complete discrete valuation fields: the decomposition into factors based on the
+slopes of the Newton polygon.
+
+The Newton polygon of a polynomial $f(X) = \sum a_i X^i$ over such a field is given by
+the convex hull of the points $(i, \val(a_i))$ and the point $(0, +\infty)$.  The lower
+boundary of this polygon consists of line segments $(x_j,y_j)$ -- $(x_{j+1},y_{j+1})$
+of slope $s_j$.  The slope factorization of $f(X)$ expresses $f(X)$ as a product
+of polynomials $g_j(X)$ with degree $x_{j+1} - x_j$ whose roots all have valuation $-s_j$.
+Our main result is a new algorithm for computing these $g_j(X)$.
+
+Any computation with $p$-adic fields must work with approximations modulo finite
+powers of $p$, and one of the key requirements in designing an algorithm is
+an analysis of how the precision of the variables evolve over the computation.
+We work with precision models developed by the same authors
+\cite{caruso-roe-vaccon:14a}*{\S 4.2}, focusing on the lattice and Newton models.
+As part of the analysis of the slope factorization algorithm, we describe
+how the precision of the quotient and remainder depend on the input polynomials
+in Euclidean division..
 
@@
 
+Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
+has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
+\begin{align*}
+A_{i+1} &= A_i + (V_i P \% A_i) \\
+B_{i+1} &= P // A_{i+1} \\
+V_{i+1} &= (2V_i - V_i^2B_{i+1}) \% A_{i+1}.
+\end{align*}
+Our main result is Theorem \ref{theo:slope-factor},
+which states that the sequence $(A_i)$ converges
+to a divisor of $P$.
+
 \medskip
@@
 {\bf Organization of the paper.}
+
+After setting notation, in
+Section \ref{sec:prec_data} we recall various models for tracking
+precision in polynomial arithmetic.  We give some background
+on Newton polygons and explain how using lattices to store
+precision can allow for extra \emph{diffuse} $p$-adic digits that
+are not localized on any single coefficient.
+
+In Section \ref{sec:quo_rem}, we consider Euclidean division.
+We describe in Theorem \ref{theo:EDivisionNP}
+how the Newton polygons of the quotient and remainder
+depend on numerator and denominator.  We use this
+result to describe in Proposition \ref{prop:NewtonprecEuclide}
+the precision evolution in Euclidean division using
+the Newton precision model.  We then compare the
+precision performance of Euclidean division in the
+jagged, Newton and lattice models experimentally,
+finding different behavior depending on the modulus.
+
+Finally, in Section \ref{sec:slope_fac} we describe
+our slope factorization algorithm, which is based on
+a Newton iteration.  Unlike other algorithms for
+slope factorization, ours does not require working
+in an extension field \todo{citation?}.  In Theorem
+\ref{theo:slope-factor} we define a sequence
+of polynomials that will converge to the factors
+determined by an extremal point in the Newton
+polygon.  We then discuss the precision behavior
+of the algorithm.
 
@@
 
-\section{Precision data}
+\section{Precision data} \label{sec:prec_data}
 
@@
 
-\section{Euclidean division}
+\section{Euclidean division} \label{sec:quo_rem}
 
@@
 
-\section{Slope factorization}
+\section{Slope factorization} \label{sec:slope_fac}
 
@@
 The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
-and its differential at at some point $P$ is the linear mapping:
+and its differential at some point $P$ is the linear mapping:
 $$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$

### 2016-01-28 polynomials/muldivfact.tex
Introduction

@@
 The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
-and its differential at some point $P$ is the linear mapping:
+and its differential at at some point $P$ is the linear mapping:
 $$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$

### 2016-01-28 polynomials/muldivfact.tex
A few corrections
Small improvements.

--- 2016-01-28 polynomials/muldivfact.tex
@@
 Our main result is Theorem \ref{theo:slope-factor},
-which states that the sequence $(A_i)$ converges
+which states that the sequence $(A_i)$ converges quadritically
 to a divisor of $P$.
@@
 The rest of this subsection is devoted to the proof of the theorem.
-If $d = n$ (resp. $d = 1$), the sequence $A_i$ is constant equal to
-$P$ (resp. to the constant coefficient of $P$) and theorem is clear. \todo{Should this be $d=0$?}
-We then assume $1 < d < n$. Without loss of generality, we may further 
+If $d = n$ (resp. $d = 0$), the sequence $A_i$ is constant equal to
+$P$ (resp. to the constant coefficient of $P$) and theorem is clear. \todo{Should this be $d=0$? (Tristan:) I agree, I have tried to change accordingly the previous and following statement.}
+We then assume $1 \leq d < n$. Without loss of generality, we may further 
 assume that the coefficient $a_d$ is equal to $1$. We set:
@@
 \Delta S_i &= -(R_i B_{i+1}) \mod A_i,  \label{eqdef:Siminus} \\
-S_{i} &= (B_i R_i + T_i S_{i-1} - T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
+S_{i} &= (B_i R_i + T_i S_{i-1} + T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
 \Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
@@
 Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
-\mod A_i$, from which \eqref{eqdef:Si2} follows directly \todo{(there is a sign error here)}.
+\mod A_i$, from which \eqref{eqdef:Si2} follows directly 
+%\todo{(there is a sign error here)}. (Tristan:fixed)
 By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.

### 2016-01-29 polynomials/muldivfact.tex
Moving comments on jagged and flat precision around

--- 2016-01-28 polynomials/muldivfact.tex
@@
 \noindent
+{\bf Flat precision.}
+The simplest method for tracking the precision of a polynomial is
+to record each coefficient modulo a fixed power of $p$.  While
+easy to analyze and implement, this method suffers when
+applied to polynomials whose Newton polygons are far from flat.
+
+\medskip
+
+\noindent
 {\bf Jagged precision.}
-The most naive precision we may consider is the so-called \emph{jagged 
-precision}. It consists in attaching an individual precision (of the
-shape $O(\pi^{N_i})$) to each coefficient separatedly.
+The next obvious approach is to record the precision of each
+coefficient individually, a method that we will refer to as \emph{jagged}
+precision.  Jagged precision is commonly implemented
+in computer algebra systems, since standard polynomial algorithms
+can be written for generic coefficient rings.  However, these
+generic implementations often have suboptimal precision behavior,
+since combining intermediate expressions into a final answer
+may lose precision.  Moreover, when compared to the Newton
+precision model, extra precision in the middle coefficients, above
+the Newton polygon of the remaining terms, will have no effect
+on any of the values of that polynomial.
 
@@
 \noindent
-{\bf Flat and Jagged precision.}
-
-The simplest method for tracking the precision of a polynomial is
-to record each coefficient modulo a fixed power of $p$, yet this
-method suffers when applied to polynomials with small roots.
-The next obvious approach is to record the precision of each
-coefficient individually, a method that we will refer to as \emph{jagged}
-precision.  Jagged precision is the easiest to implement in
-a computer algebra system, since standard polynomial algorithms,
-such as Karatsuba multiplication, can be written for generic coefficient rings.
-
-However, this ease of implementation has a cost:
-methods designed for exact rings will often have suboptimal
-precision behavior, since combining intermediate expressions
-into a final answer may lose precision.  Moreover, extra precision
-in the middle coefficients of a polynomial, above the Newton polygon
-of the remaining terms, will have no effect on any of the values
-of that polynomial.  As a consequence of these deficiencies, we
-do not present a detailed analysis of this precision model here.
-
-\medskip
-
-\noindent
 {\bf Newton precision.}
-We consider here the case where the precision is given by Newton
-functions. Concretely, we pick $A, B \in K[X]$ two polynomials which 
+We first analyze the precision behavior of Euclidean division
+in the Newton model.  Concretely, we pick $A, B \in K[X]$ two polynomials which 
 are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
@@
 when $M$ is irreducible modulo $p=2$ or Eiseistein, there is apparently 
-no benefit of using the lattice precision model. We emphasize that 
+no benefit to using the lattice precision model. We emphasize that 
 these two particular cases correspond to modulus that are usually used 
@@
 
-For other modulus, the situation is quite different and the benefit of 
+For other moduli, the situation is quite different and the benefit of 
 using the lattice precision model becomes more apparent. 

### 2016-01-29 polynomials/muldivfact.tex
Change comments on (X+1)^5+2

@@
 The modulus $(X+1)^5 + 2$ shows the advantage of working
-with lattice precision in intermediate computations.  While
-a small amount of the precision gain is diffuse, the number of
-diffuse $p$-adic digits remains constant as $n$ increases,
-and the increase in precision on the actual coefficients
-increases almost on pace with $X^5 + 2$.  In contrast,
-non-lattice methods for $(X+1)^5 + 2$ are on par
-with the other non-Eisenstein examples.
+with lattice precision in intermediate computations.  Indeed,
+the precision behavior using the lattice model closely parallels
+that of $X^5+2$, since the lattices are related by a change of
+variables.  But this structure is not detected in the Newton
+or jagged models.
 
@@
 If $d = n$ (resp. $d = 0$), the sequence $A_i$ is constant equal to
-$P$ (resp. to the constant coefficient of $P$) and theorem is clear. \todo{Should this be $d=0$? (Tristan:) I agree, I have tried to change accordingly the previous and following statement.}
-We then assume $1 \leq d < n$. Without loss of generality, we may further 
+$P$ (resp. to the constant coefficient of $P$) and theorem is clear.
+We then assume $0 < d < n$. Without loss of generality, we may further 
 assume that the coefficient $a_d$ is equal to $1$. We set:

### 2016-01-30 polynomials/muldivfact.tex
Numerical experimentation for slope factorization
Numerical results added to the article for slope factorization.
Sage implementation is given.

@@
 
+{\bf Numerical experimentations.} We have implemented the following scheme to illustrate the behaviour of precision
+for slope factorization over $\mathbb{Q}_2 [X].$
+We start with random polynomials $A$ and $B$ with given minoration on their Newton polygons, assuring
+they have no common slope.
+We compute $P=AB$ and then $A^{(1)}$, and compare the obtained precision on $A^{(1)}$
+(jagged) with the estimation given using $\Psi_{A^{(1)}}$ and Proposition \ref{prop:precA1}
+assuming~\textbf{(H)}. The mean amount of lost digits for $1000$ experiments is given in the following array.
+Two settings are possible for the initial $A$ and $B$ : relative or absolute precision equal to $100$ for their coefficients.
+
+\medskip
+
+{\small%
+\noindent \hfill%
+\renewcommand{\arraystretch}{1.2}%
+\begin{tabular}{|c|c|c|c|c|}%
+\hline
+\multirow{2}{*}{Polynomials} & 
+\multirow{2}{*}{\hspace{2mm}Precision\hspace{2mm}} & 
+\multicolumn{3}{c|}{Gain of precision} \\
+\cline{3-5}
+& &Jagged & Newton
+      & \raisebox{-0.5mm}{Lattice} \\
+\hline
+\multirow{2}{*}{
+\begin{tikzpicture}[xscale=0.1, yscale=0.05]
+\draw[->] (0,-.5) -- (0,10.5);
+\draw[->] (-0.05,0) -- (13.5,0);
+\draw (6,-.75) -- (6,.75);
+\node[below, right] at (6,-2) {$d$};
+\draw (0,5) -- (2,-2) -- (6,0) -- (10,3) -- (13,10);
+\draw[dotted] (6,0) -- (6,10);
+\node at (3,5) {$A$};
+\node at (9,5) {$B$};
+\end{tikzpicture}}
+& abs & $-14.5$ & $-14.7$ & $0$ \\
+& rel & $-1.5$ & $-3.6$ & $0$ \\
+\hline
+\multirow{2}{*}{\begin{tikzpicture}[xscale=0.11, yscale=0.05]
+\draw[->] (0,-.5) -- (0,10.5);
+\draw[->] (-0.05,0) -- (13.5,0);
+\draw (3,-.75) -- (3,.75);
+\node[below, left] at (3,-2.5) {$d$};
+\draw (0,5) -- (2,1) -- (3,0) -- (8,-2) -- (10,10);
+\draw[dotted] (3,0) -- (3,10);
+\node at (1.5,5) {$A$};
+\node at (6.5,5) {$B$};
+\end{tikzpicture}}
+& abs & $0$ & $0$ & $0$ \\
+& rel & $-0.3$ & $-1.2$ & $0$ \\
+\hline
+\end{tabular}
+\hfill \null}
+
+\medskip
+
 \bibliographystyle{plain}

### 2016-01-30 polynomials/muldivfact.tex
Wrote an algorithm for slope factorization

--- 2016-01-29 polynomials/muldivfact.tex
@@
 
+\newcommand{\softO}{O\tilde{~}}
+
 \renewcommand{\prec}{\text{\rm prec}}
@@
 based on Newton iteration.  One of its main features is that we avoid working
-over an extension of the base field.
+with fractional exponents.
 We pay particular attention to stability, and analyze the behavior
@@
 how the precision of the quotient and remainder depend on the input polynomials
-in Euclidean division..
+in Euclidean division.
 
@@
 {\bf Main Results.}
-
+%
 Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
@@
 A_{i+1} &= A_i + (V_i P \% A_i) \\
-B_{i+1} &= P // A_{i+1} \\
+B_{i+1} &= P \div A_{i+1} \\
 V_{i+1} &= (2V_i - V_i^2B_{i+1}) \% A_{i+1}.
@@
 {\bf Organization of the paper.}
-
+%
 After setting notation, in
@@
 slope factorization, ours does not require working
-in an extension field \todo{citation?}.  In Theorem
+with fractional exponents \todo{citation?}.  In Theorem
 \ref{theo:slope-factor} we define a sequence
@@
 and $\lambda_1$ are both finite.
+
+We define the function $\varphi : \R^+ \to \R 
+\cup \{+\infty\}$ by:
+\begin{equation} \label{eqdef:functionphi}
+\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
+\lambda_0 (x - d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
+%\begin{array}{r@{\,\,}ll}
+%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
+%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
+%\end{array}
+\end{equation}
 We notice that, when the polynomial is changed into $cP$ where $c$ is a 
@@
 for $cP$. As a consequence we may assume that $P$ is normalized so that 
-$\NF(P)(d) = d \lambda_0$. We define the function $\varphi : \R^+ \to \R 
-\cup \{+\infty\}$ by:
-\begin{equation} \label{eqdef:functionphi}
-\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
-\lambda_0 (x-d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
-%\begin{array}{r@{\,\,}ll}
-%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
-%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
-%\end{array}
-\end{equation}
-\todo{For $x>d$, why not say $\varphi(x) = \lambda_0 x$?} For a polynomial $Q \in K[X]$ of degree $n$, we further define:
+$\NF(P)(d) = d \lambda_0$, so that $\varphi(x) = \lambda_0 x$ for $x > d$.
+For a polynomial $Q \in K[X]$ of degree $n$, we further define:
 \begin{align}
@@
 Theorem~\ref{theo:slope-factor} is still correct if $A_0$ is any 
-polynomial of degree $d$ and leading coefficient $a_d$ and $V_0$
+polynomial of degree $d$ with leading coefficient $a_d$ and $V_0$
 is any polynomial as soon as they satisfy:
@@
 
-\todo{Answer here the second question.}
+We can now move to the second question we have raised before, 
+\emph{i.e.} the design of an algorithm for computing $A^{(1)}$ with the 
+precision given by Proposition~\ref{prop:Newtonprecslope}. Our strategy 
+consists in computing first the precision and applying then the Newton 
+iteration until the expected precision is reached. Below is the
+precise description of our algorithm.
+
+\noindent\hrulefill
+
+\noindent {\bf Algorithm} {\tt slope\_factorisation\_Newton}
+
+\noindent{\bf Input:} a monic polynomial $P + O(\varphi_P) \in K_n[X]$,
+
+\noindent\phantom{{\bf Input:} }a break point $d$ of $\NP(P)$
+
+\noindent{\bf Output:} the factor $A$ described above
+
+\smallskip\noindent 1.\ %
+Compute the functions $\NF(P)$ and $\varphi$
+
+\smallskip\noindent 2.\ %
+Compute $\varphi_A = \varphi_{|[0,d-1]} - \varphi(d) + \min_{x \in [0,n]}
+\varphi_P(x) - \varphi(x)$
+
+\smallskip\noindent 3.\ %
+Compute $\kappa = \NP(P)(d{+}1) + \NF(P)(d{-}1) - 2 \cdot \NF(P)(d)$
+
+\smallskip\noindent 4.\ %
+Set $i = 0$, $A_0 = \sum_{i=0}^d a_i X^i$
+($a_i = \text{coeffs of } P$), $V_0 = 1$
+
+\smallskip\noindent 5.\ %
+{\bf repeat until} $\varphi_{|[0,d{-}1]} + 2^i \kappa \geq \varphi_A$
+
+\smallskip\noindent 6.\ \hspace{5mm}%
+lift $A_i$, $V_i$ and $P$ at enough precision
+
+\smallskip\noindent 7.\ \hspace{5mm}%
+compute
+
+\smallskip
+
+\hspace{1cm}$\bullet$
+$A_{i+1} = A_i + (P V_i \mod A_i)$
+
+\hspace{1cm}\phantom{$\bullet$ }%
+at precision $O(\varphi_{|[0,d{-}1]} + 2^{i+1} \kappa)$
+
+\smallskip
+
+\hspace{1cm}$\bullet$
+$V_{i+1} = (2 V_i - V_i^2 \cdot (P \div A_{i+1})) \mod A_{i+1}$
+
+\hspace{1cm}\phantom{$\bullet$ }%
+at precision $O(x \mapsto \lambda_1 x + 2^{i+1}\kappa)$
+
+\smallskip\noindent 8.\ \hspace{5mm}%
+set $i = i+1$
+
+\smallskip\noindent 9.\ 
+{\bf return} $A_i$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\begin{rem}
+The precision needed at line 6 is of course governed by the computation
+performed at line 7. Note that it can be either computed \emph{a
+priori} by using Proposition~\ref{prop:NewtonprecEuclide} or dynamically 
+by using relaxed algorithms from 
+\cite{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
+cases, it is in $O(\pi^{N_i})$ with $N_i = O(2^i \kappa + \min 
+\NF(P))$.
+\end{rem}
+
+\noindent
+It follows from Theorem~\ref{theo:slope-factor}, 
+Remark~\ref{rem:slope-factor} and Proposition~\ref{prop:Newtonprecslope} 
+that Algorithm {\tt slope\_factorisation\_Newton} is correct and 
+stable. Using the standard soft-$O$ notation $\softO(\cdot)$ for
+hiding logarithmic factor, our algorithm performs at most $\softO(n)$ 
+combinatorial operations and $\softO(n)$ operations in $K$ at precision 
+$O(\pi^N)$ with $N = O(\max \varphi_P - \min \NF(P))$ if one uses 
+quasi-optimal algorithms for multiplication and Euclidean division of 
+polynomials.
 
@@
 
-\todo{Answer here the second question. (The Proposition below should
-be useful at some point.)}
+As for the second question, the discussion is similar to the case of 
+Newton precision expect that we need a new stop criterium. It is given 
+by the following proposition.
 
@@
 
+\noindent
+{\bf Numerical experimentations.} 
+%
+We have implemented the following scheme to illustrate the behaviour of 
+precision for slope factorization over $\mathbb{Q}_2 [X].$
+We start with random polynomials $A$ and $B$ with given minoration on 
+their Newton polygons, assuring they have no common slope.
+We compute $P=AB$ and then $A^{(1)}$, and compare the obtained 
+precision on $A^{(1)}$ (jagged) with the estimation given using 
+$\psi_{A^{(1)}}$ and Proposition \ref{prop:precA1} 
+assuming~\textbf{(H)}. The mean amount of lost digits for $1000$ 
+experiments is given in the following array. Two settings are possible 
+for the initial $A$ and $B$ : relative or absolute precision equal to 
+$100$ for their coefficients.
+
+\medskip
+
+{\small%
+\noindent \hfill%
+\renewcommand{\arraystretch}{1.3}%
+\begin{tabular}{|c|c|c|c|c|}%
+\hline
+\multirow{2}{*}{Polynomials} & 
+\multirow{2}{*}{\hspace{1mm}Precision\hspace{1mm}} & 
+\multicolumn{3}{c|}{Gain of precision} \\
+\cline{3-5}
+& &Jagged & Newton & Lattice \\
+\hline
+\multirow{2}{*}{%
+\begin{tikzpicture}[xscale=0.11, yscale=0.05]
+\draw[transparent] (-2,11) rectangle (15,11);
+\draw[->] (0,-.5) -- (0,10.5);
+\draw[->] (-0.05,0) -- (13.5,0);
+\draw (6,-.75) -- (6,.75);
+\node[scale=0.7,below] at (6,0) {$d$};
+\draw (0,5) -- (2,-2) -- (6,0) -- (10,3) -- (13,10);
+\draw[dotted] (6,0) -- (6,10);
+\node at (3.5,5) {$A$};
+\node at (8.5,5) {$B$};
+\end{tikzpicture}}
+& absolute & $-14.5$ & $-14.7$ & $0.0$ \\
+& relative & $\phantom{0}{-}1.5$ & $\phantom{0}{-}3.6$ & $0.0$ \\
+\hline
+\multirow{2}{*}{\begin{tikzpicture}[xscale=0.11, yscale=0.05]
+\draw[transparent] (-2,11) rectangle (15,11);
+\draw[->] (0,-.5) -- (0,10.5);
+\draw[->] (-0.05,0) -- (13.5,0);
+\draw (6,-.75) -- (6,.75);
+\node[scale=0.7,below] at (6,0) {$d$};
+\draw (0,5) -- (6,0) -- (11,-2) -- (13,10);
+\draw[dotted] (6,0) -- (6,10);
+\node at (3.5,5) {$A$};
+\node at (8.5,5) {$B$};
+\end{tikzpicture}}
+& absolute & $\phantom{-0}0.0$ & $\phantom{-0}0.0$ & $0.0$ \\
+& relative & $\phantom{0}{-}0.3$ & $\phantom{0}{-}1.2$ & $0.0$ \\
+\hline
+\end{tabular}
+\hfill \null}
+
 \bibliographystyle{plain}
-\bibliography{../roebib/Biblio}
+\bibliography{muldivfact}
 

### 2016-01-31 polynomials/muldivfact.tex
Tiny change in todo spacing
Title changed
One typo
One typo corrected...

@@
 
-\title{Division and Factorization of p-Adic Polynomials}
+\title{Division and Slope Factorization of p-Adic Polynomials}
 
@@
 \smallskip\noindent 9.\ 
-{\bf return} $A_i$
+{\bf return} $A_i + O(\varphi_A)$
 
@@
 As for the second question, the discussion is similar to the case of 
-Newton precision expect that we need a new stop criterium. It is given 
+Newton precision expect that we need a new stopping criterion. It is given 
 by the following proposition.

### 2016-01-31 polynomials/muldivfact.tex
References to Montes algorithm, add Euclidean division to Main Results

@@
 
+Polynomial factorization over local fields has seen a great deal of progress recently
+\citelist{\cite{pauli:10a} \cite{guardia-nart-pauli:12a} \cite{guardia-montes-nart:08a} \cite{montes:99a}}
+following an algorithm of Montes.  Slope factorization provides a subroutine
+in such algorithms \cite{pauli:10a}*{\S 2}.  For the most difficult inputs, it is not the dominant
+contributor to the runtime of the algorithm, but in some circumstances it will be.
+
 Any computation with $p$-adic fields must work with approximations modulo finite
@@
 Our main result is Theorem \ref{theo:slope-factor},
-which states that the sequence $(A_i)$ converges quadritically
+which states that the sequence $(A_i)$ converges quadratically
 to a divisor of $P$.
+
+In order to prove Theorem \ref{theo:slope-factor}, we also determine
+the precision of the quotient and remainder in Euclidean division,
+which may be of independent interest.  These results are found in
+Section \ref{sec:prec_track}.
 
@@
 slope factorization, ours does not require working
-with fractional exponents \todo{citation?}.  In Theorem
+with fractional exponents\todo{ citation?}.  In Theorem
 \ref{theo:slope-factor} we define a sequence
@@
 
-\subsection{Tracking precision}
+\subsection{Tracking precision} \label{sec:prec_track}
 

### 2016-01-31 polynomials/muldivfact.tex
Some remarks added. Paper fits in 8 pages.

@@
-\documentclass{sig-alternate-2013}
+\documentclass{sig-alternate-05-2015}
 
@@
 \usepackage[utf8]{inputenc}
+
+\hyphenation{regarding}
 
@@
 Polynomial factorization over local fields has seen a great deal of progress recently
-\citelist{\cite{pauli:10a} \cite{guardia-nart-pauli:12a} \cite{guardia-montes-nart:08a} \cite{montes:99a}}
+\citelist{\cite{pauli:10a}, \cite{guardia-nart-pauli:12a}, \cite{guardia-montes-nart:08a}, \cite{montes:99a}}
 following an algorithm of Montes.  Slope factorization provides a subroutine
-in such algorithms \cite{pauli:10a}*{\S 2}.  For the most difficult inputs, it is not the dominant
+in such algorithms \cite[Section~2]{pauli:10a}.  For the most difficult inputs, it is not the dominant
 contributor to the runtime of the algorithm, but in some circumstances it will be.
+We underline moreover that the methods introduced in this paper extend 
+partially to the noncommutative setting and appear this way as an
+essential building block in several decomposition algorithms of 
+$p$-adic Galois representations and $p$-adic differential 
+equations~\cite{caruso:16}.
 
@@
 We work with precision models developed by the same authors
-\cite{caruso-roe-vaccon:14a}*{\S 4.2}, focusing on the lattice and Newton models.
+\cite[Section~4.2]{caruso-roe-vaccon:14a}, focusing on the lattice and Newton models.
 As part of the analysis of the slope factorization algorithm, we describe
@@
 \begin{align*}
-A_{i+1} &= A_i + (V_i P \% A_i) \\
+A_{i+1} &= A_i + (V_i P \mod A_i) \\
 B_{i+1} &= P \div A_{i+1} \\
-V_{i+1} &= (2V_i - V_i^2B_{i+1}) \% A_{i+1}.
+V_{i+1} &= (2V_i - V_i^2B_{i+1}) \mod A_{i+1}.
 \end{align*}
@@
 which states that the sequence $(A_i)$ converges quadratically
-to a divisor of $P$.
-
-In order to prove Theorem \ref{theo:slope-factor}, we also determine
+to a divisor of $P$. This provides a quasi-optimal simple-to-implement 
+algorithm for computing slope factorizations.
+We moreover carry out a careful study of the precision and, applying
+a strategy coming from~\cite{caruso-roe-vaccon:14a}, we end up with
+an algorithm that outputs optimal results regarding to accuracy.
+
+In order to prove Theorem~\ref{theo:slope-factor}, we also determine
 the precision of the quotient and remainder in Euclidean division,
@@
 slope factorization, ours does not require working
-with fractional exponents\todo{ citation?}.  In Theorem
-\ref{theo:slope-factor} we define a sequence
+with fractional exponents. % \todo{ citation?}. 
+In Theorem \ref{theo:slope-factor} we define a sequence
 of polynomials that will converge to the factors
@@
 and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
-function whose epigraph is $\NP(P)$. It is well known \cite{dwork-geratto-sullivan:Gfunctions}*{\S 1.6} that:
+function whose epigraph is $\NP(P)$. It is well known \cite[Section~1.6]{dwork-geratto-sullivan:Gfunctions} that:
 $$\begin{array}{r@{\,\,}c@{\,\,}l}
@@
 Tracking lattice precision can be done using differentials as shown in
-\cite{caruso-roe-vaccon:14a}*{Lemma~3.4 and Proposition~3.12}: if $f : K_{\leq n}[X] 
+\cite[Lemma~3.4 and Proposition~3.12]{caruso-roe-vaccon:14a}: if $f : K_{\leq n}[X] 
 \to K_{\leq m}[X]$ denotes any strictly differentiable function with
@@
 notion of \emph{diffused digits of precision} introduced in 
-\cite{caruso-roe-vaccon:15a}*{Definition 2.3}. We shall use it repeatedly in the 
+\cite[Definition~2.3]{caruso-roe-vaccon:15a}. We shall use it repeatedly in the 
 sequel in order to compare the behaviour of the three aforementioned 
@@
 In the sequel, we use the notation $A \div B$ and $A \mod B$
-for the polynomials satisfying
-\begin{align*}
-A &= (A \div B) \cdot B + (A \mod B) \\
-\deg(A \mod B) &< \deg(B).
-\end{align*}
+for the polynomials satisfying $A = (A \div B) \cdot B + (A \mod B)$
+and $\deg(A \mod B) < \deg(B)$.
 
@@
 Both the approximations and the precision can be computing in time that is
-quasi-linear in the degree. \todo{Is this sufficiently precision complexity analysis?}
+quasi-linear in the degree. 
+%\todo{Is this sufficiently precision complexity analysis?}
 
@@
 
-The table below reports the average gain of \emph{absolute} 
-precision $G$ which is observed while executing the algorithm above 
-for various modulus and $n$. The average is taken on a sample of $1000$ 
-random inputs. We recall that $G$ is defined as follows:
+The table of Figure~\ref{fig:modularmult} reports the average gain of 
+\emph{absolute} precision $G$ which is observed while executing the 
+algorithm above for various modulus and $n$. The average is taken on a 
+sample of $1000$ random inputs. We recall that $G$ is defined as 
+follows:
 
@@
 (The term $G_d$ corresponds to diffused digits according to 
-\cite{caruso-roe-vaccon:15a}*{Definition 2.3}.)
-
-\medskip
-
+\cite[Definition~2.3]{caruso-roe-vaccon:15a}.)
+
+\begin{figure}
 {\small%
@@
 
-\medskip
+\caption{Precision for modular multiplication}
+\label{fig:modularmult}
+\end{figure}
 
@@
 
-A well-known theorem \cite{dwork-geratto-sullivan:Gfunctions}*{Thm. 6.1} asserts that each 
+A well-known theorem \cite[Theorem~6.1]{dwork-geratto-sullivan:Gfunctions} asserts that each 
 extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
@@
 The aim of this section is to design efficient and stable algorithms 
-for computing these factorizations. \todo{Compare with previous
-results...}
+for computing these factorizations: indeed the algorithm we obtain has 
+a quasi-optimal complexity (compared to the size of the input 
+polynomial) and outputs a result whose precision is (close to be) 
+optimal. One important feature of our algorithm is its simplicity 
+and its flexibility: in particular, unlike standard previous slope 
+factorization algorithms, it does not require working with
+fractional exponents.
 
@@
 \emph{i.e.} $A_\infty$ divides $P$. Finally, Eq.~\eqref{eq:rateconv}
-giving the rate of convergence follows from the writing:
-$$A_\infty - A_i = R_i + R_{i+1} + \cdots + R_j + \cdots$$
+giving the rate of convergence follows from the writing
+$A_\infty{-}A_i = \sum_{j=i}^\infty R_j$
 together with the facts that $b_\varphi(R_j) \geq 2^i \kappa$ and $\deg 
@@
 except that the constant $\kappa$ giving the rate of convergence
-should be now defined as:
-$$\kappa = \min \big( 
+should be now
+$\kappa = \min \big( 
 b_\varphi\big(V_0 P \mod A_0\big), \,
-b_0\big((1 - V_0 B_0) \mod A_0\big)\big).$$
+b_0\big((1 - V_0 B_0) \mod A_0\big)\big)$.
 \end{rem}
@@
 The precision datum is given here by a lattice $H_P$ in $K_{\leq n{-}1}
-[X]$; we shall then write:
-$$P = P_\app + O(H_P)$$
+[X]$; we shall then write
+$P = P_\app + O(H_P)$
 where $P_\app$ is a \emph{monic} approximation of the inexact polynomial 
@@
 
-\smallskip
-
-\noindent
 We make the following hypothesis \textbf{(H)}:
 
-\smallskip
+\medskip
 
@@
 The lattice $H_P$ is a first order lattice at every point
-of $P_\app + H_P$ in the sense of \cite{caruso-roe-vaccon:14a}*{Definition~3.3},
+of $P_\app + H_P$ in the sense of \cite[Definition~3.3]{caruso-roe-vaccon:14a},
 \emph{i.e.} for all $P \in P_\app + H_P$:
@@
 The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
-and its differential at some point $P$ is the linear mapping:
+and its differential at some point $P$ is the linear mapping
 $$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
@@
 \begin{cor}
-We assume~\textbf{(H)}.
-When the precision of $P$ is given by $O(\varphi_P)$ with:
-$$\varphi_P = (\NF(P) + \delta)_{|[0,d-1]}
-\quad \text{where $\delta \in \R$}$$
+We assume~\textbf{(H)}. Let $\delta \in \R$.
+When the precision of $P$ is given by $O(\NF(P)_{|[0,d-1]} + \delta)$,
 the precision of $A^{(1)}$ given by 
@@
 polynomials of degree at most $n{-}1$ (resp. at most $d{-}1$)
-whose Newton function is not less that $\varphi_P$ (resp. 
+whose Newton function is not less that $\varphi_P = \NF(P)_{|[0,d-1]} + \delta$ (resp. 
 $\varphi_{A^{(1)}}$). We have to show that $F'_A(P_\app)(H_P) = 
-H_{A^{(1)}}$. But, according to Proposition~\ref{prop:precA1}
+H_{A^{(1)}}$. According to Proposition~\ref{prop:precA1}
 the mapping $G : K_{\leq d{-}1}[X] \to K_{\leq n{-}1}[X]$,
-$$dA^{(1)} \mapsto (B_\app^{(1)} \: dA^{(1)}) \mod A_\app^{(1)}$$
+$dA^{(1)} \mapsto (B_\app^{(1)} \: dA^{(1)}) \mod A_\app^{(1)}$
 is a right inverse of $F'_A(P_\app)$. It is then enough to prove
@@
 \begin{prop}
+\label{prop:stop}
 We assume \textbf{(H)}.
 
-\vspace{-2mm}
+\vspace{-1mm}
 
@@
 Then $\tilde A^{(1)} \in A_\app^{(1)} + H_{A^{(1)}}$.
+\vspace{-1mm}
 \item Let in addition $\tilde V^{(1)} \in K_{\leq d{-}1}[X]$ such that:
@@
 the hypothesis~\textbf{(H)}.
-
-\smallskip
 
@@
 
+As a conclusion, the algorithm we propose consists in computing the 
+Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of the
+algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
+until we find a couple $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
+the requirements (i) and (ii) of Proposition~\ref{prop:stop}. Once this 
+couple has been found, one may safely output $\tilde A^{(1)} + 
+O\big(\big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}\big)$ under
+\textbf{(H)}.
+The resulting algorithm has quasi-optimal running 
+time and optimal stability.
+
+\begin{comment}
 \noindent
@@
 \hfill \null}
+\end{comment}
 

### 2016-02-01 polynomials/muldivfact.tex
Small improvements

@@
 \terms{Algorithms, Theory}
-\keywords{}
+\keywords{Algorithms, $p$-adic precision, Newton polygon, factorization}
 
@@
 Polynomial factorization over local fields has seen a great deal of progress recently
-\citelist{\cite{pauli:10a}, \cite{guardia-nart-pauli:12a}, \cite{guardia-montes-nart:08a}, \cite{montes:99a}}
+\citelist{\cite{pauli:10a} \cite{guardia-nart-pauli:12a} \cite{guardia-montes-nart:08a} \cite{montes:99a}}
 following an algorithm of Montes.  Slope factorization provides a subroutine
@@
 The aim of this section is to design efficient and stable algorithms 
-for computing these factorizations: indeed the algorithm we obtain has 
-a quasi-optimal complexity (compared to the size of the input 
+for computing these factorizations. Precisely the algorithm we obtain 
+has a quasi-optimal complexity (compared to the size of the input 
 polynomial) and outputs a result whose precision is (close to be) 
-optimal. One important feature of our algorithm is its simplicity 
-and its flexibility: in particular, unlike standard previous slope 
-factorization algorithms, it does not require working with
-fractional exponents.
+optimal. Two of its important additional features are simplicity and 
+flexibility.
 
@@
 
-In the previous references, the factor $A$ defined above is always 
-obtained \emph{via} a Newton iteration which takes place over a finite 
-extension of $K$. We introduce here a variant of this Newton iteration 
-whose main advantage is to be defined over $K$ itself.
+The factor $A$ defined above is usually obtained \emph{via} a Newton 
+iteration after having prepared our polynomial by flattening the first 
+slope using a change of variables involving possibly rational 
+exponents. We introduce here a variant of this iteration which does not 
+require the flattening step and is entirely defined over $K[X]$.
 

### 2016-02-01 polynomials/muldivfact.tex
typos

@@
 $P$ (resp. to the constant coefficient of $P$) and theorem is clear.
-We then assume $0 < d < n$. Without loss of generality, we may further 
-assume that the coefficient $a_d$ is equal to $1$. We set:
+We then assume $0 < d < n$. We set:
 \begin{align*}
@@
 \end{equation}
-We notice that, when the polynomial is changed into $cP$ where $c$ is a 
+We notice that, when the polynomial $P$ is changed into $cP$ where $c$ is a 
 nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
@@
 for $cP$. As a consequence we may assume that $P$ is normalized so that 
-$\NF(P)(d) = d \lambda_0$, so that $\varphi(x) = \lambda_0 x$ for $x > d$.
+$\NF(P)(d) = d \lambda_0$, \emph{i.e.} $\varphi(x) = \lambda_0 x$ for $x > d$.
 For a polynomial $Q \in K[X]$ of degree $n$, we further define:
@@
 $$T_0 = (1 - B_0) \mod A_0 = \big((A_0 - P) \div A_0\big) \mod A_0.$$
-Therefore $b_0(T_0) \geq \kappa$ results from $b_\varphi(A{-}P_0) \geq 
+Therefore $b_0(T_0) \geq \kappa$ results from $b_\varphi(A_0{-}P) \geq 
 \kappa$ thanks to Lemma~\ref{lem:bphib0}. 
@@
 Using again Lemma~\ref{lem:bphib0}, we then derive from 
-\eqref{eqdef:Biminus} and \eqref{eqdef:Siminus}, $b_0(\Delta B_i) \geq 
+\eqref{eqdef:Biminus} and \eqref{eqdef:Siminus} that $b_0(\Delta B_i) \geq 
 2^i \kappa$ and $b_\varphi(\Delta S_i) \geq 2^i \kappa$.

### 2016-02-23 resultants/resultants.tex
Small corrections in the paper on (sub)resultants

@@
 
-\title{Resultants and subresultants of $p$-adic polynomials}
+\title{Numerical stability of Euclide algorithm\\over ultrametric fields}
 
@@
 symbolic computation. A crucial issue when dealing with $p$-adics is 
-those of stability. Indeed, just like real numbers, $p$-adic numbers are 
+that of stability. Indeed, just like real numbers, $p$-adic numbers are 
 by nature infinite and thus need to be truncated in order to fit in the 
@@
 \emph{precision}. Usual softwares implementing $p$-adics (\emph{e.g.} 
-\textsc{magma} \cite{magma}, \textsc{pari} \cite{pari}, \textsc{sage} 
+\textsc{magma} \cite{magma}, \textsc{pari} \cite{pari}, \textsc{sagemath} 
 \cite{sage}) generally tracks the precision as follows: an individual 
@@
 The aim of this section is to introduce the setting we shall work in 
-throughout this paper (which is a bit more general than those considered 
+throughout this paper (which is a bit more general than that considered 
 in the introduction).
@@
 where the $x_i$'s lie in a fixed set $S$ of representatives of classes
-modulo $\pi$. Therefore, as an additive group, $\A$ is isomorphic to 
-the set of sequences $\N \to k$. On the contrary, the multiplicative
-structure may vary.
+modulo $\pi$ with $0 \in S$.
 
@@
 valuation of $x$ can be easily read off this writing: it is the smallest 
-integer $i$ such that $x_i \not\equiv 0 \pmod \pi$.
+integer $i$ such that $x_i \neq 0$.
 
@@
 $\A$, namely the ball of radius $e^{-N}$ centered at $\sum_{i=0}^{N-1} 
-x_i \pi^i$ (or actually any element congruent to it modulo $\pi^N$).
+x_i \pi^i$ (or actually at any element congruent to it modulo $\pi^N$).
 In other words, on a computer, we cannot work with actual $p$-adic
@@
 
-There exist simple formulae to following precision after each single
+There exist simple formulae for following precision after each single
 elementary computation. For instance, basic arithmetic operations can
@@
 any given algorithm. This is the analogue of the standard \emph{interval 
-arithmetic} over the reals. Many usual softwares (as \textsc{sage}, 
+arithmetic} over the reals. Many usual softwares (as \textsc{sagemath}, 
 \textsc{magma}) implement $p$-adic numbers and formal series this way. 
@@
 \begin{enumerate}[i)]
-\item their coefficients are, up to a sing, maximal minors of the 
+\item their coefficients are, up to a sign, maximal minors of the 
 Sylvester matrix, and
@@
 \begin{enumerate}[i)]
-\item their coefficients are, up to a sing, maximal minors of the 
+\item their coefficients are, up to a sign, maximal minors of the 
 Sylvester matrix\footnote{Indeed, observe that the matrix of $\psi_j$ is a
@@
   \SetKwInOut{Output}{Output}
-  \Input{Two polynomials $A, B \in K_d[X]$ (given at finie precision)}
+  \Input{Two polynomials $A, B \in K_d[X]$ (given at finite precision)}
   \Output{The complete sequence of subresultants of $A$ and $B$.}
@@
 title of the Section) refers to the family of
-algorithms computed \textsc{gcd}s or subresultants by means of successive
+algorithms computing \textsc{gcd}s or subresultants by means of successive
 Euclidean divisions. We believe that the stability of all algorithms in
@@
 implementations of $p$-adic numbers). Nevertheless, it defines a 
-simplified framework where computations can be performed and experiments 
+simplified framework in which computations can be carried out and experiments 
 show that it rather well reflects the behaviour of the loss of precision 
@@
 and $W_j$ be the minimum of the valuations of the coefficients of $R_j$. 
-We of course have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
+Of course we have $V_j \geq W_j$ and we set $\delta_j = V_j - W_j$.
 
@@
 the precision of $S$. Using \eqref{eq:precmul}--\eqref{eq:precdiv}, we 
-find that the computed relation precision on $r_{j+1} \cdot r_j^{-1} 
+find that the computed relative precision on $r_{j+1} \cdot r_j^{-1} 
 \cdot R_j$ is $\min(N_{j+1} - V_{j+1}, N_j - V_j)$. The absolute
 precision of this value is then
-$M = \min(N_{j+1} - \delta_j, N_j - \delta_j + V_{j+1} - V_j)$. This
-quantity is also the precision of $S$ since the other summand
+$M = \min(N_{j+1} - \delta_j, N_j - \delta_j + V_{j+1} - V_j)$. The
+latter is also the precision of $S$ since the first summand
 $R_{j+1}$ is known with higher precision. Repeating the argument, we
@@
 stands for the integer part function. In particular, we observe that 
-$(V_{d-2}, V_{d-1}) \neq (2,1)$ almost surely although the events 
+$(V_{d-1}, V_{d-2}) \neq (2,1)$ almost surely although the events 
 $\{V_{d-1} = 2\}$ and $\{V_{d-2} = 1\}$ both occur with positive 
@@
 is fixed during all the proof). Our conventions imply that $Z_i$ 
-vanishes if $i \geq d-j$. On the contrary, if $i < d-j$, let us define 
+vanishes if $i \geq d-j$. Let us define 
 $\tau(1), \ldots, \tau(d-j)$ as the numbers $\sigma(1), \ldots, 
-\sigma(d-j)$ sorted in increasing order. The random variable $Z_i$ 
-is then the minimum of $\tau(i)$ independant random variables with 
-geometric distribution of parameter $(1 - q^{-1})$ and thus its 
+\sigma(d-j)$ sorted in increasing order. For $i < d-j$, the random variable $Z_i$ 
+is the minimum of $\tau(i)$ independant random variables with 
+geometric distribution of parameter $(1 - q^{-1})$. Thus its 
 distribution is geometric of parameter $(1 - q^{-\tau(i)})$.
@@
 
-We now prove~(4). Let $(Z_i)_{i \geq 0}$ be a countable family 
-of independant random variable having all geometric distribution of 
-parameter $(1 - q^{-1})$. We set $Z = \sum_{i=1}^\infty \min(Z_1, \ldots, 
-Z_i)$. Cleary $V_j \leq Z$ and it is then enough to prove:
-$$\P[Z \geq m] \leq q^{-m + O(\sqrt m)}.$$
+We now prove~(4). Let $(G_i)_{i \geq 0}$ be a countable family 
+of independant random variables having all geometric distribution of 
+parameter $(1 - q^{-1})$. We set $G = \sum_{i=1}^\infty \min(G_1, 
+\ldots, G_i)$. Thanks to Theorem~\ref{th:lawVj}, it is enough to prove
+that $\P[G \geq m] \leq q^{-m + O(\sqrt m)}$.
 We introduce the event $E_m$ formulated as follows:
-\emph{there exists a partition $(m_1, \ldots, m_\ell)$ of $m$ such that $X_i 
-\geq m_i$ for all $i \leq \ell$}.
-Up to a measure-zero subset, $E_m$ contains the event $\{ Z \geq m\}$.
-We obtain this way:
-$$\P[Z \geq m] \leq \P[E_m] \leq
-\sum \, \prod_{i=1}^\ell \P[X_1 \geq m_i]$$
-where the latter sum runs over all partitions $(m_1, \ldots, m_\ell)$ 
-of $m$. Replacing $\P[X_1 \geq m_i]$ by $q^{-m_i}$, we get
+\emph{there exists a partition $\underline m = 
+(m_1, \ldots, m_\ell)$ of $m$ such that 
+$G_i \geq m_i$ for all $i \leq \ell$}.
+We claim that $E_m$ contains the event $\{ G \geq m\}$. Indeed assume
+$G \geq m$ and set $m'_i = \min(G_1, \ldots, G_i)$. Clearly, the 
+sequence $(m'_i)_{i \geq 1}$ is nondecreasing and $\sum_{i=1}^\infty 
+m'_i \geq m$ by assumption. Therefore there exists a partition 
+$(m_1, \ldots, m_\ell)$ of $m$ with $m_i \leq m'_i$ for all $i$.
+These $m_i$'s also satisfy $G_i \geq m_i$ and our clair is then proved. 
+We derive $\P[G \geq m] \leq \P[E_m]$ and therefore:
+$$\P[G \geq m] \leq \sum_{\underline m} \, \prod_{i=1}^\ell \P[G_i \geq m_i]$$
+where the latter sum runs over all partitions $\underline m = (m_1, \ldots, m_\ell)$ 
+of $m$. Replacing $\P[G_i \geq m_i]$ by $q^{-m_i}$, we get
 $\P[E_m] \leq p(m) \cdot q^{-m}$
@@
 formula \cite{andrews}, we know that $\log p(m)$ is equivalent to $\pi 
-\sqrt{2m/3}$. In particular it is in $q^{O(\sqrt m)}$ and~(4) is proved.
+\sqrt{2m/3}$. In particular $p(m) \in q^{O(\sqrt m)}$ and~(4) is proved.
 
@@
 m_0}$, \emph{i.e.} $m_0 - c \sqrt{m_0} \geq \log_q d$. Solving the latest
-equation, we get $m_0 = \log_q + O(\sqrt{\log_q d})$. Moreover
+equation, we get $m_0 = \log_q d + O(\sqrt{\log_q d})$. Moreover
 $\sum_{m=m_0}^\infty d\: q^{-m + c\sqrt m}$ is bounded independantly of
@@
 Let $\mathcal W_j$ and $\mathcal W_{j-1}$ denote 
-the $j$-th cofactors of $(\mathcal U_{j-1}, \mathcal U_j)$ in degree 
+the cofactors of $(\mathcal U_{j-1}, \mathcal U_j)$ in degree 
 $(d{-}j, d{-}j{-}1)$. Define $\mathcal V_j = \alpha \mathcal W_j$ and 
 $\mathcal V_{j-1} = -\alpha \mathcal W_{j-1}$ where $\alpha =
-a^{4j-4d+6}$. The relation:
+a^{2j-2d+4}$. The relation:
 \begin{equation}
@@
 A = (-1)^j \cdot a^{-2} \cdot (\mathcal V_j \mathcal R_{j-1} - \mathcal V_{j-1} \mathcal R_j) \smallskip \\
-B = (-1)^{j-1} \cdot a^{-2} \cdot (\mathcal U_j \mathcal R_{j-1} - \mathcal U_{j-1} \mathcal R_j) \\
+B = (-1)^{j-1} \cdot a^{-2} \cdot (\mathcal U_j \mathcal R_{j-1} - \mathcal U_{j-1} \mathcal R_j). \\
 \end{array}\right.
 \end{equation}
-and let $\varphi_j$ be the function mapping $(\mathcal U_j, \mathcal 
+They are both monic of degree $d$, so that we can define $\varphi_j$
+as the function taking $(\mathcal U_j, \mathcal 
 U_{j-1}, \mathcal R_j, \mathcal R_{j-1})$ to $(A,B)$. The composite 
@@
 \mathcal R_{j-1})$, the relation~\eqref{eq:missingcof} implies that 
-$\mathcal V_{j-1}$ and $\mathcal V_j$ are the missing cofactors and, 
-consequently, $A$ and $B$ have to be given by the 
+$\mathcal V_{j-1}$ and $\mathcal V_j$ are the missing cofactors (up 
+to a sign) and, consequently, $A$ and $B$ have to be given by the 
 system~\eqref{eq:systemAB}.
@@
 polynomials, we can use functoriality and assume that $\ring$ is 
-the field $\Q(c_0, c_1, \ldots, c_n)$ (with $n = 2d$) and that 
+the field $\Q(c_0, c_1, \ldots, c_n)$ (with $n = 2d$) where
 each variable $c_i$ corresponds to one coefficient of $\mathcal U_j$, 
@@
 (R_{j-1}, U_{j-1})$ for some $\mu \in \ring$. Identifying the leadings 
-coefficients, we get $\lambda = \mu$. Writing
-$\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) = 
-\Res^{d-j,d-j-1}(U_{j-1}, U_j)$,
-we get $\lambda^{2(d-j)-1} = 1$. Since the exponent is odd, this implies 
-$\lambda = 1$ and we are done.
+coefficients, we get $\lambda = \mu$. Now observe that:
+$$\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) = 
+\Res^{d-j,d-j-1}(\lambda U_{j-1}, \lambda U_j) = 
+\lambda^{2d-2j-1} \cdot \Res^{d-j,d-j-1}(U_{j-1}, U_j).$$
+Noting that $(\mathcal U_j, \mathcal U_{j-1}, \mathcal R_j, \mathcal 
+R_{j-1})$ and $(U_j, U_{j-1}, R_j, R_{j-1})$ both belong to $\mathcal
+E_j$, we derive:
+$$\mathcal R_j[j]^{2(d-j-1)} = \lambda^{2d-2j-1} \cdot R_j[j]^{2(d-j-1)}$$
+from what we finally get $\lambda = 1$ since $\mathcal R_j = \lambda
+\cdot R_j$.
 \end{proof}
@@
 0$ by convention. Given $n$ and $m$ two integers with $m < n$, we let 
-$V_{(m,n)}$ denote the set of polynomials of the form
+$V_{[m,n)}$ denote the set of polynomials of the form
 $a_m X^m + a_{m+1} X^{m+1} \cdots + a_n X^n$
-with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{(m,n)}$ has
+with $a_i \in \ring$ and $a_n \neq 0$. Clearly, $V_{[m,n)}$ has
 cardinality $(q-1) q^{n-m}$. If $P$ is any polynomial of 
 degree $n$ and $m < n$ is an integer, we further define $P[m{:}] \in 
-V_{(m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
+V_{[m,n)}$ as the polynomial obtained from $P$ by removing its monomials 
 of degree $< m$. Finally, given $(A,B)$ in $\Omega_\ring$, we 
@@
 \Omega_\ring(J) & \to & 
-V_{(n_1,n_2)} \times \cdots \times V_{(n_\ell,n_{\ell+1})} \smallskip \\
+V_{[n_1,n_2)} \times \cdots \times V_{[n_\ell,n_{\ell+1})} \smallskip \\
 (A,B) & \mapsto & 
@@
 The case $j = d-1$ is actually simpler. Indeed, the same argument works 
-expect that we know for sure that $r_{j+1} = r_d$ is invertible since it 
+except that we know for sure that $r_{j+1} = r_d$ is invertible since it 
 is equal to $1$ by convention. In that case, the probability is then 
@@
 the equidistribution of $(A^{(1)}, B^{(1)})$ shows that the 
-joint distribution $(V_j^{(0)}, V_j^{(1)})$ is the same as those
+joint distribution $(V_j^{(0)}, V_j^{(1)})$ is the same as that
 of $(Y_j^{(0)}, Y_j^{(1)})$. Repeating the argument, we see that
@@
 
-Although the defintion of a lattice is similar to that of $\Z$-lattice 
+Although the definition of a lattice is similar to that of $\Z$-lattice 
 in $\R^d$, the geometrical representation of it is quite different. 
@@
 \emph{integral polynomial} functions. In order to proceed, we assume 
-that $E$ is endowed with distingushed ``orthonormal'' 
+that $E$ is endowed with distinguished ``orthonormal'' 
 basis\footnote{One can prove that such a basis always exists.}, that is 
@@
 x_i e_i \Vert_E = \max_{1\leq i \leq n} |x_i|$ for all 
-families of $\lambda_i$'s lying in $K$. In other words, the choice of
-this distingushed ``orthonormal'' basis defines a norm-preserving 
+families of $x_i$'s lying in $K$. In other words, the choice of
+this distinguished ``orthonormal'' basis defines a norm-preserving 
 isomorphism between $E$ and $K^n$ endowed with the sup norm. We assume
-similarly that we are given a distingushed ``orthonormal'' basis
+similarly that we are given a distinguished ``orthonormal'' basis
 $(f_1, \ldots, f_m)$ of $F$. Then any function $f : E \to F$ can
@@
 Let us examine more closely the case of polynomial spaces since it will 
-be considered repeadtly in the sequel. We take $E = K_{<n}[X]$ and $F = 
+be considered repeatedly in the sequel. We take $E = K_{<n}[X]$ and $F = 
 K_{<m}[X]$ and endow both with the Gauss norm, which is defined by:
@@
 basis are the $a_i$'s and the $b_i$'s respectively. Hence, an integral
-polynomial function $f : E \to F$ is nothing but a function mapping a
+polynomial function $f : E \to F$ is nothing but a function mapping
 a polynomial $P$ to a polynomial $Q$ whose coefficients are given by
@@
 particular problem under consideration is often necessary. For now, let 
-us simply assume that we have given two lattices $H_\min$ and $H_\max$ 
+us simply assume that we are given two lattices $H_\min$ and $H_\max$ 
 with the property that:
@@
   \Input{$x$ given at precision $O(H)$}
-  \Output{$g(x)$ given at precision $O(H_{\max,n})$}
+  \Output{$f(x)$ given at precision $O(H_{\max,n})$}
 
@@
 The correctness of Algorithm~\ref{algo:stabilizedF} (under the 
-assumptions listed above) is clear after Lemma~\ref{eq:padicprec}.
+assumptions listed above) is clear after Lemma~\ref{lem:padicprec}.
 
@@
 
-     \lIf{$v_j \geq \frac N 2$}{\textbf{raise} NotImplementedError}
+     \lIf{$\val(r_j) \geq \frac N 2$}{\textbf{raise} NotImplementedError}
 
@@
 first remark that it is satisfied with high probability if $N$ is large 
-compared to $2 \cdot \log_d p$. Thus, replacing eventually $N$ by $3 
-\cdot \log_d p$ (which does not affect the complexity), the 
-assumption~(b) is harmless on average --- but maybe not on particularly 
+compared to $2 \cdot \log_d p$. Thus, replacing if necessary $N$ by $3 
+\cdot \log_d p$ (which does not affect the complexity), it is harmless
+on average --- but might be not on particularly 
 bad instances. We moreover underline that, if we are just interested in 
@@
 
-Another quite interesting question is those of designing an algorithm 
+Another quite interesting question is that of designing an algorithm 
 which combines the precision technology developed in this paper with the 
@@
 Second, experiments show that the obtained results have generally
-more much correct digits that those predicted by arithmetic interval.
-The counterpart is that, expect on small examples, obtaining proved
-results is generally intractable.
+more much correct digits than the number predicted by arithmetic interval.
+The counterpart is that, except on small examples, obtaining proved
+results is often quite difficult and even sometimes intractable.
 
@@
 floating point arithmetic may often compute much more correct digits 
-than those predicted by an analysis based on interval arithmetic. In 
+than the number predicted by a naive analysis based on interval arithmetic. In 
 order to illustrate this last assertion, let us go back to the case of 

### 2016-04-29 polynomials/slopefact.tex
Revision of the slope factorization paper

--- 2016-02-01 polynomials/muldivfact.tex
@@
 
-\permission{%
-}
+\CopyrightYear{2016}
+\setcopyright{licensedothergov}
+\conferenceinfo{ISSAC'16,}{July 19--22, 2016, Waterloo, ON, Canada} \isbn{978-1-4503-4380-0/16/07}
+\acmPrice{\$15.00}
+\doi{http://dx.doi.org/10.1145/2930889.2930897}
+
+\permission{Publication rights licensed to ACM.
+ACM acknowledges that this contribution was authored or co-authored by an employee, contractor or affiliate of a national government.
+As such, the Government retains a nonexclusive, royalty-free right to publish or reproduce this article, or to allow others to do so, for Government purposes only.}
 
@@
 
-\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulation -- \emph{Algebraic Algorithms}}
-\terms{Algorithms, Theory}
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10010147.10010148.10010149.10010150</concept_id>
+<concept_desc>Computing methodologies~Algebraic algorithms</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\vspace{-1mm}
+\ccsdesc[500]{Computing methodologies~Algebraic algorithms}
+\printccsdesc
+
+\vspace{-1.5mm}
 \keywords{Algorithms, $p$-adic precision, Newton polygon, factorization}
@@
 Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
-has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
+has an extremal point at abscissa $d$.
+Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
 \begin{align*}
@@
 \end{align*}
+where $S \div T$ and $S \mod T$ denotes the quotient and the
+remainder in the Euclidean division of $S$ by $T$ respectively.
 Our main result is Theorem \ref{theo:slope-factor},
@@
 slope factorization, ours does not require working
-with fractional exponents. % \todo{ citation?}. 
+with fractional exponents (compare for instance with
+\cite[\S~6]{poteaux-rybowicz}).
 In Theorem \ref{theo:slope-factor} we define a sequence
@@
 degree at most $n$ (resp. exactly $n$) is denoted by $K_{\leq n}[X]$ 
-(resp. $K_{=n}[X]$).
+(resp. $K_n[X]$).
 
@@
 The simplest method for tracking the precision of a polynomial is
-to record each coefficient modulo a fixed power of $p$.  While
+to record each coefficient modulo a fixed power of $\pi$.  While
 easy to analyze and implement, this method suffers when
@@
 $\varphi : [0,n] \to \R \cup \{+\infty\}$ which is piecewise affine, 
-which takes a finite value at $n$ and whose epigraph $\Epi(\varphi)$ 
+which takes a finite value at $n$ and whose epigraph\footnote{Recall
+that the epigraph is the region above the graph.} $\Epi(\varphi)$ 
 have extremal points with integral abscissa.
@@
 Let $P = P_\app + O(\varphi_P)$. We say that the Newton precision 
-$O(\varphi_P)$ on $P$ is \emph{nondegenerate} if $\varphi_P \geq 
+$O(\varphi_P)$ of $P$ is \emph{nondegenerate} if $\varphi_P \geq 
 \NF(P_\app)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
@@
 $\deg A$ and $\deg B$ respectively.
-We are interested in determining the precision on $A \mod B$ and $A
+We are interested in determining the precision of $A \mod B$ and $A
 \div B$. The following proposition gives a theoretical answer under
@@
 According to the results of \cite{caruso-roe-vaccon:14a}, in order to determine the 
-precision on $A \div B$ and $A \mod B$, we need to compute the 
+precision of $A \div B$ and $A \mod B$, we need to compute the 
 differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
@@
 \noindent $\bullet$
-in the case of jagged and Newton precision, the precision on the output 
+in the case of jagged and Newton precision, the precision of the output 
 may be written into the form $\sum_{i=0}^4 O(2^{N_i}) \: X^i$ and 
@@
 \noindent $\bullet$
-in the case of lattice precision, the precision on the output is a 
+in the case of lattice precision, the precision of the output is a 
 lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
@@
 
+\begin{rem}
+The iteration formulae of the above theorem already appeared 
+in \cite[\S~3.1]{berthomieu:hal-00642075}; nevertheless, as far
+as we know, the result of Theorem~\ref{theo:slope-factor} is
+new.
+\end{rem}
+
 %\begin{rem}
@@
 \end{proof}
-
-\begin{comment}
-\subsubsection{Tristan's proof}
-
-\begin{deftn}
-Notation:
-\[P=-d \lambda_0+d \slp(\lambda_0) + l_1 \slp (\lambda_1)+\RS \]
-($\RS$ for remaining slopes)
-
-\begin{tikzpicture}
-\draw[->] (0,-1) -- (0,2);
-
-\draw[->] (-0.05,0) -- (4,0);
-
-\node[left] at (0,0) {$0$};
-
-\draw (1.5,-.1) -- (1.5,.1);
-\node[above] at (1.5,0) {$d$};
-
-\draw (3,-.1) -- (3,.1);
-\node[above] at (2.8,0) {$d+l_1$};
-
-\node[left] at (0,1.5) {$-d \lambda_0$};
-
-\draw (0,1.5)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
-\node[below] at (.75,0.5) {$\lambda_0$};
-\node[below] at (2.25,-0.25) {$\lambda_1$};
-\node[right] at (3.5,0.25) {$\RS$};
-\end{tikzpicture}
-\end{deftn}
-
-{\bf Initial Newton polygons}
-
-\begin{lem}
-$\NP(R_0)=\NP(S_0)$:
-\[ \NP (R_0)=\NP(S_0)=(\lambda_1-\lambda_0)- d \lambda_0+(d-1)\slp( \lambda_0). \]
-
-\begin{tikzpicture}
-\draw[->] (0,-.5) -- (0,2.5);
-
-\draw[->] (-0.05,0) -- (2,0);
-
-\draw[->] (-0.05,1.75) -- (.05,1.75);
-\node[left] at (0,1.75) {$(\lambda_1-\lambda_0)$};
-\node[left] at (-.4,1.4) {$-d \lambda_0$};
-
-\node[left] at (0,0) {$0$};
-
-\draw (1,-.1) -- (1,.1);
-\node[below, right] at (1,-.2) {$d-1$};
-
-\draw (0,1.75) -- (1,.75) -- (1,2.5);
-\node[below] at (.5,1.3) {$\lambda_0$};
-\end{tikzpicture}
-
-$\NP (1-V_0B_0 \% A_0)$:
-\begin{align*}
-\NP ((1-V_0B_0)\% A_0) &\geq (\lambda_1-\lambda_0)+1\slp( \lambda_0)+(d-1) \slp(\lambda_1), \\
-&\geq (\lambda_1-\lambda_0)+d\slp( \lambda_0). \\
-\end{align*} 
-\begin{tikzpicture}
-\draw[->] (0,-1) -- (0,1);
-
-\draw[->] (-0.05,0) -- (2.5,0);
-
-\node[left] at (0,0) {$0$};
-
-\node[above] at (1.5,0) {$d-1$};
-
-\draw (.5,-.1) -- (.5,.1);
-\node[above] at (.5,0) {$1$};
-
-\draw (-.1,.25) -- (.1,.25);
-\node[left] at (0,.25) {$(\lambda_1-\lambda_0)$};
-
-\draw (0,.25) -- (.5,-.25) -- (1,-.375)-- (1,1.5);
-\draw (0,.25) -- (.5,-.25) -- (1,-.75)-- (1,1.5);
-
-\node[below] at (.25,-0.1) {$\lambda_0$};
-\node[below] at (.75,0.1) {$\lambda_1$};
-\end{tikzpicture}
-
-\label{lem:initial_NP}
-\end{lem}
-\begin{proof}
-Clear by Theorem \ref{theo:EDivisionNP}, and the fact that $coeff(P,d)=1.$
-\end{proof}
-
-\medskip
-
-{\bf Unchanged Newton polygons}
-
-\begin{lem} \label{lem_unchanged_NP}
-$\NP (V_i)$ for $i>0$ ($V_0=1$):
-\[\NP(V_i)=0+(d-1)\slp( \lambda_1). \]
-
-\begin{tikzpicture}
-\draw[->] (0,-1) -- (0,1);
-
-\draw[->] (-0.05,0) -- (2,0);
-
-\node[left] at (0,0) {$0$};
-
-\node[below, right] at (1,-.2) {$d-1$};
-
-\draw (0,0) -- (1,-.25) -- (1,1);
-\node[below] at (.5,-0.25) {$\lambda_1$};
-\end{tikzpicture}
-
-$\NP (B_i)$:
-\[\NP(B_i)=0+(l_1)\slp( \lambda_1)+\RS. \]
-
-\begin{tikzpicture}
-\draw[->] (0,-1) -- (0,1);
-
-\draw[->] (-0.05,0) -- (2.5,0);
-
-\node[left] at (0,0) {$0$};
-
-\draw (1.5,-.1) -- (1.5,.1);
-\node[above] at (1.5,0) {$l_1$};
-
-\draw (0,0) -- (1.5,-.375) -- (2,.625);
-\node[below] at (.5,-0.25) {$\lambda_1$};
-\node[below] at (2,-0.25) {$\lambda_2$};
-\end{tikzpicture}
-
-$\NP (A_i)$:
-\[\NP(A_i)=-d \lambda_0+d \slp( \lambda_0). \]
-
-\begin{tikzpicture}
-\draw[->] (0,-1) -- (0,2);
-
-\draw[->] (-0.05,0) -- (2,0);
-
-\node[left] at (0,0) {$0$};
-
-\draw (1.5,-.1) -- (1.5,.1);
-\node[above] at (1.5,0) {$d$};
-
-\node[left] at (0,1.5) {$-d \lambda_0$};
-
-\draw (0,1.5)-- (1.5,0) ;
-\node[below] at (.75,0.5) {$\lambda_0$};
-\end{tikzpicture}
-\end{lem}
-
-\medskip
-
-{\bf Newton polygons inside the loop}
-
-\begin{lem} \label{lem:iteration_inside_the_loop}
-\begin{align*}
-\NP (R_i), \NP(S_i) & \geq -\lambda_0+2^i(\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0), \\
-\NP (1-V_i B_i) & \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
-\end{align*}
-\end{lem}
-
-We prove the last two lemmas by induction.
-
-\begin{proof}
-The case $i=0$ is clear thanks to Theorem \ref{theo:EDivisionNP} and Lemma \ref{lem:initial_NP}.
-
-Now, we assume that for some $i \geq 0$ the formulae defined in the Lemmae \ref{lem_unchanged_NP} and \ref{lem:iteration_inside_the_loop} are satisfied. 
-We prove them for $i+1$.
-
-Firstly, thanks to the minoration on $\NP (R_i)$ and the fact that $R_i=A_{i+1}-A_i$, we have $\NP (A_{i+1}= \NP (A_i).$
-Since $B_{i+1} = P \div A_{i+1},$ we have $\NP (B_{i+1}) = \NP (B_i).$
-We apply \eqref{eqdef:Biminus} to obtain 
-\[\NP (B_i-B_{i+1})  \geq 2^i (\lambda_1-\lambda_0)+(\lambda_1-\lambda_0)+(l_1-1) \slp (\lambda_1)+\RS. \]
-Then, we can apply \eqref{eqdef:Siminus} to obtain 
-\[\NP (S_i-S_{i+1})  \geq -\lambda_0 d+2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
-
-As a consequence, we get that $\NP (S_{i+1}) \geq \NP (S_i)$ (pointwise).
-
-Now, we can apply \eqref{eqdef:Viminus} to obtain that 
-\[\NP (V_i-V_{i+1}) \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
-As a consequence, we get that $\NP(V_i)=\NP(V_{i+1}).$
-Similarly, we use \eqref{eqdef:Qiminus} to obtain that
-\[\NP (1-Q_i)  \geq  2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
-
-This is enough to conclude on the Newton polygon of $R_{i+1}.$
-Indeed, thanks to \eqref{eq:Riplus}, we obtain that 
-\[\NP (R_i) \geq -\lambda_0+2^i(\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0). \]
-We can proceed similarly for $1-V_{i+1} B_{i+1}$ using \eqref{eq:ViBiplus}.
-
-Finally, we can apply \eqref{eqdef:Si2} for $i+1,$
-\begin{align*}
-S_{i+1} & = (B_{i+1} R_{i+1} + (1-V_{i+1} B_{i+1}) S_{i} \\
-& + (1-V_{i+1} B_{i+1})(S_{i+1}-S_{i})) \% A_{i+1},
-\end{align*}
-to obtain the desired minoration on $\NP (S_{i+1}).$ This concludes the proof.
-\end{proof}
-
-\subsubsection{Xavier's proof}
-\end{comment}
 
@@
 chosen precision model. We address the two following questions:
-(1)~what is the precision on the factor $A$, and
+(1)~what is the precision of the factor $A$, and
 (2)~how can one compute in practice $A$ at this precision?
@@
 always assume that $P$ is monic in the sense that its leading
-coefficient is \emph{exactly}~$1$; the precision datum on $P$ then 
+coefficient is \emph{exactly}~$1$; the precision datum of $P$ then 
 only concerns the coefficients up to degree $n{-}1$. Similarly, noting
@@
 {\bf Newton precision.}
-We assume that the precision on the input $P$ has the shape 
+We assume that the precision of the input $P$ has the shape 
 $O(\varphi_P)$ where $\varphi_P$ is a Newton function of degree 
@@
 Obviously \textbf{(H)} gives an answer to the first question we have 
-raised above: the precision on the couple $A^{(1)}$ is the lattice 
+raised above: the precision of the couple $A^{(1)}$ is the lattice 
 $H_{A^{(1)}}$ defined as projection on the first component of 
@@
 
-\begin{comment}
-\noindent
-{\bf Numerical experimentations.} 
-%
-We have implemented the following scheme to illustrate the behaviour of 
-precision for slope factorization over $\mathbb{Q}_2 [X].$
-We start with random polynomials $A$ and $B$ with given minoration on 
-their Newton polygons, assuring they have no common slope.
-We compute $P=AB$ and then $A^{(1)}$, and compare the obtained 
-precision on $A^{(1)}$ (jagged) with the estimation given using 
-$\psi_{A^{(1)}}$ and Proposition \ref{prop:precA1} 
-assuming~\textbf{(H)}. The mean amount of lost digits for $1000$ 
-experiments is given in the following array. Two settings are possible 
-for the initial $A$ and $B$ : relative or absolute precision equal to 
-$100$ for their coefficients.
-
-\medskip
-
-{\small%
-\noindent \hfill%
-\renewcommand{\arraystretch}{1.3}%
-\begin{tabular}{|c|c|c|c|c|}%
-\hline
-\multirow{2}{*}{Polynomials} & 
-\multirow{2}{*}{\hspace{1mm}Precision\hspace{1mm}} & 
-\multicolumn{3}{c|}{Gain of precision} \\
-\cline{3-5}
-& &Jagged & Newton & Lattice \\
-\hline
-\multirow{2}{*}{%
-\begin{tikzpicture}[xscale=0.11, yscale=0.05]
-\draw[transparent] (-2,11) rectangle (15,11);
-\draw[->] (0,-.5) -- (0,10.5);
-\draw[->] (-0.05,0) -- (13.5,0);
-\draw (6,-.75) -- (6,.75);
-\node[scale=0.7,below] at (6,0) {$d$};
-\draw (0,5) -- (2,-2) -- (6,0) -- (10,3) -- (13,10);
-\draw[dotted] (6,0) -- (6,10);
-\node at (3.5,5) {$A$};
-\node at (8.5,5) {$B$};
-\end{tikzpicture}}
-& absolute & $-14.5$ & $-14.7$ & $0.0$ \\
-& relative & $\phantom{0}{-}1.5$ & $\phantom{0}{-}3.6$ & $0.0$ \\
-\hline
-\multirow{2}{*}{\begin{tikzpicture}[xscale=0.11, yscale=0.05]
-\draw[transparent] (-2,11) rectangle (15,11);
-\draw[->] (0,-.5) -- (0,10.5);
-\draw[->] (-0.05,0) -- (13.5,0);
-\draw (6,-.75) -- (6,.75);
-\node[scale=0.7,below] at (6,0) {$d$};
-\draw (0,5) -- (6,0) -- (11,-2) -- (13,10);
-\draw[dotted] (6,0) -- (6,10);
-\node at (3.5,5) {$A$};
-\node at (8.5,5) {$B$};
-\end{tikzpicture}}
-& absolute & $\phantom{-0}0.0$ & $\phantom{-0}0.0$ & $0.0$ \\
-& relative & $\phantom{0}{-}0.3$ & $\phantom{0}{-}1.2$ & $0.0$ \\
-\hline
-\end{tabular}
-\hfill \null}
-\end{comment}
-
 \bibliographystyle{plain}
-\bibliography{muldivfact}
+
+% \bib, bibdiv, biblist are defined by the amsrefs package.
+\begin{bibdiv}
+\begin{biblist}
+\tiny
+
+\bib{berthomieu-hoeven-lecerf:11a}{article}{
+      author={Berthomieu, J{\'e}r{\'e}my},
+      author={van~der Hoeven, Joris},
+      author={Lecerf, Gr{\'e}goire},
+       title={{Relaxed algorithms for $p$-adic numbers}},
+        date={2011},
+     journal={J. Th{\'e}orie des Nombres des Bordeaux},
+      volume={23},
+      number={3},
+       pages={{541\ndash 577}},
+}
+
+\bib{berthomieu:hal-00642075}{article}{
+  TITLE = {{Polynomial root finding over local rings and application to error correcting codes}},
+  author = {Berthomieu, J{\'e}r{\'e}my},
+  author = {Lecerf, Gr{\'e}goire},
+  author = {Quintin, Guillaume},
+  JOURNAL = {{Applicable Algebra in Engineering, Communication and Computing}},
+  PUBLISHER = {{Springer Verlag}},
+  VOLUME = {24},
+  NUMBER = {6},
+  PAGES = {413\ndash 443},
+  YEAR = {2013},
+  DOI = {10.1007/s00200-013-0200-5},
+}
+
+\bib{caruso:16}{unpublished}{
+      author={Caruso, Xavier},
+       title={Slope factorization of ore polynomials},
+        date={2016},
+        note={in preparation},
+}
+
+\bib{caruso-roe-vaccon:14a}{article}{
+      author={Caruso, Xavier},
+      author={Vaccon, Tristan},
+      author={Roe, David},
+       title={Tracking $p$-adic precision},
+        date={2014},
+     journal={LMS Journal of Computation and Mathematics},
+      volume={17 (Special issue A)},
+       pages={274\ndash 294},
+}
+
+\bib{caruso-roe-vaccon:15a}{inproceedings}{
+      author={Caruso, Xavier},
+      author={Vaccon, Tristan},
+      author={Roe, David},
+       title={$p$-adic stability in linear algebra},
+        date={2015},
+   booktitle={Proceedings of the 2015 acm on international symposium on
+  symbolic and algebraic computation},
+   publisher={ACM},
+     address={New York},
+       pages={{101\ndash 108}},
+}
+
+\bib{dwork-geratto-sullivan:Gfunctions}{book}{
+      author={Dwork, Bernard},
+      author={Gerotto, Giovanni},
+      author={Sullivan, Francis},
+       title={{An introduction to $G$-functions}},
+   publisher={Princeton U.P.},
+     address={Princeton},
+        date={1994},
+}
+
+\bib{guardia-montes-nart:08a}{article}{
+      author={Gu\`ardia, Jordi},
+      author={Montes, Jes\'us},
+      author={Nart, Enric},
+       title={Newton polygons of higher order in algebraic number theory},
+        date={2012},
+     journal={Transactions of the AMS},
+      volume={364},
+      number={1},
+       pages={361\ndash 416},
+}
+
+\bib{guardia-nart-pauli:12a}{article}{
+      author={Guàrdia, Jordi},
+      author={Nart, Enric},
+      author={Pauli, Sebastian},
+       title={Single-factor lifting and factorization of polynomials over local
+  fields},
+        date={2012},
+     journal={Journal of Symbolic Computation},
+      volume={47},
+      number={11},
+       pages={1318 \ndash  1346},
+}
+
+\bib{montes:99a}{thesis}{
+      author={Montes, Jes\'us},
+       title={Pol\'igonos de newton de orden superior y aplicaciones
+  aritm\'eticas},
+        type={Ph.D. Thesis},
+        date={1999},
+}
+
+\bib{pauli:10a}{inproceedings}{
+      author={Pauli, Sebastian},
+       title={{Factoring polynomials over local fields II}},
+        date={2010},
+   booktitle={Algorithmic number theory, 9th international symposium},
+      editor={Hanrot, Guillaume},
+      editor={Morain, Fran{\c c}ois},
+      editor={Thom\'e, Emmanuel},
+     address={Nancy, France},
+}
+
+\bib{hoeven:02a}{article}{
+      author={van~der Hoeven, Joris},
+       title={{Relax, but don't be too lazy}},
+        date={2002},
+     journal={J. Symbolic Comput.},
+      volume={34},
+      number={6},
+       pages={479\ndash 542},
+}
+
+\bib{hoeven:07a}{article}{
+      author={van~der Hoeven, Joris},
+       title={New algorithms for relaxed multiplication},
+        date={2007},
+     journal={J. Symbolic Comput.},
+      volume={42},
+      number={8},
+       pages={792\ndash 802},
+}
+
+\bib{poteaux-rybowicz}{inproceedings}{
+  author    = {Poteaux, Adrien},
+  author    = {Rybowicz, Marc},
+  title     = {Improving Complexity Bounds for the Computation of Puiseux Series over Finite Fields},
+  booktitle = {ISSAC'15},
+  pages     = {299--306},
+  year      = {2015},
+}
+
+\end{biblist}
+\end{bibdiv}
 

### 2016-04-30 polynomials/muldivfact.tex
Some word choice changes, adding definition of % and //, adding references

@@
 Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
-has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
+has a break at $i=d$.  Denote by $A \div B$ and $A \mod B$ the quotient and remainder
+in the Euclidean division of $A$ by $B$. Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
 \begin{align*}
@@
 $$F = (F_A, F_B) : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
-mapping a polynomial $P$ to the couple $(A^{(1)}, B^{(1)})$ obtained
+mapping a polynomial $P$ to the pair $(A^{(1)}, B^{(1)})$ obtained
 from it. We set $(A_\app^{(1)}, B_\app^{(1)}) = F(P_\app)$.
@@
 Obviously \textbf{(H)} gives an answer to the first question we have 
-raised above: the precision on the couple $A^{(1)}$ is the lattice 
+raised above: the precision of the pair $A^{(1)}$ is the lattice 
 $H_{A^{(1)}}$ defined as projection on the first component of 
@@
 algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
-until we find a couple $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
+until we find a pair $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
 the requirements (i) and (ii) of Proposition~\ref{prop:stop}. Once this 
-couple has been found, one may safely output $\tilde A^{(1)} + 
+pair has been found, one may safely output $\tilde A^{(1)} + 
 O\big(\big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}\big)$ under
@@
 We compute $P=AB$ and then $A^{(1)}$, and compare the obtained 
-precision on $A^{(1)}$ (jagged) with the estimation given using 
+precision of $A^{(1)}$ (jagged) with the estimation given using 
 $\psi_{A^{(1)}}$ and Proposition \ref{prop:precA1} 

### 2016-04-30 polynomials/muldivfact.tex
Insert reference for fractional exponents, add comparison to single-factor-lifting, and tweak to bring back to 8 pages

@@
 following an algorithm of Montes.  Slope factorization provides a subroutine
-in such algorithms \cite[Section~2]{pauli:10a}.  For the most difficult inputs, it is not the dominant
-contributor to the runtime of the algorithm, but in some circumstances it will be.
-We underline moreover that the methods introduced in this paper extend 
-partially to the noncommutative setting and appear this way as an
+in such algorithms \cite{pauli:10a}*{Section~2}.  Note that our algorithm lifts
+a \emph{slope factor} corresponding to a segment of the Newton polygon,
+rather than the irreducible factor found by single factor lifting \cite{guardia-nart-pauli:12a}.
+The slope factors are easier to find; as a consequence our iteration is less involved.
+We also remark that the methods introduced in this paper extend 
+partially to the non-commutative setting and appear this way as an
 essential building block in several decomposition algorithms of 
@@
 a Newton iteration.  Unlike other algorithms for
-slope factorization, ours does not require working
-with fractional exponents. % \todo{ citation?}. 
+slope factorization \cite{chistov:90a}*{Eq. (3)}, ours does not require working
+with fractional exponents.
 In Theorem \ref{theo:slope-factor} we define a sequence
@@
 We recall that one can attach a Newton function to each polynomial.
-If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
+If $P(X) = \sum_{i=0}^n a_n X^n \in K[X]$, we define its Newton
 polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
@@
 slope using a change of variables involving possibly rational 
-exponents. We introduce here a variant of this iteration which does not 
+exponents \cite{chistov:90a}*{Eq. (3)}. We introduce here a variant of this iteration which does not 
 require the flattening step and is entirely defined over $K[X]$.
@@
 If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
-is constant and the theorem is obvious. We then assume that $\lambda_0$
-and $\lambda_1$ are both finite.
+is constant and the theorem follows, so assume that $\lambda_0$
+and $\lambda_1$ are finite.
 
@@
 \end{equation}
-We notice that, when the polynomial $P$ is changed into $cP$ where $c$ is a 
-nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
-multiplied by $c$ as well whereas the $B_i$'s and the $V_i$'s remained 
-unchanged. Therefore, the theorem holds for $P$ if and only if it holds 
-for $cP$. As a consequence we may assume that $P$ is normalized so that 
-$\NF(P)(d) = d \lambda_0$, \emph{i.e.} $\varphi(x) = \lambda_0 x$ for $x > d$.
+For nonzero $c$ in a finite extension of $K$, note that the theorem holds
+for $P$ if and only if it holds for $cP$, since the $A_i$'s change to $cA_i$'s
+while the $B_i$'s and the $V_i$'s remained unchanged. We may thus assume
+that $P$ is normalized so that  $\NF(P)(d) = d \lambda_0$,
+\emph{i.e.} $\varphi(x) = \lambda_0 x$ for $x > d$.
 For a polynomial $Q \in K[X]$ of degree $n$, we further define:
@@
 \label{rem:slope-factor}
-It follows from the proof above that the sequence $(V_i)_{i \geq 0}$
+From the proof above, the sequence $(V_i)_{i \geq 0}$
 converges as well. Its limit $V_\infty$ is an inverse of $B_\infty
@@
 
-Let $P \in K_n[X]$ and $d$ be the abscissa of an extremal point of 
+Let $P \in K[X]$ have degree $n$ and $d$ be the abscissa of an extremal point of 
 $\NP(P)$. Previously (\emph{cf} Theorem~\ref{theo:slope-factor}), we 
@@
 $A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
-so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We shall also 
-always assume that $P$ is monic in the sense that its leading
-coefficient is \emph{exactly}~$1$; the precision datum on $P$ then 
-only concerns the coefficients up to degree $n{-}1$. Similarly, noting
-that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
-precision datum up to degree $d{-}1$ and $n{-}d{-}1$ respectively.
+so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We also assume
+the leading coefficient of $P$ is \emph{exactly}~$1$; the precision datum on $P$ then 
+only concerns the coefficients up to degree $n{-}1$. Similarly,
+$A^{(1)}$ and $B^{(1)}$ only carry
+precision up to degree $d{-}1$ and $n{-}d{-}1$ (resp).
 
@@
 precision given by Proposition~\ref{prop:Newtonprecslope}. Our strategy 
-consists in computing first the precision and applying then the Newton 
-iteration until the expected precision is reached. Below is the
-precise description of our algorithm.
+consists in computing first the precision and then applying the Newton 
+iteration until the expected precision is reached.
 
@@
 
-\noindent{\bf Input:} a monic polynomial $P + O(\varphi_P) \in K_n[X]$,
+\noindent{\bf Input:} a monic polynomial $P + O(\varphi_P) \in K[X]$ of degree $n$,
 
@@
 by using relaxed algorithms from 
-\cite{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
+\cites{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
 cases, it is in $O(\pi^{N_i})$ with $N_i = O(2^i \kappa + \min 
@@
 Remark~\ref{rem:slope-factor} and Proposition~\ref{prop:Newtonprecslope} 
-that Algorithm {\tt slope\_factorisation\_Newton} is correct and 
-stable. Using the standard soft-$O$ notation $\softO(\cdot)$ for
-hiding logarithmic factor, our algorithm performs at most $\softO(n)$ 
+that the above algorithm is correct and 
+stable. Using the standard notation $\softO(\cdot)$ for
+hiding logarithmic factors, our algorithm performs at most $\softO(n)$ 
 combinatorial operations and $\softO(n)$ operations in $K$ at precision 
@@
 \label{prop:precA1}
-The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
+The map $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
 and its differential at some point $P$ is the linear mapping
@@
 
-As for the second question, the discussion is similar to the case of 
-Newton precision expect that we need a new stopping criterion. It is given 
-by the following proposition.
+To compute $A^{(1)}$ in practice, we give a stopping criterion.
 
@@
 As a conclusion, the algorithm we propose consists in computing the 
-Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of the
-algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
+Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of
+Algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
 until we find a pair $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 

### 2016-04-30 polynomials/slopefact.tex
Merge David's corrections in slopefact.tex

@@
 following an algorithm of Montes.  Slope factorization provides a subroutine
-in such algorithms \cite[Section~2]{pauli:10a}.  For the most difficult inputs, it is not the dominant
-contributor to the runtime of the algorithm, but in some circumstances it will be.
-We underline moreover that the methods introduced in this paper extend 
-partially to the noncommutative setting and appear this way as an
+in such algorithms \cite[Section~2]{pauli:10a}.  Note that our algorithm lifts
+a \emph{slope factor} corresponding to a segment of the Newton polygon,
+rather than the irreducible factor found by single factor lifting \cite{guardia-nart-pauli:12a}.
+The slope factors are easier to find; as a consequence our iteration is less involved.
+We also remark that the methods introduced in this paper extend 
+partially to the non-commutative setting and appear this way as an
 essential building block in several decomposition algorithms of 
@@
 We recall that one can attach a Newton function to each polynomial.
-If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
+If $P(X) = \sum_{i=0}^n a_i X^i \in K_n[X]$, we define its Newton
 polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
@@
 $A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
-so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We shall also 
-always assume that $P$ is monic in the sense that its leading
-coefficient is \emph{exactly}~$1$; the precision datum of $P$ then 
-only concerns the coefficients up to degree $n{-}1$. Similarly, noting
-that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
-precision datum up to degree $d{-}1$ and $n{-}d{-}1$ respectively.
+so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We also assume
+the leading coefficient of $P$ is \emph{exactly}~$1$; the precision datum on $P$ then 
+only concerns the coefficients up to degree $n{-}1$. Similarly,
+$A^{(1)}$ and $B^{(1)}$ only carry
+precision up to degree $d{-}1$ and $n{-}d{-}1$ respectively.
 
@@
 Obviously \textbf{(H)} gives an answer to the first question we have 
-raised above: the precision of the couple $A^{(1)}$ is the lattice 
+raised above: the precision of $A^{(1)}$ is the lattice 
 $H_{A^{(1)}}$ defined as projection on the first component of 
@@
 As a conclusion, the algorithm we propose consists in computing the 
-Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of the
-algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
-until we find a couple $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
+Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of
+Algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
+until we find a pair $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
 the requirements (i) and (ii) of Proposition~\ref{prop:stop}. Once this 
-couple has been found, one may safely output $\tilde A^{(1)} + 
+pair has been found, one may safely output $\tilde A^{(1)} + 
 O\big(\big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}\big)$ under

### 2016-06-20 polynomials/slopefact.tex
Updates to paper for submission, staying within 8 pages

@@
 
-\CopyrightYear{2016}
-\setcopyright{licensedothergov}
-\conferenceinfo{ISSAC'16,}{July 19--22, 2016, Waterloo, ON, Canada} \isbn{978-1-4503-4380-0/16/07}
-\acmPrice{\$15.00}
+\setcopyright{acmlicensed}
+\conferenceinfo{ISSAC '16,}{July 19--22, 2016, Waterloo, ON, Canada}
+\isbn{978-1-4503-4380-0/16/07}\acmPrice{\$15.00}
 \doi{http://dx.doi.org/10.1145/2930889.2930897}
 
-\permission{Publication rights licensed to ACM.
-ACM acknowledges that this contribution was authored or co-authored by an employee, contractor or affiliate of a national government.
-As such, the Government retains a nonexclusive, royalty-free right to publish or reproduce this article, or to allow others to do so, for Government purposes only.}
-
-\begin{document}
+\clubpenalty=10000 
+\widowpenalty = 10000
+
+\title{Division and Slope Factorization of p-Adic Polynomials}
+
+\numberofauthors{3}
+\author{
+\alignauthor Xavier Caruso\\
+  \affaddr{Universit\'e Rennes 1}\\
+  \affaddr{xavier.caruso@normalesup.org}
+\alignauthor David Roe \\
+  \affaddr{University of Pittsburgh}\\
+  \affaddr{roed.math@gmail.com}
+\alignauthor Tristan Vaccon\\
+  \affaddr{JSPS--Rikkyo University}\\
+  \affaddr{vaccon@rikkyo.ac.jp}
+}
 
@@
 
-\title{Division and Slope Factorization of p-Adic Polynomials}
-
-\numberofauthors{3}
-\author{
-\alignauthor Xavier Caruso\\
-  \affaddr{Universit\'e Rennes 1}\\
-  \affaddr{\textsf{xavier.caruso@normalesup.org}}
-\alignauthor David Roe \\
-  \affaddr{University of Pittsburgh}\\
-  \affaddr{\textsf{roed.math@gmail.com}}
-\alignauthor Tristan Vaccon\\
-  \affaddr{JSPS--Rikkyo University}\\
-  \affaddr{\textsf{vaccon@rikkyo.ac.jp}}
-}
+\begin{document}
 
@@
 \end{CCSXML}
-
-\vspace{-1mm}
-\ccsdesc[500]{Computing methodologies~Algebraic algorithms}
-\printccsdesc
 
@@
 
-Any computation with $p$-adic fields must work with approximations modulo finite
-powers of $p$, and one of the key requirements in designing an algorithm is
-an analysis of how the precision of the variables evolve over the computation.
+Any $p$-adic computation must work with approximations modulo finite
+powers of $p$ and include an analysis of how the precision of the variables evolve over the computation.
 We work with precision models developed by the same authors
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
+
+\begingroup
+\raggedright
 
@@
 \begin{biblist}
-\tiny
+%\tiny
+\vspace*{0.5mm}
+\scriptsize
 
@@
 \end{bibdiv}
+\endgroup
 

### 2016-12-03 charpoly/charpoly.tex
Rough draft of the paper on characteristic polynomials

--- 2016-04-30 polynomials/slopefact.tex
@@
 
-\newcommand{\slp}{\text{\rm slp}}
-\newcommand{\RS}{\text{\rm RS}}
-
 \newcommand{\N}{\mathbb N}
@@
 \newcommand{\R}{\mathbb R}
-\renewcommand{\O}{\mathcal O}
-\newcommand{\OK}{\mathcal{O}_K}
-\newcommand{\XX}{\mathbf X}
-\newcommand{\trans}{{}^{\text t}}
-\newcommand{\T}{\mathcal{T}}
 
 \newcommand{\softO}{O\tilde{~}}
-
-\renewcommand{\prec}{\text{\rm prec}}
-\newcommand{\NF}{\text{\rm NF}}
-
-\renewcommand{\mod}{\,\%\,}
-\renewcommand{\div}{\,/\hspace{-0.3em}/\,}
-\newcommand{\nfop}[1]{%
-\raisebox{-0.1mm}{\begin{tikzpicture}
-\draw[transparent] (-0.18,0)--(0.18,0);
-\draw[very thin] (0,0) circle (0.11cm);
-\begin{scope}
-\clip (0,0) circle (0.11cm);
-\node[scale=0.75] at (0,0) { $#1$ };
-\end{scope}
-\end{tikzpicture}}}
-\newcommand{\nfplus}{\nfop+}
-\newcommand{\nftimes}{\nfop\times}
-\newcommand{\nfmod}{\nfop\mod}
-\newcommand{\nfdiv}{\nfop\div}
-
-\newcommand{\id}{\textrm{id}}
-\newcommand{\app}{\textrm{app}}
-\newcommand{\Epi}{\textrm{Epi}}
-\renewcommand{\c}{\text{\rm c}}
-
-\newcommand{\detp}{\det{'}}
-\newcommand{\low}{\text{\rm low}}
-\newcommand{\up}{\text{\rm up}}
-\newcommand{\DI}{\text{\rm DI}}
-\newcommand{\II}{\text{\rm II}}
-\DeclareMathOperator{\charpoly}{char}
-\newcommand{\charp}{\charpoly'}
-
-\newcommand{\lb}{\ensuremath{\llbracket}}
-\newcommand{\rb}{\ensuremath{\rrbracket}}
-\newcommand{\lp}{(\!(}
-\newcommand{\rp}{)\!)}
-\newcommand{\col}{\: : \:}
-
-\newcommand{\tinyplus}{\raisebox{0.4mm}{\hspace{0.4mm}\tiny +\hspace{0.4mm}}}
 
@@
 \CopyrightYear{2016}
-\setcopyright{licensedothergov}
-\conferenceinfo{ISSAC'16,}{July 19--22, 2016, Waterloo, ON, Canada} \isbn{978-1-4503-4380-0/16/07}
-\acmPrice{\$15.00}
-\doi{http://dx.doi.org/10.1145/2930889.2930897}
+\setcopyright{acmcopyright}
+\conferenceinfo{ISSAC '16,}{July 19-22, 2016, Waterloo, ON, Canada}
+\isbn{978-1-4503-4380-0/16/07}\acmPrice{\$15.00}
+\doi{http://dx.doi.org/10.1145/2930889.2930898}
 
@@
 
-\title{Division and Slope Factorization of p-Adic Polynomials}
+\title{Characteristic polynomial of p-adic matrices}
 
@@
   \affaddr{\textsf{xavier.caruso@normalesup.org}}
-\alignauthor David Roe \\
-  \affaddr{University of Pittsburgh}\\
-  \affaddr{\textsf{roed.math@gmail.com}}
+\alignauthor David Roe\\
+  \affaddr{Pittsburg University}\\
+  \affaddr{\textsf{roed@pitt.edu}}
 \alignauthor Tristan Vaccon\\
-  \affaddr{JSPS--Rikkyo University}\\
-  \affaddr{\textsf{vaccon@rikkyo.ac.jp}}
+  \affaddr{Universit\'e de Limoges}\\
+  \affaddr{\textsf{tristan.vaccon@unilim.fr}}
 }
@@
 \begin{abstract}
-We study two important operations on polynomials defined over complete 
-discrete valuation fields: Euclidean division and factorization. In particular,
-we design a simple and efficient algorithm for computing slope factorizations,
-based on Newton iteration.  One of its main features is that we avoid working
-with fractional exponents.
-We pay particular attention to stability, and analyze the behavior
-of the algorithm using several precision models.
 \end{abstract}
@@
 
-Polynomial factorization is a fundamental problem in computational algebra.
-The algorithms used to solve it depend on the ring of coefficients, with finite fields,
-local fields, number fields and rings of integers of particular interest to number theorists.
-In this article, we focus on a task that forms a building block for factorization
-algorithms over complete discrete valuation fields: the decomposition into factors based on the
-slopes of the Newton polygon.
+\todo{Example of $M$ and $M + 1$}
 
-The Newton polygon of a polynomial $f(X) = \sum a_i X^i$ over such a field is given by
-the convex hull of the points $(i, \val(a_i))$ and the point $(0, +\infty)$.  The lower
-boundary of this polygon consists of line segments $(x_j,y_j)$ -- $(x_{j+1},y_{j+1})$
-of slope $s_j$.  The slope factorization of $f(X)$ expresses $f(X)$ as a product
-of polynomials $g_j(X)$ with degree $x_{j+1} - x_j$ whose roots all have valuation $-s_j$.
-Our main result is a new algorithm for computing these $g_j(X)$.
+\section{Theoretical study}
 
-Polynomial factorization over local fields has seen a great deal of progress recently
-\citelist{\cite{pauli:10a} \cite{guardia-nart-pauli:12a} \cite{guardia-montes-nart:08a} \cite{montes:99a}}
-following an algorithm of Montes.  Slope factorization provides a subroutine
-in such algorithms \cite[Section~2]{pauli:10a}.  Note that our algorithm lifts
-a \emph{slope factor} corresponding to a segment of the Newton polygon,
-rather than the irreducible factor found by single factor lifting \cite{guardia-nart-pauli:12a}.
-The slope factors are easier to find; as a consequence our iteration is less involved.
-We also remark that the methods introduced in this paper extend 
-partially to the non-commutative setting and appear this way as an
-essential building block in several decomposition algorithms of 
-$p$-adic Galois representations and $p$-adic differential 
-equations~\cite{caruso:16}.
+\subsection{The theory of p-adic precision}
 
-Any computation with $p$-adic fields must work with approximations modulo finite
-powers of $p$, and one of the key requirements in designing an algorithm is
-an analysis of how the precision of the variables evolve over the computation.
-We work with precision models developed by the same authors
-\cite[Section~4.2]{caruso-roe-vaccon:14a}, focusing on the lattice and Newton models.
-As part of the analysis of the slope factorization algorithm, we describe
-how the precision of the quotient and remainder depend on the input polynomials
-in Euclidean division.
+\todo{Recall the Lemma (with integral version)}
 
-\medskip
+\subsection{Computation of the differential}
 
-\noindent
-{\bf Main Results.}
-%
-Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
-has an extremal point at abscissa $d$.
-Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
-\begin{align*}
-A_{i+1} &= A_i + (V_i P \mod A_i) \\
-B_{i+1} &= P \div A_{i+1} \\
-V_{i+1} &= (2V_i - V_i^2B_{i+1}) \mod A_{i+1}.
-\end{align*}
-where $S \div T$ and $S \mod T$ denotes the quotient and the
-remainder in the Euclidean division of $S$ by $T$ respectively.
-Our main result is Theorem \ref{theo:slope-factor},
-which states that the sequence $(A_i)$ converges quadratically
-to a divisor of $P$. This provides a quasi-optimal simple-to-implement 
-algorithm for computing slope factorizations.
-We moreover carry out a careful study of the precision and, applying
-a strategy coming from~\cite{caruso-roe-vaccon:14a}, we end up with
-an algorithm that outputs optimal results regarding to accuracy.
+\todo{Formula with comatrix.}
 
-In order to prove Theorem~\ref{theo:slope-factor}, we also determine
-the precision of the quotient and remainder in Euclidean division,
-which may be of independent interest.  These results are found in
-Section \ref{sec:prec_track}.
+\subsection{Properties of the image of $d \chi$}
 
-\medskip
+\todo{Surjectivity. Stability under multiplication by $X$ mod $\chi$.}
 
-\noindent
-{\bf Organization of the paper.}
-%
-After setting notation, in
-Section \ref{sec:prec_data} we recall various models for tracking
-precision in polynomial arithmetic.  We give some background
-on Newton polygons and explain how using lattices to store
-precision can allow for extra \emph{diffuse} $p$-adic digits that
-are not localized on any single coefficient.
+\todo{Image modulo $p$ and elementary divisors. 
+Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
 
-In Section \ref{sec:quo_rem}, we consider Euclidean division.
-We describe in Theorem \ref{theo:EDivisionNP}
-how the Newton polygons of the quotient and remainder
-depend on numerator and denominator.  We use this
-result to describe in Proposition \ref{prop:NewtonprecEuclide}
-the precision evolution in Euclidean division using
-the Newton precision model.  We then compare the
-precision performance of Euclidean division in the
-jagged, Newton and lattice models experimentally,
-finding different behavior depending on the modulus.
+\section{Computation of $\chi$ and $d\chi$}
 
-Finally, in Section \ref{sec:slope_fac} we describe
-our slope factorization algorithm, which is based on
-a Newton iteration.  Unlike other algorithms for
-slope factorization, ours does not require working
-with fractional exponents (compare for instance with
-\cite[\S~6]{poteaux-rybowicz}).
-In Theorem \ref{theo:slope-factor} we define a sequence
-of polynomials that will converge to the factors
-determined by an extremal point in the Newton
-polygon.  We then discuss the precision behavior
-of the algorithm.
+\subsection{Hessenberg form}
 
-\medskip
+\todo{Describe algorithm (with matrices and endormophisms?). 
+Remark about precision.}
 
-\noindent
-{\bf Notations.}
-Throughout this paper, we fix a complete discrete valuation field $K$; 
-we denote by $\val : K \to \Z \cup \{+\infty\}$ the valuation on it and 
-by $W$ its ring of integers (\emph{i.e.} the set of elements with 
-nonnegative valuation). We assume that $\val$ is normalized so that it
-is surjective and denote by $\pi$ a uniformizer of $K$, that is an 
-element of valuation $1$. Denoting by $S \subset W$ a fixed set of 
-representatives of the classes modulo $\pi$ and assuming $0 \in S$, 
-one can prove that each element in $x \in K$ can be represented 
-uniquely as a convergent series:
-\begin{equation}
-\label{eq:CDVFseries}
-x = \sum_{i=\val(x)}^{+\infty} a_i \pi^i
-\quad \text{with} \quad a_i \in S.
-\end{equation}
-The two most important examples are the field of $p$-adic numbers $K = 
-\Qp$ and the field of Laurent series $K = k((t))$ over a field $k$. The 
-valuation on them are the $p$-adic valuation and the usual valuation of 
-a Laurent series respectively. Their ring of integers are therefore 
-$\Zp$ and $k[[t]]$ respectively. A distinguished uniformizer is $p$ 
-and $t$ whereas a possible set $S$ is $\{0, \ldots, p-1\}$ and $k$
-respectively.
-The reader who is not familiar with complete discrete valuation fields
-may assume (without sacrifying too much to the generality) that $K$ is
-one of the two aforementioned examples.
+\subsection{Computation of the inverse}
 
-In what follows, the notation $K[X]$ refers to the ring of univariate 
-polynomials with coefficients in $K$. The subspace of polynomials of 
-degree at most $n$ (resp. exactly $n$) is denoted by $K_{\leq n}[X]$ 
-(resp. $K_n[X]$).
+\todo{Describe algorithm with SNF. Remark about precision.
+Application to the computation of $d\chi$}
 
-\section{Precision data} \label{sec:prec_data}
+\subsection{Jagged precision}
 
-Elements in $K$ (and \emph{a fortiori} in $K[X]$) carry an infinite 
-amount of information. They thus cannot be stored entirely in the
-memory of a computer and have to be truncated. Elements of $K$ are
-usually represented by truncating Eq.\eqref{eq:CDVFseries} as
-follows:
-\begin{equation}
-\label{eq:CDVFseriesO}
-x = \sum_{i=v}^{N-1} a_i \pi^i + O(\pi^N)
-\end{equation}
-where $N$ is an integer called the \emph{absolute precision} and 
-the notation $O(\pi^N)$ means that the coefficients $a_i$ for $i
-\geq N$ are discarded. If $N > v$ and $a_v \neq 0$, the integer $v$ 
-is the valuation of $x$ and the difference $N-v$ is called the
-\emph{relative precision}.
-Alternatively, one may think that the writing~\eqref{eq:CDVFseriesO}
-represents a subset of $K$ which consists of all elements in $K$ for
-which the $a_i$'s in the range $[v,N-1]$ are those specified. From the
-metric point of view, this is a ball (centered at any point inside it).
+\todo{Computation of optimal jagged precision.}
 
-It is worth noting that tracking precision using this representation is 
-rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
-relative) precision $N_x$ and $N_y$ respectively, one can compute the 
-sum $x+y$ (resp. the product $xy$) at absolute (resp. relative) 
-precision $\min(N_x,N_y)$. Computations with $p$-adic and Laurent
-series are often handled this way on symbolic computation softwares.
+\section{The case of diagonalizable matrices}
 
-\subsection{Precision for polynomials}
+\subsection{Legendre precision}
 
-The situation is much more subtle when we are working with a collection 
-of elements of $K$ (\emph{e.g.} a polynomial) and not just a single one.
-Indeed, several precision data may be considered and, as we shall see
-later, each of them has its own interest. Below we detail three models 
-of precision for the special case of polynomials.
+\todo{Describe Legendre precision and show that it is optimal for
+a diagonalizable matrix over $\Zp$.}
 
-\medskip
-
-\noindent
-{\bf Flat precision.}
-The simplest method for tracking the precision of a polynomial is
-to record each coefficient modulo a fixed power of $\pi$.  While
-easy to analyze and implement, this method suffers when
-applied to polynomials whose Newton polygons are far from flat.
-
-\medskip
-
-\noindent
-{\bf Jagged precision.}
-The next obvious approach is to record the precision of each
-coefficient individually, a method that we will refer to as \emph{jagged}
-precision.  Jagged precision is commonly implemented
-in computer algebra systems, since standard polynomial algorithms
-can be written for generic coefficient rings.  However, these
-generic implementations often have suboptimal precision behavior,
-since combining intermediate expressions into a final answer
-may lose precision.  Moreover, when compared to the Newton
-precision model, extra precision in the middle coefficients, above
-the Newton polygon of the remaining terms, will have no effect
-on any of the values of that polynomial.
-
-\medskip
-
-\noindent
-{\bf Newton precision.} 
-We now move to \emph{Newton precision} data. They can be actually seen 
-as particular instances of jagged precision but there exist for them 
-better representations and better algorithms.
-
-\begin{deftn}
-A \emph{Newton function of degree $n$} is a convex function 
-$\varphi : [0,n] \to \R \cup \{+\infty\}$ which is piecewise affine, 
-which takes a finite value at $n$ and whose epigraph\footnote{Recall
-that the epigraph is the region above the graph.} $\Epi(\varphi)$ 
-have extremal points with integral abscissa.
-\end{deftn}
-
-\begin{rem}
-The datum of $\varphi$ is equivalent to that of $\Epi(\varphi)$ and they 
-can easily be represented and manipulated on a computer.
-\end{rem}
-
-We recall that one can attach a Newton function to each polynomial.
-If $P(X) = \sum_{i=0}^n a_i X^i \in K_n[X]$, we define its Newton
-polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
-($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
-and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
-function whose epigraph is $\NP(P)$. It is well known \cite[Section~1.6]{dwork-geratto-sullivan:Gfunctions} that:
-$$\begin{array}{r@{\,\,}c@{\,\,}l}
-\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
-\NP(PQ) & = & \NP(P) + \NP(Q)
-\end{array}$$
-where $\text{Conv}$ denotes the convex hull and the plus sign stands 
-for the Minkowski sum. This translates to:
-$$\begin{array}{r@{\,\,}c@{\,\,}l}
-\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
-\NF(PQ) = \NF(P) \nftimes \NF(Q)
-\end{array}$$
-where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
-There exist classical algorithms for computing these two operations
-whose complexity is quasi-linear with respect to the degree.
-
-In a similar fashion, Newton functions can be used to model precision: 
-given a Newton function $\varphi$ of degree $n$, we agree that a polynomial 
-of degree at most $n$ is given at precision $O(\varphi)$ when, for all $i$,
-its $i$-th coefficient is given at precision $O\big(\pi^{\lceil \varphi(i)
-\rceil}\big)$ (where $\lceil \cdot \rceil$ is the ceiling function).
-In the sequel, we shall write
-$O(\varphi) = \sum_{i=0}^n O\big(\pi^{\lceil \varphi(i) \rceil}\big) \cdot X^i$
-and use the notation $\sum_{i=0}^n a_i X^i + O(\varphi)$ (where the
-coefficients $a_i$ are given by truncated series) to refer to a 
-polynomial given at precision $O(\varphi)$.
-
-It is easily checked that if $P$ and $Q$ are two polynomials known at 
-precision $O(\varphi_P)$ and $O(\varphi_Q)$ respectively, then $P+Q$ is 
-known at precision $O(\varphi_P \nfplus \varphi_Q)$ and $PQ$ is known at 
-precision $O\big((\varphi_P \nftimes \NF(Q)) \nfplus (\NF(P) \nftimes 
-\varphi_Q)\big)$.
-
-\begin{deftn}
-\label{def:nondeg}
-Let $P = P_\app + O(\varphi_P)$. We say that the Newton precision 
-$O(\varphi_P)$ of $P$ is \emph{nondegenerate} if $\varphi_P \geq 
-\NF(P_\app)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
-$\NP(P_\app)$.
-\end{deftn}
-
-We notice that, under the conditions of the above definition, the
-Newton polygon of $P$ is well defined. Indeed, if $\delta P$ is any
-polynomial whose Newton function is not less than $\varphi_P$, we
-have $\NP(P_\app + \delta P) = \NP(P_\app)$.
-
-\medskip
-
-\noindent
-{\bf Lattice precision.}
-The notion of \emph{lattice precision} was developed in 
-\cite{caruso-roe-vaccon:14a}. It encompasses the two previous models and has
-the decisive advantage of precision optimality.
- As a counterpart, it might be very space-consuming and time-consuming 
- for polynomials of large degree.
-
-\begin{deftn}
-Let $V$ be a finite dimensional vector space over $K$. A lattice
-in $V$ is a sub-$W$-module of $V$ generated by a $K$-basis of
-$V$.
-\end{deftn}
-
-\noindent
-We fix an integer $n$. A lattice precision datum for a polynomial of 
-degree $n$ is a lattice $H$ lying in the vector space $K_{\leq n}[X]$. 
-We shall sometimes denote it $O(H)$ in order to emphasize that it should 
-be considered as a precision datum. The notation $P_\app(X) + O(H)$ then 
-refers to any polynomial in the $W$-affine space $P_\app(X) + H$. 
-Tracking lattice precision can be done using differentials as shown in
-\cite[Lemma~3.4 and Proposition~3.12]{caruso-roe-vaccon:14a}: if $f : K_{\leq n}[X] 
-\to K_{\leq m}[X]$ denotes any strictly differentiable function with
-surjective differential, under mild assumption on $H$, we have:
-$$f(P_\app(X)+H) = f(P_\app(X)) + f'(P_\app(X))(H)$$
-where $f'(P_\app(X))$ denotes the differential of $f$ at $P_\app(X)$. 
-The equality sign reflets the optimality of the method.
-
-As already mentioned, the jagged precision model is a particular case of 
-the lattice precision. Indeed, a precision of the shape $\sum_{i=0}^n 
-O(\pi^{N_i}) X^i$ corresponds to the lattice generated by the elements 
-$\pi^{N_i} X^i$ ($0 \leq i \leq n$). This remark is the origin of the 
-notion of \emph{diffused digits of precision} introduced in 
-\cite[Definition~2.3]{caruso-roe-vaccon:15a}. We shall use it repeatedly in the 
-sequel in order to compare the behaviour of the three aforementioned 
-precision data in concrete situations.
-
-\section{Euclidean division} \label{sec:quo_rem}
-
-Euclidean division provides a building block for many algorithms associated
-to polynomials in one variable.  In order to analyze the precision behavior
-of such algorithms, we need to first understand the precision attached to
-the quotient and remainder when dividing two polynomials.
-In the sequel, we use the notation $A \div B$ and $A \mod B$
-for the polynomials satisfying $A = (A \div B) \cdot B + (A \mod B)$
-and $\deg(A \mod B) < \deg(B)$.
-
-\subsection{Euclidean division of Newton functions}
-
-\begin{deftn}
-Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
-$d$ respectively. Set $\lambda = \psi(d) - \psi(d-1)$. Letting $\Delta$
-be the greatest affine function of slope $\lambda$ with $\Delta \leq
-\varphi_{|[d,n]}$ and $\delta = \Delta(d) - \psi(d)$, we define:
-$$\begin{array}{r@{\,\,}rcl}
-\varphi \nfmod \psi & 
-\multicolumn{3}{@{}l}{= \varphi_{|[0,d{-}1]} \nfplus \big( 
-\psi_{|[0,d{-}1]} + \delta\big)} \medskip \\
-\varphi \nfdiv \psi & : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
-& x & \mapsto & \inf_{h \geq 0} \varphi(x+d+h) - \lambda h.
-\end{array}$$
-\end{deftn}
-
-\begin{figure}
-\hfill
-\begin{tikzpicture}[xscale=0.7, yscale=0.5]
-\fill[blue!30] (4,9.5)--(4,2.667)--(2,2)--(0.5,4)--(0,8)--(0,9.5)--cycle;
-\fill[yellow!50] (5,9.5)--(5,3)--(4,2.667)--(4,9.5)--cycle;
-\fill[green!30] (5,9.5)--(5,3)--(8,4)--(11,8)--(11,9.5)--cycle;
-\draw[->] (-0.5,0)--(11.5,0);
-\draw[->] (0,-0.5)--(0,9.5);
-\draw[very thick] (0,6)--(0.5,2)--(2,0)--(5,1);
-\draw[very thick] (0,9)--(3,5)--(8,4)--(11,8);
-\draw[dashed] (11,5)--(8,4);
-\draw (8,4)--(2,2)--(0.5,4)--(0,8);
-\draw[dotted,thick] (5,0)--(5,9.5);
-\draw[dotted,thick] (4,0)--(4,9.5);
-\node[scale=0.8, below left] at (0,0) { $0$ };
-\node[scale=0.8, below] at (5,0) { $d$ };
-\node[scale=0.8, below] at (4,0) { $d{-}1$ };
-\node[scale=0.9, below right] at (2.5,1.2) { $\psi$ };
-\node[scale=0.9, right] at (9.8,7.3) { $\varphi$ };
-\node[scale=0.9, right] at (10.5,4.4) { $\Delta$ };
-\end{tikzpicture}
-\hfill \null
-
-\vspace{-6mm}
-
-\caption{Euclidean division of Newton functions}
-\label{fig:NewtonEuclide}
-\end{figure}
-
-Figure~\ref{fig:NewtonEuclide} illustrates the definition: if $\varphi$ 
-and $\psi$ are the functions represented on the diagram, the epigraph of 
-$\varphi \nfmod \psi$ is the blue area whereas that of $\varphi \nfdiv \psi$ 
-is the green area translated by $(-d,0)$.
-It is an easy exercise (left to the reader) to design quasi-linear
-algorithms for computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$.
-
-\begin{theo}
-\label{theo:EDivisionNP}
-Given $A, B \in K[X]$ with $B \neq 0$, we have:
-\begin{align}
-\NF(A \mod B) & \geq \NF(A) \nfmod \NF(B) \label{eq:AmodB} \\
-\text{and} \hspace{4mm}
-\NF(A \div B) & \geq \NF(A) \nfdiv \NF(B) \label{eq:AdivB}
-\end{align}
-\end{theo}
-
-\begin{proof}
-Write $A = A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$)
-consists of monomials of $A$ of degree less than $d$ (resp. at least
-$d$). Noting that:
-$$A \mod B = A_{<d} + (A_{\geq d} \mod B)
-\quad \text{and} \quad
-A \div B = A_{\geq d} \div B$$
-we may assume that $A = A_{\geq d}$.
-
-Let us now prove Eq.~\eqref{eq:AmodB}. 
-Replacing $B$ by $c^{-1} B$ where $c$ denotes the leading coefficient
-of $B$, we may assume that $B$ is monic. Using linearity, we may further
-assume that $A$ is a monomial. Set $R_n = X^n \mod B$. The relation
-we have to prove is:
-$$\NF(R_n)(x) \geq \NF(B)(x) - \lambda(n-d)
-\quad \text{for } n\geq d.$$
-We proceed by induction. The initialisation is clear because $R_d$
-agrees with $(-B)$ up to degree $d{-}1$. We have the relation
-$R_{n+1} = X R_n - c_n B$
-where $c_n$ is the coefficient in $X^{d-1}$ of $R_n$. Thanks to the
-induction hypothesis, we have:
-\begin{align*}
-\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \lambda(n{-}d) \\
-& = - \lambda(n{+}1{-}d)
-\end{align*}
-since $\lambda = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
-$\NF(c_n B)(x) \geq \NF(B)(x) - \lambda(n{+}1{-}d)$ for all $x$. On the 
-other hand, for all $x$, we have:
-$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \lambda$$
-from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \lambda(n{+}1{-}d)$. As a
-consequence
-$\NF(R_{n+1})(x) \geq \NF(B)(x) - \lambda(n{+}1{-}d)$
-and the induction follows.
-Eq.~\eqref{eq:AdivB} is now derived from:
-$$\NF(A \div B) \nftimes \NF(B) \geq \NF(A) \nfplus \NF(A \mod B)$$
-using the estimation on $\NF(A\mod B)$ we have just proved
-(see Figure~\ref{fig:NewtonEuclide}).
-\end{proof}
-
-\subsection{Tracking precision} \label{sec:prec_track}
-
-\medskip
-
-\noindent
-{\bf Newton precision.}
-We first analyze the precision behavior of Euclidean division
-in the Newton model.  Concretely, we pick $A, B \in K[X]$ two polynomials which 
-are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
-$$A = A_\app + O(\varphi_A)
-\quad \text{and} \quad
-B = B_\app + O(\varphi_B)$$
-Here $A_\app$ and $B_\app$ are some approximations of $A$ and $B$ respectively 
-and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
-$\deg A$ and $\deg B$ respectively.
-We are interested in determining the precision of $A \mod B$ and $A
-\div B$. The following proposition gives a theoretical answer under
-mild assumptions.
-
-\begin{prop}
-\label{prop:NewtonprecEuclide}
-We keep the above notations and assume that the Newton precisions
-$O(\varphi_A)$ and $O(\varphi_B)$ on $A$ and $B$ respectively are 
-both nondegenerate (\emph{cf} Definition~\ref{def:nondeg}). Then,
-setting:
-$$\varphi = \varphi_A \nfplus 
-\big[ \varphi_B \nftimes \big(\NF(A) \nfdiv \NF(B)\big) \big]$$
-the polynomials $A \div B$ and $A \mod B$ are known at precision 
-$O(\varphi \nfdiv \NF(B))$ and $O(\varphi \nfmod \NF(B))$ respectively.
-\end{prop}
-
-\begin{proof}
-Let $\delta A$ (resp. $\delta B$) be a polynomial whose Newton 
-function is not less than $\varphi_A$ (resp. $\varphi_B$) and define
-$\delta Q$ and $\delta R$ by:
-\begin{align*}
-Q_\app + \delta Q & = (A_\app + \delta A) \div (B_\app + \delta B) \\
-R_\app + \delta R & = (A_\app + \delta A) \mod (B_\app + \delta B)
-\end{align*}
-where $Q_\app = A_\app \div B_\app$ and $R_\app = A_\app \mod B_\app$.
-We have to show that $\NF(\delta Q) \geq \varphi \nfdiv \NF(B)$ and
-$\NF(\delta R) \geq \varphi \nfmod \NF(B)$.
-Set $\delta X = \delta A - Q_\app \delta B$. 
-Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_\app) \geq
-\NF(A) \nfdiv \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
-On the other hand, an easy computation yields
-$$\delta X = (B + \delta B) \cdot \delta Q + \delta R$$
-so that $\delta Q = \delta X \div (B + \delta B)$ and 
-$\delta R = \delta X \mod (B + \delta B)$. Using again 
-Theorem~\ref{theo:EDivisionNP}, we get the desired result.
-\end{proof}
-
-With this result in hand, we may split the computation of
-Euclidean division into two pieces, first computing approximations
-$Q_\app$ and $R_\app$ and separately computing $\delta Q$ and $\delta R$.
-Both the approximations and the precision can be computing in time that is
-quasi-linear in the degree. 
-%\todo{Is this sufficiently precision complexity analysis?}
-
-\medskip
-
-\noindent
-{\bf Lattice precision.}
-We now move to lattice precision. We pick $A$ and $B$ two polynomials 
-of respective degree $n$ and $d$ and assume that they are known at 
-precision $O(H_A)$ and $O(H_B)$ respectively:
-$$A = A_\app + O(H_A)
-\quad \text{and} \quad
-B = B_\app + O(H_B).$$
-where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
-According to the results of \cite{caruso-roe-vaccon:14a}, in order to determine the 
-precision of $A \div B$ and $A \mod B$, we need to compute the 
-differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
-\mapsto X \mod Y$ at the point $(A_\app, B_\app)$. Writing $Q_\app = A_\app \div
-B_\app$ and $R_\app = A_\app \mod B_\app$, this can be done by
-expanding the relation:
-$$A_\app + dA = (B_\app + dB) (Q_\app + dQ) + (R_\app + dR)$$
-and neglecting the terms of order $\geq 2$. We get this way 
-$dA = B_\app dQ + Q_\app dB + dR$
-meaning that $dQ$ and $dR$ appears respectively as the quotient and
-the remainder of the Euclidean division of $dX = dA - Q_\app dB$ by $B_\app$.
-
-Once this has been done, the strategy is quite similar to that explained 
-for Newton precision: compute approximations and precision lattices
-separately for quotient and remainder.
-
-\subsection{An example: modular multiplication}
-
-For this example, we work over $W = \Z_2$ and fix a monic polynomial $M 
-\in \Z[X]$ (known exactly) of degree $5$. Our aim is to compare the 
-numerical stability of the multiplication in the quotient $\Z_2[X]/M$ 
-depending on the precision model we are using. In order to do so, we 
-pick $n$ random polynomials $P_1, \ldots, P_n$ in $\Z_2[X]/M(X)$ 
-(according to the Haar measure) whose coefficients are all known at 
-precision $O(2^N)$ for some large integer $N$. We then compute the 
-product of the $P_i$'s using the following quite naive algorithm.
-
-\noindent\hrulefill
-
-%\noindent {\bf Algorithm 1:} {\tt example\_product}
-%
-%\smallskip
-
-\noindent 1.\ {\bf set} $P = 1$
-
-\noindent 2.\ {\bf for} $i=1,\dots,n$ {\bf do} {\bf compute} $P = (P 
-\cdot P_i) \mod M$
-
-\noindent 3.\ {\bf return} $P$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\medskip
-
-The table of Figure~\ref{fig:modularmult} reports the average gain of 
-\emph{absolute} precision $G$ which is observed while executing the 
-algorithm above for various modulus and $n$. The average is taken on a 
-sample of $1000$ random inputs. We recall that $G$ is defined as 
-follows:
-
-\noindent $\bullet$
-in the case of jagged and Newton precision, the precision of the output 
-may be written into the form $\sum_{i=0}^4 O(2^{N_i}) \: X^i$ and 
-$G = \sum_{i=0}^4 (N_i - N)$;
-
-\noindent $\bullet$
-in the case of lattice precision, the precision of the output is a 
-lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
-$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
-$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
-index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
-generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
-(The term $G_d$ corresponds to diffused digits according to 
-\cite[Definition~2.3]{caruso-roe-vaccon:15a}.)
-
-\begin{figure}
-{\small%
-\noindent \hfill%
-\renewcommand{\arraystretch}{1.2}%
-\begin{tabular}{|c|c|c|c|c|}%
-\hline
-\multirow{3}{*}{Modulus $M$} & 
-\multirow{3}{*}{\hspace{2mm}$n$\hspace{2mm}} & 
-\multicolumn{3}{c|}{Gain of precision} \\
-\cline{3-5}
-& & \multirow{2}{*}{Jagged} & \multirow{2}{*}{Newton}
-      & \raisebox{-0.5mm}{Lattice} \\
-& & & & \raisebox{0.5mm}{\scriptsize (not dif.\tinyplus dif.)} \\
-\hline
-\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + X^2 + 1$ \\ {\scriptsize (Irred. mod $2$)} \end{tabular}}
-& $10$ & $\phantom{00}0.2$ & $\phantom{00}0.2$ & $\phantom{00}0.2\tinyplus\phantom{00}0.0$ \\
-& $50$ & $\phantom{00}4.2$ & $\phantom{00}4.2$ & $\phantom{00}4.2\tinyplus\phantom{00}0.0$ \\
-& $100$ & $\phantom{0}11.2$ & $\phantom{0}11.2$ & $\phantom{0}11.2\tinyplus\phantom{00}0.0$ \\
-\hline
-\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + 1$ \\ {\scriptsize (Sep. mod $2$)} \end{tabular}}
-& $10$ & $\phantom{00}0.4$ & $\phantom{00}0.4$ & $\phantom{00}0.9\tinyplus\phantom{00}6.0$ \\
-& $50$ & $\phantom{00}5.6$ & $\phantom{00}5.6$ & $\phantom{0}11.1\tinyplus\phantom{0}42.0$ \\
-& $100$ & $\phantom{0}13.6$ & $\phantom{0}13.6$ & $\phantom{0}27.0\tinyplus\phantom{0}87.0$ \\
-\hline
-\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + 2$ \\ {\scriptsize (Eisenstein)} \end{tabular}}
-& $10$ & $\phantom{00}6.2$ & $\phantom{00}6.2$ & $\phantom{00}6.2\tinyplus\phantom{00}0.0$ \\
-& $50$ & $\phantom{0}44.0$ & $\phantom{0}44.0$ & $\phantom{0}44.0\tinyplus\phantom{00}0.0$ \\
-& $100$ & $\phantom{0}92.5$ & $\phantom{0}92.5$ & $\phantom{0}92.5\tinyplus\phantom{00}0.0$ \\
-\hline
-\multirow{3}{*}{\begin{tabular}{@{}c@{}} $(X+1)^5 + 2$ \\ {\scriptsize (Shift Eisenstein)} \end{tabular}}
-& $10$ & $\phantom{00}0.6$ & $\phantom{00}0.6$ & $\phantom{00}4.7\tinyplus\phantom{00}1.4$ \\
-& $50$ & $\phantom{00}7.1$ & $\phantom{00}7.1$ & $\phantom{0}42.6\tinyplus\phantom{00}1.4$ \\
-& $100$ & $\phantom{0}15.1$ & $\phantom{0}15.1$ & $\phantom{0}91.8\tinyplus\phantom{00}1.4$ \\
-\hline
-\multirow{3}{*}{\begin{tabular}{@{}c@{}} $X^5 + X + 2$ \\ {\scriptsize (Two slopes)} \end{tabular}}
-& $10$ & $\phantom{00}1.7$ & $\phantom{00}1.7$ & $\phantom{00}7.9\tinyplus\phantom{00}9.8$ \\
-& $50$ & $\phantom{00}8.1$ & $\phantom{00}8.1$ & $\phantom{0}70.7\tinyplus\phantom{0}59.8$ \\
-& $100$ & $\phantom{0}16.1$ & $\phantom{0}16.1$ & $152.6\tinyplus125.9$ \\
-\hline
-\end{tabular}
-\hfill \null}
-
-\caption{Precision for modular multiplication}
-\label{fig:modularmult}
-\end{figure}
-
-We observe several interesting properties. First of all, the gains for 
-Newton precision and jagged precision always agree though one may have 
-thought at first that Newton precision is weaker. Since performing
-precision computations in the Newton framework is cheaper, it seems
-(at least on this example) that using the jagged precision model is not
-relevant.
-
-On the other hand, the lattice precision may end up with better results. 
-Nevertheless this strongly depends on the modulus $M$. For instance, 
-when $M$ is irreducible modulo $p=2$ or Eiseistein, there is apparently 
-no benefit to using the lattice precision model. We emphasize that 
-these two particular cases correspond to modulus that are usually used 
-to define (unramified and totally ramified respectively) extensions of 
-$\Q_2$.
-
-For other moduli, the situation is quite different and the benefit of 
-using the lattice precision model becomes more apparent. 
-The comparison between the gain of precision in the jagged model and 
-the number of not diffused digits in the lattice model makes sense:
-indeed the latter appears as a theoretical upper bound of the former
-and the difference between them quantifies the quality of the way we
-track precision in the jagged (or the Newton) precision model. We 
-observe that this difference is usually not negligible (\emph{cf} 
-notably the case of $M(X) = (X+1)^5 + 2$) meaning that this quality 
-is not very good in general.
-As for diffused digits, they correspond to digits that cannot be 
-``seen'' in the jagged precision model. Their number then measures the 
-intrinsic limitations of this model. We observe that it can be very 
-important as well in several cases.
-
-The modulus $(X+1)^5 + 2$ shows the advantage of working
-with lattice precision in intermediate computations.  Indeed,
-the precision behavior using the lattice model closely parallels
-that of $X^5+2$, since the lattices are related by a change of
-variables.  But this structure is not detected in the Newton
-or jagged models.
-
-\section{Slope factorization} \label{sec:slope_fac}
-
-A well-known theorem \cite[Theorem~6.1]{dwork-geratto-sullivan:Gfunctions} asserts that each 
-extremal point $M$ in the Newton polygon $\NP(P)$ of a polynomial $P \in 
-K[X]$ corresponds to a factorization $P = AB$ where the Newton polygon 
-of $A$ (resp. $B$) is given by the part of $\NP(P)$ located at the left 
-(resp. the right) of $M$. Such a factorization is often called a 
-\emph{slope factorization}.
-
-The aim of this section is to design efficient and stable algorithms 
-for computing these factorizations. Precisely the algorithm we obtain 
-has a quasi-optimal complexity (compared to the size of the input 
-polynomial) and outputs a result whose precision is (close to be) 
-optimal. Two of its important additional features are simplicity and 
-flexibility.
-
-\subsection{A Newton iteration}
-\label{ssec:Newtoniter}
-
-The factor $A$ defined above is usually obtained \emph{via} a Newton 
-iteration after having prepared our polynomial by flattening the first 
-slope using a change of variables involving possibly rational 
-exponents. We introduce here a variant of this iteration which does not 
-require the flattening step and is entirely defined over $K[X]$.
-
-\begin{theo} \label{theo:slope-factor}
-Let $P(X) = \sum_{i=0}^n a_i X^i$ be a polynomial of degree $n$ with
-coefficients in $K$. We assume that $\NP(P)$ has an extremal point 
-whose abscissa is $d$.
-We define the sequences $(A_i)_{i \geq 0}$ and $(V_i)_{i \geq 0}$
-recursively by:
-\begin{align*}
-A_0 & = \sum_{i=0}^d a_i X^i, \quad V_{0} = 1 \medskip \\
-A_{i+1} &= A_i + (V_i P \:\mod\: A_i), \smallskip \\
-V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\
-& \hspace{2cm}\text{where } B_{i+1} = P \div A_{i+1}.
-\end{align*}
-Then the sequence $(A_i)$ converges to a divisor $A_\infty$ of $P$ 
-of degree $d$ whose leading coefficient is $a_d$ and whose Newton 
-function agrees with $\NF(P)$ on $[0,d]$.
-Moreover, setting:
-$$\kappa = \NF(P)(d{+}1) + \NF(P)(d{-}1) - 2 \cdot \NF(P)(d)$$
-(with $\NF(P)(-1) = \NF(P)(n{+}1) = +\infty$ if necessary),
-we have $\kappa > 0$ and the following rate of convergence:
-\begin{equation}
-\label{eq:rateconv}
-\forall i \geq 0, \quad
-\NF(A_\infty - A_i) \geq \NF(P)_{|[0,d{-}1]} \,+\, 2^i \kappa.
-\end{equation}
-\end{theo}
-
-\begin{rem}
-The iteration formulae of the above theorem already appeared 
-in \cite[\S~3.1]{berthomieu:hal-00642075}; nevertheless, as far
-as we know, the result of Theorem~\ref{theo:slope-factor} is
-new.
-\end{rem}
-
-%\begin{rem}
-%Using $P \mapsto P^*=X^{deg(P)}P \left( \frac{1}{X} \right), $ the assumptions in Theorem \ref{theo:slope-factor} do not induce any loss in generality.
-%\end{rem}
-
-%\begin{rem}
-%Non necessity to have $A_0$ as the first slope. It is possible to use minorations.
-%
-%\begin{tikzpicture}
-%\draw[->] (0,-1) -- (0,2);
-%\draw[->] (-0.05,0) -- (4,0);
-%\node[left] at (0,0) {$0$};
-%\draw (1.5,-.1) -- (1.5,.1);
-%\node[above] at (1.5,0) {$d$};
-%\draw (0,1)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
-%\draw[green] (0,1.5)-- (1,0.33)-- (1.5,0) --(2,-.125)-- (3,0)-- (3.5,1);
-%\node[below] at (.75,-0.2) {$\lambda_0$};
-%\end{tikzpicture}
-%\end{rem}
-
-\begin{rem}
-\label{rem:unicity}
-The divisor $A$ is uniquely determined by the conditions of 
-Theorem~\ref{theo:slope-factor}. Indeed, consider two divisors
-$A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
-Then $L = \lcm(A,A')$ is a divisor of $P$ as well and the 
-slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
-\NF(P)(d{-}1)$. Therefore $\deg L = d$ and $L$ differs from $A$ and
-$A'$ by a multiplicative nonzero constant. Then, if $A$ and $A'$ 
-share in addition the same leading coefficient, they must coincide.
-\end{rem}
-
-The rest of this subsection is devoted to the proof of the theorem.
-If $d = n$ (resp. $d = 0$), the sequence $A_i$ is constant equal to
-$P$ (resp. to the constant coefficient of $P$) and theorem is clear.
-We then assume $0 < d < n$. We set:
-\begin{align*}
-\lambda_0 & = \NF(P)(d) - \NF(P)(d{-}1) \\
-\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d),
-\end{align*}
-so that $\kappa = \lambda_1 - \lambda_0$.
-The existence of an extremal point of $\NP(P)$ located at abscissa
-$d$ ensures that $\lambda_1 > \lambda_0$, \emph{i.e.} $\kappa > 0$.
-For all indices $i$, we define:
-$$\begin{array}{r@{\,\,}lr@{\,\,}l}
-Q_i &= V_i P \div A_i, &
-R_i & = V_i P \mod A_i = A_{i+1} - A_i, 
-\smallskip \\
-S_i &= P \mod A_i, &
-T_i &= (1 - V_i B_i) \mod A_i
-\end{array}$$
-and when $\square$ is some letter, we put $\Delta \square_i = \square_{i+1} 
-- \square_i$.
-%We use the next Lemma to record some useful formulae.
-
-\begin{lem} \label{lem:formulae}
-The following relations hold:
-\begin{align}
-\Delta B_i &= -(R_i B_{i+1}) \div A_i, \label{eqdef:Biminus} \\
-\Delta S_i &= -(R_i B_{i+1}) \mod A_i,  \label{eqdef:Siminus} \\
-S_{i} &= (B_i R_i + T_i S_{i-1} + T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
-\Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
-1-Q_i &= T_i - (V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
-R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
-T_{i+1} &= (T_i + V_i \: \Delta B_i)^2 \mod{A_{i+1}}. \label{eq:ViBiplus}
-\end{align}
-\end{lem}
-
-\begin{proof}
-From $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1}$, we get
-$- R_i B_{i+1}= \Delta B_i \cdot A_i + \Delta S_i$.
-Hence, by consideration of degree, we obtain \eqref{eqdef:Biminus} 
-and \eqref{eqdef:Siminus}.
-On the other hand, from $V_i P = A_i Q_i + R_i=V_i (A_i B_i + S_i)$, we 
-derive
-\begin{equation}
-\label{eq:ViBiQi} 
-(V_i B_i -Q_i) \cdot A_i=R_i-V_i S_i. 
-\end{equation} 
-Thus $R_i=V_i S_i \mod A_i$.
-Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
-\mod A_i$, from which \eqref{eqdef:Si2} follows directly 
-%\todo{(there is a sign error here)}. (Tristan:fixed)
-By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
-We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi} and
-noting that $\deg R_i < \deg A_i = d$, we 
-get \eqref{eqdef:Qiminus}.
-
-We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i$ and $V_{i+1} P=A_{i+1} 
-Q_{i+1}+R_{i+1}$. Thus:
-\begin{align*}
-R_{i+1} &= \Delta V_i \: P + (1-Q_i)R_i \\
- &= (\Delta V_i \: S_{i+1} + (1-Q_i)R_i) \mod A_{i+1},
-\end{align*} 
-and \eqref{eq:Riplus} is proved. Finally
-\begin{align*}
-T_{i+1} &\equiv 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \pmod {A_{i+1}} \\
-&\equiv (1-V_i B_{i+1})^2 \pmod {A_{i+1}} \\
-&\equiv (T_i + V_i \: \Delta B_i)^2 \pmod {A_{i+1}}
-\end{align*}
-which concludes the proof.
-\end{proof}
-
-If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
-is constant and the theorem follows, so assume that $\lambda_0$
-and $\lambda_1$ are finite.
-
-We define the function $\varphi : \R^+ \to \R 
-\cup \{+\infty\}$ by:
-\begin{equation} \label{eqdef:functionphi}
-\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
-\lambda_0 (x - d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
-%\begin{array}{r@{\,\,}ll}
-%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
-%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
-%\end{array}
-\end{equation}
-For nonzero $c$ in a finite extension of $K$, note that the theorem holds
-for $P$ if and only if it holds for $cP$, since the $A_i$'s change to $cA_i$'s
-while the $B_i$'s and the $V_i$'s remained unchanged. We may thus assume
-that $P$ is normalized so that  $\NF(P)(d) = d \lambda_0$,
-\emph{i.e.} $\varphi(x) = \lambda_0 x$ for $x > d$.
-For a polynomial $Q \in K[X]$ of degree $n$, we further define:
-\begin{align}
-b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) 
-\label{eqdef:bphi} \\
-\text{and} \quad
-b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
-\quad \text{for } i \in \{0,1\}.
-\label{eqdef:b0}
-\end{align}
-Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
-With the normalization of $P$ we chose above, we have $b_0(P)
-= b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
-$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.
-
-\begin{lem}
-\label{lem:bphib0}
-Let $b \in \{b_\varphi, b_0, b_1\}$. For $Q_1, Q_2 \in K[X]$:
-
-\smallskip
-
-a) $b(Q_1+Q_2) \geq \min \big(b(Q_1), b(Q_2)\big)$
-
-\smallskip
-
-b) $b(Q_1Q_2) \geq \min b(Q_1) + b(Q_2)$
-
-\smallskip
-
-c) $b_\varphi(Q_1 Q_2) \geq b_\varphi(Q_1) + b_0(Q_2)$
-
-\smallskip
-
-\noindent
-For $Q, A \in K[X]$ with $\deg A = d$ and $\NF(A) =
-\NF(P)_{|[0,d]}$:
-
-\smallskip
-
-d) $b(Q \mod A) \geq b(Q)$
-
-\smallskip
-
-e) $b_0(Q \div A) \geq b_\varphi(Q)$.
-\end{lem}
-
-\begin{proof}
-\emph{a)} is clear.
-
-We skip the proof of \emph{b)} which is similar to that of \emph{c)}.
-
-Let $t_1$ (resp. $t_2$) be the translation of vector $(0, b_\varphi
-(Q_1))$ (resp. $(0, b_0(Q_2))$. It follows from the definition of
-$b_\varphi$ that $\NP(Q_1)$ is a subset of $t_1(\Epi(\varphi))$ where
-$\Epi(\varphi)$ denotes the epigraph of $\varphi$. Similarly $\NP(Q_2)
-\subset t_2(C)$ where $C$ is the convex cone generated by the vectors 
-starting from $(0,0)$ to $(0,1)$ and $(1, \lambda_0)$. Thus
-$$\NP(Q_1 Q_2) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
-t_2 \circ t_1\big(\Epi(\varphi)\big)$$
-and \emph{c)} follows.
-
-Finally \emph{d)} and \emph{e)} follows from
-Theorem~\ref{theo:EDivisionNP}.
-\end{proof}
-
-We are now going to prove by induction on $i$ the conjonction of all
-equalities and inequalities below:
-\begin{equation}
-\label{eq:induction}
-\begin{array}{l}
-\NF(A_i) = \varphi_{|[0,d]}, \smallskip \\
-b_1(V_i) \geq 0, \quad
-b_\varphi(R_i) \geq 2^i \kappa \smallskip \\
-b_\varphi(S_i) \geq 0, \quad
-b_0(T_i) \geq 2^i \kappa.
-\end{array}
-\end{equation}
-
-Noting that $A_0$ and $P$ agree up to degree $d$ and that $\NP(P)$ has 
-an extremal point at abscissa $d$, we get $\NF(A_0) = \varphi_{|[0,d]}$.
-Clearly $b_1(V_0) \geq 0$ since $V_0 = 1$. It follows from the 
-definitions that $R_0 = S_0 = P \mod A_0 = (P{-}A_0) \mod A_0$. 
-We remark that $P{-}A_0 = \sum_{i=d+1}^n a_i X^i.$
-Using Theorem \ref{theo:EDivisionNP}, we obtain that
-$b_\varphi(P{-}A_0) \geq \kappa$ and then $b_\varphi(R_0) = b_\varphi(S_0)
-\geq \kappa$ (see Figure~\ref{fig:boundR0S0}).
-%
-\begin{figure}
-\hfill
-\begin{tikzpicture}[xscale=0.7, yscale=0.5]
-\fill[blue!30] (4,9.5)--(4,1.6)--(2,1.2)--(0.5,2.8)--(0,6.8)--(0,9.5)--cycle;
-\fill[yellow!50] (6,9.5)--(6,2)--(4,1.6)--(4,9.5)--cycle;
-\fill[green!30] (6,9.5)--(6,2)--(8,4)--(11,9)--(11,9.5)--cycle;
-\draw[->] (-0.5,0)--(11,0);
-\draw[->] (0,-0.5)--(0,9.5);
-\draw[very thick] (0,6)--(0.5,2)--(2,.4)--(5,1);
-\draw(5,1)--(8,4)--(11,9);
-\draw[thick] (4,9.5)-- (4,1.6)--(2,1.2)--(0.5,2.8)--(0,6.8);
-\draw[dotted] (6,2)--(4,1.6);
-%useless part?
-\draw[dotted] (7,3)--(2,2)--(0.5,3.6)--(0,7.6);
-\draw[dotted] (8,4)--(2,2.8)--(0.5,4.4)--(0,8.4);
-\draw[dotted] (9,5.667)--(2,4.267)--(0.5,5.867)--(0.046,9.5); %%(0,9.867);
-\draw[dotted] (10,7.333)--(2,5.733)--(0.5,7.333)--(0.229,9.5); %%--(0,11.333);
-\draw[dotted] (11,9)--(2,7.2)--(0.5,8.8)-- (.413,9.5);%%(0,12.8);
-\draw[very thick,dotted] (5,1)--(11,2.2);
-\draw[dotted] (0,0)-- (5,1);
-
-\draw[<->] (4,.8)--(4,1.6);
-\draw[dotted,thick] (5,0)--(5,9.5);
-\draw[dotted,thick] (4,0)--(4,9.5);
-\draw[dotted,thick] (6,0)--(6,9.5);
-\node[scale=0.8, below left] at (0,0) { $0$ };
-\node[scale=0.8, below] at (5,0) { $d$ };
-\node[scale=0.8, below] at (4,0) { $d{-}1$ };
-\node[scale=0.8, below] at (6,0) { $d{+}1$ };
-\node[scale=0.85, below left] at (1.5,1.3) { $\mathbf{A_0}$ };
-\node[scale=0.85, right] at (10.1,7.3) { $NF(P)$ };
-%\node[scale=0.9, right] at (10.5,4.4) { $\Delta$ };
-\node[scale=0.9] at (4.2,1.2) { $\kappa$ };
-\node[scale=0.8] at (3,9) { $NP(R_0)$ };
-\node[scale=0.8] at (3,8.3) { $NP(S_0)$ };
-\node[right,scale=0.9,rotate=35] at (6.2,1.7) { $\text{slope} = \lambda_1$ };
-\node[right,scale=0.9,rotate=8] at (6.8,.9) { $\text{slope} = \lambda_0$ };
-\node[thick] at (10,1.6) { $\varphi$ };
-%\draw[dashed] (11,5)--(8,4);
-%\draw (8,4)--(2,2)--(0.5,4)--(0,8);
-\end{tikzpicture}
-\hfill\null
-
-\vspace{-5mm}
-
-\caption{Bound on $\NF(R_0)$ and $\NF(S_0)$}
-\label{fig:boundR0S0}
-\end{figure}
-%
-Finally observe that:
-$$T_0 = (1 - B_0) \mod A_0 = \big((A_0 - P) \div A_0\big) \mod A_0.$$
-Therefore $b_0(T_0) \geq \kappa$ results from $b_\varphi(A_0{-}P) \geq 
-\kappa$ thanks to Lemma~\ref{lem:bphib0}. 
-We have then established~\eqref{eq:induction} when $i = 0$.
-
-We now assume~\eqref{eq:induction} for the index $i$.
-From $A_{i+1} = A_i + R_i$ and the estimation $b_\varphi(R_i) \geq 2^i 
-\kappa > 0$, we derive $\NF(A_{i+1}) = \varphi_{|[0,d]}$. Therefore,
-Lemma~\ref{lem:bphib0} applies with $A = A_i$ and $A = A_{i+1}$.
-Now coming back to the the definition of $B_{i+1}$ and using
-Theorem~\ref{theo:EDivisionNP}, we get $b_0(B_{i+1}) \geq b_1(B_{i+1}) 
-\geq 0$. As a consequence:
-$$b_\varphi(R_i B_{i+1}) \geq b_\varphi(R_i) + b_0(B_{i+1})
-\geq 2^i \kappa$$
-by Lemma~\ref{lem:bphib0} and the induction hypothesis.
-Using again Lemma~\ref{lem:bphib0}, we then derive from 
-\eqref{eqdef:Biminus} and \eqref{eqdef:Siminus} that $b_0(\Delta B_i) \geq 
-2^i \kappa$ and $b_\varphi(\Delta S_i) \geq 2^i \kappa$.
-Similarly, using \eqref{eqdef:Si2} and the estimations we already know,
-we obtain $b_\varphi(S_i) \geq 2^i \kappa$. Combining this with
-$b_\varphi(\Delta S_i) \geq 2^i \kappa$, we find
-$b_\varphi(S_{i+1}) \geq 2^i \kappa$ as well. Applying again and again
-the same strategy, we deduce successively
-$b_0(\Delta V_i) \geq 2^i \kappa$ using~\eqref{eqdef:Viminus},
-$b_0(1{-}Q_i) \geq 2^i \kappa$ using~\eqref{eqdef:Qiminus},
-$b_\varphi(R_{i+1}) \geq 2^{i+1} \kappa$ using~\eqref{eq:Riplus},
-and then $b_0(T_{i+1}) \geq 2^{i+1} \kappa$
-using~\eqref{eq:ViBiplus}. Finally, coming back to the recurrence
-defining $V_{i+1}$ and remembering that $b_1(V_i)$ and $b_1(B_{i+1})$
-are both nonnegative, we find $b_1(V_{i+1}) \geq 0$.
-The equalities and inequalities
-of Eq.~\eqref{eq:induction} have then all been established for
-the index $i+1$ and the induction goes.
-
-From the inequalities $b_\varphi(R_i) \geq 2^i \kappa$, we deduce that 
-the sequence $(A_i)$ is Cauchy and therefore converges. Its limit 
-$A_\infty$ certainly satisfies $\NF(A_\infty) = \varphi_{|[0,d]}$ because 
-all the $A_i$'s do. Moreover we know that $b_\varphi(S_i) \geq 2^i 
-\kappa$ from what we derive that the sequence $(S_i)$ goes to $0$.
-Coming back to the definition of $S_i$, we find $P \mod A_\infty = 0$,
-\emph{i.e.} $A_\infty$ divides $P$. Finally, Eq.~\eqref{eq:rateconv}
-giving the rate of convergence follows from the writing
-$A_\infty{-}A_i = \sum_{j=i}^\infty R_j$
-together with the facts that $b_\varphi(R_j) \geq 2^i \kappa$ and $\deg 
-R_j \leq d{-}1$ for all $j \geq i$.
-
-\begin{rem}
-\label{rem:slope-factor}
-From the proof above, the sequence $(V_i)_{i \geq 0}$
-converges as well. Its limit $V_\infty$ is an inverse of $B_\infty
-= P \div A_\infty$ modulo $A_\infty$ and it satisfies in addition
-$b_1(V_\infty) \geq 0$.
-
-Moreover, the conclusion of 
-Theorem~\ref{theo:slope-factor} is still correct if $A_0$ is any 
-polynomial of degree $d$ with leading coefficient $a_d$ and $V_0$
-is any polynomial as soon as they satisfy:
-$$b_\varphi\big(V_0 P \mod A_0\big) > 0
-\quad \text{and} \quad
-b_0\big((1 - V_0 B_0) \mod A_0\big) > 0$$
-except that the constant $\kappa$ giving the rate of convergence
-should be now
-$\kappa = \min \big( 
-b_\varphi\big(V_0 P \mod A_0\big), \,
-b_0\big((1 - V_0 B_0) \mod A_0\big)\big)$.
-\end{rem}
-
-\subsection{A slope factorization algorithm}
-
-Let $P \in K_n[X]$ and $d$ be the abscissa of an extremal point of 
-$\NP(P)$. Previously (\emph{cf} Theorem~\ref{theo:slope-factor}), we 
-have defined a sequence $(A_i, V_i)$ converging to $(A,V)$ where $A$ is 
-a factor of $P$ whose Newton function is $\NF(P)_{|[0,d]}$ and $V$ is the inverse of $B = P/A$ modulo $A$.
-We now assume that $P$ is known up to some finite precision: $P = P_\app + 
-O(\hspace{0.5mm}\cdots)$ where the object inside the $O$ depends on the 
-chosen precision model. We address the two following questions:
-(1)~what is the precision of the factor $A$, and
-(2)~how can one compute in practice $A$ at this precision?
-
-In the sequel, it will be convenient to use a different normalization on 
-$A$ and $B$: if $a_d$ is the coefficient of $P$ of degree $d$, we set
-$A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
-so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We also assume
-the leading coefficient of $P$ is \emph{exactly}~$1$; the precision datum on $P$ then 
-only concerns the coefficients up to degree $n{-}1$. Similarly,
-$A^{(1)}$ and $B^{(1)}$ only carry
-precision up to degree $d{-}1$ and $n{-}d{-}1$ respectively.
-
-\medskip
-
-\noindent
-{\bf Newton precision.}
-We assume that the precision of the input $P$ has the shape 
-$O(\varphi_P)$ where $\varphi_P$ is a Newton function of degree 
-$n{-}1$. From now on, we assume that the precision $O(\varphi_P)$ 
-is nondegenerate in the sense of Definition~\ref{def:nondeg}. This
-ensures in particular that the Newton polygon of $P$ is well defined. 
-We import the notations $\varphi$, $b_\varphi$ and $b_0$ from \S 
-\ref{ssec:Newtoniter} and refer to 
-Eqs.~\eqref{eqdef:functionphi}--\eqref{eqdef:b0} for the definitions.
-
-\begin{prop}
-\label{prop:Newtonprecslope}
-We keep all the above notations and assumptions. We set:
-$$\delta = \min_{x \in [0,n{-}1]} \varphi_P(x) - \varphi(x)$$
-and assume that $\delta > 0$. 
-Then the factor $A^{(1)}$ is known with precision at least 
-$O(\varphi_{A^{(1)}})$ with
-$\varphi_{A^{(1)}} = \varphi_{|[0,d{-}1]} - \varphi(d) + \delta$.
-\end{prop}
-
-\begin{proof}
-Let $\delta_P \in K[X]$ be such that $\NP(\delta_P) \geq \varphi_P$.
-Let $A_\app^{(1)}$ and $A^{(1)}$ be the monic factors of $P_\app$
-and $P_\app + \delta P$ respectively whose Newton functions are
-$\varphi_{|[0,d]} - \varphi(d)$.
-
-We define the sequences $(A_i)$ and $(V_i)$ by the recurrence: 
-\begin{align*} 
-A_0 & = A_\app, \quad V_{0} = V_\app \medskip \\ A_{i+1} 
-&= A_i + \big(V_i (P_\app + \delta_P) \:\mod\: A_i\big), \smallskip \\ 
-V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\ 
-&  \hspace{1cm}\text{where } B_{i+1} = (P_\app + \delta_P) \div A_{i+1}. 
-\end{align*} 
-where $A_\app$ and $V_\app$ are those related to $P_\app$. Note that 
-$A_\app = a_d \cdot A^{(1)}_\app$ if $a_d$ denotes the coefficient of 
-$X^d$ in $P_\app$. By Remark~\ref{rem:slope-factor}, we know that the 
-sequence $(A_i)$ converges to $A = a_d \cdot A^{(1)}$ and furthermore:
-$$\NP(A - A_\app) \geq \varphi_{|[0,d{-}1]} + 
-b_\varphi\big((V_\app \cdot \delta P) \mod A_\app\big)$$ 
-since $(V_\app P_\app) \mod A_\app = (1 - V_\app B_\app) \mod A_\app = 
-0$. Using repeatedly Lemma~\ref{lem:bphib0}, we obtain:
-$$b_\varphi\big((V_\app \cdot \delta P) \mod A_\app\big)
-  \geq b_0(V_\app) + b_\varphi(\delta P) \\
-  \geq b_\varphi(\delta P) \geq \delta.$$
-Thus $\NP(A - A_\app) \geq \varphi_{[0,d{-}1]} + \delta$. Dividing
-by $a_d$, we find
-$\NP(A^{(1)} - A^{(1)}_\app) \geq \varphi_{A^{(1)}}$ and we are done.
-\end{proof}
-
-\begin{rem}
-Under the hypothesis~\textbf{(H)} introduced below, a correct precision 
-on $A^{(1)}$ is also $O(\psi_{A^{(1)}})$ where:
-$$\psi_{A^{(1)}} = \big(\varphi_P \nftimes \big(\NF(V_\app)
-- \NF(P)(d)\big)\big) \nfmod \NF(P)_{|[0,d]}.$$
-This follows from Proposition~\ref{prop:precA1} using
-$V^{(1)}_\app = a_d^{-1} V_\app$. It follows in addition from 
-Remark~\ref{rem:slope-factor} that $\NF(V_\app)$ is bounded from below 
-by $x \mapsto \lambda_1 x$. This yields the bound
-$\psi_{A^{(1)}} \geq \big(\varphi_P \nftimes \psi\big) \nfmod 
-\NF(P)_{|[0,d]}$
-where $\psi : [0,d{-}1] \to \R$ is the affine function mapping $x$
-to $\lambda_1 x - \NF(P)(d)$. 
-\end{rem}
-
-We can now move to the second question we have raised before, 
-\emph{i.e.} the design of an algorithm for computing $A^{(1)}$ with the 
-precision given by Proposition~\ref{prop:Newtonprecslope}. Our strategy 
-consists in computing first the precision and then applying the Newton 
-iteration until the expected precision is reached.
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm} {\tt slope\_factorisation\_Newton}
-
-\noindent{\bf Input:} a monic polynomial $P + O(\varphi_P) \in K_n[X]$,
-
-\noindent\phantom{{\bf Input:} }a break point $d$ of $\NP(P)$
-
-\noindent{\bf Output:} the factor $A$ described above
-
-\smallskip\noindent 1.\ %
-Compute the functions $\NF(P)$ and $\varphi$
-
-\smallskip\noindent 2.\ %
-Compute $\varphi_A = \varphi_{|[0,d-1]} - \varphi(d) + \min_{x \in [0,n]}
-\varphi_P(x) - \varphi(x)$
-
-\smallskip\noindent 3.\ %
-Compute $\kappa = \NP(P)(d{+}1) + \NF(P)(d{-}1) - 2 \cdot \NF(P)(d)$
-
-\smallskip\noindent 4.\ %
-Set $i = 0$, $A_0 = \sum_{i=0}^d a_i X^i$
-($a_i = \text{coeffs of } P$), $V_0 = 1$
-
-\smallskip\noindent 5.\ %
-{\bf repeat until} $\varphi_{|[0,d{-}1]} + 2^i \kappa \geq \varphi_A$
-
-\smallskip\noindent 6.\ \hspace{5mm}%
-lift $A_i$, $V_i$ and $P$ at enough precision
-
-\smallskip\noindent 7.\ \hspace{5mm}%
-compute
-
-\smallskip
-
-\hspace{1cm}$\bullet$
-$A_{i+1} = A_i + (P V_i \mod A_i)$
-
-\hspace{1cm}\phantom{$\bullet$ }%
-at precision $O(\varphi_{|[0,d{-}1]} + 2^{i+1} \kappa)$
-
-\smallskip
-
-\hspace{1cm}$\bullet$
-$V_{i+1} = (2 V_i - V_i^2 \cdot (P \div A_{i+1})) \mod A_{i+1}$
-
-\hspace{1cm}\phantom{$\bullet$ }%
-at precision $O(x \mapsto \lambda_1 x + 2^{i+1}\kappa)$
-
-\smallskip\noindent 8.\ \hspace{5mm}%
-set $i = i+1$
-
-\smallskip\noindent 9.\ 
-{\bf return} $A_i + O(\varphi_A)$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\begin{rem}
-The precision needed at line 6 is of course governed by the computation
-performed at line 7. Note that it can be either computed \emph{a
-priori} by using Proposition~\ref{prop:NewtonprecEuclide} or dynamically 
-by using relaxed algorithms from 
-\cites{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
-cases, it is in $O(\pi^{N_i})$ with $N_i = O(2^i \kappa + \min 
-\NF(P))$.
-\end{rem}
-
-\noindent
-It follows from Theorem~\ref{theo:slope-factor}, 
-Remark~\ref{rem:slope-factor} and Proposition~\ref{prop:Newtonprecslope} 
-that Algorithm {\tt slope\_factorisation\_Newton} is correct and 
-stable. Using the standard soft-$O$ notation $\softO(\cdot)$ for
-hiding logarithmic factor, our algorithm performs at most $\softO(n)$ 
-combinatorial operations and $\softO(n)$ operations in $K$ at precision 
-$O(\pi^N)$ with $N = O(\max \varphi_P - \min \NF(P))$ if one uses 
-quasi-optimal algorithms for multiplication and Euclidean division of 
-polynomials.
-
-\medskip
-
-\noindent
-{\bf Lattice precision.}
-The precision datum is given here by a lattice $H_P$ in $K_{\leq n{-}1}
-[X]$; we shall then write
-$P = P_\app + O(H_P)$
-where $P_\app$ is a \emph{monic} approximation of the inexact polynomial 
-$P$ we want to factor. We assume from now that $H_P$ is sufficiently 
-small so that the Newton polygon of $P$ is well defined. We then can
-define the function:
-$$F = (F_A, F_B) : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
-mapping a polynomial $P$ to the pair $(A^{(1)}, B^{(1)})$ obtained
-from it. We set $(A_\app^{(1)}, B_\app^{(1)}) = F(P_\app)$.
-
-We make the following hypothesis \textbf{(H)}:
-
-\medskip
-
-\noindent
-\hfill
-\begin{minipage}{8cm}
-The lattice $H_P$ is a first order lattice at every point
-of $P_\app + H_P$ in the sense of \cite[Definition~3.3]{caruso-roe-vaccon:14a},
-\emph{i.e.} for all $P \in P_\app + H_P$:
-
-\vspace{-5mm}
-
-$$F(P + H_P) = F(P) + F'(P)(H_P).$$
-\end{minipage}
-\hfill \null
-
-\bigskip
-
-\noindent
-Obviously \textbf{(H)} gives an answer to the first question we have 
-raised above: the precision of $A^{(1)}$ is the lattice 
-$H_{A^{(1)}}$ defined as projection on the first component of 
-$F'(P_\app)(H_P)$. It turns out that it can be computed explicitely as 
-shown by the next proposition.
-
-\begin{prop}
-\label{prop:precA1}
-The map $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
-and its differential at some point $P$ is the linear mapping
-$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
-where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
-$B^{(1)}$ modulo $A^{(1)}$.
-\end{prop}
-
-\begin{proof}
-The function $F$ is injective and a left inverse of it is
-$G : (A^{(1)},B^{(1)}) \mapsto A^{(1)}B^{(1)}$.
-Clearly $G$ is of class $C^1$ and its differential is given by
-\begin{equation}
-\label{eq:dPdAdB}
-(dA^{(1)}, dB^{(1)}) \mapsto dP = A^{(1)} \: dB^{(1)} + B^{(1)} \: dA^{(1)}.
-\end{equation}
-Thanks to Bézout Theorem, it is invertible as soon as $A^{(1)}$ and 
-$B^{(1)}$ are coprime, which is true because $\NP(A^{(1)})$ and 
-$\NP(B^{(1)})$ do not shape a common slope. As a consequence $F$ is of 
-class $C^1$ and its differential is obtained by inverting 
-Eq.~\eqref{eq:dPdAdB}. Reducing modulo $A^{(1)}$, we get $dP \equiv 
-B^{(1)} \: dA^{(1)} \pmod {A^{(1)}}$. The claimed result follows after 
-having noticed that $dA^{(1)}$ has degree at most $d{-}1$.
-\end{proof}
-
-A remarkable Corollary of Proposition~\ref{prop:precA1} asserts the
-optimality of Proposition~\ref{prop:Newtonprecslope} in a particular 
-case.
-
-\begin{cor}
-We assume~\textbf{(H)}. Let $\delta \in \R$.
-When the precision of $P$ is given by $O(\NF(P)_{|[0,d-1]} + \delta)$,
-the precision of $A^{(1)}$ given by 
-Proposition~\ref{prop:Newtonprecslope} is optimal.
-\end{cor}
-
-\begin{proof}
-First remark that the $\delta$ defined in the statement of 
-Proposition~\ref{prop:Newtonprecslope} coincide with the $\delta$
-introduced in the Corollary. Define
-$\varphi_{A^{(1)}} = \varphi_{|[0,d{-}1]} - \varphi(d) + \delta$.
-Let $H_P$ (resp. $H_{A^{(1)}})$ be the lattice consisting of
-polynomials of degree at most $n{-}1$ (resp. at most $d{-}1$)
-whose Newton function is not less that $\varphi_P = \NF(P)_{|[0,d-1]} + \delta$ (resp. 
-$\varphi_{A^{(1)}}$). We have to show that $F'_A(P_\app)(H_P) = 
-H_{A^{(1)}}$. According to Proposition~\ref{prop:precA1}
-the mapping $G : K_{\leq d{-}1}[X] \to K_{\leq n{-}1}[X]$,
-$dA^{(1)} \mapsto (B_\app^{(1)} \: dA^{(1)}) \mod A_\app^{(1)}$
-is a right inverse of $F'_A(P_\app)$. It is then enough to prove
-that $G$ takes $H_{A^{(1)}}$ to $H_P$, which can be done easily
-using Theorem~\ref{theo:EDivisionNP}.
-\end{proof}
-
-To compute $A^{(1)}$ in practice, we give a stopping criterion.
-
-\begin{prop}
-\label{prop:stop}
-We assume \textbf{(H)}.
-
-\vspace{-1mm}
-
-\begin{enumerate}[(i)]
-\item Let $\tilde A^{(1)} \in K_{=d}[X]$ such that
-$\tilde A^{(1)} \tilde B^{(1)} \in P_\app + O(H_P)$ with
-$\tilde B^{(1)} = P_\app \div \tilde A^{(1)}$.
-Then $\tilde A^{(1)} \in A_\app^{(1)} + H_{A^{(1)}}$.
-\vspace{-1mm}
-\item Let in addition $\tilde V^{(1)} \in K_{\leq d{-}1}[X]$ such that:
-$$\big(\tilde B^{(1)} \tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}
-= H_P \mod \tilde A^{(1)}.$$
-Then $H_{A^{(1)}} = \big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}$.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}
-\emph{(i)} 
-Set $\tilde P = \tilde A^{(1)} \tilde B^{(1)}$. We know by assumption
-that $\tilde P = P_\app + O(H_P)$. Thus $F(\tilde P)$ is well defined.
-The unicity of the slope factorization (\emph{cf} 
-Remark~\ref{rem:unicity}) further implies that $F(\tilde P) = 
-(\tilde A^{(1)}, \tilde B^{(1)})$. The claimed result now follows from
-the hypothesis~\textbf{(H)}.
-
-\emph{(ii)} By applying~\textbf{(H)} with $P = \tilde P$ and replacing 
-$F'(\tilde P)$ by its expression given by Proposition~\ref{prop:precA1}, 
-we find:
-$$H_{A^{(1)}} = \big((\tilde B^{(1)})^{-1} \cdot H_P\big) \mod \tilde A^{(1)}.$$
-Dividing $\big(\tilde B^{(1)} \tilde V^{(1)} H_P\big) \mod \tilde 
-A^{(1)} = H_P \mod \tilde A^{(1)}$ by $\tilde B^{(1)}$ modulo $\tilde
-A^{(1)}$, we get
-$H_{A^{(1)}} = \big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}$
-as expected.
-\end{proof}
-
-As a conclusion, the algorithm we propose consists in computing the 
-Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of
-Algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
-until we find a pair $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
-the requirements (i) and (ii) of Proposition~\ref{prop:stop}. Once this 
-pair has been found, one may safely output $\tilde A^{(1)} + 
-O\big(\big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}\big)$ under
-\textbf{(H)}.
-The resulting algorithm has quasi-optimal running 
-time and optimal stability.
-
-\bibliographystyle{plain}
-
-% \bib, bibdiv, biblist are defined by the amsrefs package.
-\begin{bibdiv}
-\begin{biblist}
-\tiny
-
-\bib{berthomieu-hoeven-lecerf:11a}{article}{
-      author={Berthomieu, J{\'e}r{\'e}my},
-      author={van~der Hoeven, Joris},
-      author={Lecerf, Gr{\'e}goire},
-       title={{Relaxed algorithms for $p$-adic numbers}},
-        date={2011},
-     journal={J. Th{\'e}orie des Nombres des Bordeaux},
-      volume={23},
-      number={3},
-       pages={{541\ndash 577}},
-}
-
-\bib{berthomieu:hal-00642075}{article}{
-  TITLE = {{Polynomial root finding over local rings and application to error correcting codes}},
-  author = {Berthomieu, J{\'e}r{\'e}my},
-  author = {Lecerf, Gr{\'e}goire},
-  author = {Quintin, Guillaume},
-  JOURNAL = {{Applicable Algebra in Engineering, Communication and Computing}},
-  PUBLISHER = {{Springer Verlag}},
-  VOLUME = {24},
-  NUMBER = {6},
-  PAGES = {413\ndash 443},
-  YEAR = {2013},
-  DOI = {10.1007/s00200-013-0200-5},
-}
-
-\bib{caruso:16}{unpublished}{
-      author={Caruso, Xavier},
-       title={Slope factorization of ore polynomials},
-        date={2016},
-        note={in preparation},
-}
-
-\bib{caruso-roe-vaccon:14a}{article}{
-      author={Caruso, Xavier},
-      author={Vaccon, Tristan},
-      author={Roe, David},
-       title={Tracking $p$-adic precision},
-        date={2014},
-     journal={LMS Journal of Computation and Mathematics},
-      volume={17 (Special issue A)},
-       pages={274\ndash 294},
-}
-
-\bib{caruso-roe-vaccon:15a}{inproceedings}{
-      author={Caruso, Xavier},
-      author={Vaccon, Tristan},
-      author={Roe, David},
-       title={$p$-adic stability in linear algebra},
-        date={2015},
-   booktitle={Proceedings of the 2015 acm on international symposium on
-  symbolic and algebraic computation},
-   publisher={ACM},
-     address={New York},
-       pages={{101\ndash 108}},
-}
-
-\bib{dwork-geratto-sullivan:Gfunctions}{book}{
-      author={Dwork, Bernard},
-      author={Gerotto, Giovanni},
-      author={Sullivan, Francis},
-       title={{An introduction to $G$-functions}},
-   publisher={Princeton U.P.},
-     address={Princeton},
-        date={1994},
-}
-
-\bib{guardia-montes-nart:08a}{article}{
-      author={Gu\`ardia, Jordi},
-      author={Montes, Jes\'us},
-      author={Nart, Enric},
-       title={Newton polygons of higher order in algebraic number theory},
-        date={2012},
-     journal={Transactions of the AMS},
-      volume={364},
-      number={1},
-       pages={361\ndash 416},
-}
-
-\bib{guardia-nart-pauli:12a}{article}{
-      author={Guàrdia, Jordi},
-      author={Nart, Enric},
-      author={Pauli, Sebastian},
-       title={Single-factor lifting and factorization of polynomials over local
-  fields},
-        date={2012},
-     journal={Journal of Symbolic Computation},
-      volume={47},
-      number={11},
-       pages={1318 \ndash  1346},
-}
-
-\bib{montes:99a}{thesis}{
-      author={Montes, Jes\'us},
-       title={Pol\'igonos de newton de orden superior y aplicaciones
-  aritm\'eticas},
-        type={Ph.D. Thesis},
-        date={1999},
-}
-
-\bib{pauli:10a}{inproceedings}{
-      author={Pauli, Sebastian},
-       title={{Factoring polynomials over local fields II}},
-        date={2010},
-   booktitle={Algorithmic number theory, 9th international symposium},
-      editor={Hanrot, Guillaume},
-      editor={Morain, Fran{\c c}ois},
-      editor={Thom\'e, Emmanuel},
-     address={Nancy, France},
-}
-
-\bib{hoeven:02a}{article}{
-      author={van~der Hoeven, Joris},
-       title={{Relax, but don't be too lazy}},
-        date={2002},
-     journal={J. Symbolic Comput.},
-      volume={34},
-      number={6},
-       pages={479\ndash 542},
-}
-
-\bib{hoeven:07a}{article}{
-      author={van~der Hoeven, Joris},
-       title={New algorithms for relaxed multiplication},
-        date={2007},
-     journal={J. Symbolic Comput.},
-      volume={42},
-      number={8},
-       pages={792\ndash 802},
-}
-
-\bib{poteaux-rybowicz}{inproceedings}{
-  author    = {Poteaux, Adrien},
-  author    = {Rybowicz, Marc},
-  title     = {Improving Complexity Bounds for the Computation of Puiseux Series over Finite Fields},
-  booktitle = {ISSAC'15},
-  pages     = {299--306},
-  year      = {2015},
-}
-
-\end{biblist}
-\end{bibdiv}
+\subsection{Algorithm}
 

### 2016-12-03 charpoly/sage/digits_of_precision.py

@@ -1,0 +1,44 @@
+from sage.rings.infinity import Infinity
+from sage.matrix.constructor import matrix
+
+def digits_of_precision(vectors):
+    n = len(vectors[0])
+    N = len(vectors)
+    ring = vectors[0][0].parent()
+
+    # Not diffused digits of precision
+    not_diffused = n * [ Infinity ]
+    for v in vectors:
+        for i in range(n):
+            val = v[i].valuation()
+            if val < not_diffused[i]:
+                not_diffused[i] = val
+    not_diffused = sum(not_diffused)
+
+    # All digits of precision
+    all = 0
+    M = matrix(vectors)
+    for j in range(n):
+        # Find pivot
+        val = Infinity
+        piv = 0
+        for i in range(j,N):
+            tmp = M[i,j].valuation()
+            if tmp < val:
+                val = tmp
+                piv = i
+        if M[piv,j] == 0:
+            for i in range(j,N):
+                print M[i,j], M[i,j].precision_absolute()
+            raise RuntimeError
+
+        all += val
+
+        # Update matrix
+        M.swap_rows(piv,j)
+        for i in range(j+1,N):
+            scalar = ring(- M[i,j] / M[j,j])
+            M.add_multiple_of_row(i, j, scalar)
+
+    diffused = all - not_diffused
+    return all, diffused, not_diffused

### 2016-12-03 charpoly/sage/hessenberg.py

@@ -1,0 +1,56 @@
+class DynamicBasis:
+    def __init__(self, base_ring, dimension):
+        if base_ring not in DiscreteValuationRings():
+            raise TypeError("%s is not a discrete valuation ring" % base_ring)
+        self._base_ring = base_ring
+        self._dimension = dimension
+        self._space = base_ring**dimension
+        self._vectors = [ ]
+
+    def add_vector(self, v):
+        v = self._space(v)
+        coeffs = [ ]
+        for (w, piv) in self._vectors:
+            coeffs.append(v[piv])
+            v -= v[piv]*w
+        val = Infinity
+        for i in range(self._dimension):
+            n = v[i].valuation()
+            if n < v[i].precision_absolute() and n < val:
+                val = n
+                piv = i
+        if val is not Infinity:
+            scalar = v[piv]
+            v /= scalar
+            self._vectors.append((v,piv))
+            coeffs.append(scalar)
+        else:
+            v = None
+        coeffs += [ self._base_ring(0) ] * (self._dimension - len(coeffs))
+        return coeffs, v
+
+    def basis(self):
+        return [ v for (v,piv) in self._vectors ]
+
+
+def hessenberg(M):
+    if not M.is_square():
+        raise ValueError
+    dimension = M.nrows()
+    ring = M.base_ring()
+    space = ring**dimension
+    
+    basis = DynamicBasis(ring, dimension)
+    hessenberg = [ ]
+    i = 0; v = None
+    while i < dimension:
+        while v is None:
+            v = space.random_element()
+            _, v = basis.add_vector(v)
+        while v is not None:
+            v = M*v
+            coeffs, v = basis.add_vector(v)
+            hessenberg.append(coeffs)
+            i += 1
+
+    return matrix(hessenberg)

### 2016-12-03 charpoly/sage/hessenberg_char_pol.py

@@ -1,0 +1,212 @@
+def Hessenberg(Mat):
+    r"""
+    
+    Computes the Hessenberg form of the matrix M,
+    along with the passage matrix P
+
+    
+    INPUT:
+
+    - ``Mac`` -- a matrix
+    
+    OUTPUT:
+
+    H, a matrix under Hessenberg form, P the passage matrix, Pinv its inverse.
+
+    EXAMPLES::
+
+        sage: M = Matrix(Qp(2,50),4,4,[1,2,3,4,-5,-6,7,8,9,10,-11,12,13,14,15,-16])
+        sage: H,P,Pinv = Hessenberg(M)
+        sage: H
+        [  0   1   2   0]
+        [  1   0  -4  -3]
+        [  0   0   1 3/4]
+        [  0   0   0   1]
+        sage: P
+        [1, 0, 2, 3]
+        sage: P*M*Pinv
+
+    """
+    M = copy(Mat)
+    n = M.nrows()
+    m = M.ncols()
+    K = M.base_ring()
+    R = K.integer_ring()
+
+    #Beware: P could also be defined at infinite precision, but this is not our choice today
+    P = MatrixSpace(K,n,n)(1)
+    Pinv = MatrixSpace(K,n,n)(1)
+
+    for j in range(m-1):
+        i=j+1
+        imax=-1
+        #cmax= M[i,j]
+        while i<n:
+            if M[i,j] <>0:
+                if imax==-1:
+                    imax=i
+                    cmax= M[i,j]
+                else:
+                    if M[i,j].valuation()<cmax.valuation():
+                        imax=i
+                        cmax= M[i,j]
+            i=i+1
+        if imax>=0:
+            #M.rescale_row(i, M[i,j]^(-1), start_col=j)
+            M.swap_rows(j+1,imax)
+            M.swap_columns(j+1,imax)
+
+            P.swap_rows(j+1,imax)
+            Pinv.swap_columns(j+1,imax)
+
+            for l in range(j+2,n):
+                coeff = (M[l,j]*M[j+1,j]**(-1)).lift()
+                M.add_multiple_of_row(l, j+1, -coeff , start_col=j)
+                P.add_multiple_of_row(l, j+1, -coeff)
+
[…]

### 2016-12-03 charpoly/sage/test_digits.py

@@ -1,0 +1,42 @@
+#from digits_of_precision import *
+
+def digits_standard_basis(M):
+    R = M[0,0].base_ring()
+    S = PolynomialRing(R, name='X')
+    C = (M - S.gen()).adjoint()   # à remplacer par le truc de Tristan
+
+    n = M.nrows()
+    vectors = [ ]
+    for i in range(n):
+        for j in range(n):
+            vectors.append([ C[i,j][k] for k in range(n) ])
+
+    return digits_of_precision(vectors)
+
+
+def digits_legendre_diagonal(M):
+    R = M[0,0].base_ring()
+    S = PolynomialRing(R, name='X')
+    C = (M - S.gen()).adjoint()   # à remplacer par le truc de Tristan
+    diag = M.diagonal()
+
+    n = M.nrows()
+    vectors = [ ]
+    for i in range(n):
+        for j in range(n):
+            vectors.append([ C[i,j](x) for x in diag ])
+
+    return digits_of_precision(vectors)
+
+def digits_legendre(M, evaluation_points):
+    R = M[0,0].base_ring()
+    S = PolynomialRing(R, name='X')
+    C = (M - S.gen()).adjoint()   # à remplacer par le truc de Tristan
+
+    n = M.nrows()
+    vectors = [ ]
+    for i in range(n):
+        for j in range(n):
+            vectors.append([ C[i,j](x) for x in evaluation_points ])
+
+    return digits_of_precision(vectors)

### 2016-12-03 resultants/resultants.tex
Small changes

@@
-%\documentclass{sig-alternate}
-%\documentclass{lms}
-\documentclass{article}
+\documentclass{jT}
+%\documentclass{article}
 \usepackage[utf8]{inputenc}
 
-\usepackage{vmargin}
-
 \usepackage{amsmath,amssymb,amsthm}
-%\usepackage{amsrefs}
 \usepackage[usenames,dvipsnames]{color}
@@
 \usepackage{enumerate}
-\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
+\usepackage[norelsize,algoruled,vlined,english,linesnumbered]{algorithm2e}
 \usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
@@
 \renewcommand*{\theHsection}{\thesection}
+
+\numberwithin{equation}{section}
 
@@
 \DeclareMathOperator{\round}{round}
-
-\begin{document}
 
@@
 
-\def\todo#1{\ \!\!{\color{red} #1}}
 \definecolor{purple}{rgb}{0.6,0,0.6}
@@
 
-\title{Numerical stability of Euclide algorithm\\over ultrametric fields}
-
-%\numberofauthors{1}
-%\author{
-%\alignauthor Xavier Caruso\\
-%  \affaddr{Universit\'e Rennes 1}\\
-%  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
-%}
+\begin{document}
+
+\title[Numerical stability of Euclide algorithm over ultrametric fields]
+{Numerical stability of Euclide algorithm\\over ultrametric fields}
+
 \author{Xavier Caruso}
+\address{Xavier Caruso\\
+Université Rennes 1, IRMAR, 35042 Rennes Cedex\\
+France}
+\email{xavier.caruso@normalesup.org}
+\urladdr{http://perso.univ-rennes1.fr/xavier.caruso}
+
+\subjclass{11S99, 11Y99, 68W30}
+
+\keywords{Euclide algorithm, ultrametric precision, subresultants}
 
@@
 
-\begin{abstract}
+\begin{resume}
+Nous étudions le problème de la stabilité du calcul des résultants et
+sous-résultants des polynômes définis sur des anneaux de valuation
+discrète complets (\emph{e.g.} $\Zp$ ou $k[[t]]$ où $k$ est un corps).
+Nous démontrons que les algorithmes de type Euclide sont très instables
+en moyenne et, dans de nombreux cas, nous expliquons comment les rendre
+stables sans dégrader la complexité. Chemin faisant, nous déterminons
+la loi de la valuation des sous-résultants de deux polynômes $p$-adiques
+aléatoires unitaires de même degré.
+\end{resume}
+
+\begin{abstr}
 We address the problem of the stability of the computations of resultants 
@@
 same degree.
-\end{abstract}
+\end{abstr}
 
@@
 \begin{tabular}{rcl}
-$\ring$ &--- & a commutative ring (without any further assumption) \\
+$\ring$ &--- & a commutative ring \\
 $\A$ &---& a complete discrete valuation ring \\
@@
 $\ring_{<n}[X]$ &---& the free $\ring$-module consisting of
-polynomials over $\ring$ of degree $<n$ \\
+polynomials\\ & & over $\ring$ of degree $<n$ \\
 $\ring_{\leq n}[X]$ &---& the free $\ring$-module consisting of
-polynomials over $\ring$ of degree $\leq n$ \\
+polynomials\\ & & over $\ring$ of degree $\leq n$ \\
 $\ring_n[X]$ &---& the affine space consisting of \emph{monic}
-polynomials over $\ring$ of degree $n$. \smallskip \\
+polynomials\\ & & over $\ring$ of degree $n$. \smallskip \\
 
 $\Res^{d_A,d_B}(A,B)$ &---& The resultant of $A$ and
-$B$ ``computed in degree $(d_A,d_B)$'' \\
+$B$\\ & & ``computed in degree $(d_A,d_B)$'' \\
 $\Res_j^{d_A,d_B}(A,B)$ &---& The $j$-th subresultant of $A$ and
-$B$ ``computed in degree $(d_A,d_B)$'' \\
+$B$\\ & & ``computed in degree $(d_A,d_B)$'' \\
 \end{tabular}
@@
       \big(27 + O(2^5)\big) X^4 +
-      \big(11 + O(2^5)\big) X^3 +
+      \big(11 + O(2^5)\big) X^3 \\
+& \hspace{1cm} + 
       \big(5 + O(2^5)\big) X^2 +
@@
       \big(24 + O(2^5)\big) X^4 +
-      \big(25 + O(2^5)\big) X^3 +
+      \big(25 + O(2^5)\big) X^3 \\
+& \hspace{1cm} + 
       \big(12 + O(2^5)\big) X^2 +
@@
 result we get is:
-$$S_3 = 
+\begin{align*}
+S_3 & = 
       \big(3 + O(2^5)\big) X^4 +
-      \big(18 + O(2^5)\big) X^3 +
+      \big(18 + O(2^5)\big) X^3 \\
+& \hspace{1cm} +
       \big(25 + O(2^5)\big) X^2 +
       \big(15 + O(2^5)\big) X +
-      \big(15 + O(2^5)\big).$$ 
+      \big(15 + O(2^5)\big).
+\end{align*}
 In order to compute $S_4$, we have now to perform the Euclidean division 
@@
 This reinterpretation is based on the so-called \emph{subresultant 
-pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
+pseudo-re\-mainder sequence} which is defined as follows. We pick $A$ and 
 $B$ as above. Denoting by $(P \,\%\, Q)$ 
@@
   \big(14 + O(2^5)\big) X^3 + 
-  \big(5 + O(2^5)\big) X^2 + 
+  \big(5 + O(2^5)\big) X^2 \\
+& \hspace{1cm} + 
   \big(17 + O(2^5)\big) X + 
@@
 By Theorem \ref{th:lawVj}, we have $\E[V_j] =
-\sum_{i=0}^d \E[Z_i]$ with $Z_i = \min(X_{j-i}, \ldots, X_{j+i})$ ($j$ 
+\sum_{i=0}^d \E[Z_i]$ with 
+$$Z_i = \min(X_{j-i}, \ldots, X_{j+i})$$ 
+($j$ 
 is fixed during all the proof). Our conventions imply that $Z_i$ 
@@
 
-By functoriality of subresultants, we may assume that $\ring = \Z[a_0, 
+By functoriality of subresultants, we may assume that $\ring$ is the
+ring $\Z[a_0, 
 \ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$ are the two 
@@
 coefficients, we get $\lambda = \mu$. Now observe that:
-$$\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) = 
-\Res^{d-j,d-j-1}(\lambda U_{j-1}, \lambda U_j) = 
-\lambda^{2d-2j-1} \cdot \Res^{d-j,d-j-1}(U_{j-1}, U_j).$$
+\begin{align*}
+\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) & = 
+\Res^{d-j,d-j-1}(\lambda U_{j-1}, \lambda U_j) \\
+& = \lambda^{2d-2j-1} \cdot \Res^{d-j,d-j-1}(U_{j-1}, U_j).
+\end{align*}
 Noting that $(\mathcal U_j, \mathcal U_{j-1}, \mathcal R_j, \mathcal 
@@
 
-\paragraph{Tracking precision \emph{locally}}
-
+\bigskip
+
+\noindent
+\textit{Tracking precision locally.}
+%
 Assume first that we want to perform a given rather simple operation --- 
@@
 
-\paragraph{Tracking precision \emph{globally}}
-
+\bigskip
+
+\noindent
+\textit{Tracking precision globally.}
+%
 As already said, we shall use the above method for tracking precision 
@@
 
-According to Corollary \ref{cor:Vj}, the expected value of the variable 
+According to Corollary \ref{cor:Vj}, the expected value of 
 $\max(V_0, \ldots, V_{d-1})$ is in $O(\log_p d)$. Thus, the average 
-complexity of Algorithm 1 is $O(d^2 \cdot \M(N + \log d))$ bit 
+complexity of Algorithm 1 is $$O(d^2 \cdot \M(N + \log d))$$ bit 
 operations. In all usual cases (\emph{cf} Remark \ref{rem:M}), this
@@
 cheaply.
+
+\def\refname{\centerline{Bibliographie}}
 
@@
   \emph{Random matrices over a DVR and LU factorization},
-  to appear at J. Symb. Comp.
+  J. Symb. Comp. {\bf 71}, 98--123
 \bibitem{cohen}
@@
   P.~Lairez, T.~Vaccon,
-  \emph{Computation of power series solutions with $p$-adic coefficients of certain differential equations},
-  preprint (2014)
+  \emph{On $p$-adic differential equations with separation of variables},
+  preprint (2016)
 \bibitem{sage}

### 2016-12-13 charpoly/charpoly.tex
First draft for the computation of the comatrix.

@@
 
-\section{Computation of $\chi$ and $d\chi$}
+\section{Computation of $\chi$ and $d \chi$}
+
+In this section, we combine the computation of an Hessenberg form
+of a matrix and the computation of the inverse through the SNF
+over a complete discrete valuation field
+to compute $\com (X-A)$ and $d \chi.$
 
@@
 
+\begin{deftn}
+$M \in M_n (k)$ is said to be an Hessenberg matrix if
+\[\forall i \in \llbracket 3, n \rrbracket, 
+\forall j \in \llbracket 1, i-2 \rrbracket, M_{i,j}=0. \]
+If $M \in M_n (k)$ and $H \in M_n (k)$ is an Hessenberg 
+matrix such that there exists $P \in GL_n(k)$
+with $H=PMP^{-1},$ we say that H is an Hessenberg form
+of $M.$
+\end{deftn}
+
+\begin{deftn}
+In the previous definition of an Hessenberg matrix $M,$
+ if we only have $M_{i,j}= O(\pi^{n_{i,j}}),$ then we say
+ that $M$ is an approximate Hessenberg matrix.
+ This generalize directly to approximate Hessenberg form. 
+\end{deftn}
+
+It is not hard to prove that every matrix over a field admits
+an Hessenberg form.
+We prove here that over $K,$ if a matrix is 
+known at finite (jagged) precision,
+we can compute an approximate Hessenberg form of it.
+Moreover, we can provide an exact change of basis matrix.
+It relies on the following algorithm.
+
+\noindent\hrulefill
+
+\noindent {\bf Algorithm 1:} {\tt Approximate Hessenberg form computation}
+
+\noindent{\bf Input:} a matrix $M$ in $M_n(K).$
+
+\smallskip
+
+\noindent 0.\ $P:=I_n.$ \: $H:=M.$
+
+\noindent 1.\ {\bf for} $j=1,\dots,n-1$ {\bf do}
+
+\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $val(H_{i_{min},j})$ is minimal.
+
+\noindent 3.\  \:  {\bf for} $i=j+2,\dots,n$ {\bf do}
+
+\noindent 4. \ \: \:  \textbf{Eliminate} the significant digits of $H_{i,j}$ by pivoting with row $j+1$ 
+using a matrix $T.$
+
+\noindent 5. \ \: \:  $H:=H \times T^{-1}.$ \: $P:=T \times P.$
+
+\noindent 6. \textbf{Return} $H,P.$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
+
+\begin{prop} 
+Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
+$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coeffient in $M.$
+\end{prop}
+\begin{proof}
+Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
+Inside the nested \textbf{for} loop, if we want to eliminate $p^{u_y} \varepsilon_y+O(p^{n_y}$ with pivot $p^{u_x} \varepsilon_x+O(p^{n_x},$
+with the $\varepsilon$'s being units,
+the coefficient of the transvection matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
+Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
+The rest is clear.
+\end{proof}
+
 \subsection{Computation of the inverse}
 
-\todo{Describe algorithm with SNF. Remark about precision.
-Application to the computation of $d\chi$}
+In this Subsection, we prove that to compute the inverse of
+a matrix over a CDVF $K$, SNF is precision-wise optimal in the flat-precision case.
+We first recall what is the differential of the computation of the inverse of a matrix.
 
-\subsection{Jagged precision}
+\begin{lem}
+Let $u \: : \: GL_n (K) \rightarrow GL_n(K),$ $M \mapsto M^{-1}.$
+Then for $M \in GL_n (K),$ $du(M) \cdot dM=M^{-1} dM M^{-1}.$
+It is always surjective.
+\end{lem}
 
-\todo{Computation of optimal jagged precision.}
+We then have the following result about the loss in precision when computing the inverse.
+
+\begin{prop}
+Let $cond(M)$ be the valuation of the last invariant factor of $M.$
+If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
+can be computed at precision $O(\pi^{N-2cond(M)})$ by a \textbf{SNF} computation
+and this lower-bound is optimal
+at least when $N$ is large.
+\end{prop}
+\begin{proof}
+The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
+It is $-2cond(M)$ for $M^{-2}$ 2 and it is then clear that $-2cond(M)$
+can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
+and the smallest that can be achieved this way for $dM$ in a precision lattice
+of flat precision. Hence the optimality of the bound given, at least when 
+$N$ is large, thanks to \todo{Reference for precision lemma}.
+
+Now, the computation of an SNF form over a CDVF was described in \todo{reference these Vaccon}.
+From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
+known at precision at least $O(\pi^{N-cond(M)})$ with coefficients in \todo{integral ring}
+and determinant in \todo{integer ring invertible} realizing an SNF form of $M.$
+There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
+Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-cond(M),$
+we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2cond(M)}),$
+which concludes the proof.
+\end{proof}
+
+\subsection{$\com (X -H)$ and flat precision}
+
+In this subsection, we want to compute $\com (X Id -H)$
+by using the SNF computation of the previous subsection.
+As it is, in a way, optimal for the computation of the
+inverse in a CDVF, we hope to get reasonnable behaviour
+for the computation of $\com (X-M)$ even though the coefficients
+are not, \textit{a priori}, set to live in a CDVF.
+We show that we can reduce this task to the computation
+of an SNF over a CDVF, with no division in $K.$
+
+Firstly, we need a lemma to relate comatrices for
+similar matrices:
+
+\begin{lem} \label{lem:comatrix_of_similar}
+If $M_1,M_2 \in M_n(K)$ and $P \in GL_n (K)$ are such that
+$M_1=PM_2P^{-1},$ then:
+\[ \com (X-M_1)=P \com (X-M_2) P^{-1}. \] 
+\end{lem}
+
+The first ingredients are reciprocal polynomials.
+We extend their definition to matrices of polynomials.
+\begin{deftn}
+Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
+whose coefficients are the reciprocal polynomials of the
+coefficients of $A.$ 
+\end{deftn}
+We then have the following result :
+\begin{lem}
+Let $M \in M_n(K).$ Then:
+\begin{eqnarray*}
+com(1-XM)^{rec}=&com(X-M), \\
+(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
+\end{eqnarray*}
+\end{lem}
+
+This lemma allows us to compute $com(1-XM)$ instead of $com(X-M).$
+This has a remarkable advantage: the pivots during the computation of
+the SNF of $com(1-XM)$ are units of $O_K[[X]],$ and are known
+in advance to be on the diagonal. This leads to a very smooth
+precision and complexity behaviour.
+
+\noindent\hrulefill
+
+\noindent {\bf Algorithm 2:} {\tt Approximate $\com (X -H)$ }
+
+\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(O_K).$
+
+\smallskip
+
+\noindent 0.\ $U:=1-XH.$ $U_0:=1-XH.$
+
+\noindent 1.\ \textbf{Track} the following operations to get $P^{-1}$ and $Q^{-1},$
+such that $U=PU_0Q.$
+
+\noindent 2.\ {\bf for} $i=1,\dots,n-1$ {\bf do}
+
+\noindent 3.\  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,j},$ for $j\geq i+1$ 
+using the invertible pivot
+$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \Zp [X]$).
+
+\noindent 4.\    {\bf for} $i=1,\dots,n-1$ {\bf do}
+
+\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,i+1},$
+using the invertible pivot $U_{i,i}.$
+
+\noindent 6. \ $\psi:=\prod_i U_{i,i}.$
+
+\noindent 7. \ Rescale to get $U = Id \mod X^{n+1}.$
+
+\noindent 8. \ $V:=\psi \times P^{-1} \times Q^{-1}   \mod X^{n+1}.$
+
+\noindent 9. \ \textbf{Return} $V^{rec}, \psi^{rec}.$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
+
+\begin{theo}
+Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
+Then, using Algorithm 2, one can compute $\com (X -H)$ in 
+$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
+\end{theo}
+\begin{proof}
+Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
+in $O_K$ at the precision given by $H.$
+Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot,
+on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
+As a consequence, when performing this column-pivoting, only two rows ($i$ and
+$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
+Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
+Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
+Coefficient for the transvection is the opposite of the one defined by pivoting
+and the indices for the receiving and giving rows are swapped compared to the operations on columns.
+Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
+an upper-triangular matrix.
+This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.
+
+As for line 4 and 5, there are only $n-1$ eliminations, resulting again in a $\softO (n^3)$ cost
+for the computation of $P^{-1}$ and $U.$
+
+Line 6 is in $\softO (n^2)$ and 7 in $\softO (n^3).$
+
+Thanks to the fact that the $P$ only corresponds to the product of $n-1$ transvections, the
+product on line 8 is in $\softO (n^3).$
+We emphasize that no division has been done throughout the algorithm.
+Line 9 is costless, and the result is then proved.
+\end{proof}
+
+\begin{prop} \label{prop:optimal_flat}
+If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
+we can compute the optimal jagged precision on $\chi (M)$
+in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
+\end{prop}
+\begin{proof}
+Computing $M=PHP^{-1}$ an approximate Hessenberg form can be done in
+$O(n^3)$ with no division.
+Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
+Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
+thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
+$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
+and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
+\end{proof}
+
+\subsection{$\com (X -M)$ and jagged precision}
+
+\subsection{Factorization}
+
+When this time $dM$ is jagged, the trick of the proof of Proposition
+\ref{prop:optimal_flat} can not work anymore.
+Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
+are not easy as $P^{-1}  dM P$ is not jagged anymore.
+Fortunately, a factorization result of $\com (X-M)$ will be enough
+to conclude that optimal precision can still be read in $\softO (n^3)$
+operations in $O_K.$
+
+\begin{prop} \label{prop:factor_com}
+Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
+We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
+(in $K[X]$).
+Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
+both in $O_K[X]^n,$ we have: 
+\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
+\end{prop}
+\begin{proof}
+We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
+\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
+Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
+Then $A(\lambda)$ is of rank one.
+Indeed, $A(\lambda) \times (\lambda-M) =0.$
+Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
+and then $A(\lambda)$ is of rank at most one.
+Nevertheless, we can remark that
+ $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
+and thus, $A(\lambda)$ is of rank exactly one.
+As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
+It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
+By the Chinese Remainder Theorem, our first result is then proved.
+
+Now, let us take $U=col(A,1)$ the first column of $A$ and
+$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
+invertible $\mod \chi (M).$
+Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
+\end{proof}
+
+\begin{rem}
+If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
+we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
+$P\in GL_n (O_K)$ then
+\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
+In other words, factorizability of $\com(X-M)$ is stable
+by change of basis.
+\end{rem}
+
+\subsection{Computation and optimal jagged precision}
+
+Thanks to the previous results, we can now compute 
+$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
+using the following algorithm.
+
+\noindent\hrulefill
+
+\noindent {\bf Algorithm 3:} {\tt Approximate $\com (X -M)$ }
+
+\noindent{\bf Input:} an approx. $M \in M_n(O_K),$ with $discr(\chi_M) \neq 0.$
+
+\smallskip
+
+\noindent 0.\ Find $P \in GL_n(O_K)$ and $H \in M_n(O_K),$ approximate Hessenberg,
+such that $M=PHP^{-1},$ using Algorithm 1.
+
+\noindent 1.\ Compute $A=\com (X-H)$ using Algorithm 2.
+
+\noindent 2.\ Do $row(A,1) \leftarrow row(A,1)+\sum_{i=2}^n \mu_i row(A,i),$ for
+random $\mu_i \in O_K,$ by doing $T \times A$ for some $T \in GL_n(O_K).$
+Compute $B:=TAT^{-1}.$
+
+\noindent 3.\ Similarily compute $C:=S^{-1}BS$ for $S \in GL_n(O_K)$ corresponding to
+sending a random linear combination of the columns of index $j \in \llbracket 2,n \rrbracket$
+to the first column of $B.$
+
+\noindent 4.\  \textbf{If} $gcd(C_{1,1}, \chi(M)) \neq 1,$ \textbf{then} go to 2.
+
+\noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
+using Proposition \ref{prop:factor_com}.
+
+\noindent 6. \ Return $com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi (M).$
+
+\vspace{-1ex}\noindent\hrulefill
+
+\medskip
+
+\begin{theo}
+Algorithm 3 compute 
+$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
+operations in $O_K$ at initial precision.
+\end{theo}
+\begin{proof}
+As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
+Multiplying by $T$ or $S$ or their inverse corresponds
+to $n$ operations on rows or columns over a matrix with coefficients
+in $O_K[X]$ of degree at most $n.$
+Thus, it is in $\softO (n^3).$
+Step 5 and 6 are also in $\softO (n^3).$
+All that is to prove is that the set of $P$ and $S$ to avoid
+is of positive dimension.
+As any the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
+for $\lambda$ a root of $\chi (M)$ (in an algebraic closuer)
+and then apply Chinese Remainder Theorem.
+The goal of the Step $2$ is to ensure the first row of $B$ contains an
+invertible modulo $\chi (M).$
+Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
+affine hyperplane so that $row(B,1) \mod (X-\lambda)$ is a non-zero vector.
+Hence for $row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
+a finite union of affine hyperplane is to avoid.
+Similarily, the goal of Step 3 is to put an invertible coefficient (modulo
+$\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine
+hyperplane is to avoid.
+Hence, the set that the $\mu_i$'s have to avoid is a finite union
+of hyperplane, and hence, is of positive dimension.
+Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
+on Step 4, which concludes the proof.
+\end{proof}
+
+\begin{rem}
+If $dM$ is jagged, we can read the optimal
+jagged precision from $\com (X-M).$
+Indeed, we can write $\com (X-M)=\sum_{i=1}^{n-1} A_i X^i$
+for some matrices $A_i \in M_n(O_K).$
+Then the best jagged precision on the coefficient of 
+$X^i$ can be read from $tr (A_i dM).$
+What then is to do is to find the smallest 
+$val((A_i)_{j,k} dM_{k,j}).$
+\end{rem}
 

### 2016-12-14 charpoly/charpoly.tex
A few words on the computation of the char poly by Hessenberg.

@@
 to compute $\com (X-A)$ and $d \chi.$
+%The computation of an Hessenberg form provides an $O(n^3)$
+%algorithm to compute the characteristic polynomial.
 
@@
 \end{proof}
+
+\begin{rem} \label{rem:char_pol_from_hessenberg}
+From an Hessenberg form of $M,$ it is well known
+that one can compute the characteristic polynomial of 
+$M$ in $O(n^3)$ operations in $K.$ See \todo{cite Cohen} page
+55 and 56.
+However, this computation involves some divisions, and its
+behaviour regarding to precision is not easy to
+quantify.
+\end{rem}
 
@@
 
+\begin{rem}
+As a consequence of the previous remark, once the optimal jagged
+precision is known, it is possible to possible lift arbitrarily
+the precision on $M$ and then to perform
+a computation of the characteristic polynomial
+from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
+in $O(n^3).$
+By truncating afterward at the optimal jagged precision,
+we can achieve by this way this optimal precision.
+\end{rem}
+
 \section{The case of diagonalizable matrices}

### 2017-01-05 charpoly/charpoly.tex
Minor improvements.

@@
 Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
-Inside the nested \textbf{for} loop, if we want to eliminate $p^{u_y} \varepsilon_y+O(p^{n_y}$ with pivot $p^{u_x} \varepsilon_x+O(p^{n_x},$
+Inside the nested \textbf{for} loop, if we want to eliminate $p^{u_y} \varepsilon_y+O(p^{n_y})$ with pivot $p^{u_x} \varepsilon_x+O(p^{n_x}),$
 with the $\varepsilon$'s being units,
@@
 The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
-It is $-2cond(M)$ for $M^{-2}$ 2 and it is then clear that $-2cond(M)$
+It is $-2cond(M)$ for $M^{-2}$ and it is then clear that $-2cond(M)$
 can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
@@
 From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
-known at precision at least $O(\pi^{N-cond(M)})$ with coefficients in \todo{integral ring}
-and determinant in \todo{integer ring invertible} realizing an SNF form of $M.$
+known at precision at least $O(\pi^{N-cond(M)})$ with coefficients in $O_K$
+and determinant in $O_K^*$ realizing an SNF form of $M.$
 There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
@@
 
-Firstly, we need a lemma to relate comatrices for
+Firstly, we need a lemma to relate comatrices of
 similar matrices:
@@
 
-The first ingredients are reciprocal polynomials.
-We extend their definition to matrices of polynomials.
+The second ingredient we need is the reciprocal polynomial.
+We extend its definition to matrices of polynomials.
 \begin{deftn}
@@
 in $O_K$ at the precision given by $H.$
-Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot,
+Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot:
 on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
@@
 \begin{proof}
-Computing $M=PHP^{-1}$ an approximate Hessenberg form can be done in
+Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
 $O(n^3)$ with no division.
@@
 
-\subsection{$\com (X -M)$ and jagged precision}
+%\subsection{$\com (X -M)$ and jagged precision}
 
@@
 Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
-are not easy as $P^{-1}  dM P$ is not jagged anymore.
+are not easy, as $P^{-1}  dM P$ is not jagged anymore.
 Fortunately, a factorization result of $\com (X-M)$ will be enough
@@
 $V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
-invertible $\mod \chi (M).$
+invertible $\mod \chi (M)).$
 Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
@@
 All that is to prove is that the set of $P$ and $S$ to avoid
-is of positive dimension.
-As any the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
-for $\lambda$ a root of $\chi (M)$ (in an algebraic closuer)
+is of dimension at most $n-1.$
+As in the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
+for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
 and then apply Chinese Remainder Theorem.
@@
 Hence, the set that the $\mu_i$'s have to avoid is a finite union
-of hyperplane, and hence, is of positive dimension.
+of hyperplane, and hence, is of dimension at most $n-1.$
 Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
@@
 As a consequence of the previous remark, once the optimal jagged
-precision is known, it is possible to possible lift arbitrarily
+precision is known, it is possible to lift arbitrarily
 the precision on $M$ and then to perform

### 2017-01-10 charpoly/charpoly.tex
Minor improvements: shear matrices and rec. poly.

@@
 with the $\varepsilon$'s being units,
-the coefficient of the transvection matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
+the corresponding coefficient of the corresponding shear matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
 Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
@@
 \begin{deftn}
+Let $P \in K[X]$ be of degree $d \in \mathbb{N}_{>0}.$ 
+We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
 Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
@@
 Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
-Coefficient for the transvection is the opposite of the one defined by pivoting
+Corresponding coefficient for the corresponding shear matrix
+ is the opposite of the one defined by pivoting
 and the indices for the receiving and giving rows are swapped compared to the operations on columns.
@@
 
-Thanks to the fact that the $P$ only corresponds to the product of $n-1$ transvections, the
+Thanks to the fact that the $P$ only corresponds to the product of $n-1$ 
+shear matrices, the
 product on line 8 is in $\softO (n^3).$

### 2017-01-10 charpoly/charpoly.tex
Some work on Section 2

--- 2017-01-05 charpoly/charpoly.tex
@@
 \newcommand{\R}{\mathbb R}
+\newcommand{\OK}{\mathcal{O}_K}
 
@@
 
+\subsection{Notation} Throughout the paper, $K$ will refer to a complete,
+discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
+$\OK$ its ring of integers and $\pi$ a uniformizer.
+
 \section{Theoretical study}
@@
 \subsection{The theory of p-adic precision}
+
+We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
+as a foundation for our discussion of the precision for the characteristic polynomial
+of a matrix.  We will be concerned with two $K$-manifolds in what follows:
+the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
+$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
+the most general kind of precision structure we may attach to $M$ is a
+\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
+arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
+frequently work with certain classes of lattices, either \emph{jagged} lattices
+where we specify a precision for each matrix entry or \emph{flat} lattices where
+every entry is known to a fixed precision $O(p^m)$.  Similarly, precision for
+monic polynomials can be specified by giving a lattice in the tangent space
+at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
+of simplicity.
+
+The main benefit of working with lattices is the following result
+\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.
+
+%\begin{prop}
+
+%\end{prop}
 

### 2017-01-18 charpoly/charpoly.tex
Begun bibliography

--- 2017-01-10 charpoly/charpoly.tex
@@
 Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
-$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coeffient in $M.$
+$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
 \end{prop}
@@
 that one can compute the characteristic polynomial of 
-$M$ in $O(n^3)$ operations in $K.$ See \todo{cite Cohen} page
+$M$ in $O(n^3)$ operations in $K.$ See \cite{Cohen:2013} page
 55 and 56.
@@
 of flat precision. Hence the optimality of the bound given, at least when 
-$N$ is large, thanks to \todo{Reference for precision lemma}.
+$N$ is large, thanks to Lemma 3.14 of \cite{CRV:2014}
 
-Now, the computation of an SNF form over a CDVF was described in \todo{reference these Vaccon}.
+Now, the computation of an SNF form over a CDVF was described in \cite{Vaccon-these}.
 From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
@@
 
+\bibliographystyle{plain}
+\bibliography{charpoly}
+
 \end{document}

### 2017-01-25 charpoly/charpoly.tex
Added abstract, working on Section 2

--- 2017-01-10 charpoly/charpoly.tex
@@
 
-\title{Characteristic polynomial of p-adic matrices}
+\title{Characteristic polynomials of p-adic matrices}
 
@@
 \begin{abstract}
+We analyze the precision of the characteristic polynomial $\chi(A)$ of
+an $n \times n$ $p$-adic matrix $A$ using differential precision methods
+developed previously.  When $A$ is integral with precision $O(p^k)$,
+we give a criterion (checkable in time $\softO(n^\omega)$) for
+$\chi(A)$ to have precision exactly $O(p^k)$.  We also give a $\softO(n^3)$
+algorithm for determining the optimal precision when the criterion is not
+satisfied, and give examples when the precision is larger than $O(p^k)$.
 \end{abstract}
@@
 
-The main benefit of working with lattices is the following result
+Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.  For a matrix $M \in M_n(K)$
+we identify the tangent space $V$ at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
+the space $K_{<n}[X]$ of polynomials of degree less than $n$.  Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed)
+ball of radius $r$ in $V$.  
+Our analysis of the $\chi$ begins with the following result
 \citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.
 
-%\begin{prop}
-
-%\end{prop}
-
-\todo{Recall the Lemma (with integral version)}
+\begin{prop} \label{prop:mainlem}
+Suppose that $M \in M_n(K)$ and that $d\chi(M): V \to W$ is surjective.
+Let $C$ be a positive real number such that
+$B_W(1) \subset d\chi(M)(B_V(C))$.  Then, for all $\rho \in (0, 1]$ and all
+$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
+satisfies:
+\begin{equation}
+\chi(M + H) = \chi(M) + d\chi(H).
+\end{equation}
+\end{prop}
 
 \subsection{Computation of the differential}
+
+In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
+\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
+Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi(M)$ is given
+by
+\[
+d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
+\]
+Moreover, $d\chi(M)$ will be surjective exactly when the Jordan blocks of $M$ all have distinct generalized eigenvalues.
 
@@
 
-\section{Computation of $\chi$ and $d \chi$}
+Equality of lattices holds iff it does mod p, so need surjectivity modulo p.  Surjectivity of differential
+can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
+in Frobenius normal form it has only one block.
+
+\section{Computation of the characteristic polynomial and its derivative}
 
@@
 over a complete discrete valuation field
-to compute $\com (X-A)$ and $d \chi.$
+to compute $\com (X-M)$ and $d \chi.$
 %The computation of an Hessenberg form provides an $O(n^3)$

### 2017-01-25 charpoly/charpoly.tex

@@
 \newcommand{\R}{\mathbb R}
-\newcommand{\OK}{\mathcal{O}_K}
 
@@
 \todo{Example of $M$ and $M + 1$}
-
-\subsection{Notation} Throughout the paper, $K$ will refer to a complete,
-discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer.
 
@@
 \bibliographystyle{plain}
-\bibliography{../roebib/Biblio}
+\bibliography{charpoly}
 

### 2017-01-25 charpoly/charpoly.tex
Minor improvements and biblio

--- 2017-01-18 charpoly/charpoly.tex
@@
 \subsection{The theory of p-adic precision}
+
+We recall some of the definitions and results of \cite{CRV:2014}
+as a foundation for our discussion of the precision for the characteristic polynomial
+of a matrix.  We will be concerned with two $K$-manifolds in what follows:
+the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
+$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
+the most general kind of precision structure we may attach to $M$ is a
+\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
+arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
+frequently work with certain classes of lattices, either \emph{jagged} lattices
+where we specify a precision for each matrix entry or \emph{flat} lattices where
+every entry is known to a fixed precision $O(p^m)$.  Similarly, precision for
+monic polynomials can be specified by giving a lattice in the tangent space
+at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
+of simplicity.
+
+The main benefit of working with lattices is the following result
+\citelist{\cite{CRV:2014}*{Lem. 3.4} \cite{CRV:2015}*{Prop. 2.2}}.
 

### 2017-01-26 charpoly/charpoly.tex

--- 2017-01-25 charpoly/charpoly.tex
@@
 of simplicity.
+
+The main benefit of working with lattices is the following result
+\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.
 
@@
 of flat precision. Hence the optimality of the bound given, at least when 
-$N$ is large, thanks to Lemma 3.14 of \cite{CRV:2014}
+$N$ is large, thanks to Lemma 3.14 of \cite{caruso-roe-vaccon:14a}
 

### 2017-01-27 charpoly/charpoly.tex
Small reorganization of the paper of char. poly.

@@
 
-\section{Computation of the characteristic polynomial and its derivative}
+\begin{prop} \label{prop:factor_com}
+Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
+We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
+(in $K[X]$).
+Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
+both in $O_K[X]^n,$ we have: 
+\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
+\end{prop}
+\begin{proof}
+We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
+\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
+Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
+Then $A(\lambda)$ is of rank one.
+Indeed, $A(\lambda) \times (\lambda-M) =0.$
+Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
+and then $A(\lambda)$ is of rank at most one.
+Nevertheless, we can remark that
+ $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
+and thus, $A(\lambda)$ is of rank exactly one.
+As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
+It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
+By the Chinese Remainder Theorem, our first result is then proved.
+
+Now, let us take $U=col(A,1)$ the first column of $A$ and
+$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
+invertible $\mod \chi (M)).$
+Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
+\end{proof}
+
+\begin{rem}
+If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
+we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
+$P\in GL_n (O_K)$ then
+\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
+In other words, factorizability of $\com(X-M)$ is stable
+by change of basis.
+\end{rem}
+
+\section{Optimal precision\\via Hessenberg form}
+%\section{Computation of the characteristic polynomial and its derivative}
 
@@
 
-%\subsection{$\com (X -M)$ and jagged precision}
-
-\subsection{Factorization}
+\subsection{Computation and optimal jagged precision}
 
@@
 are not easy, as $P^{-1}  dM P$ is not jagged anymore.
-Fortunately, a factorization result of $\com (X-M)$ will be enough
-to conclude that optimal precision can still be read in $\softO (n^3)$
+Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
+conclude that optimal precision can still be read in $\softO (n^3)$ 
 operations in $O_K.$
 
-\begin{prop} \label{prop:factor_com}
-Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
-We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
-(in $K[X]$).
-Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
-both in $O_K[X]^n,$ we have: 
-\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
-\end{prop}
-\begin{proof}
-We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
-\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
-Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
-Then $A(\lambda)$ is of rank one.
-Indeed, $A(\lambda) \times (\lambda-M) =0.$
-Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
-and then $A(\lambda)$ is of rank at most one.
-Nevertheless, we can remark that
- $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
-and thus, $A(\lambda)$ is of rank exactly one.
-As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
-It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
-By the Chinese Remainder Theorem, our first result is then proved.
-
-Now, let us take $U=col(A,1)$ the first column of $A$ and
-$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
-invertible $\mod \chi (M)).$
-Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
-\end{proof}
-
-\begin{rem}
-If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
-we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
-$P\in GL_n (O_K)$ then
-\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
-In other words, factorizability of $\com(X-M)$ is stable
-by change of basis.
-\end{rem}
-
-\subsection{Computation and optimal jagged precision}
-
-Thanks to the previous results, we can now compute 
-$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
-using the following algorithm.
+%Thanks to the previous results, we can now compute 
+%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
+%using the following algorithm.
 
@@
 
-\section{The case of diagonalizable matrices}
+\section{Optimal precision\\via Frobenius form}
+
+\section{Eigenvalues}
+
+\subsection{Optimal precision on a simple eigenvalue}
+
+\todo{Optimal precision on $P(\lambda_i)$ and on $\lambda_i$}
 
@@
 
-\subsection{Algorithm}
+%\subsection{Algorithm}
 

### 2017-01-30 charpoly/charpoly.tex
Short form of Com(X-M)

@@
 \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
-Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi(M)$ is given
+Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi_M$ is given
 by
 \[
-d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
+d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
 \]
-Moreover, $d\chi(M)$ will be surjective exactly when the Jordan blocks of $M$ all have distinct generalized eigenvalues.
+%Moreover, $d\chi(M)$ will be surjective exactly when
 
-\todo{Formula with comatrix.}
+%\subsection{Properties of the image of $d \chi$}
 
-\subsection{Properties of the image of $d \chi$}
+\begin{prop}
+\label{prop:surjectivity}
+For $M \in M_n(K)$, the following conditions are equivalent:
+\begin{enumerate}[(i)]
+\renewcommand{\itemsep}{0pt}
+\item the differential $d\chi(M)$ is surjective
+\item the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
+to a companion matrix)
+\item the eigenspaces of $M$ associated to eigenvalues in an algebraic
+closure of $\bar K$ have all dimension $1$
+%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
+\end{enumerate}
+\end{prop}
 
-\todo{Surjectivity. Stability under multiplication by $X$ mod $\chi$.}
+By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
+However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
+to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
+often be convenient to view $d \chi_M$ as an $K$-linear mapping
+$M_n(K) \to K[X]/\chi_M(X)$.
+
+\begin{prop}
+Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
+which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
+(considered as a submodule of $K[X]/\chi_M(X)$) is  stable by 
+multiplication by the elements of $A$.
+\end{prop}
+
+\begin{proof}
+\todo{The image is generated by the entries of $C = \com(X-M)$
+and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$.}
+\end{proof}
+
+\begin{cor}
+If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
+is stable by multiplication by $X$ and hence is a module over $\OK[X]$.
+\end{cor}
 
@@
 
+\begin{comment}
 \begin{prop} \label{prop:factor_com}
@@
 \[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
+\todo{I think that the correct condition is $d \chi_M$ surjective (which
+is a bit weaker).}
 \end{prop}
+
 \begin{proof}
@@
 \end{rem}
+\end{comment}
 
-\section{Optimal precision\\via Hessenberg form}
+\section{Differential\\via Hessenberg form}
 %\section{Computation of the characteristic polynomial and its derivative}
@@
 
-\section{Optimal precision\\via Frobenius form}
+\section{Differential\\via Frobenius form}
+
+The algorithm designed in the previous section computes the differential 
+$d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
+$O(n^3)$ operations in $K$. This seems to be optimal given that the 
+(naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n
+\times n^2$. It turns out however that improvements are still possible!
+
+Precisely, we will prove in this section that $d \chi_M$ has a very 
+special form, allowing a compact encoding using only $O(n^2)$ coefficients. 
+We shall moreover design fast algorithms (with complexity $\softO
+(n^\omega)$) for computing it. The price to pay is that divisions in
+$K$ appear, which can be an issue regarding to precision in particular
+cases.
+
+From now on, we fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
+surjective.
+
+\subsection{Compact form of $d \chi_M$}
+
+Let $C$ be the companion matrix associated to $\chi_M$:
+$$N = \left( \begin{matrix}
+0 & 1 & 0 & \cdots & 0 \\
+\vdots & \ddots & 1 & \ddots & \vdots \\
+\vdots & & \ddots & \ddots & 0 \\
+0 & \cdots & \cdots & 0 & 1 \\
+-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
+\end{matrix} \right)$$
+with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
+By Proposition~\ref{prop:surjectivity}, there exists a matrix 
+$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
+result to the transpose of $M$, we find that there exists another
+invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.
+
+\begin{prop}
+\label{prop:shortcom}
+We keep the previous notations and assumptions.
+Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
+$$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
+\mod \chi_M$$
+for some $\alpha \in K[X]$.
+\end{prop}
+
+\begin{proof}
+Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
+\pmod{\chi_M}$, we deduce $(X{-}N) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
+Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}N$
+modulo $\chi_M$. On the other hand, a direct computation shows that
+every column vector $W$ lying in the right kernel of $X{-}N$ modulo 
+$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
+K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
+for some row vector $B$.
+Applying the same reasoning with $M^t$, we find that $B$ can be
+written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
+we are done.
+\end{proof}
+
+Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
+by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
+stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
+determined by $2n$ coefficients while we need $2n^2$ coefficients to 
+write down the matrices $P$ and $Q$. 
+We shall see moreover in \S \ref{sec:eigenvalues} that interesting
+informations can be read off on this short form $(\alpha, P, Q, 
+\chi_M)$.
+
+\subsection{Fast computation}
+
+\subsection{Optimal jagged precision}
 
 \section{Eigenvalues}
+\label{sec:eigenvalues}
 

### 2017-01-30 charpoly/charpoly.tex
Computation in O~(n^omega) of Com(X-M)

@@
 \newcommand{\softO}{O\tilde{~}}
+
+\newcommand{\inv}{\text{\rm inv}}
 
@@
 Let $C$ be the companion matrix associated to $\chi_M$:
-$$N = \left( \begin{matrix}
+\begin{equation}
+\label{eq:companion}
+N = \left( \begin{matrix}
 0 & 1 & 0 & \cdots & 0 \\
@@
 -a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
-\end{matrix} \right)$$
+\end{matrix} \right)
+\end{equation}
 with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
@@
 Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
-$$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
-\mod \chi_M$$
+\begin{equation}
+\label{eq:shortcom}
+\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
+\mod \chi_M
+\end{equation}
 for some $\alpha \in K[X]$.
@@
 
+We now move to algorithms and show that the quadruple $(\alpha, P, Q, 
+\chi_M)$ can be computed for a cost of $\softO(n^\omega)$ operations
+in the ground field $K$ with high probability.
+
+The matrix $P$ can be computed as follows. Pick $c \in K^n$. Define 
+$c_i = M^i c$ for all $i \geq 1$. The $c_i$'s can be computed in 
+$\softO(n^\omega)$ operations in $K$ using \cite{}. Let $P_\inv$ be the 
+$n \times n$ matrix whose rows are the $c_i$'s for $1 \leq i \leq n$. 
+Remark that $P_\inv$ is invertible if and only if $(c_0, c_1, \ldots, 
+c_{n-1})$ is a basis of $K^n$ if and only if $c$ is a cyclic vector. 
+Moreover after base change to the basis $(c_0, \ldots, c_{n-1})$, the matrix 
+$M$ takes the shape \eqref{eq:companion}. In other words, if $P_\inv$
+is invertible, then $P = P_\inv^{-1}$ is a solution of $M = P N P^{-1}$.
+Observe moreover that the condition ``$P_\inv$ is invertible'' is open
+for the Zariski topology. It then happens with high probability as soon
+as it is not empty, that is as soon as $M$ admits a cyclic vector, which
+holds by assumption.
+
+The characteristic polynomial $\chi_M$ can be recovered thanks to the
+relation $(a_0, a_1, \ldots, a_{n-1}) = -c_n \cdot P$.
+
+Now, instead of computing directly $Q$, we first compute a matrix $R$ 
+with the property that $N^t = R N R^{-1}$. For that, we apply the same
+strategy as above except that we start with the vector $e = (1, 0, 
+\ldots, 0)$ (and not with a random vector). A simple computation shows
+that, for $1 \leq i \leq n{-}1$, the vector $N^i e$ has the shape:
+$$N^i e = (0, \ldots, 0, -a_0, \star, \ldots, \star)$$
+with $n{-}i$ starting zeros. Therefore the $N^i e$'s form a basis of
+$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $N$. Once $R$ has
+been compute, we recover $Q$ using the relation $Q = P_\inv^t R$.
+
+It remains to compute the scaling factor $\alpha$. For this, we write
+the relation:
+\begin{equation}
+\label{eq:shortcomN}
+\com(X{-}N) = \alpha \cdot V^t \cdot V R^t \mod \chi_M
+\end{equation}
+which comes from Eq.~\eqref{eq:shortcom} after multiplication on the 
+left by $P^{-1}$ and multiplication on the right by $P$. We observe
+moreover that the first row of $R$ is $(1, 0, \ldots, 0)$. Evaluating
+the top left entry of Eq.~\eqref{eq:shortcomN}, we end up with the 
+relation:
+$$\alpha = a_1 + a_2 X + \cdots + a_{n-1} X^{n-2} + X^{n-1}.$$
+No further computation are then needed to derive the value of $\alpha$.
+
 \subsection{Optimal jagged precision}

### 2017-01-30 charpoly/charpoly.tex
Working on section 2

--- 2017-01-27 charpoly/charpoly.tex
@@
 
-\subsection{Notation} Throughout the paper, $K$ will refer to a complete,
-discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer.
-
 \section{Theoretical study}
@@
 
-The main benefit of working with lattices is the following result
-\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.
-
-Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.  For a matrix $M \in M_n(K)$
-we identify the tangent space $V$ at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
-the space $K_{<n}[X]$ of polynomials of degree less than $n$.  Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed)
-ball of radius $r$ in $V$.  
-Our analysis of the $\chi$ begins with the following result
-\citelist{\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} \cite{caruso-roe-vaccon:15a}*{Prop. 2.2}}.
+Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
+Our analysis of the precision behavior of $\chi$ rests upon
+the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
+For a matrix $M \in M_n(K)$ we identify the tangent space $V$
+at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
+the space $K_{<n}[X]$ of polynomials of degree less than $n$.
+Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
+we have $\com(M) = \det(M) M^{-1}$).  Recall
+\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
+that $d\chi(M)$ is given by
+\begin{equation} \label{eq:dchi}
+d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
+\end{equation}
+Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.
 
 \begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$ and that $d\chi(M): V \to W$ is surjective.
-Let $C$ be a positive real number such that
-$B_W(1) \subset d\chi(M)(B_V(C))$.  Then, for all $\rho \in (0, 1]$ and all
+Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
+let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
 $r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
@@
 
-\subsection{Computation of the differential}
+\begin{proof}
+We first show that the distinct generalized eigenvalues condition
+guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
+the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
+so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
+matrices with a single $1$ entry and all other entries $0$, it suffices to show
+that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
+have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
+$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
+are those on or below the diagonal within the Jordan blocks, and within the $i$th
+block, each entry along the $j$th subdiagonal is
+$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
+As long as the $\lambda_i$ are distinct, these form a basis for $W$.
 
-In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
-\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
-Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi(M)$ is given
-by
-\[
-d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
-\]
-Moreover, $d\chi(M)$ will be surjective exactly when the Jordan blocks of $M$ all have distinct generalized eigenvalues.
+Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
+of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
+Since the precision polygon lies below the Newton polygon \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
+and since the smallest entry of the Newton polygon occurs either with the leading
+coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(C))$.
 
-\todo{Formula with comatrix.}
+Since the coefficients of $\chi(M)$ are given by polynomials in the entries of $M$
+with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
+the conclusion.
+\end{proof}
 
@@
 
-\todo{Surjectivity. Stability under multiplication by $X$ mod $\chi$.}
+Consider the isomorphism $W \to K[X] / \chi(X)$ induced by inclusion.
+\begin{proposition}
+Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
+of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
+\end{proposition}
+\begin{proof}
+\todo{Finish proof}
+\end{proof}
+
+We can also give a criterion on $M$ that implies 
+\begin{proposition}
+Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
+a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
+\end{proposition}
+\begin{proof}
+First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
+Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
+\end{proof}
 
@@
 
-\begin{prop} \label{prop:factor_com}
-Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
-We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
-(in $K[X]$).
-Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
-both in $O_K[X]^n,$ we have: 
-\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
-\end{prop}
-\begin{proof}
-We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
-\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
-Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
-Then $A(\lambda)$ is of rank one.
-Indeed, $A(\lambda) \times (\lambda-M) =0.$
-Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
-and then $A(\lambda)$ is of rank at most one.
-Nevertheless, we can remark that
- $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
-and thus, $A(\lambda)$ is of rank exactly one.
-As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
-It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
-By the Chinese Remainder Theorem, our first result is then proved.
-
-Now, let us take $U=col(A,1)$ the first column of $A$ and
-$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
-invertible $\mod \chi (M)).$
-Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
-\end{proof}
-
-\begin{rem}
-If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
-we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
-$P\in GL_n (O_K)$ then
-\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
-In other words, factorizability of $\com(X-M)$ is stable
-by change of basis.
-\end{rem}
-
-\section{Optimal precision\\via Hessenberg form}
-%\section{Computation of the characteristic polynomial and its derivative}
+\section{Computation of the characteristic polynomial and its derivative}
 
@@
 of flat precision. Hence the optimality of the bound given, at least when 
-$N$ is large, thanks to Lemma 3.14 of \cite{caruso-roe-vaccon:14a}
+$N$ is large, thanks to Lemma 3.14 of \cite{CRV:2014}
 
@@
 
-\subsection{Computation and optimal jagged precision}
+%\subsection{$\com (X -M)$ and jagged precision}
+
+\subsection{Factorization}
 
@@
 are not easy, as $P^{-1}  dM P$ is not jagged anymore.
-Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
-conclude that optimal precision can still be read in $\softO (n^3)$ 
+Fortunately, a factorization result of $\com (X-M)$ will be enough
+to conclude that optimal precision can still be read in $\softO (n^3)$
 operations in $O_K.$
 
-%Thanks to the previous results, we can now compute 
-%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
-%using the following algorithm.
+\begin{prop} \label{prop:factor_com}
+Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
+We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
+(in $K[X]$).
+Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
+both in $O_K[X]^n,$ we have: 
+\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
+\end{prop}
+\begin{proof}
+We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
+\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
+Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
+Then $A(\lambda)$ is of rank one.
+Indeed, $A(\lambda) \times (\lambda-M) =0.$
+Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
+and then $A(\lambda)$ is of rank at most one.
+Nevertheless, we can remark that
+ $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
+and thus, $A(\lambda)$ is of rank exactly one.
+As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
+It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
+By the Chinese Remainder Theorem, our first result is then proved.
+
+Now, let us take $U=col(A,1)$ the first column of $A$ and
+$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
+invertible $\mod \chi (M)).$
+Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
+\end{proof}
+
+\begin{rem}
+If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
+we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
+$P\in GL_n (O_K)$ then
+\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
+In other words, factorizability of $\com(X-M)$ is stable
+by change of basis.
+\end{rem}
+
+\subsection{Computation and optimal jagged precision}
+
+Thanks to the previous results, we can now compute 
+$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
+using the following algorithm.
 
@@
 
-\section{Optimal precision\\via Frobenius form}
-
-\section{Eigenvalues}
-
-\subsection{Optimal precision on a simple eigenvalue}
-
-\todo{Optimal precision on $P(\lambda_i)$ and on $\lambda_i$}
+\section{The case of diagonalizable matrices}
 
@@
 
-%\subsection{Algorithm}
+\subsection{Algorithm}
 

### 2017-01-30 charpoly/charpoly.tex

--- 2017-01-30 charpoly/charpoly.tex
@@
 
-\subsection{Computation of the differential}
+\begin{proof}
+We first show that the distinct generalized eigenvalues condition
+guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
+the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
+so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
+matrices with a single $1$ entry and all other entries $0$, it suffices to show
+that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
+have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
+$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
+are those on or below the diagonal within the Jordan blocks, and within the $i$th
+block, each entry along the $j$th subdiagonal is
+$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
+As long as the $\lambda_i$ are distinct, these form a basis for $W$.
 
-In light of Proposition \ref{prop:mainlem}, we recall the computation of $d\chi$
-\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}.
-Let $\com(M)$ denote the comatrix of $M$; when $M \in \GL_n(K)$ then $\com(M) = \det(M) M^{-1}$.  Then $d\chi_M$ is given
-by
-\[
-d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
-\]
-%Moreover, $d\chi(M)$ will be surjective exactly when
+Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
+of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
+Since the precision polygon lies below the Newton polygon \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
+and since the smallest entry of the Newton polygon occurs either with the leading
+coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(C))$.
 
-%\subsection{Properties of the image of $d \chi$}
+Since the coefficients of $\chi(M)$ are given by polynomials in the entries of $M$
+with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
+the conclusion.
+\end{proof}
 
@@
 \end{prop}
-
 \begin{proof}
@@
 
+\begin{prop}
+Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
+of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
+\end{prop}
+\begin{proof}
+\todo{Finish proof}
+\end{proof}
+
+We can also give a criterion on $M$ that allows an easy description of the image of flat
+precision lattices under $d\chi(M)$.
+\begin{prop}
+Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
+a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
+\end{prop}
+\begin{proof}
+First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
+Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
+Surjectivity of differential
+can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
+in Frobenius normal form it has only one block.
+\end{proof}
+
 \todo{Image modulo $p$ and elementary divisors. 
 Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
-
-Equality of lattices holds iff it does mod p, so need surjectivity modulo p.  Surjectivity of differential
-can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
-in Frobenius normal form it has only one block.
 

### 2017-01-30 charpoly/charpoly.tex
Move 4.1 to Section 2

@@
 \todo{The image is generated by the entries of $C = \com(X-M)$
-and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$.}
+and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
+are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
 \end{proof}
@@
 \end{comment}
+
+\subsection{Compact form of $d \chi_M$}
+
+Let $C$ be the companion matrix associated to $\chi_M$:
+\begin{equation}
+\label{eq:companion}
+N = \left( \begin{matrix}
+0 & 1 & 0 & \cdots & 0 \\
+\vdots & \ddots & 1 & \ddots & \vdots \\
+\vdots & & \ddots & \ddots & 0 \\
+0 & \cdots & \cdots & 0 & 1 \\
+-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
+\end{matrix} \right)
+\end{equation}
+with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
+By Proposition~\ref{prop:surjectivity}, there exists a matrix 
+$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
+result to the transpose of $M$, we find that there exists another
+invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.
+
+\begin{prop}
+\label{prop:shortcom}
+We keep the previous notations and assumptions.
+Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
+\begin{equation}
+\label{eq:shortcom}
+\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
+\mod \chi_M
+\end{equation}
+for some $\alpha \in K[X]$.
+\end{prop}
+
+\begin{proof}
+Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
+\pmod{\chi_M}$, we deduce $(X{-}N) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
+Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}N$
+modulo $\chi_M$. On the other hand, a direct computation shows that
+every column vector $W$ lying in the right kernel of $X{-}N$ modulo 
+$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
+K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
+for some row vector $B$.
+Applying the same reasoning with $M^t$, we find that $B$ can be
+written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
+we are done.
+\end{proof}
+
+Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
+by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
+stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
+determined by $2n$ coefficients while we need $2n^2$ coefficients to 
+write down the matrices $P$ and $Q$. 
+We shall see moreover in \S \ref{sec:eigenvalues} that interesting
+informations can be read off on this short form $(\alpha, P, Q, 
+\chi_M)$.
 
@@
 
-\subsection{Compact form of $d \chi_M$}
-
-Let $C$ be the companion matrix associated to $\chi_M$:
-\begin{equation}
-\label{eq:companion}
-N = \left( \begin{matrix}
-0 & 1 & 0 & \cdots & 0 \\
-\vdots & \ddots & 1 & \ddots & \vdots \\
-\vdots & & \ddots & \ddots & 0 \\
-0 & \cdots & \cdots & 0 & 1 \\
--a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
-\end{matrix} \right)
-\end{equation}
-with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
-By Proposition~\ref{prop:surjectivity}, there exists a matrix 
-$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
-result to the transpose of $M$, we find that there exists another
-invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.
-
-\begin{prop}
-\label{prop:shortcom}
-We keep the previous notations and assumptions.
-Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
-\begin{equation}
-\label{eq:shortcom}
-\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
-\mod \chi_M
-\end{equation}
-for some $\alpha \in K[X]$.
-\end{prop}
-
-\begin{proof}
-Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
-\pmod{\chi_M}$, we deduce $(X{-}N) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
-Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}N$
-modulo $\chi_M$. On the other hand, a direct computation shows that
-every column vector $W$ lying in the right kernel of $X{-}N$ modulo 
-$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
-K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
-for some row vector $B$.
-Applying the same reasoning with $M^t$, we find that $B$ can be
-written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
-we are done.
-\end{proof}
-
-Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
-by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
-stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
-determined by $2n$ coefficients while we need $2n^2$ coefficients to 
-write down the matrices $P$ and $Q$. 
-We shall see moreover in \S \ref{sec:eigenvalues} that interesting
-informations can be read off on this short form $(\alpha, P, Q, 
-\chi_M)$.
-
 \subsection{Fast computation}

### 2017-01-31 charpoly/charpoly.tex
More on statistics

@@
 
-\subsection{$\com (X -H)$ and flat precision}
+\subsection{The comatrix of $X{-}H$}
 
@@
 
-\begin{prop} \label{prop:optimal_flat}
-If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
-we can compute the optimal jagged precision on $\chi (M)$
-in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
-\end{prop}
-\begin{proof}
-Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
-$O(n^3)$ with no division.
-Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
-Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
-thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
-$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
-and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
-\end{proof}
+\subsection{The comatrix of $X{-}M$}
 
-\subsection{Computation and optimal jagged precision}
-
-When this time $dM$ is jagged, the trick of the proof of Proposition
-\ref{prop:optimal_flat} can not work anymore.
-Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
-are not easy, as $P^{-1}  dM P$ is not jagged anymore.
-Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
-conclude that optimal precision can still be read in $\softO (n^3)$ 
-operations in $O_K.$
+\todo{Add a small intro}
 
@@
 
-\noindent {\bf Algorithm 3:} {\tt Approximate $\com (X -M)$ }
+\noindent {\bf Algorithm 3:} {\tt Approximate $\com (X{-}M)$ }
 
@@
 \end{proof}
-
-\begin{rem}
-If $dM$ is jagged, we can read the optimal
-jagged precision from $\com (X-M).$
-Indeed, we can write $\com (X-M)=\sum_{i=1}^{n-1} A_i X^i$
-for some matrices $A_i \in M_n(O_K).$
-Then the best jagged precision on the coefficient of 
-$X^i$ can be read from $tr (A_i dM).$
-What then is to do is to find the smallest 
-$val((A_i)_{j,k} dM_{k,j}).$
-\end{rem}
-
-\begin{rem}
-As a consequence of the previous remark, once the optimal jagged
-precision is known, it is possible to lift arbitrarily
-the precision on $M$ and then to perform
-a computation of the characteristic polynomial
-from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
-in $O(n^3).$
-By truncating afterward at the optimal jagged precision,
-we can achieve by this way this optimal precision.
-\end{rem}
 
@@
 
-\subsection{Optimal jagged precision}
+\section{Optimal jagged precision}
 
-\section{Eigenvalues}
-\label{sec:eigenvalues}
+\subsection{On the characteristic polynomial}
 
-\subsection{Optimal precision on a simple eigenvalue}
+\todo{Reorganize...}
+
+\begin{prop} \label{prop:optimal_flat}
+If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
+we can compute the optimal jagged precision on $\chi (M)$
+in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
+\end{prop}
+\begin{proof}
+Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
+$O(n^3)$ with no division.
+Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
+Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
+thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
+$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
+and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
+\end{proof}
+
+When this time $dM$ is jagged, the trick of the proof of Proposition
+\ref{prop:optimal_flat} can not work anymore.
+Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
+are not easy, as $P^{-1}  dM P$ is not jagged anymore.
+Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
+conclude that optimal precision can still be read in $\softO (n^3)$ 
+operations in $O_K.$
+
+\begin{rem}
+If $dM$ is jagged, we can read the optimal
+jagged precision from $\com (X-M).$
+Indeed, we can write $\com (X-M)=\sum_{i=1}^{n-1} A_i X^i$
+for some matrices $A_i \in M_n(O_K).$
+Then the best jagged precision on the coefficient of 
+$X^i$ can be read from $tr (A_i dM).$
+What then is to do is to find the smallest 
+$val((A_i)_{j,k} dM_{k,j}).$
+\end{rem}
+
+\begin{rem}
+As a consequence of the previous remark, once the optimal jagged
+precision is known, it is possible to lift arbitrarily
+the precision on $M$ and then to perform
+a computation of the characteristic polynomial
+from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
+in $O(n^3).$
+By truncating afterward at the optimal jagged precision,
+we can achieve by this way this optimal precision.
+\end{rem}
+
+{\bf Numerical experiments}
+
+\begin{figure}
+\hfill
+{\renewcommand*{\arraystretch}{1.3}
+\begin{tabular}{|c|r|r|r|}
+\cline{2-4} 
+\multicolumn{1}{l|}{} 
+  & \multicolumn{3}{c|}{Average loss of accuracy} \\
+\cline{2-4}
+\multicolumn{1}{l|}{\null\hspace{6mm}\null} 
+  & \makebox[1.5cm]{\hfill Optimal\hfill\null}
+  & \makebox[1.5cm]{\hfill CR\hfill\null}
+  & \makebox[1.5cm]{\hfill FP\hfill\null} \\
+\hline
+\rule{0pt}{2.7ex}%
+$X^0$ & $3.17$ & $196\phantom{.00}$ & $189\phantom{.00}$ \vspace{-1.5ex} \\
+{\scriptsize (det.)} 
+& {\scriptsize dev: $1.76$} & {\scriptsize dev: $240$} & {\scriptsize dev: $226$} \\
+\rule{0pt}{2.7ex}%
+$X^1$ & $2.98$ & $161\phantom{.00}$ & $156\phantom{.00}$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.69$} & {\scriptsize dev: $204$} & {\scriptsize dev: $195$} \\
+\rule{0pt}{2.7ex}%
+$X^2$ & $2.75$ & $129\phantom{.00}$ & $126\phantom{.00}$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.57$} & {\scriptsize dev: $164$} & {\scriptsize dev: $164$} \\
+\rule{0pt}{2.7ex}%
+$X^3$ & $2.74$ & $108\phantom{.00}$ & $105\phantom{.00}$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.73$} & {\scriptsize dev: $144$} & {\scriptsize dev: $143$} \\
+\rule{0pt}{2.7ex}%
+$X^4$ & $2.57$ &  $63.2\phantom{0}$ &  $60.6\phantom{0}$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.70$} & {\scriptsize dev: $85.9$} & {\scriptsize dev: $85.8$} \\
+\rule{0pt}{2.7ex}%
+$X^5$ & $2.29$ &  $51.6\phantom{0}$ &  $49.7\phantom{0}$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.66$} & {\scriptsize dev: $75.3$} & {\scriptsize dev: $74.9$} \\
+\rule{0pt}{2.7ex}%
+$X^6$ & $2.07$ &   $9.04$           &   $8.59$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.70$} & {\scriptsize dev: $26.9$} & {\scriptsize dev: $26.4$} \\
+\rule{0pt}{2.7ex}%
+$X^7$ & $1.64$ &   $5.70$           &   $5.38$ \vspace{-1.5ex} \\
+& {\scriptsize dev: $1.65$} & {\scriptsize dev: $15.3$} & {\scriptsize dev: $14.7$} \\
+\rule{0pt}{2.7ex}%
+$X^8$ & $0.99$ &   $0.99$           &   $0.99$ \vspace{-1.5ex} \\
+{\scriptsize (trace)} 
+& {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} \\
+\hline
+\end{tabular}}
+\hfill\null
+
+\medskip
+
+\hfill
+{\footnotesize Results for a sample of $1000$ instances}
+\hfill\null
+
+\caption{Average loss of accuracy on the coefficients
+of the characteristic polynomial of a random $9 \times 9$
+matrix over $\Q_2$}
+
+\end{figure}
+
+\subsection{On eigenvalues}
+\label{ssec:eigenvalues}
 

### 2017-02-01 charpoly/charpoly.tex
Working on the criteria for dchi to be surjective

--- 2017-01-30 charpoly/charpoly.tex
@@
 \end{equation}
+
+\begin{prop}
+\label{prop:surjectivity}
+For $M \in M_n(K)$, the following conditions are equivalent:
+\begin{enumerate}[(i)]
+\renewcommand{\itemsep}{0pt}
+\item \label{pt:surj} the differential $d\chi(M)$ is surjective
+\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
+to a companion matrix)
+\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
+closure $\bar{K}$ of $K$ all have dimension $1$
+%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
+\end{enumerate}
+\end{prop}
+
+\begin{proof}
+The equivalence of \eqref{pt:cyc} and \eqref{pt:1d} is standard; see
+\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1} for example.  We now show
+that \eqref{pt:surj} and \eqref{pt:cyc} are equivalent.
+
+For any $A \in \GL_n(K)$, the image of $d\chi(M)$ will be the same
+as the image of $d\chi(AMA^{-1})$, so we may assume that
+$M$ is a companion matrix.  For a companion matrix, the bottom row of
+$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi(M)$ is surjective.
+
+To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
+After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
+will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
+for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
+$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi(M)$ will not be surjective.
+\end{proof}
+
 Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.
@@
 \begin{proof}
-We first show that the distinct generalized eigenvalues condition
-guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
-the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
-so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
-matrices with a single $1$ entry and all other entries $0$, it suffices to show
-that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
-have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
-$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
-are those on or below the diagonal within the Jordan blocks, and within the $i$th
-block, each entry along the $j$th subdiagonal is
-$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
-As long as the $\lambda_i$ are distinct, these form a basis for $W$.
 
@@
 
-\begin{prop}
-\label{prop:surjectivity}
-For $M \in M_n(K)$, the following conditions are equivalent:
-\begin{enumerate}[(i)]
-\renewcommand{\itemsep}{0pt}
-\item the differential $d\chi(M)$ is surjective
-\item the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
-to a companion matrix)
-\item the eigenspaces of $M$ associated to eigenvalues in an algebraic
-closure of $\bar K$ have all dimension $1$
-%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
-\end{enumerate}
-\end{prop}
-
 By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
@@
 which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
-(considered as a submodule of $K[X]/\chi_M(X)$) is stable by 
+(considered as a submodule of $K[X]/\chi_M(X)$) is stable under 
 multiplication by the elements of $A$.
@@
 If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
-is stable by multiplication by $X$ and hence is a module over $\OK[X]$.
+is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
 \end{cor}
-
-\begin{prop}
-Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
-of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
-\end{prop}
-\begin{proof}
-\todo{Finish proof}
-\end{proof}
 

### 2017-02-01 charpoly/charpoly.tex
Minor corrections on Section 3.

--- 2017-01-31 charpoly/charpoly.tex
@@
 over a complete discrete valuation field
-to compute $\com (X-M)$ and $d \chi.$
+to compute $\com (X-M)$ and $d \chi$
+in a way such that if $M \in M_n(O_K),$
+only division by invertible elements of $O_K$
+occur.
 %The computation of an Hessenberg form provides an $O(n^3)$
@@
 
-\todo{Describe algorithm (with matrices and endormophisms?). 
-Remark about precision.}
+We begin with the study of the computation of
+an approximate Hessenberg form.
 
@@
 \begin{deftn}
-Let $P \in K[X]$ be of degree $d \in \mathbb{N}_{>0}.$ 
-We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
+$n$ being fixed, let $P \in K[X]$ be of degree $d \leq n.$ 
+We define its reciprocal polynomial as $P^{rec}=X^n P \left( 1/X \right).$
 Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
@@
 in advance to be on the diagonal. This leads to a very smooth
-precision and complexity behaviour.
+precision and complexity behaviour when the entry matrix lives in 
+$M_n(O_K).$
 
@@
 
-\todo{Add a small intro}
-
-%Thanks to the previous results, we can now compute 
-%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
-%using the following algorithm.
-
 \noindent\hrulefill
@@
 \noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
-using Proposition \ref{prop:factor_com}.
+by inverting $C_{1,1} \mod \chi(M).$
 

### 2017-02-01 charpoly/charpoly.tex
Minor

@@
 \begin{deftn}
-$n$ being fixed, let $P \in K[X]$ be of degree $d \leq n.$ 
-We define its reciprocal polynomial as $P^{rec}=X^n P \left( 1/X \right).$
+Let $P \in K[X]$ be of degree $d \leq n.$ 
+We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
 Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix

### 2017-02-02 charpoly/charpoly.tex
Working on Section 2

--- 2017-02-01 charpoly/charpoly.tex
@@
 \alignauthor David Roe\\
-  \affaddr{Pittsburg University}\\
+  \affaddr{University of Pittsburg}\\
   \affaddr{\textsf{roed@pitt.edu}}
@@
 Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
-we have $\com(M) = \det(M) M^{-1}$).  Recall
+we have $\com(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
 \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
-that $d\chi(M)$ is given by
+that $d\chi_M$ is given by
 \begin{equation} \label{eq:dchi}
-d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
+d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
 \end{equation}
@@
 \renewcommand{\itemsep}{0pt}
-\item \label{pt:surj} the differential $d\chi(M)$ is surjective
+\item \label{pt:surj} the differential $d\chi_M$ is surjective
 \item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
@@
 
-For any $A \in \GL_n(K)$, the image of $d\chi(M)$ will be the same
-as the image of $d\chi(AMA^{-1})$, so we may assume that
+For any $A \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
+as the image of $d\chi$ at $AMA^{-1}$, so we may assume that
 $M$ is a companion matrix.  For a companion matrix, the bottom row of
-$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi(M)$ is surjective.
+$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
 
@@
 for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
-$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi(M)$ will not be surjective.
+$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
+\end{proof}
+
+We also have an analogue of Proposition \ref{prop:surjectivity} for integral matrices.
+
+\begin{prop}
+\label{prop:intsurj}
+For $M \in M_n(\OK)$, the following conditions are equivalent:
+\begin{enumerate}[(i)]
+\renewcommand{\itemsep}{0pt}
+\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
+\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
+\end{enumerate}
+\end{prop}
+\begin{proof}
+The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
+\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over $\OK / \pi\OK$.
 \end{proof}
@@
 \begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
+Suppose that $M \in M_n(K)$ satisfies one of the conditions of Proposition \ref{prop:surjectivity}, and
 let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
@@
 \begin{equation}
-\chi(M + H) = \chi(M) + d\chi(H).
+\chi(M + H) = \chi(M) + d\chi_M(H).
 \end{equation}
@@
 
+The relationship between precision and the images of lattices under $d\chi_M$ allows us to
+apply Proposition \ref{prop:intsurj} to determine when the precision of the characteristic polynomial
+is the minimum possible.
+
+\begin{cor}
+Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^N)$.
+Then the characteristic polynomial of $M$ has precision larger than $O(p^N)$
+if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
+\end{cor}
+
+\todo{Image modulo $p$ and elementary divisors. 
+Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
+
+\subsection{Stability under multiplication by $X$}
+
 By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
@@
 Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
-which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
-(considered as a submodule of $K[X]/\chi_M(X)$) is stable under 
-multiplication by the elements of $A$.
+which $P(M) \in M_n(\OK)$, and $V = d \chi_M \big(M_n(\OK)\big)$
+as a submodule of $K[X]/\chi_M(X)$.  Then $V$ is stable under 
+multiplication by $A$.
 \end{prop}
 \begin{proof}
-\todo{The image is generated by the entries of $C = \com(X-M)$
-and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
-are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
+Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $V$ is given by
+the $\OK$-span of the entries of $C$.   Using the fact that the product of matrix
+with its comatrix is the determinant, $(X - M) \cdot C = \chi(M)$ and thus
+$P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
+of the left hand side is precisely $P(X) \cdot V$, while the span of the entries
+of the right hand side is contained within $V$ since $P(M) \in M_n(\OK)$.
 \end{proof}
@@
 \begin{cor}
-If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
+If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
 is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
@@
 Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
-a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
+a cyclic vector.  Then $d\chi_M\big(M_n(\OK)\big) = \OK[X] / \chi(M)$.
 \end{prop}
 \begin{proof}
-First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
-Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
+It suffices to check that $d\chi_M$ is surjective modulo $\pi$.
 Surjectivity of differential
@@
 \end{proof}
-
-\todo{Image modulo $p$ and elementary divisors. 
-Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
 

### 2017-02-02 charpoly/charpoly.tex

@@
 \begin{deftn}
-Let $P \in K[X]$ be of degree $d \in \mathbb{N}_{>0}.$ 
+Let $P \in K[X]$ be of degree $d \leq n.$ 
 We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
@@
 
-\begin{prop} \label{prop:optimal_flat}
-If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
-we can compute the optimal jagged precision on $\chi (M)$
-in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
-\end{prop}
-\begin{proof}
-Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
-$O(n^3)$ with no division.
-Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
-Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
-thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
-$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
-and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
-\end{proof}
-
-\subsection{Computation and optimal jagged precision}
-
-When this time $dM$ is jagged, the trick of the proof of Proposition
-\ref{prop:optimal_flat} can not work anymore.
-Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
-are not easy, as $P^{-1}  dM P$ is not jagged anymore.
-Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
-conclude that optimal precision can still be read in $\softO (n^3)$ 
-operations in $O_K.$
-
-%Thanks to the previous results, we can now compute 
-%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
-%using the following algorithm.
+\subsection{The comatrix of $X{-}M$}
 

### 2017-02-02 charpoly/charpoly.tex
Correction of 3.3.

@@
 
-The second ingredient we need is the reciprocal polynomial.
+The second ingredient we need is reciprocal polynomials.
 We extend its definition to matrices of polynomials.
 \begin{deftn}
-Let $P \in K[X]$ be of degree $d \leq n.$ 
-We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
-Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
-whose coefficients are the reciprocal polynomials of the
-coefficients of $A.$ 
+Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
+We define the reciprocal polynomial of order $f$ of $P$ as $P^{rec,d}=X^d P \left( 1/X \right).$
+Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
+We denote by $A^{rec,d}$ the matrix
+whose coefficients are the reciprocal polynomials of order $d$ 
+of the coefficients of $A.$ 
 \end{deftn}
@@
 \begin{eqnarray*}
-com(1-XM)^{rec}=&com(X-M), \\
-(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
+com(1-XM)^{rec,n-1}=&com(X-M), \\
+(\chi_M \times 1)^{rec,n}=&(1-XM) com(1-XM).\\
 \end{eqnarray*}
 \end{lem}
+\begin{proof}
+It all comes down to the following result:
+let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
+then $det (A^{rec,1})=det(A)^{rec,d}.$
+Indeed, one can use multilinearity of the determinant on $X^d det(A(1/X))$
+to prove this result.
+It directly implies the second part of the lemma.
+As for the first part, since coefficients of the comatrix we are interested in
+are determinants of a matrix of polynomials of degree at most 1, the result
+is also clear.
+\end{proof}
 
@@
 
-\noindent 9. \ \textbf{Return} $V^{rec}, \psi^{rec}.$
+\noindent 9. \ \textbf{Return} $V^{rec,n-1}, \psi^{rec,n}.$
 

### 2017-02-03 charpoly/charpoly.tex
add \noindent

--- 2017-02-02 charpoly/charpoly.tex
@@
 \alignauthor David Roe\\
-  \affaddr{University of Pittsburg}\\
+  \affaddr{Pittsburg University}\\
   \affaddr{\textsf{roed@pitt.edu}}
@@
 Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
-we have $\com(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
+we have $\com(M) = \det(M) M^{-1}$).  Recall
 \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
-that $d\chi_M$ is given by
+that $d\chi(M)$ is given by
 \begin{equation} \label{eq:dchi}
-d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
+d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
 \end{equation}
-
-\begin{prop}
-\label{prop:surjectivity}
-For $M \in M_n(K)$, the following conditions are equivalent:
-\begin{enumerate}[(i)]
-\renewcommand{\itemsep}{0pt}
-\item \label{pt:surj} the differential $d\chi_M$ is surjective
-\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
-to a companion matrix)
-\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
-closure $\bar{K}$ of $K$ all have dimension $1$
-%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
-\end{enumerate}
-\end{prop}
-
-\begin{proof}
-The equivalence of \eqref{pt:cyc} and \eqref{pt:1d} is standard; see
-\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1} for example.  We now show
-that \eqref{pt:surj} and \eqref{pt:cyc} are equivalent.
-
-For any $A \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
-as the image of $d\chi$ at $AMA^{-1}$, so we may assume that
-$M$ is a companion matrix.  For a companion matrix, the bottom row of
-$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
-
-To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
-After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
-will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
-for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
-$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
-\end{proof}
-
-We also have an analogue of Proposition \ref{prop:surjectivity} for integral matrices.
-
-\begin{prop}
-\label{prop:intsurj}
-For $M \in M_n(\OK)$, the following conditions are equivalent:
-\begin{enumerate}[(i)]
-\renewcommand{\itemsep}{0pt}
-\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
-\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
-\end{enumerate}
-\end{prop}
-\begin{proof}
-The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
-\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over $\OK / \pi\OK$.
-\end{proof}
-
 Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.
@@
 \begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$ satisfies one of the conditions of Proposition \ref{prop:surjectivity}, and
+Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
 let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
@@
 \begin{equation}
-\chi(M + H) = \chi(M) + d\chi_M(H).
+\chi(M + H) = \chi(M) + d\chi(H).
 \end{equation}
@@
 \begin{proof}
+We first show that the distinct generalized eigenvalues condition
+guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
+the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
+so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
+matrices with a single $1$ entry and all other entries $0$, it suffices to show
+that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
+have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
+$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
+are those on or below the diagonal within the Jordan blocks, and within the $i$th
+block, each entry along the $j$th subdiagonal is
+$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
+As long as the $\lambda_i$ are distinct, these form a basis for $W$.
 
@@
 
-The relationship between precision and the images of lattices under $d\chi_M$ allows us to
-apply Proposition \ref{prop:intsurj} to determine when the precision of the characteristic polynomial
-is the minimum possible.
-
-\begin{cor}
-Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^N)$.
-Then the characteristic polynomial of $M$ has precision larger than $O(p^N)$
-if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
-\end{cor}
-
-\todo{Image modulo $p$ and elementary divisors. 
-Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
-
-\subsection{Stability under multiplication by $X$}
+\begin{prop}
+\label{prop:surjectivity}
+For $M \in M_n(K)$, the following conditions are equivalent:
+\begin{enumerate}[(i)]
+\renewcommand{\itemsep}{0pt}
+\item the differential $d\chi(M)$ is surjective
+\item the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
+to a companion matrix)
+\item the eigenspaces of $M$ associated to eigenvalues in an algebraic
+closure of $\bar K$ have all dimension $1$
+%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
+\end{enumerate}
+\end{prop}
 
@@
 Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
-which $P(M) \in M_n(\OK)$, and $V = d \chi_M \big(M_n(\OK)\big)$
-as a submodule of $K[X]/\chi_M(X)$.  Then $V$ is stable under 
-multiplication by $A$.
+which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
+(considered as a submodule of $K[X]/\chi_M(X)$) is stable by 
+multiplication by the elements of $A$.
 \end{prop}
 \begin{proof}
-Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $V$ is given by
-the $\OK$-span of the entries of $C$.   Using the fact that the product of matrix
-with its comatrix is the determinant, $(X - M) \cdot C = \chi(M)$ and thus
-$P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
-of the left hand side is precisely $P(X) \cdot V$, while the span of the entries
-of the right hand side is contained within $V$ since $P(M) \in M_n(\OK)$.
+\todo{The image is generated by the entries of $C = \com(X-M)$
+and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
+are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
 \end{proof}
@@
 \begin{cor}
-If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
-is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
+If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
+is stable by multiplication by $X$ and hence is a module over $\OK[X]$.
 \end{cor}
+
+\begin{prop}
+Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
+of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
+\end{prop}
+\begin{proof}
+\todo{Finish proof}
+\end{proof}
 
@@
 Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
-a cyclic vector.  Then $d\chi_M\big(M_n(\OK)\big) = \OK[X] / \chi(M)$.
+a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
 \end{prop}
 \begin{proof}
-It suffices to check that $d\chi_M$ is surjective modulo $\pi$.
+First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
+Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
 Surjectivity of differential
@@
 \end{proof}
+
+\todo{Image modulo $p$ and elementary divisors. 
+Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
 
@@
 over a complete discrete valuation field
-to compute $\com (X-M)$ and $d \chi$
-in a way such that if $M \in M_n(O_K),$
-only division by invertible elements of $O_K$
-occur.
+to compute $\com (X-M)$ and $d \chi.$
 %The computation of an Hessenberg form provides an $O(n^3)$
@@
 
-We begin with the study of the computation of
-an approximate Hessenberg form.
+\todo{Describe algorithm (with matrices and endormophisms?). 
+Remark about precision.}
 
@@
 \begin{deftn}
-Let $P \in K[X]$ be of degree $d \leq n.$ 
+Let $P \in K[X]$ be of degree $d \in \mathbb{N}_{>0}.$ 
 We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
@@
 in advance to be on the diagonal. This leads to a very smooth
-precision and complexity behaviour when the entry matrix lives in 
-$M_n(O_K).$
+precision and complexity behaviour.
 
@@
 
+\todo{Add a small intro}
+
+%Thanks to the previous results, we can now compute 
+%$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3),$
+%using the following algorithm.
+
 \noindent\hrulefill
@@
 \noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
-by inverting $C_{1,1} \mod \chi(M).$
+using Proposition \ref{prop:factor_com}.
 
@@
 
+\noindent
 {\bf Numerical experiments}

### 2017-02-03 charpoly/charpoly.tex
Optimal jagged for charpoly

--- 2017-02-02 charpoly/charpoly.tex
@@
 \section{Differential\\via Hessenberg form}
-%\section{Computation of the characteristic polynomial and its derivative}
+\label{sec:diffHess}
 
@@
 \section{Differential\\via Frobenius form}
+\label{sec:diffFrob}
 
@@
 \times n^2$. It turns out however that improvements are still possible!
-
-Precisely, we will prove in this section that $d \chi_M$ has a very 
-special form, allowing a compact encoding using only $O(n^2)$ coefficients. 
-We shall moreover design fast algorithms (with complexity $\softO
-(n^\omega)$) for computing it. The price to pay is that divisions in
-$K$ appear, which can be an issue regarding to precision in particular
-cases.
+Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of 
+$d \chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
+coefficients. The aim of this short section is to design a fast 
+algorithm (with complexity $\softO(n^\omega)$) for computing this short 
+form. The price to pay is that divisions in $K$ appear, which can be an 
+issue regarding to precision in particular cases.
 
 From now on, we fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
-surjective.
-
-\subsection{Fast computation}
-
-We now move to algorithms and show that the quadruple $(\alpha, P, Q, 
-\chi_M)$ can be computed for a cost of $\softO(n^\omega)$ operations
-in the ground field $K$ with high probability.
-
+surjective. Let $(\alpha, P, Q, \chi_M)$ be the quadruaple encoding
+the short form of $d \chi_M$; we recall that they are related by the
+relations:
+\begin{align*}
+d \chi_M(dM) & =\tr(\com(X{-}M) \cdot dM) \\
+\com(X{-}M) & = \alpha \cdot P V^t \cdot V Q^t \mod \chi_M.
+\end{align*}
 The matrix $P$ can be computed as follows. Pick $c \in K^n$. Define 
@@
 \section{Optimal jagged precision}
+\label{sec:optjagged}
+
+\todo{Write a small introduction}
 
@@
 
-\todo{Reorganize...}
+For $0 \leq k < n$, let $\pi_k : K[X] \to K$ be the mapping taking a 
+polynomial to its coefficients in $X^k$. By applying 
+\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} to the composite $\pi_k 
+\circ \chi_M$, one can figure out the optimal precision on the
+$k$-th coefficient of the characteristic polynomial of $M$ (at
+least if $M$ is given at enough precision).
 
-\begin{prop} \label{prop:optimal_flat}
-If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
-we can compute the optimal jagged precision on $\chi (M)$
-in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
+Let us consider more precisely the case where $M$ is given at 
+jagged precision: the $(i,j)$ entry of $M$ is given at precision 
+$O(\pi^{N_{i,j}})$ for some given integer $N_{i,j}$. 
+Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then teaches us that
+the optimal precision on the $k$-th coefficient of $\chi_M$ is 
+$O(\pi^{M_k})$ where $M_k$ is given by the formula:
+\begin{equation}
+\label{eq:optjagged}
+M_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
+\end{equation}
+where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.
+
+\begin{prop} \label{prop:optimal_jagged}
+If $M \in M_n(O_K)$ is given at (high enough) jagged precision, 
+then we can compute the optimal jagged precision on $\chi_M$ in 
+$\softO (n^3)$ operations in $K$.
 \end{prop}
+
 \begin{proof}
-Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
-$O(n^3)$ with no division.
-Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
-Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
-thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
-$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
-and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
+We have seen in \S \ref{sec:diffHess} and \S \ref{sec:diffFrob}
+that the computation of the matrix $C = \com(X{-}M)$ can be carried out 
+within $\softO(n^3)$ operations in $K$ (either with the Hessenberg 
+method or the Frobenius method). We conclude by applying 
+Eq.~\eqref{eq:optjagged} which requires no further operation in $K$
+(but $n^3$ evaluations of valuations and $n^3$ manipulations of 
+integers).
 \end{proof}
 
-When this time $dM$ is jagged, the trick of the proof of Proposition
-\ref{prop:optimal_flat} can not work anymore.
-Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
-are not easy, as $P^{-1}  dM P$ is not jagged anymore.
-Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
-conclude that optimal precision can still be read in $\softO (n^3)$ 
-operations in $O_K.$
-
 \begin{rem}
-If $dM$ is jagged, we can read the optimal
-jagged precision from $\com (X-M).$
-Indeed, we can write $\com (X-M)=\sum_{i=1}^{n-1} A_i X^i$
-for some matrices $A_i \in M_n(O_K).$
-Then the best jagged precision on the coefficient of 
-$X^i$ can be read from $tr (A_i dM).$
-What then is to do is to find the smallest 
-$val((A_i)_{j,k} dM_{k,j}).$
+If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
+then we can avoid the final base change step in the Hessenberg method.
+Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
+we can write:
+$$tr(\com (X{-}M) \cdot dM)=tr(\com (X{-}H)\cdot P^{-1} dM P)$$
+where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
+that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
+As a consequence, the integer $M_k$ given the optimal precision on the 
+$k$-th coefficient of $M$ is also equal to
+$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
+where $C^H_{i,j}$ is the $(i,j)$ entry of the comatrix of $X{-}H$
+where $H$ is the Hessenberg form of $M$.
 \end{rem}
 
-\begin{rem}
-As a consequence of the previous remark, once the optimal jagged
-precision is known, it is possible to lift arbitrarily
-the precision on $M$ and then to perform
-a computation of the characteristic polynomial
-from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
-in $O(n^3).$
-By truncating afterward at the optimal jagged precision,
-we can achieve by this way this optimal precision.
-\end{rem}
+As a consequence of the previous discussion, once the optimal jagged 
+precision is known, it is possible to lift arbitrarily the precision on 
+$M$ and then to perform a computation of the characteristic polynomial 
+at the optimal precision. This requires $\softO(n^3)$ operations in $K$ 
+but, for several instances, may require to increase a lot the precision.
 
-{\bf Numerical experiments}
+\medskip
 
+\noindent
+{\bf Numerical experiments.}
+We have made numerical experiments in \textsc{SageMath}~\cite{sage}
+in order to compare the optimal precision obtained with the methods
+explained above with the actual precision obtained by the software.
+For doing so, we picked a sample of $1000$ random matrices $M$ in 
+$M_9(\Q_2)$ where all the entries at the same \emph{relative
+precision}.
+We recall that, in \textsc{SageMath}, random elements $x \in \Q_p$ are 
+generated as follows. We fix an integer $N$ --- the so-called relative
+precision --- and generate elements of $\Q_p$ of the shape
+$$x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)$$
+where $v$ is a random integer generated according to the distribution:
+$$\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
+\mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
+n \geq 1$$
+and $a$ is a random integer in the range $[0, p^N)$ with uniform
+
+Once this sample has been generated, we computed, for each $k \in \{0, 
+1, \ldots, 8\}$, the three following quantities:
+
+\vspace{-2mm}
+
+\begin{enumerate}[$\bullet$]
+\renewcommand{\itemsep}{0pt}
+\item the optimal precision on the $k$-th coefficient of the 
+characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}
+\item in the capped relative model\footnote{Each 
+coefficient carries its own precision which is updated after each 
+elementary arithmetical operation.},
+the precision gotten on the $k$-th coefficient of the 
+characteristic polynomial of $M$ computed \emph{via} the call:
+
+\hfill$M\texttt{.charpoly(algorithm="df")}$\hfill\null
+
+\item in the model of floating-point arithmetic (see 
+\cite{caruso-coursepadic}*{\S 2.3}), the number of correct digits of the 
+$k$-th coefficient of the characteristic polynomial of $M$.
+\end{enumerate}
+
+\noindent
+The table of Figure~\ref{fig:exp} summarizes the obtaines results. 
+%
 \begin{figure}
@@
 matrix over $\Q_2$}
+\label{fig:exp}
 
 \end{figure}
+%
+It should be read as follows. First, the acronyms CR and FP 
+refers to ``capped relative'' and ``floating-point'' respectively.
+The numbers displayed in the table are the average loss of
+\emph{relative} precision. 
+More precisely, if $N$ is the relative precision at
+which the entries of the input random matrix $M$ have been generated
+and $v$ is the valuation of the $k$-th coefficient of $\chi_M$, then:
+
+\vspace{-2mm}
+
+\begin{enumerate}[$\bullet$]
+\renewcommand{\itemsep}{0pt}
+\item the column ``Optimal'' is the average of the quantities 
+$(M_k{-}v) - N$ (where $M_k$ is defined by Eq.~\eqref{eq:optjagged}): 
+$M_k{-}v$ is the optimal \emph{relative} precision, so that the
+difference $(M_k{-}v) - N$ is the loss of relative precision;
+\item the column ``CR'' is the average of the quatities 
+$(\text{CR}_k{-}v) - N$ where $\text{CR}_k$ is the computed (absolute) 
+precision on the $k$-th coefficient of $\chi_M$;
+\item the column ``FP'' is the average of the quatities 
+$(\text{FP}_k{-}v) - N$ where $\text{FP}_k$ is the first position of
+an incorrect digit on the $k$-th coefficient of $\chi_M$.
+\end{enumerate}
+
+We observe that the loss of relative accuracy stays under control in the 
+``Optimal'' column whereas it has a very erratic behavior --- very large 
+values and very large deviation as well --- in the two other columns. 
+These experiments then strongly support the interest of applying the 
+methods developed in this paper.
 

### 2017-02-03 charpoly/charpoly.tex
Plan of the introduction

@@
 
+\todo{Explain here why computing characteric polynomial is so important}
+
+\todo{Explain also why taking care of precision is important}
+
 \todo{Example of $M$ and $M + 1$}
 
-\subsection{Notation} Throughout the paper, $K$ will refer to a complete,
-discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer.
+\medskip
+
+\noindent
+{\bf Previous contributions.}
+\todo{Many \emph{ad-hoc} for Kedlaya-like counting point algorithms.
+Our previous paper. What else?}
+
+\medskip
+
+\noindent
+{\bf The contribution of this paper.}
+
+\medskip
+
+\noindent
+{\bf Organization of the article.}
+
+\medskip
+
+\noindent
+{\bf Notations.}
+Throughout the paper, $K$ will refer to a complete, discrete valuation 
+field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its 
+valuation, $\OK$ its ring of integers and $\pi$ a uniformizer.
 
@@
 \end{enumerate}
+
+\begin{rem}
+The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
+division free algorithm of \cite{}. It is likely that, proceeding so, we 
+limit the loss of precision.
+\end{rem}
 

### 2017-02-03 charpoly/charpoly.tex
Minor improvements.

@@
 with the $\varepsilon$'s being units,
-the corresponding coefficient of the corresponding shear matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
+the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\mathbb{F}_q[X],$ $\mathbb{Q}[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
 Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
@@
 $K^n$, \emph{i.e.} $e$ is always a cyclic vector of $N$. Once $R$ has
-been compute, we recover $Q$ using the relation $Q = P_\inv^t R$.
+been computed, we recover $Q$ using the relation $Q = P_\inv^t R$.
 

### 2017-02-03 charpoly/charpoly.tex
Optimal precision on eigenvalues

--- 2017-02-03 charpoly/charpoly.tex
@@
 
+\newcommand{\calU}{\mathcal{U}}
+
 \newcommand{\softO}{O\tilde{~}}
@@
 \newcommand{\inv}{\text{\rm inv}}
+\newcommand{\app}{\text{\rm app}}
 
@@
 the optimal precision on the $k$-th coefficient of $\chi_M$ is 
-$O(\pi^{M_k})$ where $M_k$ is given by the formula:
+$O(\pi^{N'_k})$ where $N'_k$ is given by the formula:
 \begin{equation}
 \label{eq:optjagged}
-M_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
+N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
 \end{equation}
@@
 we can write:
-$$tr(\com (X{-}M) \cdot dM)=tr(\com (X{-}H)\cdot P^{-1} dM P)$$
+$$\tr(\com (X{-}M) \cdot dM)=\tr(\com (X{-}H)\cdot P^{-1} dM P)$$
 where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
 that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
-As a consequence, the integer $M_k$ given the optimal precision on the 
+As a consequence, the integer $N'_k$ given the optimal precision on the 
 $k$-th coefficient of $M$ is also equal to
@@
 \item the column ``Optimal'' is the average of the quantities 
-$(M_k{-}v) - N$ (where $M_k$ is defined by Eq.~\eqref{eq:optjagged}): 
-$M_k{-}v$ is the optimal \emph{relative} precision, so that the
-difference $(M_k{-}v) - N$ is the loss of relative precision;
+$(N'_k{-}v) - N$ (where $N'_k$ is defined by Eq.~\eqref{eq:optjagged}): 
+$N'_k{-}v$ is the optimal \emph{relative} precision, so that the
+difference $(N'_k{-}v) - N$ is the loss of relative precision;
 \item the column ``CR'' is the average of the quatities 
@@
 
-\todo{Optimal precision on $P(\lambda_i)$ and on $\lambda_i$}
-
-\subsection{Legendre precision}
-
-\todo{Describe Legendre precision and show that it is optimal for
-a diagonalizable matrix over $\Zp$.}
-
-%\subsection{Algorithm}
+Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple} eigenvalue 
+of $M$. We are interesting in quantifying the optimal precision on 
+$\lambda$ when $M$ is given with some uncertainty.
+
+For doing so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
+suppose that the uncertainty of $M$ is ``jagged'' in the sense that
+each entry of $M$ is given at some precision $O(\pi^{N_{i,j}})$.
+Let $\lambda_\app$ be the relevant eigenvalue of $M_\app$. We remark 
+that it is possible to follow the eigenvalue $\lambda_\app$ on a small 
+neighborhood $\calU$ of $M$. Precisely, there exists a unique continuous 
+function $f : \calU \to K$ such that:
+\begin{enumerate}[$\bullet$]
+\renewcommand{\itemsep}{0pt}
+\item $f(M_\app) = \lambda_\app$, and
+\item $f(M)$ is an eigenvalue of $M$ for all $M \in \calU$.
+\end{enumerate}
+
+\begin{lem}
+The function $f$ is strictly differentiable on a neighborhood of 
+$M_\app$.
+The differential of $f$ at $M$ is the linear mapping:
+$$dM \mapsto d \lambda = - \frac{\tr(\com(\lambda-M) \cdot dM)}
+{\chi'_M(\lambda)}$$
+where $\chi'_M$ is the usual derivative of $\chi_M$.
+\end{lem}
+
+\begin{proof}
+The first assertion follows from the implicit function Theorem.
+Differentiating the relation $\chi_M(\lambda) = 0$, we get
+$\chi'_M(\lambda) \cdot d \lambda + \tr(\com(X-M) \cdot dM)(\lambda) = 0$,
+from what the Lemma follows.
+\end{proof}
+
+\noindent
+Lemma 3.4 of \cite{caruso-roe-vaccon:14a} now implies that, if the
+$N_{i,j}$'s are large enough and rather well balanced, the optimal
+precision on the eigenvalue $\lambda$ is $O(p^{N'})$ with:
+$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(C_{i,j}(\lambda)) - 
+\val(\chi'_M(\lambda))\big)$$
+where $C_{i,j}$ denotes as above the $(i,j)$ entry of $\com(X{-}M)$.
+Writing
+$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t \text{ mod } \chi_M$
+as in Proposition~\ref{prop:shortcom}, we find:
+\begin{align*}
+N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
+& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
+\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big)
+\end{align*}
+where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
+denotes the $j$-th row of $Q$. Note moreover that $V(\lambda)$ is
+the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
+By the discussion of \S \ref{}, the exact value of $M$ can be 
+determined for a cost of $\softO(n^\omega)$ operations in $K$ and
+$O(n^2)$ operations on integers.
+
+When $M$ is given at flat precision, \emph{i.e.} the $N_{i,j}$'s are all 
+equal to some $N$, the formula for $N'$ rewrites:
+\begin{align*}
+N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
+& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
++ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big)
+\end{align*}
+and can therefore now be evaluated for a cost of $\softO(n^\omega)$
+operations in $K$ and only $O(n)$ operations on integers.
+
+\medskip
+
+To conclude with, let us discuss briefly the situation where we want
+to figure out the optimal jagged precision on a tuple $(\lambda_1,
+\ldots, \lambda_s)$ of eigenvalues. Applying what we have done
+previously, we find that the optimal precision on $\lambda_k$ is 
+\begin{align*}
+N'_k & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
+& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
+\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big).
+\end{align*}
+
+\begin{prop}
+\label{prop:opteigenvalues}
+The $N'_k$'s can be all computed in $\softO(n^\omega)$ operations
+in $K$ and $O(n^2 s)$ operations on integers.
+
+If the $N_{i,j}$'s are all equal, the above complexity can be 
+lowered to $\softO(n^\omega)$ operations in $K$ and $O(n s)$ operations 
+on integers.
+\end{prop}
+
+\begin{proof}
+The $\alpha(\lambda_k)$'s and the $\chi'_M(\lambda_k)$'s can be computed 
+for a cost of $\softO(ns)$ operations in $K$ using fast multipoint 
+evaluation methods~\cite{}.
+On the other hand, we observe that $P_i V(\lambda_k)^t$ is nothing but
+the $(i,k)$ entry of the matrix:
+$$P \cdot \left( \begin{matrix}
+\lambda_1 & \cdots & \lambda_s \\
+\lambda_1^2 & \cdots & \lambda_s^2 \\
+\vdots & & \vdots \\
+\lambda_1^{n-1} & \cdots & \lambda_s^{n-1}
+\end{matrix} \right).$$
+The latter product can be computed in $\softO(n^\omega)$ operations in 
+$K$\footnote{It turns out that $\softO(n^2)$ is also possible because
+the right factor is a structured matrix (a truncated Vandermonde):
+computing the above product reduces to evaluate a polynomial at the
+points $\lambda_1, \ldots, \lambda_s$.}. Therefore all the $P_i 
+V(\lambda_k)^t$'s (for $i$ and $k$ varying) 
+can be determined with the same complexity. 
+Similarly all the $V(\lambda) Q_j^t$ are computed for the same cost.
+The first assertion of Proposition~\ref{prop:opteigenvalues} follows.
+The second assertion is now proved similarly to the case of a unique
+eigenvalue.
+\end{proof}
 

### 2017-02-04 charpoly/charpoly.tex
Working on Section 1

--- 2017-02-03 charpoly/charpoly.tex
@@
 
-\todo{Explain here why computing characteric polynomial is so important}
+\todo{Explain here why computing characteristic polynomial is so important.}
 
-\todo{Explain also why taking care of precision is important}
+The characteristic polynomial is a fundamental invariant of a matrix: its roots
+give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
+In fact, the best known division-free algorithm for computing determinants
+over arbitrary rings \cite{kartofen:92a} does so using the characteristic polynomial.
+Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
+algorithms for counting points of varieties over finite fields. \todo{list some point counting references.}
+\todo{Application to characteristic polynomials of elements of p-adic field extensions?}
 
-\todo{Example of $M$ and $M + 1$}
+\todo{Explain also why taking care of precision is important.}
 
-\medskip
+When computing with $p$-adic matrices, the lack of infinite memory implies
+that the entries may only be approximated at some finite precision $O(p^m)$.  As a
+consequence, in designing algorithms for such matrices one must analyze
+not only the running time of the algorithm but also the accuracy of the result.
 
-\noindent
-{\bf Previous contributions.}
-\todo{Many \emph{ad-hoc} for Kedlaya-like counting point algorithms.
-Our previous paper. What else?}
+Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
+The simplest approach for computing the characteristic polynomial of $M$
+is to compute $\det(X-M)$ either using recursive row expansion or various division free
+algorithms \cites{\cite{seifullin:02a} \cite{kartofen:92a}}.  There are two issues
+with these methods.  First, they are slower than alternatives that allow division,
+requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
+while the lack of division implies that the result is accurate modulo $p^m$ as
+long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.
 
-\medskip
+A faster approach over a field is to compute the Frobenius normal form of $M$,
+which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
+However, the fact that it uses division frequently leads to catastrophic losses of division.
+In many examples, no precision remains at the end of the calculation.
 
-\noindent
-{\bf The contribution of this paper.}
+\todo{Example of $M$ and $M + 1$.}
 
-\medskip
+\subsection*{Previous contributions.} Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?
 
-\noindent
-{\bf Organization of the article.}
+\subsection*{The contribution of this paper.}
 
-\medskip
+\subsection*{Organization of the article.}
 
-\noindent
-{\bf Notations.}
-Throughout the paper, $K$ will refer to a complete, discrete valuation 
-field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its 
-valuation, $\OK$ its ring of integers and $\pi$ a uniformizer.
+\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
+discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
+$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}
 
@@
 the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
-For a matrix $M \in M_n(K)$ we identify the tangent space $V$
+For a matrix $M \in M_n(K)$, we identify the tangent space $V$
 at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
@@
 \begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$ satisfies one of the conditions of Proposition \ref{prop:surjectivity}, and
-let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
-$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
+Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and
+let $\alpha = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
+$r \in (0, \alpha \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
 satisfies:
@@
 of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
-Since the precision polygon lies below the Newton polygon \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
+Since the precision polygon lies below the Newton polygon of $M$ \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
 and since the smallest entry of the Newton polygon occurs either with the leading
-coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(C))$.
+coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(\alpha))$.
 
@@
 \begin{cor}
-Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^N)$.
-Then the characteristic polynomial of $M$ has precision larger than $O(p^N)$
+Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^m)$.
+Then the characteristic polynomial of $M$ has precision larger than $O(p^m)$
 if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
 \end{cor}
+
+Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field,
+the running time of matrix multiplication.
 
@@
 \section{Differential\\via Hessenberg form}
-\label{sec:diffHess}
+%\section{Computation of the characteristic polynomial and its derivative}
 
@@
 with the $\varepsilon$'s being units,
-the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\mathbb{F}_q[X],$ $\mathbb{Q}[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
+the corresponding coefficient of the corresponding shear matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
 Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
@@
 
-The second ingredient we need is reciprocal polynomials.
+The second ingredient we need is the reciprocal polynomial.
 We extend its definition to matrices of polynomials.
 \begin{deftn}
-Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
-We define the reciprocal polynomial of order $f$ of $P$ as $P^{rec,d}=X^d P \left( 1/X \right).$
-Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
-We denote by $A^{rec,d}$ the matrix
-whose coefficients are the reciprocal polynomials of order $d$ 
-of the coefficients of $A.$ 
+Let $P \in K[X]$ be of degree $d \leq n.$ 
+We define its reciprocal polynomial as $P^{rec}=X^d P \left( 1/X \right).$
+Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
+whose coefficients are the reciprocal polynomials of the
+coefficients of $A.$ 
 \end{deftn}
@@
 \begin{eqnarray*}
-com(1-XM)^{rec,n-1}=&com(X-M), \\
-(\chi_M \times 1)^{rec,n}=&(1-XM) com(1-XM).\\
+com(1-XM)^{rec}=&com(X-M), \\
+(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
 \end{eqnarray*}
 \end{lem}
-\begin{proof}
-It all comes down to the following result:
-let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
-then $det (A^{rec,1})=det(A)^{rec,d}.$
-Indeed, one can use multilinearity of the determinant on $X^d det(A(1/X))$
-to prove this result.
-It directly implies the second part of the lemma.
-As for the first part, since coefficients of the comatrix we are interested in
-are determinants of a matrix of polynomials of degree at most 1, the result
-is also clear.
-\end{proof}
 
@@
 
-\noindent 9. \ \textbf{Return} $V^{rec,n-1}, \psi^{rec,n}.$
+\noindent 9. \ \textbf{Return} $V^{rec}, \psi^{rec}.$
 
@@
 \section{Differential\\via Frobenius form}
-\label{sec:diffFrob}
 
@@
 \times n^2$. It turns out however that improvements are still possible!
-Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of 
-$d \chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
-coefficients. The aim of this short section is to design a fast 
-algorithm (with complexity $\softO(n^\omega)$) for computing this short 
-form. The price to pay is that divisions in $K$ appear, which can be an 
-issue regarding to precision in particular cases.
+
+Precisely, we will prove in this section that $d \chi_M$ has a very 
+special form, allowing a compact encoding using only $O(n^2)$ coefficients. 
+We shall moreover design fast algorithms (with complexity $\softO
+(n^\omega)$) for computing it. The price to pay is that divisions in
+$K$ appear, which can be an issue regarding to precision in particular
+cases.
 
 From now on, we fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
-surjective. Let $(\alpha, P, Q, \chi_M)$ be the quadruaple encoding
-the short form of $d \chi_M$; we recall that they are related by the
-relations:
-\begin{align*}
-d \chi_M(dM) & =\tr(\com(X{-}M) \cdot dM) \\
-\com(X{-}M) & = \alpha \cdot P V^t \cdot V Q^t \mod \chi_M.
-\end{align*}
+surjective.
+
+\subsection{Fast computation}
+
+We now move to algorithms and show that the quadruple $(\alpha, P, Q, 
+\chi_M)$ can be computed for a cost of $\softO(n^\omega)$ operations
+in the ground field $K$ with high probability.
+
 The matrix $P$ can be computed as follows. Pick $c \in K^n$. Define 
@@
 $K^n$, \emph{i.e.} $e$ is always a cyclic vector of $N$. Once $R$ has
-been computed, we recover $Q$ using the relation $Q = P_\inv^t R$.
+been compute, we recover $Q$ using the relation $Q = P_\inv^t R$.
 
@@
 \section{Optimal jagged precision}
-\label{sec:optjagged}
-
-\todo{Write a small introduction}
 
@@
 
-For $0 \leq k < n$, let $\pi_k : K[X] \to K$ be the mapping taking a 
-polynomial to its coefficients in $X^k$. By applying 
-\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} to the composite $\pi_k 
-\circ \chi_M$, one can figure out the optimal precision on the
-$k$-th coefficient of the characteristic polynomial of $M$ (at
-least if $M$ is given at enough precision).
+\todo{Reorganize...}
 
-Let us consider more precisely the case where $M$ is given at 
-jagged precision: the $(i,j)$ entry of $M$ is given at precision 
-$O(\pi^{N_{i,j}})$ for some given integer $N_{i,j}$. 
-Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then teaches us that
-the optimal precision on the $k$-th coefficient of $\chi_M$ is 
-$O(\pi^{M_k})$ where $M_k$ is given by the formula:
-\begin{equation}
-\label{eq:optjagged}
-M_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
-\end{equation}
-where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.
-
-\begin{prop} \label{prop:optimal_jagged}
-If $M \in M_n(O_K)$ is given at (high enough) jagged precision, 
-then we can compute the optimal jagged precision on $\chi_M$ in 
-$\softO (n^3)$ operations in $K$.
+\begin{prop} \label{prop:optimal_flat}
+If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
+we can compute the optimal jagged precision on $\chi (M)$
+in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
 \end{prop}
-
 \begin{proof}
-We have seen in \S \ref{sec:diffHess} and \S \ref{sec:diffFrob}
-that the computation of the matrix $C = \com(X{-}M)$ can be carried out 
-within $\softO(n^3)$ operations in $K$ (either with the Hessenberg 
-method or the Frobenius method). We conclude by applying 
-Eq.~\eqref{eq:optjagged} which requires no further operation in $K$
-(but $n^3$ evaluations of valuations and $n^3$ manipulations of 
-integers).
+Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
+$O(n^3)$ with no division.
+Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
+Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
+thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
+$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
+and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
 \end{proof}
 
+When this time $dM$ is jagged, the trick of the proof of Proposition
+\ref{prop:optimal_flat} can not work anymore.
+Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
+are not easy, as $P^{-1}  dM P$ is not jagged anymore.
+Fortunately, Proposition~\ref{prop:factom_com} will be enough to 
+conclude that optimal precision can still be read in $\softO (n^3)$ 
+operations in $O_K.$
+
 \begin{rem}
-If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
-then we can avoid the final base change step in the Hessenberg method.
-Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
-we can write:
-$$tr(\com (X{-}M) \cdot dM)=tr(\com (X{-}H)\cdot P^{-1} dM P)$$
-where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
-that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
-As a consequence, the integer $M_k$ given the optimal precision on the 
-$k$-th coefficient of $M$ is also equal to
-$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
-where $C^H_{i,j}$ is the $(i,j)$ entry of the comatrix of $X{-}H$
-where $H$ is the Hessenberg form of $M$.
+If $dM$ is jagged, we can read the optimal
+jagged precision from $\com (X-M).$
+Indeed, we can write $\com (X-M)=\sum_{i=1}^{n-1} A_i X^i$
+for some matrices $A_i \in M_n(O_K).$
+Then the best jagged precision on the coefficient of 
+$X^i$ can be read from $tr (A_i dM).$
+What then is to do is to find the smallest 
+$val((A_i)_{j,k} dM_{k,j}).$
 \end{rem}
 
-As a consequence of the previous discussion, once the optimal jagged 
-precision is known, it is possible to lift arbitrarily the precision on 
-$M$ and then to perform a computation of the characteristic polynomial 
-at the optimal precision. This requires $\softO(n^3)$ operations in $K$ 
-but, for several instances, may require to increase a lot the precision.
-
-\medskip
-
-\noindent
-{\bf Numerical experiments.}
-We have made numerical experiments in \textsc{SageMath}~\cite{sage}
-in order to compare the optimal precision obtained with the methods
-explained above with the actual precision obtained by the software.
-For doing so, we picked a sample of $1000$ random matrices $M$ in 
-$M_9(\Q_2)$ where all the entries at the same \emph{relative
-precision}.
-We recall that, in \textsc{SageMath}, random elements $x \in \Q_p$ are 
-generated as follows. We fix an integer $N$ --- the so-called relative
-precision --- and generate elements of $\Q_p$ of the shape
-$$x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)$$
-where $v$ is a random integer generated according to the distribution:
-$$\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
-\mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
-n \geq 1$$
-and $a$ is a random integer in the range $[0, p^N)$ with uniform
-
-Once this sample has been generated, we computed, for each $k \in \{0, 
-1, \ldots, 8\}$, the three following quantities:
-
-\vspace{-2mm}
-
-\begin{enumerate}[$\bullet$]
-\renewcommand{\itemsep}{0pt}
-\item the optimal precision on the $k$-th coefficient of the 
-characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}
-\item in the capped relative model\footnote{Each 
-coefficient carries its own precision which is updated after each 
-elementary arithmetical operation.},
-the precision gotten on the $k$-th coefficient of the 
-characteristic polynomial of $M$ computed \emph{via} the call:
-
-\hfill$M\texttt{.charpoly(algorithm="df")}$\hfill\null
-
-\item in the model of floating-point arithmetic (see 
-\cite{caruso-coursepadic}*{\S 2.3}), the number of correct digits of the 
-$k$-th coefficient of the characteristic polynomial of $M$.
-\end{enumerate}
-
 \begin{rem}
-The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
-division free algorithm of \cite{}. It is likely that, proceeding so, we 
-limit the loss of precision.
+As a consequence of the previous remark, once the optimal jagged
+precision is known, it is possible to lift arbitrarily
+the precision on $M$ and then to perform
+a computation of the characteristic polynomial
+from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
+in $O(n^3).$
+By truncating afterward at the optimal jagged precision,
+we can achieve by this way this optimal precision.
 \end{rem}
 
-\noindent
-The table of Figure~\ref{fig:exp} summarizes the obtaines results. 
-%
+{\bf Numerical experiments}
+
 \begin{figure}
@@
 matrix over $\Q_2$}
-\label{fig:exp}
 
 \end{figure}
-%
-It should be read as follows. First, the acronyms CR and FP 
-refers to ``capped relative'' and ``floating-point'' respectively.
-The numbers displayed in the table are the average loss of
-\emph{relative} precision. 
-More precisely, if $N$ is the relative precision at
-which the entries of the input random matrix $M$ have been generated
-and $v$ is the valuation of the $k$-th coefficient of $\chi_M$, then:
-
-\vspace{-2mm}
-
-\begin{enumerate}[$\bullet$]
-\renewcommand{\itemsep}{0pt}
-\item the column ``Optimal'' is the average of the quantities 
-$(M_k{-}v) - N$ (where $M_k$ is defined by Eq.~\eqref{eq:optjagged}): 
-$M_k{-}v$ is the optimal \emph{relative} precision, so that the
-difference $(M_k{-}v) - N$ is the loss of relative precision;
-\item the column ``CR'' is the average of the quatities 
-$(\text{CR}_k{-}v) - N$ where $\text{CR}_k$ is the computed (absolute) 
-precision on the $k$-th coefficient of $\chi_M$;
-\item the column ``FP'' is the average of the quatities 
-$(\text{FP}_k{-}v) - N$ where $\text{FP}_k$ is the first position of
-an incorrect digit on the $k$-th coefficient of $\chi_M$.
-\end{enumerate}
-
-We observe that the loss of relative accuracy stays under control in the 
-``Optimal'' column whereas it has a very erratic behavior --- very large 
-values and very large deviation as well --- in the two other columns. 
-These experiments then strongly support the interest of applying the 
-methods developed in this paper.
 

### 2017-02-04 charpoly/charpoly.tex
Minor changes to section 5

--- 2017-02-03 charpoly/charpoly.tex
@@
 For doing so, we picked a sample of $1000$ random matrices $M$ in 
-$M_9(\Q_2)$ where all the entries at the same \emph{relative
+$M_9(\Q_2)$ where all the entries are given at the same \emph{relative
 precision}.
@@
 n \geq 1$$
-and $a$ is a random integer in the range $[0, p^N)$ with uniform
+and $a$ is a random integer in the range $[0, p^N)$ with 
+taken with uniform distribution
 
@@
 the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
-By the discussion of \S \ref{}, the exact value of $M$ can be 
+By the discussion of \S \ref{}, the exact value of $N'$ can be 
 determined for a cost of $\softO(n^\omega)$ operations in $K$ and
@@
 to figure out the optimal jagged precision on a tuple $(\lambda_1,
-\ldots, \lambda_s)$ of eigenvalues. Applying what we have done
+\ldots, \lambda_s)$ of simple eigenvalues. Applying what we have done
 previously, we find that the optimal precision on $\lambda_k$ is 

### 2017-02-04 charpoly/charpoly.tex

--- 2017-02-03 charpoly/charpoly.tex
@@
 
-\todo{Explain here why computing characteric polynomial is so important}
-
-\todo{Explain also why taking care of precision is important}
-
-\todo{Example of $M$ and $M + 1$}
-
-\medskip
-
-\noindent
-{\bf Previous contributions.}
-\todo{Many \emph{ad-hoc} for Kedlaya-like counting point algorithms.
-Our previous paper. What else?}
-
-\medskip
-
-\noindent
-{\bf The contribution of this paper.}
-
-\medskip
-
-\noindent
-{\bf Organization of the article.}
-
-\medskip
-
-\noindent
-{\bf Notations.}
-Throughout the paper, $K$ will refer to a complete, discrete valuation 
-field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its 
-valuation, $\OK$ its ring of integers and $\pi$ a uniformizer.
+\todo{Explain here why computing characteristic polynomial is so important.}
+
+The characteristic polynomial is a fundamental invariant of a matrix: its roots
+give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
+In fact, the best known division-free algorithm for computing determinants
+over arbitrary rings \cite{kartofen:92a} does so using the characteristic polynomial.
+Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
+algorithms for counting points of varieties over finite fields. \todo{list some point counting references.}
+\todo{Application to characteristic polynomials of elements of p-adic field extensions?}
+
+\todo{Explain also why taking care of precision is important.}
+
+When computing with $p$-adic matrices, the lack of infinite memory implies
+that the entries may only be approximated at some finite precision $O(p^m)$.  As a
+consequence, in designing algorithms for such matrices one must analyze
+not only the running time of the algorithm but also the accuracy of the result.
+
+Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
+The simplest approach for computing the characteristic polynomial of $M$
+is to compute $\det(X-M)$ either using recursive row expansion or various division free
+algorithms \cites{\cite{seifullin:02a} \cite{kartofen:92a}}.  There are two issues
+with these methods.  First, they are slower than alternatives that allow division,
+requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
+while the lack of division implies that the result is accurate modulo $p^m$ as
+long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.
+
+A faster approach over a field is to compute the Frobenius normal form of $M$,
+which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
+However, the fact that it uses division frequently leads to catastrophic losses of division.
+In many examples, no precision remains at the end of the calculation.
+
+Instead
+
+\todo{Example of $M$ and $M + 1$.}
+
+\subsection*{Previous contributions.} Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?
+
+\subsection*{The contribution of this paper.} % Or \textbf
+
+\subsection*{Organization of the article.}
+
+\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
+discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
+$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}
 
@@
 jagged precision: the $(i,j)$ entry of $M$ is given at precision 
-$O(\pi^{N_{i,j}})$ for some given integer $N_{i,j}$. 
-Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then teaches us that
+$O(\pi^{N_{i,j}})$ for some integers $N_{i,j}$. 
+Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then shows that
 the optimal precision on the $k$-th coefficient of $\chi_M$ is 
@@
 \label{eq:optjagged}
-N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
+N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j})),
 \end{equation}
@@
 that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
-As a consequence, the integer $N'_k$ given the optimal precision on the 
+As a consequence, the integer $N'_k$ giving the optimal precision on the 
 $k$-th coefficient of $M$ is also equal to
 $N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
-where $C^H_{i,j}$ is the $(i,j)$ entry of the comatrix of $X{-}H$
+where $C^H_{i,j}$ is the $(i,j)$ entry of $\com(X{-}H)$,
 where $H$ is the Hessenberg form of $M$.
@@
 \mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
-n \geq 1$$
-and $a$ is a random integer in the range $[0, p^N)$ with uniform
+|n| \geq 1$$
+and $a$ is an integer in the range $[0, p^N)$, selected uniformly at random.
 
@@
 \noindent
-The table of Figure~\ref{fig:exp} summarizes the obtaines results. 
+The table of Figure~\ref{fig:exp} summarizes the results obtained. 
 %
@@
 values and very large deviation as well --- in the two other columns. 
-These experiments then strongly support the interest of applying the 
-methods developed in this paper.
+These experiments thus demonstrate the utility of the methods developed
+in this paper.
 
@@
 
-Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple} eigenvalue 
+Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple}
+\footnote{the corresponding generalized eigenspace has dimension $1$} eigenvalue 
 of $M$. We are interesting in quantifying the optimal precision on 
@@
 
-For doing so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
+To do so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
 suppose that the uncertainty of $M$ is ``jagged'' in the sense that
@@
 that it is possible to follow the eigenvalue $\lambda_\app$ on a small 
-neighborhood $\calU$ of $M$. Precisely, there exists a unique continuous 
+neighborhood $\calU$ of $M$. More precisely, there exists a unique continuous 
 function $f : \calU \to K$ such that:
@@
 \item $f(M_\app) = \lambda_\app$, and
-\item $f(M)$ is an eigenvalue of $M$ for all $M \in \calU$.
+\item $f(M')$ is an eigenvalue of $M'$ for all $M' \in \calU$.
 \end{enumerate}
@@
 $\chi'_M(\lambda) \cdot d \lambda + \tr(\com(X-M) \cdot dM)(\lambda) = 0$,
-from what the Lemma follows.
+from which the Lemma follows.
 \end{proof}
@@
 Lemma 3.4 of \cite{caruso-roe-vaccon:14a} now implies that, if the
-$N_{i,j}$'s are large enough and rather well balanced, the optimal
+$N_{i,j}$'s are large enough and sufficiently well balanced, the optimal
 precision on the eigenvalue $\lambda$ is $O(p^{N'})$ with:
@@
 When $M$ is given at flat precision, \emph{i.e.} the $N_{i,j}$'s are all 
-equal to some $N$, the formula for $N'$ rewrites:
+equal to some $N$, the formula for $N'$ may be rewritten:
 \begin{align*}

### 2017-02-04 charpoly/charpoly.tex
References.

@@
 the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
-By the discussion of \S \ref{}, the exact value of $N'$ can be 
+By the discussion of \S \ref{sec:diffFrob}, the exact value of $N'$ can be 
 determined for a cost of $\softO(n^\omega)$ operations in $K$ and
@@
 for a cost of $\softO(ns)$ operations in $K$ using fast multipoint 
-evaluation methods~\cite{}.
+evaluation methods (see 10.7 of \cite{gathen-gerhard:13a}).
 On the other hand, we observe that $P_i V(\lambda_k)^t$ is nothing but

### 2017-02-04 charpoly/charpoly.tex
References

@@
 In fact, the best known division-free algorithm for computing determinants
-over arbitrary rings \cite{kartofen:92a} does so using the characteristic polynomial.
+over arbitrary rings \cite{kaltofen:92a} does so using the characteristic polynomial.
 Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
@@
 is to compute $\det(X-M)$ either using recursive row expansion or various division free
-algorithms \cites{\cite{seifullin:02a} \cite{kartofen:92a}}.  There are two issues
+algorithms \cites{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
 with these methods.  First, they are slower than alternatives that allow division,
@@
 \item in the model of floating-point arithmetic (see 
-\cite{caruso-coursepadic}*{\S 2.3}), the number of correct digits of the 
+\cite{caruso:17a}*{\S 2.3}), the number of correct digits of the 
 $k$-th coefficient of the characteristic polynomial of $M$.

### 2017-02-04 charpoly/charpoly.tex
Small style fixes, updating bib entries

--- 2017-02-04 charpoly/charpoly.tex
@@
 \usepackage{amsthmnoproof}
+\usepackage{mathrsfs}
 \usepackage{amsrefs}
@@
 is to compute $\det(X-M)$ either using recursive row expansion or various division free
-algorithms \cites{\cite{seifullin:02a} \cite{kartofen:92a}}.  There are two issues
+algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
 with these methods.  First, they are slower than alternatives that allow division,
@@
 
-Instead
+Instead,
 
@@
 closure $\bar{K}$ of $K$ all have dimension $1$
-%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
+\item \label{pt:charmin} the characteristic polynomial of $M$ is equal to the minimal polynomial of $M$.
 \end{enumerate}
@@
 \begin{proof}
-The equivalence of \eqref{pt:cyc} and \eqref{pt:1d} is standard; see
+The equivalence of \eqref{pt:cyc}, \eqref{pt:1d}, and \eqref{pt:charmin} is standard; see
 \cite{hoffman-kunze:LinearAlgebra}*{\S 7.1} for example.  We now show
-that \eqref{pt:surj} and \eqref{pt:cyc} are equivalent.
+\eqref{pt:cyc} $\Rightarrow$ \eqref{pt:surj} and \eqref{pt:surj} $\Rightarrow$ \eqref{pt:1d}
 
@@
 
-To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
+Now suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
 After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
-will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
-for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
+will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
+where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
+Since $\lambda$ occurs in two Jordan blocks,
 $X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
@@
 \end{cor}
-
-We can also give a criterion on $M$ that allows an easy description of the image of flat
-precision lattices under $d\chi(M)$.
-\begin{prop}
-Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
-a cyclic vector.  Then $d\chi_M\big(M_n(\OK)\big) = \OK[X] / \chi(M)$.
-\end{prop}
-\begin{proof}
-It suffices to check that $d\chi_M$ is surjective modulo $\pi$.
-Surjectivity of differential
-can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
-in Frobenius normal form it has only one block.
-\end{proof}
 
@@
 
-Let $C$ be the companion matrix associated to $\chi_M$:
+Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$:
 \begin{equation}
 \label{eq:companion}
-N = \left( \begin{matrix}
+\mathscr{C} = \left( \begin{matrix}
 0 & 1 & 0 & \cdots & 0 \\
-\vdots & \ddots & 1 & \ddots & \vdots \\
-\vdots & & \ddots & \ddots & 0 \\
-0 & \cdots & \cdots & 0 & 1 \\
+0 & 0 & 1 & \ddots & \vdots \\
+\vdots & \vdots & \ddots & \ddots & 0 \\
+0 & 0 & \cdots & 0 & 1 \\
 -a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
@@
 By Proposition~\ref{prop:surjectivity}, there exists a matrix 
-$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
+$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Applying the same
 result to the transpose of $M$, we find that there exists another
-invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.
+invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q \mathscr{C} Q^{-1}$.
 
@@
 as in Proposition~\ref{prop:shortcom}, we find:
-\begin{align*}
-N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
+\begin{align}
+N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \nonumber \\
 & \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
-\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big)
-\end{align*}
+\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big) \label{eq:Nprime}
+\end{align}
 where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
@@
 equal to some $N$, the formula for $N'$ may be rewritten:
+\begin{align}
+N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \nonumber \\
+& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
++ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big) \label{eq:Nprimeflat}
+\end{align}
+and can therefore now be evaluated for a cost of $\softO(n^\omega)$
+operations in $K$ and only $O(n)$ operations with integers.
+
+\medskip
+
+To conclude, let us briefly discuss the situation where we want
+to figure out the optimal jagged precision on a tuple $(\lambda_1,
+\ldots, \lambda_s)$ of simple eigenvalues. Applying \eqref{eq:Nprime},
+we find that the optimal precision on $\lambda_k$ is 
 \begin{align*}
-N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
-& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
-+ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big)
-\end{align*}
-and can therefore now be evaluated for a cost of $\softO(n^\omega)$
-operations in $K$ and only $O(n)$ operations on integers.
-
-\medskip
-
-To conclude with, let us discuss briefly the situation where we want
-to figure out the optimal jagged precision on a tuple $(\lambda_1,
-\ldots, \lambda_s)$ of simple eigenvalues. Applying what we have done
-previously, we find that the optimal precision on $\lambda_k$ is 
-\begin{align*}
-N'_k & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
+N'_k & = \val(\alpha(\lambda_k)) - \val(\chi'_M(\lambda_k)) \\
 & \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
-\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big).
+\val(P_i V(\lambda_k)^t) + \val(V(\lambda_k) Q_j^t)\big).
 \end{align*}
@@
 The $N'_k$'s can be all computed in $\softO(n^\omega)$ operations
-in $K$ and $O(n^2 s)$ operations on integers.
+in $K$ and $O(n^2 s)$ operations with integers.
 
@@
 lowered to $\softO(n^\omega)$ operations in $K$ and $O(n s)$ operations 
-on integers.
+with integers.
 \end{prop}
@@
 the right factor is a structured matrix (a truncated Vandermonde):
-computing the above product reduces to evaluate a polynomial at the
+computing the above product reduces to evaluating a polynomial at the
 points $\lambda_1, \ldots, \lambda_s$.}. Therefore all the $P_i 

### 2017-02-04 charpoly/charpoly.tex
References
Working on intro, fix typesetting for com, rec, det

@@
 \DeclareMathOperator{\lcm}{lcm}
+\DeclareMathOperator{\rec}{rec}
 
@@
 
-Instead,
+Instead, we separate the computation of the precision of $\chi_M$ from the computation
+of an approximation to $\chi_M$.  Given some precision on $M$, we use \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}
+to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
+of much of this paper.  With this precision known, the actual calculation of $\chi_M$
+may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
+stable algorithm. \todo{reference a place with more detail.}
 
 \todo{Example of $M$ and $M + 1$.}
+
+One benefit of this approach is that we may account for diffuse precision:
+precision that is not localized on any single coefficient of $\chi_M$.  For example,
+let $0 \le \alpha_1 \le \alpha_2 \le \dots \le \alpha_n$, consider a
+diagonal matrix $D$ with diagonal entries $(p^{\alpha_1}, \dots, p^{\alpha_n})$,
+let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
+of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $M$ is known
+with precision $O(p^m)$ then $\chi_M$ will be known with some precision
+larger than $O(p^m)$ (see Proposition \ref{prop:mainlem}).
 
@@
 Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
-We define the reciprocal polynomial of order $f$ of $P$ as $P^{rec,d}=X^d P \left( 1/X \right).$
+We define the reciprocal polynomial of order $f$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
 Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
-We denote by $A^{rec,d}$ the matrix
+We denote by $A^{\rec,d}$ the matrix
 whose coefficients are the reciprocal polynomials of order $d$ 
@@
 \begin{eqnarray*}
-com(1-XM)^{rec,n-1}=&com(X-M), \\
-(\chi_M \times 1)^{rec,n}=&(1-XM) com(1-XM).\\
+\com(1-XM)^{\rec,n-1}=&\com(X-M), \\
+(\chi_M \times 1)^{\rec,n}=&(1-XM) \com(1-XM).\\
 \end{eqnarray*}
@@
 let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
-then $det (A^{rec,1})=det(A)^{rec,d}.$
-Indeed, one can use multilinearity of the determinant on $X^d det(A(1/X))$
+then $\det (A^{\rec,1})=\det(A)^{\rec,d}.$
+Indeed, one can use multilinearity of the determinant on $X^d \det(A(1/X))$
 to prove this result.
@@
 
-This lemma allows us to compute $com(1-XM)$ instead of $com(X-M).$
+This lemma allows us to compute $\com(1-XM)$ instead of $\com(X-M).$
 This has a remarkable advantage: the pivots during the computation of
-the SNF of $com(1-XM)$ are units of $O_K[[X]],$ and are known
+the SNF of $\com(1-XM)$ are units of $O_K[[X]],$ and are known
 in advance to be on the diagonal. This leads to a very smooth
@@
 
-\noindent 9. \ \textbf{Return} $V^{rec,n-1}, \psi^{rec,n}.$
+\noindent 9. \ \textbf{Return} $V^{\rec,n-1}, \psi^{\rec,n}.$
 
@@
 
-\noindent 6. \ Return $com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi (M).$
+\noindent 6. \ Return $\com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi (M).$
 

### 2017-02-04 charpoly/charpoly.tex
formating for cond

@@
 \DeclareMathOperator{\rec}{rec}
+\DeclareMathOperator{\cond}{cond}
 
@@
 \begin{prop}
-Let $cond(M)$ be the valuation of the last invariant factor of $M.$
+Let $\cond(M)$ be the valuation of the last invariant factor of $M.$
 If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
-can be computed at precision $O(\pi^{N-2cond(M)})$ by a \textbf{SNF} computation
+can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
 and this lower-bound is optimal
@@
 \begin{proof}
-The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
-It is $-2cond(M)$ for $M^{-2}$ and it is then clear that $-2cond(M)$
+The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M).$
+It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $-2\cond(M)$
 can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
@@
 From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
-known at precision at least $O(\pi^{N-cond(M)})$ with coefficients in $O_K$
+known at precision at least $O(\pi^{N-\cond(M)})$ with coefficients in $O_K$
 and determinant in $O_K^*$ realizing an SNF form of $M.$
 There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
-Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-cond(M),$
-we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2cond(M)}),$
+Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-\cond(M),$
+we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2\cond(M)}),$
 which concludes the proof.
@@
 $c_i = M^i c$ for all $i \geq 1$. The $c_i$'s can be computed in 
-$\softO(n^\omega)$ operations in $K$ using \cite{}. Let $P_\inv$ be the 
+$\softO(n^\omega)$ operations in $K$ using the first algorithm
+of \cite{keller-gehrig:85a}. Let $P_\inv$ be the 
 $n \times n$ matrix whose rows are the $c_i$'s for $1 \leq i \leq n$. 

### 2017-02-04 charpoly/charpoly.tex
ISSAC 2017 template, intro to Sec 5, 3.4 updated.

@@
 
-\CopyrightYear{2016}
+\CopyrightYear{2017}
 \setcopyright{acmcopyright}
-\conferenceinfo{ISSAC '16,}{July 19-22, 2016, Waterloo, ON, Canada}
-\isbn{978-1-4503-4380-0/16/07}\acmPrice{\$15.00}
-\doi{http://dx.doi.org/10.1145/2930889.2930898}
+\conferenceinfo{ISSAC '17,}{July 25-28, 2017, Kaiserslautern, Germany}
+\isbn{***-*-****-****-0/17/07}\acmPrice{\$15.00}
+\doi{http://dx.doi.org/10.1145/*******.*******}
 
@@
 write down the matrices $P$ and $Q$. 
-We shall see moreover in \S \ref{sec:eigenvalues} that interesting
+We shall see moreover in Subsection \ref{ssec:eigenvalues} that interesting
 informations can be read off on this short form $(\alpha, P, Q, 
@@
 using the invertible pivot
-$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \Zp [X]$).
+$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in O_K[X]$).
 
@@
 Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
-Then, using Algorithm 2, one can compute $\com (X -H)$ in 
+Then, using Algorithm 2, one can compute $\com (X -H)$
+and $\chi_M$ in 
 $\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
@@
 \end{proof}
+\begin{rem}
+If $M \in M_n(K)$ does not have coefficients in $O_K,$
+then, by an appropriate scaling, we can still use
+Algorithms 1 and 2 to get $\chi_M$
+in $\softO (n^3)$ operations in $O_K$ at some shift
+of the initial precision.
+\end{rem}
+
+We will see in Section \ref{sec:optjagged} that for an
+entry matrix with coefficients known at flat precision,
+Algorithms 1 and 2 are enough to
+know the optimal jagged precision on $\chi_M.$
 
 \subsection{The comatrix of $X{-}M$}
+
+In this Subsection, we are interested in a special
+case of the computation of the comatrix of $X-M,$
+when $\chi_M$ is square-free (which is the 
+generic case).
+We observe that it is possible to combine
+the result of Proposition \ref{prop:shortcom}
+with Algorithm 2 to actually compute
+the comatrix of $X-M.$
+It leads to a $\softO (n^3)$ algorithm
+where the only divisions are for
+gcd and modular inverse computations.
 
@@
 \begin{theo}
+For $M \in M_n(O_K)$ such that $discr( \chi_M) \neq 0,$
 Algorithm 3 compute 
 $\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
-operations in $O_K$ at initial precision.
+operations in $K$ at initial precision.
+The only divisions are in steps 4 and 5.
 \end{theo}
@@
 is of dimension at most $n-1.$
-As in the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
+The idea is to work modulo $X-\lambda$
 for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
@@
 Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
-on Step 4, which concludes the proof.
-\end{proof}
+on Step 4.
+Now, it is clear that the only divisions happening are on steps 4 and 5.
+This concludes the proof.
+\end{proof}
+
+\begin{rem}
+As in the previous section, it is possible
+to scale $M \in M_n(K)$ so as to
+get coefficients in $O_K$ and apply the previous algorithm.
+\end{rem}
+\begin{rem}
+We refer to \cite{caruso:15a} for the handling
+of the precision on gcd and modular inverse computations.
+In this article, ways to tame the loss of precision
+coming from divisions are explored, following
+the methods of \cite{caruso-roe-vaccon:14a}.
+\end{rem}
 
@@
 
-\todo{Write a small introduction}
+In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffHess},
+we have proposed algorithms to obtain the comatrix of 
+$X-M.$ Our motivation for these computations is to then
+be able to understand what is the optimal precision on $\chi_M.$
+In this section, we provide some answers to this question of
+the optimal precision on $\chi_M,$ along with
+numerical evidence.
+We also show that it is then possible to derive 
+optimal precision on eigenvalues of $M.$
 

### 2017-02-04 charpoly/charpoly.tex
Working on Intro

--- 2017-02-04 charpoly/charpoly.tex
@@
 
-\todo{Explain here why computing characteristic polynomial is so important.}
-
 The characteristic polynomial is a fundamental invariant of a matrix: its roots
@@
 \todo{Application to characteristic polynomials of elements of p-adic field extensions?}
-
-\todo{Explain also why taking care of precision is important.}
 
@@
 which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
-However, the fact that it uses division frequently leads to catastrophic losses of division.
+However, the fact that it uses division frequently leads to catastrophic losses of precision.
 In many examples, no precision remains at the end of the calculation.
@@
 
-\todo{Example of $M$ and $M + 1$.}
-
 One benefit of this approach is that we may account for diffuse precision:
@@
 let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
-of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $M$ is known
-with precision $O(p^m)$ then $\chi_M$ will be known with some precision
-larger than $O(p^m)$ (see Proposition \ref{prop:mainlem}).
+of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $\alpha_{n-1} > 0$
+and $M$ is known with precision $O(p^m)$ then the constant term of $\chi_M$
+will be known with precision larger than $O(p^m)$ (see
+\cite{caruso-roe-vaccon:15a}*{Prop. 3.2}).
+
+As long as none of the eigenvalues of $M$ are congruent to $-1$ modulo $p$,
+then none of the coefficients of the characteristic polynomial of $1+M$ will have
+precision larger than $O(p^m)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
+content of these two polynomials should be equivalent. The solution is that the
+extra precision in $\chi_{1+M}$ is diffuse and not visible on any individual coefficient.
+We formalize this phenomenon using lattices; see Section \ref{sec:padicprec} for
+further explanation, and \cite{caruso:17a}*{\S 3.2.2} for a specific example of the
+relationship between $\chi_M$ and $\chi_{1+M}$.
 
@@
 discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}
+$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$}
 
@@
 
-\subsection{The theory of p-adic precision}
+\subsection{The theory of p-adic precision} \label{sec:padicprec}
 
@@
 
-\begin{cor}
+\begin{cor} \label{cor:prec_gain}
 Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^m)$.
-Then the characteristic polynomial of $M$ has precision larger than $O(p^m)$
+Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(p^m)$
 if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
@@
 
-Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field,
-the running time of matrix multiplication.
-
-\todo{Image modulo $p$ and elementary divisors. 
-Consequences: (1)~computation in $\softO(n^\omega)$ and (2)~gain criterium.}
+Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field \cite{storjohann:01a}.
 
@@
 write down the matrices $P$ and $Q$. 
-We shall see moreover in \S \ref{sec:eigenvalues} that interesting
+We shall see moreover in \S \ref{sec:diffFrob} that interesting
 informations can be read off on this short form $(\alpha, P, Q, 
@@
 \begin{rem} \label{rem:char_pol_from_hessenberg}
-From an Hessenberg form of $M,$ it is well known
+From a Hessenberg form of $M,$ it is well known
 that one can compute the characteristic polynomial of 
-$M$ in $O(n^3)$ operations in $K.$ See \cite{Cohen:2013} page
-55 and 56.
-However, this computation involves some divisions, and its
-behaviour regarding to precision is not easy to
-quantify.
+$M$ in $O(n^3)$ operations in $K$ \cite{Cohen:2013}*{pp. 55--56}.
+However, this computation involves division, and its
+precision behavior is not easy to quantify.
 \end{rem}
@@
 
+\begin{rem}
 As a consequence of the previous discussion, once the optimal jagged 
-precision is known, it is possible to lift arbitrarily the precision on 
-$M$ and then to perform a computation of the characteristic polynomial 
+precision is known, it is possible to lift the entries of $M$ to a sufficiently
+large precision and then use 
+and then to perform a computation of the characteristic polynomial 
 at the optimal precision. This requires $\softO(n^3)$ operations in $K$ 
 but, for several instances, may require to increase a lot the precision.
+\end{rem}
 

### 2017-02-04 charpoly/charpoly.tex

--- 2017-02-04 charpoly/charpoly.tex
@@
 discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}
+$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$, scalar matrices and $I_n$}
 
@@
 The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
-\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over $\OK / \pi\OK$.
-\end{proof}
-
-Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.
+\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
+\end{proof}
+
+Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
+$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
 
 \begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and
-let $\alpha = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
-$r \in (0, \alpha \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
+Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and let 
+\[
+\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
+\]
+Then, for all $\rho \in (0, 1]$ and all
+$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
 satisfies:
@@
 of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
-Since the precision polygon lies below the Newton polygon of $M$ \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
-and since the smallest entry of the Newton polygon occurs either with the leading
-coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(\alpha))$.
+By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
+occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
+$B_W(1) \subset d\chi(M)(B_V(\alpha^{-1}))$.
 
@@
 Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
-\pmod{\chi_M}$, we deduce $(X{-}N) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
-Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}N$
+\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
+Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}\mathscr{C}$
 modulo $\chi_M$. On the other hand, a direct computation shows that
-every column vector $W$ lying in the right kernel of $X{-}N$ modulo 
+every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
 $\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
@@
 stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
-determined by $2n$ coefficients while we need $2n^2$ coefficients to 
+determined by $2n$ coefficients while we need $2n^2$ entries to 
 write down the matrices $P$ and $Q$. 
-We shall see moreover in Subsection \ref{ssec:eigenvalues} that interesting
-informations can be read off on this short form $(\alpha, P, Q, 
+We shall see moreover in Section \ref{sec:diffFrob} that interesting
+information can be read off of this short form $(\alpha, P, Q, 
 \chi_M)$.
@@
 
-In this section, we combine the computation of an Hessenberg form
-of a matrix and the computation of the inverse through the SNF
-over a complete discrete valuation field
-to compute $\com (X-M)$ and $d \chi$
-in a way such that if $M \in M_n(O_K),$
-only division by invertible elements of $O_K$
-occur.
-%The computation of an Hessenberg form provides an $O(n^3)$
+In this section, we combine the computation of a Hessenberg form
+of a matrix and the computation of the inverse through the Smith normal form (SNF)
+over a complete discrete valuation field (CDVF)
+to compute $\com(X-M)$ and $d \chi$.
+If $M \in M_n(O_K)$, then only division by invertible
+elements of $O_K$ will occur.
+%The computation of a Hessenberg form provides an $O(n^3)$
 %algorithm to compute the characteristic polynomial.
@@
 
-We begin with the study of the computation of
+We begin with the computation of
 an approximate Hessenberg form.
@@
 \begin{deftn}
-$M \in M_n (k)$ is said to be an Hessenberg matrix if
-\[\forall i \in \llbracket 3, n \rrbracket, 
-\forall j \in \llbracket 1, i-2 \rrbracket, M_{i,j}=0. \]
-If $M \in M_n (k)$ and $H \in M_n (k)$ is an Hessenberg 
-matrix such that there exists $P \in GL_n(k)$
-with $H=PMP^{-1},$ we say that H is an Hessenberg form
-of $M.$
+A \emph{Hessenberg matrix} is a matrix $M \in M_n(K)$ with
+\[
+M_{i,j}=0 \mbox{ for $j \le i-2$.}
+\]
+Given integers $n_{i,j}$, an \emph{approximate Hessenberg matrix}
+is a matrix $M \in M_n(K)$ with
+\[
+M_{i,j} = O(\pi^{n_{i,j}}) \mbox{ for $j \le i-2$.}
+\]
+If $M \in M_n (K)$ and $H \in M_n (K)$ is an (approximate) Hessenberg matrix
+similar to $M$, we say that H is an \emph{(approximate) Hessenberg form} of $M.$
 \end{deftn}
 
-\begin{deftn}
-In the previous definition of an Hessenberg matrix $M,$
- if we only have $M_{i,j}= O(\pi^{n_{i,j}}),$ then we say
- that $M$ is an approximate Hessenberg matrix.
- This generalize directly to approximate Hessenberg form. 
-\end{deftn}
-
 It is not hard to prove that every matrix over a field admits
-an Hessenberg form.
+a Hessenberg form.
 We prove here that over $K,$ if a matrix is 
@@
 
-\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $val(H_{i_{min},j})$ is minimal.
+\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $\val(H_{i_{min},j})$ is minimal.
 
@@
 
-In this Subsection, we prove that to compute the inverse of
+In this section, we prove that to compute the inverse of
 a matrix over a CDVF $K$, SNF is precision-wise optimal in the flat-precision case.
@@
 As it is, in a way, optimal for the computation of the
-inverse in a CDVF, we hope to get reasonnable behaviour
+inverse in a CDVF, we hope to get reasonable behavior
 for the computation of $\com (X-M)$ even though the coefficients
@@
 The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
-division free algorithm of \cite{}. It is likely that, proceeding so, we 
+division free algorithm of \cite{seifullin:02a}. It is likely that, proceeding so, we 
 limit the loss of precision.

### 2017-02-04 charpoly/charpoly.tex
Previous contribution: first version.

--- 2017-02-04 charpoly/charpoly.tex
@@
 
-\subsection*{Previous contributions.} Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?
+\subsection*{Previous contributions.} 
+%\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
+
+Since the description of Kedlaya's algorithm in
+\cite{kedlaya:01a}, the computation of characteristic polynomials
+over $p$-adic numbers
+has become a crucial point in many counting-points algorithms.
+We can cite for example \cite{harvey:07a, harvey:14a} for
+algorithms to count points on hyperelliptic curves.
+
+In most of these papers, the precision analysis usually
+deals with great details on how to obtain the matrices (\textit{e.g.} of action of Frobenius)  that are involved in the point-counting schemes.
+However, the computation of their characteristic polynomials is often
+a little bit less thoroughly studied: some
+refer to fast algorithm (using divisions), while others
+refer to division-free algorithms.
+
+In \cite{caruso-roe-vaccon:15a}, the authors have begun the application
+of the theory of differential precision of \cite{caruso-roe-vaccon:14a}
+to stable computation of characteristic polynomials.
+They have obtained a way to express the optimal precision
+on the characteristic polynomials, but have not studied 
+how to achieve this optimal precision.
 
@@
 Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
-We define the reciprocal polynomial of order $f$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
+We define the reciprocal polynomial of order $d$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
 Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$

### 2017-02-04 charpoly/charpoly.tex
Statement in Section 4.

@@
 Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
-algorithms for counting points of varieties over finite fields. \todo{list some point counting references.}
+algorithms for counting points of varieties over finite fields (see
+\cite{kedlaya:01a, harvey:07a, harvey:14a}.
+
 \todo{Application to characteristic polynomials of elements of p-adic field extensions?}
@@
 a little bit less thoroughly studied: some
-refer to fast algorithm (using divisions), while others
+refer to fast algorithms (using divisions), while others
 refer to division-free algorithms.
@@
 on the characteristic polynomials, but have not studied 
-how to achieve this optimal precision.
+how to attain this optimal precision algorithmicaly.
 
@@
 \begin{cor} \label{cor:prec_gain}
-Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^m)$.
-Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(p^m)$
+Suppose that $M \in \GL_n(\OK)$ is known with precision $O(\pi^m)$.
+Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^m)$
 if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
@@
 Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
-Inside the nested \textbf{for} loop, if we want to eliminate $p^{u_y} \varepsilon_y+O(p^{n_y})$ with pivot $p^{u_x} \varepsilon_x+O(p^{n_x}),$
+Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
 with the $\varepsilon$'s being units,
-the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\mathbb{F}_q[X],$ $\mathbb{Q}[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
+the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\mathbb{F}_q[X],$ $\mathbb{Q}[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
 Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
@@
 issue regarding to precision in particular cases.
+In this section, we only estimate the number of operations in $K$
+and not their behaviour on precision.
 
 From now on, we fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
-surjective. Let $(\alpha, P, Q, \chi_M)$ be the quadruaple encoding
+surjective. Let $(\alpha, P, Q, \chi_M)$ be the quadruple encoding
 the short form of $d \chi_M$; we recall that they are related by the
@@
 \end{align*}
+
+$\chi_M$ can be computed in $\softO(n^\omega)$ operations in $K$
+(\textit{e.g.} as a by-product of \cite{storjohann:01a}).
+
 The matrix $P$ can be computed as follows. Pick $c \in K^n$. Define 
 $c_i = M^i c$ for all $i \geq 1$. The $c_i$'s can be computed in 
-$\softO(n^\omega)$ operations in $K$ using the first algorithm
+$\softO(n^\omega)$ operations in $K,$ \textit{e.g.}
+using the first algorithm
 of \cite{keller-gehrig:85a}. Let $P_\inv$ be the 
@@
 $$\alpha = a_1 + a_2 X + \cdots + a_{n-1} X^{n-2} + X^{n-1}.$$
-No further computation are then needed to derive the value of $\alpha$.
+No further computation are then needed to derive the value of $\alpha$
+and we can conclude with the following theorem.
+
+\begin{theo}
+Given $M \in M_n(K)$ such that $d \chi_M$ is surjective,
+then one can compute $(\alpha, P, Q, \chi_M)$
+in $K[X]$ such that
+$\com(X{-}N) = \alpha \cdot V^t \cdot V R^t \mod \chi_M$
+in $\softO(n^\omega)$ operations in $K.$
+\end{theo}
 
@@
 
-In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffHess},
+In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffFrob},
 we have proposed algorithms to obtain the comatrix of 
@@
 $N_{i,j}$'s are large enough and sufficiently well balanced, the optimal
-precision on the eigenvalue $\lambda$ is $O(p^{N'})$ with:
+precision on the eigenvalue $\lambda$ is $O(\pi^{N'})$ with:
 $$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(C_{i,j}(\lambda)) - 

### 2017-02-04 charpoly/charpoly.tex
Proofreading Section 3

--- 2017-02-04 charpoly/charpoly.tex
@@
 
-\subsection*{Previous contributions.} 
-%\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
-
-Since the description of Kedlaya's algorithm in
-\cite{kedlaya:01a}, the computation of characteristic polynomials
-over $p$-adic numbers
-has become a crucial point in many counting-points algorithms.
-We can cite for example \cite{harvey:07a, harvey:14a} for
-algorithms to count points on hyperelliptic curves.
-
-In most of these papers, the precision analysis usually
-deals with great details on how to obtain the matrices (\textit{e.g.} of action of Frobenius)  that are involved in the point-counting schemes.
-However, the computation of their characteristic polynomials is often
-a little bit less thoroughly studied: some
-refer to fast algorithm (using divisions), while others
-refer to division-free algorithms.
-
-In \cite{caruso-roe-vaccon:15a}, the authors have begun the application
-of the theory of differential precision of \cite{caruso-roe-vaccon:14a}
-to stable computation of characteristic polynomials.
-They have obtained a way to express the optimal precision
-on the characteristic polynomials, but have not studied 
-how to achieve this optimal precision.
+\subsection*{Previous contributions.} Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?
 
@@
 discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$, scalar matrices and $I_n$}
+$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$, scalar matrices and $I_n$, $\sigma_1(M), \dots, \sigma_n(M)$}
 
@@
 In this section, we prove that to compute the inverse of
-a matrix over a CDVF $K$, SNF is precision-wise optimal in the flat-precision case.
-We first recall what is the differential of the computation of the inverse of a matrix.
+a matrix over a CDVF $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
+We first recall the differential of matrix inversion.
 
@@
 Let $u \: : \: GL_n (K) \rightarrow GL_n(K),$ $M \mapsto M^{-1}.$
-Then for $M \in GL_n (K),$ $du(M) \cdot dM=M^{-1} dM M^{-1}.$
+Then for $M \in GL_n (K),$ $du_M(dM)=M^{-1} dM M^{-1}.$
 It is always surjective.
@@
 \begin{prop}
-Let $\cond(M)$ be the valuation of the last invariant factor of $M.$
-If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
-can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
-and this lower-bound is optimal
-at least when $N$ is large.
+Let $\cond(M) = \val(\sigma_n(M))$.
+If $dM$ is a flat precision of $O(\pi^m)$ on $M$ then $M^{-1}$
+can be computed at precision $O(\pi^{m-2\cond(M)})$ by a \textbf{SNF} computation
+and this lower-bound is optimal,
+at least when $m$ is large.
 \end{prop}
@@
 The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M).$
-It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $-2\cond(M)$
-can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
+It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $m-2\cond(M)$
+can be obtained as the valuation of a coefficient of $du_M(dM)$
 and the smallest that can be achieved this way for $dM$ in a precision lattice
 of flat precision. Hence the optimality of the bound given, at least when 
-$N$ is large, thanks to Lemma 3.14 of \cite{caruso-roe-vaccon:14a}
-
-Now, the computation of an SNF form over a CDVF was described in \cite{Vaccon-these}.
-From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
-known at precision at least $O(\pi^{N-\cond(M)})$ with coefficients in $O_K$
-and determinant in $O_K^*$ realizing an SNF form of $M.$
+$m$ is large \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
+
+Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
+From $M$ known at flat precision $O(\pi^m),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
+known at precision at least $O(\pi^{m-\cond(M)})$, with coefficients in $O_K$
+and determinant in $O_K^\times$ realizing an Smith normal form of $M.$
 There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
-Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-\cond(M),$
-we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2\cond(M)}),$
+Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
+we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{m-2\cond(M)}),$
 which concludes the proof.
@@
 
-In this subsection, we want to compute $\com (X Id -H)$
-by using the SNF computation of the previous subsection.
-As it is, in a way, optimal for the computation of the
-inverse in a CDVF, we hope to get reasonable behavior
-for the computation of $\com (X-M)$ even though the coefficients
-are not, \textit{a priori}, set to live in a CDVF.
-We show that we can reduce this task to the computation
-of an SNF over a CDVF, with no division in $K.$
-
-Firstly, we need a lemma to relate comatrices of
+In this section, we compute $\com(X-H)$ for a Hessenberg matrix $H$
+using the Smith normal form computation of the previous section.
+The entries of $\com(X-H)$ lie in $K[X]$, which is not a CDVF,
+so we may not directly apply the methods of the previous section.
+However, we may relate $\com(X-H)$ to $\com(1-XH)$, whose
+entries lie in the CDVF $K((X))$.  In this way, we compute $\com(X-H)$
+using an SNF method, with no division in $K$.
+
+First, we need a lemma relating comatrices of
 similar matrices:
@@
 Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
-We define the reciprocal polynomial of order $d$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
+We define the reciprocal polynomial of order $f$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
 Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$

### 2017-02-04 charpoly/charpoly.tex
Organization of the paper.

--- 2017-02-04 charpoly/charpoly.tex
@@
 
+In Section \ref{sec:theo_study}, we study the
+differential of the characteristic polynomial so as
+to get optimal information on precision 
+and a compact writing of this differential.
+Then, in Sections \ref{sec:diffHess} and \ref{sec:diffFrob}
+we propose algorithms to compute this differential
+either through Hessenberg or going through
+some Frobenius normal form.
+Finally, we propose in Section \ref{sec:optjagged}
+algorithms to compute the optimal precision
+coefficient-wise on the characteristic polynomial
+and on some eigenvalues along with
+numerical evidence of the behaviour of the computation
+of the characteristic polynomial.
+
 \subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
@@
 \section{Theoretical study}
+\label{sec:theo_study}
 
@@
 \chi_M)$.
+
+\begin{rem}
+Keeping the previous notation, if $U \in GL_n(K),$
+then the corresponding quadruple for 
+$UMU^{-1}$ is
+$(\alpha, UP, U^{t,-1}Q, \chi_M),$
+which can be computed in $O(n^\omega)$ operations in $K.$
+This is faster than computing $U com(X-M) U^{-1},$
+which is, at first sight, in $O(n^4)$ operations in $K.$
+\end{rem}
 

### 2017-02-04 charpoly/charpoly.tex
Contributions, notations

@@
 
-\subsection*{The contribution of this paper.} % Or \textbf
+\subsection*{The contribution of this paper.}
+
+Thanks to the application the framework of differential precision of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
+we known that
+the precision on the characteristic polynomial
+$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
+is determined by the comatrix $com(X I_n -M).$
+In this article, we provide:
+\begin{enumerate}
+\item A factorization of $com(X I_n -M)$ (when $M$ has a cyclic vector).
+\item A simple criterion to decide whether there can be a gain in precision
+on $\chi_M$ when $M \in M_n(\mathbb{Z}_p).$
+\item A $\softO (n^3)$ algorithm with operations in $O_K$
+to compute the optimal precision on each coefficient of $\chi_M$ when $M$
+is given with uniform precision on its entries.
+\item A $\softO (n^\omega)$ algorithm to compute the 
+optimal precision on each coefficient of $\chi_M$
+for $M \in M_n (K),$
+using divisions.
+\end{enumerate}
 
@@
 discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$, scalar matrices and $I_n$}
+$\OK$ its ring of integers and $\pi$ a uniformizer. 
+We will write that $f(n)=\softO (g(n))$ if there exists some
+$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k.$
+We fix an $\omega \in \mathbb{R}$ such that the multiplication
+of two matrices over a ring is in $O(n^\omega)$ operations in the
+ring. Currently, the smallest proved $\omega$ is less than
+$2.3728639 $ thanks to \cite{legall:14a}. 
+$I_n$ will denote the identity matrix of rank $n$ in $M_n(K).$
+When there is no ambiguity, we will drop this $I_n$
+for scalar matrices, \textit{e.g.} for $\lambda \in K$
+and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n -M.$
 

### 2017-02-04 charpoly/charpoly.tex
Working on Section 3

--- 2017-02-04 charpoly/charpoly.tex
@@
 \DeclareMathOperator{\cond}{cond}
+\DeclareMathOperator{\disc}{Disc}
+\DeclareMathOperator{\row}{row}
 
@@
 \newcommand{\OK}{\mathcal{O}_K}
+
+\newcommand{\llb}{[\mkern-2.5mu[}
+\newcommand{\rrb}{]\mkern-2.5mu]}
+\newcommand{\llp}{(\mkern-2.5mu(}
+\newcommand{\rrp}{)\mkern-2.5mu)}
 
@@
 \begin{prop} \label{prop:factor_com}
-Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
+Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
 We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
@@
 Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
-both in $O_K[X]^n,$ we have: 
+both in $\OK[X]^n,$ we have: 
 \[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
@@
 \begin{rem}
-If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
+If $M \in M_n (\OK)$ is such that for some $U,V \in O_{K}[X]^n$
 we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
-$P\in GL_n (O_K)$ then
+$P\in GL_n (\OK)$ then
 \[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
@@
 to compute $\com(X-M)$ and $d \chi$.
-If $M \in M_n(O_K)$, then only division by invertible
-elements of $O_K$ will occur.
+If $M \in M_n(\OK)$, then only division by invertible
+elements of $\OK$ will occur.
 %The computation of a Hessenberg form provides an $O(n^3)$
@@
 Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
-$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
+$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
 \end{prop}
 \begin{proof}
-Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
+Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
 Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
@@
 From $M$ known at flat precision $O(\pi^m),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
-known at precision at least $O(\pi^{m-\cond(M)})$, with coefficients in $O_K$
-and determinant in $O_K^\times$ realizing an Smith normal form of $M.$
+known at precision at least $O(\pi^{m-\cond(M)})$, with coefficients in $\OK$
+and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
 There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
@@
 However, we may relate $\com(X-H)$ to $\com(1-XH)$, whose
-entries lie in the CDVF $K((X))$.  In this way, we compute $\com(X-H)$
+entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\com(X-H)$
 using an SNF method, with no division in $K$.
@@
 Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
-We denote by $A^{\rec,d}$ the matrix
-whose coefficients are the reciprocal polynomials of order $d$ 
-of the coefficients of $A.$ 
+We denote by $A^{\rec,d}$ the matrix with $(A^{\rec,d})_{i,j} = (A_{i,j})^{\rec,d}$.
 \end{deftn}
@@
 \com(1-XM)^{\rec,n-1}=&\com(X-M), \\
-(\chi_M \times 1)^{\rec,n}=&(1-XM) \com(1-XM).\\
+(\chi_M I_n)^{\rec,n}=&(1-XM) \com(1-XM).\\
 \end{eqnarray*}
@@
 to prove this result.
-It directly implies the second part of the lemma.
-As for the first part, since coefficients of the comatrix we are interested in
-are determinants of a matrix of polynomials of degree at most 1, the result
-is also clear.
+It directly implies the second part of the lemma; the first part follows
+from the fact that the entries of $\com(X-M)$ and of $\com(1-XM)$
+are determinants of size $n-1$.
 \end{proof}
@@
 This has a remarkable advantage: the pivots during the computation of
-the SNF of $\com(1-XM)$ are units of $O_K[[X]],$ and are known
+the SNF of $\com(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
 in advance to be on the diagonal. This leads to a very smooth
-precision and complexity behaviour when the entry matrix lives in 
-$M_n(O_K).$
+precision and complexity behaviour when the input matrix lives in 
+$M_n(\OK).$
 
@@
 
-\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(O_K).$
+\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(\OK).$
 
@@
 
-\noindent 1.\ \textbf{Track} the following operations to get $P^{-1}$ and $Q^{-1},$
-such that $U=PU_0Q.$
+\noindent 1.\ While updating $U$, \textbf{track} $P$ and $Q$ so that $U_0=PUQ$ is always satisfied.
 
@@
 using the invertible pivot
-$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in O_K[X]$).
+$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \OK[X]$).
 
@@
 
-\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,i+1},$
+\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i+1,i},$
 using the invertible pivot $U_{i,i}.$
@@
 
-\noindent 7. \ Rescale to get $U = Id \mod X^{n+1}.$
-
-\noindent 8. \ $V:=\psi \times P^{-1} \times Q^{-1}   \mod X^{n+1}.$
+\noindent 7. \ Rescale to get $U = I_n \mod X^{n+1}.$
+
+\noindent 8. \ $V:=\psi \times P \times Q   \mod X^{n+1}.$ \footnote{The product $P \times Q$
+should be implemented by sequential row operations corresponding to the eliminations in Step 5
+in order to avoid a product of two matrices in $M_n(\OK[X])$.}
 
@@
 \begin{theo}
-Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
+Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix.
 Then, using Algorithm 2, one can compute $\com (X -H)$
-and $\chi_M$ in 
-$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
+and $\chi_H$ in 
+$\softO (n^3)$ operations in $\OK$ at the precision given by $H.$
 \end{theo}
 \begin{proof}
-Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
-in $O_K$ at the precision given by $H.$
-Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot:
-on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
+First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
+in $\OK$ at the precision given by $H.$
+Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
+the only other nonzero coefficient in its column is $U_{i+1,i}$.
 As a consequence, when performing this column-pivoting, only two rows ($i$ and
-$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
+$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
 Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
-Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
-Corresponding coefficient for the corresponding shear matrix
- is the opposite of the one defined by pivoting
-and the indices for the receiving and giving rows are swapped compared to the operations on columns.
-Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
-an upper-triangular matrix.
-This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.
-
-As for line 4 and 5, there are only $n-1$ eliminations, resulting again in a $\softO (n^3)$ cost
-for the computation of $P^{-1}$ and $U.$
+Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
+The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
+no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
+for lines 2 and 3 is indeed $\softO (n^3)$ operations.
+
+For lines 4 and 5, there are only $n-1$ eliminations, resulting in a $\softO (n^2)$ cost
+for the computation of $U.$. Rather than actually construct $P$, we just track the eliminations
+performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.
 
@@
 \begin{rem}
-If $M \in M_n(K)$ does not have coefficients in $O_K,$
-then, by an appropriate scaling, we can still use
-Algorithms 1 and 2 to get $\chi_M$
-in $\softO (n^3)$ operations in $O_K$ at some shift
-of the initial precision.
+If $M \in M_n(K)$ does not have coefficients in $\OK,$
+we may apply Algorithms 1 and 2 to $p^v M \in M_n(\OK)$
+in $\softO (n^3)$ operations in $\OK$, and then divide
+the coefficient of $X^k$ in the resulting polynomial by $p^{kv}$.
 \end{rem}
@@
 
-In this Subsection, we are interested in a special
-case of the computation of the comatrix of $X-M,$
-when $\chi_M$ is square-free (which is the 
-generic case).
-We observe that it is possible to combine
-the result of Proposition \ref{prop:shortcom}
-with Algorithm 2 to actually compute
-the comatrix of $X-M.$
-It leads to a $\softO (n^3)$ algorithm
-where the only divisions are for
-gcd and modular inverse computations.
+In this section, we combine Proposition \ref{prop:shortcom}
+with Algorithm 2 to compute the comatrix of $X-M$
+when $\chi_M$ is squarefree.   Note that this condition on
+$\chi_M$ is equivalent to $M$ being diagonalizable under
+the assumption that $d\chi_M$ is surjective.  The result is
+the following $\softO(n^3)$ algorithm, where the only
+divisions are for gcd and modular inverse computations.
 
@@
 
-\noindent{\bf Input:} an approx. $M \in M_n(O_K),$ with $discr(\chi_M) \neq 0.$
+\noindent{\bf Input:} an approx. $M \in M_n(\OK),$ with $\disc(\chi_M) \neq 0.$
 
@@
 
-\noindent 0.\ Find $P \in GL_n(O_K)$ and $H \in M_n(O_K),$ approximate Hessenberg,
+\noindent 0.\ Find $P \in GL_n(\OK)$ and $H \in M_n(\OK),$ approximate Hessenberg,
 such that $M=PHP^{-1},$ using Algorithm 1.
 
-\noindent 1.\ Compute $A=\com (X-H)$ using Algorithm 2.
-
-\noindent 2.\ Do $row(A,1) \leftarrow row(A,1)+\sum_{i=2}^n \mu_i row(A,i),$ for
-random $\mu_i \in O_K,$ by doing $T \times A$ for some $T \in GL_n(O_K).$
+\noindent 1.\ Compute $A=\com (X-H)$ and $\chi(M) = \chi(H)$ using Algorithm 2.
+
+\noindent 2.\ Do $\row(A,1) \leftarrow \row(A,1)+\sum_{i=2}^n \mu_i \row(A,i),$ for
+random $\mu_i \in \OK,$ by doing $T \times A$ for some $T \in GL_n(\OK).$
 Compute $B:=TAT^{-1}.$
 
-\noindent 3.\ Similarily compute $C:=S^{-1}BS$ for $S \in GL_n(O_K)$ corresponding to
-sending a random linear combination of the columns of index $j \in \llbracket 2,n \rrbracket$
+\noindent 3.\ Similarily compute $C:=S^{-1}BS$ for $S \in GL_n(\OK)$ corresponding to
+adding a random linear combination of the columns of index $j \ge 2$
 to the first column of $B.$
@@
 
-\noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
+\noindent 5. \ Compute $U,V \in \OK[X]^n$ such that $C=U {}^t V \mod \chi (M),$
 by inverting $C_{1,1} \mod \chi(M).$
@@
 \begin{theo}
-For $M \in M_n(O_K)$ such that $discr( \chi_M) \neq 0,$
-Algorithm 3 compute 
+For $M \in M_n(\OK)$ such that $\disc( \chi_M) \neq 0,$
+Algorithm 3 computes
 $\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
-operations in $K$ at initial precision.
+operations in $K$.
 The only divisions are in steps 4 and 5.
@@
 \begin{proof}
-As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
+As we have already seen, completing Steps 0 and 1 is in $\softO (n^3).$
 Multiplying by $T$ or $S$ or their inverse corresponds
 to $n$ operations on rows or columns over a matrix with coefficients
-in $O_K[X]$ of degree at most $n.$
+in $\OK[X]$ of degree at most $n.$
 Thus, it is in $\softO (n^3).$
@@
 to scale $M \in M_n(K)$ so as to
-get coefficients in $O_K$ and apply the previous algorithm.
+get coefficients in $\OK$ and apply the previous algorithm.
 \end{rem}
@@
 \begin{prop} \label{prop:optimal_jagged}
-If $M \in M_n(O_K)$ is given at (high enough) jagged precision, 
+If $M \in M_n(\OK)$ is given at (high enough) jagged precision, 
 then we can compute the optimal jagged precision on $\chi_M$ in 
@@
 \begin{rem}
-If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
+If $M \in M_n(\OK)$ is given at (high enough) \emph{flat} precision, 
 then we can avoid the final base change step in the Hessenberg method.

### 2017-02-04 charpoly/charpoly.tex

@@
 
-\subsection*{The contribution of this paper.} % Or \textbf
+\subsection*{The contribution of this paper.}
+
+Thanks to the application the framework of differential precision of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
+we known that
+the precision on the characteristic polynomial
+$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
+is determined by the comatrix $com(X I_n -M).$
+In this article, we provide:
+\begin{enumerate}
+\item A factorization of $com(X I_n -M)$ (when $M$ has a cyclic vector).
+\item A simple criterion to decide whether there can be a gain in precision
+on $\chi_M$ when $M \in M_n(\mathbb{Z}_p).$
+\item A $\softO (n^3)$ algorithm with operations in $O_K$
+to compute the optimal precision on each coefficient of $\chi_M$ when $M$
+is given with uniform precision on its entries.
+\item A $\softO (n^\omega)$ algorithm to compute the 
+optimal precision on each coefficient of $\chi_M$
+for $M \in M_n (K),$
+using divisions.
+\end{enumerate}
 
 \subsection*{Organization of the article.}
+
+In Section \ref{sec:theo_study}, we study the
+differential of the characteristic polynomial so as
+to get optimal information on precision 
+and a compact writing of this differential.
+Then, in Sections \ref{sec:diffHess} and \ref{sec:diffFrob}
+we propose algorithms to compute this differential
+either through Hessenberg or going through
+some Frobenius normal form.
+Finally, we propose in Section \ref{sec:optjagged}
+algorithms to compute the optimal precision
+coefficient-wise on the characteristic polynomial
+and on some eigenvalues along with
+numerical evidence of the behaviour of the computation
+of the characteristic polynomial.
 
@@
 discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$, scalar matrices and $I_n$, $\sigma_1(M), \dots, \sigma_n(M)$}
+$\OK$ its ring of integers and $\pi$ a uniformizer. 
+We will write that $f(n)=\softO (g(n))$ if there exists some
+$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k.$
+We fix an $\omega \in \mathbb{R}$ such that the multiplication
+of two matrices over a ring is in $O(n^\omega)$ operations in the
+ring. Currently, the smallest proved $\omega$ is less than
+$2.3728639 $ thanks to \cite{legall:14a}. 
+$I_n$ will denote the identity matrix of rank $n$ in $M_n(K).$
+When there is no ambiguity, we will drop this $I_n$
+for scalar matrices, \textit{e.g.} for $\lambda \in K$
+and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n -M.$
 
 \section{Theoretical study}
+\label{sec:theo_study}
 

### 2017-02-04 charpoly/charpoly.tex
Not compiling

@@
 \DeclareMathOperator{\row}{row}
+\DeclareMathOperator{\col}{col}
 
@@
 \subsection*{Previous contributions.} 
-%\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
+\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
 
@@
 and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n -M.$
+Finally, we write $\sigma_1(M), \dots, \sigma_n(M)$ for the
+elementary divisors of $M$, sorted in increasing order of valuation.
 
@@
 \item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
-\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
+\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
 \end{enumerate}
@@
 \[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
-\todo{I think that the correct condition is $d \chi_M$ surjective (which
-is a bit weaker).}
+\todo{I think that the correct condition is $d \chi_M$ surjective (which is a bit weaker).}
 \end{prop}
@@
 which can be computed in $O(n^\omega)$ operations in $K.$
-This is faster than computing $U com(X-M) U^{-1},$
+This is faster than computing $U \com(X-M) U^{-1},$
 which is, at first sight, in $O(n^4)$ operations in $K.$
@@
 
-\noindent 1.\ Compute $A=\com (X-H)$ and $\chi(M) = \chi(H)$ using Algorithm 2.
+\noindent 1.\ Compute $A=\com (X-H)$ and $\chi_M = \chi_H$ using Algorithm 2.
 
@@
 
-\noindent 4.\  \textbf{If} $gcd(C_{1,1}, \chi(M)) \neq 1,$ \textbf{then} go to 2.
-
-\noindent 5. \ Compute $U,V \in \OK[X]^n$ such that $C=U {}^t V \mod \chi (M),$
-by inverting $C_{1,1} \mod \chi(M).$
-
-\noindent 6. \ Return $\com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi (M).$
+\noindent 4.\  \textbf{If} $gcd(C_{1,1}, \chi_M) \neq 1,$ \textbf{then} go to 2.
+
+\noindent 5. \ Let $F$ be the inverse of $C_{1,1} \mod \chi_M$.
+
+\noindent 6. \ Let $U = \col(C,1)$ and $V = F \cdot \row(C,1) \mod \chi_M$.
+
+\noindent 7. \ Return $\com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi_M.$
 
@@
 operations in $K$.
-The only divisions are in steps 4 and 5.
+The only divisions occur in taking gcds and inverses modulo $\chi_M$.
 \end{theo}
@@
 Thus, it is in $\softO (n^3).$
-Step 5 and 6 are also in $\softO (n^3).$
+Step 5 and 6 are in $\softO (n).$
 All that is to prove is that the set of $P$ and $S$ to avoid
@@
 Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
-affine hyperplane so that $row(B,1) \mod (X-\lambda)$ is a non-zero vector.
-Hence for $row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
+affine hyperplane so that $\row(B,1) \mod (X-\lambda)$ is a non-zero vector.
+Hence for $\row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
 a finite union of affine hyperplane is to avoid.
-Similarily, the goal of Step 3 is to put an invertible coefficient (modulo
+Similarly, the goal of Step 3 is to put an invertible coefficient (modulo
 $\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine

### 2017-02-04 charpoly/charpoly.tex
Remark 5.3

--- 2017-02-04 charpoly/charpoly.tex
@@
 \DeclareMathOperator{\cond}{cond}
-\DeclareMathOperator{\disc}{Disc}
-\DeclareMathOperator{\row}{row}
 
@@
 \newcommand{\OK}{\mathcal{O}_K}
-
-\newcommand{\llb}{[\mkern-2.5mu[}
-\newcommand{\rrb}{]\mkern-2.5mu]}
-\newcommand{\llp}{(\mkern-2.5mu(}
-\newcommand{\rrp}{)\mkern-2.5mu)}
 
@@
 
-\todo{Application to characteristic polynomials of elements of p-adic field extensions?}
+%\todo{Application to characteristic polynomials of elements of p-adic field extensions?}
 
@@
 may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
-stable algorithm. \todo{reference a place with more detail.}
+stable algorithm (see Remark \ref{rem:lift_for_optimal}).
 
@@
 \begin{prop} \label{prop:factor_com}
-Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
+Let $M \in M_n (O_K)$ be such that $discr(\chi (M)) \neq 0.$
 We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
@@
 Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
-both in $\OK[X]^n,$ we have: 
+both in $O_K[X]^n,$ we have: 
 \[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
@@
 \begin{rem}
-If $M \in M_n (\OK)$ is such that for some $U,V \in O_{K}[X]^n$
+If $M \in M_n (O_K)$ is such that for some $U,V \in O_{K}[X]^n$
 we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
-$P\in GL_n (\OK)$ then
+$P\in GL_n (O_K)$ then
 \[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
@@
 to compute $\com(X-M)$ and $d \chi$.
-If $M \in M_n(\OK)$, then only division by invertible
-elements of $\OK$ will occur.
+If $M \in M_n(O_K)$, then only division by invertible
+elements of $O_K$ will occur.
 %The computation of a Hessenberg form provides an $O(n^3)$
@@
 Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
-$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
+$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
 \end{prop}
 \begin{proof}
-Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
+Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
 Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
@@
 In this section, we prove that to compute the inverse of
-a matrix over a CDVF $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
-We first recall the differential of matrix inversion.
+a matrix over a CDVF $K$, SNF is precision-wise optimal in the flat-precision case.
+We first recall what is the differential of the computation of the inverse of a matrix.
 
@@
 Let $u \: : \: GL_n (K) \rightarrow GL_n(K),$ $M \mapsto M^{-1}.$
-Then for $M \in GL_n (K),$ $du_M(dM)=M^{-1} dM M^{-1}.$
+Then for $M \in GL_n (K),$ $du(M) \cdot dM=M^{-1} dM M^{-1}.$
 It is always surjective.
@@
 \begin{prop}
-Let $\cond(M) = \val(\sigma_n(M))$.
-If $dM$ is a flat precision of $O(\pi^m)$ on $M$ then $M^{-1}$
-can be computed at precision $O(\pi^{m-2\cond(M)})$ by a \textbf{SNF} computation
-and this lower-bound is optimal,
-at least when $m$ is large.
+Let $\cond(M)$ be the valuation of the last invariant factor of $M.$
+If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
+can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
+and this lower-bound is optimal
+at least when $N$ is large.
 \end{prop}
@@
 The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M).$
-It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $m-2\cond(M)$
-can be obtained as the valuation of a coefficient of $du_M(dM)$
+It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $-2\cond(M)$
+can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
 and the smallest that can be achieved this way for $dM$ in a precision lattice
 of flat precision. Hence the optimality of the bound given, at least when 
-$m$ is large \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
-
-Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
-From $M$ known at flat precision $O(\pi^m),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
-known at precision at least $O(\pi^{m-\cond(M)})$, with coefficients in $\OK$
-and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
+$N$ is large, thanks to Lemma 3.14 of \cite{caruso-roe-vaccon:14a}
+
+Now, the computation of an SNF form over a CDVF was described in \cite{Vaccon-these}.
+From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
+known at precision at least $O(\pi^{N-\cond(M)})$ with coefficients in $O_K$
+and determinant in $O_K^*$ realizing an SNF form of $M.$
 There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
-Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
-we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{m-2\cond(M)}),$
+Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-\cond(M),$
+we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2\cond(M)}),$
 which concludes the proof.
@@
 
-In this section, we compute $\com(X-H)$ for a Hessenberg matrix $H$
-using the Smith normal form computation of the previous section.
-The entries of $\com(X-H)$ lie in $K[X]$, which is not a CDVF,
-so we may not directly apply the methods of the previous section.
-However, we may relate $\com(X-H)$ to $\com(1-XH)$, whose
-entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\com(X-H)$
-using an SNF method, with no division in $K$.
-
-First, we need a lemma relating comatrices of
+In this subsection, we want to compute $\com (X Id -H)$
+by using the SNF computation of the previous subsection.
+As it is, in a way, optimal for the computation of the
+inverse in a CDVF, we hope to get reasonable behavior
+for the computation of $\com (X-M)$ even though the coefficients
+are not, \textit{a priori}, set to live in a CDVF.
+We show that we can reduce this task to the computation
+of an SNF over a CDVF, with no division in $K.$
+
+Firstly, we need a lemma to relate comatrices of
 similar matrices:
@@
 Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
-We denote by $A^{\rec,d}$ the matrix with $(A^{\rec,d})_{i,j} = (A_{i,j})^{\rec,d}$.
+We denote by $A^{\rec,d}$ the matrix
+whose coefficients are the reciprocal polynomials of order $d$ 
+of the coefficients of $A.$ 
 \end{deftn}
@@
 \com(1-XM)^{\rec,n-1}=&\com(X-M), \\
-(\chi_M I_n)^{\rec,n}=&(1-XM) \com(1-XM).\\
+(\chi_M \times 1)^{\rec,n}=&(1-XM) \com(1-XM).\\
 \end{eqnarray*}
@@
 to prove this result.
-It directly implies the second part of the lemma; the first part follows
-from the fact that the entries of $\com(X-M)$ and of $\com(1-XM)$
-are determinants of size $n-1$.
+It directly implies the second part of the lemma.
+As for the first part, since coefficients of the comatrix we are interested in
+are determinants of a matrix of polynomials of degree at most 1, the result
+is also clear.
 \end{proof}
@@
 This has a remarkable advantage: the pivots during the computation of
-the SNF of $\com(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
+the SNF of $\com(1-XM)$ are units of $O_K[[X]],$ and are known
 in advance to be on the diagonal. This leads to a very smooth
-precision and complexity behaviour when the input matrix lives in 
-$M_n(\OK).$
+precision and complexity behaviour when the entry matrix lives in 
+$M_n(O_K).$
 
@@
 
-\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(\OK).$
+\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(O_K).$
 
@@
 
-\noindent 1.\ While updating $U$, \textbf{track} $P$ and $Q$ so that $U_0=PUQ$ is always satisfied.
+\noindent 1.\ \textbf{Track} the following operations to get $P^{-1}$ and $Q^{-1},$
+such that $U=PU_0Q.$
 
@@
 using the invertible pivot
-$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \OK[X]$).
+$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in O_K[X]$).
 
@@
 
-\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i+1,i},$
+\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,i+1},$
 using the invertible pivot $U_{i,i}.$
@@
 
-\noindent 7. \ Rescale to get $U = I_n \mod X^{n+1}.$
-
-\noindent 8. \ $V:=\psi \times P \times Q   \mod X^{n+1}.$ \footnote{The product $P \times Q$
-should be implemented by sequential row operations corresponding to the eliminations in Step 5
-in order to avoid a product of two matrices in $M_n(\OK[X])$.}
+\noindent 7. \ Rescale to get $U = Id \mod X^{n+1}.$
+
+\noindent 8. \ $V:=\psi \times P^{-1} \times Q^{-1}   \mod X^{n+1}.$
 
@@
 \begin{theo}
-Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix.
+Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
 Then, using Algorithm 2, one can compute $\com (X -H)$
-and $\chi_H$ in 
-$\softO (n^3)$ operations in $\OK$ at the precision given by $H.$
+and $\chi_M$ in 
+$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
 \end{theo}
 \begin{proof}
-First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
-in $\OK$ at the precision given by $H.$
-Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
-the only other nonzero coefficient in its column is $U_{i+1,i}$.
+Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
+in $O_K$ at the precision given by $H.$
+Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot:
+on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
 As a consequence, when performing this column-pivoting, only two rows ($i$ and
-$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
+$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
 Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
-Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
-The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
-no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
-for lines 2 and 3 is indeed $\softO (n^3)$ operations.
-
-For lines 4 and 5, there are only $n-1$ eliminations, resulting in a $\softO (n^2)$ cost
-for the computation of $U.$. Rather than actually construct $P$, we just track the eliminations
-performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.
+Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
+Corresponding coefficient for the corresponding shear matrix
+ is the opposite of the one defined by pivoting
+and the indices for the receiving and giving rows are swapped compared to the operations on columns.
+Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
+an upper-triangular matrix.
+This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.
+
+As for line 4 and 5, there are only $n-1$ eliminations, resulting again in a $\softO (n^3)$ cost
+for the computation of $P^{-1}$ and $U.$
 
@@
 \begin{rem}
-If $M \in M_n(K)$ does not have coefficients in $\OK,$
-we may apply Algorithms 1 and 2 to $p^v M \in M_n(\OK)$
-in $\softO (n^3)$ operations in $\OK$, and then divide
-the coefficient of $X^k$ in the resulting polynomial by $p^{kv}$.
+If $M \in M_n(K)$ does not have coefficients in $O_K,$
+then, by an appropriate scaling, we can still use
+Algorithms 1 and 2 to get $\chi_M$
+in $\softO (n^3)$ operations in $O_K$ at some shift
+of the initial precision.
 \end{rem}
@@
 
-In this section, we combine Proposition \ref{prop:shortcom}
-with Algorithm 2 to compute the comatrix of $X-M$
-when $\chi_M$ is squarefree.   Note that this condition on
-$\chi_M$ is equivalent to $M$ being diagonalizable under
-the assumption that $d\chi_M$ is surjective.  The result is
-the following $\softO(n^3)$ algorithm, where the only
-divisions are for gcd and modular inverse computations.
+In this Subsection, we are interested in a special
+case of the computation of the comatrix of $X-M,$
+when $\chi_M$ is square-free (which is the 
+generic case).
+We observe that it is possible to combine
+the result of Proposition \ref{prop:shortcom}
+with Algorithm 2 to actually compute
+the comatrix of $X-M.$
+It leads to a $\softO (n^3)$ algorithm
+where the only divisions are for
+gcd and modular inverse computations.
 
@@
 
-\noindent{\bf Input:} an approx. $M \in M_n(\OK),$ with $\disc(\chi_M) \neq 0.$
+\noindent{\bf Input:} an approx. $M \in M_n(O_K),$ with $discr(\chi_M) \neq 0.$
 
@@
 
-\noindent 0.\ Find $P \in GL_n(\OK)$ and $H \in M_n(\OK),$ approximate Hessenberg,
+\noindent 0.\ Find $P \in GL_n(O_K)$ and $H \in M_n(O_K),$ approximate Hessenberg,
 such that $M=PHP^{-1},$ using Algorithm 1.
 
-\noindent 1.\ Compute $A=\com (X-H)$ and $\chi(M) = \chi(H)$ using Algorithm 2.
-
-\noindent 2.\ Do $\row(A,1) \leftarrow \row(A,1)+\sum_{i=2}^n \mu_i \row(A,i),$ for
-random $\mu_i \in \OK,$ by doing $T \times A$ for some $T \in GL_n(\OK).$
+\noindent 1.\ Compute $A=\com (X-H)$ using Algorithm 2.
+
+\noindent 2.\ Do $row(A,1) \leftarrow row(A,1)+\sum_{i=2}^n \mu_i row(A,i),$ for
+random $\mu_i \in O_K,$ by doing $T \times A$ for some $T \in GL_n(O_K).$
 Compute $B:=TAT^{-1}.$
 
-\noindent 3.\ Similarily compute $C:=S^{-1}BS$ for $S \in GL_n(\OK)$ corresponding to
-adding a random linear combination of the columns of index $j \ge 2$
+\noindent 3.\ Similarily compute $C:=S^{-1}BS$ for $S \in GL_n(O_K)$ corresponding to
+sending a random linear combination of the columns of index $j \in \llbracket 2,n \rrbracket$
 to the first column of $B.$
@@
 
-\noindent 5. \ Compute $U,V \in \OK[X]^n$ such that $C=U {}^t V \mod \chi (M),$
+\noindent 5. \ Compute $U,V \in O_K[X]^n$ such that $C=U {}^t V \mod \chi (M),$
 by inverting $C_{1,1} \mod \chi(M).$
@@
 \begin{theo}
-For $M \in M_n(\OK)$ such that $\disc( \chi_M) \neq 0,$
-Algorithm 3 computes
+For $M \in M_n(O_K)$ such that $discr( \chi_M) \neq 0,$
+Algorithm 3 compute 
 $\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
-operations in $K$.
+operations in $K$ at initial precision.
 The only divisions are in steps 4 and 5.
@@
 \begin{proof}
-As we have already seen, completing Steps 0 and 1 is in $\softO (n^3).$
+As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
 Multiplying by $T$ or $S$ or their inverse corresponds
 to $n$ operations on rows or columns over a matrix with coefficients
-in $\OK[X]$ of degree at most $n.$
+in $O_K[X]$ of degree at most $n.$
 Thus, it is in $\softO (n^3).$
@@
 to scale $M \in M_n(K)$ so as to
-get coefficients in $\OK$ and apply the previous algorithm.
+get coefficients in $O_K$ and apply the previous algorithm.
 \end{rem}
@@
 \begin{prop} \label{prop:optimal_jagged}
-If $M \in M_n(\OK)$ is given at (high enough) jagged precision, 
+If $M \in M_n(O_K)$ is given at (high enough) jagged precision, 
 then we can compute the optimal jagged precision on $\chi_M$ in 
@@
 \begin{rem}
-If $M \in M_n(\OK)$ is given at (high enough) \emph{flat} precision, 
+If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
 then we can avoid the final base change step in the Hessenberg method.
@@
 
-\begin{rem}
+\begin{rem} \label{rem:lift_for_optimal}
 As a consequence of the previous discussion, once the optimal jagged 
 precision is known, it is possible to lift the entries of $M$ to a sufficiently
-large precision and then use 
-and then to perform a computation of the characteristic polynomial 
-at the optimal precision. This requires $\softO(n^3)$ operations in $K$ 
-but, for several instances, may require to increase a lot the precision.
+large precision, rescale them to have entries in $O_K$
+and then use Algorithm 2 to
+compute the characteristic polynomial.
+The output might then need to be rescaled and
+ truncated at the optimal precision. 
+This requires $\softO(n^3)$ operations in $O_K$ 
+and unfortunately, for several instances, may require to increase a lot the precision.
 \end{rem}

### 2017-02-04 charpoly/charpoly.tex
Minor stuffs.

@@
 on $\chi_M$ when $M \in M_n(\mathbb{Z}_p).$
-\item A $\softO (n^3)$ algorithm with operations in $O_K$
+\item A $\softO (n^3)$ algorithm with operations in $\Zp$
 to compute the optimal precision on each coefficient of $\chi_M$ when $M$
@@
 optimal precision on each coefficient of $\chi_M$
-for $M \in M_n (K),$
+for $M \in M_n (\Qp),$
 using divisions.

### 2017-02-04 charpoly/charpoly.tex
Saturday evening upload

--- 2017-02-04 charpoly/charpoly.tex
@@
 
-\noindent 5. \ Let $F$ be the inverse of $C_{1,1} \mod \chi_M$.
-
-\noindent 6. \ Let $U = \col(C,1)$ and $V = F \cdot \row(C,1) \mod \chi_M$.
-
-\noindent 7. \ Return $\com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi_M.$
+\noindent 5. Let $F$ be the inverse of $C_{1,1} \mod \chi_M$.
+
+\noindent 6. Let $U := \col(C,1)$ and $V := F \cdot \row(C,1) \mod \chi_M$.
+
+\noindent 7. Return $\com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi_M.$
 
@@
 Algorithm 3 computes
-$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
+$\com (X-M) \pmod{\chi_M}$ in average complexity $\softO (n^3)$
 operations in $K$.
@@
 Thus, it is in $\softO (n^3).$
-Step 5 and 6 are in $\softO (n).$
+Step 5 is in $\softO (n),$ Step 6 in $\softO(n^2)$ and Step 7 in $\softO(n^3)$.
 All that is to prove is that the set of $P$ and $S$ to avoid
@@
 The goal of the Step $2$ is to ensure the first row of $B$ contains an
-invertible modulo $\chi (M).$
+invertible entry modulo $\chi_M.$
 Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
@@
 of hyperplane, and hence, is of dimension at most $n-1.$
-Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
-on Step 4.
-Now, it is clear that the only divisions happening are on steps 4 and 5.
+Thus, almost any choice of $\mu_i$ leads to a matrix $C$ passing the test
+in Step 4.
 This concludes the proof.
@@
 We refer to \cite{caruso:15a} for the handling
-of the precision on gcd and modular inverse computations.
+of the precision of gcd and modular inverse computations.
 In this article, ways to tame the loss of precision

### 2017-02-04 charpoly/charpoly.tex
Comment \todo

@@
 \subsection*{Previous contributions.} 
-\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
+%\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
 
@@
 
-Thanks to the application the framework of differential precision of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
-we known that
-the precision on the characteristic polynomial
-$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
-is determined by the comatrix $com(X I_n -M).$
+Thanks to the application the framework of differential precision of 
+\cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a}, we known 
+that the precision on the characteristic polynomial $\chi_M$ of a matrix 
+$M \in M_n(\mathbb{Q}_p)$ is determined by the comatrix $com(X I_n -M).$ 
 In this article, we provide:
 \begin{enumerate}
-\item A factorization of $com(X I_n -M)$ (when $M$ has a cyclic vector).
+\renewcommand{\itemsep}{0pt}
+%\item A factorization of $com(X I_n -M)$ (when $M$ has a cyclic vector).
 \item A simple criterion to decide whether there can be a gain in precision
 on $\chi_M$ when $M \in M_n(\mathbb{Z}_p).$
-\item A $\softO (n^3)$ algorithm with operations in $\Zp$
-to compute the optimal precision on each coefficient of $\chi_M$ when $M$
-is given with uniform precision on its entries.
+\item A $\softO (n^3)$ algorithm with operations in $\Zp$ to compute the 
+optimal precision on each coefficient of $\chi_M$ when $M$ is given with 
+uniform precision on its entries.
 \item A $\softO (n^\omega)$ algorithm to compute the 
 optimal precision on each coefficient of $\chi_M$
-for $M \in M_n (\Qp),$
-using divisions.
+for $M \in M_n (\Qp)$, using divisions.
+\item A $\softO (n^\omega)$ algorithm to compute the 
+optimal precision on an eigenvalue of $M$.
 \end{enumerate}

### 2017-02-04 charpoly/charpoly.tex
Additional updates before submission

--- 2017-02-04 charpoly/charpoly.tex
@@
 \subsection*{Previous contributions.}
-\todo{Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?}
 
@@
 over $p$-adic numbers
-has become a crucial point in many counting-points algorithms.
-We can cite for example \cite{harvey:07a, harvey:14a} for
-algorithms to count points on hyperelliptic curves.
+has become a crucial ingredient in many counting-points algorithms.
+For example, \cite{harvey:07a, harvey:14a} use $p$-adic cohomology
+and the characteristic polynomial of Frobenius to compute zeta functions
+of hyperelliptic curves.
 
 In most of these papers, the precision analysis usually
-deals with great details on how to obtain the matrices (\textit{e.g.} of action of Frobenius)  that are involved in the point-counting schemes.
+deals with great details on how to obtain the matrices (\textit{e.g.} of action of Frobenius)
+that are involved in the point-counting schemes.
 However, the computation of their characteristic polynomials is often
 a little bit less thoroughly studied: some
-refer to fast algorithms (using divisions), while others
-refer to division-free algorithms.
+refer to fast algorithms (using division), while others
+apply division-free algorithms.
 
@@
 of the theory of differential precision of \cite{caruso-roe-vaccon:14a}
-to stable computation of characteristic polynomials.
+to the stable computation of characteristic polynomials.
 They have obtained a way to express the optimal precision
-on the characteristic polynomials, but have not studied 
-how to attain this optimal precision algorithmicaly.
+on the characteristic polynomial, but have not given practical algorithms
+to attain this optimal precision.
 
@@
 
-Thanks to the application the framework of differential precision of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
-we known that
-the precision on the characteristic polynomial
+Thanks to the application the framework of differential precision
+of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
+we know that the precision of the characteristic polynomial
 $\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
-is determined by the comatrix $com(X I_n -M).$
+is determined by the comatrix $\com(X -M).$
 In this article, we provide:
 \begin{enumerate}
-\item A factorization of $com(X I_n -M)$ (when $M$ has a cyclic vector).
-\item A simple criterion to decide whether there can be a gain in precision
-on $\chi_M$ when $M \in M_n(\mathbb{Z}_p).$
-\item A $\softO (n^3)$ algorithm with operations in $\Zp$
-to compute the optimal precision on each coefficient of $\chi_M$ when $M$
-is given with uniform precision on its entries.
-\item A $\softO (n^\omega)$ algorithm to compute the 
-optimal precision on each coefficient of $\chi_M$
-for $M \in M_n (\Qp),$
-using divisions.
+\renewcommand{\itemsep}{0pt}
+\item Proposition \ref{prop:shortcom}: a factorization of $\com(X -M)$
+as a product of two rank-$1$ matrices (when $M$ has a cyclic vector),
+computable in $\softO(n^\omega)$ operations by Theorem \ref{thm:compute_shortcom}
+\item Corollary \ref{cor:prec_gain}: a simple, $\softO(n^\omega)$ criterion
+to decide whether $\chi_M$ is defined at precision
+higher than the precision of $M$ (when $M \in M_n(\mathbb{Z}_p)$).
+\item Theorem \ref{thm:alg3}: a $\softO(n^3)$ algorithm with operations in $\Zp$
+to compute the optimal precision on each coefficient of $\chi_M$ (when $M$
+is given with uniform precision on its entries).
+\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
+algorithm to compute the optimal precision on each eigenvalue of $M$
 \end{enumerate}
@@
 
-In Section \ref{sec:theo_study}, we study the
-differential of the characteristic polynomial so as
-to get optimal information on precision 
-and a compact writing of this differential.
-Then, in Sections \ref{sec:diffHess} and \ref{sec:diffFrob}
-we propose algorithms to compute this differential
-either through Hessenberg or going through
-some Frobenius normal form.
+In Section \ref{sec:theo_study}, we review the differential
+theory of precision developed in \cite{caruso-roe-vaccon:14a}
+and apply it to the specific case of the characteristic polynomial,
+giving conditions under which the differential will be surjective
+(and thus provide a good measure of precision).  We also give
+a condition based on reduction modulo $p$ that determines
+whether the characteristic polynomial will have a higher
+precision that the input matrix, and show that the image
+of the set of integral matrices has the structure of an $\OK[X]$-module
+when $M$ is itself integral. Finally, we give a compact description
+of $\com(X-M)$, the main ingredient in the differential.
+
+In Section \ref{sec:diffHess}, we develop $\softO(n^3)$ algorithms to approximate
+the Hessenberg form of $M$, and through it to find $\com(X-M)$ and thus
+find the precision of the characteristic polynomial of $M$.  In Section \ref{sec:diffFrob},
+we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\com(X-M)$.
+
 Finally, we propose in Section \ref{sec:optjagged}
-algorithms to compute the optimal precision
-coefficient-wise on the characteristic polynomial
-and on some eigenvalues along with
-numerical evidence of the behaviour of the computation
-of the characteristic polynomial.
+algorithms to compute the optimal coefficient-wise precision
+for the characteristic polynomial.  We also give the results
+of some experiments demonstrating that these methods can lead
+to dramatic gains in precision over standard interval arithmetic.
+We close with results describing the precision associated to
+eigenvalues of a matrix.
 
@@
 We will write that $f(n)=\softO (g(n))$ if there exists some
-$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k.$
+$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k).$
 We fix an $\omega \in \mathbb{R}$ such that the multiplication
 of two matrices over a ring is in $O(n^\omega)$ operations in the
-ring. Currently, the smallest proved $\omega$ is less than
+ring. Currently, the smallest known $\omega$ is less than
 $2.3728639 $ thanks to \cite{legall:14a}. 
-$I_n$ will denote the identity matrix of rank $n$ in $M_n(K).$
+We will denote by $I_n$ the identity matrix of rank $n$ in $M_n(K).$
 When there is no ambiguity, we will drop this $I_n$
 for scalar matrices, \textit{e.g.} for $\lambda \in K$
-and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n -M.$
+and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n - M.$
 Finally, we write $\sigma_1(M), \dots, \sigma_n(M)$ for the
@@
 Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
-We assume that $A=\com (X-M)$ is such that $gcd(A_{1,1}, \chi (M))=1$
+We assume that $A=\com (X-M)$ is such that $\gcd(A_{1,1}, \chi (M))=1$
 (in $K[X]$).
@@
 Nevertheless, we can remark that
- $gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
+ $\gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
 and thus, $A(\lambda)$ is of rank exactly one.
@@
 \begin{rem}
-Keeping the previous notation, if $U \in GL_n(K),$
-then the corresponding quadruple for 
+With the previous notation, if $U \in GL_n(K),$
+the quadruple for 
 $UMU^{-1}$ is
-$(\alpha, UP, U^{t,-1}Q, \chi_M),$
+$(\alpha, UP, (U^t)^{-1}Q, \chi_M),$
 which can be computed in $O(n^\omega)$ operations in $K.$
@@
 
-\noindent 4.\  \textbf{If} $gcd(C_{1,1}, \chi_M) \neq 1,$ \textbf{then} go to 2.
+\noindent 4.\  \textbf{If} $\gcd(C_{1,1}, \chi_M) \neq 1,$ \textbf{then} go to 2.
 
@@
 \begin{theo}
+\label{thm:alg3}
 For $M \in M_n(\OK)$ such that $\disc( \chi_M) \neq 0,$
@@
 \end{align*}
-
-$\chi_M$ can be computed in $\softO(n^\omega)$ operations in $K$
+An approximation to $\chi_M$ can be computed in $\softO(n^\omega)$ operations in $K$
 (\textit{e.g.} as a by-product of \cite{storjohann:01a}).
@@
 $M$ takes the shape \eqref{eq:companion}. In other words, if $P_\inv$
-is invertible, then $P = P_\inv^{-1}$ is a solution of $M = P N P^{-1}$.
-Observe moreover that the condition ``$P_\inv$ is invertible'' is open
+is invertible, then $P = P_\inv^{-1}$ is a solution of $M = P \mathscr{C} P^{-1}$,
+where $\mathscr{C}$ is the companion matrix similar to $M$.
+Moreover, observe that the condition ``$P_\inv$ is invertible'' is open
 for the Zariski topology. It then happens with high probability as soon
@@
 The characteristic polynomial $\chi_M$ can be recovered thanks to the
-relation $(a_0, a_1, \ldots, a_{n-1}) = -c_n \cdot P$.
-
-Now, instead of computing directly $Q$, we first compute a matrix $R$ 
-with the property that $N^t = R N R^{-1}$. For that, we apply the same
-strategy as above except that we start with the vector $e = (1, 0, 
-\ldots, 0)$ (and not with a random vector). A simple computation shows
-that, for $1 \leq i \leq n{-}1$, the vector $N^i e$ has the shape:
-$$N^i e = (0, \ldots, 0, -a_0, \star, \ldots, \star)$$
-with $n{-}i$ starting zeros. Therefore the $N^i e$'s form a basis of
-$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $N$. Once $R$ has
+relation $a_0c_0 + a_1c_1 + \dots + a_{n-1}c_{n-1} = -c_{n-1} \cdot P$.
+
+Now, instead of directly computing $Q$, we first compute a matrix $R$ 
+with the property that $\mathscr{C}^t = R \mathscr{C} R^{-1}$. To do so,
+we apply the same strategy as above except that we start with the vector
+$e = (1, 0, \ldots, 0)$ (and not with a random vector). A simple computation shows
+that, for $1 \leq i \leq n{-}1$, the vector $\mathscr{C}^i e$ has the shape:
+$$\mathscr{C}^i e = (0, \ldots, 0, -a_0, \star, \ldots, \star)$$
+with $n{-}i$ starting zeros. Therefore the $\mathscr{C}^i e$'s form a basis of
+$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $\mathscr{C}$. Once $R$ has
 been computed, we recover $Q$ using the relation $Q = P_\inv^t R$.
@@
 \label{eq:shortcomN}
-\com(X{-}N) = \alpha \cdot V^t \cdot V R^t \mod \chi_M
+\com(X{-}\mathscr{C}) = \alpha \cdot V^t \cdot V R^t \mod \chi_M
 \end{equation}
@@
 $$\alpha = a_1 + a_2 X + \cdots + a_{n-1} X^{n-2} + X^{n-1}.$$
-No further computation are then needed to derive the value of $\alpha$
-and we can conclude with the following theorem.
+No further computation are then needed to derive the value of $\alpha$.
+We summarize this section with the following theorem:
 
 \begin{theo}
+\label{thm:compute_shortcom}
 Given $M \in M_n(K)$ such that $d \chi_M$ is surjective,
@@
 be able to understand what is the optimal precision on $\chi_M.$
-In this section, we provide some answers to this question of
-the optimal precision on $\chi_M,$ along with
-numerical evidence.
+In this section, we provide some answers to this question,
+along with numerical evidence.
 We also show that it is then possible to derive 
-optimal precision on eigenvalues of $M.$
+optimal precision of eigenvalues of $M.$
 

### 2017-02-04 charpoly/charpoly.tex

@@
 $k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k).$
+We will write $M$ for an $n \times n$ matrix over $K$,
+and $\chi$ the characteristic polynomial map, $\chi_M \in K[X]$ for
+the characteristic polynomial of $M$ and $d\chi_M$ for the differential
+of $\chi$ at $M$, as a linear map from $M_n(K)$ to the space
+of polynomials of degree less than $n$.
 We fix an $\omega \in \mathbb{R}$ such that the multiplication
@@
 For a matrix $M \in M_n(K)$, we identify the tangent space $V$
-at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
+at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi_M$ with
 the space $K_{<n}[X]$ of polynomials of degree less than $n$.
@@
 occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
-$B_W(1) \subset d\chi(M)(B_V(\alpha^{-1}))$.
-
-Since the coefficients of $\chi(M)$ are given by polynomials in the entries of $M$
+$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.
+
+Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
 with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
@@
 the $\OK$-span of the entries of $C$.   Using the fact that the product of matrix
-with its comatrix is the determinant, $(X - M) \cdot C = \chi(M)$ and thus
+with its comatrix is the determinant, $(X - M) \cdot C = \chi_M$ and thus
 $P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
@@
 $P\in GL_n (\OK)$ then
-\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi(M).\]
+\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi_M.\]
 In other words, factorizability of $\com(X-M)$ is stable
@@
 Similarly, the goal of Step 3 is to put an invertible coefficient (modulo
-$\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine
+$\chi_M$) on $C_{1,1},$ and again, only a finite union of affine
 hyperplane is to avoid.

### 2017-02-12 book/course.tex
Course of p-adics (given at JNCF)

\documentclass[11pt]{article}

\usepackage[utf8x]{inputenc}
\usepackage[english]{babel}
\usepackage[a4paper,top=2.5cm,left=2.5cm,right=2.5cm,bottom=2.5cm]{geometry}
\usepackage{charter}
\usepackage{microtype}
\usepackage{tikz,pgfplots}
\usepackage{url}
\usepackage[plainpages=false,pdfpagelabels,colorlinks=true,citecolor=blue,hypertexnames=false]{hyperref}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{stmaryrd,dsfont}
\usepackage{enumerate,array,multirow}
\usepackage{listings}
\usepackage{comment}
\usepackage{xspace}
%\usepackage{showkeys}

\numberwithin{equation}{section}
\numberwithin{figure}{section}

\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{digit}{rgb}{0.3,0,0.5}
\definecolor{beige}{rgb}{0.9,0.9,0.7}
\definecolor{comment}{rgb}{0.8,0.3,0.3}

\makeatletter
\let\commentfullflexible\lst@column@fullflexible
\makeatother
\lstset{
  basicstyle=\tt\small,
  commentstyle=\commentfullflexible\color{comment}\rm,
  keywordstyle=\color{blue}\bf,
  escapeinside={(*}{*)},
  language=python,
  morekeywords={lift,curry,uncurry},
  emph={self},emphstyle={\color{black!50}},
  emph={[2]digits,carry},emphstyle={[2]\color{darkgreen}},
}

\def\todo#1{{\color{red}#1}}

\def\ph{\vphantom{$A^A_A$}}

\makeatletter
\setcounter{secnumdepth}{3}
\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
                                    {2ex \@plus1ex \@minus.2ex}%
                                    {1ex}%
                                    {\normalfont\normalsize\bfseries}}
\renewcommand\subparagraph{\@startsection{subparagraph}{5}{\z@}%
                                       {1.5ex \@plus1ex \@minus .2ex}%
                                       {-0.5em}%
                                      {\normalfont\normalsize\it}}
\makeatother

\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}

\newtheorem{theointro}{Theorem}
\newtheorem{theo}{Theorem}[section]
\newtheorem{prop}[theo]{Proposition}
\newtheorem{lem}[theo]{Lemma}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{conj}[theo]{Conjecture}
\theoremstyle{definition}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem*{ex*}{Example}

\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\F}{\mathbb F}
\newcommand{\Fp}{\F_p}
\newcommand{\R}{\mathbb R}
\newcommand{\m}{\mathfrak m}
\renewcommand{\P}{\mathbb P}
\newcommand{\E}{\mathbb E}

\newcommand{\calB}{\mathcal B}
\newcommand{\calF}{\mathcal F}
\newcommand{\calH}{\mathcal H}
\newcommand{\calL}{\mathcal L}
\newcommand{\calN}{\mathcal N}
\newcommand{\calR}{\mathcal R}
\newcommand{\calT}{\mathcal T}
\newcommand{\calU}{\mathcal U}
\newcommand{\calZ}{\mathcal Z}

\newcommand{\GL}{\text{\rm GL}}
\newcommand{\SL}{\text{\rm SL}}

\newcommand{\Frac}{\text{\rm Frac}\:}
\newcommand{\spec}{\text{\rm Spec}\:}
\newcommand{\Diag}{\text{\rm Diag}}
\newcommand{\Span}{\text{\rm Span}}
\renewcommand{\min}{\text{\rm min}}
\renewcommand{\max}{\text{\rm max}}
\newcommand{\pr}{\text{\rm pr}}
\newcommand{\symm}{\text{\rm s}}

\newcommand{\id}{\text{\rm id}}
\newcommand{\ev}{\text{\rm ev}}
\newcommand{\rec}{\text{\rm rec}}
\newcommand{\val}{\text{\rm val}}
\newcommand{\triv}{\text{\rm triv}}
\newcommand{\abs}{\text{\rm abs}}
\newcommand{\rel}{\text{\rm rel}}
\newcommand{\Res}{\text{\rm Res}}
\newcommand{\Com}{\text{\rm Comatrix}}

\newcommand{\zealous}{\text{\rm Zeal}}

\newcommand{\Value}{\text{\rm value}}
\newcommand{\ttx}{\texttt{\rm x}\xspace}
\newcommand{\tty}{\texttt{\rm y}\xspace}
\newcommand{\ttF}{\texttt{\rm F}\xspace}
\newcommand{\ttN}{\texttt{\rm N}\xspace}

\newcommand{\nan}{\text{\rm NaN}}
\newcommand{\FP}{\text{\rm FP}}
\newcommand{\QpFP}{\Qp^{\FP}}
\newcommand{\QpNan}{\Qp'}

\newcommand{\cond}{\text{\rm cond}}

\newcommand{\softO}{\tilde O}

\newcommand{\sage}{\textsc{SageMath}~\cite{sage}\xspace}
\newcommand{\sagenoref}{\textsc{SageMath}\xspace}
\newcommand{\magma}{\textsc{Magma}~\cite{magma}\xspace}
\newcommand{\pari}{\textsc{Pari}~\cite{pari}\xspace}
\newcommand{\maple}{\textsc{Maple}\xspace}
\newcommand{\mathemagix}{\textsc{Mathemagix}~\cite{mathemagix}\xspace}

\title{Computations with $p$-adic numbers}
\author{Xavier Caruso}
\date\today

\begin{document}

\maketitle

\begin{abstract}
This document contains the notes of a lecture I gave at the ``Journées 
Nationales du Calcul Formel\footnote{(French) National Computer Algebra 
Days}'' (JNCF) on January 2017. The aim of the lecture was to discuss 
low-level algorithmics for $p$-adic numbers. It is divided into two main 
parts: first, we present various implementations of $p$-adic numbers and 
compare them and second, we introduce a general framework for studying 
precision issues and apply it in several concrete situations.
\end{abstract}

\setcounter{tocdepth}{2}
\tableofcontents

\section*{Introduction}

The field of $p$-adic numbers, $\Qp$, was first introduced by 
Kurt Hensel at the end of the 19th century in a short paper written in 
German~\cite{He97}. From that time, the popularity of $p$-adic numbers 
has grown without interruption throughout the 20th century. Their first 
success was materialized by the famous Hasse--Minkowski's Theorem 
\cite{Se70} that 
states that a Diophantine equation of the form $P(x_1, \ldots, x_n) = 0$ 
where $P$ is a polynomial of total degree at most $2$ has a solution over $\Q$ 
if and only if it has a solution over $\R$ and a solution over $\Qp$ for 
all prime numbers $p$. This characterization is quite interesting because 
testing whether a polynomial equation has a $p$-adic solution can be 
carried out in a very efficient way using \emph{analytic} methods just 
like over the reals.
This kind of strategy is nowadays ubiquitous in many areas of Number 
Theory and Arithmetic Geometry. After Diophantine equations, other 
typical examples come from the study of number fields: we hope deriving 
interesting information about a number field $K$ by studying carefully
all its $p$-adic incarnations $K \otimes_\Q \Qp$. The ramification of
$K$, its Galois properties, \emph{etc.} can be --- and are very often 
--- studied in this manner~\cite{Sa72,Ne99}.
The class field theory, which provides a precise description of all
Abelian extensions\footnote{An abelian extension is a Galois extension
whose Galois group is abelian.} of a given number field, is also 
formulated in this language~\cite{Ne12}.
The importance of $p$-adic numbers is so prominent today that there is 
still nowadays very active research on theories which are dedicated to 
purely $p$-adic objects: one can mention for instance the
study of $p$-adic geometry and $p$-adic cohomologies~\cite{BO78,LS07},
the theory of $p$-adic differential equations~\cite{Ke10}, Coleman's theory 
of $p$-adic integration~\cite{Co98}, the $p$-adic Hodge theory~\cite{BrCo09}, 
the $p$-adic Langlands correspondence~\cite{Be11}, the study of $p$-adic 
modular forms~\cite{Go88}, $p$-adic $\zeta$-functions~\cite{Ko84} and 
$L$-functions~\cite{Co89}, \emph{etc.} The proof of Fermat's last 
Theorem by Wiles and Taylor~\cite{Wi95,TaWi95} is stamped with many of 
these ideas and developments.

Over the last decades, $p$-adic methods have taken some importance in 
Symbolic Computation as well. For a long time, $p$-adic methods have 
been used for factoring polynomials over $\Q$~\cite{LeLeLo82}. More recently, 
there has been a wide diversification of the use of $p$-adic numbers for 
effective computations: Bostan et al.~\cite{BoGoPeSc05} used Newton sums 
for polynomials over $\Zp$ to compute composed products for polynomials 
over $\Fp$; Gaudry et al.~\cite{GaHoWeRiKo06} used $p$-adic lifting 
methods to generate genus 2 CM hyperelliptic curves; Kedlaya 
\cite{Ke01}, Lauder~\cite{La04} and many followers used $p$-adic 
cohomology to count points on hyperelliptic curves over finite fields; 
Lercier and Sirvent~\cite{LeSi08} computed isogenies between elliptic 
curves over finite fields using $p$-adic differential equations.

The need to build solid foundations to the algorithmics of $p$-adic 
numbers has then emerged. This is however not straightforward because a 
single $p$-adic number encompasses an infinite amount of information 
(the infinite sequence of its digits) and then necessarily needs to be 
truncated in order to fit in the memory of a computer. From this point 
of view, $p$-adic numbers behave very similarly to real numbers and the 
questions that emerge when we are trying to implement $p$-adic numbers 
are often the same as the questions arising when dealing with rounding 
errors in the real setting~\cite{Mu89,DaMu97,Mu09}. The algorithmic study 
of $p$-adic numbers is then located at the frontier between Symbolic 
Computation and Numerical Analysis and imports ideas and results coming 
from both of these domains.

\subparagraph{Content and organization of this course.}

This course focuses on the low-level implementation of $p$-adic numbers 
(and then voluntarily omits high-level algorithms making use of $p$-adic 
numbers) and pursues two main objectives.
The first one is to introduce and discuss the most standard strategies 
for implementing $p$-adic numbers on computers. We shall detail three of 
them, each of them having its own spirit: (1)~the \emph{zealous arithmetic} 
which is inspired by interval arithmetic in the real setting, (2)~the 
\emph{lazy arithmetic} with its \emph{relaxed} improvement and (3)~the 
\emph{$p$-adic floating-point arithmetic}, the last two being inspired
by the eponym approaches in the real setting.

The second aim of this course is to develop a general theory giving 
quite powerful tools to study the propagation of accuracy in the 
$p$-adic world. The basic underlying idea is to linearize the situation 
(and then model the propagation of accuracy using differentials); it is 
once again inspired from classical methods in the real case. %~\cite{}. 
However, it turns out that the non-archimedean nature of $\Qp$ 
(\emph{i.e.} the fact that $\Z$ is bounded in $\Qp$) is the source of 
many simplifications which will allow us to state much more accurate 
results and to go much further in the $p$-adic setting. As an example,
we shall see that the theory of $p$-adic precision yields a general
strategy for increasing the numerical stability of any given algorithm
(assuming that the problem it solves is well-conditioned).

\medskip

This course is organized as follows. 
\S \ref{sec:padicnumbers} is devoted to the introduction of $p$-adic 
numbers: we define them, prove their main properties and discuss in
more details their place in Number Theory, Arithmetic Geometry and 
Symbolic Computation.
The presentation of the standard implementations of $p$-adic numbers 
mentioned above is achieved in \S \ref{sec:implementations}. 
A careful comparison between them is moreover proposed and supported
by many examples coming from linear algebra and commutative algebra.
Finally, in \S \ref{sec:precision}, we expose the aforementioned theory 
of $p$-adic precision. We then detail its applications: we will notably
examine many very concrete situations and, for each of them, we will
explain how the theory of $p$-adic precision helps us either in quantifying 
the qualities of a given algorithm regarding to numerical stability or,
even better, in improving them.

\subparagraph{Acknowledgments.}

This document contains the (augmented) notes of a lecture I gave at the 
``Journées Nationales du Calcul Formel\footnote{(French) 
National Computer Algebra Days}'' (JNCF) on January 2017. I heartily thank 
the organizers and the scientific committee of the JNCF for giving me 
the opportunity to give these lectures and for encouraging me to write 
down these notes. I am very grateful to Delphine Boucher, 
Nicolas Brisebarre, Claude-Pierre 
Jeannerod, Marc Mezzarobba and Tristan Vaccon for their careful reading
and their helpful comments on an earlier version of these notes.

\subparagraph{Notation.}

We use standard notation for the set of numbers: $\N$ is the set of 
natural integers (\emph{including $0$}), $\Z$ is the set of relative 
integers, $\Q$ is the set of rational numbers and $\R$ is the set of 
real numbers.
We will sometimes use the soft-$O$ notation $\softO(-)$ for writing 
complexities; we recall that, given a sequence of positive real numbers 
$(u_n)$, $\softO(u_n)$ is defined as the union of the sets $O(u_n \log^k 
u_n)$ for $k$ varying in $\N$.

Throughout this course, the letter $p$ always refers to a fixed prime 
number.

\section{Introduction to $p$-adic numbers}
\label{sec:padicnumbers}

In this first section, we define $p$-adic numbers, discuss their basic 
properties and try to explain, by selecting a few relevant examples, 
their place in Number Theory, Algebraic Geometry and Symbolic 
Computation. The presentation below is voluntarily very summarized; we 
refer the interested reader to~\cite{Am75,Go97} for a more complete 
exposition of the theory of $p$-adic numbers.

\subsection{Definition and first properties}

$p$-adic numbers are very ambivalent objects which can be thought of 
under many different angles: computational, algebraic, analytic. It 
turns out that each point of view leads to its own definition of 
$p$-adic numbers: computer scientists often prefer viewing a $p$-adic 
number as a sequence of digits while algebraists prefer speaking of 
projective limits and analysts are more comfortable with Banach spaces 
and completions. Of course all these approaches have their own interest 
and understanding the intersections between them is often the key behind 
the most important advances.

In this subsection, we briefly survey all the standard definitions of 
$p$-adic numbers and provide several mental representations in order to 
try as much as possible to help the reader to develop a good $p$-adic 
intuition.

\subsubsection{Down-to-earth definition}
\label{sssec:downtoearth}

Recall that each positive integer $n$ can be written in \emph{base $p$},
that is as a finite sum:
$$n = a_0 + a_1 p + a_2 p^2 + \cdots + a_\ell p^\ell$$
where the $a_i$'s are integers between $0$ and $p{-}1$, the so-called
\emph{digits}. This writing is moreover unique assuming that the most
significant digit $a_\ell$ does not vanish.
A possible strategy to compute the expansion in base $p$ goes as 
follows. We first compute $a_0$ by noting that it is necessarily the 
remainder in the Euclidean division of $n$ by $p$: indeed it is 
congruent to $n$ modulo $p$ and lies in the range $[0,p{-}1]$ by 
definition. Once $a_0$ is known, we compute $n_1 = \frac{n-a_0}p$, which 
is also the quotient in the Euclidean division of $n$ by $p$. Clearly
$n_1 = a_1 + a_2 p + \cdots + a_\ell p^{\ell-1}$
and we can now compute $a_1$ repeating the same strategy. 
Figure~\ref{fig:decbasis} shows a simple execution of this algorithm.

\begin{figure}
\hfill$\begin{array}{rr@{\hspace{1ex}}c@{\hspace{1ex}}r@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c}
a_0: & 1742 & = & 248 & \times & 7 & + & \mathbf{6} \\
a_1: & 248 & = & 35 & \times & 7 & + & \mathbf{3} \\
a_2: & 35 & = & 5 & \times & 7 & + & \mathbf{0} \\
a_3: & 5 & = & 0 & \times & 7 & + & \mathbf{5} \\
\end{array}$
\hfill\null

\medskip

\hfill $\Longrightarrow$ \quad $1742 = \overline{5036}^7$\hfill\null

\caption{Expansion of $1742$ in base $7$}
\label{fig:decbasis}
\end{figure}

By definition, a \emph{$p$-adic integer} is an infinite formal sum of 
the shape:
$$x = a_0 + a_1 p + a_2 p^2 + \cdots + a_i p^i + \cdots$$
where the $a_i$'s are integers between $0$ and $p{-}1$. In other words,
a $p$-adic integer 
is an integer written in base $p$ with an infinite number of 
digits. We will sometimes alternatively write $x$ as follows:
$$x = \overline{\ldots a_i \ldots a_3 a_2 a_1 a_0}^p$$
or simply
$$x = \ldots a_i \ldots a_3 a_2 a_1 a_0$$
when no confusion can arise.
The set of $p$-adic integers is denoted by $\Z_p$. It is endowed with a 
natural structure of commutative ring. Indeed, we can add, subtract and 
multiply $p$-adic integers using the schoolbook method; note that 
handling carries is possible since they propagate on the left.
%
\begin{figure}
\hfill
\raisebox{-\height}{
$\begin{array}{cc@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c}
  & \ldots & 2 & 3 & 0 & 6 & 2 & 4 & 4 \\{}
+ & \ldots & 1 & 6 & 5 & 2 & 3 & 3 & 2 \\
\hline
  & \mathbf \ldots & \mathbf 4 & \mathbf 2 & \mathbf 6 & \mathbf 1 & \mathbf 6 & \mathbf 0 & \mathbf 6 
\end{array}$}
\hspace{2cm}
\raisebox{-\height}{
$\begin{array}{cc@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c@{\hspace{1ex}}c}
  & \ldots & 2 & 3 & 0 & 6 & 2 & 4 & 4 \\{}
\times & \ldots & 1 & 6 & 5 & 2 & 3 & 3 & 2 \\
\hline
  & \ldots & 4 & 6 & 1 & 5 & 5 & 2 & 1 \\
  & \ldots & 2 & 2 & 5 & 0 & 6 & 5 & \\
  & \ldots & 2 & 5 & 0 & 6 & 5 & & \\
  & \ldots & 5 & 5 & 2 & 1 & & & \\
  & \ldots & 6 & 1 & 6 & & & & \\
  & \ldots & 6 & 3 & & & & & \\
  & \ldots & 4 & & & & & & \\
\cline{2-9}
  & \mathbf \ldots & \mathbf 4 & \mathbf 3 & \mathbf 2 & \mathbf 0 & \mathbf 3 & \mathbf 0 & \mathbf 1 
\end{array}$}
\hfill\null

\caption{Addition and multiplication in $\Z_7$}
\end{figure}
%
The ring of natural integers $\N$ appears naturally as a subring of 
$\Z_p$: it consists of $p$-adic integers $\ldots a_i \ldots a_3 a_2 a_1 
a_0$ for which $a_i = 0$ when $i$ is large enough. Note in particular 
that the integer $p$ writes $\ldots 0010$ in $\Z_p$ and more generally 
$p^n$ writes $\ldots 0010\ldots 0$ with $n$ ending zeros.
As a consequence, a $p$-adic integer is a multiple of $p^n$ if and 
only if it ends with (at least) $n$ zeros.
Remark that negative integers are $p$-adic integers as well: the 
opposite of $n$ is, by definition, the result of the subtraction 
$0{-}n$.

\bigskip

Similarly, we define a \emph{$p$-adic number} as a formal infinite sum of 
the shape:
$$x = a_{-n} p^{-n} + a_{-n+1} p^{-n+1} + \cdots + a_i p^i + \cdots$$
where $n$ is an integer which may depend on $x$. 
Alternatively, we will write:
$$x = \overline{\ldots a_i \ldots a_2 a_1 a_0\,.\, 
a_{-1} a_{-2} \ldots a_{-n}}^p$$
and, when no confusion may arise, we will freely remove the bar and the 
trailing $p$.
A $p$-adic number is then nothing but a ``decimal'' number written in 
base $p$ with an infinite number of digits before the decimal mark 
and a
finite amount of digits after the decimal mark. Addition and multiplication
extend to $p$-adic numbers as well.

The set of $p$-adic numbers is denoted by $\Q_p$. Clearly $\Q_p = 
\Z_p[\frac 1 p]$. We shall see later (\emph{cf} Proposition 
\ref{prop:invZp}, page \pageref{prop:invZp}) that $\Q_p$ is actually the 
fraction field of $\Z_p$; in particular it is a field and $\Q$, which is 
the fraction field of $\Z$, naturally embeds into $\Q_p$.

\subsubsection{Second definition: projective limits}

From the point of view of addition and multiplication, the last 
digit of a $p$-adic integer behaves like an integer modulo $p$, that is 
an element of the finite field $\Fp = \Z/p\Z$.
In other words, the application $\pi_1 : \Z_p \to \Z/p\Z$ taking a 
$p$-adic integer $x = a_0 + a_1 p + a_2 p^2 + \cdots$ to the class of 
$a_0$ modulo $p$ is a ring homomorphism.
More generally, given a positive integer $n$, the map:
$$\begin{array}{rcl}
\pi_n :\quad \Z_p & \to & \Z/p^n\Z \\
a_0 + a_1 p + a_2 p^2 + \cdots & \mapsto &
(a_0 + a_1 p + \cdots + a_{n-1} p^{n-1}) \text{ mod } p^n
\end{array}$$
is a ring homomorphism. These morphisms are compatible in the following 
sense: for all $x \in \Z_p$, we have $\pi_{n+1}(x) \equiv \pi_n(x) 
\pmod{p^n}$ (and more generally $\pi_m(x) \equiv \pi_n(x) \pmod{p^n}$ 
provided that $m \geq n$). Putting the $\pi_n$'s all together, we end
up with a ring homomorphism:
$$\begin{array}{rcl}
\pi: \quad \Z_p & \to & \varprojlim_n \Z/p^n\Z \smallskip \\
x & \mapsto & (\pi_1(x), \pi_2(x), \ldots)
\end{array}$$
where $\varprojlim_n \Z/p^n\Z$ is by definition the subring of 
$\prod_{n=1}^\infty \Z/p^n\Z$ consisting of sequences $(x_1, x_2,
\ldots)$ for which $x_{n+1} \equiv x_n \pmod{p^n}$ for all $n$: it
is called the \emph{projective limit} of the $\Z/p^n\Z$'s.

Conversely, consider a sequence $(x_1, x_2, \ldots) \in \varprojlim_n 
\Z/p^n\Z$. In a slight abuse of notation, continue to write $x_n$ for 
the unique integer of the range $\llbracket 0, p^n{-}1 \rrbracket$ 
which is congruent to 
$x_n$ modulo $p^n$ and write it in base $p$:
$$x_n = a_{n,0} + a_{n,1} p + \cdots + a_{n,n-1} p^{n-1}$$
(the expansion stops at $(n{-}1)$ since $x_n < p^n$ by construction).
The condition $x_{n+1} \equiv x_n \pmod{p^n}$ implies that 
$a_{n+1,i} = a_{n,i}$ for all $i \in \llbracket 0, n{-}1\rrbracket$.
In other words, when $i$ remains fixed, the sequence $(a_{n,i})_{n > i}$ 
is constant and thus converges to some $a_i$. Set:
$$\psi(x_1, x_2, \ldots) = \ldots a_i \ldots a_2 a_1 a_0 \in \Z_p.$$
We define this way an application $\psi : \varprojlim_n \Z/p^n\Z \to 
\Z_p$ which is by construction a left and a right inverse of $\pi$. In
other words, $\pi$ and $\psi$ are isomorphisms which are inverses of
each other.

The above discussion allows us to give an alternative definition of 
$\Z_p$, which is:
$$\Z_p = \varprojlim_n \, \Z/p^n\Z.$$
The map $\pi_n$ then corresponds to the projection onto the $n$-th factor.
This definition is more abstract and it seems more difficult to 
handle as well. However it has the enormous advantage of making the ring 
structure appear clearly and, for this reason, it is often much more
useful and powerful than the down-to-earth definition of \S 
\ref{sssec:downtoearth}.
As a typical example, let us prove the following proposition.

\begin{prop}
\label{prop:invZp}
\begin{enumerate}[(a)]
\renewcommand{\itemsep}{0pt}
\item An element $x \in \Z_p$ is invertible in $\Z_p$ if and only
if $\pi_1(x)$ does not vanish.
\item The ring $\Q_p$ is the fraction field of $\Z_p$; in particular,
it is a field.
\end{enumerate}
\end{prop}

\begin{proof}
\emph{(a)}~Let $x \in \Z_p$.
Viewing $\Z_p$ as $\varprojlim_n \, \Z/p^n\Z$, we find that $x$ 
is invertible in $\Z_p$ if and only if $\pi_n(x)$ is invertible in
$\Z/p^n\Z$ for all $n$. The latest condition is equivalent to
requiring that $\pi_n(x)$ and $p^n$ are coprime for all $n$. 
Noting that $p$ is prime, this is further equivalent to the fact that 
$\pi_n(x) \text{ mod } p = \pi_1(x)$ does not vanish in $\Z/p\Z$.

\smallskip

\noindent
\emph{(b)}~By definition $\Q_p = \Z_p[\frac 1 p]$. It is then enough
to prove that any nonzero $p$-adic integer $x$ can be written as a
product $x = p^n u$ where $n$ is a nonnegative integer and $u$ is a
unit in $\Z_p$. Let $n$ be the number of zeros at the end of the 
$p$-adic expansion of $x$ (or, equivalently, the largest integer $n$
such that $\pi_n(x) = 0$). Then $x$ can be written $p^n u$ where $u$
is a $p$-adic integer whose last digit does not vanish. By the first
part of the proposition, $u$ is then invertible in $\Z_p$ and we are
done.
\end{proof}

We note that the first statement of Proposition~\ref{prop:invZp}
shows that the subset of non-invertible elements of $\Zp$ is exactly
the kernel of $\pi_1$. We deduce from this that $\Zp$ is a local ring
with maximal ideal $\ker \pi_1$.

\subsubsection{Valuation and norm}
\label{sssec:padicnorm}

We define the \emph{$p$-adic valuation} of the nonzero $p$-adic 
number 
$$x = \ldots a_i \ldots a_2 a_1 a_0\,.\, a_{-1} a_{-2} \ldots a_{-n}$$
as the smallest (possibly negative) integer $v$ for which $a_v$ 
does not vanish. We denote it $\val_p(x)$ or simply $\val(x)$ if no
confusion may arise. Alternatively $\val(x)$ 
can be defined as the largest integer $v$ such that $x \in p^v \Zp$. 
When $x = 0$, we put $\val(0) = +\infty$.
We define this way a function $\val : \Q_p \to \Z \cup \{+\infty\}$.
Writing down the computations (and remembering that $p$ is prime),
we immediately check the following compatibility properties for all
$x,y \in \Qp$:
\begin{enumerate}[(1)]
\renewcommand{\itemsep}{0pt}
\item $\val(x+y) \geq \min\big(\val(x), \val(y)\big)$,
\item $\val(xy) = \val(x) + \val(y)$.
\end{enumerate}
Note moreover that the equality $\val(x+y) 
= \min\big(\val(x), \val(y)\big)$ does hold as soon as $\val(x)
\neq \val(y)$. As we shall see later, this property reflects the
tree structure of $\Z_p$ (see \S \ref{sssec:tree}).

The \emph{$p$-adic norm} $|\cdot|_p$ is defined by
$|x|_p = p^{-\val(x)}$ for $x \in \Q_p$. In the sequel, when no
confusion can arise, we shall often write $|\cdot|$ instead of
$|\cdot|_p$. The properties (1) and (2) above immediately translate 
as follows:
\begin{enumerate}[(1)]
\renewcommand{\itemsep}{0pt}
\item[(1')] $|x+y| \leq \max\big(|x|, |y|\big)$ and equality holds
if $|x| \neq |y|$,
\item[(2')] $|xy| = |x| \cdot |y|$.
\end{enumerate}
Remark that (1') implies that $|\cdot|$ satisfies the triangular 
inequality, that is $|x+y| \leq |x| + |y|$ for all $x,y \in \Q_p$. 
It is however much stronger: we say that the $p$-adic norm is 
\emph{ultrametric} or \emph{non Archimedean}. We will see later that
ultrametricity has strong consequences on the topology of $\Q_p$ 
(see for example Corollary~\ref{cor:convseries} below) and strongly 
influences the calculus with $p$-adic (univariate and multivariate) 
functions as well (see \S \ref{sssec:preclemma}).
This is far from being anecdotic; on the contrary, this will be the 
starting point of the theory of $p$-adic precision we will
develop in \S \ref{sec:precision}.

The $p$-adic norm defines a natural distance $d$ on $\Q_p$ as follows: 
we agree that the distance between two $p$-adic numbers $x$ and $y$ is 
$|x-y|_p$. Again this distance is ultrametric in the sense that:
$$d(x,z) \leq \max\big( d(x,y), d(y,z)\big).$$
Moreover the equality holds as soon as $d(x,y) \neq d(y,z)$: all 
triangles in $\Qp$ are isosceles!
Observe also that $d$ takes its values in a proper subset of $\R^+$
(namely $\{0\} \cup \{p^n : n \in \Z\}$) whose unique accumulation point 
is $0$. This property has surprising consequences; for example, closed
balls of positive radius are also open balls and \emph{vice et versa}. 
In particular $\Zp$ is open (in $\Qp$) and compact according to the 
topology defined by the distance. From now on, we endow $\Qp$ with this 
topology.

Clearly, a $p$-adic number lies in $\Z_p$ if and only if its $p$-adic 
valuation is nonnegative, that is if and only if its $p$-adic norm is at 
most $1$. In other words, $\Z_p$ appears as the closed unit ball in 
$\Q_p$. Viewed this way, it is remarkable that it is stable under 
addition (compare with $\R$); it is however a direct consequence of the 
ultrametricity. Similarly, by Proposition \ref{prop:invZp}, a $p$-adic 
integer is invertible in $\Zp$ if and only if it has norm $1$, meaning 
that the group of units of $\Zp$ is then the unit sphere in $\Qp$. As 
for the maximal ideal of $\Z_p$, it consists of elements of positive 
valuation and then appears as the open unit ball in $\Q_p$ (which is
also the closed ball of radius $p^{-1}$).

\subsubsection{Completeness}

The following important proposition shows that $\Qp$ is nothing but the 
completion of $\Q$ according to the $p$-adic distance. In that sense, 
$\Qp$ arises in a very natural way... just as does~$\R$.

\begin{prop}
The space $\Q_p$ equipped with its natural distance is complete
(in the sense that every Cauchy sequence converges). Moreover $\Q$
is dense in $\Q_p$.
\end{prop}

\begin{proof}
We first prove that $\Qp$ is complete. Let $(u_n)_{n \geq 0}$ be
a $\Qp$-valued Cauchy sequence. It is then bounded and rescaling the
$u_n$'s by a uniform scalar, we may assume that $|u_n| \leq 1$ 
(\emph{i.e.} $u_n \in \Zp$) for all $n$. 
For each $n$, write :
$$u_n = \sum_{i=0}^\infty a_{n,i} p^i$$
with $a_{n,i} \in \{0, 1, \ldots, p{-}1\}$. Fix an integer $i_0$ and
set $\varepsilon = p^{-i_0}$. Since $(u_n)$ is a Cauchy sequence, there
exists a rank $N$ with the property that $|u_n - u_m| \leq \varepsilon$
for all $n,m \geq N$. Coming back to the definition of the $p$-adic
norm, we find that $u_n - u_m$ is divisible by $p^{i_0}$. Writing $u_n = u_m 
+ (u_n - u_m)$ and computing the sum, we get $a_{n,i} = a_{m,i}$ for all 
$i \leq i_0$. In particular the sequence $(a_{n,i_0})_{n \geq 0}$ is 
ultimately constant. Let $a_{i_0} \in \{0, 1, \ldots, p{-}1\}$ denote its 
limit.
Now define $\ell = \sum_{i=0}^\infty a_i p^i \in \Zp$ and
consider again $\varepsilon > 0$. Let $i_0$ be an integer such that
$p^{-i_0} \leq \varepsilon$. By construction, there exists a rank $N$
for which $a_{n,i} = a_i$ whenever $n \geq N$ and $i \leq i_0$. For
$n \geq N$, the difference $u_n - \ell$ is then divisible by $p^{i_0}$
and hence has norm at most $\varepsilon$. Hence $(u_n)$ converges to 
$\ell$.

We now prove that $\Q$ is dense in $\Qp$. Since $\Qp = \Zp[\frac 1 p]$, 
it is enough to prove that $\Z$ is dense in $\Zp$. Pick $a \in \Zp$ and
write $a = \sum_{i \geq 0} a_i p^i$. For a nonnegative integer $n$, set
$b_n = \sum_{i=0}^{n-1} a_i p^i$. Clearly $b_n$ is an integer and the
sequence $(b_n)_{n \geq 0}$ converges to $a$. The density follows.
\end{proof}

\begin{cor}
\label{cor:convseries}
Let $(u_n)_{n \geq 0}$ be a sequence of $p$-adic numbers. The series 
$\sum_{n \geq 0} u_n$ converges in $\Qp$ if and only if its general
term $u_n$ converges to $0$.
\end{cor}

\begin{proof}
Set $s_n = \sum_{i=0}^{n-1} u_i$. Clearly $u_n = s_{n+1} - s_n$ for
all $n$. If $(s_n)$ converges to a limit $s \in \Qp$, then $u_n$ converges 
to $s-s = 0$.
We now assume that $(u_n)$ goes to $0$. We claim that
$(s_n)$ is a Cauchy sequence (and therefore converges). 
Indeed, let $\varepsilon > 0$ and pick an integer $N$ for which $|u_i| \leq
\varepsilon$ for all $i \geq N$. Given two integers $m$ and $n$
with $m > n \geq N$, we have:
$$|s_m - s_n| = \left|\sum_{i=n}^{m-1} u_i \right| \leq
\max \big(|u_n|, |u_{n+1}|, \ldots, |u_{m-1}|\big)$$
thanks to ultrametricity. Therefore $|s_m - s_n| \leq \varepsilon$
and we are done.
\end{proof}

\subsubsection{Tree representation}
\label{sssec:tree}

\begin{figure}
\hfill
\begin{tikzpicture}[xscale=9,yscale=-2.5]
\begin{scope}[inner sep=1mm,outer sep=0pt]
\node[scale=0.8] at (1.1, 0.2) { height };
\node[scale=0.8] at (1.1, 0.5) { $0$ };
\node[scale=0.8] at (1.1, 1) { $1$ };
\node[scale=0.8] at (1.1, 5/3) { $2$ };
\node[scale=0.8] at (1.1, 19/9) { $3$ };
\draw[thin,dotted] (1.1,2.4)--(1.1,2.8);
\node[draw,fill=digit!30,circle,inner sep=2.5mm] (void) at (1/2,0.5) { };
\node[draw,circle] (0) at (1/4,1) { $0$ };
\node[draw,fill=digit!30,circle] (1) at (3/4,1) { $1$ };
\node[scale=0.8,draw,circle] (00) at (1/8,5/3) { $00$ };
\node[scale=0.8,draw,circle] (10) at (3/8,5/3) { $10$ };
\node[scale=0.8,draw,fill=digit!30,circle] (01) at (5/8,5/3) { $01$ };
\node[scale=0.8,draw,circle] (11) at (7/8,5/3) { $11$ };
\node[scale=0.6,draw,circle] (000) at (1/16,19/9) { $000$ };
\node[scale=0.6,draw,circle] (100) at (3/16,19/9) { $100$ };
\node[scale=0.6,draw,circle] (010) at (5/16,19/9) { $010$ };
\node[scale=0.6,draw,circle] (110) at (7/16,19/9) { $110$ };
\node[scale=0.6,draw,fill=digit!30,circle] (001) at (9/16,19/9) { $001$ };
\node[scale=0.6,draw,circle] (101) at (11/16,19/9) { $101$ };
\node[scale=0.6,draw,circle] (011) at (13/16,19/9) { $011$ };
\node[scale=0.6,draw,circle] (111) at (15/16,19/9) { $111$ };
\end{scope}
\draw (void)--(0); \draw[very thick,digit] (void)--(1);
\draw (0)--(00); \draw (0)--(10);
\draw[very thick,digit] (1)--(01); \draw (1)--(11);
\draw (00)--(000); \draw (00)--(100);
\draw[very thick,digit] (01)--(001); \draw (01)--(101);
\draw (10)--(010); \draw (10)--(110);
\draw (11)--(011); \draw (11)--(111);
\draw (000)--(1/32,65/27); \draw (000)--(3/32,65/27);
\draw (100)--(5/32,65/27); \draw (100)--(7/32,65/27);
\draw (010)--(9/32,65/27); \draw (010)--(11/32,65/27);
\draw (110)--(13/32,65/27); \draw (110)--(15/32,65/27);
\draw (001)--(17/32,65/27); \draw (001)--(19/32,65/27);
\draw (101)--(21/32,65/27); \draw (101)--(23/32,65/27);
\draw (011)--(25/32,65/27); \draw (011)--(27/32,65/27);
\draw (111)--(29/32,65/27); \draw (111)--(31/32,65/27);
\draw (0,3)--(1,3);
\draw[thin,dotted] (1/32,2.6)--(1/32,2.8);
\draw[thin,dotted] (3/32,2.6)--(3/32,2.8);
\draw[thin,dotted] (5/32,2.6)--(5/32,2.8);
\draw[thin,dotted] (7/32,2.6)--(7/32,2.8);
\draw[thin,dotted] (9/32,2.6)--(9/32,2.8);
\draw[thin,dotted] (11/32,2.6)--(11/32,2.8);
\draw[thin,dotted] (13/32,2.6)--(13/32,2.8);
\draw[thin,dotted] (15/32,2.6)--(15/32,2.8);
\draw[thin,dotted] (19/32,2.6)--(19/32,2.8);
\draw[thin,dotted] (21/32,2.6)--(21/32,2.8);
\draw[thin,dotted] (23/32,2.6)--(23/32,2.8);
\draw[thin,dotted] (25/32,2.6)--(25/32,2.8);
\draw[thin,dotted] (27/32,2.6)--(27/32,2.8);
\draw[thin,dotted] (29/32,2.6)--(29/32,2.8);
\draw[thin,dotted] (31/32,2.6)--(31/32,2.8);
\draw[very thick,digit] (001)--(17/32,65/27)--(35/64,211/81)
                 --(69/128,665/243)--(137/256,2059/729)
                 --(275/512,6305/2187)--(549/1024,2.9219);
\draw[digit] (549/1024,2.9219)--(0.536,3);
\node[digit,scale=0.8,below] at (0.536,3) { $...010010001$ };
\node[below] at (1,3) { $\Z_2$ };
\end{tikzpicture}
\hfill\null

\caption{Tree representation of $\Z_2$}
\label{fig:treeZ2}
\end{figure}

Geometrically, it is often convenient and meaningful to represent $\Zp$ 
as the infinite full $p$-ary tree. In order to explain this 
representation, we need a definition.

\begin{deftn}
\label{def:intervalZp}
For $h \in \N$ and $a \in\Zp$, we set:
$$I_{h,a} = \big\{ \, x \in \Zp 
\quad \text{s.t.} \quad 
x \equiv a \pmod{p^h} \, \big\}.$$
An \emph{interval} of $\Zp$ is a subset of $\Zp$ of the form $I_{h,a}$
for some $h$ and $a$.
\end{deftn}

If $a$ decomposes in base $p$ as
$a = a_0 + a_1 p + a_2 p^2 + \cdots + a_{h-1} p^{h-1} + \cdots$,
the interval $I_{h,a}$ consists exactly of the $p$-adic integers whose
last digits are $a_{h-1} \ldots a_1 a_0$ in this order.
On the other hand, from the analytic point of view, the condition $x 
\equiv a \pmod {p^h}$ is equivalent to $|x-a| \leq p^{-h}$. Thus the
interval $I_{h,a}$ is nothing but the closed ball of centre $a$ and 
radius $p^{-h}$. Even better, the intervals of $\Zp$ are exactly the 
closed balls of $\Zp$.

Clearly $I_{h,a} = I_{h,a'}$ if and only if $a \equiv a' \pmod{p^h}$. 
In particular, given an interval $I$ of $\Zp$, there is exactly one
integer $h$ such that $I = I_{h,a}$. We will denote it by $h(I)$ and
call it the \emph{height} of $I$. We note that there exist exactly
$p^h$ intervals of $\Zp$ of height $h$ since these intervals are
indexed by the classes modulo $p^h$ (or equivalently by the sequences
of $h$ digits between $0$ and $p{-}1$).

From the topological point of view, intervals behave like $\Zp$:
they are at the same time open and compact.

\medskip

We now define the \emph{tree} of $\Zp$, denoted by $\calT(\Zp)$, as follows:
its vertices are the intervals of $\Zp$ and we put an edge $I \to J$ 
whenever $h(J) = h(I) + 1$ and $J \subset I$. 
A picture of $\calT(\Z_2)$ is represented on Figure~\ref{fig:treeZ2}.
The labels indicated on the vertices are the last $h$ digits of $a$.
Coming back to a general $p$, 
we observe that the height of an interval $I$ corresponds to the 
usual height function in the tree $\calT(\Zp)$. Moreover, given two 
intervals $I$ and $J$, the inclusion $J \subset I$ holds if and only 
if there exists a path from $I$ to $J$.

Elements of $\Zp$ bijectively correspond to infinite paths of 
$\calT(\Zp)$ starting from the root through the following 
correspondence: an element $x \in \Zp$ is encoded by the path
$$I_{0,x} \to I_{1,x} \to I_{2,x} \to \cdots \to I_{h,x} \to \cdots.$$
Under this encoding, an infinite path of $\calT(\Zp)$ starting from the 
root 
$$I_0 \to I_1 \to I_2 \to \cdots \to I_h \to \cdots$$ 
corresponds to a uniquely determined $p$-adic integer, which is the 
unique element lying in the decreasing intersection $\bigcap_{h \in \N} 
I_h$. Concretely each new $I_h$ determines a new digit of $x$; the whole 
collection of the $I_h$'s then defines $x$ entirely.
The distance on $\Zp$ can be visualized on $\calT(\Zp)$ as well:
given $x, y \in \Zp$, we have $|x-y| = p^{-h}$ where $h$ is the height
where the paths attached to $x$ and $y$ separate.

\medskip

The above construction easily extends to $\Qp$.

\begin{deftn}
\label{def:intervalQp}
For $h \in \Z$ and $a \in\Qp$, we set:
$$I_{h,a} = \big\{ \, x \in \Qp 
\quad \text{s.t.} \quad 
|x - a| \leq p^{-h} \, \big\}.$$
A \emph{bounded interval} of $\Qp$ is a subset of $\Qp$ of the form
$I_{h,a}$ for some $h$ and $a$.
\end{deftn}

Similarly to the case of $\Zp$, a bounded interval of $\Qp$ of height 
$h$ is a subset of $\Qp$ consisting of $p$-adic numbers whose digits at 
the positions $< h$ are fixed (they have to agree with the digits
of $a$ at the same positions).

The graph $\calT(\Qp)$ is defined as follows: its vertices are the 
intervals of $\Qp$ while there is an edge $I \to J$ if $h(J) = h(I) 
+ 1$ and $J \subset I$. We draw the attention of the reader to the
fact that $\calT(\Qp)$ is a tree but it is not rooted: there does not
exist a largest bounded interval in $\Qp$.
To understand better the structure of $\calT(\Qp)$, let us define, 
for any integer $v$, the subgraph $\calT(p^{-v} \Zp)$ of $\calT(\Qp)$ 
consisting of intervals which are contained in $p^{-v}\Zp$. From the
fact that $\Qp$ is the union of all $p^{-v}\Zp$, we derive that
$\calT(\Qp) = \bigcup_{v \geq 0} \calT(p^{-v}\Zp)$. Moreover, for all
$v$, $\calT(p^{-v} \Zp)$ is a rooted tree (with root $p^{-v}\Zp$)
which is isomorphic to $\calT(\Zp)$ except that the height function is 
shifted by $-v$. The tree $\calT(p^{-v-1}\Zp)$ is thus obtained by 
juxtaposing $p$ copies of $\calT(p^{-v}\Zp)$ and linking the roots of 
them to a common parent $p^{-v}\Zp$ (which then becomes the new root).

\subsection{Newton iteration over the $p$-adic numbers}

Newton iteration is a well-known tool in Numerical Analysis for 
approximating a zero of a ``nice'' function defined on a real interval.
More precisely, given a differentiable function $f : [a,b] \to \R$, we
define a recursive sequence $(x_i)_{i \geq 0}$ by:
\begin{equation}
\label{eq:Newtoniter}
x_0 \in [a,b] \quad ; \quad 
x_{i+1} = x_i - \frac{f(x_i)}{f'(x_i)}, \, i = 0, 1, 2, \ldots
\end{equation}
Under some assumptions, one can prove that the sequence $(x_i)$ 
converges to a zero of $f$, namely $x_\infty$. Moreover the convergence is 
very rapid since, assuming that $f$ is twice differentiable, we usually 
have an inequality of the shape $|x_\infty - x_i| \leq \rho^{2^i}$ for 
some $\rho \in (0,1)$. In other words, the number of correct digits 
roughly doubles at each iteration. The Newton recurrence 
\eqref{eq:Newtoniter} has a nice geometrical interpretation as well: the 
value $x_{i+1}$ is the $x$-coordinate of the intersection point of the 
$x$-axis with the tangent to the curve $y = f(x)$ at the point $x_i$ 
(see Figure~\ref{fig:Newtoniter}).

\begin{figure}
\hfill
\begin{tikzpicture}[xscale=2,yscale=0.2]
\draw[thick,red] plot[smooth] coordinates { 
  (0, -1)
  (0.20, -0.96)
  (0.40, -0.84)
  (0.60, -0.64)
  (0.80, -0.36)
  (1.0, 0)
  (1.2, 0.44)
  (1.4, 0.96)
  (1.6, 1.56)
  (1.8, 2.24)
  (2.0, 3)
  (2.2, 3.84)
  (2.4, 4.76)
  (2.6, 5.76)
  (2.8, 6.84)
  (3.0, 8.0)
  (3.2, 9.24)
  (3.4, 10.56)
  (3.6, 11.96)
  (3.8, 13.44)
  (4.0, 15)
  (4.2, 16.64)
  (4.4, 18.36)
  (4.6, 20.16)
  (4.8, 22.04)
  (5, 24)
  (5.2, 26.04)
};
\draw[->] (0,0)--(5.2,0);
\draw[dotted] (5,0)--(5,24);
\draw (5,24)--(2.6,0);
\draw[dotted] (2.6,0)--(2.6,5.76);
\draw (2.6,5.76)--(1.493,0);
\draw[dotted] (1.493,0)--(1.493,1.229);
\draw (1.493,1.229)--(1.081,0);
\node[below, scale=0.8] at (5,0) { $x_0$ };
\node[below, scale=0.8] at (2.6,0) { $x_1$ };
\node[below, scale=0.8] at (1.493,0) { $x_2$ };
\node[below, scale=0.8] at (1.1,0) { $x_3$ };
\node[red, scale=0.6] at (1,0) { $\bullet$ };
\node[red,above, scale=0.8] at (0.9,0) { $x_\infty$ };
\end{tikzpicture}
\hfill\null

\caption{Newton iteration over the reals}
\label{fig:Newtoniter}
\end{figure}

\subsubsection{Hensel's Lemma}
\label{sssec:hensel}

It is quite remarkable that the above discussion extends almost 
\emph{verbatim} when $\R$ is replaced by $\Qp$. Actually, extending the 
notion of differentiability to $p$-adic functions is quite subtle and 
probably the most difficult part. This will be achieved in \S 
\ref{sssec:classC1} (for functions of class $C^1$) and \S 
\ref{sssec:NewtonC2} (for functions of class $C^2$). For now, we prefer 
avoiding these technicalities and restricting ourselves to the simpler 
(but still interesting) case of polynomials. For this particular case, 
the Newton iteration is known as \emph{Hensel's Lemma} and already 
appears in Hensel's seminal paper~\cite{He97} in which $p$-adic 
numbers are introduced.

Let $f(X) = a_0 + a_1 X + \cdots + a_n X^n$ be a polynomial in the 
variable $X$ with coefficients in $\Qp$. Recall that the derivative 
of $f$ can be defined in a purely algebraic way as $f'(X) = a_1 + 
2 a_2 X + \cdots + n a_n X^{n-1}$.

\begin{theo}[Hensel's Lemma]
\label{th:Hensel}
Let $f \in \Zp[X]$ be a polynomial with coefficients in $\Zp$.
We suppose that we are given some $a \in \Zp$ with the property that
$|f(a)| < |f'(a)|^2$. 
Then the sequence $(x_i)_{i \geq 0}$
defined by the recurrence:
$$x_0 = a \quad ; \quad x_{i+1} = x_i - \frac{f(x_i)}{f'(x_i)}$$
is well defined and converges to $x_\infty \in \Zp$ with $f(x_\infty) = 
0$.
The rate of convergence is given by:
$$|x_\infty - x_i| \leq |f'(a)| \cdot 
\left(\frac{|f(a)|}{|f'(a)|^2}\right)^{2^i}.$$
Moreover $x_\infty$ is the unique root of $f$ in the open ball of
centre $a$ and radius $|f'(a)|$.
\end{theo}

The proof of the above theorem is based on the next lemma:

\begin{lem}
\label{lem:polyC2}
Given $f \in \Zp[X]$ and $x, h \in \Zp$, we have:
\begin{enumerate}[(i)]
\item $|f(x+h) - f(x)| \leq |h|$.
\item $|f(x+h) - f(x) - h f'(x)| \leq |h|^2$.
\end{enumerate}
\end{lem}

\begin{proof}
For any nonnegative integer $i$, define $f^{[i]} = \frac 1{i!} \: 
f^{(i)}$ where $f^{(i)}$ stands for the $i$-th derivative of $f$.
Taylor's formula then reads:
\begin{equation}
\label{eq:taylor2}
f(x+h) - f(x) = h f'(x) + h^2 f^{[2]}(x) + \cdots + h^n f^{[n]}(x).
\end{equation}
Moreover, a direct computation shows that the coefficients of $f^{[i]}$ 
are obtained from that of $f$ by multiplying by binomial coefficients. 
Therefore $f^{[i]}$ has coefficients in $\Zp$. Hence $f^{[i]}(x) \in 
\Zp$, \emph{i.e.} $|f^{[i]}(x)| \leq 1$, for all $i$.
We deduce that each summand of the right hand
side of \eqref{eq:taylor2} has norm at most $|h|$. The first assertion
follows while the second is proved similarly.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{th:Hensel}]
Define $\rho = \frac{|f(a)|}{|f'(a)|^2}$. 
We first prove by induction on $i$ the following conjunction:
$$(H_i) : \quad |f'(x_i)| = |f'(a)|
\quad \text{and} \quad |f(x_i)| \leq |f'(a)|^2 \cdot \rho^{2^i}.$$
Clearly $(H_0)$ holds. We assume now that $(H_i)$ holds for some $i 
\geq 0$. We put $h
= -\frac{f(x_i)}{f'(x_i)}$ so that $x_{i+1} = x_i + h$. We write
$f'(x_{i+1}) = \big(f'(x_{i+1}) - f'(x_i)\big) + f'(x_i)$.
Observe that the first summand $f'(x_{i+1}) - f'(x_i)$ has norm at 
most $|h|$ by the first assertion of Lemma~\ref{lem:polyC2}, while 
$|f'(x_i)| \geq |h| \cdot \rho^{-2^i} > |h|$ by the induction 
hypothesis.
The norm of $f'(x_{i+1})$ is then the maximum of the norms of the two
summands, which is $|f'(x_i)| = |f'(a)|$.
Now, applying again Lemma~\ref{lem:polyC2}, we get $|f(x_{i+1})| \leq 
|h|^2 \leq |f'(a)|^2 \cdot \rho^{2^i}$ and the induction goes.

Coming back to the recurrence defining the $x_i$'s, we get:
\begin{equation}
\label{eq:rateNewton}
|x_{i+1} - x_i| = \frac{|f(x_i)|}{|f'(x_i)|} \leq |f'(a)| \cdot 
\rho^{2^i}.
\end{equation}
By Corollary~\ref{cor:convseries}, this implies the convergence
of the sequence $(x_i)_{i \geq 0}$. Its limit $x_\infty$ is a 
solution to the equation $x_\infty = x_\infty + 
\frac{f(x_\infty)}{f'(x_\infty)}$. Thus $f(x_\infty)$ has to vanish. 
The announced rate of convergence follows from Eq.~\eqref{eq:rateNewton} 
thanks to ultrametricity.

It remains to prove uniqueness. For this, consider $y \in \Zp$
with $f(y) = 0$ and $|y - x_0| < |f'(a)|$. Since $|x_\infty - x_0|
\leq |f'(a)| \cdot \rho < |f'(a)|$, we deduce $|x_\infty - y| <
|f'(a)|$ as well. Applying Lemma~\ref{lem:polyC2} with $x = x_\infty$
and $h = y - x_\infty$, we find $|h f'(x_\infty)| \leq |h|^2$. Since 
$|h| < |f'(a)| = |f'(x_\infty)|$, this implies $|h| = 0$, 
\emph{i.e.} $x = x_\infty$. Uniqueness is proved.
\end{proof}

\begin{rem}
\label{rem:Hensel}
All conclusions of Theorem~\ref{th:Hensel} are still valid for any
sequence $(x_i)$ satisfying the weaker assumption:
$$\left|x_{i+1} - x_i + \frac{f(x_i)}{f'(x_i)}\right| 
\leq |f'(a)| \cdot \rho^{2^{i+1}}$$
(the proof is entirely similar). Roughly speaking, this stronger version 
allows us to work with approximations at \emph{each} iteration. It will 
play a quite important role for algorithmic purpose (notably in \S 
\ref{sssec:intervalNewton}) since computers cannot handle exact $p$-adic 
numbers but always need to work with truncations.
\end{rem}

\subsubsection{Computation of the inverse}
\label{sssec:inverse}

A classical application of Newton iteration is the computation of the 
inverse: for computing the inverse of a real number $c$, we introduce 
the function $x \mapsto \frac 1 x - c$ and the associated Newton scheme. 
This leads to the recurrence $x_{i+1} = 2 x_i - c x_i^2$ with initial 
value $x_0 = a$ where $a$ is sufficiently close to $\frac 1 c$.

In the $p$-adic setting, the same strategy applies (although it does not 
appear as a direct consequence of Hensel's Lemma since the mapping $x 
\mapsto \frac 1 x - c$ is \emph{not} polynomial). Anyway, let us pick 
an invertible element $c \in \Zp$, \emph{i.e.} $|c| = 1$, and define the
sequence $(x_i)_{i \geq 0}$ by:
$$x_0 = a \quad ; \quad x_{i+1} = 2 x_i - c x_i^2, \, i = 0, 1, 2, \ldots$$
where $a$ is any $p$-adic number whose last digit is the inverse modulo
$p$ of the last digit of $c$. Computing such an element $a$ reduces to
computing a modular inverse and thus is efficiently feasible.

\begin{prop}
\label{prop:Newtoninverse}
The sequence $(x_i)_{i \geq 0}$ defined above converges to $\frac 1 c 
\in \Zp$. More precisely, we have $|c x_i - 1| \leq p^{-2^i}$ for all 
$i$.
\end{prop}

\begin{proof}
We prove the last statement of the proposition by induction on $i$. By
construction of $a$, it holds for $i = 0$. Now observe that
$c x_{i+1} - 1 = 2 c x_i - c^2 x_i^2 - 1 = - (c x_i - 1)^2$.
Taking norms on both sides, we get $|c x_{i+1} - 1| = |c x_i - 1|^2$
and the induction goes this way.
\end{proof}

Quite interestingly, this method extends readily to 
the non-commutative case. Let us illustrate this by showing how it can be 
used to compute inverses of matrices with coefficients in $\Zp$. Assume
then that we are starting with a matrix $C \in M_n(\Zp)$ whose reduction
modulo $p$ is invertible, \emph{i.e.} there exists a matrix $A$ such
that $AC \equiv I_n \pmod p$ where $I_n$ is the identity matrix of 
size $n$. Note that the computation of $A$ reduces to the inversion 
of an $n \times n$ matrix over the finite field $\Fp$ and so can be 
done efficiently.
We now define the sequence $(X_i)_{i \geq 0}$ by the recurrence:
$$X_0 = A \quad ; \quad X_{i+1} = 2 X_i - X_i C X_i$$
(be careful with the order in the last term). Mimicking the proof of
Proposition~\ref{prop:Newtoninverse}, we write:
$$C X_{i+1} - I_n = 2 C X_i - C X_i C X_i - I_n = - (C X_i - I_n)^2$$
and obtain this way that each entry of $C X_i - 1$ has norm at most
$p^{2^i}$. Therefore $X_i$ converges to a matrix $X_\infty$ satisfying
$C X_\infty = I_n$, \emph{i.e.} $X_\infty = C^{-1}$ (which in particular
implies that $C$ is invertible). A similar argument works for $p$-adic 
skew polynomials and $p$-adic differential operators as well.

\subsubsection{Square roots in $\Qp$}
\label{sssec:sqrt}

Another important application of Newton iteration is the computation
of square roots. Again, the classical scheme over $\R$ applies without
(substantial) modification over $\Qp$.

Let $c$ be a $p$-adic number. If the $p$-adic valuation of $c$ is odd, 
then $c$ is clearly not a square in $\Qp$ and it does not make sense to 
compute its square root. On the contrary, if the $p$-adic valuation of 
$c$ is even, then we can write $c = p^{2v} c'$ where $v$ is an integer and 
$c'$ is a unit in $\Zp$. Computing a square root then reduces to 
computing a square root of $c'$; in other words, we may assume that $c$ 
is invertible in $\Zp$, \emph{i.e.} $|c| = 1$.

We now introduce the polynomial function $f(x) = x^2 - c$. In order to 
apply Hensel's Lemma (Theorem~\ref{th:Hensel}), we need a first rough
approximation $a$ of $\sqrt c$. Precisely, we need to find some $a \in 
\Zp$ with the property that $|f(a)| < |f'(a)|^2 = |2a|^2$. From $|c| = 
1$, the above inequality implies $|a| = 1$ as well and therefore can be
rewritten as 
$a^2 \equiv c \pmod q$ where $q = 8$ if $p = 2$ and $q = p$ otherwise.
We then first need to compute a square root of $c$ modulo $q$. If $p=2$,
this can be achieved simply by looking at the table of squares modulo $8$:
$$\begin{array}{|c||c|c|c|c|c|c|c|c|}
\hline
a & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 \\
\hline
a^2 & 0 & 1 & 4 & 1 & 0 & 1 & 4 & 1 \\
\hline
\end{array}$$

Observe moreover that $c$ is necessarily odd (since it is assumed to be
invertible in $\Z_2$). If it is congruent to $1$
modulo $8$, $a = 1$; otherwise, there is no solution
and $c$ has no square root in $\Q_2$.
When $p > 2$, we have to compute a square root in the finite field $\Fp$
for which efficient algorithms are known~\cite[\S 1.5]{Co93}. If 
$c \text{ mod }p$ is not a square in $\Fp$, then $c$ does not admit a square root 
in $\Qp$ either.

Once we have computed the initial value $a$, we consider the recursive
sequence $(x_i)_{i \geq 0}$ defined by $x_0 = a$ and
$$x_{i+1} = x_i - \frac{f(x_i)}{f'(x_i)} = \frac 1 2  \left( x_i
+ \frac c {x_i} \right), \, i = 0, 1, 2, \ldots$$
By Hensel's Lemma, it converges to a limit $x_\infty$ which is a square
root of $c$. Moreover the rate of convergence is given by:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
|x_\infty - x_i| & \leq p^{-2^i} & \text{if } p > 2 \smallskip \\
|x_\infty - x_i| & \leq 2^{-(2^i+1)} & \text{if } p = 2
\end{array}$$
meaning that the number of correct digits of $x_i$ is at least $2^i$ 
(resp. $2^i + 1$) when $p > 2$ (resp. $p = 2$).

\subsection{Similarities with formal and Laurent series}

According to the very first definition of $\Zp$ we have given (see \S 
\ref{sssec:downtoearth}), $p$-adic integers have formal 
similitudes with formal series over $\Fp$: they are both described as
infinite series with coefficients between $0$ and $p{-}1$, the main
difference being that additions and multiplications involve carries
in the $p$-adic case while they do not for formal series. 
From a more abstract point of view, the parallel between $\Zp$ and 
$\Fp[\![t]\!]$ is also apparent. For example $\Fp [\![t]\!]$ is also 
endowed with a valuation; this valuation defines a distance on 
$\Fp[\![t]\!]$ for which $\Fp[\![t]\!]$ is complete and Hensel's Lemma 
(together with Newton iteration) extends \emph{verbatim} to formal 
series. In the same fashion, the analogue of $\Qp$ is the field of 
Laurent series $\Fp(\!(t)\!)$. Noticing further that $\Fp[t]$ is dense 
in $\Fp[\![t]\!]$ and similarly that $\Fp(t)$ is dense in $\Fp(\!(t)\!)$ 
\footnote{The natural embedding $\Fp(t) \to \Fp(\!(t)\!)$ takes a 
rational function to its Taylor expansion at $0$.}, we can even 
supplement the correspondence between the characteristic $0$ and the 
characteristic $p$, ending up with the ``dictionary'' of 
Figure~\ref{fig:char0p}.

\begin{figure}
\hfill
\begin{tabular}{ccc}
char. $0$ & & char. $p$ \smallskip \\
$\Z$ & $\longleftrightarrow$ & $\Fp[t]$ \\
$\Q$ & $\longleftrightarrow$ & $\Fp(t)$ \\
$\Z_p$ & $\longleftrightarrow$ & $\Fp[\![t]\!]$ \\
$\Q_p$ & $\longleftrightarrow$ & $\Fp(\!(t)\!)$ 
\end{tabular}
\hfill\null

\caption{Similarities between characteristic $0$ and characteristic $p$}
\label{fig:char0p}
\end{figure}

Algebraists have actually managed to capture the essence of there 
resemblances in the notion of \emph{discrete valuation rings/fields} and 
\emph{completion} of those~\cite{Se62}. Recently, Scholze defined the 
notion of \emph{perfectoid} which allows us (under several additional 
assumptions) to build an actual bridge between the characteristic $0$ 
and the characteristic $p$. It is not the purpose of this lecture to go 
further in this direction; we nevertheless refer interested people to 
\cite{Sc12,Sc14,Bh14} for the exposition of the theory.

\subsubsection{The point of view of algebraic geometry}

Algebraic Geometry has an original and interesting point of view 
on the various rings and fields introduced before, showing in 
particular how $p$-adic numbers can arise in many natural problems in 
arithmetics. The underlying ubiquitous idea in algebraic geometry is
to associate to any ring $A$ a geometrical space $\spec A$ --- its 
so-called \emph{spectrum} --- on which the functions are the elements
of $A$. We will not detail here the general construction of $\spec A$
but just try to explain informally what it looks like when $A$ is 
one of the rings appearing in Figure~\ref{fig:char0p}.

Let us start with $k[t]$ where $k$ is a field (of any characteristic). 
For the simplicity of the exposition, let us assume further that $k$ is 
algebraically closed.
One thinks of elements of $k[t]$ as (polynomial) functions over $k$, 
meaning that this spectrum should be thought of as $k$ itself. $\spec 
k[t]$ is called the affine line over $k$ and is usually drawn as a 
straight line. The spectrum of $k(t)$ can be understood in similar 
terms: a rational fraction $f \in k(t)$ defines a function on $k$ as 
well, except that it can be undefined at some points. Therefore $\spec 
k(t)$ might be thought as the affine line over $k$ with a ``moving'' 
finite set of points removed. It is called the \emph{generic point} of 
the affine line.

What about $k[\![t]\!]$? If $k$ has characteristic $0$ (which we assume 
for the simplicity of the discussion), the datum of $f \in k[\![t]\!]$ 
is equivalent to the datum of the values of the $f^{(n)}(0)$'s (for $n$ 
varying in $\N$); we sometimes say that $f$ defines a function on a 
\emph{formal neighborhood} of $0$. This formal neighborhood is the 
spectrum of $k[\![t]\!]$; it should then be thought as a kind of 
thickening of the point $0 \in \spec k[t]$ which does not include any 
other point (since a formal series $f \in k[\![t]\!]$ cannot be 
evaluated in general at any other point than $0$). 
Finally $\spec k(\!(t)\!)$ is the \emph{punctured formal neighborhood} 
of $0$; it is obtained from $\spec k[\![t]\!]$ by removing the point
$0$ but not its neighborhood!

The embedding $k[t] \to k[\![t]\!]$ (resp. $k(t) \to k(\!(t)\!)$) given 
by Taylor expansion at $0$ corresponds to the restriction of a function 
$f$ defined on the affine line (resp. the generic point of the affine 
line) to the formal neighborhood (resp. the punctured formal 
neighborhood) of $0$.

Of course, one can define similarly formal neighborhoods and punctured 
formal neighborhoods around other points: for $a\in k$, the 
corresponding rings are respectively $k[\![h_a]\!]$ and $k(\!(h_a)\!)$ 
where $h_a$ is a formal variable which plays the role $t{-}a$. The 
algebraic incarnation of the restrictions to $\spec k[\![h_a]\!]$ and 
$k(\!(h_a)\!)$ are the Taylor expansions at $a$.

The above discussion is illustrated by the drawing at the top of 
Figure~\ref{fig:alggeom}; it represents more precisely the rational 
fraction $f(t) = \frac{t^2}{1-t}$ viewed as a function defined on $\spec 
k(t)$ (with $1$ as point of indeterminacy) together its Taylor expansion 
around $0$ (in blue) and $2$ (in purple) which are defined on formal 
neighborhoods.

\begin{figure}
\hfill
\begin{tikzpicture}[xscale=1.5,yscale=0.3]
% f(t) = t^2 / (t-1)
\draw plot[smooth] coordinates { 
  (-1.40000000000000,0.483333333333333)
  (-1.30000000000000,0.473913043478261)
  (-1.20000000000000,0.472727272727273)
  (-1.10000000000000,0.480952380952381)
  (-1.00000000000000,0.500000000000000)
  (-0.900000000000000,0.531578947368421)
  (-0.800000000000000,0.577777777777778)
  (-0.700000000000000,0.641176470588235)
  (-0.600000000000000,0.725000000000000)
  (-0.500000000000000,0.833333333333333)
  (-0.400000000000000,0.971428571428571)
  (-0.300000000000000,1.14615384615385)
  (-0.200000000000000,1.36666666666667)
  (-0.100000000000000,1.64545454545455)
  (0,2.00000000000000)
  (0.0999999999999999,2.45555555555555)
  (0.200000000000000,3.05000000000000)
  (0.300000000000000,3.84285714285714)
  (0.400000000000000,4.93333333333333)
  (0.500000000000000,6.50000000000000)
};
\draw plot[smooth] coordinates { 
  (1.50000000000000,-7.50000000000000)
  (1.60000000000000,-5.93333333333333)
  (1.70000000000000,-4.84285714285714)
  (1.80000000000000,-4.05000000000000)
  (1.90000000000000,-3.45555555555555)
  (2.00000000000000,-3.00000000000000)
  (2.10000000000000,-2.64545454545454)
  (2.20000000000000,-2.36666666666667)
  (2.30000000000000,-2.14615384615384)
  (2.40000000000000,-1.97142857142857)
  (2.50000000000000,-1.83333333333333)
  (2.60000000000000,-1.72500000000000)
  (2.70000000000000,-1.64117647058823)
  (2.80000000000000,-1.57777777777778)
  (2.90000000000000,-1.53157894736842)
  (3.00000000000000,-1.50000000000000)
  (3.10000000000000,-1.48095238095238)
  (3.20000000000000,-1.47272727272727)
  (3.30000000000000,-1.47391304347826)
  (3.40000000000000,-1.48333333333333)
};
\node[above] at (3.8,-1.483) { $\frac{t^2}{1-t} \in k(t)$ };

\draw[dotted] (0,-9)--(0,2);
\draw[ultra thick, blue] plot[smooth] coordinates { 
  (-0.200000000000000,1.36666666666667)
  (-0.100000000000000,1.64545454545455)
  (0,2.00000000000000)
  (0.0999999999999999,2.45555555555555)
  (0.200000000000000,3.05000000000000)
};
\node[blue,scale=0.9,right] at (0.1,2) { $t^2 + t^3 + t^4 + \cdots \in k[\![t]\!]$ };

\draw[dotted] (2,-9)--(2,-3);
\draw[ultra thick, purple] plot[smooth] coordinates { 
  (1.80000000000000,-4.05000000000000)
  (1.90000000000000,-3.45555555555555)
  (2.00000000000000,-3.00000000000000)
  (2.10000000000000,-2.64545454545454)
  (2.20000000000000,-2.36666666666667)
};
\node[purple,scale=0.9,below right] at (2,-3) { $-4 - (t{-}2)^2 + (t{-}2)^3 - (t{-}2)^4 + \cdots \in k[\![t{-}2]\!]$ };

\begin{scope}[yshift=-9cm]
\draw (-1.4,0)--(3.4,0);
\node[below right,scale=0.8] at (3.4,0) { $\spec k[t]$ };
\draw (-1,0.3)--(-1,-0.3);
\node[scale=0.7] at (-1,-1) { ${-}1$ };
\draw (0,0.3)--(0,-0.3);
\node[scale=0.7] at (0,-1) { $0$ };
\draw (1,0.3)--(1,-0.3);
\node[scale=0.7] at (1,-1) { $1$ };
\draw (2,0.3)--(2,-0.3);
\node[scale=0.7] at (2,-1) { $2$ };
\draw (3,0.3)--(3,-0.3);
\node[scale=0.7] at (3,-1) { $3$ };
\draw[ultra thick, blue] (-0.2,0)--(0.2,0);
\node[above right,blue,scale=0.7] at (0,0) { $\spec k[\![t]\!]$ };
\draw[ultra thick, purple] (1.8,0)--(2.2,0);
\node[above right,purple,scale=0.7] at (2,0) { $\spec k[\![t{-}2]\!]$ };
\end{scope}

\begin{scope}[yshift=-20cm]
% f(t) = t^2 / (t-1)
\draw plot[smooth] coordinates { 
  (-1.40000000000000,0.483333333333333)
  (-1.30000000000000,0.473913043478261)
  (-1.20000000000000,0.472727272727273)
  (-1.10000000000000,0.480952380952381)
  (-1.00000000000000,0.500000000000000)
  (-0.900000000000000,0.531578947368421)
  (-0.800000000000000,0.577777777777778)
  (-0.700000000000000,0.641176470588235)
  (-0.600000000000000,0.725000000000000)
  (-0.500000000000000,0.833333333333333)
  (-0.400000000000000,0.971428571428571)
  (-0.300000000000000,1.14615384615385)
  (-0.200000000000000,1.36666666666667)
  (-0.100000000000000,1.64545454545455)
  (0,2.00000000000000)
  (0.0999999999999999,2.45555555555555)
  (0.200000000000000,3.05000000000000)
  (0.300000000000000,3.84285714285714)
  (0.400000000000000,4.93333333333333)
  (0.500000000000000,6.50000000000000)
};
\draw plot[smooth] coordinates { 
  (1.50000000000000,-7.50000000000000)
  (1.60000000000000,-5.93333333333333)
  (1.70000000000000,-4.84285714285714)
  (1.80000000000000,-4.05000000000000)
  (1.90000000000000,-3.45555555555555)
  (2.00000000000000,-3.00000000000000)
  (2.10000000000000,-2.64545454545454)
  (2.20000000000000,-2.36666666666667)
  (2.30000000000000,-2.14615384615384)
  (2.40000000000000,-1.97142857142857)
  (2.50000000000000,-1.83333333333333)
  (2.60000000000000,-1.72500000000000)
  (2.70000000000000,-1.64117647058823)
  (2.80000000000000,-1.57777777777778)
  (2.90000000000000,-1.53157894736842)
  (3.00000000000000,-1.50000000000000)
  (3.10000000000000,-1.48095238095238)
  (3.20000000000000,-1.47272727272727)
  (3.30000000000000,-1.47391304347826)
  (3.40000000000000,-1.48333333333333)
};
\node[above] at (3.7,-1.483) { $\frac{909}{5} \in \Q$ };

\draw[dotted] (0,-9)--(0,2);
\draw[ultra thick, blue] plot[smooth] coordinates { 
  (-0.200000000000000,1.36666666666667)
  (-0.100000000000000,1.64545454545455)
  (0,2.00000000000000)
  (0.0999999999999999,2.45555555555555)
  (0.200000000000000,3.05000000000000)
};
\node[blue,scale=0.9,right] at (0.1,2) { $\ldots102001100 \in \Z_3$ };

\draw[dotted] (2,-9)--(2,-3);
\draw[ultra thick, purple] plot[smooth] coordinates { 
  (1.80000000000000,-4.05000000000000)
  (1.90000000000000,-3.45555555555555)
  (2.00000000000000,-3.00000000000000)
  (2.10000000000000,-2.64545454545454)
  (2.20000000000000,-2.36666666666667)
};
\node[purple,scale=0.9,below right] at (2,-3) { $\ldots412541634 \in \Z_7$ };

\begin{scope}[yshift=-9cm]
\draw (-1.4,0)--(3.4,0);
\node[below right,scale=0.8] at (3.4,0) { $\spec \Z$ };
\draw (-1,0.3)--(-1,-0.3);
\node[scale=0.7] at (-1,-1) { $2$ };
\draw (0,0.3)--(0,-0.3);
\node[scale=0.7] at (0,-1) { $3$ };
\draw (1,0.3)--(1,-0.3);
\node[scale=0.7] at (1,-1) { $5$ };
\draw (2,0.3)--(2,-0.3);
\node[scale=0.7] at (2,-1) { $7$ };
\draw (3,0.3)--(3,-0.3);
\node[scale=0.7] at (3,-1) { $11$ };
\draw[ultra thick, blue] (-0.2,0)--(0.2,0);
\node[above right,blue,scale=0.7] at (0,0) { $\spec \Z_3$ };
\draw[ultra thick, purple] (1.8,0)--(2.2,0);
\node[above right,purple,scale=0.7] at (2,0) { $\spec \Z_7$ };
\end{scope}
\end{scope}
\end{tikzpicture}
\hfill \null

\caption{The point of view of algebraic geometry}
\label{fig:alggeom}
\end{figure}

\medskip

We now move to the arithmetical rings ($\Z$, $\Q$, $\Zp$ and $\Qp$) for 
which the picture is surprisingly very similar. In order to explain what 
is the spectrum of $\Z$, we first need to reinterpret $\spec k[t]$ in a 
more algebraic way. Given $f \in k[t]$, its evaluation at $a$, namely 
$f(a)$, can alternatively be defined as the remainder in the Euclidean 
division of $a$ by $t{-}a$. The affine line over $k$ then appears more 
intrinsically as the set of monic polynomials of degree $1$ over $k$ or, 
even better, as the set of irreducible monic polynomials (recall that we 
had assumed that $k$ is algebraically closed).

Translating this to the arithmetical world, we are inclined to think at 
the spectrum of $\Z$ as the set of prime numbers. Moreover, the 
evaluation of an integer $a \in \Z$ at the prime $p$ should be nothing 
but the remainder in the Euclidean division of $a$ by $p$. The integer 
$a$ defines this way a function over the spectrum of $\Z$.

A rational number $\frac a b$ can be reduced modulo $p$ for all prime 
$p$ not dividing $b$. It then defines a function on the spectrum of $\Z$ 
except that there are points of indeterminacy. $\spec \Q$ then appears as 
the generic point of $\spec \Z$ exactly as $\spec k(t)$ appeared as the
generic point of $\spec k[t]$.

Now sending an integer or a rational number to $\Qp$ is the exact
algebraic analogue of writing the Taylor expansion of a polynomial or
a rational fraction. That is the reason why $\spec \Zp$ should be
thought of as a formal neighborhood of the point $p \in \spec \Z$.
Similarly $\spec \Qp$ is the punctured formal neighborhood around $p$.

The second drawing of Figure~\ref{fig:alggeom} (which really looks like 
the first one) displays the rational number $\frac{909} 5$ viewed as a 
function over $\spec \Z$ (with $5$ as point of indeterminacy) together 
with its local $p$-adic/Taylor expansion at $p = 3$ and $p = 7$.

\subsubsection{Local-global principle}
\label{sssec:localglobal}

When studying equations where the unknown is a function, it is often 
interesting to look at local properties of a potential solution.
Typically, if we have to solve a differential equation:
$$a_d(t) y^{(d)}(t) + a_{d-1}(t) y^{(d-1)}(t) + \cdots + a_1(t) y'(t)
+ a_0 y(t) = b(t)$$
a standard strategy consists in looking for analytic solutions of the 
shape $y(t) = \sum_n c_n (t{-}a)^n$ for some $a$ lying in the base 
field. The differential equation then rewrites as a recurrence on the 
coefficients $c_n$ which sometimes can be solved. This reasoning yields
local solutions which have to be glued afterwards.

Keeping in mind the analogy between functions and integers/rationals, we 
would like to use a similar strategy for studying Diophantine equations
over $\Q$. Consider then a Diophantine equation of the shape:
\begin{equation}
\label{eq:polyeq}
P(x_1, \ldots, x_n) = 0
\end{equation}
where $P$ is a polynomial with rational coefficients and the $x_i$'s are
the unknowns. If Eq.~\eqref{eq:polyeq} has a global solution, \emph{i.e.}
a solution in $\Q^n$, then it must have local solutions everywhere,
\emph{i.e.} a solution in $\Qp^n$ for each prime number $p$. Indeed $\Q$
embeds into each $\Qp$. We are interested in the converse: assuming that
Eq.~\eqref{eq:polyeq} has local solutions everywhere, can we glue them 
in order to build a global solution? 
Unfortunately, the answer is negative in general. There is nevertheless
one remarkable case for which this principle works well.

\begin{theo}[Hasse--Minkowski]
Let $P(x_1, \ldots, x_n)$ be a multivariate polynomial. We assume that
$P$ is quadratic, \emph{i.e.} that the total degree of $P$ is $2$.
Then,
the equation \eqref{eq:polyeq} has a solution in $\Q^n$ if and only if
it has a solution in $\R^n$ and in $\Qp^n$ for all prime numbers $p$.
\end{theo}

We refer to~\cite{Se70} for the proof of this theorem (which is not
really the purpose of this course).
Understanding the local-global principle beyond the case of quadratic 
polynomials has motivated a lot of research for more than 50 years.
In 1970, at the International Congress of Mathematicians in Nice, Manin 
highlighted a new obstruction of cohomological nature to the possibility of glueing local 
solutions~\cite{Ma71}. This obstruction is called nowadays the 
\emph{Brauer--Manin obstruction}\footnote{The naming comes from the fact 
this obstruction is written in the language of Brauer groups... the 
latter being defined by Grothendieck in the context we are interested 
in.}. Exhibiting situations where it can explain, on its own, the 
non-existence of rational solutions is still an active domain of 
research today, in which very recent breakthrough progresses have been 
done~\cite{HaWi16,Ha16}.

\subsection{Why should we implement $p$-adic numbers?}

We have seen that $p$-adic numbers are a wonderful mathematical object 
which might be quite useful for arithmeticians. However it is still not 
clear that it is worth implementing them in mathematical software.
The aim of this subsection is to convince the reader that it is
definitely worth it. Here are three strong arguments supporting this 
thesis:
\begin{enumerate}[(A)]
\renewcommand{\itemsep}{0pt}
\item \emph{$p$-adic numbers provide sometimes better numerical stability;}
\item \emph{$p$-adic numbers provide a solution for ``allowing for
division by $p$ in $\Fp$'';}
\item \emph{$p$-adic numbers really appear in nature.}
\end{enumerate}
In the next paragraphs (\S\S \ref{sssec:Hilbert}--\ref{sssec:nature}), 
we detail several examples illustrating the above arguments and showing 
that $p$-adic numbers appear as an essential tool in many questions 
of algorithmic nature.

\subsubsection{The Hilbert matrix}
\label{sssec:Hilbert}

The first aforementioned argument, namely the possibility using $p$-adic 
numbers to have better numerical stability, is used in several contexts 
as factorization of polynomials over $\Q$ or computation of Galois 
groups of number fields.
In order to avoid having to introduce too advanced concepts here, we 
have chosen to focus on a simpler example which was pointed out in 
Vaccon's PhD thesis \cite[\S 1.3.4]{Va15}: the computation of 
the inverse of the Hilbert matrix. Although this example is not directly 
related to the most concrete concerns of arithmeticians, it already very 
well highlights the potential of $p$-adic numbers when we are willing
to apply numerical methods to a problem which is initially stated over 
the rationals.

We recall that the Hilbert matrix of size $n$ is the square matrix $H_n$ 
whose $(i,j)$ entry is $\frac 1 {i+j-1}$ ($1 \leq i,j \leq n$). For 
example:
$$H_4 = \left( \begin{matrix}
1 & 1/2 & 1/3 & 1/4 \\
1/2 & 1/3 & 1/4 & 1/5 \\
1/3 & 1/4 & 1/5 & 1/6 \\
1/4 & 1/5 & 1/6 & 1/7 
\end{matrix}\right).$$
Hilbert matrices are famous for many reasons. One of them is that 
they are very ill-conditioned, meaning that numerical computations 
involving Hilbert matrices may lead to important numerical errors.
A typical example is the computation of the inverse of $H_n$. Let us
first mention that an exact formula giving the entries of $H_n^{-1}$
is known:
\begin{equation}
\label{eq:Hninv}
(H_n^{-1})_{i,j} = (-1)^{i+j} \cdot (i+j-1) 
\cdot \binom{n+i-1}{n-j}
\cdot \binom{n+j-1}{n-i}
\cdot \binom{i+j-2}{i-1}^2.
\end{equation}
(see for example~\cite{Ch83}).
We observe in particular than $H_n^{-1}$ has integral coefficients.

We now move to the numerical approach: we consider $H_n$ as a matrix 
with real coefficients and compute its inverse using standard Gaussian 
elimination (with choice of pivot) and IEEE floating-point arithmetics 
(with $53$ bits of precision)~\cite{IEEE08}. Here is the result we get 
with \sage for $n = 4$:
$$\small
H_4^{-1} \approx \left( \begin{matrix}
\phantom{+}15.9999999999998 & -119.999999999997 & \phantom{+}239.999999999992 & -139.999999999995 \\
-119.999999999997 & \phantom{+}1199.99999999996 & -2699.99999999989 & \phantom{+}1679.99999999993 \\
\phantom{+}239.999999999992 & -2699.99999999989 & \phantom{+}6479.99999999972 & -4199.99999999981 \\
-139.999999999995 & \phantom{+}1679.99999999993 & -4199.99999999981 & \phantom{+}2799.99999999987
\end{matrix} \right).$$
We observe that the accuracy of the computed result is acceptable but not so
high: the number of correct binary digits is about $45$ (on average 
on each entry of the matrix), meaning then that the number of incorrect 
digits is about $8$. Let us now increase the size of the matrix and observe 
how the accuracy behaves:

\begin{center}
\begin{tabular}{|c||c|c|c|c|c|c|c|c|c|c|}
\hline
{\small \begin{tabular}{@{}c@{}} size of \vspace{-0.3ex}\\the matrix\end{tabular}} 
& $5$ & $6$ & $7$ & $8$ & $9$ & $10$ & $11$ & $12$ & $13$ \\
\hline
{\small \begin{tabular}{@{}c@{}} number of \vspace{-0.3ex}\\ correct digits\end{tabular}}
& $40$ & $34$ & $28$ & $25$ & $19$ & $14$ & $9$ & $4$ & $0$ \\
\hline
\end{tabular}
\end{center}

\noindent
We see that the losses of accuracy are enormous.

\smallskip

On the other hand, let us examine now the computation goes when 
$H_n^{-1}$ is viewed as a matrix over $\Q_2$. Making experimentations in 
\sagenoref using again Gaussian elimination and the straightforward 
analogue of floating-point arithmetic 
(see \S \ref{ssec:pfloat}) with $53$ bits of precision, we 
observe the following behavior:
\begin{center}
\begin{tabular}{|c||c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
{\small \begin{tabular}{@{}c@{}} size of \vspace{-0.3ex}\\the matrix\end{tabular}} 
& $5$ & $6$ & $7$ & $8$ & $9$ & $10$ & $11$ & $12$ & $13$ & $50$ & $100$ \\
\hline
{\small \begin{tabular}{@{}c@{}} number of \vspace{-0.3ex}\\ correct digits\end{tabular}}
& $52$ & $52$ & $51$ & $51$ & $51$ & $51$ & $51$ & $51$ & $51$ & $49$ & $48$ \\
\hline
\end{tabular}
\end{center}

\noindent
The computation of $H_n^{-1}$ seems quite accurate over $\Q_2$ whereas it 
was on the contrary highly inaccurate over $\R$. As a consequence, if we 
want to use numerical methods to compute the exact inverse of $H_n$ over 
$\Q$, it is much more interesting to go through the $2$-adic numbers. Of 
course this approach does not make any sense if we want a \emph{real} 
approximation of the entries of $H_n^{-1}$; in particular, if we are 
only interesting in the size of the entries of $H_n^{-1}$ (but not in
their exact values) passing through the $p$-adics is absurd since two 
integers of different sizes might be very close in the $p$-adic world.

\medskip

The phenomena occurring here are actually easy to analyze. The accuracy
of the inversion of a real matrix is governed by the 
condition number which is defined by:
$$\cond_\R(H_n) = \Vert H_n \Vert_\R \cdot \Vert H_n^{-1} \Vert_\R$$
where $\Vert \cdot \Vert_\R$ is some norm on $M_n(\R)$. 
According to Eq.~\eqref{eq:Hninv}, the entries of $H_n$ are large: 
as an example, the bottom right entry of $H_n$ is equal to $(2n-1) \cdot 
\binom{2n-2}{n-1}^2$ and thus grows exponentially fast with respect to 
$n$. As a consequence the condition number $\cond_\R(H_n)$ is quite 
large as well; this is the source of the loss of accuracy observed
for the computation of $H_n^{-1}$ over $\R$.

Over $\Qp$, one can argue similarly and consider the 
$p$-adic condition number:
$$\cond_{\Qp}(H_n) = 
  \Vert H_n \Vert_{\Qp} \cdot \Vert H_n^{-1} \Vert_{\Qp}$$
where $\Vert \cdot \Vert_{\Qp}$ is the infinite norm over $M_n(\Qp)$
(other norms do not make sense over $\Qp$ because of the ultrametricity).
Since $H_n^{-1}$ has integral coefficients, all its entries have their
$p$-adic norm bounded by $1$. Thus $\Vert H_n^{-1} \Vert_{\Qp} \leq 1$. 
As for the Hilbert matrix $H_n$ itself, its norm is equal to $p^v$ where 
$v$ is the highest power of $p$ appearing in a denominator of an entry 
of $H_n$, \emph{i.e.} $v$ is the unique integer such that $p^v \leq 2n-1 
< p^{v+1}$. Therefore $\Vert H_n \Vert_{\Qp} \leq 2n$ and 
$\cond_{\Qp}(H_n) = O(n)$. The growth of the $p$-adic condition number 
in then rather slow, explaining why the computation of $H_n^{-1}$ is 
accurate over $\Qp$.

\subsubsection{Lifting the positive characteristic}
\label{sssec:diffeq}

In this paragraph, we give more details about the mysterious sentence: 
\emph{$p$-adic numbers provide a solution for ``allowing for divisions by $p$ 
in $\Fp$''}.
Let us assume that we are working on a problem that makes sense over any 
field and always admits a unique solution (we will give examples later). 
To fix ideas, let us agree that our problem consists in designing a fast 
algorithm for computing the aforementioned unique solution. Assume 
further that such an algorithm is available when $k$ has characteristic 
$0$ but does not extend to the positive characteristic (because it
involves a division by $p$ at some point).
In this situation, one can sometimes take advantage of $p$-adic numbers to attack the 
problem over the finite field $\Fp$, proceeding concretely in three
steps as follows:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item we lift our 
problem over $\Qp$ (meaning that we introduce a $p$-adic instance of 
our problem whose reduction modulo $p$ is the problem we have started
with),
\item we solve the problem over $\Qp$ (which has characteristic $0$),
\item we finally reduce the solution modulo $p$. 
\end{enumerate}
The existence and uniqueness of the solution together ensure that the
solution of the $p$-adic problem is defined over $\Zp$ and reduces 
modulo $p$ to the correct answer.
Concretely, here are two significant examples where the above strategy 
was used with success: the fast computation of composed products 
\cite{BoGoPeSc05}
and the fast computation of isogenies in positive characteristic 
\cite{LeSi08,LaVa16}.

\medskip

In order to give more substance to our thesis, we briefly detail the 
example of composed products. Let $k$ be a field. Given two monic 
polynomials $P$ and $Q$ with coefficients in $k$, we recall that the 
composed product of $P$ and $Q$ is defined by:
$$P \otimes Q = \sum_\alpha \sum_\beta (X - \alpha \beta)$$
where $\alpha$ (resp. $\beta$) runs over the roots of $P$ (resp. 
of $Q$) in an algebraic closure of $k$, with the convention that a
root is repeated a number of times equal to its multiplicity. Note
that $P \otimes Q$ is always defined over $k$ since all its coefficients
are written as symmetric expressions in the $\alpha$'s and the $\beta$'s.
We address the question of the fast multiplication of composed products.

If $k$ has characteristic $0$, Dvornicich and Traverso~\cite{DvTr89} proposed
a solution based on Newton sums. Indeed, letting $S_{P,n}$ and $S_{Q,n}$ 
be the Newton sums of $P$ and $Q$ respectively:
$$S_{P,n} = \sum_\alpha \alpha^n \quad and \quad 
S_{Q,n} = \sum_\beta \beta^n.$$
It is obvious that the product $S_{P,n} \cdot S_{Q,n}$ is the $n$-th 
Newton sum of $P \otimes Q$. Therefore, if we design fast algorithms for 
going back and forth between the coefficients of a polynomial and its 
Newton sums, the problem of fast computation of the composed product 
will be solved.

Schönhage~\cite{Sc82} proposed a nice solution to the latter problem. 
It relies on a remarkable differential equation relating a polynomial
and the generating function of the sequence of its Newton sums.
Precisely, let $A$ be a monic polynomial of degree $d$ over $k$. 
We define the formal series:
$$S_A(t) = \sum_{n=0}^\infty S_{A,n+1} t^n.$$
A straightforward computation then leads to the following remarkable relation:
$$(E) : \quad A_\rec'(t) = - S_A(t) \cdot A_\rec(t)$$
where $A_\rec$ is the reciprocal polynomial of $A$ defined by $A_\rec(X) 
= X^d \cdot A(\frac 1 X)$.
If the polynomial $A$ is known, then one can compute $S_A(t)$ 
by performing a simple division in $k[\![t]\!]$. Using fast 
multiplication and Newton iteration for computing the inverse (see \S 
\ref{sssec:inverse}), this method leads to an algorithm that computes 
$S_1, S_2, \ldots, S_N$ for a cost of $\softO(N)$ operations in $k$.

Going in the other direction reduces to solving the differential equation 
$(E)$ where the unknown is now the polynomial $A_\rec$. When $k$ has 
characteristic $0$, Newton iteration can be used in this context (see 
for instance~\cite{BoChLeSaSc07}). This leads to the following 
recurrence:
$$B_1(t) = 1 \quad ; \quad
B_{i+1}(t) = B_i(t) - B_i(t) 
\int \left(\frac{B'_i(t)}{B_i(t)} + S_A(t) \right)dt$$
where the integral sign stands for the linear operator acting on 
$k[\![t]\!]$ by $t^n \mapsto \frac{t^{n+1}}{n+1}$.
One proves that $B_i(t) \equiv A_\rec(t) \pmod {t^{2^i}}$ for all 
$i$. Since $A_\rec(t)$ is a polynomial of degree $d$, it is then enough 
to compute it modulo $t^{d+1}$. Hence after $\log_2 (d{+}1)$ iterations 
of the Newton scheme, we have computed all the relevant coefficients. 
Moreover all the computations can be carried by omitting all terms of 
degree $> d$. This leads to an algorithm that computes $A_\rec$ (and 
hence $A$) for a cost of $\softO(d)$ operations in $k$.

In positive characteristic, this strategy no longer works because the 
integration involves divisions by integers. When $k$ is the finite 
field with $p$ elements\footnote{The same strategy actually extends to 
all finite fields.}, one can tackle this issue by lifting the problem
over $\Zp$, following the general strategy discussed here. First, we 
choose two polynomials $\hat P$ and $\hat Q$ with coefficients in $\Zp$ 
with the property that $\hat P$ (resp. $\hat Q$) reduces to $P$ (resp. 
$Q$) modulo $p$. Second, applying the characteristic $0$ method to
$\hat P$ and $\hat Q$, we compute $\hat P \otimes \hat Q$. Third, we
reduce the obtained polynomial modulo $p$, thus recovering $P \otimes
Q$. This strategy was concretized by the work of Bostan, Gonzalez-Vega, 
Perdry and Schost~\cite{BoGoPeSc05}.
The main difficulty in their work is to keep control on the $p$-adic 
precision we need in order to carry out the computation and be sure to 
get the correct result at the end; this analysis is quite subtle and was 
done by hand in~\cite{BoGoPeSc05}. Very general and powerful tools are now
available for dealing with such questions; they will be presented in
\S \ref{sec:precision}.

\subsubsection{$p$-adic numbers in nature}
\label{sssec:nature}

We have already seen in the introduction of this course that $p$-adic 
numbers are involved in many recent developments in Number Theory and 
Arithmetic Geometry (see also \S \ref{sssec:localglobal}). Throughout 
the 20th century, many $p$-adic objects have been defined and studied as 
well: $p$-adic modular/automorphic forms, $p$-adic differential 
equations, $p$-adic cohomologies, $p$-adic Galois representations, 
\emph{etc.} It turns out that mathematicians, more and more often, feel 
the need to make ``numerical'' experimentations on these objects for 
which having a nice implementation of $p$-adic numbers is of course a 
prerequisite.

Moreover $p$-adic theories sometimes have direct consequences in other 
areas of mathematics. A good example in this direction is the famous 
problem of counting points on algebraic curves defined over finite 
fields (\emph{i.e.} roughly speaking counting the number of solutions in 
finite fields of equations of the shape $P(X,Y) = 0$ where $P$ is 
bivariate polynomial). This question has received much attention during
the last decade because of its applications to cryptography (it serves
as a primitive in the selection of secure elliptic curves).
Since the brilliant intuition of Weil~\cite{We49} followed by the revolution 
of Algebraic Geometry conducted by Grothendieck in the 20th century 
\cite{SGA}, the approach to counting problems is now often cohomological.
Roughly speaking, if $C$ is an algebraic variety (with some additional 
assumptions) defined over a finite field $\F_q$, then the number of points
of $C$ is related to the traces of the Frobenius map acting on the
cohomology of $C$. Counting points on $C$ then ``reduces'' to computing
the Frobenius map on the cohomology. Now it turns out that traditional
algebraic cohomology theory yields vector spaces over $\Q_\ell$ (for an 
auxiliary prime number $\ell$ which does not need to be equal to the 
characteristic of the finite field we are working with).
This is the way $\ell$-adic numbers enter naturally into the scene.

\section{Several implementations of $p$-adic numbers}
\label{sec:implementations}

Now we are all convinced that it is worth implementing $p$-adic numbers, 
we need to discuss the details of the implementation. The main problem 
arising when we are trying to put $p$-adic numbers on computers is the 
precision. Indeed, remember that a $p$-adic number is defined as an 
infinite sequence of digits, so that it \emph{a priori} cannot be stored 
entirely in the memory of a computer. From this point of view, $p$-adic 
numbers really behave like real numbers; the reader should therefore not 
be surprised if he/she often detects similarities between the solutions 
we are going to propose for the implementation of $p$-adic numbers and 
the usual implementation of reals.

In this section, we design and discuss three totally different 
paradigms: zealous arithmetic (\S \ref{ssec:zealous}), lazy 
arithmetic together with the relaxed improvement (\S \ref{ssec:lazy}) and 
$p$-adic floating-point arithmetic (\S \ref{ssec:pfloat}). Each of 
these ways of thinking has of course its own advantages and 
disadvantages; we will try to compare them fairly in \S
\ref{ssec:compparadigm}.

\subsection{Zealous arithmetic}
\label{ssec:zealous}

Zealous arithmetic is by far the most common implementation of 
$p$-adic numbers in usual mathematical software: \magma, \sage and 
\pari use it for instance. It appears as the exact analogue of the 
interval arithmetic of the real setting: we replace $p$-adic numbers by 
intervals. The benefit is of course that intervals can be represented 
without errors by finite objects and then can be stored and manipulated 
by computers.

\subsubsection{Intervals and the big-$O$ notation}

Recall that we have defined in \S \ref{sssec:tree} (see Definitions 
\ref{def:intervalZp} and \ref{def:intervalQp}) the notion of interval: 
a bounded interval of $\Qp$ is by definition a closed ball lying inside
$\Qp$, \emph{i.e.} a subset of the form:
$$I_{N,a} = \big\{ \, x \in \Qp 
\quad \text{s.t.} \quad 
|x - a| \leq p^{-N} \, \big\}.$$
The condition $|x - a| \leq p^{-N}$ can be rephrased in a more algebraic 
way since it is equivalent to the condition $x \equiv a \pmod{p^N \Zp}$, 
\emph{i.e.} $x \in a + p^N \Zp$. In short $I_{N,a} = a + p^N \Zp$. In 
symbolic computation, we generally prefer using the big-$O$ notation and 
write:
$$I_{N,a} = a + O(p^N).$$
The following result is easy but fundamental for our propose.

\begin{prop}
\label{prop:intervalcomputer}
Each bounded interval $I$ of $\Qp$ may be written as:
$$I = p^v s + O(p^N)$$
where $(N,v,s)$ is either of the form $(N,N,0)$ or a triple of relative 
integers with $v < N$, $s \in [0, p^{N-v})$ and $\gcd(s,p) = 1$.
Moreover this writing is unique.

In particular, bounded intervals of $\Qp$ are representable by exact
values on computers.
\end{prop}

\begin{proof}
It is a direct consequence of the fact that
$I_{N,a}$ depends only on the class of $a$ modulo $p^N$.
\end{proof}

The interval $a + O(p^N)$ 
has a very suggestive interpretation if we are thinking at $p$-adic 
numbers in terms of infinite sequences of digits.
Indeed, write 
$$a = a_v p^v + a_{v+1} p^{v+1} + \cdots + a_{N-1} p^{N-1} + \cdots$$ 
with $0 \leq a_i < p$ and agree to define $a_i = 0$ for 
$i < v$. A $p$-adic number $x$ then lies in $a + O(p^N)$ if its $i$-th 
digit is $a_i$ for all $i < N$. Therefore one may think of the notation 
$a + O(p^N)$ as a $p$-adic number of the shape:
$$a_v p^v + a_{v+1} p^{v+1} + \cdots + a_{N-1} p^{N-1} + {}?\:p^N 
+{} ?\:p^{N+1} + \cdots$$
where the digits at all positions $\geq N$ are unspecified. This 
description enlightens Proposition \ref{prop:intervalcomputer} 
which should become absolutely obvious now.

\medskip

We conclude this paragraph by introducing some additional vocabulary.

\begin{deftn}
Let $I = a + O(p^N)$ be an interval. We define:

\medskip

\noindent
\begin{tabular}{c@{\hspace{1ex}}lll}
$\bullet$ &
the \emph{absolute precision} of $I$: & $\abs(I) = N$; \smallskip \\
$\bullet$ &
the \emph{valuation} of $I$: & $\val(I) = \val(a)$ & if $0 \not\in I$, \\
 && $\val(I) = N$ & otherwise; \smallskip \\
$\bullet$ &
the \emph{relative precision} of $I$: & 
\multicolumn{2}{l}{$\rel(I) = \abs(I) - \val(I)$.}
\end{tabular}
\end{deftn}

\smallskip

We remark that the valuation of $I$ is the integer $v$ of 
Proposition~\ref{prop:intervalcomputer}. It is also always the smallest 
valuation of an element of $I$. Moreover, coming back to the 
interpretation of $p$-adic numbers as infinite sequences of digits, we 
observe that the relative precision of $a + O(p^N)$ is the number of 
known digits (with rightmost zeroes omitted) of the family of $p$-adic 
numbers it represents.

\subsubsection{The arithmetics of intervals}

Since intervals are defined as subsets of $\Qp$, basic arithmetic
operations on them are defined in a straightforward way. For example, 
the sum of the intervals $I$ and $J$ is the subset of $\Qp$ consisting of 
all the elements of the form $a + b$ with $a \in I$ and $b \in J$. As with real 
intervals, it is easy to write down explicit formulas giving the results
of the basic operations performed on $p$-adic intervals.

\begin{prop}
\label{prop:arithinterval}
Let $I = a + O(p^N)$ and $I' = a' + O(p^{N'})$ be two bounded intervals
of $\Qp$. Set $v = \val(I)$ and $v' = \val(I')$. We have:
\begin{align}
I + I' & = a + a' + O\big(p^{\min(N,\,N')}\big) \label{eq:precadd} \\
I - I' & = a - a' + O\big(p^{\min(N,\,N')}\big) \label{eq:precsub} \\
I \times I' & = aa' + O\big(p^{\min(v+N',\,N+v')}\big) \label{eq:precmul} \\
I \div I' & = \frac a{a'} + O\big(p^{\min(v + N' -2v',\, N-v')}\big) \label{eq:precdiv}
\end{align}
where in the last equality, we have further assumed that $0 \not\in I'$.
\end{prop}

\begin{rem}
\label{rem:arithinterval}
Focusing on the precision, we observe that the two first 
formulae stated in Proposition \ref{prop:arithinterval} immediately
imply $\abs(I + I') = \abs(I - I') = \min(\abs(I), \, \abs(I'))$.
Concerning multiplication and division, the results look ugly at 
first glance. They are however much more pretty if we translate them 
in terms of relative precision; indeed, they become simply
$\rel(I\times I') = \rel(I \div I') = \min(\rel(I), \, \rel(I'))$
which is parallel to the case of addition and subtraction and
certainly easier to remember.
\end{rem}

\begin{proof}[Proof of Proposition~\ref{prop:arithinterval}]
The proofs of Eq.~\eqref{eq:precadd} and Eq.~\eqref{eq:precsub} are easy
and left as an exercise to the reader. We then move directly to
multiplication. Let $x \in I \times I'$. By definition, there exist
$h \in p^N \Zp$ and $h' \in p^{N'}\Zp$ such that:
$$x = (a + h) (a' + h')
= aa' + ah' + a'h + hh'.$$
Moreover, coming back to the definition of the valuation of an
interval, we find $v \leq \val(a)$. The term $ah'$ is then divisible 
by $p^{v+N'}$. Similarly $a'h$ is divisible by $p^{N+v'}$. Finally
$hh'$ is divisible by $p^{v+v'}$; it is then \emph{a fortiori} also
divisible by $p^{v+N'}$ because $v \leq N$. Thus $ah'+ a'h + hh'$
is divisible by $p^{\min(v+N',\,N+v')}$ and we have proved one
inclusion:
$$I \times I' \subset aa' + O\big(p^{\min(v+N',\,N+v')}\big).$$
Conversely, up to swapping $I$ and $I'$, we may assume that $\min
(v+N',\,N+v') = v+N'$. Up to changing $a$, we may further suppose
that $\val(a) = v$. Pick now $y \in aa' + O(p^{v+N'})$, \emph{i.e.}
$y = aa' + h$ for some $h$ divisible by $p^{v+N'}$. 
Thanks to our assumption on $a$, we have $\val(\frac h a) = \val(h) - 
\val(a) \geq v + N' - v = N'$, \emph{i.e.} $p^{N'}$ divides
$\frac h a$. Writing $y = a \times (a' + \frac h a)$ then shows 
that $y \in I \times I'$. The converse inclusion follows and
Eq.~\eqref{eq:precmul} is established.

We now assume that $0 \not\in I'$ and start the proof of 
Eq.~\eqref{eq:precdiv}. We define $1 \div I'$ as the set of 
inverses of elements of $I'$. We claim that:
\begin{equation}
\label{eq:oneoverIp}
1 \div I' = \frac 1{a'} + O(p^{N'-2v'}).
\end{equation}
In other to prove the latter relation, we write $I' = a' \times 
(1 + O(p^{N-v'}))$. We then notice that the function $x \mapsto
x^{-1}$ induces a involution from $1 + O(p^{N-v'})$ to itself; in
particular $1 \div (1 + O(p^{N-v'})) = 1 + O(p^{N-v'})$. Dividing
both sides by $a'$, we get \eqref{eq:oneoverIp}. We finally
derive that
$I \div I' = I \times \big(\frac 1{a'} + O(p^{N'-2v'})\big)$.
Eq.~\eqref{eq:precdiv} then follows from Eq.~\eqref{eq:precmul}.
\end{proof}

Using Proposition~\ref{prop:arithinterval}, it is more or less 
straightforward to implement addition, subtraction, multiplication and 
division on intervals when they are represented as triples $(N,v,s)$ as 
in Proposition~\ref{prop:intervalcomputer} (be careful however at the 
conditions on $v$ and $s$). This yields the basis of the \emph{zealous
arithmetic}. % which is implemented in \sage, \magma and \pari.

\subsubsection{Newton iteration}
\label{sssec:intervalNewton}

A tricky point when dealing with zealous arithmetic concerns Newton
iteration. To illustrate it, let us examine the example of the
computation of a square root of $c$ over $\Q_2$ for
$$c = 1 + 2^{3} + 2^{4} + 2^{5} + 2^{10} + 2^{13} + 2^{16} + 2^{17} + 
      2^{18} + 2^{19} + O(2^{20}).$$
We observe that $c \equiv 1 \pmod 8$. By Hensel's Lemma, $c$
has then a unique square root in $\Q_2$ which is congruent to $1$ 
modulo $4$. Let us denote it by $\sqrt c$. Following \S \ref{sssec:sqrt}, 
we compute $\sqrt c$ using Newton iteration: if $(x_i)_{i \geq 0}$ is
the recursive sequence defined by
$$x_0 = 1 \quad ; \quad x_{i+1} = \frac 1 2 \cdot \left( x_i + \frac 
c{x_i}\right), \, i = 0, 1, 2, \ldots$$
we know that $x_i$ and $\sqrt c$ share the same $(2^i + 1)$ final
digits. Here is the result we get if we compute the first $x_i$'s
using zealous arithmetic:

$$\begin{array}{lr}
x_1 = & % 3 correct digits
 {\color{black!70}\ldots1111001001000011}{\color{purple}101} \\
x_2 = & % 5 correct digits
 {\color{black!70}\ldots0011101000111}{\color{purple}10101} \\
x_3 = & % 9 correct digits
 {\color{black!70}\ldots10111010}{\color{purple}001010101} \\
x_4 = & % 17 correct digits
 {\color{black!70}\ldots}{\color{purple}0111010001010101} \\
x_5 = & % 33 correct digits
 {\color{black!70}\ldots}{\color{purple}111010001010101} \\
x_6 = & % 33 correct digits
 {\color{black!70}\ldots}{\color{purple}11010001010101} 
\end{array}$$

\noindent
Here the $2^i{+}1$ rightmost digits of $x_i$ (which are the correct
digits of $\sqrt c$) are colored in purple and the dots represent
the unknown digits, that is the digits which are absorbed by the
$O(-)$. We observe that the precision decreases by $1$ digit at 
each iteration; this is of course due to the division by $2$ in the 
recurrence defining the $x_i$'s.
The maximal number of correct digits is obtained for $x_4$ with 
$16$ correct digits. After this step, the result stabilizes but 
the precision continues to decrease. Combining naively zealous 
arithmetic and Newton iteration, we have then managed to compute 
$\sqrt c$ at precision $O(2^{16})$.

\medskip

It turns out that the losses of precision we have just highlighted has 
no intrinsic meaning but is just a consequence of zealous 
arithmetic. We will now explain that it is possible to slightly modify 
Newton iteration in order to completely eliminate this 
unpleasant phenomenon. The starting point of your argument is 
Remark~\ref{rem:Hensel} which tells us that Newton iteration still 
converges at the same rate to $\sqrt c$ as soon as the sequence $(x_i)$ 
satisfies the weaker condition:
$$\left|x_{i+1} - 
\frac 1 2 \cdot \left( x_i + \frac
c{x_i}\right) \right| \leq 2^{2^{i+1}+1}.$$
In other words, we have a complete freedom on the choice of the 
``non-purple'' digits of $x_{i+1}$. In particular, if some digit of 
$x_{i+1}$ was not computed because the precision on $x_i$ was too poor, 
we can just assign freely our favorite value (\emph{e.g.} $0$) to this 
digit without having to fear unpleasant consequences. Even better, we 
can assign $0$ to each ``non-purple'' digit of $x_i$ as well; this will
not affect the final result and leads to computation with smaller
integers. 
Proceeding this way, we obtain the following sequence:

$$\begin{array}{lrr}
x_1: & % 3 correct digits
 {\color{black!70}\ldots00000000000000000}{\color{purple}101} \\
x_2: & % 5 correct digits
 {\color{black!70}\ldots000000000000000}{\color{purple}10101} \\
x_3: & % 9 correct digits
 {\color{black!70}\ldots00000000000}{\color{purple}001010101} \\
x_4: & % 17 correct digits
 {\color{black!70}\ldots000}{\color{purple}10111010001010101} \\
x_5: & % 33 correct digits
 {\color{black!70}\ldots}{\color{purple}1010111010001010101}
\end{array}$$

\noindent
and we obtain the final result with $19$ correct digits instead of $16$. 
Be very careful: we cannot assign arbitrarily the $20$-th digit of $x_5$ 
because it is a ``purple'' digit and not a ``non-purple'' one.
More precisely if we do it, we have to write it in black (of course, we 
cannot decide randomly what are the correct digits of $\sqrt c$) and a 
new iteration of the Newton scheme again loses this $20$-th digit
because of the division by $2$.
In that case, one may wonder why we did not try to assign a $21$-st digit to 
$x_4$ in order to get one more correct digit in $x_5$. This sounds like
a good idea but does not work because the input $c$ --- on which we do 
not have any freedom --- appears in the recurrence formula defining the 
$x_i$'s and is given at precision $O(2^{20})$. For this reason, each 
$x_i$ cannot be computed with a better precision than $O(2^{19})$.

In fact, we can easily convince ourselves that $O(2^{19})$ is the 
optimal precision for $\sqrt c$. Indeed $c$ can be lifted at precision
$O(2^{21})$ either to:
\begin{align*}
c_1 & = 1 + 2^{3} + 2^{4} + 2^{5} + 2^{10} + 2^{13} + 2^{16} + 2^{17} + 
        2^{18} + 2^{19} + O(2^{21}) \\
\text{or} \quad 
c_2 & = 1 + 2^{3} + 2^{4} + 2^{5} + 2^{10} + 2^{13} + 2^{16} + 2^{17} + 
        2^{18} + 2^{19} + 2^{20} + O(2^{21}).
\end{align*}
The square roots of these liftings are:
\begin{align*}
\sqrt{c_1} & = 1 + 2^2 + 2^4 + 2^6 + 2^{10} + 2^{12} + 2^{13} + 2^{14} + 
               2^{16} + 2^{18} + O(2^{20}) \\
\sqrt{c_2} & = 1 + 2^2 + 2^4 + 2^6 + 2^{10} + 2^{12} + 2^{13} + 2^{14} + 
               2^{16} + 2^{18} + 2^{19} + O(2^{20})
\end{align*}
and we now see clearly that the $19$-th digit of $\sqrt c$ is
affected by the $20$-th digit of $c$.

\subsection{Lazy and relaxed arithmetic}
\label{ssec:lazy}

The very basic idea behind lazy arithmetic is the following: in every 
day life, we are not working with all $p$-adic numbers but only with 
\emph{computable} $p$-adic numbers, \emph{i.e.} $p$-adic numbers for
which there exists a program that outputs the sequence of its digits.
A very natural idea is then to use these programs to represent $p$-adic
numbers. By operating on programs, one should then be able to perform
additions, multiplications, \emph{etc}. of $p$-adic numbers.

In this subsection, we examine further this idea. In \S 
\ref{sssec:lazynaive}, we adopt a very naive point of view, insisting on 
ideas and not taking care of doability/performances. We hope that this 
will help the reader to understand more quickly the mechanisms behind the 
notion of lazy $p$-adic numbers. We will then move to complexity 
questions and will focus on the problem of designing a framework 
allowing our algorithms to take advantage of fast multiplication 
algorithms for integers (as 
Karatsuba's algorithm, Schönhage--Strassen's algorithm 
\cite[\S 8]{GaGe03} or Fürer's algorithm and its improvements 
\cite{Fu09,HaHoLe15}). This will lead us to report on the theory of 
\emph{relaxed algorithms} introduced recently by Van der Hoeven and his
followers~\cite{Ho97,Ho02,Ho07,BeHoLe11,BeLe12,Le13}.

Lazy $p$-adic numbers have been implemented in the software \mathemagix.

\subsubsection{Lazy $p$-adic numbers}
\label{sssec:lazynaive}

\begin{deftn}
A \emph{lazy $p$-adic number} is a program \ttx that takes as
input an integer $N$ and outputs a \emph{rational number} $\ttx
(N)$ with the property that:
$$\big|\ttx(N+1) - \ttx(N)\big|_p \leq p^{-N}$$
for all $N$.
\end{deftn}

The above condition implies that the sequence $\ttx(N)$ is a Cauchy 
sequence in $\Qp$ and therefore converges. We call its limit the 
\emph{value} of $\ttx$ and denote it by $\Value(\ttx)$. Thanks to 
ultrametricity, we obtain $|\Value(\ttx) - \ttx(N)| \leq p^{-N}$ for all 
$N$. In other words, $\ttx(N)$ is nothing but an approximation of 
$\Value(\ttx)$ at precision $O(p^N)$.

The $p$-adic numbers that arise as values of lazy $p$-adic numbers
are called \emph{computable}.
We observe that there are only countably many lazy $p$-adic numbers;
the values they take in $\Qp$ then form a countable set as well. Since $\Qp$
is uncountable (it is equipotent to the set of functions $\N \to
\{0, 1, \ldots, p{-}1\}$), it then exists many uncomputable $p$-adic
numbers.

Our aim is to use lazy $p$-adic numbers to implement actual (computable) 
$p$-adic numbers. In order to do so, we need (at least) to answer the 
following two questions:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item Given (the source code of) two lazy $p$-adic numbers \ttx and 
\tty, can we decide algorithmically whether $\Value(\ttx) = \Value(\tty)$
or not?
\item Can we lift standard operations on $p$-adic numbers at the level
of lazy $p$-adic numbers?
\end{itemize}
Unfortunately, the first question admits a negative answer; it is yet
another consequence of Turing's halting problem. For our purpose, this 
means that it is impossible to implement equality at the level of
lazy $p$-adic numbers. Inequalities however can always be detected. In
order to explain this, we need a lemma.

\begin{lem}
\label{lem:vallazy}
Let \ttx be a lazy $p$-adic number and set $x = \Value(\ttx)$.
For all integers $N$, we have:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
\val_p(x) & = \val_p(\ttx(N)) & \text{if } \val_p(\ttx(N)) < N \smallskip \\
\val_p(x) & \geq N & \text{otherwise.}
\end{array}$$
\end{lem}

\begin{proof}
Recall that the norm is related to the valuation through the formula
$|a| = p^{-\val_p(a)}$ for all $a \in \Qp$. Moreover by definition,
we know that $|x - \ttx(N)| \leq p^{-N}$. If $\val_p
(\ttx(N)) \geq N$, we derive $|\ttx(N)| \leq p^{-N}$ and thus $|x|
\leq p^{-N}$ by the ultrametric inequality.
On the contrary if $\val_p (\ttx(N)) < N$, write $x = (x - \ttx(N))
+ \ttx(N)$ and observe that the two summands have different norms.
Hence $|x| = \min\big(|x - \ttx(N)|, |\ttx(N)|) = \ttx(N)$ and we
are done.
\end{proof}

\noindent
Lemma \ref{lem:vallazy} implies that $\Value(\ttx) = \Value(\tty)$ if and only if 
$\val_p(\ttx(N) - \tty(N)) \geq N$ for all integers $N$. Thus, assuming 
that we have access to a routine $\val_p$ computing the $p$-adic 
valuation of a rational number, we can write down the function 
\texttt{is\_equal} as follows (Python style\footnote{We emphasize that all the 
procedures we are going to write are written in pseudo-code in Python 
\emph{style} but certainly not in pure Python: they definitely do not
``compile'' in Python.}):

\begin{lstlisting}
    def is_equal(x,y):
        for N in (*$\N$*):
            v = (*$\val_p$*)(x(N) - y(N))
            if v < N: return False
        return True
\end{lstlisting}

\noindent
We observe that this function stops if and only if it answers False,
\emph{i.e.} inequalities are detectable but equalities are not.

\medskip
%\subsubsection{Operation on lazy $p$-adic numbers}

We now move to the second question and try to define standard operations 
at the level of lazy $p$-adic numbers. Let us start with addition. 
Given two computable $p$-adic numbers $x$ and $y$ represented by the 
programs \ttx and \tty respectively, it is actually easy to build a 
third program \texttt{x\_plus\_y} representing the sum $x+y$. Here it is:

\begin{lstlisting}
    def x_plus_y(N):
        return x(N) + y(N)
\end{lstlisting}

The case of multiplication is a bit more subtle because of precision.
Indeed going back to Eq.~\eqref{eq:precmul}, we see that, in order to
compute the product $xy$ at precision $O(p^N)$, we need to know $x$
and $y$ at precision $O(p^{N-\val(y)})$ and $O(p^{N-\val(x)})$
respectively. We thus need to compute first the valuation of $x$ and
$y$. Following Proposition~\ref{lem:vallazy}, we write down the
following procedure:

\begin{lstlisting}
    def val(x):
        for N in (*$\N$*):
            v = (*$\val_p$*)(x(N))
            if v < N: return v
        return Infinity
\end{lstlisting}

\noindent
However, we observe one more time that the function \texttt{val} can 
never stop; precisely it stops if and only if $x$ does not vanish.
For application to multiplication, it is nevertheless not an issue.
Recall that we needed to know $\val(x)$ because we wanted to compute
$y$ at precision $O(p^{N-\val(x)})$. But obviously computing $y$ at
higher precision will do the job as well. Hence it is in fact enough
to compute an integer $v_x$ with the guarantee that $\val(x) \geq
v_x$. By Lemma~\ref{lem:vallazy}, an acceptable value of $v_x$ is
$\min(0, \val_p(\ttx(0)))$ (or more generally $\min(i, \val_p(\ttx(i)))$ 
for any value of $i$).
A lazy $p$-adic number whose value if $xy$ is then given by the
following program:

\begin{lstlisting}
    def x_mul_y(N):
        vx = min(0, (*$\val_p$*)(x(0))
        vy = min(0, (*$\val_p$*)(y(0))
        return x(N-vy) * y(N-vx)
\end{lstlisting}

The case of division is similar except that we cannot do the same
trick for the denominator: looking at Eq.~\eqref{eq:precdiv}, we 
conclude that we do not need a lower bound on the valuation of
the denominator but an upper bound.
This is not that surprising since we easily imagine that the division
becomes more and more ``difficult'' when the denominator gets closer 
and closer to $0$. Taking the argument to its limit, we notice that
a program which is supposed to perform a division should be able, as 
a byproduct, to detect whether the divisor is zero or not. But we 
have already seen that the latter is impossible.
We are then reduced to code the division as follows:

\begin{lstlisting}
    def x_div_y(N):
        vx = min(0, (*$\val_p$*)(x(0))
        vy = val(y)  # This step might never stop
        return x(N + vy) / y(N + 2*vy - vx)
\end{lstlisting}

\noindent
keeping in mind that the routine never stops when the denominator 
vanishes.

\subsubsection{Relaxed arithmetic}

The algorithms we have sketched in \S\ref{sssec:lazynaive} are very 
naive and inefficient. In particular, they redo many times a lot of 
computations. For example, consider the procedure \texttt{x\_plus\_y} we 
have designed previously and observe that if we had already computed the 
sum $x + y$ at precision $O(p^N)$ and we now ask for the same sum $x + 
y$ at precision $O(p^{N+1})$, the computation restarts from scratch 
without taking advantage of the fact that only one digit of the result 
remains unknown.

One option for fixing this issue consists basically in implementing a 
``sparse cache'': we decide in advance to compute and cache the 
$\ttx(N)$'s only for a few values of $N$'s (typically the powers of $2$) 
and, on the input $N$, we output $\ttx(N')$ for a ``good'' $N' \geq N$. 
Concretely (a weak form of) this idea is implemented by maintaining two 
global variables \texttt{current\_precision[x]} and 
\texttt{current\_approximation[x]} (attached to each lazy $p$-adic 
number $x$) which are always related by the equation:
\begin{center}
\tt
current\_approximation[x] = x(current\_precision[x])
\end{center}
If we are asking for $\ttx(N)$ for some $N$ which is not greater than 
\texttt{current\_approximation[x]}, we output 
\texttt{current\_precision[x]} without launching a new computation. 
If, instead, $N$ is greater than \texttt{current\_approximation[x]}, we are 
obliged to redo the computation but we take the opportunity to double
the current precision (at least). Here is the corresponding code:

\begin{lstlisting}
    def x_with_sparse_cache(N):
        if current_precision[x] < N:
            current_precision[x] = max(N, 2*current_precision[x])
            current_approximation[x] = x(current_precision[x])
        return current_approximation[x]
\end{lstlisting}

\noindent
This solution is rather satisfying but it has nevertheless several 
serious disadvantages: it often outputs too large results\footnote{Of 
course it is always possible to reduce the result modulo $p^N$ but this 
extra operation has a non-negligible cost.} (because they are too 
accurate) and often does unnecessary computations.

Another option was developed more recently first by van der Hoeven in 
the context of formal power series~\cite{Ho97,Ho02,Ho07} and then by 
Berthomieu, Lebreton, Lecerf and van der Hoeven for $p$-adic numbers 
\cite{BeHoLe11,BeLe12,Le13}. It is the so-called \emph{relaxed 
arithmetic} that we are going to expose now. For simplicity we shall 
only cover the case of $p$-adic \emph{integers}, \emph{i.e.} $\Zp$. 
Extending the theory to $\Qp$ is more or less straightforward but needs 
an additional study of the precision which makes the exposition more 
technical without real benefit.

\paragraph{Data structure and addition}

The framework in which relaxed arithmetic takes place is a bit different 
and more sophisticated than the framework we have used until now. First 
of all, relaxed arithmetic does not view a $p$-adic number as a sequence 
of more and more accurate approximations but as the sequence of its 
digits. Moreover it needs a richer data structure in order to provide
enough facilities for designing its algorithms. For this reason, it is
preferable to encode $p$-adic numbers by classes which may handle its own
internal variables and its own methods.

We first introduce the class \texttt{RelaxedPAdicInteger} which is an 
abstract class for all relaxed $p$-adic integers (\emph{i.e.} relaxed 
$p$-adic integers must be all instances of a derived class of 
\texttt{RelaxedPAdicInteger}). The class \texttt{RelaxedPAdicInteger} is 
defined as follows:

\begin{lstlisting}
    class RelaxedPAdicInteger:
        # (*\color{comment}\textrm{Variable}*)
        digits           # (*\color{comment}\textrm{list of (already computed) digits}*)

        # (*\color{comment}\textrm{Virtual method}*)
        def next(self)   # (*\color{comment}\textrm{compute the next digit and append it to the list}*)
\end{lstlisting}

\noindent
We insist on the fact that the method \texttt{next} is an internal
(private) method which is not supposed to be called outside the class.
Of course, although it does not appear above, we assume that the class 
\texttt{RelaxedPAdicInteger} is endowed with several additional public methods 
making the interface user-friendly. For instance if \ttx is a relaxed 
$p$-adic integer inheriting from \texttt{RelaxedPAdicInteger}, we shall often
use the construction $\ttx[N]$ to access to the $N$-th digit of \ttx.
In pure Python, this functionality can be implemented by adding to the
class a method \texttt{\_\_getitem\_\_} written as follows:

\begin{lstlisting}
        def __getitem__(self,N):
            n = len(self.digits)   # (*\color{comment}\textrm{Index of the first uncomputed digit}*)
            while n < N+1:         # (*\color{comment}\textrm{We compute the digits until the position \ttN}*)
                self.next()
                n += 1
            return self.digits[N]  # (*\color{comment}\textrm{We return the \ttN-th digit}*)
\end{lstlisting}

In order to highlight the flexibility of the construction, let us 
explain as a warm-up how addition is implemented in this framework.
One actually just follows the schoolbook algorithm: the $n$-th digit
of a sum is obtained by adding the $n$-th digit of the summands plus
possibly a carry. We add an additional local variable to the class in
order to store the current carry and end up with the following
implementation:

\begin{lstlisting}
    class x_plus_y(RelaxedPAdicInteger):
        # (*\color{comment}\textrm{Additional variable}*)
        carry            # (*\color{comment}\textrm{variable for storing the current carry}*)

        def next(self):  # (*\color{comment}\textrm{compute the next digit and append it to the list}*)
            n = len(self.digits)         # (*\color{comment}\textrm{index of the first uncomputed digit}*)
            s = x[n] + y[n] + self.carry # (*\color{comment}\textrm{perform the addition}*)
            self.digits[n] = s % p       # (*\color{comment}\textrm{compute and store the new digit}*)
            self.carry = s // p          # (*\color{comment}\textrm{update the carry}*)
\end{lstlisting}

\noindent
That's all. Although this does not appear in the pseudo-code above, the 
relaxed $p$-adic integers \ttx and \tty (which are supposed to be 
instances of the class \texttt{RelaxedPAdicInteger}) should be passed as 
attributes to the constructor of the class.

The subtraction is performed similarly (exercise left to the reader).

\paragraph{Multiplication}

\emph{A priori}, multiplication can be treated in a similar fashion. 
Given $x$ and $y$ two $p$-adic integers with digits $x_i$'s and $y_j$'s, 
the $n$-th digit of a product $xy$ is obtained by adding the 
contribution of all cross products $x_i y_j$ for $i+j=n$ plus a possible 
carry. This approach leads to the following straightforward 
implementation:

\begin{lstlisting}
    class x_mul_y(RelaxedPAdicInteger):
        # (*\color{comment}\textrm{Additional variable}*)
        carry            # (*\color{comment}\textrm{variable for storing the current carry}*)

        def next(self):  # (*\color{comment}\textrm{compute the next digit and append it to the list}*)
            n = len(self.digits)
            s = self.carry
            for i in 0,1,...,n:
                s += x[i] * y[n-i]
            self.digits[n] = s % p
            self.carry = s // p
\end{lstlisting}

\noindent
Nevertheless this strategy has one important defect: it does not take 
advantage of the fast multiplication algorithms for integers, that is, it 
computes the first $N$ digits of $xy$ in $O(N^2 \log p)$ 
bit operations while we would have expected only $\softO(N \log p)$.

\medskip

The relaxed multiplication algorithm from~\cite{BeHoLe11} fixes this drawback. The 
rough idea behind it is to gather digits in $x$ and $y$ as follows:
\begin{align*}
x & = x_0 + (x_1 + x_2 p) p + (x_3 + x_4 p + x_5 p^2 + x_6 p^3) p^3 + \cdots \\
y & = y_0 + (y_1 + y_2 p) p + (y_3 + y_4 p + y_5 p^2 + y_6 p^3) p^3 + \cdots
\end{align*}
and to implement a kind of block multiplication. More precisely, we 
materialize the situation by a grid drawn in the half plane (see 
Figure~\ref{fig:relaxedmul}).
%
\begin{figure}
\hfill
\begin{tikzpicture}[scale=0.5]
\fill[black!20] (0,0)--(11,0)--(11,7)--(7,7)--(7,11)--(0,11)--cycle;
\draw[very thin,black!30] (0,0) grid (15,15);
\draw (0,0) rectangle (15,15);
\draw (0,0) grid (15,1);
\draw (0,1) grid (1,15);
\draw[xshift=1cm,yshift=1cm,step=2] (0,0) grid (14,2);
\draw[xshift=1cm,yshift=1cm,step=2] (0,2) grid (2,14);
\draw[xshift=3cm,yshift=3cm,step=4] (0,0) grid (12,4);
\draw[xshift=3cm,yshift=3cm,step=4] (0,4) grid (4,12);
\fill (10.5,0.5) circle (1.5mm);
\fill (9.5,1.5) circle (1.5mm);
\fill (8.5,2.5) circle (1.5mm);
\fill (7.5,3.5) circle (1.5mm);
\fill (6.5,4.5) circle (1.5mm);
\fill (5.5,5.5) circle (1.5mm);
\fill (4.5,6.5) circle (1.5mm);
\fill (3.5,7.5) circle (1.5mm);
\fill (2.5,8.5) circle (1.5mm);
\fill (1.5,9.5) circle (1.5mm);
\fill (0.5,10.5) circle (1.5mm);

\node[below,scale=0.8] at (0.5,0) { $0$ };
\node[below,scale=0.8] at (1.5,0) { $1$ };
\node[below,scale=0.8] at (2.5,0) { $2$ };
\node[below,scale=0.8] at (3.5,0) { $3$ };
\node[below,scale=0.8] at (4.5,0) { $4$ };
\node[below,scale=0.8] at (5.5,0) { $5$ };
\node[below,scale=0.8] at (6.5,0) { $6$ };
\node[below,scale=0.8] at (7.5,0) { $7$ };
\node[below,scale=0.8] at (8.5,0) { $8$ };
\node[below,scale=0.8] at (9.5,0) { $9$ };
\node[below,scale=0.8] at (10.5,0) { $10$ };
\node[below,scale=0.8] at (11.5,0) { $11$ };
\node[below,scale=0.8] at (12.5,0) { $12$ };
\node[below,scale=0.8] at (13.5,0) { $13$ };
\node[below,scale=0.8] at (14.5,0) { $14$ };
\node[left,scale=0.8] at (0,0.5) { $0$ };
\node[left,scale=0.8] at (0,1.5) { $1$ };
\node[left,scale=0.8] at (0,2.5) { $2$ };
\node[left,scale=0.8] at (0,3.5) { $3$ };
\node[left,scale=0.8] at (0,4.5) { $4$ };
\node[left,scale=0.8] at (0,5.5) { $5$ };
\node[left,scale=0.8] at (0,6.5) { $6$ };
\node[left,scale=0.8] at (0,7.5) { $7$ };
\node[left,scale=0.8] at (0,8.5) { $8$ };
\node[left,scale=0.8] at (0,9.5) { $9$ };
\node[left,scale=0.8] at (0,10.5) { $10$ };
\node[left,scale=0.8] at (0,11.5) { $11$ };
\node[left,scale=0.8] at (0,12.5) { $12$ };
\node[left,scale=0.8] at (0,13.5) { $13$ };
\node[left,scale=0.8] at (0,14.5) { $14$ };
\end{tikzpicture}
\hfill\null

\caption{Relaxed multiplication scheme}
\label{fig:relaxedmul}
\end{figure}
%
The coordinate on the $x$-axis (resp. the $y$-axis) corresponds to the 
integers $i$'s (resp. $j$'s) that serve as indices for the positions of 
the digits of the $p$-adic integer $x$ (resp. $y$). As for the cell 
$(i,j)$, it represents the product $x_i y_j$.
The computation of the $n$-th digit of $xy$ needs to know all the digits 
appearing on the anti-diagonal $i+j = n$; we refer again to
Figure~\ref{fig:relaxedmul} where this anti-diagonal is displayed for $n=10$.

We now pave the grid using larger and larger squares as follows. We pave 
the first row\footnote{It is of course the row corresponding to $j=1$ 
which is drawn on the bottom in Figure~\ref{fig:relaxedmul}.} and the 
first column by squares of size $1$. Once this has been done, we forget
temporarily the cells that have been already paved (\emph{i.e.} the 
cells lying on the first row or the first column), we 
group the two next lines and the two next columns together and we pave 
them with squares of size $2$. We continue this process and double the
size of the squares at each iteration. The obtained result is displayed
on Figure~\ref{fig:relaxedmul}.
This construction exhibits the following remarkable property:

\begin{lem}
\label{lem:paving}
If $(i,j)$ is a cell located on a square of size $2^\ell$ of the paving, 
then $2^\ell < i$ and $2^\ell < j$.
\end{lem}

\begin{proof}
The squares of size $2^\ell$ are all located by construction on the 
region where both coordinates $i$ and $j$ are greater than or equal to 
$1 + 2 + 2^2 + \cdots + 2^{\ell-1} = 2^\ell - 1$. The Lemma follows from 
this observation.
\end{proof}

\noindent
As an example, look at the cell $(7,3)$ on Figure~\ref{fig:relaxedmul}; 
it is the bottom left corner of a square of size $4$.
Let us denote by
$S_{i,j}$ the square of the paving with bottom left corner located at
position $(i,j)$. For such any $(i,j)$, we define:
$$C_{i,j}(t) = \big(x_i + t x_{i+1} + \cdots + t^\ell x_{i+2^\ell-1}\big)
\times \big(y_j + t y_{j+1} + \cdots + t^\ell y_{j+2^\ell-1}\big) \in
\Z[t]$$
where $2^\ell$ is the size of $S_{i,j}$ and $t$ is a new formal
variable.
We are now ready to explain the concrete idea behind the relaxed 
multiplication algorithm. In the \texttt{next} procedure, when we 
encounter a pair $(i,j=n{-}i)$ we distinguish between two cases:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item if $(i,j)$ is the position of a cell located at the bottom left
corner of square of the paving, we add the contribution of $C_{i,j}(p)$
(instead of adding the sole contribution of $x_i y_j$),
\item otherwise, we do nothing (since the contribution of $x_i y_j$ 
has already been taken into account previously).
\end{itemize}
Here are several several important remarks. First, by 
Lemma~\ref{lem:paving}, the computation of the $C_{i,j}(p)$'s in the 
first step only requires the digits of $x$ and $y$ up to the position 
$n$. Second, notice that listing all the pairs $(i,j)$ fitting into 
the first case is easy. Indeed, notice that $(i,j)$ is the bottom left 
corner of a paving square of size $2^\ell$ if and only if $i$ and $j$ 
are both congruent to $2^\ell - 1$ modulo $2^\ell$ and one of them is 
actually equal to $2^\ell - 1$. Hence, for a given nonnegative integer 
$\ell$, there exist at most two such cells $(i,j)$ for which $i+j = n$ 
and they are moreover given by explicit formulas.

At that point, one may wonder why we have introduced the polynomials 
$C_{i,j}(t)$'s instead of working only with the values $C_{i,j}(p)$'s. 
The reason is technical: this modification is needed to get a good 
control on the size of the carries\footnote{Another option, used in 
\cite{BeHoLe11}, would be to build an algorithm for performing 
operations on integers written in base $p$. This can be achieved for 
instance using Kronecker substitution~\cite[Corollary~8.27]{GaGe03}.}. 
We will elaborate on this later (see Remark \ref{rem:carry} below).
By introducing the construction $\ttx[i, \ldots, j]$ for giving access
to the polynomial
$$x_i + x_{i+1} t + x_{i+2} t^2 + \cdots + x_j t^{j-i} \in \Z[t]$$
the above ideas translate into the concrete algorithm of 
Figure~\ref{fig:relaxedmulimpl}.
%
\begin{figure}
\begin{lstlisting}
    class x_mul_y(RelaxedPAdicInteger):
        # (*\rm\color{comment}Additional variable*)
        carry            # (*\rm\color{comment}variable in $\Z[t]$ for storing the current carry*)

        def next(self):  # (*\rm\color{comment}compute the next digit and append it to the list*)
            n = len(self.digits)
            m = n + 2; (*\rm$\ell$*) = 0; s = 0
            while m > 1:
                # (*\rm\color{comment}The contribution of the first square of size $2^\ell$*)
                s += x[2^(*\rm$\ell$*) - 1, ..., 2^((*\rm$\ell$*)+1) - 2]
                   * y[(m-1)*2^(*\rm$\ell$*) - 1, ..., m*2^(*\rm$\ell$*) - 2]
                # (*\rm\color{comment}The contribution of the second square*)
                if m > 2:   # (*\rm\color{comment}Case where the two squares are indeed not the same*)
                    s += y[2^(*\rm$\ell$*) - 1, ..., 2^((*\rm$\ell$*)+1) - 2]
                       * x[(m-1)*2^(*\rm$\ell$*) - 1, ..., m*2^(*\rm$\ell$*) - 2]
                if m (*\rm is odd*): break
                m = m // 2
                (*\rm$\ell$*) += 1
            s += self.carry
            self.digits[n] = s(0) % p
            self.carry = (s(0) // p) + (s // t)
\end{lstlisting}

\caption{An implementation of the relaxed multiplication}
\label{fig:relaxedmulimpl}
\end{figure}

We now briefly sketch its complexity analysis.
The first observation is that the while loop of the \texttt{next} method 
of Figure \ref{fig:relaxedmulimpl} is repeated until $\ell$ reaches the 
$2$-adic valuation of $n+2$. Writing $v = \val_2(n+2)$, the total 
complexity for the execution of the while loop therefore stays within: 
$$\sum_{\ell=0}^v \softO(2^\ell \log p) = \softO(2^v \log p)$$
if we use fast algorithms for polynomial multiplication~\cite[\S 8]{GaGe03}.
Moreover at the end of the while loop, the degree of $s$ remains 
bounded by $2^v$ and its coefficients have all at most $O(\log(np))$
binary digits.
The second step, which is more technical, consists in bounding the 
size of the carry. The key point is to notice that after the computation
of the $(n{-}1)$-st digit, the carry takes the form:
$$c + (c_0 + c_1 t + c_2 t^2 + \cdots + c_n t^n)$$
where the $c_k$'s come from the contribution of the cells $(i',j')$ with 
$i' + j' \geq n$ that have been already discovered (\emph{i.e.} belonging 
to a paving square $S_{i,j}$ for which $C_{i,j}(t)$ has already been
computed) and $c$ comes from the carries appearing while computing the
$t^{n-1}$-term of the product
$(x_0 + x_1 t + x_2 t^2 + \cdots + x_n t^n) \times
(y_0 + y_1 t + y_2 t^2 + \cdots + y_n t^n)$.
Once this has been noticed, we deduce that the $c_k$'s are
all bounded from above by $n\cdot (p{-}1)^2$ while:
$$c \leq \frac 1 {p^n} \sum_{i'+j' \leq n} x_{i'} y_{j'}
\leq np\cdot (p{-}1).$$
Hence the carry itself is a polynomial of degree at most $n$ and each
of its coefficients has at most $O(\log(pn))$ binary digits.

\begin{rem}
\label{rem:carry}
While the growth of the coefficients of the carry remains under 
control, its value at $p$ may --- and does --- grow exponentially fast 
with respect to $n$. That is the reason why we had to switch to 
polynomials.
\end{rem}

Noticing finally that the division by $t$ on polynomials is free (it is 
just a shift of the coefficients), we find that the last three lines 
of the \texttt{next} method displayed on Figure
\ref{fig:relaxedmulimpl} have a cost of $\softO(2^v \log(np))$ bit
operations. The total complexity of the computation of the $n$-th
digit then stays within $\softO(2^v \log(np))$ bit operations as well.

Summing up all these contributions up to $N$, we find that the relaxed 
multiplication algorithm computes the first $N$ digits of a product in 
$\softO(N \log p)$ bit operations.
We refer to~\cite{BeHoLe11} for a more precise complexity analysis (of a 
slightly different version of the algorithm we have presented here).

\paragraph{Computation of fixed points}

A quite interesting application of the relaxed approach is the 
possibility to define and compute very easily fixed points of 
contraction mappings defined over $\Zp$ (or more generally $\Zp^d$ for 
some $d$). Before going into this, we need to define the notion of
\emph{function} over relaxed $p$-adic numbers. 
Recall that a relaxed $p$-adic integer is an instance of 
\texttt{RelaxedPAdicInteger}. A function over relaxed $p$-adic numbers 
can then be defined as a class deriving from 
\texttt{RelaxedPAdicInteger} endowed with a constructor accepting \ttx 
as parameter and constructing $F(\ttx)$. Actually we have already seen
such examples: the classes \texttt{x\_plus\_y} and \texttt{x\_mul\_y}
we have designed before fit exactly in this framework.
Below is yet another easy example modeling the function $s : x \mapsto 
1 + px$.

\begin{lstlisting}
    class S(RelaxedPAdicInteger):
        def next(self):
            n = len(self.digits)
            if n == 0: self.digits[0] = 1
            else:      self.digits[n] = x[n-1]
\end{lstlisting}

\begin{deftn}
Let $F$ be a function defined over relaxed $p$-adic integers.
We say that $F$ is a \emph{contraction} if its \texttt{next} method 
computes the $n$-th digit by making only use of the first $n{-}1$ digits 
of \ttx.
\end{deftn}

\noindent
Concretely $F$ is a contraction if its \texttt{next} method never
calls $\ttx[i]$ for some $i \geq n$ when $n$ denotes (as usual) the 
position of the digit the \texttt{next} method is currently computing.
For example, the function \texttt{S} introduced above is a contraction.

\begin{rem}
If $F$ is a contraction modeling a function $f : \Zp \to \Zp$,
the first $n$ digits of $f(x)$ depends only on the first $n{-}1$ digits
of $x$. This implies that $f$ is a $\frac 1 p$-contraction in the 
usual sense:
$$\textstyle 
|f(x) - f(y)| \leq \frac 1 p \cdot |x-y|.$$
Note that the converse is not necessarily true: it may happen that $f$
is indeed a contraction but is modeled by the function $F$ which misses
this property.
\end{rem}

If $F$ is a contraction modeling a function $f : \Zp \to \Zp$, the 
Banach fixed point Theorem implies that $f$ has a unique fixed point in 
$\Zp$. Moreover this sequence is the limit of any sequence 
$(x_n)_{n\geq0}$ solution to the recurrence $x_{n+1} = f(x_n)$. This 
property translates immediately in the world of programs: it says that 
we can compute the fixed point of $f$ just by replacing each occurrence 
of \ttx by the current instance of the class (\texttt{self}) in the 
\texttt{next} function of $F$. 
As an example, applying this treatment to the function $S$, we get:

\begin{lstlisting}
    class SFixed(RelaxedPAdicInteger):
        def next(self):
            n = len(self.digits)
            if n == 0: self.digits[0] = 1
            else:      self.digits[n] = self[n-1]
\end{lstlisting}

\noindent
which is\footnote{Or more precisely: ``whose any instance is''} a 
relaxed $p$-adic number representing the fixed point of $s$, namely
$\frac 1{1-p}$.

\paragraph{Division}

The above toy example has a real interest because it can be generalized 
to handle arbitrary divisions $a \div b$ where $b$ is invertible in $\Zp$.
We assume first that $b$ is congruent to $1$ modulo $p$, \emph{i.e} $b = 
1 + pb'$ for some $b' \in \Zp$. The assumption means that the $0$-th 
digit of $b$ is 1; as for the digits of $b'$, they are those of $b$ 
shifted by one position. We introduce the affine function $f : \Zp \to 
\Zp$ defined by:
$$f(x) = a + (1-b) x = a - p b' x.$$
It is a contraction which can be modeled (using relaxed multiplication 
and relaxed subtraction) by a contraction $F$ defined over relaxed
$p$-adic integers. Applying the recipe discussed above, we then build
a new relaxed $p$-adic integer that models the unique fixed point of
$x$, which is $\frac a {1+pb'} = \frac a b$. We have therefore computed
the division of $a$ by $b$ in the relaxed world.

For a general $b$, one can proceed as follows. Let $b_0$ be the $0$-th 
digit of $b$. We compute an integer $c$ such that $b_0 c \equiv 1 \pmod 
p$ and view $c$ as a $p$-adic number. Since $b \equiv b_0 \pmod p$, we 
get $bc \equiv 1 \pmod p$ as well. Moreover the fraction $\frac a b$ 
rewrites $\frac a b = \frac{ac}{bc}$ and we can now apply the method 
above. Using this techniques, a general relaxed division reduces to one
inversion modulo $p$ and two relaxed multiplication for the preparation 
plus one more relaxed multiplication and one relaxed subtraction for 
the computation of the fixed point.

\subsection{Floating-point arithmetic}
\label{ssec:pfloat}

The most common representation of real numbers on computers is
by far floating-point arithmetic which is normalized by the
IEEE 754-2008 Standard~\cite{IEEE08,Mu09}.
Let us recall very briefly that the rough idea behind the floating-point 
arithmetic is to select a finite fixed subset $\calR$ of $\R$ 
(the so-called set of \emph{floating-point numbers}) on which we
define the operators $\oplus$ and $\otimes$ which are supposed to mimic
the usual addition and multiplication of reals.

In the $p$-adic world, this approach has an analogue which was first 
proposed (as far as we know) in a unpublished note by Kedlaya and Roe 
in 2008~\cite{KeRo08}. Unfortunately it seems that it has never been implemented 
yet\footnote{Although an implementation was very recently proposed by 
Roe for integration in \sagenoref; see 
\url{https://trac.sagemath.org/ticket/20348}.}. The aim of the 
subsection is to report on Kedlaya and Roe's proposal.

\subsubsection{Definition of $p$-adic floating-point numbers}

The construction of $p$-adic floating-point numbers depends on the 
initial choice of three positive integers $N$, $e_\min$, $e_\max$ that 
we fix now until the end of \S \ref{ssec:pfloat}. We assume $e_\min < 
e_\max$ and $N \geq 1$.
The integer $N$ is called the \emph{precision} of the system while 
$e_\min$ and $e_\max$ are respectively the minimal and the maximal 
\emph{exponent} of the system.

\begin{deftn}
A \emph{$p$-adic floating-point number} is either 
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item a $p$-adic number $x \in \Qp$ of the shape $p^e s$ with the 
conditions:
\begin{equation}
\label{eq:condpFP}
\textstyle
e_\min \leq e \leq e_\max \quad ; \quad
-\frac{p^N-1} 2 < s \leq \frac{p^N-1} 2 \quad ; \quad
\gcd(s,p) = 1
\end{equation}
\item the number $0 \in \Qp$,
\item the special symbol $\infty$,
\item the special symbol $\nan$ (not a number).
\end{itemize}
\end{deftn}

\noindent
Note that the conditions~\eqref{eq:condpFP} imply that a $p$-adic 
floating-point number $x \in \Qp$ of the shape $p^e s$ cannot be zero
and moreover that $e$ and $s$ are uniquely determined by $x$. Indeed, 
since $s$ is prime with $p$, it cannot vanish and $e$ has to be equal
to the $p$-adic valuation of $x$. It is thus determined by $x$ and,
consequently, so is $s = p^{-e} \: x$.

\begin{deftn}
Given a $p$-adic floating-point number $x$ of the shape $x = p^e s$,
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item the integer $e$ is called the \emph{exponent} of $x$, 
\item the integer $s$ is called the \emph{significand} (or sometimes
the \emph{mentissa}) of $x$.
\end{itemize}
\end{deftn}

\paragraph{Representation of $p$-adic floating-point numbers.}

We represent $p$-adic floating-point numbers on machine by pairs
of integers as follows:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item the $p$-adic number $p^e s$ is represented by $(e,s)$,
\item the $p$-adic number $0$ is represented by $(e_\max,0)$,
\item the symbol $\infty$ is represented by $(e_\min{-}1, 1)$,
\item the symbol $\nan$ is represented by $(e_\min{-}1, 0)$.
\end{itemize}

\begin{rem}
The above definitions omit several components of the IEEE 
standard. First, they do not include subnormal numbers. These are 
meant to allow for gradual underflow to zero, which is reasonable to 
provide when working in base $2$ with relatively small $e$. However, for 
$e$ as large as the bit length of a modern machine word (\emph{i.e.} no 
less than 32), this benefit is sufficiently minor that it does not 
outweigh the added complexity needed to handle subnormal numbers 
properly.
Second, the above definitions do not include signed infinities, because $p$-adic numbers 
do not have a meaningful notion of sign. For similar reasons, the comparison
operators $<, \leq, >, \geq$ are not defined for $p$-adic numbers.
Third, they do not provide for multiple types of \nan's. For instance, in 
the IEEE specification, a distinction is made between signaling \nan's 
(those which raise an exception upon any operation) and quiet \nan's.
\end{rem}

For some computations, it may be useful to allow some $p$-adic floating-point 
numbers to be represented in more than one way. We thus define an 
operation called \emph{normalization} on the set of pairs of integers 
$(e,s)$ as follows:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item if $e < e_\min$ and $s=0$, return $(e_\min{-}1,0)$ (which represents $\nan$);
\item if $e < e_\min$ and $s \neq 0$, return $(e_\min{-}1,1)$ (which represents $\infty$);
\item if $e > e_\max$, return $(e_\max,0) = 0$ (which represents $0$);
\item if $e_\min \leq e \leq e_\max$ and $s=0$, return $(e_\max,0) = 0$;
\item if $e_\min \leq e \leq e_\max$ and $s \neq 0$, let $k \in 
\{0,\dots,m-1\}$ be the largest integer such that $p^k$ divides $s$ and 
return the normalization of $(e+k, s')$ where $s'$ is the unique integer
such that $-\frac{p^N-1} 2< s' \leq \frac{p^N-1} 2$ and
$s' \equiv p^{-k} m \pmod{p^N}$.
\end{itemize}

\noindent
For many operations, it is safe to leave an unnormalized pair $(e,s)$ 
as long as $s \neq 0$ and $e_\min \leq e \leq e_\max - \val_p(s)$ since
$p^e s$ is then guaranteed to equal the value of the normalization of 
$(e,s)$. However, one should normalize before testing for equality.

\subsubsection{Operations on $p$-adic floats}

Let $\QpFP$ be the set of $p$-adic floating-point numbers. Define also 
$\QpNan = \Qp \sqcup \{\infty, \nan\}$. Clearly $\QpFP \subset \QpNan$.

\begin{lem}
\label{lem:roundingFP}
Given $x \in \Qp$ with $e_\min \leq \val(x) \leq e_\max$, there
exists a unique element $y \in \Qp \cap \QpFP$ such that
$|x-y| \leq |x| \cdot p^{-N}$.
\end{lem}

\begin{proof}
We write $x = p^v \cdot x_0$ where $v = \val(x)$ and $x_0$ is
invertible in $\Zp$. Set $e = v$ and let $s$ be the unique integer
of the range $\big(\!-\frac{p^N-1} 2,  \frac{p^N-1} 2\big]$ which is
congruent to $x_0$ modulo $p^N$. Define $y = p^e s$; it is a $p$-adic
floating-point number by the assumptions of the Lemma. Moreover:
$$|x - y| = p^{-v} \cdot |x_0 - s| \leq p^{-v-N} = |x| \cdot p^{-N}.$$
The existence is proved.

As for uniqueness, remark first that $y = 0$ cannot do the job. Assume now 
that $y' = p^{e'} s'$ is a $p$-adic floating-point number written in 
normalized form which satisfies the required inequality. First notice 
that $e' = v$ necessarily. Indeed, from $e' > v$, we would deduce $|x - 
y'| = |x| > |x| \cdot p^{-N}$. Similarly from $e_i < v$, we 
would get $|x - y'| = p^{-e_1} > |x|$.
Consequently $|x-y'| = |x| \cdot |x_0 - s'|$ and our assumption on $y'$
translates to the congruence $x_0 \equiv s' \pmod {p^N}$. Therefore we
derive $s \equiv s' \pmod{p^N}$ and then $s = s'$ since both $s$ and $s'$
have to lie in the interval $\big(\!-\frac{p^N-1} 2,  \frac{p^N-1} 2\big]$.
Hence $y = y'$ and we are done.
\end{proof}

\begin{deftn}
The \emph{rounding function} is the function 
$o : \QpNan \to \QpFP$ defined as follows:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item if $x \in \{\infty, \nan\}$, then $o(x) = x$;
\item if $x \in \Qp$ and $\val(x) < e_\min$,
then $o(x) = \infty$ (\emph{overflow});
\item if $x \in \Qp$ and $\val(x) > e_\max$,
then $o(x) = 0$ (\emph{underflow});
\item if $x \in \Qp$ and $e_\min \leq \val(x) \leq e_\max$,
then $o(x)$ is the unique element of $\QpFP$ such that
$|x - o(x)| \leq |x| \cdot p^{-N}$.
\end{itemize}
\end{deftn}

\begin{rem}
We emphasize that overflow (resp. underflow) occurs when $e$ is 
small (resp. large) which is the exact opposite of the real case.
This is \emph{not} a typo; the reason behind that is just that, in
the $p$-adic world, $p^N$ goes to $0$ when $N$ goes to $+\infty$.
\end{rem}

\begin{rem}
Note that the uniqueness in the last case means that there is no 
analogue of the notion of a rounding mode in real arithmetic.
\end{rem}

Given any $k$-ary function $f : (\QpNan)^k \to \QpNan$, we define its 
\emph{compliant approximation} $f_\FP : (\QpFP)^k \to \QpFP$ by:
$$f_\FP(x_1, \ldots, x_k) = o \big(f(x_1, \ldots, x_k)\big).$$
We are now in position to specify the four basic arithmetic operations 
(addition, subtraction, multiplication, division) on $\QpFP$. In order 
to do so, we need first to extend them on $\QpNan$. Let us agree to set
(here $\div$ stands for the division operator):

$$\begin{array}{ll}
\forall x \in \QpNan, 
  & x \pm \nan = \nan \pm x = \nan \smallskip \\
\forall x \in \QpNan \backslash \{\nan\}, 
  & x \pm \infty = \infty \pm x = \infty \medskip \\

\forall x \in \QpNan,
  & x \times \nan = \nan \times x = \nan \smallskip \\
\forall x \in \QpNan \backslash \{0, \nan\}, 
  & x \times \infty = \infty \times x = \infty \smallskip \\
  & 0 \times \infty = \infty \times 0 = \nan \medskip \\

\forall x \in \QpNan,
  & x \div \nan = \nan \div x = \nan \smallskip \\
\forall x \in \QpNan \backslash \{0, \nan\}, 
  & x \div 0 = \infty, \quad 0 \div 0 = \nan \smallskip \\
\forall x \in \QpNan \backslash \{\infty, \nan\}, 
  & x \div \infty = 0, \quad \infty \div \infty = \nan \smallskip \\
\end{array}$$
Addition, subtraction, multiplication and division 
on floating-point $p$-adic numbers are, by definition, the operations
$+_\FP$, $-_\FP$, $\times_\FP$ and $\div_\FP$ respectively.

\smallskip

Let us examine concretely how the computation goes at the level of
$p$-adic floating-point numbers. We start with multiplication with
is the easiest operator to handle. The formula we derive immediately
from the definition is:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
(p^{e_1} s_1) \: \times_\FP \: (p^{e_2} s_2)
  & = p^{e_1 + e_2} \cdot (s_1 s_2 \text{ mod } p^N)
  & \text{if } e_1 + e_2 \leq e_\max \smallskip \\
  & = 0
  & \text{otherwise}
\end{array}$$
where $a \text{ mod } p^N$ is the unique representation of $a$ modulo
$p^N$ lying in the interval $\big(\!-\frac{p^N-1} 2,  \frac{p^N-1} 2\big]$.
We emphasize moreover that if $(e_1, s_1)$ and $(e_2, s_2)$ are both
written in normalized form then so is $(e_1 + e_2, s_1 s_2 \text{ mod } p^N)$ (under the 
assumption $e_1 + e_2 \leq e_\max$). Indeed the product of two numbers
which are coprime to $p$ is coprime to $p$ as well.
The case of the division is similar:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
(p^{e_1} s_1) \: \div_\FP \: (p^{e_2} s_2)
  & = p^{e_1 - e_2} \cdot (s_1 s_2^{-1} \text{ mod } p^N)
  & \text{if } e_1 - e_2 \geq e_\min \smallskip \\
  & = \infty
  & \text{otherwise.}
\end{array}$$
The cases of addition and subtraction are more subtle because they 
can introduce cancellations of the rightmost digits. When $e_1
\neq e_2$, this cannot happen and the formula writes as follows:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
(p^{e_1} s_1) \: +_\FP \: (p^{e_2} s_2)
  & = p^{e_1} \cdot (s_1 + p^{e_2 - e_1} s_2 \text{ mod } p^N)
  & \text{if } e_1 < e_2 \smallskip \\
  & = p^{e_2} \cdot (p^{e_1-e_2} s_1 + s_2 \text{ mod } p^N)
  & \text{if } e_2 < e_1.
\end{array}$$
On the contrary, when $e_1 = e_2$, the naive formula
$(p^e s_1) \: +_\FP \: (p^e s_2)
  = p^e \cdot (s_1 + s_2 \text{ mod } p^N)$
might fail because $s_1 + s_2$ can have a huge valuation. Instead 
we introduce $v = \val_p(s_1 + s_2)$ and the correct formula writes:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
(p^e s_1) \: +_\FP \: (p^e s_2)
  & = p^{e+v} \cdot \big(\frac{s_1 + s_2}{p^v} \text{ mod } p^N\big)
  & \text{if } v \leq e_\max - e \smallskip \\
  & = 0
  & \text{otherwise.}
\end{array}$$
As an alternative, we may prefer using non-normalized representations
of $p$-adic floating-point numbers; it is indeed an option but it
requires to be \emph{very} careful with underflows.

\subsection{Comparison between paradigms}
\label{ssec:compparadigm}

We have just presented three quite different ways of thinking of
$p$-adic numbers from the computational point of view. Of course, 
each of them has its own advantages and disadvantages regarding
flexibility, rapidity, accuracy, \emph{etc.}
In this subsection we compare them, trying as much as possible to
support our arguments with many examples.

\subsubsection{Zealous arithmetic \textsc{vs.} lazy/relaxed arithmetic}
\label{sssec:zealousVSrelaxed}

The main difference between zealous and lazy/relaxed arithmetic is
certainly that they are designed for different uses! In the zealous
perspective, the inputs are given with a certain precision which is
supposed to be fixed once for all: one can imagine for instance that
the inputs come from measures\footnote{Through I have to confess that
I have never seen a relevant physical $p$-adic measure yet.} or that
computing them requires a lot of effort and is then not an option.
So the precision on the inputs is immutable and the goal of zealous
arithmetic is to propagate the precision to the output (trying to
be as sharp as possible).

On the other hand, the point of view of lazy/relaxed arithmetic is the 
exact opposite: instead of fixing the precision on the inputs, we fix a 
\emph{target} precision and we propagate it back to the inputs. Indeed 
suppose, as an easy example, that we have to evaluate the product $xy$. In 
the lazy/relaxed world, this is done by creating and returning the 
program \texttt{x\_mul\_y} \emph{without running it}.
The execution (\emph{i.e.} the actual computation of the product) is 
postponed until somebody asks for the value of $xy$ at a certain 
precision $O(p^N)$ (through the call $\texttt{x\_mul\_y}[N]$). At that 
moment, the program runs and infers the needed precision on the inputs 
$x$ and $y$.

Except maybe in very particular situations, the point of view of 
lazy/relaxed arithmetic is certainly much closer to the needs of the 
mathematician. Indeed, it is quite rare to have to work with $p$-adic
numbers that do not come from a previous computation. Mathematicians
are more relaxed than zealous!

Apart from that, still regarding precision, the zealous and the 
lazy/relaxed approaches are both based on the arithmetic of intervals 
and, for this reason, they exhibit similar behaviors in the following 
sense. Assume that we are given a certain function $f : \Zp \to \Zp$ to 
be evaluated and define the two numbers $\ell_1$ and $\ell_2$ (depending 
\emph{a priori} on extra parameters) as follows:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item when we ask for the value of $f(a + O(p^N))$, the zealous 
approach answers $b + O(p^{N-\ell_1})$ (loss of $\ell_1$ digits of 
precision),
\item when we ask for the value of $f(a)$ at precision $O(p^N)$, 
the lazy/relaxed technique requires the computation of $a$ at precision 
$O(p^{N+\ell_2})$.
\end{itemize}
It turns out that $\ell_1$ and $\ell_2$ do not very much depend 
neither on $a$, nor on $N$. Moreover they are almost 
equal\footnote{We refer to \S \ref{sssec:preclemmaatwork} for a 
clarification of this statement based on a theoretical study (which
is itself based on the theory of $p$-adic precision we shall develop
in \S \ref{ssec:foundpadicprec}).}. On the 
other hand, we underline that this value $\ell_1 \approx \ell_2$ is very 
dependent on the function $f$ and even on the algorithm we use for the 
evaluation. We will present and discuss this last remark in \S 
\ref{sssec:compexamples} below.

\medskip

The lazy/relaxed arithmetic has nevertheless several more or less 
annoying disadvantages. First of all, it is
% more conceptual and so 
certainly more difficult to implement efficiently (although the 
implementation in \mathemagix is very well polished and quite 
competitive; we refer to~\cite{BeHoLe11} for timings). It is also 
supposed to be a bit slower; for instance, the relaxed multiplication 
looses asymptotically a factor $\log N$ (for the precision $O(p^N)$) 
compared to the zealous approach.
Another more serious issue is memory. Indeed the relaxed arithmetic 
needs to keep stored all intermediate results by design. As an easy 
toy example, let us have a quick look at the following function

\begin{lstlisting}
    def nth_term(n)
        u = (*$a$*)
        for i in (*$1, 2, \ldots, n$*):
            u = (*$f$*)(u)
        return u
\end{lstlisting}

\noindent
that computes the $n$-th term of a recursive sequence defined by its 
initial value $u_0 = a$ and the recurrence $u_{i+1} = f(u_i)$. Here $a$ 
and $f$ are given data.
In zealous arithmetic, the above implementation requires to store only 
one single value of the sequence $(u_i)_{i \geq 0}$ at the same time 
(assuming that we can rely on a good garbage collector); indeed at the 
$i$-th iteration of the loop, the value of $u_i$ is computed and stored 
in the variable \texttt{u} while the previous value of \texttt{u}, 
\emph{i.e.} the value of $u_{i-1}$, is destroyed.
On the other hand, in relaxed arithmetic, the variable representing 
$u_i$ stores the definition of $u_i$, including then the value of 
$u_{i-1}$. The relaxed $p$-adic number returns by the function 
\texttt{nth\_term} is then a huge structure in which all the relaxed 
$p$-adic numbers $u_1, u_2, \ldots, u_n$ have to appear. When we will
then ask for the computation of the first $N$ digits of $u_n$, the
relaxed machinery will start to work and compute --- and store --- 
the values of all the $u_i$'s at the desired precision. This is the price 
to pay in order to be able to compute one more digit of $u_n$ without 
having to redo many calculations.

\subsubsection{Interval arithmetic \textsc{vs.} floating-point arithmetic}
\label{sssec:intervalVSfloat}

$p$-adic floating-point arithmetic has a very serious limitation for 
its use in mathematics: the results it outputs are \emph{not proved} 
and even often wrong! Precisely, the most significand digits of the 
output are very likely to be correct while the least significand digits are 
very likely to be incorrect... and we have \emph{a priori} no way to know 
which digits are correct and which ones are not.
On the other hand, at least in the real setting, interval arithmetic 
is known for its tendency to yield pessimistic enclosures. 
What about the $p$-adic 
case? At first, one might have expected that ultrametricity may help. 
Indeed ultrametricity seems to tell that rounding errors do not 
accumulate as it does in the real world. Unfortunately this simple 
rationale is too naive: in practice, $p$-adic interval arithmetic
does overestimate the losses of precision exactly as real interval
arithmetic does.

The causes are multiple and complex but some of them can be isolated. 
The first source of loss of precision comes from the situation where we 
add two $p$-adic numbers known at different precision 
(\emph{e.g.} two $p$-adic numbers of different sizes). As a toy example, 
consider the function $f : \Zp^2 \to \Zp^2$ mapping $(x,y)$ to $(x+y, 
x-y)$ (it is a similarity in the $p$-adic plane) and suppose that we
want to evaluate $f \circ f$ on the entry $x = 1 + O(p^2)$, $y = 1
+ O(p^{20})$. We then code the following function:
\begin{lstlisting}
    def F(x,y):
        return x+y, x-y
\end{lstlisting}
and call $\ttF(\ttF(\ttx,\tty))$.
Let us have a look at precision. According to 
Proposition~\ref{prop:arithinterval}, the first call $\ttF(\ttx,\tty)$ 
returns the pair $(2 + O(p^2), O(p^2))$ and the final result is then 
$(2 + O(p^2), 2 + O(p^2))$. On the other hand, observing that $f \circ f$
is the mapping taking $(x,y)$ to $(2x,2y)$. Using this, we end up with:
$$\begin{array}{r@{\hspace{0.5ex}}ll}
f \circ f(x,y) & = \big(2 + O(p^2), 2 + O(p^{20})\big) 
  & \text{if } p > 2 \smallskip \\
& = \big(2 + O(2^3), 2 + O(2^{21})\big) 
  & \text{if } p = 2
\end{array}$$
which is much more accurate. 
Interval arithmetic misses the simplification and consequently loses 
accuracy.

A second more subtle source of inaccuracy comes from the fact that 
interval arithmetic often misses \emph{gain} of precision. A very basic 
example is the computation of $px$; written this way, the absolute 
precision increases by $1$. However if, for some good 
reason\footnote{For example, imagine that all the values $x, 2x, 3x, 
\ldots, px$ are needed.}, the product $px$ does not appear explicitly 
but instead is computed by adding $x$ to itself $p$ times, 
interval arithmetic will not see the gain of precision.
It turns out that similar phenomena appear for the multiplicative 
analogue of this example, \emph{i.e.} the computation of $x^p$ for $x = 
a + O(p^N)$. For simplicity assume that $\val(x) = 0$ and $N \geq 
1$. Using again Proposition~\ref{prop:arithinterval}, we find that 
zealous arithmetic leads to the result $x^p = a^p + O(p^N)$. However, 
surprisingly, this result is not optimal (regarding precision) as 
shown by the next lemma.

\begin{lem}
\label{lem:ppower}
If $a \equiv b \pmod{p^N}$, then $a^p \equiv b^p \pmod{p^{N+1}}$.
\end{lem}

\begin{proof}
Write $b = a + p^Nc$ with $c \in \Zp$ and compute:
$$b^p = (a + p^Nc)^p = \sum_{i=0}^p \binom p i a^{p-i} (p^Nc)^{i}.$$
When $i \geq 2$, the corresponding term is divisible by $p^{2N}$ while 
when $i=1$, it equals $p^{N+1} a^{p-1} c$ and is therefore apparently
divisible by $p^{N+1}$. As a consequence $b^p \equiv a^p \pmod{p^{N+1}}$
as claimed.
\end{proof}

\noindent
According to Lemma \ref{lem:ppower}, the correct precision for $x^p$ is (at least) 
$O(p^{N+1})$ which is not detected by arithmetic interval. \sage fixes
this issue by an \emph{ad-hoc} implementation of the power function 
which knows about Lemma~\ref{lem:ppower}. However similar behaviors 
happen in quite a lot of different other situations and they of course 
cannot be all fixed by \emph{ad-hoc} patches.

\begin{rem}
Recall that we have seen in \S \ref{sssec:intervalNewton} that the 
computation of square roots in $\Q_2$ looses one digit of precision. 
It is absolutely coherent with the result above which says that the 
computation of squares gains one digit of precision.
\end{rem}

As a conclusion, a very common situation where $p$-adic floating-point 
arithmetic can be very helpful is that situation where the mathematician 
is experimenting, trying to understand how the new mathematical 
objects he has just introduced behave. At the moment, he does not really need 
proofs\footnote{Well, it is of course often better to have proofs... but 
it is maybe too early.}; he needs fast computations and plausible 
accuracy, exactly what $p$-adic floating-point arithmetic can offer.

\subsubsection{Comparison by everyday life examples}
\label{sssec:compexamples}

We analyze many examples and compare for each of them the accuracy we 
get with $p$-adic floating-point arithmetic on the one hand and with interval 
arithmetic (restricting ourselves to the zealous approach for 
simplicity) on the other hand. Examples are picked as basic and very
common primitives in linear algebra and commutative algebra.

\paragraph{Determinant}

Our first example concerns the computation of the determinant of a 
square matrix with entries in $\Zp$. In order to be as accurate as
possible, we shall always use a division-free algorithm (see for
instance~\cite{Bi11} or~\cite{KaVi05} and the references therein for 
faster solutions).
When $M$ is a generic matrix there is actually not so much to say:
if the entries of the matrix are given with $N$ significand digits,
the same holds for the determinant and this precision is optimal.
Nonetheless, quite interesting phenomena show up for matrices having 
a special form. 
In the sequel, we will examine the case of matrices $M$ of the form $M 
= P D Q$ where $P, Q \in \GL_d(\Zp)$ and $D$ is a diagonal matrix with 
diagonal entries $p^{a_1}, \ldots, p^{a_d}$. We assume that the three
matrices $P$, $D$ and $Q$ are given at precision $O(p^N)$ for some $N$.
Here is a concrete example (picked at random) with $p=2$, $d=4$ and
$N = 10$:
$$\begin{array}{r@{\hspace{0.5ex}}l}
\multicolumn{2}{c}{
  a_1 = 0 \quad ; \quad
  a_2 = 2 \quad ; \quad
  a_3 = 3 \quad ; \quad
  a_4 = 5} \medskip \\
P & = 
\small \left(\begin{array}{rrrr}
\ldots1111100100 & \ldots0110110101 & \ldots0101011000 & \ldots1101010001 \\
\ldots1010001101 & \ldots0110011001 & \ldots1101111000 & \ldots1010100100 \\
\ldots1011101111 & \ldots0100100111 & \ldots0000111101 & \ldots0010010001 \\
\ldots1011111001 & \ldots1000100011 & \ldots1100100110 & \ldots0111100011
\end{array}\right) \medskip \\
Q & = 
\small \left(\begin{array}{rrrr}
\ldots1010110001 & \ldots0010011111 & \ldots1010010010 & \ldots1010001001 \\
\ldots1111101111 & \ldots0111100101 & \ldots0110101000 & \ldots0111100000 \\
\ldots0111110100 & \ldots0010010101 & \ldots0000101111 & \ldots1001100010 \\
\ldots0101111111 & \ldots0101110111 & \ldots1110000011 & \ldots1110000110
\end{array}\right)
\end{array}$$
so that:
\begin{equation}
\label{eq:matrixM}
\begin{array}{r@{\hspace{0.5ex}}l}
M & = 
\small \left(\begin{array}{rrrr}
\ldots0101110000 & \ldots0011100000 & \ldots1011001000 & \ldots0011000100 \\
\ldots1101011001 & \ldots1101000111 & \ldots0111001010 & \ldots0101110101 \\
\ldots0111100011 & \ldots1011100101 & \ldots0010100110 & \ldots1111110111 \\
\ldots0000111101 & \ldots1101110011 & \ldots0011010010 & \ldots1001100001
\end{array}\right)
\end{array}
\end{equation}

\begin{rem}
Each entry of $M$ is known at precision $O(2^{10})$ as well. Indeed any 
permutation of $M$ by an element $H \in 2^{10} M_4(\Z_2)$ can be induced 
by a perturbation of $D$ by the element $P^{-1} H Q^{-1}$ which lies in 
$2^{10} M_4(\Z_2)$ as well because $P$ and $Q$ have their inverses in 
$M_4(\Z_2)$.
\end{rem}

Here are the values for the determinant of $M$ computed according to
the two strategies we want to compare:
\begin{center}
\begin{tabular}{ll}
Interval (zealous) arithmetic: & $\det M = O(2^{10})$ \smallskip \\
Floating-point arithmetic: & $\det M = 2^{10} \times \ldots0001001101$
\end{tabular}
\end{center}
We observe the interval arithmetic does not manage to decide whether
$\det M$ vanishes or not. On the other hand, the floating-point approach
outputs a result with $10$ significand digits by design; the point is
that we do not know \emph{a priori} whether these digits are correct or not.
In our particular case, we can however answer this question by computing 
$\det M$ as the product $\det P \cdot \det D \cdot \det Q = 2^{10} 
\det P \cdot \det Q$ using zealous arithmetic. We find this way:
$$\det M = 2^{10} \times \ldots 01101$$
which means that the result computed by floating-point arithmetic has (at
least) $5$ correct digits. This is actually optimal as we shall see
later in \S \ref{sssec:diffexamples}.

\paragraph{Characteristic polynomial}

We now move to the characteristic polynomial keeping first the same
matrix $M$. Here are the results we get:
$$\begin{array}{cr@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
\multicolumn{6}{l}{\text{Interval (zealous) arithmetic:}}\smallskip\\
\hspace{5mm}\null
& \chi_M(X) = X^4 + {}& (\ldots 0001000010) & X^3 
      + {}& (\ldots 1000101100) & X^2 \\
  & {}+ {}& (\ldots 0011100000) & X^{\phantom{1}} 
      + {}& (\ldots 0000000000) \medskip \\
\multicolumn{6}{l}{\text{Floating-point arithmetic:}}\smallskip\\
& \chi_M(X) = X^4 + {}& (\ldots 0{\color{purple}0001000010}) & X^3 
      + {}& (\ldots 11{\color{purple}1000101100}) & X^2 \\
  & {}+ {}& (\ldots 110{\color{purple}100011100000}) & X^{\phantom{1}} 
      + {}& (2^{10} \times \ldots 00010{\color{purple}01101}) \\
\end{array}$$
where, in the second case, the correct digits are written in 
purple\footnote{We decided which digits are correct simply
by computing at higher precision (with zealous arithmetic in order
to get a guaranteed result).}.
We observe again that, although the computation is not proved 
mathematically, the floating-point arithmetic outputs more accurate
results. As we shall see later (see \S \ref{sssec:diffexamples}) 
the accuracy it gives is even optimal for this particular example.

We consider now the matrix $N = I_4 + M$ where $I_4$ is the identity
matrix of size $4$.
The computation of $\chi_N$ leads to the following results:
$$\begin{array}{cr@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
\multicolumn{6}{l}{\text{Interval (zealous) arithmetic:}}\smallskip\\
\hspace{5mm}\null
& \chi_N(X) = X^4 + {}& (\ldots 0000111110) & X^3 
      + {}& (\ldots 0101101100) & X^2 \\
  & {}+ {}& (\ldots 0101001010) & X^{\phantom{1}} 
      + {}& (\ldots 0100001011) \medskip \\
\multicolumn{6}{l}{\text{Floating-point arithmetic:}}\smallskip\\
& \chi_N(X) = X^4 + {}& (\ldots0{\color{purple}0000111110}) & X^3 
      + {}& (\ldots00{\color{purple}0101101100}) & X^2 \\
  & {}+ {}& (\ldots1{\color{purple}0101001010}) & X^{\phantom{1}} 
      + {}& (\ldots {\color{purple}0100001011}) \\
\end{array}$$
On that example, we remark that interval arithmetic is as accurate
as floating-point arithmetic.
More interesting is the evaluation of $\chi_N$ at $1$, which is 
nothing but the determinant of $M$ we have already computed before.
The values we get are the following:
\begin{center}
\begin{tabular}{ll}
Interval (zealous) arithmetic: & $\chi_N(1) = O(2^{10})$ \smallskip \\
Floating-point arithmetic: & $\chi_N(1) = 0$
\end{tabular}
\end{center}
Although there is no surprise with interval arithmetic, we observe
that floating-point arithmetic now fails to produce any correct digit.

\paragraph{LU factorization}

LU factorization is a classical tool in linear algebra which serves
as primitives for many problems.
We refer to~\cite[\S 2]{AbLo04} for an introduction to the topic. 
Recall briefly that a LU factorization of a matrix $M$ is a 
decomposition of the form $M = LU$ where $L$ (resp. $U$) is a lower triangular 
(resp. upper triangular) matrix. In the sequel, we require moreover that 
the diagonal entries of $L$ are all equal to $1$.
With the normalization, one can prove that any matrix $M$ defined over a 
field admits a \emph{unique} LU factorization as soon as all its principal 
minors\footnote{Recall the $i$-th principal minor of $M$ is the 
determinant of the submatrix of $M$ obtained by selecting the first $i$
rows and first $i$ columns.} do not vanish.

LU factorizations can be computed using standard Gaussian elimination: 
starting from $M$, we first multiply the first column by the appropriate 
scalar in order to make the top left entry equal to $1$ and use it as 
pivot to cancel all coefficients on the first line by operating on 
columns. We then get a matrix of the shape:
$$\left(\begin{matrix}
1 & 0 & \cdots & 0 \\
\star & \star & \cdots & \star \\
\vdots & \vdots & & \vdots \\
\star & \star & \cdots & \star 
\end{matrix}\right)$$
and we continue this process with the submatrix obtained by deleting
the first row and the first column. The matrix we get at the end is
the $L$-part of the LU factorization of $M$.

We propose to explore the numerical stability of LU factorization 
\emph{via} Gaussian elimination in the $p$-adic case. Let us start with 
an example. Consider first the same matrix $M$ as above and write its LU 
factorization $M = LU$ (one can check that its principal minors do not 
vanish). Performing Gaussian elimination as described above within the 
framework of zealous arithmetic on the one hand and within the framework 
of floating-point arithmetic on the other hand, we end up with the 
following matrices:
\begin{equation}
\label{eq:resultLU}
\begin{array}{cr@{\hspace{0.5ex}}l}
\multicolumn{3}{l}{\text{Interval (zealous) arithmetic:}}\smallskip\\
\hspace{5mm}\null 
& L & = 
\small \left(\begin{array}{rrr@{\hspace{2em}}r}
1 & 0 & 0 & 0 \\
2^{-4} \times \ldots001111 & 1 & 0 & 0 \\
2^{-4} \times \ldots010101 & \ldots100011 & 1 & 0 \\
2^{-4} \times \ldots001011 & \ldots010101 & \ldots110 & 1
\end{array}\right) \medskip \\
\multicolumn{3}{l}{\text{Floating-point arithmetic:}}\smallskip\\
\hspace{5mm}\null 
& L & = 
\small \left(\begin{array}{rrr@{\hspace{2em}}r}
1 & 0 & 0 & 0 \\
2^{-4} \times \ldots1010{\color{purple}001111} & 1 & 0 & 0 \\
2^{-4} \times \ldots0110{\color{purple}010101} & \ldots{0\color{purple}011100011} & 1 & 0 \\
2^{-4} \times \ldots0101{\color{purple}001011} & \ldots{\color{purple}0111010101} & 0001{\color{purple}0110110} & 1
\end{array}\right)
\end{array}
\end{equation}
As before the correct digits are displayed in purple. We remark once
again that the accuracy of $p$-adic floating-point arithmetic is better 
than that of interval arithmetic (it is actually even optimal on that
particular example as we shall see in \S \ref{sssec:diffexamples}).
More precisely, we note that the
zealous precision is sharp on the first column but the gap increases
when we are moving to the right.

\begin{rem}
It is somehow classical~\cite[\S 1.4]{Ho75} that the entries of $L$ and $U$ can 
all be expressed as the quotient of two appropriate minors of $M$ 
(Cramer-like formulas). Evaluating such expressions, it is possible to 
compute the LU factorization and stay sharp on precision within the 
zealous framework. The drawback is of course complexity since evaluating 
two determinants for each entry of $L$ and $U$ is clearly very 
time-consuming. A stable and efficient algorithm, combining the 
advantages of the two approaches, is designed in~\cite{Ca12}.
\end{rem}

\paragraph{Bézout coefficients and Euclidean algorithm}

We now move to polynomials and examine the computation of
Bézout coefficients \emph{via} the Euclidean algorithm. We pick at
random two monic polynomials $P$ and $Q$ of degree $4$ over $\Z_2$:

$$\begin{array}{r@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
P = X^4 + {}& (\ldots 1101111111) & X^3 
        + {}& (\ldots 0011110011) & X^2 \\
      {}+ {}& (\ldots 1001001100) & X^{\phantom{1}} 
        + {}& (\ldots 0010111010) \medskip \\
Q = X^4 + {}& (\ldots 0101001011) & X^3 
        + {}& (\ldots 0111001111) & X^2 \\
      {}+ {}& (\ldots 0100010000) & X^{\phantom{1}} 
        + {}& (\ldots 1101000111) 
\end{array}$$

We can check that $P$ and $Q$ are coprime modulo $p$ (observe that
$P + Q \equiv 1 \pmod p$); they are therefore \emph{a fortiori} coprime
in $\Qp[X]$. By Bézout Theorem, there exist two polynomials 
$U, V \in \Z_2[X]$ such that $UP + VQ = 1$. Moreover these polynomials 
are uniquely determined if we require $\deg U, \deg V \leq 3$. Computing 
them with the extended Euclidean algorithm (without any kind of 
algorithmic optimization), we get:
$$\begin{array}{cr@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
\multicolumn{6}{l}{\text{Interval (zealous) arithmetic:}}\smallskip\\
\hspace{1cm}\null
&U = {}& (\ldots 101100) & X^3 
   + {}& (\ldots 101100) & X^2 \\
&{}+ {}& (\ldots 100) & X^{\phantom{1}} 
   + {}& (\ldots 1011) \medskip \\
&V = {}& (\ldots 010100) & X^3 
   + {}& (\ldots 100100) & X^2 \\
&{}+ {}& (\ldots 100) & X^{\phantom{1}} 
   + {}& (\ldots 101) \medskip \\
\multicolumn{6}{l}{\text{Floating-point arithmetic:}}\smallskip\\
&U = {}& (\ldots 101{\color{purple}011101100}) & X^3 
   + {}& (\ldots 111{\color{purple}100101100}) & X^2 \\
&{}+ {}& (\ldots 100000{\color{purple}110100}) & X^{\phantom{1}} 
   + {}& (\ldots 01{\color{purple}10001011}) \medskip \\
&V = {}& (\ldots 010{\color{purple}100010100}) & X^3 
   + {}& (\ldots 00{\color{purple}1011100100}) & X^2 \\
&{}+ {}& (\ldots 00010{\color{purple}0111100}) & X^{\phantom{1}} 
   + {}& (\ldots 111{\color{purple}1100101})
\end{array}$$
Floating-point arithmetic provides again better accuracy, though far for 
being optimal this time. Indeed the theory of subresultants 
\cite[\S 4.1]{Wi96}
shows that the coefficients of $U$ and $V$ can be all expressed as 
quotients of some minors of the Sylvester matrix of $(P,Q)$ by the 
resultant of $P$ and $Q$, denoted hereafter by $\Res(P,Q)$. From the 
fact that $P$ and $Q$ are coprime modulo $p$, we deduce that $\Res(P,Q)$ 
does not vanish modulo $p$. Thus $\val_p(\Res(P,Q)) = 0$ and dividing by
$\Res(P,Q)$ does not decrease the absolute precision according to
Proposition~\ref{prop:arithinterval}. As a consequence, following this
path and staying within the zealous framework, we can calculate
the values of $U$ and $V$ at precision $O(2^{10})$. Here is the result
we get:
$$\begin{array}{r@{\hspace{0.2ex}}r@{\hspace{0.5ex}}
l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}
l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}
l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}}
U = {}& (\ldots 0011101100) & X^3 
  + {}& (\ldots 0100101100) & X^2 
{}+ {}& (\ldots 1101110100) & X^{\phantom{1}} 
  + {}& (\ldots 0010001011) \medskip \\
V = {}& (\ldots 1100010100) & X^3 
  + {}& (\ldots 1011100100) & X^2 
{}+ {}& (\ldots 0110111100) & X^{\phantom{1}} 
  + {}& (\ldots 0001100101).
\end{array}$$
We observe that the latter values (in addition to being proved) are 
even more accurate that the result which was computed ``naively'' using 
floating-point arithmetic. The drawback is complexity since evaluating
many determinants requires a lot of time. 
The theory of $p$-adic precision we are going to introduce in the next
section (see \S \ref{sec:precision} and, more especially, \S
\ref{sssec:adaptiveprecision}) provides the tools for writing a
stabilized version of Euclidean algorithm that computes provably $U$ 
and $V$ at (almost) optimal precision. We refer to~\cite{Ca17} for more
details (the material developed in  \S \ref{sec:precision} is necessary 
to read this reference).

\paragraph{Polynomial evaluation and interpolation}

Polynomial evaluation and polynomial interpolation are very classical 
and useful primitives involved in many algorithms in symbolic 
computation. In this last paragraph, we examine how they behave from
the point of view of precision. We focus on a very basic (but already
very instructive) example.
We consider the following two procedures:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item \texttt{evaluation}: it takes as input a polynomial $P \in \Qp[X]$ 
of degree at most $d$ and outputs $P(0), P(1), \ldots, P(d)$;
\item \texttt{interpolation}: it takes as input a list of values $y_0, 
\ldots, y_d \in \Qp$ and returns the interpolation polynomial $P \in 
\Qp[X]$ of degree at most $d$ such that $P(i) = y_i$ for $i \in \{0,1,\ldots, d\}$.
\end{itemize}
Algorithms (and notably fast algorithms) for these tasks abound in the
literature (see for instance~\cite[\S 10]{GaGe03}). For our propose, we choose naive algorithms: we implement
\texttt{evaluation} by evaluating separately the $P(i)$'s (using Hörner
scheme say) and we implement \texttt{interpolation} using the method
of divided differences~\cite[\S 2]{Hi56}. Under our assumptions 
(interpolation at the first integers), it turns out that it takes a 
particularly simple form that we make explicit now.

Define the \emph{difference operator} $\Delta$ on $\Qp[X]$ by $\Delta 
A(X) = A(X+1) - A(X)$. The values taken by $A$ at the integers are 
related to the values $\Delta^n A(0)$ by a simple closed formula, as
shown by the next lemma.

\begin{lem}
\label{lem:mahler}
For all polynomials $A \in \Qp[X]$ of degree at most $d$, we have:
$$A(X) = \sum_{n=0}^d \Delta^n A(0) \cdot \binom X n
\quad \text{where} \quad
\binom X n = \frac{X(X-1)\cdots (X-n+1)}{n!}.$$
\end{lem}

\begin{proof}
We proceed by induction on $d$. When $d = 0$, the lemma is trivial.
We assume now that it holds for all polynomials $A$ of degree at most
$d$ and consider a polynomial $A \in \Qp[X]$ with $\deg A = d{+}1$.
We define $B(X) = \sum_{n=0}^d \Delta^n A(0) \cdot \binom X n$. 
Remarking that $\Delta \binom X n = 
\binom X{n-1}$ for all positive integers $n$, we derive:
$$\Delta B (X) = \sum_{n=1}^d \Delta^n A(0) \cdot \binom X {n-1}
= \sum_{n=0}^{d-1} \Delta^{n+1} A(0) \cdot \binom X n.$$
From the induction hypothesis, we deduce $\Delta B = \Delta A$ (since 
$\Delta A$ has degree at most $d$). Furthermore, going back to the
definition of $B$, we find $A(0) = B(0)$. All together, these two
relations imply $A = B$ and the induction goes.
\end{proof}

\begin{rem}
Lemma~\ref{lem:mahler} extends by continuity to all continuous
functions $f$ on $\Zp$. In this generality, it states that any
continuous function $f : \Zp \to \Zp$ can be uniquely written as
a convergent series of the shape:
$$f(x) = \sum_{n=0}^\infty a_n \cdot \binom x n$$
where the $a_n$'s lie in $\Zp$ and converge to $0$ when $i$ goes
to infinity. The $a_n$'s are moreover uniquely determined: we have
$a_n = \Delta^n f(0)$. They are called the \emph{Mahler coefficients}
of $f$. We refer to~\cite{Ma58} for much more details on this topic.
\end{rem}

From Lemma~\ref{lem:mahler}, we easily derive an algorithm for our 
interpolation problem. Given $y_0, \ldots, y_d$, we define the 
``divided'' differences $y_{n,i}$ for $0 \leq i \leq n \leq d$ by 
(decreasing) induction on $n$ by $y_{d,i} = y_i$ and $y_{n-1,i} = 
y_{n,i+1} - y_{n,i}$. These quantities can be easily computed.
Moreover, thanks to Lemma~\ref{lem:mahler}, the interpolation 
polynomial $P$ we are looking for writes
$P(X) = \sum_{n=0}^d y_{n,0} \binom x n$.
This provides an algorithm for computing the interpolation polynomial.

Let us now try to apply successively \texttt{evaluation} and
\texttt{interpolation} taking as input a random polynomial $P$ of 
degree $d=8$ with coefficients in $\Zp$ (for $p = 2$ as usual):
$$\begin{array}{r@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
P = {}& (\ldots 0111001110) & X^8 
  + {}& (\ldots 0101010001) & X^7 
  + {}& (\ldots 1000001100) & X^6 \\
{}+ {}& (\ldots 1010001101) & X^5 
  + {}& (\ldots 1111000100) & X^4
  + {}& (\ldots 0011101101) & X^3 \\
{}+ {}& (\ldots 1010010111) & X^2
  + {}& (\ldots 0011011010) & X^{\phantom{1}} 
  + {}& (\ldots 0001011110) 
\end{array}$$
The results we get are:
$$\begin{array}{cr@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
\multicolumn{8}{l}{\text{Interval (zealous) arithmetic:}}\smallskip\\
\hspace{1cm}\null
&    {}& (\ldots 110) & X^8 
   + {}& (\ldots 001) & X^7 
   + {}& (\ldots 100) & X^6 \\
&{}+ {}& (\ldots 101) & X^5 
   + {}& (\ldots 100) & X^4
   + {}& (\ldots 1101) & X^3 \\
&{}+ {}& (\ldots 10111) & X^2
   + {}& (\ldots 1011010) & X^{\phantom{1}} 
   + {}& (\ldots 0001011110) \medskip \\
\multicolumn{8}{l}{\text{Floating-point arithmetic:}}\smallskip\\
\hspace{1cm}\null
&    {}& (\ldots 0000{\color{purple}1001110}) & X^8 
   + {}& (\ldots 101{\color{purple}1010001}) & X^7 
   + {}& (\ldots 00101{\color{purple}0001100}) & X^6 \\
&{}+ {}& (\ldots 0{\color{purple}010001101}) & X^5 
   + {}& (\ldots 0000{\color{purple}11000100}) & X^4
   + {}& (\ldots 010{\color{purple}11011101}) & X^3 \\
&{}+ {}& (\ldots 0{\color{purple}010010111}) & X^2
   + {}& (\ldots 11{\color{purple}011011010}) & X^{\phantom{1}} 
   + {}& (\ldots 0{\color{purple}0001011110}) 
\end{array}$$
The result computed by floating-point arithmetic is again a bit more
accurate. However this observation is not valid anymore where the degree
$d$ gets larger.
%
\begin{figure}
$$\scriptsize
\begin{array}{r@{\hspace{0.2ex}}
r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}
r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}
r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
\multicolumn{9}{l}{\text{Initial polynomial:}}\smallskip\\
     {}& (\ldots 0101101001) & X^{19}
   + {}& (\ldots 1101000011) & X^{18} 
   + {}& (\ldots 0011001110) & X^{17} 
   + {}& (\ldots 1001011010) & X^{16} \\
 {}+ {}& (\ldots 0011100111) & X^{15} 
   + {}& (\ldots 0110101110) & X^{14}
   + {}& (\ldots 0111111001) & X^{13}
   + {}& (\ldots 1011010111) & X^{12} \\
 {}+ {}& (\ldots 0100000100) & X^{11}
   + {}& (\ldots 0000110000) & X^{10}
   + {}& (\ldots 1110101010) & X^{9\phantom{0}}
   + {}& (\ldots 1111101100) & X^{8} \\
 {}+ {}& (\ldots 0100010001) & X^{7\phantom{0}}
   + {}& (\ldots 0101010000) & X^{6\phantom{0}}
   + {}& (\ldots 0111101111) & X^{5\phantom{0}}
   + {}& (\ldots 1100010011) & X^{4} \\
 {}+ {}& (\ldots 0100000001) & X^{3\phantom{0}}
   + {}& (\ldots 1000010010) & X^{2\phantom{0}}
   + {}& (\ldots 0000100000) & X^{\phantom{01}} 
   + {}& (\ldots 0001111110) \medskip \\
\multicolumn{9}{l}{\text{Interval (zealous) arithmetic:}}\smallskip\\
       & O(2^{-6}) & X^{19}
   + {}& O(2^{-6}) & X^{18}
   + {}& O(2^{-6}) & X^{17}
   + {}& O(2^{-5}) & X^{16} \\
 {}+ {}& O(2^{-6}) & X^{15}
   + {}& O(2^{-6}) & X^{14}
   + {}& O(2^{-6}) & X^{13}
   + {}& O(2^{-5}) & X^{12} \\
 {}+ {}& O(2^{-6}) & X^{11}
   + {}& O(2^{-6}) & X^{10}
   + {}& O(2^{-6}) & X^{9\phantom{0}}
   + {}& O(2^{-5}) & X^{8} \\
 {}+ {}& O(2^{-4}) & X^{7\phantom{0}}
   + {}& O(2^{-3}) & X^{6\phantom{0}}
   + {}& O(2^{-2}) & X^{5\phantom{0}}
   + {}& O(2^{-1}) & X^{4} \\
 {}+ {}& (\ldots 1) & X^{3\phantom{0}}
   + {}& (\ldots 010) & X^{2\phantom{0}}
   + {}& (\ldots 100000) & X^{\phantom{01}} 
   + {}& (\ldots 0001111110) \medskip \\
\multicolumn{9}{l}{\text{Floating-point arithmetic:}}\smallskip\\
     {}& (2^{-3} \times \ldots 1110011011) & X^{19}
   + {}& (2^{-5} \times \ldots 0000000011) & X^{18} 
   + {}& (2^{-3} \times \ldots 0001011111) & X^{17} 
   + {}& (2^{-5} \times \ldots 1100111101) & X^{16} \\ 
 {}+ {}& (\ldots 11111100110) & X^{15}
   + {}& (2^{-4} \times \ldots 0110100011) & X^{14} 
   + {}& (2^{-2} \times \ldots 0000010011) & X^{13} 
   + {}& (2^{-4} \times \ldots 1010001101) & X^{12} \\ 
 {}+ {}& (2^{-3} \times \ldots 0010000011) & X^{11}
   + {}& (2^{-5} \times \ldots 0100101111) & X^{10} 
   + {}& (2^{-3} \times \ldots 0000110011) & X^{9\phantom{0}} 
   + {}& (2^{-5} \times \ldots 1010101001) & X^{8} \\ 
 {}+ {}& (2^{-2} \times \ldots 0010000101) & X^{7\phantom{0}}
   + {}& (2^{-2} \times \ldots 1101100111) & X^{6\phantom{0}} 
   + {}& (\ldots 110{\color{purple}1101111}) & X^{5\phantom{0}} 
   + {}& (2^{-1} \times \ldots 0011100111) & X^{4} \\ 
 {}+ {}& (\ldots 01011101{\color{purple}01}) & X^{3\phantom{0}}
   + {}& (\ldots 11011101{\color{purple}010}) & X^{2\phantom{0}} 
   + {}& (\ldots 000000001{\color{purple}100000}) & X^{\phantom{01}}
   + {}& (\ldots 0{\color{purple}0001111110})
\end{array}$$

\caption{Evaluation and re-interpolation of a polynomial of degree $19$
over $\Z_2$}
\label{fig:evalinterpol}
\end{figure}
%
Figure~\ref{fig:evalinterpol} shows an example with a polynomial
(picked at random) of degree $d = 19$. We see that almost all digits
are incorrect, many coefficients have negative valuations, \emph{etc.}
For this problem, floating-point arithmetic is then not well suited.

One may wonder whether another algorithm, specially designed for 
stability, would lead to better results. Unfortunately, the answer is 
negative: we shall see later (in \S \ref{sssec:diffexamples}) that 
the problem of polynomial evaluation and interpolation is \emph{very} 
ill-conditioned in the $p$-adic setting, so that numerical methods are 
ineffective.

\section{The art of tracking $p$-adic precision}
\label{sec:precision}

In many examples presented in \S \ref{sssec:compexamples}, we have often 
recognized similar behaviors: interval arithmetic often overestimates 
the losses of precision while floating-point arithmetic provides more 
accurate --- but unproved --- results. Understanding precisely the
origin of these phenomena is a quite stimulating question (that has
been widely studied in the real setting).

Recently Caruso, Roe and Vaccon~\cite{CaRoVa14} proposed a general theory for 
dealing with precision in the $p$-adic setting and this way provided
powerful tools for attacking the aforementioned question. Their rough 
idea was to develop an analogue of interval arithmetic in higher 
dimensions. In other words, instead of attaching a precision $O(p^N)$ to 
each $p$-adic variable, they group variables and attach to the 
collection of all of them a \emph{unique} \emph{global} precision datum 
materialized by an ``ellipsoid'' in some normed $p$-adic vector space. 
The magic of ultrametricity then operates: ellipsoids are rather easy to 
deal with and behave very well with respect to tracking of precision.

In this section, we report on Caruso, Roe and Vaccon's work. 
\S \ref{ssec:foundpadicprec} is dedicated to the foundations of their 
theory; it is mostly of mathematical nature and deals with $p$-adic 
analysis in several variables. It culminates with the statement (and
the proof) of the precision Lemma (Theorem \ref{theo:preclemma}).
Some first applications are discussed in \S \ref{ssec:optimalprec}
where the precision Lemma is used for finding the maximal precision
one can expect in many concrete situations. Finally we propose in
\S \ref{ssec:latticearith} general methods for reaching this optimal 
precision and apply them in concrete cases.

\subsection{Foundations of the theory of $p$-adic precision}
\label{ssec:foundpadicprec}

The aforementioned theory of $p$-adic precision is based on a single 
result of $p$-adic analysis --- the so-called \emph{precision Lemma} --- 
controlling how ellipsoids transform under mappings of class $C^1$. In 
fact the terminology ``ellipsoid'' is not quite appropriate to the 
$p$-adic setting (though very suggestive for the comparison with the 
real setting) because vector spaces over $\Qp$ are not equipped with 
some $L^2$-norm. Mathematicians then prefer using the term ``lattice'' 
because, as we shall see below, the $p$-adic lattices behave like usual 
$\Z$-lattices in $\R$-vector spaces.

This subsection is organized as follows. We first introduce the notion 
of $p$-adic lattice (\S\S 
\ref{sssec:lattices}--\ref{sssec:complattices}) together with the 
necessary material of $p$-adic analysis (\S \ref{sssec:classC1}). After 
this preparation, \S \ref{sssec:preclemma} is devoted to the precision 
Lemma: we state it and prove it. Applications to $p$-adic precision 
will be discussed in the next subsections (\S \ref{ssec:optimalprec}
and \S \ref{ssec:latticearith}).

\subsubsection{Lattices in finite-dimensional $p$-adic vector spaces}
\label{sssec:lattices}

Let $E$ be a finite-dimensional vector space over $\Qp$.
A (ultrametric) \emph{norm} on $E$ is a mapping $\Vert \cdot \Vert_E :
E \to \R^+$ satisfying the usual requirements:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item $\Vert x \Vert_E = 0$ if and only if $x = 0$;
\item $\Vert \lambda x \Vert_E = |\lambda| \cdot \Vert x \Vert_E$;
\item \label{item:triangleineq}
$\Vert x + y \Vert_E \leq \max\big( \Vert x \Vert_E, 
\Vert y \Vert_E \big)$.
\end{enumerate}
Here $x$ and $y$ refer to elements of $E$ while $\lambda$ refers
to a scalar in $\Qp$. We notice that, without further assumption, 
one can prove that equality holds in \eqref{item:triangleineq} as
soon as $\Vert x \Vert_E \neq \Vert y \Vert_E$: all triangles are 
isosceles in all normed $p$-adic vector spaces! Indeed, by symmetry, we 
may assume $\Vert x \Vert_E < \Vert y \Vert_E$. Now we remark that from 
$\Vert x + y \Vert_E < \Vert y \Vert_E$, we would deduce:
$$\Vert y \Vert_E 
  \leq \max\big(\Vert x+y\Vert_E, \Vert {-}x \Vert_E\big) 
  = \max\big(\Vert x+y\Vert_E, \Vert x \Vert_E\big) 
  < \Vert y \Vert_E$$
which is a contradiction. Our assumption was then absurd, meaning that 
$\Vert x + y \Vert_E = \Vert y \Vert_E$.

Given a real number $r$, we let $B_E(r)$ denote the closed ball in $E$ 
of centre $0$ and radius $r$, \emph{i.e.}:
$$B_E(r) = \big\{\, x \in E \,\text{ s.t. } \Vert x \Vert_E \leq r 
\,\big\}.$$
It is worth remarking that $B_E(r)$ is a module over $\Zp$. Indeed,
on the one hand, multiplying by a scalar in $\Zp$ does not increase
the norm (since elements of $\Zp$ have norm at most $1$) and, on the 
other hand, the ultrametric triangular inequality implies that $E$ is 
stable under addition. Balls in $p$-adic vector spaces have then two
faces: one is analytic and one is algebraic.
Being able to switch between these two points of view is often very
powerful.

The very basic (but still very important) example of a normed 
$\Qp$-vector space is $\Qp^d$ itself endowed with the infinite norm
defined by
$\Vert (x_1, x_2, \ldots, x_d) \Vert_\infty 
= \max \big(|x_1|, |x_2|, \ldots, |x_d|\big)$ for $x_1, \ldots, x_d
\in \Qp$.
The unit ball of $\big(\Qp^d, \Vert \cdot \Vert_\infty\big)$ is $\Zp^d$
and, more generally, $B_{\Qp^d}(r) = p^n \Zp^d$ where $n$ is the unique
relative integer defined by $p^{-n} \leq r < p^{-(n{-}1)}$. We notice
that they are indeed $\Zp$-modules.
Other standard norms over $\R^d$ do not have a direct $p$-adic analogue
since they violate the ultrametric triangular inequality.

\medskip

The general notion of lattice is modeled on balls:

\begin{deftn}
\label{def:latticeana}
Let $E$ be a finite-dimensional vector space over $\Qp$.
A \emph{lattice} in $E$ is a subset of $E$ of the form $B_E(1)$ 
for some $p$-adic norm on $E$.
\end{deftn}

\noindent
Remark that if $H$ is a lattice in $E$ and $a$ is a non zero scalar
in $\Qp$, then $aH$ is a lattice in $E$ as well. Indeed if $H$ is
the closed unit ball for some norm $\Vert \cdot \Vert_E$, then $aH$ 
is the closed unit ball for $\frac 1{|a|} \Vert \cdot \Vert_E$.
More generally, if $H \subset E$ is a lattice and $f : E \to E$ is
a bijective linear transformation, then $f(H)$ is also a lattice.
A consequence of Proposition \ref{prop:nakayama} below is that all
lattices take the form $f(H_0)$ where $H_0$ is a fixed lattice and
$f$ vary in $\GL(E)$.

Lattices might be thought of as ``ellipsoids'' centered at $0$. 
They form a class of special neighborhoods of $0$ that we will use 
afterwards to model the precision (see \S \ref{ssec:latticearith}).
From this perspective, they will 
appear as the natural generalization to higher dimension of the notion 
of bounded interval (centered at $0$) of $\Qp$ we have introduced in \S 
\ref{sssec:tree} and already widely used in \S \ref{ssec:zealous} to 
model precision in zealous arithmetic.

Figure~\ref{fig:lattice} shows a possible picture of a lattice drawn in 
the $p$-adic plane $\Qp^2$ (endowed with the infinite norm). 
Nevertheless, we need of course to be very careful with such 
representations because the topology of $\Qp$ has nothing to do with the 
topology of the paper sheet (or the screen). In particular, it is quite 
difficult to reflect ultrametricity.

\begin{figure}
\hfill
\begin{tikzpicture}
\begin{scope}[beige, rounded corners=10pt]
\fill (0.1,0.2) rectangle (3.9,5.8);
\end{scope}

\begin{scope}[xshift=2cm,yshift=3cm,rotate=20]
\draw[black,fill=green!50] (-0.7,-1.5) rectangle (0.7,1.5);
\draw[black,->] (-1,0)--(1,0);
\draw[black,->] (0,-2)--(0,2);
\end{scope}

\begin{scope}[very thick, rounded corners=10pt]
\draw (0.1,0.2) rectangle (3.9,5.8);
\end{scope}
\end{tikzpicture}
\hfill \null

\caption{Picture of a lattice in the ultrametric world}
\label{fig:lattice}
\end{figure}

\medskip

Here is another purely algebraic definition of lattices which justifies
the wording (compare with the case of $\Z$-lattice in $\R$-vector
spaces).

\begin{deftn}
\label{def:latticealg}
Let $E$ be a finite-dimensional vector space over $\Qp$.
A \emph{lattice} in $E$ is a $\Zp$-module generated by a basis of $E$ 
over $\Qp$.
\end{deftn}

\noindent
The fact that a lattice in the sense of Definition \ref{def:latticealg}
is a lattice in the sense of Definition \ref{def:latticeana} is rather
easy: if $(e_1, \ldots, e_d)$ is a basis of $E$ over $\Qp$, the 
$\Zp$-span of the $e_i$'s is the closed unit ball for the norm 
$\Vert \cdot \Vert_E$ defined by:
$$\Vert \lambda_1 e_i + \cdots + \lambda_d e_d \Vert_E =
\max\big(|\lambda_1|, \ldots, |\lambda_d|\big).$$
As for the converse, it follows from the next proposition.

\begin{prop}
\label{prop:nakayama}
Let $E$ be a $d$-dimensional normed vector space over $\Qp$.
There exists a basis $(e_1, \ldots, e_d)$ of $E$ over $\Qp$ such 
that $B_E(1)$ is the $\Zp$-module generated by the $e_i$'s.
\end{prop}

\begin{proof}
Set $L = B_E(1)$. We have already seen that $L$ is a module over $\Zp$; 
thus the quotient $L/pL$ makes sense and is a vector space over $\Fp = 
\Z/p\Z$. Consider $(\bar e_i)_{i \in I}$ a basis of $L/pL$. (We shall 
see later that $L/pL$ is finite dimensional of dimension $d$ over $\Fp$ 
but, for now, we do not know this and we do not assume anything on the 
set $I$ indexing the basis.) For all $i \in I$, consider $e_i \in L$
which reduces to $\bar e_i$ modulo $p$.

We first claim that the family $(e_i)_{i \in I}$ (where $e_i$ is
considered as an element
of $E$) is free over $\Qp$. Indeed consider a relation of the form
$\sum_{i \in I} \lambda_i e_i = 0$ with $\lambda_i \in \Qp$ and
$\lambda_i = 0$ for almost all $i$. Assume by contradiction that
this relation is non trivial. Then $v = \min_i \, \val_p(\lambda_i)$ is
finite. Up to multiplying the $\lambda_i$'s by $p^{-v}$, we may assume 
that $\lambda_i \in \Zp$ for all $i \in I$ and that there exists
at least one index $i$ for which $\lambda_i$ does not reduce to $0$
modulo $p$.
Now reducing our dependency relation modulo $p$, we get $\sum_{i \in I} 
\bar \lambda_i \bar e_i = 0$ where $\bar \lambda_i \in \Fp$ is the class 
of $\lambda_i$ modulo $p$. Since the family $(\bar e_i)_{i \in I}$ is free 
over $\Fp$ by construction, we derive $\bar \lambda_i = 0$ which
contradicts our assumption.

It follows from the freedom of the $e_i$'s that $I$ is finite of 
cardinality at most $d$. Let us then write $I = \{1, 2, \ldots, d'\}$ 
with $d' \leq d$. We now prove that $(e_1, \ldots, e_{d'})$ generates $L$ 
over $\Zp$. Let then $x \in L$. Using that the $\bar e_i$'s generate 
$L/pL$, we find $p$-adic integers $\lambda_{i,1}$ ($0 \leq i \leq d'$)
such that:
\begin{align}
x & \equiv \lambda_{1,1} e_1 + 
\lambda_{2,1} e_2 + \cdots + \lambda_{d',1} e_{d'} \pmod{pL} \nonumber\\
\text{\emph{i.e.}} \quad 
x & = \lambda_{1,1} e_1 +
\lambda_{2,1} e_2 + \cdots + \lambda_{d',1} e_{d'} + p x_1
\label{eq:xmodp}
\end{align}
for some $x_1 \in L$. Applying the same reasoning to $x_1$ and 
re-injecting in Eq.~\eqref{eq:xmodp}, we end up with an equality of
the form $x = \lambda_{1,2} e_1 +
\lambda_{2,2} e_2 + \cdots + \lambda_{d',2} e_{d'} + p x_2$
with $x_2 \in L$ and $\lambda_{i,2} \equiv \lambda_{i,1} \pmod p$
for all $i$. Continuing this process we construct $d'$ sequences 
$(\lambda_{i,n})_{n \geq 0}$ with the property that
$$x \equiv \lambda_{1,n} e_1 + 
\lambda_{2,n} e_2 + \cdots + \lambda_{d',n} e_{d'} \pmod{p^nL}$$
and $\lambda_{i,n+1} \equiv \lambda_{i,n} \pmod {p^n}$ for all $n$
and $i$. The latter congruence shows that, for all $i$, the sequence 
$(\lambda_{i,n})_{n \geq 0}$ is Cauchy and then converges
to some $\lambda_i \in \Zp$. Passing to the limit, we find that these 
$\lambda_i$'s furthermore satisfy $x = \lambda_1 e_1 + \cdots + 
\lambda_{d'} e_{d'}$.

It now remains to prove that the $e_i$'s generate $E$ over $\Qp$.
For this, remark that any vector $x \in E$ can be written as $x =
p^v y$ with $\Vert y \Vert_E \leq 1$, \emph{i.e.} $y \in B_E(1)$.
By what we have done before, we know that $y$ can be written as a
linear combination of the $e_i$'s. Hence the same holds for $x$.
\end{proof}

\begin{rem}
Proposition~\ref{prop:nakayama} is a particular case of Nakayama's
Lemma (which is a classical result of commutative
algebra). We refer to~\cite{Ei95} for a general introduction to 
commutative algebra including an exposition of Nakayama's
Lemma in a much more general context.
\end{rem}

Proposition~\ref{prop:nakayama} has other important consequences. It 
shows for instance that the inclusion of a lattice $H$ in the ambient 
space $E$ is homeomorphic to the inclusion of $\Zp^d$ in $\Qp^d$ 
(where $d$ is the dimension of $E$). In particular lattices are all
open and compact at the same time.

Three other consequences are enumerated in the next corollary.

\begin{cor}
\label{cor:complete}
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item
All finite dimension normed vector spaces over $\Qp$ are complete.
\item
All norms over a given finite dimension vector space over $\Qp$
are equivalent.
\item
All linear applications between finite dimensional vector spaces over 
$\Qp$ are continuous.
\end{enumerate}
\end{cor}

\begin{proof}
Let $E$ be a finite dimension normal vector space over $\Qp$. Use
Proposition \ref{prop:nakayama} to 
pick a basis $e_1, \ldots, e_d$ of $E$ whose $\Zp$-span
is $B_E(1)$. We claim that an element $x \in E$ lies in the ball
$B_E(p^{-n})$ if and only if all its coordinates on the basis $(e_1,
\ldots, e_d)$ are divisible by $p^n$. Indeed the latter assertion
implies easily the former by factoring out $p^n$. Conversely, let
$x \in B_E(p^{-n})$. Then $p^{-n} x$ has norm at most $1$ and thus 
can be written as
$p^{-n} x = \lambda_1 e_1 + \cdots + \lambda_d e_d$
with $\lambda_i \in \Zp$ for all $i$. Multiplying by $p^n$ on both 
side and identifying coefficients, we get the claim.

Let $(x_n)_{n \geq 0}$ be a Cauchy sequence with values in $E$. For
all $n$, write $x_n = \lambda_{n,1} e_1 + \cdots + \lambda_{n,d} e_d$
with $\lambda_{n,i} \in \Qp$. It follows from the result we have 
proved in the previous paragraph that the sequences $(\lambda_{n,i})
_{n \geq 0}$ are Cauchy for all $i$. They thus converge and hence so 
does $(x_n)_{n \geq 0}$. This proves (i).

The two other assertions are easy (after Proposition 
\ref{prop:nakayama}) and left to the reader.
\end{proof}

\subsubsection{Computation with lattices}
\label{sssec:complattices}

The algebraic side of lattices provides the tools for representing and 
manipulating $p$-adic lattices on computers (at least if the underlying 
vector space $E$ is reasonable) in a quite similar fashion as usual
integral lattices are represented and manipulated \emph{via} integral
matrices.

For simplicity, let us expose the theory in the case where $E = \Qp^d$ 
(endowed with the infinite norm). We then represent a lattice $H \subset 
E$ by the matrix whose row vectors form a 
basis of $L$ (in the sense of Definition \ref{def:latticealg}). 
Equivalently, we can take the matrix, in the canonical basis, of the 
linear transformation $f$ mentioned just above Definition~\ref{def:latticeana}.

For example, if $d = 4$ and $H$ is generated by the vectors:
\begin{align*}
e_1 & = 
\big(\ldots0101110000,\, \ldots0011100000,\, \ldots1011001000,\, \ldots0011000100\big) \\
e_2 & = 
\big(\ldots1101011001,\, \ldots1101000111,\, \ldots0111001010,\, \ldots0101110101\big) \\
e_3 & = 
\big(\ldots0111100011,\, \ldots1011100101,\, \ldots0010100110,\, \ldots1111110111\big) \\
e_4 & = 
\big(\ldots0000111101,\, \ldots1101110011,\, \ldots0011010010,\, \ldots1001100001\big)
\end{align*}
we build the matrix
\begin{equation}
\label{eq:examplelattice}
\begin{array}{r@{\hspace{0.5ex}}l}
M & = 
\small \left(\begin{array}{rrrr}
\ldots0101110000 & \ldots0011100000 & \ldots1011001000 & \ldots0011000100 \\
\ldots1101011001 & \ldots1101000111 & \ldots0111001010 & \ldots0101110101 \\
\ldots0111100011 & \ldots1011100101 & \ldots0010100110 & \ldots1111110111 \\
\ldots0000111101 & \ldots1101110011 & \ldots0011010010 & \ldots1001100001
\end{array}\right)
\end{array}
\end{equation}

\begin{rem}
By convention, our vectors will always be row vectors.
\end{rem}

\noindent
The matrix $M$ we obtain this way is rather nice but we can further 
simplify it using Hermite reduction~\cite[\S 2.4]{Co93}. 
In the $p$-adic setting, Hermite reduction takes the following form.

\begin{theo}[$p$-adic Hermite normal form]
\label{theo:hermitered}
Any matrix $M \in \GL_d(\Qp)$ can be uniquely written as a product
$M = U A$ where $U \in \GL_d(\Zp)$ and $A$ has the shape:
$$A = \left(\begin{matrix}
p^{n_1} & a_{1,2} & \cdots & \cdots & a_{1,d} \\
0 & p^{n_2} & \ddots & & \vdots \\
\vdots & \ddots & \ddots & \ddots & \vdots \\
\vdots & & \ddots & p^{n_{d-1}} & a_{d-1,d} \\
0 & \cdots & \cdots & 0 & p^{n_d} 
\end{matrix} \right)$$
where the $n_i$'s are relative integers and the $a_{i,j}$'s are 
rational numbers of the form $a_{i,j} = \frac {b_{i,j}}{p^{v_{i,j}}}$
with $0 \leq b_{i,j} < p^{n_j + v_{i,j}}$.

The matrix $A$ is called the \emph{Hermite normal form} of $M$.
\end{theo}

\begin{rem}
The left multiplication by the matrix $U$ corresponds to operations on
the rows of $M$, that are operations on the vectors $e_i$.
The invertibility of $U$ over $\Zp$ ensures that the row vectors of
$A$ continue to generate the lattice $H$ we have started with.
\end{rem}

The proof of Theorem~\ref{theo:hermitered} is constructive and can be 
done by row-echelonizing the matrix $M$. Instead of writing it down for 
a general $M$, let us just show how it works on the example 
\eqref{eq:examplelattice}. We first select in the first column an entry 
with minimal valuation, we then move it to the top left corner by swapping rows 
and we normalize it so that it becomes a power of $p$ by rescaling the 
first row by the appropriate invertible element of $\Zp$.
In our example, one can for instance choose the second entry of the 
first column. After swap and renormalization, we get:
$$\small \left(\begin{array}{rrrr}
               1 & \ldots1110011111 & \ldots0011011010 & \ldots1101111101 \\
\ldots0101110000 & \ldots0011100000 & \ldots1011001000 & \ldots0011000100 \\
\ldots0111100011 & \ldots1011100101 & \ldots0010100110 & \ldots1111110111 \\
\ldots0000111101 & \ldots1101110011 & \ldots0011010010 & \ldots1001100001
\end{array}\right)$$
We now use the top left entry as pivot to clear the other entries of
the first column, obtaining this way the new matrix:
$$\small \left(\begin{array}{rrrr}
1 & \ldots1110011111 & \ldots0011011010 & \ldots1101111101 \\
0 & \ldots0001010000 & \ldots0101101000 & \ldots0100010100 \\
0 & \ldots0111101000 & \ldots0101011000 & \ldots1100100000 \\
0 & \ldots1010010000 & \ldots0011100000 & \ldots0011011000
\end{array}\right)$$
Remark that these row operations do not affect the $\Zp$-span
of the row vectors (\emph{i.e.} they correspond to a transformation
matrix $U$ which lies in $\GL_d(\Zp)$).
We continue this process with the $3 \times 3$ matrix obtained by erasing the first row
and the first column. Since we have forgotten the first row, the smallest
valuation of an entry of the second column is now $3$ and the corresponding
entry is located on the third row. We then swap the second and the third
rows, rescale the (new) second row in order to put $2^3$ on the diagonal
and use this value $2^3$ as pivot to cancel the remaining entries on the
second column. After these operations, we find:
$$\small \left(\begin{array}{rrrr}
1 & \ldots1110011111 & \ldots0011011010 & \ldots1101111101 \\
0 & 2^3 & \ldots0000111000 & \ldots0110100000 \\
0 &   0 & \ldots1100111000 & \ldots0011010100 \\
0 &   0 & \ldots1011110000 & \ldots0001011000
\end{array}\right)$$
Iterating again one time this process, we arrive at:
$$\small \left(\begin{array}{rrrr}
1 & \ldots1110011111 & \ldots0011011010 & \ldots1101111101 \\
0 & 2^3 & \ldots0000111000 & \ldots0110100000 \\
0 &   0 & 2^3 & \ldots000001100 \\
0 &   0 &   0 & \ldots111110000
\end{array}\right)$$
Interestingly, observe that the precision on the last two entries of the
last column has decreased by one digit. This is due to the fact that the
pivot $2^3$ was not the element with the smallest valuation on its
\emph{row}. This loss of precision may cause troubles only when the 
initial matrix $M$ is known at precision $O(p^N)$ with $N \leq \max_i\, 
n_i$; in practice such a situation very rarely happen and will never
appear in this course.

The next step of Hermite reduction is the normalization of the bottom
right entry:
$$\small \left(\begin{array}{rrrr}
1 & \ldots1110011111 & \ldots0011011010 & \ldots1101111101 \\
0 & 2^3 & \ldots0000111000 & \ldots0110100000 \\
0 &   0 & 2^3 & \ldots000001100 \\
0 &   0 &   0 & 2^4
\end{array}\right)$$
It remains now to clean up the upper triangular part of the matrix. For this, we
proceed again column by column by using the pivots on the diagonal. Of
course there is nothing to do for the first column. We now reduce the 
$(1,2)$ entry modulo $2^3$ which is the pivot located on the same column.
In order to do so, we add to the first row the appropriate multiple 
of the second row. We obtain this way the new matrix:
$$\small \left(\begin{array}{rrrr}
1 &   7 & \ldots1110110010 & \ldots0010011101 \\
0 & 2^3 & \ldots0000111000 & \ldots0110100000 \\
0 &   0 & 2^3 & \ldots000001100 \\
0 &   0 &   0 & 2^4
\end{array}\right)$$
We emphasize that the $(1,2)$ entry of the matrix is now \emph{exact}!
Repeating this procedure several times we end up finally with
$$\small \left(\begin{array}{rrrr}
1 &   7 &   2 &   5 \\
0 & 2^3 &   0 &  12 \\
0 &   0 & 2^3 &  12 \\
0 &   0 &   0 & 2^4
\end{array}\right)$$
which is the expected Hermite normal form.

\begin{rem}
The algorithmic problem of computing the Hermite normal form has been 
widely studied over the integers and we now know much more efficient 
algorithms (taking advantage of fast matrix multiplication) to solve it 
\cite{KaVi05}; these algorithms extend without trouble to the case 
of $\Zp$ (with the same complexity). Other algorithms specially 
designed for the $p$-adic setting are also available~\cite{Ca12}.
\end{rem}

Here is a remarkable corollary of Theorem~\ref{theo:hermitered}
(compare with Proposition \ref{prop:intervalcomputer}):

\begin{cor}
Lattices in $\Qp^d$ are representable on computers by exact data.
\end{cor}

Operations on lattices can be performed on Hermite forms without 
difficulty. The sum of two lattices, for example, is computed by 
concatenating the two corresponding Hermite matrices and re-echelonizing. 
In a similar fashion, one can compute the image of a lattice under a 
surjective\footnote{Surjectivity ensures that the image remains a 
lattice in the codomain.} linear mapping $f$: we apply $f$ to each 
generator and echelonize the (possibly rectangular) matrix obtained this 
way.

\subsubsection{A small excursion into $p$-adic analysis}
\label{sssec:classC1}

Another important ingredient we will need is the notion of 
differentiable functions in the $p$-adic world. All the material 
presented in this section is classical. We refer to~\cite[\S I.4]{Sc11} 
and~\cite{Co10} for a much more detailed and systematic exposition.

\paragraph{The case of univariate functions}

The notion of differentiability at a given point comes with no surprise: 
given an open subset $U \subset \Qp$, $x \in U$ and a function $f : U 
\to \Qp$, we say that $f$ is differentiable at $x$ if $\frac{f(x) - 
f(y)} {x-y}$ has a limit when $y$ tends to $x$. 
When $f$ is differentiable at $x$, we define:
$$f'(x) = \lim_{y \to x} \, \frac{f(x) - f(y)} {x-y}.$$
Alternatively one can define the derivative using Taylor expansion: one
says that $f$ is differentiable at $x$ is there exists $f'(x) \in \Qp$
for which:
\begin{equation}
\label{eq:defdiff}
\big|f(y) - f(x) - (y{-}x) f'(x)\big| = o\big(|y{-}x|\big)
\quad \text{for } y \to x.
\end{equation}
Usual formulas for differentiating sums, products, composed functions,
\emph{etc.} extend \emph{verbatim} to the $p$-adic case.

The fact that $\Qp$ is (highly) disconnected has however unpleasant 
consequences. For instance, the vanishing of $f'$ on an interval does
not imply the constancy of $f$. Indeed, consider as an example the
indicator function of $\Zp$: it is a locally constant function defined 
on $\Qp$. It is then differentiable with zero derivative but it
is not constant on $\Qp$.
One can cook up even worse examples. Look for instance at the 
function $f : \Zp \to \Zp$ defined by:
$$a_0 + a_1 p + \cdots + a_n p^n + \cdots \quad \mapsto \quad
a_0 + a_1 p^2 + \cdots + a_n p^{2n} + \cdots$$
where the $a_i$'s are the digits, \emph{i.e.} they lie in the range
$[0,p{-}1]$. A straightforward computation shows that $|f(x) - f(y)| 
= |x-y|^2$, \emph{i.e.} $f$ is $2$-Hölder. In particular, $f$ is
differentiable everywhere and $f'$ vanishes on $\Zp$. Yet $f$ is 
apparently injective and \emph{a fortiori} not constant on any
interval.

\medskip

The notion of $p$-adic function of class $C^1$ is more subtle. Indeed 
the notion of differentiable function with continuous derivative is of 
course well defined but not that interesting in the $p$-adic setting. 
Precisely, this definition is too weak and encompasses many 
``irregular'' functions. A more flexible definition is the
following.

\begin{deftn}
\label{def:classC1}
Let $U$ be an open subset of $\Qp$ and $f : U \to \Qp$ be a function.
We say that $f$ is of class $C^1$ on $U$ if there exists a function
$f' : U \to \Qp$ satisfying the following property: for all $a \in U$ 
and all $\varepsilon > 0$, there exists a neighborhood $U_{a,\varepsilon}
\subset U$ of $a$ on which:
\begin{equation}
\label{eq:defC1uni}
\big|f(y) - f(x) - (y{-}x) f'(a)\big| \leq  \varepsilon \cdot 
|y{-}x|.
\end{equation}
\end{deftn}

Given a function $f : U \to \Qp$ of class $C^1$ on $U$, it is clear
that $f$ is differentiable on $U$ and that the function $f'$ appearing in
Definition \ref{def:classC1} has to be the derivative of $f$. Moreover
$f'$ is necessarily continuous on $U$. Indeed consider $a \in U$ and
$\varepsilon > 0$. Let $b \in U_{a,\varepsilon}$. 
The intersection $U = U_{a,\varepsilon} \cap U_{b,\varepsilon}$ is 
open and not empty. It then contains at least two different points,
say $x$ and $y$. It follows from the definition that:
\begin{align*}
\big|f(y) - f(x) - (y{-}x) f'(a)\big| 
 & \leq \varepsilon \cdot |y-x| \smallskip \\ 
\text{and} \quad
\big|f(y) - f(x) - (y{-}x) f'(b)\big| 
 & \leq \varepsilon \cdot |y-x|. 
\end{align*}
Combining these inequalities, we derive
$\big| f'(b) - f'(a) \big| \leq \varepsilon$
which proves the continuity of $f'$ at $x$.
In the real setting, the continuity of $f'$ implies conversely the inequality 
\eqref{eq:defC1uni}\footnote{Indeed, by the mean value theorem we can write 
$f(y) - f(x) = (y-x)f'(g(x,y))$ for some $g(x,y) \in [x,y]$. Thus $|f(y) 
- f(x) - (y-x) f'(a)| = |y-x| \cdot |f'(g(x,y)) - f'(a)|$ and the 
conclusion follows from Heine--Cantor Theorem.} but this implication
fails in the $p$-adic world. A counter-example is given by the function
$f : \Zp \to \Qp$ defined by:
\begin{align*}
x & 
= a_v p^v + a_{v+1} p^{v+1} + \cdots + a_{2v} p^{2v} + a_{2v+1} p^{2v+1} + \cdots \\
\mapsto \quad f(x) & 
= a_v p^{2v} + \big(a_{2v} p^{2v} + a_{2v+1} p^{2v+2} + a_{2v+2} p^{2v+4} + \cdots \big)
\end{align*}
where the $a_i$'s are integers between $0$ and $p{-}1$ with $a_v \neq 0$.
One checks that $|f(x)| \leq |x|^2$ and $|f(x) - f(y)| = 
\frac{|x-y|^2}{|x|^2}$ if $|x-y| \leq |x|^2 < 1$. These inequalities 
ensure that $f$ is differentiable on $\Zp$ and that its derivative 
vanishes everywhere (and thus is continuous). On the other hand when 
$|x-y| = |x|^2 < 1$, we have $|f(x) - f(y)| = |x - y|$, showing that $f$ 
cannot be of class $C^1$.

\begin{rem}
\label{rem:altdefC1uni}
Alternatively, following~\cite{Co10}, one may define the notion of class 
$C^1$ for a function $f : U \to F$ by requiring the existence of a 
covering $(U_i)_{i \in I}$ of $U$ and of \emph{continuous} real-valued 
functions $\varepsilon_i : \R^+ \to \R^+$ with 
$\varepsilon_i(0) = 0$ such that:
\begin{equation}
\label{eq:defC1uni2}
\forall i \in I, \quad \forall x, y \in U_i, \quad 
\big|f(y) - f(x) - (y{-}x) f'(x)\big| \leq  |y{-}x| \cdot
\varepsilon_i\big(|y{-}x|\big)
\end{equation}
(compare with \eqref{eq:defC1uni}). In brief, a function $f$ is of class 
$C^1$ when the estimation \eqref{eq:defdiff} is \emph{locally uniform} 
on $x$. When the domain $U$ is compact (\emph{e.g.} $U = \Zp$), one 
can remove the word ``locally'', \emph{i.e.} one can forget about the 
covering and just take $I = \{\star\}$ and $U_\star = U$.
\end{rem}

\begin{proof}[Proof of the equivalence between the two definitions]
Assume first that $f$ satisfies the definition of Remark
\ref{rem:altdefC1uni}. Let $a \in U$ and $\varepsilon > 0$. We have
$a \in U_i$ for some $i$. Let $\delta$ be a positive real number such 
that $\varepsilon_i < \varepsilon$ on the interval $[0, \delta)$.
Define $U_{a, \varepsilon}$ as the intersection of $U_i$ with the
open ball of centre $a$ and radius $\delta$. On $U_{a,\varepsilon}$,
the estimation:
\begin{equation}
\label{eq:C1equiv}
\big|f(y) - f(x) - (y{-}x) f'(x)\big| \leq  \varepsilon \cdot 
|y{-}x|
\end{equation}
holds. It remains then to relate $f'(x)$ to $f'(a)$. In order to
do so, we write:
\begin{align*}
\big|f(x) - f(a) - (x{-}a) f'(a)\big| & \leq \varepsilon\cdot |x{-}a| \\
\big|f(a) - f(x) - (a{-}x) f'(x)\big| & \leq \varepsilon\cdot |a{-}x|.
\end{align*}
Combining these inequalities, we obtain $|f'(x) - f'(a)| \leq 
\varepsilon$. Re-injecting this new input in \eqref{eq:C1equiv}, 
we finally get \eqref{eq:defC1uni} as desired.

Conversely, assume that $f$ is of class $C^1$ in the sense of Definition
\ref{def:classC1}. Since the problem is local on the domain, we may
assume that $U$ is an open ball. Up to translating and rescaling $f$,
one may further suppose without loss of generality that $U = \Zp$. In
particular, note that $U$ is compact. We define the function 
$\varepsilon$ on $(0,\infty)$ by:
$$\varepsilon(\delta) = \sup_{\substack{x,y \in \Zp \\ 0 < |y{-}x| \leq
\delta}} \frac{|f(y)-f(x)-f'(x)(y{-}x)|}{|y{-}x|}.$$
Compacity ensures that the supremum is finite, so that $\varepsilon$
is well defined. We have to show that $\varepsilon$ goes to $0$ when 
$\delta$ goes to $0$. Let $\varepsilon' > 0$. By assumption, for all
$a \in \Zp$, there exists an open neighborhood $U_{a,\varepsilon'}$ of 
$a$ on which $\frac{|f(y)-f(x)-f'(x)(y{-}x)|}{|y{-}x|} \leq \varepsilon'$.
Up to shrinking $U_{a,\varepsilon'}$, one may assume that
$U_{a,\varepsilon'} = a + p^{n_a} \Zp$ for some positive integer $n_a$.
Now observe that the family of all $U_{a,\varepsilon'}$ when $a$ varies 
is a covering of $\Zp$; by compacity, one can extract from it a finite 
subcovering $(U_{a_i,\varepsilon'})_{1 \leq i \leq m}$ that continues 
to cover $\Zp$. If $n$ denotes the supremum of the $n_{a_i}$'s ($1 
\leq i \leq m$), we thus derive $\varepsilon(\delta) \leq \varepsilon'$
for $\delta \leq p^{-n}$ and we are done.
\end{proof}

\paragraph{The case of multivariate functions}

Let $E$ and $F$ be two finite dimensional normed vector spaces over
$\Qp$. We denote by $\calL(E,F)$ the space of $\Qp$-linear mappings 
from $E$ to $F$. The definition of differentiability and ``of class
$C^1$'' is mimicked from the univariate case.

\begin{deftn}
Let $U$ be an open subset of $E$.

A function $f : U \to F$ is \emph{differentiable} at the point 
$x \in U$ if there exists a linear mapping $df_x \in \calL(E,F)$
such that:
$$\Vert f(y) - f(x) - df_x(y{-}x)\Vert_F = o\big(\Vert y{-}x\Vert_E\big)
\quad \text{when} \quad y \to x.$$

A function $f : U \to F$ is \emph{of class $C^1$} on $U$ if there exists 
a function $df : U \to \calL(E,F)$, $x \mapsto df_x$ satisfying the 
following property: for all $v \in U$ and all $\varepsilon > 0$, there 
exists a neighborhood $U_{a,\varepsilon} \subset U$ of $a$ on which:
\begin{equation}
\label{eq:defC1}
\Vert f(y) - f(x) - df_a(y{-}x)\Vert_F \leq  \varepsilon \cdot 
\Vert y{-}x\Vert_E.
\end{equation}
\end{deftn}

Of course, if $f$ is of class $C^1$ on $U$, it is differentiable at 
every point $x \in U$. When $E = \Qp^n$ and $F = \Qp^m$ (or more
generally when $E$ and $F$ are equipped with distinguished bases)
the matrix of the linear mapping $df_x$ is
the Jacobian matrix $J(f)_x$ defined by:
$$J(f)_x = 
\left(\frac{\partial f_j}{\partial x_i}(x)\right)_
{1 \leq i \leq n, \, 1 \leq j \leq m}$$
where $x_1, \ldots, x_n$ are the coordinates of $E$ and $f_1, \ldots,
f_m$ are the components of $f$.

\begin{rem}
\label{rem:altdefC1multi}
Similarly to the univariate case (see Remark \ref{rem:altdefC1uni}), a function 
$f : U \to F$ is of class $C^1$ if and only if there exist a covering 
$(U_i)_{i \in I}$ of $U$ and some \emph{continuous} real-valued
functions $\varepsilon_i : \R^+ \to \R^+$ with $\varepsilon_i(0) = 0$ 
such that:
\begin{equation}
\label{eq:defC1multi2}
\forall i \in I, \quad \forall x, y \in U_i, \quad
\Vert f(y) - f(x) - (y{-}x) f'(x)\Vert_F \leq  \Vert y{-}x\Vert_E \cdot
\varepsilon_i\big(\Vert y{-}x\Vert_E \big).
\end{equation}
and one can just take $I = \{\star\}$ and $U_\star = U$ when $U$ is compact.
\end{rem}

\subsubsection{The precision Lemma}
\label{sssec:preclemma}

The precision Lemma is a result of $p$-adic analysis controlling how 
lattices transform under sufficiently regular mappings. Before stating 
it, we need a definition.

\begin{deftn}
\label{def:roundedsmall}
Let $E$ be a normed vector space over $\Qp$. 
Given $\rho \in (0,1]$ and $r > 0$, we say that a lattice $H \subset E$ 
is $\rho$-rounded, $r$-small if
$B_E(\rho r) \subset H \subset B_E(r)$.
\end{deftn}

When $E = \Qp^d$ (endowed with the infinite norm), one can determine 
$\rho$ and $r$ satisfying the conditions of Definition 
\ref{def:roundedsmall} by looking at any matrix $M$ representing the 
lattice $H$ (see \S \ref{sssec:complattices}). Indeed, if $n$ is an 
integer for which the matrix $p^n M$ has all its entries in $\Zp$, the 
lattice $p^n H$ is included in $\Zp^n = B_E(1)$, meaning that $H \subset 
B_E(p^n)$. Similarly $B_E(p^{-m}) \subset H$ whenever $m$ is an integer 
such that $p^m M^{-1} \in M_d(\Zp)$.
Therefore if $n$ is the smallest valuation of an entry of $M$ and $m$
is the smallest valuation of an entry of $M^{-1}$, the corresponding
lattice $H$ is $(p^{n+m})$-bounded, $p^n$-small.

\begin{theo}[Precision Lemma]
\label{theo:preclemma}
Let $E$ and $F$ be two finite-dimensional $p$-adic normed vector spaces
and let $f : U \rightarrow F$ be a function of class $C^1$ defined on an 
open subset $U$ of $E$.

Let $v \in U$ be such that $df_v$ is surjective.
Then, for all $\rho \in (0, 1]$, there exists a positive real number 
$\delta$ such that, for any $r \in (0,\delta)$, any lattice 
$\rho$-bounded $r$-small lattice $H$ satisfies:
\begin{equation}
\label{eq:preclemma}
f(v+H) = f(v) + df_v(H).
\end{equation}
\end{theo}

Unfortunately the precision Lemma is a bit technical; understanding its 
precise content is then probably not easy at first glance. In order to 
help the reader, let us say that the most important part of the 
precision Lemma is the conclusion, namely Eq.~\eqref{eq:preclemma}. This 
equation explains how $f$ transforms a ``shifted'' lattice (\emph{i.e.} 
a lattice translated by some vectors) and teaches us that $f$ transforms 
a shifted lattice into another shifted lattice! From people coming from 
the real world, this result should appear as really amazing: in the real 
case, the image of an ellipsoid under a function $f$ is in general 
definitely \emph{not} another ellipsoid (unless $f$ is affine). In the 
$p$-adic case, this happens for any function $f$ of class $C^1$ with 
\emph{surjective differential} (this assumption is important) and almost 
any lattice $H$ (the assumptions on $H$ are actually rather weak though 
technical). This is the magic of ultrametricity.

Below, we give the proof of the precision Lemma and discuss several 
extensions. We advise the reader who is more interested by applications 
(than by computations with $\varepsilon$) to skip the end of the 
subsection and go directly to \S \ref{ssec:optimalprec}, page 
\pageref{ssec:optimalprec}.

\begin{proof}[Proof of Theorem~\ref{theo:preclemma}]
Without loss of generality, we may assume $v=0$ and $f(0)=0$. 
From the surjectivity of $df_0$, we derive that there exists a positive 
constant $C$ such that $B_F(1) \subset df_0(B_E(C))$.
Pick $\varepsilon < \frac\rho C$, and choose a positive real number 
$\delta$ for which
\begin{equation}
\label{eq:C1preclemma}
\Vert f(b) - f(a) - df_0(b{-}a) \Vert_F \leq 
\varepsilon \cdot \Vert b{-}a\Vert_E
\end{equation}
on the ball $B_E(\delta)$.
Let $r \in (0, \delta)$. We suppose that $H$ is a lattice with
$B_E(\rho r) \subset H \subset B_E(r).$
We seek to show that $f$ maps $H$ surjectively onto $df_0 (H)$. We first 
prove that $f(H) \subset df_0 (H)$. Suppose $x \in H$. Applying
Eq.~\eqref{eq:C1preclemma} with $a = 0$ and $b = x$, we get
$\Vert f(x)-df_0(x) \Vert_F \leq \varepsilon \Vert x \Vert_E$.
Setting $y=f(x)-df_0(x)$, we have $\Vert y \Vert_F \leq
\varepsilon r$. The definition of $C$ implies that $B_F(\varepsilon r)
\subset df_0 (B_E(\rho r))$. Thus there exists $x' \in B_E(\rho r)$ such 
that $df_0 (x') =y$. Then $f(x)= df_0 (x-x') \in df_0 (H)$.

We now prove surjectivity. Let $y \in df_0 (H)$. Let $x_0 \in H$
be such that $y = df_0 (x_0)$. We inductively define two sequences
$(x_n)$ and $(z_n)$ by the following cross requirements:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item $z_n$ is an element of $E$ satisfying $df_0(z_n) = y -
f(x_n)$ and $\Vert z_n \Vert_E \leq C \cdot \Vert y - f(x_n) \Vert_F$, and
\item $x_{n+1}=x_n+z_n$.
\end{itemize}
For convenience, let us also define $x_{-1} = 0$ and $z_{-1}=x_0.$ We claim that the
sequences $(x_n)$ and $(z_n)$ are well defined and take their values in
$H$. We do so by induction, assuming that $x_{n-1}$ and $x_n$ belong to $H$
and showing that $z_n$ and $x_{n+1}$ do as well. Noticing that
\begin{equation}
\label{eq:mainlemma}
\begin{aligned}
y - f(x_n) &= f(x_{n-1}) + df_0(z_{n-1}) - f(x_n) \\
&= f(x_{n-1}) - f(x_n) - df_0(x_{n-1} - x_n)
\end{aligned}
\end{equation}
we deduce using differentiability that
$\Vert y - f(x_n) \Vert_F \leq \varepsilon \cdot \Vert x_n - x_{n-1}
\Vert_E$.
Since we are assuming that $x_{n-1}$ and $x_n$ lie in $H \subset
B_E(r)$, we find $\Vert y - f(x_n) \Vert \leq \varepsilon r$. Thus
$\Vert z_n \Vert \leq C \cdot \varepsilon r < \rho r$ and then
$z_n \in H$. From the relation $x_{n+1} = x_n + z_n$, we finally
deduce $x_{n+1} \in H$.

Using \eqref{eq:mainlemma} and differentiability at $0$ once more,
we get
$$\Vert y  - f(x_n) \Vert
\leq \varepsilon \cdot \Vert z_{n-1} \Vert \leq \varepsilon C \cdot \Vert
y - f(x_{n-1}) \Vert,$$
for all $n > 0$.  Therefore, $\Vert y - f(x_n) \Vert = O(a^n)$ and
$\Vert z_n \Vert = O(a^n)$ for $a = \varepsilon C < \rho \leq 1$.
These conditions show that $(x_n)_{n \geq 0}$ is a Cauchy sequence, which converges 
since $E$ is complete (see Corollary \ref{cor:complete}). Write $x$ for the 
limit of the $x_n$; we have $x \in H$ because $H$ is closed (see again
Corollary \ref{cor:complete}). Moreover, 
$f$ is continuous on $H \subseteq U_\varepsilon$ since it is 
differentiable, and thus $y=f(x)$.
\end{proof}

For the applications we have in mind, Theorem \ref{theo:preclemma} is 
actually a bit too weak because it is not effective: the value of 
$\delta$ is not explicit whereas we will often need it in concrete 
applications. In the most general case, it seems difficult to say much 
more about $\delta$. Nevertheless, there are many cases of interest 
where the dependence on $\delta$ can be made explicit.
The simplest such case is that of multivariate polynomials and is
covered by the next proposition.

\begin{prop}
\label{prop:preclemmapoly}
Let $f : \Qp^n \to \Qp^m$ be a function whose coordinates are all 
multivariate polynomials with coefficients in $\Zp$. Let $v \in \Zp^n$ 
such that $df_v$ is surjective. Then Eq.~\eqref{eq:preclemma} holds as 
soon as $H \subset B_E(r)$ and $B_F(p r^2) \subset df_v(H)$ for some 
positive real number $r$.
\end{prop}

\begin{rem}
The case where $f$ is a multivariate polynomial with coefficients in
$\Qp$ reduces to the above proposition by multiplying $f$ by an
appropriate constant. Similarly, if the point $v$ does not lie in
$\Zp^n$ but in $\Qp^n$, we can shift and rescale the polynomial $f$ 
in order to place ourselves within the scope of application of
Proposition \ref{prop:preclemmapoly}.
\end{rem}

The proof of Proposition \ref{prop:preclemmapoly} is based on the
following Lemma.

\begin{lem}
\label{lem:multipolyC1}
Let $g : \Qp^n \to \Qp^m$ be a function whose coordinates are all
multivariate polynomials with coefficients in $\Zp$. Then
$$\Vert g(b) - g(a) - dg_0(b{-}a) \Vert \leq \max 
  \big(\Vert a \Vert, \Vert b \Vert \big) \cdot \Vert b{-}a \Vert$$
for all $a,b \in \Zp^n$.
\end{lem}

\begin{proof}
We may assume $m = 1$. By linearity, we may further assume that
$g(x_1, \ldots, x_n) = x_1^{\alpha_1} \cdots x_n^{\alpha_n}$. If the
sum of the $\alpha_i$'s is at most $1$, the quantity $g(b) - g(a) - 
dg_0(b{-}a)$ vanishes and the lemma is clear. Otherwise the differential
$dg_0$ vanishes and we write:
$$\begin{array}{l}
g(b_1, \ldots, b_n) - g(a_1, \ldots, a_n) \smallskip \\
\hspace{2em}
=  (b_1^{\alpha_1} - a_1^{\alpha_1}) b_2^{\alpha_2} \cdots b_n^{\alpha_n} 
 + a_1^{\alpha_1} (b_2^{\alpha_2} - a_2^{\alpha_2}) b_3^{\alpha_3} \cdots b_n^{\alpha_n} 
 + \cdots 
 + a_1^{\alpha_1} \cdots a_{n-1}^{\alpha_{n-1}} (a_n^{\alpha_n} - b_n^{\alpha_n}).
\end{array}$$
Noting that $b_i^{\alpha_i} - a_i^{\alpha_i} = 
(b_i - a_i) \sum_{k=0}^{\alpha_i-1} b_i^k a_i^{\alpha_i-1-k}$ and
remembering that $\alpha_1 + \cdots + \alpha_n \geq 2$, we get 
the announced result by applying the ultrametric triangular
inequality.
\end{proof}

\begin{proof}[Proof of Proposition \ref{prop:preclemmapoly}]
Let $H$ be a lattice satisfying the assumptions of the proposition. 
Let $n$ be the largest relative integer for which $p^n \leq r$. 
Then $B_E(r) = B_E(p^n)$ and $B_F(p^{2n+1}) \subset B_F(p r^2)$, so 
that $H \subset B_E(p^n)$ and $B_F(p^{2n+1}) \subset df_v(H)$.

Set $H' = p^n H$; clearly $H'$ is a lattice contained in $B_E(1)$. Let 
$\varphi : \Qp^n \to \Qp^n$ be a linear mapping taking bijectively 
$B_E(1)$ to $H'$. Consider the composite $g = f \circ (v + \varphi)$. 
From $H' \subset B_E(1)$, we deduce that the entries of the matrix of 
$\varphi$ (in the canonical basis) lie in $\Zp$. Since moreover $v \in
\Zp^n$, we deduce that all the coordinates of $g$ are given by
multivariate polynomials with coefficients in $\Zp$. Furthermore, we
have:
\begin{align*}
f(v+H) & = f(v + p^{-n} H') = f(v + \varphi(B_E(p^n)) = g(B_E(p^n)) \\
\text{and}\quad
f(v) + df_v(H) & = f(v) + df_v \circ \varphi(B_E(p^n)) = g(0) + dg_0(B_E(p^n)).
\end{align*}
We then have to prove that $g(B_E(p^n)) = g(0) + dg_0(B_E(p^n))$
knowing that $g$ is a multivariate polynomial function with coefficients
in $\Zp$ and $B_F(p^{n+1}) \subset dg_0(B_E(1))$. This can be done by
following the lines of the proof of Theorem \ref{theo:preclemma} and
refining Eq.~\eqref{eq:C1preclemma} by the estimation of Lemma
\ref{lem:multipolyC1}.
\end{proof}

Other results concerning more general $f$ and building on the machinery 
of Newton polygons are available in the literature. We refer the reader 
to~\cite{CaRoVa14} for the case of locally analytic functions and to 
\cite{CaRoVa15} for the case of solutions of partial differential equations of 
order $1$.

\subsection{Optimal precision and stability of algorithms}
\label{ssec:optimalprec}

The precision Lemma (Theorem \ref{theo:preclemma}) provides the 
mathematical tools for finding the intrinsic optimal loss/gain of 
precision of a given problem and then for studying the stability 
of numerical $p$-adic algorithms.
% add a sentence here?

\subsubsection{The precision Lemma at work}
\label{sssec:preclemmaatwork}

Consider an ``abstract $p$-adic problem'' encoded by a mathematical 
function $\varphi : U \to F$ where $U$ is an open subset in a finite 
dimensional $p$-adic vector space $E$ and $F$ is another finite
dimensional $p$-adic vector space.
The modelling is straightforward: the domain $U$ is the space of 
inputs, the codomain $F$ is the space of outputs and $\varphi$ is 
the mapping sending an input to the expected output.
Many concrete problems fit into this framework: for instance $\varphi$ 
can be the function mapping a pair of $p$-adic numbers to their sum
or their product, but it can also be the function mapping a polynomial
(of a given degree) to its derivative, a matrix (of given dimensions) 
to its inverse, a family of multivariate polynomials (of given size and 
degrees) to its reduced Gröbner basis, \emph{etc.}

In order to apply the precision Lemma, we shall assume that $\varphi$ is 
of class $C^1$. This hypothesis is not restrictive at all in practice 
since in many common cases, the function $\varphi$ has much more 
regularity than that: it is often obtained by composing sums, products, 
divisions and if-statements so that it is locally given by 
multivariate rational fractions.
We suppose furthermore that $E$ and $F$ are endowed with distinguished 
bases $(e_1, \ldots, e_n)$ and $(f_1, \ldots, f_m)$ respectively. This
choice models the way the elements of $E$ and $F$ are represented on the 
computer. For instance if $E = \Qp[X]_{< n}$ (the vector space of
polynomials of degree less than $n$) and polynomials are represented
internally by the list of their coefficients, we will just choose the
canonical basis. On the contrary, if polynomials are represented as
linear combinations of the shape:
$$\lambda_0 + \lambda_1 X + \lambda_2 X(X{-}1) + \cdots +
\lambda_{n-1} X (X{-}1) \cdots (X{-}n{+}2)$$
then the distinguished basis we will choose is $(1, X, X(X{-}1), \ldots,
X (X{-}1) \cdots (X{-}n{+}2))$.

\paragraph{Context of zealous arithmetic}

Recall that, in the zealous point of view, $p$-adic numbers are 
modeled by intervals of the form $a + O(p^N)$. As a consequence the 
input of the problem we are studying will not be an actual element of 
$U$ but a quantity of the form:
$$\big(a_1 + O\big(p^{N_1}\big) \big) \cdot e_1 + 
\big(a_2 + O\big(p^{N_2}\big) \big) \cdot e_2 + \cdots +
\big(a_n + O\big(p^{N_n}\big) \big) \cdot e_n$$
which can be rewritten as the shifted lattice $v + H$ with:
\begin{align*}
v & = a_1 e_1 + a_2 e_2 + \cdots + a_n e_n \\
H & = \Span_{\Zp} \big( p^{N_1} e_1, \, p^{N_2} e_2, \cdots, \,
p^{N_n} e_n \big).
\end{align*}
Similarly the output takes the form:
$$\big(b_1 + O\big(p^{M_1}\big) \big) \cdot f_1 + 
\big(b_2 + O\big(p^{M_2}\big) \big) \cdot f_2 + \cdots +
\big(b_m + O\big(p^{M_m}\big) \big) \cdot f_m$$
where the $b_j$'s are the coordinates of $\varphi(v)$ and then do not 
depend (fortunately) on the algorithm we are using to evaluate 
$\varphi$. On the other hand, the $M_j$'s may --- and do --- depend on
it. The question we address here can be formulated as follows: what is 
the maximal value one can expect for the $M_j$'s? In other words: what 
is the optimal precision one can expect on the output (in terms of the 
initial precision we had on the inputs)?

The precision Lemma provides a satisfying answer to this question.
Indeed, for $j \in \{1, \ldots, m\}$, let $\pr_{F,j} : F \to \Qp$ be
the linear mapping taking a vector to its $j$-th coordinate (in the
distinguished basis $(f_1, \ldots, f_m)$ we have fixed once for all). 
From the precision Lemma applied to the composite $\varphi_j = \pr_{F,j} 
\circ \varphi$, we get:
\begin{equation}
\label{eq:preclemmazealous}
\varphi_j(v+H) =
\varphi_j(v) + d\varphi_{j,v}(H) = b_j + d\varphi_{j,v}(H) 
\end{equation}
as soon as $d\varphi_{j,v}$ is surjective (and $H$ satisfies some
additional mild assumptions that we will ignore for now). 
Under these assumptions, Eq.~\eqref{eq:preclemmazealous} holds and 
the equality signs in it show that the optimal precision $O(p^{M_j})$ we 
were looking for is nothing but $d\varphi_{j,v}(H)$; note that the 
latter is a lattice in $\Qp$ and then necessarily takes the form 
$p^{M_j}\Zp = O(p^{M_j})$ for some relative integer $M_j$.
Note moreover that the surjectivity of $d\varphi_{j,v}$ is equivalent
to its non-vanishing (since $d\varphi_{j,v}$ takes its value in a
one-dimensional vector space). 
Furthermore, by the chain rule, we have $d\varphi_{j,v} = \pr_{F,j} 
\circ d\varphi_v$ because $\pr_{F,j}$ is linear and then agrees with
its differential at any point.

Finding the optimal precision $O(p^{M_j})$ can then be done along the
following lines: we first compute the Jacobian matrix $J(\varphi)_v$
of $\varphi$ at the point $v$ and form the product:
\begin{equation}
\label{eq:jacobianzealous}
A = \left( \begin{matrix}
p^{N_1} & & \\ & \ddots & \\ & & p^{N_n}
\end{matrix}\right) \cdot J(\varphi)_v
\end{equation}
whose rows generate $d\varphi_{j,v}(H)$. The integer $M_j$ appears
then as the smallest valuation of an entry of the $j$-th column of
$A$ (unless this column vanishes in which case the precision Lemma
cannot be applied). Of course, the computation of the Jacobian matrix 
can be boring and/or time-consuming; we shall see however in
\S \ref{sssec:diffexamples} below that, in many situations, simple
mathematical arguments provide closed formulas for $J(\varphi)_v$,
avoiding this way their direct computation.

\subparagraph{The notion of diffused digits of precision.}

One important remark is that the lattice:
$$H' = \Span_{\Zp} \big( p^{M_1} f_1, \, p^{M_2} f_2, \cdots, \,
p^{M_m} e_m \big) \subset F$$
(for the $M_j$'s we have constructed) does satisfy $f(v+H) \subset f(v) 
+ H'$ but does \emph{not} satisfy $f(v+H) = f(v) + H'$ in general.
In other words, although we cannot expect more correct digits on each 
component of the output separately, the precision on the output is 
globally not optimal. Below is a very simple example illustrating this 
point.

\begin{ex*}
Assume that we want to evaluate an affine polynomial at the two points 
$0$ and $p$. The function $\varphi$ modeling this problem is:
$$\varphi : \Qp[X]_{\leq 1} \to \Qp^2, 
\quad P(X) = a X + b \mapsto (P(0), P(p)) = (b,\,ap{+}b).$$
It is a linear function so its Jacobian is easy to compute. We find
$J(\varphi) = \Big( {\small \begin{matrix} 0 & p \\ 1 & 1
\end{matrix}} \Big)$.
We assume furthermore that the inputs $a$ and $b$ are both given at
precision $O(p^N)$. We then find that the optimal precision on $P(0)$ and 
$P(p)$ is $O(p^N)$ as well. Nevertheless the determinant of $J(\varphi)$
is apparently $p$; hence $J(\varphi)$ is not invertible in $M_2(\Zp)$ 
and its row vectors generate a lattice which is strictly smaller than 
$\Zp^2$. 
What does happen concretely? The main observation is that the difference 
$P(p) - P(0) = ap$ can be computed at precision $O(p^{N+1})$ because
of the multiplication by $p$. There is thus one more digit that we know
but this digit is ``diffused'' between $P(0)$ and $P(p)$ and only appears
when we look at the difference.
\end{ex*}

The phenomenon enlighten in the above example is quite general: \emph{we 
cannot in general reach the optimal precision by just giving individual 
precision on each coordinate of the output (we say that some digits of 
precision are diffused over several coordinates)... but this becomes 
always\footnote{At least when the precision Lemma applies...} possible 
after a suitable base change (which recombines the diffused digits of 
precision into actual digits).} In order to make this sentence more
precise, we 
first formulate a rigorous definition concerning the diffusion of the 
digits of precision.

\begin{deftn}
\label{def:diffuseddigits}
Let $F$ be a finite dimensional vector space over $\Qp$ with basis
$(f_1, \ldots, f_m)$.

\smallskip

\noindent
We say that a lattice $H$ is \emph{diagonal} with respect to 
$(f_1, \ldots, f_m)$ if it has the shape
$H = \Span_{\Zp} \big(p^{\nu_1} f_1, \ldots, p^{\nu_m} f_m\big)$ for
some relative integers $\nu_1, \ldots, \nu_m$.

\smallskip

\noindent
The \emph{number of diffused digits} (with respect to $(f_1, \ldots,
f_m)$) of a given lattice $H$ is the 
logarithm in base $p$ of the index of $H$ in the smallest diagonal
lattice containing $H$.
\end{deftn}

Roughly speaking, the number of diffused digits of a lattice $H$ with 
respect to the basis $(f_1, \ldots, f_m)$ quantifies the quality of the 
basis for expressing the precision encoded by $H$: if $H$ has 
no diffused digit, it is represented by a diagonal matrix and the 
precision splits properly into each coordinate. On the contrary, if 
$H$ has diffused digits, writing the precision in the system of 
coordinates associated with $(f_1, \ldots, f_m)$ leads to non optimal 
results.
We remark furthermore that there always exists a basis in which the 
number of diffused digits of a given lattice $H$ is zero: it suffices to 
take a basis whose $\Zp$-span is $H$ (such a basis always exists by 
Proposition \ref{prop:nakayama}).

Before going further, let us emphasize that, given a set of generators 
of $H$, it is not difficult to compute effectively its number of 
diffused digits of precision. Indeed, let $M$ be a matrix whose row
vectors span $H$.
Pick in addition a diagonal lattice $H' = \Span_{\Zp} \big(p^{\nu_1} f_1, \ldots, 
p^{\nu_m} f_m\big)$ with $H \subset H'$. By projecting on the $j$-th 
coordinate we see that $\nu_j$ is at least the smallest valuation $v_j$ of 
an entry of the $j$-th column of $M$. Conversely, one easily checks that 
the diagonal lattice $\Span_{\Zp} \big(p^{v_1} f_1, \ldots, p^{v_m} 
f_m\big)$ contains $H$. Thus the smallest diagonal matrix containing $H$ 
is $\Span_{\Zp} \big(p^{v_1} f_1, \ldots, p^{v_m} f_m\big)$. It follows 
that the number of diffused digits of precision of $H$ with respect to 
the basis $(f_1, \ldots, f_m)$ is given by:
\begin{equation}
\label{eq:numberdiffused}
(v_1 + v_2 + \cdots + v_m) - \val_p(\det M).
\end{equation}
If $M$ is in Hermite normal form (see \S \ref{sssec:complattices}) this 
quantity vanishes if and only if $M$ is indeed diagonal.

\medskip

We now go back to our initial question: we want to analyze the optimal 
precision for the computation of $\varphi(v)$ for $v$ is given at 
precision $H = \Span_{\Zp} \big( p^{N_1} e_1, \, p^{N_2} e_2, \cdots, \, 
p^{N_n} e_n \big)$. The precision Lemma applied with $\varphi$ itself
(assuming that the hypotheses of the precision Lemma are fulfilled
of course) yields the equality:
$$\varphi(v+H) = \varphi(v) + d\varphi_v(H).$$
The number of diffused digits of precision resulting from the computation 
of $\varphi(v)$ is then the number of diffused digits of the lattice
$d\varphi_v(H)$. If we are working in the basis $(f_1, \ldots, f_n)$,
this number can be large, meaning that writing down the precision in 
this basis can be weak. However, if we are allowing a change of basis on 
the codomain, we can always reduce the number of diffused digits to $0$; 
in such a basis, the precision can be made optimal!

\paragraph{Context of lazy/relaxed arithmetic}

As we have explained in \S \ref{sssec:zealousVSrelaxed}, the point of 
view of lazy/relaxed arithmetic differs from the zealous one on the 
following point: instead of fixing the precision on the input and 
propagating it to the outputs, it fixes a target precision and infers 
from it the precision needed on the inputs.
Of course, in order to save time and memory, we would like to avoid 
the computation of unnecessary digits. Regarding precision, the
question then becomes: what is the minimal number of digits we have
to compute on the inputs in order to ensure a correct output at the
requested precision?

This question translates into our precision language as follows: we fix 
some relative integers $M_1, \ldots, M_m$ and look for relative
integers $N_1, \ldots, N_n$ for which:
\begin{equation}
\label{eq:lazypreclemma}
\varphi(v + H) \subset \varphi(v) + H'
\end{equation}
with $H = \Span_{\Zp} \big(p^{N_1} f_1, \ldots, p^{N_n} e_n\big)$
(the unknown) and 
$H' = \Span_{\Zp} \big(p^{M_1} f_1, \ldots, p^{M_m} f_m\big)$.
Assuming that the precision Lemma applies for $\varphi$ and $H$, 
Eq.~\eqref{eq:lazypreclemma} is equivalent to $d\varphi_v(H) \subset H'$,
\emph{i.e.} $H \subset d\varphi_v^{-1}(H')$. The problem then reduces
to find the largest diagonal lattice sitting in $d\varphi_v^{-1}(H')$.

The comparison with the zealous situation is instructive: in the zealous 
case, one had the lattice $d \varphi_v(H_{\zealous})$ and we were 
looking for the smallest diagonal lattice containing it. The problem is
now ``reversed'' and can actually be related to the zealous problem
using duality. Before going further, we need to recall basic facts about
duality and lattices.

\medskip

Given $V$ a finite dimensional $\Qp$-vector space, we define its 
\emph{dual} $V^\star$ as the space of $\Qp$-linear forms on $V$. We
recall that there is a canonical isomorphism between $V$ and its
bidual $V^{\star\star}$; it maps an element $x \in V$ to the linear
form taking $\ell \in V^\star$ to $\ell(x)$. 
We recall also that each basis $(v_1, \ldots, v_d)$ of $V$ has a
dual basis $(v_1^\star, \ldots, v_d^\star)$ defined by $v_j^\star(v_i)
= 1$ if $i = j$ and $v_j^\star(v_i) = 0$ otherwise.

If $L \subset V$ is any $\Zp$-module, its dual $L^\star$ is defined by 
as the subset of $V^\star$ consisting of linear forms $\ell : V \to \Qp$ 
for which $\ell(L) \subset \Zp$. Clearly $L^\star$ is a sub-$\Zp$-module 
of $V^\star$. One can check that the construction $L \mapsto L^\star$ is 
involutive (\emph{i.e.} $L^{\star\star} = L$), order-reversing 
(\emph{i.e.} if $L_1 \subset L_2$ then $L_2^\star \subset L_1^\star$) 
and preserves diagonal lattices (\emph{i.e.} if $L$ is a diagonal 
lattice with respect to some basis then $L^\star$ is a diagonal lattice 
with respect to the dual basis).

\begin{rem}
\label{rem:duallattice}
If $L$ is a lattice, so is $L^\star$. More precisely, if
$(v_1, \ldots, v_d)$ is a basis of $V$ and $M$ is a square matrix whose 
rows span $L$ (with respect to the above basis), it is easily checked 
that $L^\star$ is spanned by the rows of ${}^{\text{t}}\! M^{-1}$ (with 
respect to the dual basis). On the contrary if $L$ does not generate 
$V$ as a $\Qp$-vector space, the dual space $L^\star$ contains a
$\Qp$-line; conversely if $L$ contains a $\Qp$-line then $L^\star$
does not generate $V^\star$ over $\Qp$.
\end{rem}

Applying the above formalism to our situation, we find that the 
condition $H \subset d\varphi_v^{-1}(H')$ we want to ensure is 
equivalent to $d\varphi_v^{-1}(H')^\star \subset H^\star$. Our problem 
then reduces to finding the smallest diagonal lattice containing 
$d\varphi_v^{-1}(H')^\star$. By the analysis we have done in the
zealous context, the 
integer $N_i$ we are looking for then appears as the smallest valuation 
of an entry of $i$-th column of a matrix whose rows span the space 
$d\varphi_v^{-1}(H')^\star$ (with respect to the dual basis $(e_1^\star, 
\ldots, e_n^\star)$). Computing $d\varphi_v^{-1}(H')^\star$ using
Remark \ref{rem:duallattice} (and taking care of the possible kernel of 
$d\varphi_v$), we finally find that $N_i$ is the opposite of the 
smallest valuation of an entry of the $i$-th \emph{row} of the matrix:
$$J(\varphi)_v \cdot
\left( \begin{matrix}
p^{-M_1} & & \\ & \ddots & \\ & & p^{-M_m}
\end{matrix}\right)$$
where $J(\varphi)_v$ is the Jacobian matrix of $\varphi$ at $v$ (compare 
with \eqref{eq:jacobianzealous}).

\subsubsection{Everyday life examples revisited}
\label{sssec:diffexamples}

We revisit the examples of \S \ref{sssec:compexamples} in light of the 
theory of $p$-adic precision.

\paragraph{The $p$-power map}

As a training example, let us first examine the computation of the 
$p$-power map (though technically it was not treated in \S 
\ref{sssec:compexamples} but at the end of \S 
\ref{sssec:intervalVSfloat}). Recall that we have seen that raising an 
invertible $p$-adic integer $x$ at the power $p$ gains one digit of 
precision: if $x$ is given at precision $O(p^N)$, one can compute $x^p$ 
at precision $O(p^{N+1})$.

We introduce the function $\varphi : \Qp \to \Qp$ taking $x$ to $x^p$. 
It is apparently of class $C^1$ and its differential at $x$ is the 
multiplication by $\varphi'(x) = p x^{p-1}$. The Jacobian matrix of
$\varphi$ at $x$ is then the $1 \times 1$ matrix whose unique entry
is $p x^{p-1}$. By the results of \S \ref{sssec:preclemmaatwork} 
(see particularly Eq.~\eqref{eq:preclemmazealous}), the optimal 
precision on $x^p$ is the valuation of 
$$(p x^{p-1}) \times p^N = p^{N+1} x^{p-1}.$$
We recover this way the exponent $N{+}1$ thanks to the factor $p$ 
which appears in the Jacobian. Unlike Lemma \ref{lem:ppower},
the method we have just applied here teaches us in addition that
the exponent $N{+}1$ cannot be improved.

\paragraph{Determinant}

We take here $\varphi = \det : M_d(\Qp) \to \Qp$. We endow $M_d(\Qp)$ 
with its canonical basis and let $x_{i,j}$ denote the corresponding 
coordinate functions: for $A \in M_d(\Zp)$, $x_{i,j}(A)$ is the $(i,j)$ 
entry of $A$.
Fix temporarily a pair $(i,j)$. Expanding the determinant according 
to the $i$-th row, we find that $\varphi$ is an affine function of
$x_{i,j}$ whose linear term is $(-1)^{i+j} x_{i,j} \cdot \det G_{i,j}$
where $G_{i,j}$ is the matrix obtained from the ``generic'' matrix
$G = (x_{i,j})_{1 \leq i,j \leq d}$ by erasing the $i$-th row and the 
$j$-th column. Therefore
$\frac{\partial \varphi}{\partial x_{i,j}} = (-1)^{i+j} \det G_{i,j}$.
Evaluating this identity at some matrix $M \in M_d(\Qp)$, we get
the well-known formula:
\begin{equation}
\label{eq:partialdet}
\frac{\partial \varphi}{\partial x_{i,j}}(M) = (-1)^{i+j} \det M_{i,j}
\end{equation}
where $M_{i,j}$ is the matrix obtained from $M$ by erasing the $i$-th
row and the $j$-th column. The Jacobian matrix of $\varphi$ at $M$ is 
then the column matrix (whose rows are indexed by the pairs $(i,j)$
for $1 \leq i,j \leq d$) with entries $(-1)^{i+j} \det M_{i,j}$.
According to the results of \S \ref{sssec:preclemmaatwork} (see 
especially Eq.~\eqref{eq:preclemmazealous}), the optimal precision
for $\det M$, when the $(i,j)$ entry of $M$ is given at precision
$O(p^{N_{i,j}})$, is:
\begin{equation}
\label{eq:optdet}
O\big(p^{N'}\big) \quad \text{for} \quad
N' = \min_{i,j} \big( N_{i,j} + \val(M_{i,j})\big).
\end{equation}
Remember of course that this conclusion is only valid when the
assumptions of the precision Lemma are all fulfilled. Nonetheless,
relying on Proposition \ref{prop:preclemmapoly} (after having noticed that the 
determinant is a multivariate polynomial function), we can further 
write down explicit sufficient conditions for this to hold. These
conditions take a particularly simple form when all the entries of
$M$ are given at the \emph{same} precision $O(p^N)$: in this case,
we find that the precision \eqref{eq:optdet} is indeed the optimal
precision as soon as $N' > \val(\det M)$, \emph{i.e.} as soon as it 
is possible to ensure that $\det M$ does not vanish.

\medskip

The general result above applies in particular to the matrix $M$ 
considered as a running example in \S \ref{sssec:compexamples} (see 
Eq.~\eqref{eq:matrixM}):
\begin{equation}
\label{eq:matrixM2}
\begin{array}{r@{\hspace{0.5ex}}l}
M & = 
\small \left(\begin{array}{rrrr}
\ldots0101110000 & \ldots0011100000 & \ldots1011001000 & \ldots0011000100 \\
\ldots1101011001 & \ldots1101000111 & \ldots0111001010 & \ldots0101110101 \\
\ldots0111100011 & \ldots1011100101 & \ldots0010100110 & \ldots1111110111 \\
\ldots0000111101 & \ldots1101110011 & \ldots0011010010 & \ldots1001100001
\end{array}\right) \in M_4(\Z_2).
\end{array}
\end{equation}
Recall that all the entries of $M$ were given at the same precision
$O(2^{10})$.
A simple computation (using zealous arithmetic) shows that the comatrix 
of $M$ (\emph{i.e.} the matrix whose $(i,j)$ entry is $(-1)^{i+j} \det 
M_{i,j}$) is:
$$\begin{array}{r@{\hspace{0.5ex}}l}
\Com(M) & =
\small \left(\begin{array}{rrrr}
\ldots1111000000 & \ldots0001100000 & \ldots1101000000 & \ldots1001100000 \\
\ldots0111000000 & \ldots0011100000 & \ldots1001000000 & \ldots1011100000 \\
\ldots0111000000 & \ldots0001100000 & \ldots0111000000 & \ldots0011100000 \\
\ldots1010000000 & \ldots0111000000 & \ldots1110000000 & \ldots0011000000
\end{array}\right)
\end{array}$$
We observe that the minimal valuation of its entries is $5$ (it
is reached on the second column), meaning that the optimal precision
on $\det M$ is $O(2^{15})$. Coming back to \S \ref{sssec:compexamples}, 
we see that the interval floating-point arithmetic reached this
accuracy whereas zealous interval arithmetic
missed many digits (without further help).

\subparagraph{Computation of determinants with optimal precision.}

It is worth remarking that one can design a simple algorithm that computes 
the determinant of a $p$-adic matrix at the optimal precision 
\emph{within the framework of zealous arithmetic} when the coefficients 
of the input matrix are all given at the same precision $O(p^N)$. This 
algorithm is based on a variation of the Smith normal form that we state 
now.

\begin{prop}
\label{prop:Smith}
Any matrix $M \in M_d(\Qp)$ admits a factorization of the following
form:
\begin{equation}
\label{eq:Smith}
M = P \cdot 
\left(\begin{matrix} a_1 & & \\ & \ddots & \\ & & a_d \end{matrix}\right)
\cdot Q
\end{equation}
where $P$ and $Q$ are have determinant $\pm 1$, the matrix in the middle 
is diagonal and its diagonal entries satisfy $\val(a_1) \leq \cdots \leq 
\val(a_d)$.
\end{prop}

\noindent
The factorization \eqref{eq:Smith} is quite interesting for us for two 
reasons. First of all, it reveals the smallest valuation of an entry
of the comatrix of $M$: it is 
\begin{equation}
\label{eq:optprecdet}
v = \val(a_1) + \val(a_2) + \cdots + \val(a_{d-1}).
\end{equation}
Indeed, taking exterior powers, one proves the entries of $\Com(M)$ 
and $\Com(P^{-1} M Q^{-1})$ span the same $\Zp$-submodule of $\Qp$.
Moreover, since $P^{-1} M Q^{-1}$ is diagonal, its comatrix is easily
computed; we find:
$$\Com(P^{-1} M Q^{-1}) =
\left(\begin{matrix} b_1 & & \\ & \ddots & \\ & & b_d \end{matrix}\right)
\quad \text{with} \quad
b_i = a_1 \cdots a_{i-1} a_{i+1} \cdots a_d$$
(recall that $\Com(A) = \det A \cdot A^{-1}$ if $A$ is invertible)
and we are done. 
The direct consequence of this calculation is the following: if the 
entries of $M$ are all given at precision $O(p^N)$, the optimal 
precision on $\det M$ is $O(p^{N+v})$ (assuming $v < N$).

The second decisive advantage of the factorization \eqref{eq:Smith} is 
that it can be computed without any loss of precision.
The underlying algorithm basically follows the lines of the Hermite
reduction we have presented in \S \ref{sssec:complattices}, except on
the three following points:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item instead of choosing the pivot as an entry of smallest valuation 
on the working column, we choose as an entry of smallest valuation on
the whole matrix,
\item we do not only clear the entries behind the pivot but also the
entries located on the right of the pivot,
\item in order to keep the transformation matrices of determinant
$\pm 1$, we do not rescale rows.
\end{itemize}
For example, starting with our favorite matrix $M$ (see 
Eq.~\eqref{eq:matrixM2} above), we first select the $(2,1)$ entry
(which has valuation $0$), we put it on the top left corner and use
it as pivot to clear the entries on the first row and the first
column. Doing so, we obtain the intermediate form:
$$\small \left(\begin{array}{rrrr}
\ldots1101011001 & 0 & 0 & 0 \\
0 & \ldots0001010000 & \ldots0101101000 & \ldots0100010100 \\
0 & \ldots0111101000 & \ldots0101011000 & \ldots1100100000 \\
0 & \ldots1010010000 & \ldots0011100000 & \ldots0110011000
\end{array}\right)$$
and we retain that we have swapped two rows, implying that the
determinant of the above matrix is the opposite to the determinant
of $M$. We now select the $(2,4)$ entry (which as valuation $2$) 
and continue the reduction:
$$\small \left(\begin{array}{rrrr}
\ldots1101011001 & 0 & 0 & 0 \\
0 & \ldots0100010100 & 0 & 0 \\
0 & 0 & \ldots0100011000 & \ldots0101101000 \\
0 & 0 & \ldots0000110000 & \ldots0000110000
\end{array}\right)$$
We observe that we have not lost any precision in this step (all 
the entries of the matrix above are known at precision $O(2^{10})$),
and this 
even if our pivot had positive valuation. It is actually a general
phenomenon due to the fact that we always choose the pivot of
smallest valuation. Continuing this process, we end up with the
matrix:
$$\small \left(\begin{array}{rrrr}
\ldots1101011001 & 0 & 0 & 0 \\
0 & \ldots0100010100 & 0 & 0 \\
0 & 0 & \ldots0100011000 & 0 \\
0 & 0 & 0 & \ldots1101100000
\end{array}\right)$$
Now multiplying the diagonal entries, we find
$\det M = 2^{10} \times ...01101$ with precision $O(2^{15})$!

\medskip

For a general $M$, the same precision analysis works. Indeed, when the 
entries of $M$ are all given at precision $O(p^N)$, the reduction 
algorithm we have sketched above outputs a diagonal matrix whose 
determinant is the same as the one of $M$ and whose diagonal entries 
$a_1, \ldots, a_d$ are all known at precision $O(p^N)$. By Proposition 
\ref{prop:arithinterval} (see also Remark \ref{rem:arithinterval}),
zealous arithmetic can compute the product $a_1 a_2 \cdots a_d$ 
at precision $O(p^{N+w})$ with 
$$w = \val(a_1) + \cdots + \val(a_d) - \max_i \val(a_i).$$
We then recover the optimal precision given by Eq.~\eqref{eq:optprecdet}.

\paragraph{Characteristic polynomial}

The case of characteristic polynomials has many similarities with
that of determinants.
We introduce the function $\varphi : M_d(\Qp) \to \Qp[X]_{< d}$
that maps a matrix $M$ to $\varphi(M) = \det(X I_d - M) - X^d$ where
$I_d$ is the identity matrix of size $d$. The renormalization 
(consisting in subtracting $X^d$) is harmless but needed in order to 
ensure that $\varphi$ takes its values in a vector space (and not an
affine space) and has a surjective differential almost everywhere.
The partial derivatives of $\varphi$ are:
$$\frac{\partial \varphi}{\partial x_{i,j}}(M) = (-1)^{i+j} \cdot
\det (X I_d - M)_{i,j}$$
where $(X I_d - M)_{i,j}$ is the matrix obtained from $X I_d {-} M$ 
by deleting the $i$-th row and the $j$-th column (compare with 
Eq.~\eqref{eq:partialdet}). The Jacobian matrix $J(\varphi)_M$ of 
$\varphi$ at $M$
is then a matrix with $d^2$ rows, indexed by pairs $(i,j)$ with
$1 \leq i,j \leq d$, and $d$ columns whose row with index $(i,j)$
contains the coefficients of the polynomial $(-1)^{i+j} \cdot
\det (X I_d - M)_{i,j}$. 
Using again the results of \S\ref{sssec:preclemmaatwork}, the optimal
precision on each coefficient of the characteristic polynomial of
$M$ can be read off from $J(\varphi)_M$.

\medskip

Let us explore further our running example: the matrix $M$ given by 
Eq.~\eqref{eq:matrixM2} whose entries are all known at precision 
$O(2^{10})$. A direct computation leads to the Jacobian matrix 
displayed on the left of Figure~\ref{fig:jaccharpoly}.
%
\begin{figure}
{\scriptsize\hfill
$\begin{array}{cc@{}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{0.2ex}}c}
(i,j) && X^3 & X^2 & X & 1 \\
\rule[3ex]{0pt}{\baselineskip}
(1,1) &
\raisebox{0.2ex}{\multirow{16}{*}{$\left(\parbox[c][39ex]{0cm}{}\right.$}}
 &1 & \ldots0110110010 & \ldots0111111000 & \ldots0001000000 &
\raisebox{0.2ex}{\multirow{16}{*}{$\left.\parbox[c][39ex]{0cm}{}\right)$}} \\
(1,2) &&0 & \ldots0011100000 & \ldots1011010100 & \ldots1110100000 \\
(1,3) &&0 & \ldots1011001000 & \ldots1001001000 & \ldots0011000000 \\
(1,4) &&0 & \ldots0011000100 & \ldots1111100100 & \ldots0110100000 \\
(2,1) &&0 & \ldots1101011001 & \ldots1010010000 & \ldots1001000000 \\
(2,2) &&1 & \ldots1110001001 & \ldots1001001100 & \ldots1100100000 \\
(2,3) &&0 & \ldots0111001010 & \ldots0110011000 & \ldots0111000000 \\
(2,4) &&0 & \ldots0101110101 & \ldots0111111100 & \ldots0100100000 \\
(3,1) &&0 & \ldots0111100011 & \ldots1010000000 & \ldots1001000000 \\
(3,2) &&0 & \ldots1011100101 & \ldots1110100000 & \ldots1110100000 \\
(3,3) &&1 & \ldots0011101000 & \ldots1001000100 & \ldots1001000000 \\
(3,4) &&0 & \ldots1111110111 & \ldots1111100100 & \ldots1100100000 \\
(4,1) &&0 & \ldots0000111101 & \ldots0010111000 & \ldots0110000000 \\
(4,2) &&0 & \ldots1101110011 & \ldots1101011000 & \ldots1001000000 \\
(4,3) &&0 & \ldots0011010010 & \ldots1101001000 & \ldots0010000000 \\
(4,4) &&1 & \ldots1010100011 & \ldots0111010000 & \ldots1101000000
\end{array}$
\quad $\stackrel{\text{\tiny Hermite}}{\longrightarrow}$ \quad
$\begin{array}{c@{}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{0.5ex}}c}
& X^3 & X^2 & X & 1 \\
\rule[3ex]{0pt}{\baselineskip}
\raisebox{0.2ex}{\multirow{16}{*}{$\left(\parbox[c][39ex]{0cm}{}\right.$}}
& 1 & 0 & 0 & 0 &
\raisebox{0.2ex}{\multirow{16}{*}{$\left.\parbox[c][39ex]{0cm}{}\right)$}} \\
&{\color{black!50} 0} & 1 & 0 & 0 \\
&{\color{black!50} 0} & {\color{black!50} 0} & 2^2 & 0 \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & 2^5 \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} 
\end{array}$}
\hfill\null

\caption{The Jacobian of the characteristic polynomial at $M$}
\label{fig:jaccharpoly} 
\end{figure} 
% 
We observe that the minimal valuation of an entry of the column of $X^j$ 
is $0$, $0$, $2$ and $5$ when $j$ is $3$, $2$, $1$ and $0$ respectively. 
Consequently the optimal precision on the coefficients of $X^3$, $X^2$, 
$X$ and on the constant coefficient are $O(2^{10})$, $O(2^{10})$, 
$O(2^{12})$ and $O(2^{15})$ respectively. Coming back to \S 
\ref{sssec:compexamples}, we notice that floating-point arithmetic
found all the correct digits.
We observe moreover that the Hermite normal form of the Jacobian matrix
(shown on the right on Figure~\ref{fig:jaccharpoly}) is diagonal. There
is therefore no diffused digit of precision for this example; in other
words the precision written as
$O(p^{10}) X^3 + O(p^{10}) X^2 + O(p^{12}) X + O(p^{15})$
is sharp.

\begin{rem}
Making more extensive tests, we conclude that $p$-adic floating point
arithemtic was pretty lucky with the above example: in general, it 
indeed sometimes finds all (or almost all) relevant digits but it may
also sometimes not be more accurate than zealous arithmetic. We refer
to \cite{CaRoVa17} for precise statistics thereupon.
\end{rem}

Let us now have a look at the matrix $N = I_4 + M$. Playing the same
game as before, we obtain the Jacobian matrix displayed on the left of
Figure~\ref{fig:jaccharpoly2}.
%
\begin{figure}
\hfill
{\scriptsize
$\begin{array}{cc@{}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{0.2ex}}c}
(i,j) && X^3 & X^2 & X & 1 \\
\rule[3ex]{0pt}{\baselineskip}
(1,1) &
\raisebox{0.2ex}{\multirow{16}{*}{$\left(\parbox[c][39ex]{0cm}{}\right.$}}
 &1 & \ldots0110101111 & \ldots1010010111 & \ldots1111111001 &
\raisebox{0.2ex}{\multirow{16}{*}{$\left.\parbox[c][39ex]{0cm}{}\right)$}} \\
(1,2)&&0 & \ldots0011100000 & \ldots0100010100 & \ldots0110101100 \\
(1,3)&&0 & \ldots1011001000 & \ldots0010111000 & \ldots0101000000 \\
(1,4)&&0 & \ldots0011000100 & \ldots1001011100 & \ldots1010000000 \\
(2,1)&&0 & \ldots1101011001 & \ldots1111011110 & \ldots1100001001 \\
(2,2)&&1 & \ldots1110000110 & \ldots1100111101 & \ldots0001011100 \\
(2,3)&&0 & \ldots0111001010 & \ldots1000000100 & \ldots0111110010 \\
(2,4)&&0 & \ldots0101110101 & \ldots1100010010 & \ldots0010011001 \\
(3,1)&&0 & \ldots0111100011 & \ldots1010111010 & \ldots0110100011 \\
(3,2)&&0 & \ldots1011100101 & \ldots0111010110 & \ldots1011100101 \\
(3,3)&&1 & \ldots0011100101 & \ldots0001110111 & \ldots0011100011 \\
(3,4)&&0 & \ldots1111110111 & \ldots1111110110 & \ldots1100110011 \\
(4,1)&&0 & \ldots0000111101 & \ldots0000111110 & \ldots0100000101 \\
(4,2)&&0 & \ldots1101110011 & \ldots0001110010 & \ldots1001011011 \\
(4,3)&&0 & \ldots0011010010 & \ldots0110100100 & \ldots1000001010 \\
(4,4)&&1 & \ldots1010100000 & \ldots0010001101 & \ldots0000010010
\end{array}$
\quad $\stackrel{\text{\tiny Hermite}}{\longrightarrow}$ \quad
$\begin{array}{c@{}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{1.5em}}r@{\hspace{0.5ex}}c}
& X^3 & X^2 & X & 1 \\
\rule[3ex]{0pt}{\baselineskip}
\raisebox{0.2ex}{\multirow{16}{*}{$\left(\parbox[c][39ex]{0cm}{}\right.$}}
& 1 & 0 & 1 & 30 &
\raisebox{0.2ex}{\multirow{16}{*}{$\left.\parbox[c][39ex]{0cm}{}\right)$}} \\
&{\color{black!50} 0} & 1 & 2 & 29 \\
&{\color{black!50} 0} & {\color{black!50} 0} & 2^2 & 28 \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & 2^5 \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} \\
&{\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} 
\end{array}$}
\hfill\null

\caption{The Jacobian of the characteristic polynomial at $I_4{+}M$}
\label{fig:jaccharpoly2}
\end{figure}
%
Each column of this matrix contains an entry of valuation zero. The 
optimal precision on each individual coefficient of the characteristic 
polynomial of $N$ is then no more than $O(p^{10})$. However the Hermite 
reduced form of the Jacobian is now no longer diagonal, meaning that 
diffused digits of precision do appear. One can moreover count them 
using Eq.~\eqref{eq:numberdiffused} (applied to the Hermite normal
form of the Jacobian): we find $7$. This means that the precision
$$O(p^{10}) X^3 + O(p^{10}) X^2 + O(p^{10}) X + O(p^{10})$$
is not sharp; more precisely, after a suitable base change on $\Qp[X]_{< 
d}$, it should be possible to visualize seven more digits. 
Given that $\chi_N(X) = \chi_M(X{-}1)$, this base change is of course
the one which is induced by the change of variable $X \mapsto X{-}1$.

\paragraph{LU factorization}

Given a positive integer $d$, define $L_d(\Zp)$ as the subspace of 
$M_d(\Zp)$ consisting of lower triangular matrices with zero diagonal.
Similarly let $U_d(\Zp)$ be the subspace space of $M_d(\Zp)$ 
consisting of upper triangular matrices. Clearly $L_d(\Zp)$ and 
$U_d(\Zp)$ are $p$-adic vector spaces of respective dimensions 
$\frac{d(d{-}1)}2$ and $\frac{d(d{+}1)}2$. Let $\calU$ be the open 
subset of $M_d(\Zp)$ defined by the non-vanishing of all principal 
minors. We define two mappings:
$$\varphi : \calU \to L_d(\Zp), \, M \mapsto L{-}I_d 
\quad \text{and} \quad 
\psi : \calU \to L_d(\Zp), \, M \mapsto U$$
where $M = LU$ is the (unique) LU factorization of $M$.
In order to compute the differentials of $\varphi$ and 
$\psi$\footnote{These two functions are indeed differentiable. One can 
prove this by noticing for instance that the entries of $L$ and $U$ are 
given by explicit multivariate rational fractions~\cite[\S 1.4]{Ho75}.}, we
simply differentiate the defining relation $M = LU$. We get this
way the relation $dM = dL \cdot U + L \cdot dU$. We rewrite it as:
$$L^{-1} \cdot dM \cdot U^{-1} = 
\big(L^{-1} \cdot dL\big) + \big(dU \cdot U^{-1}\big)$$
and observe that the first summand of the right hand side is strictly lower 
triangular whereas the second summand of upper triangular. We
therefore derive
$dL = L \cdot \text{Lo}\big(L^{-1} \cdot dM \cdot U^{-1}\big)$ and
$dU = \text{Up}\big(L^{-1} \cdot dM \cdot U^{-1}\big) \cdot U$
where $\text{Lo}$ (resp. $\text{Up}$) is the projection on the first
factor (resp. on the second factor) of the decomposition $M_d(\Zp)
= L_d(\Zp) \oplus U_d(\Zp)$. 
The differentials of $\varphi$ and $\psi$ at $M \in \calU$ are 
then given by the linear mappings:
$$\begin{array}{rl}
d \varphi_M : M_d(\Zp) \to L_d(\Zp), &
dM \mapsto L \cdot \text{Lo}\big(L^{-1} \cdot dM \cdot U^{-1}\big) \smallskip \\
d \psi_M : M_d(\Zp) \to U_d(\Zp), &
dM \mapsto \text{Up}\big(L^{-1} \cdot dM \cdot U^{-1}\big) \cdot U 
\end{array}$$
where $L$ and $U$ are the ``L-part'' and the ``U-part'' of the LU 
factorization of $M$ respectively.

\medskip

As an example, take again the matrix $M$ given by 
Eq.~\eqref{eq:matrixM2}. The Jacobian matrix of $\varphi$ at this point
is the matrix displayed on Figure~\ref{fig:jacLU}.
%
\begin{figure}
\hfill{\tiny
$\begin{array}{cc@{\hspace{1ex}}r@{\hspace{1ex}}r@{\hspace{1ex}}r@{\hspace{2ex}}r@{\hspace{0ex}}r@{\hspace{4ex}}r@{\hspace{2ex}}c}
 && (2,1) & (3,1) & (3,2) & (4,1) & (4,2) & (4,3) \\
\rule[3ex]{0pt}{\baselineskip}
(1,1) &
\raisebox{0.2ex}{\multirow{16}{*}{$\left(\parbox[c][39ex]{0cm}{}\right.$}} &
 \ldots11,\!00110111 & \ldots11,\!01001101 & \ldots1100101\phantom{,\!0} & \ldots00,\!10010011 & \ldots1100010 & \ldots001,\!01\phantom{0} &
\raisebox{0.2ex}{\multirow{16}{*}{$\left.\parbox[c][39ex]{0cm}{}\right)$}} \\
(1,2) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & \ldots110101,\!1 & 0\phantom{,\!00000000} & \ldots011111 & \ldots100,\!01\phantom{0} \\
(1,3) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & \ldots1011,\!1\phantom{00} \\
(1,4) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & 0\phantom{,\!000} \\
(2,1) && \ldots111010,\!0111\phantom{0000} & \ldots000000\phantom{,\!00000000} & \ldots0111001110\phantom{,\!0} & \ldots000\phantom{,\!00000000} & \ldots11100010 & \ldots0110,\!111 \\
(2,2) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & \ldots0100001001\phantom{,\!0} & 0\phantom{,\!00000000} & \ldots0011111 & \ldots1111,\!011 \\
(2,3) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & \ldots10111,\!01\phantom{0} \\
(2,4) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & 0\phantom{,\!000} \\
(3,1) && 0\phantom{,\!00000000} & \ldots111010,\!0111\phantom{0000} & \ldots00110100110\phantom{,\!0} & \ldots000\phantom{,\!00000000} & \ldots00000000 & \ldots1111,\!11\phantom{0} \\
(3,2) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & \ldots0111011101\phantom{,\!0} & 0\phantom{,\!00000000} & \ldots0000000 & \ldots1000,\!11\phantom{0} \\
(3,3) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & \ldots11010,\!1\phantom{00} \\
(3,4) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & 0\phantom{,\!000} \\
(4,1) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & \ldots111010,\!0111\phantom{0000} & \ldots00110100110 & \ldots1010,\!011 \\
(4,2) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & \ldots0111011101 & \ldots0100,\!111 \\
(4,3) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & \ldots00100,\!01\phantom{0} \\
(4,4) && 0\phantom{,\!00000000} & 0\phantom{,\!00000000} & 0\phantom{,\!0} & 0\phantom{,\!00000000} & 0 & 0\phantom{,\!000} \\ \\
&&\raisebox{0.2ex}{
 \multirow{6}{*}{$\stackrel{\text{\tiny Hermite}}{\longrightarrow}\quad\left(\parbox[c][17ex]{0cm}{}\right.$}}\hspace{2ex}
 2^{-8} & 2^{-8} \times 11 & 0 & 2^{-8} \times 5 & 0 & 0 &
\raisebox{0.2ex}{\multirow{6}{*}{$\left.\parbox[c][17ex]{0cm}{}\right)$}} \\
&& {\color{black!50} 0} & 2^{-4} & 0 & 0 & 0 & 0 \\
&& {\color{black!50} 0} & {\color{black!50} 0} & 2^{-1} & 0 & 0 & 2^{-3} \\
&& {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & 2^{-4} & 0 & 2^{-3} \\
&& {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & 1 & 2^{-3} \\
&& {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & {\color{black!50} 0} & 2^{-2} 
\end{array}$}
\hfill\null

\caption{The Jacobian of the L-part of the LU factorization at $M$}
\label{fig:jacLU}
\end{figure}
%
Looking at the minimal valuation of the entries of $J(\varphi)_M$ column 
by column, we find the optimal precision for each entry of the matrix
$L$ (defined as the L-part $L$ of the LU factorization of $M$):
\begin{equation}
\label{eq:optprecLU}
\left(\begin{matrix}
{\color{black!50} -} & {\color{black!50} -} & {\color{black!50} -} & {\color{black!50} -} \\
O(2^2) & {\color{black!50} -} & {\color{black!50} -} & {\color{black!50} -} \\
O(2^2) & O(2^9) & {\color{black!50} -} & {\color{black!50} -} \\
O(2^2) & O(2^{10}) & O(2^7) & {\color{black!50} -} \\
\end{matrix}\right)
\end{equation}
(we recall that the entries of $M$ were all initially given at precision 
$O(2^{10})$). Comparing with the numerical results obtained in \S 
\ref{sssec:compexamples} (see Eq.~\eqref{eq:resultLU}), we see that 
floating-point arithmetic, one more time, found all the relevant 
digits.

A closer look at Figure~\ref{fig:jacLU} shows that the lattice 
$d \varphi_M(2^{10} M_4(\Z_2)) = 
2^{10} \cdot d \varphi_M(M_4(\Z_2))$
has exactly $9$ diffused digits (apply Eq.~\eqref{eq:numberdiffused}).
The precision \eqref{eq:optprecLU} is then globally not optimal although 
it is on each entry separately. For example, if $\ell_{i,j}$ denotes the 
$(i,j)$ entry of $L$, the linear combination $\ell_{3,2} - 11 \cdot 
\ell_{3,1}$ can be known at precision $O(2^6)$ although $\ell_{3,2}$ and
$\ell_{3,1}$ cannot be known at higher precision than $O(2^2)$.
It is remarkable that floating-point arithmetic actually ``saw'' 
these diffused digits; indeed, taking the values computed in
floating-point arithmetic for $\ell_{3,2}$ and $\ell_{3,1}$, we compute:
$$\ell_{3,2} - 11 \cdot \ell_{3,1} = \ldots0000010111.$$
It turns out that the last six digits of the latter value are 
correct!

\begin{rem}
\label{rem:precLU}
For a general input $d \times d$ matrix $M$ whose principal minors have 
valuation $v_1, \ldots, v_d$, we expect that zealous arithmetic looses 
about $\Omega(v_1 + \cdots + v_d)$ significant digits (for the 
computation of the LU factorization of $M$ using standard Gaussian
elimination) while floating-point arithmetics looses only $O(\max(v_1, 
\ldots, v_d))$ significant digits. If $M$ is picked at random in the
space $M_d(\Zp)$ (equipped with its natural Haar measure), the former 
is $O(\frac d{p-1})$ on average while the second is $O(\log_p d)$ on
average~\cite{Ca12}.
\end{rem}

\paragraph{Bézout coefficients}

We now move to commutative algebra. We address the question of the 
computation of the Bézout coefficients of two monic polynomials of 
degree $d$ which are supposed to be coprime.
Let $\mathcal U$ be the subset of $\Qp[X]_{< d} \times \Qp[X]_{< d}$ 
consisting of pairs $(\tilde P, \tilde Q)$ for which $P = X^d + \tilde 
P$ and $Q = X^d + \tilde Q$ are coprime. Observe that $\mathcal U$ is defined by
the non-vanishing of some resultant and so is open. We introduce the
function:
$$\begin{array}{rcl}
\varphi : \quad \calU & \longrightarrow & \Qp[X]_{< d} \times \Qp[X]_{< d} \smallskip \\
(\tilde P, \tilde Q) & \mapsto & (U,V) \quad \text{s.t.} \quad UP + VQ = 1
\end{array}$$
for $P = X^d + \tilde P$ and $Q = X^d + \tilde Q$. (Note that $U$
and $V$ are uniquely determined thanks to the conditions on the degree.)
Once again, the differential of $\varphi$ can be easily computed
by differentiating the defining relation $UP + VQ = 1$; indeed doing so,
we immediately get:
\begin{equation}
\label{eq:diffBezoutrel}
\big(dU \cdot P\big) + \big(dV \cdot Q\big) = dR
\end{equation}
for $dR = - U {\cdot} dP - V {\cdot} dQ$. Eq~\eqref{eq:diffBezoutrel} gives
$dU {\cdot} P \equiv dR \pmod Q$ for what we derive $dU \equiv U {\cdot} dR
\pmod Q$. Comparing degrees we find $dU = U {\cdot} dR \text{ mod } Q$.
Similarly $dV = V {\cdot} dR \text{ mod } P$. 
The differential of $\varphi$ at a pair $(\tilde P, \tilde Q)$ is 
then the linear mapping:
$$\begin{array}{rcl}
d \varphi_{(\tilde P, \tilde Q)} :
\quad \Qp[X]_{< d} \times \Qp[X]_{< d} & \longrightarrow & \Qp[X]_{< d} \times \Qp[X]_{< d} \smallskip \\
(dP, dQ) & \mapsto & (dU, dV) = \big(U{\cdot}dR \text{ mod } Q, \, V{\cdot}dR \text{ mod } P\big) \smallskip \\
& & \hspace{5em}\text{where} \quad dR = - U \cdot dP - V \cdot dQ.
\end{array}$$
It is important to note that $d \varphi_{(\tilde P, \tilde Q)}$ is never
injective because it maps $(-V, U)$ to $0$. Consequently, it is never
surjective either and one cannot apply the precision Lemma to it. The
easiest way to fix this issue is to decompose $\varphi$ as $\varphi = 
(\varphi_U, \varphi_V)$ and to apply the precision Lemma to each 
component separately (though this option leads to less accurate results).

\medskip

For the particular example we have studied in \S 
\ref{sssec:compexamples}, namely:
$$\begin{array}{r@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l@{\hspace{0.2ex}}r@{\hspace{0.5ex}}l}
P = X^4 + {}& (\ldots 1101111111) & X^3 
        + {}& (\ldots 0011110011) & X^2 \\
      {}+ {}& (\ldots 1001001100) & X^{\phantom{1}} 
        + {}& (\ldots 0010111010) \medskip \\
Q = X^4 + {}& (\ldots 0101001011) & X^3 
        + {}& (\ldots 0111001111) & X^2 \\
      {}+ {}& (\ldots 0100010000) & X^{\phantom{1}} 
        + {}& (\ldots 1101000111) 
\end{array}$$
the Jacobian matrix we find is shown on Figure~\ref{fig:jacBezout}.
%
\begin{figure}
\hfill{\tiny
$\begin{array}{r@{\hspace{0.3ex}}lc@{\hspace{1ex}}rrrr@{\hspace{3ex}}|@{\hspace{3ex}}rrrr@{\hspace{2ex}}c}
&&& \multicolumn{4}{c|@{\hspace{3ex}}}{dU} & \multicolumn{4}{c}{dV} \\
\rule[3ex]{0pt}{\baselineskip}
(dP,&dQ) && X^3 & X^2 & X & 1 & X^3 & X^2 & X & 1 \\
\rule[3ex]{0pt}{\baselineskip}
(X^3,&0) &
\raisebox{0.2ex}{\multirow{8}{*}{$\left(\parbox[c][22ex]{0cm}{}\right.$}} &
           \ldots0100111 & \ldots0001000 & \ldots1000000 & \ldots1010000 & \ldots1011001 & \ldots0100000 & \ldots1000000 & \ldots0100000 &
\raisebox{0.2ex}{\multirow{8}{*}{$\left.\parbox[c][22ex]{0cm}{}\right)$}} \\
(X^2,&0) &&\ldots1010000 & \ldots0010111 & \ldots0111000 & \ldots1000000 & \ldots0110000 & \ldots0101001 & \ldots0110000 & \ldots0000000 \\
(X,&0)   &&\ldots1000000 & \ldots0010000 & \ldots1010111 & \ldots0111000 & \ldots1000000 & \ldots1110000 & \ldots1101001 & \ldots0110000 \\
(1,&0)   &&\ldots1111000 & \ldots1101000 & \ldots0011000 & \ldots1010111 & \ldots0001000 & \ldots0111000 & \ldots0001000 & \ldots1001001 \\
\rule[3ex]{0pt}{\baselineskip}
(0,&X^3) &&\ldots1011001 & \ldots1111000 & \ldots0010000 & \ldots1110000 & \ldots0100111 & \ldots1100000 & \ldots1100000 & \ldots1100000 \\
(0,&X^2) &&\ldots1110000 & \ldots0101001 & \ldots0001000 & \ldots0010000 & \ldots0010000 & \ldots0010111 & \ldots0010000 & \ldots0100000 \\
(0,&X)   &&\ldots0010000 & \ldots0100000 & \ldots0011001 & \ldots0001000 & \ldots1110000 & \ldots0100000 & \ldots1100111 & \ldots1010000 \\
(0,&1)   &&\ldots1001000 & \ldots0101000 & \ldots1011000 & \ldots0011001 & \ldots0111000 & \ldots0111000 & \ldots1001000 & \ldots0000111 
\end{array}$}
\hfill\null

\caption{The Jacobian of the ``Bézout function'' at $(P,Q)$}
\label{fig:jacBezout}
\end{figure}
%
Each column of this matrix contains an entry of valuation zero, meaning 
that the optimal precision on each coefficient of $U$ and $V$ is 
$O(2^{10})$. Observe that it was not reached by interval arithmetic or 
floating-point arithmetic!
Continuing our analysis, we further observe that the lattice generated 
by the $dU$-part (resp. the $dV$-part) of $J(\varphi)_{(\tilde P, \tilde 
Q)}$ has no diffused digit. The precision
$$O(2^{10}) X^3 + O(2^{10}) X^2 + O(2^{10}) X + O(2^{10})$$
on both $U$ and $V$ is then optimal (but the joint precision induced 
on the pair $(U,V)$ is not).

Another option for fixing the default of surjectivity of $d 
\varphi_{(P,Q)}$ is to observe that $\varphi$ in fact takes its value in 
the hyperplane $H$ of $\Qp[X]_{< d} \times \Qp[X]_{< d}$ consisting of 
pairs of polynomials $(P,Q)$ whose coefficients in degree $d{-}1$ agree. 
Restricting the codomain of $\varphi$ to $H$, we then obtain a 
well-defined mapping whose differential is almost everywhere surjective. 
Applying the precision Lemma, we find that the pair $(U,V)$ has $16$ 
diffused digits!

\begin{rem}
A careful study of the precision in Euclidean algorithm is presented 
in~\cite{Ca17}: we prove in this reference that if $P$ and $Q$ are two
monic polynomials of degree $d$, Euclidean algorithm executed within the 
framework of zealous arithmetic (resp. floating-point arithmetic) looses 
$\Omega(v_0 + \cdots + v_{d{-}1})$ (resp. $O(\max(v_0, \ldots, 
v_{d{-}1}))$) digits on each coefficient where $v_j$ is the valuation of 
the $j$-th scalar subresultant of $P$ and $Q$. Moreover, for random 
polynomials, we have by \cite[Corollary~3.6]{Ca17}:
\begin{align*}
\E\big[v_0 + \cdots + v_{d{-}1}\big] & 
\textstyle \geq \frac d{p-1} \smallskip \\
\E\big[\max(v_0, \ldots, v_{d{-}1})\big] & 
\textstyle \leq \log_p d + O\big(\sqrt{\log_p d}\big).
\end{align*}
Compare this result with the case of LU factorization
(see Remark~\ref{rem:precLU}).
\end{rem}

\paragraph{Polynomial evaluation and interpolation}

Recall that the last example considered in \S \ref{sssec:compexamples}
was about evaluation and interpolation of polynomials: precisely, 
starting with a polynomial $P \in \Zp[X]$ of degree $d$, we first
evaluated it at the points $0, 1, \ldots, d$ and then reconstructed
it by interpolation from the values $P(0), P(1), \ldots, P(d)$.
We have observed that this problem seemed to be numerically highly 
unstable: for example, for $d = 19$ and an initial polynomial $P$ given 
at precision $O(2^{10})$, we were only able to reconstruct a few
number of digits and even found not integral coefficients in many 
places (see Figure~\ref{fig:evalinterpol}). We propose here to give
a theoretical explanation of this phenomenon.

Consider the function $\varphi : \Qp[X]_{\leq d} \to \Qp^{d+1}$, $P 
\mapsto (P(0), P(1), \ldots, P(d))$. Note that it is linear, so that $d 
\varphi_P = \varphi$ for all $P$. 
The numerical stability of our evaluation problem is governed by
the lattice
$H = \varphi(p^N \Zp[X]_{\leq d}) = p^N \cdot \varphi(\Zp[X]_{\leq d})$
where $O(p^N)$ is the initial precision we have on each coefficient of
$P$ (assuming for simplicity that it is the same for all coefficients).
Applying $\varphi$ to the standard basis of $\Zp[X]_{\leq d}$, we find
that $H$ is generated by the row vectors of the Vandermonde matrix:
$$M = \left( \begin{matrix}
1 & 1 & 1 & \cdots & 1 \\
0 & 1 & 2 & \cdots & d \\
0 & 1 & 2^2 & \cdots & d^2 \\
\vdots & \vdots & \vdots & & \vdots \\
0 & 1 & 2^d & \cdots & d^d
\end{matrix} \right).$$
According to Eq.~\eqref{eq:numberdiffused}, the number of diffused
digits of $H$ is then equal to the $p$-adic valuation of the
determinant of $M$ whose value is:
$$\det M = \prod_{0 \leq i < j \leq d} (j-i) = 1! \times 2! \times 
\cdots \times d!.$$
In order to estimate its $p$-adic valuation, we recall the following
result.

\begin{prop}[Legendre's formula]
For all positive integer $n$, we have:
$$\val_p(n!) = \Big\lfloor\frac n p\Big\rfloor + \Big\lfloor\frac n {p^2}\Big\rfloor + \cdots +
\Big\lfloor\frac n {p^k}\Big\rfloor + \cdots$$
where $\left\lfloor\cdot\right\rfloor$ is the usual floor function.
\end{prop}

\begin{proof}
Exercise.
\end{proof}

It follows in particular from Legendre's formula that $\val_p(i!) \geq
\frac i p - 1$, from what we derive:
$$\val_p(\det M) \geq \sum_{i=1}^d {\textstyle \big(\frac i p - 1\big)}
= \frac{d(d+1)}{2p} - d.$$
When $d$ is large compared to $p$, the number of diffused digits
is much larger than the 
number of $p$-adic numbers on which these diffused digits ``diffuse'' 
(which are the $d{+}1$ values $P(0), \ldots, P(d)$). They
will then need to have an influence at high 
precision\footnote{Roughly speaking, if we want to dispatch $n$ 
(diffused) digits 
between $m$ places, we have to increase the precision by at least $\lceil
\frac n m\rceil$ digits; this is the pigeonhole principle.}. This is why it is so 
difficult to get a sharp precision on the tuple $(P(0), \ldots, P(d))$. Concretely the 
$n$-th digits of the coefficients of $P$ may influence the $(n+\frac 
d{2p})$-th digit of some linear combination of the $P(i)$'s and 
conversely the $n$-th digits of the $P(i)$'s may influence the $(n-\frac 
d{2p})$-th digits of the coefficients of $P$. In other words, in order 
to be able to recover all the coefficients of $P$ at the initial 
precision $O(p^N)$, one should have used at least $N+\frac d{2p}$ digits 
of precision (in the model of floating-point arithmetic).

\subparagraph{Evaluation at general points.}

Until now, we have only considered evaluation at the first integers. We 
may wonder whether to what extend the instability we have observed above 
is related to this particular points. 
It turns out that it is totally independant and that similar behaviors 
show up with any set of evaluation points. Indeed, let $a_0, \ldots, 
a_d$ be pairwise distinct elements of $\Zp$ and consider the function 
$\varphi : \Qp[X]_{\leq d} \to \Qp^{d+1}$, $P \mapsto (P(a_0), P(a_1), 
\ldots, P(a_d))$.
It is linear and its Jacobian at any point is the following Vandermonde 
matrix:
$$\left( \begin{matrix}
1 & 1 & 1 & \cdots & 1 \\
a_0 & a_1 & a_2 & \cdots & a_d \\
a_0^2 & a_1^2 & a_2^2 & \cdots & a_d^2 \\
\vdots & \vdots & \vdots & & \vdots \\
a_0^d & a_1^d & a_2^d & \cdots & a_d^d
\end{matrix} \right)$$
whose determinant is
$V(a_0, \ldots, a_d) = \prod_{0\leq i < j \leq d} (a_i - a_j)$.
Thanks to Eq.~\eqref{eq:numberdiffused}, the number of diffused digits 
of the lattice $\varphi(p^N \Zp[X]_{\leq d})$ is the valuation of 
$V(a_0, \ldots, a_d)$. The next lemma ensures that this number is always
as large as it was in the particular case we have considered at first
(\emph{i.e.} $a_i = i$ for all $i$).

\begin{lem}
For all $a_0, \ldots, a_d \in \Zp$, we have:
$$\val\big(V(a_0, \ldots, a_d)\big) \geq 
\val\big(1! \times 2! \times \cdots \times d!\big).$$
\end{lem}

\begin{proof}
Given a finite subset $A \subset \Zp$, we denote by $\nu(A)$ the 
valuation of $V(x_1, \ldots, x_n)$ where the $x_i$'s are the elements of 
$A$. We note that $V(x_1, \ldots, x_n)$ depends up to a sign to the way 
the elements of $A$ are enumerated but $\nu(A)$ depends only on $A$. We
are going to prove by induction on the cardinality $n$ of $A$ that
$\nu(A) \geq \val\big(1! \times 2! \times \cdots \times (n{-}1)!\big)$.

If $A$ has cardinality $2$, the result is obvious. Let us now consider 
a finite subset $A$ of $\Zp$ of cardinality $n$. For $i \in \{1, \ldots,
p\}$, let $A_i$ be the subset of $A$ consisting of elements which are
congruent to $i$ modulo $p$ and write $n_i = \text{Card } A_i$. Define
in addition $B_i$ as the set of $\frac{x-r} p$ for $x$ varying in $A_i$.
Clearly $B_i$ is a subset of $\Zp$ of cardinality $n_i$.
Moreover, we easily check that: 
\begin{equation}
\label{eq:valvdm}
\nu(A) = \sum_{i=1}^p \nu(A_i) 
= \sum_{i=1}^p \left(\frac{n_i \: (n_i{-}1)} 2 + \nu(B_i)\right).
\end{equation}
Define the function $v : \N \to \N$,
$n \mapsto \frac{n(n{-}1)} 2 + \val(1!) + \cdots + \val((n{-}1)!)$.
It follows from the induction hypothesis that $\nu(A) \geq v(n_1) + 
\cdots + v(n_p)$. Observe that $f(n{+}1) - f(n) = n + \val(n!)$ and
deduce from this that $f(n) + f(m) \leq f(n{+}1) + f(m{-}1)$ as soon
as $m \geq n+2$. Applying again and again this inequality and noting
in addition that $n_1 + \cdots + n_p = n$, we end up with:
\begin{equation}
\label{eq:convexityv}
v(n_1) + \cdots + v(n_p) \geq r \cdot v(q{+}1) + (p{-}r) \cdot v(q)
\end{equation}
where $q$ and $r$ are the quotient and the remainder of the Euclidean
division of $n$ by $p$ respectively. 
It is enough to prove that the right hand side of 
Eq.~\eqref{eq:convexityv} is equal to $\val(1!) + \cdots + 
\val((n{-}1)!)$. This follows from Eq.~\eqref{eq:valvdm} applied with
the particular set $A = \{0, 1, \ldots, n{-}1\}$.
\end{proof}

\begin{rem}
The above proof shows in a similar fashion 
that if $a_0, \ldots, a_d$ are integers, then $V(a_0, \ldots, a_d)$
is divisible by $1! \times 2! \times \cdots \times d!$.
\end{rem}

The conclusion of the above analysis is that the 
evaluation-interpolation strategy --- which is used for instance in FFT 
--- cannot be used in the $p$-adic setting without further care. 
Precisely, from this point of view, one should really think of $\Zp$ as 
if it were a ring of characteristic $p$ and reuse all the techniques 
developed in the case of finite fields for avoiding divisions by $p$.

\subsubsection{Newton iteration}
\label{sssec:NewtonC2}

We have already seen that Hensel's Lemma is a powerful tool in the 
$p$-adic context. However, as highlighted in \S 
\ref{sssec:intervalNewton}, it may have a strange behavior regarding 
precision. Below we use the precision Lemma in order to study carefully 
this phenomenon. We also take the opportunity to extend the Newton 
scheme (on which Hensel's Lemma is based) to the framework of 
multivariate functions of class $C^2$.

\paragraph{Functions of class $C^2$ in the $p$-adic setting}

Recall that, given two $\Qp$-vector spaces $E$ and $F$, we denote by
$\calL(E,F)$ the space of linear mappings from $E$ to $F$. Similarly,
Given three $\Qp$-vector spaces $E_1$, $E_2$ and $F$, we denote by 
$\calB(E_1{\times}E_2, F)$ the space of bilinear functions 
$E_1{\times}E_2 \to F$. We recall that there exist canonical 
isomorphisms --- the so-called \emph{curryfication isomorphisms} --- 
between $\calB(E_1{\times}E_2, F)$, $\calL(E_1, \calL(E_2,F))$ and 
$\calL(E_2, \calL(E_1,F))$. When $E_1 = E_2 = E$, we define further 
$\calB^\symm(E{\times}E)$ as the subspace of $\calB(E{\times}E, F)$ 
consisting of \emph{symmetric} bilinear functions.

We recall also that when $E$ and $F$ are endowed with norms, the
space $\calL(E,F)$ inherits the norm operator defined by 
$\Vert f \Vert_{\calL(E,F)} = \sup_{x \in B_E(1)} \Vert f(x) \Vert_F$.
Similarly, we endow $\calB(E_1{\times}E_2,F)$ with the norm
$\Vert \cdot \Vert_{\calB(E_1{\times}E_2,F)}$ defined by
$\Vert b \Vert_{\calB(E_1{\times}E_2,F)} = \sup_{x_1 \in B_{E_1}(1), 
x_2 \in B_{E_2}(1)} \Vert b(x_1,x_2) \Vert_F$. The curryfication
isomorphisms do preserve the norm.

The definition of ``being of class $C^2$'' is inspired from the 
alternative definition of ``being of class $C^1$'' provided by Remark 
\ref{rem:altdefC1multi} (see also Remark \ref{rem:altdefC1uni}).

\begin{deftn}
\label{def:classC2}
Let $E$ and $F$ be two normed finite dimensional vector spaces and let 
$U$ be an open subset of $E$.
A function $f : U \to F$ is of \emph{class $C^2$} if there exist:
\begin{itemize}
\renewcommand{\itemsep}{0pt}
\item a function $df : U \to \calL(E,F)$, $x \mapsto df_x$
\item a function $d^2 f: U \to \calB^\symm(E{\times}E, F)$, $x \mapsto d^2 f_x$
\item a covering $(U_i)_{i \in I}$ of $U$, and
\item for all $i \in I$, a \emph{continuous} real-valued function 
$\varepsilon_i : \R^+ \to \R^+$ vanishing at $0$
\end{itemize}
such that, for all $i \in I$ and all $x,y \in U_i$:
\begin{equation}
\label{eq:defC2}
\textstyle
\Vert f(y) - f(x) - df_x(y{-}x) - \frac 1 2 d^2 f_x(y{-}x, y{-}x) \Vert_F \leq 
\Vert y{-}x \Vert_E^2 \cdot \varepsilon_i\big(\Vert y{-}x \Vert_E\big)
\end{equation}
\end{deftn}

\begin{prop}
Let $E$ and $F$ be two normed finite dimensional vector spaces and let 
$U$ be an open subset of $E$. Let $f : U \to F$ be a function of class 
$C^2$. Then:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item
The function $f$ is of class $C^1$ on $U$ and its differential is the
function $df$ of Definition \ref{def:classC2};
\item
the function $df$ is of class $C^1$ on $U$ as well and its 
differential is the function $d^2 f$ (viewed as a element of $\calL(E, 
\calL(E,F))$) of Definition \ref{def:classC2}.
\end{enumerate}
\end{prop}

\begin{proof}
We consider the data $df$, $d^2 f$, $U_i$, $\varepsilon_i$ given by 
Definition \ref{def:classC2}. Without loss of generality, we may assume 
that the $U_i$'s are all balls and that the $\varepsilon_i$'s are all 
non-decreasing functions. By this additional assumption, $U_i$ is
compact; the function $x \mapsto \Vert d^2 f_x \Vert_{\calB(E{\times}E,
F)}$ is then bounded on $U_i$ by a constant $C_i$. Therefore
Eq.~\eqref{eq:defC2} implies:
$$\Vert f(y) - f(x) - df_x(y{-}x) \Vert_F \leq 
\Vert y{-}x \Vert_E^2 \cdot \max\big(C_i, \varepsilon_i\big(\Vert y{-}x \Vert_E\big)\big)$$
for all $x,y \in U_i$. The first assertion of the Proposition follows.

We now prove the second assertion. We fix $i \in I$. Let $x,y \in U_i$. 
We set $\delta = \Vert y{-}x \Vert_E$,
\begin{align*}
\ell & = df_x - df_y - d^2 f_x(y-x) \in \calL(E,F) \\
\text{and} \quad
b & = \textstyle \frac 1 2 \big(d^2 f_x - d^2 f_y\big) \in \calB^\symm(E{\times}E, F).
\end{align*}
Observe that, thanks to our assumption on $U_i$ (and ultrametricity), 
the ball $B_E(y,\delta)$ of centre $y$ and radius $\delta$ is included 
in $U_i$. For $z \in B_E(y,\delta)$, we define:
\begin{align*}
\delta_{z,x} & \textstyle 
  = f(z) - f(x) - df_x(z{-}x) - \frac 1 2 d^2 f_x(z{-}x, z{-}x) \smallskip \\
\delta_{z,y} & \textstyle
  = f(z) - f(y) - df_y(z{-}y) - \frac 1 2 d^2 f_y(z{-}y, z{-}y) \smallskip \\
\delta_{y,x} & \textstyle
  = f(y) - f(x) - df_x(y{-}x) - \frac 1 2 d^2 f_x(y{-}x, y{-}x).
\end{align*}
By our assumptions, these three vectors have norm at most $\delta^2 
\cdot \varepsilon_i(\delta)$.
A simple computation using linearity and bilinearity yields 
$\delta_{z,x} - \delta_{z,y} - \delta_{y,x} = \ell(z{-}y) - 
b(z{-}y,z{-}y)$. Consequently, we get the estimation:
$$\Vert \ell(h) - b(h,h) \Vert_F \leq \delta^2 \cdot \varepsilon_i(\delta)$$
for all $h \in E$ with $\Vert h \Vert_E \leq \delta$. Applying it with
$(p{+}1)h$ in place of $h$, we find:
$$\Vert \ell(h) - (p{+}1) \cdot b(h,h) \Vert_F \leq \delta^2 \cdot \varepsilon_i(\delta)$$
and combining now the two above inequalities, we end up with
$\Vert b(h,h) \Vert_F \leq p \delta^2 \cdot \varepsilon_i(\delta)$ and
so $\Vert \ell(h) \Vert_F \leq p \delta^2 \cdot \varepsilon_i(\delta)$
as well. This inequality being true for all $h \in B_E(\alpha)$, we 
derive $\Vert \ell \Vert_{\calL(E,F)} \leq p^2 \delta \cdot 
\varepsilon_i(\delta)$. This proves the differentiability of $df$
together with the fact that its differential is $d^2 f$.
\end{proof}

\paragraph{Hensel's Lemma for functions of class $C^2$}

We want to develop an analogue of Hensel's Lemma (see Theorem 
\ref{th:Hensel}) for a function $f : U \to F$ of class $C^2$ whose 
domain $U$ is an open subset of a finite dimensional $p$-adic vector 
space $E$ and whose codomain $F$ is another $p$-adic vector space of the same 
dimension. For simplicity, we assume further that $U = B_E(1)$ (the
general case is deduced from this one by introducing coverings).
Under this additional assumption, the function $d^2 f$ is bounded, 
\emph{i.e.} there exists a constant $C$ such that
$\Vert d^2 f_x \Vert_{\calB(E{\times}E, F)} \leq C$ for all $x \in B_E(1)$.
Plugging this into Eq.~\eqref{eq:defC2} and possibly enlarging a bit
$C$, we derive the two estimations:
\begin{align}
\forall x, y \in B_E(1), \quad 
& \Vert f(y) - f(x) - df_x(y{-}x) \Vert_F \leq 
  C \cdot \Vert y{-}x \Vert_E^2 \label{eq:Calmost2} \medskip \\
& \Vert df_y - df_x \Vert_{\calL(E,F)} \leq 
  C \cdot \Vert y{-}x \Vert_E \label{eq:Calmost2diff}
\end{align}
which are the key points for making Newton iteration work. 
We consider a point $v \in B_E(1)$ for which:
\begin{equation}
\label{eq:hypHensel}
df_v \text{ is invertible}
\quad \text{and} \quad
\Vert f(v) \Vert_F \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2 
\,<\, C^{-1} \,\leq\, 
\Vert df_v^{-1} \Vert_{\calL(F,E)}.
\end{equation}
We set $r = \big(C \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}\big)^{-1}$ 
and denote by $V$ the open ball of centre $v$ and radius $r$. Observe
that $r \leq 1$, so that $V \subset B_E(1)$.

\begin{lem}
\label{lem:HenselC2}
Under the assumptions \eqref{eq:Calmost2}, \eqref{eq:Calmost2diff}
and \eqref{eq:hypHensel}, the linear function $df_x$ is invertible
and $\Vert df_x^{-1} \Vert_{\calL(F,E)} = \Vert df_v^{-1} \Vert_{\calL(F,E)}$
for all $x \in V$. Moreover, the Newton iteration mapping 
$\calN : V \to E$, $x \mapsto x - df_x^{-1}(f(x))$
takes its values in $V$ and satisfies:
\begin{equation}
\label{eq:estimNewton}
\Vert f(\calN(x)) \Vert_F \leq
C \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2 \cdot \Vert f(x) \Vert_F^2.
\end{equation}
\end{lem}

\begin{proof}
Let $x \in V$. Set $g = df_v - df_x$.
By the inequality \eqref{eq:Calmost2diff}, we have:
$$\Vert g \Vert_{\calL(E,F)} = 
\Vert df_v - df_x \Vert_{\calL(E,F)} \leq C \cdot
\Vert v{-}x \Vert_E < C r = \big(\Vert df_v^{-1} \Vert_{\calL(F,E)}\big)^{-1}.$$
Write
$df_x = df_v - g = df_v \circ (\id_E - df_v^{-1} 
\circ g)$. The above estimation shows that the norm of $df_v^{-1} 
\circ g$ is strictly less than $1$. Therefore the series
$\sum_{n=0}^\infty (df_v^{-1} \circ g)^{\circ n}$ converges to
an inverse of $(\id_E - df_v^{-1} \circ g)$. In particular 
$(\id_E - df_v^{-1} \circ g)$ is invertible and so is $df_x$.
Remark in addition that the norm of $\sum_{n=0}^\infty (df_v^{-1} 
\circ g)^{\circ n}$ is at most $1$ and so
$\Vert df_x^{-1} \Vert_{\calL(F,E)} \leq \Vert df_v^{-1} \Vert_{\calL(F,E)}$.
Reverting the roles of $v$ and $x$, we find even better that equality 
does hold.
Applying now \eqref{eq:Calmost2}, we get:
$$\Vert f(v) - f(x) - df_x(v{-}x) \Vert_F 
  \leq C \cdot \Vert v{-}x \Vert_E^2
  < C r^2 = \frac r {\Vert df_v^{-1} \Vert_{\calL(F,E)}^2}.$$
Applying $df_x^{-1}$, we deduce $\Vert df_x^{-1}(f(v)) - df_x^{-1}(f(x)) 
- (v{-}x) \Vert_F < r$. Notice furthermore that $\Vert v{-}x\Vert_E <
r$ by assumption and $\Vert df_x^{-1}(f(v)) \Vert_F \leq \Vert df_v^{-1} 
\Vert_{\calL(F,E)} \cdot \Vert f(v) \Vert_F < r$.
Consequently we find $\Vert df_x^{-1}(f(x)) \Vert_F < r$ as well, for
what we derive $\calN(x) \in V$. Eq.~\eqref{eq:estimNewton} finally
follows by applying again Eq.~\eqref{eq:Calmost2} with $x$ and 
$\calN(x)$.
\end{proof}

\begin{cor}[Hensel's Lemma for function of class $C^2$]
\label{cor:HenselC2}
Under the assumptions \eqref{eq:Calmost2}, \eqref{eq:Calmost2diff}
and \eqref{eq:hypHensel}, the sequence $(x_i)_{i \geq 0}$ defined by 
the recurrence
$$x_0 = v \quad ; 
\quad x_{i+1} = \calN(x_i) = x_i - df_{x_i}^{-1}(f(x_i))$$
is well defined and converges to $x_\infty \in V$ such that
$f(x_\infty) = 0$.
The rate of convergence is given by:
$$\Vert x_i - x_\infty \Vert_E \leq 
r \cdot \Big(C \cdot \Vert f(v) \Vert_F \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2 \Big)^{2^i} =
\frac {\Big(C \cdot \Vert f(v) \Vert_F \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2 \Big)^{2^i}}
{C \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}}.$$
Moreover $x_\infty$ is the unique solution in $V$ to the equation $f(x) 
= 0$.
\end{cor}

\begin{proof}
We define $\rho = C \cdot 
\Vert f(v) \Vert_F \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2$
By Lemma \eqref{lem:HenselC2}, all the $x_i$'s lie in $V$ and:
\begin{equation}
\label{eq:normfxi}
\Vert f(x_{i+1}) \Vert_F \leq 
C \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2 \cdot \Vert f(x_i) \Vert_F^2
\end{equation}
for all $i$. By induction, we derive $\Vert f(x_i) \Vert_F \leq
\big(C \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)}^2\big)^{-1} 
\cdot \rho^{2^i}$. Therefore:
\begin{equation}
\label{eq:diffxi}
\Vert x_{i+1} - x_i \Vert_E =
\Vert df_{x_i}^{-1}(f(x_i)) \Vert_E \leq
\Vert df_v^{-1} \Vert_{\calL(F,E)} \cdot \Vert f(x_i) \Vert_F \leq
r \cdot \rho^{2^i}.
\end{equation}
The sequence $(x_i)_{i \geq 0}$ is a Cauchy sequence and therefore 
converges to some $x_\infty$ for which $f(x_\infty) = 0$.
Eq.~\eqref{eq:diffxi} implies moreover the announced rate of convergence 
together with the fact that $x_\infty$ belongs to $V$.
It then only remains to prove the uniqueness of the solution to the 
equation $f(x) = 0$ in $V$. For 
this, assume that $y \in V$ satisfies $f(y) = 0$. 
Instantiating Eq.~\eqref{eq:Calmost2} with $x = x_\infty$, we obtain
$\Vert df_{x_\infty}(y{-}x_\infty) \Vert_F \leq C \cdot 
\Vert y{-}x_\infty \Vert_E^2$. Hence:
$$\Vert y{-}x_\infty \Vert_E 
= \Vert df_{x_\infty}^{-1}\big(df_{x_\infty}(y{-}x_\infty)\big) \Vert_E 
\leq C \cdot \Vert df_v^{-1} \Vert_{\calL(F,E)} \cdot
  \Vert y{-}x_\infty \Vert_E^2 
= r^{-1} \cdot \Vert y{-}x_\infty \Vert_E^2.$$
Since moreover $\Vert y - x_\infty \Vert_E$ has to be strictly less
than $r$ by our assumptions, we derive $y = x_\infty$ and uniqueness
follows.
\end{proof}

\begin{rem}
\label{rem:HenselC2}
Hensel's Lemma is stable in the sense that its conclusion remains
correct for a sequence $(x_i)_{i \geq 0}$ satisfying the weaker
recurrence:
$$\quad x_{i+1} = x_i - df_{x_i}^{-1}(f(x_i))\, +\, (\text{some
 small perturbation})$$
as soon as the perturbation is small enough to continue to ensure 
that the estimation \eqref{eq:normfxi} holds.
\end{rem}

\paragraph{Precision}

In practice, the function $f$ of which we want to find a root is often 
not exact but given with some uncertainty; think typically of the case
where $f$ is a polynomial (of given degree) with coefficients in $\Qp$
given at some finite precision.
In order to model this, we introduce $C^2(B_E(1),F)$, the set 
of functions $f : B_E(1) \to F$ of class $C^2$. We endow it with the
$C^2$-norm defined by
$$\Vert f \Vert_{C^2} = \max \big( 
\Vert f \Vert_\infty,
\Vert df \Vert_\infty,
\Vert d^2f \Vert_\infty\big)$$
where the infinite norm of a function is defined as usual as the
supremum of the norms of its values.
We consider in addition a finite dimensional subspace $\calF$ of 
$C^2(B_E(1),F)$. The uncertainty on $f$ will be modeled by some lattice 
in $\calF$.

We fix $v \in B_E(1)$ together with two positive real numbers $C$ and 
$r$. We assume $r \leq 1$. Let $V$ be the open ball in $E$ of centre $v$ 
and radius $r$; clearly $V \subset B_E(1)$.
Let $\calU$ be the open subset of $\calF$ consisting of function $f$
for which Eqs.~\eqref{eq:Calmost2}, \eqref{eq:Calmost2diff} and 
\eqref{eq:hypHensel} hold and $\Vert df_v^{-1} \Vert_{\calL(F,E)}
= \frac r C$. By Hensel's Lemma (Corollary \ref{cor:HenselC2}), any
$f \in \calU$ has a unique zero in $V$. The function:
$$\begin{array}{rcl}
\calZ : \quad \calU & \longrightarrow & V \\
f & \mapsto & x \quad \text{s.t.} \quad x\in V \text{ and } f(x) = 0
\end{array}$$
is then well defined. In the idea of applying the precision Lemma, we
study its differentiability.

\begin{lem}
\label{lem:diffZ}
The function $\calZ$ is differentiable on $\calU$. 
Moreover, its differential at $f \in \calU$ is the linear mapping 
$d\calZ_f : \varphi \mapsto - df_{\calZ(f)}^{-1}(\varphi(\calZ(f)))$.
\end{lem}

\begin{proof}
The lemma can be seen as an application of the $p$-adic implicit 
functions Theorem (see~\cite[Proposition~4.3]{Sc11}). Below, we give 
a direct proof avoiding it.
Let $f, g \in \calU$. We set $x = \calZ(f)$, $\varphi = g{-}f$
and $y = x - df_x^{-1}(\varphi(x))$.
Since $f(x)$ vanishes, Eq.~\eqref{eq:Calmost2} reads:
\begin{equation}
\label{eq:diffZ1}
\Vert g(y) - (\id_E - dg_x \circ df_x^{-1})(\varphi(x)) \Vert_F \leq
C \cdot \Vert y{-}x \Vert_E^2 \leq 
C \cdot \Vert df_x^{-1} \Vert_{\calL(F,E)}^2 \cdot \Vert \varphi \Vert_\infty^2.
\end{equation}
From the identity $\id_E - dg_x \circ df_x^{-1} = (df_x - dg_x) \circ 
df_x^{-1}$, we deduce moreover that 
\begin{equation}
\label{eq:diffZ2}
\Vert \id_E - dg_x \circ df_x^{-1}\Vert_{\calL(E,E)} \leq
\Vert df_x^{-1} \Vert_{\calL(F,E)} \cdot 
\Vert df_x - dg_x \Vert_{\calL(E,F)} \leq \Vert df_x^{-1} \Vert_{\calL(F,E)}
\cdot \Vert d \varphi \Vert_\infty.
\end{equation}
Combining \eqref{eq:diffZ1} and \eqref{eq:diffZ2}, we derive
$\Vert g(y) \Vert_F \leq \Vert df_x^{-1} \Vert_{\calL(F,E)}
\cdot \Vert \varphi \Vert_{C^2}^2$. Corollary \ref{cor:HenselC2}
then implies 
$\Vert y - \calZ(g) \Vert_E \leq \Vert df_x^{-1} \Vert_{\calL(F,E)}^2
\cdot \Vert \varphi \Vert_{C^2}^2 =  \Vert df_v^{-1} \Vert_{\calL(F,E)}^2
\cdot \Vert \varphi \Vert_{C^2}^2$, which proves the lemma.
\end{proof}

Applying the precision Lemma, we end up with the next corollary.

\begin{cor}
If the uncertainty on $f$ is given by a (sufficiently rounded and small) 
lattice $\calH \subset \calU$, then the optimal precision on $x = \calZ(f)$ 
is $df_x^{-1} \circ \ev_x(\calH)$ where $\ev_x : \calH \to F$ is the 
evaluation morphism at $x$.
\end{cor}

\noindent
It is quite instructive to compare this result with the optimal
precision we get after a single iteration of the Newton scheme.
In order to do so, we introduce the map:
$$\begin{array}{rcl}
\calN : \quad \calU \times V & \longrightarrow & V \\
(f,x) & \mapsto & f(x) - df_x^{-1}(f(x))
\end{array}$$
which is well defined thanks to Lemma~\ref{lem:HenselC2}. It is moreover 
clear that $\calN$ is differentiable since it appears as a composite of 
differentiable functions.
A straightforward (but tedious) calculation shows that its differential
at $(f,x)$ is given by:
$$d \calN_{(f,x)} : (\varphi, \xi) \mapsto
-df_x^{-1}\big(\varphi(x)\big) + df_x^{-1}\big(d^2 f_x(\xi, f(x))\big)
+ df_x^{-1}\big( d\varphi_x(f(x))\big).$$
In particular, we observe that if $f(x) = 0$, the last two terms
vanish as well, so that the linear mapping $d \calN_{(f,x)}$ does 
not depend on the second variable $\xi$ and agrees with $d \calZ_f$.
In the language of precision, this means that the optimal precision 
on $\calZ(f)$ is governed by the last iteration of the Newton scheme.

In practice, this result suggests the following strategy: \emph{in order 
to compute $\calZ(f)$, we start with some $v$ satisfying the 
requirements of Hensel's Lemma (Corollary \ref{cor:HenselC2}), we 
iterate the Newton scheme \emph{without taking care (so much) of 
precision} until the obtained approximation looks correct and we finally 
perform a last iteration reintroducing the machinery for tracking 
precision.}
By Remark~\ref{rem:HenselC2}, the omission of precision tracking is 
harmless while the equality $d \calZ_f = d \calN_{(f, \calZ(f))}$ shows 
(under mild assumptions on the lattice $\calH$) that the precision we 
will get at the end (using the above strategy) will be very good because 
the loss of precision will just come from the last iteration and so will 
be limited.

This behavior is very well illustrated by the example of the 
computation of square roots detailed in \S \ref{sssec:intervalNewton}.
(We then encourage the reader to read it and study it again in light of 
the discussion of this paragraph.)
More recently Lairez and Vaccon applied this strategy for the 
computation of the solutions of $p$-adic differential equations with 
separation of variables~\cite{LaVa16}; they managed to improve this way a
former algorithm by Lercier and Sirvent for computing isogenies between 
elliptic curves in positive characteristic~\cite{LeSi08}.

\subsection{Lattice-based methods for tracking precision}
\label{ssec:latticearith}

Until now, we have explained how the precision Lemma can be used for 
finding the optimal precision and analyzing this way the stability of 
algorithms. It turns out that the precision Lemma is also useful for 
\emph{stabilizing} algorithms, \emph{i.e.} for producing a stable 
procedure by altering slightly another given procedure which might be 
highly unstable.

\subsubsection{The method of adaptive precision}
\label{sssec:adaptiveprecision}

We place ourselves in the following general setting. Let 
\texttt{Phi} be a routine that computes a mathematical function $\varphi 
: U \to V$ whose domain $U$ is an open subset of a finite dimensional 
$\Qp$-vector space $E$ (\emph{i.e.} \texttt{Phi} takes as input a tuple 
of $p$-adic numbers but may possibly fails for particular instances 
lying a closed subset) and whose codomain $V$ is an open subset of
another finite dimensional $\Qp$-vector space (\emph{i.e.} \texttt{Phi} 
outputs a tuple of $p$-adic numbers as well).
We assume that $f$ is of class $C^1$ on $U$. We moreover fix an input $x 
\in U$ for which the differential $df_x$ is surjective.

\begin{rem}
While the $C^1$-assumption is usually harmless, the surjectivity
assumption is often more serious. 
For instance, observe that it implies that the dimension of $F$ 
(\emph{i.e.} the size of the output) is not greater than the dimension 
of $E$ (\emph{i.e.} the size of the input). Clearly there exist many 
interesting algorithms that do not satisfy this requirement and 
therefore do not fit into our framework.
One can always however workaround this issue as follows. We write the 
space of outputs $F$ as a direct sum $F = F_1 \oplus \cdots \oplus F_m$ 
and decompose the procedure \texttt{Phi} accordingly, \emph{i.e.} for 
each $i$, we introduce the algorithm $\texttt{Phi}_i$ that outputs only 
the $i$-th part of $\texttt{Phi}$. Clearly $\texttt{Phi}_i$ is modeled 
by the mathematical function $\varphi_i = \pr_i \circ \varphi$ where 
$\pr_i : F \to F_i$ is the canonical projection. As a consequence 
$d\varphi_{i,x} = \pr_i \circ d \varphi_x$; if the $F_i$'s are small 
enough, it will then be unlikely that one of them is not surjective.
For instance, in the special case where the $F_i$'s are all lines,
the writing $F = F_1 \oplus \cdots \oplus F_m$ corresponds to the
choice of a basis of $F$ and the surjectivity of the $d\varphi_{i,x}$'s
is equivalent to the fact that each column of the Jacobian matrix has
a nonzero entry (which is clearly much weaker than requiring its
surjectivity).
\end{rem}

Our aim is to compute $\varphi(x)$ using the procedure \texttt{Phi}; we 
would like moreover to be sharp in terms of precision and get proved results. We 
have seen that neither zealous arithmetic, nor floating-point arithmetic 
can ensure these two requirements at the same time: zealous arithmetic is 
usually not sharp (see \S \ref{sssec:compexamples}, \S 
\ref{sssec:diffexamples} for many examples) while floating-point 
arithmetic is not proved.
The case of lazy/relaxed arithmetic is a bit aside but it is not quite 
satisfactory either. Indeed recall that in the lazy approach, we are 
fixing a target precision; the output of \texttt{Phi} will then be sharp 
by design. The issue is elsewhere and comes from the fact that the 
lazy/relaxed machinery will generally compute the input $x$ at a higher 
precision than needed, consuming then more resources than necessary (see 
\S \ref{sssec:preclemmaatwork} for a theoretical discussion about this
and/or \S \ref{sssec:somos} below for a concrete example).

\paragraph{Context of zealous arithmetic: precision on inputs}

We assume that the input $x$ is given at some precision which is 
represented by a lattice $H$ in $E$: if the $i$-th coordinate of $x$ is 
given at precision $O(p^{N_i})$, the lattice $H$ is the diagonal lattice
$$H = p^{N_1} \Zp \oplus p^{N_2} \Zp \oplus \cdots \oplus
p^{N_d} \Zp \quad \text{(with $d = \dim E$)}$$
(see also \S \ref{sssec:preclemmaatwork}). We underline nevertheless 
that non diagonal lattices $H$ (corresponding to precision data with
diffused digits, see Definition \ref{def:diffuseddigits}) are also 
permitted. We assume that $H$ is nice enough so that the precision 
Lemma applies, giving:
\begin{equation}
\label{eq:adapt}
\varphi(x+H) = \varphi(x) + d\varphi_x(H).
\end{equation}

\subparagraph{Separation of approximation and precision.}

The formula above strongly suggests to split the computation of 
$\varphi(x{+}H)$ into two independent parts corresponding to the two 
summands. Concerning precision, we have to compute the lattice $d 
\varphi_x(H)$. 
One option for this 
%(that will be discussed further in \S %\ref{sssec:implementation})
is to rely on automatic differentiation techniques~\cite{BaBrChDi00,Ne10}; 
this approach however often leads to costly computations and 
really affects the complexity of our algorithm. 
For this reason, alternative strategies are often preferable (when 
available). The simplest one is probably to precompute $d \varphi_x$ by 
hand (we have seen in \S \ref{sssec:diffexamples} that it is tractable 
in many situations) and plug the obtained result into our algorithm.
On the other hand, we observe that $d \varphi_x(H)$ is a lattice in $F$ 
and is therefore encoded by a $\dim F \times \dim F$ matrix, which might 
be quite a large object (compare with $\dim F$ which is the size of 
the output). Therefore, if we do not need to be \emph{extremely} careful
on precision, we could prefer replacing $d \varphi_x(H)$ by a slightly
larger lattice $H_\max$ which is more easily representable, \emph{e.g.}
$H_\max$ could be a diagonal lattice.

For now on, we assume that we have at our disposal a lattice $H_\max$ 
containing $d \varphi_x(H)$ and we focus on the computation of the first 
term of Eq.~\eqref{eq:adapt}, that is the approximation.
Since $d \varphi_x(H)$ is stable under addition, 
Eq.~\eqref{eq:adapt} yields $\varphi(x+H) = y + d\varphi_x(H) 
\subset y + H_\max$ for any $y$ lying in $\varphi(x{+}H)$. Using now
that $H_\max$ is stable under addition as well, we end up with 
$\varphi(x+H) \subset y + H_\max$ as soon as $y \in \varphi(x{+}H) + 
H_\max$.
To conclude, it is then enough to 
compute the value of \emph{any} element of $\varphi(x{+}H)$ at precision 
$H_\max$. For doing so, we can rely on zealous arithmetic: we increase 
sufficiently the precision on $x$ (picking arbitrary values for digits) 
so that $\varphi(x)$ can be computed at precision at least $H_\max$ and 
we output $\varphi(x)$ at precision $H_\max$. 
In more conceptual terms, what we do is to choose an element $x' \in 
x{+}H$ together with a diagonal lattice $H' \subset H$ having the 
property that zealous arithmetic computes $y$ and $H'_\zealous$ 
such that $f(x'+H') \subset y + H'_\zealous$ and 
$H'_\zealous \subset H_\max$. We thus have:
$$y \, \in \, \varphi(x'+H') + H'_\zealous 
\, \subset \, \varphi(x+H) + H_\max$$
as wanted (see also Figure~\ref{fig:adaptive1}).
%
\begin{figure}
\hfill
\begin{tikzpicture}[scale=1.2]

\begin{scope}[beige, rounded corners=10pt]
\fill (0.1,0.2) rectangle (3.9,5.8);
\fill (6.1,0.2) rectangle (9.9,5.8);
\end{scope}

\draw[->,very thick] (4.1,3)--(5.9,3)
  node[midway,above] { $\varphi$ };

\begin{scope}[xshift=2cm,yshift=3.5cm,rotate=20]
\draw[black,fill=green!50] (-0.8,-1.8) rectangle (0.8,1.8);
\node[scale=0.75] at (0,0) { $x{+}H$ };
\end{scope}
\draw[black,opacity=0.8,fill=purple!30] (2,2.8) rectangle (2.8,2);
\node[scale=0.75] at (2.4,2.4) { $x'{+}H'$ };

\begin{scope}[xshift=8cm,yshift=3cm]
\draw[black,fill=orange!50] (-1.7,-2.3) rectangle (1.7,1.8);
\node[above left,scale=0.75] at (1.7,-2.3) { $\varphi(x){+}H_\max$ };
\draw[black,fill=green!50,rotate=-40] (-1,-1.3) rectangle (1,1.3);
\node[scale=0.75] at (0.15,1.1) { $\varphi(x){+}{}$ };
\node[scale=0.75] at (0.1,0.8) { $d\varphi_x(H)$ };
%\draw[black,fill=red!50] (-0.8,-0.8) rectangle (0.4,0.4);
%\node[scale=0.75] at (-0.2,0.1) { $\varphi(x){+}H_\min$ };
\draw[black,opacity=0.8,fill=purple!30] (-1,-1.4) rectangle (0.4,0.2);
\node[scale=0.75] at (-0.3,-0.6) { $y{+}H'_\zealous$ };
\end{scope}

\begin{scope}[very thick, rounded corners=10pt]
\draw (0.1,0.2) rectangle (3.9,5.8);
\draw (6.1,0.2) rectangle (9.9,5.8);
\end{scope}
\end{tikzpicture}
\hfill \null

\caption{The method of adaptive precision: first attempt}
\label{fig:adaptive1}
\end{figure}
%
These ideas lead to the following implementation:

\begin{lstlisting}
    def Phi_stabilized_v1(x):
        lift x(*\textrm{ at precision }$H'$*)
        y = Phi(x)    # (*\color{comment}\textrm{here }$\tty$\textrm{ is computed at precision }$H'_{\zealous}$*)
        return y(*\textrm{ at precision} $H_{\max}$*)
\end{lstlisting}

\noindent
The method presented above works but has a serious drawback: in 
practice, if often happens that the precision encoded by $H'$ 
is very high, leading then to very costly computations. In the next 
paragraph, we explain a strategy for dealing with this issue.

\subparagraph{Viewing \textrm{\tt Phi} as a sequence of steps.}

Roughly speaking, the problem with the above approach comes from the 
fact that the precision is tracked using interval arithmetic all along 
the execution of \texttt{Phi}. In order to avoid this, the idea consists 
in decomposing \texttt{Phi} into several steps; we then let interval 
arithmetic do the job of tracking precision within each individual step 
but we readjust the precision (using informations coming from the 
precision Lemma) each time we switch from one step to the next one. 
Concretely the readjustment is made by applying the method of the
previous paragraph to each individual step.

Let us now go further into the details. As said above, we view the routine 
\texttt{Phi} as a finite sequence of steps $\texttt{Step}_1, \ldots, 
\texttt{Step}_n$. Let $\sigma_i : U_{i-1} \to U_i$ be the mathematical 
function modeling the $i$-th step; its codomain $U_i$ is the space of 
outputs of the $i$-th step, that is the space of ``active'' variables 
after the execution of $i$-th step. Mathematically, it is an open 
subset in a finite dimensional $\Qp$-vector space $E_i$. For $i \in \{0, 
\ldots, n\}$, we set $\varphi_i = \sigma_i \circ \cdots \circ \sigma_1$ 
and $x_i = \varphi_i(x)$; they are respectively the function modeling 
the execution of the first $i$ steps of \texttt{Phi} and the state of 
the memory after the execution of the $i$-th step of \texttt{Phi} on the 
input $x$.
In what follows, we always assume that $d\varphi_{i,x}$ is 
surjective\footnote{This assumption is rather restrictive but really 
simplifies the discussion.} and, even more, that the precision Lemma 
applies with $\varphi_i$, $x$ and $H$, meaning that, for all $i$:
\begin{equation}
\label{eq:adapt2}
\varphi_i(x+H) = x_i + d \varphi_{i,x}(H).
\end{equation}
We define $H_i = d \varphi_{i,x}(H)$ and we assume that we
are given a ``simple'' lattice $H_{i,\max}$ containing $H_i$.
However, for the application we have in mind, this datum will not be 
sufficient; indeed, we shall need to be \emph{extremely} careful with
precision at intermediate levels. In order to do so \emph{without having 
to manipulate the ``complicated'' lattice $H_i$}, we assume that we are 
given in addition another ``simple'' lattice $H_{i,\min}$ which is 
\emph{contained} in $H_i$. With these notations, the stabilized version
of \texttt{Phi} takes the following schematic form:

\begin{lstlisting}[label={algo:phistabzealousv2}]
    def Phi_stabilized_v2(x):
        (*$\tty_0$*) = x
        for (*$i$*) in (*$1, 2, \ldots, n$*):
            lift (*$\tty_{i-1}$\textrm{ at enough precision}*)
            (*$\tty_i$*) = (*$\texttt{Step}_i(\tty_{i-1})$*)  # (*\color{comment}\textrm{here we want }$\tty_i$\textrm{ to be computed at precision at least }$H_{i,\min}$*)
        return (*$\tty_n$\textrm{ at precision} $H_{n,\max}$*)
\end{lstlisting}

\noindent
The locution ``enough precision'' in the above code means that we
want to ensure that zealous arithmetics is able to compute $\tty_i$ 
at precision $H_{i,\min}$. Let $H'_{i-1}$ be a lattice encoding such
an acceptable precision. Rigorously, it can be defined as follows. We
set $y_0 = x$ and for $i \in \{1,\ldots, n\}$, we define inductively 
$H'_{i-1}$, $y_i$ and $H'_{i,\zealous}$ by requiring that the routine 
$\texttt{Step}_i$ called on the input $y_{i-1}$ given at precision 
$H'_{i-1}$ computes $y_i$ at precision $H'_{i,\zealous}$ with 
$H'_{i,\zealous} \subset H_{i,\min}$ (see Figure~\ref{fig:adaptive2}). 
%
\begin{figure}
\hfill
\begin{tikzpicture}[scale=1.2]

\begin{scope}[beige, rounded corners=10pt]
\fill (0.1,0.2) rectangle (3.9,5.8);
\fill (5.1,0.2) rectangle (8.9,5.8);
\end{scope}

\draw[->,very thick] (-0.9,3)--(-0.1,3)
  node[midway,above] { $\sigma_{i-1}$ };
\draw[->,very thick] (4.1,3)--(4.9,3)
  node[midway,above] { $\sigma_i$ };
\draw[->,very thick] (9.1,3)--(9.9,3)
  node[midway,above] { $\sigma_{i+1}$ };

\begin{scope}[xshift=2cm,yshift=3cm,rotate=10]
\draw[black,fill=green!50] (-1.3,-2.3) rectangle (1.3,2.3);
\node[scale=0.75] at (0,1.9) { $x_{i-1}{+}H_{i-1}$ };
\end{scope}

\draw[black,fill=red!50] (1.2,1) rectangle (3,4);
\node[scale=0.75] at (2.1,3.6) { $y_{i-1}{+}H_{i-1,\min}$ };
\draw[black,opacity=0.8,fill=purple!30] (1.5,1.5) rectangle (2.9,2.5);
\node[scale=0.75] at (2.2,2) { $y_{i-1}{+}H'_{i-1}$ };

\begin{scope}[xshift=7cm,yshift=3cm]
%\draw[black,fill=orange!50] (-1.7,-2.3) rectangle (1.7,1.8);
%\node[above left,scale=0.75] at (1.7,-2.3) { $g(x){+}H_\max$ };
\draw[black,fill=green!50,rotate=-40] (-1,-1.3) rectangle (1,1.3);
\node[scale=0.75] at (0.15,1.1) { $x_i{+}H_i$ };
\draw[black,fill=red!50] (-0.8,-0.8) rectangle (0.6,0.4);
\node[scale=0.75] at (-0.1,0.1) { $y_i{+}H_{i,\min}$ };
\draw[black,opacity=0.8,fill=purple!30] (-0.7,-0.7) rectangle (0.5,-0.3);
\node[scale=0.75] at (-0.1,-0.5) { $y_i{+}H'_{i,\zealous}$ };
\end{scope}

\begin{scope}[very thick, rounded corners=10pt]
\draw (0.1,0.2) rectangle (3.9,5.8);
\draw (5.1,0.2) rectangle (8.9,5.8);
\end{scope}
\end{tikzpicture}
\hfill \null

\caption{The method of adaptive precision: second attempt}
\label{fig:adaptive2}
\end{figure}
%
Of course the value $y_i$ we have just defined corresponds 
exactly to the variable $\tty_i$ of the procedure 
\texttt{Phi\_stabilized}. The following lemma is the key for proving the 
correctness of the algorithm \texttt{Phi\_stabilized\_v2}.

\begin{lem}
\label{lem:adaptcorrect}
For all $i \in \{1,\ldots,n\}$, we have $y_i \in x_i + H_i$.
\end{lem}

\begin{proof}
We argue by induction on $i$. The initialization is obvious given that 
$y_0 = x = x_0$. We now assume that $y_{i-1} \in x_{i-1} + H_{i-1}$. 
Applying $\sigma_i$, we derive
\begin{equation}
\label{eq:adapt2step}
\sigma_i(y_{i-1}) \in \sigma_i(x_{i-1} + H_{i-1}) = 
\sigma_i \circ \varphi_{i-1}(x + H) = \varphi_i (x+H) = x_i + H_i
\end{equation}
thanks to Eq.~\eqref{eq:adapt2} applied twice. Moreover, by
construction, we know that $\sigma_i(y_{i-1}) \in y_i + H'_{i,\zealous}
\subset y_i + H_{i,\min} \subset y_i + H_i$. As a consequence the
difference $\sigma_i(y_{i-1}) - y_i$ lies in the lattice $H_i$; this
can be rewritten again as $y_i \in \sigma_i(y_{i-1}) + H_i$. Combining
with Eq.~\eqref{eq:adapt2step}, we get $y_i \in x_i + H_i$ as desired.
\end{proof}

\noindent
We derive from Lemma~\ref{lem:adaptcorrect} that $x_n + H_n = y_n +
H_n$. Combining this equality with Eq.~\eqref{eq:adapt2}, we obtain
$\varphi(x+H) = x_n + H_n = y_n + H_n \subset y_n + H_{\max,n}$.
This inclusion exactly means that \texttt{Phi\_stabilized} called on 
the input $x$ given at precision $H$ outputs an approximation of 
$\varphi(x)$ at precision $H_{\max,n}$; in other words, the algorithm 
\texttt{Phi\_stabilized} is correct.

\smallskip

The main advantage of the algorithm \texttt{Phi\_stabilized\_v2} 
(compared to \texttt{Phi\_stabilized\_v1}) is that the precision encoded 
by the lattices $H'_{i-1}$ are usually acceptable (compared to the 
precision encoded by $H'$). Actual computations are then carried out
at a reasonable precision and then consume moderated resources.

\begin{rem}
Interestingly, observe that the spirit of algorithm 
\texttt{Phi\_stabilized\_v2} is close to the behavior of $p$-adic 
floating-point arithmetic; precisely, in $p$-adic floating-point 
arithmetic, we are decomposing \texttt{Phi} as a sequence of elementary 
steps (each of them consists of a single arithmetic operation) and the 
lattices $H_{i,\min}$ are chosen in such a way that the number of 
significand digits of each $p$-adic variable is kept constant. For 
this reason, the method of adaptive precision can be also used to 
derive \emph{proofs} on the results obtained \emph{via} $p$-adic 
floating-point arithmetic.
\end{rem}

\paragraph{Context of lazy/relaxed arithmetic: target precision}

We recall that the question addressed in the context of lazy/relaxed 
arithmetic is formulated as follows: we fix a target precision on the 
output and we want to avoid as much as possible the computation with
unnecessary digits of the 
input, that are the digits that do not influence the output at the 
requested precision. For simplicity, it is convenient to assume that the 
output of \texttt{Phi} consists of a single $p$-adic 
number\footnote{This assumption is harmless since one can always reduce 
to this case by projection on each coordinate.}, that is $F = \Qp$. The 
target precision is then given by a single integer $N$.

We recall from \S \ref{sssec:preclemmaatwork} that, if $H'$ is a lattice 
in $E$, then the condition:
\begin{equation}
\label{eq:adaptlazy}
\varphi(x+H') \subset \varphi(x) + p^N \Zp
\end{equation}
ensures that the knowledge of $x$ at precision $H'$ is enough to compute 
$\varphi(x)$ at precision $O(p^N)$. 
Moreover, assuming that $H'$ is nice enough so that the precision Lemma 
applies, the inclusion \eqref{eq:adaptlazy} is equivalent to $H' \subset 
d \varphi_x^{-1}(p^N \Zp)$. We draw the attention of the reader to the
fact that $d \varphi_x^{-1}(p^N \Zp)$ itself is usually \emph{not} a 
lattice because $d \varphi_x^{-1}$ often has a nontrivial kernel.
Anyway, 
the above analysis gives a theoretical answer to the question we are 
interested in: the unnecessary digits on the inputs are those lying in 
some nice enough lattice contained in $d\varphi_x^{-1}(p^N \Zp)$.

Mimicking what we have done before in the zealous context, we can use the 
above result to write down a first stabilized version of \texttt{Phi}. 
Set $H = d\varphi_x^{-1} (\Zp)$ and choose a lattice $H_\min$ contained 
in $H$. Clearly $p^N H_\min$ is a lattice contained in $d\varphi_x^{-1} 
(p^N \Zp)$. Moreover, examining closely the statement of the precision 
Lemma, we find that there exists an integer $N_0$ such that:
\begin{equation}
\label{eq:adaptlazyv1}
\varphi(x + p^N H) = \varphi(x) + p^N d \varphi_x(H) \subset
\varphi(x) + p^N \Zp
\end{equation}
for all $N \geq N_0$. The first stabilized version of \texttt{Phi}
then can be described as follows.

\begin{lstlisting}
    def Phi_stabilized_v1(x):
        def Phi_of_x(N):
            xapp = x((*$p^{\max(N,N_0)} H$*))    # (*\color{comment}\textrm{Evaluation of $\ttx$ at precision $p^{\min(N,N_0)} H$}*)
            return Phi(xapp) % (*$p^N$*)
        return Phi_of_x
\end{lstlisting}

\begin{rem}
In the code above, the variable \texttt{x} is a tuple of lazy $p$-adic 
numbers (as described in \S \ref{ssec:lazy}), not a tuple of relaxed 
$p$-adic numbers. Making the method of adaptive precision work within 
the framework of relaxed arithmetic is possible but tedious; for 
simplicity, we shall omit these technicalities and restrict 
ourselves to lazy $p$-adic numbers in this course.

The attentive reader has certainly noticed that the value $N_0$ has 
appeared explicitly in the function \texttt{Phi\_stabilized\_v1} above. 
This implies that we need to precompute it. This has to be done by hand 
\emph{a priori} by explicating the constants in the precision Lemma in 
the special case we are considering; we do not hide that this may 
require some effort (through general techniques for that are 
available~\cite{CaRoVa14,CaRoVa15}).
\end{rem}

Eq.~\eqref{eq:adaptlazyv1} shows that the nested function 
\texttt{Phi\_of\_x} returns the correct answer.
The main benefit of \texttt{Phi\_stabilized\_v1} is that it asks for the 
computation of the input at sharp precision. Nevertheless, it carries 
all the computations with exact rational numbers and so still consumes 
a lot of resources.

In order to get rid of this disadvantage, we can follow the strategy 
already used with success in the zealous context: we decompose 
\texttt{Phi} into $n$ steps $\texttt{Step}_1, \ldots, \texttt{Step}_n$. 
Again we call $\sigma_i : U_{i-1} \to U_i$ the mathematical function 
modeling $\texttt{Step}_i$. We set $\varphi_i = \sigma_i \circ \cdots 
\circ \sigma_1$, $x_i = \varphi_i(x)$ and introduce moreover the 
function $\psi_i = \sigma_n \circ \cdots \circ \sigma_{i+1}$. Clearly 
$\varphi = \psi_i \circ \varphi_i$ for all $i$. Interestingly, observe 
that the relation $d \varphi_x = d \psi_{i,x_i} \circ d \varphi_{i,x}$ 
implies that all the differentials $d \psi_{i,x_i}$'s are surjective 
without any further assumption. For each $i \in \{0,\ldots, n\}$, we 
define $H_i = d \psi_{i,x_i}^{-1} (\Zp)$ (which is generally \emph{not} 
a lattice) and choose a lattice $H_{i,\min}$ contained in $H_i$. The 
optimized form of the stabilized version of \texttt{Phi} then writes as 
follows.

\begin{lstlisting}
    def Phi_stabilized_v2(x):
        def Phi_of_x(N):
            (*$\tty_0$*) = x((*$p^N H_{0,\min}$*))   # (*\color{comment}\textrm{Evaluation of $\ttx$ at precision $p^N H_{0,\min}$}*)
            for (*$i$*) in (*$1, 2, \ldots, n$*):
                (*$\tty_i$*) = (*$\texttt{Step}_i(\tty_{i-1})$*) % (*$p^N H_{i,\min}$*)
            return (*$\tty_n$*) % (*$p^N$*)
        return Phi_of_x
\end{lstlisting}

\noindent
Roughly speaking, the above optimization allows us to ``do the modulo'' 
after each step, avoiding then the explosion of the size of the 
intermediate values.

\begin{prop}
\label{prop:adaptlazy}
There exists an integer $N_0$ (which can be made explicit with some
effort) such that the nested function \textrm{\tt Phi\_of\_x} outputs a
correct answer on the inputs $N \geq N_0$.
\end{prop}

\begin{rem}
We underline that Proposition~\ref{prop:adaptlazy} does not assume the 
surjectivity of the differentials $d \varphi_{i,x}$; recall that this 
assumption was unavoidable in the case of zealous arithmetic and 
appeared as a serious limitation of the method. We have to mention 
however that the analysis of the algorithm \texttt{Phi\_stabilized\_v2} 
is easier under this additional hypothesis (see \S \ref{sssec:somos} 
below for a concrete example).
\end{rem}

\begin{proof}[Sketch of the proof of Proposition~\ref{prop:adaptlazy}]
Given an integer $N$, we let $(y_{N,i})_{0 \leq i \leq n}$ be a 
sequence such that $y_{N,0} = x$ and $y_{N,i} = \sigma_i(y_{N,i-1}) +
p^N H_{i,\min}$.
The value $y_{N,i}$ corresponds to the value of the variable $\tty_i$ 
computed by the function \texttt{Phi\_of\_x} called on the input $N$. 
Moreover, for all fixed $i$, the sequence $(y_{N,i})_{N \geq 0}$ 
converges to $x_i$ when $N$ goes to infinity. Since $\psi_i$ is a 
function of class $C^1$, this implies that $d\psi_{i,y_{N,i}}$ converges 
to $d\psi_{i,x_i}$ when $N$ goes to infinity. From this, we derive that
$d \psi_{i,y_{N,i}}(H_{i,\min}) = d \psi_{i,x_i}(H_{i,\min})$
for $N$ large enough. On the other hand, analyzing closely the proof of the precision 
Lemma, we find that there exists an integer $N_0$ such that:
\begin{align}
\psi_i\big(y_{N,i} + p^N H_{i,\min}\big) 
& = \psi_i(y_{N,i}) + d \psi_{i,y_{N,i}}\big(p^N H_{i,\min}\big) \nonumber \\
& = \psi_i(y_{N,i}) + p^N d \psi_{i,x_i}\big(H_{i,\min}\big)
\label{eq:adaptlazyv2}
\end{align}
for all $i$ and all $N \geq N_0$. We claim that \textrm{\tt Phi\_of\_x} 
outputs a correct answer (\emph{i.e.} that $y_{N,n} \equiv \varphi(x) 
\pmod{p^N}$) whenever $N \geq N_0$. More precisely, we are going to prove by
induction on $i$ that $\psi_i(y_{i,N}) \equiv \varphi(x) \pmod{p^N}$
for $i \in \{0, 1, \ldots, n\}$ and $N \geq N_0$. This is obvious
when $i = 0$ given that $\psi_0 = \varphi$ and $y_{0,N} = x$. Let us
now assume $\psi_{i-1}(y_{i-1,N}) \equiv \varphi(x) \pmod{p^N}$. 
Noting than $\psi_i \circ \sigma_i = \psi_{i-1}$, we derive from
$\sigma_i(y_{N,i-1}) \in y_{N,i} + p^N H_{i,\min}$ that:
\begin{align*}
\psi_{i-1}(y_{N,i-1}) \in \psi_i\big(y_{N,i} + p^N H_{i,\min}\big)
& = \psi_i(y_{N,i}) + p^N d \psi_{i,x_i}\big(H_{i,\min}\big) \\
& \subset \psi_i(y_{N,i}) + p^N \Zp
\end{align*}
using Eq.~\eqref{eq:adaptlazyv2}.
Consequently $\psi_i(y_{N,i}) \equiv \psi_{i-1}(y_{N,i-1}) 
\pmod{p^N}$ and we are done thanks to the induction hypothesis.
\end{proof}

\subsubsection{Example: The Somos 4 sequence}
\label{sssec:somos}

The method of adaptive precision (presented in \S 
\ref{sssec:adaptiveprecision}) has been used with success in several 
concrete contexts: the computation of the solutions of some $p$-adic 
differential equations~\cite{LaVa16} and the computation of 
\textsc{gcd}'s and Bézout coefficients \emph{via} the usual extended 
Euclidean algorithm~\cite{Ca17}.
Below, we detail another example, initially pointed out by Buhler and 
Kedlaya~\cite{BuKe12}, which is simpler (it actually looks like a toy 
example) but already shows up all the interesting phenomena: it is the 
evaluation of the $p$-adic Somos 4 sequences.

\paragraph{The Somos 4 sequence: definition, properties and calculations}

The Somos 4 sequences were first introduced in~\cite{So89} by Somos.
It is a recursive sequence defined by the recurrence:
$$\begin{array}{l}
u_1 = a \quad ; \quad 
u_2 = b \quad ; \quad 
u_3 = c \quad ; \quad 
u_4 = d \medskip \\
\displaystyle
u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n} 
\quad \text{for} \quad n \geq 1
\end{array}$$
where $a$, $b$, $c$ and $d$ are given initial values.
%
\begin{figure}
\tiny \hfill
$\begin{array}{r@{\hspace{0.5ex}}r@{\hspace{0.2ex}}l}
u_5 = & %\frac{u_2 u_4 + u_3^2}{u_1} = &
\frac 1 a &  \big(c^{2} + b d\big)
\medskip \\
u_6 = & %\frac{u_3 u_5 + u_4^2}{u_2} = &
\frac 1{ab} &  \big(c^{3} + b c d + a d^{2}\big)
\medskip \\
u_7 = & %\frac{u_4 u_6 + u_5^2}{u_3} = &
\frac 1{a^2 bc} &  \big(b c^{4} + 2 b^{2} c^{2} d + a c^{3} d + b^{3} d^{2} + a b c d^{2} + a^{2} d^{3}\big)
\medskip \\
u_8 = & %\frac{u_5 u_7 + u_6^2}{u_4} = &
\frac 1{a^3b^2cd} &  \big(b^{2} c^{6} + a c^{7} + 3 b^{3} c^{4} d + 3 a b c^{5} d + 3 b^{4} c^{2} d^{2} + 3 a b^{2} c^{3} d^{2} + 2 a^{2} c^{4} d^{2} + b^{5} d^{3} + a b^{3} c d^{3} + 3 a^{2} b c^{2} d^{3} + a^{2} b^{2} d^{4} + a^{3} c d^{4}\big)
\medskip \\
u_9 = & %\frac{u_6 u_8 + u_7^2}{u_5} = &
\frac 1{a^3b^2c^2d} &  \big(b^{2} c^{8} + a c^{9} + 4 b^{3} c^{6} d + 3 a b c^{7} d + 6 b^{4} c^{4} d^{2} + 6 a b^{2} c^{5} d^{2} + 3 a^{2} c^{6} d^{2} + 4 b^{5} c^{2} d^{3} + 7 a b^{3} c^{3} d^{3} + 6 a^{2} b c^{4} d^{3} + b^{6} d^{4} \\
& & \hspace{5mm}
{}+ 3 a b^{4} c d^{4} + 5 a^{2} b^{2} c^{2} d^{4} + 3 a^{3} c^{3} d^{4} + 2 a^{2} b^{3} d^{5} + 3 a^{3} b c d^{5} + a^{4} d^{6}\big)
\medskip \\
u_{10} = & %\frac{u_7 u_9 + u_8^2}{u_6} = &
\frac 1{a^5b^3c^3d^2} & \big(b^{4} c^{10} + 2 a b^{2} c^{11} + a^{2} c^{12} + 5 b^{5} c^{8} d + 11 a b^{3} c^{9} d + 6 a^{2} b c^{10} d + 10 b^{6} c^{6} d^{2} + 24 a b^{4} c^{7} d^{2} + 17 a^{2} b^{2} c^{8} d^{2} + 4 a^{3} c^{9} d^{2} \\
& \phantom{\frac 1{a^5b^3c^3d^2}} & \hspace{5mm}
{}+ 10 b^{7} c^{4} d^{3} + 26 a b^{5} c^{5} d^{3} + 28 a^{2} b^{3} c^{6} d^{3} + 15 a^{3} b c^{7} d^{3} + 5 b^{8} c^{2} d^{4} + 14 a b^{6} c^{3} d^{4} + 27 a^{2} b^{4} c^{4} d^{4} + 24 a^{3} b^{2} c^{5} d^{4} \\
& \phantom{\frac 1{a^5b^3c^3d^2}} & \hspace{5mm}
{}+ 6 a^{4} c^{6} d^{4} + b^{9} d^{5} + 3 a b^{7} c d^{5} + 14 a^{2} b^{5} c^{2} d^{5} + 19 a^{3} b^{3} c^{3} d^{5} + 12 a^{4} b c^{4} d^{5} + 3 a^{2} b^{6} d^{6} + 6 a^{3} b^{4} c d^{6} + 9 a^{4} b^{2} c^{2} d^{6} \\
& \phantom{\frac 1{a^5b^3c^3d^2}} & \hspace{5mm}
{}+ 4 a^{5} c^{3} d^{6} + 3 a^{4} b^{3} d^{7} + 3 a^{5} b c d^{7} + a^{6} d^{8}\big)
\end{array}$
\hfill \null

\caption{The first terms of the generic SOMOS 4 sequence}
\label{fig:somos}
\end{figure}
%
The first values of the generic Somos 4 sequence (that is the Somos
4 sequence where the initial values are left as indeterminates) are
shown on Figure~\ref{fig:somos}.
We observe a quite surprising property: the denominators of the first 
$u_i$'s are all monomials. We insist on the fact that this behavior
is \emph{a priori} not expected at all. For instance the definition of
$u_9$ involves a division by $u_5$ but a miraculous simplification
occurs.
This phenomenon is in fact general. Fomin and Zelevinsky proved in 
\cite{FoZe02} (as an application of their theory of cluster algebras) 
that the Somos 4 sequence exhibits the \emph{Laurent phenomenon}: for all $n$, 
the term $u_n$ of the generic Somos 4 sequence is a Laurent polynomial
in $a$, $b$, $c$, $d$, that is a polynomial in $a$, $b$, $c$, $d$ and
their inverses.

We address the question of the computation of the $n$-th term (for $n$ 
large) of a Somos 4 sequence whose initial values $a$, $b$, $c$ and $d$ 
are invertible elements of $\Zp$. At least two options are available: 
(1)~we unroll the recurrence starting from the given initial values 
and (2)~we precompute the $n$-th term of the generic Somos 4 
sequence and, in a second time, we plug the values of $a$, $b$, $c$ and 
$d$. Concretely, they correspond to the following codes respectively:

\begin{lstlisting}
    def somos_option1(a,b,c,d,n):     # (*\color{comment}\textrm{We assume }$n \geq 5$*)
        x,y,z,t = a,b,c,d
        for i in (*$1, \, 2, \, \ldots, \, \texttt{n}{-}4$*):
            x,y,z,t = y, z, t, (y*t + z*z)/x
        return t
\end{lstlisting}

\begin{lstlisting}
    def somos_option2(a,b,c,d,n):     # (*\color{comment}\textrm{We assume }$n \geq 5$*)
        X,Y,Z,T = A,B,C,D             # (*\color{comment}\textrm{{\tt A}, {\tt B}, {\tt C}, {\tt D} are formal variables}*)
        for i in (*$1, \, 2, \, \ldots, \, \texttt{n}{-}4$*):
            X,Y,Z,T = Y, Z, T, (Y*T + Z*Z)/X
        return T(A=a, B=b, C=c, D=d)
\end{lstlisting}

The second option has of course a terrible complexity because the size 
of the terms of the generic Somos 4 sequence grows very rapidly (see 
Figure~\ref{fig:somos}). However, if we are mostly interested in 
numerical stability, this option is rather attractive; indeed, we deduce 
from the Laurent phenomenon that if the initial values $a$, $b$, $c$ and 
$d$ (which are supposed to be invertible in $\Zp$) are given at 
precision $O(p^N)$ then all the next terms will be computed at precision 
$O(p^N)$ as well. 
On the contrary, the first option may lead to quite important losses of
precision. 
%
\begin{figure}
\renewcommand{\arraystretch}{1.2}
\hfill \small
$\begin{array}{|c|p{2.6cm}|p{2.6cm}|p{2.6cm}||p{2.6cm}|}
\cline{2-5} 
\multicolumn{1}{c|}{\phantom{\Big(}}
& \hfill\text{Laurent method}\hfill\null 
& \hfill\text{Zealous arith.}\hfill\null 
& \hfill\text{Float. arith.}\hfill\null 
& \hfill\text{Relaxed arith.}\hfill\null \\
\hline
u_1 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_2 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_3 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_4 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_5 & \hfill$\ldots0000000010$ & \hfill$\ldots0000000010$ &  \hfill$\ldots0000000010$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_6 & \hfill$\ldots0000000011$ & \hfill$\ldots0000000011$ &  \hfill$\ldots0000000011$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_7 & \hfill$\ldots0000000111$ & \hfill$\ldots0000000111$ &  \hfill$\ldots0000000111$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_8 & \hfill$\ldots0000010111$ & \hfill$\ldots0000010111$ &  \hfill$\ldots0000010111$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_9 & \hfill$\ldots0000111011$ &  \hfill$\ldots000111011$ &  \hfill$\ldots0000111011$ & {\scriptsize precision on $u_1$:} \hfill $11$ \\
u_{10}
    & \hfill$\ldots0100111010$ &  \hfill$\ldots100111010$ & \hfill$\ldots10100111010$ & {\scriptsize precision on $u_1$:} \hfill $11$ \\
u_{50} 
    & \hfill$\ldots0000011010$ &          \hfill$\ldots0$ & \hfill$\ldots11000011010$ & {\scriptsize precision on $u_1$:} \hfill $19$ \\
u_{54} 
    & \hfill$\ldots0100101001$ &  \hfill \textsc{boom} \hfill\null &  \hfill$\ldots1100101001$ & {\scriptsize precision on $u_1$:} \hfill $20$ \\
u_{500} 
    & \hfill$\ldots1111110010$ &  \hfill --- \hfill\null  & \hfill$\ldots01111110010$ & {\scriptsize precision on $u_1$:} \hfill $109$ \\
\hline
\end{array}$
\hfill\null

\caption{The Somos 4 sequence with initial values $1, 1, 1, 1 \in \Z_2$}
\label{fig:somos1111}
\end{figure}
%
\begin{figure}
\renewcommand{\arraystretch}{1.2}
\hfill \small
$\begin{array}{|c|p{2.6cm}|p{2.6cm}|p{2.6cm}||p{2.6cm}|}
\cline{2-5}
\multicolumn{1}{c|}{\phantom{\Big(}}
& \hfill\text{Laurent method}\hfill\null 
& \hfill\text{Zealous arith.}\hfill\null 
& \hfill\text{Float. arith.}\hfill\null 
& \hfill\text{Relaxed arith.}\hfill\null \\
\hline
u_1 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_2 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_3 & \hfill$\ldots0000000001$ & \hfill$\ldots0000000001$ &  \hfill$\ldots0000000001$ & \hfill --- \hfill\null \\
u_4 & \hfill$\ldots0000000011$ & \hfill$\ldots0000000011$ &  \hfill$\ldots0000000011$ & \hfill --- \hfill\null \\
u_5 & \hfill$\ldots0000000100$ & \hfill$\ldots0000000100$ &  \hfill$\ldots000000000100$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_6 & \hfill$\ldots0000001101$ & \hfill$\ldots0000001101$ &  \hfill$\ldots0000001101$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_7 & \hfill$\ldots0000110111$ & \hfill$\ldots0000110111$ &  \hfill$\ldots0000110111$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_8 & \hfill$\ldots0111010111$ & \hfill$\ldots0111010111$ &  \hfill$\ldots0111010111$ & {\scriptsize precision on $u_1$:} \hfill $10$ \\
u_9 & \hfill$\ldots0111101111$ & \hfill$\ldots11101111$ &  \hfill$\ldots1111101111$ & {\scriptsize precision on $u_1$:} \hfill $12$ \\
u_{10}
    & \hfill$\ldots0000010010$ & \hfill$\ldots00010010$ &  \hfill$\ldots11000010010$ & {\scriptsize precision on $u_1$:} \hfill $12$ \\
u_{11} 
    & \hfill$\ldots1000111001$ & \hfill$\ldots00111001$ &  \hfill$\ldots0000111001$ & {\scriptsize precision on $u_1$:} \hfill $12$ \\
u_{12}
    & \hfill$\ldots0011111101$ & \hfill$\ldots11111101$ &  \hfill$\ldots0011111101$ & {\scriptsize precision on $u_1$:} \hfill $12$ \\
u_{13}
    & \hfill$\ldots0000110101$ & \hfill$\ldots00110101$ &  \hfill$\ldots0000110101$ & {\scriptsize precision on $u_1$:} \hfill $12$ \\
u_{14}
    & \hfill$\ldots1101010011$ & \hfill$\ldots1010011$ &  \hfill$\ldots1101010011$ & {\scriptsize precision on $u_1$:} \hfill $13$ \\
u_{15}
    & \hfill$\ldots0000000000$ & \hfill$\ldots0000000$ &  \hfill$0$ & {\scriptsize precision on $u_1$:} \hfill $13$ \\
u_{16}
    & \hfill$\ldots0101011101$ & \hfill$\ldots1011101$ &  \hfill$\ldots0101011101$ & {\scriptsize precision on $u_1$:} \hfill $13$ \\
u_{17}
    & \hfill$\ldots1001101011$ & \hfill$\ldots1101011$ &  \hfill$\ldots1001101011$ & {\scriptsize precision on $u_1$:} \hfill $13$ \\
u_{18}
    & \hfill$\ldots0011110011$ & \hfill$\ldots1110011$ &  \hfill$\ldots0011110011$ & {\scriptsize precision on $u_1$:} \hfill $13$ \\
u_{19}
    & \hfill$\ldots0000000111$ & \hfill\textsc{boom}\hfill\null &  \hfill\textsc{boom}\hfill\null & {\scriptsize precision on $u_1$:} \hfill $23$ \\
\hline
\end{array}$
\hfill\null

\caption{The Somos 4 sequence with initial values $1, 1, 1, 3 \in \Z_2$}
\label{fig:somos1113}
\end{figure}

The table of Figure~\ref{fig:somos1111} shows the results obtained for 
the particular Somos 4 sequence initialized with $a = b = c = d = 1 \in 
\Z_2$ using various methods. The first column corresponds to the second 
option discussed above; as expected, no losses of precision have to be 
reported. On the second column (resp. the third column), we have 
displayed the results obtained by unrolling the recurrence using the 
algorithm \texttt{somos\_option1} within the framework of zealous 
arithmetic (resp. $p$-adic floating-point arithmetic). We observe that 
the precision decreases rapidly in the second column. For example, one 
digit of precision is lost when computing $u_9$ because of the division 
by $u_5$ which has valuation $1$. These losses of precision continue 
then regularly and, at some point, we find a value which becomes 
indistinguishable from $0$. Dividing by it then produces an error and 
the computations cannot continue. The third column exhibits much better
results though certain digits are still wrong (compare with the first
column).
Finally, the last column concerned lazy/relaxed arithmetic; it shows the 
number of digits of $u_1$ that have been asked for in order to compute 
the term $u_n$ at precision $O(2^{10})$. We observe that this number
increases regularly in parallel with the losses of precision observed
in the zealous context. For instance, the relaxed technology needs one 
more digit of precision on $u_1$ to compute $u_9$ while the zealous
approach loses one digit of precision on $u_9$. There phenomena have of 
course a common explanation: they are both due to the division by $u_5$ 
which has valuation $1$.

The Somos 4 sequence starting with $(a, b, c, d) = (1,1,1,3)$ is even 
more unstable (see Figure~\ref{fig:somos1113}); indeed its $15$-th term 
is divisible by $2^{10}$ and so vanishes at the precision $O(2^{10})$ we 
are considering. Zealous arithmetic and floating-point arithmetic are 
both puzzled after this underflow and crash on the computation of 
$u_{19}$ (which requires a division by $u_{15}$).

\paragraph{Stabilization}

We now apply the theory of adaptive precision, trying to stabilize 
this way the algorithm \texttt{somos\_version1}. The decomposition of 
the routine \texttt{somos\_version1} into steps in straightforward: each 
step corresponds to an interaction of the for loop. They are all modeled 
by the same mathematical function:
$$\begin{array}{rcl}
\sigma : \qquad \Qp^4 & \longrightarrow & \Qp^4 \smallskip \\
(x,y,z,t) & \mapsto & \Big(y,z,t,\frac{yt+z^2}x\Big).
\end{array}$$

\begin{rem}
Of course $\sigma$ is not defined at the points of the form $(0,y,z,t)$. 
In the sequel, it will be nevertheless more convenient to regard 
$\sigma$ as a partially defined function over $\Qp^4$ (than as a 
well-defined function over $\Qp^\star \times \Qp^3$).
\end{rem}

\subparagraph{Differential computations.}

For $i \geq 0$, we set $\varphi_i = \sigma \circ \cdots \circ \sigma$
($i$ times). Clearly $\sigma$ and the $\varphi_i$'s are all functions
of class $C^1$ on their domain of definition. The Jacobian matrix of 
$\sigma$ at the point $(x,y,z,t)$ is easily seen to be:
$$J(\sigma)_{(x,y,z,t)} = 
\left( \begin{array}{c@{\hspace{4ex}}c@{\hspace{4ex}}c@{\hspace{2.5ex}}c}
0 & 0 & 0 & -\frac{yt+z^2}{x^2} \smallskip \\
1 & 0 & 0 & \frac t x \smallskip \\
0 & 1 & 0 & \frac{2z} x \smallskip \\
0 & 0 & 1 & \frac y x 
\end{array} \right).$$
The computation of the Jacobian matrix of $\varphi_i$ is more 
complicated. It is however remarkable that we can easily have access to 
its determinant. Indeed, observe that the determinant of $J(\sigma) 
_{(x,y,z,t)}$ is equal to $\frac {t'} x$ where $t' = \frac{yt+z^2}x$ is 
the last coordinate of $\sigma(x,y,z,t)$. By the chain rule, we then
deduce that:
\begin{equation}
\label{eq:detjacsomos}
\det J(\varphi_i)_{(a,b,c,d)} = 
\frac{u_5}{u_1} \cdot \frac {u_6}{u_2} \cdots \frac{u_{i+4}}{u_i} =
\frac{u_{i+1} \cdot u_{i+2} \cdot u_{i+3} \cdot u_{i+4}}{u_1 \cdot u_2 \cdot u_3 \cdot u_4} =
\frac{u_{i+1} \cdot u_{i+2} \cdot u_{i+3} \cdot u_{i+4}}{abcd}
\end{equation}
where the $u_i$'s are the terms of the Somos 4 sequence initialized 
by $(u_1, u_2, u_3, u_4) = (a,b,c,d)$.
Apart from that, the matrix $J(\varphi_i)_{(a,b,c,d)}$ has another 
remarkable property: if $a,b,c,d$ are invertible in $\Zp$, all the 
entries of $J(\varphi_i)_{(a,b,c,d)}$ lie in $\Zp$; indeed they can
be obtained alternatively as the partial derivatives of the terms of 
the generic Somos 4 sequence in which only divisions by $a$, $b$,
$c$ and $d$ occur. As a consequence, we derive the double inclusion:
\begin{align}
\label{eq:latticesomos}
& p^{v(i)} \cdot \Zp^4 \, \subset \, 
  d \varphi_{i,(a,b,c,d)}\big(\Zp^4\big) \, \subset \, \Zp^4\\
\text{with} \quad &
  v(i) = \val(u_{i+1}) + \val(u_{i+2}) + \val(u_{i+3}) + \val(u_{i+4})
\nonumber
\end{align}
which is the key for making the method of adaptive precision work.

\begin{lem}
\label{lem:somosvi}
For all $i$, there is at most one non invertible element among four
consecutive terms of the Somos 4 sequence.
In particular:
$$v(i) = \max\big(\val(u_{i+1}), \val(u_{i+2}), \val(u_{i+3}),
\val(u_{i+4})\big).$$
\end{lem}

\begin{proof}
It is enough to prove that if $\val(u_{i-3}) = \val(u_{i-2}) = 
\val(u_{i-1}) = 0$ and $\val(u_i) > 0$, then $\val(u_{i+1}) = 
\val(u_{i+2}) = \val(u_{i+3}) = 0$. This can be derived easily from
the defining formula of the Somos 4 sequence.
\end{proof}

\begin{ex*}
As an instructive example, we take the values $(a,b,c,d) = 
(1,1,1,3)$ (given at precision $O(2^{10})$) and $i = 14$ corresponding 
to the annoying situation reported on Figure~\ref{fig:somos1113} where a 
term of the Somos sequence vanishes at the working precision.
An explicit computation yields:
$$\small J(\varphi_{14})_{(1,1,1,3)} = 
\left(\begin{array}{rrrr}
\ldots0001110110 & \ldots0100010110 & \ldots1110111100 & \ldots0111110000 \\
\ldots1100101001 & \ldots1110001010 & \ldots1101010100 & \ldots0111111101 \\
\ldots0001001100 & \ldots0101001110 & \ldots1010110101 & \ldots0011101100 \\
\ldots0000000111 & \ldots0100100101 & \ldots1011100010 & \ldots0101011110
\end{array}\right).$$
According to the results of \S \ref{sssec:preclemmaatwork}, the $j$-th 
column of this matrix (for $1 \leq j \leq 4$) encodes the loss/gain of
precision on the computation of $u_{j+14}$. Precisely, we observe that
each column of $J(\varphi_{14})_{(1,1,1,3)}$ contains an element of 
valuation zero, meaning that the optimal precision on $u_{15}$, 
$u_{16}$, $u_{17}$ and $u_{18}$ separately is $O(2^{10})$. In 
particular, we cannot decide whether $u_{15}$ actually vanishes or 
does not. This sounds quite weird because we were able in any case to 
compute $u_{19}$ using algorithm \texttt{somos\_option1}. The point is 
that the formula given the value of $u_{19}$ has the shape $u_{19} = 
\frac{\text{numerator}}{u_{15}}$ where the numerator is divisible by 
$u_{15}$. The approach based on the Laurent phenomenon views this
simplification and is therefore able to compute $u_{19}$ even when
$u_{15}$ vanishes!

The analysis of the optional precision on the quadruple $(u_{15}, 
u_{16}, u_{17}, u_{18})$ is interesting as well. However, according to 
Eq.~\eqref{eq:detjacsomos}, the determinant of $J(\varphi_{14})_{(1,1, 
1,3)}$ vanishes at the working precision; we cannot then certify its 
surjectivity and it seems that we are in a deadlock. In order to go 
further, we increase a bit the precision on the initial values: for now 
on, we assume that $a$, $b$, $c$ and $d$ (whose values are $1$, $1$, $1$ 
and $3$ respectively) are given at precision $O(2^N)$ with $N > 10$. 
The computation then shows that the determinant of $J(\varphi_{14})_ 
{(1,1,1,3)}$ has valuation $10$. 
Thanks to Eq.~\eqref{eq:numberdiffused} that the quadruple $(u_{15}, 
u_{16}, u_{17}, u_{18})$ has $10$ diffused digits of precision. We
can even be more precise by computing the Hermite normal form of 
$J(\varphi_{14})_{(1,1,1,3)}$, which is:
$$\left( \begin{array}{c@{\hspace{4ex}}c@{\hspace{4ex}}c@{\hspace{4ex}}c}
1 & 0 & 0 & 179 \\
0 & 1 & 0 & 369 \\
0 & 0 & 1 & 818 \\
0 & 0 & 0 & 2^{10} \\
\end{array}\right)$$
meaning that the particular linear combination $179 \: u_{15} + 369 \: u_{16} 
+ 818 \: u_{17} - u_{18}$ could in principle be determined at precision 
$O(2^{N+10})$. Using this, it becomes possible to determine $u_{19}$ at
precision $O(2^N)$ (while a naive computation will lead to precision
$O(2^{N-10})$ because of the division by $u_{15}$).
\end{ex*}

\subparagraph{Explicit precision Lemma.}

Another task we have to do in order to use the theory of adaptive
precision is to make explicit the constants in the precision Lemma
for the particular problem we are studying. In the case of the Somos
4 sequence, the result we need takes the following form.

\begin{prop}
\label{prop:preclemmasomos}
We keep the above notations and assumptions\footnote{In particular,
we continue to assume that the initial values $a$, $b$, $c$ and $d$
are invertible in $\Zp$.}.
For all $N > v(i)$, we have:
$$\varphi_i\big((a,b,c,d) + p^N \Zp^4\big) =
(u_{i+1}, u_{i+2}, u_{i+3}, u_{i+4}) + 
p^N \: d \varphi_{i,(a,b,c,d)}\big(\Zp^4\big).$$
\end{prop}

\begin{proof}
The proof is a slight generalization (to the case of multivariate 
\emph{Laurent} polynomials) of Proposition \ref{prop:preclemmapoly}. 
In fact, the only missing input is the generalization of Lemma 
\ref{lem:multipolyC1} to our setting. Concretely, we then just have 
to establish that: 
\begin{equation}
\label{eq:multilaurentC1}
\Vert \varphi_i(v_2) - \varphi_i(v_1) - d\varphi_{i,v}(v_2{-}v_1) \Vert 
  \leq \max
  \big(\Vert v_1{-}v \Vert, \Vert v_2{-}v \Vert \big) \cdot \Vert v_2{-}v_1 \Vert
\end{equation}
for all integers $i$ and
all vectors $v, v_1, v_2 \in \Zp^4$ whose all coordinates are
invertible. (Here $\Vert \cdot \Vert$ denotes as usual the infinite
norm.) For doing so, we write $\varphi_i$ as the compositum
$\varphi_i = \tilde \varphi_i \circ \iota$ where $\iota$ is the
(partially defined) embedding:
$$\iota : \Qp^4 \to \Qp^4, \quad (x,y,z,t) \mapsto (x,y,z,t, x^{-1},
y^{-1}, z^{-1}, t^{-1})$$
and $\tilde \varphi_i : \Qp^8 \to \Qp^4$ is a function whose each
coordinate is given by a multivariate polynomial with coefficients 
in $\Zp$. We set $w = \iota(v)$, $w_1 = \iota(v_1)$, $w_2 = \iota
(v_2)$ and compute:
\begin{align*}
\varphi_i(v_2) - \varphi_i(v_1) - d\varphi_{i,v}(v_2{-}v_1)
  & = \tilde \varphi_i(w_2) - \tilde \varphi_i(w_1) - d\tilde \varphi_{i,w}
\circ d \iota_v (v_2{-}v_1) \\
  & = \tilde \varphi_i(w_2) - \tilde \varphi_i(w_1) - d\tilde \varphi_{i,w} (w_2{-}w_1) \\
  & \hspace{1.5cm} {}+ d\tilde \varphi_{i,w}( w_2 - w_1 - d \iota_v (v_2{-}v_1)\big).
\end{align*}
From the shape of $\tilde \varphi_i$, we deduce that the norm of its
differential $d\tilde \varphi_{i,w}$ is at most $1$. Therefore:
\begin{align}
&\Vert \varphi_i(v_2) - \varphi_i(v_1) - d\varphi_{i,v}(v_2{-}v_1) \Vert \smallskip \nonumber \\
&\hspace{1cm} \leq \max\big(
  \Vert \tilde \varphi_i(w_2) - \tilde \varphi_i(w_1) - d\tilde \varphi_{i,w} (w_2{-}w_1) \Vert, \,
  \Vert w_2 - w_1 - d \iota_v (v_2{-}v_1) \Vert\big) \smallskip \nonumber \\
&\hspace{1cm} \leq \max\big(
  \Vert w_1{-}w \Vert {\cdot} \Vert w_2{-}w_1 \Vert,\,
  \Vert w_2{-}w \Vert {\cdot} \Vert w_2{-}w_1 \Vert,\,
  \Vert w_2{-}w_1 - d \iota_v (v_2{-}v_1) \Vert\big).
  \label{eq:multilaurentC1step}
\end{align}
the last inequality coming from Lemma \ref{lem:multipolyC1} applied to a 
shift of the function $\tilde \varphi_i$. Given three invertible 
elements $x, x_1, x_2$ in $\Zp$, we have:
$$\left|\frac 1{x_1} - \frac 1{x_2} - \Big(\frac{-1}{x^2}\Big) \cdot (x_1{-}x_2) \right| 
= \left| \frac{(x_1{-}x_2) \cdot (x_1 x_2 - x^2)}{x^2 x_1 x_2} \right|
= |x_1{-}x_2| \cdot |x_1 x_2 - x^2|.$$
Writing $x_1 x_2 - x^2 = x (x_1 - x) + x_1 (x_2 - x)$ shows
that $|x_1 x_2 - x^2| \leq \max\big(|x_1{-}x|, |x_2{-}x|\big)$. Thus:
$$\Vert w_2{-}w_1 - d \iota_v (v_2{-}v_1) \Vert \leq
\Vert v_2{-}v_1 \Vert \cdot \max\big(\Vert v_1{-}v \Vert, \,
\Vert v_2{-}v \Vert\big).$$
Finally, a straightforward computation shows that $\Vert w_1{-}w \Vert = 
\Vert v_1{-}v \Vert$, $\Vert w_2{-}w \Vert = \Vert v_2{-}v \Vert$ and 
$\Vert w_2{-}w_1 \Vert = \Vert v_2{-}v_1 \Vert$. Inserting these inputs
into Eq.~\eqref{eq:multilaurentC1step}, we get Eq.~\eqref{eq:multilaurentC1} 
as desired.
\end{proof}

\subparagraph{Context of zealous arithmetic.}

We fix a positive integer $N$. We assume that the initial values $a$, 
$b$, $c$ and $d$ are given at precision $O(p^N)$. The corresponding 
lattice is $H = p^n \Zp^4$. By Proposition \ref{prop:preclemmasomos}, 
the precision Lemma applies with $\varphi_i$, $x$ and $H$ as soon as $N 
> v(i)$.

We now follow step by step the method of adaptive precision (see \S 
\ref{sssec:adaptiveprecision}). 
We first set $H_i = d\varphi_{i,(a,b,c,d)}(H) = p^N \: 
d\varphi_{i,(a,b,c,d)} (\Zp^4)$. We then have to exhibit two ``simple'' 
lattices $H_{i,\min}$ and $H_{i,\max}$ that approximates $H_i$ from 
below and from above respectively. The answer is given by the inclusions 
\eqref{eq:latticesomos}: we take $H_{i,\min} = p^{N+v(i)} \Zp^4$ and 
$H_{i,\max} = p^N \Zp^4$. Concerning the lattice $H'_{i-1}$, it can be any 
lattice for which the following requirement holds: if the quadruple 
$(x,y,z,t)$ is known at precision $H'_{i-1}$, then zealous 
arithmetic is able to compute $\sigma(x,y,z,t)$ at precision 
$H'_{i,\min} = p^{N+v(i)} \Zp^4$. Since the division by $x$ induces
a loss of precision of at most $\val(x)$ digits, we can safely take
$H'_{i-1} = p^{N+v(i)+\val(u_i)} \Zp^4$.
Instantiating the routine \texttt{Phi\_stabilized\_v2} (page 
\pageref{algo:phistabzealousv2}) to our setting, we end up with the following stabilized 
version of the procedure \texttt{somos\_option1}.

\begin{lstlisting}
    def somos_stabilized(a,b,c,d,n):
        # (*\color{comment}\textrm{We assume }$\texttt{n} \geq 5$*)
        # (*\color{comment}a,b,c,d\textrm{ are \emph{units} in $\Zp$ given at precision $O(p^N)$}*)
        x,y,z,t = a,b,c,d
        for i in (*$1, \, 2, \, \ldots, \, \texttt{n}{-}4$*):
            u = (y*t + z*z)/x   # (*\color{comment}\textrm{Precomputation at small precision}*)
            v = (*$\val_p$*)(y) + (*$\val_p$*)(z) + (*$\val_p$*)(t) + (*$\val_p$*)(u)
            if v(*${}\geq N$*): raise PrecisionError
            lift x,y,z,t (*\textrm{at precision $O(p^{N+\texttt{v}+\val_p(\ttx)})$}*)
            x,y,z,t = y, z, t, (y*t + z*z)/x
        return t (*\textrm{at precision $O(p^N)$}*)
\end{lstlisting}

\noindent
This algorithm always returns a correct answer at precision $O(p^N)$. 
It might fail --- and then it raises an error --- if the precision on the 
entries is not sufficient; by Lemma \ref{lem:somosvi}, this occurs 
exactly when one term of the Somos 4 sequence we are computing is 
indistinguishable from $0$. One possible solution in that case is to 
increase arbitrarily the precision on $a$, $b$, $c$ and $d$, rerun the 
algorithm and finally truncate the obtained result back to precision 
$O(p^N)$.

\subparagraph{Context of lazy/relaxed arithmetic.}

The literal translation to the lazy world of the stabilized algorithm 
we have designed just above is:

\begin{lstlisting}
    def somos_stabilized(a,b,c,d,n):
        # (*\color{comment}\textrm{We assume }$\texttt{n} \geq 5$*)
        # (*\color{comment}a,b,c,d\textrm{ are lazy \emph{invertible} $p$-adic numbers}*)
        def nth_term(N):
            x,y,z,t = a(N),b(N),c(N),d(N)
            for i in (*$1, \, 2, \, \ldots, \, \texttt{n}{-}4$*):
                u = (y*t + z*z) / x
                v = (*$\val_p$*)(y) + (*$\val_p$*)(z) + (*$\val_p$*)(t) + (*$\val_p$*)(u)
                if v(*${}\geq N$*): raise PrecisionError
                x,y,z,t = y % (*$p^{\texttt{N+v}}$*), z % (*$p^{\texttt{N+v}}$*), t % (*$p^{\texttt{N+v}}$*), u % (*$p^{\texttt{N+v}}$*)
            return t % (*$p^{\texttt N}$*)
        return nth_term
\end{lstlisting}

\begin{rem}
Instead of raising an error if $N$ is too small, it is more clever to
rerun the computation at higher precision. We can implement this idea
simply by replacing
the statement ``\texttt{{\color{blue} raise} PrecisionError}'' by
``\texttt{{\color{blue} return} ntn\_term(2*N)}'' in the code above.
\end{rem}

\noindent 
The correctness of this algorithm is proved similarly to the case of 
zealous arithmetic by applying the precision Lemma to the function 
$\varphi_i$ (see Proposition \ref{prop:preclemmasomos}). We notice, in 
particular, that we do not need here Proposition \ref{prop:adaptlazy}; 
this is nice because making explicit the constant $N_0$ of 
Proposition \ref{prop:adaptlazy} is tedious. This simplification is
due to the fact that the differential $d \varphi_{i,(a,b,c,d)}$ is
surjective (actually even bijective).

%\subsubsection{Towards automization and implementation}
%\label{sssec:implementation}

\addcontentsline{toc}{section}{\refname}
\begin{thebibliography}{99}
\small
\renewcommand{\itemsep}{0pt}
\bibitem{AbLo04}
  Jounaïdi Abdeljaoued and Henri Lombardi.
  \emph{Méthodes matricielles, introduction à la complexité algébrique}.
  Springer Verlag, Berlin, Heidelberg (2004)

\bibitem{Am75}
  Yvette Amice.
  \emph{Les nombres p-adiques}.
  PUF (1975)

\bibitem{BaBrChDi00}
  Micheal Bartholomew-Biggs, Steven Brown, Bruce Christianson and Laurence Dixon.
  \emph{Automatic differentiation of algorithms}.
  Journal of Comp. and App. Math. \textbf{124} (2000), 171--190

\bibitem{pari}
  Christian Batut, Karim Belabas, Dominique Benardi, Henri Cohen and Michel Olivier.
  \emph{User's guide to {PARI-GP}}. {(1985--2013)}.

\bibitem{Be11}
  Laurent Berger.
  \emph{La correspondance de Langlands locale $p$-adique pour $\GL_2(\Qp)$}.
  Astérisque \textbf{339} (2011), 157--180

\bibitem{BO78}
  Pierre Berthelot and Arthur Ogus.
  \emph{Notes on Crystalline Cohomology}.
  Princeton University Press (1978)

\bibitem{BeHoLe11}
  Jérémy Berthomieu, Joris van~der Hoeven, and Grégoire Lecerf.
  \emph{Relaxed algorithms for $p$-adic numbers}.
  {J. Number Theor. Bordeaux} \textbf{23} (2011), 541--577

\bibitem{BeLe12}
  Jérémy Berthomieu and Romain Lebreton. 
  \emph{Relaxed $p$-adic Hensel lifting for algebraic systems}.
  In ISSAC'12 (2012), 59--66

\bibitem{Bh14}
  Bhargav Bhatt.
  \emph{What is... a Perfectoid Space?}
  Notices of the Amer. Math. Soc. \textbf{61} (2014), 1082--1084

\bibitem{Bi11}
  Richard Bird.
  \emph{A simple division-free algorithm for computing determinants}.
  Information Processing Letters \textbf{111} (2011), 1072--1074

\bibitem{magma}
  Wieb Bosma, John Cannon, and Catherine Payoust.
  \emph{The Magma algebra system. I. The user language.}
  J. Symbolic Comput. \textbf{24} (1997), 235--265

\bibitem{BoChLeSaSc07}
  Alin Bostan, Frédéric Chyzak, Grégoire Lecerf, Bruno Salvy and Éric Schost.
  \emph{Differential equations for algebraic functions}.
  In ISSAC'07 (2007), 25--32

\bibitem{BoGoPeSc05}
  Alin Bostan, Laureano Gonz{\'a}lez-Vega, Hervé Perdry and Éric Schost.
  \emph{From {N}ewton sums to coefficients: complexity issues in
  characteristic $p$}.
  In {\em {MEGA'05}} (2005)

%\bibitem{BoMoSaSc08}
%  Alin Bostan, François Morain, Bruno Salvy and Éric Schost. 
%  \emph{Fast algorithms for computing isogenies between elliptic curves}.
%  Math. Comput. \textbf{77} (2008), 1755--1778

\bibitem{BrCo09}
  Olivier Brinon and Brian Conrad.
  \emph{CMI Summer School notes on p-adic Hodge theory}.
  Available at \url{http://math.stanford.edu/~conrad/papers/notes.pdf} (2009)

\bibitem{BuKe12}
  Joe Buhler and Kiran Kedlaya.
  \emph{Condensation of determinants and the Robbins phenomenon}.
  Microsoft Research Summer Number Theory Day, Redmond (2012),
  available at \url{http://kskedlaya.org/slides/microsoft2012.pdf}

\bibitem{Ca12}
  Xavier Caruso.
  \emph{Random matrices over a DVR and LU factorization}.
  J. Symbolic Comput. \textbf{71} (2015), 98--123

\bibitem{Ca17}
  Xavier Caruso.
  \emph{Numerical stability of Euclidean algorithm over ultrametric fields}.
  to appear at J. Number Theor. Bordeaux

\bibitem{CaRoVa14}
  Xavier Caruso, David Roe and Tristan Vaccon.
  \emph{Tracking $p$-adic precision}.
  LMS J. Comp. and Math. {\bf 17} (2014), 274--294

\bibitem{CaRoVa15}
  Xavier Caruso, David Roe and Tristan Vaccon.
  \emph{$p$-adic stability in linear algebra}.
  In ISSAC'15 (2015)

\bibitem{CaRoVa17}
  Xavier Caruso, David Roe and Tristan Vaccon.
  \emph{Characteristic polynomials of $p$-adic matrices}.
  In preparation

\bibitem{Ch83}
  Man-Duen Choi. 
  \emph{Tricks or treats with the Hilbert matrix.},
  Amer. Math. Monthly (1983), 301--312

\bibitem{Co89}
  John Coates.
  \emph{On $p$-adic $L$-functions}.
  Astérisque \textbf{177} (1989), 33--59

\bibitem{Co93}
  Henri Cohen,
  \emph{A course in computational algebraic number theory},
  Springer Verlag, Berlin (1993)

\bibitem{Co98}
  Pierre Colmez.
  \emph{Integration sur les variétés $p$-adiques}.
  Astérisque \textbf{248} (1998)

\bibitem{Co10}
  Pierre Colmez,
  \emph{Fonctions d'une variable $p$-adique},
  Astérisque \textbf{330} (2010), 13--59

\bibitem{DaMu97}
  Marc Daumas, Jean-Michel Muller~et~al.
  \emph{Qualité des Calculs sur Ordinateur}.
  Masson, Paris (1997)

\bibitem{DvTr89}
  Roberto Dvornicich and Carlo Traverso. 
  \emph{Newton symmetric functions and the arithmetic of algebraically closed fields}. 
  In AAECC-5, LNCS \textbf{356}, Springer, Berlin (1989), 216--224

\bibitem{Ei95}
  David Eisenbud.
  \emph{Commutative Algebra: with a view toward algebraic geometry}.
  Springer Science \& Business Media \textbf{150} (1995)

\bibitem{FoZe02}
  Sergey Fomin and Andrei Zelevinsky.
  \emph{The Laurent phenomenon}.
  {Advances in Applied Math.} \textbf{28} (2002), 119--144

\bibitem{Fu09}
  Martin Fürer. 
  \emph{Faster integer multiplication}.
  SIAM J. Comput. \textbf{39} (2009), 979--1005

\bibitem{SGA}
  Alexander Grothendieck~et~al.
  \emph{Séminaire de géométrie algébrique du Bois-Marie}.
  (1971--1977)

\bibitem{GaHoWeRiKo06}
  Pierrick Gaudry, Thomas Houtmann, Annegret Weng, Christophe Ritzenthaler and David Kohel.
  \emph{The $2$-adic CM method for genus $2$ curves with application to cryptography}.
  In {Asiacrypt 2006}, {LNCS 4284}, 114--129

\bibitem{GaGe03}
  Joachim von zur Gathen and Jürgen Gerhard.
  \emph{Modern Computer Algebra}.
  Cambridge University Press, Cambridge (2003)

%\bibitem{Go91}
%  David Goldberg.
%  \emph{What every computer scientist should know about floating-point arithmetic}.
%  ACM Computing Surveys \textbf{23} (1991)

\bibitem{Go88}
  Fernando Gouvêa.
  \emph{Arithmetic of $p$-adic modular forms}.
  Lecture Notes in Math. \textbf{1304}, Springer-Verlag, Berlin, New-York (1988)

\bibitem{Go97}
  Fernando Gouvêa.
  \emph{$p$-adic Numbers: An Introduction}.
  Springer (1997)

\bibitem{He97}
  Kurt Hensel. 
  \emph{Über eine neue Begründung der Theorie der algebraischen Zahlen.}
  Jahresbericht der Deutschen Mathematiker-Vereinigung \textbf{6} (1897), 83--88

\bibitem{HaMC91}
  James Hafner and Kevin McCurley.
  \emph{Asymptotically fast triangularization of matrices over rings}. 
  SIAM Journal of Comp. \textbf{20} (1991), 1068--1083

\bibitem{Ha16}
  David Harari.
  \emph{Zéro-cycles et points rationnels sur les fibrations en 
  variétés rationnellement connexes (d'après Harpaz et Wittenberg)}.
  Séminaire Bourbaki, Exp. 1096, Astérisque \textbf{380} (2016), 231--262

\bibitem{HaWi16}
  Yonatan Harpaz and Olivier Wittenberg.
  \emph{On the fibration method for zero-cycles and rational points}.
  Ann. of Math. \textbf{183} (2016), 229--295.

\bibitem{HaHoLe15}
  David Harvey, Joris van der Hoeven and Grégoire Lecerf. 
  \emph{Even faster integer multiplication}.
  J. Complexity \textbf{36} (2015), 1--30

\bibitem{Hi56}
  Francis Hilbebrand.
  \emph{Introduction to Numerical Analysis}.
  McGraw-Hill, New York (1956)

\bibitem{Ho97}
  Joris van~der Hoeven.
  \emph{Lazy multiplication of formal power series}. 
  In ISSAC'97 (1997), 17--20

\bibitem{Ho02}
  Joris van~der Hoeven.
  \emph{Relax, but don't be too lazy}.
  {J. Symbolic Comput.} \textbf{34} (2002), 479--542

\bibitem{Ho07}
  Joris van~der Hoeven.
  \emph{New algorithms for relaxed multiplication}.
  {J. Symbolic Comput.} \textbf{42} (2007), 792--802

\bibitem{mathemagix}
  Joris van der Hoeven, Grégoire Lecerf and Bernard Mourrain 
  \emph{The Mathemagix Language}, 2002--2012

\bibitem{Ho75}
  Alston Householder, 
  \emph{The Theory of Matrices in Numerical Analysis}. 
  (1975)

\bibitem{IEEE08}
  IEEE Computer Society.
  \emph{{IEEE} Standard for Floating-Point Arithmetic, {IEEE} Standard 754-2008}.
  IEEE Computer Society, New York (2008)

\bibitem{KaVi05}
  Erich Kaltofen and Gilles Villard.
  \emph{On the complexity of computing determinants}.
  Comp. Complexity \textbf{13} (2005), 91--130

%\bibitem{KaOf62}
%  Anatolii Karatsuba and Yuri Ofman.
%  \emph{Multiplication of many-digital numbers by automatic computers}.
%  {Proceedings of the USSR Academy of Sciences} \textbf{145} (1962), 293--294

\bibitem{Ke01}
  Kiran Kedlaya.
  \emph{Counting points on hyperelliptic curves using {M}onsky--{W}ashnitzer cohomology}.
  {J. Ramanujan Math. Soc.} \textbf{16} (2001), 323--338

\bibitem{Ke10}
  Kiran Kedlaya.
  \emph{$p$-adic Differential Equations}.
  Cambridge University Press (2010)

\bibitem{KeRo08}
  Kiran Kedlaya and David Roe.
  \emph{Two specifications for $p$-adic floating-point arithmetic: a
  Sage enhancement proposal}.
  Personal communication

\bibitem{Ko84}
  Neal Koblitz.
  \emph{$p$-adic Numbers, $p$-adic Analysis, and Zeta-Functions}.
  Graduate Texts in Math. \textbf{58}, Berlin, New York, Springer-Verlag (1984)

\bibitem{LaVa16}
  Pierre Lairez and Tristan Vaccon,
  \emph{On $p$-adic differential equations with separation of variables}.
  In ISSAC'16 (2016)

\bibitem{La04}
  Alan Lauder.
  \emph{Deformation theory and the computation of zeta functions}.
  {Proc. London Math. Soc.} \textbf{88} (2004), 565--602

\bibitem{Le13}
  Romain Lebreton. 
  \emph{Relaxed Hensel lifting of triangular sets}. 
  In MEGA’13 (2013)

\bibitem{LeLeLo82}
  Arjen Lenstra, Hendrik Lenstra and László Lovász.
  \emph{Factoring polynomials with rational coefficients}.
  Math. Ann. \textbf{261} (1982), 515--534

\bibitem{LeSi08}
  Reynald Lercier and Thomas Sirvent.
  \emph{On Elkies subgroups of $\ell$-torsion points in elliptic curves
  defined over a finite field}.
  {J. Number Theor. Bordeaux} \textbf{20} (2008), 783--797

\bibitem{LS07}
  Bernard Le Stum.
  \emph{Rigid cohomology}.
  Cambridge University Press (2007)

\bibitem{LiPi94}
  Carla Limongelli and Roberto Pirastu. 
  \emph{Exact solution of linear systems over rational
  numbers by parallel $p$-adic arithmetic}. 
  In Parallel Processing: CONPAR 94-VAPP VI (1994), 313--323

\bibitem{Ma58}
  Kurt Mahler.
  \emph{An interpolation series for continuous functions of a $p$-adic variable}.
  J. reine und angew. Mathematik \textbf{199} (1958), 23--34

\bibitem{Ma71}
  Yuri Manin.
  \emph{Le groupe de Brauer-Grothendieck en géométrie diophantienne}.
  In Actes du Congrès International des Mathématiciens (Nice, 1970), Tome 1, 
  Gauthier-Villars, Paris (1971), 401--411

\bibitem{Mu89}
  Jean-Michel Muller.
  \emph{Arithmétique des ordinateurs}.
  Masson, Paris (1989)

\bibitem{Mu09}
  Jean-Michel Muller~et~al.
  \emph{Handbook of floating-point arithmetic}.
  Birkhauser, Boston (2009)

\bibitem{Ne10}
  Richard Neidinger,
  \emph{Introduction to Automatic Differentiation and MATLAB Object-Oriented Programming}.
  SIAM Review \textbf{52} (2010), 545--563

\bibitem{Ne99}
  Jurgen Neurkich.
  \emph{Algebraic Number Theory}.
  Springer, Berlin, New York (1999)

\bibitem{Ne12}
  Jurgen Neurkich,
  \emph{Class Field Theory. The Bonn lectures}.
  Edited by Alexander Schmidt. Springer, Berlin, London (2013)

\bibitem{Ro20}
  Alain Robert.
  \emph{A course in $p$-adic analysis}.
  Springer Science \& Business Media \textbf{198} (2000)

\bibitem{Ro97}
  R.~Tyrell Rockafellar.
  \emph{Variational Analysis}.
  Grundlehren der Mathematischen Wissenschaften \textbf{317}, Springer-Verlag
  (1997)

\bibitem{Sa72}
  Pierre Samuel.
  \emph{Algebraic theory of numbers}.
  Paris, Hermann (1972)

\bibitem{Sc11}
  Peter Schneider.
  \emph{$p$-Adic Lie groups}.
  {Grundlehren der mathematischen Wissenschaften 344}. Springer,
  Berlin (2011)

\bibitem{Sc12}
  Peter Scholze.
  \emph{Perfectoid spaces: A survey}.
  Current Developments in Mathematics (2012)

\bibitem{Sc14}
  Peter Scholze.
  \emph{Perfectoid spaces and their Applications}.
  Proceedings of the ICM 2014 (2014)

\bibitem{Sc82}
  Arnold Schönhage,
  \emph{The fundamental theorem of algebra in terms of computational complexity}.
  Technical report, Univ. Tübingen (1982)

\bibitem{Se62}
  Jean-Pierre Serre.
  \emph{Corps locaux}.
  Hermann Paris (1962)

\bibitem{Se70}
  Jean-Pierre Serre.
  \emph{Cours d'arithmétique}.
  PUF (1970)

\bibitem{So89}
  Michael Somos.
  \emph{Problem 1470}.
  {Crux Mathematicorum} \textbf{15} (1989), 208--208.

\bibitem{sage}
  William Stein~et~al.
  \emph{Sage Mathematics Software},
  The Sage Development Team, {2005--2017}.

\bibitem{TaWi95}
  Richard Taylor and Andrew Wiles.
  \emph{Ring theoretic properties of certain Hecke algebras}.
  Ann. of Math. \textbf{141} (1995), 553--572

\bibitem{Va15}
  Tristan Vaccon.
  \emph{Précision $p$-adique}.
  PhD Thesis (2015). Available at \url{https://tel.archives-ouvertes.fr/tel-01205269}

\bibitem{We49}
  André Weil.
  \emph{Numbers of solutions of equations in finite fields}.
  Bull. Amer. Math. Soc. \textbf{55} (1949), 497--508

\bibitem{Wi95}
  Andrew Wiles.
  \emph{Modular elliptic curves and Fermat's Last Theorem}.
  Ann. of Math. \textbf{141} (1995), 443--551

\bibitem{Wi96}
  Franz Winkler.
  \emph{Polynomial Algorithms in Computer Algebra}.
  Springer Wien New Work (1996)
\end{thebibliography}

\end{document}

### 2017-04-14 charpoly/charpoly.tex
Addressing easiest of the referee comments, adding list of things to address to introduction

@@
 
+\newcommand{\done}[1]{\textcolor{blue}{#1}}
+\newcommand{\tdo}[1]{\textcolor{red}{#1}}
+
 \def\binom#1#2{\Big(\begin{array}{cc} #1 \\ #2 \end{array}\Big)}
@@
 \section{Introduction}
+\vspace{0.1in}
+\noindent \textbf{Reviewer 1:}
+\begin{enumerate}
+\item \tdo{What are the implications, if any, of your work on well-known methods such as Dixon's algorithm?}
+\item \tdo{In section 3, Algorithm 1: what is the cost of step 5, and how did you avoid quartic complexity?}
+\item \tdo{In section 4, give a concrete construction for approximating the characteristic polynomial as a ``by product'' of the work of [15].}
+\item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
+\item \done{We clarified why we chose the distribution in Figure 1.}
+\item \tdo{Think about whether Proposition 5.6 can be sped up using the work of Kedlaya and Umans.}
+\end{enumerate}
+\textbf{Reviewer 2:}
+\begin{enumerate}
+\item \done{We adjusted the reference as suggested, as also pointed out by Reviewers 3 and 5.}
+\end{enumerate}
+\textbf{Reviewer 3:}
+\begin{enumerate}
+\item \tdo{Convince reviewer that the complexity claimed in Theorem 4.1 is accurate: ``it lacks a precision loss analysis and relies on classic division based algorithms}
+\item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
+\item \tdo{The definition of comatrix is usually the matrix formed by the cofactors such that $A^{-1} = \com(A)^T / \det(A)$.}
+\item \tdo{The definitions of companion matrix in the proof of Proposition 2.1 and in Section 2.3 are inconsistent (coefficients on the last column or last row)}
+\item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
+\item \done{We have switched from $A^t$ to $A^T$ to denote the transpose.}
+\item \tdo{A few explanations on where does the formula for the differential comes from
+would be most welcome (those of [3 Appendix B] and some connection of the trace
+with the fact that we do degree one approx of a charpoly)}
+\item \tdo{In the experiments, (section 5.1), it is not clear which of the two methods is being used to compute the comatrix.}
+\item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing com(A) differ), would be welcome.}
+\item \tdo{Clarify the intro paragraph just before the Previous contributions section.}
+\item \tdo{Make the relationship between $P$, $Q$ and $R$ in Theorem 4.1 explicit.}
+\item \done{We have carried out the other minor corrections suggested.}
+\end{enumerate}
+\textbf{Reviewer 5:}
+\begin{enumerate}
+\item \tdo{In the abstract, clarify what kind of operations are being counted.}
+\item \tdo{In the abstract, clarify what we mean by ``precision exactly $O(p^k)$.''}
+\item \tdo{In the introduction, we say ``Let $M$ be known at precision $O(p^m)$.''  What does this mean?  Coefficient-wise?  Overall?  This needs to be said clearly.}
+\item \tdo{In the introduction, clarify what kind of operations are being counted.}
+\item \tdo{In the discussion of existing methods, clarify what we mean by the claim that they do not yield optimal precision.}
+\item \tdo{Define comatrix properly, as requested by reviewer 3.}
+\item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}
+\item \tdo{The definition of ``flat precision'' and ``jagged precision'' both need to be more clearly defined.}
+\end{enumerate}
+\textbf{General clarity comments:}
+\begin{enumerate}
+\item \tdo{The writing is rather dense and relies too heavily on external material, which makes the reading sometimes unpleasant. (3)}
+\item \tdo{Overall the presentation lacks a minimal set of definitions and terminology regarding the notions of p-adic rings and their precision to make it sufficiently self-contained for an non-expert reader.
+I believe only a one or two additional paragraphs could be sufficient to drastically improve the presentation of section 2, in particular recalling the various precision models.
+The proofs are very concise and rely too heavily on a good command of p-adic precision terminology.
+(3)}
+\item \tdo{The paper itself is unnecessarily difficult to read, to the extent that it sometimes unclear what is being proven or to verify correctness.
+A number of key definitions are not given and there are too few motivating/clarifying examples.
+(5)}
+\item \tdo{The definitions are unclear and the comparison with previous work is deficient.
+If this is to appear the definitions and results must be made very clear, and the results must be demonstrated to be superior to the division-free algorithms cited, both in theory and in practice.}
+\end{enumerate}
 
@@
 In fact, the best known division-free algorithm for computing determinants
-over arbitrary rings \cite{kaltofen:92a} does so using the characteristic polynomial.
+over arbitrary rings \cite{kaltofen-villard:04a} does so using the characteristic polynomial.
 Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
@@
 A faster approach over a field is to compute the Frobenius normal form of $M$,
-which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
+which is achievable in running time $\softO(n^\omega)$ \citelist{\cite{keller-gehrig:85a} \cite{pernet-storjohann:07a}}.
 However, the fact that it uses division frequently leads to catastrophic losses of precision.
@@
 
-Thanks to the application the framework of differential precision
+Thanks to the application of the framework of differential precision
 of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
@@
 whether the characteristic polynomial will have a higher
-precision that the input matrix, and show that the image
+precision than the input matrix, and show that the image
 of the set of integral matrices has the structure of an $\OK[X]$-module
@@
 
-Now suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
+Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
 After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
@@
 Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
-which $P(M) \in M_n(\OK)$, and $V = d \chi_M \big(M_n(\OK)\big)$
-as a submodule of $K[X]/\chi_M(X)$.  Then $V$ is stable under 
+which $P(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
+as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
 multiplication by $A$.
@@
 \begin{proof}
-Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $V$ is given by
-the $\OK$-span of the entries of $C$.   Using the fact that the product of matrix
+Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $\Lambda$ is given by
+the $\OK$-span of the entries of $C$.   Using the fact that the product of a matrix
 with its comatrix is the determinant, $(X - M) \cdot C = \chi_M$ and thus
 $P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
-of the left hand side is precisely $P(X) \cdot V$, while the span of the entries
-of the right hand side is contained within $V$ since $P(M) \in M_n(\OK)$.
+of the left hand side is precisely $P(X) \cdot \Lambda$, while the span of the entries
+of the right hand side is contained within $\Lambda$ since $P(M) \in M_n(\OK)$.
 \end{proof}
@@
 \end{cor}
-
-\begin{comment}
-\begin{prop} \label{prop:factor_com}
-Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
-We assume that $A=\com (X-M)$ is such that $\gcd(A_{1,1}, \chi (M))=1$
-(in $K[X]$).
-Then for $U=col(A,1)$ and $V=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
-both in $\OK[X]^n,$ we have: 
-\[ \com (X-M)= U \cdot {}^t V \mod \chi (M).\]
-\todo{I think that the correct condition is $d \chi_M$ surjective (which is a bit weaker).}
-\end{prop}
-
-\begin{proof}
-We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
-\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
-Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
-Then $A(\lambda)$ is of rank one.
-Indeed, $A(\lambda) \times (\lambda-M) =0.$
-Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
-and then $A(\lambda)$ is of rank at most one.
-Nevertheless, we can remark that
- $\gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
-and thus, $A(\lambda)$ is of rank exactly one.
-As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
-It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
-By the Chinese Remainder Theorem, our first result is then proved.
-
-Now, let us take $U=col(A,1)$ the first column of $A$ and
-$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
-invertible $\mod \chi (M)).$
-Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
-\end{proof}
-
-\begin{rem}
-If $M \in M_n (\OK)$ is such that for some $U,V \in O_{K}[X]^n$
-we have $\com (X-M)= U \cdot {}^t V \mod \chi (M)$ and if 
-$P\in GL_n (\OK)$ then
-\[\com (X-PMP^{-1})=(PU) \cdot ( {}^tV P^{-1}) \mod \chi_M.\]
-In other words, factorizability of $\com(X-M)$ is stable
-by change of basis.
-\end{rem}
-\end{comment}
 
@@
 with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
-By Proposition~\ref{prop:surjectivity}, there exists a matrix 
-$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Applying the same
-result to the transpose of $M$, we find that there exists another
-invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q \mathscr{C} Q^{-1}$.
+Assume one of the equivalent conditions of Proposition~\ref{prop:surjectivity}, giving a matrix 
+$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Then Proposition~\ref{prop:surjectivity}
+also holds for $M^T$, yielding an
+invertible matrix $Q \in \GL_n(K)$ with $M^T = Q \mathscr{C} Q^{-1}$.
 
@@
 We keep the previous notations and assumptions.
-Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
+Let $Y$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
 \begin{equation}
 \label{eq:shortcom}
-\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
+\com(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
 \mod \chi_M
@@
 every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
-$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
-K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
+$\chi_M$ can be written as $W = w \cdot Y^T$ for some $w \in 
+K[X]/\chi_M$. We deduce that $C \equiv P \cdot Y^T B \pmod{\chi_M}$
 for some row vector $B$.
-Applying the same reasoning with $M^t$, we find that $B$ can be
-written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
+Applying the same reasoning with $M^T$, we find that $B$ can be
+written $B = \alpha Y Q^T$ for some $\alpha \in K[X]/\chi_M$ and
 we are done.
@@
 $UMU^{-1}$ is
-$(\alpha, UP, (U^t)^{-1}Q, \chi_M),$
+$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
 which can be computed in $O(n^\omega)$ operations in $K.$
@@
 
-\noindent 8. \ $V:=\psi \times P \times Q   \mod X^{n+1}.$ \footnote{The product $P \times Q$
+\noindent 8. \ $Y:=\psi \times P \times Q   \mod X^{n+1}.$ \footnote{The product $P \times Q$
 should be implemented by sequential row operations corresponding to the eliminations in Step 5
@@
 
-\noindent 9. \ \textbf{Return} $V^{\rec,n-1}, \psi^{\rec,n}.$
+\noindent 9. \ \textbf{Return} $Y^{\rec,n-1}, \psi^{\rec,n}.$
 
@@
 
-\noindent 6. Let $U := \col(C,1)$ and $V := F \cdot \row(C,1) \mod \chi_M$.
-
-\noindent 7. Return $\com(X-M):=(PT^{-1}S U \times V S^{-1} T P^{-1}) \mod \chi_M.$
+\noindent 6. Let $U := \col(C,1)$ and $Y := F \cdot \row(C,1) \mod \chi_M$.
+
+\noindent 7. Return $\com(X-M):=(PT^{-1}S U \times Y S^{-1} T P^{-1}) \mod \chi_M.$
 
@@
 d \chi_M(dM) & =\tr(\com(X{-}M) \cdot dM) \\
-\com(X{-}M) & = \alpha \cdot P V^t \cdot V Q^t \mod \chi_M.
+\com(X{-}M) & = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M.
 \end{align*}
@@
 Now, instead of directly computing $Q$, we first compute a matrix $R$ 
-with the property that $\mathscr{C}^t = R \mathscr{C} R^{-1}$. To do so,
+with the property that $\mathscr{C}^T = R \mathscr{C} R^{-1}$. To do so,
 we apply the same strategy as above except that we start with the vector
@@
 with $n{-}i$ starting zeros. Therefore the $\mathscr{C}^i e$'s form a basis of
-$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $\mathscr{C}$. Once $R$ has
-been computed, we recover $Q$ using the relation $Q = P_\inv^t R$.
+$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $\mathscr{C}$.
+We may then let $R$ have columns $e, \mathscr{C}e, \dots, \mathscr{C}^{n-1}e$
+and recover $Q$ using the relation $Q = P_\inv^T R$.
 
@@
 \label{eq:shortcomN}
-\com(X{-}\mathscr{C}) = \alpha \cdot V^t \cdot V R^t \mod \chi_M
+\com(X{-}\mathscr{C}) = \alpha \cdot Y^T \cdot Y R^T \mod \chi_M
 \end{equation}
@@
 in $K[X]$ such that
-$\com(X{-}N) = \alpha \cdot V^t \cdot V R^t \mod \chi_M$
+$\com(X{-}N) = \alpha \cdot Y^T \cdot Y R^T \mod \chi_M$
 in $\softO(n^\omega)$ operations in $K.$
@@
 
-For $0 \leq k < n$, let $\pi_k : K[X] \to K$ be the mapping taking a 
+For $0 \leq k < n$, let $\beta_k : K[X] \to K$ be the mapping taking a 
 polynomial to its coefficients in $X^k$. By applying 
-\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} to the composite $\pi_k 
+\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} to the composite $\beta_k 
 \circ \chi_M$, one can figure out the optimal precision on the
@@
 \label{eq:optjagged}
-N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j})),
+N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\beta_k(C_{i,j})),
 \end{equation}
@@
 $k$-th coefficient of $M$ is also equal to
-$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
+$N + \min_{1 \leq i, j\leq n} \val(\beta_k(C^H_{i,j}))$
 where $C^H_{i,j}$ is the $(i,j)$ entry of $\com(X{-}H)$,
@@
 explained above with the actual precision obtained by the software.
-For doing so, we picked a sample of $1000$ random matrices $M$ in 
-$M_9(\Q_2)$ where all the entries are given at the same \emph{relative
-precision}.
-We recall that, in \textsc{SageMath}, random elements $x \in \Q_p$ are 
-generated as follows. We fix an integer $N$ --- the so-called relative
-precision --- and generate elements of $\Q_p$ of the shape
-$$x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)$$
+For doing so, we picked a sample of $1000$ matrices $M$ in 
+$M_9(\Q_2)$ where all the entries are chosen randomly as follows.
+We fix an integer $N$, the relative precision, and generate elements of $\Q_p$ of the shape
+\[
+x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)
+\]
 where $v$ is a random integer generated according to the distribution:
-$$\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
+\[
+\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
 \mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
-|n| \geq 1$$
+|n| \geq 1
+\]
 and $a$ is an integer in the range $[0, p^N)$, selected uniformly at random.
+This distribution is chosen merely because it is the default in \textsc{SageMath}.
 
@@
 Writing
-$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t \text{ mod } \chi_M$
+$\com(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \text{ mod } \chi_M$
 as in Proposition~\ref{prop:shortcom}, we find:
@@
 & \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
-\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big) \label{eq:Nprime}
+\val(P_i Y(\lambda)^T) + \val(Y(\lambda) Q_j^T)\big) \label{eq:Nprime}
 \end{align}
 where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
-denotes the $j$-th row of $Q$. Note moreover that $V(\lambda)$ is
+denotes the $j$-th row of $Q$. Note moreover that $Y(\lambda)$ is
 the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
@@
 N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \nonumber \\
-& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
-+ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big) \label{eq:Nprimeflat}
+& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i Y(\lambda)^T)
++ \min_{1 \leq j \leq n} \val(Y(\lambda) Q_j^T)\big) \label{eq:Nprimeflat}
 \end{align}
@@
 & \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
-\val(P_i V(\lambda_k)^t) + \val(V(\lambda_k) Q_j^t)\big).
+\val(P_i Y(\lambda_k)^T) + \val(Y(\lambda_k) Q_j^T)\big).
 \end{align*}
@@
 evaluation methods (see 10.7 of \cite{gathen-gerhard:13a}).
-On the other hand, we observe that $P_i V(\lambda_k)^t$ is nothing but
+On the other hand, we observe that $P_i Y(\lambda_k)^T$ is nothing but
 the $(i,k)$ entry of the matrix:
@@
 points $\lambda_1, \ldots, \lambda_s$.}. Therefore all the $P_i 
-V(\lambda_k)^t$'s (for $i$ and $k$ varying) 
+Y(\lambda_k)^T$'s (for $i$ and $k$ varying) 
 can be determined with the same complexity. 
-Similarly all the $V(\lambda) Q_j^t$ are computed for the same cost.
+Similarly all the $Y(\lambda) Q_j^T$ are computed for the same cost.
 The first assertion of Proposition~\ref{prop:opteigenvalues} follows.
@@
 \end{document}
+
+\begin{comment}
+\begin{prop} \label{prop:factor_com}
+Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
+We assume that $A=\com (X-M)$ is such that $\gcd(A_{1,1}, \chi (M))=1$
+(in $K[X]$).
+Then for $U=col(A,1)$ and $\Lambda=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
+both in $\OK[X]^n,$ we have: 
+\[ \com (X-M)= U \cdot ^T \mod \chi (M).\]
+\todo{I think that the correct condition is $d \chi_M$ surjective (which is a bit weaker).}
+\end{prop}
+
+\begin{proof}
+We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
+\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
+Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
+Then $A(\lambda)$ is of rank one.
+Indeed, $A(\lambda) \times (\lambda-M) =0.$
+Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
+and then $A(\lambda)$ is of rank at most one.
+Nevertheless, we can remark that
+ $\gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
+and thus, $A(\lambda)$ is of rank exactly one.
+As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
+It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
+By the Chinese Remainder Theorem, our first result is then proved.
+
+Now, let us take $U=col(A,1)$ the first column of $A$ and
+$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
+invertible $\mod \chi (M)).$
+Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
+\end{proof}
+
+\begin{rem}
+If $M \in M_n (\OK)$ is such that for some $U,V \in O_{K}[X]^n$
+we have $\com (X-M)= U \cdot V^T \mod \chi (M)$ and if 
+$P\in GL_n (\OK)$ then
+\[\com (X-PMP^{-1})=(PU) \cdot ( V^T P^{-1}) \mod \chi_M.\]
+In other words, factorizability of $\com(X-M)$ is stable
+by change of basis.
+\end{rem}
+\end{comment}

### 2017-04-15 charpoly/charpoly.tex
Notation for precision uniformized

--- 2017-02-04 charpoly/charpoly.tex
@@
 an $n \times n$ $p$-adic matrix $A$ using differential precision methods
-developed previously.  When $A$ is integral with precision $O(p^k)$,
+developed previously.  When $A$ is integral with precision $O(p^N)$,
 we give a criterion (checkable in time $\softO(n^\omega)$) for
-$\chi(A)$ to have precision exactly $O(p^k)$.  We also give a $\softO(n^3)$
+$\chi(A)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
 algorithm for determining the optimal precision when the criterion is not
-satisfied, and give examples when the precision is larger than $O(p^k)$.
+satisfied, and give examples when the precision is larger than $O(p^N)$.
 \end{abstract}
@@
 \vspace{-1.5mm}
-\keywords{Algorithms, $p$-adic precision, Newton polygon, factorization}
+\keywords{Algorithms, $p$-adic precision, characteristic polynomial,
+eigenvalue}
 
@@
 When computing with $p$-adic matrices, the lack of infinite memory implies
-that the entries may only be approximated at some finite precision $O(p^m)$.  As a
+that the entries may only be approximated at some finite precision $O(p^N)$.  As a
 consequence, in designing algorithms for such matrices one must analyze
@@
 
-Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
+Let $M \in M_n(\Qp)$ be known at precision $O(p^N)$.
 The simplest approach for computing the characteristic polynomial of $M$
@@
 requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
-while the lack of division implies that the result is accurate modulo $p^m$ as
+while the lack of division implies that the result is accurate modulo $p^N$ as
 long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.
@@
 of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $\alpha_{n-1} > 0$
-and $M$ is known with precision $O(p^m)$ then the constant term of $\chi_M$
-will be known with precision larger than $O(p^m)$ (see
+and $M$ is known with precision $O(p^N)$ then the constant term of $\chi_M$
+will be known with precision larger than $O(p^N)$ (see
 \cite{caruso-roe-vaccon:15a}*{Prop. 3.2}).
@@
 then none of the coefficients of the characteristic polynomial of $1+M$ will have
-precision larger than $O(p^m)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
+precision larger than $O(p^N)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
 content of these two polynomials should be equivalent. The solution is that the
@@
 where we specify a precision for each matrix entry or \emph{flat} lattices where
-every entry is known to a fixed precision $O(p^m)$.  Similarly, precision for
+every entry is known to a fixed precision $O(p^N)$.  Similarly, precision for
 monic polynomials can be specified by giving a lattice in the tangent space

### 2017-04-16 charpoly/charpoly.tex
Comments on referees' report

@@
 \newcommand{\tdo}[1]{\textcolor{red}{#1}}
+\definecolor{answer}{rgb}{0,0.5,0.2}
+\newcommand{\xavier}[1]{\textcolor{answer}{{\bf Xavier:} #1}}
+\newcommand{\tristan}[1]{\textcolor{answer}{{\bf Tristan:} #1}}
+\newcommand{\david}[1]{\textcolor{answer}{{\bf David:} #1}}
 
@@
 \item \tdo{What are the implications, if any, of your work on well-known methods such as Dixon's algorithm?}
-\item \tdo{In section 3, Algorithm 1: what is the cost of step 5, and how did you avoid quartic complexity?}
-\item \tdo{In section 4, give a concrete construction for approximating the characteristic polynomial as a ``by product'' of the work of [15].}
+\item \tdo{In section 3, Algorithm 1: what is the cost of step 5, and how did you avoid quartic complexity?}\\
+      \xavier{I think that the cost is $O(n^2)$ because it corresponds to an elementary operation of rows/columns. Is that correct?}
+\item \tdo{In section 4, give a concrete construction for approximating the characteristic polynomial as a ``by product'' of the work of [15].}\\
+      \xavier{I do not really understand this remark. I mean: once we have the Frobenius invariants, the characteristic polynomial is just the product of them, isn't it?}
 \item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
 \item \done{We clarified why we chose the distribution in Figure 1.}
-\item \tdo{Think about whether Proposition 5.6 can be sped up using the work of Kedlaya and Umans.}
+\item \tdo{Think about whether Proposition 5.6 can be sped up using the work of Kedlaya and Umans.}\\
+      \xavier{I cannot see how Kedlaya--Umans can help: the cost $\softO(n^\omega)$ comes from the computation of a Frobenius normal form...}
 \end{enumerate}
@@
 \begin{enumerate}
-\item \tdo{Convince reviewer that the complexity claimed in Theorem 4.1 is accurate: ``it lacks a precision loss analysis and relies on classic division based algorithms}
+\item \tdo{Convince reviewer that the complexity claimed in Theorem 4.1 is accurate: ``it lacks a precision loss analysis and relies on classic division based algorithms}\\
+      \xavier{One can argue that we do not need \emph{a priori} much accuracy on the precision of $\com(A-X)$ because only the valuations of the entries of the latter matrix are relevant.
+So counting operations in $K$ makes sense.
+By the way, we could say at some point that a good idea would be to use relaxed algorithms and increase the precision until we are able to read the valuations.}
 \item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
 \item \tdo{The definition of comatrix is usually the matrix formed by the cofactors such that $A^{-1} = \com(A)^T / \det(A)$.}
-\item \tdo{The definitions of companion matrix in the proof of Proposition 2.1 and in Section 2.3 are inconsistent (coefficients on the last column or last row)}
+      \xavier{Ok pour switching $\com$ to $\text{Adj}$ (and adding a sentence explaining the difference with the previous paper).}
+\item \tdo{The definitions of companion matrix in the proof of Proposition 2.1 and in Section 2.3 are inconsistent (coefficients on the last column or last row)}\\
+      \xavier{Is that enough to replace ``the bottom row'' by ``the last column'' in the proof of Proposition 2.1?}
 \item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
@@
 with the fact that we do degree one approx of a charpoly)}
-\item \tdo{In the experiments, (section 5.1), it is not clear which of the two methods is being used to compute the comatrix.}
+      \xavier{FWIW, I wrote a proof in my lecture at JNCF. Maybe we can just cite it.}
+\item \tdo{In the experiments, (section 5.1), it is not clear which of the two methods is being used to compute the comatrix.}\\
+      \xavier{Ok but it is not important (since we are just interested in the result).
+Should we really say more about that?
+(In practice, I think that we used the \texttt{adjoint} method of \textsc{sage}.)}
 \item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing com(A) differ), would be welcome.}
 \item \tdo{Clarify the intro paragraph just before the Previous contributions section.}
-\item \tdo{Make the relationship between $P$, $Q$ and $R$ in Theorem 4.1 explicit.}
+\item \tdo{Make the relationship between $P$, $Q$ and $R$ in Theorem 4.1 explicit.}\\
+      \xavier{\done{Done.} I've rewritten everything in terms of $P$ and $Q$ only.}
 \item \done{We have carried out the other minor corrections suggested.}
@@
 \item \tdo{In the abstract, clarify what kind of operations are being counted.}
-\item \tdo{In the abstract, clarify what we mean by ``precision exactly $O(p^k)$.''}
-\item \tdo{In the introduction, we say ``Let $M$ be known at precision $O(p^m)$.''  What does this mean?  Coefficient-wise?  Overall?  This needs to be said clearly.}
+\item \tdo{In the abstract, clarify what we mean by ``precision exactly $O(p^k)$.''}\\
+      \xavier{By the way, we should also mention in the abstract the $\softO(n^\omega)$ algorithm.}
+\item \tdo{In the introduction, we say ``Let $M$ be known at precision $O(p^m)$.''  What does this mean?  Coefficient-wise?  Overall?  This needs to be said clearly.}\\
+      \xavier{Well, it is the same, isn't it? Anyway, we may add ``coefficient-wise'' if it's not clear.}
 \item \tdo{In the introduction, clarify what kind of operations are being counted.}
@@
 \item \tdo{Define comatrix properly, as requested by reviewer 3.}
-\item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}
+\item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}\\
+      \xavier{Maybe we should instead remove any mention to the tangent space: it's not really necessary.}
 \item \tdo{The definition of ``flat precision'' and ``jagged precision'' both need to be more clearly defined.}
@@
 \end{enumerate}
+\xavier{I propose to expand the first paragraph of Section 2.1 (and even
+maybe make a separate subsection for it). Concretely, I suggest to 
+(1)~do not talk about the tangent space (but instead consider lattices
+in $M_n(K)$),
+(2)~give precision definition of flat, jagged precisions and relate them 
+with lattices,
+(3)~maybe cite the notes of my recent lecture at JNCF in which all of
+this is detailed.}
+
+\bigskip
 
@@
 in $K[X]$ such that
-$\com(X{-}N) = \alpha \cdot Y^T \cdot Y R^T \mod \chi_M$
+$\com(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$
 in $\softO(n^\omega)$ operations in $K.$

### 2017-04-16 charpoly/charpoly.tex
Switch comatrix to adjugate and adjust notation

@@
 \DeclareMathOperator{\tr}{Tr}
-\DeclareMathOperator{\com}{Com}
+\DeclareMathOperator{\adj}{Adj}
 \DeclareMathOperator{\Grass}{Grass}
@@
 \newcommand{\Fp}{\mathbb{F}_p}
+\newcommand{\Fq}{\mathbb{F}_q}
 \newcommand{\R}{\mathbb R}
@@
 \begin{abstract}
-We analyze the precision of the characteristic polynomial $\chi(A)$ of
-an $n \times n$ $p$-adic matrix $A$ using differential precision methods
-developed previously.  When $A$ is integral with precision $O(p^N)$,
+We analyze the precision of the characteristic polynomial $\chi(M)$ of
+an $n \times n$ $p$-adic matrix $M$ using differential precision methods
+developed previously.  When $M$ is integral with precision $O(p^N)$,
 we give a criterion (checkable in time $\softO(n^\omega)$) for
-$\chi(A)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
+$\chi(M)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
 algorithm for determining the optimal precision when the criterion is not
@@
 \item \tdo{Convince reviewer that the complexity claimed in Theorem 4.1 is accurate: ``it lacks a precision loss analysis and relies on classic division based algorithms}\\
-      \xavier{One can argue that we do not need \emph{a priori} much accuracy on the precision of $\com(A-X)$ because only the valuations of the entries of the latter matrix are relevant.
+      \xavier{One can argue that we do not need \emph{a priori} much accuracy on the precision of $\adj(M-X)$ because only the valuations of the entries of the latter matrix are relevant.
 So counting operations in $K$ makes sense.
@@
 \item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
-\item \tdo{The definition of comatrix is usually the matrix formed by the cofactors such that $A^{-1} = \com(A)^T / \det(A)$.}
-      \xavier{Ok pour switching $\com$ to $\text{Adj}$ (and adding a sentence explaining the difference with the previous paper).}
+\item \done{We have switched to using the term ``adjugate'' rather than ``comatrix.''}
 \item \tdo{The definitions of companion matrix in the proof of Proposition 2.1 and in Section 2.3 are inconsistent (coefficients on the last column or last row)}\\
@@
 \item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
-\item \done{We have switched from $A^t$ to $A^T$ to denote the transpose.}
+\item \done{We have switched from $M^t$ to $M^T$ to denote the transpose.}
 \item \tdo{A few explanations on where does the formula for the differential comes from
@@
 (In practice, I think that we used the \texttt{adjoint} method of \textsc{sage}.)}
-\item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing com(A) differ), would be welcome.}
+\item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing adj(M) differ), would be welcome.}
 \item \tdo{Clarify the intro paragraph just before the Previous contributions section.}
@@
 \item \tdo{In the discussion of existing methods, clarify what we mean by the claim that they do not yield optimal precision.}
-\item \tdo{Define comatrix properly, as requested by reviewer 3.}
+\item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
 \item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}\\
@@
 $\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
-is determined by the comatrix $\com(X -M).$
+is determined by the adjugate $\adj(X -M).$
 In this article, we provide:
@@
 \renewcommand{\itemsep}{0pt}
-\item Proposition \ref{prop:shortcom}: a factorization of $\com(X -M)$
+\item Proposition \ref{prop:shortcom}: a factorization of $\adj(X -M)$
 as a product of two rank-$1$ matrices (when $M$ has a cyclic vector),
@@
 when $M$ is itself integral. Finally, we give a compact description
-of $\com(X-M)$, the main ingredient in the differential.
+of $\adj(X-M)$, the main ingredient in the differential.
 
 In Section \ref{sec:diffHess}, we develop $\softO(n^3)$ algorithms to approximate
-the Hessenberg form of $M$, and through it to find $\com(X-M)$ and thus
+the Hessenberg form of $M$, and through it to find $\adj(X-M)$ and thus
 find the precision of the characteristic polynomial of $M$.  In Section \ref{sec:diffFrob},
-we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\com(X-M)$.
+we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\adj(X-M)$.
 
@@
 of polynomials of degree less than $n$.
-We fix an $\omega \in \mathbb{R}$ such that the multiplication
+We fix an $\omega \in \R$ such that the multiplication
 of two matrices over a ring is in $O(n^\omega)$ operations in the
@@
 the space $K_{<n}[X]$ of polynomials of degree less than $n$.
-Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
-we have $\com(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
+Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
+we have $\adj(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
 \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
@@
 \begin{equation} \label{eq:dchi}
-d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
+d\chi_M: dM \mapsto \tr(\adj(X-M) \cdot dM).
 \end{equation}
@@
 
-For any $A \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
-as the image of $d\chi$ at $AMA^{-1}$, so we may assume that
+For any $P \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
+as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
 $M$ is a companion matrix.  For a companion matrix, the bottom row of
-$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
+$\adj(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
 
 Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
-After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
+After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X-M)$
 will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
@@
 Since $\lambda$ occurs in two Jordan blocks,
-$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
+$X - \lambda$ will divide every entry of $\adj(X-M)$ and $d\chi_M$ will not be surjective.
 \end{proof}
@@
 Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
-of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
+of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
 By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
@@
 \begin{prop}
-Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
-which $P(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
+Let $R$ be the subring of $K[X]$ consisting of polynomials $f$ for
+which $f(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
 as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
-multiplication by $A$.
+multiplication by $R$.
 \end{prop}
 \begin{proof}
-Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $\Lambda$ is given by
-the $\OK$-span of the entries of $C$.   Using the fact that the product of a matrix
-with its comatrix is the determinant, $(X - M) \cdot C = \chi_M$ and thus
-$P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
-of the left hand side is precisely $P(X) \cdot \Lambda$, while the span of the entries
-of the right hand side is contained within $\Lambda$ since $P(M) \in M_n(\OK)$.
+Let $A = \adj(X-M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
+the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
+with its adjugate is the determinant, $(X - M) \cdot A = \chi_M$ and thus
+$f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
+of the left hand side is precisely $f(X) \cdot \Lambda$, while the span of the entries
+of the right hand side is contained within $\Lambda$ since $f(M) \in M_n(\OK)$.
 \end{proof}
@@
 \label{eq:shortcom}
-\com(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
+\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
 \mod \chi_M
@@
 \begin{proof}
-Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
-\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
-Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}\mathscr{C}$
+Write $A = \adj(X{-}M)$. From $(X{-}M) \cdot A \equiv 0 
+\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} A \equiv 0 \pmod{\chi_M}$. 
+Therefore each column of $P^{-1} A$ lies in the right kernel of $X{-}\mathscr{C}$
 modulo $\chi_M$. On the other hand, a direct computation shows that
@@
 $\chi_M$ can be written as $W = w \cdot Y^T$ for some $w \in 
-K[X]/\chi_M$. We deduce that $C \equiv P \cdot Y^T B \pmod{\chi_M}$
+K[X]/\chi_M$. We deduce that $A \equiv P \cdot Y^T B \pmod{\chi_M}$
 for some row vector $B$.
@@
 
-Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
+Proposition~\ref{prop:shortcom} shows that $\adj(X{-}M)$ can be encoded 
 by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
@@
 which can be computed in $O(n^\omega)$ operations in $K.$
-This is faster than computing $U \com(X-M) U^{-1},$
+This is faster than computing $U \adj(X-M) U^{-1},$
 which is, at first sight, in $O(n^4)$ operations in $K.$
@@
 over a complete discrete valuation field (CDVF)
-to compute $\com(X-M)$ and $d \chi$.
+to compute $\adj(X-M)$ and $d \chi$.
 If $M \in M_n(\OK)$, then only division by invertible
@@
 with the $\varepsilon$'s being units,
-the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\mathbb{F}_q[X],$ $\mathbb{Q}[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
+the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
 Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
@@
 
-\subsection{The comatrix of $X{-}H$}
-
-In this section, we compute $\com(X-H)$ for a Hessenberg matrix $H$
+\subsection{The adjugate of $X{-}H$}
+
+In this section, we compute $\adj(X-H)$ for a Hessenberg matrix $H$
 using the Smith normal form computation of the previous section.
-The entries of $\com(X-H)$ lie in $K[X]$, which is not a CDVF,
+The entries of $\adj(X-H)$ lie in $K[X]$, which is not a CDVF,
 so we may not directly apply the methods of the previous section.
-However, we may relate $\com(X-H)$ to $\com(1-XH)$, whose
-entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\com(X-H)$
+However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
+entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
 using an SNF method, with no division in $K$.
@@
 
-\begin{lem} \label{lem:comatrix_of_similar}
+\begin{lem} \label{lem:adjugate_of_similar}
 If $M_1,M_2 \in M_n(K)$ and $P \in GL_n (K)$ are such that
 $M_1=PM_2P^{-1},$ then:
-\[ \com (X-M_1)=P \com (X-M_2) P^{-1}. \] 
+\[ \adj (X-M_1)=P \adj (X-M_2) P^{-1}. \] 
 \end{lem}
@@
 \begin{deftn}
-Let $d \in \mathbb{N}$ and $P \in K[X]$ of degree at most $d.$ 
-We define the reciprocal polynomial of order $d$ of $P$ as $P^{\rec,d}=X^d P \left( 1/X \right).$
-Let $A \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
-We denote by $A^{\rec,d}$ the matrix with $(A^{\rec,d})_{i,j} = (A_{i,j})^{\rec,d}$.
+Let $d \in \mathbb{N}$ and $f \in K[X]$ of degree at most $d.$ 
+We define the reciprocal polynomial of order $d$ of $f$ as $f^{\rec,d}=X^d f \left( 1/X \right).$
+Let $F \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
+We denote by $F^{\rec,d}$ the matrix with $(F^{\rec,d})_{i,j} = (F_{i,j})^{\rec,d}$.
 \end{deftn}
@@
 \begin{eqnarray*}
-\com(1-XM)^{\rec,n-1}=&\com(X-M), \\
-(\chi_M I_n)^{\rec,n}=&(1-XM) \com(1-XM).\\
+\adj(1-XM)^{\rec,n-1}=&\adj(X-M), \\
+(\chi_M I_n)^{\rec,n}=&(1-XM) \adj(1-XM).\\
 \end{eqnarray*}
@@
 It all comes down to the following result:
-let $A \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
-then $\det (A^{\rec,1})=\det(A)^{\rec,d}.$
-Indeed, one can use multilinearity of the determinant on $X^d \det(A(1/X))$
+let $F \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
+then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
+Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$
 to prove this result.
 It directly implies the second part of the lemma; the first part follows
-from the fact that the entries of $\com(X-M)$ and of $\com(1-XM)$
+from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
 are determinants of size $n-1$.
@@
 
-This lemma allows us to compute $\com(1-XM)$ instead of $\com(X-M).$
+This lemma allows us to compute $\adj(1-XM)$ instead of $\adj(X-M).$
 This has a remarkable advantage: the pivots during the computation of
-the SNF of $\com(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
+the SNF of $\adj(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
 in advance to be on the diagonal. This leads to a very smooth
@@
 
-\noindent {\bf Algorithm 2:} {\tt Approximate $\com (X -H)$ }
+\noindent {\bf Algorithm 2:} {\tt Approximate $\adj (X -H)$ }
 
@@
 Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix.
-Then, using Algorithm 2, one can compute $\com (X -H)$
+Then, using Algorithm 2, one can compute $\adj (X -H)$
 and $\chi_H$ in 
@@
 
-\subsection{The comatrix of $X{-}M$}
+\subsection{The adjugate of $X{-}M$}
 
 In this section, we combine Proposition \ref{prop:shortcom}
-with Algorithm 2 to compute the comatrix of $X-M$
+with Algorithm 2 to compute the adjugate of $X-M$
 when $\chi_M$ is squarefree.   Note that this condition on
@@
 
-\noindent {\bf Algorithm 3:} {\tt Approximate $\com (X{-}M)$ }
+\noindent {\bf Algorithm 3:} {\tt Approximate $\adj (X{-}M)$ }
 
@@
 
-\noindent 1.\ Compute $A=\com (X-H)$ and $\chi_M = \chi_H$ using Algorithm 2.
+\noindent 1.\ Compute $A=\adj (X-H)$ and $\chi_M = \chi_H$ using Algorithm 2.
 
@@
 
-\noindent 7. Return $\com(X-M):=(PT^{-1}S U \times Y S^{-1} T P^{-1}) \mod \chi_M.$
+\noindent 7. Return $\adj(X-M):=(PT^{-1}S U \times Y S^{-1} T P^{-1}) \mod \chi_M.$
 
@@
 Algorithm 3 computes
-$\com (X-M) \pmod{\chi_M}$ in average complexity $\softO (n^3)$
+$\adj (X-M) \pmod{\chi_M}$ in average complexity $\softO (n^3)$
 operations in $K$.
@@
 \begin{align*}
-d \chi_M(dM) & =\tr(\com(X{-}M) \cdot dM) \\
-\com(X{-}M) & = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M.
+d \chi_M(dM) & =\tr(\adj(X{-}M) \cdot dM) \\
+\adj(X{-}M) & = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M.
 \end{align*}
@@
 \label{eq:shortcomN}
-\com(X{-}\mathscr{C}) = \alpha \cdot Y^T \cdot Y R^T \mod \chi_M
+\adj(X{-}\mathscr{C}) = \alpha \cdot Y^T \cdot Y R^T \mod \chi_M
 \end{equation}
@@
 in $K[X]$ such that
-$\com(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$
+$\adj(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$
 in $\softO(n^\omega)$ operations in $K.$
@@
 In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffFrob},
-we have proposed algorithms to obtain the comatrix of 
+we have proposed algorithms to obtain the adjugate of 
 $X-M.$ Our motivation for these computations is to then
@@
 \label{eq:optjagged}
-N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\beta_k(C_{i,j})),
+N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\beta_k(A_{i,j})),
 \end{equation}
-where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.
+where $A_{i,j}$ is the $(i,j)$ entry of the adjugate $\adj(X{-}M)$.
 
@@
 We have seen in \S \ref{sec:diffHess} and \S \ref{sec:diffFrob}
-that the computation of the matrix $C = \com(X{-}M)$ can be carried out 
+that the computation of the matrix $A = \adj(X{-}M)$ can be carried out 
 within $\softO(n^3)$ operations in $K$ (either with the Hessenberg 
@@
 then we can avoid the final base change step in the Hessenberg method.
-Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
+Indeed, observe that, thanks to Lemma \ref{lem:adjugate_of_similar}, 
 we can write:
-$$\tr(\com (X{-}M) \cdot dM)=\tr(\com (X{-}H)\cdot P^{-1} dM P)$$
+$$\tr(\adj (X{-}M) \cdot dM)=\tr(\adj (X{-}H)\cdot P^{-1} dM P)$$
 where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
@@
 $k$-th coefficient of $M$ is also equal to
-$N + \min_{1 \leq i, j\leq n} \val(\beta_k(C^H_{i,j}))$
-where $C^H_{i,j}$ is the $(i,j)$ entry of $\com(X{-}H)$,
+$N + \min_{1 \leq i, j\leq n} \val(\beta_k(A^H_{i,j}))$
+where $A^H_{i,j}$ is the $(i,j)$ entry of $\adj(X{-}H)$,
 where $H$ is the Hessenberg form of $M$.
@@
 The differential of $f$ at $M$ is the linear mapping:
-$$dM \mapsto d \lambda = - \frac{\tr(\com(\lambda-M) \cdot dM)}
+$$dM \mapsto d \lambda = - \frac{\tr(\adj(\lambda-M) \cdot dM)}
 {\chi'_M(\lambda)}$$
@@
 Differentiating the relation $\chi_M(\lambda) = 0$, we get
-$\chi'_M(\lambda) \cdot d \lambda + \tr(\com(X-M) \cdot dM)(\lambda) = 0$,
+$\chi'_M(\lambda) \cdot d \lambda + \tr(\adj(X-M) \cdot dM)(\lambda) = 0$,
 from which the Lemma follows.
@@
 precision on the eigenvalue $\lambda$ is $O(\pi^{N'})$ with:
-$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(C_{i,j}(\lambda)) - 
+$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(A_{i,j}(\lambda)) - 
 \val(\chi'_M(\lambda))\big)$$
-where $C_{i,j}$ denotes as above the $(i,j)$ entry of $\com(X{-}M)$.
+where $A_{i,j}$ denotes as above the $(i,j)$ entry of $\adj(X{-}M)$.
 Writing
-$\com(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \text{ mod } \chi_M$
+$\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \text{ mod } \chi_M$
 as in Proposition~\ref{prop:shortcom}, we find:
@@
 Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
-We assume that $A=\com (X-M)$ is such that $\gcd(A_{1,1}, \chi (M))=1$
+We assume that $A=\adj (X-M)$ is such that $\gcd(A_{1,1}, \chi (M))=1$
 (in $K[X]$).
@@
 both in $\OK[X]^n,$ we have: 
-\[ \com (X-M)= U \cdot ^T \mod \chi (M).\]
+\[ \adj (X-M)= U \cdot ^T \mod \chi (M).\]
 \todo{I think that the correct condition is $d \chi_M$ surjective (which is a bit weaker).}
@@
 If $M \in M_n (\OK)$ is such that for some $U,V \in O_{K}[X]^n$
-we have $\com (X-M)= U \cdot V^T \mod \chi (M)$ and if 
+we have $\adj (X-M)= U \cdot V^T \mod \chi (M)$ and if 
 $P\in GL_n (\OK)$ then
-\[\com (X-PMP^{-1})=(PU) \cdot ( V^T P^{-1}) \mod \chi_M.\]
-In other words, factorizability of $\com(X-M)$ is stable
+\[\adj (X-PMP^{-1})=(PU) \cdot ( V^T P^{-1}) \mod \chi_M.\]
+In other words, factorizability of $\adj(X-M)$ is stable
 by change of basis.

### 2017-04-16 charpoly/charpoly.tex
Fix companion matrix convention

@@
 \item \done{We have switched to using the term ``adjugate'' rather than ``comatrix.''}
-\item \tdo{The definitions of companion matrix in the proof of Proposition 2.1 and in Section 2.3 are inconsistent (coefficients on the last column or last row)}\\
-      \xavier{Is that enough to replace ``the bottom row'' by ``the last column'' in the proof of Proposition 2.1?}
+\item \done{We have given a definition of the companion matrix in our notation section and adjusted the proof of Proposition 2.1 accordingly.}
 \item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
@@
 and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n - M.$
-Finally, we write $\sigma_1(M), \dots, \sigma_n(M)$ for the
+We write $\sigma_1(M), \dots, \sigma_n(M)$ for the
 elementary divisors of $M$, sorted in increasing order of valuation.
+Finally, given a polynomial $\chi(X) = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$,
+the \emph{companion matrix} $\mathscr{C}$ associated to $\chi$ is
+\begin{equation}
+\label{eq:companion}
+\mathscr{C} = \left( \begin{matrix}
+0 & 1 & 0 & \cdots & 0 \\
+0 & 0 & 1 & \ddots & \vdots \\
+\vdots & \vdots & \ddots & \ddots & 0 \\
+0 & 0 & \cdots & 0 & 1 \\
+-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
+\end{matrix} \right).
+\end{equation}
 
@@
 as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
-$M$ is a companion matrix.  For a companion matrix, the bottom row of
+$M$ is a companion matrix.  For a companion matrix, the last column of
 $\adj(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
@@
 
-Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$:
-\begin{equation}
-\label{eq:companion}
-\mathscr{C} = \left( \begin{matrix}
-0 & 1 & 0 & \cdots & 0 \\
-0 & 0 & 1 & \ddots & \vdots \\
-\vdots & \vdots & \ddots & \ddots & 0 \\
-0 & 0 & \cdots & 0 & 1 \\
--a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
-\end{matrix} \right)
-\end{equation}
-with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
+Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$ as in \eqref{eq:companion}.
 Assume one of the equivalent conditions of Proposition~\ref{prop:surjectivity}, giving a matrix 

### 2017-04-16 charpoly/charpoly.tex
Adding some responses to the reviewer notes

@@
 \item \tdo{In section 3, Algorithm 1: what is the cost of step 5, and how did you avoid quartic complexity?}\\
-      \xavier{I think that the cost is $O(n^2)$ because it corresponds to an elementary operation of rows/columns. Is that correct?}
+      \xavier{I think that the cost is $O(n^2)$ because it corresponds to an elementary operation of rows/columns. Is that correct?} \\
+      \david{I agree.  I think we just need to explain that the matrix products are just notational, not actually computed.}
 \item \tdo{In section 4, give a concrete construction for approximating the characteristic polynomial as a ``by product'' of the work of [15].}\\
-      \xavier{I do not really understand this remark. I mean: once we have the Frobenius invariants, the characteristic polynomial is just the product of them, isn't it?}
+      \xavier{I do not really understand this remark. I mean: once we have the Frobenius invariants, the characteristic polynomial is just the product of them, isn't it?}\\
+      \david{Agreed.  I think I can clarify the language somewhat, but there's no real work to be done.}
 \item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
@@
 with the fact that we do degree one approx of a charpoly)}
-      \xavier{FWIW, I wrote a proof in my lecture at JNCF. Maybe we can just cite it.}
+      \xavier{FWIW, I wrote a proof in my lecture at JNCF. Maybe we can just cite it.}\\
+      \david{Citing it seems reasonable.}
 \item \tdo{In the experiments, (section 5.1), it is not clear which of the two methods is being used to compute the comatrix.}\\
@@
 Should we really say more about that?
-(In practice, I think that we used the \texttt{adjoint} method of \textsc{sage}.)}
+(In practice, I think that we used the \texttt{adjoint} method of \textsc{sage}.)}\\
+      \david{I had the same reaction when I first saw the results.  I think we can say something to make clearer that it doesn't matter how it is computed.}
 \item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing adj(M) differ), would be welcome.}
 \item \tdo{Clarify the intro paragraph just before the Previous contributions section.}
-\item \tdo{Make the relationship between $P$, $Q$ and $R$ in Theorem 4.1 explicit.}\\
-      \xavier{\done{Done.} I've rewritten everything in terms of $P$ and $Q$ only.}
 \item \done{We have carried out the other minor corrections suggested.}
@@
 \item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}\\
-      \xavier{Maybe we should instead remove any mention to the tangent space: it's not really necessary.}
+      \xavier{Maybe we should instead remove any mention to the tangent space: it's not really necessary.}\\
+      \david{That's probably a good idea.  We'll have to make a remark about ignoring the leading monic term.}
 \item \tdo{The definition of ``flat precision'' and ``jagged precision'' both need to be more clearly defined.}

### 2017-04-17 charpoly/charpoly.tex
Addressed some more reviewer suggestions

@@
 \begin{enumerate}
-\item \tdo{What are the implications, if any, of your work on well-known methods such as Dixon's algorithm?}
-\item \tdo{In section 3, Algorithm 1: what is the cost of step 5, and how did you avoid quartic complexity?}\\
-      \xavier{I think that the cost is $O(n^2)$ because it corresponds to an elementary operation of rows/columns. Is that correct?} \\
-      \david{I agree.  I think we just need to explain that the matrix products are just notational, not actually computed.}
-\item \tdo{In section 4, give a concrete construction for approximating the characteristic polynomial as a ``by product'' of the work of [15].}\\
-      \xavier{I do not really understand this remark. I mean: once we have the Frobenius invariants, the characteristic polynomial is just the product of them, isn't it?}\\
-      \david{Agreed.  I think I can clarify the language somewhat, but there's no real work to be done.}
+\item \tdo{What are the implications, if any, of your work on well-known methods such as Dixon's algorithm?}\\
+     \david{Do either of you know what Dixon's algorithm refers to in this context?}
+\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time} \\
+      \david{Tristan, can you check that my clarification before Proposition 3.2 is correct?}
+\item \done{In section 4, we clarified how the characteristic polynomial is a by-product of [15].}
 \item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
@@
 would be most welcome (those of [3 Appendix B] and some connection of the trace
-with the fact that we do degree one approx of a charpoly)}
+with the fact that we do degree one approx of a charpoly)}\\
       \xavier{FWIW, I wrote a proof in my lecture at JNCF. Maybe we can just cite it.}\\
       \david{Citing it seems reasonable.}
-\item \tdo{In the experiments, (section 5.1), it is not clear which of the two methods is being used to compute the comatrix.}\\
-      \xavier{Ok but it is not important (since we are just interested in the result).
-Should we really say more about that?
-(In practice, I think that we used the \texttt{adjoint} method of \textsc{sage}.)}\\
-      \david{I had the same reaction when I first saw the results.  I think we can say something to make clearer that it doesn't matter how it is computed.}
+\item \done{In section 5.1, we added a remark noting that the method for computing the adjugate matrix has no effect on the extent of the precision loss, just on the runtime.}
 \item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing adj(M) differ), would be welcome.}
@@
 
+Note that we use the notation of matrix multiplication in line 5, but actual updating of
+$H$ and $P$ uses row and column operations.
+
 \begin{prop} 
@@
 An approximation to $\chi_M$ can be computed in $\softO(n^\omega)$ operations in $K$
-(\textit{e.g.} as a by-product of \cite{storjohann:01a}).
+by computing an approximation to the Frobenius normal form of $M$ \cite{storjohann:01a}.
 
@@
 \item the optimal precision on the $k$-th coefficient of the 
-characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}
+characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}.
+Note that it does not matter which method we use to compute $\adj(X-M)$,
+since we are only interested in the precision loss and not the runtime.
 \item in the capped relative model\footnote{Each 
@@
 
-\hfill$M\texttt{.charpoly(algorithm="df")}$\hfill\null
+\hfill$M\texttt{.charpoly(algorithm="df")}.$\hfill\null
 

### 2017-04-18 charpoly/charpoly.tex
Rewrote abstract and "The contribution of this paper"

@@
 an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
-developed previously.  When $M$ is integral with precision $O(p^N)$,
-we give a criterion (checkable in time $\softO(n^\omega)$) for
-$\chi(M)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
-algorithm for determining the optimal precision when the criterion is not
-satisfied, and give examples when the precision is larger than $O(p^N)$.
+developed previously.  When $M$ is an integral matrix whose entries are 
+all given at the same precision $O(p^N)$, we give a criterion (checkable 
+within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
+coefficient of $\chi(M)$ with more accuracy that $O(p^N)$.
+In general, we provide two algorithms for determining the optimal 
+precision on the coefficients of $\chi(M)$ and on their eigenvalues.
+We provide evidences showing that classical algorithms do not reach
+this optimal precision in general.
 \end{abstract}
@@
 
-Thanks to the application of the framework of differential precision
-of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
-we know that the precision of the characteristic polynomial
-$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
-is determined by the adjugate $\adj(X -M).$
+This paper provides a theoretical and concrete study of the propagation 
+of the precision during the computation of the characteristic polynomial 
+of matrix defined over the $p$-adics. Thanks to the application of the 
+general framework of differential precision developed in 
+\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a}, we know that the 
+optimal precision of the characteristic polynomial $\chi_M$ of a matrix 
+$M \in M_n(\mathbb{Q}_p)$ is controled by the adjugate $\adj(X{-}M).$
 In this article, we provide:
@@
 \renewcommand{\itemsep}{0pt}
-\item Proposition \ref{prop:shortcom}: a factorization of $\adj(X -M)$
-as a product of two rank-$1$ matrices (when $M$ has a cyclic vector),
-computable in $\softO(n^\omega)$ operations by Theorem \ref{thm:compute_shortcom}
-\item Corollary \ref{cor:prec_gain}: a simple, $\softO(n^\omega)$ criterion
-to decide whether $\chi_M$ is defined at precision
-higher than the precision of $M$ (when $M \in M_n(\mathbb{Z}_p)$).
-\item Theorem \ref{thm:alg3}: a $\softO(n^3)$ algorithm with operations in $\Zp$
-to compute the optimal precision on each coefficient of $\chi_M$ (when $M$
-is given with uniform precision on its entries).
+\item Proposition \ref{prop:shortcom}: a short representation of
+the matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
+field (instead to $O(n^3)$ for a dense representation);
+\item two algorithms to compute $\adj(X{-}M)$: the first one is 
+division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
+the second one may perform divisions and has complexity $\tilde 
+O(n^\omega)$;
+\end{enumerate}
+and deduce from this the following applications to the question
+we are interested in:
+\begin{enumerate}
+\renewcommand{\itemsep}{0pt}
+\setcounter{enumi}{2}
+\item Corollary \ref{cor:prec_gain}: a simple criterion to decide 
+whether $\chi_M$ is defined at precision higher than the precision of 
+$M$ (when $M$ lies in $M_n(\mathbb{Z}_p)$ and all its entries are given 
+at the same precision);
+\item two algorithms to compute the optimal precision on each 
+coefficient of the characteristic polynomial of a matrix $M \in 
+M_n(\Qp)$ (whose entries may be given at different precision): the first 
+one is division-free and has complexity $\softO(n^3)$ while the second 
+one may perform divisions but has complexity $\softO(n^\omega)$;
 \item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
-algorithm to compute the optimal precision on each eigenvalue of $M$
+algorithm to compute the optimal precision on each eigenvalue of $M$.
 \end{enumerate}
+Moreover, we provide evidences showing that classical algorithms for 
+computing the characteristic polynomials do not reach the optimal 
+precision on average.
 

### 2017-04-18 charpoly/charpoly.tex
Added a remark explaining why we count operations in K

@@
 
+\todo{Rewrite this subsection with more details.}
+
 We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
@@
 
+\xavier{I propose to move Proposition \ref{prop:mainlem} here.}
+
+\subsection{Several facts about the differential of $\chi$}
+
 \begin{prop}
@@
 
-In this section, we combine the computation of a Hessenberg form
-of a matrix and the computation of the inverse through the Smith normal form (SNF)
-over a complete discrete valuation field (CDVF)
-to compute $\adj(X-M)$ and $d \chi$.
-If $M \in M_n(\OK)$, then only division by invertible
-elements of $\OK$ will occur.
+In this section, we combine the computation of a Hessenberg form of a 
+matrix and the computation of the inverse through the Smith normal form 
+(SNF) over a complete discrete valuation field (CDVF) to compute 
+$\adj(X{-}M)$ and $d \chi$. If $M \in M_n(\OK)$, then only division 
+by invertible elements of $\OK$ will occur.
 %The computation of a Hessenberg form provides an $O(n^3)$
 %algorithm to compute the characteristic polynomial.
+
+\begin{rem}
+In what follows, we will count operations in $K$ (regardless to 
+precision) for expressing our complexities. This choice makes sense 
+because we will not usually need much precision on the entries of 
+$\adj(X{-}M)$ (roughly only their valuations matter).
+% should we argue more than that?
+\end{rem}
 

### 2017-04-19 charpoly/charpoly.tex
Small improvement.

@@
 \item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time} \\
-      \david{Tristan, can you check that my clarification before Proposition 3.2 is correct?}
+      \david{Tristan, can you check that my clarification before Proposition 3.2 is correct?} \\
+      \tristan{This is perfect.}
 \item \done{In section 4, we clarified how the characteristic polynomial is a by-product of [15].}
@@
 Note that we use the notation of matrix multiplication in line 5, but actual updating of
-$H$ and $P$ uses row and column operations.
+$H$ and $P$ uses row and column operations (individually using $O(n)$ operations in $K$).
 
@@
 
-First, we need a lemma relating comatrices of
+First, we need a lemma relating adjugates of
 similar matrices:
@@
 This has a remarkable advantage: the pivots during the computation of
-the SNF of $\adj(1-XM)$ are units of $\OK\llb X\rrb,$ and are known
+the SNF of $\adj(1-XM)$ are units of $K\llb X\rrb$ (of $\OK\llb X\rrb$ if $M \in M_n(\OK)$) and are known
 in advance to be on the diagonal. This leads to a very smooth
@@
 \begin{theo}
-Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix.
+Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix with
+coefficient-wise precision $O(\pi^M).$
 Then, using Algorithm 2, one can compute $\adj (X -H)$
@@
 For lines 4 and 5, there are only $n-1$ eliminations, resulting in a $\softO (n^2)$ cost
-for the computation of $U.$. Rather than actually construct $P$, we just track the eliminations
+for the computation of $U.$ Rather than actually construct $P$, we just track the eliminations
 performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.
@@
 $d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
-$O(n^3)$ operations in $K$. This seems to be optimal given that the 
+$\softO (n^3)$ operations in $K$. This seems to be optimal given that the 
 (naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n
@@
 If $M \in M_n(\OK)$ is given at (high enough) \emph{flat} precision, 
-then we can avoid the final base change step in the Hessenberg method.
+then we can avoid the final base change step in the Hessenberg 
+and thus any division by a non-invertible.
 Indeed, observe that, thanks to Lemma \ref{lem:adjugate_of_similar}, 

### 2017-04-20 charpoly/charpoly.tex
Deux remarques sur O(n^4).

@@
 with these methods.  First, they are slower than alternatives that allow division,
-requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
+requiring $O(n!)$, $O(n^4)$\tristan{Shouldn't it be $\softO(n^4)$ ?} and $\softO(n^{2 + \omega/2})$ operations.  Second,
 while the lack of division implies that the result is accurate modulo $p^N$ as
@@
 This is faster than computing $U \adj(X-M) U^{-1},$
-which is, at first sight, in $O(n^4)$ operations in $K.$
+which is, at first sight, in $O(n^4)$\tristan{Shouldn't it be $\softO (n^4)$?} operations in $K.$
 \end{rem}

### 2017-04-21 charpoly/charpoly.tex
Section 2.1 rewritten

@@
 
+\newcommand{\famN}{\mathcal{N}}
+
 \newcommand{\llb}{[\mkern-2.5mu[}
@@
      \david{Do either of you know what Dixon's algorithm refers to in this context?}
-\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time} \\
-      \david{Tristan, can you check that my clarification before Proposition 3.2 is correct?} \\
-      \tristan{This is perfect.}
+\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time}
 \item \done{In section 4, we clarified how the characteristic polynomial is a by-product of [15].}
@@
 \begin{enumerate}
-\item \tdo{In the abstract, clarify what kind of operations are being counted.}
-\item \tdo{In the abstract, clarify what we mean by ``precision exactly $O(p^k)$.''}\\
-      \xavier{By the way, we should also mention in the abstract the $\softO(n^\omega)$ algorithm.}
+\item \done{The abstract has been rewritten. In particular, the kind of operations we are counted and the locution ``precision exactly $O(p^k)$'' have been clarified.}
 \item \tdo{In the introduction, we say ``Let $M$ be known at precision $O(p^m)$.''  What does this mean?  Coefficient-wise?  Overall?  This needs to be said clearly.}\\
@@
 \item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
-\item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}\\
-      \xavier{Maybe we should instead remove any mention to the tangent space: it's not really necessary.}\\
-      \david{That's probably a good idea.  We'll have to make a remark about ignoring the leading monic term.}
+\item \done{The notion of lattice (in the tangent space) has been defined.}
 \item \tdo{The definition of ``flat precision'' and ``jagged precision'' both need to be more clearly defined.}
@@
 \end{enumerate}
-\xavier{I propose to expand the first paragraph of Section 2.1 (and even
-maybe make a separate subsection for it). Concretely, I suggest to 
-(1)~do not talk about the tangent space (but instead consider lattices
-in $M_n(K)$),
-(2)~give precision definition of flat, jagged precisions and relate them 
-with lattices,
-(3)~maybe cite the notes of my recent lecture at JNCF in which all of
-this is detailed.}
+
+\done{
+What we have done so far:
+\begin{enumerate}
+\item Section \emph{The contribution of this paper} of the introduction
+has been rewritten and detailed. We split our contributions into two 
+parts: the first part is about the (fast) computation of the adjugate
+of $X{-}M$ while the second part concerns applications to precision.
+We also added a sentence about numerical experiments.
+\item Section 2.1 has been rewritten. We no longer talk about tangent
+spaces. Jagged and flat precisions, together with their relationships with
+lattice precision, are introduced and discussed. 
+Proposition \ref{prop:mainlem} has been moved to Section 2.1 and stated
+in terms of valuations.
+\end{enumerate}
+}
 
@@
 stable algorithm (see Remark \ref{rem:lift_for_optimal}).
+\xavier{Maybe, we should say that it is not completely satisfying and (maybe) that we plan to come back to this question later.}
 
@@
 
-\todo{Rewrite this subsection with more details.}
-
-We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
-as a foundation for our discussion of the precision for the characteristic polynomial
-of a matrix.  We will be concerned with two $K$-manifolds in what follows:
-the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
-$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
+We recall some of the definitions and results of 
+\cite{caruso-roe-vaccon:14a} as a foundation for our discussion of the 
+precision for the characteristic polynomial of a matrix. We will be 
+concerned with two spaces in what follows: the space $M_n(K)$ of $n 
+\times n$ matrices with entries in $K$ and the space $K_n[X]$ of monic 
+degree $n$ polynomials over $K$. We define as well $K_{<n}[X]$ (resp. 
+${\OK}_{<n}[X]$) as the space of polynomials over $K$ (resp. over $\OK$) 
+of degree strictly less than $n$.
+
+There are several choices for modeling precision on $p$-adic matrices. 
+The first one consists in representing each entry of a matrix as a 
+quantity of the form $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
+is called the \emph{absolute precision} and may depend on $(i,j)$. The 
+way of representing precision will be called \emph{jagged} in the 
+sequel. When the $N_{i,j}$ are all equal (\emph{i.e.} when each entry 
+is given at the same precision), we say that the precision datum is 
+\emph{flat}.
+A more flexible way of modeling precision consists in using lattices. We 
+recall that a lattice in a $M_n(K)$ the image of $M_n(\OK)$ under a 
+$\Qp$-linear transformation of $M_n(K)$. Since $\OK$ appears as the 
+closed unit ball of $K$ (for the $p$-adic norm), cosets of the form
+$M + H$ (where $M$ lies in $M_n(K)$ and $H$ is a lattice) are special
+neighborhood of $M$; we will use them for modeling precision on $M$. 
+Given a family of integers $\famN = (N_{i,j})_{i,j}$ the lattice:
+$$H_{\famN} = 
+\big\{ \, M = (a_{i,j})_{i,j} \in M_n(\Qp) \,\, | \,\, \forall i,\!j\!:
+\pi^{N_{i,j}} \text{ divides } a_{i,j} \, \big\}$$
+corresponds to the jagged precision introduced above.
+
+Jagged and flat precisions are defining similarly for monic polynomials 
+(taking care that no precision datum is needed for the leading 
+coefficient which is supposed to be exactly~$1$). Simiarly, the notion 
+of \emph{lattice precision} extends to polynomials. The underlying 
+vector space is now $K_{<n}[X]$ and we define a lattice in it as the
+image of ${\OK}_{<n}[X]$ under some linear mapping.
+
+The benefit of working with lattices is that they behave well under 
+differentiable maps; this is the content of 
+\cite{caruso-roe-vaccon:14a}*{Lem. 3.4}. Write $\chi : M_n(K) \to 
+K_n[X]$ for the characteristic polynomial map and $d \chi_M$ for its 
+differential at $M$. Thanks to 
+Recall from \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} 
+\cite{caruso-roe-vaccon:15a}*{\S 3.3}} that $d\chi_M$ is given by
+\begin{equation} \label{eq:dchi}
+\begin{array}{rcl}
+d\chi_M: \quad M_n(K) & \to & K_{<n}[X] \smallskip \\
+ dM & \mapsto & \tr(\adj(X{-}M) \cdot dM).
+\end{array}
+\end{equation}
+where $\adj(M)$ denotes the adjugate\footnote{In 
+\citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the 
+adjugate was referred to using the non-standard term \emph{comatrix}} of 
+$M$ (when $M \in \GL_n(K)$, we have $\adj(M) = \det(M) M^{-1}$) and 
+$d\chi_M$ the differential at $M$. In this setting, 
+\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} says that
+\begin{equation}
+\label{eq:mainlem}
+\chi(M + H) = \chi(M) + d\chi_M(H)
+\end{equation}
+whenever $H$ is a sufficiently ``small'' and ``regular'' lattice.
+More precisely, let $\sigma_1(M), \dots, \sigma_n(M)$ denote the 
+elementary divisors of $M$.
+
+\begin{prop} \label{prop:mainlem}
+Let $M \in M_n(K)$ and assume that $d \chi_M$ is surjective. Let:
+\[
+\nu = \max\left(\sum_{i=1}^{n-1} \val(\sigma_i(M)), 0\right).
+\]
+Then, for all integers $N, M \geq 0$ with $M + \nu < 2N$
+any lattice $H$ such that $M_n(\pi^N \OK) \subset H \subset
+M_n(\pi^M \OK)$
+satisfies Eq.~\eqref{eq:mainlem}.
+\end{prop}
+
+\begin{proof}
+Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
+of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
+By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
+occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
+${\OK}_{<n}[X] \subset d\chi_M(M_n(p^\nu \Zp))$.
+
+Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
+with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
+the conclusion.
+\end{proof}
+
+\begin{comment}
+Given a matrix $M \in M_n(K)$,
 the most general kind of precision structure we may attach to $M$ is a
@@
 Our analysis of the precision behavior of $\chi$ rests upon
-the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
+the computation of its derivative $d\chi$, using 
+\cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
 For a matrix $M \in M_n(K)$, we identify the tangent space $V$
@@
 the space $K_{<n}[X]$ of polynomials of degree less than $n$.
-Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
-we have $\adj(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
-\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
-that $d\chi_M$ is given by
-\begin{equation} \label{eq:dchi}
-d\chi_M: dM \mapsto \tr(\adj(X-M) \cdot dM).
-\end{equation}
-
-\xavier{I propose to move Proposition \ref{prop:mainlem} here.}
+\end{comment}
 
 \subsection{Several facts about the differential of $\chi$}
+
+\smallskip
+
+\noindent
+{\bf About surjectivity.}
+The surjectivity of $d \chi_M$ appeared as an assumption in 
+Proposition~\ref{prop:mainlem}. The next Proposition make it
+entirely explicit.
 
@@
 \renewcommand{\itemsep}{0pt}
-\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
+\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is 
+${\OK}_{<n}[X]$.
 \item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
@@
 
-Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
-$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
-
-\begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and let 
-\[
-\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
-\]
-Then, for all $\rho \in (0, 1]$ and all
-$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
-satisfies:
-\begin{equation}
-\chi(M + H) = \chi(M) + d\chi_M(H).
-\end{equation}
-\end{prop}
-
-\begin{proof}
-
-Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
-of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
-By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
-occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
-$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.
-
-Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
-with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
-the conclusion.
-\end{proof}
-
 The relationship between precision and the images of lattices under $d\chi_M$ allows us to
@@
 
-\subsection{Stability under multiplication by $X$}
-
+\medskip
+
+\noindent
+{\bf Stability under multiplication by $X$.}
 By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
@@
 \begin{proof}
-Let $A = \adj(X-M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
+Let $A = \adj(X{-}M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
 the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
-with its adjugate is the determinant, $(X - M) \cdot A = \chi_M$ and thus
+with its adjugate is the determinant, $(X{-}M) \cdot A = \chi_M$ and thus
 $f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
@@
 
-\subsection{Compact form of $d \chi_M$}
-
+\medskip
+
+\noindent
+{\bf Compact form of $d \chi_M$.}
 Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$ as in \eqref{eq:companion}.
@@
 This is faster than computing $U \adj(X-M) U^{-1},$
-which is, at first sight, in $O(n^4)$\tristan{Shouldn't it be $\softO (n^4)$?} operations in $K.$
+which is, at first sight, in $\softO(n^4)$
+%\tristan{Shouldn't it be $\softO (n^4)$?}
+operations in $K.$
 \end{rem}

### 2017-04-21 charpoly/charpoly.tex
Working on section 2.1

--- 2017-04-19 charpoly/charpoly.tex
@@
 within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
-coefficient of $\chi(M)$ with more accuracy that $O(p^N)$.
+coefficient of $\chi(M)$ with more accuracy than $O(p^N)$.
 In general, we provide two algorithms for determining the optimal 
-precision on the coefficients of $\chi(M)$ and on their eigenvalues.
-We provide evidences showing that classical algorithms do not reach
+precision of the coefficients of $\chi(M)$ and of their eigenvalues.
+We provide evidence showing that classical algorithms do not reach
 this optimal precision in general.
@@
 \item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
-\item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}\\
-      \xavier{Maybe we should instead remove any mention to the tangent space: it's not really necessary.}\\
-      \david{That's probably a good idea.  We'll have to make a remark about ignoring the leading monic term.}
-\item \tdo{The definition of ``flat precision'' and ``jagged precision'' both need to be more clearly defined.}
+\item \done{We have expanded upon the discussion of lattices, and removed any reference to tangent spaces (since they were a distraction for matrices and polynomials).}
+\item \done{We have fleshed out the definitions of flat and jagged precision.}
 \end{enumerate}
@@
 the matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
-field (instead to $O(n^3)$ for a dense representation);
+field (instead of $O(n^3)$ for a dense representation);
 \item two algorithms to compute $\adj(X{-}M)$: the first one is 
@@
 \end{enumerate}
-Moreover, we provide evidences showing that classical algorithms for 
+Moreover, we provide evidence showing that classical algorithms for 
 computing the characteristic polynomials do not reach the optimal 
-precision on average.
+precision in many contexts.
 
@@
 
-\todo{Rewrite this subsection with more details.}
+\todo{Rewrite this subsection with more details.  Add references to JNCF notes.}
 
@@
 the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
-$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
-the most general kind of precision structure we may attach to $M$ is a
-\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
-arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
-frequently work with certain classes of lattices, either \emph{jagged} lattices
-where we specify a precision for each matrix entry or \emph{flat} lattices where
-every entry is known to a fixed precision $O(p^N)$.  Similarly, precision for
-monic polynomials can be specified by giving a lattice in the tangent space
-at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
-of simplicity.
+$K_n[X]$ of monic degree $n$ polynomials over $K$.  One cannot represent
+matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
+together with a measure of the approximation's accuracy.  Recall that a
+\emph{lattice} in $M_n(K)$ is an open $\OK$-submodule of $M_n(K)$:
+the $\OK$-span of $n^2$ basis vectors \cite{caruso-roe-vaccon:14a}*{\S 2.2}.
+We record the precision of $M$ with lattice $H$: the matrix $M$ is understood
+as some matrix of the form $M + h$, where $h \in H$.
+
+One problem with representing precision in this way is that storing $H$
+requires $n^2$ basis vectors, each with $n^2$ entries.  We can shrink the
+space required by working only with certain classes of lattices.  When
+using \emph{jagged} precision, the basis matrices for $H$ can only have
+one nonzero entry; this corresponds to specifying a precision for each
+entry of the matrix separately.  Going further, $M$ is said to have \emph{flat}
+precision $O(p^N)$ if every entry has precision $O(p^N)$, or equivalently,
+if $H = p^N M_n(\OK)$.  Similarly, precision for
+monic polynomials can be specified by giving a lattice in $K_{<n}[X]$,
+or specifying precision coefficient-wise in the interest of simplicity.
 
@@
 the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
-For a matrix $M \in M_n(K)$, we identify the tangent space $V$
-at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi_M$ with
-the space $K_{<n}[X]$ of polynomials of degree less than $n$.
 Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
@@
 
-\xavier{I propose to move Proposition \ref{prop:mainlem} here.}
+When $d\chi_M$ is surjective and $H$ is small enough, we may determine the precision of $\chi(M)$ precisely.
+Write $V$ for $M_n(K)$, $W$ for $K_n[X]$, 
+$B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
+$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
+
+\begin{prop} \label{prop:mainlem}
+Let $M \in M_n(K)$ and suppose that $d\chi_M$ is surjective.  Let 
+\[
+\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
+\]
+Then, for all $\rho \in (0, 1]$ and all
+$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
+satisfies:
+\begin{equation}
+\chi(M + H) = \chi(M) + d\chi_M(H).
+\end{equation}
+\end{prop}
+
+\begin{proof}
+Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
+of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
+By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
+occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
+$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.
+
+Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
+with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
+the conclusion.
+\end{proof}
 
@@
 
+We may rephrase the condition that $d\chi_M$ is surjective using more familiar invariants attached to $M$.
 \begin{prop}
@@
 \eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
-\end{proof}
-
-Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
-$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
-
-\begin{prop} \label{prop:mainlem}
-Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and let 
-\[
-\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
-\]
-Then, for all $\rho \in (0, 1]$ and all
-$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
-satisfies:
-\begin{equation}
-\chi(M + H) = \chi(M) + d\chi_M(H).
-\end{equation}
-\end{prop}
-
-\begin{proof}
-
-Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
-of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
-By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
-occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
-$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.
-
-Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
-with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
-the conclusion.
 \end{proof}

### 2017-04-21 charpoly/charpoly.tex

@@
 \newcommand{\OK}{\mathcal{O}_K}
+
+\newcommand{\famN}{\mathcal{N}}
 
@@
 as a foundation for our discussion of the precision for the characteristic polynomial
-of a matrix.  We will be concerned with two $K$-manifolds in what follows:
+of a matrix.  We will be concerned with two spaces in what follows:
 the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
-$K_n[X]$ of monic degree $n$ polynomials over $K$.  One cannot represent
+$K_n[X]$ of monic degree $n$ polynomials over $K$.  We also define $K_{<n}[X]$ (resp. 
+${\OK}_{<n}[X]$) as the space of polynomials over $K$ (resp. over $\OK$) 
+of degree strictly less than $n$.
+
+One cannot represent
 matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
-together with a measure of the approximation's accuracy.  Recall that a
-\emph{lattice} in $M_n(K)$ is an open $\OK$-submodule of $M_n(K)$:
-the $\OK$-span of $n^2$ basis vectors \cite{caruso-roe-vaccon:14a}*{\S 2.2}.
-We record the precision of $M$ with lattice $H$: the matrix $M$ is understood
-as some matrix of the form $M + h$, where $h \in H$.
-
-One problem with representing precision in this way is that storing $H$
-requires $n^2$ basis vectors, each with $n^2$ entries.  We can shrink the
-space required by working only with certain classes of lattices.  When
-using \emph{jagged} precision, the basis matrices for $H$ can only have
-one nonzero entry; this corresponds to specifying a precision for each
-entry of the matrix separately.  Going further, $M$ is said to have \emph{flat}
-precision $O(p^N)$ if every entry has precision $O(p^N)$, or equivalently,
-if $H = p^N M_n(\OK)$.  Similarly, precision for
-monic polynomials can be specified by giving a lattice in $K_{<n}[X]$,
-or specifying precision coefficient-wise in the interest of simplicity.
-
-Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
+together with a measure of the approximation's accuracy; there are several ways to do so.
+When using \emph{jagged} precision, on gives a precision on each entry,
+representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
+is called the \emph{absolute precision} and may depend on $(i,j)$.
+When the $N_{i,j}$ are all equal, we say that the precision datum is 
+\emph{flat}.
+
+We may also model precision more flexibly by using lattices.  We 
+recall that a \emph{lattice} $H$ in $M_n(K)$ is the $\OK$-span of a vector
+space basis, or equivalently the image of $M_n(\OK)$ under a 
+$K$-linear transformation of $M_n(K)$ \cite{caruso-roe-vaccon:14a}*{\S 2.2}.
+For $M \in M_n(K)$, cosets of the form $M + H$ are special
+neighborhoods of $M$; we will use them for modeling precision on $M$. 
+Given a family of integers $\famN = (N_{i,j})_{i,j}$ the lattice
+\[
+H_{\famN} = \big\{ \, M = (a_{i,j})_{i,j} \in M_n(\Qp) \,\, | \,\, \forall i,\!j\!:
+\pi^{N_{i,j}} \text{ divides } a_{i,j} \, \big\}
+\]
+corresponds to the jagged precision introduced above.
+The benefit of working with lattices is that they behave well under 
+differentiable maps \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}, but they
+require more space and time to compute.
+
+Jagged and flat precision is defined similarly for monic polynomials, 
+taking care that no precision is specified for the leading 
+coefficient which is supposed to be exactly~$1$. The notion 
+of \emph{lattice precision} also extends to polynomials. The underlying 
+vector space is now $K_{<n}[X]$ and we define a lattice in it as the
+image of ${\OK}_{<n}[X]$ under some linear mapping.
+
+Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map
+and $d \chi_M$ for its differential at $M$.
 Our analysis of the precision behavior of $\chi$ rests upon
@@
 Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
-we have $\adj(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
+we have $\adj(M) = \det(M) M^{-1}$).  Recall
 \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
@@
 \begin{equation} \label{eq:dchi}
-d\chi_M: dM \mapsto \tr(\adj(X-M) \cdot dM).
+\begin{array}{rcl}
+d\chi_M: \quad M_n(K) & \to & K_{<n}[X] \smallskip \\
+ dM & \mapsto & \tr(\adj(X{-}M) \cdot dM).
+\end{array}
 \end{equation}
 
-When $d\chi_M$ is surjective and $H$ is small enough, we may determine the precision of $\chi(M)$ precisely.
-Write $V$ for $M_n(K)$, $W$ for $K_n[X]$, 
-$B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
+When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may determine
+the precision of $\chi(M)$ precisely as
+\begin{equation} \label{eq:mainlem}
+\chi(M + H) = \chi(M) + d\chi_M(H).
+\end{equation}
+More precisely, let
 $\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
@@
 \begin{prop} \label{prop:mainlem}
-Let $M \in M_n(K)$ and suppose that $d\chi_M$ is surjective.  Let 
+Let $M \in M_n(K)$ and assume that $d \chi_M$ is surjective. Let:
+\[
+\nu = \max\left(\sum_{i=1}^{n-1} \val(\sigma_i(M)), 0\right).
+\]
+Then, for all integers $a, b \geq 0$ with $a + \nu < 2b$,
+any lattice $H$ such that $M_n(\pi^b \OK) \subset H \subset
+M_n(\pi^a \OK)$
+satisfies Eq.~\eqref{eq:mainlem}
+\end{prop}
+
+\begin{prop} \label{prop:mainlem}
+Let $M \in M_n(K)$ and assume that $d\chi_M$ is surjective.  Let 
 \[
@@
 occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
-$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.
+${\OK}_{<n}[X] \subset d\chi_M(M_n(\pi^\nu \OK))$.
 
@@
 We may rephrase the condition that $d\chi_M$ is surjective using more familiar invariants attached to $M$.
+
 \begin{prop}
@@
 This is faster than computing $U \adj(X-M) U^{-1},$
-which is, at first sight, in $O(n^4)$ operations in $K.$
+which is, at first sight, in $\softO(n^4)$
+%\tristan{Shouldn't it be $\softO (n^4)$?}
+operations in $K.$
 \end{rem}

### 2017-04-25 charpoly/charpoly.tex
Removing duplicate version of Prop 2.1
Finishing up addressing reviewer comments, many small changes to make result fit into 8 pages

@@
 \section{Introduction}
+%\begin{comment}
 \vspace{0.1in}
@@
 \begin{enumerate}
-\item \tdo{What are the implications, if any, of your work on well-known methods such as Dixon's algorithm?}\\
-     \david{Do either of you know what Dixon's algorithm refers to in this context?}
+\item \done{We do not see a direct application of this paper to Dixon's algorithm, since his algorithm solves a $p$-adic system $p$-adically rather than computing a characteristic polynomial.
+Moreover, since the prime can be chosen arbitrarily, it will usually not have repeated eigenvalues modulo $p$ and thus there will be no gain in precision.}
 \item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time}
@@
 \item \done{We clarified why we chose the distribution in Figure 1.}
-\item \tdo{Think about whether Proposition 5.6 can be sped up using the work of Kedlaya and Umans.}\\
-      \xavier{I cannot see how Kedlaya--Umans can help: the cost $\softO(n^\omega)$ comes from the computation of a Frobenius normal form...}
+\item \done{We do not see how to apply Kedlaya-Umans.
+Yes, there is multipoint evaluation used in Proposition 5.6, but it is dominated by the computation of the matrix $P$, which requires $\softO(n^\omega)$ operations in $K$.}
 \end{enumerate}
@@
 \begin{enumerate}
-\item \tdo{Convince reviewer that the complexity claimed in Theorem 4.1 is accurate: ``it lacks a precision loss analysis and relies on classic division based algorithms}\\
-      \xavier{One can argue that we do not need \emph{a priori} much accuracy on the precision of $\adj(M-X)$ because only the valuations of the entries of the latter matrix are relevant.
-So counting operations in $K$ makes sense.
-By the way, we could say at some point that a good idea would be to use relaxed algorithms and increase the precision until we are able to read the valuations.}
+\item \done{We have added Remark 3.1 outlining why we do not need much accuracy in the computation of $\adj(M{-}X)$: it's roughly true that only the valuations of the entries are relevant, since we only care about the lattice that the entries generate.
+We hope that this remark is sufficient to convince the reviewer that the lack of precision analysis in the discussion preceding Theorem 4.1 does not invalidate the claimed complexity.}
 \item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
@@
 \item \done{We have switched from $M^t$ to $M^T$ to denote the transpose.}
-\item \tdo{A few explanations on where does the formula for the differential comes from
-would be most welcome (those of [3 Appendix B] and some connection of the trace
-with the fact that we do degree one approx of a charpoly)}\\
-      \xavier{FWIW, I wrote a proof in my lecture at JNCF. Maybe we can just cite it.}\\
-      \david{Citing it seems reasonable.}
+\item \done{We have added a reference to course notes of Caruso (now reference [2]) which go into more detail on how the formula for the differential of $\chi_M$ is derived.}
 \item \done{In section 5.1, we added a remark noting that the method for computing the adjugate matrix has no effect on the extent of the precision loss, just on the runtime.}
-\item \tdo{Further experiments (with higher dimension, or showing how the two methods for computing adj(M) differ), would be welcome.}
-\item \tdo{Clarify the intro paragraph just before the Previous contributions section.}
+\item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).
+We note that we did perform the current experiment in higher dimension with similar results: dimension $8$ was chosen to be representative without the table taking up too much space in the paper.}
+\item \done{We have removed the two paragraphs before the Previous contributions section.
+Some of the material discussed therein has been moved to Section 2.1, and some is now referenced in the course notes [2]}
 \item \done{We have carried out the other minor corrections suggested.}
@@
 \item \done{The abstract has been rewritten. In particular, the kind of operations we are counted and the locution ``precision exactly $O(p^k)$'' have been clarified.}
-\item \tdo{In the introduction, we say ``Let $M$ be known at precision $O(p^m)$.''  What does this mean?  Coefficient-wise?  Overall?  This needs to be said clearly.}\\
-      \xavier{Well, it is the same, isn't it? Anyway, we may add ``coefficient-wise'' if it's not clear.}
-\item \tdo{In the introduction, clarify what kind of operations are being counted.}
-\item \tdo{In the discussion of existing methods, clarify what we mean by the claim that they do not yield optimal precision.}
+\item \done{We have also clarified what kind of operations are being counted and what is meant by ``precision $O(p^N)$`` in the introduction.}
+\item \done{We have added a discussion of the term \emph{optimal} precision in Section 2.1, after equation (3).}
 \item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
@@
 \textbf{General clarity comments:}
-\begin{enumerate}
-\item \tdo{The writing is rather dense and relies too heavily on external material, which makes the reading sometimes unpleasant. (3)}
-\item \tdo{Overall the presentation lacks a minimal set of definitions and terminology regarding the notions of p-adic rings and their precision to make it sufficiently self-contained for an non-expert reader.
+%\begin{enumerate}
+%\item \tdo{The writing is rather dense and relies too heavily on external material, which makes the reading sometimes unpleasant. (3)}
+%\item \tdo{Overall the presentation lacks a minimal set of definitions and terminology regarding the notions of p-adic rings and their precision to make it sufficiently self-contained for an non-expert reader.
 I believe only a one or two additional paragraphs could be sufficient to drastically improve the presentation of section 2, in particular recalling the various precision models.
@@
 (3)}
-\item \tdo{The paper itself is unnecessarily difficult to read, to the extent that it sometimes unclear what is being proven or to verify correctness.
+%\item \tdo{The paper itself is unnecessarily difficult to read, to the extent that it sometimes unclear what is being proven or to verify correctness.
 A number of key definitions are not given and there are too few motivating/clarifying examples.
 (5)}
-\item \tdo{The definitions are unclear and the comparison with previous work is deficient.
+%\item \tdo{The definitions are unclear and the comparison with previous work is deficient.
 If this is to appear the definitions and results must be made very clear, and the results must be demonstrated to be superior to the division-free algorithms cited, both in theory and in practice.}
-\end{enumerate}
-
-\done{
-What we have done so far:
+%\end{enumerate}
+
 \begin{enumerate}
-\item Section \emph{The contribution of this paper} of the introduction
-has been rewritten and detailed. We split our contributions into two 
+\item \done{Section \emph{The contribution of this paper} of the introduction
+has been rewritten with additional detail. We split our contributions into two 
 parts: the first part is about the (fast) computation of the adjugate
 of $X{-}M$ while the second part concerns applications to precision.
-We also added a sentence about numerical experiments.
-\item Section 2.1 has been rewritten. We no longer talk about tangent
-spaces. Jagged and flat precisions, together with their relationships with
+We also added a sentence about numerical experiments.}
+\item \done{Section 2.1 has been rewritten. We no longer talk about tangent
+spaces. Jagged and flat precision, together with their relationships with
 lattice precision, are introduced and discussed. 
-Proposition \ref{prop:mainlem} has been moved to Section 2.1 and stated
-in terms of valuations.
+Proposition 2.1 has been moved to Section 2.1 and stated
+in terms of valuations.}
+\item \done{Small adjustments have been made throughout the paper to attempt
+to make the prose more readable, though we are constrained by the page
+limit, which hinders us from adding more examples or details.}
 \end{enumerate}
-}
 
 \bigskip
+%\end{comment}
 
@@
 
-When computing with $p$-adic matrices, the lack of infinite memory implies
-that the entries may only be approximated at some finite precision $O(p^N)$.  As a
+When computing with $p$-adic matrices, entries may only be approximated
+at some finite precision $O(p^N)$.  As a
 consequence, in designing algorithms for such matrices one must analyze
@@
 
-Let $M \in M_n(\Qp)$ be known at precision $O(p^N)$.
+Let $M \in M_n(\Qp)$, with every entry given at precision $O(p^N)$.
 The simplest approach for computing the characteristic polynomial of $M$
@@
 with these methods.  First, they are slower than alternatives that allow division,
-requiring $O(n!)$, $O(n^4)$\tristan{Shouldn't it be $\softO(n^4)$ ?} and $\softO(n^{2 + \omega/2})$ operations.  Second,
+requiring $O((n+1)!)$, $\softO(n^5)$ and $\softO(n^{3 + \omega/2})$ operations in $K$.  Second,
 while the lack of division implies that the result is accurate modulo $p^N$ as
@@
 which is achievable in running time $\softO(n^\omega)$ \citelist{\cite{keller-gehrig:85a} \cite{pernet-storjohann:07a}}.
-However, the fact that it uses division frequently leads to catastrophic losses of precision.
+However, the use of division frequently leads to catastrophic losses of precision.
 In many examples, no precision remains at the end of the calculation.
@@
 stable algorithm (see Remark \ref{rem:lift_for_optimal}).
-\xavier{Maybe, we should say that it is not completely satisfying and (maybe) that we plan to come back to this question later.}
-
-One benefit of this approach is that we may account for diffuse precision:
-precision that is not localized on any single coefficient of $\chi_M$.  For example,
-let $0 \le \alpha_1 \le \alpha_2 \le \dots \le \alpha_n$, consider a
-diagonal matrix $D$ with diagonal entries $(p^{\alpha_1}, \dots, p^{\alpha_n})$,
-let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
-of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $\alpha_{n-1} > 0$
-and $M$ is known with precision $O(p^N)$ then the constant term of $\chi_M$
-will be known with precision larger than $O(p^N)$ (see
-\cite{caruso-roe-vaccon:15a}*{Prop. 3.2}).
-
-As long as none of the eigenvalues of $M$ are congruent to $-1$ modulo $p$,
-then none of the coefficients of the characteristic polynomial of $1+M$ will have
-precision larger than $O(p^N)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
-content of these two polynomials should be equivalent. The solution is that the
-extra precision in $\chi_{1+M}$ is diffuse and not visible on any individual coefficient.
-We formalize this phenomenon using lattices; see Section \ref{sec:padicprec} for
-further explanation, and \cite{caruso:17a}*{\S 3.2.2} for a specific example of the
-relationship between $\chi_M$ and $\chi_{1+M}$.
 
@@
 In most of these papers, the precision analysis usually
-deals with great details on how to obtain the matrices (\textit{e.g.} of action of Frobenius)
+dwells on how to obtain the matrices (\textit{e.g.} of action of Frobenius)
 that are involved in the point-counting schemes.
-However, the computation of their characteristic polynomials is often
-a little bit less thoroughly studied: some
+However, the computation of their characteristic polynomials 
+attracts less attention: some
 refer to fast algorithms (using division), while others
@@
 to the stable computation of characteristic polynomials.
-They have obtained a way to express the optimal precision
-on the characteristic polynomial, but have not given practical algorithms
+They describe the optimal precision
+for the characteristic polynomial, but do not give practical algorithms
 to attain this optimal precision.
@@
 to dramatic gains in precision over standard interval arithmetic.
-We close with results describing the precision associated to
+We close by describing the precision associated to
 eigenvalues of a matrix.
@@
 
-\todo{Rewrite this subsection with more details.  Add references to JNCF notes.}
-
 We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
@@
 
-Jagged and flat precision is defined similarly for monic polynomials, 
+Jagged, flat and lattice precision is defined similarly for monic polynomials, 
 taking care that no precision is specified for the leading 
-coefficient which is supposed to be exactly~$1$. The notion 
-of \emph{lattice precision} also extends to polynomials. The underlying 
+coefficient which must be exactly~$1$. The underlying 
 vector space is now $K_{<n}[X]$ and we define a lattice in it as the
@@
 the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
-Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
-we have $\adj(M) = \det(M) M^{-1}$).  Recall
+Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to as the \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
+we have $\adj(M) {=} \det(M) M^{-1}$).  Recall
 \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
@@
 
-When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may determine
+When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may give
 the precision of $\chi(M)$ precisely as
@@
 \end{equation}
-More precisely, let
+The equality in \eqref{eq:mainlem} justifies our use of the phrase \emph{optimal} precision
+to denote the lattice precision $d\chi_M(H)$. To make the conditions on $H$ under
+which \eqref{eq:mainlem} holds more precise, let
 $\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
@@
 
-\begin{prop} \label{prop:mainlem}
-Let $M \in M_n(K)$ and assume that $d\chi_M$ is surjective.  Let 
-\[
-\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
-\]
-Then, for all $\rho \in (0, 1]$ and all
-$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
-satisfies:
-\begin{equation}
-\chi(M + H) = \chi(M) + d\chi_M(H).
-\end{equation}
-\end{prop}
-
 \begin{proof}
@@
 \end{proof}
+
+For further discussion of using lattices to measure precision,
+details on the computation of $d\chi_M$ in \eqref{eq:dchi}, and 
+examples of additional precision in $\chi(M)$,
+see \cite{caruso:17a}*{\S 2.4.3 and \S 3.2.2}
 
@@
 \begin{proof}
-The equivalence of \eqref{pt:cyc}, \eqref{pt:1d}, and \eqref{pt:charmin} is standard; see
-\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1} for example.  We now show
+The equivalence of \eqref{pt:cyc}, \eqref{pt:1d}, and \eqref{pt:charmin} is standard; see \emph{e.g.}
+\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1}.  We now show
 \eqref{pt:cyc} $\Rightarrow$ \eqref{pt:surj} and \eqref{pt:surj} $\Rightarrow$ \eqref{pt:1d}
@@
 
-We also have an analogue of Proposition \ref{prop:surjectivity} for integral matrices.
+We also have an integral analogue of Proposition \ref{prop:surjectivity}.
 
@@
 The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
-\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
+\eqref{pt:cycp} is as in Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
 \end{proof}
@@
 If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
-is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
+is stable under multiplication by $X$, so is a module over $\OK[X]$.
 \end{cor}
@@
 
-\section{Differential\\via Hessenberg form}
+\section{Hessenberg form}
 \label{sec:diffHess}
@@
 a Hessenberg form.
-We prove here that over $K,$ if a matrix is 
-known at finite (jagged) precision,
-we can compute an approximate Hessenberg form of it.
-Moreover, we can provide an exact change of basis matrix.
-It relies on the following algorithm.
+We prove here that we may compute an approximate Hessenberg form 
+for any matrix known at finite (jagged) precision over $K$.
+Moreover, we provide an exact change of basis matrix.
 
@@
 \begin{prop} 
-Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
-$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficent in $M.$
+Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M$.
+$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
 \end{prop}
@@
 can be obtained as the valuation of a coefficient of $du_M(dM)$
-and the smallest that can be achieved this way for $dM$ in a precision lattice
-of flat precision. Hence the optimality of the bound given, at least when 
-$m$ is large \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
+and the smallest that can be achieved this way for $dM$ in a flat precision lattice.
+Hence the optimality of the bound given, for large 
+$m$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
 
@@
 
-First, we need a lemma relating adjugates of
+We begin by relating adjugates of
 similar matrices:
@@
 This has a remarkable advantage: the pivots during the computation of
-the SNF of $\adj(1-XM)$ are units of $K\llb X\rrb$ (of $\OK\llb X\rrb$ if $M \in M_n(\OK)$) and are known
+the SNF of $\adj(1-XM)$ are units in $K\llb X\rrb$ (in $\OK\llb X\rrb$ if $M \in M_n(\OK)$) and are known
 in advance to be on the diagonal. This leads to a very smooth
-precision and complexity behaviour when the input matrix lives in 
-$M_n(\OK).$
+precision and complexity behavior when $M \in M_n(\OK).$
 
@@
 product on line 8 is in $\softO (n^3).$
-We emphasize that no division has been done throughout the algorithm.
-Line 9 is costless, and the result is then proved.
+We emphasize that no division has been done.
+Line 9 is costless, and the result is proven.
 \end{proof}
@@
 with Algorithm 2 to compute the adjugate of $X-M$
-when $\chi_M$ is squarefree.   Note that this condition on
-$\chi_M$ is equivalent to $M$ being diagonalizable under
-the assumption that $d\chi_M$ is surjective.  The result is
+when $\chi_M$ is squarefree.   Note that, under
+the assumption that $d\chi_M$ is surjective,
+$\chi_M$ is squarefree if and only if $M$ is diagonalizable.  The result is
 the following $\softO(n^3)$ algorithm, where the only
@@
 to $n$ operations on rows or columns over a matrix with coefficients
-in $\OK[X]$ of degree at most $n.$
-Thus, it is in $\softO (n^3).$
-Step 5 is in $\softO (n),$ Step 6 in $\softO(n^2)$ and Step 7 in $\softO(n^3)$.
-All that is to prove is that the set of $P$ and $S$ to avoid
-is of dimension at most $n-1.$
+in $\OK[X]$ of degree at most $n$.
+Thus, it is in $\softO(n^3).$
+Step 5 is in $\softO(n)$, Step 6 in $\softO(n^2)$ and Step 7 in $\softO(n^3)$.
+We need only prove that the set of $P$ and $S$ to avoid
+is of dimension at most $n-1$.
+
 The idea is to work modulo $X-\lambda$
@@
 The goal of the Step $2$ is to ensure the first row of $B$ contains an
-invertible entry modulo $\chi_M.$
+invertible entry modulo $\chi_M$.
 Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
 affine hyperplane so that $\row(B,1) \mod (X-\lambda)$ is a non-zero vector.
-Hence for $\row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
-a finite union of affine hyperplane is to avoid.
-Similarly, the goal of Step 3 is to put an invertible coefficient (modulo
-$\chi_M$) on $C_{1,1},$ and again, only a finite union of affine
-hyperplane is to avoid.
-Hence, the set that the $\mu_i$'s have to avoid is a finite union
-of hyperplane, and hence, is of dimension at most $n-1.$
-Thus, almost any choice of $\mu_i$ leads to a matrix $C$ passing the test
-in Step 4.
-This concludes the proof.
+Hence we need only avoid a finite union of affine hyperplanes
+in order for $\row(B,1) \mod \chi (M)$ to contain an invertible coefficient.
+
+Similarly, the goal of Step 3 is make $C_{1,1}$ invertible modulo
+$\chi_M$.  Again, only a finite union of affine
+hyperplane need be avoided.
+Thus, almost any choice of $\mu_i$ leads to a matrix $C$ passing
+Step 4, concluding the proof.
 \end{proof}
@@
 
-\section{Differential\\via Frobenius form}
+\section{Frobenius form}
 \label{sec:diffFrob}
@@
 In this section, we only estimate the number of operations in $K$
-and not their behaviour on precision.
-
-From now on, we fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
+and not their behavior on precision.
+
+We now fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
 surjective. Let $(\alpha, P, Q, \chi_M)$ be the quadruple encoding
@@
 \end{align*}
-An approximation to $\chi_M$ can be computed in $\softO(n^\omega)$ operations in $K$
-by computing an approximation to the Frobenius normal form of $M$ \cite{storjohann:01a}.
+We may approximate $\chi_M$ in $\softO(n^\omega)$ operations in $K$
+by approximating the Frobenius normal form of $M$ \cite{storjohann:01a}.
 
@@
 
-It remains to compute the scaling factor $\alpha$. For this, we write
-the relation:
+It remains to compute the scaling factor $\alpha$. We write
 \begin{equation}
@@
 \end{equation}
-which comes from Eq.~\eqref{eq:shortcom} after multiplication on the 
-left by $P^{-1}$ and multiplication on the right by $P$. We observe
+by multiplying Eq.~\eqref{eq:shortcom} on the 
+left by $P^{-1}$ and on the right by $P$. We observe
 moreover that the first row of $R$ is $(1, 0, \ldots, 0)$. Evaluating
@@
 \label{thm:compute_shortcom}
-Given $M \in M_n(K)$ such that $d \chi_M$ is surjective,
-then one can compute $(\alpha, P, Q, \chi_M)$
-in $K[X]$ such that
-$\adj(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$
-in $\softO(n^\omega)$ operations in $K.$
+Given $M \in M_n(K)$ with $d \chi_M$ surjective,
+one can compute $\alpha, P, Q, \chi_M) \in K[X]$ with
+$\softO(n^\omega)$ operations in $K$ such that
+$\adj(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$.
 \end{theo}
@@
 The output might then need to be rescaled and
- truncated at the optimal precision. 
+ truncated to the optimal precision. 
 This requires $\softO(n^3)$ operations in $O_K$ 
-and unfortunately, for several instances, may require to increase a lot the precision.
+and may require a large intermediate precision.
+Better methods for approximating $\chi_M$ with
+lower intermediate precision would be a valuable
+future contribution.
 \end{rem}
@@
 \]
-where $v$ is a random integer generated according to the distribution:
+where $v$ is an integer generated according to the distribution:
 \[
@@
 characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}.
-Note that it does not matter which method we use to compute $\adj(X-M)$,
-since we are only interested in the precision loss and not the runtime.
+Note that the method for computing $\adj(X-M)$ is irrelevant
+since we only record the precision loss, not the runtime.
 \item in the capped relative model\footnote{Each 
 coefficient carries its own precision which is updated after each 
-elementary arithmetical operation.},
-the precision gotten on the $k$-th coefficient of the 
-characteristic polynomial of $M$ computed \emph{via} the call:
+elementary arithmetic operation.},
+the precision on the $k$-th coefficient of the 
+characteristic polynomial of $M$ is computed \emph{via} the call:
 
@@
 
-\begin{rem}
-The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
-division free algorithm of \cite{seifullin:02a}. It is likely that, proceeding so, we 
-limit the loss of precision.
-\end{rem}
+%\begin{rem}
+%The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
+%division free algorithm of \cite{seifullin:02a}. It is likely that, proceeding so, we 
+%limit the loss of precision.
+%\end{rem}
 
@@
 
-\caption{Average loss of accuracy on the coefficients
-of the characteristic polynomial of a random $9 \times 9$
+\caption{Average precision loss on 
+the characteristic polynomial of a random $9 \times 9$
 matrix over $\Q_2$}
@@
 %
-It should be read as follows. First, the acronyms CR and FP 
-refers to ``capped relative'' and ``floating-point'' respectively.
-The numbers displayed in the table are the average loss of
+It should be read as follows. The acronyms CR and FP 
+refer to ``capped relative'' and ``floating-point'' respectively.
+The numbers displayed are the average loss of
 \emph{relative} precision. 
-More precisely, if $N$ is the relative precision at
-which the entries of the input random matrix $M$ have been generated
-and $v$ is the valuation of the $k$-th coefficient of $\chi_M$, then:
+More precisely, if $N$ is the relative precision of
+the entries of $M$ and $v$ is the valuation of
+the $k$-th coefficient of $\chi_M$, then:
 
@@
 We observe that the loss of relative accuracy stays under control in the 
-``Optimal'' column whereas it has a very erratic behavior --- very large 
-values and very large deviation as well --- in the two other columns. 
-These experiments thus demonstrate the utility of the methods developed
-in this paper.
+``Optimal'' column whereas it has very erratic behavior in the two other columns,
+demonstrating the utility of the methods developed in this paper.
 
@@
 The function $f$ is strictly differentiable on a neighborhood of 
-$M_\app$.
-The differential of $f$ at $M$ is the linear mapping:
-$$dM \mapsto d \lambda = - \frac{\tr(\adj(\lambda-M) \cdot dM)}
-{\chi'_M(\lambda)}$$
+$M_\app$, with differential at $M$ the map
+\[
+dM \mapsto d \lambda = - \frac{\tr(\adj(\lambda-M) \cdot dM)}
+{\chi'_M(\lambda)},
+\]
 where $\chi'_M$ is the usual derivative of $\chi_M$.
@@
 \begin{proof}
-The first assertion follows from the implicit function Theorem.
-Differentiating the relation $\chi_M(\lambda) = 0$, we get
-$\chi'_M(\lambda) \cdot d \lambda + \tr(\adj(X-M) \cdot dM)(\lambda) = 0$,
-from which the Lemma follows.
+The first assertion follows from the implicit function theorem,
+the second from differentiating $\chi_M(\lambda) = 0$, which gives
+$\chi'_M(\lambda) \cdot d \lambda + \tr(\adj(X-M) \cdot dM)(\lambda) = 0$.
+%from which the Lemma follows.
 \end{proof}
@@
 evaluation methods (see 10.7 of \cite{gathen-gerhard:13a}).
-On the other hand, we observe that $P_i Y(\lambda_k)^T$ is nothing but
+On the other hand, we observe that $P_i Y(\lambda_k)^T$ is
 the $(i,k)$ entry of the matrix:

### 2017-04-25 charpoly/charpoly.tex
Small fixes (typos, notations...)

@@
 \begin{abstract}
-We analyze the precision of the characteristic polynomial $\chi(M)$ of 
+We analyze the precision of the characteristic polynomial $\chi_M$ of 
 an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
@@
 within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
-coefficient of $\chi(M)$ with more accuracy than $O(p^N)$.
+coefficient of $\chi_M$ with more accuracy than $O(p^N)$.
 In general, we provide two algorithms for determining the optimal 
-precision of the coefficients of $\chi(M)$ and of their eigenvalues.
+precision of the coefficients of $\chi_M$ and of their eigenvalues.
 We provide evidence showing that classical algorithms do not reach
@@
 \section{Introduction}
-%\begin{comment}
+\begin{comment}
 \vspace{0.1in}
@@
 \item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).
-We note that we did perform the current experiment in higher dimension with similar results: dimension $8$ was chosen to be representative without the table taking up too much space in the paper.}
+We note that we did perform the current experiment in higher dimension with similar results: dimension $9$ was chosen to be representative without the table taking up too much space in the paper.}
 \item \done{We have removed the two paragraphs before the Previous contributions section.
@@
 \bigskip
-%\end{comment}
+\end{comment}
 
@@
 The simplest approach for computing the characteristic polynomial of $M$
-is to compute $\det(X-M)$ either using recursive row expansion or various division free
+is to compute $\det(X{-}M)$ either using recursive row expansion or various division free
 algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
@@
 Instead, we separate the computation of the precision of $\chi_M$ from the computation
-of an approximation to $\chi_M$.  Given some precision on $M$, we use \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}
+of an approximation to $\chi_M$.  Given some precision on $M$, we use the theory developed in \cite{caruso-roe-vaccon:14a}
 to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
@@
 division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
-the second one may perform divisions and has complexity $\tilde 
-O(n^\omega)$;
+the second one may perform divisions and has complexity $\softO(n^\omega)$;
 \end{enumerate}
@@
 M_n(\Qp)$ (whose entries may be given at different precision): the first 
-one is division-free and has complexity $\softO(n^3)$ while the second 
+one is division-free and has complexity $\softO(n^3)$ (operations in
+$\Zp$) while the second 
 one may perform divisions but has complexity $\softO(n^\omega)$;
@@
 together with a measure of the approximation's accuracy; there are several ways to do so.
-When using \emph{jagged} precision, on gives a precision on each entry,
+When using \emph{jagged} precision, one gives a precision on each entry,
 representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
@@
 When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may give
-the precision of $\chi(M)$ precisely as
+the precision of $\chi_M$ precisely as
 \begin{equation} \label{eq:mainlem}
@@
 details on the computation of $d\chi_M$ in \eqref{eq:dchi}, and 
-examples of additional precision in $\chi(M)$,
-see \cite{caruso:17a}*{\S 2.4.3 and \S 3.2.2}
+examples of additional precision in $\chi_M$,
+see \cite{caruso:17a}*{\S 2.4.3 and \S 3.2.2}.
 
@@
 $M$ is a companion matrix.  For a companion matrix, the last column of
-$\adj(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
+$\adj(X{-}M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
 
 Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
-After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X-M)$
+After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X{-}M)$
 will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
@@
 Since $\lambda$ occurs in two Jordan blocks,
-$X - \lambda$ will divide every entry of $\adj(X-M)$ and $d\chi_M$ will not be surjective.
+$X - \lambda$ will divide every entry of $\adj(X{-}M)$ and $d\chi_M$ will not be surjective.
 \end{proof}
@@
 \begin{cor} \label{cor:prec_gain}
-Suppose that $M \in \GL_n(\OK)$ is known with precision $O(\pi^m)$.
-Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^m)$
+Suppose that $M \in \GL_n(\OK)$ is known with flat precision $O(\pi^N)$.
+Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^N)$
 if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
@@
 to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
-often be convenient to view $d \chi_M$ as an $K$-linear mapping
+often be convenient to view $d \chi_M$ as a $K$-linear mapping
 $M_n(K) \to K[X]/\chi_M(X)$.
@@
 by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
-stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
+stays within $O(n^2)$: the polynomials $\alpha$ and $\chi_M$ are 
 determined by $2n$ coefficients while we need $2n^2$ entries to 
@@
 We prove here that we may compute an approximate Hessenberg form 
-for any matrix known at finite (jagged) precision over $K$.
+for any matrix known at finite jagged precision over $K$.
 Moreover, we provide an exact change of basis matrix.
@@
 Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M$.
-$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
+$P$ is exact 
+%over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, 
+and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
 \end{prop}
 \begin{proof}
-Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
+%Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
 Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
 with the $\varepsilon$'s being units,
-the corresponding coefficient of the corresponding shear matrix is the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of $\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}.$
-Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
+the corresponding coefficient of the corresponding shear matrix is 
+%the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of 
+any exact element of $K$ congruent to 
+$\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}$.
+Exactness follows directly. 
+%Over other fields, we can not lift, but the computations are still valid.
 The rest is clear.
@@
 In this section, we prove that to compute the inverse of
-a matrix over a CDVF $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
+a matrix over $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
 We first recall the differential of matrix inversion.
@@
 Let $\cond(M) = \val(\sigma_n(M))$.
-If $dM$ is a flat precision of $O(\pi^m)$ on $M$ then $M^{-1}$
-can be computed at precision $O(\pi^{m-2\cond(M)})$ by a \textbf{SNF} computation
+If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
+can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
 and this lower-bound is optimal,
-at least when $m$ is large.
+at least when $N$ is large.
 \end{prop}
+
 \begin{proof}
-The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M).$
-It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that $m-2\cond(M)$
+The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M)$.
+%It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that 
+Consequently $N - 2\cond(M)$
 can be obtained as the valuation of a coefficient of $du_M(dM)$
-and the smallest that can be achieved this way for $dM$ in a flat precision lattice.
+and it is the smallest that can be achieved this way for $dM$ in a flat precision lattice.
 Hence the optimality of the bound given, for large 
-$m$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
+$N$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
 
 Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
-From $M$ known at flat precision $O(\pi^m),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
-known at precision at least $O(\pi^{m-\cond(M)})$, with coefficients in $\OK$
+From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
+known at precision at least $O(\pi^{N-\cond(M)})$, with coefficients in $\OK$
 and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
@@
 Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
-we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{m-2\cond(M)}),$
+we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{N-2\cond(M)}),$
 which concludes the proof.
@@
 using the Smith normal form computation of the previous section.
-The entries of $\adj(X-H)$ lie in $K[X]$, which is not a CDVF,
+The entries of $\adj(X-H)$ lie in $K[X]$, which is not a complete
+discrete valuation field,
 so we may not directly apply the methods of the previous section.
 However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
-entries lie in the CDVF $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
+entries lie in $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
 using an SNF method, with no division in $K$.
@@
 Let $M \in M_n(K).$ Then:
-\begin{eqnarray*}
-\adj(1-XM)^{\rec,n-1}=&\adj(X-M), \\
-(\chi_M I_n)^{\rec,n}=&(1-XM) \adj(1-XM).\\
-\end{eqnarray*}
+\begin{align*}
+\adj(1-XM)^{\rec,n-1}&=\adj(X-M), \\
+(\chi_M I_n)^{\rec,n}&=(1-XM) \adj(1-XM).\\
+\end{align*}
 \end{lem}
@@
 then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
-Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$
-to prove this result.
+%Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$ to prove this result.
 It directly implies the second part of the lemma; the first part follows
@@
 Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix with
-coefficient-wise precision $O(\pi^M).$
-Then, using Algorithm 2, one can compute $\adj (X -H)$
+flat precision $O(\pi^N).$
+Then, using Algorithm 2, one can compute $\adj (X{-}H)$
 and $\chi_H$ in 
-$\softO (n^3)$ operations in $\OK$ at the precision given by $H.$
+$\softO (n^3)$ operations in $\OK$ at precision $O(\pi^N).$
 \end{theo}
@@
 First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
-in $\OK$ at the precision given by $H.$
+in $\OK$ at precision $O(\pi^N).$
 Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
@@
 
-For lines 4 and 5, there are only $n-1$ eliminations, resulting in a $\softO (n^2)$ cost
-for the computation of $U.$ Rather than actually construct $P$, we just track the eliminations
+For lines 4 and 5, there are only $n{-}1$ eliminations, resulting in a $\softO (n^2)$ cost
+for the computation of $U.$ Rather than actually constructing $P$, we just track the eliminations
 performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.
 
-Line 6 is in $\softO (n^2)$ and 7 in $\softO (n^3).$
-
-Thanks to the fact that the $P$ only corresponds to the product of $n-1$ 
+Line 6 is in $\softO (n^2)$ and line 7 in $\softO (n^3).$
+
+Thanks to the fact that the $P$ only corresponds to the product of $n{-}1$ 
 shear matrices, the
@@
 In this section, we combine Proposition \ref{prop:shortcom}
-with Algorithm 2 to compute the adjugate of $X-M$
+with Algorithm 2 to compute the adjugate of $X{-}M$
 when $\chi_M$ is squarefree.   Note that, under
@@
 The idea is to work modulo $X-\lambda$
-for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
+for $\lambda$ a root of $\chi_M$ (in an algebraic closure)
 and then apply Chinese Remainder Theorem.
@@
 $d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
-$\softO (n^3)$ operations in $K$. This seems to be optimal given that the 
-(naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n
+$\softO (n^3)$ operations in $K$. This seems to be optimal given that 
+the (naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n 
 \times n^2$. It turns out however that improvements are still possible! 
-Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of 
-$d \chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
+Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of $d 
+\chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
 coefficients. The aim of this short section is to design a fast 
-algorithm (with complexity $\softO(n^\omega)$) for computing this short 
-form. The price to pay is that divisions in $K$ appear, which can be an 
-issue regarding to precision in particular cases.
-In this section, we only estimate the number of operations in $K$
+algorithm (with complexity $\softO(n^\omega)$ operations in $K$) for 
+computing this short form. The price to pay is that divisions in $K$ 
+appear, which can be an issue regarding to precision in particular 
+cases. In this section, we only estimate the number of operations in $K$ 
 and not their behavior on precision.
@@
 Given $M \in M_n(K)$ with $d \chi_M$ surjective,
-one can compute $\alpha, P, Q, \chi_M) \in K[X]$ with
+one can compute $(\alpha, P, Q, \chi_M) \in K[X]$ with
 $\softO(n^\omega)$ operations in $K$ such that
-$\adj(X{-}N) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$.
+$\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$.
 \end{theo}
@@
 we have proposed algorithms to obtain the adjugate of 
-$X-M.$ Our motivation for these computations is to then
+$X{-}M.$ Our motivation for these computations is to then
 be able to understand what is the optimal precision on $\chi_M.$
@@
 characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}.
-Note that the method for computing $\adj(X-M)$ is irrelevant
-since we only record the precision loss, not the runtime.
-\item in the capped relative model\footnote{Each 
+(Note that the method for computing $\adj(X{-}M)$ is irrelevant
+since we only record the precision loss, not the runtime.)
+\item in the capped relative mode (each 
 coefficient carries its own precision which is updated after each 
-elementary arithmetic operation.},
+elementary arithmetic operation),
 the precision on the $k$-th coefficient of the 
@@
 To do so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
-suppose that the uncertainty of $M$ is ``jagged'' in the sense that
+suppose that the uncertainty of $M$ is jagged:
 each entry of $M$ is given at some precision $O(\pi^{N_{i,j}})$.
@@
 neighborhood $\calU$ of $M$. More precisely, there exists a unique continuous 
-function $f : \calU \to K$ such that:
-\begin{enumerate}[$\bullet$]
-\renewcommand{\itemsep}{0pt}
-\item $f(M_\app) = \lambda_\app$, and
-\item $f(M')$ is an eigenvalue of $M'$ for all $M' \in \calU$.
-\end{enumerate}
+function $f : \calU \to K$ such that
+$f(M_\app) = \lambda_\app$, and
+$f(M')$ is an eigenvalue of $M'$ for all $M' \in \calU$.
 
@@
 for a cost of $\softO(ns)$ operations in $K$ using fast multipoint 
-evaluation methods (see 10.7 of \cite{gathen-gerhard:13a}).
+evaluation methods (see \S 10.7 of \cite{gathen-gerhard:13a}).
 On the other hand, we observe that $P_i Y(\lambda_k)^T$ is
@@
 \end{matrix} \right).$$
-The latter product can be computed in $\softO(n^\omega)$ operations in 
-$K$\footnote{It turns out that $\softO(n^2)$ is also possible because
-the right factor is a structured matrix (a truncated Vandermonde):
+The latter product can be computed in $\softO(n^2)$ operations in 
+$K$. Indeed,
+the right factor is a truncated Vandermonde matrix, so that
 computing the above product reduces to evaluating a polynomial at the
-points $\lambda_1, \ldots, \lambda_s$.}. Therefore all the $P_i 
+points $\lambda_1, \ldots, \lambda_s$. Therefore all the $P_i 
 Y(\lambda_k)^T$'s (for $i$ and $k$ varying) 

### 2017-04-25 charpoly/ReviewerResponse.tex
Move reviewer comments to new file.

\documentclass{article}
\usepackage{amsmath}
\usepackage{fullpage}

\DeclareMathOperator{\adj}{Adj}

\newcommand{\softO}{O\tilde{~}}
\newcommand{\done}[1]{#1}

\title{Response to Reviewer Comments for \\ ``Characteristic polynomials of $p$-adic matrices''}
\author{Xavier Caruso, David Roe and Tristan Vaccon}

\begin{document}
\maketitle

\vspace{0.1in}
\noindent \textbf{Reviewer 1:}
\begin{enumerate}
\item \done{We do not see a direct application of this paper to Dixon's algorithm, since his algorithm solves a $p$-adic system $p$-adically rather than computing a characteristic polynomial.  Moreover, since the prime can be chosen arbitrarily, it will usually not have repeated eigenvalues modulo $p$ and thus there will be no gain in precision.}
\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time}
\item \done{In section 4, we clarified how the characteristic polynomial is a by-product of [15].}
\item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
\item \done{We clarified why we chose the distribution in Figure 1.}
\item \done{We do not see how to apply Kedlaya-Umans.  Yes, there is multipoint evaluation used in Proposition 5.6, but it is dominated by the computation of the matrix $P$, which requires $\softO(n^\omega)$ operations in $K$.}
\end{enumerate}
\textbf{Reviewer 2:}
\begin{enumerate}
\item \done{We adjusted the reference as suggested, as also pointed out by Reviewers 3 and 5.}
\end{enumerate}
\textbf{Reviewer 3:}
\begin{enumerate}
\item \done{We have added Remark 3.1 outlining why we do not need much accuracy in the computation of $\adj(M{-}X)$: it's roughly true that only the valuations of the entries are relevant, since we only care about the lattice that the entries generate.  We hope that this remark is sufficient to convince the reviewer that the lack of precision analysis in the discussion preceding Theorem 4.1 does not invalidate the claimed complexity.}
\item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
\item \done{We have switched to using the term ``adjugate'' rather than ``comatrix.''}
\item \done{We have given a definition of the companion matrix in our notation section and adjusted the proof of Proposition 2.1 accordingly.}
\item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
\item \done{We have switched from $M^t$ to $M^T$ to denote the transpose.}
\item \done{We have added a reference to course notes of Caruso (now reference [2]) which go into more detail on how the formula for the differential of $\chi_M$ is derived.}
\item \done{In section 5.1, we added a remark noting that the method for computing the adjugate matrix has no effect on the extent of the precision loss, just on the runtime.}
\item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).  We note that we did perform the current experiment in higher dimension with similar results: dimension $8$ was chosen to be representative without the table taking up too much space in the paper.}
\item \done{We have removed the two paragraphs before the Previous contributions section.  Some of the material discussed therein has been moved to Section 2.1, and some is now referenced in the course notes [2]}
\item \done{We have carried out the other minor corrections suggested.}
\end{enumerate}
\textbf{Reviewer 5:}
\begin{enumerate}
\item \done{The abstract has been rewritten. In particular, the kind of operations we are counted and the locution ``precision exactly $O(p^k)$'' have been clarified.}
\item \done{We have also clarified what kind of operations are being counted and what is meant by ``precision $O(p^N)$`` in the introduction.}
\item \done{We have added a discussion of the term \emph{optimal} precision in Section 2.1, after equation (3).}
\item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
\item \done{We have expanded upon the discussion of lattices, and removed any reference to tangent spaces (since they were a distraction for matrices and polynomials).}
\item \done{We have fleshed out the definitions of flat and jagged precision.}
\end{enumerate}
\textbf{General clarity comments:}
%\begin{enumerate}
%\item \tdo{The writing is rather dense and relies too heavily on external material, which makes the reading sometimes unpleasant. (3)}
%\item \tdo{Overall the presentation lacks a minimal set of definitions and terminology regarding the notions of p-adic rings and their precision to make it sufficiently self-contained for an non-expert reader. I believe only a one or two additional paragraphs could be sufficient to drastically improve the presentation of section 2, in particular recalling the various precision models.  The proofs are very concise and rely too heavily on a good command of p-adic precision terminology. (3)}
%\item \tdo{The paper itself is unnecessarily difficult to read, to the extent that it sometimes unclear what is being proven or to verify correctness.  A number of key definitions are not given and there are too few motivating/clarifying examples. (5)}
%\item \tdo{The definitions are unclear and the comparison with previous work is deficient.  If this is to appear the definitions and results must be made very clear, and the results must be demonstrated to be superior to the division-free algorithms cited, both in theory and in practice.}
%\end{enumerate}

\begin{enumerate}
\item \done{Section \emph{The contribution of this paper} of the introduction
has been rewritten with additional detail. We split our contributions into two 
parts: the first part is about the (fast) computation of the adjugate
of $X{-}M$ while the second part concerns applications to precision.
We also added a sentence about numerical experiments.}
\item \done{Section 2.1 has been rewritten. We no longer talk about tangent
spaces. Jagged and flat precision, together with their relationships with
lattice precision, are introduced and discussed. 
Proposition 2.1 has been moved to Section 2.1 and stated
in terms of valuations.}
\item \done{Small adjustments have been made throughout the paper to attempt
to make the prose more readable, though we are constrained by the page
limit, which hinders us from adding more examples or details.}
\end{enumerate}

\end{document}

### 2017-04-25 charpoly/charpoly.tex
Move reviewer comments to new file.

--- 2017-04-25 charpoly/charpoly.tex
@@
 \section{Introduction}
-%\begin{comment}
-\vspace{0.1in}
-\noindent \textbf{Reviewer 1:}
-\begin{enumerate}
-\item \done{We do not see a direct application of this paper to Dixon's algorithm, since his algorithm solves a $p$-adic system $p$-adically rather than computing a characteristic polynomial.
-Moreover, since the prime can be chosen arbitrarily, it will usually not have repeated eigenvalues modulo $p$ and thus there will be no gain in precision.}
-\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time}
-\item \done{In section 4, we clarified how the characteristic polynomial is a by-product of [15].}
-\item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
-\item \done{We clarified why we chose the distribution in Figure 1.}
-\item \done{We do not see how to apply Kedlaya-Umans.
-Yes, there is multipoint evaluation used in Proposition 5.6, but it is dominated by the computation of the matrix $P$, which requires $\softO(n^\omega)$ operations in $K$.}
-\end{enumerate}
-\textbf{Reviewer 2:}
-\begin{enumerate}
-\item \done{We adjusted the reference as suggested, as also pointed out by Reviewers 3 and 5.}
-\end{enumerate}
-\textbf{Reviewer 3:}
-\begin{enumerate}
-\item \done{We have added Remark 3.1 outlining why we do not need much accuracy in the computation of $\adj(M{-}X)$: it's roughly true that only the valuations of the entries are relevant, since we only care about the lattice that the entries generate.
-We hope that this remark is sufficient to convince the reviewer that the lack of precision analysis in the discussion preceding Theorem 4.1 does not invalidate the claimed complexity.}
-\item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
-\item \done{We have switched to using the term ``adjugate'' rather than ``comatrix.''}
-\item \done{We have given a definition of the companion matrix in our notation section and adjusted the proof of Proposition 2.1 accordingly.}
-\item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
-\item \done{We have switched from $M^t$ to $M^T$ to denote the transpose.}
-\item \done{We have added a reference to course notes of Caruso (now reference [2]) which go into more detail on how the formula for the differential of $\chi_M$ is derived.}
-\item \done{In section 5.1, we added a remark noting that the method for computing the adjugate matrix has no effect on the extent of the precision loss, just on the runtime.}
-\item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).
-We note that we did perform the current experiment in higher dimension with similar results: dimension $8$ was chosen to be representative without the table taking up too much space in the paper.}
-\item \done{We have removed the two paragraphs before the Previous contributions section.
-Some of the material discussed therein has been moved to Section 2.1, and some is now referenced in the course notes [2]}
-\item \done{We have carried out the other minor corrections suggested.}
-\end{enumerate}
-\textbf{Reviewer 5:}
-\begin{enumerate}
-\item \done{The abstract has been rewritten. In particular, the kind of operations we are counted and the locution ``precision exactly $O(p^k)$'' have been clarified.}
-\item \done{We have also clarified what kind of operations are being counted and what is meant by ``precision $O(p^N)$`` in the introduction.}
-\item \done{We have added a discussion of the term \emph{optimal} precision in Section 2.1, after equation (3).}
-\item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
-\item \done{We have expanded upon the discussion of lattices, and removed any reference to tangent spaces (since they were a distraction for matrices and polynomials).}
-\item \done{We have fleshed out the definitions of flat and jagged precision.}
-\end{enumerate}
-\textbf{General clarity comments:}
-%\begin{enumerate}
-%\item \tdo{The writing is rather dense and relies too heavily on external material, which makes the reading sometimes unpleasant. (3)}
-%\item \tdo{Overall the presentation lacks a minimal set of definitions and terminology regarding the notions of p-adic rings and their precision to make it sufficiently self-contained for an non-expert reader.
-I believe only a one or two additional paragraphs could be sufficient to drastically improve the presentation of section 2, in particular recalling the various precision models.
-The proofs are very concise and rely too heavily on a good command of p-adic precision terminology.
-(3)}
-%\item \tdo{The paper itself is unnecessarily difficult to read, to the extent that it sometimes unclear what is being proven or to verify correctness.
-A number of key definitions are not given and there are too few motivating/clarifying examples.
-(5)}
-%\item \tdo{The definitions are unclear and the comparison with previous work is deficient.
-If this is to appear the definitions and results must be made very clear, and the results must be demonstrated to be superior to the division-free algorithms cited, both in theory and in practice.}
-%\end{enumerate}
-
-\begin{enumerate}
-\item \done{Section \emph{The contribution of this paper} of the introduction
-has been rewritten with additional detail. We split our contributions into two 
-parts: the first part is about the (fast) computation of the adjugate
-of $X{-}M$ while the second part concerns applications to precision.
-We also added a sentence about numerical experiments.}
-\item \done{Section 2.1 has been rewritten. We no longer talk about tangent
-spaces. Jagged and flat precision, together with their relationships with
-lattice precision, are introduced and discussed. 
-Proposition 2.1 has been moved to Section 2.1 and stated
-in terms of valuations.}
-\item \done{Small adjustments have been made throughout the paper to attempt
-to make the prose more readable, though we are constrained by the page
-limit, which hinders us from adding more examples or details.}
-\end{enumerate}
-
-\bigskip
-%\end{comment}
 

### 2017-04-26 charpoly/charpoly.tex

--- 2017-04-25 charpoly/charpoly.tex
@@
 
-%\vspace{1mm}
-% \noindent
-% {\bf Categories and Subject Descripto\RS:} \\
-%\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
-%  Manipulation -- \emph{Algebraic Algorithms}
-%
-% \vspace{1mm}
-% \noindent
-% {\bf General Terms:} Algorithms, Theory
-%
-% \vspace{1mm}
-% \noindent
-% {\bf Keywords:} $p$-adic precision, linear algebra, ultrametric analysis
-%\medskip
-
 \section{Introduction}
-\begin{comment}
-\vspace{0.1in}
-\noindent \textbf{Reviewer 1:}
-\begin{enumerate}
-\item \done{We do not see a direct application of this paper to Dixon's algorithm, since his algorithm solves a $p$-adic system $p$-adically rather than computing a characteristic polynomial.
-Moreover, since the prime can be chosen arbitrarily, it will usually not have repeated eigenvalues modulo $p$ and thus there will be no gain in precision.}
-\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time}
-\item \done{In section 4, we clarified how the characteristic polynomial is a by-product of [15].}
-\item \done{In section 5.1, we replaced $\pi$ by $\beta$.}
-\item \done{We clarified why we chose the distribution in Figure 1.}
-\item \done{We do not see how to apply Kedlaya-Umans.
-Yes, there is multipoint evaluation used in Proposition 5.6, but it is dominated by the computation of the matrix $P$, which requires $\softO(n^\omega)$ operations in $K$.}
-\end{enumerate}
-\textbf{Reviewer 2:}
-\begin{enumerate}
-\item \done{We adjusted the reference as suggested, as also pointed out by Reviewers 3 and 5.}
-\end{enumerate}
-\textbf{Reviewer 3:}
-\begin{enumerate}
-\item \done{We have added Remark 3.1 outlining why we do not need much accuracy in the computation of $\adj(M{-}X)$: it's roughly true that only the valuations of the entries are relevant, since we only care about the lattice that the entries generate.
-We hope that this remark is sufficient to convince the reviewer that the lack of precision analysis in the discussion preceding Theorem 4.1 does not invalidate the claimed complexity.}
-\item \done{We have updated the references for characteristic polynomial using Frobenius normal form as suggested.}
-\item \done{We have switched to using the term ``adjugate'' rather than ``comatrix.''}
-\item \done{We have given a definition of the companion matrix in our notation section and adjusted the proof of Proposition 2.1 accordingly.}
-\item \done{We have clarified the use of ``repeated eigenvalue'' in the proof of Proposition 2.1.}
-\item \done{We have switched from $M^t$ to $M^T$ to denote the transpose.}
-\item \done{We have added a reference to course notes of Caruso (now reference [2]) which go into more detail on how the formula for the differential of $\chi_M$ is derived.}
-\item \done{In section 5.1, we added a remark noting that the method for computing the adjugate matrix has no effect on the extent of the precision loss, just on the runtime.}
-\item \done{We agree that further experiments would be beneficial, but we are constrained by space (due to the 8 page limit) and time (due to the limited time to make revisions).
-We note that we did perform the current experiment in higher dimension with similar results: dimension $9$ was chosen to be representative without the table taking up too much space in the paper.}
-\item \done{We have removed the two paragraphs before the Previous contributions section.
-Some of the material discussed therein has been moved to Section 2.1, and some is now referenced in the course notes [2]}
-\item \done{We have carried out the other minor corrections suggested.}
-\end{enumerate}
-\textbf{Reviewer 5:}
-\begin{enumerate}
-\item \done{The abstract has been rewritten. In particular, the kind of operations we are counted and the locution ``precision exactly $O(p^k)$'' have been clarified.}
-\item \done{We have also clarified what kind of operations are being counted and what is meant by ``precision $O(p^N)$`` in the introduction.}
-\item \done{We have added a discussion of the term \emph{optimal} precision in Section 2.1, after equation (3).}
-\item \done{We've switched from using ``comatrix'' to using ``adjugate,'' as also requested by reviewer 3.}
-\item \done{We have expanded upon the discussion of lattices, and removed any reference to tangent spaces (since they were a distraction for matrices and polynomials).}
-\item \done{We have fleshed out the definitions of flat and jagged precision.}
-\end{enumerate}
-\textbf{General clarity comments:}
-%\begin{enumerate}
-%\item \tdo{The writing is rather dense and relies too heavily on external material, which makes the reading sometimes unpleasant. (3)}
-%\item \tdo{Overall the presentation lacks a minimal set of definitions and terminology regarding the notions of p-adic rings and their precision to make it sufficiently self-contained for an non-expert reader.
-I believe only a one or two additional paragraphs could be sufficient to drastically improve the presentation of section 2, in particular recalling the various precision models.
-The proofs are very concise and rely too heavily on a good command of p-adic precision terminology.
-(3)}
-%\item \tdo{The paper itself is unnecessarily difficult to read, to the extent that it sometimes unclear what is being proven or to verify correctness.
-A number of key definitions are not given and there are too few motivating/clarifying examples.
-(5)}
-%\item \tdo{The definitions are unclear and the comparison with previous work is deficient.
-If this is to appear the definitions and results must be made very clear, and the results must be demonstrated to be superior to the division-free algorithms cited, both in theory and in practice.}
-%\end{enumerate}
-
-\begin{enumerate}
-\item \done{Section \emph{The contribution of this paper} of the introduction
-has been rewritten with additional detail. We split our contributions into two 
-parts: the first part is about the (fast) computation of the adjugate
-of $X{-}M$ while the second part concerns applications to precision.
-We also added a sentence about numerical experiments.}
-\item \done{Section 2.1 has been rewritten. We no longer talk about tangent
-spaces. Jagged and flat precision, together with their relationships with
-lattice precision, are introduced and discussed. 
-Proposition 2.1 has been moved to Section 2.1 and stated
-in terms of valuations.}
-\item \done{Small adjustments have been made throughout the paper to attempt
-to make the prose more readable, though we are constrained by the page
-limit, which hinders us from adding more examples or details.}
-\end{enumerate}
-
-\bigskip
-\end{comment}
 
@@
 \cite{kedlaya:01a, harvey:07a, harvey:14a}.
-
-%\todo{Application to characteristic polynomials of elements of p-adic field extensions?}
 
@@
 for the characteristic polynomial, but do not give practical algorithms
-to attain this optimal precision.
+to compute this optimal precision.
 
@@
 precision in many contexts.
+Note that we focus on computing the precision of $\chi_M$: we do not give a stable
+algorithm for computing the characteristic polynomial itself.
 
@@
 theory of precision developed in \cite{caruso-roe-vaccon:14a}
-and apply it to the specific case of the characteristic polynomial,
-giving conditions under which the differential will be surjective
-(and thus provide a good measure of precision).  We also give
-a condition based on reduction modulo $p$ that determines
-whether the characteristic polynomial will have a higher
-precision than the input matrix, and show that the image
-of the set of integral matrices has the structure of an $\OK[X]$-module
-when $M$ is itself integral. Finally, we give a compact description
+and apply it to the specific case of the characteristic polynomial.
+We also describe when the characteristic polynomial will have a higher
+precision than the input matrix and give a compact description
 of $\adj(X-M)$, the main ingredient in the differential.
@@
 which can be computed in $O(n^\omega)$ operations in $K.$
-This is faster than computing $U \adj(X-M) U^{-1},$
-which is, at first sight, in $\softO(n^4)$
-%\tristan{Shouldn't it be $\softO (n^4)$?}
-operations in $K.$
+This is faster than computing $U \adj(X-M) U^{-1}$ using the dense
+representation, which requires $\softO(n^{\omega+1})$ operations in $K.$
 \end{rem}
@@
 by invertible elements of $\OK$ will occur.
-%The computation of a Hessenberg form provides an $O(n^3)$
-%algorithm to compute the characteristic polynomial.
 
@@
 $\adj(X{-}M)$ (roughly only their valuations matter).
-% should we argue more than that?
 \end{rem}
@@
 $P$ is exact  
-%over finite extensions of $\mathbb{Q}_p$ and $k\llp X\rrp$, 
 and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
@@
 \begin{proof}
-%Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k\llp X \rrp.$
 Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
@@
 the corresponding coefficient of the corresponding shear matrix is 
-%the lift(in $\mathbb{Z}, $  $\Fq[X],$ $\Q[X]$ or adequate extension) of 
 any exact element of $K$ congruent to 
@@
 Exactness follows directly. 
-%Over other fields, we can not lift, but the computations are still valid.
 The rest is clear.
@@
 The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M)$.
-%It is $-2\cond(M)$ for $M^{-2}$ and it is then clear that 
-Consequently $N - 2\cond(M)$
-can be obtained as the valuation of a coefficient of $du_M(dM)$
+Consequently, $N - 2\cond(M)$
+can be obtained as the valuation of a coefficient of $du_M(dM)$,
 and it is the smallest that can be achieved this way for $dM$ in a flat precision lattice.
-Hence the optimality of the bound given, for large 
+Hence the optimality of the bound given for large 
 $N$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
@@
 then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
-%Indeed, one can use multilinearity of the determinant on $X^d \det(F(1/X))$ to prove this result.
 It directly implies the second part of the lemma; the first part follows
@@
 
+Note that the precision behavior described persists when using algorithms
+other than the division-free method.
+
 \item in the model of floating-point arithmetic (see 
@@
 
-%\begin{rem}
-%The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
-%division free algorithm of \cite{seifullin:02a}. It is likely that, proceeding so, we 
-%limit the loss of precision.
-%\end{rem}
-
 \noindent
 The table of Figure~\ref{fig:exp} summarizes the results obtained. 
-%
 \begin{figure}
@@
 $\chi'_M(\lambda) \cdot d \lambda + \tr(\adj(X-M) \cdot dM)(\lambda) = 0$.
-%from which the Lemma follows.
 \end{proof}

### 2017-05-09 charpoly/charpoly.tex
Using acmart class, but citations are weird

@@
-\documentclass{sig-alternate-05-2015}
+\documentclass[sigconf]{acmart}
 
@@
 \usepackage{amsmath,amssymb}
-\usepackage{amsthmnoproof}
+%\usepackage{amsthmnoproof}
+\usepackage{amsthm}
 \usepackage{mathrsfs}
+\let\bibsection\relax
 \usepackage{amsrefs}
-\usepackage[usenames,dvipsnames]{color}
+%\usepackage[usenames,dvipsnames]{color}
 \usepackage{stmaryrd}
@@
 \usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
-\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
+%\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
 \usepackage{comment}
@@
 
-\CopyrightYear{2017}
-\setcopyright{acmcopyright}
-\conferenceinfo{ISSAC '17,}{July 25-28, 2017, Kaiserslautern, Germany}
-\isbn{***-*-****-****-0/17/07}\acmPrice{\$15.00}
-\doi{http://dx.doi.org/10.1145/*******.*******}
-
-\permission{Publication rights licensed to ACM.
-ACM acknowledges that this contribution was authored or co-authored by an employee, contractor or affiliate of a national government.
-As such, the Government retains a nonexclusive, royalty-free right to publish or reproduce this article, or to allow others to do so, for Government purposes only.}
-
-\begin{document}
+\clubpenalty=10000
+\widowpenalty = 10000
 
@@
 
+\begin{document}
+
 \title{Characteristic polynomials of p-adic matrices}
 
-\numberofauthors{3}
-\author{
-\alignauthor Xavier Caruso\\
-  \affaddr{Universit\'e Rennes 1}\\
-  \affaddr{\textsf{xavier.caruso@normalesup.org}}
-\alignauthor David Roe\\
-  \affaddr{University of Pittsburg}\\
-  \affaddr{\textsf{roed@pitt.edu}}
-\alignauthor Tristan Vaccon\\
-  \affaddr{Universit\'e de Limoges}\\
-  \affaddr{\textsf{tristan.vaccon@unilim.fr}}
-}
-
-\maketitle
-
-\begin{abstract}
-We analyze the precision of the characteristic polynomial $\chi_M$ of 
-an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
-developed previously.  When $M$ is an integral matrix whose entries are 
-all given at the same precision $O(p^N)$, we give a criterion (checkable 
-within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
-coefficient of $\chi_M$ with more accuracy than $O(p^N)$.
-In general, we provide two algorithms for determining the optimal 
-precision of the coefficients of $\chi_M$ and of their eigenvalues.
-We provide evidence showing that classical algorithms do not reach
-this optimal precision in general.
-\end{abstract}
+\author{Xavier Caruso}
+\affiliation{Universit\'e Rennes 1}
+\email{xavier.caruso@normalesup.org}
+\author{David Roe}
+\affiliation{University of Pittsburg}
+\email{roed@pitt.edu}
+\author{Tristan Vaccon}
+\affiliation{Universit\'e de Limoges}
+\email{tristan.vaccon@unilim.fr}
+
+\copyrightyear{2017}
+\acmYear{2017}
+\setcopyright{acmlicensed}
+\acmConference{ISSAC '17}{July 25-28, 2017}{Kaiserslautern, Germany}
+\acmPrice{15.00}
+\acmDOI{http://dx.doi.org/10.1145/3087604.3087618}
+\acmISBN{978-1-4503-5064-8/17/07}
 
@@
 
-\vspace{-1mm}
 \ccsdesc[500]{Computing methodologies~Algebraic algorithms}
-\printccsdesc
-
-\vspace{-1.5mm}
+
 \keywords{Algorithms, $p$-adic precision, characteristic polynomial,
 eigenvalue}
+
+\begin{abstract}
+We analyze the precision of the characteristic polynomial $\chi_M$ of 
+an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
+developed previously.  When $M$ is an integral matrix whose entries are 
+all given at the same precision $O(p^N)$, we give a criterion (checkable 
+within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
+coefficient of $\chi_M$ with more accuracy than $O(p^N)$.
+In general, we provide two algorithms for determining the optimal 
+precision of the coefficients of $\chi_M$ and of $M$'s eigenvalues.
+We provide evidence showing that classical algorithms do not reach
+this optimal precision in general.
+\end{abstract}
+
+\maketitle
 

### 2017-05-09 charpoly/charpoly.tex
Remove amsrefs

@@
 \usepackage{mathrsfs}
-\let\bibsection\relax
-\usepackage{amsrefs}
+%\let\bibsection\relax
+%\usepackage{amsrefs}
 %\usepackage[usenames,dvipsnames]{color}
@@
 \usepackage{enumerate}
-\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
+%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
 %\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
@@
 \usepackage{multirow}
-\usepackage{tikz}
+%\usepackage{tikz}
 
@@
 algorithms for counting points of varieties over finite fields (see
-\cite{kedlaya:01a, harvey:07a, harvey:14a}.
+\cite{kedlaya:01a, harvey:07a, harvey:14a}).
 
@@
 is to compute $\det(X{-}M)$ either using recursive row expansion or various division free
-algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
+algorithms \cite{seifullin:02a,kaltofen:92a}.  There are two issues
 with these methods.  First, they are slower than alternatives that allow division,
@@
 A faster approach over a field is to compute the Frobenius normal form of $M$,
-which is achievable in running time $\softO(n^\omega)$ \citelist{\cite{keller-gehrig:85a} \cite{pernet-storjohann:07a}}.
+which is achievable in running time $\softO(n^\omega)$ \cite{keller-gehrig:85a,pernet-storjohann:07a}.
 However, the use of division frequently leads to catastrophic losses of precision.
@@
 
-\subsection*{Previous contributions.}
-
+\medskip
+
+\noindent
+{\bf Previous contributions.}
 Since the description of Kedlaya's algorithm in
@@
 
-\subsection*{The contribution of this paper.}
-
+\medskip
+
+\noindent
+{\bf The contribution of this paper.}
 This paper provides a theoretical and concrete study of the propagation 
@@
 
-\subsection*{Organization of the article.}
-
+\medskip
+
+\noindent
+{\bf Organization of the article.}
 In Section \ref{sec:theo_study}, we review the differential
@@
 
-\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
+\medskip
+
+\noindent
+{\bf Notations.} 
+Throughout the paper, $K$ will refer to a complete,
 discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
@@
 space basis, or equivalently the image of $M_n(\OK)$ under a 
-$K$-linear transformation of $M_n(K)$ \cite{caruso-roe-vaccon:14a}*{\S 2.2}.
+$K$-linear transformation of $M_n(K)$ \cite[\S 2.2]{caruso-roe-vaccon:14a}.
 For $M \in M_n(K)$, cosets of the form $M + H$ are special
@@
 The benefit of working with lattices is that they behave well under 
-differentiable maps \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}, but they
+differentiable maps \cite[Lemma 3.4]{caruso-roe-vaccon:14a}, but they
 require more space and time to compute.
@@
 Our analysis of the precision behavior of $\chi$ rests upon
-the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
-Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to as the \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
-we have $\adj(M) {=} \det(M) M^{-1}$).  Recall
-\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
+the computation of its derivative $d\chi$, using \cite[Lemma 3.4]{caruso-roe-vaccon:14a}
+Let $\adj(M)$ denote the adjugate\footnote{In \cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a} the adjugate was referred to as the \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
+we have $\adj(M) {=} \det(M) M^{-1}$).  Recall from
+\cite[Appendix B]{caruso-roe-vaccon:14a} and \cite[\S 3.3]{caruso-roe-vaccon:15a}
 that $d\chi_M$ is given by
@@
 \begin{proof}
-Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
+Recall \cite[Definition 3.3]{caruso-roe-vaccon:15a} that the \emph{precision polygon}
 of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
-By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
+By \cite[Proposition 3.4]{caruso-roe-vaccon:15a}, the endpoints of the precision polygon
 occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
@@
 Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
-with integral coefficients, \cite{caruso-roe-vaccon:15a}*{Prop. 2.2} implies
+with integral coefficients, \cite[Proposition 2.2]{caruso-roe-vaccon:15a} implies
 the conclusion.
@@
 examples of additional precision in $\chi_M$,
-see \cite{caruso:17a}*{\S 2.4.3 and \S 3.2.2}.
+see \cite[\S\S 2.4.3, 3.2.2]{caruso:17a}.
 
@@
 The equivalence of \eqref{pt:cyc}, \eqref{pt:1d}, and \eqref{pt:charmin} is standard; see \emph{e.g.}
-\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1}.  We now show
+\cite[\S 7.1]{hoffman-kunze:LinearAlgebra}.  We now show
 \eqref{pt:cyc} $\Rightarrow$ \eqref{pt:surj} and \eqref{pt:surj} $\Rightarrow$ \eqref{pt:1d}
@@
 that one can compute the characteristic polynomial of 
-$M$ in $O(n^3)$ operations in $K$ \cite{Cohen:2013}*{pp. 55--56}.
+$M$ in $O(n^3)$ operations in $K$ \cite[pp. 55--56]{Cohen:2013}
 However, this computation involves division, and its
@@
 Hence the optimality of the bound given for large 
-$N$ \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
+$N$ \cite[Lemma 3.4]{caruso-roe-vaccon:14a}.
 
@@
 polynomial to its coefficients in $X^k$. By applying 
-\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} to the composite $\beta_k 
+\cite[Lemma 3.4]{caruso-roe-vaccon:14a} to the composite $\beta_k 
 \circ \chi_M$, one can figure out the optimal precision on the
@@
 
-\vspace{-2mm}
+%\vspace{-2mm}
 
@@
 
+\smallskip
+
 \hfill$M\texttt{.charpoly(algorithm="df")}.$\hfill\null
 
+\smallskip
+
+\noindent 
 Note that the precision behavior described persists when using algorithms
@@
 \item in the model of floating-point arithmetic (see 
-\cite{caruso:17a}*{\S 2.3}), the number of correct digits of the 
+\cite[\S 2.3]{caruso:17a}), the number of correct digits of the 
 $k$-th coefficient of the characteristic polynomial of $M$.

### 2017-06-12 charpoly/charpoly.tex
Complied with editor's requests.
Fixing small typos

@@
 \item Proposition \ref{prop:shortcom}: a short representation of
-the matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
+the adjugate matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
 field (instead of $O(n^3)$ for a dense representation);
@@
 
-\subsection{Several facts about the differential of $\chi$}
+\subsection{Several facts about the differential of \texorpdfstring{$\chi$}{x}}
 
@@
 \begin{proof}
-Inside the nested \textbf{for} loop, if we want to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
-with the $\varepsilon$'s being units,
+Inside the nested \textbf{for} loop, to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
 the corresponding coefficient of the corresponding shear matrix is 
@@
 \begin{proof}
-The smallest valuation of a coefficient of $M^{-1}$ is $-\cond(M)$.
+The smallest valuation of an entry of $M^{-1}$ is $-\cond(M)$.
 Consequently, $N - 2\cond(M)$
-can be obtained as the valuation of a coefficient of $du_M(dM)$,
+can be obtained as the valuation of an entry of $du_M(dM)$,
 and it is the smallest that can be achieved this way for $dM$ in a flat precision lattice.
@@
 
-\subsection{The adjugate of $X{-}H$}
+\subsection{The adjugate of \texorpdfstring{$X{-}H$}{X-H}}
 
@@
 \begin{proof}
-It all comes down to the following result:
-let $F \in M_d(K[X])$ a matrix of polynomials of degree at most $1,$
-then $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
-It directly implies the second part of the lemma; the first part follows
+For any matrix $F \in M_d(K[X])$ of polynomials of degree at most $1$,
+we have $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
+This result directly implies the second part of the lemma; the first part follows
 from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
@@
 
-\subsection{The adjugate of $X{-}M$}
+\subsection{The adjugate of \texorpdfstring{$X{-}M$}{X-M}}
 

### 2017-11-16 ZpL/ZpL.tex
Plan (first tentative) of the article on the package ZpL

--- 2017-06-12 charpoly/charpoly.tex
@@
 
+\newcommand{\ZpL}{\texttt{ZpL}}
+\newcommand{\ZpLCA}{\texttt{ZpLCA}}
+\newcommand{\ZpLCR}{\texttt{ZpLCR}}
+\newcommand{\ZpLF}{\texttt{ZpLF}}
+
 \newcommand{\famN}{\mathcal{N}}
@@
 
+\fancyhead{}
+
 \begin{document}
 
-\title{Characteristic polynomials of p-adic matrices}
+\title{ZpL: a p-adic precision package}
 
 \author{Xavier Caruso}
-\affiliation{Universit\'e Rennes 1}
+  \affiliation{Universit\'e Rennes 1; \\
+  \institution{IRMAR}
+  \city{Rennes, France}
+  \postcode{35042}
+}
 \email{xavier.caruso@normalesup.org}
 \author{David Roe}
-\affiliation{University of Pittsburg}
+  \affiliation{University of Pittsburg; \\
+  \institution{Department of Mathematics}
+  \city{Pittsburgh, PA, USA}
+  \postcode{15260}
+}
 \email{roed@pitt.edu}
 \author{Tristan Vaccon}
-\affiliation{Universit\'e de Limoges}
+  \affiliation{Universit\'e de Limoges; \\
+  \institution{CNRS, XLIM UMR 7252}
+  \city{Limoges, France}  
+  \postcode{87060}  
+  }
 \email{tristan.vaccon@unilim.fr}
-
-\copyrightyear{2017}
-\acmYear{2017}
-\setcopyright{acmlicensed}
-\acmConference{ISSAC '17}{July 25-28, 2017}{Kaiserslautern, Germany}
-\acmPrice{15.00}
-\acmDOI{http://dx.doi.org/10.1145/3087604.3087618}
-\acmISBN{978-1-4503-5064-8/17/07}
-
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10010147.10010148.10010149.10010150</concept_id>
-<concept_desc>Computing methodologies~Algebraic algorithms</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
 
@@
 \begin{abstract}
-We analyze the precision of the characteristic polynomial $\chi_M$ of 
-an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
-developed previously.  When $M$ is an integral matrix whose entries are 
-all given at the same precision $O(p^N)$, we give a criterion (checkable 
-within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
-coefficient of $\chi_M$ with more accuracy than $O(p^N)$.
-In general, we provide two algorithms for determining the optimal 
-precision of the coefficients of $\chi_M$ and of $M$'s eigenvalues.
-We provide evidence showing that classical algorithms do not reach
-this optimal precision in general.
 \end{abstract}
@@
 
-The characteristic polynomial is a fundamental invariant of a matrix: its roots
-give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
-In fact, the best known division-free algorithm for computing determinants
-over arbitrary rings \cite{kaltofen-villard:04a} does so using the characteristic polynomial.
-Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
-algorithms for counting points of varieties over finite fields (see
-\cite{kedlaya:01a, harvey:07a, harvey:14a}).
+\begin{itemize}
+\item Interval arithmetic, floating point arithmetic
+\item short introduction of the precision lemma and lattices
+\item goal: tracking precision using this theory
+\item short presentation of the package
+\item plan of the article
+\end{itemize}
 
-When computing with $p$-adic matrices, entries may only be approximated
-at some finite precision $O(p^N)$.  As a
-consequence, in designing algorithms for such matrices one must analyze
-not only the running time of the algorithm but also the accuracy of the result.
+\section{Examples}
 
-Let $M \in M_n(\Qp)$, with every entry given at precision $O(p^N)$.
-The simplest approach for computing the characteristic polynomial of $M$
-is to compute $\det(X{-}M)$ either using recursive row expansion or various division free
-algorithms \cite{seifullin:02a,kaltofen:92a}.  There are two issues
-with these methods.  First, they are slower than alternatives that allow division,
-requiring $O((n+1)!)$, $\softO(n^5)$ and $\softO(n^{3 + \omega/2})$ operations in $K$.  Second,
-while the lack of division implies that the result is accurate modulo $p^N$ as
-long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.
+\begin{itemize}
+\item Elementary arithmetic
+\item Determinant + matrix multiplication
+\item gcd of polynomials
+\item some benchmarks on precision
+\end{itemize}
 
-A faster approach over a field is to compute the Frobenius normal form of $M$,
-which is achievable in running time $\softO(n^\omega)$ \cite{keller-gehrig:85a,pernet-storjohann:07a}.
-However, the use of division frequently leads to catastrophic losses of precision.
-In many examples, no precision remains at the end of the calculation.
+\section{Theory}
 
-Instead, we separate the computation of the precision of $\chi_M$ from the computation
-of an approximation to $\chi_M$.  Given some precision on $M$, we use the theory developed in \cite{caruso-roe-vaccon:14a}
-to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
-of much of this paper.  With this precision known, the actual calculation of $\chi_M$
-may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
-stable algorithm (see Remark \ref{rem:lift_for_optimal}).
+\subsection{The precision Lemma}
 
-\medskip
+\subsection{The way the precision is tracked with \ZpL}
 
-\noindent
-{\bf Previous contributions.}
-Since the description of Kedlaya's algorithm in
-\cite{kedlaya:01a}, the computation of characteristic polynomials
-over $p$-adic numbers
-has become a crucial ingredient in many counting-points algorithms.
-For example, \cite{harvey:07a, harvey:14a} use $p$-adic cohomology
-and the characteristic polynomial of Frobenius to compute zeta functions
-of hyperelliptic curves.
+We follow the precision lattice.
+Closely related to automatic differentiation.
 
-In most of these papers, the precision analysis usually
-dwells on how to obtain the matrices (\textit{e.g.} of action of Frobenius)
-that are involved in the point-counting schemes.
-However, the computation of their characteristic polynomials 
-attracts less attention: some
-refer to fast algorithms (using division), while others
-apply division-free algorithms.
+\begin{itemize}
+\item \ZpLCR, \ZpLCA: we use a cap to ensure that we have a lattice at
+each step.
+\item \ZpLF: we work with $\Zp$-submodules of possible positive codimension
+and represent them using matrices of floating point $p$-adics.
+\end{itemize}
 
-In \cite{caruso-roe-vaccon:15a}, the authors have begun the application
-of the theory of differential precision of \cite{caruso-roe-vaccon:14a}
-to the stable computation of characteristic polynomials.
-They describe the optimal precision
-for the characteristic polynomial, but do not give practical algorithms
-to compute this optimal precision.
+\subsection{The notion of working precision}
 
-\medskip
+\begin{itemize}
+\item what is it?
+\item how can we compute it? (Answer: Hermite reduction of the transpose)
+\item how can we estimate it?
+\end{itemize}
 
-\noindent
-{\bf The contribution of this paper.}
-This paper provides a theoretical and concrete study of the propagation 
-of the precision during the computation of the characteristic polynomial 
-of matrix defined over the $p$-adics. Thanks to the application of the 
-general framework of differential precision developed in 
-\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a}, we know that the 
-optimal precision of the characteristic polynomial $\chi_M$ of a matrix 
-$M \in M_n(\mathbb{Q}_p)$ is controled by the adjugate $\adj(X{-}M).$
-In this article, we provide:
-\begin{enumerate}
-\renewcommand{\itemsep}{0pt}
-\item Proposition \ref{prop:shortcom}: a short representation of
-the adjugate matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
-field (instead of $O(n^3)$ for a dense representation);
-\item two algorithms to compute $\adj(X{-}M)$: the first one is 
-division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
-the second one may perform divisions and has complexity $\softO(n^\omega)$;
-\end{enumerate}
-and deduce from this the following applications to the question
-we are interested in:
-\begin{enumerate}
-\renewcommand{\itemsep}{0pt}
-\setcounter{enumi}{2}
-\item Corollary \ref{cor:prec_gain}: a simple criterion to decide 
-whether $\chi_M$ is defined at precision higher than the precision of 
-$M$ (when $M$ lies in $M_n(\mathbb{Z}_p)$ and all its entries are given 
-at the same precision);
-\item two algorithms to compute the optimal precision on each 
-coefficient of the characteristic polynomial of a matrix $M \in 
-M_n(\Qp)$ (whose entries may be given at different precision): the first 
-one is division-free and has complexity $\softO(n^3)$ (operations in
-$\Zp$) while the second 
-one may perform divisions but has complexity $\softO(n^\omega)$;
-\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
-algorithm to compute the optimal precision on each eigenvalue of $M$.
-\end{enumerate}
-Moreover, we provide evidence showing that classical algorithms for 
-computing the characteristic polynomials do not reach the optimal 
-precision in many contexts.
-Note that we focus on computing the precision of $\chi_M$: we do not give a stable
-algorithm for computing the characteristic polynomial itself.
+\subsection{Complexity}
 
-\medskip
+\begin{tabular}{l|c|c|}
+& \ZpLCA/\ZpLCR & \ZpLF \\
+\hline
+Creation of a variable & $O(n)$ & $O(n_0)$ \\
+\hline
+Deletion of a variable & $O(m^2)$ & $-$ \\
+\hline
+\end{tabular}
 
-\noindent
-{\bf Organization of the article.}
-In Section \ref{sec:theo_study}, we review the differential
-theory of precision developed in \cite{caruso-roe-vaccon:14a}
-and apply it to the specific case of the characteristic polynomial.
-We also describe when the characteristic polynomial will have a higher
-precision than the input matrix and give a compact description
-of $\adj(X-M)$, the main ingredient in the differential.
+discussion on temporal locality
 
-In Section \ref{sec:diffHess}, we develop $\softO(n^3)$ algorithms to approximate
-the Hessenberg form of $M$, and through it to find $\adj(X-M)$ and thus
-find the precision of the characteristic polynomial of $M$.  In Section \ref{sec:diffFrob},
-we give a $\softO(n^\omega)$ algorithm to compute the compact description of $\adj(X-M)$.
+\section{Conclusion}
 
-Finally, we propose in Section \ref{sec:optjagged}
-algorithms to compute the optimal coefficient-wise precision
-for the characteristic polynomial.  We also give the results
-of some experiments demonstrating that these methods can lead
-to dramatic gains in precision over standard interval arithmetic.
-We close by describing the precision associated to
-eigenvalues of a matrix.
-
-\medskip
-
-\noindent
-{\bf Notations.} 
-Throughout the paper, $K$ will refer to a complete,
-discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
-$\OK$ its ring of integers and $\pi$ a uniformizer. 
-We will write that $f(n)=\softO (g(n))$ if there exists some
-$k \in \mathbb{N}$ such that $f(n)=O(g(n) \log (n)^k).$
-We will write $M$ for an $n \times n$ matrix over $K$,
-and $\chi$ the characteristic polynomial map, $\chi_M \in K[X]$ for
-the characteristic polynomial of $M$ and $d\chi_M$ for the differential
-of $\chi$ at $M$, as a linear map from $M_n(K)$ to the space
-of polynomials of degree less than $n$.
-We fix an $\omega \in \R$ such that the multiplication
-of two matrices over a ring is in $O(n^\omega)$ operations in the
-ring. Currently, the smallest known $\omega$ is less than
-$2.3728639 $ thanks to \cite{legall:14a}. 
-We will denote by $I_n$ the identity matrix of rank $n$ in $M_n(K).$
-When there is no ambiguity, we will drop this $I_n$
-for scalar matrices, \textit{e.g.} for $\lambda \in K$
-and $M \in M_n (K),$ $\lambda-M$ denotes $\lambda I_n - M.$
-We write $\sigma_1(M), \dots, \sigma_n(M)$ for the
-elementary divisors of $M$, sorted in increasing order of valuation.
-Finally, given a polynomial $\chi(X) = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$,
-the \emph{companion matrix} $\mathscr{C}$ associated to $\chi$ is
-\begin{equation}
-\label{eq:companion}
-\mathscr{C} = \left( \begin{matrix}
-0 & 1 & 0 & \cdots & 0 \\
-0 & 0 & 1 & \ddots & \vdots \\
-\vdots & \vdots & \ddots & \ddots & 0 \\
-0 & 0 & \cdots & 0 & 1 \\
--a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
-\end{matrix} \right).
-\end{equation}
-
-\section{Theoretical study}
-\label{sec:theo_study}
-
-\subsection{The theory of p-adic precision} \label{sec:padicprec}
-
-We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
-as a foundation for our discussion of the precision for the characteristic polynomial
-of a matrix.  We will be concerned with two spaces in what follows:
-the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
-$K_n[X]$ of monic degree $n$ polynomials over $K$.  We also define $K_{<n}[X]$ (resp. 
-${\OK}_{<n}[X]$) as the space of polynomials over $K$ (resp. over $\OK$) 
-of degree strictly less than $n$.
-
-One cannot represent
-matrices $M \in M_n(K)$ exactly, but instead must specify an approximation
-together with a measure of the approximation's accuracy; there are several ways to do so.
-When using \emph{jagged} precision, one gives a precision on each entry,
-representing the $(i,j)$ entry as $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
-is called the \emph{absolute precision} and may depend on $(i,j)$.
-When the $N_{i,j}$ are all equal, we say that the precision datum is 
-\emph{flat}.
-
-We may also model precision more flexibly by using lattices.  We 
-recall that a \emph{lattice} $H$ in $M_n(K)$ is the $\OK$-span of a vector
-space basis, or equivalently the image of $M_n(\OK)$ under a 
-$K$-linear transformation of $M_n(K)$ \cite[\S 2.2]{caruso-roe-vaccon:14a}.
-For $M \in M_n(K)$, cosets of the form $M + H$ are special
-neighborhoods of $M$; we will use them for modeling precision on $M$. 
-Given a family of integers $\famN = (N_{i,j})_{i,j}$ the lattice
-\[
-H_{\famN} = \big\{ \, M = (a_{i,j})_{i,j} \in M_n(\Qp) \,\, | \,\, \forall i,\!j\!:
-\pi^{N_{i,j}} \text{ divides } a_{i,j} \, \big\}
-\]
-corresponds to the jagged precision introduced above.
-The benefit of working with lattices is that they behave well under 
-differentiable maps \cite[Lemma 3.4]{caruso-roe-vaccon:14a}, but they
-require more space and time to compute.
-
-Jagged, flat and lattice precision is defined similarly for monic polynomials, 
-taking care that no precision is specified for the leading 
-coefficient which must be exactly~$1$. The underlying 
-vector space is now $K_{<n}[X]$ and we define a lattice in it as the
-image of ${\OK}_{<n}[X]$ under some linear mapping.
-
-Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map
-and $d \chi_M$ for its differential at $M$.
-Our analysis of the precision behavior of $\chi$ rests upon
-the computation of its derivative $d\chi$, using \cite[Lemma 3.4]{caruso-roe-vaccon:14a}
-Let $\adj(M)$ denote the adjugate\footnote{In \cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a} the adjugate was referred to as the \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
-we have $\adj(M) {=} \det(M) M^{-1}$).  Recall from
-\cite[Appendix B]{caruso-roe-vaccon:14a} and \cite[\S 3.3]{caruso-roe-vaccon:15a}
-that $d\chi_M$ is given by
-\begin{equation} \label{eq:dchi}
-\begin{array}{rcl}
-d\chi_M: \quad M_n(K) & \to & K_{<n}[X] \smallskip \\
- dM & \mapsto & \tr(\adj(X{-}M) \cdot dM).
-\end{array}
-\end{equation}
-
-When $d\chi_M$ is surjective and $H$ is sufficiently ``small'' and ``regular,'' we may give
-the precision of $\chi_M$ precisely as
-\begin{equation} \label{eq:mainlem}
-\chi(M + H) = \chi(M) + d\chi_M(H).
-\end{equation}
-The equality in \eqref{eq:mainlem} justifies our use of the phrase \emph{optimal} precision
-to denote the lattice precision $d\chi_M(H)$. To make the conditions on $H$ under
-which \eqref{eq:mainlem} holds more precise, let
-$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.
-
-\begin{prop} \label{prop:mainlem}
-Let $M \in M_n(K)$ and assume that $d \chi_M$ is surjective. Let:
-\[
-\nu = \max\left(\sum_{i=1}^{n-1} \val(\sigma_i(M)), 0\right).
-\]
-Then, for all integers $a, b \geq 0$ with $a + \nu < 2b$,
-any lattice $H$ such that $M_n(\pi^b \OK) \subset H \subset
-M_n(\pi^a \OK)$
-satisfies Eq.~\eqref{eq:mainlem}
-\end{prop}
-
-\begin{proof}
-Recall \cite[Definition 3.3]{caruso-roe-vaccon:15a} that the \emph{precision polygon}
-of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
-By \cite[Proposition 3.4]{caruso-roe-vaccon:15a}, the endpoints of the precision polygon
-occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
-${\OK}_{<n}[X] \subset d\chi_M(M_n(\pi^\nu \OK))$.
-
-Since the coefficients of $\chi_M$ are given by polynomials in the entries of $M$
-with integral coefficients, \cite[Proposition 2.2]{caruso-roe-vaccon:15a} implies
-the conclusion.
-\end{proof}
-
-For further discussion of using lattices to measure precision,
-details on the computation of $d\chi_M$ in \eqref{eq:dchi}, and 
-examples of additional precision in $\chi_M$,
-see \cite[\S\S 2.4.3, 3.2.2]{caruso:17a}.
-
-\subsection{Several facts about the differential of \texorpdfstring{$\chi$}{x}}
-
-We may rephrase the condition that $d\chi_M$ is surjective using more familiar invariants attached to $M$.
-
-\begin{prop}
-\label{prop:surjectivity}
-For $M \in M_n(K)$, the following conditions are equivalent:
-\begin{enumerate}[(i)]
-\renewcommand{\itemsep}{0pt}
-\item \label{pt:surj} the differential $d\chi_M$ is surjective
-\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
-to a companion matrix)
-\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
-closure $\bar{K}$ of $K$ all have dimension $1$
-\item \label{pt:charmin} the characteristic polynomial of $M$ is equal to the minimal polynomial of $M$.
-\end{enumerate}
-\end{prop}
-
-\begin{proof}
-The equivalence of \eqref{pt:cyc}, \eqref{pt:1d}, and \eqref{pt:charmin} is standard; see \emph{e.g.}
-\cite[\S 7.1]{hoffman-kunze:LinearAlgebra}.  We now show
-\eqref{pt:cyc} $\Rightarrow$ \eqref{pt:surj} and \eqref{pt:surj} $\Rightarrow$ \eqref{pt:1d}
-
-For any $P \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
-as the image of $d\chi$ at $PMP^{-1}$, so we may assume that
-$M$ is a companion matrix.  For a companion matrix, the last column of
-$\adj(X{-}M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.
-
-Now suppose that $M$ has an eigenvalue $\lambda$ over $\bar{K}$ occurring in multiple Jordan blocks.
-After conjugating into Jordan normal form over $\bar{K}$, the entries of $\adj(X{-}M)$
-will also be block diagonal, and divisible within each block by the product of $(X-\mu)^{d_\mu}$,
-where $\mu, d_\mu$ ranges over the eigenvalues and dimensions of the other Jordan blocks.
-Since $\lambda$ occurs in two Jordan blocks,
-$X - \lambda$ will divide every entry of $\adj(X{-}M)$ and $d\chi_M$ will not be surjective.
-\end{proof}
-
-We also have an integral analogue of Proposition \ref{prop:surjectivity}.
-
-\begin{prop}
-\label{prop:intsurj}
-For $M \in M_n(\OK)$, the following conditions are equivalent:
-\begin{enumerate}[(i)]
-\renewcommand{\itemsep}{0pt}
-\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is 
-${\OK}_{<n}[X]$.
-\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
-\end{enumerate}
-\end{prop}
-\begin{proof}
-The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
-\eqref{pt:cycp} is as in Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
-\end{proof}
-
-The relationship between precision and the images of lattices under $d\chi_M$ allows us to
-apply Proposition \ref{prop:intsurj} to determine when the precision of the characteristic polynomial
-is the minimum possible.
-
-\begin{cor} \label{cor:prec_gain}
-Suppose that $M \in \GL_n(\OK)$ is known with flat precision $O(\pi^N)$.
-Then the characteristic polynomial of $M$ has precision lattice strictly contained in $O(\pi^N)$
-if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
-\end{cor}
-
-Note that this criterion is checkable using $\softO(n^\omega)$ operations in the residue field \cite{storjohann:01a}.
-
-\medskip
-
-\noindent
-{\bf Stability under multiplication by $X$.}
-By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
-However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
-to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
-often be convenient to view $d \chi_M$ as a $K$-linear mapping
-$M_n(K) \to K[X]/\chi_M(X)$.
-
-\begin{prop}
-Let $R$ be the subring of $K[X]$ consisting of polynomials $f$ for
-which $f(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
-as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
-multiplication by $R$.
-\end{prop}
-\begin{proof}
-Let $A = \adj(X{-}M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
-the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
-with its adjugate is the determinant, $(X{-}M) \cdot A = \chi_M$ and thus
-$f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
-of the left hand side is precisely $f(X) \cdot \Lambda$, while the span of the entries
-of the right hand side is contained within $\Lambda$ since $f(M) \in M_n(\OK)$.
-\end{proof}
-
-\begin{cor}
-If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
-is stable under multiplication by $X$, so is a module over $\OK[X]$.
-\end{cor}
-
-\medskip
-
-\noindent
-{\bf Compact form of $d \chi_M$.}
-Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$ as in \eqref{eq:companion}.
-Assume one of the equivalent conditions of Proposition~\ref{prop:surjectivity}, giving a matrix 
-$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Then Proposition~\ref{prop:surjectivity}
-also holds for $M^T$, yielding an
-invertible matrix $Q \in \GL_n(K)$ with $M^T = Q \mathscr{C} Q^{-1}$.
-
-\begin{prop}
-\label{prop:shortcom}
-We keep the previous notations and assumptions.
-Let $Y$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
-\begin{equation}
-\label{eq:shortcom}
-\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T
-\mod \chi_M
-\end{equation}
-for some $\alpha \in K[X]$.
-\end{prop}
-
-\begin{proof}
-Write $A = \adj(X{-}M)$. From $(X{-}M) \cdot A \equiv 0 
-\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} A \equiv 0 \pmod{\chi_M}$. 
-Therefore each column of $P^{-1} A$ lies in the right kernel of $X{-}\mathscr{C}$
-modulo $\chi_M$. On the other hand, a direct computation shows that
-every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
-$\chi_M$ can be written as $W = w \cdot Y^T$ for some $w \in 
-K[X]/\chi_M$. We deduce that $A \equiv P \cdot Y^T B \pmod{\chi_M}$
-for some row vector $B$.
-Applying the same reasoning with $M^T$, we find that $B$ can be
-written $B = \alpha Y Q^T$ for some $\alpha \in K[X]/\chi_M$ and
-we are done.
-\end{proof}
-
-Proposition~\ref{prop:shortcom} shows that $\adj(X{-}M)$ can be encoded 
-by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
-stays within $O(n^2)$: the polynomials $\alpha$ and $\chi_M$ are 
-determined by $2n$ coefficients while we need $2n^2$ entries to 
-write down the matrices $P$ and $Q$. 
-We shall see moreover in Section \ref{sec:diffFrob} that interesting
-information can be read off of this short form $(\alpha, P, Q, 
-\chi_M)$.
-
-\begin{rem}
-With the previous notation, if $U \in GL_n(K),$
-the quadruple for 
-$UMU^{-1}$ is
-$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
-which can be computed in $O(n^\omega)$ operations in $K.$
-This is faster than computing $U \adj(X-M) U^{-1}$ using the dense
-representation, which requires $\softO(n^{\omega+1})$ operations in $K.$
-\end{rem}
-
-\section{Hessenberg form}
-\label{sec:diffHess}
-
-In this section, we combine the computation of a Hessenberg form of a 
-matrix and the computation of the inverse through the Smith normal form 
-(SNF) over a complete discrete valuation field (CDVF) to compute 
-$\adj(X{-}M)$ and $d \chi$. If $M \in M_n(\OK)$, then only division 
-by invertible elements of $\OK$ will occur.
-
-\begin{rem}
-In what follows, we will count operations in $K$ (regardless to 
-precision) for expressing our complexities. This choice makes sense 
-because we will not usually need much precision on the entries of 
-$\adj(X{-}M)$ (roughly only their valuations matter).
-\end{rem}
-
-\subsection{Hessenberg form}
-
-We begin with the computation of
-an approximate Hessenberg form.
-
-\begin{deftn}
-A \emph{Hessenberg matrix} is a matrix $M \in M_n(K)$ with
-\[
-M_{i,j}=0 \mbox{ for $j \le i-2$.}
-\]
-Given integers $n_{i,j}$, an \emph{approximate Hessenberg matrix}
-is a matrix $M \in M_n(K)$ with
-\[
-M_{i,j} = O(\pi^{n_{i,j}}) \mbox{ for $j \le i-2$.}
-\]
-If $M \in M_n (K)$ and $H \in M_n (K)$ is an (approximate) Hessenberg matrix
-similar to $M$, we say that H is an \emph{(approximate) Hessenberg form} of $M.$
-\end{deftn}
-
-It is not hard to prove that every matrix over a field admits
-a Hessenberg form.
-We prove here that we may compute an approximate Hessenberg form 
-for any matrix known at finite jagged precision over $K$.
-Moreover, we provide an exact change of basis matrix.
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm 1:} {\tt Approximate Hessenberg form computation}
-
-\noindent{\bf Input:} a matrix $M$ in $M_n(K).$
-
-\smallskip
-
-\noindent 0.\ $P:=I_n.$ \: $H:=M.$
-
-\noindent 1.\ {\bf for} $j=1,\dots,n-1$ {\bf do}
-
-\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $\val(H_{i_{min},j})$ is minimal.
-
-\noindent 3.\  \:  {\bf for} $i=j+2,\dots,n$ {\bf do}
-
-\noindent 4. \ \: \:  \textbf{Eliminate} the significant digits of $H_{i,j}$ by pivoting with row $j+1$ 
-using a matrix $T.$
-
-\noindent 5. \ \: \:  $H:=H \times T^{-1}.$ \: $P:=T \times P.$
-
-\noindent 6. \textbf{Return} $H,P.$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\medskip
-
-Note that we use the notation of matrix multiplication in line 5, but actual updating of
-$H$ and $P$ uses row and column operations (individually using $O(n)$ operations in $K$).
-
-\begin{prop} 
-Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M$.
-$P$ is exact  
-and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coefficient in $M.$
-\end{prop}
-\begin{proof}
-Inside the nested \textbf{for} loop, to eliminate $\pi^{u_y} \varepsilon_y+O(\pi^{n_y})$ with pivot $\pi^{u_x} \varepsilon_x+O(\pi^{n_x}),$
-the corresponding coefficient of the corresponding shear matrix is 
-any exact element of $K$ congruent to 
-$\pi^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{u_y-u_x\min (n_x-u_x,n_y-u_y)}$.
-Exactness follows directly. 
-The rest is clear.
-\end{proof}
-
-\begin{rem} \label{rem:char_pol_from_hessenberg}
-From a Hessenberg form of $M,$ it is well known
-that one can compute the characteristic polynomial of 
-$M$ in $O(n^3)$ operations in $K$ \cite[pp. 55--56]{Cohen:2013}
-However, this computation involves division, and its
-precision behavior is not easy to quantify.
-\end{rem}
-
-\subsection{Computation of the inverse}
-
-In this section, we prove that to compute the inverse of
-a matrix over $K$, the Smith normal form is precision-wise optimal in the flat-precision case.
-We first recall the differential of matrix inversion.
-
-\begin{lem}
-Let $u \: : \: GL_n (K) \rightarrow GL_n(K),$ $M \mapsto M^{-1}.$
-Then for $M \in GL_n (K),$ $du_M(dM)=M^{-1} dM M^{-1}.$
-It is always surjective.
-\end{lem}
-
-We then have the following result about the loss in precision when computing the inverse.
-
-\begin{prop}
-Let $\cond(M) = \val(\sigma_n(M))$.
-If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
-can be computed at precision $O(\pi^{N-2\cond(M)})$ by a \textbf{SNF} computation
-and this lower-bound is optimal,
-at least when $N$ is large.
-\end{prop}
-
-\begin{proof}
-The smallest valuation of an entry of $M^{-1}$ is $-\cond(M)$.
-Consequently, $N - 2\cond(M)$
-can be obtained as the valuation of an entry of $du_M(dM)$,
-and it is the smallest that can be achieved this way for $dM$ in a flat precision lattice.
-Hence the optimality of the bound given for large 
-$N$ \cite[Lemma 3.4]{caruso-roe-vaccon:14a}.
-
-Now, the computation of the Smith normal form was described in \cite{Vaccon-these}.
-From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$, and $P$ and $Q$ 
-known at precision at least $O(\pi^{N-\cond(M)})$, with coefficients in $\OK$
-and determinant in $\OK^\times$ realizing an Smith normal form of $M.$
-There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
-Since the smallest valuation occurring in $\Delta^{-1}$ is $-\cond(M),$
-we see that $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ is known at precision at least $O(\pi^{N-2\cond(M)}),$
-which concludes the proof.
-\end{proof}
-
-\subsection{The adjugate of \texorpdfstring{$X{-}H$}{X-H}}
-
-In this section, we compute $\adj(X-H)$ for a Hessenberg matrix $H$
-using the Smith normal form computation of the previous section.
-The entries of $\adj(X-H)$ lie in $K[X]$, which is not a complete
-discrete valuation field,
-so we may not directly apply the methods of the previous section.
-However, we may relate $\adj(X-H)$ to $\adj(1-XH)$, whose
-entries lie in $K\llp X\rrp$.  In this way, we compute $\adj(X-H)$
-using an SNF method, with no division in $K$.
-
-We begin by relating adjugates of
-similar matrices:
-
-\begin{lem} \label{lem:adjugate_of_similar}
-If $M_1,M_2 \in M_n(K)$ and $P \in GL_n (K)$ are such that
-$M_1=PM_2P^{-1},$ then:
-\[ \adj (X-M_1)=P \adj (X-M_2) P^{-1}. \] 
-\end{lem}
-
-The second ingredient we need is reciprocal polynomials.
-We extend its definition to matrices of polynomials.
-\begin{deftn}
-Let $d \in \mathbb{N}$ and $f \in K[X]$ of degree at most $d.$ 
-We define the reciprocal polynomial of order $d$ of $f$ as $f^{\rec,d}=X^d f \left( 1/X \right).$
-Let $F \in M_n(K[X])$ a matrix of polynomials of degree at most $d.$
-We denote by $F^{\rec,d}$ the matrix with $(F^{\rec,d})_{i,j} = (F_{i,j})^{\rec,d}$.
-\end{deftn}
-We then have the following result :
-\begin{lem}
-Let $M \in M_n(K).$ Then:
-\begin{align*}
-\adj(1-XM)^{\rec,n-1}&=\adj(X-M), \\
-(\chi_M I_n)^{\rec,n}&=(1-XM) \adj(1-XM).\\
-\end{align*}
-\end{lem}
-\begin{proof}
-For any matrix $F \in M_d(K[X])$ of polynomials of degree at most $1$,
-we have $\det (F^{\rec,1})=\det(F)^{\rec,d}.$
-This result directly implies the second part of the lemma; the first part follows
-from the fact that the entries of $\adj(X-M)$ and of $\adj(1-XM)$
-are determinants of size $n-1$.
-\end{proof}
-
-This lemma allows us to compute $\adj(1-XM)$ instead of $\adj(X-M).$
-This has a remarkable advantage: the pivots during the computation of
-the SNF of $\adj(1-XM)$ are units in $K\llb X\rrb$ (in $\OK\llb X\rrb$ if $M \in M_n(\OK)$) and are known
-in advance to be on the diagonal. This leads to a very smooth
-precision and complexity behavior when $M \in M_n(\OK).$
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm 2:} {\tt Approximate $\adj (X -H)$ }
-
-\noindent{\bf Input:} an approximate Hessenberg matrix $H$ in $M_n(\OK).$
-
-\smallskip
-
-\noindent 0.\ $U:=1-XH.$ $U_0:=1-XH.$
-
-\noindent 1.\ While updating $U$, \textbf{track} $P$ and $Q$ so that $U_0=PUQ$ is always satisfied.
-
-\noindent 2.\ {\bf for} $i=1,\dots,n-1$ {\bf do}
-
-\noindent 3.\  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,j},$ for $j\geq i+1$ 
-using the invertible pivot
-$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \OK[X]$).
-
-\noindent 4.\    {\bf for} $i=1,\dots,n-1$ {\bf do}
-
-\noindent 5. \  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i+1,i},$
-using the invertible pivot $U_{i,i}.$
-
-\noindent 6. \ $\psi:=\prod_i U_{i,i}.$
-
-\noindent 7. \ Rescale to get $U = I_n \mod X^{n+1}.$
-
-\noindent 8. \ $Y:=\psi \times P \times Q   \mod X^{n+1}.$ \footnote{The product $P \times Q$
-should be implemented by sequential row operations corresponding to the eliminations in Step 5
-in order to avoid a product of two matrices in $M_n(\OK[X])$.}
-
-\noindent 9. \ \textbf{Return} $Y^{\rec,n-1}, \psi^{\rec,n}.$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\medskip
-
-\begin{theo}
-Let $H \in M_n(\OK)$ be an approximate Hessenberg matrix with
-flat precision $O(\pi^N).$
-Then, using Algorithm 2, one can compute $\adj (X{-}H)$
-and $\chi_H$ in 
-$\softO (n^3)$ operations in $\OK$ at precision $O(\pi^N).$
-\end{theo}
-\begin{proof}
-First, the operations of the lines 2 and 3 use $\softO (n^3)$ operations
-in $\OK$ at precision $O(\pi^N).$
-Indeed, since $H$ is an approximate Heisenberg matrix, when we use $U_{i,i}$ as pivot
-the only other nonzero coefficient in its column is $U_{i+1,i}$.
-As a consequence, when performing this column-pivoting, only two rows ($i$ and
-$i+1$) lead to operations in $\OK \llb X \rrb$ other than checking precision.
-Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
-Following line 1, the computation of $Q$ is done by operations on rows, starting from the identity matrix.
-The order in which the entries of $U$ are cleared implies that $Q$ is just filled in as an upper triangular matrix:
-no additional operations in $\OK\llb X\rrb$ are required. Thus the total cost
-for lines 2 and 3 is indeed $\softO (n^3)$ operations.
-
-For lines 4 and 5, there are only $n{-}1$ eliminations, resulting in a $\softO (n^2)$ cost
-for the computation of $U.$ Rather than actually constructing $P$, we just track the eliminations
-performed in order to do the corresponding row operations on $Q$, since we only need the product $P \times Q$.
-
-Line 6 is in $\softO (n^2)$ and line 7 in $\softO (n^3).$
-
-Thanks to the fact that the $P$ only corresponds to the product of $n{-}1$ 
-shear matrices, the
-product on line 8 is in $\softO (n^3).$
-We emphasize that no division has been done.
-Line 9 is costless, and the result is proven.
-\end{proof}
-\begin{rem}
-If $M \in M_n(K)$ does not have coefficients in $\OK,$
-we may apply Algorithms 1 and 2 to $p^v M \in M_n(\OK)$
-in $\softO (n^3)$ operations in $\OK$, and then divide
-the coefficient of $X^k$ in the resulting polynomial by $p^{kv}$.
-\end{rem}
-
-We will see in Section \ref{sec:optjagged} that for an
-entry matrix with coefficients known at flat precision,
-Algorithms 1 and 2 are enough to
-know the optimal jagged precision on $\chi_M.$
-
-\subsection{The adjugate of \texorpdfstring{$X{-}M$}{X-M}}
-
-In this section, we combine Proposition \ref{prop:shortcom}
-with Algorithm 2 to compute the adjugate of $X{-}M$
-when $\chi_M$ is squarefree.   Note that, under
-the assumption that $d\chi_M$ is surjective,
-$\chi_M$ is squarefree if and only if $M$ is diagonalizable.  The result is
-the following $\softO(n^3)$ algorithm, where the only
-divisions are for gcd and modular inverse computations.
-
-\noindent\hrulefill
-
-\noindent {\bf Algorithm 3:} {\tt Approximate $\adj (X{-}M)$ }
-
-\noindent{\bf Input:} an approx. $M \in M_n(\OK),$ with $\disc(\chi_M) \neq 0.$
-
-\smallskip
-
-\noindent 0.\ Find $P \in GL_n(\OK)$ and $H \in M_n(\OK),$ approximate Hessenberg,
-such that $M=PHP^{-1},$ using Algorithm 1.
-
-\noindent 1.\ Compute $A=\adj (X-H)$ and $\chi_M = \chi_H$ using Algorithm 2.
-
-\noindent 2.\ Do $\row(A,1) \leftarrow \row(A,1)+\sum_{i=2}^n \mu_i \row(A,i),$ for
-random $\mu_i \in \OK,$ by doing $T \times A$ for some $T \in GL_n(\OK).$
-Compute $B:=TAT^{-1}.$
-
-\noindent 3.\ Similarily compute $C:=S^{-1}BS$ for $S \in GL_n(\OK)$ corresponding to
-adding a random linear combination of the columns of index $j \ge 2$
-to the first column of $B.$
-
-\noindent 4.\  \textbf{If} $\gcd(C_{1,1}, \chi_M) \neq 1,$ \textbf{then} go to 2.
-
-\noindent 5. Let $F$ be the inverse of $C_{1,1} \mod \chi_M$.
-
-\noindent 6. Let $U := \col(C,1)$ and $Y := F \cdot \row(C,1) \mod \chi_M$.
-
-\noindent 7. Return $\adj(X-M):=(PT^{-1}S U \times Y S^{-1} T P^{-1}) \mod \chi_M.$
-
-\vspace{-1ex}\noindent\hrulefill
-
-\medskip
-
-\begin{theo}
-\label{thm:alg3}
-For $M \in M_n(\OK)$ such that $\disc( \chi_M) \neq 0,$
-Algorithm 3 computes
-$\adj (X-M) \pmod{\chi_M}$ in average complexity $\softO (n^3)$
-operations in $K$.
-The only divisions occur in taking gcds and inverses modulo $\chi_M$.
-\end{theo}
-\begin{proof}
-As we have already seen, completing Steps 0 and 1 is in $\softO (n^3).$
-Multiplying by $T$ or $S$ or their inverse corresponds
-to $n$ operations on rows or columns over a matrix with coefficients
-in $\OK[X]$ of degree at most $n$.
-Thus, it is in $\softO(n^3).$
-Step 5 is in $\softO(n)$, Step 6 in $\softO(n^2)$ and Step 7 in $\softO(n^3)$.
-We need only prove that the set of $P$ and $S$ to avoid
-is of dimension at most $n-1$.
-
-The idea is to work modulo $X-\lambda$
-for $\lambda$ a root of $\chi_M$ (in an algebraic closure)
-and then apply Chinese Remainder Theorem.
-The goal of the Step $2$ is to ensure the first row of $B$ contains an
-invertible entry modulo $\chi_M$.
-Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
-affine hyperplane so that $\row(B,1) \mod (X-\lambda)$ is a non-zero vector.
-Hence we need only avoid a finite union of affine hyperplanes
-in order for $\row(B,1) \mod \chi (M)$ to contain an invertible coefficient.
-
-Similarly, the goal of Step 3 is make $C_{1,1}$ invertible modulo
-$\chi_M$.  Again, only a finite union of affine
-hyperplane need be avoided.
-Thus, almost any choice of $\mu_i$ leads to a matrix $C$ passing
-Step 4, concluding the proof.
-\end{proof}
-
-\begin{rem}
-As in the previous section, it is possible
-to scale $M \in M_n(K)$ so as to
-get coefficients in $\OK$ and apply the previous algorithm.
-\end{rem}
-\begin{rem}
-We refer to \cite{caruso:15a} for the handling
-of the precision of gcd and modular inverse computations.
-In this article, ways to tame the loss of precision
-coming from divisions are explored, following
-the methods of \cite{caruso-roe-vaccon:14a}.
-\end{rem}
-
-\section{Frobenius form}
-\label{sec:diffFrob}
-
-The algorithm designed in the previous section computes the differential 
-$d \chi_M$ of $\chi$ at a given matrix $M \in M_n(K)$ for a cost of 
-$\softO (n^3)$ operations in $K$. This seems to be optimal given that 
-the (naive) size of the $d \chi_M$ is $n^3$: it is a matrix of size $n 
-\times n^2$. It turns out however that improvements are still possible! 
-Indeed, thanks to Proposition~\ref{prop:shortcom}, the matrix of $d 
-\chi_M$ admits a compact form which can be encoded using only $O(n^2)$ 
-coefficients. The aim of this short section is to design a fast 
-algorithm (with complexity $\softO(n^\omega)$ operations in $K$) for 
-computing this short form. The price to pay is that divisions in $K$ 
-appear, which can be an issue regarding to precision in particular 
-cases. In this section, we only estimate the number of operations in $K$ 
-and not their behavior on precision.
-
-We now fix a matrix $M \in M_n(K)$ for which $d \chi_M$ is 
-surjective. Let $(\alpha, P, Q, \chi_M)$ be the quadruple encoding
-the short form of $d \chi_M$; we recall that they are related by the
-relations:
-\begin{align*}
-d \chi_M(dM) & =\tr(\adj(X{-}M) \cdot dM) \\
-\adj(X{-}M) & = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M.
-\end{align*}
-We may approximate $\chi_M$ in $\softO(n^\omega)$ operations in $K$
-by approximating the Frobenius normal form of $M$ \cite{storjohann:01a}.
-
-The matrix $P$ can be computed as follows. Pick $c \in K^n$. Define 
-$c_i = M^i c$ for all $i \geq 1$. The $c_i$'s can be computed in 
-$\softO(n^\omega)$ operations in $K,$ \textit{e.g.}
-using the first algorithm
-of \cite{keller-gehrig:85a}. Let $P_\inv$ be the 
-$n \times n$ matrix whose rows are the $c_i$'s for $1 \leq i \leq n$. 
-Remark that $P_\inv$ is invertible if and only if $(c_0, c_1, \ldots, 
-c_{n-1})$ is a basis of $K^n$ if and only if $c$ is a cyclic vector. 
-Moreover after base change to the basis $(c_0, \ldots, c_{n-1})$, the matrix 
-$M$ takes the shape \eqref{eq:companion}. In other words, if $P_\inv$
-is invertible, then $P = P_\inv^{-1}$ is a solution of $M = P \mathscr{C} P^{-1}$,
-where $\mathscr{C}$ is the companion matrix similar to $M$.
-Moreover, observe that the condition ``$P_\inv$ is invertible'' is open
-for the Zariski topology. It then happens with high probability as soon
-as it is not empty, that is as soon as $M$ admits a cyclic vector, which
-holds by assumption.
-
-The characteristic polynomial $\chi_M$ can be recovered thanks to the
-relation $a_0c_0 + a_1c_1 + \dots + a_{n-1}c_{n-1} = -c_{n-1} \cdot P$.
-
-Now, instead of directly computing $Q$, we first compute a matrix $R$ 
-with the property that $\mathscr{C}^T = R \mathscr{C} R^{-1}$. To do so,
-we apply the same strategy as above except that we start with the vector
-$e = (1, 0, \ldots, 0)$ (and not with a random vector). A simple computation shows
-that, for $1 \leq i \leq n{-}1$, the vector $\mathscr{C}^i e$ has the shape:
-$$\mathscr{C}^i e = (0, \ldots, 0, -a_0, \star, \ldots, \star)$$
-with $n{-}i$ starting zeros. Therefore the $\mathscr{C}^i e$'s form a basis of
-$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $\mathscr{C}$.
-We may then let $R$ have columns $e, \mathscr{C}e, \dots, \mathscr{C}^{n-1}e$
-and recover $Q$ using the relation $Q = P_\inv^T R$.
-
-It remains to compute the scaling factor $\alpha$. We write
-\begin{equation}
-\label{eq:shortcomN}
-\adj(X{-}\mathscr{C}) = \alpha \cdot Y^T \cdot Y R^T \mod \chi_M
-\end{equation}
-by multiplying Eq.~\eqref{eq:shortcom} on the 
-left by $P^{-1}$ and on the right by $P$. We observe
-moreover that the first row of $R$ is $(1, 0, \ldots, 0)$. Evaluating
-the top left entry of Eq.~\eqref{eq:shortcomN}, we end up with the 
-relation:
-$$\alpha = a_1 + a_2 X + \cdots + a_{n-1} X^{n-2} + X^{n-1}.$$
-No further computation are then needed to derive the value of $\alpha$.
-We summarize this section with the following theorem:
-
-\begin{theo}
-\label{thm:compute_shortcom}
-Given $M \in M_n(K)$ with $d \chi_M$ surjective,
-one can compute $(\alpha, P, Q, \chi_M) \in K[X]$ with
-$\softO(n^\omega)$ operations in $K$ such that
-$\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \mod \chi_M$.
-\end{theo}
-
-\section{Optimal jagged precision}
-\label{sec:optjagged}
-
-In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffFrob},
-we have proposed algorithms to obtain the adjugate of 
-$X{-}M.$ Our motivation for these computations is to then
-be able to understand what is the optimal precision on $\chi_M.$
-In this section, we provide some answers to this question,
-along with numerical evidence.
-We also show that it is then possible to derive 
-optimal precision of eigenvalues of $M.$
-
-\subsection{On the characteristic polynomial}
-
-For $0 \leq k < n$, let $\beta_k : K[X] \to K$ be the mapping taking a 
-polynomial to its coefficients in $X^k$. By applying 
-\cite[Lemma 3.4]{caruso-roe-vaccon:14a} to the composite $\beta_k 
-\circ \chi_M$, one can figure out the optimal precision on the
-$k$-th coefficient of the characteristic polynomial of $M$ (at
-least if $M$ is given at enough precision).
-
-Let us consider more precisely the case where $M$ is given at 
-jagged precision: the $(i,j)$ entry of $M$ is given at precision 
-$O(\pi^{N_{i,j}})$ for some integers $N_{i,j}$. 
-Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then shows that
-the optimal precision on the $k$-th coefficient of $\chi_M$ is 
-$O(\pi^{N'_k})$ where $N'_k$ is given by the formula:
-\begin{equation}
-\label{eq:optjagged}
-N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\beta_k(A_{i,j})),
-\end{equation}
-where $A_{i,j}$ is the $(i,j)$ entry of the adjugate $\adj(X{-}M)$.
-
-\begin{prop} \label{prop:optimal_jagged}
-If $M \in M_n(\OK)$ is given at (high enough) jagged precision, 
-then we can compute the optimal jagged precision on $\chi_M$ in 
-$\softO (n^3)$ operations in $K$.
-\end{prop}
-
-\begin{proof}
-We have seen in \S \ref{sec:diffHess} and \S \ref{sec:diffFrob}
-that the computation of the matrix $A = \adj(X{-}M)$ can be carried out 
-within $\softO(n^3)$ operations in $K$ (either with the Hessenberg 
-method or the Frobenius method). We conclude by applying 
-Eq.~\eqref{eq:optjagged} which requires no further operation in $K$
-(but $n^3$ evaluations of valuations and $n^3$ manipulations of 
-integers).
-\end{proof}
-
-\begin{rem}
-If $M \in M_n(\OK)$ is given at (high enough) \emph{flat} precision, 
-then we can avoid the final base change step in the Hessenberg 
-and thus any division by a non-invertible.
-Indeed, observe that, thanks to Lemma \ref{lem:adjugate_of_similar}, 
-we can write:
-$$\tr(\adj (X{-}M) \cdot dM)=\tr(\adj (X{-}H)\cdot P^{-1} dM P)$$
-where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
-that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
-As a consequence, the integer $N'_k$ giving the optimal precision on the 
-$k$-th coefficient of $M$ is also equal to
-$N + \min_{1 \leq i, j\leq n} \val(\beta_k(A^H_{i,j}))$
-where $A^H_{i,j}$ is the $(i,j)$ entry of $\adj(X{-}H)$,
-where $H$ is the Hessenberg form of $M$.
-\end{rem}
-
-\begin{rem} \label{rem:lift_for_optimal}
-As a consequence of the previous discussion, once the optimal jagged 
-precision is known, it is possible to lift the entries of $M$ to a sufficiently
-large precision, rescale them to have entries in $O_K$
-and then use Algorithm 2 to
-compute the characteristic polynomial.
-The output might then need to be rescaled and
- truncated to the optimal precision. 
-This requires $\softO(n^3)$ operations in $O_K$ 
-and may require a large intermediate precision.
-Better methods for approximating $\chi_M$ with
-lower intermediate precision would be a valuable
-future contribution.
-\end{rem}
-
-\medskip
-
-\noindent
-{\bf Numerical experiments.}
-We have made numerical experiments in \textsc{SageMath}~\cite{sage}
-in order to compare the optimal precision obtained with the methods
-explained above with the actual precision obtained by the software.
-For doing so, we picked a sample of $1000$ matrices $M$ in 
-$M_9(\Q_2)$ where all the entries are chosen randomly as follows.
-We fix an integer $N$, the relative precision, and generate elements of $\Q_p$ of the shape
-\[
-x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)
-\]
-where $v$ is an integer generated according to the distribution:
-\[
-\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
-\mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
-|n| \geq 1
-\]
-and $a$ is an integer in the range $[0, p^N)$, selected uniformly at random.
-This distribution is chosen merely because it is the default in \textsc{SageMath}.
-
-Once this sample has been generated, we computed, for each $k \in \{0, 
-1, \ldots, 8\}$, the three following quantities:
-
-%\vspace{-2mm}
-
-\begin{enumerate}[$\bullet$]
-\renewcommand{\itemsep}{0pt}
-\item the optimal precision on the $k$-th coefficient of the 
-characteristic polynomial of $M$ given by Eq.~\eqref{eq:optjagged}.
-(Note that the method for computing $\adj(X{-}M)$ is irrelevant
-since we only record the precision loss, not the runtime.)
-\item in the capped relative mode (each 
-coefficient carries its own precision which is updated after each 
-elementary arithmetic operation),
-the precision on the $k$-th coefficient of the 
-characteristic polynomial of $M$ is computed \emph{via} the call:
-
-\smallskip
-
-\hfill$M\texttt{.charpoly(algorithm="df")}.$\hfill\null
-
-\smallskip
-
-\noindent 
-Note that the precision behavior described persists when using algorithms
-other than the division-free method.
-
-\item in the model of floating-point arithmetic (see 
-\cite[\S 2.3]{caruso:17a}), the number of correct digits of the 
-$k$-th coefficient of the characteristic polynomial of $M$.
-\end{enumerate}
-
-\noindent
-The table of Figure~\ref{fig:exp} summarizes the results obtained. 
-\begin{figure}
-\hfill
-{\renewcommand*{\arraystretch}{1.3}
-\begin{tabular}{|c|r|r|r|}
-\cline{2-4} 
-\multicolumn{1}{l|}{} 
-  & \multicolumn{3}{c|}{Average loss of accuracy} \\
-\cline{2-4}
-\multicolumn{1}{l|}{\null\hspace{6mm}\null} 
-  & \makebox[1.5cm]{\hfill Optimal\hfill\null}
-  & \makebox[1.5cm]{\hfill CR\hfill\null}
-  & \makebox[1.5cm]{\hfill FP\hfill\null} \\
-\hline
-\rule{0pt}{2.7ex}%
-$X^0$ & $3.17$ & $196\phantom{.00}$ & $189\phantom{.00}$ \vspace{-1.5ex} \\
-{\scriptsize (det.)} 
-& {\scriptsize dev: $1.76$} & {\scriptsize dev: $240$} & {\scriptsize dev: $226$} \\
-\rule{0pt}{2.7ex}%
-$X^1$ & $2.98$ & $161\phantom{.00}$ & $156\phantom{.00}$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.69$} & {\scriptsize dev: $204$} & {\scriptsize dev: $195$} \\
-\rule{0pt}{2.7ex}%
-$X^2$ & $2.75$ & $129\phantom{.00}$ & $126\phantom{.00}$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.57$} & {\scriptsize dev: $164$} & {\scriptsize dev: $164$} \\
-\rule{0pt}{2.7ex}%
-$X^3$ & $2.74$ & $108\phantom{.00}$ & $105\phantom{.00}$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.73$} & {\scriptsize dev: $144$} & {\scriptsize dev: $143$} \\
-\rule{0pt}{2.7ex}%
-$X^4$ & $2.57$ &  $63.2\phantom{0}$ &  $60.6\phantom{0}$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.70$} & {\scriptsize dev: $85.9$} & {\scriptsize dev: $85.8$} \\
-\rule{0pt}{2.7ex}%
-$X^5$ & $2.29$ &  $51.6\phantom{0}$ &  $49.7\phantom{0}$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.66$} & {\scriptsize dev: $75.3$} & {\scriptsize dev: $74.9$} \\
-\rule{0pt}{2.7ex}%
-$X^6$ & $2.07$ &   $9.04$           &   $8.59$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.70$} & {\scriptsize dev: $26.9$} & {\scriptsize dev: $26.4$} \\
-\rule{0pt}{2.7ex}%
-$X^7$ & $1.64$ &   $5.70$           &   $5.38$ \vspace{-1.5ex} \\
-& {\scriptsize dev: $1.65$} & {\scriptsize dev: $15.3$} & {\scriptsize dev: $14.7$} \\
-\rule{0pt}{2.7ex}%
-$X^8$ & $0.99$ &   $0.99$           &   $0.99$ \vspace{-1.5ex} \\
-{\scriptsize (trace)} 
-& {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} \\
-\hline
-\end{tabular}}
-\hfill\null
-
-\medskip
-
-\hfill
-{\footnotesize Results for a sample of $1000$ instances}
-\hfill\null
-
-\caption{Average precision loss on 
-the characteristic polynomial of a random $9 \times 9$
-matrix over $\Q_2$}
-\label{fig:exp}
-
-\end{figure}
-%
-It should be read as follows. The acronyms CR and FP 
-refer to ``capped relative'' and ``floating-point'' respectively.
-The numbers displayed are the average loss of
-\emph{relative} precision. 
-More precisely, if $N$ is the relative precision of
-the entries of $M$ and $v$ is the valuation of
-the $k$-th coefficient of $\chi_M$, then:
-
-%\vspace{-2mm}
-
-\begin{enumerate}[$\bullet$]
-\renewcommand{\itemsep}{0pt}
-\item the column ``Optimal'' is the average of the quantities 
-$(N'_k{-}v) - N$ (where $N'_k$ is defined by Eq.~\eqref{eq:optjagged}): 
-$N'_k{-}v$ is the optimal \emph{relative} precision, so that the
-difference $(N'_k{-}v) - N$ is the loss of relative precision;
-\item the column ``CR'' is the average of the quatities 
-$(\text{CR}_k{-}v) - N$ where $\text{CR}_k$ is the computed (absolute) 
-precision on the $k$-th coefficient of $\chi_M$;
-\item the column ``FP'' is the average of the quatities 
-$(\text{FP}_k{-}v) - N$ where $\text{FP}_k$ is the first position of
-an incorrect digit on the $k$-th coefficient of $\chi_M$.
-\end{enumerate}
-
-We observe that the loss of relative accuracy stays under control in the 
-``Optimal'' column whereas it has very erratic behavior in the two other columns,
-demonstrating the utility of the methods developed in this paper.
-
-\subsection{On eigenvalues}
-\label{ssec:eigenvalues}
-
-Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple}
-\footnote{the corresponding generalized eigenspace has dimension $1$} eigenvalue 
-of $M$. We are interesting in quantifying the optimal precision on 
-$\lambda$ when $M$ is given with some uncertainty.
-
-To do so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
-suppose that the uncertainty of $M$ is jagged:
-each entry of $M$ is given at some precision $O(\pi^{N_{i,j}})$.
-Let $\lambda_\app$ be the relevant eigenvalue of $M_\app$. We remark 
-that it is possible to follow the eigenvalue $\lambda_\app$ on a small 
-neighborhood $\calU$ of $M$. More precisely, there exists a unique continuous 
-function $f : \calU \to K$ such that
-$f(M_\app) = \lambda_\app$, and
-$f(M')$ is an eigenvalue of $M'$ for all $M' \in \calU$.
-
-\begin{lem}
-The function $f$ is strictly differentiable on a neighborhood of 
-$M_\app$, with differential at $M$ the map
-\[
-dM \mapsto d \lambda = - \frac{\tr(\adj(\lambda-M) \cdot dM)}
-{\chi'_M(\lambda)},
-\]
-where $\chi'_M$ is the usual derivative of $\chi_M$.
-\end{lem}
-
-\begin{proof}
-The first assertion follows from the implicit function theorem,
-the second from differentiating $\chi_M(\lambda) = 0$, which gives
-$\chi'_M(\lambda) \cdot d \lambda + \tr(\adj(X-M) \cdot dM)(\lambda) = 0$.
-\end{proof}
-
-\noindent
-Lemma 3.4 of \cite{caruso-roe-vaccon:14a} now implies that, if the
-$N_{i,j}$'s are large enough and sufficiently well balanced, the optimal
-precision on the eigenvalue $\lambda$ is $O(\pi^{N'})$ with:
-$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(A_{i,j}(\lambda)) - 
-\val(\chi'_M(\lambda))\big)$$
-where $A_{i,j}$ denotes as above the $(i,j)$ entry of $\adj(X{-}M)$.
-Writing
-$\adj(X{-}M) = \alpha \cdot P Y^T \cdot Y Q^T \text{ mod } \chi_M$
-as in Proposition~\ref{prop:shortcom}, we find:
-\begin{align}
-N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \nonumber \\
-& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
-\val(P_i Y(\lambda)^T) + \val(Y(\lambda) Q_j^T)\big) \label{eq:Nprime}
-\end{align}
-where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
-denotes the $j$-th row of $Q$. Note moreover that $Y(\lambda)$ is
-the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
-By the discussion of \S \ref{sec:diffFrob}, the exact value of $N'$ can be 
-determined for a cost of $\softO(n^\omega)$ operations in $K$ and
-$O(n^2)$ operations on integers.
-
-When $M$ is given at flat precision, \emph{i.e.} the $N_{i,j}$'s are all 
-equal to some $N$, the formula for $N'$ may be rewritten:
-\begin{align}
-N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \nonumber \\
-& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i Y(\lambda)^T)
-+ \min_{1 \leq j \leq n} \val(Y(\lambda) Q_j^T)\big) \label{eq:Nprimeflat}
-\end{align}
-and can therefore now be evaluated for a cost of $\softO(n^\omega)$
-operations in $K$ and only $O(n)$ operations with integers.
-
-\medskip
-
-To conclude, let us briefly discuss the situation where we want
-to figure out the optimal jagged precision on a tuple $(\lambda_1,
-\ldots, \lambda_s)$ of simple eigenvalues. Applying \eqref{eq:Nprime},
-we find that the optimal precision on $\lambda_k$ is 
-\begin{align*}
-N'_k & = \val(\alpha(\lambda_k)) - \val(\chi'_M(\lambda_k)) \\
-& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
-\val(P_i Y(\lambda_k)^T) + \val(Y(\lambda_k) Q_j^T)\big).
-\end{align*}
-
-\begin{prop}
-\label{prop:opteigenvalues}
-The $N'_k$'s can be all computed in $\softO(n^\omega)$ operations
-in $K$ and $O(n^2 s)$ operations with integers.
-
-If the $N_{i,j}$'s are all equal, the above complexity can be 
-lowered to $\softO(n^\omega)$ operations in $K$ and $O(n s)$ operations 
-with integers.
-\end{prop}
-
-\begin{proof}
-The $\alpha(\lambda_k)$'s and the $\chi'_M(\lambda_k)$'s can be computed 
-for a cost of $\softO(ns)$ operations in $K$ using fast multipoint 
-evaluation methods (see \S 10.7 of \cite{gathen-gerhard:13a}).
-On the other hand, we observe that $P_i Y(\lambda_k)^T$ is
-the $(i,k)$ entry of the matrix:
-$$P \cdot \left( \begin{matrix}
-\lambda_1 & \cdots & \lambda_s \\
-\lambda_1^2 & \cdots & \lambda_s^2 \\
-\vdots & & \vdots \\
-\lambda_1^{n-1} & \cdots & \lambda_s^{n-1}
-\end{matrix} \right).$$
-The latter product can be computed in $\softO(n^2)$ operations in 
-$K$. Indeed,
-the right factor is a truncated Vandermonde matrix, so that
-computing the above product reduces to evaluating a polynomial at the
-points $\lambda_1, \ldots, \lambda_s$. Therefore all the $P_i 
-Y(\lambda_k)^T$'s (for $i$ and $k$ varying) 
-can be determined with the same complexity. 
-Similarly all the $Y(\lambda) Q_j^T$ are computed for the same cost.
-The first assertion of Proposition~\ref{prop:opteigenvalues} follows.
-The second assertion is now proved similarly to the case of a unique
-eigenvalue.
-\end{proof}
-
-\bibliographystyle{plain}
-\bibliography{charpoly}
+\begin{itemize}
+\item In which case should we use this package? Mainly for exploration
+\item example of $p$-adic differential equations (especially for $p=2$)
+\end{itemize}
 
 \end{document}
-
-\begin{comment}
-\begin{prop} \label{prop:factor_com}
-Let $M \in M_n (\OK)$ be such that $discr(\chi (M)) \neq 0.$
-We assume that $A=\adj (X-M)$ is such that $\gcd(A_{1,1}, \chi (M))=1$
-(in $K[X]$).
-Then for $U=col(A,1)$ and $\Lambda=A_{1,1}^{-1} row(A,1) \mod \chi (M),$
-both in $\OK[X]^n,$ we have: 
-\[ \adj (X-M)= U \cdot ^T \mod \chi (M).\]
-\todo{I think that the correct condition is $d \chi_M$ surjective (which is a bit weaker).}
-\end{prop}
-
-\begin{proof}
-We first prove that for all $i,j \in \llbracket 1, n \rrbracket,$ we have
-\begin{equation}A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod \chi (M). \label{eqn:2-minors-com} \end{equation}
-Let $\lambda$ be a root of $\chi (M)$ in an algebraic closure of $K.$
-Then $A(\lambda)$ is of rank one.
-Indeed, $A(\lambda) \times (\lambda-M) =0.$
-Since $\chi (M)$ is square-free, then $\lambda-M$ is of rank $n-1$
-and then $A(\lambda)$ is of rank at most one.
-Nevertheless, we can remark that
- $\gcd(A_{1,1}, \chi (M))=1,$ $A_{1,1} \neq 0 \mod (X-\lambda)$
-and thus, $A(\lambda)$ is of rank exactly one.
-As a consequence, all $2 \times 2$-minors of $A(\lambda)$ are zero.
-It means that \[A_{1,1} A_{i,j}=A_{1,i}A_{1,j} \mod (X-\lambda). \]
-By the Chinese Remainder Theorem, our first result is then proved.
-
-Now, let us take $U=col(A,1)$ the first column of $A$ and
-$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
-invertible $\mod \chi (M)).$
-Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
-\end{proof}
-
-\begin{rem}
-If $M \in M_n (\OK)$ is such that for some $U,V \in O_{K}[X]^n$
-we have $\adj (X-M)= U \cdot V^T \mod \chi (M)$ and if 
-$P\in GL_n (\OK)$ then
-\[\adj (X-PMP^{-1})=(PU) \cdot ( V^T P^{-1}) \mod \chi_M.\]
-In other words, factorizability of $\adj(X-M)$ is stable
-by change of basis.
-\end{rem}
-\end{comment}

### 2017-11-17 ZpL/ZpL.tex
Added subsection "Correctness"
Added Section 4 to the plan

@@
 \usepackage{multirow}
+\usepackage{xspace}
 %\usepackage{tikz}
@@
 
-\newcommand{\ZpL}{\texttt{ZpL}}
-\newcommand{\ZpLCA}{\texttt{ZpLCA}}
-\newcommand{\ZpLCR}{\texttt{ZpLCR}}
-\newcommand{\ZpLF}{\texttt{ZpLF}}
+\newcommand{\ZpL}{\texttt{ZpL}\xspace}
+\newcommand{\ZpLCA}{\texttt{ZpLCA}\xspace}
+\newcommand{\ZpLCR}{\texttt{ZpLCR}\xspace}
+\newcommand{\ZpLF}{\texttt{ZpLF}\xspace}
 
@@
 
-\title{ZpL: a p-adic precision package}
+\title{\ZpL: a p-adic precision package}
 
@@
 
+\subsection{Elementary arithmetic}
+
 \begin{itemize}
-\item Elementary arithmetic
-\item Determinant + matrix multiplication
+\item Compare $px$ and $x + x + \cdots +x$ ($p$ times)
+\item Compare $x^p$ and $x \cdot x \cdots x$ ($p$ times)
+\item $x$ and $y$ are given at different precision, compute $u = x+y$,
+$v = x-y$ and $u+v$
+\end{itemize}
+
+\subsection{Linear algebra}
+
+\begin{itemize}
+\item determinant
+\item characteristic polynomial
+\end{itemize}
+
+\subsection{Commutative algebra}
+
+\begin{itemize}
 \item gcd of polynomials
-\item some benchmarks on precision
 \end{itemize}
@@
 
+\subsection{Correctness}
+
+\begin{itemize}
+\item prove that \ZpLCA and \ZpLCR are correct under some assumptions
+(to be determined)
+\item explain that \ZpLF is not meant to provide proved results
+\end{itemize}
+
 \subsection{Complexity}
@@
 
+\section{Presentation of the package}
+
+\begin{itemize}
+\item How to access to the precision
+\item Diffused digits
+\item Historic of variables creation/deletion, etc...
+\item Other features?
+\end{itemize}
+
+\subsection{More examples}
+
+\begin{itemize}
+\item SOMOS sequence
+\item many multiplications of matrices
+\item Dodgson algorithm
+\item subresultants
+\item Grobner basis
+\item $p$-adic differential equations
+\end{itemize}
+
 \section{Conclusion}

### 2017-11-17 ZpL/txttotex.pl
First examples

@@ -1,0 +1,40 @@
+#!/usr/bin/perl -w
+
+use strict;
+
+my @columns;
+my $nlines = 0;
+
+while(1) {
+  my @lines;
+  my $last = 1;
+  while(<>) {
+    chomp; s/\s*$//;
+    last if /^\s*$/;
+    push @lines, $_;
+    $last = 0;
+  }
+  last if $last == 1;
+  push @columns, \@lines;
+  my $n = 1 + $#lines;
+  $nlines = $n if $n > $nlines;
+}
+
+my $ncol = 1 + $#columns;
+my $coldef = "|" . ("l|" x $ncol);
+
+print "\\begin{tabular}{$coldef}\n";
+for (my $i = 0; $i < $nlines; $i++) {
+  for (my $j = 0; $j < $ncol; $j++) {
+    my $t = $columns[$j][$i];
+    if ($t =~ s/^\|\s*//) {
+      print "\\hfill\\verb?" . $t . "?";
+    } else {
+      print "\\verb?" . $t . "?";
+    }
+    if ($j < $ncol-1) { print " & "; }
+    else { print " \\\\\n"; }
+  }
+}
+
+print "\\end{tabular}\n";

### 2018-01-07 ZpL/ZpL.tex
First examples in ZpL.tex
First version of 5.1

@@
 \usepackage{xspace}
+\usepackage{tabularx,multicol}
 %\usepackage{tikz}
@@
 
-\begin{itemize}
-\item Compare $px$ and $x + x + \cdots +x$ ($p$ times)
-\item Compare $x^p$ and $x \cdot x \cdots x$ ($p$ times)
-\item $x$ and $y$ are given at different precision, compute $u = x+y$,
-$v = x-y$ and $u+v$
-\end{itemize}
+\begin{tabular}{|l|l|}
+\verb?R = Zp(3,5)? & \verb?R = ZpL(3,5)? \\
+\verb?x = R(143,5); x? & \verb?x = R(143,5); x? \\
+\hfill\verb?...12022? & \hfill\verb?...12022? \\
+\verb?3*x? & \verb?3*x? \\
+\hfill\verb?...120220? & \hfill\verb?...120220? \\
+\verb?x + x + x? & \verb?x + x + x? \\
+\hfill\verb?...20220? & \hfill\verb?...120220? \\
+\verb?x^3? & \verb?x^3? \\
+\hfill\verb?...020222? & \hfill\verb?...020222? \\
+\verb?x * x * x? & \verb?x * x * x? \\
+\hfill\verb?...20222? & \hfill\verb?...020222? \\
+\end{tabular}
+
+\begin{tabular}{|l|l|}
+\verb?R = Zp(2,10)? & \verb?R = ZpL(2,10)? \\
+\verb?x = R(987,10); x? & \verb?x = R(987,10); x? \\
+\hfill\verb?...1111011011? & \hfill\verb?...1111011011? \\
+\verb?y = R(21,5); y? & \verb?y = R(21,5); y? \\
+\hfill\verb?...10101? & \hfill\verb?...10101? \\
+\verb?u = x + y; u? & \verb?u = x + y; u? \\
+\hfill\verb?...10000? & \hfill\verb?...10000? \\
+\verb?v = x - y; v? & \verb?v = x - y; v? \\
+\hfill\verb?...00110? & \hfill\verb?...00110? \\
+\verb?u + v? & \verb?u + v? \\
+\hfill\verb?...10110? & \hfill\verb?...11110110110? \\
+\end{tabular}
 
@@
 \begin{itemize}
+\item Discuss coercion/conversion: \texttt{Zp} coerces to \ZpL but no coersion
+in the other direction
 \item How to access to the precision
@@
 \subsection{More examples}
+
+\begin{tabular}{|l|l|}
+\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
+\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
+\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
+\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
+\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
+\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
+\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
+\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
+\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
+\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
+\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
+\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
+\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
+\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
+\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
+\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
+\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
+\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
+\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
+\hfill\verb?...11? & \hfill\verb?...0000000111? \\
+\hfill\verb?...10? & \hfill\verb?...1010001110? \\
+\hfill\verb?...01? & \hfill\verb?...1110110001? \\
+\hfill\verb?...01? & \hfill\verb?...011110111001? \\
+\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
+\hfill\verb?...1? & \hfill\verb?...11111000011? \\
+\end{tabular}
 
@@
 
+\subsection{Exploration}
+
+The purpose of this package is that of exploration
+of the $p$-adic precision.
+We conclude with a last example coming
+from the field of the computation of
+isogenies between elliptic curves over finite fields.
+Techniques to compute these isogenies
+through the solving of $p$-adic differential equations
+have a long history, and a first
+analysis of the behaviour of precision using
+lattices has been investigated in \cite{LV16}.
+
+In that article, the authors were interested in
+solving $y'^2=g \times h(y)$ with
+$g,h,y \in \mathbb{Z}_p \llbracket x \rrbracket$
+such that $g(0)=h(0)=1$ and $y(0)=0.$
+
+The main result was that the intrinsic loss in precision
+when computing the coefficient $x^n$ of $y$
+from $g$ and $h$ was in $\log_p(n)$
+even though a naive analysis of
+the Newton method for solving the equation
+yield a loss in $\log_p(n)^2.$
+
+However, this does not encompass the case of 
+elliptic curves defined and isogeneous 
+over $\mathbb{Q},$ which can yield a differential 
+equation with a solution $y$ whose coefficients
+have negative valuation.
+
+For instance, the curves 
+$E: \: y^2=x^3+1$
+and $E': \: y^2=+ (m^2-1)x^2/4 + m^6$ are isogeneous
+over $\mathbb{Q}$ if  $m$ is an odd integer.
+We take $m=67.$ The corresponding isogeny has coefficients with possibly
+negative $3$-adic valuation.
+
+We can solve numerically 
+the corresponding differential equation
+using the algorithm of \cite{Lercier-Sirvent:08}
+over $\mathbb{Q}_3$ with lattice precision.
+Any generic implementation of the Newton
+method of \cite{Lercier-Sirvent:08} is enough.
+
+\begin{tabular}{|l|}
+\verb?    A = QpL(3)? \\
+\verb?    R.<x> = PolynomialRing(A)?\\
+\verb?    m=67?\\
+\verb?    g=1+(1/4)*x^2+x^6?\\
+\verb?    h=1+((m^2)/4)*x^2+m^6* x^6?\\
+\verb?    y=Lercier_Sirvent(g,h,64)?\\
+\verb?    y[31]?\\
+\verb?(467995903278061284942799/3^14 + O(3^36))*x^31?\\
+\end{tabular}
+
+We can then see that up to $63,$ the behaviour
+of the loss in relative precision is the same
+as for the coefficient in $x^{31}$: \textbf{no loss
+occured}.
+Thus, those simple manipulations
+hint at a different behaviour for the loss 
+in precision in at least some special cases of
+of isogenies defined over $\mathbb{Q}$
+but not necessarily over $\mathbb{Z}_p.$
+
+\bibliographystyle{plain}
+\bibliography{ZpL}
+
 \end{document}

### 2018-01-15 ZpL/ZpL.tex
More differential equations

@@
 
+\subsubsection{$p$-adic differential equations}
+
+In \cite{LV16}, the behaviour of the precision when solving
+$p$-adic differential equations with separation of 
+variables has been studied.
+The authors have investigated the gap when that appears
+when applying a Newton-method solver between
+the theoretic loss in precision and the 
+actual loss in precision for a naive implemenationin \verb?Zp(p)?.
+We can reach this theoretical loss in precision using \verb?ZpL?.
+We use a generic \verb?Newton_Iteration_Solver(g,h,N)?
+that applies \verb?N? steps of the Newton method for 
+$y'=g \times h(y)$ as described in \citep{LV16}.
+
+\begin{tabular}{|l|l|}
+\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
+\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
+\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
+\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
+\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
+\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
+\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
+\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
+\hfill\verb?9? & \hfill\verb?4? \\
+\end{tabular}
+The theorical loss in precision for the coefficient of $t^{31}$
+is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
+loss in precision for a naive implementation is clearly more severe.
+
 \section{Conclusion}
@@
 \begin{tabular}{|l|}
-\verb?    A = QpL(3)? \\
-\verb?    R.<x> = PolynomialRing(A)?\\
+\verb?    R = QpL(3)? \\
+\verb?    A.<x> = PolynomialRing(R)?\\
 \verb?    m=67?\\

### 2018-01-17 ZpL/ZpL.tex
First version of 3.1

@@
 \subsection{The precision Lemma}
+In \cite{caruso-roe-vaccon:14a}, we have suggested the 
+use of lattices to represent the precision of elements in 
+$K$-vector spaces.  It contrasts with the
+\emph{coordinate-wise method} (of \textit{e.g.}
+ \verb?Zp(5)?) that is of use in Sage prior to
+ our package, and  
+where the precision of an element is specified by giving the precision
+of each coordinate separately and is updated after each basic
+operation.
+
+Consider a finite 
+dimensional normed vector space $E$ 
+defined over $K$. 
+We use the notation $\Vert \cdot \Vert_E$ for the norm 
+on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
+closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
+E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
+Because of ultrametricity, the balls $B^{\phantom 
+-}_E(r)$ and $B^-_E(r)$ are examples of lattices. 
+Lattices can be thought of as special neighborhoods of $0,$
+ and therefore are good 
+candidates to model precision data. Moreover, as revealed in 
+\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
+differentiable maps:
+
+\begin{prop}
+\label{prop:precision}
+Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
+and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
+$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
+that the differential $f'(v_0)$ is surjective.
+Then, for all $\rho \in (0, 1]$, there exists a positive real
+number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
+$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
+satisfies:
+\begin{equation}
+\label{eq:firstorder}
+f(v_0 + H) = f(v_0) + f'(v_0) (H).
+\end{equation}
+\end{prop}
+
+This proposition enables the \emph{lattice method} of tracking precision,
+where the precision of the input is specified as a lattice $H$ and precision
+is tracked via differentials of the steps within a given algorithm.
+The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
+yields the optimum possible precision. 
+We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
+exposition.
+We emphasize that polynomials, rational fractions (outside of their poles)
+and converging power series are strictly differentiable.
+
+The lemma is more precise in the context of so-called
+integral polynomial functions.
+A function $f : E \to F$ is said to be
+\emph{integral polynomial} if it is given by multivariate polynomial 
+functions with coefficients in $\O_K$ in any (equivalently all) system 
+of coordinates associated to a $\OK$-basis of $B_E(1)$.
+
+\begin{prop}
+\label{prop:precision2}
+We keep the notation of Proposition \ref{prop:precision} and assume 
+in addition that $f$ is integral polynomial. Let $C$ be a positive real
+number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
+Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
+\rho^{-1}$.
+\end{prop}
+
+We will use the following definition to measure the gap
+between lattice and coordinate-wise methods. 
+Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.
+
+\begin{deftn} \label{def:diffused}
+Let $H \subset E$ be a lattice. The number of 
+\emph{diffused digits of precision} of $H$ is the length of
+$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
+\end{deftn}
+If $H$ represents the actual precision on some object, then
+$H_0$ is the smallest diagonal lattice containing $H$.
+Coordinate-wise methods, with one precision datum per coordinate,
+corresponds to diagonal lattices.
+Hence, they cannot yield a precision better than $H_0$, and
+then $k$ provides a lower bound on the number of $p$-adic digits
+gained by lattice methods over standard methods.
 
@@
 \end{tabular}
+The last line provides the loss in precision between the
+original solution of the differential equation and its approximation
+obtained through computation.
 The theorical loss in precision for the coefficient of $t^{31}$
@@
 over $\mathbb{Q}_3$ with lattice precision.
+We use an initial precision of 50 digits.
 Any generic implementation of the Newton
@@
 \begin{tabular}{|l|}
-\verb?    R = QpL(3)? \\
+\verb?    R = QpL(3,50)? \\
 \verb?    A.<x> = PolynomialRing(R)?\\

### 2018-01-23 ZpL/ZpL.tex
ZpLC: how precision is tracked

@@
 \usepackage{tabularx,multicol}
-%\usepackage{tikz}
+\usepackage{tikz}
 
@@
 
+\newcommand{\calV}{\mathcal{V}}
+\newcommand{\ttv}{\texttt{v}\xspace}
+\newcommand{\ttw}{\texttt{w}\xspace}
+\newcommand{\tttmp}{\texttt{tmp}\xspace}
+
+\newcommand{\rem}{\,\%\,}
+
+\newcommand{\sage}{\textsc{SageMath}\xspace}
 \newcommand{\ZpL}{\texttt{ZpL}\xspace}
-\newcommand{\ZpLCA}{\texttt{ZpLCA}\xspace}
-\newcommand{\ZpLCR}{\texttt{ZpLCR}\xspace}
+\newcommand{\ZpLC}{\texttt{ZpLC}\xspace}
 \newcommand{\ZpLF}{\texttt{ZpLF}\xspace}
+
+\newcommand{\card}{\text{Card}\:}
+\newcommand{\ind}{\text{ind}}
+\newcommand{\coind}{\text{coind}}
 
@@
 \theoremstyle{definition}
-\newtheorem{rem}[theo]{Remark}
+\newtheorem{rmq}[theo]{Remark}
 \newtheorem{ex}[theo]{Example}
@@
 
+\todo{Convention: in matrices, vectors are represented by row-vector.}
+
 \section{Examples}
@@
 
-\section{Theory}
+\section{Presentation of the package}
+
+In this Section, we explain how our package \ZpL works and analyze
+its performances.
+The main theoretical result on which our package is based in the 
+ultrametric precision theory developed in \cite{padicprec}, which 
+suggests to track precision \emph{via} lattices and differential 
+computations. For this reason, our approach is very inspired by 
+Automatic Differentiation techniques \cite{} and our implementation 
+follows the usual Operator Overloading strategy \cite{}.
+
+We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
+this former is safer while the latter is faster. They are presented in 
+\S \ref{ssec:ZpLC} and \S \ref{ssec:ZpLF} respectively. As for \S 
+\ref{ssec:preclemma} below, it serves as a short survey on the result of 
+\cite{padicprec} on which our developments are built.
+
+\smallskip
+
+\noindent \textit{Remark about the naming.}
+%
+The letter \texttt{L}, which appears in the name of the package, 
+comes from ``lattices''. The letters \texttt{C} (in \ZpLC) and 
+\texttt{F} (in \ZpLF) stand for ``cap'' and ``float'' respectively.
 
 \subsection{The precision Lemma}
+\label{ssec:preclemma}
+
 In \cite{caruso-roe-vaccon:14a}, we have suggested the 
@@
 \emph{coordinate-wise method} (of \textit{e.g.}
- \verb?Zp(5)?) that is of use in Sage prior to
- our package, and  
+ \verb?Zp(5)?) that is used traditionally in \sage and
 where the precision of an element is specified by giving the precision
@@
 $E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
-that the differential $f'(v_0)$ is surjective.
+that the differential $df_{v_0}$ is surjective.
 Then, for all $\rho \in (0, 1]$, there exists a positive real
@@
 \label{eq:firstorder}
-f(v_0 + H) = f(v_0) + f'(v_0) (H).
+f(v_0 + H) = f(v_0) + df_{v_0} (H).
 \end{equation}
@@
 in addition that $f$ is integral polynomial. Let $C$ be a positive real
-number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
+number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
 Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
@@
 
-\subsection{The way the precision is tracked with \ZpL}
+\subsection{\ZpLC}
+\label{ssec:ZpLC}
 
-We follow the precision lattice.
-Closely related to automatic differentiation.
+\todo{Maybe, add a short intro.}
 
-\begin{itemize}
-\item \ZpLCR, \ZpLCA: we use a cap to ensure that we have a lattice at
-each step.
-\item \ZpLF: we work with $\Zp$-submodules of possible positive codimension
+\medskip
+
+\noindent \textbf{Tracking precision.}
+%
+In what follows, it will be convenient to use a notion of discrete time 
+represented by the letter $t$. Rigorously, it is defined as follows: 
+$t=0$ when the $p$-adic ring $\ZpLC(\:\cdots)$ is created and increases by 
+$1$ each time a variable is created, deleted\footnote{The deletion can
+be explicit (through a call to the \texttt{del} operator) or implicit
+(handled by the garbage collector).} or updated.
+
+Let $\calV_t$ be the set of alive variables at time $t$. Set $E_t = 
+\Qp^{\calV_t}$; it is a finite dimensional vector space over $\Qp$ which 
+should be thought as the set of all possible values that can be taken by 
+the variables in $\calV_t$. For $\ttv \in \calV_t$, let $e_\ttv \in
+\Qp^{\calV_t}$ be the vector whose all coordinates vanished except 
+that at position $\ttv$ which takes the value $1$. The family 
+$(e_\ttv)_{\ttv \in \calV_t}$ is obviously a basis of $E_t$; we will 
+refer to it as the \emph{canonical basis}.
+
+The package \ZpLC follows precision by keeping track of a lattice 
+$H_t$ in $E_t$, which is a unique object whose purpose is to model the 
+precision on all the variables in $\calV_t$ together.
+Concretely, this lattice is represented by a matrix $M_t$ in row-echelon 
+form whose lines form a set of generators.
+Below, we explain how the matrices $M_t$ are updated each time
+$t$ increases.
+
+\smallskip
+
+\noindent \textit{Creating a variable.}
+%
+This happens when we encounter an instruction having one of the
+two following forms:
+
+\medskip
+
+\noindent \hspace{5mm} \makebox[2.5cm]{[Computation]\hfill\null}
+\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?
+
+\smallskip
+
+\noindent \hspace{5mm} \makebox[2.5cm]{[New value]\hfill\null}
+\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?
+
+\medskip
+
+\noindent
+In both cases, $\ttw$ is the newly created variable. 
+The $\ttv_i$'s stand for already defined variables and $f$ is some 
+$n$-ary builtin function (in most cases it is just addition, 
+subtraction, multiplication or division). On the contrary, the terms 
+``value'' and ``prec'' refer to user-specified constants or integral 
+values which was computed earlier.
+
+\smallskip
+
+Let us first examine the first construction [Computation].
+With our conventions, if $t$ is the time just before the execution of 
+the instruction we are interested in, the $\ttv_i$'s lie in $\calV_t$ 
+while $\ttw$ does not. Moreover $\calV_{t+1} = \calV_t \sqcup \{\ttw\}$, 
+so that $E_{t+1} = E_t \oplus \Qp e_\ttw$.
+The mapping taking the values of variables at time $t$ to that at 
+time $t{+}1$ is:
+$$\begin{array}{rcl}
+F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
+\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n)
+\end{array}$$
+where $x_i$ is the $\ttv_i$-th coordinate of the vector $\underline x$.
+The Jacobian matrix of $F$ at $\underline x$ is easily computed; it 
+is the block matrix $J_{\underline x}(F) = \big( \begin{matrix} I & L
+\end{matrix} \big)$ where $I$ is the identity matrix of size $\text{Card}
+\:\calV_t$ and $L$ is the column vector whose $\ttv$-th entry is
+$\frac{\partial f}{\partial \ttv} (\underline x)$ if $\ttv$ is one of 
+the $\ttv_i$'s and $0$ otherwise.
+Therefore, the image of $H_t$ under $dF_{\underline x}$ is represented 
+by the matrix $J_{\underline x}(F) \cdot M_t = \big( \begin{matrix} M_t & C
+\end{matrix} \big)$ where $C$ is the column vector:
+$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
+\cdot C_i$$
+where $C_i$ is the column vector of $M_t$ corresponding to the
+variable $\ttv_i$.
+Observe that the matrix $J_{\underline x}(F) \cdot M_t$ is no longer a 
+square matrix; it has one extra column. This reflects the fact that 
+$\dim E_{t+1} = \dim E_t + 1$. Rephasing this in a different language, 
+the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
+$E_{t+1}$ but is included in an hyperplane.
+
+The package \ZpLC tackles this issue by introducing a cap: we do not 
+work with $dF_{\underline x}(H_t)$ but instead define the lattice: 
+$$H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$$
+where $N_{t+1}$ is an integer, the so-called \emph{cap}.
+Alternatively, one may introduce the application:
+$$\begin{array}{rcl}
+\tilde F: \quad E_t \oplus \Qp & \longrightarrow & E_{t+1} \smallskip \\
+\underline x \oplus c & \mapsto & \underline x \:\oplus 
+\big(f(x_1, \ldots, x_n) + c\big).
+\end{array}$$
+The lattice $H_{t+1}$ is then the image of $H_t \oplus p^{N_{t+1}}
+\Zp$ under $d\tilde F_{(\underline x, \star)}$ for any value of $\star$.
+The choice of the 
+cap is of course a sensitive question. \ZpLC proceeds as follows. When a 
+ring is created, it comes with two constants (which can be specified by 
+the user): a relative cap \textsc{relcap} and an absolute cap 
+\textsc{abscap}. With these predefined values, the chosen cap is:
+$$N_{t+1} = 
+\min\big(\textsc{abscap},\, \textsc{relcap} + v_p(y)\big)$$
+with $y = f(x_1, \ldots, x_n)$.
+
+In concrete terms, the lattice $H_{t+1}$ is represented by the block 
+matrix:
+$$\left(\begin{matrix}
+M_t & C \smallskip \\ 0 & p^{N_{t+1}}
+\end{matrix}\right).$$
+Performing row operations, we see then the entries of $C$ can be
+reduced modulo $p^{N_{t+1}}$ without changing the lattice. In order
+to optimize the size of the objects, we perform this reduction and
+define $M_{t+1}$ by:
+$$M_{t+1} = \left(\begin{matrix}
+M_t & C \text{ mod } p^{N_{t+1}} \smallskip \\ 0 & p^{N_{t+1}}
+\end{matrix}\right).$$
+We observe in particular that $M_{t+1}$ is still in row-echelon form.
+
+\smallskip
+
+Finally, we need to precise which value is set to the newly created 
+variable $\ttw$. We observe that it cannot be exactly $f(x_1, \ldots, 
+x_n)$ because the latter is \emph{a priori} a $p$-adic number which 
+cannot be computed exactly. For this reason, we have to truncate it at 
+some finite precision. Again we choose the precision $O(p^{N_{t+1}})$,
+\emph{i.e.} we define $x_\ttw$ as $f(x_1, \ldots, x_n) \text{ mod } 
+p^{N_{t+1}}$. This makes sense because:
+$$\bar x \oplus f(x_1, \ldots, x_n) \equiv
+\bar x \oplus x_\ttw \pmod{H_{t+1}}$$
+thanks to the extra generator we have added.
+
+\medskip
+
+\noindent \textit{Deleting a variable.}
+%
+Let us now examine the case where a variable $\ttw$ is deleted (or 
+collected by the garbage collector). Just after the deletion, at
+time $t{+}1$, we then have $\calV_{t+1} = \calV_t \backslash \{\ttw\}$.
+Thus $E_t = E_{t+1} \oplus \Qp e_\ttw$. Moreover, the deletion of $\ttw$
+is modeled by the canonical projection $f : E_t \to E_{t+1}$. Since $f$
+is linear, it is its own differential (at each point) and we set 
+$H_{t+1} = f(H_t)$.
+
+A matrix representing $H_{t+1}$ is deduced from $M_t$ by erasing the
+column corresponding to $\ttw$. However the matrix we get this way
+has the following shape:
+$$\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
+\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
+    --(3,-3)--(3,-5)--(4,-5)--(4,-6)--(5,-6)--(5,-7)--(6,-7)--(6,0)--cycle;
+\draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
+    --(3,-3)--(3,-5)--(4,-5)--(4,-6)--(5,-6)--(5,-7)--(6,-7);
+\draw[thin] (6,-7)--(6,0)--(0,0);
+\end{tikzpicture}}\right)$$
+where a cell is colored when it can contain a non-vanishing entry.
+At this point, we encounter an issue: the above matrix is no longer
+echelonized (except the $\ttw$ corresponded to the last column, 
+\emph{i.e.} was the latest created variable which was still alive). 
+We then need to re-echelonize it before proceeding to the next
+instruction. However, we notice that only the left bottom corner
+needs to be echelonized. We will come back to this observation later
+when we will study complexity.
+
+\medskip
+
+\noindent \textit{Updating a variable.}
+%
+Just like for creation, this happens when the program reaches an
+affectation \verb?w = ...? where the variable \ttw is already 
+defined.
+
+This situation can actually be decomposed into three steps: first, we 
+compute the right-hand-side and store it in a temporary variable \tttmp; 
+second, we delete \ttw; third, we rename \tttmp in \ttw. The first step 
+corresponds to the creation of a variable and was then already studied. 
+Similarly, the second step is a variable deletion and again has already 
+been discussed. As for the third step, it just changes the name of an
+element of $\calV_t$ but does not affect the lattice $H_t$, nor the
+matrix $M_t$.
+
+\medskip
+
+\noindent \textbf{Complexity.}
+%
+We now briefly discuss the cost of the above operations. For this,
+it is convenient to introduce a total order on $\calV_t$: for $\ttv, 
+\ttw \in \calV_t$, we convene that $\ttv <_t \ttw$ is \ttv was created 
+before~\ttw. By construction, the columns of the matrix $M_t$ are 
+ordered with respect to $<_t$.
+For $\ttv \in \calV_t$, we define the \emph{index} of \ttv, denoted by 
+$\ind_t(\ttv)$ as the number of elements of $\calV_t$ which are not 
+greater than \ttv. If we sort the elements of $\calV_t$ by increasing 
+order, \ttv then appears in $\ind_t(\ttv)$-th position.
+We also define the \emph{coindex} of \ttv by
+$\coind_t(\ttv) = \card \calV_t - \ind_t(\ttv)$.
+
+\medskip
+
+\noindent \textit{Creating a variable.}
+
+$$O\bigg(\sum_{i=1}^n \ind_t(\ttv_i)\bigg) \subset 
+O\big(n \cdot \card \calV_t\big)$$
+
+\medskip
+
+\noindent \textit{Deleting a variable.}
+
+$$O\big(\coind_t(\ttw)^2\big)$$
+
+\medskip
+
+\noindent \textit{Temporal locality.}
+
+\medskip
+
+\noindent \textbf{Correctness.}
+
+\subsection{\ZpLF}
+\label{ssec:ZpLF}
+
+we work with $\Zp$-submodules of possible positive codimension
 and represent them using matrices of floating point $p$-adics.
-\end{itemize}
 
-\subsection{The notion of working precision}
+\subsubsection*{Tracking precision}
 
-\begin{itemize}
-\item what is it?
-\item how can we compute it? (Answer: Hermite reduction of the transpose)
-\item how can we estimate it?
-\end{itemize}
-
-\subsection{Correctness}
-
-\begin{itemize}
-\item prove that \ZpLCA and \ZpLCR are correct under some assumptions
-(to be determined)
-\item explain that \ZpLF is not meant to provide proved results
-\end{itemize}
-
-\subsection{Complexity}
-
-\begin{tabular}{l|c|c|}
-& \ZpLCA/\ZpLCR & \ZpLF \\
-\hline
-Creation of a variable & $O(n)$ & $O(n_0)$ \\
-\hline
-Deletion of a variable & $O(m^2)$ & $-$ \\
-\hline
-\end{tabular}
-
-discussion on temporal locality
-
-\section{Presentation of the package}
-
-\begin{itemize}
-\item Discuss coercion/conversion: \texttt{Zp} coerces to \ZpL but no coersion
-in the other direction
-\item How to access to the precision
-\item Diffused digits
-\item Historic of variables creation/deletion, etc...
-\item Other features?
-\end{itemize}
+\subsubsection*{Complexity}
 

### 2018-01-29 ZpL/ZpL.tex
ZpLF + absolute precision

@@
 
+\newcommand{\ph}{\vphantom{$A^A_A$}}
+
 \newcommand{\N}{\mathbb N}
@@
 \newcommand{\tttmp}{\texttt{tmp}\xspace}
+\newcommand{\calU}{\mathcal{U}}
+\newcommand{\calK}{\mathcal{K}}
 
@@
 \newcommand{\sage}{\textsc{SageMath}\xspace}
-\newcommand{\ZpL}{\texttt{ZpL}\xspace}
-\newcommand{\ZpLC}{\texttt{ZpLC}\xspace}
-\newcommand{\ZpLF}{\texttt{ZpLF}\xspace}
+\newcommand{\ZpL}{\text{\rm \tt ZpL}\xspace}
+\newcommand{\ZpLC}{\text{\rm \tt ZpLC}\xspace}
+\newcommand{\ZpLF}{\text{\rm \tt ZpLF}\xspace}
 
@@
 
-\newcommand{\calU}{\mathcal{U}}
-
 \newcommand{\softO}{O\tilde{~}}
@@
 \newcommand{\app}{\text{\rm app}}
+
+\newcommand{\cIn}{{\color{blue} \tt \phantom{Zp}In}:}
+\newcommand{\cOut}{{\color{red} \tt \phantom{Z}Out}:}
+\newcommand{\cZpCR}{{\color{red} \tt ZpCR}:}
+\newcommand{\cZpFP}{{\color{red} \tt ZpFP}:}
+\newcommand{\cZpLC}{{\color{red} \tt ZpLC}:}
+\newcommand{\cZpLF}{{\color{red} \tt ZpLF}:}
 
@@
 
-\keywords{Algorithms, $p$-adic precision, characteristic polynomial,
-eigenvalue}
+\keywords{Algorithms, $p$-adic precision, Automatic Differentiation}
 
 \begin{abstract}
+We present a new package \ZpL for the Computer Algebra system \sage. It 
+implements a sharp tracking of precision on $p$-adic numbers, following 
+the theory of ultrametric precision introduced in \cite{padicprec}. The 
+underlying algorithms are mostly based on Automatic Differentiation 
+techniques. We introduce them, study their complexity and discuss our 
+design choices.
+We illustrate the benefits of our package (in comparison with previous 
+implementations) with a large sample of examples coming from Linear 
+Algebra, Commutative Algebra and Differential Equations.
 \end{abstract}
@@
 follows the usual Operator Overloading strategy \cite{}.
+We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
+this former is safer while the latter is faster.
 
-We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
-this former is safer while the latter is faster. They are presented in 
-\S \ref{ssec:ZpLC} and \S \ref{ssec:ZpLF} respectively. As for \S 
-\ref{ssec:preclemma} below, it serves as a short survey on the result of 
-\cite{padicprec} on which our developments are built.
+This Section is organized as follows.
+\todo{[...]}.
 
@@
 
-\subsection{\ZpLC}
-\label{ssec:ZpLC}
+\subsection{Tracking precision}
 
@@
 
-\medskip
-
-\noindent \textbf{Tracking precision.}
-%
 In what follows, it will be convenient to use a notion of discrete time 
@@
 
-The package \ZpLC follows precision by keeping track of a lattice 
-$H_t$ in $E_t$, which is a unique object whose purpose is to model the 
-precision on all the variables in $\calV_t$ together.
+\subsubsection{The case of \ZpLC.}
+
+Following Proposition~\ref{prop:precision},
+the package \ZpLC follows the precision by keeping track of a lattice
+$H_t$ in $E_t$, which is a global object whose purpose is to model the 
+precision on all the variables in $\calV_t$ all together.
 Concretely, this lattice is represented by a matrix $M_t$ in row-echelon 
-form whose lines form a set of generators.
+form whose rows form a set of generators.
 Below, we explain how the matrices $M_t$ are updated each time
@@
 Alternatively, one may introduce the application:
-$$\begin{array}{rcl}
+\begin{equation}
+\label{eq:tildeF}
+\begin{array}{rcl}
 \tilde F: \quad E_t \oplus \Qp & \longrightarrow & E_{t+1} \smallskip \\
@@
 \big(f(x_1, \ldots, x_n) + c\big).
-\end{array}$$
+\end{array}
+\end{equation}
 The lattice $H_{t+1}$ is then the image of $H_t \oplus p^{N_{t+1}}
@@
 \textsc{abscap}. With these predefined values, the chosen cap is:
-$$N_{t+1} = 
-\min\big(\textsc{abscap},\, \textsc{relcap} + v_p(y)\big)$$
+\begin{equation}
+\label{eq:cap}
+N_{t+1} = 
+\min\big(\textsc{abscap},\, \textsc{relcap} + v_p(y)\big)
+\end{equation}
 with $y = f(x_1, \ldots, x_n)$.
@@
 
-\smallskip
-
 Finally, we need to precise which value is set to the newly created 
@@
 thanks to the extra generator we have added.
+
+\smallskip
+
+The second construction
+``\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?''
+is easier to handle since, roughly speaking, it corresponds to the
+case $n = 0$.
+In this situation, keeping in mind the cap, the lattice $H_{t+1}$ is 
+defined by:
+$$H_{t+1} = H_t + p^{\min(\text{prec}, N_{t+1})} \Zp e_\ttw$$
+where the cap $N_{t+1}$ is:
+$$N_{t+1} = \min\big(\textsc{abscap},\, \textsc{relcap} + 
+v_p(\text{value})\big).$$
+The corresponding matrix $M_{t+1}$ is then given by:
+$$M_{t+1} = \left(\begin{matrix}
+M_t & 0 \smallskip \\ 0 & p^{\min(\text{prec}, N_{t+1})}
+\end{matrix}\right).$$
 
@@
 $H_{t+1} = f(H_t)$.
+A matrix representing $H_{t+1}$ is deduced from $M_t$ by erasing the 
+column corresponding to $\ttw$. However the matrix we get this way is 
+no longer in row-echelon form. We then need to re-echelonize it.
 
-A matrix representing $H_{t+1}$ is deduced from $M_t$ by erasing the
-column corresponding to $\ttw$. However the matrix we get this way
-has the following shape:
+\smallskip
+
+More precisely, in the \ZpLC case, the obtained matrix has this 
+shape:
 $$\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
+\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
+    --(3,-3)--(3,-4)--(4,-4)--(4,-5)--(5,-5)--(5,-6)--(6,-6)
+    --(6,-7)--(7,-7)--(7,0)--cycle;
+\draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
+    --(3,-3)--(3,-4)--(4,-4)--(4,-5)--(5,-5)--(5,-6)--(6,-6)
+    --(6,-7)--(7,-7);
+\draw[thin] (7,-7)--(7,0)--(0,0);
+\fill (3,0) rectangle (4,-4);
+\draw[->] (2.9,-4.8)--(3.2,-4.2);
+\node[scale=0.8] at (2,-5.2) { deleted };
+\node[scale=0.8] at (2,-6) { column };
+\end{tikzpicture}}\right)
+\quad = \quad
+\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
 \fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
@@
 where a cell is colored when it can contain a non-vanishing entry.
-At this point, we encounter an issue: the above matrix is no longer
-echelonized (except the $\ttw$ corresponded to the last column, 
-\emph{i.e.} was the latest created variable which was still alive). 
-We then need to re-echelonize it before proceeding to the next
-instruction. However, we notice that only the left bottom corner
-needs to be echelonized. We will come back to this observation later
-when we will study complexity.
+The top part of the matrix is then already echenolized, so that we
+only have to re-echelonize the bottom right corner whose size is
+the distance from the column corresponding to the erased variable
+to the end. 
+Thanks to the particular shape of the matrix, the echelonization can be 
+performed efficiently: we combine the first rows (of the bottom right 
+part) in order to clear the first unwanted nonzero entry and then 
+proceed recursively.
 
@@
 Just like for creation, this happens when the program reaches an
-affectation \verb?w = ...? where the variable \ttw is already 
+affectation ``\verb?w = ...?'' where the variable \ttw is already 
 defined.
@@
 
+\subsubsection{The case of \ZpLF.}
+
+The way the package \ZpLF tracks precision is based on similar 
+techniques but differs from \ZpLC in that it does not introduce a cap 
+but instead allows $H_t$ to be a sub-$\Zp$-module of $E_t$ of any 
+codimension.
+This point of view is nice because it implies smaller objects and 
+consequently leads to faster algorithms. However, it has a huge 
+drawback; indeed, unlike lattices, submodules of $E_t$ of arbitrary 
+codimensions are \emph{not} exact objects, in the sense that they cannot 
+be represented by integral matrices in full generality. Consequently,
+they cannot be encoded on a computer.
+We work around this drawback by replacing everywhere exact $p$-adic 
+numbers by Floating Point $p$-adic numbers (at some given precision) 
+\cite{}.
+
+The fact that the lattice $H_t$ can now have arbitrary codimension
+translates to the fact the matrix $M_t$ can be rectangular.
+Precisely, we will maintain matrices $M_t$ of the shape:
+\begin{equation}
+\label{eq:shapeM-ZpLF}
+\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
+\fill[black!20] (0,0)--(0,-1)--(3,-1)--(3,-2)--(5,-2)--(5,-3)
+    --(6,-3)--(6,-4)--(8,-4)--(8,-5)--(9,-5)--(9,-6)--(12,-6)
+    --(12,-7)--(14,-7)--(14,0)--cycle;
+\draw (0,0)--(0,-1)--(3,-1)--(3,-2)--(5,-2)--(5,-3)
+    --(6,-3)--(6,-4)--(8,-4)--(8,-5)--(9,-5)--(9,-6)--(12,-6)
+    --(12,-7)--(14,-7);
+\fill (0,0) rectangle (1,-1);
+\fill (3,-1) rectangle (4,-2);
+\fill (5,-2) rectangle (6,-3);
+\fill (6,-3) rectangle (7,-4);
+\fill (8,-4) rectangle (9,-5);
+\fill (9,-5) rectangle (10,-6);
+\fill (12,-6) rectangle (13,-7);
+\draw[thin] (14,-7)--(14,0)--(0,0);
+\end{tikzpicture}}\right)
+\end{equation}
+where only the colored cells may contain a nonzero value and the 
+black cells ---the so-called \emph{pivots}--- do not vanish. A
+variable whose corresponding column contains a pivot will be called 
+a \emph{pivot variable at time $t$}.
+
+Below, we explain in more details how the package \ZpLF updates 
+the matrix $M_t$ when a variable is created or deleted.
+
 \medskip
 
-\noindent \textbf{Complexity.}
+\noindent \textit{Creating a variable.}
 %
-We now briefly discuss the cost of the above operations. For this,
+We assume first that the newly created variable is defined through an
+affectation of the form:
+
+\medskip
+
+\hfill\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?\hfill\null
+
+\medskip
+
+\noindent
+As already explained in the case of \ZpLC, this line of code is 
+modeled by the mathematical mapping:
+$$\begin{array}{rcl}
+F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
+\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n).
+\end{array}$$
+Here $\underline x$ represents the state of memory at time $t$, 
+and $x_i$ is the coordinate of $\underline x$ corresponding to the
+variable $\ttv_i$.
+
+In the \ZpLF framework, $H_{t+1}$ is defined as the image of $H_t$
+under the differential $dF_{\underline x}$. Accordingly, the matrix
+$M_{t+1}$ is defined as $M_{t+1} = \left(\begin{matrix}
+M_t & C \end{matrix}\right)$
+where $C$ is the column vector:
+$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
+\cdot C_i.$$
+However, we insist on the fact that all the above computations take
+place in the ``ring'' of Floating Point $p$-adic numbers. Therefore,
+we cannot guarantee that the rows of $M_{t+1}$ generate $H_{t+1}$.
+Nonetheless, they generate a module which is expected to be closed 
+to $H_{t+1}$.
+
+\smallskip
+
+If \ttw is created by the code
+``\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?'',
+we define $H_{t+1} = H_t \oplus p^{\text{prec}} \Zp e_\ttw$ and consequently:
+$$M_{t+1} = \left(\begin{matrix}
+M_t & 0 \smallskip \\ 0 & p^{\text{prec}}
+\end{matrix}\right)$$
+If $\text{prec}$ is $+\infty$ (or, equivalently, not specified), we
+agree that $H_{t+1} = H_t$ and $M_{t+1} = (\begin{matrix} M_t & 0 
+\end{matrix})$.
+
+\medskip
+
+\noindent \textit{Deleting a variable.} 
+% 
+As for \ZpLC, the matrix operation implied by the deletion of the 
+variable \ttw is modeled by the deletion of the corresponding column 
+of $M_t$. If \ttw is not a pivot variable at time $t$, the matrix 
+$M_t$ keeps the form \eqref{eq:shapeM-ZpLF} after erasure; therefore
+no more treatment is needed in this case.
+
+Otherwise, we re-echelonize the matrix as follows. After the deletion 
+of the column $C_\ttw$, we examine the first column $C$ which was 
+located on the right of $C_\ttw$. Two situations may occur (depending
+on the fact that $C$ was or was not a pivot column):
+
+\noindent \hfill
+\begin{tikzpicture}[scale=0.5]
+\fill[black!20] (0,0)--(1,0)--(1,-2)--(2,-2)--(2,2)--(0,2)--(0,0);
+\draw (0,0)--(1,0)--(1,-2)--(2,-2);
+\node at (1.5,2.8) { $C$ };
+\draw[->] (1.5,2.4)--(1.5,2.1);
+\node at (1.5,-0.5) { \ph $x$ };
+\node at (1.5,-1.5) { \ph $y$ };
+\node[scale=0.9] at (1,-2.8) { First case };
+\begin{scope}[xshift=5cm]
+\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,2)--(0,2)--(0,0);
+\draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2);
+\node at (1.5,2.8) { $C$ };
+\draw[->] (1.5,2.4)--(1.5,2.1);
+\node at (1.5,-1.5) { \ph $y$ };
+\node[scale=0.9] at (1,-2.8) { Second case };
+\end{scope}
+\end{tikzpicture}
+\hfill\null
+
+\noindent
+
+In the first case, we perform row operations in order to replace the 
+couple $(x,y)$ by $(d, 0)$ where $d$ is element of smallest valuation 
+among $x$ and $y$. Observe that $y$ is necessarily nonzero in this case, 
+so that $d$ does not vanish as well. After this operation, we move to
+the next column and repeat the same process.
+
+The second case is divided into two subcases. First, if $y$ does not 
+vanish, it can serve as a pivot and the obtained matrix has the desired 
+shape. When this occurs, the echelonization stops. On the contrary, if 
+$y = 0$, we just untint the corresponding cell and move to the next 
+column without modifying the matrix.
+
+\subsection{Visualizing the precision}
+
+Our package implements several methods giving access to the precision 
+structure. In the subsection, we present and discuss the most relevant 
+features in this direction.
+
+\medskip
+
+\noindent \textbf{Absolute precision of one element.}
+%
+This is the simplest accessible precision datum.
+It is encapsulated in the notation when an element is printed. For
+example, the (partial) session:
+
+\smallskip
+
+\noindent
+\begin{tabular}{rl}
+\cIn  & \verb?R = ZpLC(2, print_mode='digits')? \\
+      & \verb?v = R(173,10); v? \\
+\cOut & \verb?...0010101101?
+\end{tabular}
+
+\smallskip
+
+\noindent
+indicates that the absolute precision on \ttv is $10$ since
+exactly $10$ digits are printed.
+The method \verb?precision_absolute? provides a more easy-to-use
+access to the absolute precision.
+
+\smallskip
+
+\noindent
+\begin{tabular}{rl}
+\cIn  & \verb?v.precision_absolute()? \\
+\cOut & \verb?10?
+\end{tabular}
+
+\smallskip
+
+\noindent
+Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at
+time $t$) as the smallest valuation of an entry of $C_{t,\ttv}$ 
+defined as the column of $M_t$ corresponding to the variable \ttv.
+This sounds revelant because, if we believe that the submodule $H_t
+\subset E_t$ is supposed to encode the precision on the variables
+in $\calV_t$, Proposition~\ref{prop:precision} applied with the 
+projection onto the $\ttv$-component $\pi_\ttv : E_t \to \Qp e_\ttv$ 
+indicates that a good candidate for the precision on $e_\ttv$ is 
+$\pi_\ttv(H_t)$. Identifying $\Qp e_\ttv$ with $\Qp$, $\pi_\ttv(H_t)$ is 
+the $\Zp$-module generated by the entries of $C_{t,\ttv}$; it is then 
+equal to $p^N \Zp$ where $N$ is the smallest valuation of an entry of 
+$C_{t,\ttv}$.
+
+\medskip
+
+\noindent \textit{About correctness.}
+%
+We emphasize that the absolute precision computed this way is \emph{not} 
+proved, either for \ZpLF or \ZpLC. However, in the case of \ZpLC, one
+can be slightly more precise. Let $\calU_t$ be the vector space of 
+user-defined variables before time $t$ and $U_t$ be the lattice
+modeling the precision on them. The pair $(\calU_t, U_t)$ is defined 
+inductively as follows: we set $\calU_0 = U_0 = 0$ and 
+$\calU_{t+1} = \calU_t \oplus \Qp e_\ttw$,
+$U_{t+1} = U_t \oplus p^{\text{prec}} \Zp e_\ttw$
+when a new variable \ttw is created by
+``\verb?w = R(?value\verb?, ?prec\verb?)?''; otherwise, we put 
+$\calU_{t+1} = \calU_t$ and $U_{t+1} = U_t$. Moreover the values
+entered by the user defines a vector (with integral coordinates)
+$\underline u_t \in \calU_t$.
+
+Similarly, in order to model the caps, we define a pair $(\calK_t, 
+K_t)$ by the recurrence $\calK_{t+1} = \calK_t \oplus \Qp e_\ttw$,
+$K_{t+1} = K_t \oplus p^{N_{t+1}} \Zp e_\ttw$
+each time a new variable \ttw is created. 
+Here, the exponent $N_{t+1}$ is the cap defined by Eq.~\eqref{eq:cap}.
+In case of deletion, we put $\calK_{t+1} = \calK_t$ and $K_{t+1} = K_t$.
+
+Taking the compositum of all the functions $\tilde F$ (\emph{cf} 
+Eq.~\eqref{eq:tildeF}) from time $0$ to $t$, we find that the execution 
+of the session until time $t$ is modeled by a mathematical function
+$\Phi_t : \calU_t \oplus \calK_t \to E_t$.
+From the design of \ZpLC, we deduce further that there exists a 
+vector $\underline k_t \in K_t$ such that:
+$$\Phi_t(\underline u_t \oplus \underline k_t) = \underline x_t 
+\quad \text{and} \quad
+d\Phi_t (U_t \oplus K_t) = H_t$$
+where the differential of $\Phi_t$ is taken at the point
+$\underline u_t \oplus \underline k_t$.
+
+\medskip
+
+\noindent \textbf{Precision on a subset of elements.}
+
+\medskip
+
+\noindent \textbf{Diffused digits of precision.}
+
+\subsection{Complexity}
+
+We now discuss the cost of the above operations.
+For this,
 it is convenient to introduce a total order on $\calV_t$: for $\ttv, 
@@
 
-\noindent \textit{Creating a variable.}
+\noindent \textbf{Creating a variable.}
 
-$$O\bigg(\sum_{i=1}^n \ind_t(\ttv_i)\bigg) \subset 
-O\big(n \cdot \card \calV_t\big)$$
+\ZpLC: $O\bigg(\sum_{i=1}^n \ind_t(\ttv_i)\bigg) \subset 
+O\big(n \cdot \card \calV_t\big)$
+
+\ZpLF:
 
@@
 
-\subsection{\ZpLF}
-\label{ssec:ZpLF}
-
-we work with $\Zp$-submodules of possible positive codimension
-and represent them using matrices of floating point $p$-adics.
-
 \subsubsection*{Tracking precision}
@@
 
-\subsection{More examples}
+\section{More examples}
 

### 2018-01-29 ZpL/ZpL.tex
More on absolute precision of an element

@@
 \theoremstyle{definition}
-\newtheorem{rmq}[theo]{Remark}
+\newtheorem{rmk}[theo]{Remark}
 \newtheorem{ex}[theo]{Example}
@@
 \noindent
-Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at
-time $t$) as the smallest valuation of an entry of $C_{t,\ttv}$ 
-defined as the column of $M_t$ corresponding to the variable \ttv.
-This sounds revelant because, if we believe that the submodule $H_t
-\subset E_t$ is supposed to encode the precision on the variables
-in $\calV_t$, Proposition~\ref{prop:precision} applied with the 
-projection onto the $\ttv$-component $\pi_\ttv : E_t \to \Qp e_\ttv$ 
+
+Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at time 
+$t$) as the smallest valuation of an entry of the column of $M_t$ 
+corresponding to the variable \ttv. Alternatively, it can be defined as 
+the unique integer $N$ for which $\pi_\ttv(H_t) = p^N \Zp$ where 
+$\pi_\ttv : E_t \to \Qp$ takes a vector to its \ttv-coordinate.
+This definition of the absolute precision sounds revelant because, if we 
+believe that the submodule $H_t \subset E_t$ is supposed to encode the 
+precision on the variables in $\calV_t$, 
+Proposition~\ref{prop:precision} applied with the mapping $\pi_\ttv$ 
 indicates that a good candidate for the precision on $e_\ttv$ is 
-$\pi_\ttv(H_t)$. Identifying $\Qp e_\ttv$ with $\Qp$, $\pi_\ttv(H_t)$ is 
-the $\Zp$-module generated by the entries of $C_{t,\ttv}$; it is then 
-equal to $p^N \Zp$ where $N$ is the smallest valuation of an entry of 
-$C_{t,\ttv}$.
+$\pi_\ttv(H_t) = p^N \Zp$.
 
@@
 $\underline u_t \oplus \underline k_t$. 
-
-\medskip
+Set $\Phi_{t,\ttv} = \pi_\ttv \circ \Phi_t$. It maps 
+$\underline u_t \oplus \underline k_t$ to the \ttv-coordinate 
+$x_{t,\ttv}$ of $\underline x_t$ and satisfies:
+$$d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$$
+where $N$ is the value returned by \verb?precision_absolute?. Thus, 
+as soon as the assumptions of Proposition \ref{prop:precision} are
+fulfilled, we derive
+$\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus (\underline k_t + K_t)\big) = 
+x_{t,\ttv} + p^N \Zp$.
+Noting that $k_t \in K_t$, we finally get:
+\begin{equation}
+\label{eq:absprec}
+\Phi_{t,\ttv} (\underline u_t + U_t) \subset
+\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus K_t\big) = 
+x_{t,\ttv} + p^N \Zp.
+\end{equation}
+The latter inclusion means that the computed value $x_{t,\ttv}$ is
+accurate at precision $O(p^N)$, \emph{i.e.} that the output absolute
+precision is correct.
+
+Unfortunately, checking automatically the assumptions of Proposition 
+\ref{prop:precision} in full generality seems to be difficult, though it 
+can be done by hand for many particular examples \cite{}.
+
+\begin{rmk}
+Assuming that Proposition \ref{prop:precision} applies, 
+the absolute precision computed as above is optimal if and only if
+the inclusion of \eqref{eq:absprec} is an equality. Applying again
+Proposition \ref{prop:precision}
+with the restricted mapping $\Phi_{t,\ttv} : \calU_t \to \Qp$ and the
+lattice $U_t$, we find that this happens if and only if
+$d\Phi_{t,\ttv}(U_t) = p^N \Zp$.
+
+Unfortunately, the latter condition cannot be checked on the matrix 
+$M_t$ (because of reductions). 
+However it is possible (and easy) to check whether the weaker condition 
+$d\Phi_{t,\ttv}(K_t) \subsetneq p^N \Zp$. This checking is achieved by 
+the method
+\verb?is_precision_capped? (provided by our package) which returns true 
+if $d\Phi_{t,\ttv}(K_t) = p^N \Zp$. As a consequence, when this method
+answers false, the absolute precision computed by the software is
+likely optimal.
+\end{rmk}
 
@@
 
-\noindent \textbf{Diffused digits of precision.}
+\noindent \textit{Diffused digits of precision.}
 

### 2018-01-29 ZpL/ZpL.tex
Precision on a subset of elements and diffused digits

@@
 
-The lemma is more precise in the context of so-called
-integral polynomial functions.
-A function $f : E \to F$ is said to be
-\emph{integral polynomial} if it is given by multivariate polynomial 
-functions with coefficients in $\O_K$ in any (equivalently all) system 
-of coordinates associated to a $\OK$-basis of $B_E(1)$.
-
-\begin{prop}
-\label{prop:precision2}
-We keep the notation of Proposition \ref{prop:precision} and assume 
-in addition that $f$ is integral polynomial. Let $C$ be a positive real
-number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
-Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
-\rho^{-1}$.
-\end{prop}
-
-We will use the following definition to measure the gap
-between lattice and coordinate-wise methods. 
-Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.
-
-\begin{deftn} \label{def:diffused}
-Let $H \subset E$ be a lattice. The number of 
-\emph{diffused digits of precision} of $H$ is the length of
-$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
-\end{deftn}
-If $H$ represents the actual precision on some object, then
-$H_0$ is the smallest diagonal lattice containing $H$.
-Coordinate-wise methods, with one precision datum per coordinate,
-corresponds to diagonal lattices.
-Hence, they cannot yield a precision better than $H_0$, and
-then $k$ provides a lower bound on the number of $p$-adic digits
-gained by lattice methods over standard methods.
+%The lemma is more precise in the context of so-called
+%integral polynomial functions.
+%A function $f : E \to F$ is said to be
+%\emph{integral polynomial} if it is given by multivariate polynomial 
+%functions with coefficients in $\O_K$ in any (equivalently all) system 
+%of coordinates associated to a $\OK$-basis of $B_E(1)$.
+%
+%\begin{prop}
+%\label{prop:precision2}
+%We keep the notation of Proposition \ref{prop:precision} and assume 
+%in addition that $f$ is integral polynomial. Let $C$ be a positive real
+%number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
+%Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
+%\rho^{-1}$.
+%\end{prop}
 
@@
 \noindent
-
 Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at time 
@@
 \noindent \textbf{Precision on a subset of elements.}
+%
+Our package implements the method \verb?precision_lattice? through which 
+we can have access to the joint precision on a set of variables: it 
+outputs a matrix (in echelon form) whose rows generates a lattice 
+representing the precision on the subset of given variables.
+
+When the variables are ``independant'', the precision lattice is
+split and the method \verb?precision_lattice? outputs a diagonal
+matrix:
+
+\smallskip
+
+\noindent
+\begin{tabular}{rl}
+\cIn  & \verb?R = ZpLC(2, print_mode='digits')? \\
+      & \verb?x = R(987,10); y = R(21,5)? \\
+      & \verb?R.precision().precision_lattice([x,y])? \\
+\cOut & \verb?[1024    0]? \\
+      & \verb?[   0   32]?
+\end{tabular}
+
+\smallskip
+
+\noindent
+However, after some computations, the precision matrix evolves and 
+does not remain diagonal in general (though it is always triangular
+because it is displayed in row-echelon form):
+
+\smallskip
+
+\noindent
+\begin{tabular}{rl}
+\cIn  & \verb?u, v = x+y, x-y? \\
+      & \verb?R.precision().precision_lattice([u,v])? \\
+\cOut & \verb?[  32 2016]? \\
+      & \verb?[   0 2048]?
+\end{tabular}
+
+\smallskip
+
+\noindent
+The fact that the precision matrix is no longer diagonal indicates
+that some well-chosen linear combinations of $u$ and $v$ are known
+with more digits than $u$ and $v$ themselves. In this particular
+example, the sum $u+v$ is known at precision $O(2^{11})$ while the
+(optimal) precision on $u$ and $v$ separately is only $O(2^5)$.
+
+\smallskip
+
+\noindent
+\begin{tabular}{rl}
+\cIn  & \verb?u? \\
+\cOut & \verb?...10000? \\
+\cIn  & \verb?v? \\
+\cOut & \verb?...00110? \\
+\cIn  & \verb?u + v? \\
+\cOut & \verb?...11110110110? \\
+\end{tabular}
 
@@
 \noindent \textit{Diffused digits of precision.}
+%
+The phenomenon observed above is formalized by the notion of
+diffused digits of precision introduced in \cite{}. We recall
+briefly its definition.
+
+\begin{deftn}
+\label{def:diffused}
+Let $E$ be a $\Qp$-vector space endowed with a distinguished basis 
+$(e_1, \ldots, e_n)$ and write $\pi_i : E \to \Qp e_i$ for the 
+projections.
+Let $H \subset E$ be a lattice. The number of \emph{diffused digits of 
+precision} of $H$ is the length of $H_0/H$ where $H_0 = \pi_1(H) \oplus 
+\cdots \oplus \pi_n(H)$.
+\end{deftn}
+
+If $H$ represents the actual precision on some object, then
+$H_0$ is the smallest diagonal lattice containing $H$. It then
+corresponds to the maximal \emph{coordinate-wise} precision we can 
+reach on the set of $n$ variables corresponding to the basis $(e_1,
+\ldots, e_n)$.
+The diffused digits of precision only appear on linear combinations
+of those variables.
+
+The method \verb?number_of_diffused_digits? computes the number of
+diffused digits of precision on a set of variables. Observe:
+
+\smallskip
+
+\noindent
+\begin{tabular}{rl}
+\cIn  & \verb?R.precision().number_of_diffused_digits([x,y])? \\
+\cOut & \verb?0? \\
+\cIn  & \verb?R.precision().number_of_diffused_digits([u,v])? \\
+\cOut & \verb?6? \\
+\end{tabular}
+
+\smallskip
+
+\noindent
+For the last example, we recall that the relevant precision lattice 
+$H$ is generated by the $2 \times 2$ matrix:
+$$\left(\begin{matrix} 2^5 & 2016 \\ 0 & 2^{11} \end{matrix}\right).$$
+The minimal diagonal suplattice $H_0$ is generated by the scalar matrix 
+$2^5 \cdot \text{I}_2$ and contains $H$ with index $2^6$ in it. This is 
+where the six digits of precision comes from.
+There are easily visible here: the sum $u+v$ is known with $11$ digits,
+that is exactly $6$ digits more than the summands $u$ and $v$.
 

### 2018-01-29 ZpL/ZpL.tex
Done part of the examples

--- 2018-01-29 ZpL/ZpL.tex
@@
 
-\begin{itemize}
-\item determinant
-\item characteristic polynomial
-\end{itemize}
+In linear algebra, determinant and 
+characteristic polynomial are notoriously
+hard to compute: see \cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We illustrate this with the following
+example (using a division-free algorithm
+for the computation of the characteristic polynomial) :
+
+\bigskip
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?Q2 = Zp(2, prec=10, print_mode="digits")? \\
+ & \verb?MS = MatrixSpace(Q2,3,3)? \\
+ & \verb?Mat = MS.random_element()? \\
+ & \verb?chi = Mat.charpoly()? \\
+ & \verb?deto = Mat.determinant()? \\
+ & \verb?deto? \\
+\cZpCR
+ & \verb?...000100000100? \\
+\cZpFP
+ & \verb? ...11010101010? \\
+\cZpLC
+ & \verb?  ...0110110001? \\
+\cZpLF
+ & \verb?...001110111100? \\
+\end{tabular}
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?chi? \\
+\cZpCR
+ & \verb?(...0000000001)*x^3 + (...1001111111)*x^2 ?\\
+ & \verb?+ (...01110101110)*x + (...111011111100)? \\
+\cZpFP
+ & \verb?...0000000001*x^3 + ...1101001001*x^2 + ?\\
+ & \verb?...0100110011*x + ...00101010110? \\
+\cZpLC
+ & \verb?...0000000001*x^3 + ...1011000111*x^2 + ?\\
+ & \verb?...0011101000*x + ...1001001111? \\
+\cZpLF
+ & \verb?...0000000001*x^3 + ...1111000101*x^2 + ?\\
+ & \verb?...0011111111*x + ...110001000100? \\
+\end{tabular}
 
@@
 
-\begin{itemize}
-\item gcd of polynomials
-\end{itemize}
+Our package can be applied to
+computation over $p$-adic polynomials.
+A natural example is that of the computation of GCD,
+whose stability has been studied in \cite{caruso-2017}.
+
+A naive implementation of the Euclide algorithm
+can produce different behaviour depending 
+on the type of implementation of the field
+of $p$-adic coefficients.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?Z2 = Zp(2, prec=15, print_mode="digits")? \\
+  & \verb?S.<x> = PolynomialRing(Z2)? \\
+ & \verb?P = random_element(S, degree=10, prec=5)? \\
+ & \verb?Q = random_element(S, degree=10, prec=5)? \\
+ & \verb?D = x^5 + random_element(S, degree=4, prec=8)? \\
+ & \verb?def euclidean(A,B):? \\
+ & \verb?    while B != 0:? \\
+ & \verb?        A, B = B, A % B? \\
+ & \verb?    return A.monic()? \\
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{rl}
+\cIn
+  & \verb?euclidean(D*P, D*Q)? \\
+\cZpCR
+ & \verb?(...1)*x^9 + (0)*x^8 + (...,.1)*x^7 ?\\
+ & \verb?+ (...,.1)*x^6 + (...1)*x^5 + (...1)*x^4?\\
+ & \verb? + (...,.1)*x^3 + (0)*x^2 + (...,.1)*x + (...10)? \\
+\cZpFP
+ & \verb?...000000000000000000000000000001? \\
+\cZpLC
+ & \verb?...1*x^5+...100*x^3+...100*x^2+...1*x+...1? \\
+\end{tabular}
+
+With high probability, $P$ and $Q$ are coprime, implying that the gcd of $DP$ is $DQ$ is $D$.  
+However, we observe that we do not always get this expected result.
+More precisely, ZpCR often founds a gcd of higher degree.
+Indeed the Euclidean algorithm stops prematurely because the software believes that some early remainder vanishes due to the lack of precision.  
+The situation with ZpFP is the exact opposite:
+since the precision is too high (always the maximal cap), 
+the software sometimes does not notice that the remainder vanishes.
+Hence, it happens that the Euclidean algorithm 
+obtains a last non-zero remainder of degree less than that of
+$D.$
 
@@
 
-\subsubsection{$p$-adic differential equations}
+\subsection{Gröbner bases}
+Our package can be applied 
+on complex computations like that of
+Gröbner bases using generic
+Gröbner bases algorithms.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?Q2 = Qp(2, prec=50, print_mode="digits")? \\
+ & \verb?R.<x,y,z> = PolynomialRing(Q2, order = 'invlex')? \\
+ & \verb?F = [Q2(2,10)*x + Q2(1,10)*z, ? \\
+ & \verb?     Q2(1,10)*x^2 + Q2(1,10)*y^2 - Q2(2,10)*z^2,? \\
+ & \verb?     Q2(4,10)*y^2 + Q2(1,10)*y*z + Q2(8,10)*z^2]? \\
+ & \verb?from sage.rings.polynomial.toy_buchberger?\\
+ &\verb?  import buchberger_improved? \\
+ & \verb?g = buchberger_improved(ideal(F)); g.sort()? \\
+ & \verb?g? \\
+\cZpCR
+ & \verb?[x^3, x*y + ...1100010*x^2, y^2 + ...11001*x^2,? \\
+ & \verb? z + ...0000000010*x]? \\
+\cZpLC
+ & \verb?[[x^3, x*y + ...111100010*x^2, y^2 + ...1111111001*x^2,? \\
+ & \verb? z + ...0000000010*x]]? \\
+\end{tabular}
+
+As we can see, some loss in precision occurs in the
+Buchberger algorithm.
+Had we used \cZpFP , many padded zeroes or one would have appeared.
+
+\subsection{$p$-adic differential equations}
 

### 2018-01-30 ZpL/ZpL.tex
Small improvement in one tikz picture
Most examples done

@@
 \usepackage{tikz}
+\usetikzlibrary{patterns}
 
@@
 \newcommand{\sage}{\textsc{SageMath}\xspace}
+\newcommand{\ZpCR}{\text{\rm \tt ZpCR}\xspace}
+\newcommand{\ZpFP}{\text{\rm \tt ZpFP}\xspace}
 \newcommand{\ZpL}{\text{\rm \tt ZpL}\xspace}
@@
 
-\begin{tabular}{|l|l|}
-\verb?R = Zp(3,5)? & \verb?R = ZpL(3,5)? \\
-\verb?x = R(143,5); x? & \verb?x = R(143,5); x? \\
-\hfill\verb?...12022? & \hfill\verb?...12022? \\
-\verb?3*x? & \verb?3*x? \\
-\hfill\verb?...120220? & \hfill\verb?...120220? \\
-\verb?x + x + x? & \verb?x + x + x? \\
-\hfill\verb?...20220? & \hfill\verb?...120220? \\
-\verb?x^3? & \verb?x^3? \\
-\hfill\verb?...020222? & \hfill\verb?...020222? \\
-\verb?x * x * x? & \verb?x * x * x? \\
-\hfill\verb?...20222? & \hfill\verb?...020222? \\
-\end{tabular}
-
-\begin{tabular}{|l|l|}
-\verb?R = Zp(2,10)? & \verb?R = ZpL(2,10)? \\
-\verb?x = R(987,10); x? & \verb?x = R(987,10); x? \\
-\hfill\verb?...1111011011? & \hfill\verb?...1111011011? \\
-\verb?y = R(21,5); y? & \verb?y = R(21,5); y? \\
-\hfill\verb?...10101? & \hfill\verb?...10101? \\
-\verb?u = x + y; u? & \verb?u = x + y; u? \\
-\hfill\verb?...10000? & \hfill\verb?...10000? \\
-\verb?v = x - y; v? & \verb?v = x - y; v? \\
-\hfill\verb?...00110? & \hfill\verb?...00110? \\
-\verb?u + v? & \verb?u + v? \\
-\hfill\verb?...10110? & \hfill\verb?...11110110110? \\
-\end{tabular}
+We begin our tour of the features of the \ZpL package
+with some basic arithmetic computations.
+The first step is the definition of the $p$-adic ring or field.
+\begin{tabular}{rl}
+\cIn
+ & \verb?Z3 = Zp(3, prec=20, print_mode="digits")? \\
+\end{tabular}
+It means that we will work in various implementation of the 
+ring of $3$-adic integers with $20$ digits
+of precision. 
+We show the results for \ZpCR, \ZpFP and \ZpLC.
+Results for \ZpLF are only shown when they differ from \ZpLC.
+We first take some random element $x$.
+\begin{tabular}{rl}
+\cIn
+ & \verb?x = random_element(Z3,5)? \\
+ & \verb?x? \\
+\cZpCR
+ & \verb?...11111? \\
+\cZpFP
+ & \verb?...00000000000000011111? \\
+\cZpLC
+ & \verb?...11111? \\
+\end{tabular}
+Multiplication by $3$ is a shift on the digits, whether it 
+is done by addition or multiplication.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?3*x? \\
+\cZpCR
+ & \verb?...111110? \\
+\cZpFP
+ & \verb?...000000000000000111110? \\
+\cZpLC
+ & \verb?...111110? \\
+\cIn
+ & \verb?x + x + x? \\
+\cZpCR
+ & \verb?...11110? \\
+\cZpFP
+ & \verb?...000000000000000111110? \\
+\cZpLC
+ & \verb?...111110? \\
+\end{tabular}
+
+Something is lost however in \ZpCR when performing
+$x+2x.$
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?y = 2*x? \\
+ & \verb?y? \\
+\cZpCR
+ & \verb?...22222? \\
+\cZpFP
+ & \verb?...00000000000000022222? \\
+\cZpLC
+ & \verb?...22222? \\
+\cIn
+ & \verb?x + y? \\
+\cZpCR
+ & \verb?...11110? \\
+\cZpFP
+ & \verb?...000000000000000111110? \\
+\cZpLC
+ & \verb?...111110? \\
+\end{tabular}
+
+The same phenomenon occurs when comparing the computation
+of $x^3$ and $x \times x^2.$
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?x^3? \\
+\cZpCR
+ & \verb?...010101? \\
+\cZpFP
+ & \verb?...00000010100000010101? \\
+\cZpLC
+ & \verb?...010101? \\
+\cIn
+ & \verb?y = x^2? \\
+ & \verb?x * y? \\
+\cZpCR
+ & \verb?...10101? \\
+\cZpFP
+ & \verb?...00000010100000010101? \\
+\cZpLC
+ & \verb?...010101? \\
+
+\end{tabular}
+
+\ZpL is also well suited for working with coefficients 
+with unbalanced precision.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?x = random_element(Z2,10)? \\
+ & \verb?y = random_element(Z2,5)? \\
+\cIn
+ & \verb?u = x+y? \\
+ & \verb?u? \\
+\cZpCR
+ & \verb?...10111? \\
+\cZpFP
+ & \verb?...000000000000000000000011010111? \\
+\cZpLC
+ & \verb?...10111? \\
+\cIn
+ & \verb?v = x-y? \\
+ & \verb?v? \\
+\cZpCR
+ & \verb?...01111? \\
+\cZpFP
+ & \verb?...000000000000000000000011001111? \\
+\cZpLC
+ & \verb?...01111? \\
+\end{tabular}
+
+Now, let us compute $u+v=2 \times x.$
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?u+v? \\
+\cZpCR
+ & \verb?...00110? \\
+\cZpFP
+ & \verb?...0000000000000000000000110100110? \\
+\cZpLC
+ & \verb?...00110100110? \\
+\cIn
+ & \verb?2*x? \\
+\cZpCR
+ & \verb?...00110100110? \\
+\cZpFP
+ & \verb?...0000000000000000000000110100110? \\
+\cZpLC
+ & \verb?...00110100110? \\
+\end{tabular}
+
+No loss in precision appears in \ZpLC, contrary to \ZpCR.
+We remark that \ZpFP also provides all the correct digits.
 
@@
 for the computation of the characteristic polynomial) :
-
-\bigskip
 
@@
 \end{tabular}
-
-\bigskip
 
@@
 \draw[thin] (7,-7)--(7,0)--(0,0);
-\fill (3,0) rectangle (4,-4);
+\begin{scope}
+\clip (3,0.2) rectangle (4,-4.2);
+\draw[pattern=north east lines] (2.5,0.5) rectangle (4.5,-4.5);
+\end{scope}
 \draw[->] (2.9,-4.8)--(3.2,-4.2);
@@
 
-\begin{tabular}{|l|l|}
-\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
-\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
-\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
-\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
-\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
-\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
-\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
-\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
-\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
-\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
-\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
-\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
-\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
-\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
-\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
-\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
-\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
-\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
-\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
-\hfill\verb?...11? & \hfill\verb?...0000000111? \\
-\hfill\verb?...10? & \hfill\verb?...1010001110? \\
-\hfill\verb?...01? & \hfill\verb?...1110110001? \\
-\hfill\verb?...01? & \hfill\verb?...011110111001? \\
-\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
-\hfill\verb?...1? & \hfill\verb?...11111000011? \\
-\end{tabular}
-
-\begin{itemize}
-\item SOMOS sequence
-\item many multiplications of matrices
-\item Dodgson algorithm
-\item subresultants
-\item Grobner basis
-\item $p$-adic differential equations
-\end{itemize}
+\subsection{SOMOS-4}
+
+The SOMOS-4 sequence is a well-known example
+of a computation with a large gap between
+the theoretical loss in precision
+and that of a naive computation (see \cite{caruso-roe-vaccon:14a}).
+We show here that precision can be saved
+even when using a generic implementation
+of the SOMOS iteration.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?def somos4(u0, u1, u2, u3, n):? \\
+ & \verb?    a, b, c, d = u0, u1, u2, u3? \\
+ & \verb?    for _ in range(4, n+1):? \\
+ & \verb?        a, b, c, d = b, c, d, (b*d + c*c) / a? \\
+ & \verb?    return d? \\
+\end{tabular}
+\begin{tabular}{rl}
+\cIn
+ & \verb?somos4(Z2(1,15), Z2(1,15), Z2(1,15), Z2(3,15), 18)? \\
+\cZpCR
+ & \verb?...11? \\
+\cZpFP
+ & \verb?...110111001000100100000000000111? \\
+\cZpLC
+ & \verb?...100000000000111? \\
+\cIn
+ & \verb?somos4(Z2(1,15), Z2(1,15), Z2(1,15), Z2(3,15), 100)? \\
+\cZpCR
+ & \verb?PrecisionError: cannot divide by something?\\
+ &\verb?indistinguishable from zero.? \\
+\cZpFP
+ & \verb?...110000101000011001001001110001? \\
+\cZpLC
+ & \verb?...001001001110001? \\
+\end{tabular}
+%
+%\begin{tabular}{|l|l|}
+%\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
+%\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
+%\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
+%\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
+%\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
+%\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
+%\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
+%\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
+%\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
+%\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
+%\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
+%\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
+%\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
+%\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
+%\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
+%\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
+%\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
+%\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
+%\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
+%\hfill\verb?...11? & \hfill\verb?...0000000111? \\
+%\hfill\verb?...10? & \hfill\verb?...1010001110? \\
+%\hfill\verb?...01? & \hfill\verb?...1110110001? \\
+%\hfill\verb?...01? & \hfill\verb?...011110111001? \\
+%\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
+%\hfill\verb?...1? & \hfill\verb?...11111000011? \\
+%\end{tabular}
+%
+%\begin{itemize}
+%\item SOMOS sequence
+%\item many multiplications of matrices
+%\item Dodgson algorithm
+%\item subresultants
+%\item Grobner basis
+%\item $p$-adic differential equations
+%\end{itemize}
+
+\subsection{Matrix multiplication}
+The multiplication of many matrices has been studied
+in \citep{caruso-roe-vaccon:15} and is known to often 
+provide an heavy diffused digits phenomenon.
+Here is a sample example. the result for \ZpFP has not been
+added, but its first digits coincide with that of \ZpLC.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?Z2 = Zp(2, prec=100)? \\
+ & \verb?MS2 = MatrixSpace(Z2,2,2)? \\
+ & \verb?Mat = random_element(MS2,prec=5)? \\
+ & \verb?for _ in range(100):? \\
+ & \verb?    Mat = Mat*random_element(MS2, prec=5)? \\
+ & \verb?Mat? \\
+\cZpCR
+ & \verb?[ O(2^9) O(2^11)]? \\
+ & \verb?[ O(2^9) O(2^11)]? \\
+\cZpLC
+ & \verb?[2^9 + 2^10 + 2^11 + O(2^12)  O(2^14)]? \\
+ & \verb?[       2^9 + 2^11 + O(2^12)  O(2^14)]? \\
+\end{tabular}
+
+\ZpL can count the number of diffusede digits.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?Z2.precision().number_of_diffused_digits()? \\
+\cZpLC
+ & \verb?604? \\
+\end{tabular}
 
@@
 \cZpLC
- & \verb?[[x^3, x*y + ...111100010*x^2, y^2 + ...1111111001*x^2,? \\
- & \verb? z + ...0000000010*x]]? \\
+ & \verb?[[x^3, x*y + ...111100010*x^2,? \\
+ & \verb?y^2 + ...1111111001*x^2, z + ...0000000010*x]]? \\
 \end{tabular}
@@
 As we can see, some loss in precision occurs in the
-Buchberger algorithm.
+Buchberger algorithm and is avoided thanks to \ZpL.
 Had we used \cZpFP , many padded zeroes or one would have appeared.
@@
 
-\begin{tabular}{|l|l|}
-\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
-\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
-\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
-\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
-\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
-\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
-\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
-\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
-\hfill\verb?9? & \hfill\verb?4? \\
-\end{tabular}
-The last line provides the loss in precision between the
-original solution of the differential equation and its approximation
-obtained through computation.
-The theorical loss in precision for the coefficient of $t^{31}$
-is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
-loss in precision for a naive implementation is clearly more severe.
+\begin{tabular}{rl}
+\cIn
+ & \verb?N = 4? \\
+ & \verb?Z2 = Zp(2, prec=50, print_mode="digits")? \\
+ & \verb?Q2=Z2.fraction_field()? \\
+ & \verb?S.<t> = PowerSeriesRing(Q2, 2^N)? \\
+ & \verb?h = 1 + t + t^3? \\
+ & \verb?y = t + t^2 * random_element(S,10)? \\
+ & \verb?g = y.derivative() / h(y)? \\
+ & \verb?u = Newton_Iteration_Solver(g, h, N)? \\
+ & \verb?u[15]? \\
+\cZpCR
+ & \verb?...1101? \\
+\cZpLC
+ & \verb?...11011101? \\
+\end{tabular}
+
+%\begin{tabular}{|l|l|}
+%\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
+%\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
+%\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
+%\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
+%\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
+%\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
+%\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
+%\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
+%\hfill\verb?9? & \hfill\verb?4? \\
+%\end{tabular}
+%The last line provides the loss in precision between the
+%original solution of the differential equation and its approximation
+%obtained through computation.
+%The theorical loss in precision for the coefficient of $t^{31}$
+%is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
+%loss in precision for a naive implementation is clearly more severe.
 
@@
 over $\mathbb{Q}_3$ with lattice precision.
-We use an initial precision of 50 digits.
-Any generic implementation of the Newton
-method of \cite{Lercier-Sirvent:08} is enough.
-
-\begin{tabular}{|l|}
-\verb?    R = QpL(3,50)? \\
-\verb?    A.<x> = PolynomialRing(R)?\\
-\verb?    m=67?\\
-\verb?    g=1+(1/4)*x^2+x^6?\\
-\verb?    h=1+((m^2)/4)*x^2+m^6* x^6?\\
-\verb?    y=Lercier_Sirvent(g,h,64)?\\
-\verb?    y[31]?\\
-\verb?(467995903278061284942799/3^14 + O(3^36))*x^31?\\
+We use an initial precision of 50 digits
+and some generic implementation of the Newton
+method of \cite{Lercier-Sirvent:08}.
+
+\begin{tabular}{rl}
+\cIn
+ & \verb?R = Qp(3,50)? \\
+ & \verb?A.<x> = PolynomialRing(R)? \\
+ & \verb?m=67? \\
+ & \verb?g=1+(1/4)*x^2+x^6? \\
+ & \verb?h=1+((m^2)/4)*x^2+m^6* x^6? \\
+ & \verb?y=Lercier_Sirvent(g,h,6)? \\
+ & \verb?y[31]? \\
+ & \verb?u = Newton_Iteration_Solver(g, h, N)? \\
+ & \verb?u[15]? \\
+\cZpCR
+ & \verb?(467995903278061284942799/3^14 + O(3^36))*x^31? \\
+\cZpLC
+ & \verb?(467995903278061284942799/3^14 + O(3^36))*x^31? \\
 \end{tabular}
@@
 as for the coefficient in $x^{31}$: \textbf{no loss
-occured}.
-Thus, those simple manipulations
+occured}. This happens for \ZpCR as for \ZpLC,
+and thus, \ZpCR seems to attain the optimal precision.
+consequently, those simple manipulations
 hint at a different behaviour for the loss 

### 2018-01-30 ZpL/ZpL.tex
Complexity of ZpL

--- 2018-01-29 ZpL/ZpL.tex
@@
 \newcommand{\ZpLF}{\text{\rm \tt ZpLF}\xspace}
+\newcommand{\ZpFP}{\text{\rm \tt ZpFP}\xspace}
 
@@
 \newcommand{\coind}{\text{coind}}
+\newcommand{\hgt}{\text{hgt}}
+\newcommand{\cohgt}{\text{cohgt}}
+\renewcommand{\in}{\text{in}}
+\newcommand{\out}{\text{out}}
 
@@
 \subsection{Tracking precision}
+\label{ssec:trackprec}
 
@@
 As for \ZpLC, the matrix operation implied by the deletion of the 
-variable \ttw is modeled by the deletion of the corresponding column 
-of $M_t$. If \ttw is not a pivot variable at time $t$, the matrix 
-$M_t$ keeps the form \eqref{eq:shapeM-ZpLF} after erasure; therefore
-no more treatment is needed in this case.
+variable \ttw is the deletion of the corresponding column of $M_t$. If 
+\ttw is not a pivot variable at time $t$, the matrix $M_t$ keeps the 
+form \eqref{eq:shapeM-ZpLF} after erasure; therefore no more treatment 
+is needed in this case.
 
@@
 $2^5 \cdot \text{I}_2$ and contains $H$ with index $2^6$ in it. This is 
-where the six digits of precision comes from.
+where the $6$ digits of precision comes from.
 There are easily visible here: the sum $u+v$ is known with $11$ digits,
-that is exactly $6$ digits more than the summands $u$ and $v$.
+that is exactly $6$ more digits than the summands $u$ and $v$.
 
@@
 before~\ttw. By construction, the columns of the matrix $M_t$ are 
-ordered with respect to $<_t$.
+ordered with respect to $<_t$. We denote by $r_t$ (resp. $c_t$) the
+number of rows (resp. of columns) of $M_t$. By construction $r_t$ is
+also the cardinality of $\calV_t$. We have $c_t \leq r_t$ and the 
+equality always holds in the \ZpLC case.
+
 For $\ttv \in \calV_t$, we define the \emph{index} of \ttv, denoted by 
@@
 order, \ttv then appears in $\ind_t(\ttv)$-th position.
-We also define the \emph{coindex} of \ttv by
-$\coind_t(\ttv) = \card \calV_t - \ind_t(\ttv)$.
+We also define the \emph{co-index} of \ttv by
+$\coind_t(\ttv) = r_t - \ind_t(\ttv)$.
+
+Similarly, for any variable $\ttv \in \calV_t$, we define the 
+\emph{height} (resp. the \emph{co-height}) of \ttv at time $t$ as the 
+number of pivot variables \ttw such that $\ttw \leq_t \ttv$ (resp. such 
+that $\ttw >_t \ttv$). We denote it by $\hgt_t(\ttv)$ (resp. by 
+$\cohgt_t(\ttv)$). Clearly $\hgt_t(\ttv) + \cohgt_t(\ttv) = c_t$.
+The height of \ttv is the height of the significant 
+part of the column of $M_t$ which corresponds to \ttv. In the case of
+\ZpLC, all variables are pivot variables and thus $\hgt_t(\ttv) = 
+\ind_t(\ttv)$ and $\cohgt_t(\ttv) = \coind_t(\ttv)$ for all \ttv.
 
@@
 \noindent \textbf{Creating a variable.}
-
-\ZpLC: $O\bigg(\sum_{i=1}^n \ind_t(\ttv_i)\bigg) \subset 
-O\big(n \cdot \card \calV_t\big)$
-
-\ZpLF:
+%
+With the notations of \S \ref{ssec:trackprec}, it is obvious that 
+creating a new variable \ttw requires:
+$$O\bigg(\sum_{i=1}^n \hgt_i(\ttv_i)\bigg) \subset O(n \: c_t)$$ 
+operations in $\Qp$. Here, we recall that $n$ is the arity of the 
+operation defining \ttw. In most cases it is $2$; thus the above
+complexity reduces to $O(c_t)$.
+
+In the \ZpLF context, $c_t$ counts the number of user-defined variables. 
+It is then expected to be constant (roughly equal to the size of the 
+input) while running a given algorithm.
+
+On the contrary, in the \ZpLC context, $c_t$ counts the number of
+variables which are alive at time $t$. It is no longer expected to
+be constant but evolves continuously when the algorithm runs.
+%
+\begin{figure}
+% Characteristic polynomial
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\noindent\hfill%
+\begin{tabular}{|l|r|r|r|r|r|}
+\hline
+\textbf{Dimension} &
+ 2 &   5 &   10 &    20 &      50 \\
+\hline
+\textbf{Total} &
+35 & 424 & 5\:539 & 83\:369 & 3\:170\:657 \\
+\hline
+\textbf{Simult.} &
+17 &  65 &  225 &   845 &    5\:101 \\
+\hline
+\end{tabular}%
+\hfill\null
+
+\vspace{1mm}
+
+\noindent\hfill%
+{\small Computation of characteristic polynomial}%
+\hfill\null
+
+\bigskip
+
+% Euclidean algorithm
+
+\noindent\hfill%
+\begin{tabular}{|l|r|r|r|r|r|r|}
+\hline
+\textbf{Degree} &
+ 2 &   5 &   10 &   20 &   50 &   100 \\
+\hline
+\textbf{Total} &
+54 & 130 &  332 & 1\:036 & 4\:110 & 10\:578 \\ 
+\hline
+\textbf{Simult.} &
+18 &  31 &   56 &  106 &  256 &   507 \\
+\hline
+\end{tabular}%
+\hfill\null
+
+\vspace{1mm}
+
+\noindent\hfill%
+{\small Naive Euclidean algorithm}%
+\hfill\null
+
+\caption{Numbers of involved variables}
+\label{fig:variables}
+\end{figure}
+%
+The tables of Figure~\ref{fig:variables} show the total number of 
+created variables (which reflects the complexity) together with the 
+maximum number of variables involved at the same time (which reflects 
+the memory occupation) while executing two basic computations.
+The first one is the computation of the characteristic polynomial of a 
+square matrix by the default algorithm used by \sage for $p$-adic fields 
+(which is a division-free algorithm of quartic complexity) while the
+second one is the computation of the gdc of two polynomials using a
+naive Euclidean algorithm (of quadratic complexity). We can observe 
+that, for both of them, the memory occupation is roughly equal to the 
+square root of the complexity.
 
@@
 
-\noindent \textit{Deleting a variable.}
-
-$$O\big(\coind_t(\ttw)^2\big)$$
+\noindent \textbf{Deleting a variable.}
+%
+The deletion of the variable \ttw induces the deletion of the 
+corresponding column of $M_t$, possibly followed by a partial 
+row-echelonization. In terms of algebraic complexity, the deletion is 
+free. As for the cost of the echelonization, it stays within:
+$$O\big(\coind_t(\ttw) \cdot \cohgt_t(\ttw) \big)$$
+operations in $\Qp$.
+
+In the \ZpLF case, we expect that, most of the time, the deleted
+variables were created after all initial variables were set by the
+user. This means that we expect $\cohgt_t(\ttw)$ to vanish and so,
+the corresponding cost to be negligible.
+
+In the \ZpLC case, we always have $\cohgt_t(\ttw) = \coind_t(\ttw)$, so 
+that the cost becomes $O\big(\coind_t(\ttw)^2\big)$. This does not look 
+nice \emph{a priori}. However, the principle of temporal locality 
+asserts that $\cohgt_t(\ttw)$ tends to be small in general: destroyed 
+variables are often variables that were created recently. As a basic 
+exemple, variables which are local to a small piece of code (\emph{e.g.} 
+a short function or a loop) do not survive for a long time. It turns
+out that this behaviour is typical in many implementations!
+%
+\begin{figure}
+\noindent\hfill%
+\begin{tikzpicture}[xscale=0.35,yscale=0.08]
+\draw[fill=blue] (-0.45,0) rectangle (0.45,46);
+\draw[fill=blue] (0.65,0) rectangle (1.45,69);
+\draw[fill=blue] (1.65,0) rectangle (2.45,5);
+\draw[fill=blue] (2.65,0) rectangle (3.45,7);
+\draw[fill=blue] (3.65,0) rectangle (4.45,7);
+\draw[fill=blue] (4.65,0) rectangle (5.45,12);
+\draw[fill=blue] (5.65,0) rectangle (6.45,7);
+\draw[fill=blue] (6.65,0) rectangle (7.45,5);
+\draw[fill=blue] (7.65,0) rectangle (8.45,1);
+\draw[fill=blue] (8.65,0) rectangle (9.45,1);
+\draw (9.65,0)--(10.45,0);
+\draw[fill=blue] (10.65,0) rectangle (11.45,1);
+\draw[fill=blue] (11.65,0) rectangle (12.45,1);
+\draw[fill=blue] (12.65,0) rectangle (13.45,2);
+\draw[fill=blue] (13.65,0) rectangle (14.45,2);
+\draw[fill=blue] (14.65,0) rectangle (15.45,2);
+\draw[fill=blue] (15.65,0) rectangle (16.45,2);
+\draw[fill=blue] (16.65,0) rectangle (17.45,1);
+\draw[fill=blue] (17.65,0) rectangle (18.45,1);
+\draw[fill=blue] (18.65,0) rectangle (19.45,1);
+\node[scale=0.7] at (0,-2.5) { $0$ };
+\node[scale=0.7] at (5,-2.5) { $5$ };
+\node[scale=0.7] at (10,-2.5) { $10$ };
+\node[scale=0.7] at (15,-2.5) { $15$ };
+\draw (-2,0)--(-2,72);
+\draw (-2.2,0)--(-1.8,0);
+\node[scale=0.7,left] at (-2.1,0) { $0$ };
+\draw (-2.2,10)--(-1.8,10);
+\node[scale=0.7,left] at (-2.1,10) { $10$ };
+\draw (-2.2,20)--(-1.8,20);
+\node[scale=0.7,left] at (-2.1,20) { $20$ };
+\draw (-2.2,30)--(-1.8,30);
+\node[scale=0.7,left] at (-2.1,30) { $30$ };
+\draw (-2.2,40)--(-1.8,40);
+\node[scale=0.7,left] at (-2.1,40) { $40$ };
+\draw (-2.2,50)--(-1.8,50);
+\node[scale=0.7,left] at (-2.1,50) { $50$ };
+\draw (-2.2,60)--(-1.8,60);
+\node[scale=0.7,left] at (-2.1,60) { $60$ };
+\draw (-2.2,70)--(-1.8,70);
+\node[scale=0.7,left] at (-2.1,70) { $70$ };
+\end{tikzpicture}%
+\hfill\null
+
+\caption{The distribution of $\cohgt_t(\ttw)$}
+\label{fig:deletion}
+\end{figure}
+%
+The histogram of Figure~\ref{fig:deletion} shows 
+the distribution of $\cohgt_t(\ttw)$ while executing the Euclidean
+algorithm (naive implementation) with two polynomials of degree $7$
+as input. The bias is evident: most of the time $\cohgt_t(\ttw) \leq
+1$.
 
@@
 
-\noindent \textit{Temporal locality.}
-
-\medskip
-
-\noindent \textbf{Correctness.}
-
-\subsubsection*{Tracking precision}
-
-\subsubsection*{Complexity}
+\noindent \textbf{Summary: Impact on complexity.}
+% 
+We consider the case of an algorithm with the following characteristics: 
+its complexity is $c$ operations in $\Qp$ (without any tracking of 
+precision), its memory usage is $m$ elements of $\Qp$, its input and its 
+output have size $s_\in$ and $s_\out$ (elements of $\Qp$) respectively.
+
+In the case of \ZpLF, creating a variable has a cost $O(s_\in)$ 
+whereas deleting a variable is free. 
+Thus when executed with the \ZpLF mecanism, the complexity of our
+algorithm becomes $O(s_\in c)$.
+
+In the \ZpLC framework, creating a variable has a cost $O(m)$. The case 
+of deletion is more difficult to handle. However, by the temporal 
+locality principle, it seems safe to assume that it is not the 
+bottleneck (which is the case in practice). Therefore, when executed 
+with the \ZpLF mecanism, the cost of our algorithm is expected to be 
+roughly $O(mc)$. Going further in speculation, we might estimate the 
+magnitude of $m$ by $m \simeq O(s + \sqrt c)$ with $s = \max(s_\in, 
+s_\out)$, leading to a complexity of $O(c^{3/2} + sc)$. For 
+quasi-optimal algorithms, the term $sc \simeq c^2$ dominates. However, 
+as soon as the complexity is at least quadratic in $s$, the dominant 
+term is $c^{3/2}$ and the impact on the complexity is then diminished.
 
@@
 Buchberger algorithm.
-Had we used \cZpFP , many padded zeroes or one would have appeared.
+Had we used \ZpFP, many padded zeroes or one would have appeared.
 

### 2018-01-30 ZpL/ZpL.tex

--- 2018-01-30 ZpL/ZpL.tex
@@
 Buchberger algorithm and is avoided thanks to \ZpL.
-Had we used \cZpFP , many padded zeroes or one would have appeared.
+Had we used \ZpFP, many padded zeroes or one would have appeared.
 

### 2018-01-30 ZpL/ZpL.tex
Some words about the complexity model

@@
 \newcommand{\ZpLF}{\text{\rm \tt ZpLF}\xspace}
-\newcommand{\ZpFP}{\text{\rm \tt ZpFP}\xspace}
 
@@
 We now discuss the cost of the above operations.
-For this,
-it is convenient to introduce a total order on $\calV_t$: for $\ttv, 
+In what follows, we shall count operations in $\Qp$. Although $\Qp$
+is an inexact field, our model of complexity makes sense because the
+size of the $p$-adic numbers we manipulate will always be roughly the
+same size: for \ZpLF, it is the precision we use for Floating Point 
+Arithmetic while, for \ZpLC, it is bounded by the absolute cap which
+was fixed at the beginning.
+
+It is convenient to introduce a total order on $\calV_t$: for $\ttv, 
 \ttw \in \calV_t$, we convene that $\ttv <_t \ttw$ is \ttv was created 

### 2018-01-30 ZpL/ZpL.tex
redefine \in is not a good idea!

@@
 \newcommand{\cohgt}{\text{cohgt}}
-\renewcommand{\in}{\text{in}}
+\newcommand{\inp}{\text{in}}
 \newcommand{\out}{\text{out}}
@@
 precision), its memory usage is $m$ elements of $\Qp$, its input and its 
-output have size $s_\in$ and $s_\out$ (elements of $\Qp$) respectively.
-
-In the case of \ZpLF, creating a variable has a cost $O(s_\in)$ 
+output have size $s_\inp$ and $s_\out$ (elements of $\Qp$) respectively.
+
+In the case of \ZpLF, creating a variable has a cost $O(s_\inp)$ 
 whereas deleting a variable is free. 
 Thus when executed with the \ZpLF mecanism, the complexity of our
-algorithm becomes $O(s_\in c)$.
+algorithm becomes $O(s_\inp c)$.
 
@@
 roughly $O(mc)$. Going further in speculation, we might estimate the 
-magnitude of $m$ by $m \simeq O(s + \sqrt c)$ with $s = \max(s_\in, 
+magnitude of $m$ by $m \simeq O(s + \sqrt c)$ with $s = \max(s_\inp, 
 s_\out)$, leading to a complexity of $O(c^{3/2} + sc)$. For 

### 2018-01-31 ZpL/ZpL.tex
Examples shortened: small font + ZpFP removed

@@
 
-\subsection{Elementary arithmetic}
-
 We begin our tour of the features of the \ZpL package
@@
 The first step is the definition of the $p$-adic ring or field.
-\begin{tabular}{rl}
-\cIn
- & \verb?Z3 = Zp(3, prec=20, print_mode="digits")? \\
-\end{tabular}
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+ & \verb?Z2 = Zp(2, print_mode="digits")? \\
+ & \verb?Q2 = Qp(2, print_mode="digits")? \\
+\end{tabular}}
+
+\noindent
 It means that we will work in various implementation of the 
-ring of $3$-adic integers with $20$ digits
-of precision. 
-We show the results for \ZpCR, \ZpFP and \ZpLC.
+ring of $2$-adic integers with $20$ digits of precision (which
+is the default).
+The rings \verb?Z3? and \verb?Q3? are defined similarly.
+
+We show the results for \ZpCR%, \ZpFP 
+and \ZpLC.
 Results for \ZpLF are only shown when they differ from \ZpLC.
+
+\subsection{Elementary arithmetic}
+
 We first take some random element $x$.
+
+{\noindent \small
 \begin{tabular}{rl}
@@
  & \verb?...11111? \\
-\cZpFP
- & \verb?...00000000000000011111? \\
+%\cZpFP
+% & \verb?...00000000000000011111? \\
 \cZpLC
  & \verb?...11111? \\
-\end{tabular}
+\end{tabular}}
+
 Multiplication by $3$ is a shift on the digits, whether it 
@@
 
+{\noindent \small
 \begin{tabular}{rl}
@@
  & \verb?...111110? \\
-\cZpFP
- & \verb?...000000000000000111110? \\
+%\cZpFP
+% & \verb?...000000000000000111110? \\
 \cZpLC
@@
  & \verb?...11110? \\
-\cZpFP
- & \verb?...000000000000000111110? \\
+%\cZpFP
+% & \verb?...000000000000000111110? \\
 \cZpLC
  & \verb?...111110? \\
-\end{tabular}
-
-Something is lost however in \ZpCR when performing
-$x+2x.$
-
-\begin{tabular}{rl}
-\cIn
- & \verb?y = 2*x? \\
- & \verb?y? \\
+\end{tabular}}
+
+Something is lost however in \ZpCR when performing $x+2x$.
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+ & \verb?y = 2*x; y? \\
 \cZpCR
  & \verb?...22222? \\
-\cZpFP
- & \verb?...00000000000000022222? \\
+%\cZpFP
+% & \verb?...00000000000000022222? \\
 \cZpLC
@@
  & \verb?...11110? \\
-\cZpFP
- & \verb?...000000000000000111110? \\
+%\cZpFP
+% & \verb?...000000000000000111110? \\
 \cZpLC
  & \verb?...111110? \\
-\end{tabular}
+\end{tabular}}
 
 The same phenomenon occurs when comparing the computation
-of $x^3$ and $x \times x^2.$
-
+of $x^3$ and $x \times x^2$.
+
+{\noindent \small
 \begin{tabular}{rl}
@@
  & \verb?...010101? \\
-\cZpFP
- & \verb?...00000010100000010101? \\
+%\cZpFP
+% & \verb?...00000010100000010101? \\
 \cZpLC
@@
  & \verb?...10101? \\
-\cZpFP
- & \verb?...00000010100000010101? \\
+%\cZpFP
+% & \verb?...00000010100000010101? \\
 \cZpLC
  & \verb?...010101? \\
-
-\end{tabular}
+\end{tabular}}
 
@@
  & \verb?...10111? \\
-\cZpFP
- & \verb?...000000000000000000000011010111? \\
+%\cZpFP
+% & \verb?...000000000000000000000011010111? \\
 \cZpLC
@@
  & \verb?...01111? \\
-\cZpFP
- & \verb?...000000000000000000000011001111? \\
+%\cZpFP
+% & \verb?...000000000000000000000011001111? \\
 \cZpLC
  & \verb?...01111? \\
-\end{tabular}
-
-Now, let us compute $u+v=2 \times x.$
-
+\end{tabular}}
+
+Now, let us compute $u+v$ (and observe that it is actually equal to $2x$).
+
+{\noindent \small
 \begin{tabular}{rl}
@@
  & \verb?...00110? \\
-\cZpFP
- & \verb?...0000000000000000000000110100110? \\
+%\cZpFP
+% & \verb?...0000000000000000000000110100110? \\
 \cZpLC
@@
  & \verb?...00110100110? \\
-\cZpFP
- & \verb?...0000000000000000000000110100110? \\
+%\cZpFP
+% & \verb?...0000000000000000000000110100110? \\
 \cZpLC
  & \verb?...00110100110? \\
-\end{tabular}
+\end{tabular}}
 
 No loss in precision appears in \ZpLC, contrary to \ZpCR.
-We remark that \ZpFP also provides all the correct digits.
+%We remark that \ZpFP also provides all the correct digits.
 
@@
 
-\begin{tabular}{rl}
-\cIn
- & \verb?Q2 = Zp(2, prec=10, print_mode="digits")? \\
- & \verb?MS = MatrixSpace(Q2,3,3)? \\
- & \verb?Mat = MS.random_element()? \\
- & \verb?chi = Mat.charpoly()? \\
- & \verb?deto = Mat.determinant()? \\
- & \verb?deto? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+ & \verb?M = random_element(MatrixSpace(2,2), prec=10)? \\
+ & \verb?M.determinant()? \\
 \cZpCR
  & \verb?...000100000100? \\
-\cZpFP
- & \verb? ...11010101010? \\
+%\cZpFP
+% & \verb? ...11010101010? \\
 \cZpLC
@@
  & \verb?...001110111100? \\
-\end{tabular}
-
-\begin{tabular}{rl}
-\cIn
- & \verb?chi? \\
+\cIn
+ & \verb?M.charpoly()? \\
 \cZpCR
@@
  & \verb?+ (...01110101110)*x + (...111011111100)? \\
-\cZpFP
- & \verb?...0000000001*x^3 + ...1101001001*x^2 + ?\\
- & \verb?...0100110011*x + ...00101010110? \\
+%\cZpFP
+% & \verb?...0000000001*x^3 + ...1101001001*x^2 + ?\\
+% & \verb?...0100110011*x + ...00101010110? \\
 \cZpLC
@@
  & \verb?...0011111111*x + ...110001000100? \\
-\end{tabular}
+\end{tabular}}
 
@@
 
-Our package can be applied to
-computation over $p$-adic polynomials.
-A natural example is that of the computation of GCD,
-whose stability has been studied in \cite{caruso-2017}.
+Our package can be applied to computation over $p$-adic polynomials. A 
+natural example is that of the computation of GCD, whose stability has 
+been studied in \cite{caruso-2017}.
 
@@
 
-\begin{tabular}{rl}
-\cIn
- & \verb?Z2 = Zp(2, prec=15, print_mode="digits")? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
  & \verb?S.<x> = PolynomialRing(Z2)? \\
@@
  & \verb?D = x^5 + random_element(S, degree=4, prec=8)? \\
+\cIn
  & \verb?def euclidean(A,B):? \\
@@
  & \verb?    return A.monic()? \\
-\end{tabular}
-
-\begin{tabular}{rl}
 \cIn
@@
  & \verb? + (...,.1)*x^3 + (0)*x^2 + (...,.1)*x + (...10)? \\
-\cZpFP
- & \verb?...000000000000000000000000000001? \\
+%\cZpFP
+% & \verb?...000000000000000000000000000001? \\
 \cZpLC
  & \verb?...1*x^5+...100*x^3+...100*x^2+...1*x+...1? \\
-\end{tabular}
+\end{tabular}}
 
@@
  & \verb?    return d? \\
-\end{tabular}
-\begin{tabular}{rl}
-\cIn
- & \verb?somos4(Z2(1,15), Z2(1,15), Z2(1,15), Z2(3,15), 18)? \\
+\cIn
+ & \verb?u0 = u1 = u2 = Z2(1,15); u3 = Z2(3,15)? \\
+ & \verb?somos4(u0, u1, u2, u3, 18)? \\
 \cZpCR
  & \verb?...11? \\
-\cZpFP
- & \verb?...110111001000100100000000000111? \\
+%\cZpFP
+% & \verb?...110111001000100100000000000111? \\
 \cZpLC
@@
 \cIn
- & \verb?somos4(Z2(1,15), Z2(1,15), Z2(1,15), Z2(3,15), 100)? \\
+ & \verb?somos4(u0, u1, u2, u3, 100)? \\
 \cZpCR
@@
  &\verb?indistinguishable from zero.? \\
-\cZpFP
- & \verb?...110000101000011001001001110001? \\
+%\cZpFP
+% & \verb?...110000101000011001001001110001? \\
 \cZpLC
  & \verb?...001001001110001? \\
-\end{tabular}
+\end{tabular}}
+
 %
@@
 
-\begin{tabular}{rl}
-\cIn
- & \verb?Z2 = Zp(2, prec=100)? \\
- & \verb?MS2 = MatrixSpace(Z2,2,2)? \\
- & \verb?Mat = random_element(MS2,prec=5)? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+% & \verb?Z2 = Zp(2, prec=100)? \\
+ & \verb?MS = MatrixSpace(Z2,2)? \\
+ & \verb?M = random_element(MS, prec=5)? \\
  & \verb?for _ in range(100):? \\
- & \verb?    Mat = Mat*random_element(MS2, prec=5)? \\
- & \verb?Mat? \\
+ & \verb?    M *= random_element(MS, prec=5)? \\
+ & \verb?M? \\
 \cZpCR
@@
  & \verb?[       2^9 + 2^11 + O(2^12)  O(2^14)]? \\
-\end{tabular}
-
-\ZpL can count the number of diffusede digits.
-
-\begin{tabular}{rl}
-\cIn
- & \verb?Z2.precision().number_of_diffused_digits()? \\
-\cZpLC
- & \verb?604? \\
-\end{tabular}
+\end{tabular}}
+
+%\ZpL can count the number of diffused digits.
+%\begin{tabular}{rl}
+%\cIn
+% & \verb?Z2.precision().number_of_diffused_digits()? \\
+%\cZpLC
+% & \verb?604? \\
+%\end{tabular}
 
@@
 
-\begin{tabular}{rl}
-\cIn
- & \verb?Q2 = Qp(2, prec=50, print_mode="digits")? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+% & \verb?Q2 = Qp(2, prec=50, print_mode="digits")? \\
  & \verb?R.<x,y,z> = PolynomialRing(Q2, order = 'invlex')? \\
- & \verb?F = [Q2(2,10)*x + Q2(1,10)*z, ? \\
+ & \verb?F = [ Q2(2,10)*x + Q2(1,10)*z, ? \\
  & \verb?      Q2(1,10)*x^2 + Q2(1,10)*y^2 - Q2(2,10)*z^2,? \\
- & \verb?     Q2(4,10)*y^2 + Q2(1,10)*y*z + Q2(8,10)*z^2]? \\
- & \verb?from sage.rings.polynomial.toy_buchberger?\\
- &\verb?  import buchberger_improved? \\
+ & \verb?      Q2(4,10)*y^2 + Q2(1,10)*y*z + Q2(8,10)*z^2 ]? \\
+\cIn
+ & \verb?from sage.rings.polynomial.toy_buchberger\?\\
+ & \verb?    import buchberger_improved? \\
  & \verb?g = buchberger_improved(ideal(F)); g.sort()? \\
@@
 \cZpCR
- & \verb?[x^3, x*y + ...1100010*x^2, y^2 + ...11001*x^2,? \\
- & \verb? z + ...0000000010*x]? \\
-\cZpLC
- & \verb?[[x^3, x*y + ...111100010*x^2,? \\
- & \verb?y^2 + ...1111111001*x^2, z + ...0000000010*x]]? \\
-\end{tabular}
+ & \verb?[x^3, x*y + ...1100010*x^2,? \\
+ & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
+\cZpLC
+ & \verb?[x^3, x*y + ...111100010*x^2,? \\
+ & \verb? y^2 + ...1111111001*x^2, z + ...0000000010*x]? \\
+\end{tabular}}
 
@@
 Buchberger algorithm and is avoided thanks to \ZpL.
-Had we used \ZpFP, many padded zeroes or one would have appeared.
+%Had we used \ZpFP, many padded zeroes or one would have appeared.
 
@@
 
-\begin{tabular}{rl}
-\cIn
- & \verb?N = 4? \\
- & \verb?Z2 = Zp(2, prec=50, print_mode="digits")? \\
- & \verb?Q2=Z2.fraction_field()? \\
- & \verb?S.<t> = PowerSeriesRing(Q2, 2^N)? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+% & \verb?N = 4? \\
+% & \verb?Z2 = Zp(2, prec=50, print_mode="digits")? \\
+% & \verb?Q2=Z2.fraction_field()? \\
+ & \verb?S.<t> = PowerSeriesRing(Q2, 16)? \\
  & \verb?h = 1 + t + t^3? \\
@@
  & \verb?g = y.derivative() / h(y)? \\
- & \verb?u = Newton_Iteration_Solver(g, h, N)? \\
+ & \verb?u = Newton_Iteration_Solver(g, h, 4)? \\
  & \verb?u[15]? \\
@@
  & \verb?...11011101? \\
-\end{tabular}
+\end{tabular}}
 
@@
  & \verb?A.<x> = PolynomialRing(R)? \\
- & \verb?m=67? \\
- & \verb?g=1+(1/4)*x^2+x^6? \\
- & \verb?h=1+((m^2)/4)*x^2+m^6* x^6? \\
- & \verb?y=Lercier_Sirvent(g,h,6)? \\
+ & \verb?m = 67? \\
+ & \verb?g = 1 + (1/4)*x^2 + x^6? \\
+ & \verb?h = 1 + ((m^2)/4)*x^2 + m^6*x^6? \\
+ & \verb?y = Lercier_Sirvent(g, h, 6)? \\
  & \verb?y[31]? \\
- & \verb?u = Newton_Iteration_Solver(g, h, N)? \\
+ & \verb?u = Newton_Iteration_Solver(g, h, N)? \\  % ???
  & \verb?u[15]? \\
@@
  & \verb?(467995903278061284942799/3^14 + O(3^36))*x^31? \\
-\end{tabular}
+\end{tabular}}
 

### 2018-01-31 ZpL/ZpL.tex
Small modifications

@@
 
+\smallskip
+
 {\noindent \small
@@
 \end{tabular}}
+
+\smallskip
 
@@
 
+\smallskip
+
 Multiplication by $3$ is a shift on the digits, whether it 
 is done by addition or multiplication.
+
+\smallskip
 
@@
 
+\smallskip
+
 Something is lost however in \ZpCR when performing $x+2x$.
+
+\smallskip
 
@@
 
+\smallskip
+
 The same phenomenon occurs when comparing the computation
 of $x^3$ and $x \times x^2$.
+
+\smallskip
 
@@
 
+\smallskip
+
 \ZpL is also well suited for working with coefficients 
 with unbalanced precision.
+
+\smallskip
 
@@
 
+\smallskip
+
 Now, let us compute $u+v$ (and observe that it is actually equal to $2x$).
+
+\smallskip
 
@@
 
+\smallskip
+
 No loss in precision appears in \ZpLC, contrary to \ZpCR.
@@
 
-In linear algebra, determinant and 
-characteristic polynomial are notoriously
-hard to compute: see \cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We illustrate this with the following
-example (using a division-free algorithm
-for the computation of the characteristic polynomial) :
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?M = random_element(MatrixSpace(2,2), prec=10)? \\
- & \verb?M.determinant()? \\
-\cZpCR
- & \verb?...000100000100? \\
+In linear algebra, determinant and characteristic polynomial are 
+notoriously hard to compute: see \cite{caruso-roe-vaccon:15, 
+caruso-roe-vaccon:17}. We illustrate this with the following example 
+(using a division-free algorithm for the computation of the 
+characteristic polynomial) :
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+ & \verb?M = random_element(MatrixSpace(Q2,3), prec=10)? \\
+ & \verb?M.charpoly()? \\
+\cZpCR
+ & \verb?...00000000000000000001*x^3 + ? \\
+ & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
+\cZpLC
+ & \verb?...00000000000000000001*x^3 + ? \\
+ & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
+ & \verb?...010000010? \\
+\end{tabular}}
+
+%\begin{tabular}{rl}
+%\cIn
+% & \verb?M = random_element(MatrixSpace(Q2,3), prec=10)? \\
+% & \verb?M.determinant()? \\
+%\cZpCR
+% & \verb?...000100000100? \\
 %\cZpFP
 % & \verb? ...11010101010? \\
-\cZpLC
- & \verb?  ...0110110001? \\
-\cZpLF
- & \verb?...001110111100? \\
-\cIn
- & \verb?M.charpoly()? \\
-\cZpCR
- & \verb?(...0000000001)*x^3 + (...1001111111)*x^2 ?\\
- & \verb?+ (...01110101110)*x + (...111011111100)? \\
+%\cZpLC
+% & \verb?  ...0110110001? \\
+%\cZpLF
+% & \verb?...001110111100? \\
+%\cIn
+% & \verb?M.charpoly()? \\
+%\cZpCR
+% & \verb?(...0000000001)*x^3 + (...1001111111)*x^2 ?\\
+% & \verb?+ (...01110101110)*x + (...111011111100)? \\
 %\cZpFP
@@
 % & \verb?...0100110011*x + ...00101010110? \\
-\cZpLC
- & \verb?...0000000001*x^3 + ...1011000111*x^2 + ?\\
- & \verb?...0011101000*x + ...1001001111? \\
-\cZpLF
- & \verb?...0000000001*x^3 + ...1111000101*x^2 + ?\\
- & \verb?...0011111111*x + ...110001000100? \\
-\end{tabular}}
+%\cZpLC
+% & \verb?...0000000001*x^3 + ...1011000111*x^2 + ?\\
+% & \verb?...0011101000*x + ...1001001111? \\
+%\cZpLF
+% & \verb?...0000000001*x^3 + ...1111000101*x^2 + ?\\
+% & \verb?...0011111111*x + ...110001000100? \\
+%\end{tabular}
 
@@
  & \verb?Q = random_element(S, degree=10, prec=5)? \\
- & \verb?D = x^5 + random_element(S, degree=4, prec=8)? \\
+ & \verb?D = x^5 + random_element(S, degree=4, prec=8); D? \\
+\cZpCR
+ & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
+ & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
+ & \verb?...10000110*x + ...11100010? \\
+\cZpLC
+ & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
+ & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
+ & \verb?...10000110*x + ...11100010? \\
 \cIn
@@
  & \verb?    return A.monic()? \\
-\cIn
  & \verb?euclidean(D*P, D*Q)? \\
 \cZpCR
- & \verb?(...1)*x^9 + (0)*x^8 + (...,.1)*x^7 ?\\
- & \verb?+ (...,.1)*x^6 + (...1)*x^5 + (...1)*x^4?\\
- & \verb? + (...,.1)*x^3 + (0)*x^2 + (...,.1)*x + (...10)? \\
-%\cZpFP
-% & \verb?...000000000000000000000000000001? \\
-\cZpLC
- & \verb?...1*x^5+...100*x^3+...100*x^2+...1*x+...1? \\
-\end{tabular}}
-
-With high probability, $P$ and $Q$ are coprime, implying that the gcd of $DP$ is $DQ$ is $D$.  
-However, we observe that we do not always get this expected result.
-More precisely, ZpCR often founds a gcd of higher degree.
-Indeed the Euclidean algorithm stops prematurely because the software believes that some early remainder vanishes due to the lack of precision.  
-The situation with ZpFP is the exact opposite:
-since the precision is too high (always the maximal cap), 
-the software sometimes does not notice that the remainder vanishes.
-Hence, it happens that the Euclidean algorithm 
-obtains a last non-zero remainder of degree less than that of
-$D.$
+ & \verb?0*x^9 + ...1*x^8 + 0*x^7 + 0*x^6 + 0*x^5 +? \\
+ & \verb?0*x^4 + 0*x^3 + ...1*x^2 + ...10*x + ...10? \\
+\cZpLC
+ & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
+ & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
+ & \verb?...10000110*x + ...11100010? \\
+\end{tabular}}
+
+\smallskip
+
+\noindent
+With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
+$DP$ is $DQ$ is $D$.  However, we observe that we do not always get this 
+expected result. More precisely, \ZpCR often founds a gcd of higher 
+degree. Indeed the Euclidean algorithm stops prematurely because the 
+software believes that some early remainder vanishes due to the lack of 
+precision.
+%The situation with ZpFP is the exact opposite:
+%since the precision is too high (always the maximal cap), 
+%the software sometimes does not notice that the remainder vanishes.
+%Hence, it happens that the Euclidean algorithm 
+%obtains a last non-zero remainder of degree less than that of
+%$D.$
 
@@
 
-In this Section, we explain how our package \ZpL works and analyze
+In this Section, we explain how our package \ZpL works and we analyze
 its performances.
@@
 This Section is organized as follows.
-\todo{[...]}.
+\S \ref{ssec:preclemma} contains a brief overview on the theory of
+$p$-adic precision of \cite{padicprec}. 
+In the next two subsections, we explain in more details how \ZpLC and 
+\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
+Automatic Differentiation leading to the actual computation of a
+global object modeling the precision (which is the aforementioned 
+lattice). 
+In \S \ref{ssec:viewprec}, we explain how precision on any individual number
+can be recoved from the above computations and discuss the validity
+of our results.
+The overhead complexity induced by our package is analyzed in
+\S \ref{ssec:complexity}.
 
@@
 
-\todo{Maybe, add a short intro.}
+We now explain in more details the internal mecanisms \ZpLC and \ZpLF 
+use for tracking precision.
 
@@
 \end{matrix} \big)$ where $C$ is the column vector:
-$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
-\cdot C_i$$
+\begin{equation}
+\label{eq:defC}
+C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
+\cdot C_i
+\end{equation}
 where $C_i$ is the column vector of $M_t$ corresponding to the
@@
 The package \ZpLC tackles this issue by introducing a cap: we do not 
-work with $dF_{\underline x}(H_t)$ but instead define the lattice: 
-$$H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$$
+work with $dF_{\underline x}(H_t)$ but instead define the lattice
+$H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$
 where $N_{t+1}$ is an integer, the so-called \emph{cap}.
@@
 with $y = f(x_1, \ldots, x_n)$.
-
 In concrete terms, the lattice $H_{t+1}$ is represented by the block 
@@
 \emph{i.e.} we define $x_\ttw$ as $f(x_1, \ldots, x_n) \text{ mod } 
-p^{N_{t+1}}$. This makes sense because:
-$$\bar x \oplus f(x_1, \ldots, x_n) \equiv
-\bar x \oplus x_\ttw \pmod{H_{t+1}}$$
-thanks to the extra generator we have added.
+p^{N_{t+1}}$. 
+The congruence
+$\bar x \oplus f(x_1, \ldots, x_n) \equiv
+\bar x \oplus x_\ttw \pmod{H_{t+1}}$
+(which holds thanks to the extra generator we have added) justifies
+this choice.
 
@@
 In this situation, keeping in mind the cap, the lattice $H_{t+1}$ is 
-defined by:
-$$H_{t+1} = H_t + p^{\min(\text{prec}, N_{t+1})} \Zp e_\ttw$$
+defined by
+$H_{t+1} = H_t + p^{\min(\text{prec}, N_{t+1})} \Zp e_\ttw$
 where the cap $N_{t+1}$ is:
@@
 affectation ``\verb?w = ...?'' where the variable \ttw is already 
-defined.
-
-This situation can actually be decomposed into three steps: first, we 
-compute the right-hand-side and store it in a temporary variable \tttmp; 
-second, we delete \ttw; third, we rename \tttmp in \ttw. The first step 
-corresponds to the creation of a variable and was then already studied. 
-Similarly, the second step is a variable deletion and again has already 
-been discussed. As for the third step, it just changes the name of an
-element of $\calV_t$ but does not affect the lattice $H_t$, nor the
-matrix $M_t$.
+defined. This situation reduces to the creation of the temporary
+variable (the value of the right-hand-size), the deletion of the
+old variable \ttw and a renaming. It can then be handled using the
+methods discussed previously.
 
@@
 
-Below, we explain in more details how the package \ZpLF updates 
-the matrix $M_t$ when a variable is created or deleted.
+%Below, we explain in more details how the package \ZpLF updates 
+%the matrix $M_t$ when a variable is created or deleted.
 
@@
 affectation of the form:
-
-\medskip
-
-\hfill\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?\hfill\null
-
-\medskip
-
-\noindent
-As already explained in the case of \ZpLC, this line of code is 
+``\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?''.
+As already explained in the case of \ZpLC, this code is 
 modeled by the mathematical mapping:
@@
 M_t & C \end{matrix}\right)$
-where $C$ is the column vector:
-$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
-\cdot C_i.$$
-However, we insist on the fact that all the above computations take
+where $C$ is the column vector defined by Eq.~\eqref{eq:defC}.
+%$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
+%\cdot C_i.$$
+However, we insist on the fact that all the computations now take
 place in the ``ring'' of Floating Point $p$-adic numbers. Therefore,
 we cannot guarantee that the rows of $M_{t+1}$ generate $H_{t+1}$.
-Nonetheless, they generate a module which is expected to be closed 
+Nonetheless, they generate a module which is expected to be close
 to $H_{t+1}$.
@@
 \begin{tikzpicture}[scale=0.5]
-\fill[black!20] (0,0)--(1,0)--(1,-2)--(2,-2)--(2,2)--(0,2)--(0,0);
+\fill[black!20] (0,0)--(1,0)--(1,-2)--(2,-2)--(2,1)--(0,1)--(0,0);
 \draw (0,0)--(1,0)--(1,-2)--(2,-2);
-\node at (1.5,2.8) { $C$ };
-\draw[->] (1.5,2.4)--(1.5,2.1);
+\node at (1.5,1.8) { $C$ };
+\draw[->] (1.5,1.4)--(1.5,1.1);
 \node at (1.5,-0.5) { \ph $x$ };
@@
 \begin{scope}[xshift=5cm]
-\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,2)--(0,2)--(0,0);
+\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,1)--(0,1)--(0,0);
 \draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2);
-\node at (1.5,2.8) { $C$ };
-\draw[->] (1.5,2.4)--(1.5,2.1);
+\node at (1.5,1.8) { $C$ };
+\draw[->] (1.5,1.4)--(1.5,1.1);
 \node at (1.5,-1.5) { \ph $y$ };
@@
 \subsection{Visualizing the precision}
+\label{ssec:viewprec}
 
@@
 
+{\noindent \small
 \noindent
 \begin{tabular}{rl}
-\cIn  & \verb?R = ZpLC(2, print_mode='digits')? \\
-      & \verb?v = R(173,10); v? \\
+\cIn  & \verb?v = Z2(173,10); v? \\
 \cOut & \verb?...0010101101?
-\end{tabular}
+\end{tabular}}
 
@@
 
+{\noindent \small
 \noindent
@@
 \cOut & \verb?10?
-\end{tabular}
+\end{tabular}}
 
@@
 indicates that a good candidate for the precision on $e_\ttv$ is 
-$\pi_\ttv(H_t) = p^N \Zp$.
+$\pi_\ttv(H_t)$, that is $p^N \Zp$.
 
@@
 $\underline u_t \oplus \underline k_t$. 
-Set $\Phi_{t,\ttv} = \pi_\ttv \circ \Phi_t$. It maps 
+Set $\Phi_{t,\ttv} = \pi_\ttv \circ \Phi_t$; it maps 
 $\underline u_t \oplus \underline k_t$ to the \ttv-coordinate 
-$x_{t,\ttv}$ of $\underline x_t$ and satisfies:
-$$d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$$
+$x_{t,\ttv}$ of $\underline x_t$ and satisfies
+$d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$
 where $N$ is the value returned by \verb?precision_absolute?. Thus, 
@@
 we can have access to the joint precision on a set of variables: it 
-outputs a matrix (in echelon form) whose rows generates a lattice 
+outputs a matrix (in echelon form) whose rows generate a lattice 
 representing the precision on the subset of given variables.
@@
 
-\noindent
-\begin{tabular}{rl}
-\cIn  & \verb?R = ZpLC(2, print_mode='digits')? \\
-      & \verb?x = R(987,10); y = R(21,5)? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn  & \verb?x = Z2(987,10); y = Z2(21,5)? \\
       & \verb?R.precision().precision_lattice([x,y])? \\
@@
       & \verb?[   0   32]?
-\end{tabular}
+\end{tabular}}
 
@@
 
-\noindent
+{\noindent \small
 \begin{tabular}{rl}
@@
       & \verb?[   0 2048]?
-\end{tabular}
+\end{tabular}}
 
@@
 
-\noindent
-\begin{tabular}{rl}
-\cIn  & \verb?u? \\
-\cOut & \verb?...10000? \\
-\cIn  & \verb?v? \\
-\cOut & \verb?...00110? \\
+{\noindent \small
+\begin{tabular}{rl}
+\cIn  & \verb?u, v? \\
+\cOut & \verb?(...10000, ...00110)? \\
 \cIn  & \verb?u + v? \\
 \cOut & \verb?...11110110110? \\
-\end{tabular}
+\end{tabular}}
 
@@
 \cOut & \verb?6? \\
-\end{tabular}
+\end{tabular}}
 
@@
 $$\left(\begin{matrix} 2^5 & 2016 \\ 0 & 2^{11} \end{matrix}\right).$$
-The minimal diagonal suplattice $H_0$ is generated by the scalar matrix 
+The minimal diagonal suplattice $H_0$ of $H$ is generated by the scalar matrix 
 $2^5 \cdot \text{I}_2$ and contains $H$ with index $2^6$ in it. This is 
@@
 \subsection{Complexity}
+\label{ssec:complexity}
 
@@
 is an inexact field, our model of complexity makes sense because the
-size of the $p$-adic numbers we manipulate will always be roughly the
+size of the $p$-adic numbers we manipulate will all have roughly the
 same size: for \ZpLF, it is the precision we use for Floating Point 
-Arithmetic while, for \ZpLC, it is bounded by the absolute cap which
+Arithmetic while, for \ZpLC, it is the absolute cap which
 was fixed at the beginning.
@@
 ordered with respect to $<_t$. We denote by $r_t$ (resp. $c_t$) the
-number of rows (resp. of columns) of $M_t$. By construction $r_t$ is
+number of rows (resp. columns) of $M_t$. By construction $r_t$ is
 also the cardinality of $\calV_t$. We have $c_t \leq r_t$ and the 
@@
 \emph{height} (resp. the \emph{co-height}) of \ttv at time $t$ as the 
-number of pivot variables \ttw such that $\ttw \leq_t \ttv$ (resp. such 
-that $\ttw >_t \ttv$). We denote it by $\hgt_t(\ttv)$ (resp. by 
+number of pivot variables \ttw such that $\ttw \leq_t \ttv$ (resp. 
+$\ttw >_t \ttv$). We denote it by $\hgt_t(\ttv)$ (resp. by 
 $\cohgt_t(\ttv)$). Clearly $\hgt_t(\ttv) + \cohgt_t(\ttv) = c_t$.
@@
 created variables (which reflects the complexity) together with the 
-maximum number of variables involved at the same time (which reflects 
+maximum number of variables alive at the same time (which reflects 
 the memory occupation) while executing two basic computations.
@@
 (which is a division-free algorithm of quartic complexity) while the
-second one is the computation of the gdc of two polynomials using a
+second one is the computation of the gcd of two polynomials using a
 naive Euclidean algorithm (of quadratic complexity). We can observe 
@@
 row-echelonization. In terms of algebraic complexity, the deletion is 
-free. As for the cost of the echelonization, it stays within:
-$$O\big(\coind_t(\ttw) \cdot \cohgt_t(\ttw) \big)$$
+free. The cost of the echelonization is within
+$O\big(\coind_t(\ttw) \cdot \cohgt_t(\ttw) \big)$
 operations in $\Qp$.
@@
 that the cost becomes $O\big(\coind_t(\ttw)^2\big)$. This does not look 
-nice \emph{a priori}. However, the principle of temporal locality 
+nice \emph{a priori}. However, the principle of temporal locality \cite{}
 asserts that $\cohgt_t(\ttw)$ tends to be small in general: destroyed 
@@
 roughly $O(mc)$. Going further in speculation, we might estimate the 
-magnitude of $m$ by $m \simeq O(s + \sqrt c)$ with $s = \max(s_\inp, 
+magnitude of $m$ as about $s + \sqrt c$ with $s = \max(s_\inp, 
 s_\out)$, leading to a complexity of $O(c^{3/2} + sc)$. For 
@@
 as soon as the complexity is at least quadratic in $s$, the dominant 
-term is $c^{3/2}$ and the impact on the complexity is then diminished.
+term is $c^{3/2}$ and the impact on the complexity is then moderated.
 
@@
 of the SOMOS iteration.
+
+\smallskip
 
@@
 $y'=g \times h(y)$ as described in \citep{LV16}.
+
+\smallskip
 

### 2018-01-31 ZpL/ZpL.tex
Section 2.1 condensed

@@
 \cIn
- & \verb?x = random_element(Z3,5)? \\
- & \verb?x? \\
+ & \verb?x = random_element(Z3,5); x? \\
 \cZpCR
@@
 {\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?3*x? \\
+\begin{tabular}{rl@{\hspace{1.5cm}}rl}
+\cIn
+ & \verb?3*x? &
+\cIn
+ & \verb?x + x + x? \\
 \cZpCR
- & \verb?...111110? \\
+ & \verb?...111110? &
+\cZpCR
+ & \verb? ...11110? \\
+%\cZpFP
+% & \verb?...000000000000000111110? \\
 %\cZpFP
@@
 \cZpLC
- & \verb?...111110? \\
-\cIn
- & \verb?x + x + x? \\
-\cZpCR
- & \verb?...11110? \\
-%\cZpFP
-% & \verb?...000000000000000111110? \\
+ & \verb?...111110? &
 \cZpLC
@@
 
-Something is lost however in \ZpCR when performing $x+2x$.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?y = 2*x; y? \\
+%Something is lost however in \ZpCR when performing $x+2x$.
+%
+%\smallskip
+%
+%{\noindent \small
+%\begin{tabular}{rl}
+%\cIn
+% & \verb?y = 2*x; y? \\
+%\cZpCR
+% & \verb?...22222? \\
+%%\cZpFP
+%% & \verb?...00000000000000022222? \\
+%\cZpLC
+% & \verb?...22222? \\
+%\cIn
+% & \verb?x + y? \\
+%\cZpCR
+% & \verb?...11110? \\
+%%\cZpFP
+%% & \verb?...000000000000000111110? \\
+%\cZpLC
+% & \verb?...111110? \\
+%\end{tabular}}
+%
+%\smallskip
+
+\noindent
+\todo{Add a short explanation?}
+
+The same phenomenon occurs for multiplication.
+%when comparing the computation of $x^3$ and $x \times x^2$.
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl@{\hspace{1.5cm}}rl}
+\cIn
+ & \verb?x^3? &
+\cIn
+ & \verb?x * x * x? \\
 \cZpCR
- & \verb?...22222? \\
+ & \verb?...010101? &
+\cZpCR
+ & \verb? ...10101? \\
 %\cZpFP
-% & \verb?...00000000000000022222? \\
-\cZpLC
- & \verb?...22222? \\
-\cIn
- & \verb?x + y? \\
-\cZpCR
- & \verb?...11110? \\
-%\cZpFP
-% & \verb?...000000000000000111110? \\
-\cZpLC
- & \verb?...111110? \\
-\end{tabular}}
-
-\smallskip
-
-The same phenomenon occurs when comparing the computation
-of $x^3$ and $x \times x^2$.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?x^3? \\
-\cZpCR
- & \verb?...010101? \\
+% & \verb?...00000010100000010101? \\
 %\cZpFP
@@
 \cZpLC
- & \verb?...010101? \\
-\cIn
- & \verb?y = x^2? \\
- & \verb?x * y? \\
-\cZpCR
- & \verb?...10101? \\
-%\cZpFP
-% & \verb?...00000010100000010101? \\
+ & \verb?...010101? &
 \cZpLC
@@
 \cIn
- & \verb?u = x+y? \\
- & \verb?u? \\
+ & \verb?u, v = x+y, x-y? \\
+ & \verb?u, v? \\
 \cZpCR
- & \verb?...10111? \\
-%\cZpFP
-% & \verb?...000000000000000000000011010111? \\
+ & \verb?(...10111, ...01111)? \\
 \cZpLC
- & \verb?...10111? \\
-\cIn
- & \verb?v = x-y? \\
- & \verb?v? \\
+ & \verb?(...10111, ...01111)? \\
+\end{tabular}}
+
+\smallskip
+
+\noindent
+Now, let us compute $u+v$ and compare it with $2x$.
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl@{\hphantom{\hspace{0.8cm}}}rl}
+\cIn
+ & \verb?u + v? &
+\cIn
+ & \verb?2*x? \\
 \cZpCR
- & \verb?...01111? \\
-%\cZpFP
-% & \verb?...000000000000000000000011001111? \\
+ & \verb?...00110? &
+\cZpCR
+ & \verb?...00110100110? \\
 \cZpLC
- & \verb?...01111? \\
-\end{tabular}}
-
-\smallskip
-
-Now, let us compute $u+v$ (and observe that it is actually equal to $2x$).
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?u+v? \\
-\cZpCR
- & \verb?...00110? \\
-%\cZpFP
-% & \verb?...0000000000000000000000110100110? \\
+ & \verb?...00110100110? &
 \cZpLC
  & \verb?...00110100110? \\
-\cIn
- & \verb?2*x? \\
-\cZpCR
- & \verb?...00110100110? \\
-%\cZpFP
-% & \verb?...0000000000000000000000110100110? \\
-\cZpLC
- & \verb?...00110100110? \\
-\end{tabular}}
-
-\smallskip
-
+\end{tabular}}
+
+\smallskip
+
+\noindent
 No loss in precision appears in \ZpLC, contrary to \ZpCR.

### 2018-01-31 ZpL/ZpL.tex
Syntax highlighting

@@
 \usetikzlibrary{patterns}
+
+\definecolor{input}{HTML}{303060}
+\definecolor{output}{HTML}{A00000}
+\definecolor{string}{HTML}{A02020}
+\definecolor{ring}{HTML}{A020A0}
+\definecolor{function}{HTML}{A06020}
+\definecolor{constructor}{HTML}{205080}
+\definecolor{method}{HTML}{205080}
+\definecolor{keyword}{HTML}{008000}
 
@@
 \newcommand{\sage}{\textsc{SageMath}\xspace}
-\newcommand{\ZpCR}{\text{\rm \tt ZpCR}\xspace}
-\newcommand{\ZpFP}{\text{\rm \tt ZpFP}\xspace}
-\newcommand{\ZpL}{\text{\rm \tt ZpL}\xspace}
-\newcommand{\ZpLC}{\text{\rm \tt ZpLC}\xspace}
-\newcommand{\ZpLF}{\text{\rm \tt ZpLF}\xspace}
+\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
+\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
+\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
+\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
+\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
 
@@
 
-\newcommand{\cIn}{{\color{blue} \tt \phantom{Zp}In}:}
+\newcommand{\cIn}{{\color{input} \tt \phantom{Zp}In}:}
 \newcommand{\cOut}{{\color{red} \tt \phantom{Z}Out}:}
-\newcommand{\cZpCR}{{\color{red} \tt ZpCR}:}
-\newcommand{\cZpFP}{{\color{red} \tt ZpFP}:}
-\newcommand{\cZpLC}{{\color{red} \tt ZpLC}:}
-\newcommand{\cZpLF}{{\color{red} \tt ZpLF}:}
+\newcommand{\cZpCR}{{\color{output} \tt ZpCR}:}
+\newcommand{\cZpFP}{{\color{output} \tt ZpFP}:}
+\newcommand{\cZpLC}{{\color{output} \tt ZpLC}:}
+\newcommand{\cZpLF}{{\color{output} \tt ZpLF}:}
 
@@
 \cIn
- & \verb?Z2 = Zp(2, print_mode="digits")? \\
- & \verb?Q2 = Qp(2, print_mode="digits")? \\
+ & {\color{ring}\verb?Z2?}\verb? = Zp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
+ & {\color{ring}\verb?Q2?}\verb? = Qp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
 \end{tabular}}
@@
 is the default).
-The rings \verb?Z3? and \verb?Q3? are defined similarly.
+The rings {\color{ring}\verb?Z3?} and {\color{ring}\verb?Q3?} are defined similarly.
 
@@
 \cIn
- & \verb?x = random_element(Z3,5); x? \\
+ & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z3?}\verb?,5); x? \\
 \cZpCR
@@
 \cIn
- & \verb?x = random_element(Z2,10)? \\
- & \verb?y = random_element(Z2,5)? \\
+ & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,10)? \\
+ & \verb?y = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,5)? \\
 \cIn
@@
 \noindent
-Now, let us compute $u+v$ and compare it with $2x$.
+Now, let us compute $u+v$ and compare it with $2x$ (observe that
+they should be equal).
 
@@
 \cIn
- & \verb?M = random_element(MatrixSpace(Q2,3), prec=10)? \\
+ & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
  & \verb?M.charpoly()? \\
@@
 %\cIn
-% & \verb?M = random_element(MatrixSpace(Q2,3), prec=10)? \\
+% & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
 % & \verb?M.determinant()? \\
@@
 \cIn
- & \verb?S.<x> = PolynomialRing(Z2)? \\
- & \verb?P = random_element(S, degree=10, prec=5)? \\
- & \verb?Q = random_element(S, degree=10, prec=5)? \\
- & \verb?D = x^5 + random_element(S, degree=4, prec=8); D? \\
+ & \verb?S.<x> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Z2?}\verb?)? \\
+ & \verb?P = ?{\color{function}\verb?random_element?}\verb?(S, degree=10, prec=5)? \\
+ & \verb?Q = ?{\color{function}\verb?random_element?}\verb?(S, degree=10, prec=5)? \\
+ & \verb?D = x^5 + ?{\color{function}\verb?random_element?}\verb?(S, degree=4, prec=8); D? \\
 \cZpCR
@@
 \cIn
- & \verb?def euclidean(A,B):? \\
- & \verb?    while B != 0:? \\
+ & {\color{keyword}\verb?def?}\verb? ?{\color{function}\verb?euclidean?}\verb?(A,B):? \\
+ & \verb?    ?{\color{keyword}\verb?while?}\verb? B != 0:? \\
  & \verb?        A, B = B, A % B? \\
- & \verb?    return A.monic()? \\
- & \verb?euclidean(D*P, D*Q)? \\
+ & \verb?    ?{\color{keyword}\verb?return?}\verb? A.monic()? \\
+ & \verb??{\color{function}\verb?euclidean?}\verb?(D*P, D*Q)? \\
 \cZpCR
@@
 \begin{tabular}{rl}
-\cIn  & \verb?v = Z2(173,10); v? \\
-\cOut & \verb?...0010101101?
+\cIn   & \verb?v = ?{\color{ring}\verb?Z2?}\verb?(173,10); v? \\
+\cZpLC & \verb?...0010101101?
 \end{tabular}}
@@
 exactly $10$ digits are printed.
-The method \verb?precision_absolute? provides a more easy-to-use
+The method \verb??{\color{method}\verb?precision_absolute?}\verb?? provides a more easy-to-use
 access to the absolute precision.
@@
 \begin{tabular}{rl}
-\cIn  & \verb?v.precision_absolute()? \\
-\cOut & \verb?10?
+\cIn   & \verb?v.?{\color{method}\verb?precision_absolute?}\verb?()? \\
+\cZpLC & \verb?10?
 \end{tabular}}
@@
 $d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$
-where $N$ is the value returned by \verb?precision_absolute?. Thus, 
+where $N$ is the value returned by \verb??{\color{method}\verb?precision_absolute?}\verb??. Thus, 
 as soon as the assumptions of Proposition \ref{prop:precision} are
@@
 %
-Our package implements the method \verb?precision_lattice? through which 
+Our package implements the method \verb??{\color{method}\verb?precision_lattice?}\verb?? through which 
 we can have access to the joint precision on a set of variables: it 
@@
 When the variables are ``independant'', the precision lattice is
-split and the method \verb?precision_lattice? outputs a diagonal
+split and the method \verb??{\color{method}\verb?precision_lattice?}\verb?? outputs a diagonal
 matrix:
@@
 \begin{tabular}{rl}
-\cIn  & \verb?x = Z2(987,10); y = Z2(21,5)? \\
-      & \verb?R.precision().precision_lattice([x,y])? \\
-\cOut & \verb?[1024    0]? \\
+\cIn   & \verb?x = ?{\color{ring}\verb?Z2?}\verb?(987,10); y = ?{\color{ring}\verb?Z2?}\verb?(21,5)? \\
+       & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision_lattice?}\verb?([x,y])? \\
+\cZpLC & \verb?[1024    0]? \\
        & \verb?[   0   32]?
@@
 \cIn   & \verb?u, v = x+y, x-y? \\
-      & \verb?R.precision().precision_lattice([u,v])? \\
-\cOut & \verb?[  32 2016]? \\
+       & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision_lattice?}\verb?([u,v])? \\
+\cZpLC & \verb?[  32 2016]? \\
        & \verb?[   0 2048]?
@@
 \cIn   & \verb?u, v? \\
-\cOut & \verb?(...10000, ...00110)? \\
+\cZpLC & \verb?(...10000, ...00110)? \\
 \cIn   & \verb?u + v? \\
-\cOut & \verb?...11110110110? \\
+\cZpLC & \verb?...11110110110? \\
 \end{tabular}}
@@
 
-The method \verb?number_of_diffused_digits? computes the number of
+The method \verb??{\color{method}\verb?number_of_diffused_digits?}\verb?? computes the number of
 diffused digits of precision on a set of variables. Observe:
@@
 \begin{tabular}{rl}
-\cIn  & \verb?R.precision().number_of_diffused_digits([x,y])? \\
-\cOut & \verb?0? \\
-\cIn  & \verb?R.precision().number_of_diffused_digits([u,v])? \\
-\cOut & \verb?6? \\
+\cIn   & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?number_of_diffused_digits?}\verb?([x,y])? \\
+\cZpLC & \verb?0? \\
+\cIn   & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?number_of_diffused_digits?}\verb?([u,v])? \\
+\cZpLC & \verb?6? \\
 \end{tabular}}
@@
 \cIn
- & \verb?def somos4(u0, u1, u2, u3, n):? \\
+ & {\color{keyword}\verb?def?}\verb? somos4(u0, u1, u2, u3, n):? \\
  & \verb?    a, b, c, d = u0, u1, u2, u3? \\
@@
  & \verb?        a, b, c, d = b, c, d, (b*d + c*c) / a? \\
- & \verb?    return d? \\
-\cIn
- & \verb?u0 = u1 = u2 = Z2(1,15); u3 = Z2(3,15)? \\
+ & \verb?    ?{\color{keyword}\verb?return?}\verb? d? \\
+\cIn
+ & \verb?u0 = u1 = u2 = ?{\color{ring}\verb?Z2?}\verb?(1,15); u3 = ?{\color{ring}\verb?Z2?}\verb?(3,15)? \\
  & \verb?somos4(u0, u1, u2, u3, 18)? \\
@@
 \cIn
-% & \verb?Z2 = Zp(2, prec=100)? \\
- & \verb?MS = MatrixSpace(Z2,2)? \\
- & \verb?M = random_element(MS, prec=5)? \\
+% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=100)? \\
+ & \verb?MS = ?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Z2?}\verb?,2)? \\
+ & \verb?M = ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
  & \verb?for _ in range(100):? \\
- & \verb?    M *= random_element(MS, prec=5)? \\
+ & \verb?    M *= ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
  & \verb?M? \\
@@
 %\cIn
-% & \verb?Z2.precision().number_of_diffused_digits()? \\
-%\cZpLC
+% & {\color{ring}\verb?Z2?}\verb?.precision().?{\color{method}\verb?number_of_diffused_digits?}\verb?()? \\
+%\ZpLC
 % & \verb?604? \\
@@
 {\noindent \small
-\begin{tabular}{rl}
-\cIn
-% & \verb?Q2 = Qp(2, prec=50, print_mode="digits")? \\
- & \verb?R.<x,y,z> = PolynomialRing(Q2, order = 'invlex')? \\
- & \verb?F = [ Q2(2,10)*x + Q2(1,10)*z, ? \\
- & \verb?      Q2(1,10)*x^2 + Q2(1,10)*y^2 - Q2(2,10)*z^2,? \\
- & \verb?      Q2(4,10)*y^2 + Q2(1,10)*y*z + Q2(8,10)*z^2 ]? \\
-\cIn
- & \verb?from sage.rings.polynomial.toy_buchberger\?\\
- & \verb?    import buchberger_improved? \\
- & \verb?g = buchberger_improved(ideal(F)); g.sort()? \\
- & \verb?g? \\
+\begin{tabular}{@{}rl}
+\cIn
+% & {\color{ring}\verb?Q2?}\verb? = Qp(2, prec=50, print_mode="digits")? \\
+ & \verb?R.<x,y,z> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, order=?{\color{string}\verb?'invlex'?}\verb?)? \\
+ & \verb?F = [ ?{\color{ring}\verb?Q2?}\verb?(2,10)*x + ?{\color{ring}\verb?Q2?}\verb?(1,10)*z, ? \\
+ & \verb?      ?{\color{ring}\verb?Q2?}\verb?(1,10)*x^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y^2 - ?{\color{ring}\verb?Q2?}\verb?(2,10)*z^2,? \\
+ & \verb?      ?{\color{ring}\verb?Q2?}\verb?(4,10)*y^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y*z + ?{\color{ring}\verb?Q2?}\verb?(8,10)*z^2 ]? \\
+\cIn
+ & {\color{keyword}\verb?from?}\verb? sage.rings.polynomial.toy_buchberger\?\\
+ & \verb?    ?{\color{keyword}\verb?import?}\verb? ?{\color{function}\verb?buchberger_improved?} \\
+ & \verb?g = ?{\color{function}\verb?buchberger_improved?}\verb?(ideal(F))? \\
+ & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
 \cZpCR
@@
 We can reach this theoretical loss in precision using \verb?ZpL?.
-We use a generic \verb?Newton_Iteration_Solver(g,h,N)?
+We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
 that applies \verb?N? steps of the Newton method for 
@@
 % & \verb?N = 4? \\
-% & \verb?Z2 = Zp(2, prec=50, print_mode="digits")? \\
-% & \verb?Q2=Z2.fraction_field()? \\
- & \verb?S.<t> = PowerSeriesRing(Q2, 16)? \\
+% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=50, print_mode="digits")? \\
+% & {\color{ring}\verb?Q2?}\verb?=?{\color{ring}\verb?Z2?}\verb?.fraction_field()? \\
+ & \verb?S.<t> = ?{\color{constructor}\verb?PowerSeriesRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, 16)? \\
  & \verb?h = 1 + t + t^3? \\
- & \verb?y = t + t^2 * random_element(S,10)? \\
+ & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S,10)? \\
  & \verb?g = y.derivative() / h(y)? \\
- & \verb?u = Newton_Iteration_Solver(g, h, 4)? \\
+ & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4)? \\
  & \verb?u[15]? \\
@@
  & \verb?R = Qp(3,50)? \\
- & \verb?A.<x> = PolynomialRing(R)? \\
+ & \verb?A.<x> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(R)? \\
  & \verb?m = 67? \\

### 2018-01-31 ZpL/ZpL.tex
Typo

@@
 
-A naive implementation of the Euclide algorithm
+A naive implementation of the Euclidean algorithm
 can produce different behaviour depending 
@@
 {\noindent \small
-\begin{tabular}{rl}
+\begin{tabular}{@{}rl}
 \cIn

### 2018-02-01 ZpL/ZpL.tex
More syntax highlighting
More syntax highlighting (again)

@@
 \definecolor{input}{HTML}{303060}
-\definecolor{output}{HTML}{A00000}
+\definecolor{output}{HTML}{804000}
 \definecolor{string}{HTML}{A02020}
 \definecolor{ring}{HTML}{A020A0}
-\definecolor{function}{HTML}{A06020}
+\definecolor{function}{HTML}{205080} % {A06020}
 \definecolor{constructor}{HTML}{205080}
@@
 \definecolor{keyword}{HTML}{008000}
+\definecolor{error}{HTML}{B01010}
 
@@
 We first take some random element $x$.
+\todo{Should we say a word about the function \texttt{random\_element}?}
 
@@
  & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
- & \verb?M.charpoly()? \\
+ & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
 \cZpCR
@@
  & \verb?    ?{\color{keyword}\verb?return?}\verb? A.monic()? \\
- & \verb??{\color{function}\verb?euclidean?}\verb?(D*P, D*Q)? \\
+ & {\color{function}\verb?euclidean?}\verb?(D*P, D*Q)? \\
 \cZpCR
@@
 exactly $10$ digits are printed.
-The method \verb??{\color{method}\verb?precision_absolute?}\verb?? provides a more easy-to-use
+The method {\color{method}\verb?precision_absolute?} provides a more easy-to-use
 access to the absolute precision.
@@
 $d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$
-where $N$ is the value returned by \verb??{\color{method}\verb?precision_absolute?}\verb??. Thus, 
+where $N$ is the value returned by {\color{method}\verb?precision_absolute?}. Thus, 
 as soon as the assumptions of Proposition \ref{prop:precision} are
@@
 %
-Our package implements the method \verb??{\color{method}\verb?precision_lattice?}\verb?? through which 
+Our package implements the method {\color{method}\verb?precision_lattice?} through which 
 we can have access to the joint precision on a set of variables: it 
@@
 When the variables are ``independant'', the precision lattice is
-split and the method \verb??{\color{method}\verb?precision_lattice?}\verb?? outputs a diagonal
+split and the method {\color{method}\verb?precision_lattice?} outputs a diagonal
 matrix:
@@
 
-The method \verb??{\color{method}\verb?number_of_diffused_digits?}\verb?? computes the number of
+The method {\color{method}\verb?number_of_diffused_digits?} computes the number of
 diffused digits of precision on a set of variables. Observe:
@@
 \cIn
- & {\color{keyword}\verb?def?}\verb? somos4(u0, u1, u2, u3, n):? \\
+ & {\color{keyword}\verb?def?}\verb? ?{\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, n):? \\
  & \verb?    a, b, c, d = u0, u1, u2, u3? \\
- & \verb?    for _ in range(4, n+1):? \\
+ & \verb?    ?{\color{keyword}\verb?for?}\verb? _ in ?{\color{keyword}\verb?range?}\verb?(4, n+1):? \\
  & \verb?        a, b, c, d = b, c, d, (b*d + c*c) / a? \\
@@
  & \verb?u0 = u1 = u2 = ?{\color{ring}\verb?Z2?}\verb?(1,15); u3 = ?{\color{ring}\verb?Z2?}\verb?(3,15)? \\
- & \verb?somos4(u0, u1, u2, u3, 18)? \\
+ & {\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, 18)? \\
 \cZpCR
@@
 \cIn
- & \verb?somos4(u0, u1, u2, u3, 100)? \\
+ & {\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, 100)? \\
 \cZpCR
- & \verb?PrecisionError: cannot divide by something?\\
+ & {\color{error}\verb?PrecisionError?}\verb?: cannot divide by something?\\
  &\verb?indistinguishable from zero.? \\
@@
 provide an heavy diffused digits phenomenon.
-Here is a sample example. the result for \ZpFP has not been
-added, but its first digits coincide with that of \ZpLC.
+Here is a simple example.
+% The result for \ZpFP has not been
+% added, but its first digits coincide with that of \ZpLC.
 
@@
  & \verb?M = ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
- & \verb?for _ in range(100):? \\
+ & \verb?for _ in range(25):? \\
  & \verb?    M *= ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
@@
 \cZpCR
- & \verb?[ O(2^9) O(2^11)]? \\
- & \verb?[ O(2^9) O(2^11)]? \\
+ & \verb?[0 0]? \\
+ & \verb?[0 0]? \\
 \cZpLC
- & \verb?[2^9 + 2^10 + 2^11 + O(2^12)  O(2^14)]? \\
- & \verb?[       2^9 + 2^11 + O(2^12)  O(2^14)]? \\
+ & \verb?[...100000000000   ...1000000000]? \\
+ & \verb?[   ...010000000     ...00100000]? \\
 \end{tabular}}
@@
  & \verb?h = 1 + ((m^2)/4)*x^2 + m^6*x^6? \\
- & \verb?y = Lercier_Sirvent(g, h, 6)? \\
+ & \verb?y = ?{\color{function}\verb?Lercier_Sirvent?}\verb?(g, h, 6)? \\
  & \verb?y[31]? \\
- & \verb?u = Newton_Iteration_Solver(g, h, N)? \\  % ???
- & \verb?u[15]? \\
 \cZpCR

### 2018-02-02 ZpL/ZpL.tex
Adding introduction, fixing a few typos

--- 2018-01-31 ZpL/ZpL.tex
@@
 \usepackage{tabularx,multicol}
-\usepackage{tikz}
-\usetikzlibrary{patterns}
-
-\definecolor{input}{HTML}{303060}
-\definecolor{output}{HTML}{A00000}
-\definecolor{string}{HTML}{A02020}
-\definecolor{ring}{HTML}{A020A0}
-\definecolor{function}{HTML}{A06020}
-\definecolor{constructor}{HTML}{205080}
-\definecolor{method}{HTML}{205080}
-\definecolor{keyword}{HTML}{008000}
+%\usepackage{tikz}
 
@@
 
-\newcommand{\ph}{\vphantom{$A^A_A$}}
-
 \newcommand{\N}{\mathbb N}
@@
 
-\newcommand{\calV}{\mathcal{V}}
-\newcommand{\ttv}{\texttt{v}\xspace}
-\newcommand{\ttw}{\texttt{w}\xspace}
-\newcommand{\tttmp}{\texttt{tmp}\xspace}
-\newcommand{\calU}{\mathcal{U}}
-\newcommand{\calK}{\mathcal{K}}
-
-\newcommand{\rem}{\,\%\,}
-
-\newcommand{\sage}{\textsc{SageMath}\xspace}
-\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
-\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
-\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
-\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
-\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
-
-\newcommand{\card}{\text{Card}\:}
-\newcommand{\ind}{\text{ind}}
-\newcommand{\coind}{\text{coind}}
-\newcommand{\hgt}{\text{hgt}}
-\newcommand{\cohgt}{\text{cohgt}}
-\newcommand{\inp}{\text{in}}
-\newcommand{\out}{\text{out}}
+\newcommand{\ZpL}{\texttt{ZpL}\xspace}
+\newcommand{\ZpLCA}{\texttt{ZpLCA}\xspace}
+\newcommand{\ZpLCR}{\texttt{ZpLCR}\xspace}
+\newcommand{\ZpLF}{\texttt{ZpLF}\xspace}
 
@@
 \newcommand{\app}{\text{\rm app}}
-
-\newcommand{\cIn}{{\color{input} \tt \phantom{Zp}In}:}
-\newcommand{\cOut}{{\color{red} \tt \phantom{Z}Out}:}
-\newcommand{\cZpCR}{{\color{output} \tt ZpCR}:}
-\newcommand{\cZpFP}{{\color{output} \tt ZpFP}:}
-\newcommand{\cZpLC}{{\color{output} \tt ZpLC}:}
-\newcommand{\cZpLF}{{\color{output} \tt ZpLF}:}
 
@@
 \theoremstyle{definition}
-\newtheorem{rmk}[theo]{Remark}
+\newtheorem{rem}[theo]{Remark}
 \newtheorem{ex}[theo]{Example}
@@
   \affiliation{Universit\'e Rennes 1; \\
-  \institution{IRMAR}
-  \city{Rennes, France}
-  \postcode{35042}
+  %\institution{IRMAR}
+  %\city{Rennes, France}
+  %\postcode{35042}
 }
@@
 \author{David Roe}
-  \affiliation{University of Pittsburg; \\
-  \institution{Department of Mathematics}
-  \city{Pittsburgh, PA, USA}
-  \postcode{15260}
+  \affiliation{MIT; \\
+  %\institution{Department of Mathematics}
+  %\city{Pittsburgh, PA, USA}
+  %\postcode{15260}
 }
-\email{roed@pitt.edu}
+\email{roed@mit.edu}
 \author{Tristan Vaccon}
   \affiliation{Universit\'e de Limoges; \\
-  \institution{CNRS, XLIM UMR 7252}
-  \city{Limoges, France}  
-  \postcode{87060}  
+  %\institution{CNRS, XLIM UMR 7252}
+  %\city{Limoges, France}  
+  %\postcode{87060}  
   }
@@
 
-\keywords{Algorithms, $p$-adic precision, Automatic Differentiation}
+\keywords{Algorithms, $p$-adic precision, characteristic polynomial,
+eigenvalue}
 
 \begin{abstract}
-We present a new package \ZpL for the Computer Algebra system \sage. It 
-implements a sharp tracking of precision on $p$-adic numbers, following 
-the theory of ultrametric precision introduced in \cite{padicprec}. The 
-underlying algorithms are mostly based on Automatic Differentiation 
-techniques. We introduce them, study their complexity and discuss our 
-design choices.
-We illustrate the benefits of our package (in comparison with previous 
-implementations) with a large sample of examples coming from Linear 
-Algebra, Commutative Algebra and Differential Equations.
 \end{abstract}
@@
 
-\begin{itemize}
-\item Interval arithmetic, floating point arithmetic
-\item short introduction of the precision lemma and lattices
-\item goal: tracking precision using this theory
-\item short presentation of the package
-\item plan of the article
-\end{itemize}
+When computing with real and $p$-adic fields, exact results are usually impossible, since most elements have infinite decimal or $p$-adic expansions.
+Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.
+In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \todo{include citations}.
+The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.
 
-\todo{Convention: in matrices, vectors are represented by row-vector.}
+The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.
+At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \todo{add reference to floating point arithmetic}.
+For example, the representable numbers in $53$-bit floating point arithmetic are those of the form $m \cdot 2^e$ \todo{describe valid $m$ and $e$}.
+Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.
+Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.
+
+If provably correct results are desired, interval arithmetic provides an alternative to floating point.
+Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.
+This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
+Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision \todo{include reference to SOMOS section}.
+
+For better control of precision, we may use ellipsoids rather than balls.
+This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.
+For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.
+From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Lemma \todo{add number} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.
+
+In this paper, we describe an implementation of this idea in Sage \todo{cite Sage}.  \todo{Add short description of package features.}
+
+\todo{Plan of article}
 
@@
 
-We begin our tour of the features of the \ZpL package
-with some basic arithmetic computations.
-The first step is the definition of the $p$-adic ring or field.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & {\color{ring}\verb?Z2?}\verb? = Zp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
- & {\color{ring}\verb?Q2?}\verb? = Qp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
-\end{tabular}}
-
-\smallskip
-
-\noindent
-It means that we will work in various implementation of the 
-ring of $2$-adic integers with $20$ digits of precision (which
-is the default).
-The rings {\color{ring}\verb?Z3?} and {\color{ring}\verb?Q3?} are defined similarly.
-
-We show the results for \ZpCR%, \ZpFP 
-and \ZpLC.
-Results for \ZpLF are only shown when they differ from \ZpLC.
-
 \subsection{Elementary arithmetic}
 
-We first take some random element $x$.
+\begin{tabular}{|l|l|}
+\verb?R = Zp(3,5)? & \verb?R = ZpL(3,5)? \\
+\verb?x = R(143,5); x? & \verb?x = R(143,5); x? \\
+\hfill\verb?...12022? & \hfill\verb?...12022? \\
+\verb?3*x? & \verb?3*x? \\
+\hfill\verb?...120220? & \hfill\verb?...120220? \\
+\verb?x + x + x? & \verb?x + x + x? \\
+\hfill\verb?...20220? & \hfill\verb?...120220? \\
+\verb?x^3? & \verb?x^3? \\
+\hfill\verb?...020222? & \hfill\verb?...020222? \\
+\verb?x * x * x? & \verb?x * x * x? \\
+\hfill\verb?...20222? & \hfill\verb?...020222? \\
+\end{tabular}
 
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z3?}\verb?,5); x? \\
-\cZpCR
- & \verb?...11111? \\
-%\cZpFP
-% & \verb?...00000000000000011111? \\
-\cZpLC
- & \verb?...11111? \\
-\end{tabular}}
-
-\smallskip
-
-Multiplication by $3$ is a shift on the digits, whether it 
-is done by addition or multiplication.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl@{\hspace{1.5cm}}rl}
-\cIn
- & \verb?3*x? &
-\cIn
- & \verb?x + x + x? \\
-\cZpCR
- & \verb?...111110? &
-\cZpCR
- & \verb? ...11110? \\
-%\cZpFP
-% & \verb?...000000000000000111110? \\
-%\cZpFP
-% & \verb?...000000000000000111110? \\
-\cZpLC
- & \verb?...111110? &
-\cZpLC
- & \verb?...111110? \\
-\end{tabular}}
-
-\smallskip
-
-%Something is lost however in \ZpCR when performing $x+2x$.
-%
-%\smallskip
-%
-%{\noindent \small
-%\begin{tabular}{rl}
-%\cIn
-% & \verb?y = 2*x; y? \\
-%\cZpCR
-% & \verb?...22222? \\
-%%\cZpFP
-%% & \verb?...00000000000000022222? \\
-%\cZpLC
-% & \verb?...22222? \\
-%\cIn
-% & \verb?x + y? \\
-%\cZpCR
-% & \verb?...11110? \\
-%%\cZpFP
-%% & \verb?...000000000000000111110? \\
-%\cZpLC
-% & \verb?...111110? \\
-%\end{tabular}}
-%
-%\smallskip
-
-\noindent
-\todo{Add a short explanation?}
-
-The same phenomenon occurs for multiplication.
-%when comparing the computation of $x^3$ and $x \times x^2$.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl@{\hspace{1.5cm}}rl}
-\cIn
- & \verb?x^3? &
-\cIn
- & \verb?x * x * x? \\
-\cZpCR
- & \verb?...010101? &
-\cZpCR
- & \verb? ...10101? \\
-%\cZpFP
-% & \verb?...00000010100000010101? \\
-%\cZpFP
-% & \verb?...00000010100000010101? \\
-\cZpLC
- & \verb?...010101? &
-\cZpLC
- & \verb?...010101? \\
-\end{tabular}}
-
-\medskip
-
-\ZpL is also well suited for working with coefficients 
-with unbalanced precision.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,10)? \\
- & \verb?y = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,5)? \\
-\cIn
- & \verb?u, v = x+y, x-y? \\
- & \verb?u, v? \\
-\cZpCR
- & \verb?(...10111, ...01111)? \\
-\cZpLC
- & \verb?(...10111, ...01111)? \\
-\end{tabular}}
-
-\smallskip
-
-\noindent
-Now, let us compute $u+v$ and compare it with $2x$ (observe that
-they should be equal).
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl@{\hphantom{\hspace{0.8cm}}}rl}
-\cIn
- & \verb?u + v? &
-\cIn
- & \verb?2*x? \\
-\cZpCR
- & \verb?...00110? &
-\cZpCR
- & \verb?...00110100110? \\
-\cZpLC
- & \verb?...00110100110? &
-\cZpLC
- & \verb?...00110100110? \\
-\end{tabular}}
-
-\smallskip
-
-\noindent
-No loss in precision appears in \ZpLC, contrary to \ZpCR.
-%We remark that \ZpFP also provides all the correct digits.
+\begin{tabular}{|l|l|}
+\verb?R = Zp(2,10)? & \verb?R = ZpL(2,10)? \\
+\verb?x = R(987,10); x? & \verb?x = R(987,10); x? \\
+\hfill\verb?...1111011011? & \hfill\verb?...1111011011? \\
+\verb?y = R(21,5); y? & \verb?y = R(21,5); y? \\
+\hfill\verb?...10101? & \hfill\verb?...10101? \\
+\verb?u = x + y; u? & \verb?u = x + y; u? \\
+\hfill\verb?...10000? & \hfill\verb?...10000? \\
+\verb?v = x - y; v? & \verb?v = x - y; v? \\
+\hfill\verb?...00110? & \hfill\verb?...00110? \\
+\verb?u + v? & \verb?u + v? \\
+\hfill\verb?...10110? & \hfill\verb?...11110110110? \\
+\end{tabular}
 
@@
 
-In linear algebra, determinant and characteristic polynomial are 
-notoriously hard to compute: see \cite{caruso-roe-vaccon:15, 
-caruso-roe-vaccon:17}. We illustrate this with the following example 
-(using a division-free algorithm for the computation of the 
-characteristic polynomial) :
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
- & \verb?M.charpoly()? \\
-\cZpCR
- & \verb?...00000000000000000001*x^3 + ? \\
- & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
-\cZpLC
- & \verb?...00000000000000000001*x^3 + ? \\
- & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
- & \verb?...010000010? \\
-\end{tabular}}
-
-%\begin{tabular}{rl}
-%\cIn
-% & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
-% & \verb?M.determinant()? \\
-%\cZpCR
-% & \verb?...000100000100? \\
-%\cZpFP
-% & \verb? ...11010101010? \\
-%\cZpLC
-% & \verb?  ...0110110001? \\
-%\cZpLF
-% & \verb?...001110111100? \\
-%\cIn
-% & \verb?M.charpoly()? \\
-%\cZpCR
-% & \verb?(...0000000001)*x^3 + (...1001111111)*x^2 ?\\
-% & \verb?+ (...01110101110)*x + (...111011111100)? \\
-%\cZpFP
-% & \verb?...0000000001*x^3 + ...1101001001*x^2 + ?\\
-% & \verb?...0100110011*x + ...00101010110? \\
-%\cZpLC
-% & \verb?...0000000001*x^3 + ...1011000111*x^2 + ?\\
-% & \verb?...0011101000*x + ...1001001111? \\
-%\cZpLF
-% & \verb?...0000000001*x^3 + ...1111000101*x^2 + ?\\
-% & \verb?...0011111111*x + ...110001000100? \\
-%\end{tabular}
+\begin{itemize}
+\item determinant
+\item characteristic polynomial
+\end{itemize}
 
@@
 
-Our package can be applied to computation over $p$-adic polynomials. A 
-natural example is that of the computation of GCD, whose stability has 
-been studied in \cite{caruso-2017}.
+\begin{itemize}
+\item gcd of polynomials
+\end{itemize}
 
-A naive implementation of the Euclide algorithm
-can produce different behaviour depending 
-on the type of implementation of the field
-of $p$-adic coefficients.
+\section{Theory}
 
-\smallskip
+\subsection{The precision Lemma}
 
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?S.<x> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Z2?}\verb?)? \\
- & \verb?P = ?{\color{function}\verb?random_element?}\verb?(S, degree=10, prec=5)? \\
- & \verb?Q = ?{\color{function}\verb?random_element?}\verb?(S, degree=10, prec=5)? \\
- & \verb?D = x^5 + ?{\color{function}\verb?random_element?}\verb?(S, degree=4, prec=8); D? \\
-\cZpCR
- & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
- & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
- & \verb?...10000110*x + ...11100010? \\
-\cZpLC
- & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
- & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
- & \verb?...10000110*x + ...11100010? \\
-\cIn
- & {\color{keyword}\verb?def?}\verb? ?{\color{function}\verb?euclidean?}\verb?(A,B):? \\
- & \verb?    ?{\color{keyword}\verb?while?}\verb? B != 0:? \\
- & \verb?        A, B = B, A % B? \\
- & \verb?    ?{\color{keyword}\verb?return?}\verb? A.monic()? \\
- & \verb??{\color{function}\verb?euclidean?}\verb?(D*P, D*Q)? \\
-\cZpCR
- & \verb?0*x^9 + ...1*x^8 + 0*x^7 + 0*x^6 + 0*x^5 +? \\
- & \verb?0*x^4 + 0*x^3 + ...1*x^2 + ...10*x + ...10? \\
-\cZpLC
- & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
- & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
- & \verb?...10000110*x + ...11100010? \\
-\end{tabular}}
+\subsection{The way the precision is tracked with \ZpL}
 
-\smallskip
+We follow the precision lattice.
+Closely related to automatic differentiation.
 
-\noindent
-With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
-$DP$ is $DQ$ is $D$.  However, we observe that we do not always get this 
-expected result. More precisely, \ZpCR often founds a gcd of higher 
-degree. Indeed the Euclidean algorithm stops prematurely because the 
-software believes that some early remainder vanishes due to the lack of 
-precision.
-%The situation with ZpFP is the exact opposite:
-%since the precision is too high (always the maximal cap), 
-%the software sometimes does not notice that the remainder vanishes.
-%Hence, it happens that the Euclidean algorithm 
-%obtains a last non-zero remainder of degree less than that of
-%$D.$
+\begin{itemize}
+\item \ZpLCR, \ZpLCA: we use a cap to ensure that we have a lattice at
+each step.
+\item \ZpLF: we work with $\Zp$-submodules of possible positive codimension
+and represent them using matrices of floating point $p$-adics.
+\end{itemize}
+
+\subsection{The notion of working precision}
+
+\begin{itemize}
+\item what is it?
+\item how can we compute it? (Answer: Hermite reduction of the transpose)
+\item how can we estimate it?
+\end{itemize}
+
+\subsection{Correctness}
+
+\begin{itemize}
+\item prove that \ZpLCA and \ZpLCR are correct under some assumptions
+(to be determined)
+\item explain that \ZpLF is not meant to provide proved results
+\end{itemize}
+
+\subsection{Complexity}
+
+\begin{tabular}{l|c|c|}
+& \ZpLCA/\ZpLCR & \ZpLF \\
+\hline
+Creation of a variable & $O(n)$ & $O(n_0)$ \\
+\hline
+Deletion of a variable & $O(m^2)$ & $-$ \\
+\hline
+\end{tabular}
+
+discussion on temporal locality
 
@@
 
-In this Section, we explain how our package \ZpL works and we analyze
-its performances.
-The main theoretical result on which our package is based in the 
-ultrametric precision theory developed in \cite{padicprec}, which 
-suggests to track precision \emph{via} lattices and differential 
-computations. For this reason, our approach is very inspired by 
-Automatic Differentiation techniques \cite{} and our implementation 
-follows the usual Operator Overloading strategy \cite{}.
-We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
-this former is safer while the latter is faster.
+\begin{itemize}
+\item Discuss coercion/conversion: \texttt{Zp} coerces to \ZpL but no coersion
+in the other direction
+\item How to access to the precision
+\item Diffused digits
+\item Historic of variables creation/deletion, etc...
+\item Other features?
+\end{itemize}
 
-This Section is organized as follows.
-\S \ref{ssec:preclemma} contains a brief overview on the theory of
-$p$-adic precision of \cite{padicprec}. 
-In the next two subsections, we explain in more details how \ZpLC and 
-\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
-Automatic Differentiation leading to the actual computation of a
-global object modeling the precision (which is the aforementioned 
-lattice). 
-In \S \ref{ssec:viewprec}, we explain how precision on any individual number
-can be recoved from the above computations and discuss the validity
-of our results.
-The overhead complexity induced by our package is analyzed in
-\S \ref{ssec:complexity}.
+\subsection{More examples}
 
-\smallskip
+\begin{tabular}{|l|l|}
+\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
+\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
+\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
+\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
+\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
+\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
+\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
+\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
+\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
+\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
+\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
+\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
+\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
+\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
+\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
+\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
+\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
+\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
+\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
+\hfill\verb?...11? & \hfill\verb?...0000000111? \\
+\hfill\verb?...10? & \hfill\verb?...1010001110? \\
+\hfill\verb?...01? & \hfill\verb?...1110110001? \\
+\hfill\verb?...01? & \hfill\verb?...011110111001? \\
+\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
+\hfill\verb?...1? & \hfill\verb?...11111000011? \\
+\end{tabular}
 
-\noindent \textit{Remark about the naming.}
-%
-The letter \texttt{L}, which appears in the name of the package, 
-comes from ``lattices''. The letters \texttt{C} (in \ZpLC) and 
-\texttt{F} (in \ZpLF) stand for ``cap'' and ``float'' respectively.
+\begin{itemize}
+\item SOMOS sequence
+\item many multiplications of matrices
+\item Dodgson algorithm
+\item subresultants
+\item Grobner basis
+\item $p$-adic differential equations
+\end{itemize}
 
-\subsection{The precision Lemma}
-\label{ssec:preclemma}
-
-In \cite{caruso-roe-vaccon:14a}, we have suggested the 
-use of lattices to represent the precision of elements in 
-$K$-vector spaces.  It contrasts with the
-\emph{coordinate-wise method} (of \textit{e.g.}
- \verb?Zp(5)?) that is used traditionally in \sage and
-where the precision of an element is specified by giving the precision
-of each coordinate separately and is updated after each basic
-operation.
-
-Consider a finite 
-dimensional normed vector space $E$ 
-defined over $K$. 
-We use the notation $\Vert \cdot \Vert_E$ for the norm 
-on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
-closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
-E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
-Because of ultrametricity, the balls $B^{\phantom 
--}_E(r)$ and $B^-_E(r)$ are examples of lattices. 
-Lattices can be thought of as special neighborhoods of $0,$
- and therefore are good 
-candidates to model precision data. Moreover, as revealed in 
-\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
-differentiable maps:
-
-\begin{prop}
-\label{prop:precision}
-Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
-and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
-$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
-that the differential $df_{v_0}$ is surjective.
-Then, for all $\rho \in (0, 1]$, there exists a positive real
-number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
-$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
-satisfies:
-\begin{equation}
-\label{eq:firstorder}
-f(v_0 + H) = f(v_0) + df_{v_0} (H).
-\end{equation}
-\end{prop}
-
-This proposition enables the \emph{lattice method} of tracking precision,
-where the precision of the input is specified as a lattice $H$ and precision
-is tracked via differentials of the steps within a given algorithm.
-The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
-yields the optimum possible precision. 
-We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
-exposition.
-We emphasize that polynomials, rational fractions (outside of their poles)
-and converging power series are strictly differentiable.
-
-%The lemma is more precise in the context of so-called
-%integral polynomial functions.
-%A function $f : E \to F$ is said to be
-%\emph{integral polynomial} if it is given by multivariate polynomial 
-%functions with coefficients in $\O_K$ in any (equivalently all) system 
-%of coordinates associated to a $\OK$-basis of $B_E(1)$.
-%
-%\begin{prop}
-%\label{prop:precision2}
-%We keep the notation of Proposition \ref{prop:precision} and assume 
-%in addition that $f$ is integral polynomial. Let $C$ be a positive real
-%number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
-%Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
-%\rho^{-1}$.
-%\end{prop}
-
-\subsection{Tracking precision}
-\label{ssec:trackprec}
-
-We now explain in more details the internal mecanisms \ZpLC and \ZpLF 
-use for tracking precision.
-
-In what follows, it will be convenient to use a notion of discrete time 
-represented by the letter $t$. Rigorously, it is defined as follows: 
-$t=0$ when the $p$-adic ring $\ZpLC(\:\cdots)$ is created and increases by 
-$1$ each time a variable is created, deleted\footnote{The deletion can
-be explicit (through a call to the \texttt{del} operator) or implicit
-(handled by the garbage collector).} or updated.
-
-Let $\calV_t$ be the set of alive variables at time $t$. Set $E_t = 
-\Qp^{\calV_t}$; it is a finite dimensional vector space over $\Qp$ which 
-should be thought as the set of all possible values that can be taken by 
-the variables in $\calV_t$. For $\ttv \in \calV_t$, let $e_\ttv \in
-\Qp^{\calV_t}$ be the vector whose all coordinates vanished except 
-that at position $\ttv$ which takes the value $1$. The family 
-$(e_\ttv)_{\ttv \in \calV_t}$ is obviously a basis of $E_t$; we will 
-refer to it as the \emph{canonical basis}.
-
-\subsubsection{The case of \ZpLC.}
-
-Following Proposition~\ref{prop:precision},
-the package \ZpLC follows the precision by keeping track of a lattice
-$H_t$ in $E_t$, which is a global object whose purpose is to model the 
-precision on all the variables in $\calV_t$ all together.
-Concretely, this lattice is represented by a matrix $M_t$ in row-echelon 
-form whose rows form a set of generators.
-Below, we explain how the matrices $M_t$ are updated each time
-$t$ increases.
-
-\smallskip
-
-\noindent \textit{Creating a variable.}
-%
-This happens when we encounter an instruction having one of the
-two following forms:
-
-\medskip
-
-\noindent \hspace{5mm} \makebox[2.5cm]{[Computation]\hfill\null}
-\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?
-
-\smallskip
-
-\noindent \hspace{5mm} \makebox[2.5cm]{[New value]\hfill\null}
-\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?
-
-\medskip
-
-\noindent
-In both cases, $\ttw$ is the newly created variable. 
-The $\ttv_i$'s stand for already defined variables and $f$ is some 
-$n$-ary builtin function (in most cases it is just addition, 
-subtraction, multiplication or division). On the contrary, the terms 
-``value'' and ``prec'' refer to user-specified constants or integral 
-values which was computed earlier.
-
-\smallskip
-
-Let us first examine the first construction [Computation].
-With our conventions, if $t$ is the time just before the execution of 
-the instruction we are interested in, the $\ttv_i$'s lie in $\calV_t$ 
-while $\ttw$ does not. Moreover $\calV_{t+1} = \calV_t \sqcup \{\ttw\}$, 
-so that $E_{t+1} = E_t \oplus \Qp e_\ttw$.
-The mapping taking the values of variables at time $t$ to that at 
-time $t{+}1$ is:
-$$\begin{array}{rcl}
-F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
-\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n)
-\end{array}$$
-where $x_i$ is the $\ttv_i$-th coordinate of the vector $\underline x$.
-The Jacobian matrix of $F$ at $\underline x$ is easily computed; it 
-is the block matrix $J_{\underline x}(F) = \big( \begin{matrix} I & L
-\end{matrix} \big)$ where $I$ is the identity matrix of size $\text{Card}
-\:\calV_t$ and $L$ is the column vector whose $\ttv$-th entry is
-$\frac{\partial f}{\partial \ttv} (\underline x)$ if $\ttv$ is one of 
-the $\ttv_i$'s and $0$ otherwise.
-Therefore, the image of $H_t$ under $dF_{\underline x}$ is represented 
-by the matrix $J_{\underline x}(F) \cdot M_t = \big( \begin{matrix} M_t & C
-\end{matrix} \big)$ where $C$ is the column vector:
-\begin{equation}
-\label{eq:defC}
-C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
-\cdot C_i
-\end{equation}
-where $C_i$ is the column vector of $M_t$ corresponding to the
-variable $\ttv_i$.
-Observe that the matrix $J_{\underline x}(F) \cdot M_t$ is no longer a 
-square matrix; it has one extra column. This reflects the fact that 
-$\dim E_{t+1} = \dim E_t + 1$. Rephasing this in a different language, 
-the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
-$E_{t+1}$ but is included in an hyperplane.
-
-The package \ZpLC tackles this issue by introducing a cap: we do not 
-work with $dF_{\underline x}(H_t)$ but instead define the lattice
-$H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$
-where $N_{t+1}$ is an integer, the so-called \emph{cap}.
-Alternatively, one may introduce the application:
-\begin{equation}
-\label{eq:tildeF}
-\begin{array}{rcl}
-\tilde F: \quad E_t \oplus \Qp & \longrightarrow & E_{t+1} \smallskip \\
-\underline x \oplus c & \mapsto & \underline x \:\oplus 
-\big(f(x_1, \ldots, x_n) + c\big).
-\end{array}
-\end{equation}
-The lattice $H_{t+1}$ is then the image of $H_t \oplus p^{N_{t+1}}
-\Zp$ under $d\tilde F_{(\underline x, \star)}$ for any value of $\star$.
-The choice of the 
-cap is of course a sensitive question. \ZpLC proceeds as follows. When a 
-ring is created, it comes with two constants (which can be specified by 
-the user): a relative cap \textsc{relcap} and an absolute cap 
-\textsc{abscap}. With these predefined values, the chosen cap is:
-\begin{equation}
-\label{eq:cap}
-N_{t+1} = 
-\min\big(\textsc{abscap},\, \textsc{relcap} + v_p(y)\big)
-\end{equation}
-with $y = f(x_1, \ldots, x_n)$.
-In concrete terms, the lattice $H_{t+1}$ is represented by the block 
-matrix:
-$$\left(\begin{matrix}
-M_t & C \smallskip \\ 0 & p^{N_{t+1}}
-\end{matrix}\right).$$
-Performing row operations, we see then the entries of $C$ can be
-reduced modulo $p^{N_{t+1}}$ without changing the lattice. In order
-to optimize the size of the objects, we perform this reduction and
-define $M_{t+1}$ by:
-$$M_{t+1} = \left(\begin{matrix}
-M_t & C \text{ mod } p^{N_{t+1}} \smallskip \\ 0 & p^{N_{t+1}}
-\end{matrix}\right).$$
-We observe in particular that $M_{t+1}$ is still in row-echelon form.
-
-Finally, we need to precise which value is set to the newly created 
-variable $\ttw$. We observe that it cannot be exactly $f(x_1, \ldots, 
-x_n)$ because the latter is \emph{a priori} a $p$-adic number which 
-cannot be computed exactly. For this reason, we have to truncate it at 
-some finite precision. Again we choose the precision $O(p^{N_{t+1}})$,
-\emph{i.e.} we define $x_\ttw$ as $f(x_1, \ldots, x_n) \text{ mod } 
-p^{N_{t+1}}$. 
-The congruence
-$\bar x \oplus f(x_1, \ldots, x_n) \equiv
-\bar x \oplus x_\ttw \pmod{H_{t+1}}$
-(which holds thanks to the extra generator we have added) justifies
-this choice.
-
-\smallskip
-
-The second construction
-``\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?''
-is easier to handle since, roughly speaking, it corresponds to the
-case $n = 0$.
-In this situation, keeping in mind the cap, the lattice $H_{t+1}$ is 
-defined by
-$H_{t+1} = H_t + p^{\min(\text{prec}, N_{t+1})} \Zp e_\ttw$
-where the cap $N_{t+1}$ is:
-$$N_{t+1} = \min\big(\textsc{abscap},\, \textsc{relcap} + 
-v_p(\text{value})\big).$$
-The corresponding matrix $M_{t+1}$ is then given by:
-$$M_{t+1} = \left(\begin{matrix}
-M_t & 0 \smallskip \\ 0 & p^{\min(\text{prec}, N_{t+1})}
-\end{matrix}\right).$$
-
-\medskip
-
-\noindent \textit{Deleting a variable.}
-%
-Let us now examine the case where a variable $\ttw$ is deleted (or 
-collected by the garbage collector). Just after the deletion, at
-time $t{+}1$, we then have $\calV_{t+1} = \calV_t \backslash \{\ttw\}$.
-Thus $E_t = E_{t+1} \oplus \Qp e_\ttw$. Moreover, the deletion of $\ttw$
-is modeled by the canonical projection $f : E_t \to E_{t+1}$. Since $f$
-is linear, it is its own differential (at each point) and we set 
-$H_{t+1} = f(H_t)$.
-A matrix representing $H_{t+1}$ is deduced from $M_t$ by erasing the 
-column corresponding to $\ttw$. However the matrix we get this way is 
-no longer in row-echelon form. We then need to re-echelonize it.
-
-\smallskip
-
-More precisely, in the \ZpLC case, the obtained matrix has this 
-shape:
-$$\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
-\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
-    --(3,-3)--(3,-4)--(4,-4)--(4,-5)--(5,-5)--(5,-6)--(6,-6)
-    --(6,-7)--(7,-7)--(7,0)--cycle;
-\draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
-    --(3,-3)--(3,-4)--(4,-4)--(4,-5)--(5,-5)--(5,-6)--(6,-6)
-    --(6,-7)--(7,-7);
-\draw[thin] (7,-7)--(7,0)--(0,0);
-\begin{scope}
-\clip (3,0.2) rectangle (4,-4.2);
-\draw[pattern=north east lines] (2.5,0.5) rectangle (4.5,-4.5);
-\end{scope}
-\draw[->] (2.9,-4.8)--(3.2,-4.2);
-\node[scale=0.8] at (2,-5.2) { deleted };
-\node[scale=0.8] at (2,-6) { column };
-\end{tikzpicture}}\right)
-\quad = \quad
-\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
-\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
-    --(3,-3)--(3,-5)--(4,-5)--(4,-6)--(5,-6)--(5,-7)--(6,-7)--(6,0)--cycle;
-\draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,-3)
-    --(3,-3)--(3,-5)--(4,-5)--(4,-6)--(5,-6)--(5,-7)--(6,-7);
-\draw[thin] (6,-7)--(6,0)--(0,0);
-\end{tikzpicture}}\right)$$
-where a cell is colored when it can contain a non-vanishing entry.
-The top part of the matrix is then already echenolized, so that we
-only have to re-echelonize the bottom right corner whose size is
-the distance from the column corresponding to the erased variable
-to the end. 
-Thanks to the particular shape of the matrix, the echelonization can be 
-performed efficiently: we combine the first rows (of the bottom right 
-part) in order to clear the first unwanted nonzero entry and then 
-proceed recursively.
-
-\medskip
-
-\noindent \textit{Updating a variable.}
-%
-Just like for creation, this happens when the program reaches an
-affectation ``\verb?w = ...?'' where the variable \ttw is already 
-defined. This situation reduces to the creation of the temporary
-variable (the value of the right-hand-size), the deletion of the
-old variable \ttw and a renaming. It can then be handled using the
-methods discussed previously.
-
-\subsubsection{The case of \ZpLF.}
-
-The way the package \ZpLF tracks precision is based on similar 
-techniques but differs from \ZpLC in that it does not introduce a cap 
-but instead allows $H_t$ to be a sub-$\Zp$-module of $E_t$ of any 
-codimension.
-This point of view is nice because it implies smaller objects and 
-consequently leads to faster algorithms. However, it has a huge 
-drawback; indeed, unlike lattices, submodules of $E_t$ of arbitrary 
-codimensions are \emph{not} exact objects, in the sense that they cannot 
-be represented by integral matrices in full generality. Consequently,
-they cannot be encoded on a computer.
-We work around this drawback by replacing everywhere exact $p$-adic 
-numbers by Floating Point $p$-adic numbers (at some given precision) 
-\cite{}.
-
-The fact that the lattice $H_t$ can now have arbitrary codimension
-translates to the fact the matrix $M_t$ can be rectangular.
-Precisely, we will maintain matrices $M_t$ of the shape:
-\begin{equation}
-\label{eq:shapeM-ZpLF}
-\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
-\fill[black!20] (0,0)--(0,-1)--(3,-1)--(3,-2)--(5,-2)--(5,-3)
-    --(6,-3)--(6,-4)--(8,-4)--(8,-5)--(9,-5)--(9,-6)--(12,-6)
-    --(12,-7)--(14,-7)--(14,0)--cycle;
-\draw (0,0)--(0,-1)--(3,-1)--(3,-2)--(5,-2)--(5,-3)
-    --(6,-3)--(6,-4)--(8,-4)--(8,-5)--(9,-5)--(9,-6)--(12,-6)
-    --(12,-7)--(14,-7);
-\fill (0,0) rectangle (1,-1);
-\fill (3,-1) rectangle (4,-2);
-\fill (5,-2) rectangle (6,-3);
-\fill (6,-3) rectangle (7,-4);
-\fill (8,-4) rectangle (9,-5);
-\fill (9,-5) rectangle (10,-6);
-\fill (12,-6) rectangle (13,-7);
-\draw[thin] (14,-7)--(14,0)--(0,0);
-\end{tikzpicture}}\right)
-\end{equation}
-where only the colored cells may contain a nonzero value and the 
-black cells ---the so-called \emph{pivots}--- do not vanish. A
-variable whose corresponding column contains a pivot will be called 
-a \emph{pivot variable at time $t$}.
-
-%Below, we explain in more details how the package \ZpLF updates 
-%the matrix $M_t$ when a variable is created or deleted.
-
-\medskip
-
-\noindent \textit{Creating a variable.}
-%
-We assume first that the newly created variable is defined through an
-affectation of the form:
-``\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?''.
-As already explained in the case of \ZpLC, this code is 
-modeled by the mathematical mapping:
-$$\begin{array}{rcl}
-F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
-\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n).
-\end{array}$$
-Here $\underline x$ represents the state of memory at time $t$, 
-and $x_i$ is the coordinate of $\underline x$ corresponding to the
-variable $\ttv_i$.
-
-In the \ZpLF framework, $H_{t+1}$ is defined as the image of $H_t$
-under the differential $dF_{\underline x}$. Accordingly, the matrix
-$M_{t+1}$ is defined as $M_{t+1} = \left(\begin{matrix}
-M_t & C \end{matrix}\right)$
-where $C$ is the column vector defined by Eq.~\eqref{eq:defC}.
-%$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
-%\cdot C_i.$$
-However, we insist on the fact that all the computations now take
-place in the ``ring'' of Floating Point $p$-adic numbers. Therefore,
-we cannot guarantee that the rows of $M_{t+1}$ generate $H_{t+1}$.
-Nonetheless, they generate a module which is expected to be close
-to $H_{t+1}$.
-
-\smallskip
-
-If \ttw is created by the code
-``\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?'',
-we define $H_{t+1} = H_t \oplus p^{\text{prec}} \Zp e_\ttw$ and consequently:
-$$M_{t+1} = \left(\begin{matrix}
-M_t & 0 \smallskip \\ 0 & p^{\text{prec}}
-\end{matrix}\right)$$
-If $\text{prec}$ is $+\infty$ (or, equivalently, not specified), we
-agree that $H_{t+1} = H_t$ and $M_{t+1} = (\begin{matrix} M_t & 0 
-\end{matrix})$.
-
-\medskip
-
-\noindent \textit{Deleting a variable.} 
-% 
-As for \ZpLC, the matrix operation implied by the deletion of the 
-variable \ttw is the deletion of the corresponding column of $M_t$. If 
-\ttw is not a pivot variable at time $t$, the matrix $M_t$ keeps the 
-form \eqref{eq:shapeM-ZpLF} after erasure; therefore no more treatment 
-is needed in this case.
-
-Otherwise, we re-echelonize the matrix as follows. After the deletion 
-of the column $C_\ttw$, we examine the first column $C$ which was 
-located on the right of $C_\ttw$. Two situations may occur (depending
-on the fact that $C$ was or was not a pivot column):
-
-\noindent \hfill
-\begin{tikzpicture}[scale=0.5]
-\fill[black!20] (0,0)--(1,0)--(1,-2)--(2,-2)--(2,1)--(0,1)--(0,0);
-\draw (0,0)--(1,0)--(1,-2)--(2,-2);
-\node at (1.5,1.8) { $C$ };
-\draw[->] (1.5,1.4)--(1.5,1.1);
-\node at (1.5,-0.5) { \ph $x$ };
-\node at (1.5,-1.5) { \ph $y$ };
-\node[scale=0.9] at (1,-2.8) { First case };
-\begin{scope}[xshift=5cm]
-\fill[black!20] (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2)--(2,1)--(0,1)--(0,0);
-\draw (0,0)--(0,-1)--(1,-1)--(1,-2)--(2,-2);
-\node at (1.5,1.8) { $C$ };
-\draw[->] (1.5,1.4)--(1.5,1.1);
-\node at (1.5,-1.5) { \ph $y$ };
-\node[scale=0.9] at (1,-2.8) { Second case };
-\end{scope}
-\end{tikzpicture}
-\hfill\null
-
-\noindent
-
-In the first case, we perform row operations in order to replace the 
-couple $(x,y)$ by $(d, 0)$ where $d$ is element of smallest valuation 
-among $x$ and $y$. Observe that $y$ is necessarily nonzero in this case, 
-so that $d$ does not vanish as well. After this operation, we move to
-the next column and repeat the same process.
-
-The second case is divided into two subcases. First, if $y$ does not 
-vanish, it can serve as a pivot and the obtained matrix has the desired 
-shape. When this occurs, the echelonization stops. On the contrary, if 
-$y = 0$, we just untint the corresponding cell and move to the next 
-column without modifying the matrix.
-
-\subsection{Visualizing the precision}
-\label{ssec:viewprec}
-
-Our package implements several methods giving access to the precision 
-structure. In the subsection, we present and discuss the most relevant 
-features in this direction.
-
-\medskip
-
-\noindent \textbf{Absolute precision of one element.}
-%
-This is the simplest accessible precision datum.
-It is encapsulated in the notation when an element is printed. For
-example, the (partial) session:
-
-\smallskip
-
-{\noindent \small
-\noindent
-\begin{tabular}{rl}
-\cIn   & \verb?v = ?{\color{ring}\verb?Z2?}\verb?(173,10); v? \\
-\cZpLC & \verb?...0010101101?
-\end{tabular}}
-
-\smallskip
-
-\noindent
-indicates that the absolute precision on \ttv is $10$ since
-exactly $10$ digits are printed.
-The method \verb??{\color{method}\verb?precision_absolute?}\verb?? provides a more easy-to-use
-access to the absolute precision.
-
-\smallskip
-
-{\noindent \small
-\noindent
-\begin{tabular}{rl}
-\cIn   & \verb?v.?{\color{method}\verb?precision_absolute?}\verb?()? \\
-\cZpLC & \verb?10?
-\end{tabular}}
-
-\smallskip
-
-\noindent
-Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at time 
-$t$) as the smallest valuation of an entry of the column of $M_t$ 
-corresponding to the variable \ttv. Alternatively, it can be defined as 
-the unique integer $N$ for which $\pi_\ttv(H_t) = p^N \Zp$ where 
-$\pi_\ttv : E_t \to \Qp$ takes a vector to its \ttv-coordinate.
-This definition of the absolute precision sounds revelant because, if we 
-believe that the submodule $H_t \subset E_t$ is supposed to encode the 
-precision on the variables in $\calV_t$, 
-Proposition~\ref{prop:precision} applied with the mapping $\pi_\ttv$ 
-indicates that a good candidate for the precision on $e_\ttv$ is 
-$\pi_\ttv(H_t)$, that is $p^N \Zp$.
-
-\medskip
-
-\noindent \textit{About correctness.}
-%
-We emphasize that the absolute precision computed this way is \emph{not} 
-proved, either for \ZpLF or \ZpLC. However, in the case of \ZpLC, one
-can be slightly more precise. Let $\calU_t$ be the vector space of 
-user-defined variables before time $t$ and $U_t$ be the lattice
-modeling the precision on them. The pair $(\calU_t, U_t)$ is defined 
-inductively as follows: we set $\calU_0 = U_0 = 0$ and 
-$\calU_{t+1} = \calU_t \oplus \Qp e_\ttw$,
-$U_{t+1} = U_t \oplus p^{\text{prec}} \Zp e_\ttw$
-when a new variable \ttw is created by
-``\verb?w = R(?value\verb?, ?prec\verb?)?''; otherwise, we put 
-$\calU_{t+1} = \calU_t$ and $U_{t+1} = U_t$. Moreover the values
-entered by the user defines a vector (with integral coordinates)
-$\underline u_t \in \calU_t$.
-
-Similarly, in order to model the caps, we define a pair $(\calK_t, 
-K_t)$ by the recurrence $\calK_{t+1} = \calK_t \oplus \Qp e_\ttw$,
-$K_{t+1} = K_t \oplus p^{N_{t+1}} \Zp e_\ttw$
-each time a new variable \ttw is created. 
-Here, the exponent $N_{t+1}$ is the cap defined by Eq.~\eqref{eq:cap}.
-In case of deletion, we put $\calK_{t+1} = \calK_t$ and $K_{t+1} = K_t$.
-
-Taking the compositum of all the functions $\tilde F$ (\emph{cf} 
-Eq.~\eqref{eq:tildeF}) from time $0$ to $t$, we find that the execution 
-of the session until time $t$ is modeled by a mathematical function
-$\Phi_t : \calU_t \oplus \calK_t \to E_t$.
-From the design of \ZpLC, we deduce further that there exists a 
-vector $\underline k_t \in K_t$ such that:
-$$\Phi_t(\underline u_t \oplus \underline k_t) = \underline x_t 
-\quad \text{and} \quad
-d\Phi_t (U_t \oplus K_t) = H_t$$
-where the differential of $\Phi_t$ is taken at the point
-$\underline u_t \oplus \underline k_t$. 
-Set $\Phi_{t,\ttv} = \pi_\ttv \circ \Phi_t$; it maps 
-$\underline u_t \oplus \underline k_t$ to the \ttv-coordinate 
-$x_{t,\ttv}$ of $\underline x_t$ and satisfies
-$d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$
-where $N$ is the value returned by \verb??{\color{method}\verb?precision_absolute?}\verb??. Thus, 
-as soon as the assumptions of Proposition \ref{prop:precision} are
-fulfilled, we derive
-$\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus (\underline k_t + K_t)\big) = 
-x_{t,\ttv} + p^N \Zp$.
-Noting that $k_t \in K_t$, we finally get:
-\begin{equation}
-\label{eq:absprec}
-\Phi_{t,\ttv} (\underline u_t + U_t) \subset
-\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus K_t\big) = 
-x_{t,\ttv} + p^N \Zp.
-\end{equation}
-The latter inclusion means that the computed value $x_{t,\ttv}$ is
-accurate at precision $O(p^N)$, \emph{i.e.} that the output absolute
-precision is correct.
-
-Unfortunately, checking automatically the assumptions of Proposition 
-\ref{prop:precision} in full generality seems to be difficult, though it 
-can be done by hand for many particular examples \cite{}.
-
-\begin{rmk}
-Assuming that Proposition \ref{prop:precision} applies, 
-the absolute precision computed as above is optimal if and only if
-the inclusion of \eqref{eq:absprec} is an equality. Applying again
-Proposition \ref{prop:precision}
-with the restricted mapping $\Phi_{t,\ttv} : \calU_t \to \Qp$ and the
-lattice $U_t$, we find that this happens if and only if
-$d\Phi_{t,\ttv}(U_t) = p^N \Zp$.
-
-Unfortunately, the latter condition cannot be checked on the matrix 
-$M_t$ (because of reductions). 
-However it is possible (and easy) to check whether the weaker condition 
-$d\Phi_{t,\ttv}(K_t) \subsetneq p^N \Zp$. This checking is achieved by 
-the method
-\verb?is_precision_capped? (provided by our package) which returns true 
-if $d\Phi_{t,\ttv}(K_t) = p^N \Zp$. As a consequence, when this method
-answers false, the absolute precision computed by the software is
-likely optimal.
-\end{rmk}
-
-\noindent \textbf{Precision on a subset of elements.}
-%
-Our package implements the method \verb??{\color{method}\verb?precision_lattice?}\verb?? through which 
-we can have access to the joint precision on a set of variables: it 
-outputs a matrix (in echelon form) whose rows generate a lattice 
-representing the precision on the subset of given variables.
-
-When the variables are ``independant'', the precision lattice is
-split and the method \verb??{\color{method}\verb?precision_lattice?}\verb?? outputs a diagonal
-matrix:
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn   & \verb?x = ?{\color{ring}\verb?Z2?}\verb?(987,10); y = ?{\color{ring}\verb?Z2?}\verb?(21,5)? \\
-       & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision_lattice?}\verb?([x,y])? \\
-\cZpLC & \verb?[1024    0]? \\
-       & \verb?[   0   32]?
-\end{tabular}}
-
-\smallskip
-
-\noindent
-However, after some computations, the precision matrix evolves and 
-does not remain diagonal in general (though it is always triangular
-because it is displayed in row-echelon form):
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn   & \verb?u, v = x+y, x-y? \\
-       & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision_lattice?}\verb?([u,v])? \\
-\cZpLC & \verb?[  32 2016]? \\
-       & \verb?[   0 2048]?
-\end{tabular}}
-
-\smallskip
-
-\noindent
-The fact that the precision matrix is no longer diagonal indicates
-that some well-chosen linear combinations of $u$ and $v$ are known
-with more digits than $u$ and $v$ themselves. In this particular
-example, the sum $u+v$ is known at precision $O(2^{11})$ while the
-(optimal) precision on $u$ and $v$ separately is only $O(2^5)$.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn   & \verb?u, v? \\
-\cZpLC & \verb?(...10000, ...00110)? \\
-\cIn   & \verb?u + v? \\
-\cZpLC & \verb?...11110110110? \\
-\end{tabular}}
-
-\medskip
-
-\noindent \textit{Diffused digits of precision.}
-%
-The phenomenon observed above is formalized by the notion of
-diffused digits of precision introduced in \cite{}. We recall
-briefly its definition.
-
-\begin{deftn}
-\label{def:diffused}
-Let $E$ be a $\Qp$-vector space endowed with a distinguished basis 
-$(e_1, \ldots, e_n)$ and write $\pi_i : E \to \Qp e_i$ for the 
-projections.
-Let $H \subset E$ be a lattice. The number of \emph{diffused digits of 
-precision} of $H$ is the length of $H_0/H$ where $H_0 = \pi_1(H) \oplus 
-\cdots \oplus \pi_n(H)$.
-\end{deftn}
-
-If $H$ represents the actual precision on some object, then
-$H_0$ is the smallest diagonal lattice containing $H$. It then
-corresponds to the maximal \emph{coordinate-wise} precision we can 
-reach on the set of $n$ variables corresponding to the basis $(e_1,
-\ldots, e_n)$.
-The diffused digits of precision only appear on linear combinations
-of those variables.
-
-The method \verb??{\color{method}\verb?number_of_diffused_digits?}\verb?? computes the number of
-diffused digits of precision on a set of variables. Observe:
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn   & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?number_of_diffused_digits?}\verb?([x,y])? \\
-\cZpLC & \verb?0? \\
-\cIn   & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?number_of_diffused_digits?}\verb?([u,v])? \\
-\cZpLC & \verb?6? \\
-\end{tabular}}
-
-\smallskip
-
-\noindent
-For the last example, we recall that the relevant precision lattice 
-$H$ is generated by the $2 \times 2$ matrix:
-$$\left(\begin{matrix} 2^5 & 2016 \\ 0 & 2^{11} \end{matrix}\right).$$
-The minimal diagonal suplattice $H_0$ of $H$ is generated by the scalar matrix 
-$2^5 \cdot \text{I}_2$ and contains $H$ with index $2^6$ in it. This is 
-where the $6$ digits of precision comes from.
-There are easily visible here: the sum $u+v$ is known with $11$ digits,
-that is exactly $6$ more digits than the summands $u$ and $v$.
-
-\subsection{Complexity}
-\label{ssec:complexity}
-
-We now discuss the cost of the above operations.
-In what follows, we shall count operations in $\Qp$. Although $\Qp$
-is an inexact field, our model of complexity makes sense because the
-size of the $p$-adic numbers we manipulate will all have roughly the
-same size: for \ZpLF, it is the precision we use for Floating Point 
-Arithmetic while, for \ZpLC, it is the absolute cap which
-was fixed at the beginning.
-
-It is convenient to introduce a total order on $\calV_t$: for $\ttv, 
-\ttw \in \calV_t$, we convene that $\ttv <_t \ttw$ is \ttv was created 
-before~\ttw. By construction, the columns of the matrix $M_t$ are 
-ordered with respect to $<_t$. We denote by $r_t$ (resp. $c_t$) the
-number of rows (resp. columns) of $M_t$. By construction $r_t$ is
-also the cardinality of $\calV_t$. We have $c_t \leq r_t$ and the 
-equality always holds in the \ZpLC case.
-
-For $\ttv \in \calV_t$, we define the \emph{index} of \ttv, denoted by 
-$\ind_t(\ttv)$ as the number of elements of $\calV_t$ which are not 
-greater than \ttv. If we sort the elements of $\calV_t$ by increasing 
-order, \ttv then appears in $\ind_t(\ttv)$-th position.
-We also define the \emph{co-index} of \ttv by
-$\coind_t(\ttv) = r_t - \ind_t(\ttv)$.
-
-Similarly, for any variable $\ttv \in \calV_t$, we define the 
-\emph{height} (resp. the \emph{co-height}) of \ttv at time $t$ as the 
-number of pivot variables \ttw such that $\ttw \leq_t \ttv$ (resp. 
-$\ttw >_t \ttv$). We denote it by $\hgt_t(\ttv)$ (resp. by 
-$\cohgt_t(\ttv)$). Clearly $\hgt_t(\ttv) + \cohgt_t(\ttv) = c_t$.
-The height of \ttv is the height of the significant 
-part of the column of $M_t$ which corresponds to \ttv. In the case of
-\ZpLC, all variables are pivot variables and thus $\hgt_t(\ttv) = 
-\ind_t(\ttv)$ and $\cohgt_t(\ttv) = \coind_t(\ttv)$ for all \ttv.
-
-\medskip
-
-\noindent \textbf{Creating a variable.}
-%
-With the notations of \S \ref{ssec:trackprec}, it is obvious that 
-creating a new variable \ttw requires:
-$$O\bigg(\sum_{i=1}^n \hgt_i(\ttv_i)\bigg) \subset O(n \: c_t)$$ 
-operations in $\Qp$. Here, we recall that $n$ is the arity of the 
-operation defining \ttw. In most cases it is $2$; thus the above
-complexity reduces to $O(c_t)$.
-
-In the \ZpLF context, $c_t$ counts the number of user-defined variables. 
-It is then expected to be constant (roughly equal to the size of the 
-input) while running a given algorithm.
-
-On the contrary, in the \ZpLC context, $c_t$ counts the number of
-variables which are alive at time $t$. It is no longer expected to
-be constant but evolves continuously when the algorithm runs.
-%
-\begin{figure}
-% Characteristic polynomial
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\noindent\hfill%
-\begin{tabular}{|l|r|r|r|r|r|}
-\hline
-\textbf{Dimension} &
- 2 &   5 &   10 &    20 &      50 \\
-\hline
-\textbf{Total} &
-35 & 424 & 5\:539 & 83\:369 & 3\:170\:657 \\
-\hline
-\textbf{Simult.} &
-17 &  65 &  225 &   845 &    5\:101 \\
-\hline
-\end{tabular}%
-\hfill\null
-
-\vspace{1mm}
-
-\noindent\hfill%
-{\small Computation of characteristic polynomial}%
-\hfill\null
-
-\bigskip
-
-% Euclidean algorithm
-
-\noindent\hfill%
-\begin{tabular}{|l|r|r|r|r|r|r|}
-\hline
-\textbf{Degree} &
- 2 &   5 &   10 &   20 &   50 &   100 \\
-\hline
-\textbf{Total} &
-54 & 130 &  332 & 1\:036 & 4\:110 & 10\:578 \\ 
-\hline
-\textbf{Simult.} &
-18 &  31 &   56 &  106 &  256 &   507 \\
-\hline
-\end{tabular}%
-\hfill\null
-
-\vspace{1mm}
-
-\noindent\hfill%
-{\small Naive Euclidean algorithm}%
-\hfill\null
-
-\caption{Numbers of involved variables}
-\label{fig:variables}
-\end{figure}
-%
-The tables of Figure~\ref{fig:variables} show the total number of 
-created variables (which reflects the complexity) together with the 
-maximum number of variables alive at the same time (which reflects 
-the memory occupation) while executing two basic computations.
-The first one is the computation of the characteristic polynomial of a 
-square matrix by the default algorithm used by \sage for $p$-adic fields 
-(which is a division-free algorithm of quartic complexity) while the
-second one is the computation of the gcd of two polynomials using a
-naive Euclidean algorithm (of quadratic complexity). We can observe 
-that, for both of them, the memory occupation is roughly equal to the 
-square root of the complexity.
-
-\medskip
-
-\noindent \textbf{Deleting a variable.}
-%
-The deletion of the variable \ttw induces the deletion of the 
-corresponding column of $M_t$, possibly followed by a partial 
-row-echelonization. In terms of algebraic complexity, the deletion is 
-free. The cost of the echelonization is within
-$O\big(\coind_t(\ttw) \cdot \cohgt_t(\ttw) \big)$
-operations in $\Qp$.
-
-In the \ZpLF case, we expect that, most of the time, the deleted
-variables were created after all initial variables were set by the
-user. This means that we expect $\cohgt_t(\ttw)$ to vanish and so,
-the corresponding cost to be negligible.
-
-In the \ZpLC case, we always have $\cohgt_t(\ttw) = \coind_t(\ttw)$, so 
-that the cost becomes $O\big(\coind_t(\ttw)^2\big)$. This does not look 
-nice \emph{a priori}. However, the principle of temporal locality \cite{}
-asserts that $\cohgt_t(\ttw)$ tends to be small in general: destroyed 
-variables are often variables that were created recently. As a basic 
-exemple, variables which are local to a small piece of code (\emph{e.g.} 
-a short function or a loop) do not survive for a long time. It turns
-out that this behaviour is typical in many implementations!
-%
-\begin{figure}
-\noindent\hfill%
-\begin{tikzpicture}[xscale=0.35,yscale=0.08]
-\draw[fill=blue] (-0.45,0) rectangle (0.45,46);
-\draw[fill=blue] (0.65,0) rectangle (1.45,69);
-\draw[fill=blue] (1.65,0) rectangle (2.45,5);
-\draw[fill=blue] (2.65,0) rectangle (3.45,7);
-\draw[fill=blue] (3.65,0) rectangle (4.45,7);
-\draw[fill=blue] (4.65,0) rectangle (5.45,12);
-\draw[fill=blue] (5.65,0) rectangle (6.45,7);
-\draw[fill=blue] (6.65,0) rectangle (7.45,5);
-\draw[fill=blue] (7.65,0) rectangle (8.45,1);
-\draw[fill=blue] (8.65,0) rectangle (9.45,1);
-\draw (9.65,0)--(10.45,0);
-\draw[fill=blue] (10.65,0) rectangle (11.45,1);
-\draw[fill=blue] (11.65,0) rectangle (12.45,1);
-\draw[fill=blue] (12.65,0) rectangle (13.45,2);
-\draw[fill=blue] (13.65,0) rectangle (14.45,2);
-\draw[fill=blue] (14.65,0) rectangle (15.45,2);
-\draw[fill=blue] (15.65,0) rectangle (16.45,2);
-\draw[fill=blue] (16.65,0) rectangle (17.45,1);
-\draw[fill=blue] (17.65,0) rectangle (18.45,1);
-\draw[fill=blue] (18.65,0) rectangle (19.45,1);
-\node[scale=0.7] at (0,-2.5) { $0$ };
-\node[scale=0.7] at (5,-2.5) { $5$ };
-\node[scale=0.7] at (10,-2.5) { $10$ };
-\node[scale=0.7] at (15,-2.5) { $15$ };
-\draw (-2,0)--(-2,72);
-\draw (-2.2,0)--(-1.8,0);
-\node[scale=0.7,left] at (-2.1,0) { $0$ };
-\draw (-2.2,10)--(-1.8,10);
-\node[scale=0.7,left] at (-2.1,10) { $10$ };
-\draw (-2.2,20)--(-1.8,20);
-\node[scale=0.7,left] at (-2.1,20) { $20$ };
-\draw (-2.2,30)--(-1.8,30);
-\node[scale=0.7,left] at (-2.1,30) { $30$ };
-\draw (-2.2,40)--(-1.8,40);
-\node[scale=0.7,left] at (-2.1,40) { $40$ };
-\draw (-2.2,50)--(-1.8,50);
-\node[scale=0.7,left] at (-2.1,50) { $50$ };
-\draw (-2.2,60)--(-1.8,60);
-\node[scale=0.7,left] at (-2.1,60) { $60$ };
-\draw (-2.2,70)--(-1.8,70);
-\node[scale=0.7,left] at (-2.1,70) { $70$ };
-\end{tikzpicture}%
-\hfill\null
-
-\caption{The distribution of $\cohgt_t(\ttw)$}
-\label{fig:deletion}
-\end{figure}
-%
-The histogram of Figure~\ref{fig:deletion} shows 
-the distribution of $\cohgt_t(\ttw)$ while executing the Euclidean
-algorithm (naive implementation) with two polynomials of degree $7$
-as input. The bias is evident: most of the time $\cohgt_t(\ttw) \leq
-1$.
-
-\medskip
-
-\noindent \textbf{Summary: Impact on complexity.}
-% 
-We consider the case of an algorithm with the following characteristics: 
-its complexity is $c$ operations in $\Qp$ (without any tracking of 
-precision), its memory usage is $m$ elements of $\Qp$, its input and its 
-output have size $s_\inp$ and $s_\out$ (elements of $\Qp$) respectively.
-
-In the case of \ZpLF, creating a variable has a cost $O(s_\inp)$ 
-whereas deleting a variable is free. 
-Thus when executed with the \ZpLF mecanism, the complexity of our
-algorithm becomes $O(s_\inp c)$.
-
-In the \ZpLC framework, creating a variable has a cost $O(m)$. The case 
-of deletion is more difficult to handle. However, by the temporal 
-locality principle, it seems safe to assume that it is not the 
-bottleneck (which is the case in practice). Therefore, when executed 
-with the \ZpLF mecanism, the cost of our algorithm is expected to be 
-roughly $O(mc)$. Going further in speculation, we might estimate the 
-magnitude of $m$ as about $s + \sqrt c$ with $s = \max(s_\inp, 
-s_\out)$, leading to a complexity of $O(c^{3/2} + sc)$. For 
-quasi-optimal algorithms, the term $sc \simeq c^2$ dominates. However, 
-as soon as the complexity is at least quadratic in $s$, the dominant 
-term is $c^{3/2}$ and the impact on the complexity is then moderated.
-
-\section{More examples}
-
-\subsection{SOMOS-4}
-
-The SOMOS-4 sequence is a well-known example
-of a computation with a large gap between
-the theoretical loss in precision
-and that of a naive computation (see \cite{caruso-roe-vaccon:14a}).
-We show here that precision can be saved
-even when using a generic implementation
-of the SOMOS iteration.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & {\color{keyword}\verb?def?}\verb? somos4(u0, u1, u2, u3, n):? \\
- & \verb?    a, b, c, d = u0, u1, u2, u3? \\
- & \verb?    for _ in range(4, n+1):? \\
- & \verb?        a, b, c, d = b, c, d, (b*d + c*c) / a? \\
- & \verb?    ?{\color{keyword}\verb?return?}\verb? d? \\
-\cIn
- & \verb?u0 = u1 = u2 = ?{\color{ring}\verb?Z2?}\verb?(1,15); u3 = ?{\color{ring}\verb?Z2?}\verb?(3,15)? \\
- & \verb?somos4(u0, u1, u2, u3, 18)? \\
-\cZpCR
- & \verb?...11? \\
-%\cZpFP
-% & \verb?...110111001000100100000000000111? \\
-\cZpLC
- & \verb?...100000000000111? \\
-\cIn
- & \verb?somos4(u0, u1, u2, u3, 100)? \\
-\cZpCR
- & \verb?PrecisionError: cannot divide by something?\\
- &\verb?indistinguishable from zero.? \\
-%\cZpFP
-% & \verb?...110000101000011001001001110001? \\
-\cZpLC
- & \verb?...001001001110001? \\
-\end{tabular}}
-
-%
-%\begin{tabular}{|l|l|}
-%\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
-%\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
-%\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
-%\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
-%\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
-%\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
-%\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
-%\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
-%\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
-%\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
-%\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
-%\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
-%\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
-%\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
-%\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
-%\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
-%\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
-%\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
-%\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
-%\hfill\verb?...11? & \hfill\verb?...0000000111? \\
-%\hfill\verb?...10? & \hfill\verb?...1010001110? \\
-%\hfill\verb?...01? & \hfill\verb?...1110110001? \\
-%\hfill\verb?...01? & \hfill\verb?...011110111001? \\
-%\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
-%\hfill\verb?...1? & \hfill\verb?...11111000011? \\
-%\end{tabular}
-%
-%\begin{itemize}
-%\item SOMOS sequence
-%\item many multiplications of matrices
-%\item Dodgson algorithm
-%\item subresultants
-%\item Grobner basis
-%\item $p$-adic differential equations
-%\end{itemize}
-
-\subsection{Matrix multiplication}
-The multiplication of many matrices has been studied
-in \citep{caruso-roe-vaccon:15} and is known to often 
-provide an heavy diffused digits phenomenon.
-Here is a sample example. the result for \ZpFP has not been
-added, but its first digits coincide with that of \ZpLC.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
-% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=100)? \\
- & \verb?MS = ?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Z2?}\verb?,2)? \\
- & \verb?M = ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
- & \verb?for _ in range(100):? \\
- & \verb?    M *= ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
- & \verb?M? \\
-\cZpCR
- & \verb?[ O(2^9) O(2^11)]? \\
- & \verb?[ O(2^9) O(2^11)]? \\
-\cZpLC
- & \verb?[2^9 + 2^10 + 2^11 + O(2^12)  O(2^14)]? \\
- & \verb?[       2^9 + 2^11 + O(2^12)  O(2^14)]? \\
-\end{tabular}}
-
-%\ZpL can count the number of diffused digits.
-%\begin{tabular}{rl}
-%\cIn
-% & {\color{ring}\verb?Z2?}\verb?.precision().?{\color{method}\verb?number_of_diffused_digits?}\verb?()? \\
-%\ZpLC
-% & \verb?604? \\
-%\end{tabular}
-
-\subsection{Gröbner bases}
-Our package can be applied 
-on complex computations like that of
-Gröbner bases using generic
-Gröbner bases algorithms.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{@{}rl}
-\cIn
-% & {\color{ring}\verb?Q2?}\verb? = Qp(2, prec=50, print_mode="digits")? \\
- & \verb?R.<x,y,z> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, order=?{\color{string}\verb?'invlex'?}\verb?)? \\
- & \verb?F = [ ?{\color{ring}\verb?Q2?}\verb?(2,10)*x + ?{\color{ring}\verb?Q2?}\verb?(1,10)*z, ? \\
- & \verb?      ?{\color{ring}\verb?Q2?}\verb?(1,10)*x^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y^2 - ?{\color{ring}\verb?Q2?}\verb?(2,10)*z^2,? \\
- & \verb?      ?{\color{ring}\verb?Q2?}\verb?(4,10)*y^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y*z + ?{\color{ring}\verb?Q2?}\verb?(8,10)*z^2 ]? \\
-\cIn
- & {\color{keyword}\verb?from?}\verb? sage.rings.polynomial.toy_buchberger\?\\
- & \verb?    ?{\color{keyword}\verb?import?}\verb? ?{\color{function}\verb?buchberger_improved?} \\
- & \verb?g = ?{\color{function}\verb?buchberger_improved?}\verb?(ideal(F))? \\
- & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
-\cZpCR
- & \verb?[x^3, x*y + ...1100010*x^2,? \\
- & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
-\cZpLC
- & \verb?[x^3, x*y + ...111100010*x^2,? \\
- & \verb? y^2 + ...1111111001*x^2, z + ...0000000010*x]? \\
-\end{tabular}}
-
-\smallskip
-
-As we can see, some loss in precision occurs in the
-Buchberger algorithm and is avoided thanks to \ZpL.
-%Had we used \ZpFP, many padded zeroes or one would have appeared.
-
-\subsection{$p$-adic differential equations}
+\subsubsection{$p$-adic differential equations}
 
@@
 We can reach this theoretical loss in precision using \verb?ZpL?.
-We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
+We use a generic \verb?Newton_Iteration_Solver(g,h,N)?
 that applies \verb?N? steps of the Newton method for 
-$y'=g \times h(y)$ as described in \citep{LV16}.
+$y'=g \times h(y)$ as described in \cite{LV16}.
 
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
-% & \verb?N = 4? \\
-% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=50, print_mode="digits")? \\
-% & {\color{ring}\verb?Q2?}\verb?=?{\color{ring}\verb?Z2?}\verb?.fraction_field()? \\
- & \verb?S.<t> = ?{\color{constructor}\verb?PowerSeriesRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, 16)? \\
- & \verb?h = 1 + t + t^3? \\
- & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S,10)? \\
- & \verb?g = y.derivative() / h(y)? \\
- & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4)? \\
- & \verb?u[15]? \\
-\cZpCR
- & \verb?...1101? \\
-\cZpLC
- & \verb?...11011101? \\
-\end{tabular}}
-
-%\begin{tabular}{|l|l|}
-%\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
-%\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
-%\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
-%\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
-%\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
-%\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
-%\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
-%\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
-%\hfill\verb?9? & \hfill\verb?4? \\
-%\end{tabular}
-%The last line provides the loss in precision between the
-%original solution of the differential equation and its approximation
-%obtained through computation.
-%The theorical loss in precision for the coefficient of $t^{31}$
-%is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
-%loss in precision for a naive implementation is clearly more severe.
+\begin{tabular}{|l|l|}
+\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
+\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
+\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
+\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
+\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
+\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
+\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
+\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
+\hfill\verb?9? & \hfill\verb?4? \\
+\end{tabular}
+The theorical loss in precision for the coefficient of $t^{31}$
+is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
+loss in precision for a naive implementation is clearly more severe.
 
@@
 $E: \: y^2=x^3+1$
-and $E': \: y^2=+ (m^2-1)x^2/4 + m^6$ are isogeneous
+and $E': \: y^2= x^3+ (m^2-1)x^2/4 + m^6$ are isogeneous
 over $\mathbb{Q}$ if  $m$ is an odd integer.
@@
 over $\mathbb{Q}_3$ with lattice precision.
-We use an initial precision of 50 digits
-and some generic implementation of the Newton
-method of \cite{Lercier-Sirvent:08}.
+Any generic implementation of the Newton
+method of \cite{Lercier-Sirvent:08} is enough.
 
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & \verb?R = Qp(3,50)? \\
- & \verb?A.<x> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(R)? \\
- & \verb?m = 67? \\
- & \verb?g = 1 + (1/4)*x^2 + x^6? \\
- & \verb?h = 1 + ((m^2)/4)*x^2 + m^6*x^6? \\
- & \verb?y = Lercier_Sirvent(g, h, 6)? \\
- & \verb?y[31]? \\
- & \verb?u = Newton_Iteration_Solver(g, h, N)? \\  % ???
- & \verb?u[15]? \\
-\cZpCR
- & \verb?(467995903278061284942799/3^14 + O(3^36))*x^31? \\
-\cZpLC
- & \verb?(467995903278061284942799/3^14 + O(3^36))*x^31? \\
-\end{tabular}}
-
-\smallskip
+\begin{tabular}{|l|}
+\verb?    R = QpL(3)? \\
+\verb?    A.<x> = PolynomialRing(R)?\\
+\verb?    m=67?\\
+\verb?    g=1+(1/4)*x^2+x^6?\\
+\verb?    h=1+((m^2)/4)*x^2+m^6* x^6?\\
+\verb?    y=Lercier_Sirvent(g,h,64)?\\
+\verb?    y[31]?\\
+\verb?(467995903278061284942799/3^14 + O(3^36))*x^31?\\
+\end{tabular}
 
@@
 as for the coefficient in $x^{31}$: \textbf{no loss
-occured}. This happens for \ZpCR as for \ZpLC,
-and thus, \ZpCR seems to attain the optimal precision.
-consequently, those simple manipulations
+occured}.
+Thus, those simple manipulations
 hint at a different behaviour for the loss 

### 2018-02-02 ZpL/ZpL.tex

--- 2018-02-01 ZpL/ZpL.tex
@@
 
-\begin{itemize}
-\item Interval arithmetic, floating point arithmetic
-\item short introduction of the precision lemma and lattices
-\item goal: tracking precision using this theory
-\item short presentation of the package
-\item plan of the article
-\end{itemize}
+When computing with real and $p$-adic fields, exact results are usually impossible, since most elements have infinite decimal or $p$-adic expansions.
+Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.
+In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \todo{include citations}.
+The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.
+
+The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.
+At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \todo{add reference to floating point arithmetic}.
+For example, the representable numbers in $53$-bit floating point arithmetic are those of the form $m \cdot 2^e$ \todo{describe valid $m$ and $e$}.
+Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.
+Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.
+
+If provably correct results are desired, interval arithmetic provides an alternative to floating point.
+Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.
+This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
+Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision \todo{include reference to SOMOS section}.
+
+For better control of precision, we may use ellipsoids rather than balls.
+This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.
+For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.
+From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Lemma \todo{add number} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.
+
+In this paper, we describe an implementation of this idea in Sage \todo{cite Sage}.  \todo{Add short description of package features.}
+
+\todo{Plan of article}
 
@@
 that applies \verb?N? steps of the Newton method for 
-$y'=g \times h(y)$ as described in \citep{LV16}.
+$y'=g \times h(y)$ as described in \cite{LV16}.
 

### 2018-02-02 ZpL/ZpL.tex
Bibliography

--- 2018-02-01 ZpL/ZpL.tex
@@
 implements a sharp tracking of precision on $p$-adic numbers, following 
-the theory of ultrametric precision introduced in \cite{padicprec}. The 
+the theory of ultrametric precision introduced in \cite{caruso-roe-vaccon:14a}. The 
 underlying algorithms are mostly based on Automatic Differentiation 
@@
 natural example is that of the computation of GCD, whose stability has 
-been studied in \cite{caruso-2017}.
+been studied in \cite{caruso:2017}.
 
@@
 The main theoretical result on which our package is based in the 
-ultrametric precision theory developed in \cite{padicprec}, which 
+ultrametric precision theory developed in \cite{caruso-roe-vaccon:14a}, which 
 suggests to track precision \emph{via} lattices and differential 
 computations. For this reason, our approach is very inspired by 
-Automatic Differentiation techniques \cite{} and our implementation 
-follows the usual Operator Overloading strategy \cite{}.
+Automatic Differentiation techniques \cite{rall:1981} and our implementation 
+follows the usual Operator Overloading strategy.
 We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
@@
 \S \ref{ssec:preclemma} contains a brief overview on the theory of
-$p$-adic precision of \cite{padicprec}. 
+$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
 In the next two subsections, we explain in more details how \ZpLC and 
@@
 use of lattices to represent the precision of elements in 
-$K$-vector spaces.  It contrasts with the
+$\Qp$-vector spaces.  It contrasts with the
 \emph{coordinate-wise method} (of \textit{e.g.}
@@
 
-Consider a finite 
-dimensional normed vector space $E$ 
-defined over $K$. 
-We use the notation $\Vert \cdot \Vert_E$ for the norm 
-on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
-closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
-E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
-Because of ultrametricity, the balls $B^{\phantom 
--}_E(r)$ and $B^-_E(r)$ are examples of lattices. 
-Lattices can be thought of as special neighborhoods of $0,$
- and therefore are good 
-candidates to model precision data. Moreover, as revealed in 
-\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
-differentiable maps:
+Consider a finite dimensional normed vector space $E$ defined over 
+$\Qp$. We use the notation $\Vert \cdot \Vert_E$ for the norm on $E$ and 
+$B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. closed) 
+ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
+E$ is a sub-$\Zp$-module which generates $E$ over $\Qp$. Because of 
+ultrametricity, the balls $B^{\phantom -}_E(r)$ and $B^-_E(r)$ are 
+examples of lattices. Lattices can be thought of as special 
+neighborhoods of $0$, and therefore are good candidates to model 
+precision data. Moreover, as revealed in \cite{caruso-roe-vaccon:14a}, 
+they behave quite well under (strictly) differentiable maps:
 
@@
 \label{prop:precision}
-Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
+Let $E$ and $F$ be two finite dimensional normed vector spaces over $\Qp$ 
 and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
@@
 %\emph{integral polynomial} if it is given by multivariate polynomial 
-%functions with coefficients in $\O_K$ in any (equivalently all) system 
-%of coordinates associated to a $\OK$-basis of $B_E(1)$.
+%functions with coefficients in $\Zp$ in any (equivalently all) system 
+%of coordinates associated to a $\Zp$-basis of $B_E(1)$.
 %
@@
 represented by the letter $t$. Rigorously, it is defined as follows: 
-$t=0$ when the $p$-adic ring $\ZpLC(\:\cdots)$ is created and increases by 
-$1$ each time a variable is created, deleted\footnote{The deletion can
-be explicit (through a call to the \texttt{del} operator) or implicit
-(handled by the garbage collector).} or updated.
+$t=0$ when the $p$-adic ring $\ZpLC(\:\cdots)$ or $\ZpLF(\:\cdots)$ is 
+created and increases by $1$ each time a variable is created, 
+deleted\footnote{The deletion can be explicit (through a call to the 
+\texttt{del} operator) or implicit (handled by the garbage collector).} 
+or updated.
 
@@
 numbers by Floating Point $p$-adic numbers (at some given precision) 
-\cite{}.
+\cite{course-padic}.
 
@@
 \ref{prop:precision} in full generality seems to be difficult, though it 
-can be done by hand for many particular examples \cite{}.
+can be done by hand for many particular examples
+\cite{caruso-roe-vaccon:14a, caruso:2017, LV16}.
 
@@
 %
-The phenomenon observed above is formalized by the notion of
-diffused digits of precision introduced in \cite{}. We recall
-briefly its definition.
+The phenomenon observed above is formalized by the notion of diffused
+digits of precision introduced in \cite{caruso-roe-vaccon:15}.
+We recall briefly its definition.
 
@@
 that the cost becomes $O\big(\coind_t(\ttw)^2\big)$. This does not look 
-nice \emph{a priori}. However, the principle of temporal locality \cite{}
+nice \emph{a priori}. However, the principle of temporal locality 
+\cite{denning:2005}
 asserts that $\cohgt_t(\ttw)$ tends to be small in general: destroyed 
@@
 actual loss in precision for a naive implemenationin \verb?Zp(p)?.
-We can reach this theoretical loss in precision using \verb?ZpL?.
+We can reach this theoretical loss in precision using \ZpL.
 We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
@@
  & \verb?g = y.derivative() / h(y)? \\
- & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4)? \\
- & \verb?u[15]? \\
+ & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4); u[15]? \\
 \cZpCR

### 2018-02-03 ZpL/ZpL.tex
Fixing small typos

--- 2018-02-02 ZpL/ZpL.tex
@@
 $DP$ is $DQ$ is $D$.  However, we observe that we do not always get this 
-expected result. More precisely, \ZpCR often founds a gcd of higher 
+expected result. More precisely, \ZpCR often finds a gcd of higher 
 degree. Indeed the Euclidean algorithm stops prematurely because the 
@@
 closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
-E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
+E$ is a sub-$\OK$-module which generates $E$ over $K$. 
 Because of ultrametricity, the balls $B^{\phantom 
@@
 exposition.
-We emphasize that polynomials, rational fractions (outside of their poles)
-and converging power series are strictly differentiable.
+We emphasize that polynomials, rational functions (away from their poles)
+and convergent power series are strictly differentiable.
 
@@
 \Qp^{\calV_t}$; it is a finite dimensional vector space over $\Qp$ which 
-should be thought as the set of all possible values that can be taken by 
+should be thought of as the set of all possible values that can be taken by 
 the variables in $\calV_t$. For $\ttv \in \calV_t$, let $e_\ttv \in
-\Qp^{\calV_t}$ be the vector whose all coordinates vanished except 
-that at position $\ttv$ which takes the value $1$. The family 
+\Qp^{\calV_t}$ be the vector whose coordinates all vanish except 
+at position $\ttv$ which takes the value $1$. The family 
 $(e_\ttv)_{\ttv \in \calV_t}$ is obviously a basis of $E_t$; we will 
@@
 where $N_{t+1}$ is an integer, the so-called \emph{cap}.
-Alternatively, one may introduce the application:
+Alternatively, one may introduce the map:
 \begin{equation}
@@
 
-Finally, we need to precise which value is set to the newly created 
+Finally, we need to explain which value is set to the newly created 
 variable $\ttw$. We observe that it cannot be exactly $f(x_1, \ldots, 
@@
 We assume first that the newly created variable is defined through an
-affectation of the form:
+statement of the form:
 ``\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?''.
@@
 In the first case, we perform row operations in order to replace the 
-couple $(x,y)$ by $(d, 0)$ where $d$ is element of smallest valuation 
-among $x$ and $y$. Observe that $y$ is necessarily nonzero in this case, 
+pair $(x,y)$ by $(d, 0)$ where $d$ is an element of valuation 
+$\min(v_p(x), v_p(y))$. Observe that $y$ is necessarily nonzero in this case, 
 so that $d$ does not vanish as well. After this operation, we move to
@@
 
-When the variables are ``independant'', the precision lattice is
+When the variables are ``independent'', the precision lattice is
 split and the method {\color{method}\verb?precision_lattice?} outputs a diagonal
@@
 It is convenient to introduce a total order on $\calV_t$: for $\ttv, 
-\ttw \in \calV_t$, we convene that $\ttv <_t \ttw$ is \ttv was created 
+\ttw \in \calV_t$, we say that $\ttv <_t \ttw$ if \ttv was created 
 before~\ttw. By construction, the columns of the matrix $M_t$ are 
@@
 naive Euclidean algorithm (of quadratic complexity). We can observe 
-that, for both of them, the memory occupation is roughly equal to the 
+that, for both of them, the memory usage is roughly equal to the 
 square root of the complexity.
@@
 The multiplication of many matrices has been studied
-in \citep{caruso-roe-vaccon:15} and is known to often 
+in \cite{caruso-roe-vaccon:15} and is known to often 
 provide an heavy diffused digits phenomenon.
@@
 
-We can solve numerically 
+We can numerically solve 
 the corresponding differential equation

### 2018-02-03 ZpL/ZpL.tex

@@
 
-Consider a finite 
-dimensional normed vector space $E$ 
-defined over $K$. 
-We use the notation $\Vert \cdot \Vert_E$ for the norm 
-on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
-closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
-E$ is a sub-$\OK$-module which generates $E$ over $K$. 
-Because of ultrametricity, the balls $B^{\phantom 
--}_E(r)$ and $B^-_E(r)$ are examples of lattices. 
-Lattices can be thought of as special neighborhoods of $0,$
- and therefore are good 
-candidates to model precision data. Moreover, as revealed in 
-\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
-differentiable maps:
+Consider a finite dimensional normed vector space $E$ defined over 
+$\Qp$. We use the notation $\Vert \cdot \Vert_E$ for the norm on $E$ and 
+$B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. closed) 
+ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
+E$ is a sub-$\Zp$-module which generates $E$ over $\Qp$. Because of 
+ultrametricity, the balls $B^{\phantom -}_E(r)$ and $B^-_E(r)$ are 
+examples of lattices. Lattices can be thought of as special 
+neighborhoods of $0$, and therefore are good candidates to model 
+precision data. Moreover, as revealed in \cite{caruso-roe-vaccon:14a}, 
+they behave quite well under (strictly) differentiable maps:
 

### 2018-02-03 ZpL/ZpL.tex
Working on intro and bibliographyy

@@
 Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.
-In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \todo{include citations}.
+In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15,caruso-roe-vaccon:16,caruso-roe-vaccon:17}.
 The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.
@@
 The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.
-At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \todo{add reference to floating point arithmetic}.
-For example, the representable numbers in $53$-bit floating point arithmetic are those of the form $m \cdot 2^e$ \todo{describe valid $m$ and $e$}.
+At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \cite{ieee754:2008}.
 Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.
@@
 This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
-Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision \todo{include reference to SOMOS section}.
+Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision, (see Section \ref{SOMOS-4} for example).
 
@@
 For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.
-From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Lemma \todo{add number} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.
-
-In this paper, we describe an implementation of this idea in Sage \todo{cite Sage}.  \todo{Add short description of package features.}
+From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Proposition \ref{prop:precision} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.
+
+In this paper, we describe an implementation of this idea in \sage \cite{sage}.  \todo{Add short description of package features.}
 

### 2018-02-03 ZpL/ZpL.tex
More comments in Section 2

@@
 \newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
+\newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
+\newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}
 
@@
 
-We begin our tour of the features of the \ZpL package
-with some basic arithmetic computations.
-The first step is the definition of the $p$-adic ring or field.
+The first step is to define the parents, that are the rings of $p$-adic
+numbers we will work with.
 
@@
 \cIn
- & {\color{ring}\verb?Z2?}\verb? = Zp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
- & {\color{ring}\verb?Q2?}\verb? = Qp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
+ & {\color{ring}\verb?Z2?}\verb? = ?{\color{output}\verb?ZpXX?}\verb?(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
+ & {\color{ring}\verb?Q2?}\verb? = ?{\color{output}\verb?QpXX?}\verb?(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
 \end{tabular}}
@@
 \noindent
-It means that we will work in various implementation of the 
-ring of $2$-adic integers with $20$ digits of precision (which
-is the default).
-The rings {\color{ring}\verb?Z3?} and {\color{ring}\verb?Q3?} are defined similarly.
-
-We show the results for \ZpCR%, \ZpFP 
-and \ZpLC.
+\ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
+The first one \ZpCR is the usual constructor for $p$-adic parents in 
+\sage. It tracks precision using interval arithmetics. On the contrary
+\ZpLC and \ZpLF are provided by our package. In the sequel, we will
+compare the results output with the three 
 Results for \ZpLF are only shown when they differ from \ZpLC.
@@
 
-We first take some random element $x$.
-\todo{Should we say a word about the function \texttt{random\_element}?}
+We begin our tour of the features of the \ZpL package with some basic 
+arithmetic computations.
+We first pick some random element $x$.
+The function {\color{function}\verb?random_element?} is designed so
+that it guarantees that the picked random element is the same for
+each constructor \ZpCR, \ZpLC and \ZpLF.
 
@@
 
-Multiplication by $3$ is a shift on the digits, whether it 
-is done by addition or multiplication.
+\noindent
+Multiplication by $p$ (here $3$) is a shift on the digits and thus
+leads to a gain of one digit in absolute precision.
+On the example below, we observe that when this multiplication is
+split into several steps, \ZpCR does not see the gain of precision
+while \ZpLC and \ZpLF do.
 
@@
 \noindent
-\todo{Add a short explanation?}
-
 The same phenomenon occurs for multiplication.
@@
 \noindent
-No loss in precision appears in \ZpLC, contrary to \ZpCR.
-%We remark that \ZpFP also provides all the correct digits.
+Again \ZpCR does not output the optimal precision when the computation
+is split into several steps whereas \ZpLC and \ZpLF do.
+Actually, these toy examples illustrate quite common situations which 
+often occur during the execution of many algorithms. For this reason,
+interval arithmetics often overestimates the losses of precision. 
+Roughly speaking, the aim of our package is to ``fix this misfeature''.
+In the next subsections, we present a bunch of examples showing the
+benefit of \ZpL in various contexts.
 
@@
 
-In linear algebra, determinant and characteristic polynomial are 
-notoriously hard to compute: see \cite{caruso-roe-vaccon:15, 
-caruso-roe-vaccon:17}. We illustrate this with the following example 
-(using a division-free algorithm for the computation of the 
-characteristic polynomial) :
+In linear algebra, characteristic polynomials are notoriously hard to 
+compute: see \cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We 
+illustrate this with the following example (using a division-free 
+algorithm for the computation of the characteristic polynomial) :
 
@@
  & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
+ & \verb?M.?{\color{method}\verb?determinant?}\verb?()? \\
+\cZpCR
+ & \verb?...010000010? \\
+\cZpLC
+ & \verb?...010000010? \\
+\cIn
  & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
@@
 
+\smallskip
+
+\noindent
+We observe that \ZpLC can guarantee $4$ more digits on the $x$ 
+coefficient. Moreover, it recovers the correct precision on the
+constant coefficient (which is the determinant) whereas \ZpCR is 
+confused and cannot even certify that it does not vanish.
+
 \subsection{Commutative algebra}
 
-Our package can be applied to computation over $p$-adic polynomials. A 
+Our package can be applied to computation with $p$-adic polynomials. A 
 natural example is that of the computation of GCD, whose stability has 
@@
 With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
-$DP$ is $DQ$ is $D$.  However, we observe that we do not always get this 
-expected result. More precisely, \ZpCR often finds a gcd of higher 
-degree. Indeed the Euclidean algorithm stops prematurely because the 
-software believes that some early remainder vanishes due to the lack of 
-precision.
+$DP$ is $DQ$ is $D$.  However, we observe that \ZpCR output a quite 
+different result. The point is that, in the \ZpCR case, Euclidean 
+algorithm stops prematurely because the test \verb?B != 0? fails 
+too early due to the lack of precision.
 %The situation with ZpFP is the exact opposite:

### 2018-02-03 ZpL/ZpL.tex
Reorganization of examples (all in Section 2)

@@
 This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
-Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision, (see Section \ref{SOMOS-4} for example).
+Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).
 
@@
 
-\section{Examples}
+\section{Short demonstration}
+\label{sec:demo}
 
@@
 
+\medskip
+
+\noindent \textbf{SOMOS 4.}
+%
+A first example is the SOMOS-4 sequence. It is defined by the recurrence: 
+$$u_{n+4} = \frac{u_{n+1} u_{n+3} + u_{n+2}^2}{u_n}$$
+and is known for its high numerical instability (see 
+\cite{caruso-roe-vaccon:14a}). Nevertheless, the \ZpL package saves 
+precision even when using a generic unstable implementation of the SOMOS 
+iteration.
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+ & {\color{keyword}\verb?def?}\verb? ?{\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, n):? \\
+ & \verb?    a, b, c, d = u0, u1, u2, u3? \\
+ & \verb?    ?{\color{keyword}\verb?for?}\verb? _ in ?{\color{keyword}\verb?range?}\verb?(4, n+1):? \\
+ & \verb?        a, b, c, d = b, c, d, (b*d + c*c) / a? \\
+ & \verb?    ?{\color{keyword}\verb?return?}\verb? d? \\
+\cIn
+ & \verb?u0 = u1 = u2 = ?{\color{ring}\verb?Z2?}\verb?(1,15); u3 = ?{\color{ring}\verb?Z2?}\verb?(3,15)? \\
+ & {\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, 18)? \\
+\cZpCR
+ & \verb?...11? \\
+%\cZpFP
+% & \verb?...110111001000100100000000000111? \\
+\cZpLC
+ & \verb?...100000000000111? \\
+\cIn
+ & {\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, 100)? \\
+\cZpCR
+ & {\color{error}\verb?PrecisionError?}\verb?: cannot divide by something?\\
+ &\verb?indistinguishable from zero.? \\
+%\cZpFP
+% & \verb?...110000101000011001001001110001? \\
+\cZpLC
+ & \verb?...001001001110001? \\
+\end{tabular}}
+
+%
+%\begin{tabular}{|l|l|}
+%\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
+%\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
+%\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
+%\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
+%\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
+%\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
+%\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
+%\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
+%\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
+%\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
+%\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
+%\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
+%\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
+%\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
+%\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
+%\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
+%\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
+%\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
+%\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
+%\hfill\verb?...11? & \hfill\verb?...0000000111? \\
+%\hfill\verb?...10? & \hfill\verb?...1010001110? \\
+%\hfill\verb?...01? & \hfill\verb?...1110110001? \\
+%\hfill\verb?...01? & \hfill\verb?...011110111001? \\
+%\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
+%\hfill\verb?...1? & \hfill\verb?...11111000011? \\
+%\end{tabular}
+%
+%\begin{itemize}
+%\item SOMOS sequence
+%\item many multiplications of matrices
+%\item Dodgson algorithm
+%\item subresultants
+%\item Grobner basis
+%\item $p$-adic differential equations
+%\end{itemize}
+
 \subsection{Linear algebra}
 
-In linear algebra, characteristic polynomials are notoriously hard to 
-compute: see \cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We 
-illustrate this with the following example (using a division-free 
-algorithm for the computation of the characteristic polynomial) :
+Many generic algorithms of linear algebra lead to quite important 
+instability when they are used with $p$-adic numbers. In many cases, 
+our package \ZpL rubs this instability without having to change the 
+algorithm, nor the implementation.
+
+\medskip
+
+\noindent \textbf{Matrix multiplication.}
+%
+As revealed in \cite{caruso-roe-vaccon:15}, a first simple example where 
+this phenomenon appears is simply matrix multiplication.
+This might be surprising because no division occurs in this situation.
+Observe nevertheless the difference between \ZpCR and \ZpLC.
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=100)? \\
+ & \verb?MS = ?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Z2?}\verb?,2)? \\
+ & \verb?M = ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
+ & \verb?for _ in range(25):? \\
+ & \verb?    M *= ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
+ & \verb?M? \\
+\cZpCR
+ & \verb?[0 0]? \\
+ & \verb?[0 0]? \\
+\cZpLC
+ & \verb?[...100000000000   ...1000000000]? \\
+ & \verb?[   ...010000000     ...00100000]? \\
+\end{tabular}}
+
+%\ZpL can count the number of diffused digits.
+%\begin{tabular}{rl}
+%\cIn
+% & {\color{ring}\verb?Z2?}\verb?.precision().?{\color{method}\verb?number_of_diffused_digits?}\verb?()? \\
+%\ZpLC
+% & \verb?604? \\
+%\end{tabular}
+
+\smallskip
+
+\noindent
+On the aforementioned example, we notice that \ZpCR is unable to decide 
+whether the product vanishes or not. Having good estimates on the 
+precision is therefore very important in such situations.
+
+\medskip
+
+\noindent \textbf{Characteristic polynomials.}
+%
+Characteristic polynomials are notoriously hard to compute 
+\cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We illustrate this 
+with the following example (using the default algorithm of \sage for the 
+computation of the characteristic polynomial, which is a division free 
+algorithm in this setting) :
 
@@
 
-Our package can be applied to computation with $p$-adic polynomials. A 
-natural example is that of the computation of GCD, whose stability has 
+Our package can be applied to computation with $p$-adic polynomials.
+
+\medskip
+
+\noindent \textbf{Euclidean algorithm.}
+%
+A natural example is that of the computation of GCD, whose stability has 
 been studied in \cite{caruso:2017}.
-
-A naive implementation of the Euclidean algorithm
-can produce different behaviour depending 
-on the type of implementation of the field
-of $p$-adic coefficients.
+A naive implementation of the Euclidean algorithm can produce different 
+behaviour depending on the type of implementation of the field of 
+$p$-adic coefficients.
 
@@
 
-\section{Presentation of the package}
+\medskip
+
+\noindent \textbf{Gröbner bases.}
+%
+Our package can be applied on complex computations like that of Gröbner 
+bases using generic Gröbner bases algorithms.
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{@{}rl}
+\cIn
+% & {\color{ring}\verb?Q2?}\verb? = Qp(2, prec=50, print_mode="digits")? \\
+ & \verb?R.<x,y,z> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, order=?{\color{string}\verb?'invlex'?}\verb?)? \\
+ & \verb?F = [ ?{\color{ring}\verb?Q2?}\verb?(2,10)*x + ?{\color{ring}\verb?Q2?}\verb?(1,10)*z, ? \\
+ & \verb?      ?{\color{ring}\verb?Q2?}\verb?(1,10)*x^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y^2 - ?{\color{ring}\verb?Q2?}\verb?(2,10)*z^2,? \\
+ & \verb?      ?{\color{ring}\verb?Q2?}\verb?(4,10)*y^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y*z + ?{\color{ring}\verb?Q2?}\verb?(8,10)*z^2 ]? \\
+\cIn
+ & {\color{keyword}\verb?from?}\verb? sage.rings.polynomial.toy_buchberger\?\\
+ & \verb?    ?{\color{keyword}\verb?import?}\verb? ?{\color{function}\verb?buchberger_improved?} \\
+ & \verb?g = ?{\color{function}\verb?buchberger_improved?}\verb?(ideal(F))? \\
+ & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
+\cZpCR
+ & \verb?[x^3, x*y + ...1100010*x^2,? \\
+ & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
+\cZpLC
+ & \verb?[x^3, x*y + ...111100010*x^2,? \\
+ & \verb? y^2 + ...1111111001*x^2, z + ...0000000010*x]? \\
+\end{tabular}}
+
+\smallskip
+
+\noindent
+As we can see, some loss in precision occurs in the
+Buchberger algorithm and is avoided thanks to \ZpL.
+%Had we used \ZpFP, many padded zeroes or one would have appeared.
+
+\subsection{$p$-adic differential equations}
+
+In \cite{LV16}, the behaviour of the precision when solving
+$p$-adic differential equations with separation of 
+variables has been studied.
+The authors have investigated the gap when that appears
+when applying a Newton-method solver between
+the theoretic loss in precision and the 
+actual loss in precision for a naive implemenationin \verb?Zp(p)?.
+We can reach this theoretical loss in precision using \ZpL.
+We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
+that applies \verb?N? steps of the Newton method for 
+$y'=g \times h(y)$ as described in \cite{LV16}.
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn
+% & \verb?N = 4? \\
+% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=50, print_mode="digits")? \\
+% & {\color{ring}\verb?Q2?}\verb?=?{\color{ring}\verb?Z2?}\verb?.fraction_field()? \\
+ & \verb?S.<t> = ?{\color{constructor}\verb?PowerSeriesRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, 16)? \\
+ & \verb?h = 1 + t + t^3? \\
+ & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S,10)? \\
+ & \verb?g = y.derivative() / h(y)? \\
+ & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4); u[15]? \\
+\cZpCR
+ & \verb?...1101? \\
+\cZpLC
+ & \verb?...11011101? \\
+\end{tabular}}
+
+%\begin{tabular}{|l|l|}
+%\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
+%\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
+%\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
+%\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
+%\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
+%\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
+%\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
+%\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
+%\hfill\verb?9? & \hfill\verb?4? \\
+%\end{tabular}
+%The last line provides the loss in precision between the
+%original solution of the differential equation and its approximation
+%obtained through computation.
+%The theorical loss in precision for the coefficient of $t^{31}$
+%is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
+%loss in precision for a naive implementation is clearly more severe.
+
+\section{What's behind the scene?}
 
@@
 
-\section{More examples}
-
-\subsection{SOMOS-4}
-
-The SOMOS-4 sequence is a well-known example
-of a computation with a large gap between
-the theoretical loss in precision
-and that of a naive computation (see \cite{caruso-roe-vaccon:14a}).
-We show here that precision can be saved
-even when using a generic implementation
-of the SOMOS iteration.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
- & {\color{keyword}\verb?def?}\verb? ?{\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, n):? \\
- & \verb?    a, b, c, d = u0, u1, u2, u3? \\
- & \verb?    ?{\color{keyword}\verb?for?}\verb? _ in ?{\color{keyword}\verb?range?}\verb?(4, n+1):? \\
- & \verb?        a, b, c, d = b, c, d, (b*d + c*c) / a? \\
- & \verb?    ?{\color{keyword}\verb?return?}\verb? d? \\
-\cIn
- & \verb?u0 = u1 = u2 = ?{\color{ring}\verb?Z2?}\verb?(1,15); u3 = ?{\color{ring}\verb?Z2?}\verb?(3,15)? \\
- & {\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, 18)? \\
-\cZpCR
- & \verb?...11? \\
-%\cZpFP
-% & \verb?...110111001000100100000000000111? \\
-\cZpLC
- & \verb?...100000000000111? \\
-\cIn
- & {\color{function}\verb?somos4?}\verb?(u0, u1, u2, u3, 100)? \\
-\cZpCR
- & {\color{error}\verb?PrecisionError?}\verb?: cannot divide by something?\\
- &\verb?indistinguishable from zero.? \\
-%\cZpFP
-% & \verb?...110000101000011001001001110001? \\
-\cZpLC
- & \verb?...001001001110001? \\
-\end{tabular}}
-
-%
-%\begin{tabular}{|l|l|}
-%\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
-%\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
-%\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
-%\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
-%\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
-%\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
-%\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
-%\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
-%\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
-%\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
-%\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
-%\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
-%\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
-%\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
-%\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
-%\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
-%\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
-%\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
-%\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
-%\hfill\verb?...11? & \hfill\verb?...0000000111? \\
-%\hfill\verb?...10? & \hfill\verb?...1010001110? \\
-%\hfill\verb?...01? & \hfill\verb?...1110110001? \\
-%\hfill\verb?...01? & \hfill\verb?...011110111001? \\
-%\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
-%\hfill\verb?...1? & \hfill\verb?...11111000011? \\
-%\end{tabular}
-%
-%\begin{itemize}
-%\item SOMOS sequence
-%\item many multiplications of matrices
-%\item Dodgson algorithm
-%\item subresultants
-%\item Grobner basis
-%\item $p$-adic differential equations
-%\end{itemize}
-
-\subsection{Matrix multiplication}
-The multiplication of many matrices has been studied
-in \cite{caruso-roe-vaccon:15} and is known to often 
-provide an heavy diffused digits phenomenon.
-Here is a simple example.
-% The result for \ZpFP has not been
-% added, but its first digits coincide with that of \ZpLC.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
-% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=100)? \\
- & \verb?MS = ?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Z2?}\verb?,2)? \\
- & \verb?M = ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
- & \verb?for _ in range(25):? \\
- & \verb?    M *= ?{\color{function}\verb?random_element?}\verb?(MS, prec=5)? \\
- & \verb?M? \\
-\cZpCR
- & \verb?[0 0]? \\
- & \verb?[0 0]? \\
-\cZpLC
- & \verb?[...100000000000   ...1000000000]? \\
- & \verb?[   ...010000000     ...00100000]? \\
-\end{tabular}}
-
-%\ZpL can count the number of diffused digits.
-%\begin{tabular}{rl}
-%\cIn
-% & {\color{ring}\verb?Z2?}\verb?.precision().?{\color{method}\verb?number_of_diffused_digits?}\verb?()? \\
-%\ZpLC
-% & \verb?604? \\
-%\end{tabular}
-
-\subsection{Gröbner bases}
-Our package can be applied 
-on complex computations like that of
-Gröbner bases using generic
-Gröbner bases algorithms.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{@{}rl}
-\cIn
-% & {\color{ring}\verb?Q2?}\verb? = Qp(2, prec=50, print_mode="digits")? \\
- & \verb?R.<x,y,z> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, order=?{\color{string}\verb?'invlex'?}\verb?)? \\
- & \verb?F = [ ?{\color{ring}\verb?Q2?}\verb?(2,10)*x + ?{\color{ring}\verb?Q2?}\verb?(1,10)*z, ? \\
- & \verb?      ?{\color{ring}\verb?Q2?}\verb?(1,10)*x^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y^2 - ?{\color{ring}\verb?Q2?}\verb?(2,10)*z^2,? \\
- & \verb?      ?{\color{ring}\verb?Q2?}\verb?(4,10)*y^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y*z + ?{\color{ring}\verb?Q2?}\verb?(8,10)*z^2 ]? \\
-\cIn
- & {\color{keyword}\verb?from?}\verb? sage.rings.polynomial.toy_buchberger\?\\
- & \verb?    ?{\color{keyword}\verb?import?}\verb? ?{\color{function}\verb?buchberger_improved?} \\
- & \verb?g = ?{\color{function}\verb?buchberger_improved?}\verb?(ideal(F))? \\
- & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
-\cZpCR
- & \verb?[x^3, x*y + ...1100010*x^2,? \\
- & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
-\cZpLC
- & \verb?[x^3, x*y + ...111100010*x^2,? \\
- & \verb? y^2 + ...1111111001*x^2, z + ...0000000010*x]? \\
-\end{tabular}}
-
-\smallskip
-
-As we can see, some loss in precision occurs in the
-Buchberger algorithm and is avoided thanks to \ZpL.
-%Had we used \ZpFP, many padded zeroes or one would have appeared.
-
-\subsection{$p$-adic differential equations}
-
-In \cite{LV16}, the behaviour of the precision when solving
-$p$-adic differential equations with separation of 
-variables has been studied.
-The authors have investigated the gap when that appears
-when applying a Newton-method solver between
-the theoretic loss in precision and the 
-actual loss in precision for a naive implemenationin \verb?Zp(p)?.
-We can reach this theoretical loss in precision using \ZpL.
-We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
-that applies \verb?N? steps of the Newton method for 
-$y'=g \times h(y)$ as described in \cite{LV16}.
-
-\smallskip
-
-{\noindent \small
-\begin{tabular}{rl}
-\cIn
-% & \verb?N = 4? \\
-% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=50, print_mode="digits")? \\
-% & {\color{ring}\verb?Q2?}\verb?=?{\color{ring}\verb?Z2?}\verb?.fraction_field()? \\
- & \verb?S.<t> = ?{\color{constructor}\verb?PowerSeriesRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, 16)? \\
- & \verb?h = 1 + t + t^3? \\
- & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S,10)? \\
- & \verb?g = y.derivative() / h(y)? \\
- & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4); u[15]? \\
-\cZpCR
- & \verb?...1101? \\
-\cZpLC
- & \verb?...11011101? \\
-\end{tabular}}
-
-%\begin{tabular}{|l|l|}
-%\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
-%\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
-%\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
-%\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
-%\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
-%\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
-%\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
-%\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
-%\hfill\verb?9? & \hfill\verb?4? \\
-%\end{tabular}
-%The last line provides the loss in precision between the
-%original solution of the differential equation and its approximation
-%obtained through computation.
-%The theorical loss in precision for the coefficient of $t^{31}$
-%is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
-%loss in precision for a naive implementation is clearly more severe.
-
 \section{Conclusion}

### 2018-02-03 ZpL/ZpL.tex
Add explicity the keyword "prec" in examples

@@
 \cIn
- & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z3?}\verb?,5); x? \\
+ & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z3?}\verb?, prec=5); x? \\
 \cZpCR
@@
 \cIn
- & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,10)? \\
- & \verb?y = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,5)? \\
+ & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?, prec=10)? \\
+ & \verb?y = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?, prec=5)? \\
 \cIn
@@
 As revealed in \cite{caruso-roe-vaccon:15}, a first simple example where 
-this phenomenon appears is simply matrix multiplication.
+unstability appears is simply matrix multiplication.
 This might be surprising because no division occurs in this situation.
@@
  & \verb?h = 1 + t + t^3? \\
- & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S,10)? \\
+ & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S, prec=10)? \\
  & \verb?g = y.derivative() / h(y)? \\
@@
 the method
-\verb?is_precision_capped? (provided by our package) which returns true 
+{\color{method}\verb?is_precision_capped?} (provided by our package) which returns true 
 if $d\Phi_{t,\ttv}(K_t) = p^N \Zp$. As a consequence, when this method

### 2018-02-03 ZpL/ZpL.tex
Add a conclusion

@@
 This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
-Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).
+Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision (see Section \ref{sec:demo} for many examples).
 
@@
 if $d\Phi_{t,\ttv}(K_t) = p^N \Zp$. As a consequence, when this method
-answers false, the absolute precision computed by the software is
-likely optimal.
+answers \textsc{false}, the absolute precision computed by the software 
+is likely optimal.
 \end{rmk}
@@
 as soon as the complexity is at least quadratic in $s$, the dominant 
-term is $c^{3/2}$ and the impact on the complexity is then moderated.
+term is $c^{3/2}$ and the impact on the complexity is then limited.
 
@@
 
-\begin{itemize}
-\item In which case should we use this package? Mainly for exploration
-\item example of $p$-adic differential equations (especially for $p=2$)
-\end{itemize}
-
+The package \ZpL provides powerful tools (based on Automatic 
+Differentiation) to track precision in the $p$-adic setting. In many 
+concrete situations, it heavily outdoes usual interval arithmetics 
+which is currently widely used by Computer Algebra softwares (see
+\S \ref{sec:demo} for many examples).
+The impact of complexity is controlled but nevertheless non-negligible 
+(see \S \ref{ssec:complexity}). For this reason, it is not really 
+reasonable to believe that a fast algorithm could rely \emph{directly} 
+on the machinery proposed by \ZpL. At least for now, bringing together 
+rapidity and stability still requires a substantial human contribution 
+and a careful special study of all parameters.
+
+Nevertheless, we do believe that \ZpL can be extremely helpful to 
+whoever wants to design a fast and stable $p$-adic algorihm for a couple 
+of reasons. First, it provides solutions 
+to automatically detect which steps of a given algorithm are stable and 
+which ones are not. This way, it highlights the parts of the algorithm 
+on which the researcher has to concentrate its efforts.
+Second, recall that a classical strategy to improve stability consists 
+in working internally at higher precision. The point is that adjusting 
+at best the internal precision (\emph{i.e.} finding the good compromise 
+between efficiency and accuracy) is not an easy task in general. It 
+turns out that the knowledge of the number of diffused digits of 
+precision gives very useful hints in this direction. Indeed, as an 
+extreme example, consider the case when there is no diffused digit of 
+precision. Then, the optimal precision completely splits over the
+variables and there is no need to increase internally the precision. On 
+the contrary, when there are many diffused digits of precision, a large 
+increment is often required.
+Since \ZpL gives a direct access to the number of diffused digits of 
+precision, it could be very useful to the designer who is concerned
+with the balance between efficiency and accuracy.
+
+\begin{comment}
 \subsection{Exploration}
@@
 but not necessarily over $\mathbb{Z}_p.$ 
+\end{comment}
 

### 2018-02-03 ZpL/ZpL.tex
Comment Figure 2 in order to save space

@@
 variables has been studied.
-The authors have investigated the gap when that appears
+The authors have investigated the gap that appears
 when applying a Newton-method solver between
 the theoretic loss in precision and the 
-actual loss in precision for a naive implemenationin \verb?Zp(p)?.
+actual loss in precision for a naive implemenation in \verb?Zp(p)?.
 We can reach this theoretical loss in precision using \ZpL.
@@
 exposition.
-We emphasize that polynomials, rational functions (away from their poles)
-and convergent power series are strictly differentiable.
+%We emphasize that polynomials, rational functions (away from their poles)
+%and convergent power series are strictly differentiable.
 
@@
 $H_{t+1} = H_t + p^{\min(\text{prec}, N_{t+1})} \Zp e_\ttw$
-where the cap $N_{t+1}$ is:
-$$N_{t+1} = \min\big(\textsc{abscap},\, \textsc{relcap} + 
-v_p(\text{value})\big).$$
+for the cap
+$N_{t+1} = \min\big(\textsc{abscap},\, \textsc{relcap} + 
+v_p(\text{value})\big)$.
 The corresponding matrix $M_{t+1}$ is then given by:
@@
 %
+\begin{comment}
 \begin{figure}
@@
 1$.
+\end{comment}
+For example, while executing the Euclidean algorithm (naive 
+implementation) with two polynomials of degree $7$, most of the 
+time $\cohgt_t(\ttw) \leq 1$.
 

### 2018-02-04 ZpL/ZpL.tex
Adding summary of paper

--- 2018-02-03 ZpL/ZpL.tex
@@
   %\institution{Department of Mathematics}
-  %\city{Pittsburgh, PA, USA}
-  %\postcode{15260}
+  %\city{Cambridge, MA, USA}
+  %\postcode{02139}
 }
@@
 \begin{abstract}
-We present a new package \ZpL for the Computer Algebra system \sage. It 
+We present a new package \ZpL for the mathematical software system \sage. It 
 implements a sharp tracking of precision on $p$-adic numbers, following 
 the theory of ultrametric precision introduced in \cite{caruso-roe-vaccon:14a}. The 
-underlying algorithms are mostly based on Automatic Differentiation 
+underlying algorithms are mostly based on automatic differentiation 
 techniques. We introduce them, study their complexity and discuss our 
@@
 We illustrate the benefits of our package (in comparison with previous 
-implementations) with a large sample of examples coming from Linear 
-Algebra, Commutative Algebra and Differential Equations.
+implementations) with a large sample of examples coming from linear 
+algebra, commutative algebra and differential equations.
 \end{abstract}
@@
 
-When computing with real and $p$-adic fields, exact results are usually impossible, since most elements have infinite decimal or $p$-adic expansions.
-Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.
-In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15,caruso-roe-vaccon:16,caruso-roe-vaccon:17}.
-The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.
-
-The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.
-At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \cite{ieee754:2008}.
-Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.
-Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.
-
-If provably correct results are desired, interval arithmetic provides an alternative to floating point.
-Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.
-This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
-Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).
-
-For better control of precision, we may use ellipsoids rather than balls.
-This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.
-For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.
-From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Proposition \ref{prop:precision} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.
-
-In this paper, we describe an implementation of this idea in \sage \cite{sage}.  \todo{Add short description of package features.}
-
-\todo{Plan of article}
-
-\todo{Convention: in matrices, vectors are represented by row-vector.}
+When computing with real and $p$-adic fields, exact results are usually impossible,
+since most elements have infinite decimal or $p$-adic expansions.
+Working with these fields thus requires an analysis of how precision evolves
+through the sequence of steps involved in carrying out a computation.
+In this paper, we describe a package for computing with $p$-adic rings and fields,
+based on a series of papers by the same authors
+\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15,caruso-roe-vaccon:16,caruso-roe-vaccon:17}.
+The core of the package is a method for tracking precision using $p$-adic lattices which
+can yield dramatically more precise results, at the cost of increased runtime and memory usage.
+
+The standard method for handling precision when computing with real numbers
+is floating point arithmetic, which may also be used in $p$-adic computation.
+At a given precision level, a finite set of representable numbers are chosen,
+and arithmetic operations are defined to give a representable number
+that is close to the true result \cite{ieee754:2008}.  Floating point arithmetic
+has the benefit of efficient arithmetic operations, but users are responsible
+for tracking the precision of the results.  Numerically unstable algorithms can
+lead to very inaccurate answers \cite{higham:2002}.
+
+If provably correct results are desired, interval arithmetic provides
+an alternative to floating point.  Instead of just tracking an approximation to the answer,
+the package also tracks a radius within which the true result lies.
+This method is commonly used for $p$-adic computations since
+the ultrametric property of $p$-adic fields frequently keeps the radius small.
+Computations remain fairly efficient with this approach, but numerical instability
+can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).
+Tracking the precision of multiple variables concurrently, the set of possible
+true values associated to an inexact value takes the form of an ellipsoid
+with axes parallel to the coordinate axes.
+
+For better control of precision, we may allow arbitrary axes.  This change would have
+little utility for real numbers, since such ellipsoids are not preserved by most functions.
+For $p$-adic fields, in contrast, sufficiently small ellipsoids are mapped to other
+ellipsoids by differentiable maps.  From an algebraic perspective,
+these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space,
+and Proposition \ref{prop:precision} describes how the image of such a coset
+under a map $f$ is given exactly by applying the differential of $f$ to $H$.
+
+In this paper, we describe an implementation of this idea in \sage \cite{sage}.
+Rather than attaching a precision to each element, we store the precision of
+many elements together by tracking a precision lattice for the whole collection
+of variables.  As variables are created and destroyed, we update a matrix whose
+rows represent the vectors in the lattice.  Information about the precision
+of elements is extracted from the matrix as necessary.
+
+The article is structured as follows. In \S \ref{sec:demo} we provide a demonstration
+of the package, showing how it can provide more precise answers than the
+traditional methods for tracking $p$-adic precision.  In particular, \S \ref{ssec:elem}
+describes elementary arithmetic and the SOMOS-4 sequence, \S \ref{ssec:linalg}
+gives examples from linear algebra, \S \ref{ssec:comalg} examples using polynomials,
+and \S \ref{ssec:diffeq} examples of differential equations.
+
+In \S \ref{sec:implementation} we give more details on the implementation.
+\S \ref{ssec:preclemma} contains a brief overview on the theory of
+$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
+In the next two subsections, we explain in more details how \ZpLC and 
+\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
+automatic differentiation leading to the actual computation of the
+lattice that models the precision.
+In \S \ref{ssec:viewprec}, we explain how precision on any individual number
+can be recovered and discuss the validity
+of our results.
+The complexity overhead induced by our package is analyzed in
+\S \ref{ssec:complexity}.
+
+Finally, \S \ref{sec:conclusion} contains a discussion of how we see this
+package fitting into the existing $p$-adic implementations.  While these
+methods do introduce overhead, they are well suited to exploring
+precision behavior when designing algorithms, and can provide
+hints as to when further precision analysis would be useful.
 
@@
 The first one \ZpCR is the usual constructor for $p$-adic parents in 
-\sage. It tracks precision using interval arithmetics. On the contrary
+\sage. It tracks precision using interval arithmetic. On the contrary
 \ZpLC and \ZpLF are provided by our package. In the sequel, we will
@@
 \subsection{Elementary arithmetic}
+\label{ssec:elem}
 
@@
 leads to a gain of one digit in absolute precision.
-On the example below, we observe that when this multiplication is
+In the example below, we observe that when this multiplication is
 split into several steps, \ZpCR does not see the gain of precision
@@
 \subsection{Linear algebra}
+\label{ssec:linalg}
 
@@
 \subsection{Commutative algebra}
+\label{ssec:comalg}
 
@@
 \subsection{$p$-adic differential equations}
+\label{ssec:diffeq}
 
@@
 \section{What's behind the scene?}
+\label{sec:implementation}
 
@@
 this former is safer while the latter is faster.
-
-This Section is organized as follows.
-\S \ref{ssec:preclemma} contains a brief overview on the theory of
-$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
-In the next two subsections, we explain in more details how \ZpLC and 
-\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
-Automatic Differentiation leading to the actual computation of a
-global object modeling the precision (which is the aforementioned 
-lattice). 
-In \S \ref{ssec:viewprec}, we explain how precision on any individual number
-can be recoved from the above computations and discuss the validity
-of our results.
-The overhead complexity induced by our package is analyzed in
-\S \ref{ssec:complexity}.
 
@@
 whereas deleting a variable is free. 
-Thus when executed with the \ZpLF mecanism, the complexity of our
+Thus when executed with the \ZpLF mechanism, the complexity of our
 algorithm becomes $O(s_\inp c)$.
@@
 \section{Conclusion}
-
-\begin{itemize}
-\item In which case should we use this package? Mainly for exploration
-\item example of $p$-adic differential equations (especially for $p=2$)
-\end{itemize}
-
-\subsection{Exploration}
+\label{sec:conclusion}
+
+%\begin{itemize}
+%\item In which case should we use this package? Mainly for exploration
+%\item example of $p$-adic differential equations (especially for $p=2$)
+%\end{itemize}
+
+%\subsection{Exploration}
 

### 2018-02-04 ZpL/ZpL.tex

@@
 Computations remain fairly efficient with this approach, but numerical instability
-can still lead to dramatic losses in precision, (see Section \ref{sec:demo} for many examples).
+can still lead to dramatic losses in precision (see Section \ref{sec:demo} for many examples).
 Tracking the precision of multiple variables concurrently, the set of possible
@@
 1$.
+%For example, while executing the Euclidean algorithm (naive 
+%implementation) with two polynomials of degree $7$, most of the 
+%time $\cohgt_t(\ttw) \leq 1$.
 
@@
 
-%\begin{itemize}
-%\item In which case should we use this package? Mainly for exploration
-%\item example of $p$-adic differential equations (especially for $p=2$)
-%\end{itemize}
-
-%\subsection{Exploration}
+The package \ZpL provides powerful tools (based on Automatic 
+Differentiation) to track precision in the $p$-adic setting. In many 
+concrete situations, it heavily outdoes usual interval arithmetics 
+which is currently widely used by Computer Algebra softwares (see
+\S \ref{sec:demo} for many examples).
+The impact of complexity is controlled but nevertheless non-negligible 
+(see \S \ref{ssec:complexity}). For this reason, it is not really 
+reasonable to believe that a fast algorithm could rely \emph{directly} 
+on the machinery proposed by \ZpL. At least for now, bringing together 
+rapidity and stability still requires a substantial human contribution 
+and a careful special study of all parameters.
+
+Nevertheless, we do believe that \ZpL can be extremely helpful to 
+whoever wants to design a fast and stable $p$-adic algorihm for a couple 
+of reasons. First, it provides solutions 
+to automatically detect which steps of a given algorithm are stable and 
+which ones are not. This way, it highlights the parts of the algorithm 
+on which the researcher has to concentrate its efforts.
+Second, recall that a classical strategy to improve stability consists 
+in working internally at higher precision. The point is that adjusting 
+at best the internal precision (\emph{i.e.} finding the good compromise 
+between efficiency and accuracy) is not an easy task in general. It 
+turns out that the knowledge of the number of diffused digits of 
+precision gives very useful hints in this direction. Indeed, as an 
+extreme example, consider the case when there is no diffused digit of 
+precision. Then, the optimal precision completely splits over the
+variables and there is no need to increase internally the precision. On 
+the contrary, when there are many diffused digits of precision, a large 
+increment is often required.
+Since \ZpL gives a direct access to the number of diffused digits of 
+precision, it could be very useful to the designer who is concerned
+with the balance between efficiency and accuracy.
+
+\begin{comment}
+\subsection{Exploration}
 

### 2018-02-04 ZpL/ZpL.tex
Fixing typos

@@
 
-\title{\ZpL: a p-adic precision package}
+\title{\texorpdfstring{\ZpL}{ZpL}: a p-adic precision package}
 
@@
 little utility for real numbers, since such ellipsoids are not preserved by most functions.
-For $p$-adic fields, in contrast, sufficiently small ellipsoids are mapped to other
-ellipsoids by differentiable maps.  From an algebraic perspective,
+For $p$-adic fields, in contrast, differentiable maps with surjective differential will send
+sufficiently small ellipsoids to other ellipsoids.  From an algebraic perspective,
 these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space,
@@
 
-The first step is to define the parents, that are the rings of $p$-adic
+The first step is to define the parents: the rings of $p$-adic
 numbers we will work with.
@@
 \ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
-The first one \ZpCR is the usual constructor for $p$-adic parents in 
+The first, \ZpCR, is the usual constructor for $p$-adic parents in 
 \sage. It tracks precision using interval arithmetic. On the contrary
 \ZpLC and \ZpLF are provided by our package. In the sequel, we will
-compare the results output with the three 
+compare the outputs provided by each parent. 
 Results for \ZpLF are only shown when they differ from \ZpLC.
@@
 often occur during the execution of many algorithms. For this reason,
-interval arithmetics often overestimates the losses of precision. 
+interval arithmetic often overestimates the losses of precision. 
 Roughly speaking, the aim of our package is to ``fix this misfeature''.
@@
 As revealed in \cite{caruso-roe-vaccon:15}, a first simple example where 
-unstability appears is simply matrix multiplication.
+instability appears is simply matrix multiplication.
 This might be surprising because no division occurs in this situation.
@@
 A naive implementation of the Euclidean algorithm can produce different 
-behaviour depending on the type of implementation of the field of 
+behavior depending on the type of implementation of the field of 
 $p$-adic coefficients.
@@
 
-In \cite{LV16}, the behaviour of the precision when solving
+In \cite{LV16}, the behavior of the precision when solving
 $p$-adic differential equations with separation of 
@@
 the theoretic loss in precision and the 
-actual loss in precision for a naive implemenation in \verb?Zp(p)?.
+actual loss in precision for a naive implementation in \verb?Zp(p)?.
 We can reach this theoretical loss in precision using \ZpL.
@@
 
-\section{What's behind the scene?}
+\section{Behind the scenes}
 \label{sec:implementation}
 
-In this Section, we explain how our package \ZpL works and we analyze
-its performances.
-The main theoretical result on which our package is based in the 
+In this section, we explain how our package \ZpL works and analyze
+its performance.
+The main theoretical result on which our package is based is the 
 ultrametric precision theory developed in \cite{caruso-roe-vaccon:14a}, which 
-suggests to track precision \emph{via} lattices and differential 
+suggests tracking precision \emph{via} lattices and differential 
 computations. For this reason, our approach is very inspired by 
-Automatic Differentiation techniques \cite{rall:1981} and our implementation 
-follows the usual Operator Overloading strategy.
+automatic differentiation techniques \cite{rall:1981} and our implementation 
+follows the usual operator overloading strategy.
 We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
@@
 
-In \cite{caruso-roe-vaccon:14a}, we have suggested the 
+In \cite{caruso-roe-vaccon:14a}, we suggest the 
 use of lattices to represent the precision of elements in 
-$\Qp$-vector spaces.  It contrasts with the
+$\Qp$-vector spaces.  This approach contrasts with the
 \emph{coordinate-wise method} (of \textit{e.g.}
- \verb?Zp(5)?) that is used traditionally in \sage and
+ \verb?Zp(5)?) that is traditionally used in \sage
 where the precision of an element is specified by giving the precision
@@
 
-We now explain in more details the internal mecanisms \ZpLC and \ZpLF 
+We now explain in more details the internal mechanisms \ZpLC and \ZpLF 
 use for tracking precision.
@@
 square matrix; it has one extra column. This reflects the fact that 
-$\dim E_{t+1} = \dim E_t + 1$. Rephasing this in a different language, 
+$\dim E_{t+1} = \dim E_t + 1$. Rephrasing this in a different language, 
 the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
@@
 where a cell is colored when it can contain a non-vanishing entry.
-The top part of the matrix is then already echenolized, so that we
+The top part of the matrix is then already echelonized, so that we
 only have to re-echelonize the bottom right corner whose size is
@@
 variables are often variables that were created recently. As a basic 
-exemple, variables which are local to a small piece of code (\emph{e.g.} 
+example, variables which are local to a small piece of code (\emph{e.g.} 
 a short function or a loop) do not survive for a long time. It turns
-out that this behaviour is typical in many implementations!
+out that this behavior is typical in many implementations!
 %
@@
 bottleneck (which is the case in practice). Therefore, when executed 
-with the \ZpLF mecanism, the cost of our algorithm is expected to be 
+with the \ZpLF mechanism, the cost of our algorithm is expected to be 
 roughly $O(mc)$. Going further in speculation, we might estimate the 
@@
 
-The package \ZpL provides powerful tools (based on Automatic 
-Differentiation) to track precision in the $p$-adic setting. In many 
-concrete situations, it heavily outdoes usual interval arithmetics 
-which is currently widely used by Computer Algebra softwares (see
-\S \ref{sec:demo} for many examples).
-The impact of complexity is controlled but nevertheless non-negligible 
-(see \S \ref{ssec:complexity}). For this reason, it is not really 
-reasonable to believe that a fast algorithm could rely \emph{directly} 
-on the machinery proposed by \ZpL. At least for now, bringing together 
-rapidity and stability still requires a substantial human contribution 
+The package \ZpL provides powerful tools (based on automatic 
+differentiation) to track precision in the $p$-adic setting. In many 
+concrete situations, it greatly outperforms standard interval arithmetic,
+as shown in \S \ref{sec:demo}.
+The impact on complexity is controlled but nevertheless non-negligible 
+(see \S \ref{ssec:complexity}). For this reason, it is unlikely that a fast
+algorithm will rely \emph{directly} on the machinery proposed by \ZpL,
+though it might do so for a specific part of a computation.
+At least for now, bringing together rapidity and stability
+still requires a substantial human contribution 
 and a careful special study of all parameters.
 
-Nevertheless, we do believe that \ZpL can be extremely helpful to 
-whoever wants to design a fast and stable $p$-adic algorihm for a couple 
-of reasons. First, it provides solutions 
+Nevertheless, we believe that \ZpL can be extremely helpful to 
+anyone designing a fast and stable $p$-adic algorithm for a couple 
+of reasons. First, it provides mechanisms 
 to automatically detect which steps of a given algorithm are stable and 
-which ones are not. This way, it highlights the parts of the algorithm 
-on which the researcher has to concentrate its efforts.
+which ones are not. In this way, it highlights the parts of the algorithm 
+on which the researcher has to concentrate their effort.
 Second, recall that a classical strategy to improve stability consists 
-in working internally at higher precision. The point is that adjusting 
-at best the internal precision (\emph{i.e.} finding the good compromise 
-between efficiency and accuracy) is not an easy task in general. It 
-turns out that the knowledge of the number of diffused digits of 
-precision gives very useful hints in this direction. Indeed, as an 
-extreme example, consider the case when there is no diffused digit of 
-precision. Then, the optimal precision completely splits over the
-variables and there is no need to increase internally the precision. On 
+in working internally at higher precision. Finding the internal increase
+in precision that best balances efficiency and accuracy is not an easy
+task in general. Understanding the number of diffused digits of 
+precision gives very useful hints in this direction. For example, when there are
+no diffused digits of precision then the optimal precision completely splits over the
+variables and there is no need to internally increase the precision. On 
 the contrary, when there are many diffused digits of precision, a large 
@@
 have a long history, and a first
-analysis of the behaviour of precision using
+analysis of the behavior of precision using
 lattices has been investigated in \cite{LV16}.
@@
 However, this does not encompass the case of 
-elliptic curves defined and isogeneous 
+elliptic curves defined and isogenous 
 over $\mathbb{Q},$ which can yield a differential 
@@
 $E: \: y^2=x^3+1$
-and $E': \: y^2= x^3+ (m^2-1)x^2/4 + m^6$ are isogeneous
+and $E': \: y^2= x^3+ (m^2-1)x^2/4 + m^6$ are isogenous
 over $\mathbb{Q}$ if  $m$ is an odd integer.
@@
 
-We can then see that up to $63,$ the behaviour
+We can then see that up to $63,$ the behavior
 of the loss in relative precision is the same
 as for the coefficient in $x^{31}$: \textbf{no loss
-occured}. This happens for \ZpCR as for \ZpLC,
+occurred}. This happens for \ZpCR as for \ZpLC,
 and thus, \ZpCR seems to attain the optimal precision.
 consequently, those simple manipulations
-hint at a different behaviour for the loss 
+hint at a different behavior for the loss 
 in precision in at least some special cases of

### 2018-02-04 ZpL/ZpL.tex
Small fixes

--- 2018-02-04 ZpL/ZpL.tex
@@
 \definecolor{error}{HTML}{B01010}
+\definecolor{comment}{HTML}{606060}
 
@@
   %\institution{Department of Mathematics}
-  %\city{Cambridge, MA, USA}
-  %\postcode{02139}
+  %\city{Pittsburgh, PA, USA}
+  %\postcode{15260}
 }
@@
 \begin{abstract}
-We present a new package \ZpL for the mathematical software system \sage. It 
+We present a new package \ZpL for the Computer Algebra system \sage. It 
 implements a sharp tracking of precision on $p$-adic numbers, following 
 the theory of ultrametric precision introduced in \cite{caruso-roe-vaccon:14a}. The 
-underlying algorithms are mostly based on automatic differentiation 
+underlying algorithms are mostly based on Automatic Differentiation 
 techniques. We introduce them, study their complexity and discuss our 
@@
 We illustrate the benefits of our package (in comparison with previous 
-implementations) with a large sample of examples coming from linear 
-algebra, commutative algebra and differential equations.
+implementations) with a large sample of examples coming from Linear 
+Algebra, Commutative Algebra and Differential Equations.
 \end{abstract}
@@
 
-When computing with real and $p$-adic fields, exact results are usually impossible,
-since most elements have infinite decimal or $p$-adic expansions.
-Working with these fields thus requires an analysis of how precision evolves
-through the sequence of steps involved in carrying out a computation.
-In this paper, we describe a package for computing with $p$-adic rings and fields,
-based on a series of papers by the same authors
-\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15,caruso-roe-vaccon:16,caruso-roe-vaccon:17}.
-The core of the package is a method for tracking precision using $p$-adic lattices which
-can yield dramatically more precise results, at the cost of increased runtime and memory usage.
-
-The standard method for handling precision when computing with real numbers
-is floating point arithmetic, which may also be used in $p$-adic computation.
-At a given precision level, a finite set of representable numbers are chosen,
-and arithmetic operations are defined to give a representable number
-that is close to the true result \cite{ieee754:2008}.  Floating point arithmetic
-has the benefit of efficient arithmetic operations, but users are responsible
-for tracking the precision of the results.  Numerically unstable algorithms can
-lead to very inaccurate answers \cite{higham:2002}.
-
-If provably correct results are desired, interval arithmetic provides
-an alternative to floating point.  Instead of just tracking an approximation to the answer,
-the package also tracks a radius within which the true result lies.
-This method is commonly used for $p$-adic computations since
-the ultrametric property of $p$-adic fields frequently keeps the radius small.
-Computations remain fairly efficient with this approach, but numerical instability
-can still lead to dramatic losses in precision (see Section \ref{sec:demo} for many examples).
-Tracking the precision of multiple variables concurrently, the set of possible
-true values associated to an inexact value takes the form of an ellipsoid
-with axes parallel to the coordinate axes.
-
-For better control of precision, we may allow arbitrary axes.  This change would have
-little utility for real numbers, since such ellipsoids are not preserved by most functions.
-For $p$-adic fields, in contrast, sufficiently small ellipsoids are mapped to other
-ellipsoids by differentiable maps.  From an algebraic perspective,
-these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space,
-and Proposition \ref{prop:precision} describes how the image of such a coset
-under a map $f$ is given exactly by applying the differential of $f$ to $H$.
-
-In this paper, we describe an implementation of this idea in \sage \cite{sage}.
-Rather than attaching a precision to each element, we store the precision of
-many elements together by tracking a precision lattice for the whole collection
-of variables.  As variables are created and destroyed, we update a matrix whose
-rows represent the vectors in the lattice.  Information about the precision
-of elements is extracted from the matrix as necessary.
-
-The article is structured as follows. In \S \ref{sec:demo} we provide a demonstration
-of the package, showing how it can provide more precise answers than the
-traditional methods for tracking $p$-adic precision.  In particular, \S \ref{ssec:elem}
-describes elementary arithmetic and the SOMOS-4 sequence, \S \ref{ssec:linalg}
-gives examples from linear algebra, \S \ref{ssec:comalg} examples using polynomials,
-and \S \ref{ssec:diffeq} examples of differential equations.
-
-In \S \ref{sec:implementation} we give more details on the implementation.
-\S \ref{ssec:preclemma} contains a brief overview on the theory of
-$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
-In the next two subsections, we explain in more details how \ZpLC and 
-\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
-automatic differentiation leading to the actual computation of the
-lattice that models the precision.
-In \S \ref{ssec:viewprec}, we explain how precision on any individual number
-can be recovered and discuss the validity
-of our results.
-The complexity overhead induced by our package is analyzed in
-\S \ref{ssec:complexity}.
-
-Finally, \S \ref{sec:conclusion} contains a discussion of how we see this
-package fitting into the existing $p$-adic implementations.  While these
-methods do introduce overhead, they are well suited to exploring
-precision behavior when designing algorithms, and can provide
-hints as to when further precision analysis would be useful.
+When computing with real and $p$-adic fields, exact results are usually impossible, since most elements have infinite decimal or $p$-adic expansions.
+Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.
+In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15,caruso-roe-vaccon:16,caruso-roe-vaccon:17}.
+The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.
+
+The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.
+At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \cite{ieee754:2008}.
+Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.
+Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.
+
+If provably correct results are desired, interval arithmetic provides an alternative to floating point.
+Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.
+This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.
+Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision (see Section \ref{sec:demo} for many examples).
+
+For better control of precision, we may use ellipsoids rather than balls.
+This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.
+For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.
+From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Proposition \ref{prop:precision} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.
+
+In this paper, we describe an implementation of this idea in \sage \cite{sage}.  \todo{Add short description of package features.}
+
+\todo{Plan of article}
+
+\todo{Convention: in matrices, vectors are represented by row-vector.}
 
@@
 The first one \ZpCR is the usual constructor for $p$-adic parents in 
-\sage. It tracks precision using interval arithmetic. On the contrary
+\sage. It tracks precision using interval arithmetics. On the contrary
 \ZpLC and \ZpLF are provided by our package. In the sequel, we will
-compare the results output with the three 
+compare the results output with the three constructors.
 Results for \ZpLF are only shown when they differ from \ZpLC.
@@
 \subsection{Elementary arithmetic}
-\label{ssec:elem}
 
@@
 leads to a gain of one digit in absolute precision.
-In the example below, we observe that when this multiplication is
+On the example below, we observe that when this multiplication is
 split into several steps, \ZpCR does not see the gain of precision
@@
 \subsection{Linear algebra}
-\label{ssec:linalg}
 
@@
 \subsection{Commutative algebra}
-\label{ssec:comalg}
 
@@
 \subsection{$p$-adic differential equations}
-\label{ssec:diffeq}
 
@@
 \section{What's behind the scene?}
-\label{sec:implementation}
 
@@
 this former is safer while the latter is faster.
+
+This Section is organized as follows.
+\S \ref{ssec:preclemma} contains a brief overview on the theory of
+$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
+In the next two subsections, we explain in more details how \ZpLC and 
+\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
+Automatic Differentiation leading to the actual computation of a
+global object modeling the precision (which is the aforementioned 
+lattice). 
+In \S \ref{ssec:viewprec}, we explain how precision on any individual number
+can be recoved from the above computations and discuss the validity
+of our results.
+The overhead complexity induced by our package is analyzed in
+\S \ref{ssec:complexity}.
 
@@
 \cIn   & \verb?x = ?{\color{ring}\verb?Z2?}\verb?(987,10); y = ?{\color{ring}\verb?Z2?}\verb?(21,5)? \\
-       & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision_lattice?}\verb?([x,y])? \\
+\cIn   & {\color{comment}\verb?#? \textit{We first retrieve the precision object}} \\
+       & \verb?L = ?{\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision?}\verb?()?\\
+\cIn   & \verb?L.?{\color{method}\verb?precision_lattice?}\verb?([x,y])? \\
 \cZpLC & \verb?[1024    0]? \\
@@
 \cIn   & \verb?u, v = x+y, x-y? \\
-       & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?precision_lattice?}\verb?([u,v])? \\
+       & \verb?L.?{\color{method}\verb?precision_lattice?}\verb?([u,v])? \\
 \cZpLC & \verb?[  32 2016]? \\
@@
 \begin{tabular}{rl}
-\cIn   & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?number_of_diffused_digits?}\verb?([x,y])? \\
+\cIn   & \verb?L.?{\color{method}\verb?number_of_diffused_digits?}\verb?([x,y])? \\
 \cZpLC & \verb?0? \\
-\cIn   & {\color{ring}\verb?Z2?}\verb?.?{\color{method}\verb?number_of_diffused_digits?}\verb?([u,v])? \\
+\cIn   & \verb?L.?{\color{method}\verb?number_of_diffused_digits?}\verb?([u,v])? \\
 \cZpLC & \verb?6? \\
@@
 1$.
-%For example, while executing the Euclidean algorithm (naive 
-%implementation) with two polynomials of degree $7$, most of the 
-%time $\cohgt_t(\ttw) \leq 1$.
+\end{comment}
+For example, while executing the Euclidean algorithm (naive 
+implementation) with two polynomials of degree $7$, most of the 
+time $\cohgt_t(\ttw) \leq 1$.
 
@@
 whereas deleting a variable is free. 
-Thus when executed with the \ZpLF mechanism, the complexity of our
+Thus when executed with the \ZpLF mecanism, the complexity of our
 algorithm becomes $O(s_\inp c)$.
@@
 bottleneck (which is the case in practice). Therefore, when executed 
-with the \ZpLF mecanism, the cost of our algorithm is expected to be 
+with the \ZpLC mecanism, the cost of our algorithm is expected to be 
 roughly $O(mc)$. Going further in speculation, we might estimate the 
@@
 \section{Conclusion}
-\label{sec:conclusion}
 

### 2018-02-04 ZpL/ZpL.tex

--- 2018-02-04 ZpL/ZpL.tex
@@
 Computations remain fairly efficient with this approach, but numerical instability
-can still lead to dramatic losses in precision (see Section \ref{sec:demo} for many examples).
+can still lead to dramatic losses in precision (see \S \ref{sec:demo} for many examples).
 Tracking the precision of multiple variables concurrently, the set of possible
@@
 
-For better control of precision, we may allow arbitrary axes.  This change would have
-little utility for real numbers, since such ellipsoids are not preserved by most functions.
-For $p$-adic fields, in contrast, differentiable maps with surjective differential will send
-sufficiently small ellipsoids to other ellipsoids.  From an algebraic perspective,
-these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space,
-and Proposition \ref{prop:precision} describes how the image of such a coset
-under a map $f$ is given exactly by applying the differential of $f$ to $H$.
+For better control of precision, we may allow arbitrary axes.  This 
+change would have little utility for real numbers, since such ellipsoids 
+are not preserved by most functions. For $p$-adic fields, in contrast, 
+differentiable maps with surjective differential will send sufficiently 
+small ellipsoids to other ellipsoids.  From an algebraic perspective, 
+these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic 
+vector space, and the main result of \cite{caruso-roe-vaccon:14a} (see 
+also Proposition \ref{prop:precision} below) describes how the image of 
+such a coset under a map $f$ is given exactly by applying the 
+differential of $f$ to $H$.
 
@@
 Rather than attaching a precision to each element, we store the precision of
-many elements together by tracking a precision lattice for the whole collection
+many elements together by tracking a precision module for the whole collection
 of variables.  As variables are created and destroyed, we update a matrix whose
-rows represent the vectors in the lattice.  Information about the precision
+rows represent the vectors in the module.  Information about the precision
 of elements is extracted from the matrix as necessary.
@@
 automatic differentiation leading to the actual computation of the
-lattice that models the precision.
+module that models the precision.
 In \S \ref{ssec:viewprec}, we explain how precision on any individual number
@@
 compare the outputs provided by each parent.
-Results for \ZpLF are only shown when they differ from \ZpLC.
+Results for \ZpLF are only displayed when they differ from \ZpLC.
 
@@
 split into several steps, \ZpCR does not see the gain of precision
-while \ZpLC and \ZpLF do.
+while \ZpL does.
 
@@
 Again \ZpCR does not output the optimal precision when the computation
-is split into several steps whereas \ZpLC and \ZpLF do.
+is split into several steps whereas \ZpL does.
 Actually, these toy examples illustrate quite common situations which 
@@
 the variables in $\calV_t$. For $\ttv \in \calV_t$, let $e_\ttv \in
-\Qp^{\calV_t}$ be the vector whose coordinates all vanish except 
+E_t$ be the vector whose coordinates all vanish except 
 at position $\ttv$ which takes the value $1$. The family 
@@
 
-\smallskip
-
-More precisely, in the \ZpLC case, the obtained matrix has this 
-shape:
+More precisely, the obtained matrix has this shape:
 $$\left(\raisebox{-0.5\height}{\begin{tikzpicture}[scale=0.3]
@@
 We work around this drawback by replacing everywhere exact $p$-adic 
-numbers by Floating Point $p$-adic numbers (at some given precision) 
+numbers by floating point $p$-adic numbers (at some given precision) 
 \cite{course-padic}.
@@
 %
-We assume first that the newly created variable is defined through an
+We assume first that the newly created variable is defined through a
 statement of the form:
@@
 However, we insist on the fact that all the computations now take
-place in the ``ring'' of Floating Point $p$-adic numbers. Therefore,
+place in the ``ring'' of floating point $p$-adic numbers. Therefore,
 we cannot guarantee that the rows of $M_{t+1}$ generate $H_{t+1}$.
@@
 $t$) as the smallest valuation of an entry of the column of $M_t$ 
-corresponding to the variable \ttv. Alternatively, it can be defined as 
+corresponding to the variable \ttv. Alternatively, it is
 the unique integer $N$ for which $\pi_\ttv(H_t) = p^N \Zp$ where 
@@
 \ldots, e_n)$.
-The diffused digits of precision only appear on linear combinations
-of those variables.
+%The diffused digits of precision only appear on linear combinations
+%of those variables.
 
@@
 $2^5 \cdot \text{I}_2$ and contains $H$ with index $2^6$ in it. This is 
-where the $6$ digits of precision comes from.
+where the $6$ digits of precision come from.
 There are easily visible here: the sum $u+v$ is known with $11$ digits,
@@
 size of the $p$-adic numbers we manipulate will all have roughly the
-same size: for \ZpLF, it is the precision we use for Floating Point 
-Arithmetic while, for \ZpLC, it is the absolute cap which
+same size: for \ZpLF, it is the precision we use for floating point 
+arithmetic while, for \ZpLC, it is the absolute cap which
 was fixed at the beginning.
@@
 \cite{denning:2005}
-asserts that $\cohgt_t(\ttw)$ tends to be small in general: destroyed 
+asserts that $\coind_t(\ttw)$ tends to be small in general: destroyed 
 variables are often variables that were created recently. As a basic 
@@
 
-\caption{The distribution of $\cohgt_t(\ttw)$}
+\caption{The distribution of $\coind_t(\ttw)$}
 \label{fig:deletion}
@@
 The histogram of Figure~\ref{fig:deletion} shows 
-the distribution of $\cohgt_t(\ttw)$ while executing the Euclidean
+the distribution of $\coind_t(\ttw)$ while executing the Euclidean
 algorithm (naive implementation) with two polynomials of degree $7$
-as input. The bias is evident: most of the time $\cohgt_t(\ttw) \leq
+as input. The bias is evident: most of the time $\coind_t(\ttw) \leq
 1$.
@@
 %implementation) with two polynomials of degree $7$, most of the 
-%time $\cohgt_t(\ttw) \leq 1$.
+%time $\coind_t(\ttw) \leq 1$.
 

### 2018-02-05 ZpL/ZpL.tex
Trivial change
Addressing Tristan's comment on y'^2 and adding comment on diffused digits in matrices

@@
 through the sequence of steps involved in carrying out a computation.
-In this paper, we describe a package for computing with $p$-adic rings and fields,
+In this paper, we describe a package for computing with $p$-adic rings and fields \cite{trac:23505},
 based on a series of papers by the same authors
@@
 
-In \cite{LV16}, the behavior of the precision when solving
-$p$-adic differential equations with separation of 
-variables has been studied.
-The authors have investigated the gap that appears
-when applying a Newton-method solver between
-the theoretic loss in precision and the 
-actual loss in precision for a naive implementation in \verb?Zp(p)?.
-We can reach this theoretical loss in precision using \ZpL.
-We use a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?
-that applies \verb?N? steps of the Newton method for 
-$y'=g \times h(y)$ as described in \cite{LV16}.
+In \cite{LV16}, the authors studied the computation of isogenies between elliptic curves
+over finite fields by applying the lattice precision model to $p$-adic
+differential equations.  Specifically, they considered the equation
+$y'=g(x) \times h(y)$ with
+$g,h,y \in \mathbb{Z}_p \llbracket x \rrbracket$
+such that $g(0)=h(0)=1$ and $y(0)=0.$
+
+Their main result was that the intrinsic loss in precision
+when computing the coefficient $x^n$ of $y$
+from $g$ and $h$ was in $\log_p(n)$
+even though a naive analysis of
+the Newton method for solving the equation
+yield a loss in $\log_p(n)^2.$
+
+We can reach this theoretical loss in precision using \ZpL, while \ZpCR does not perform as well.
+We apply \verb?N? steps of the Newton method for 
+$y'=g \times h(y)$ as described in \cite{LV16}, using a generic {\color{function}\verb?Newton_Iteration_Solver?}\verb?(g,h,N)?.
 
@@
 that is exactly $6$ more digits than the summands $u$ and $v$.
+
+Diffused digits frequently arise in practice.  In the context of the matrix multiplication
+example of \S \ref{ssec:linalg}, we get
+
+\smallskip
+
+{\noindent \small
+\begin{tabular}{rl}
+\cIn & \verb?L.?{\color{method}\verb?number_of_diffused_digits?}\verb?(M.?{\color{method}\verb?list?}\verb?())? \\
+\cZpLC & \verb?11? \\
+\end{tabular}}
+
+\smallskip
 

### 2018-04-30 ZpL/ZpL.tex
Began to take care of the remarks of the reviewers.

@@
 The first, \ZpCR, is the usual constructor for $p$-adic parents in 
-\sage. It tracks precision using interval arithmetic. On the contrary
+\sage.\footnote{It stands for \textit{capped relative}.} It tracks precision using interval arithmetic. On the contrary
 \ZpLC and \ZpLF are provided by our package. In the sequel, we will
 compare the outputs provided by each parent.
-Results for \ZpLF are only displayed when they differ from \ZpLC.
+As the results for \ZpLF and \ZpLC agree on all examples in this section, so we only write \ZpLC for them. 
+More details about what are precisely \ZpLC and \ZpLF are provided in Section \ref{sec:implementation}.
 
@@
 arithmetic computations.
-We first pick some random element $x$.
+We first pick some random element $x$ (using the natural
+Haar measure on $\mathbb{Z}_p$, as is default in \sage).
 The function {\color{function}\verb?random_element?} is designed so
@@
 With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
-$DP$ is $DQ$ is $D$.  However, we observe that \ZpCR output a quite 
+$DP$ and $DQ$ is $D$.  However, we observe that \ZpCR output a quite 
 different result. The point is that, in the \ZpCR case, Euclidean 
@@
 the Newton method for solving the equation
-yield a loss in $\log_p(n)^2.$
+yields a loss in $\log_p(n)^2.$
 
@@
 ``value'' and ``prec'' refer to user-specified constants or integral 
-values which was computed earlier.
+values which were computed earlier.
 
@@
 \end{matrix}\right).$$
-Performing row operations, we see then the entries of $C$ can be
+Performing row operations, we see then that the entries of $C$ can be
 reduced modulo $p^{N_{t+1}}$ without changing the lattice. In order
@@
 exactly $10$ digits are printed.
-The method {\color{method}\verb?precision_absolute?} provides a more easy-to-use
+The method {\color{method}\verb?precision_absolute?} provides an easier-to-use
 access to the absolute precision.

### 2018-05-04 ZpL/ZpL.tex
Address some reviewers' remarks

@@
 \newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
+\newcommand{\QpLC}{\text{\color{output} \rm \tt QpLC}\xspace}
+\newcommand{\QpLF}{\text{\color{output} \rm \tt QpLF}\xspace}
 \newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}
+
+\newcommand{\LC}{\text{\rm \tt LC}\xspace}
+\newcommand{\LF}{\text{\rm \tt LF}\xspace}
 
@@
 of elements is extracted from the matrix as necessary.
+Concretely, our package provides two new \sage 
+parents\footnote{Parents in \sage are the analogues of container objects 
+in computer science. Basically it corresponds to the implementation of a 
+concrete mathematical structure, as the ring of $p$-adic integers.}, 
+namely \ZpLC and \ZpLF (and their fraction fields \QpLC and \QpLF). The 
+acronyms \LC and \LF stand for ``lattice-cap'' and ``lattice-float''
+respectively (see \S \ref{sec:implementation} for details). Our package 
+is merged in the standard distribution of \sage since version 8.2.
 
@@
 $p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
-In the next two subsections, we explain in more details how \ZpLC and 
-\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
+In the next two subsections, we explain in more details how our package
+works. \S \ref{ssec:trackprec} is devoted to the implementation of 
 automatic differentiation leading to the actual computation of the
@@
 The first, \ZpCR, is the usual constructor for $p$-adic parents in 
-\sage.\footnote{It stands for \textit{capped relative}.} It tracks precision using interval arithmetic. On the contrary
+\sage\footnote{It stands for \textit{capped relative}.}. It tracks 
+precision using interval arithmetic. On the contrary
 \ZpLC and \ZpLF are provided by our package. In the sequel, we will
 compare the outputs provided by each parent.
-As the results for \ZpLF and \ZpLC agree on all examples in this section, so we only write \ZpLC for them. 
+As the results for \ZpLF and \ZpLC agree on all examples in this section, 
+we shall only mention \ZpLC in the sequel.
 More details about what are precisely \ZpLC and \ZpLF are provided in Section \ref{sec:implementation}.
+
+In what follows, random elements in $\Zp$ are always picked with respect 
+to the Haar measure. Random elements in $\Qp$ are constructed as a 
+product $p^V \cdot U$ where $U$ is a random element in $\Zp$ 
+(distributed according to the Haar measure) and $V$ is a random variable
+taking values in $\Z$ which is independant from $U$ and distributed as
+follows:
+$$\textstyle
+\text{Prob}[V{=}0] = \frac 1 5, \quad
+\text{Prob}[V{=}v] = \frac 2 {5|v|\cdot (|v|+1)}, \, \forall v \neq 0.$$
+This distribution is the default in \sage.
 
@@
 arithmetic computations.
-We first pick some random element $x$ (using the natural
-Haar measure on $\mathbb{Z}_p$, as is default in \sage).
+We first pick some random element $x$.
 The function {\color{function}\verb?random_element?} is designed so
@@
 %
-A natural example is that of the computation of GCD, whose stability has 
-been studied in \cite{caruso:2017}.
-A naive implementation of the Euclidean algorithm can produce different 
+A natural example is that of the computation of GCD in the Euclidean 
+ring $\Qp[X]$, whose stability has been studied in \cite{caruso:2017}. A 
+naive implementation of the Euclidean algorithm can produce different 
 behavior depending on the type of implementation of the field of 
@@
 exactly $10$ digits are printed.
-The method {\color{method}\verb?precision_absolute?} provides an easier-to-use
-access to the absolute precision.
+The method {\color{method}\verb?precision_absolute?} provides an 
+easier-to-use access to the absolute precision.
 
@@
 
-Taking the compositum of all the functions $\tilde F$ (\emph{cf} 
+Taking the composition of all the functions $\tilde F$ (\emph{cf} 
 Eq.~\eqref{eq:tildeF}) from time $0$ to $t$, we find that the execution 

### 2018-05-05 ZpL/ZpL.tex

@@
 Rather than attaching a precision to each element, we store the precision of
-many elements together by tracking a precision module for the whole collection
-of variables.  As variables are created and destroyed, we update a matrix whose
-rows represent the vectors in the module.  Information about the precision
-of elements is extracted from the matrix as necessary.
+many elements together by tracking a precision module\footnote{Throughout
+the paper, the word ``module'' has its mathematical meaning: it denotes 
+an additive group endowed with an external product by scalars in $\Zp$.} 
+for the whole collection of variables.  As variables are created and 
+destroyed, we update a matrix whose rows represent the vectors in the 
+module.  Information about the precision of elements is extracted from 
+the matrix as necessary.
 Concretely, our package provides two new \sage 

### 2018-05-07 ZpL/ZpL.tex
Working on referee comments

@@
 The first, \ZpCR, is the usual constructor for $p$-adic parents in 
-\sage\footnote{It stands for \textit{capped relative}.}. It tracks 
-precision using interval arithmetic. On the contrary
-\ZpLC and \ZpLF are provided by our package. In the sequel, we will
-compare the outputs provided by each parent.
+\sage. It tracks 
+precision using interval arithmetic. The parents 
+\ZpLC and \ZpLF provided by our package\footnote{\text{\color{output} \rm \tt CR} stands for \textit{capped relative},
+\text{\color{output} \rm \tt LC} for \textit{lattice-cap}, and \text{\color{output} \rm \tt LF} for \textit{lattice-float}.}
+track precision differently; more details will be provided in Section \ref{sec:implementation}.
+In the sequel, we will compare the outputs provided by each parent.
 As the results for \ZpLF and \ZpLC agree on all examples in this section, 
-we shall only mention \ZpLC in the sequel.
-More details about what are precisely \ZpLC and \ZpLF are provided in Section \ref{sec:implementation}.
+we will only give those for \ZpLC.
 
@@
 (distributed according to the Haar measure) and $V$ is a random variable
-taking values in $\Z$ which is independant from $U$ and distributed as
+taking values in $\Z$ which is independent from $U$ and distributed as
 follows:
@@
 \text{Prob}[V{=}v] = \frac 2 {5|v|\cdot (|v|+1)}, \, \forall v \neq 0.$$
-This distribution is the default in \sage.
+This distribution is the default in \sage, and the function {\color{function}\verb?random_element?}
+used below picks the same element for each of \ZpCR, \ZpLC and \ZpLF.
 
@@
 
-We begin our tour of the features of the \ZpL package with some basic 
-arithmetic computations.
-We first pick some random element $x$.
-The function {\color{function}\verb?random_element?} is designed so
-that it guarantees that the picked random element is the same for
-each constructor \ZpCR, \ZpLC and \ZpLF.
+We begin our tour of the \ZpL package with some basic 
+arithmetic.  Elements are approximated by giving a valuation
+and a unit, where the unit is specified modulo a power of $p$ determined
+by the global lattice precision.
 
@@
 maximum number of variables alive at the same time (which reflects 
-the memory occupation) while executing two basic computations.
+the memory usage) while executing two basic computations.
 The first one is the computation of the characteristic polynomial of a 
@@
 its complexity is $c$ operations in $\Qp$ (without any tracking of 
-precision), its memory usage is $m$ elements of $\Qp$, its input and its 
+precision), its memory usage is $m$ elements of $\Qp$, its input and 
 output have size $s_\inp$ and $s_\out$ (elements of $\Qp$) respectively.

### 2018-05-07 ZpL/ZpL.tex
More referee changes

@@
 \newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
+\newcommand{\QpCR}{\text{\color{output} \rm \tt QpCR}\xspace}
 \newcommand{\QpLC}{\text{\color{output} \rm \tt QpLC}\xspace}
@@
 \newcommand{\cZpLF}{{\color{output} \tt ZpLF}:}
+\newcommand{\cQpCR}{{\color{output} \tt QpCR}:}
+\newcommand{\cQpLC}{{\color{output} \tt QpLC}:}
 
@@
 an alternative to floating point.  Instead of just tracking an approximation to the answer,
-the package also tracks a radius within which the true result lies.
+one also tracks a radius within which the true result lies.
 This method is commonly used for $p$-adic computations since
@@
 respectively (see \S \ref{sec:implementation} for details). Our package 
-is merged in the standard distribution of \sage since version 8.2.
+has been included in the standard distribution of \sage since version 8.2.
 
@@
 is split into several steps whereas \ZpL does.
-Actually, these toy examples illustrate quite common situations which 
-often occur during the execution of many algorithms. For this reason,
-interval arithmetic often overestimates the losses of precision. 
-Roughly speaking, the aim of our package is to ``fix this misfeature''.
+These basic examples illustrate situations 
+that occur during the execution of many algorithms. 
+As a result, interval arithmetic frequently overestimates the loss of precision. 
+The aim of our package is to reduce this precision loss.
 In the next subsections, we present a bunch of examples showing the
@@
 
-Many generic algorithms of linear algebra lead to quite important 
-instability when they are used with $p$-adic numbers. In many cases, 
-our package \ZpL rubs this instability without having to change the 
-algorithm, nor the implementation.
+Many generic algorithms of linear algebra exhibit
+instability when used with $p$-adic numbers.
+Our package \ZpL frequently eliminates this instability without having to change either the 
+algorithm or the implementation.
 
@@
 Characteristic polynomials are notoriously hard to compute 
-\cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We illustrate this 
-with the following example (using the default algorithm of \sage for the 
-computation of the characteristic polynomial, which is a division free 
-algorithm in this setting) :
+\cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}.  By default,
+\sage uses a division-free algorithm over $\Qp$, this choice is not enough to achieve optimal
+precision in the \QpCR case since negative valuations may appear.  We illustrate the different
+precision behavior in the following example.
 
@@
  & \verb?M.?{\color{method}\verb?determinant?}\verb?()? \\
-\cZpCR
+\cQpCR
  & \verb?...010000010? \\
-\cZpLC
+\cQpLC
  & \verb?...010000010? \\
@@
  & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
-\cZpCR
+\cQpCR
  & \verb?...00000000000000000001*x^3 + ? \\
  & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
-\cZpLC
+\cQpLC
  & \verb?...00000000000000000001*x^3 + ? \\
@@
  & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
-\cZpCR
+\cQpCR
  & \verb?[x^3, x*y + ...1100010*x^2,? \\
  & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
-\cZpLC
+\cQpLC
  & \verb?[x^3, x*y + ...111100010*x^2,? \\
@@
  & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4); u[15]? \\
-\cZpCR
+\cQpCR
  & \verb?...1101? \\
-\cZpLC
+\cQpLC
  & \verb?...11011101? \\
@@
 the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
-$E_{t+1}$ but is included in an hyperplane.
-
-The package \ZpLC tackles this issue by introducing a cap: we do not 
-work with $dF_{\underline x}(H_t)$ but instead define the lattice
-$H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$
-where $N_{t+1}$ is an integer, the so-called \emph{cap}.
+$E_{t+1}$ but is included in an hyperplane, violating our definition of
+a precision lattice.
+
+The package \ZpLC tackles this issue by introducing a cap, analogous
+to adding $O(p^N)$ when creating a $p$-adic number from an integer.
+Specifically, we choose an integer $N_{t+1}$ and replace the $\Zp$-submodule
+$dF_{\underline x}(H_t)$ with the lattice $H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$.
 Alternatively, one may introduce the map:
@@
 \Zp$ under $d\tilde F_{(\underline x, \star)}$ for any value of $\star$.
-The choice of the 
-cap is of course a sensitive question. \ZpLC proceeds as follows. When a 
-ring is created, it comes with two constants (which can be specified by 
-the user): a relative cap \textsc{relcap} and an absolute cap 
-\textsc{abscap}. With these predefined values, the chosen cap is:
+In order to carry out this strategy, one must choose $N_{t+1}$.  We do
+so by associating two constants to each ring: a relative cap \textsc{relcap}
+and an absolute cap \textsc{abscap}. We then set
 \begin{equation}
@@
 with $y = f(x_1, \ldots, x_n)$.
+Larger caps mean that the precision of the final result is less likely to be affected,
+but require more space.  See Remark \ref{rmk:cap_optimality} for more details.
+
 In concrete terms, the lattice $H_{t+1}$ is represented by the block 
@@
 reduced modulo $p^{N_{t+1}}$ without changing the lattice. In order
-to optimize the size of the objects, we perform this reduction and
+save space, we perform this reduction and
 define $M_{t+1}$ by:
@@
 We work around this drawback by replacing everywhere exact $p$-adic 
-numbers by floating point $p$-adic numbers (at some given precision) 
-\cite{course-padic}.
+numbers by floating point $p$-adic numbers (at some given relative precision $k$) 
+\cite{course-padic}.  In the floating point model, we approximate all
+numbers in the disc $p^v(u + O(p^k))$ by a single rational number
+$p^vu$.  Operations on these numbers are then not exact, but instead
+produce the representative in the correct disc.
 
@@
 %\cdot C_i.$$
-However, we insist on the fact that all the computations now take
-place in the ``ring'' of floating point $p$-adic numbers. Therefore,
-we cannot guarantee that the rows of $M_{t+1}$ generate $H_{t+1}$.
-Nonetheless, they generate a module which is expected to be close
-to $H_{t+1}$.
+However, since operations defining $M_{t+1}$ are performed with floating point numbers,
+the matrix $M_{t+1}$ gives only an approximate basis for $H_{t+1}$.
 
@@
 
-Unfortunately, checking automatically the assumptions of Proposition 
-\ref{prop:precision} in full generality seems to be difficult, though it 
-can be done by hand for many particular examples
-\cite{caruso-roe-vaccon:14a, caruso:2017, LV16}.
-
-\begin{rmk}
+Unfortunately, automatically checking the assumptions of Proposition 
+\ref{prop:precision} in full generality is difficult because it requires bounds
+on higher derivatives.  To produce provably correct results, the assumptions
+can be checked by hand \cite{caruso-roe-vaccon:14a, caruso:2017, LV16},
+or \ZpCR may be used.
+
+\begin{rmk} \label{rmk:cap_optimality}
 Assuming that Proposition \ref{prop:precision} applies, 
@@
 In the \ZpLC case, we always have $\cohgt_t(\ttw) = \coind_t(\ttw)$, so 
-that the cost becomes $O\big(\coind_t(\ttw)^2\big)$. This does not look 
-nice \emph{a priori}. However, the principle of temporal locality 
-\cite{denning:2005}
-asserts that $\coind_t(\ttw)$ tends to be small in general: destroyed 
-variables are often variables that were created recently. As a basic 
+that the cost becomes $O\big(\coind_t(\ttw)^2\big)$, which seems high \emph{a priori}.
+However, the principle of temporal locality \cite{denning:2005}
+asserts that $\coind_t(\ttw)$ tends to be small: destroyed 
+variables are often recently created ones.  As a simple 
 example, variables which are local to a small piece of code (\emph{e.g.} 
-a short function or a loop) do not survive for a long time. It turns
-out that this behavior is typical in many implementations!
+a short function or a loop) have a short lifetime.
 %
@@
 The package \ZpL provides powerful tools (based on automatic 
-differentiation) to track precision in the $p$-adic setting. In many 
-concrete situations, it greatly outperforms standard interval arithmetic,
-as shown in \S \ref{sec:demo}.
+differentiation) to track precision in the $p$-adic setting.
+It frequently outperforms standard interval arithmetic in terms of the
+precision of the output, as shown in \S \ref{sec:demo}.
 The impact on complexity is controlled but nevertheless non-negligible 
@@
 still requires a substantial human contribution 
-and a careful special study of all parameters.
+and a careful study of all parameters.
 
@@
 Second, recall that a classical strategy to improve stability consists 
-in working internally at higher precision. Finding the internal increase
+of working internally at higher precision. Finding the internal increase
 in precision that best balances efficiency and accuracy is not an easy
-task in general. Understanding the number of diffused digits of 
+task in general. Understanding the diffused 
 precision gives very useful hints in this direction. For example, when there are
@@
 the contrary, when there are many diffused digits of precision, a large 
-increment is often required.
+increment is required.
 Since \ZpL gives a direct access to the number of diffused digits of 
-precision, it could be very useful to the designer who is concerned
+precision, it can be very useful to the designer who is concerned
 with the balance between efficiency and accuracy.

### 2018-05-07 ZpL/ZpL.tex
Typo + remark

@@
 parents\footnote{Parents in \sage are the analogues of container objects 
-in computer science. Basically it corresponds to the implementation of a 
-concrete mathematical structure, as the ring of $p$-adic integers.}, 
+in computer science. Basically a parent corresponds to an implementation
+of a concrete mathematical structure, as the ring of $p$-adic integers.}, 
 namely \ZpLC and \ZpLF (and their fraction fields \QpLC and \QpLF). The 
@@
 or \ZpCR may be used.
+\todo{NDXavier: I would not mention \ZpCR here but say instead that we
+can use \cite[Proposition~3.12]{caruso-roe-vaccon:14a} and that we plan
+to include this in a next release of the package.}
 

### 2018-05-07 ZpL/ZpL.tex
Adding references to Kedlaya and Lauder

--- 2018-05-07 ZpL/ZpL.tex
@@
 through the sequence of steps involved in carrying out a computation.
+Such analysis can be carried out automatically by the software, or via theorems specific
+to a particular application \cite{kedlaya:01a, lauder:04a}.
 In this paper, we describe a package for computing with $p$-adic rings and fields \cite{trac:23505},
@@
 The core of the package is a method for tracking precision using $p$-adic lattices which
-can yield dramatically more precise results, at the cost of increased runtime and memory usage.
+can yield dramatically more precise results than traditional methods, at the cost of increased runtime and memory usage.
 

### 2018-05-08 ZpL/ZpL.tex
number_of_diffused_digits -> diffused_digits

--- 2018-05-07 ZpL/ZpL.tex
@@
 %\cIn
-% & {\color{ring}\verb?Z2?}\verb?.precision().?{\color{method}\verb?number_of_diffused_digits?}\verb?()? \\
+% & {\color{ring}\verb?Z2?}\verb?.precision().?{\color{method}\verb?diffused_digits?}\verb?()? \\
 %\ZpLC
@@
 
-The method {\color{method}\verb?number_of_diffused_digits?} computes the number of
+The method {\color{method}\verb?diffused_digits?} computes the number of
 diffused digits of precision on a set of variables. Observe:
@@
 \begin{tabular}{rl}
-\cIn   & \verb?L.?{\color{method}\verb?number_of_diffused_digits?}\verb?([x,y])? \\
+\cIn   & \verb?L.?{\color{method}\verb?diffused_digits?}\verb?([x,y])? \\
 \cZpLC & \verb?0? \\
-\cIn   & \verb?L.?{\color{method}\verb?number_of_diffused_digits?}\verb?([u,v])? \\
+\cIn   & \verb?L.?{\color{method}\verb?diffused_digits?}\verb?([u,v])? \\
 \cZpLC & \verb?6? \\
@@
 \begin{tabular}{rl}
-\cIn & \verb?L.?{\color{method}\verb?number_of_diffused_digits?}\verb?(M.?{\color{method}\verb?list?}\verb?())? \\
+\cIn & \verb?L.?{\color{method}\verb?diffused_digits?}\verb?(M.?{\color{method}\verb?list?}\verb?())? \\
 \cZpLC & \verb?11? \\

### 2018-05-08 ZpL/ZpL.tex
Remark on the correctness

@@
 Unfortunately, automatically checking the assumptions of Proposition 
-\ref{prop:precision} in full generality is difficult because it requires bounds
-on higher derivatives.  To produce provably correct results, the assumptions
-can be checked by hand \cite{caruso-roe-vaccon:14a, caruso:2017, LV16},
-or \ZpCR may be used.
-\todo{NDXavier: I would not mention \ZpCR here but say instead that we
-can use \cite[Proposition~3.12]{caruso-roe-vaccon:14a} and that we plan
-to include this in a next release of the package.}
+\ref{prop:precision} in full generality is not straightforward because 
+it requires bounds on higher derivatives. For now, our package does not 
+perform this inspection but we plan to include it in a forthcoming 
+release, along the lines of 
+\cite[Proposition~3.12]{caruso-roe-vaccon:14a}. 
+Meanwhile, the assumptions can be checked by hand using the results
+of \cite{caruso-roe-vaccon:14a, caruso:2017, LV16}.
 

### 2018-05-08 ZpL/ZpL.tex
Add in ACM info

@@
 \newtheorem{deftn}[theo]{Definition}
+\copyrightyear{2018}
+\acmYear{2018}
+\setcopyright{acmlicensed}
+\acmConference[ISSAC'18]{2018 ACM International Symposium on Symbolic and Algebraic Computation}{July 16--19, 2018}{New York, NY, USA}
+\acmBooktitle{ISSAC'18: 2018 ACM Int'l Symposium on Symbolic \& Algebraic Computation, July 16--19, 2018, NY, NY, USA}
+\acmPrice{15.00}
+\acmDOI{10.1145/XXXXXX.XXXXXX}
+\acmISBN{978-1-4503-5550-6/18/07}
 

### 2018-05-08 ZpL/ZpL.tex
copyright
ACM info after \begin{document}

@@
 \documentclass[sigconf]{acmart}
+%\documentclass{acmart}
 
@@
 \newtheorem{deftn}[theo]{Definition}
+
+\fancyhead{}
+
+\begin{document}
+
 \copyrightyear{2018}
@@
 \setcopyright{acmlicensed}
-\acmConference[ISSAC'18]{2018 ACM International Symposium on Symbolic and Algebraic Computation}{July 16--19, 2018}{New York, NY, USA}
-\acmBooktitle{ISSAC'18: 2018 ACM Int'l Symposium on Symbolic \& Algebraic Computation, July 16--19, 2018, NY, NY, USA}
+\acmConference[ISSAC '18]{2018 ACM International Symposium on Symbolic and
+Algebraic Computation}{July 16--19, 2018}{New York, NY, USA}
+\acmBooktitle{ISSAC '18: 2018 ACM International Symposium on Symbolic and
+Algebraic Computation, July 16--19, 2018, New York, NY, USA}
 \acmPrice{15.00}
-\acmDOI{10.1145/XXXXXX.XXXXXX}
+\acmDOI{10.1145/3208976.3208995}
 \acmISBN{978-1-4503-5550-6/18/07}
 
-\fancyhead{}
-
-\begin{document}
+%\CopyrightYear{2018}
+%\setcopyright{acmlicensed}
+%\conferenceinfo{ISSAC '18,}{July 16--19, 2018, New York, NY, USA}
+%\isbn{978-1-4503-5550-6/18/07}\acmPrice{$15.00}
+%\doi{https://doi.org/10.1145/3208976.3208995}
 
@@
 \author{Xavier Caruso}
-  \affiliation{Universit\'e Rennes 1; \\
+  \affiliation{CNRS, Universit\'e Rennes 1; \\
   %\institution{IRMAR}
@@
 \S \ref{ssec:complexity}.
-
 Finally, \S \ref{sec:conclusion} contains a discussion of how we see this

### 2018-05-08 ZpL/ZpL.tex
8 pages

@@
 and \S \ref{ssec:diffeq} examples of differential equations.
-
 In \S \ref{sec:implementation} we give more details on the implementation.
@@
 $p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
-In the next two subsections, we explain in more details how our package
-works. \S \ref{ssec:trackprec} is devoted to the implementation of 
+%In the next two subsections, we explain in more details how our package
+%works.
+\S \ref{ssec:trackprec} is devoted to the implementation of 
 automatic differentiation leading to the actual computation of the
@@
 precision using interval arithmetic. The parents 
-\ZpLC and \ZpLF provided by our package\footnote{\text{\color{output} \rm \tt CR} stands for \textit{capped relative},
-\text{\color{output} \rm \tt LC} for \textit{lattice-cap}, and \text{\color{output} \rm \tt LF} for \textit{lattice-float}.}
+\ZpLC and \ZpLF provided by our package
+%\footnote{\text{\color{output} \rm \tt CR} stands for \textit{capped relative},
+%\text{\color{output} \rm \tt LC} for \textit{lattice-cap}, and \text{\color{output} \rm \tt LF} for \textit{lattice-float}.}
 track precision differently; more details will be provided in Section \ref{sec:implementation}.
